### 2017-03-15 main.tex
Update on Overleaf.

\documentclass[a4paper]{article}

%% Language and font encodings
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}

%% Sets page size and margins
\usepackage[a4paper,top=3cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

%% Useful packages
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}

\title{Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States}
\author{Shaw, Matsen, Minin}

\begin{document}
\maketitle

%comments from Erick:
% >  This is great!
% > - would be great to have all the calculations showing the "refinement" structure written out for the various tree structures (just done by hand and scanned is just fine) 
% > - I think of the "refinement" as a partition, so perhaps rather than "empty refinement" we have a "single-element partition"
% > - in order to interpret the site patterns I need to figure out in what order the leaves are labeled
% > - how is ell calculated?
% > - on the numerical sanity check, it doesn't seem clear that the branch lengths can be different than the inferred ones-- they can!
% > - re "general case", agreed though I don't think that we need numerical optimization of branch lengths in a set of a branch length partition-- we can just get the optimal branch length directly. Namely, for each branch, we know all hidden states (given that we are in a set of a partition) and so can count the number of mutations across that branch. From there we have a simple likelihood function, which either has an optimal in the interior or on the boundary of the partition.

\renewcommand{\arraystretch}{1.2} % because otherwise exponents get eaten by \hline

\section{Ancestral state reconstruction using maximum likelihood}

\subsection{The usual story}

We have observed character data $\mathbf{y}=[y_1,\ldots,y_n]\in\mathcal{Y}^n$ and corresponding unknown ancestral states $\boldsymbol\xi=[\xi_1,\ldots,\xi_n]\in\Xi^n$.
Both $\mathcal{Y}$ and $\Xi$ are discrete sets.
Without loss of generality, assume $\mathcal{Y}=\Xi=\{0,1\}$.
For a topology $\tau$ and branch lengths $t$ we form the likelihood
\begin{equation}
\label{eq:full_likelihood}
L_n(\mathbf{y};\boldsymbol\xi, \tau, t) = \prod_{i=1}^{n} \ P(y_i, \xi_i | \tau, t).
\end{equation}
In particular, we are interested in
$$
(\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
$$
which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.

Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$, the typical approach to estimate these parameters involves computing the marginal likelihood
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_n(\mathbf{y}; \tau, t) = \sum_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t).
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_n(\mathbf{y}; \tau, t).
$$
The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

\subsection{Joint maximization}

One tantalizing approach is to do away with the marginalization and the two-step maximization, directly estimating the maximum likelihood parameters from the full likelihood in \eqref{eq:full_likelihood}.
We can perform this iteratively, now computing a profile likelihood
\begin{equation}
\label{eq:profile_likelihood}
L_n'(\mathbf{y};\tau, t) = \max_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t)
\end{equation}
and estimating the topology and branch lengths via
$$
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ L_n'(\mathbf{y};\tau, t)
$$
while using $\hat{\boldsymbol\xi}$ from \eqref{eq:profile_likelihood} as an estimate for $\boldsymbol\xi$.

\textbf{Claim to prove}: Is it obvious that this particular iterative approach can be done? In other words, do $(\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t})$ from \eqref{eq:profile_likelihood} correspond to those of \eqref{eq:full_likelihood}? I believe since $\Xi$ is a discrete set that this is true, but it may be worth a more rigorous argument.

\subsection{Site pattern formulation}

Now, since $\mathcal{Y}$ and $\Xi$ are discrete sets, we define unique site patterns $\mathbf{s}=[s_1,\ldots,s_q]\in\mathcal{S}$ and ancestral state categories $\mathbf{z}=[z_1,\ldots,z_q]\in\mathcal{Z}$ such that
\begin{align}
L_n'(\mathbf{y};\tau, t) &= \max_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t) \\
    &= \max_{\boldsymbol\xi} \ \prod_{i=1}^{n} \ P(y_i, \xi_i | \tau, t) \\
    &= \max_{\mathbf{z}} \ \prod_{j=1}^{q} \ P(s_j, z_j | \tau, t)^{n_j} \label{eq:site_pattern_likelihood}
\end{align}
where $n_j$ is the number of observations in the sample where site pattern $s_j$ was seen.
Since $\mathcal{S}$ and $\mathcal{Z}$ are also discrete sets, the value of each factor $P(y_i, \xi_i | \tau, t)$ will be maximized at one or more values of $\xi_i$, though independent of $i$ given $j$.
This independence allows us to group factors by those where $\xi_i$ equals $z_j$ so that $P(s_j, z_j | \tau, t)$ is maximized.

\textbf{Example:} This might be a good time for a toy example showing this for a very simple topology.

\subsection{The case of infinite data}

We are now interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood}.
Let
$$
L_n''(\mathbf{s};\mathbf{z},\tau,t) = \prod_{j=1}^q \ P(s_j, z_j | \tau, t)^{n_j}
$$
so that
$$
\max_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t) = \max_{\mathbf{z}} \ L_n''(\mathbf{s};\mathbf{z},\tau,t).
$$
Now assume $n$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
\frac{1}{n} \log L_n''(\mathbf{s};\mathbf{z},\tau,t) &= \sum_{j} \frac{n_j}{n}\cdot \log P(s_j, z_j | \tau, t) \\
  &= \sum_{j} \frac{n_j}{n}\cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)].
\end{align}
In the limit of infinite data we see
$$
\frac{1}{n} \log L_n''(\mathbf{s};\mathbf{z},\tau,t) \rightarrow \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)]
$$
and that, since $z_j$ appears only in the last term, the log of \eqref{eq:site_pattern_likelihood} in the same limit is proportional to
\begin{align}
\frac{1}{n} \log L_n'(\mathbf{y};\tau, t) &= \max_{\mathbf{z}} \ \frac{1}{n} \log L_n''(\mathbf{s};\mathbf{z},\tau,t) \\
    &\rightarrow \max_{z_1, \ldots, z_q} \ \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)] \\
    &= \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \max_{z_j} \ \log P(z_j | s_j, \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{align}

\textbf{A note on terminology:}  While we are passing to the limit of infinite data, we don't actually \emph{need} infinite data for these results to be striking. Another way of considering the above is what would happen if we knew the ``truth,'' i.e., the true generating topology and branch lengths.

\subsection{Showing inconsistency}

For clarity, define \eqref{eq:site_pattern_profile_likelihood_mean} as
$$
J(\tau, t; \tau^*, t^*) = \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \max_{z_j} \ \log P(z_j | s_j, \tau, t)]
$$
where $(\tau, t)$ are the estimands and $(\tau^*, t^*)$ are the fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if there exist estimands $(\tau,t)\neq(\tau^*,t^*)$ where
$$
J(\tau, t; \tau^*, t^*) > J(\tau^*, t^*; \tau^*, t^*)
$$
then we have a possible inconsistency, and the maximum likelihood parameters obtained in this manner may not conform to the truth in the limit.
The issue with this approach is that we do not guarantee any estimation procedure will recover the values $(\tau, t)$.
We may show a stronger, more suggestive result if for $\tau'\neq\tau^*$ we have
\begin{equation}
\label{eq:inconsistency_inequality}
J(\tau', \hat{t}_1; \tau^*, t^*) > J(\tau^*, \hat{t}_2; \tau^*, t^*)
\end{equation}
with $\hat{t}_1$ and $\hat{t}_2$ estimated by maximizing \eqref{eq:profile_likelihood}.

\section{Inconsistency of joint maximum likelihood}

\subsection{Parameter setting}

Define the ``Farris zone'' topology $\tau_1$ as \cite{farris}
\begin{verbatim}
1\ /3
  | 
2/ \4
\end{verbatim}
and the ``Felsenstein zone'' topology $\tau_2$ as \cite{felsenstein}
\begin{verbatim}
1\      /3
   >--< 
2/      \4
\end{verbatim}
where, for generating data $t^*$, the probability of a change in character along the top two branches and the middle branch is $p_x$ and the probability of a change in character along the bottom two branches is $p_y$.
In our estimand $t$, we modify the classical topologies by allowing the middle branch to be unequal to either $p_x$ or $p_y$ and denote this probability $p_w$; the parameters $p_x, p_y$ and $p_w$ all lie in interval $[0,1/2]$.
To obtain branch length parameters, we transform these probabilities using the function $f(p)=1-2p$ so that $t=\{x,y,w\}:=\{f(p_x), f(p_y), f(p_w)\}$; the parameters $x, y$ and $w$ all lie in the interval $[0,1]$.

\textbf{Justification for estimand:} It might be worth arguing why this form of the estimand $t$ is necessary.
I think the idea is that for either of these topologies, any maximum obtained will occur where the top two and bottom two branches are equal, though not necessarily the middle branch.
This would be good to show explicitly.

Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies under these two topologies, calculated using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{semplesteel}.
Table~\ref{tab:likelihoods} contains calculations of likelihood values for fixed site patterns and topologies, with the maximum values over ancestral state patterns noted.

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
\multicolumn{3}{c}{Classical}\\
    \hline
$s_j$   &$P(s_j|\tau_1,t)$&$P(s_j|\tau_2,t)$\\
    \hline
0000&$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
0001&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
0010&$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
0100&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
1000&$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
0011&$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
0101&$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
1001&$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
    \hline
\multicolumn{3}{c}{Modified}\\
    \hline
$s_j$   &$P(s_j|\tau_1,t)$&$P(s_j|\tau_2,t)$\\
    \hline
0000&$1+x^2+w^2+4xyw+x^2w^2$&$1+2xw+2xyw+x^2y+w^2y+x^2w^2$\\
0001&$1+x^2-w^2-x^2w^2$&$1+x^2y-w^2y-x^2w^2$\\
0010&$1-x^2+w^2-x^2w^2$&$1-x^2y+w^2y-x^2w^2$\\
0100&$1+x^2-w^2-x^2w^2$&$1+x^2y-w^2y-x^2w^2$\\
1000&$1-x^2+w^2-x^2w^2$&$1-x^2y+w^2y-x^2w^2$\\
0011&$1-x^2-w^2+x^2w^2$&$1+2xw-2xyw-x^2y-w^2y+x^2w^2$\\
0101&$1+x^2+w^2-4xyw+x^2w^2$&$1-2xw-2xyw+x^2y+w^2y+x^2w^2$\\
1001&$1-x^2-w^2+x^2w^2$&$1-2xw+2xyw-x^2y-w^2y+x^2w^2$\\
    \hline
\end{tabular}    
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$P(z_j|s_j,\tau_1,t)$}\\
\hline
& \multicolumn{2}{|c|}{$z_j$}\\
    \hline
$s_j$    &00                              &01\\
    \hline
0000&$(1+x)^2(1+y)(1+w)^{2*}$          &$(1+x)^2(1-y)(1-w)^2$\\
0001&$(1+x)^2(1+y)(1+w)(1-w)^*$       &$(1+x)^2(1-y)(1+w)(1-w)$\\
0010&$(1+x)^2(1+y)(1+w)(1-w)^*$       &$(1+x)^2(1-y)(1+w)(1-w)$\\
0100&$(1+x)(1-x)(1+y)(1+w)^{2*}$       &$(1+x)(1-x)(1-y)(1-w)^2$\\
1000&$(1+x)(1-x)(1+y)(1+w)^{2*}$       &$(1+x)(1-x)(1-y)(1-w)^2$\\
0011&$(1+x)(1-x)(1+y)(1+w)(1-w)^{\dagger}$    &$(1+x)(1-x)(1-y)(1+w)(1-w)$\\
0101&$(1+x)^2(1+y)(1-w)^{2\ddagger}$          &$(1+x)^2(1-y)(1+w)^{2\ddagger}$\\
1001&$(1+x)(1-x)(1+y)(1+w)(1-w)^{\dagger}$    &$(1+x)(1-x)(1-y)(1+w)(1-w)$\\
    \hline
    \hline
&10                           &11\\
    \hline
0000&$(1-x)^2(1-y)(1+w)^2$        &$(1-x)^2(1+y)(1-w)^2$\\
0001&$(1-x)^2(1-y)(1+w)(1-w)$     &$(1-x)^2(1+y)(1+w)(1-w)$\\
0010&$(1-x)^2(1-y)(1+w)(1-w)$     &$(1-x)^2(1+y)(1+w)(1-w)$\\
0100&$(1+x)(1-x)(1-y)(1+w)^2$     &$(1+x)(1-x)(1+y)(1-w)^2$\\
1000&$(1+x)(1-x)(1-y)(1+w)^2$     &$(1+x)(1-x)(1+y)(1-w)^2$\\
0011&$(1+x)(1-x)(1-y)(1+w)(1-w)$  &$(1+x)(1-x)(1+y)(1+w)(1-w)^{\dagger}$\\
0101&$(1-x)^2(1-y)(1-w)^2$       &$(1-x)^2(1+y)(1+w)^{2\ddagger}$\\
1001&$(1+x)(1-x)(1-y)(1+w)(1-w)$  &$(1+x)(1-x)(1+y)(1+w)(1-w)^{\dagger}$\\
    \hline
    \multicolumn{3}{c}{$P(z_j|s_j,\tau_2,t)$}\\
\hline
& \multicolumn{2}{|c|}{$z_j$}\\
    \hline
$s_j$    &00                              &01\\
    \hline
0000&$(1+x)^2(1+y)(1+w)^{2*}$          &$(1+x)(1-x)(1-y)(1+w)(1-w)$\\
0001&$(1+x)^2(1+y)(1+w)(1-w)^{\ddagger}$       &$(1+x)(1-x)(1-y)(1+w)^{2\ddagger}$\\
0010&$(1+x)(1-x)(1+y)(1+w)^{2\ddagger}$       &$(1+x)^2(1-y)(1+w)(1-w)^{\ddagger}$\\
0100&$(1+x)^2(1+y)(1+w)(1-w)^{\ddagger}$       &$(1+x)(1-x)(1-y)(1-w)^2$\\
1000&$(1+x)(1-x)(1+y)(1+w)^{2\ddagger}$       &$(1-x)^2(1-y)(1+w)(1-w)$\\
0011&$(1+x)(1-x)(1+y)(1+w)(1-w)^{\ddagger}$          &$(1+x)^2(1-y)(1+w)^{2\ddagger}$\\
0101&$(1+x)^2(1+y)(1-w)^{2\ddagger}$    &$(1+x)(1-x)(1-y)(1+w)(1-w)^{\ddagger}$\\
1001&$(1+x)(1-x)(1+y)(1+w)(1-w)^{\ddagger}$    &$(1-x)^2(1-y)(1+w)^{2\ddagger}$\\
    \hline
    \hline
&10                           &11\\
    \hline
0000&$(1+x)(1-x)(1-y)(1+w)(1-w)$        &$(1-x)^2(1+y)(1-w)^2$\\
0001&$(1+x)(1-x)(1-y)(1-w)^2$     &$(1-x)^2(1+y)(1+w)(1-w)$\\
0010&$(1-x)^2(1-y)(1+w)(1-w)$     &$(1+x)(1-x)(1+y)(1-w)^2$\\
0100&$(1+x)(1-x)(1-y)(1+w)^{2\ddagger}$     &$(1-x)^2(1+y)(1+w)(1-w)$\\
1000&$(1+x)^2(1-y)(1+w)(1-w)^{\ddagger}$     &$(1+x)(1-x)(1+y)(1-w)^2$\\
0011&$(1-x)^2(1-y)(1-w)^2$       &$(1+x)(1-x)(1+y)(1+w)(1-w)^{\ddagger}$\\
0101&$(1+x)(1-x)(1-y)(1+w)(1-w)^{\ddagger}$  &$(1-x)^2(1+y)(1+w)^{2\ddagger}$\\
1001&$(1+x)^2(1-y)(1-w)^{2\ddagger}$  &$(1+x)(1-x)(1+y)(1+w)(1-w)^{\ddagger}$\\
\hline
\end{tabular}    
\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
Maxima determined row-wise (i.e., by site pattern).
All values multiplied by $1/32$.
Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
\label{tab:likelihoods}
\end{table}

\subsection{Attempts to bound the likelihoods}

Ideally we could show for some generating parameters $t^*$ that there are simple bounds such that
$$
J(\tau', \hat{t}_1; \tau^*, t^*) \geq \underline{J}_1 > \overline{J}_2 \geq J(\tau^*, \hat{t}_2; \tau^*, t^*)
$$
where the fact that $\underline{J}_1 > \overline{J}_2$ will be obvious.
This approach could work since we have very particular forms for the prior probabilities and the likelihoods; we would need to find values of $t^*$ that minimize the probability (from the generating parameters) placed on those cases where the likelihood for the generating parameters is higher than those for a candidate parameter.
The frustrating aspect of this would be deriving a form for $J(\tau, \hat{t}; \tau^*, t^*)$ since it involves constrained maximization over three separate parameters: $x, y$ and $w$.

Since $x$ and $y$ are both in $[0,1]$ we have ready-made bounds available, though they may not be as tight as we want.
For example, in the case where both $x,y > 1/2$ we can compute upper and lower bounds on the generating probabilities by plugging in the appropriate value to Table~\ref{tab:sitepatprob}.

This is still ad hoc in that we need to fix an upper or lower bound for $x$ and $y$ to obtain these bounds.
Ideally we would seek tight bounds $(\underline{x}, \overline{x})$ and $(\underline{y}, \overline{y})$, where ``tight'' refers to how close they approximate the likelihoods they are bounding and not the parameters.

\subsection{Simplifying the likelihoods}

To simplify matters, we will first focus on bounding the the terms with the generating probabilities as they are more complicated.
Let
\begin{align}
\label{eq:information}
H(\mathbf{p};x, y, w) &= \sum_{j} P(s_j | \tau^*, \{x^*,y^*\}) \cdot \log P(s_j | \tau', \{x,y,w\}) \nonumber \\ 
&:= \sum_{j} p(s_j) \cdot \log P(s_j | \tau', \{x,y,w\}).
\end{align}
By Gibbs inequality,
$$
H(\mathbf{p};x, y, w) \le H(\mathbf{p};x^*,y^*, y^*),
$$
though this bound can be made tighter.
In our simplifications we will make use of the constants
$$
a = p(0100) + p(1000) + p(0011) + p(1001) = \left\{
	\begin{array}{ll}
		(4-2x^{*2}-2y^{*2})/8  & \mbox{Farris zone,} \\
		(4-2x^{*2}y^{*}-2y^{*3})/8  & \mbox{Felsenstein zone,}
	\end{array}
\right.
$$
$$
b = p(0101) = \left\{
	\begin{array}{ll}
		(1+x^{*2}+y^{*2}-4x^{*}y^{*2}+x^{*2}y^{*2})/8  & \mbox{Farris zone,} \\
		(1-2x^{*}y^{*}-2x^{*}y^{*2}+x^{*2}y^{*}+y^{*3}+x^{*2}y^{*2})/8  & \mbox{Felsenstein zone,}
	\end{array}
\right.
$$
$$
c = p(0001) + p(0010) + p(0100) + p(1000) = (4-4x^{*2}y^{*2})/8
$$
and
$$
d = p(0101) + p(1001) = \left\{
	\begin{array}{ll}
		(2-4x^{*}y^{*2}+2x^{*2}y^{*2})/8  & \mbox{Farris zone} \\
		(2-4x^{*}y^{*}+2x^{*2}y^{*2})/8  & \mbox{Felsenstein zone}
	\end{array}
\right.
$$
with corresponding bounds
$$
0 \le a,c \le \frac{1}{2}
$$
and
$$
0 \le b,d \le \frac{1}{4}.
$$

Our key approach will be to decompose the likelihood as
\begin{equation}
\label{eq:partial-likelihood}
L^{(i)}(x, y, w) = H(\mathbf{p}; x, y, w) + L^{p,(i)}(x,y,w)
\end{equation}
where the first term is \eqref{eq:information} and the second term will be a weighted sum of the $\log$s of values from Table~\ref{tab:likelihoods}, the superscript indicating which choices for ancestral state likelihood values we choose.
Define
$$
\{\hat{x}, \hat{y}, \hat{w}, \hat{i}\} = \arg\max_{\{x,y,w,i\}} \ L^{p,(i)}(x,y,w).
$$
We have upper and lower bounds for \eqref{eq:partial-likelihood} in the form
$$
H(\mathbf{p}; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(\mathbf{p}; x^{*}, y^{*}, y^{*}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
$$

\textbf{Argument for lower bound}: This makes intuitive sense, though may be difficult to show in general.
Can we not just say $\max_u f(u) + g(u)$ is greater than $f(u') + g(u')$ for any $u'$ that doesn't maximize $f(u) + g(u)$ just by definition?
There could be multiple maxima, but that's why we have greater than or equal to.
I think the only thing we'd need to show is that a maximum exists, but since $\{x,y,w\}$ are supported on a closed set this is certainly true.
I must be missing something.

\textbf{Argument for upper bound}: I think this is true mostly because $\max_u f(u) + g(u)$ should be less than $\max_u f(u) + \max_u g(u)$ by some triangle inequality argument.
If we were to actually maximize the full likelihood, we should get something smaller than maximizing some form of it that we consider by looking at the parts separately.
For a fixed $\{x^*, y^*\}$ we can calculate this upper bound directly---can we obtain it as a useful-looking function of these parameters in the general case?

\subsubsection{A simplified Farris likelihood}

The likelihood for the Farris zone topology is less complicated than that of the Felsenstein zone since only one site pattern has an ambiguous likelihood---all site patterns except \texttt{0101} have a fixed value for their likelihood when taking the maximum.
We decompose the full likelihood

where $L_{fa}^{(i)}(x,y,w)$ takes one of three forms:
$$
L_{fa}^{(1)} = (2-a)\log(1+x)+a\log(1-x)+(2-a-2b)\log(1+w)+(a+2b)\log(1-w)+\log(1+y),
$$
$$
L_{fa}^{(2)} = (2-a-2b)\log(1+x)+(a+2b)\log(1-x)+(2-a)\log(1+w)+a\log(1-w)+\log(1+y),
$$
$$
L_{fa}^{(3)} = (2-a)\log(1+x)+a\log(1-x)+(2-a)\log(1+w)+a\log(1-w)+(1-b)\log(1+y)+b\log(1-y).
$$
These likelihoods have maxima
$$
\hat{L}_{fa}^{(1)} = \hat{L}_{fa}^{(2)} = (2-a)\log(2-a)+a\log(a)+(2-a-2b)\log(2-a-2b)+(a+2b)\log(a+2b)+\log(2),
$$
$$
\hat{L}_{fa}^{(3)} = (2-a)\log(2-a)+a\log(a)+(2-a)\log(2-a)+a\log(a)+(1-b)\log(1-b)+b\log(b)+\log(2).
$$

\textbf{TODO}: cases for $\{x^{*}, y^{*}\}$ where each of the above is the corresponding maximum.
I think the first is greater than or equal to the second in most cases, though not all (i.e., when $b$ is large but $a$ is small).

\subsubsection{A simplified Felsenstein likelihood}

We construct similar bounds on the Felsenstein likelihood, though to make matters simpler we replace all $(1+y)$ terms in Table~\ref{tab:likelihoods} with $(1-y)$ to obtain a lower bound, performing the reverse substitution to obtain upper bounds.
As in \eqref{eq:partial-likelihood} we see that the likelihood in this case has one of six forms:
$$
L_{fe}^{(1)} = 2\log(1+x)+(2-c-2d)\log(1+w)+(c+2d)\log(1-w)+\log(1-y),
$$
$$
L_{fe}^{(2)} = (2-c-2d)\log(1+x)+(c+2d)\log(1-x)+2\log(1+w)+\log(1-y),
$$
$$
L_{fe}^{(3)} = (2-2d)\log(1+x)+2d\log(1-x)+(2-c)\log(1+w)+c\log(1-w)+\log(1-y),
$$
$$
L_{fe}^{(4)} = (2-c)\log(1+x)+c\log(1-x)+(2-2d)\log(1+w)+2d\log(1-w)+\log(1-y),
$$
$$
L_{fe}^{(5)} = (2-d)\log(1+x)+d\log(1-x)+(2-c-d)\log(1+w)+(c+d)\log(1-w)+\log(1-y)
$$
and
$$
L_{fe}^{(6)} = (2-c-d)\log(1+x)+(c+d)\log(1-x)+(2-d)\log(1+w)+d\log(1-w)+\log(1-y)
$$
with maxima
$$
\hat{L}_{fe}^{(1)} = \hat{L}_{fe}^{(2)} = 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d)
$$
$$
\hat{L}_{fe}^{(3)} = \hat{L}_{fe}^{(4)} = (2-2d)\log(2-2d)+2d\log(2d)+(2-c)\log(2-c)+c\log(c)
$$
and
$$
\hat{L}_{fe}^{(5)} = \hat{L}_{fe}^{(6)} = (2-d)\log(2-d)+d\log(d)+(2-c-d)\log(2-c-d)+(c+d)\log(c+d).
$$

\textbf{TODO}: cases for $\{x^{*}, y^{*}\}$ where each of the above is the corresponding maximum.
I think the first is always greater than or equal to the other two, but I haven't been able to show it yet.
It may take a little information theory--fu, or just standard application of the bounds in the introduction to this section.

\paragraph{Example}

Set $\{x^*, y^*\} = \{0.5, 0.5\}$.
For a Farris zone generating topology, we have
$$
a =  0.375, \ b =  0.1328125,
$$
resulting in
$$
L_{fa}^{(1)} = L_{fa}^{(2)} \approx 1.24637119863,
$$
$$
L_{fa}^{(3)} \approx 1.14372724688
$$
and
$$
H_{fa}(\mathbf{p};x^*,y^*, y^*) \approx -1.99591752302.
$$
Thus computing the upper bound yields
$$
\max_{x,y,w} \ L_{fa}(x,y,w) \le -0.749546324382
$$
and it remains to show that 
$$
\max_{x,y,w} \ L_{fe}(x,y,w) \ge -0.749546324382.
$$

For the same generating parameters,
$$
c = 0.46875, \ d = 0.203125,
$$
resulting in
$$
L_{fe}^{(1)} = L_{fe}^{(2)} \approx 1.40196030769, 
$$
$$
L_{fe}^{(3)} = L_{fe}^{(4)} \approx 0.674162530673
$$
and
$$
L_{fe}^{(5)} = L_2^{(6)} \approx 0.838975450706.
$$
Calculating shows
$$
H_{fe}(\mathbf{p}; \hat{x}, \hat{y}, \hat{w}) = -2.06377559511,
$$
and
$$
\max_{x,y,w} \ L_{fe}(x,y,w) \ge -0.661815287426 > -0.749546324382 \ge \max_{x,y,w} \ L_{fa}(x,y,w)
$$
where site patterns are generated from the Farris zone topology.
Wait, really?
Did I just show something???

\paragraph{Example}

Say we have found values for $\{x^{*}, y^{*}\}$ where the likelihoods we need to consider are
$$
\hat{L}_{fa}(a,b) = (2-a)\log(2-a)+a\log(a)+(2-a-2b)\log(2-a-2b)+(a+2b)\log(a+2b)+\log(2)
$$
and
$$
\hat{L}_{fe}(c,d) = 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d).
$$
This is true in the case of $\{.5, .5\}$ so there is at least measurable support for this case.
What we need to show now is the support of this intersects with the support of
$$
H(\mathbf{p}; 1, 0, 1-c-2d) + \hat{L}_{fe}(c,d) \ge H(\mathbf{p}; x^{*}, y^{*}, y^{*}) + \hat{L}_{fa}(a,b)
$$
in some measurable set where data are generated on a Farris topology.
All of these inequalities are functions of $\{x^{*}, y^{*}\}$ only, and we can cheat a bit and plot the values at which this is true.
This may give us some insight into what form these regions of inconsistency take---see Fig.~\ref{fig:inconsistency-farris}.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{ineqs-max-far-gen}
\caption{Regions of inconsistency}
\label{fig:inconsistency-farris}
\end{figure}

Some scratch work letting $\{x, y\}=\{x^*,y^*\}$:
$$
c+2d = 1-xy^2, \ 2-c-2d = 1+xy^2
$$
$$
H(\mathbf{p}; 1, 0, 1-c-2d) \propto (2-4xy^2+2x^2y^2)\log(1-2xy^2+x^2y^4) + (2+4xy^2+2x^2y^2)\log(1+2xy^2+x^2y^4)+(4-4x^2y^2)\log(1-x^2y^4)
$$
$$
\hat{L}_{fe}(c,d) = 2\log(2) + (1+xy^2)\log(1+xy^2) + (1-xy^2)\log(1-xy^2)
$$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section{Temporarily unused text I may want later to show more general results on the likelihood estimates}

\subsubsection{A counterexample in branch lengths}

Let $(\tau^*, t^*)=(\tau_1, \{x,y,x\})$.
We claim there exist $\{x,y\}$ and $\{x, \tilde{y}\}$ such that
$$
L'(\tau^*, t^*; \tau^*, \{x, \tilde{y}\}) > L'(\tau^*, t^*; \tau^*, t^*).
$$
The model is given by values in the Table.
Set $x=0$ to obtain
$$
L'(\tau^*, t^*; \tau^*, t^*) \propto 8\log(1+y)
$$
and
$$
L'(\tau^*, t^*; \tau^*, \{0, \tilde{y}\}) \propto 8\log(1+\tilde{y}).
$$
Therefore, if $\tilde{y} > y$ we have an inconsistency.

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
    \hline
$s_j$   &$P(s_j|\tau_1,t)$&$\max_{z_j} \ P(s_j,z_j|\tau_1,t)$\\
    \hline
0000&$1+2x^2+4x^2y+x^4$&$(1+x)^4(1+y)$\\
0001&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
0010&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
0100&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
1000&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
0011&$1+2x^2-4x^2y+x^4$&max$\{(1+x)^2(1+y)(1-x)^2,(1+x)^4(1-y)\}$\\
0101&$1-2x^2+x^4$&$(1+x)^2(1+y)(1-x)^2$\\
1001&$1-2x^2+x^4$&$(1+x)^2(1+y)(1-x)^2$\\
    \hline
\end{tabular}    
\caption{Counterexample 1.}
\label{tab:sitepatprob_case1}
\end{table}

\subsubsection{A counterexample in topologies}

Let $(\tau^*, t^*)=(\tau_1, \{x,y,y\})$.
We claim there exist $\{x,y\}$ such that
$$
L'(\tau^*, t^*; \tau_2, t^*) > L'(\tau^*, t^*; \tau^*, t^*).
$$
Set $y=0$ to obtain parameters given in Table.
We have
$$
L'(\tau^*, t^*; \tau^*, t^*) \propto 4(1+x^2)\log(1+x^2)+4(1-x^2)\log(1-x^2)+4(1+x^2)\log((1+x)^2)+4(1-x^2)\log((1+x)(1-x))
$$
and
$$
L'(\tau^*, t^*; \tau_2, t^*) \propto 3(1+x^2)\log((1+x)^2)+3(1-x^2)\log((1+x)^2)+(1+x^2)\log((1+x)(1-x))+(1-x^2)\log((1+x)(1-x)).
$$
The difference being greater than zero, i.e.,
$$
L'(\tau^*, t^*; \tau_2, t^*) - L'(\tau^*, t^*; \tau^*, t^*) > 0,
$$
simplifies to
$$
4x^2(\log(1-x^2)-\log(1+x^2)+\log(1-x)-\log(1+x)) - 2(2\log(1+x^2)+2\log(1-x^2)-\log(1+x)+\log(1-x)) > 0.
$$
Let $0 < x < 1/\sqrt{2}$ so that the above inequality becomes
$$
-\log(1-x^2)-3\log(1+x^2) > 0
$$
or
$$
(1+x^2)^3(1-x^2) < 1.
$$
Applying $x<1/\sqrt{2}$ shows if
$$
x < \sqrt{-1+2^{1/3}} \approx 0.5098
$$
we have an inconsistency.

\begin{table}
\centering
\begin{tabular}{|l|l|l||l|l|}
    \hline
$s_j$   &$P(s_j|\tau_1,t)$&$\max_{z_j} \ P(s_j,z_j|\tau_1,t)$&$P(s_j|\tau_2,t)$&$\max_{z_j} \ P(s_j,z_j|\tau_2,t)$\\
    \hline
0000&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)^2$\\
0001&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)^2$\\
0010&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)^2$\\
0100&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)^2$\\
1000&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)^2$\\
0011&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)(1-x)$\\
0101&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)^2$\\
1001&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)(1-x)$\\
    \hline
\end{tabular}    
\caption{Counterexample 2.}
\label{tab:sitepatprob_case2}
\end{table}

\subsubsection{More fleshed out counterexamples}

The above counterexamples are highly pathological, but they show there exist whole spaces of parameters yielding inconsistencies.
It seems as long as $x$ and $y$ are ``small enough'' we see inconsistency.

\subsection{start}

We can calculate the probability of a particular site pattern under this topology from the probability that all character states are equal as
$$
P(0000|\tau^*,t) = \frac{1}{8}(1+x^2+z^2+4xyz+x^2z^2)
$$
where the probability of any site pattern will be the same as above with the corresponding branch length parameter replaced with its negative.
For example, if we have the pattern $0001$, since the last branch has length $z$ we replace it with $-z$ and obtain
$$
P(0001|\tau^*,t) = \frac{1}{8}(1+x^2-z^2-x^2z^2).
$$
All possible site pattern probabilities are given in Table~\ref{tab:sitepatprob}.
All likelihoods and their maximal values are given in Table~\ref{tab:likelihoods}.

It can be shown (I'll do it later) that for our $t^*$ the right-hand term above is $-16\log 8$.
Moreover, for $t=\{x,y,z\}$
\begin{equation}
\sum_{s} \log P(s | \tau^*, t) + \log P(\xi | s, \tau^*, t) = -16\log 8 + f(x,y,z) + g(x,y,z)
\end{equation}
where
\begin{align}
f(x,y,z) &= 4\log(1+x)+4\log(1-x)+4\log(1+z)+4\log(1-z)+\\
& 3\log(1+x^2)+3\log(1+z^2)+\\
& \log((1+x^2)(1+z^2)+4xyz) + \log((1+x^2)(1+z^2)-4xyz).
\end{align}
and
\begin{align}
g(x,y,z) &= 10\log(1+x)+10\log(1+z)+4\log(1-x)+4\log(1-z)+7\log(1+y)+\\
& \left\{ 
  \begin{array}{l}
  2\log(1+x)+2\log(1-z)+\log(1+y)\\
  2\log(1+x)+2\log(1+z)+\log(1-y)\\
  2\log(1-x)+2\log(1-z)+\log(1+y)\\
  \end{array} \right.
\end{align}
whose component will be chosen based on whichever yields the maximum value for $\log L''$.

\subsection{Two fixed topologies}

\subsection{Special case: $x=z=0$}

Consider now setting all data-generating branch length parameters $t^*$ to zero; this yields equal probabilities of $1/8$ for each site pattern.
By the inequality of Gibbs we have
$$
\sum_{s} P(s|\tau^*,t^*)\log P(s|\tau^*,t^*) \geq \sum_{s} P(s|\tau^*,t^*)\log P(s|\tau^*,t)
$$
summing over all site patterns, which is equivalent to
$$
\sum_{s} \log P(s|\tau^*,t^*) \geq \sum_{s} \log P(s|\tau^*,t).
$$
We must show, then, that there exist $t\neq t^*$ such that
$$
\max_{\xi_s} \ \sum_{s} \log P(s | \tau^*, t) + \log P(\xi_s | s, \tau^*, t) > \max_{\xi_s} \sum_{s} \log P(s | \tau^*, t^*) + \log P(\xi_s | s, \tau^*, t^*).
$$
Assume for $t$ that $x=z=0$ but $y\neq 0$.
Clearly site pattern probabilities are still all equal, and the first summand of each side of the inequality can be ignored in this case.
Call
$$
f(x,y,z) = \sum_{s} \log P(\xi_s | s, \tau^*, t=\{x,y,z\}).
$$
For branch lengths $\{x,y,z\}$, by considering all eight cases (see Table~\ref{tab:likelihoods}) of site patterns we can show that in terms of branch length \emph{probabilities}%
	%
    \footnote{The function $f$ has inverse $f^{-1}(u)=(1-u)/2$ so that the probability of a change along a branch of length $x$ maps to $(1-x)/2$ and the probability of no change maps to $(1+x)/2$.}
    %
\begin{align}
f(x,y,z) \propto & \ 10\log(1+x)+10\log(1+z)+4\log(1-x)+4\log(1-z)+7\log(1+y)+\\
& \left\{ 
  \begin{array}{l}
  2\log(1+x)+2\log(1-z)+\log(1+y)\\
  2\log(1+x)+2\log(1+z)+\log(1-y)\\
  2\log(1-x)+2\log(1-z)+\log(1+y)\\
  \end{array} \right.
\end{align}
whose last summand will be chosen based on whichever yields the maximum value and where the proportionality is due to requiring branch length parameters $\{x,y,z\}$ to be multiplied by $1/2$ to map them to probabilities.
We see that
$$
f(x,y,z)|_{x=z=0} = 8\log(1+y)
$$
and thus require the constraint
$$
8\log(1+y) > 0
$$
or that $y$ is any value bounded away from zero, and will obtain a maximum at $y=1$ or $p_y=0$.

\subsection{Special case: $x=z\neq 0$}

Fixing $x=z=0$ is a little too pathological---anything can happen when we're flipping coins to determine terminal characters.
Let's investigate a relaxed yet still special case where branches to the tips are all equal but allowed to be nonzero.
So we set $\tau^*=\tau$, $t^*=\{x,x,y\}$ and $t=\{x,x,\tilde{y}\}$ where the same terminal branches are used for both parameter sets with only the internal branch different.
Removing terms involving only $x$ and redundant $y$ shows that we need only concern ourselves with the objective functions
\begin{align*}
L_1(u) =& 8\log(1+u) + \\
  &(1+2x^2+4x^2y+x^4)\log(1+2x^2+4x^2u+x^4) + \\
  &(1+2x^2-4x^2y+x^4)\log(1+2x^2-4x^2u+x^4)
\end{align*}
if $(1+u)(1-x)^2 > (1-u)(1+x)^2$---or, more simply, if $u > 2x / (1+x^2)$---and
\begin{align*}
L_2(u) =& (7-2x^2+4x^2y-x^4)\log(1+u) + \\
  &(1+2x^2-4x^2y+x^4)\log(1-u) + \\
  &(1+2x^2+4x^2y+x^4)\log(1+2x^2+4x^2u+x^4) + \\
  &(1+2x^2-4x^2y+x^4)\log(1+2x^2-4x^2u+x^4)
\end{align*}
otherwise.
We can rule out the case where $\tilde{y} < y$ in $L_1$ by the inequality of Gibbs and using a similar argument along with the inequality
$$
1+2x^2-4x^2y+x^4 \le 7-2x^2+4x^2y-x^4
$$
implies we can rule out $\tilde{y} < y$ in $L_2$ as well (actually I'm not sure if I believe this anymore; the case of $L_1$ definitely holds, though).
This leaves us with four sets of inequalities where an inconsistency will arise,
$$
\left\{\tilde{y} > y > \frac{2x}{1+x^2}, \ L_1(\tilde{y}) > L_1(y)\right\}, 
$$
$$
\left\{\tilde{y} > \frac{2x}{1+x^2} > y, \ L_1(\tilde{y}) > L_2(y)\right\}, 
$$
$$
\left\{\frac{2x}{1+x^2} > \tilde{y} > y, \ L_2(\tilde{y}) > L_2(y)\right\},
$$
and
$$
\left\{\frac{2x}{1+x^2} > y > \tilde{y}, \ L_2(\tilde{y}) > L_2(y)\right\}.
$$
Simplifying shows
$$
L_2(u) = L_1(u) + (1+2x^2-4x^2y+x^4)\log\frac{1-u}{1+u} \le L_1(u).
$$
Can we use this fact?
I believe in my heart of hearts, perhaps foolishly, that the first set of inequalities holds categorically.
The fact above will then imply the second set of inequalities must also hold.
The tricky bits are the last two sets of inequalities.

If you'll allow me to indulge in a bit of cheating, I provide a few plots showing---for fixed $x$---the regions where inconsistencies arise.
It looks like there is some regularity that we must be able to exploit somehow, if only to show that for a fixed $x$ there are inequalities in $y$ and $\tilde{y}$ governing consistency.

%\begin{figure}
%\centering
%\includegraphics[width=.3\textwidth]{ineqs-x-0_25}\hfill
%\includegraphics[width=.3\textwidth]{ineqs-x-0_75}\hfill
%\includegraphics[width=.3\textwidth]{ineqs-x-0_9}
%\caption{Various regions of inconsistency for fixed $x$}
%\end{figure}

Say $\tau_1$ is the ``Felsenstein zone'' tree and $\tau_2$ is the ``Farris zone'' tree, but both trees have a different value for its middle, connecting branch.
%EM widdershins!! This still depends on how we draw them, no?
The tips are labeled widdershins proceeding from the top-left tip.
\begin{verbatim}
Felsenstein:
1\      /3
   >--< 
2/      \4

Farris:
1\ /3
  | 
2/ \4
\end{verbatim}
Our branch length parameters $t=\{x,y,z\}$ will consist of transformed probabilities where $y$ corresponds to this middle branch.
For $p_x\in[0,1/2]$ defined as
$$p_x = Pr(\mbox{change occurred along branch})$$
then $x = 1-2p_x$.
The parameter $x$ takes values in $[0,1]$ and will appear in likelihood calculations as either $(1+x)$ or $(1-x)$.
Properties for $y$ and $z$ are identical.
Assume for the moment that $\tau^*=\tau_1$ and $t^*=\{x,y,z\}$.

\subsubsection{Example site pattern: \texttt{0101}}

It can be shown that the only $A$ for which there is more than one partition for $\tau_2$ is for the site pattern \texttt{0101}.
In this case
$$A_1(\tau_1) = \{(1+x)^2(1-z)^2 = (1-x)^2(1+z)^2\},$$
$$A_2(\tau_1) = \{(1+x)(1+y)(1-z) = (1-x)(1-y)(1+z)\},$$
$$A_3(\tau_1) = \{(1-x)(1+y)(1+z) = (1+x)(1-y)(1-z)\}$$
and
$$A_1(\tau_2) = \{(1-z)^2(1+y) = (1+z)^2(1-y)\},$$
$$A_2(\tau_2) = \{(1-z)^2(1+x)^2 = (1+z)^2(1-x)^2\},$$
$$A_3(\tau_2) = \{(1+x)^2(1-y) = (1-x)^2(1+y)\}.$$

Now, the probability of generating this site pattern under $\tau^*$ is given as
$$
Pr(\texttt{0101} | \tau^*, t^*) = \frac{1}{8}(1-2xz+y(x^2+z^2-2xz)+x^2z^2).
$$

\subsubsection{Example site pattern: \texttt{0000}}

In the case of site pattern \texttt{0000} the refinements $A$ for both topologies are single-element partitions.
In other words, no configuration of branch lengths will result in an ancestral state other than \texttt{00}.
Here we have the probability
$$
Pr(\texttt{0000} | \tau^*, t^*) = \frac{1}{8}(1+2xz+y(x^2+z^2+2xz)+x^2z^2)
$$
and the maximum value for $Pr(\xi | \texttt{0000}, \tau, t^*)$ for both $\tau_1$ and $\tau_2$ is
$$
Pr(\xi | \texttt{0000}, \tau, t^*) = (1+x)^2(1+z)^2(1+y).
$$

\subsubsection{All refinements with multi-element partitions}

(show calculations for all refinements that have multiple partitions)

\subsection{Numerical sanity check}

Let's suppose site pattern \texttt{0101} is the only one that drives some disconcert between the generating tree and the maximum likelihood tree.
For a number of combinations of $\{x,y,z\}$ we need to calculate first
$$
Pr(\texttt{0101} | \tau^*, t^*) = \frac{1}{8}(1-2xz+y(x^2+z^2-2xz)+x^2z^2)
$$
and the maximum value over each $A_i(\tau)$ of $Pr(\xi | \texttt{0101}, \tau, t^*)$ for $\tau\in\{\tau_1, \tau_2\}$.
In other words, say we have the triplet $\{x,y,z\}=\{0.8,0.5,0.8\}$.
Then we have
$$
\max_{\xi} Pr(\xi | \texttt{0101}, \tau_1, t^*) = 0.0243, \ \max_{\xi} Pr(\xi | \texttt{0101}, \tau_2, t^*) = 0.6561, \ Pr(\texttt{0101} | \tau^*, t^*) = 0.0162.
$$
This doesn't yet tell us anything, as the probabilities for other site patterns may overwhelm this particular one, but gives a general flavor of how these computations will be performed.
Moreover, a different triplet of $\{x,y,z\}$ could obtain a different maximum value for $Pr(\xi | s, \tau, t^*)$.
This can be seen with the pattern \texttt{0011} where
$$
\max_{\xi} Pr(\xi | \texttt{0011}, \tau_1, t^*) = 0.6561, \ \max_{\xi} Pr(\xi | \texttt{0011}, \tau_2, t^*) = 0.0243, \ Pr(\texttt{0011} | \tau^*, t^*) = 0.1762
$$
meaning that if all else were equal---i.e., $\max_{\xi} Pr(\xi | s, \tau, t^*)$ were the same for $\tau_1$ and $\tau_2$ for all $s$---and that the maximum value actually occurs at this particular triplet $\{x,y,z\}$ then
$$
L(\tau_1) = 0.928 > 0.119 = L(\tau_2)
$$
and we have no contradiction.
That's a lot of ``if'' statements, so let's investigate in a little more generality.

\subsection{Differing branch lengths}

First, we do not wish to assume $\{x,y,z\}$ are the same for both trees.

\subsubsection{General case}

The generating probability $Pr(s | \tau^*, t^*)$ has a specific form given a Hadamard transformation of the branch length parameters, while the term proportional to $Pr(\xi | s, \tau, t^*)$ relies on taking the maximum over a function with multiple components.
All of this is easy to write into a computer program, and we can numerically find the maximum value over $\{\theta_1, \ldots, \theta_5\}\in[0,1]$ of each $L(\tau_1)$ and $L(\tau_2)$.
If $L(\tau_1) < L(\tau_2)$ we see that jointly maximizing branch lengths and ancestral states results in a contradiction.

In sufficient enough generality, we would look for any combination of $\{\tau_1,\tau_2\}$ that would result in this contradiction, but let's first focus on the two from the previous section.

\subsubsection{Classic formulations of topologies}

\begin{table}
\centering
\begin{tabular}{|l|l|}
    \hline
s   &$P(s)$\\
    \hline
0000&$1+x^2+y^2+4xy^2+x^2y^2$\\
0001&$1+x^2-y^2-x^2y^2$\\
0010&$1+x^2-y^2-x^2y^2$\\
0100&$1-x^2+y^2-x^2y^2$\\
1000&$1-x^2+y^2-x^2y^2$\\
0011&$1+x^2+y^2-4xy^2+x^2y^2$\\
0101&$1-x^2-y^2+x^2y^2$\\
1001&$1-x^2-y^2+x^2y^2$\\
    \hline
\end{tabular}    
\caption{Site pattern probabilities for Farris zone topology.
All values are multiplied by $1/8$.}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|l|}
    \hline
s   &$P(s)$\\
    \hline
0000&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
0001&$1+x^2y-y^3-x^2y^2$\\
0010&$1-x^2y+y^3-x^2y^2$\\
0100&$1+x^2y-y^3-x^2y^2$\\
1000&$1-x^2y+y^3-x^2y^2$\\
0011&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
0101&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
1001&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
    \hline
\end{tabular}    
\caption{Site pattern probabilities for Felsenstein zone topology.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob-fels}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \hline
    &00                              &01\\
    \hline
0000&$(1+x)^2(1+y)^{3*}$          &$(1+x)^2(1-y)^3$\\
0001&$(1+x)^2(1+y)^2(1-y)^*$       &$(1+x)^2(1+y)(1-y)^2$\\
0010&$(1+x)^2(1+y)^2(1-y)^*$       &$(1+x)^2(1+y)(1-y)^2$\\
0100&$(1+x)(1-x)(1+y)^{3*}$       &$(1+x)(1-x)(1-y)^3$\\
1000&$(1+x)(1-x)(1+y)^{3*}$       &$(1+x)(1-x)(1-y)^3$\\
0011&$(1+x)^2(1+y)(1-y)^{2}$          &$(1+x)^2(1-y)(1+y)^{2\ddagger}$\\
0101&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
1001&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
    \hline
    \hline
&10                           &11\\
    \hline
0000&$(1-x)^2(1+y)^2(1-y)$        &$(1-x)^2(1+y)(1-y)^2$\\
0001&$(1-x)^2(1+y)(1-y)^2$     &$(1-x)^2(1+y)^2(1-y)$\\
0010&$(1-x)^2(1+y)(1-y)^2$     &$(1-x)^2(1+y)^2(1-y)$\\
0100&$(1+x)(1-x)(1+y)^2(1-y)$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
1000&$(1+x)(1-x)(1+y)^2(1-y)$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
0011&$(1-x)^2(1-y)^3$       &$(1-x)^2(1+y)^{3\ddagger}$\\
0101&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
1001&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
    \hline
\end{tabular}    
\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
Maxima determined row-wise (i.e., by site pattern).
Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \hline
    &00                              &01\\
    \hline
0000&$(1+x)^2(1+y)^{3*}$          &$(1+x)(1-x)(1+y)(1-y)^2$\\
0001&$(1+x)^2(1+y)^2(1-y)^{*}$       &$(1+x)(1-x)(1+y)^2(1-y)$\\
0010&$(1+x)(1-x)(1+y)^{3\ddagger}$       &$(1+x)^2(1+y)(1-y)^{2\ddagger}$\\
0100&$(1+x)^2(1+y)^2(1-y)^*$       &$(1+x)(1-x)(1-y)^3$\\
1000&$(1+x)(1-x)(1+y)^{3\ddagger}$       &$(1-x)^2(1+y)(1-y)^2$\\
0011&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$          &$(1+x)^2(1+y)^2(1-y)$\\
0101&$(1+x)^2(1+y)(1-y)^{2\ddagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
1001&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
    \hline
    \hline
&10                           &11\\
    \hline
0000&$(1+x)(1-x)(1+y)(1-y)^2$        &$(1-x)^2(1+y)(1-y)^2$\\
0001&$(1+x)(1-x)(1-y)^3$     &$(1-x)^2(1+y)^2(1-y)$\\
0010&$(1-x)^2(1+y)(1-y)^2$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
0100&$(1+x)(1-x)(1+y)^2(1-y)^{\ddagger}$     &$(1-x)^2(1+y)^2(1-y)$\\
1000&$(1+x)^2(1+y)(1-y)^{2\ddagger}$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
0011&$(1-x)^2(1-y)^3$       &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
0101&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1-x)^2(1+y)^{3\ddagger}$\\
1001&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
    \hline
\end{tabular}    
\caption{Likelihood calculations for all site patterns and internal states of Felsenstein topology.
Maxima determined row-wise (i.e., by site pattern).
Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
\label{tab:likelihoods-fels}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|l|}
    \hline
s   &$P(s)$\\
    \hline
0000&$1+x^2+z^2+2xyz+x^2y+z^2y+x^2z^2$\\
0001&$1+x^2-z^2+x^2y-z^2y-x^2z^2$\\
0010&$1-x^2+z^2-x^2y+z^2y-x^2z^2$\\
0100&$1+x^2-z^2+x^2y-z^2y-x^2z^2$\\
1000&$1-x^2+z^2-x^2y+z^2y-x^2z^2$\\
0011&$1-x^2-z^2-2xyz-x^2y-z^2y+x^2z^2$\\
0101&$1+x^2+z^2-2xyz+x^2y+z^2y+x^2z^2$\\
1001&$1-x^2-z^2-2xyz-x^2y-z^2y+x^2z^2$\\
    \hline
\end{tabular}    
\caption{Site pattern probabilities for Felsenstein zone topology.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob-fels}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \hline
    &00                              &01\\
    \hline
0000&$(1+x)^2(1+y)(1+z)^{2*}$          &$(1+x)(1-x)(1-y)(1+z)(1-z)$\\
0001&$(1+x)^2(1+y)(1+z)(1-z)^{\ddagger}$       &$(1+x)(1-x)(1-y)(1+z)^{2\ddagger}$\\
0010&$(1+x)(1-x)(1+y)(1+z)^{2\ddagger}$       &$(1+x)^2(1-y)(1+z)(1-z)^{\ddagger}$\\
0100&$(1+x)^2(1+y)(1+z)(1-z)^{\ddagger}$       &$(1+x)(1-x)(1-y)(1-z)^2$\\
1000&$(1+x)(1-x)(1+y)(1+z)^{2\ddagger}$       &$(1-x)^2(1-y)(1+z)(1-z)$\\
0011&$(1+x)(1-x)(1+y)(1+z)(1-z)^{\ddagger}$          &$(1+x)^2(1-y)(1+z)^{2\ddagger}$\\
0101&$(1+x)^2(1+y)(1-z)^{2\ddagger}$    &$(1+x)(1-x)(1-y)(1+z)(1-z)^{\ddagger}$\\
1001&$(1+x)(1-x)(1+y)(1+z)(1-z)^{\dagger}$    &$(1+x)(1-x)(1-y)(1+z)(1-z)$\\
    \hline
    \hline
&10                           &11\\
    \hline
0000&$(1+x)(1-x)(1-y)(1+z)(1-z)$        &$(1-x)^2(1+y)(1-z)^2$\\
0001&$(1+x)(1-x)(1-y)(1-z)^2$     &$(1-x)^2(1+y)(1+z)(1-z)$\\
0010&$(1-x)^2(1-y)(1+z)(1-z)$     &$(1+x)(1-x)(1+y)(1-z)^2$\\
0100&$(1+x)(1-x)(1-y)(1+z)^{2\ddagger}$     &$(1-x)^2(1+y)(1+z)(1-z)$\\
1000&$(1+x)^2(1-y)(1+z)(1-z)^{\ddagger}$     &$(1+x)(1-x)(1+y)(1-z)^2$\\
0011&$(1-x)^2(1-y)(1-z)^2$       &$(1+x)(1-x)(1+y)(1+z)(1-z)^{\ddagger}$\\
0101&$(1+x)(1-x)(1-y)(1+z)(1-z)^{\ddagger}$  &$(1-x)^2(1+y)(1+z)^{2\ddagger}$\\
1001&$(1+x)(1-x)(1-y)(1+z)(1-z)$  &$(1+x)(1-x)(1+y)(1+z)(1-z)^{\dagger}$\\
    \hline
\end{tabular}    
\caption{Likelihood calculations for all site patterns and internal states of Felsenstein topology.
Maxima determined row-wise (i.e., by site pattern).
Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
\label{tab:likelihoods-fels}
\end{table}

In the first case we are in a similar situation to the case of $x=z=0$ where $\tilde{y}>y$ implies $L(\tilde{y})>L(y)$.
The second case is more interesting.
We still require $\tilde{y}>y$, but we now have a further restriction.
We are interested in the simultaneous inequalities
$$
\frac{2x}{1+x^2} > \tilde{y} > y,
$$
$$
[7-2x^2+4x^2y-x^4]\log\frac{1+\tilde{y}}{1+y}+[1+2x^2-4x^2y+x^4]\log\frac{1-\tilde{y}}{1-y} > 0
$$
which can be written
$$
8\log\frac{1+\tilde{y}}{1+y}+[1+2x^2-4x^2y+x^4]\log\frac{(1-\tilde{y})(1+y)}{(1+\tilde{y})(1-y)} > 0.
$$
What can we say about this?

Now terms involving only $P(s|\tau,t)$ differ between $t^*$ and $t$ for the site patterns 0000 and 0101 but no others, allowing us to ignore them.
Additionally, all terms involving $P(\xi_s|s,\tau,t)$ will differ between $t^*$ and $t$ yielding a search for parameters such that
$$
\sum_{s\in\{0000,0101\}} P(s|\tau^*, t^*)\log P(s|\tau^*, t) + \sum_{s} P(s | \tau^*, t^*) \log P(\xi_s | s, \tau^*, t)
$$
is maximized at $t\neq t^*$.
Since $\log P(\xi_s|s,\tau,t)$ breaks up into a sum of terms involving $x$ and $y$ we have
$$
(6-4x^2-2x^4)\log(1+y)+\sum_{s\in\{0000,0101\}} P(s|\tau^*, t^*)[\log P(s|\tau^*, t)+\log P(\xi_s|s, \tau^*, t)].
$$
For a bit of clarity, let
$$
a(x,y) = (1+2x^2+4x^2y+x^4),
$$
$$
b(x,y) = (1+2x^2-4x^2y+x^4)
$$
so that we have
$$
(6-4x^2-2x^4)\log(1+y)+a(x,y)\log a(x,y)+b(x,y)\log b(x,y) + \sum_{s\in\{0000,0101\}} P(s|\tau^*, t^*)\log P(\xi_s|s, \tau^*, t).
$$
Canceling more terms shows the right-hand sum is
$$
(1+2x^2+4x^2y+x^4)\log(1+y) + 16x^2y\log(1+x) + P(0101|\tau^*, t^*)\log P(\xi_s|0101, \tau^*, t).
$$
That final right-hand term above is actually a component function since we are maximizing a likelihood, and the 0101 row in Table~\ref{tab:likelihoods} does not have a unique maximum value.
This value is either
$$
c_1(x,y) = (1+x)^4(1-y),
$$
$$
c_2(x,y) = (1+x)^2(1-x)^2(1+y).
$$
We are interested in two inequalities; if
$$
c_2(x,y) > c_1(x,y)
$$
then our whole objective function simplifies since many terms with $\log(1+y)$ cancel.
We can then focus on a second inequality, namely whether
$$
8\log(1+y) + x^2y(4\log(1+y)+16\log(1+x)) + a(x,y)\log a(x,y) + b(x,y)\log b(x,y) + c_2(x,y)...blah
$$
is greater than
$$
8\log(1+\tilde{y}) + x^2y(4\log(1+\tilde{y})+16\log(1+x)) + a(x,y) + b(x,y) + c_2(x,y)...
$$

\subsection{Dashed intuition}

I am having a tough time seeing the intuition behind this result.
If $x=z=0$ then we are flipping a coin for what happens at the internal node proceeding from the leaf.
Why would effectively constraining the interior nodes to be equal result in a larger value for the likelihood?
It doesn't have anything to do with efficiency, does it (i.e., by constraining the parameters to be equal we are using the same amount of data to estimate fewer parameters)?
Can we show that constraining any $\xi_i,\xi_j$ to be equal in $L_n$ will also result in problems?

I think what may help my sanity is proceeding with the above for the usual maximum likelihood story, that is, for
$$
\log L''' = \sum_{s} P(s | \tau^*, t^*) \cdot [\log P(s | \tau^*, t) + \log\sum_{\xi} P(\xi_s | s, \tau^*, t)]
$$
show
$$
L'''(\tau^*, t^*) > L'''(\tau^*, t)
$$
for $t\neq t^*$.
At least for $x=z=0$ we can see that these two quantities will be equal for all $y$ due to the symmetry of choosing the internal nodes.
Is this satisfying?

UPDATE: thinking about this more, it actually makes sense.
If we are maximizing over internal states, we will always prefer a branch length probability that will indicate no change occurs between them, effectively constraining all internal states to be the same given a particular site pattern.
As we can see in the table, the maximum value almost always occurs for internal states that are equivalent.
Seeking harmony between internal states will always increase the likelihood except in cases where branches to terminating leaves are sufficiently small.
So the question might be, now, whether or not we can prove a general result: since internal states seem to always maximize the likelihood by being constrained to be equal, what conditions are needed on the terminal branches (here $x$ and $z$) that will yield (in)consistent estimates?

\subsection{Plan of attack}

We recall \eqref{eq:inconsistency_inequality} and set $\tau^*=\tau_1$ and $\tau'=\tau_2$.
In other words, data are generated from the ``Farris zone'' topology, and we seek all values for $t^*$ where we can obtain a higher likelihood by proceeding with joint maximization assuming the ``Felsenstein zone'' topology.

\subsection{The cheater's approach}

Using \texttt{python}, we can numerically maximize \eqref{eq:site_pattern_profile_likelihood_mean} over $t$ separately for fixed topologies $\tau^*$ and $\tau'$.
For $t^*$, we can calculate these values over a grid of values for $(x^*, y^*)\in[0,1]^2$.
At worst, this will show us if we're on a fool's errand, but if not then we could benefit by gaining insight into what form a refinement of branch lengths yielding inconsistency would take.

\end{document}

### 2017-03-17 main.tex
update

@@
 \usepackage{graphicx}
+\usepackage{subcaption}
 \usepackage[colorinlistoftodos]{todonotes}
 \usepackage[colorlinks=true, allcolors=blue]{hyperref}
+
+\graphicspath{ {figures/} }
 
@@
 
-Define the ``Farris zone'' topology $\tau_1$ as \cite{farris}
-\begin{verbatim}
-1\ /3
-  | 
-2/ \4
-\end{verbatim}
-and the ``Felsenstein zone'' topology $\tau_2$ as \cite{felsenstein}
-\begin{verbatim}
-1\      /3
-   >--< 
-2/      \4
-\end{verbatim}
-where, for generating data $t^*$, the probability of a change in character along the top two branches and the middle branch is $p_x$ and the probability of a change in character along the bottom two branches is $p_y$.
-In our estimand $t$, we modify the classical topologies by allowing the middle branch to be unequal to either $p_x$ or $p_y$ and denote this probability $p_w$; the parameters $p_x, p_y$ and $p_w$ all lie in interval $[0,1/2]$.
-To obtain branch length parameters, we transform these probabilities using the function $f(p)=1-2p$ so that $t=\{x,y,w\}:=\{f(p_x), f(p_y), f(p_w)\}$; the parameters $x, y$ and $w$ all lie in the interval $[0,1]$.
+Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
+
+\begin{figure}
+\centering
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[width=.95\textwidth]{farris_blank}
+\caption[short]{Farris topology $\tau_1$}
+\end{subfigure}
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[width=.95\textwidth]{felsenstein_blank}
+\caption[short]{Felsenstein topology $\tau_2$}
+\end{subfigure}
+\caption{Two simple topologies}
+\end{figure}
+
+Call $t=\{x,y,w\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ is the case where the bottom three branches all share the same parameter, the classical construction of this topology.
+The branch length parameters are such that the probability of a change in character along the top two branches is $p_x=1-2x$, with corresponding equalities for the other two branches.
+Here, $\{p_x,p_y,p_z\}\in[0,1/2]^3$ and $\{x,y,z\}\in[0,1]^3$.
 
@@
     \hline
-$s_j$   &$P(s_j|\tau_1,t)$&$P(s_j|\tau_2,t)$\\
+$s_j$   &$P(s_j|\tau_1,t^*)$&$P(s_j|\tau_2,t^*)$\\
     \hline
@@
     \hline
-0000&$1+x^2+w^2+4xyw+x^2w^2$&$1+2xw+2xyw+x^2y+w^2y+x^2w^2$\\
-0001&$1+x^2-w^2-x^2w^2$&$1+x^2y-w^2y-x^2w^2$\\
-0010&$1-x^2+w^2-x^2w^2$&$1-x^2y+w^2y-x^2w^2$\\
-0100&$1+x^2-w^2-x^2w^2$&$1+x^2y-w^2y-x^2w^2$\\
-1000&$1-x^2+w^2-x^2w^2$&$1-x^2y+w^2y-x^2w^2$\\
-0011&$1-x^2-w^2+x^2w^2$&$1+2xw-2xyw-x^2y-w^2y+x^2w^2$\\
-0101&$1+x^2+w^2-4xyw+x^2w^2$&$1-2xw-2xyw+x^2y+w^2y+x^2w^2$\\
-1001&$1-x^2-w^2+x^2w^2$&$1-2xw+2xyw-x^2y-w^2y+x^2w^2$\\
+0000&$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
+0001&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
+0010&$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
+0100&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
+1000&$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
+0011&$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
+0101&$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
+1001&$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
     \hline
@@
     \hline
-0000&$(1+x)^2(1+y)(1+w)^{2*}$          &$(1+x)^2(1-y)(1-w)^2$\\
-0001&$(1+x)^2(1+y)(1+w)(1-w)^*$       &$(1+x)^2(1-y)(1+w)(1-w)$\\
-0010&$(1+x)^2(1+y)(1+w)(1-w)^*$       &$(1+x)^2(1-y)(1+w)(1-w)$\\
-0100&$(1+x)(1-x)(1+y)(1+w)^{2*}$       &$(1+x)(1-x)(1-y)(1-w)^2$\\
-1000&$(1+x)(1-x)(1+y)(1+w)^{2*}$       &$(1+x)(1-x)(1-y)(1-w)^2$\\
-0011&$(1+x)(1-x)(1+y)(1+w)(1-w)^{\dagger}$    &$(1+x)(1-x)(1-y)(1+w)(1-w)$\\
-0101&$(1+x)^2(1+y)(1-w)^{2\ddagger}$          &$(1+x)^2(1-y)(1+w)^{2\ddagger}$\\
-1001&$(1+x)(1-x)(1+y)(1+w)(1-w)^{\dagger}$    &$(1+x)(1-x)(1-y)(1+w)(1-w)$\\
+0000&$(1+x)^2   (1+w)(1+y)^{2*}$          &$(1+x)^2   (1-w)(1-y)^2$\\
+0001&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
+0010&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
+0100&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
+1000&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
+0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$   &$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
+1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     \hline
@@
     \hline
-0000&$(1-x)^2(1-y)(1+w)^2$        &$(1-x)^2(1+y)(1-w)^2$\\
-0001&$(1-x)^2(1-y)(1+w)(1-w)$     &$(1-x)^2(1+y)(1+w)(1-w)$\\
-0010&$(1-x)^2(1-y)(1+w)(1-w)$     &$(1-x)^2(1+y)(1+w)(1-w)$\\
-0100&$(1+x)(1-x)(1-y)(1+w)^2$     &$(1+x)(1-x)(1+y)(1-w)^2$\\
-1000&$(1+x)(1-x)(1-y)(1+w)^2$     &$(1+x)(1-x)(1+y)(1-w)^2$\\
-0011&$(1+x)(1-x)(1-y)(1+w)(1-w)$  &$(1+x)(1-x)(1+y)(1+w)(1-w)^{\dagger}$\\
-0101&$(1-x)^2(1-y)(1-w)^2$       &$(1-x)^2(1+y)(1+w)^{2\ddagger}$\\
-1001&$(1+x)(1-x)(1-y)(1+w)(1-w)$  &$(1+x)(1-x)(1+y)(1+w)(1-w)^{\dagger}$\\
+0000&$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
+0001&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+0010&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+0100&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
+1000&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
+0011&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
+0101&$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
+1001&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
     \hline
@@
     \hline
-0000&$(1+x)^2(1+y)(1+w)^{2*}$          &$(1+x)(1-x)(1-y)(1+w)(1-w)$\\
-0001&$(1+x)^2(1+y)(1+w)(1-w)^{\ddagger}$       &$(1+x)(1-x)(1-y)(1+w)^{2\ddagger}$\\
-0010&$(1+x)(1-x)(1+y)(1+w)^{2\ddagger}$       &$(1+x)^2(1-y)(1+w)(1-w)^{\ddagger}$\\
-0100&$(1+x)^2(1+y)(1+w)(1-w)^{\ddagger}$       &$(1+x)(1-x)(1-y)(1-w)^2$\\
-1000&$(1+x)(1-x)(1+y)(1+w)^{2\ddagger}$       &$(1-x)^2(1-y)(1+w)(1-w)$\\
-0011&$(1+x)(1-x)(1+y)(1+w)(1-w)^{\ddagger}$          &$(1+x)^2(1-y)(1+w)^{2\ddagger}$\\
-0101&$(1+x)^2(1+y)(1-w)^{2\ddagger}$    &$(1+x)(1-x)(1-y)(1+w)(1-w)^{\ddagger}$\\
-1001&$(1+x)(1-x)(1+y)(1+w)(1-w)^{\ddagger}$    &$(1-x)^2(1-y)(1+w)^{2\ddagger}$\\
+0000&$(1+x)^2   (1+w)(1+y)^{2*}$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+0001&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$\\
+0010&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$\\
+0100&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1-y)^2$\\
+1000&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1-x)^2   (1-w)(1+y)(1-y)$\\
+0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
+0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$    &$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$\\
+1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1-x)^2   (1-w)(1+y)^{2\ddagger}$\\
     \hline
@@
     \hline
-0000&$(1+x)(1-x)(1-y)(1+w)(1-w)$        &$(1-x)^2(1+y)(1-w)^2$\\
-0001&$(1+x)(1-x)(1-y)(1-w)^2$     &$(1-x)^2(1+y)(1+w)(1-w)$\\
-0010&$(1-x)^2(1-y)(1+w)(1-w)$     &$(1+x)(1-x)(1+y)(1-w)^2$\\
-0100&$(1+x)(1-x)(1-y)(1+w)^{2\ddagger}$     &$(1-x)^2(1+y)(1+w)(1-w)$\\
-1000&$(1+x)^2(1-y)(1+w)(1-w)^{\ddagger}$     &$(1+x)(1-x)(1+y)(1-w)^2$\\
-0011&$(1-x)^2(1-y)(1-w)^2$       &$(1+x)(1-x)(1+y)(1+w)(1-w)^{\ddagger}$\\
-0101&$(1+x)(1-x)(1-y)(1+w)(1-w)^{\ddagger}$  &$(1-x)^2(1+y)(1+w)^{2\ddagger}$\\
-1001&$(1+x)^2(1-y)(1-w)^{2\ddagger}$  &$(1+x)(1-x)(1+y)(1+w)(1-w)^{\ddagger}$\\
+0000&$(1+x)(1-x)(1-w)(1+y)(1-y)$             &$(1-x)^2   (1+w)(1-y)^2$\\
+0001&$(1+x)(1-x)(1-w)(1-y)^2$                &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+0010&$(1-x)^2   (1-w)(1+y)(1-y)$             &$(1+x)(1-x)(1+w)(1-y)^2$\\
+0100&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$      &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+1000&$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$  &$(1+x)(1-x)(1+w)(1-y)^2$\\
+0011&$(1-x)^2   (1-w)(1-y)^2$                &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
+0101&$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$  &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
+1001&$(1+x)^2   (1-w)(1-y)^{2\ddagger}$      &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
 \hline
@@
 To simplify matters, we will first focus on bounding the the terms with the generating probabilities as they are more complicated.
-Let
+Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$ so that
 \begin{align}
 \label{eq:information}
-H(\mathbf{p};x, y, w) &= \sum_{j} P(s_j | \tau^*, \{x^*,y^*\}) \cdot \log P(s_j | \tau', \{x,y,w\}) \nonumber \\ 
-&:= \sum_{j} p(s_j) \cdot \log P(s_j | \tau', \{x,y,w\}).
+H_{\tau_1}(x, y;x', y', w') &= \sum_{j} P(s_j | \tau_1, \{x,y,y\}) \cdot \log P(s_j | \tau', \{x',y',w'\}) \nonumber \\ 
+&:= \sum_{j} p(s_j) \cdot \log P(s_j | \tau', \{x',y',w'\}).
 \end{align}
@@
 $$
-H(\mathbf{p};x, y, w) \le H(\mathbf{p};x^*,y^*, y^*),
+H_{\tau_1}(x, y;x', y', w') \le H_{\tau_1}(x, y;x, y, y),
 $$
@@
 $$
-a = p(0100) + p(1000) + p(0011) + p(1001) = \left\{
-	\begin{array}{ll}
-		(4-2x^{*2}-2y^{*2})/8  & \mbox{Farris zone,} \\
-		(4-2x^{*2}y^{*}-2y^{*3})/8  & \mbox{Felsenstein zone,}
-	\end{array}
-\right.
-$$
-$$
-b = p(0101) = \left\{
-	\begin{array}{ll}
-		(1+x^{*2}+y^{*2}-4x^{*}y^{*2}+x^{*2}y^{*2})/8  & \mbox{Farris zone,} \\
-		(1-2x^{*}y^{*}-2x^{*}y^{*2}+x^{*2}y^{*}+y^{*3}+x^{*2}y^{*2})/8  & \mbox{Felsenstein zone,}
-	\end{array}
-\right.
-$$
-$$
-c = p(0001) + p(0010) + p(0100) + p(1000) = (4-4x^{*2}y^{*2})/8
+a = p(0100) + p(1000) + p(0011) + p(1001) = (4-2x^2-2y^2)/8,
+$$
+$$
+b = p(1000) + p(0010) + p(0011) + p(1001) = (4-2x^2-2y^2)/8,
+$$
+$$
+c = p(0001) + p(0010) + p(0100) + p(1000) = (4-4x^2y^2)/8
 $$
@@
 $$
-d = p(0101) + p(1001) = \left\{
-	\begin{array}{ll}
-		(2-4x^{*}y^{*2}+2x^{*2}y^{*2})/8  & \mbox{Farris zone} \\
-		(2-4x^{*}y^{*}+2x^{*2}y^{*2})/8  & \mbox{Felsenstein zone}
-	\end{array}
-\right.
+d = p(0101) + p(1001) = (2-4xy^2+2x^2y^2)/8
 $$
@@
 $$
-0 \le a,c \le \frac{1}{2}
+0 \le a,b,c \le \frac{1}{2}
 $$
@@
 $$
-0 \le b,d \le \frac{1}{4}.
+0 \le d \le \frac{1}{4}.
 $$
@@
 \label{eq:partial-likelihood}
-L^{(i)}(x, y, w) = H(\mathbf{p}; x, y, w) + L^{p,(i)}(x,y,w)
+L^{(i)}(x', y', w') = H_{\tau_1}(x,y;x',y',w') + L^{p,(i)}(x',y',w')
 \end{equation}
@@
 $$
-H(\mathbf{p}; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(\mathbf{p}; x^{*}, y^{*}, y^{*}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
+H_{\tau_1}(x,y; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(x,y;x,y,y) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
 $$
@@
 
-The likelihood for the Farris zone topology is less complicated than that of the Felsenstein zone since only one site pattern has an ambiguous likelihood---all site patterns except \texttt{0101} have a fixed value for their likelihood when taking the maximum.
-We decompose the full likelihood
-
-where $L_{fa}^{(i)}(x,y,w)$ takes one of three forms:
-$$
-L_{fa}^{(1)} = (2-a)\log(1+x)+a\log(1-x)+(2-a-2b)\log(1+w)+(a+2b)\log(1-w)+\log(1+y),
-$$
-$$
-L_{fa}^{(2)} = (2-a-2b)\log(1+x)+(a+2b)\log(1-x)+(2-a)\log(1+w)+a\log(1-w)+\log(1+y),
-$$
-$$
-L_{fa}^{(3)} = (2-a)\log(1+x)+a\log(1-x)+(2-a)\log(1+w)+a\log(1-w)+(1-b)\log(1+y)+b\log(1-y).
-$$
-These likelihoods have maxima
-$$
-\hat{L}_{fa}^{(1)} = \hat{L}_{fa}^{(2)} = (2-a)\log(2-a)+a\log(a)+(2-a-2b)\log(2-a-2b)+(a+2b)\log(a+2b)+\log(2),
-$$
-$$
-\hat{L}_{fa}^{(3)} = (2-a)\log(2-a)+a\log(a)+(2-a)\log(2-a)+a\log(a)+(1-b)\log(1-b)+b\log(b)+\log(2).
-$$
-
-\textbf{TODO}: cases for $\{x^{*}, y^{*}\}$ where each of the above is the corresponding maximum.
-I think the first is greater than or equal to the second in most cases, though not all (i.e., when $b$ is large but $a$ is small).
+The likelihood for the Farris topology is less complicated than that of the Felsenstein since only one site pattern has an ambiguous likelihood---all site patterns except \texttt{0101} have a fixed value for their likelihood when taking the maximum.
+To fix this value, we will consider an upper bound on the partial likelihood by setting the value for site pattern 0011 to $(1+x)^2(1+w)(1+y)^2$.
+Here the partial likelihood $L_{fa}^{p,(i)}(x,y,w)$ takes the form:
+$$
+L_{fa}^{p} = (2-a)\log(1+x)+a\log(1-x)+(2-b)\log(1+y)+b\log(1-y)+\log(1+w),
+$$
+and this has a maximum
+$$
+\hat{L}_{fa}^{p} = (2-a)\log(2-a)+a\log(a)+(2-b)\log(2-b)+b\log(b)+\log(2).
+$$
 
@@
 
-We construct similar bounds on the Felsenstein likelihood, though to make matters simpler we replace all $(1+y)$ terms in Table~\ref{tab:likelihoods} with $(1-y)$ to obtain a lower bound, performing the reverse substitution to obtain upper bounds.
-As in \eqref{eq:partial-likelihood} we see that the likelihood in this case has one of six forms:
-$$
-L_{fe}^{(1)} = 2\log(1+x)+(2-c-2d)\log(1+w)+(c+2d)\log(1-w)+\log(1-y),
-$$
-$$
-L_{fe}^{(2)} = (2-c-2d)\log(1+x)+(c+2d)\log(1-x)+2\log(1+w)+\log(1-y),
-$$
-$$
-L_{fe}^{(3)} = (2-2d)\log(1+x)+2d\log(1-x)+(2-c)\log(1+w)+c\log(1-w)+\log(1-y),
-$$
-$$
-L_{fe}^{(4)} = (2-c)\log(1+x)+c\log(1-x)+(2-2d)\log(1+w)+2d\log(1-w)+\log(1-y),
-$$
-$$
-L_{fe}^{(5)} = (2-d)\log(1+x)+d\log(1-x)+(2-c-d)\log(1+w)+(c+d)\log(1-w)+\log(1-y)
+We construct similar bounds on the Felsenstein likelihood, though to make matters simpler we replace all $(1+y)$ terms in Table~\ref{tab:likelihoods} with $(1-y)$ to obtain a lower bound.
+Unlike the Farris partial likelihood, the likelihood in this case has one of six forms:
+$$
+L_{fe}^{(1)} = 2\log(1+x)+(2-c-2d)\log(1+y)+(c+2d)\log(1-y)+\log(1-w),
+$$
+$$
+L_{fe}^{(2)} = (2-c-2d)\log(1+x)+(c+2d)\log(1-x)+2\log(1+y)+\log(1-w),
+$$
+$$
+L_{fe}^{(3)} = (2-2d)\log(1+x)+2d\log(1-x)+(2-c)\log(1+y)+c\log(1-y)+\log(1-w),
+$$
+$$
+L_{fe}^{(4)} = (2-c)\log(1+x)+c\log(1-x)+(2-2d)\log(1+y)+2d\log(1-y)+\log(1-w),
+$$
+$$
+L_{fe}^{(5)} = (2-d)\log(1+x)+d\log(1-x)+(2-c-d)\log(1+y)+(c+d)\log(1-y)+\log(1-w)
 $$
@@
 $$
-L_{fe}^{(6)} = (2-c-d)\log(1+x)+(c+d)\log(1-x)+(2-d)\log(1+w)+d\log(1-w)+\log(1-y)
+L_{fe}^{(6)} = (2-c-d)\log(1+x)+(c+d)\log(1-x)+(2-d)\log(1+y)+d\log(1-y)+\log(1-w)
 $$
@@
 
-\textbf{TODO}: cases for $\{x^{*}, y^{*}\}$ where each of the above is the corresponding maximum.
+\textbf{TODO}: cases for $\{x, y\}$ where each of the above is the corresponding maximum.
 I think the first is always greater than or equal to the other two, but I haven't been able to show it yet.
@@
 
-Set $\{x^*, y^*\} = \{0.5, 0.5\}$.
+Set $\{x, y\} = \{0.25, 0.25\}$.
 For a Farris zone generating topology, we have
 $$
-a =  0.375, \ b =  0.1328125,
+a = b = 0.46875
 $$
@@
 $$
-L_{fa}^{(1)} = L_{fa}^{(2)} \approx 1.24637119863,
-$$
-$$
-L_{fa}^{(3)} \approx 1.14372724688
+\hat{L}_{fa}^{p}(\hat{x}, \hat{y}, \hat{w}) \approx -2.17803560694
 $$
@@
 $$
-H_{fa}(\mathbf{p};x^*,y^*, y^*) \approx -1.99591752302.
-$$
-Thus computing the upper bound yields
-$$
-\max_{x,y,w} \ L_{fa}(x,y,w) \le -0.749546324382
+H_{fa}(x,y;x,y,y) \approx -2.07509999739.
+$$
+Computing the upper bound yields
+$$
+\max_{x,y,w} \ L_{fa}(x,y,w) \le -4.25313560433
 $$
@@
 $$
-\max_{x,y,w} \ L_{fe}(x,y,w) \ge -0.749546324382.
+\max_{x,y,w} \ L_{fe}(x,y,w) \ge -4.25313560433
 $$
@@
 $$
-c = 0.46875, \ d = 0.203125,
+c = 0.498046875, \ d = 0.2431640625,
 $$
@@
 $$
-L_{fe}^{(1)} = L_{fe}^{(2)} \approx 1.40196030769, 
-$$
-$$
-L_{fe}^{(3)} = L_{fe}^{(4)} \approx 0.674162530673
+\hat{L}_{fe}(\hat{x}, \hat{y}, \hat{w}) \approx -2.07919739112
+$$
+Calculating shows%
+    %
+    \footnote{It's not a typo that $\hat{L}_{fe}$ and $H_{fe}(x,y; \hat{x}, \hat{y}, \hat{w})$ are the same---at least not in the code.
+        I should try to show why this is the case.
+        }
+    %
+$$
+H_{fe}(x,y; \hat{x}, \hat{y}, \hat{w}) \approx -2.07919739112
 $$
@@
 $$
-L_{fe}^{(5)} = L_2^{(6)} \approx 0.838975450706.
-$$
-Calculating shows
-$$
-H_{fe}(\mathbf{p}; \hat{x}, \hat{y}, \hat{w}) = -2.06377559511,
+\max_{x,y,w} \ L_{fe}(x,y,w) \ge -4.15839478224 > -4.25313560433 \ge \max_{x,y,w} \ L_{fa}(x,y,w)
+$$
+where site patterns are generated from the Farris zone topology.
+Confirming, run \texttt{python}'s numerical optimization solver for $\{.25, .25\}$.
+Since we are computing a constrained maximization, we can find a global maximum using basin hopping, and doing so yields
+$$
+\max_{x,y,w} \ L_{fa}(x,y,w) \approx -4.5101771294218045
 $$
@@
 $$
-\max_{x,y,w} \ L_{fe}(x,y,w) \ge -0.661815287426 > -0.749546324382 \ge \max_{x,y,w} \ L_{fa}(x,y,w)
-$$
-where site patterns are generated from the Farris zone topology.
-Wait, really?
-Did I just show something???
-
-\paragraph{Example}
-
-Say we have found values for $\{x^{*}, y^{*}\}$ where the likelihoods we need to consider are
-$$
-\hat{L}_{fa}(a,b) = (2-a)\log(2-a)+a\log(a)+(2-a-2b)\log(2-a-2b)+(a+2b)\log(a+2b)+\log(2)
-$$
-and
-$$
-\hat{L}_{fe}(c,d) = 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d).
-$$
-This is true in the case of $\{.5, .5\}$ so there is at least measurable support for this case.
-What we need to show now is the support of this intersects with the support of
-$$
-H(\mathbf{p}; 1, 0, 1-c-2d) + \hat{L}_{fe}(c,d) \ge H(\mathbf{p}; x^{*}, y^{*}, y^{*}) + \hat{L}_{fa}(a,b)
-$$
-in some measurable set where data are generated on a Farris topology.
-All of these inequalities are functions of $\{x^{*}, y^{*}\}$ only, and we can cheat a bit and plot the values at which this is true.
+\max_{x,y,w} \ L_{fe}(x,y,w) \approx -4.1544859851260281.
+$$
+Woohoo!
+
+\paragraph{General example}
+
+We want to find values for $\{x,y\}$ where
+$$
+H_{fe}(x,y; 1, 0, 1-c-2d) + \hat{L}_{fe}(c,d) \ge H_{fa}(x,y; x, y, y) + \hat{L}_{fa}(a,b).
+$$
+We can cheat a bit and plot the values at which this is true.
 This may give us some insight into what form these regions of inconsistency take---see Fig.~\ref{fig:inconsistency-farris}.
@@
 
-Some scratch work letting $\{x, y\}=\{x^*,y^*\}$:
-$$
-c+2d = 1-xy^2, \ 2-c-2d = 1+xy^2
-$$
-$$
-H(\mathbf{p}; 1, 0, 1-c-2d) \propto (2-4xy^2+2x^2y^2)\log(1-2xy^2+x^2y^4) + (2+4xy^2+2x^2y^2)\log(1+2xy^2+x^2y^4)+(4-4x^2y^2)\log(1-x^2y^4)
-$$
-$$
-\hat{L}_{fe}(c,d) = 2\log(2) + (1+xy^2)\log(1+xy^2) + (1-xy^2)\log(1-xy^2)
-$$
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \newpage
+
 \section{Temporarily unused text I may want later to show more general results on the likelihood estimates}
+
+Some scratch work:
+$$
+a = b = (4-2x^2-2y^2)/8,
+$$
+$$
+c+2d = 1-xy^2, \ 2-c-2d = 1+xy^2
+$$
+$$
+H_{fe}(x,y; 1, 0, 1-c-2d) \propto (2-4xy^2+2x^2y^2)\log(1-2xy^2+x^2y^4) + (2+4xy^2+2x^2y^2)\log(1+2xy^2+x^2y^4)+(4-4x^2y^2)\log(1-x^2y^4)
+$$
+$$
+H_{fa}(x,y; x, y, y) \propto (2-4xy^2+2x^2y^2)\log(1-2xy^2+x^2y^4) + (2+4xy^2+2x^2y^2)\log(1+2xy^2+x^2y^4)+(4-4x^2y^2)\log(1-x^2y^4)
+$$
+$$
+\hat{L}_{fe} \propto 2\log(2) + (1+xy^2)\log(1+xy^2) + (1-xy^2)\log(1-xy^2)
+$$
+$$
+\hat{L}_{fa} \propto \log(2) + (4-2a)\log(2-a)+2a\log(a).
+$$
 
@@
 
+\bibliographystyle{plain}
+\bibliography{sample}
+
 \end{document}

### 2017-03-18 main.tex
update

@@
 
-\textbf{A note on terminology:}  While we are passing to the limit of infinite data, we don't actually \emph{need} infinite data for these results to be striking.
-Another way of considering the above is what would happen if we knew the ``truth,'' i.e., the true generating topology and branch lengths.
-
 \subsection{Showing inconsistency}
@@
 $$
-where $(\tau, t)$ are the estimands and $(\tau^*, t^*)$ are the fixed, generating parameters.
-For data generated from $(\tau^*, t^*)$, if there exist estimands $(\tau,t)\neq(\tau^*,t^*)$ where
-$$
-J(\tau, t; \tau^*, t^*) > J(\tau^*, t^*; \tau^*, t^*)
-$$
-then we have a possible inconsistency, and the maximum likelihood parameters obtained in this manner may not conform to the truth in the limit.
-The issue with this approach is that we do not guarantee any estimation procedure will recover the values $(\tau, t)$.
-We may show a stronger, more suggestive result if for $\tau'\neq\tau^*$ we have
+where $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
+For data generated from $(\tau^*, t^*)$, if
 \begin{equation}
@@
 \end{equation}
-with $\hat{t}_1$ and $\hat{t}_2$ estimated by maximizing \eqref{eq:profile_likelihood}.
+for $\tau'\neq\tau^*$ with $\hat{t}_1$ and $\hat{t}_2$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
 
@@
 \subsection{Parameter setting}
-
-Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
 
@@
 \caption{Two simple topologies}
+\label{fig:farris-fels-top}
 \end{figure}
 
+Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
 Call $t=\{x,y,w\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ is the case where the bottom three branches all share the same parameter, the classical construction of this topology.
@@
 
-\subsection{Attempts to bound the likelihoods}
-
-Ideally we could show for some generating parameters $t^*$ that there are simple bounds such that
-$$
-J(\tau', \hat{t}_1; \tau^*, t^*) \geq \underline{J}_1 > \overline{J}_2 \geq J(\tau^*, \hat{t}_2; \tau^*, t^*)
-$$
-where the fact that $\underline{J}_1 > \overline{J}_2$ will be obvious.
-This approach could work since we have very particular forms for the prior probabilities and the likelihoods; we would need to find values of $t^*$ that minimize the probability (from the generating parameters) placed on those cases where the likelihood for the generating parameters is higher than those for a candidate parameter.
-The frustrating aspect of this would be deriving a form for $J(\tau, \hat{t}; \tau^*, t^*)$ since it involves constrained maximization over three separate parameters: $x, y$ and $w$.
-
-Since $x$ and $y$ are both in $[0,1]$ we have ready-made bounds available, though they may not be as tight as we want.
-For example, in the case where both $x,y > 1/2$ we can compute upper and lower bounds on the generating probabilities by plugging in the appropriate value to Table~\ref{tab:sitepatprob}.
-
-This is still ad hoc in that we need to fix an upper or lower bound for $x$ and $y$ to obtain these bounds.
-Ideally we would seek tight bounds $(\underline{x}, \overline{x})$ and $(\underline{y}, \overline{y})$, where ``tight'' refers to how close they approximate the likelihoods they are bounding and not the parameters.
-
-\subsection{Simplifying the likelihoods}
+\subsection{Bounding the likelihoods}
 
@@
 \label{eq:information}
-H_{\tau_1}(x, y;x', y', w') &= \sum_{j} P(s_j | \tau_1, \{x,y,y\}) \cdot \log P(s_j | \tau', \{x',y',w'\}) \nonumber \\ 
+H_{fa}(x, y;x', y', w') &= \sum_{j} P(s_j | \tau_1, \{x,y,y\}) \cdot \log P(s_j | \tau', \{x',y',w'\}) \nonumber \\ 
 &:= \sum_{j} p(s_j) \cdot \log P(s_j | \tau', \{x',y',w'\}).
@@
 $$
-H_{\tau_1}(x, y;x', y', w') \le H_{\tau_1}(x, y;x, y, y),
+H_{fa}(x, y;x', y', w') \le H_{fa}(x, y;x, y, y),
 $$
@@
 \label{eq:partial-likelihood}
-L^{(i)}(x', y', w') = H_{\tau_1}(x,y;x',y',w') + L^{p,(i)}(x',y',w')
+L^{(i)}(x', y', w') = H_{fa}(x,y;x',y',w') + L^{p,(i)}(x',y',w')
 \end{equation}
@@
 $$
-H_{\tau_1}(x,y; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(x,y;x,y,y) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
+H_{fa}(x,y; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(x,y;x,y,y) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
 $$
@@
 
-\paragraph{Example}
+\subsubsection{Example}
 
@@
 
-\paragraph{General example}
+\subsubsection{General example}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-
-\section{Temporarily unused text I may want later to show more general results on the likelihood estimates}
-
-Some scratch work:
-$$
-a = b = (4-2x^2-2y^2)/8,
-$$
-$$
-c+2d = 1-xy^2, \ 2-c-2d = 1+xy^2
-$$
-$$
-H_{fe}(x,y; 1, 0, 1-c-2d) \propto (2-4xy^2+2x^2y^2)\log(1-2xy^2+x^2y^4) + (2+4xy^2+2x^2y^2)\log(1+2xy^2+x^2y^4)+(4-4x^2y^2)\log(1-x^2y^4)
-$$
-$$
-H_{fa}(x,y; x, y, y) \propto (2-4xy^2+2x^2y^2)\log(1-2xy^2+x^2y^4) + (2+4xy^2+2x^2y^2)\log(1+2xy^2+x^2y^4)+(4-4x^2y^2)\log(1-x^2y^4)
-$$
-$$
-\hat{L}_{fe} \propto 2\log(2) + (1+xy^2)\log(1+xy^2) + (1-xy^2)\log(1-xy^2)
-$$
-$$
-\hat{L}_{fa} \propto \log(2) + (4-2a)\log(2-a)+2a\log(a).
-$$
-
-\subsubsection{A counterexample in branch lengths}
-
-Let $(\tau^*, t^*)=(\tau_1, \{x,y,x\})$.
-We claim there exist $\{x,y\}$ and $\{x, \tilde{y}\}$ such that
-$$
-L'(\tau^*, t^*; \tau^*, \{x, \tilde{y}\}) > L'(\tau^*, t^*; \tau^*, t^*).
-$$
-The model is given by values in the Table.
-Set $x=0$ to obtain
-$$
-L'(\tau^*, t^*; \tau^*, t^*) \propto 8\log(1+y)
-$$
-and
-$$
-L'(\tau^*, t^*; \tau^*, \{0, \tilde{y}\}) \propto 8\log(1+\tilde{y}).
-$$
-Therefore, if $\tilde{y} > y$ we have an inconsistency.
-
-\begin{table}
-\centering
-\begin{tabular}{|l|l|l|}
-    \hline
-$s_j$   &$P(s_j|\tau_1,t)$&$\max_{z_j} \ P(s_j,z_j|\tau_1,t)$\\
-    \hline
-0000&$1+2x^2+4x^2y+x^4$&$(1+x)^4(1+y)$\\
-0001&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
-0010&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
-0100&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
-1000&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
-0011&$1+2x^2-4x^2y+x^4$&max$\{(1+x)^2(1+y)(1-x)^2,(1+x)^4(1-y)\}$\\
-0101&$1-2x^2+x^4$&$(1+x)^2(1+y)(1-x)^2$\\
-1001&$1-2x^2+x^4$&$(1+x)^2(1+y)(1-x)^2$\\
-    \hline
-\end{tabular}    
-\caption{Counterexample 1.}
-\label{tab:sitepatprob_case1}
-\end{table}
-
-\subsubsection{A counterexample in topologies}
-
-Let $(\tau^*, t^*)=(\tau_1, \{x,y,y\})$.
-We claim there exist $\{x,y\}$ such that
-$$
-L'(\tau^*, t^*; \tau_2, t^*) > L'(\tau^*, t^*; \tau^*, t^*).
-$$
-Set $y=0$ to obtain parameters given in Table.
-We have
-$$
-L'(\tau^*, t^*; \tau^*, t^*) \propto 4(1+x^2)\log(1+x^2)+4(1-x^2)\log(1-x^2)+4(1+x^2)\log((1+x)^2)+4(1-x^2)\log((1+x)(1-x))
-$$
-and
-$$
-L'(\tau^*, t^*; \tau_2, t^*) \propto 3(1+x^2)\log((1+x)^2)+3(1-x^2)\log((1+x)^2)+(1+x^2)\log((1+x)(1-x))+(1-x^2)\log((1+x)(1-x)).
-$$
-The difference being greater than zero, i.e.,
-$$
-L'(\tau^*, t^*; \tau_2, t^*) - L'(\tau^*, t^*; \tau^*, t^*) > 0,
-$$
-simplifies to
-$$
-4x^2(\log(1-x^2)-\log(1+x^2)+\log(1-x)-\log(1+x)) - 2(2\log(1+x^2)+2\log(1-x^2)-\log(1+x)+\log(1-x)) > 0.
-$$
-Let $0 < x < 1/\sqrt{2}$ so that the above inequality becomes
-$$
--\log(1-x^2)-3\log(1+x^2) > 0
-$$
-or
-$$
-(1+x^2)^3(1-x^2) < 1.
-$$
-Applying $x<1/\sqrt{2}$ shows if
-$$
-x < \sqrt{-1+2^{1/3}} \approx 0.5098
-$$
-we have an inconsistency.
-
-\begin{table}
-\centering
-\begin{tabular}{|l|l|l||l|l|}
-    \hline
-$s_j$   &$P(s_j|\tau_1,t)$&$\max_{z_j} \ P(s_j,z_j|\tau_1,t)$&$P(s_j|\tau_2,t)$&$\max_{z_j} \ P(s_j,z_j|\tau_2,t)$\\
-    \hline
-0000&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)^2$\\
-0001&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)^2$\\
-0010&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)^2$\\
-0100&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)^2$\\
-1000&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)^2$\\
-0011&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)(1-x)$\\
-0101&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)^2$\\
-1001&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)(1-x)$\\
-    \hline
-\end{tabular}    
-\caption{Counterexample 2.}
-\label{tab:sitepatprob_case2}
-\end{table}
-
-\subsubsection{More fleshed out counterexamples}
-
-The above counterexamples are highly pathological, but they show there exist whole spaces of parameters yielding inconsistencies.
-It seems as long as $x$ and $y$ are ``small enough'' we see inconsistency.
-
-\subsection{start}
-
-We can calculate the probability of a particular site pattern under this topology from the probability that all character states are equal as
-$$
-P(0000|\tau^*,t) = \frac{1}{8}(1+x^2+z^2+4xyz+x^2z^2)
-$$
-where the probability of any site pattern will be the same as above with the corresponding branch length parameter replaced with its negative.
-For example, if we have the pattern $0001$, since the last branch has length $z$ we replace it with $-z$ and obtain
-$$
-P(0001|\tau^*,t) = \frac{1}{8}(1+x^2-z^2-x^2z^2).
-$$
-All possible site pattern probabilities are given in Table~\ref{tab:sitepatprob}.
-All likelihoods and their maximal values are given in Table~\ref{tab:likelihoods}.
-
-It can be shown (I'll do it later) that for our $t^*$ the right-hand term above is $-16\log 8$.
-Moreover, for $t=\{x,y,z\}$
-\begin{equation}
-\sum_{s} \log P(s | \tau^*, t) + \log P(\xi | s, \tau^*, t) = -16\log 8 + f(x,y,z) + g(x,y,z)
-\end{equation}
-where
-\begin{align}
-f(x,y,z) &= 4\log(1+x)+4\log(1-x)+4\log(1+z)+4\log(1-z)+\\
-& 3\log(1+x^2)+3\log(1+z^2)+\\
-& \log((1+x^2)(1+z^2)+4xyz) + \log((1+x^2)(1+z^2)-4xyz).
-\end{align}
-and
-\begin{align}
-g(x,y,z) &= 10\log(1+x)+10\log(1+z)+4\log(1-x)+4\log(1-z)+7\log(1+y)+\\
-& \left\{ 
-  \begin{array}{l}
-  2\log(1+x)+2\log(1-z)+\log(1+y)\\
-  2\log(1+x)+2\log(1+z)+\log(1-y)\\
-  2\log(1-x)+2\log(1-z)+\log(1+y)\\
-  \end{array} \right.
-\end{align}
-whose component will be chosen based on whichever yields the maximum value for $\log L''$.
-
-\subsection{Two fixed topologies}
-
-\subsection{Special case: $x=z=0$}
-
-Consider now setting all data-generating branch length parameters $t^*$ to zero; this yields equal probabilities of $1/8$ for each site pattern.
-By the inequality of Gibbs we have
-$$
-\sum_{s} P(s|\tau^*,t^*)\log P(s|\tau^*,t^*) \geq \sum_{s} P(s|\tau^*,t^*)\log P(s|\tau^*,t)
-$$
-summing over all site patterns, which is equivalent to
-$$
-\sum_{s} \log P(s|\tau^*,t^*) \geq \sum_{s} \log P(s|\tau^*,t).
-$$
-We must show, then, that there exist $t\neq t^*$ such that
-$$
-\max_{\xi_s} \ \sum_{s} \log P(s | \tau^*, t) + \log P(\xi_s | s, \tau^*, t) > \max_{\xi_s} \sum_{s} \log P(s | \tau^*, t^*) + \log P(\xi_s | s, \tau^*, t^*).
-$$
-Assume for $t$ that $x=z=0$ but $y\neq 0$.
-Clearly site pattern probabilities are still all equal, and the first summand of each side of the inequality can be ignored in this case.
-Call
-$$
-f(x,y,z) = \sum_{s} \log P(\xi_s | s, \tau^*, t=\{x,y,z\}).
-$$
-For branch lengths $\{x,y,z\}$, by considering all eight cases (see Table~\ref{tab:likelihoods}) of site patterns we can show that in terms of branch length \emph{probabilities}%
-	%
-    \footnote{The function $f$ has inverse $f^{-1}(u)=(1-u)/2$ so that the probability of a change along a branch of length $x$ maps to $(1-x)/2$ and the probability of no change maps to $(1+x)/2$.}
-    %
-\begin{align}
-f(x,y,z) \propto & \ 10\log(1+x)+10\log(1+z)+4\log(1-x)+4\log(1-z)+7\log(1+y)+\\
-& \left\{ 
-  \begin{array}{l}
-  2\log(1+x)+2\log(1-z)+\log(1+y)\\
-  2\log(1+x)+2\log(1+z)+\log(1-y)\\
-  2\log(1-x)+2\log(1-z)+\log(1+y)\\
-  \end{array} \right.
-\end{align}
-whose last summand will be chosen based on whichever yields the maximum value and where the proportionality is due to requiring branch length parameters $\{x,y,z\}$ to be multiplied by $1/2$ to map them to probabilities.
-We see that
-$$
-f(x,y,z)|_{x=z=0} = 8\log(1+y)
-$$
-and thus require the constraint
-$$
-8\log(1+y) > 0
-$$
-or that $y$ is any value bounded away from zero, and will obtain a maximum at $y=1$ or $p_y=0$.
-
-\subsection{Special case: $x=z\neq 0$}
-
-Fixing $x=z=0$ is a little too pathological---anything can happen when we're flipping coins to determine terminal characters.
-Let's investigate a relaxed yet still special case where branches to the tips are all equal but allowed to be nonzero.
-So we set $\tau^*=\tau$, $t^*=\{x,x,y\}$ and $t=\{x,x,\tilde{y}\}$ where the same terminal branches are used for both parameter sets with only the internal branch different.
-Removing terms involving only $x$ and redundant $y$ shows that we need only concern ourselves with the objective functions
-\begin{align*}
-L_1(u) =& 8\log(1+u) + \\
-  &(1+2x^2+4x^2y+x^4)\log(1+2x^2+4x^2u+x^4) + \\
-  &(1+2x^2-4x^2y+x^4)\log(1+2x^2-4x^2u+x^4)
-\end{align*}
-if $(1+u)(1-x)^2 > (1-u)(1+x)^2$---or, more simply, if $u > 2x / (1+x^2)$---and
-\begin{align*}
-L_2(u) =& (7-2x^2+4x^2y-x^4)\log(1+u) + \\
-  &(1+2x^2-4x^2y+x^4)\log(1-u) + \\
-  &(1+2x^2+4x^2y+x^4)\log(1+2x^2+4x^2u+x^4) + \\
-  &(1+2x^2-4x^2y+x^4)\log(1+2x^2-4x^2u+x^4)
-\end{align*}
-otherwise.
-We can rule out the case where $\tilde{y} < y$ in $L_1$ by the inequality of Gibbs and using a similar argument along with the inequality
-$$
-1+2x^2-4x^2y+x^4 \le 7-2x^2+4x^2y-x^4
-$$
-implies we can rule out $\tilde{y} < y$ in $L_2$ as well (actually I'm not sure if I believe this anymore; the case of $L_1$ definitely holds, though).
-This leaves us with four sets of inequalities where an inconsistency will arise,
-$$
-\left\{\tilde{y} > y > \frac{2x}{1+x^2}, \ L_1(\tilde{y}) > L_1(y)\right\}, 
-$$
-$$
-\left\{\tilde{y} > \frac{2x}{1+x^2} > y, \ L_1(\tilde{y}) > L_2(y)\right\}, 
-$$
-$$
-\left\{\frac{2x}{1+x^2} > \tilde{y} > y, \ L_2(\tilde{y}) > L_2(y)\right\},
-$$
-and
-$$
-\left\{\frac{2x}{1+x^2} > y > \tilde{y}, \ L_2(\tilde{y}) > L_2(y)\right\}.
-$$
-Simplifying shows
-$$
-L_2(u) = L_1(u) + (1+2x^2-4x^2y+x^4)\log\frac{1-u}{1+u} \le L_1(u).
-$$
-Can we use this fact?
-I believe in my heart of hearts, perhaps foolishly, that the first set of inequalities holds categorically.
-The fact above will then imply the second set of inequalities must also hold.
-The tricky bits are the last two sets of inequalities.
-
-If you'll allow me to indulge in a bit of cheating, I provide a few plots showing---for fixed $x$---the regions where inconsistencies arise.
-It looks like there is some regularity that we must be able to exploit somehow, if only to show that for a fixed $x$ there are inequalities in $y$ and $\tilde{y}$ governing consistency.
-
-%\begin{figure}
-%\centering
-%\includegraphics[width=.3\textwidth]{ineqs-x-0_25}\hfill
-%\includegraphics[width=.3\textwidth]{ineqs-x-0_75}\hfill
-%\includegraphics[width=.3\textwidth]{ineqs-x-0_9}
-%\caption{Various regions of inconsistency for fixed $x$}
-%\end{figure}
-
-Say $\tau_1$ is the ``Felsenstein zone'' tree and $\tau_2$ is the ``Farris zone'' tree, but both trees have a different value for its middle, connecting branch.
-%EM widdershins!! This still depends on how we draw them, no?
-The tips are labeled widdershins proceeding from the top-left tip.
-\begin{verbatim}
-Felsenstein:
-1\      /3
-   >--< 
-2/      \4
-
-Farris:
-1\ /3
-  | 
-2/ \4
-\end{verbatim}
-Our branch length parameters $t=\{x,y,z\}$ will consist of transformed probabilities where $y$ corresponds to this middle branch.
-For $p_x\in[0,1/2]$ defined as
-$$p_x = Pr(\mbox{change occurred along branch})$$
-then $x = 1-2p_x$.
-The parameter $x$ takes values in $[0,1]$ and will appear in likelihood calculations as either $(1+x)$ or $(1-x)$.
-Properties for $y$ and $z$ are identical.
-Assume for the moment that $\tau^*=\tau_1$ and $t^*=\{x,y,z\}$.
-
-\subsubsection{Example site pattern: \texttt{0101}}
-
-It can be shown that the only $A$ for which there is more than one partition for $\tau_2$ is for the site pattern \texttt{0101}.
-In this case
-$$A_1(\tau_1) = \{(1+x)^2(1-z)^2 = (1-x)^2(1+z)^2\},$$
-$$A_2(\tau_1) = \{(1+x)(1+y)(1-z) = (1-x)(1-y)(1+z)\},$$
-$$A_3(\tau_1) = \{(1-x)(1+y)(1+z) = (1+x)(1-y)(1-z)\}$$
-and
-$$A_1(\tau_2) = \{(1-z)^2(1+y) = (1+z)^2(1-y)\},$$
-$$A_2(\tau_2) = \{(1-z)^2(1+x)^2 = (1+z)^2(1-x)^2\},$$
-$$A_3(\tau_2) = \{(1+x)^2(1-y) = (1-x)^2(1+y)\}.$$
-
-Now, the probability of generating this site pattern under $\tau^*$ is given as
-$$
-Pr(\texttt{0101} | \tau^*, t^*) = \frac{1}{8}(1-2xz+y(x^2+z^2-2xz)+x^2z^2).
-$$
-
-\subsubsection{Example site pattern: \texttt{0000}}
-
-In the case of site pattern \texttt{0000} the refinements $A$ for both topologies are single-element partitions.
-In other words, no configuration of branch lengths will result in an ancestral state other than \texttt{00}.
-Here we have the probability
-$$
-Pr(\texttt{0000} | \tau^*, t^*) = \frac{1}{8}(1+2xz+y(x^2+z^2+2xz)+x^2z^2)
-$$
-and the maximum value for $Pr(\xi | \texttt{0000}, \tau, t^*)$ for both $\tau_1$ and $\tau_2$ is
-$$
-Pr(\xi | \texttt{0000}, \tau, t^*) = (1+x)^2(1+z)^2(1+y).
-$$
-
-\subsubsection{All refinements with multi-element partitions}
-
-(show calculations for all refinements that have multiple partitions)
-
-\subsection{Numerical sanity check}
-
-Let's suppose site pattern \texttt{0101} is the only one that drives some disconcert between the generating tree and the maximum likelihood tree.
-For a number of combinations of $\{x,y,z\}$ we need to calculate first
-$$
-Pr(\texttt{0101} | \tau^*, t^*) = \frac{1}{8}(1-2xz+y(x^2+z^2-2xz)+x^2z^2)
-$$
-and the maximum value over each $A_i(\tau)$ of $Pr(\xi | \texttt{0101}, \tau, t^*)$ for $\tau\in\{\tau_1, \tau_2\}$.
-In other words, say we have the triplet $\{x,y,z\}=\{0.8,0.5,0.8\}$.
-Then we have
-$$
-\max_{\xi} Pr(\xi | \texttt{0101}, \tau_1, t^*) = 0.0243, \ \max_{\xi} Pr(\xi | \texttt{0101}, \tau_2, t^*) = 0.6561, \ Pr(\texttt{0101} | \tau^*, t^*) = 0.0162.
-$$
-This doesn't yet tell us anything, as the probabilities for other site patterns may overwhelm this particular one, but gives a general flavor of how these computations will be performed.
-Moreover, a different triplet of $\{x,y,z\}$ could obtain a different maximum value for $Pr(\xi | s, \tau, t^*)$.
-This can be seen with the pattern \texttt{0011} where
-$$
-\max_{\xi} Pr(\xi | \texttt{0011}, \tau_1, t^*) = 0.6561, \ \max_{\xi} Pr(\xi | \texttt{0011}, \tau_2, t^*) = 0.0243, \ Pr(\texttt{0011} | \tau^*, t^*) = 0.1762
-$$
-meaning that if all else were equal---i.e., $\max_{\xi} Pr(\xi | s, \tau, t^*)$ were the same for $\tau_1$ and $\tau_2$ for all $s$---and that the maximum value actually occurs at this particular triplet $\{x,y,z\}$ then
-$$
-L(\tau_1) = 0.928 > 0.119 = L(\tau_2)
-$$
-and we have no contradiction.
-That's a lot of ``if'' statements, so let's investigate in a little more generality.
-
-\subsection{Differing branch lengths}
-
-First, we do not wish to assume $\{x,y,z\}$ are the same for both trees.
-
-\subsubsection{General case}
-
-The generating probability $Pr(s | \tau^*, t^*)$ has a specific form given a Hadamard transformation of the branch length parameters, while the term proportional to $Pr(\xi | s, \tau, t^*)$ relies on taking the maximum over a function with multiple components.
-All of this is easy to write into a computer program, and we can numerically find the maximum value over $\{\theta_1, \ldots, \theta_5\}\in[0,1]$ of each $L(\tau_1)$ and $L(\tau_2)$.
-If $L(\tau_1) < L(\tau_2)$ we see that jointly maximizing branch lengths and ancestral states results in a contradiction.
-
-In sufficient enough generality, we would look for any combination of $\{\tau_1,\tau_2\}$ that would result in this contradiction, but let's first focus on the two from the previous section.
-
-\subsubsection{Classic formulations of topologies}
-
-\begin{table}
-\centering
-\begin{tabular}{|l|l|}
-    \hline
-s   &$P(s)$\\
-    \hline
-0000&$1+x^2+y^2+4xy^2+x^2y^2$\\
-0001&$1+x^2-y^2-x^2y^2$\\
-0010&$1+x^2-y^2-x^2y^2$\\
-0100&$1-x^2+y^2-x^2y^2$\\
-1000&$1-x^2+y^2-x^2y^2$\\
-0011&$1+x^2+y^2-4xy^2+x^2y^2$\\
-0101&$1-x^2-y^2+x^2y^2$\\
-1001&$1-x^2-y^2+x^2y^2$\\
-    \hline
-\end{tabular}    
-\caption{Site pattern probabilities for Farris zone topology.
-All values are multiplied by $1/8$.}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{|l|l|}
-    \hline
-s   &$P(s)$\\
-    \hline
-0000&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
-0001&$1+x^2y-y^3-x^2y^2$\\
-0010&$1-x^2y+y^3-x^2y^2$\\
-0100&$1+x^2y-y^3-x^2y^2$\\
-1000&$1-x^2y+y^3-x^2y^2$\\
-0011&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
-0101&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
-1001&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
-    \hline
-\end{tabular}    
-\caption{Site pattern probabilities for Felsenstein zone topology.
-All values are multiplied by $1/8$.}
-\label{tab:sitepatprob-fels}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{|l|ll|}
-    \hline
-    &00                              &01\\
-    \hline
-0000&$(1+x)^2(1+y)^{3*}$          &$(1+x)^2(1-y)^3$\\
-0001&$(1+x)^2(1+y)^2(1-y)^*$       &$(1+x)^2(1+y)(1-y)^2$\\
-0010&$(1+x)^2(1+y)^2(1-y)^*$       &$(1+x)^2(1+y)(1-y)^2$\\
-0100&$(1+x)(1-x)(1+y)^{3*}$       &$(1+x)(1-x)(1-y)^3$\\
-1000&$(1+x)(1-x)(1+y)^{3*}$       &$(1+x)(1-x)(1-y)^3$\\
-0011&$(1+x)^2(1+y)(1-y)^{2}$          &$(1+x)^2(1-y)(1+y)^{2\ddagger}$\\
-0101&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
-1001&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
-    \hline
-    \hline
-&10                           &11\\
-    \hline
-0000&$(1-x)^2(1+y)^2(1-y)$        &$(1-x)^2(1+y)(1-y)^2$\\
-0001&$(1-x)^2(1+y)(1-y)^2$     &$(1-x)^2(1+y)^2(1-y)$\\
-0010&$(1-x)^2(1+y)(1-y)^2$     &$(1-x)^2(1+y)^2(1-y)$\\
-0100&$(1+x)(1-x)(1+y)^2(1-y)$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
-1000&$(1+x)(1-x)(1+y)^2(1-y)$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
-0011&$(1-x)^2(1-y)^3$       &$(1-x)^2(1+y)^{3\ddagger}$\\
-0101&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
-1001&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
-    \hline
-\end{tabular}    
-\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
-Maxima determined row-wise (i.e., by site pattern).
-Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{|l|ll|}
-    \hline
-    &00                              &01\\
-    \hline
-0000&$(1+x)^2(1+y)^{3*}$          &$(1+x)(1-x)(1+y)(1-y)^2$\\
-0001&$(1+x)^2(1+y)^2(1-y)^{*}$       &$(1+x)(1-x)(1+y)^2(1-y)$\\
-0010&$(1+x)(1-x)(1+y)^{3\ddagger}$       &$(1+x)^2(1+y)(1-y)^{2\ddagger}$\\
-0100&$(1+x)^2(1+y)^2(1-y)^*$       &$(1+x)(1-x)(1-y)^3$\\
-1000&$(1+x)(1-x)(1+y)^{3\ddagger}$       &$(1-x)^2(1+y)(1-y)^2$\\
-0011&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$          &$(1+x)^2(1+y)^2(1-y)$\\
-0101&$(1+x)^2(1+y)(1-y)^{2\ddagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
-1001&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
-    \hline
-    \hline
-&10                           &11\\
-    \hline
-0000&$(1+x)(1-x)(1+y)(1-y)^2$        &$(1-x)^2(1+y)(1-y)^2$\\
-0001&$(1+x)(1-x)(1-y)^3$     &$(1-x)^2(1+y)^2(1-y)$\\
-0010&$(1-x)^2(1+y)(1-y)^2$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
-0100&$(1+x)(1-x)(1+y)^2(1-y)^{\ddagger}$     &$(1-x)^2(1+y)^2(1-y)$\\
-1000&$(1+x)^2(1+y)(1-y)^{2\ddagger}$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
-0011&$(1-x)^2(1-y)^3$       &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
-0101&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1-x)^2(1+y)^{3\ddagger}$\\
-1001&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
-    \hline
-\end{tabular}    
-\caption{Likelihood calculations for all site patterns and internal states of Felsenstein topology.
-Maxima determined row-wise (i.e., by site pattern).
-Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
-\label{tab:likelihoods-fels}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{|l|l|}
-    \hline
-s   &$P(s)$\\
-    \hline
-0000&$1+x^2+z^2+2xyz+x^2y+z^2y+x^2z^2$\\
-0001&$1+x^2-z^2+x^2y-z^2y-x^2z^2$\\
-0010&$1-x^2+z^2-x^2y+z^2y-x^2z^2$\\
-0100&$1+x^2-z^2+x^2y-z^2y-x^2z^2$\\
-1000&$1-x^2+z^2-x^2y+z^2y-x^2z^2$\\
-0011&$1-x^2-z^2-2xyz-x^2y-z^2y+x^2z^2$\\
-0101&$1+x^2+z^2-2xyz+x^2y+z^2y+x^2z^2$\\
-1001&$1-x^2-z^2-2xyz-x^2y-z^2y+x^2z^2$\\
-    \hline
-\end{tabular}    
-\caption{Site pattern probabilities for Felsenstein zone topology.
-All values are multiplied by $1/8$.}
-\label{tab:sitepatprob-fels}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{|l|ll|}
-    \hline
-    &00                              &01\\
-    \hline
-0000&$(1+x)^2(1+y)(1+z)^{2*}$          &$(1+x)(1-x)(1-y)(1+z)(1-z)$\\
-0001&$(1+x)^2(1+y)(1+z)(1-z)^{\ddagger}$       &$(1+x)(1-x)(1-y)(1+z)^{2\ddagger}$\\
-0010&$(1+x)(1-x)(1+y)(1+z)^{2\ddagger}$       &$(1+x)^2(1-y)(1+z)(1-z)^{\ddagger}$\\
-0100&$(1+x)^2(1+y)(1+z)(1-z)^{\ddagger}$       &$(1+x)(1-x)(1-y)(1-z)^2$\\
-1000&$(1+x)(1-x)(1+y)(1+z)^{2\ddagger}$       &$(1-x)^2(1-y)(1+z)(1-z)$\\
-0011&$(1+x)(1-x)(1+y)(1+z)(1-z)^{\ddagger}$          &$(1+x)^2(1-y)(1+z)^{2\ddagger}$\\
-0101&$(1+x)^2(1+y)(1-z)^{2\ddagger}$    &$(1+x)(1-x)(1-y)(1+z)(1-z)^{\ddagger}$\\
-1001&$(1+x)(1-x)(1+y)(1+z)(1-z)^{\dagger}$    &$(1+x)(1-x)(1-y)(1+z)(1-z)$\\
-    \hline
-    \hline
-&10                           &11\\
-    \hline
-0000&$(1+x)(1-x)(1-y)(1+z)(1-z)$        &$(1-x)^2(1+y)(1-z)^2$\\
-0001&$(1+x)(1-x)(1-y)(1-z)^2$     &$(1-x)^2(1+y)(1+z)(1-z)$\\
-0010&$(1-x)^2(1-y)(1+z)(1-z)$     &$(1+x)(1-x)(1+y)(1-z)^2$\\
-0100&$(1+x)(1-x)(1-y)(1+z)^{2\ddagger}$     &$(1-x)^2(1+y)(1+z)(1-z)$\\
-1000&$(1+x)^2(1-y)(1+z)(1-z)^{\ddagger}$     &$(1+x)(1-x)(1+y)(1-z)^2$\\
-0011&$(1-x)^2(1-y)(1-z)^2$       &$(1+x)(1-x)(1+y)(1+z)(1-z)^{\ddagger}$\\
-0101&$(1+x)(1-x)(1-y)(1+z)(1-z)^{\ddagger}$  &$(1-x)^2(1+y)(1+z)^{2\ddagger}$\\
-1001&$(1+x)(1-x)(1-y)(1+z)(1-z)$  &$(1+x)(1-x)(1+y)(1+z)(1-z)^{\dagger}$\\
-    \hline
-\end{tabular}    
-\caption{Likelihood calculations for all site patterns and internal states of Felsenstein topology.
-Maxima determined row-wise (i.e., by site pattern).
-Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
-\label{tab:likelihoods-fels}
-\end{table}
-
-In the first case we are in a similar situation to the case of $x=z=0$ where $\tilde{y}>y$ implies $L(\tilde{y})>L(y)$.
-The second case is more interesting.
-We still require $\tilde{y}>y$, but we now have a further restriction.
-We are interested in the simultaneous inequalities
-$$
-\frac{2x}{1+x^2} > \tilde{y} > y,
-$$
-$$
-[7-2x^2+4x^2y-x^4]\log\frac{1+\tilde{y}}{1+y}+[1+2x^2-4x^2y+x^4]\log\frac{1-\tilde{y}}{1-y} > 0
-$$
-which can be written
-$$
-8\log\frac{1+\tilde{y}}{1+y}+[1+2x^2-4x^2y+x^4]\log\frac{(1-\tilde{y})(1+y)}{(1+\tilde{y})(1-y)} > 0.
-$$
-What can we say about this?
-
-Now terms involving only $P(s|\tau,t)$ differ between $t^*$ and $t$ for the site patterns 0000 and 0101 but no others, allowing us to ignore them.
-Additionally, all terms involving $P(\xi_s|s,\tau,t)$ will differ between $t^*$ and $t$ yielding a search for parameters such that
-$$
-\sum_{s\in\{0000,0101\}} P(s|\tau^*, t^*)\log P(s|\tau^*, t) + \sum_{s} P(s | \tau^*, t^*) \log P(\xi_s | s, \tau^*, t)
-$$
-is maximized at $t\neq t^*$.
-Since $\log P(\xi_s|s,\tau,t)$ breaks up into a sum of terms involving $x$ and $y$ we have
-$$
-(6-4x^2-2x^4)\log(1+y)+\sum_{s\in\{0000,0101\}} P(s|\tau^*, t^*)[\log P(s|\tau^*, t)+\log P(\xi_s|s, \tau^*, t)].
-$$
-For a bit of clarity, let
-$$
-a(x,y) = (1+2x^2+4x^2y+x^4),
-$$
-$$
-b(x,y) = (1+2x^2-4x^2y+x^4)
-$$
-so that we have
-$$
-(6-4x^2-2x^4)\log(1+y)+a(x,y)\log a(x,y)+b(x,y)\log b(x,y) + \sum_{s\in\{0000,0101\}} P(s|\tau^*, t^*)\log P(\xi_s|s, \tau^*, t).
-$$
-Canceling more terms shows the right-hand sum is
-$$
-(1+2x^2+4x^2y+x^4)\log(1+y) + 16x^2y\log(1+x) + P(0101|\tau^*, t^*)\log P(\xi_s|0101, \tau^*, t).
-$$
-That final right-hand term above is actually a component function since we are maximizing a likelihood, and the 0101 row in Table~\ref{tab:likelihoods} does not have a unique maximum value.
-This value is either
-$$
-c_1(x,y) = (1+x)^4(1-y),
-$$
-$$
-c_2(x,y) = (1+x)^2(1-x)^2(1+y).
-$$
-We are interested in two inequalities; if
-$$
-c_2(x,y) > c_1(x,y)
-$$
-then our whole objective function simplifies since many terms with $\log(1+y)$ cancel.
-We can then focus on a second inequality, namely whether
-$$
-8\log(1+y) + x^2y(4\log(1+y)+16\log(1+x)) + a(x,y)\log a(x,y) + b(x,y)\log b(x,y) + c_2(x,y)...blah
-$$
-is greater than
-$$
-8\log(1+\tilde{y}) + x^2y(4\log(1+\tilde{y})+16\log(1+x)) + a(x,y) + b(x,y) + c_2(x,y)...
-$$
-
-\subsection{Dashed intuition}
-
-I am having a tough time seeing the intuition behind this result.
-If $x=z=0$ then we are flipping a coin for what happens at the internal node proceeding from the leaf.
-Why would effectively constraining the interior nodes to be equal result in a larger value for the likelihood?
-It doesn't have anything to do with efficiency, does it (i.e., by constraining the parameters to be equal we are using the same amount of data to estimate fewer parameters)?
-Can we show that constraining any $\xi_i,\xi_j$ to be equal in $L_n$ will also result in problems?
-
-I think what may help my sanity is proceeding with the above for the usual maximum likelihood story, that is, for
-$$
-\log L''' = \sum_{s} P(s | \tau^*, t^*) \cdot [\log P(s | \tau^*, t) + \log\sum_{\xi} P(\xi_s | s, \tau^*, t)]
-$$
-show
-$$
-L'''(\tau^*, t^*) > L'''(\tau^*, t)
-$$
-for $t\neq t^*$.
-At least for $x=z=0$ we can see that these two quantities will be equal for all $y$ due to the symmetry of choosing the internal nodes.
-Is this satisfying?
-
-UPDATE: thinking about this more, it actually makes sense.
-If we are maximizing over internal states, we will always prefer a branch length probability that will indicate no change occurs between them, effectively constraining all internal states to be the same given a particular site pattern.
-As we can see in the table, the maximum value almost always occurs for internal states that are equivalent.
-Seeking harmony between internal states will always increase the likelihood except in cases where branches to terminating leaves are sufficiently small.
-So the question might be, now, whether or not we can prove a general result: since internal states seem to always maximize the likelihood by being constrained to be equal, what conditions are needed on the terminal branches (here $x$ and $z$) that will yield (in)consistent estimates?
-
-\subsection{Plan of attack}
-
-We recall \eqref{eq:inconsistency_inequality} and set $\tau^*=\tau_1$ and $\tau'=\tau_2$.
-In other words, data are generated from the ``Farris zone'' topology, and we seek all values for $t^*$ where we can obtain a higher likelihood by proceeding with joint maximization assuming the ``Felsenstein zone'' topology.
-
-\subsection{The cheater's approach}
-
-Using \texttt{python}, we can numerically maximize \eqref{eq:site_pattern_profile_likelihood_mean} over $t$ separately for fixed topologies $\tau^*$ and $\tau'$.
-For $t^*$, we can calculate these values over a grid of values for $(x^*, y^*)\in[0,1]^2$.
-At worst, this will show us if we're on a fool's errand, but if not then we could benefit by gaining insight into what form a refinement of branch lengths yielding inconsistency would take.
-
 \bibliographystyle{plain}

### 2017-03-18 deprecated.tex
also post deprecated

\section{Temporarily unused text I may want later to show more general results on the likelihood estimates}

Some scratch work:
$$
a = b = (4-2x^2-2y^2)/8,
$$
$$
c+2d = 1-xy^2, \ 2-c-2d = 1+xy^2
$$
$$
H_{fe}(x,y; 1, 0, 1-c-2d) \propto (2-4xy^2+2x^2y^2)\log(1-2xy^2+x^2y^4) + (2+4xy^2+2x^2y^2)\log(1+2xy^2+x^2y^4)+(4-4x^2y^2)\log(1-x^2y^4)
$$
$$
H_{fa}(x,y; x, y, y) \propto (2-4xy^2+2x^2y^2)\log(1-2xy^2+x^2y^4) + (2+4xy^2+2x^2y^2)\log(1+2xy^2+x^2y^4)+(4-4x^2y^2)\log(1-x^2y^4)
$$
$$
\hat{L}_{fe} \propto 2\log(2) + (1+xy^2)\log(1+xy^2) + (1-xy^2)\log(1-xy^2)
$$
$$
\hat{L}_{fa} \propto \log(2) + (4-2a)\log(2-a)+2a\log(a).
$$

\subsubsection{A counterexample in branch lengths}

Let $(\tau^*, t^*)=(\tau_1, \{x,y,x\})$.
We claim there exist $\{x,y\}$ and $\{x, \tilde{y}\}$ such that
$$
L'(\tau^*, t^*; \tau^*, \{x, \tilde{y}\}) > L'(\tau^*, t^*; \tau^*, t^*).
$$
The model is given by values in the Table.
Set $x=0$ to obtain
$$
L'(\tau^*, t^*; \tau^*, t^*) \propto 8\log(1+y)
$$
and
$$
L'(\tau^*, t^*; \tau^*, \{0, \tilde{y}\}) \propto 8\log(1+\tilde{y}).
$$
Therefore, if $\tilde{y} > y$ we have an inconsistency.

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
    \hline
$s_j$   &$P(s_j|\tau_1,t)$&$\max_{z_j} \ P(s_j,z_j|\tau_1,t)$\\
    \hline
0000&$1+2x^2+4x^2y+x^4$&$(1+x)^4(1+y)$\\
0001&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
0010&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
0100&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
1000&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
0011&$1+2x^2-4x^2y+x^4$&max$\{(1+x)^2(1+y)(1-x)^2,(1+x)^4(1-y)\}$\\
0101&$1-2x^2+x^4$&$(1+x)^2(1+y)(1-x)^2$\\
1001&$1-2x^2+x^4$&$(1+x)^2(1+y)(1-x)^2$\\
    \hline
\end{tabular}    
\caption{Counterexample 1.}
\label{tab:sitepatprob_case1}
\end{table}

\subsubsection{A counterexample in topologies}

Let $(\tau^*, t^*)=(\tau_1, \{x,y,y\})$.
We claim there exist $\{x,y\}$ such that
$$
L'(\tau^*, t^*; \tau_2, t^*) > L'(\tau^*, t^*; \tau^*, t^*).
$$
Set $y=0$ to obtain parameters given in Table.
We have
$$
L'(\tau^*, t^*; \tau^*, t^*) \propto 4(1+x^2)\log(1+x^2)+4(1-x^2)\log(1-x^2)+4(1+x^2)\log((1+x)^2)+4(1-x^2)\log((1+x)(1-x))
$$
and
$$
L'(\tau^*, t^*; \tau_2, t^*) \propto 3(1+x^2)\log((1+x)^2)+3(1-x^2)\log((1+x)^2)+(1+x^2)\log((1+x)(1-x))+(1-x^2)\log((1+x)(1-x)).
$$
The difference being greater than zero, i.e.,
$$
L'(\tau^*, t^*; \tau_2, t^*) - L'(\tau^*, t^*; \tau^*, t^*) > 0,
$$
simplifies to
$$
4x^2(\log(1-x^2)-\log(1+x^2)+\log(1-x)-\log(1+x)) - 2(2\log(1+x^2)+2\log(1-x^2)-\log(1+x)+\log(1-x)) > 0.
$$
Let $0 < x < 1/\sqrt{2}$ so that the above inequality becomes
$$
-\log(1-x^2)-3\log(1+x^2) > 0
$$
or
$$
(1+x^2)^3(1-x^2) < 1.
$$
Applying $x<1/\sqrt{2}$ shows if
$$
x < \sqrt{-1+2^{1/3}} \approx 0.5098
$$
we have an inconsistency.

\begin{table}
\centering
\begin{tabular}{|l|l|l||l|l|}
    \hline
$s_j$   &$P(s_j|\tau_1,t)$&$\max_{z_j} \ P(s_j,z_j|\tau_1,t)$&$P(s_j|\tau_2,t)$&$\max_{z_j} \ P(s_j,z_j|\tau_2,t)$\\
    \hline
0000&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)^2$\\
0001&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)^2$\\
0010&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)^2$\\
0100&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)^2$\\
1000&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)^2$\\
0011&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)(1-x)$\\
0101&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)^2$\\
1001&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)(1-x)$\\
    \hline
\end{tabular}    
\caption{Counterexample 2.}
\label{tab:sitepatprob_case2}
\end{table}

\subsubsection{More fleshed out counterexamples}

The above counterexamples are highly pathological, but they show there exist whole spaces of parameters yielding inconsistencies.
It seems as long as $x$ and $y$ are ``small enough'' we see inconsistency.

\subsection{start}

We can calculate the probability of a particular site pattern under this topology from the probability that all character states are equal as
$$
P(0000|\tau^*,t) = \frac{1}{8}(1+x^2+z^2+4xyz+x^2z^2)
$$
where the probability of any site pattern will be the same as above with the corresponding branch length parameter replaced with its negative.
For example, if we have the pattern $0001$, since the last branch has length $z$ we replace it with $-z$ and obtain
$$
P(0001|\tau^*,t) = \frac{1}{8}(1+x^2-z^2-x^2z^2).
$$
All possible site pattern probabilities are given in Table~\ref{tab:sitepatprob}.
All likelihoods and their maximal values are given in Table~\ref{tab:likelihoods}.

It can be shown (I'll do it later) that for our $t^*$ the right-hand term above is $-16\log 8$.
Moreover, for $t=\{x,y,z\}$
\begin{equation}
\sum_{s} \log P(s | \tau^*, t) + \log P(\xi | s, \tau^*, t) = -16\log 8 + f(x,y,z) + g(x,y,z)
\end{equation}
where
\begin{align}
f(x,y,z) &= 4\log(1+x)+4\log(1-x)+4\log(1+z)+4\log(1-z)+\\
& 3\log(1+x^2)+3\log(1+z^2)+\\
& \log((1+x^2)(1+z^2)+4xyz) + \log((1+x^2)(1+z^2)-4xyz).
\end{align}
and
\begin{align}
g(x,y,z) &= 10\log(1+x)+10\log(1+z)+4\log(1-x)+4\log(1-z)+7\log(1+y)+\\
& \left\{ 
  \begin{array}{l}
  2\log(1+x)+2\log(1-z)+\log(1+y)\\
  2\log(1+x)+2\log(1+z)+\log(1-y)\\
  2\log(1-x)+2\log(1-z)+\log(1+y)\\
  \end{array} \right.
\end{align}
whose component will be chosen based on whichever yields the maximum value for $\log L''$.

\subsection{Two fixed topologies}

\subsection{Special case: $x=z=0$}

Consider now setting all data-generating branch length parameters $t^*$ to zero; this yields equal probabilities of $1/8$ for each site pattern.
By the inequality of Gibbs we have
$$
\sum_{s} P(s|\tau^*,t^*)\log P(s|\tau^*,t^*) \geq \sum_{s} P(s|\tau^*,t^*)\log P(s|\tau^*,t)
$$
summing over all site patterns, which is equivalent to
$$
\sum_{s} \log P(s|\tau^*,t^*) \geq \sum_{s} \log P(s|\tau^*,t).
$$
We must show, then, that there exist $t\neq t^*$ such that
$$
\max_{\xi_s} \ \sum_{s} \log P(s | \tau^*, t) + \log P(\xi_s | s, \tau^*, t) > \max_{\xi_s} \sum_{s} \log P(s | \tau^*, t^*) + \log P(\xi_s | s, \tau^*, t^*).
$$
Assume for $t$ that $x=z=0$ but $y\neq 0$.
Clearly site pattern probabilities are still all equal, and the first summand of each side of the inequality can be ignored in this case.
Call
$$
f(x,y,z) = \sum_{s} \log P(\xi_s | s, \tau^*, t=\{x,y,z\}).
$$
For branch lengths $\{x,y,z\}$, by considering all eight cases (see Table~\ref{tab:likelihoods}) of site patterns we can show that in terms of branch length \emph{probabilities}%
	%
    \footnote{The function $f$ has inverse $f^{-1}(u)=(1-u)/2$ so that the probability of a change along a branch of length $x$ maps to $(1-x)/2$ and the probability of no change maps to $(1+x)/2$.}
    %
\begin{align}
f(x,y,z) \propto & \ 10\log(1+x)+10\log(1+z)+4\log(1-x)+4\log(1-z)+7\log(1+y)+\\
& \left\{ 
  \begin{array}{l}
  2\log(1+x)+2\log(1-z)+\log(1+y)\\
  2\log(1+x)+2\log(1+z)+\log(1-y)\\
  2\log(1-x)+2\log(1-z)+\log(1+y)\\
  \end{array} \right.
\end{align}
whose last summand will be chosen based on whichever yields the maximum value and where the proportionality is due to requiring branch length parameters $\{x,y,z\}$ to be multiplied by $1/2$ to map them to probabilities.
We see that
$$
f(x,y,z)|_{x=z=0} = 8\log(1+y)
$$
and thus require the constraint
$$
8\log(1+y) > 0
$$
or that $y$ is any value bounded away from zero, and will obtain a maximum at $y=1$ or $p_y=0$.

\subsection{Special case: $x=z\neq 0$}

Fixing $x=z=0$ is a little too pathological---anything can happen when we're flipping coins to determine terminal characters.
Let's investigate a relaxed yet still special case where branches to the tips are all equal but allowed to be nonzero.
So we set $\tau^*=\tau$, $t^*=\{x,x,y\}$ and $t=\{x,x,\tilde{y}\}$ where the same terminal branches are used for both parameter sets with only the internal branch different.
Removing terms involving only $x$ and redundant $y$ shows that we need only concern ourselves with the objective functions
\begin{align*}
L_1(u) =& 8\log(1+u) + \\
  &(1+2x^2+4x^2y+x^4)\log(1+2x^2+4x^2u+x^4) + \\
  &(1+2x^2-4x^2y+x^4)\log(1+2x^2-4x^2u+x^4)
\end{align*}
if $(1+u)(1-x)^2 > (1-u)(1+x)^2$---or, more simply, if $u > 2x / (1+x^2)$---and
\begin{align*}
L_2(u) =& (7-2x^2+4x^2y-x^4)\log(1+u) + \\
  &(1+2x^2-4x^2y+x^4)\log(1-u) + \\
  &(1+2x^2+4x^2y+x^4)\log(1+2x^2+4x^2u+x^4) + \\
  &(1+2x^2-4x^2y+x^4)\log(1+2x^2-4x^2u+x^4)
\end{align*}
otherwise.
We can rule out the case where $\tilde{y} < y$ in $L_1$ by the inequality of Gibbs and using a similar argument along with the inequality
$$
1+2x^2-4x^2y+x^4 \le 7-2x^2+4x^2y-x^4
$$
implies we can rule out $\tilde{y} < y$ in $L_2$ as well (actually I'm not sure if I believe this anymore; the case of $L_1$ definitely holds, though).
This leaves us with four sets of inequalities where an inconsistency will arise,
$$
\left\{\tilde{y} > y > \frac{2x}{1+x^2}, \ L_1(\tilde{y}) > L_1(y)\right\}, 
$$
$$
\left\{\tilde{y} > \frac{2x}{1+x^2} > y, \ L_1(\tilde{y}) > L_2(y)\right\}, 
$$
$$
\left\{\frac{2x}{1+x^2} > \tilde{y} > y, \ L_2(\tilde{y}) > L_2(y)\right\},
$$
and
$$
\left\{\frac{2x}{1+x^2} > y > \tilde{y}, \ L_2(\tilde{y}) > L_2(y)\right\}.
$$
Simplifying shows
$$
L_2(u) = L_1(u) + (1+2x^2-4x^2y+x^4)\log\frac{1-u}{1+u} \le L_1(u).
$$
Can we use this fact?
I believe in my heart of hearts, perhaps foolishly, that the first set of inequalities holds categorically.
The fact above will then imply the second set of inequalities must also hold.
The tricky bits are the last two sets of inequalities.

If you'll allow me to indulge in a bit of cheating, I provide a few plots showing---for fixed $x$---the regions where inconsistencies arise.
It looks like there is some regularity that we must be able to exploit somehow, if only to show that for a fixed $x$ there are inequalities in $y$ and $\tilde{y}$ governing consistency.

%\begin{figure}
%\centering
%\includegraphics[width=.3\textwidth]{ineqs-x-0_25}\hfill
%\includegraphics[width=.3\textwidth]{ineqs-x-0_75}\hfill
%\includegraphics[width=.3\textwidth]{ineqs-x-0_9}
%\caption{Various regions of inconsistency for fixed $x$}
%\end{figure}

Say $\tau_1$ is the ``Felsenstein zone'' tree and $\tau_2$ is the ``Farris zone'' tree, but both trees have a different value for its middle, connecting branch.
%EM widdershins!! This still depends on how we draw them, no?
The tips are labeled widdershins proceeding from the top-left tip.
\begin{verbatim}
Felsenstein:
1\      /3
   >--< 
2/      \4

Farris:
1\ /3
  | 
2/ \4
\end{verbatim}
Our branch length parameters $t=\{x,y,z\}$ will consist of transformed probabilities where $y$ corresponds to this middle branch.
For $p_x\in[0,1/2]$ defined as
$$p_x = Pr(\mbox{change occurred along branch})$$
then $x = 1-2p_x$.
The parameter $x$ takes values in $[0,1]$ and will appear in likelihood calculations as either $(1+x)$ or $(1-x)$.
Properties for $y$ and $z$ are identical.
Assume for the moment that $\tau^*=\tau_1$ and $t^*=\{x,y,z\}$.

\subsubsection{Example site pattern: \texttt{0101}}

It can be shown that the only $A$ for which there is more than one partition for $\tau_2$ is for the site pattern \texttt{0101}.
In this case
$$A_1(\tau_1) = \{(1+x)^2(1-z)^2 = (1-x)^2(1+z)^2\},$$
$$A_2(\tau_1) = \{(1+x)(1+y)(1-z) = (1-x)(1-y)(1+z)\},$$
$$A_3(\tau_1) = \{(1-x)(1+y)(1+z) = (1+x)(1-y)(1-z)\}$$
and
$$A_1(\tau_2) = \{(1-z)^2(1+y) = (1+z)^2(1-y)\},$$
$$A_2(\tau_2) = \{(1-z)^2(1+x)^2 = (1+z)^2(1-x)^2\},$$
$$A_3(\tau_2) = \{(1+x)^2(1-y) = (1-x)^2(1+y)\}.$$

Now, the probability of generating this site pattern under $\tau^*$ is given as
$$
Pr(\texttt{0101} | \tau^*, t^*) = \frac{1}{8}(1-2xz+y(x^2+z^2-2xz)+x^2z^2).
$$

\subsubsection{Example site pattern: \texttt{0000}}

In the case of site pattern \texttt{0000} the refinements $A$ for both topologies are single-element partitions.
In other words, no configuration of branch lengths will result in an ancestral state other than \texttt{00}.
Here we have the probability
$$
Pr(\texttt{0000} | \tau^*, t^*) = \frac{1}{8}(1+2xz+y(x^2+z^2+2xz)+x^2z^2)
$$
and the maximum value for $Pr(\xi | \texttt{0000}, \tau, t^*)$ for both $\tau_1$ and $\tau_2$ is
$$
Pr(\xi | \texttt{0000}, \tau, t^*) = (1+x)^2(1+z)^2(1+y).
$$

\subsubsection{All refinements with multi-element partitions}

(show calculations for all refinements that have multiple partitions)

\subsection{Numerical sanity check}

Let's suppose site pattern \texttt{0101} is the only one that drives some disconcert between the generating tree and the maximum likelihood tree.
For a number of combinations of $\{x,y,z\}$ we need to calculate first
$$
Pr(\texttt{0101} | \tau^*, t^*) = \frac{1}{8}(1-2xz+y(x^2+z^2-2xz)+x^2z^2)
$$
and the maximum value over each $A_i(\tau)$ of $Pr(\xi | \texttt{0101}, \tau, t^*)$ for $\tau\in\{\tau_1, \tau_2\}$.
In other words, say we have the triplet $\{x,y,z\}=\{0.8,0.5,0.8\}$.
Then we have
$$
\max_{\xi} Pr(\xi | \texttt{0101}, \tau_1, t^*) = 0.0243, \ \max_{\xi} Pr(\xi | \texttt{0101}, \tau_2, t^*) = 0.6561, \ Pr(\texttt{0101} | \tau^*, t^*) = 0.0162.
$$
This doesn't yet tell us anything, as the probabilities for other site patterns may overwhelm this particular one, but gives a general flavor of how these computations will be performed.
Moreover, a different triplet of $\{x,y,z\}$ could obtain a different maximum value for $Pr(\xi | s, \tau, t^*)$.
This can be seen with the pattern \texttt{0011} where
$$
\max_{\xi} Pr(\xi | \texttt{0011}, \tau_1, t^*) = 0.6561, \ \max_{\xi} Pr(\xi | \texttt{0011}, \tau_2, t^*) = 0.0243, \ Pr(\texttt{0011} | \tau^*, t^*) = 0.1762
$$
meaning that if all else were equal---i.e., $\max_{\xi} Pr(\xi | s, \tau, t^*)$ were the same for $\tau_1$ and $\tau_2$ for all $s$---and that the maximum value actually occurs at this particular triplet $\{x,y,z\}$ then
$$
L(\tau_1) = 0.928 > 0.119 = L(\tau_2)
$$
and we have no contradiction.
That's a lot of ``if'' statements, so let's investigate in a little more generality.

\subsection{Differing branch lengths}

First, we do not wish to assume $\{x,y,z\}$ are the same for both trees.

\subsubsection{General case}

The generating probability $Pr(s | \tau^*, t^*)$ has a specific form given a Hadamard transformation of the branch length parameters, while the term proportional to $Pr(\xi | s, \tau, t^*)$ relies on taking the maximum over a function with multiple components.
All of this is easy to write into a computer program, and we can numerically find the maximum value over $\{\theta_1, \ldots, \theta_5\}\in[0,1]$ of each $L(\tau_1)$ and $L(\tau_2)$.
If $L(\tau_1) < L(\tau_2)$ we see that jointly maximizing branch lengths and ancestral states results in a contradiction.

In sufficient enough generality, we would look for any combination of $\{\tau_1,\tau_2\}$ that would result in this contradiction, but let's first focus on the two from the previous section.

\subsubsection{Classic formulations of topologies}

\begin{table}
\centering
\begin{tabular}{|l|l|}
    \hline
s   &$P(s)$\\
    \hline
0000&$1+x^2+y^2+4xy^2+x^2y^2$\\
0001&$1+x^2-y^2-x^2y^2$\\
0010&$1+x^2-y^2-x^2y^2$\\
0100&$1-x^2+y^2-x^2y^2$\\
1000&$1-x^2+y^2-x^2y^2$\\
0011&$1+x^2+y^2-4xy^2+x^2y^2$\\
0101&$1-x^2-y^2+x^2y^2$\\
1001&$1-x^2-y^2+x^2y^2$\\
    \hline
\end{tabular}    
\caption{Site pattern probabilities for Farris zone topology.
All values are multiplied by $1/8$.}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|l|}
    \hline
s   &$P(s)$\\
    \hline
0000&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
0001&$1+x^2y-y^3-x^2y^2$\\
0010&$1-x^2y+y^3-x^2y^2$\\
0100&$1+x^2y-y^3-x^2y^2$\\
1000&$1-x^2y+y^3-x^2y^2$\\
0011&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
0101&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
1001&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
    \hline
\end{tabular}    
\caption{Site pattern probabilities for Felsenstein zone topology.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob-fels}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \hline
    &00                              &01\\
    \hline
0000&$(1+x)^2(1+y)^{3*}$          &$(1+x)^2(1-y)^3$\\
0001&$(1+x)^2(1+y)^2(1-y)^*$       &$(1+x)^2(1+y)(1-y)^2$\\
0010&$(1+x)^2(1+y)^2(1-y)^*$       &$(1+x)^2(1+y)(1-y)^2$\\
0100&$(1+x)(1-x)(1+y)^{3*}$       &$(1+x)(1-x)(1-y)^3$\\
1000&$(1+x)(1-x)(1+y)^{3*}$       &$(1+x)(1-x)(1-y)^3$\\
0011&$(1+x)^2(1+y)(1-y)^{2}$          &$(1+x)^2(1-y)(1+y)^{2\ddagger}$\\
0101&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
1001&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
    \hline
    \hline
&10                           &11\\
    \hline
0000&$(1-x)^2(1+y)^2(1-y)$        &$(1-x)^2(1+y)(1-y)^2$\\
0001&$(1-x)^2(1+y)(1-y)^2$     &$(1-x)^2(1+y)^2(1-y)$\\
0010&$(1-x)^2(1+y)(1-y)^2$     &$(1-x)^2(1+y)^2(1-y)$\\
0100&$(1+x)(1-x)(1+y)^2(1-y)$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
1000&$(1+x)(1-x)(1+y)^2(1-y)$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
0011&$(1-x)^2(1-y)^3$       &$(1-x)^2(1+y)^{3\ddagger}$\\
0101&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
1001&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
    \hline
\end{tabular}    
\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
Maxima determined row-wise (i.e., by site pattern).
Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \hline
    &00                              &01\\
    \hline
0000&$(1+x)^2(1+y)^{3*}$          &$(1+x)(1-x)(1+y)(1-y)^2$\\
0001&$(1+x)^2(1+y)^2(1-y)^{*}$       &$(1+x)(1-x)(1+y)^2(1-y)$\\
0010&$(1+x)(1-x)(1+y)^{3\ddagger}$       &$(1+x)^2(1+y)(1-y)^{2\ddagger}$\\
0100&$(1+x)^2(1+y)^2(1-y)^*$       &$(1+x)(1-x)(1-y)^3$\\
1000&$(1+x)(1-x)(1+y)^{3\ddagger}$       &$(1-x)^2(1+y)(1-y)^2$\\
0011&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$          &$(1+x)^2(1+y)^2(1-y)$\\
0101&$(1+x)^2(1+y)(1-y)^{2\ddagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
1001&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
    \hline
    \hline
&10                           &11\\
    \hline
0000&$(1+x)(1-x)(1+y)(1-y)^2$        &$(1-x)^2(1+y)(1-y)^2$\\
0001&$(1+x)(1-x)(1-y)^3$     &$(1-x)^2(1+y)^2(1-y)$\\
0010&$(1-x)^2(1+y)(1-y)^2$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
0100&$(1+x)(1-x)(1+y)^2(1-y)^{\ddagger}$     &$(1-x)^2(1+y)^2(1-y)$\\
1000&$(1+x)^2(1+y)(1-y)^{2\ddagger}$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
0011&$(1-x)^2(1-y)^3$       &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
0101&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1-x)^2(1+y)^{3\ddagger}$\\
1001&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
    \hline
\end{tabular}    
\caption{Likelihood calculations for all site patterns and internal states of Felsenstein topology.
Maxima determined row-wise (i.e., by site pattern).
Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
\label{tab:likelihoods-fels}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|l|}
    \hline
s   &$P(s)$\\
    \hline
0000&$1+x^2+z^2+2xyz+x^2y+z^2y+x^2z^2$\\
0001&$1+x^2-z^2+x^2y-z^2y-x^2z^2$\\
0010&$1-x^2+z^2-x^2y+z^2y-x^2z^2$\\
0100&$1+x^2-z^2+x^2y-z^2y-x^2z^2$\\
1000&$1-x^2+z^2-x^2y+z^2y-x^2z^2$\\
0011&$1-x^2-z^2-2xyz-x^2y-z^2y+x^2z^2$\\
0101&$1+x^2+z^2-2xyz+x^2y+z^2y+x^2z^2$\\
1001&$1-x^2-z^2-2xyz-x^2y-z^2y+x^2z^2$\\
    \hline
\end{tabular}    
\caption{Site pattern probabilities for Felsenstein zone topology.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob-fels}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \hline
    &00                              &01\\
    \hline
0000&$(1+x)^2(1+y)(1+z)^{2*}$          &$(1+x)(1-x)(1-y)(1+z)(1-z)$\\
0001&$(1+x)^2(1+y)(1+z)(1-z)^{\ddagger}$       &$(1+x)(1-x)(1-y)(1+z)^{2\ddagger}$\\
0010&$(1+x)(1-x)(1+y)(1+z)^{2\ddagger}$       &$(1+x)^2(1-y)(1+z)(1-z)^{\ddagger}$\\
0100&$(1+x)^2(1+y)(1+z)(1-z)^{\ddagger}$       &$(1+x)(1-x)(1-y)(1-z)^2$\\
1000&$(1+x)(1-x)(1+y)(1+z)^{2\ddagger}$       &$(1-x)^2(1-y)(1+z)(1-z)$\\
0011&$(1+x)(1-x)(1+y)(1+z)(1-z)^{\ddagger}$          &$(1+x)^2(1-y)(1+z)^{2\ddagger}$\\
0101&$(1+x)^2(1+y)(1-z)^{2\ddagger}$    &$(1+x)(1-x)(1-y)(1+z)(1-z)^{\ddagger}$\\
1001&$(1+x)(1-x)(1+y)(1+z)(1-z)^{\dagger}$    &$(1+x)(1-x)(1-y)(1+z)(1-z)$\\
    \hline
    \hline
&10                           &11\\
    \hline
0000&$(1+x)(1-x)(1-y)(1+z)(1-z)$        &$(1-x)^2(1+y)(1-z)^2$\\
0001&$(1+x)(1-x)(1-y)(1-z)^2$     &$(1-x)^2(1+y)(1+z)(1-z)$\\
0010&$(1-x)^2(1-y)(1+z)(1-z)$     &$(1+x)(1-x)(1+y)(1-z)^2$\\
0100&$(1+x)(1-x)(1-y)(1+z)^{2\ddagger}$     &$(1-x)^2(1+y)(1+z)(1-z)$\\
1000&$(1+x)^2(1-y)(1+z)(1-z)^{\ddagger}$     &$(1+x)(1-x)(1+y)(1-z)^2$\\
0011&$(1-x)^2(1-y)(1-z)^2$       &$(1+x)(1-x)(1+y)(1+z)(1-z)^{\ddagger}$\\
0101&$(1+x)(1-x)(1-y)(1+z)(1-z)^{\ddagger}$  &$(1-x)^2(1+y)(1+z)^{2\ddagger}$\\
1001&$(1+x)(1-x)(1-y)(1+z)(1-z)$  &$(1+x)(1-x)(1+y)(1+z)(1-z)^{\dagger}$\\
    \hline
\end{tabular}    
\caption{Likelihood calculations for all site patterns and internal states of Felsenstein topology.
Maxima determined row-wise (i.e., by site pattern).
Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
\label{tab:likelihoods-fels}
\end{table}

In the first case we are in a similar situation to the case of $x=z=0$ where $\tilde{y}>y$ implies $L(\tilde{y})>L(y)$.
The second case is more interesting.
We still require $\tilde{y}>y$, but we now have a further restriction.
We are interested in the simultaneous inequalities
$$
\frac{2x}{1+x^2} > \tilde{y} > y,
$$
$$
[7-2x^2+4x^2y-x^4]\log\frac{1+\tilde{y}}{1+y}+[1+2x^2-4x^2y+x^4]\log\frac{1-\tilde{y}}{1-y} > 0
$$
which can be written
$$
8\log\frac{1+\tilde{y}}{1+y}+[1+2x^2-4x^2y+x^4]\log\frac{(1-\tilde{y})(1+y)}{(1+\tilde{y})(1-y)} > 0.
$$
What can we say about this?

Now terms involving only $P(s|\tau,t)$ differ between $t^*$ and $t$ for the site patterns 0000 and 0101 but no others, allowing us to ignore them.
Additionally, all terms involving $P(\xi_s|s,\tau,t)$ will differ between $t^*$ and $t$ yielding a search for parameters such that
$$
\sum_{s\in\{0000,0101\}} P(s|\tau^*, t^*)\log P(s|\tau^*, t) + \sum_{s} P(s | \tau^*, t^*) \log P(\xi_s | s, \tau^*, t)
$$
is maximized at $t\neq t^*$.
Since $\log P(\xi_s|s,\tau,t)$ breaks up into a sum of terms involving $x$ and $y$ we have
$$
(6-4x^2-2x^4)\log(1+y)+\sum_{s\in\{0000,0101\}} P(s|\tau^*, t^*)[\log P(s|\tau^*, t)+\log P(\xi_s|s, \tau^*, t)].
$$
For a bit of clarity, let
$$
a(x,y) = (1+2x^2+4x^2y+x^4),
$$
$$
b(x,y) = (1+2x^2-4x^2y+x^4)
$$
so that we have
$$
(6-4x^2-2x^4)\log(1+y)+a(x,y)\log a(x,y)+b(x,y)\log b(x,y) + \sum_{s\in\{0000,0101\}} P(s|\tau^*, t^*)\log P(\xi_s|s, \tau^*, t).
$$
Canceling more terms shows the right-hand sum is
$$
(1+2x^2+4x^2y+x^4)\log(1+y) + 16x^2y\log(1+x) + P(0101|\tau^*, t^*)\log P(\xi_s|0101, \tau^*, t).
$$
That final right-hand term above is actually a component function since we are maximizing a likelihood, and the 0101 row in Table~\ref{tab:likelihoods} does not have a unique maximum value.
This value is either
$$
c_1(x,y) = (1+x)^4(1-y),
$$
$$
c_2(x,y) = (1+x)^2(1-x)^2(1+y).
$$
We are interested in two inequalities; if
$$
c_2(x,y) > c_1(x,y)
$$
then our whole objective function simplifies since many terms with $\log(1+y)$ cancel.
We can then focus on a second inequality, namely whether
$$
8\log(1+y) + x^2y(4\log(1+y)+16\log(1+x)) + a(x,y)\log a(x,y) + b(x,y)\log b(x,y) + c_2(x,y)...blah
$$
is greater than
$$
8\log(1+\tilde{y}) + x^2y(4\log(1+\tilde{y})+16\log(1+x)) + a(x,y) + b(x,y) + c_2(x,y)...
$$

\subsection{Dashed intuition}

I am having a tough time seeing the intuition behind this result.
If $x=z=0$ then we are flipping a coin for what happens at the internal node proceeding from the leaf.
Why would effectively constraining the interior nodes to be equal result in a larger value for the likelihood?
It doesn't have anything to do with efficiency, does it (i.e., by constraining the parameters to be equal we are using the same amount of data to estimate fewer parameters)?
Can we show that constraining any $\xi_i,\xi_j$ to be equal in $L_n$ will also result in problems?

I think what may help my sanity is proceeding with the above for the usual maximum likelihood story, that is, for
$$
\log L''' = \sum_{s} P(s | \tau^*, t^*) \cdot [\log P(s | \tau^*, t) + \log\sum_{\xi} P(\xi_s | s, \tau^*, t)]
$$
show
$$
L'''(\tau^*, t^*) > L'''(\tau^*, t)
$$
for $t\neq t^*$.
At least for $x=z=0$ we can see that these two quantities will be equal for all $y$ due to the symmetry of choosing the internal nodes.
Is this satisfying?

UPDATE: thinking about this more, it actually makes sense.
If we are maximizing over internal states, we will always prefer a branch length probability that will indicate no change occurs between them, effectively constraining all internal states to be the same given a particular site pattern.
As we can see in the table, the maximum value almost always occurs for internal states that are equivalent.
Seeking harmony between internal states will always increase the likelihood except in cases where branches to terminating leaves are sufficiently small.
So the question might be, now, whether or not we can prove a general result: since internal states seem to always maximize the likelihood by being constrained to be equal, what conditions are needed on the terminal branches (here $x$ and $z$) that will yield (in)consistent estimates?

\subsection{Plan of attack}

We recall \eqref{eq:inconsistency_inequality} and set $\tau^*=\tau_1$ and $\tau'=\tau_2$.
In other words, data are generated from the ``Farris zone'' topology, and we seek all values for $t^*$ where we can obtain a higher likelihood by proceeding with joint maximization assuming the ``Felsenstein zone'' topology.

\subsection{The cheater's approach}

Using \texttt{python}, we can numerically maximize \eqref{eq:site_pattern_profile_likelihood_mean} over $t$ separately for fixed topologies $\tau^*$ and $\tau'$.
For $t^*$, we can calculate these values over a grid of values for $(x^*, y^*)\in[0,1]^2$.
At worst, this will show us if we're on a fool's errand, but if not then we could benefit by gaining insight into what form a refinement of branch lengths yielding inconsistency would take.

### 2017-03-20 main.tex
also post deprecated
update

@@
 \label{eq:information}
-H_{fa}(x, y;x', y', w') &= \sum_{j} P(s_j | \tau_1, \{x,y,y\}) \cdot \log P(s_j | \tau', \{x',y',w'\}) \nonumber \\ 
+H_{\tau'}(x, y;x', y', w') &= \sum_{j} P(s_j | \tau_1, \{x,y,y\}) \cdot \log P(s_j | \tau', \{x',y',w'\}) \nonumber \\ 
 &:= \sum_{j} p(s_j) \cdot \log P(s_j | \tau', \{x',y',w'\}).
@@
 $$
-H_{fa}(x, y;x', y', w') \le H_{fa}(x, y;x, y, y),
+H_{\tau'}(x, y;x', y', w') \le H_{\tau'}(x, y;x, y, y),
 $$
 though this bound can be made tighter.
+
+\textbf{Show}: where these simplifications come about.
+It's just adding/subtracting $\log$ terms, but could be good to walk through.
+
 In our simplifications we will make use of the constants
@@
 \label{eq:partial-likelihood}
-L^{(i)}(x', y', w') = H_{fa}(x,y;x',y',w') + L^{p,(i)}(x',y',w')
+L^{(i)}(x', y', w') = H_{\tau'}(x,y;x',y',w') + L^{p,(i)}(x',y',w')
 \end{equation}
@@
 $$
-H_{fa}(x,y; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(x,y;x,y,y) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
+H_{\tau'}(x,y; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(x,y;x,y,y) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
 $$

### 2017-03-21 deprecated.tex
getting close

@@
 $$
+
+\subsubsection{Refinements}
+
+It will be an undertaking, but if we want to approach this using the refinements we might get something pretty nifty.
+Everything will be quite messy, but really the entire approach boils down to a series of nested inequalities in two variables.
+We may need a program that can reduce symbolic equations, but this will at least get us somewhere that isn't a lot of loose approximations---initial results using numerical optimization show the region in~\ref{fig:inconsistency-farris} is conservative, and very much so.
+See Fig.~\ref{fig:numerical-inconsistency-farris}.
+
+Perhaps something along the lines of a more general results would be interesting?
+Like, could we calculate the measure/area of inconsistency using cooler techniques?
+Could we do this for a general topology?
+Would that even be interesting since we've already found a counterexample?
+
+It also seems that, in the $\{x, y\}=\{.25, .25\}$ example, the lower bound is tight while the upper bound is looser.
+This surprises me since we replaced a lot in the likelihood, but this might indicate that the $H$ term is actually the one that varies more than the $L$ term.
+I guess this is intuitive: the difference between the lower and upper bound on $L$ is $\log2\approx0.69$.
+Maybe getting a tighter upper bound will be useful here?
+
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{ineqs-num-far-gen}
+\caption{Regions of inconsistency---numerically obtained}
+\label{fig:numerical-inconsistency-farris}
+\end{figure}
+
+\subsubsection{Tighter upper bound}
+
+Call
+$$
+e = p(0001) + p(0100) = \frac{1+x^2-y^2-x^2y^2}{4},
+$$
+$$
+f = p(0010) + p(1000) = \frac{1-x^2+y^2-x^2y^2}{4}.
+$$
+Then some algebra and standard $\log$ bounds can be used to get an upper bound for the full Farris likelihood $L^{(i)}$ as
+$$
+(2-a+e)\log(2-\frac{2a}{2+e}) + a\log\frac{2a}{2+e} + (2-b+f)\log(2-\frac{2b}{2+f}) + a\log\frac{2b}{2+f}.
+$$
+In our example of $\{x,y\}=\{.25,.25\}$ this bound is $a$ compared to $-4.25313560433$ before.
 

### 2017-03-21 main.tex
getting close

@@
 $$
-H_{\tau'}(x,y; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(x,y;x,y,y) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
+H_{\tau'}(x,y; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H_{\tau'}(x,y;x,y,y) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
 $$
@@
 $$
-H_{fe}(x,y; 1, 0, 1-c-2d) + \hat{L}_{fe}(c,d) \ge H_{fa}(x,y; x, y, y) + \hat{L}_{fa}(a,b).
+H_{fe}(x,y; 1, 1-c-2d, 0) + \hat{L}_{fe}(c,d) \ge H_{fa}(x,y; x, y, y) + \hat{L}_{fa}(a,b).
 $$
-We can cheat a bit and plot the values at which this is true.
-This may give us some insight into what form these regions of inconsistency take---see Fig.~\ref{fig:inconsistency-farris}.
+Since this is a single inequality in two variables, we can plot it analytically; we need the region where
+\begin{align*}
+    0 &\ge H_{fa}(x,y; x, y, y) + \hat{L}_{fa}(a,b) - H_{fe}(x,y; 1, 1-c-2d, 0) - \hat{L}_{fe}(c,d) \\
+      & = 2(12+2x^2+2y^2)\log((12+2x^2+2y^2)/8) \\
+      &\qquad + 2(4-2x^2-2y^2)\log((4-2x^2-2y^2)/8)  + 8\log2 \\
+      &\qquad + 2(1+x^2-y^2-x^2y^2)\log(1+x^2-y^2-x^2y^2) \\
+      &\qquad + 2(1-x^2+y^2-x^2y^2)\log(1-x^2+y^2-x^2y^2) \\
+      &\qquad + 2(1-x^2-y^2+x^2y^2)\log(1-x^2-y^2+x^2y^2) \\
+      &\qquad + (1+x^2+y^2+4xy^2+x^2y^2)\log(1+x^2+y^2+4xy^2+x^2y^2) \\
+      &\qquad + (1+x^2+y^2-4xy^2+x^2y^2)\log(1+x^2+y^2-4xy^2+x^2y^2) \\
+      &\qquad - 16\log2 - 8(1+xy^2)\log(1+xy^2) - 8(1-xy^2)\log(1-xy^2) \\
+      &\qquad - (2-4xy^2+2x^2y^2)\log(1-2xy^2+x^2y^4) \\
+      &\qquad - (2+4xy^2+2x^2y^2)\log(1+2xy^2+x^2y^4) \\
+      &\qquad - (4-4x^2y^2)\log(1-x^2y^4).
+\end{align*}
+The \texttt{sage} function in listing~\ref{list:sage} will generate Fig.~\ref{fig:inconsistency-farris}.
+
+\begin{table}
+    \begin{center}
+\begin{verbatim}
+var('x,y')
+inconsistency = 2*(12+2*x^2+2*y^2)*log((12+2*x^2+2*y^2)/8) \
+    + 2*(4-2*x^2-2*y^2)*log((4-2*x^2-2*y^2)/8) + 8*log(2.) \
+    + 2*(1+x^2-y^2-x^2*y^2)*log(1+x^2-y^2-x^2*y^2) \
+    + 2*(1-x^2+y^2-x^2*y^2)*log(1-x^2+y^2-x^2*y^2) \
+    + 2*(1-x^2-y^2+x^2*y^2)*log(1-x^2-y^2+x^2*y^2) \
+    + (1+x^2+y^2+4*x*y^2+x^2*y^2)*log(1+x^2+y^2+4*x*y^2+x^2*y^2) \
+    + (1+x^2+y^2-4*x*y^2+x^2*y^2)*log(1+x^2+y^2-4*x*y^2+x^2*y^2) \
+    - 16*log(2.) - 8*(1+x*y^2)*log(1+x*y^2) - 8*(1-x*y^2)*log(1-x*y^2) \
+    - (2-4*x*y^2+2*x^2*y^2)*log(1-2*x*y^2+x^2*y^4) \
+    - (2+4*x*y^2+2*x^2*y^2)*log(1+2*x*y^2+x^2*y^4) - (4-4*x^2*y^2)*log(1-x^2*y^4)
+
+inconsistency_plot = implicit_plot(
+    inconsistency(x,y)==0,
+    (x,0,1),
+    (y,0,1),
+    axes_labels=['$x$', '$y$'],
+    title='Region of inconsistency for Farris-generating topology',
+    color='black'
+)
+\end{verbatim}
+    \end{center}
+\caption{Sage code to generate Fig.~\ref{fig:inconsistency-farris}}
+\label{list:sage}
+\end{table}
 
@@
 \centering
-\includegraphics[width=.9\textwidth]{ineqs-max-far-gen}
+\includegraphics[width=.9\textwidth]{analytic-inconsistency}
 \caption{Regions of inconsistency}

### 2017-03-21 main.tex
rearranging

@@
 
-To simplify matters, we will first focus on bounding the the terms with the generating probabilities as they are more complicated.
-Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$ so that
+Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
+Our approach will be to decompose the likelihood in \eqref{eq:site_pattern_profile_likelihood_mean} as
+\begin{equation}
+\label{eq:partial-likelihood}
+L^{(i)}(x', y', w') = H(x,y,\tau_1;x',y',w', \tau') + L^{p,(i)}(x',y',w')
+\end{equation}
+where the first term is
 \begin{align}
 \label{eq:information}
-H_{\tau'}(x, y;x', y', w') &= \sum_{j} P(s_j | \tau_1, \{x,y,y\}) \cdot \log P(s_j | \tau', \{x',y',w'\}) \nonumber \\ 
+H(x,y,\tau_1;x',y',w',\tau') &= \sum_{j} P(s_j | \tau_1, \{x,y,y\}) \cdot \log P(s_j | \tau', \{x',y',w'\}) \nonumber \\ 
 &:= \sum_{j} p(s_j) \cdot \log P(s_j | \tau', \{x',y',w'\}).
 \end{align}
-By Gibbs inequality,
+and the second term will be a weighted sum of the $\log$s of values from Table~\ref{tab:likelihoods}, the superscript indicating which choices for ancestral state likelihood values we choose.
+
+To bound the first term, we can make use of Gibbs inequality
 $$
-H_{\tau'}(x, y;x', y', w') \le H_{\tau'}(x, y;x, y, y),
+H(x, y, \tau_1;x', y', w', \tau') \le H(x, y, \tau_1;x, y, y, \tau_1),
 $$
 though this bound can be made tighter.
-
-\textbf{Show}: where these simplifications come about.
-It's just adding/subtracting $\log$ terms, but could be good to walk through.
-
-In our simplifications we will make use of the constants
-$$
-a = p(0100) + p(1000) + p(0011) + p(1001) = (4-2x^2-2y^2)/8,
-$$
-$$
-b = p(1000) + p(0010) + p(0011) + p(1001) = (4-2x^2-2y^2)/8,
-$$
-$$
-c = p(0001) + p(0010) + p(0100) + p(1000) = (4-4x^2y^2)/8
-$$
-and
-$$
-d = p(0101) + p(1001) = (2-4xy^2+2x^2y^2)/8
-$$
-with corresponding bounds
-$$
-0 \le a,b,c \le \frac{1}{2}
-$$
-and
-$$
-0 \le d \le \frac{1}{4}.
-$$
-
-Our key approach will be to decompose the likelihood as
-\begin{equation}
-\label{eq:partial-likelihood}
-L^{(i)}(x', y', w') = H_{\tau'}(x,y;x',y',w') + L^{p,(i)}(x',y',w')
-\end{equation}
-where the first term is \eqref{eq:information} and the second term will be a weighted sum of the $\log$s of values from Table~\ref{tab:likelihoods}, the superscript indicating which choices for ancestral state likelihood values we choose.
+For the second term, we seek an upper bound for the $\tau_1$ case and a lower bound for the $\tau_2$ case.
 Define
 $$
-\{\hat{x}, \hat{y}, \hat{w}, \hat{i}\} = \arg\max_{\{x,y,w,i\}} \ L^{p,(i)}(x,y,w).
+\{\hat{x}, \hat{y}, \hat{w}, \hat{i}\} = \arg\max_{\{x,y,w,i\}} \ L^{p,(i)}(x,y,w,\tau').
 $$
@@
 $$
-H_{\tau'}(x,y; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H_{\tau'}(x,y;x,y,y) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
+H(x,y,\tau_1; \hat{x}, \hat{y}, \hat{w}, \tau') + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w},\tau_1) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(x,y,\tau_1;x,y,y,\tau_1) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w},\tau_1).
 $$
@@
 
-\subsubsection{A simplified Farris likelihood}
+\subsubsection{Upper bound for Farris likelihood}
 
-The likelihood for the Farris topology is less complicated than that of the Felsenstein since only one site pattern has an ambiguous likelihood---all site patterns except \texttt{0101} have a fixed value for their likelihood when taking the maximum.
-To fix this value, we will consider an upper bound on the partial likelihood by setting the value for site pattern 0011 to $(1+x)^2(1+w)(1+y)^2$.
-Here the partial likelihood $L_{fa}^{p,(i)}(x,y,w)$ takes the form:
+To obtain an upper bound for the Farris likelihood---the terms in the top half of Table~\ref{tab:likelihoods}---we use the term $(1+x)^2(1+w)(1+y)^2$ for the $0101$ site pattern.
+Since all cases have a $(1+x)^2$ term except $\{0100, 1000, 0011, 1001\}$, and all cases except $\{0001, 0010, 0011, 1001\}$ have a $(1+y)^2$, we can add and subtract the $\log$ of these terms weighted by each of their generating probabilities.
+In other words, define
 $$
-L_{fa}^{p} = (2-a)\log(1+x)+a\log(1-x)+(2-b)\log(1+y)+b\log(1-y)+\log(1+w),
-$$
-and this has a maximum
-$$
-\hat{L}_{fa}^{p} = (2-a)\log(2-a)+a\log(a)+(2-b)\log(2-b)+b\log(b)+\log(2).
-$$
-
-\subsubsection{A simplified Felsenstein likelihood}
-
-We construct similar bounds on the Felsenstein likelihood, though to make matters simpler we replace all $(1+y)$ terms in Table~\ref{tab:likelihoods} with $(1-y)$ to obtain a lower bound.
-Unlike the Farris partial likelihood, the likelihood in this case has one of six forms:
-$$
-L_{fe}^{(1)} = 2\log(1+x)+(2-c-2d)\log(1+y)+(c+2d)\log(1-y)+\log(1-w),
-$$
-$$
-L_{fe}^{(2)} = (2-c-2d)\log(1+x)+(c+2d)\log(1-x)+2\log(1+y)+\log(1-w),
-$$
-$$
-L_{fe}^{(3)} = (2-2d)\log(1+x)+2d\log(1-x)+(2-c)\log(1+y)+c\log(1-y)+\log(1-w),
-$$
-$$
-L_{fe}^{(4)} = (2-c)\log(1+x)+c\log(1-x)+(2-2d)\log(1+y)+2d\log(1-y)+\log(1-w),
-$$
-$$
-L_{fe}^{(5)} = (2-d)\log(1+x)+d\log(1-x)+(2-c-d)\log(1+y)+(c+d)\log(1-y)+\log(1-w)
+a = p(0100) + p(1000) + p(0011) + p(1001)
 $$
@@
 $$
-L_{fe}^{(6)} = (2-c-d)\log(1+x)+(c+d)\log(1-x)+(2-d)\log(1+y)+d\log(1-y)+\log(1-w)
+b = p(1000) + p(0010) + p(0011) + p(1001).
 $$
-with maxima
+Then $L^{p,(i)}(x',y',w',\tau_1)$ takes the form
 $$
-\hat{L}_{fe}^{(1)} = \hat{L}_{fe}^{(2)} = 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d)
+L^{p}(x',y',w') = (2-a)\log(1+x')+a\log(1-x')+(2-b)\log(1+y')+b\log(1-y')+\log(1+w')
 $$
+with a maximum
 $$
-\hat{L}_{fe}^{(3)} = \hat{L}_{fe}^{(4)} = (2-2d)\log(2-2d)+2d\log(2d)+(2-c)\log(2-c)+c\log(c)
+\hat{L}^{p}(a,b,\tau_1) = (2-a)\log(2-a)+a\log(a)+(2-b)\log(2-b)+b\log(b)+\log(2)
+$$
+where $a$ and $b$ are functions of the true generating branch parameters $\{x,y\}$.
+
+\subsubsection{Lower bound for Felsenstein likelihood}
+
+We construct a similar lower bound on the Felsenstein likelihood, though here we replace all $(1+w)$ terms with $(1-w)$.
+Let
+$$
+c = p(0001) + p(0010) + p(0100) + p(1000)
 $$
@@
 $$
-\hat{L}_{fe}^{(5)} = \hat{L}_{fe}^{(6)} = (2-d)\log(2-d)+d\log(d)+(2-c-d)\log(2-c-d)+(c+d)\log(c+d).
+d = p(0101) + p(1001).
 $$
+The likelihood in this case takes one of six forms:
+$$
+L^{(1)}(x',y',w') = 2\log(1+x')+(2-c-2d)\log(1+y')+(c+2d)\log(1-y')+\log(1-w'),
+$$
+$$
+L^{(2)}(x',y',w') = (2-c-2d)\log(1+x')+(c+2d)\log(1-x')+2\log(1+y')+\log(1-w'),
+$$
+$$
+L^{(3)}(x',y',w') = (2-2d)\log(1+x')+2d\log(1-x')+(2-c)\log(1+y')+c\log(1-y')+\log(1-w'),
+$$
+$$
+L^{(4)}(x',y',w') = (2-c)\log(1+x')+c\log(1-x')+(2-2d)\log(1+y')+2d\log(1-y')+\log(1-w'),
+$$
+$$
+L^{(5)}(x',y',w') = (2-d)\log(1+x')+d\log(1-x')+(2-c-d)\log(1+y')+(c+d)\log(1-y')+\log(1-w')
+$$
+or
+$$
+L^{(6)}(x',y',w') = (2-c-d)\log(1+x')+(c+d)\log(1-x')+(2-d)\log(1+y')+d\log(1-y')+\log(1-w').
+$$
+These have maxima
+$$
+\hat{L}^{(1)}(c,d) = \hat{L}^{(2)}(c,d) = 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d),
+$$
+$$
+\hat{L}^{(3)}(c,d) = \hat{L}^{(4)}(c,d) = (2-2d)\log(2-2d)+2d\log(2d)+(2-c)\log(2-c)+c\log(c)
+$$
+and
+$$
+\hat{L}^{(5)}(c,d) = \hat{L}^{(6)}(c,d) = (2-d)\log(2-d)+d\log(d)+(2-c-d)\log(2-c-d)+(c+d)\log(c+d)
+$$
+where $c$ and $d$ are also functions of the true generating branch parameters $\{x,y\}$.
 
@@
 
-\subsubsection{Example}
-
-Set $\{x, y\} = \{0.25, 0.25\}$.
-For a Farris zone generating topology, we have
+Thus, let
 $$
-a = b = 0.46875
-$$
-resulting in
-$$
-\hat{L}_{fa}^{p}(\hat{x}, \hat{y}, \hat{w}) \approx -2.17803560694
-$$
-and
-$$
-H_{fa}(x,y;x,y,y) \approx -2.07509999739.
-$$
-Computing the upper bound yields
-$$
-\max_{x,y,w} \ L_{fa}(x,y,w) \le -4.25313560433
-$$
-and it remains to show that 
-$$
-\max_{x,y,w} \ L_{fe}(x,y,w) \ge -4.25313560433
+\hat{L}^{p}(c,d,\tau_2) = 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d).
 $$
 
-For the same generating parameters,
-$$
-c = 0.498046875, \ d = 0.2431640625,
-$$
-resulting in
-$$
-\hat{L}_{fe}(\hat{x}, \hat{y}, \hat{w}) \approx -2.07919739112
-$$
-Calculating shows%
-    %
-    \footnote{It's not a typo that $\hat{L}_{fe}$ and $H_{fe}(x,y; \hat{x}, \hat{y}, \hat{w})$ are the same---at least not in the code.
-        I should try to show why this is the case.
-        }
-    %
-$$
-H_{fe}(x,y; \hat{x}, \hat{y}, \hat{w}) \approx -2.07919739112
-$$
-and
-$$
-\max_{x,y,w} \ L_{fe}(x,y,w) \ge -4.15839478224 > -4.25313560433 \ge \max_{x,y,w} \ L_{fa}(x,y,w)
-$$
-where site patterns are generated from the Farris zone topology.
-Confirming, run \texttt{python}'s numerical optimization solver for $\{.25, .25\}$.
-Since we are computing a constrained maximization, we can find a global maximum using basin hopping, and doing so yields
-$$
-\max_{x,y,w} \ L_{fa}(x,y,w) \approx -4.5101771294218045
-$$
-and
-$$
-\max_{x,y,w} \ L_{fe}(x,y,w) \approx -4.1544859851260281.
-$$
-Woohoo!
-
-\subsubsection{General example}
+\subsubsection{Region of inconsistency}
 
@@
 
+\paragraph{Example}
+
+Set $\{x, y\} = \{0.25, 0.25\}$.
+For a Farris zone generating topology, we have
+$$
+a = b = 0.46875
+$$
+resulting in
+$$
+\hat{L}^{p}(\hat{x}, \hat{y}, \hat{w},\tau_1) \approx -2.17803560694
+$$
+and
+$$
+H(x,y,\tau_1;x,y,y,\tau_1) \approx -2.07509999739.
+$$
+Computing the upper bound yields
+$$
+\max_{x,y,w} \ L(x,y,w,\tau_1) \le -4.25313560433
+$$
+and it remains to show that 
+$$
+\max_{x,y,w} \ L(x,y,w,\tau_2) \ge -4.25313560433
+$$
+
+For the same generating parameters,
+$$
+c = 0.498046875, \ d = 0.2431640625,
+$$
+resulting in
+$$
+\hat{L}(\hat{x}, \hat{y}, \hat{w},\tau_2) \approx -2.07919739112
+$$
+Calculating shows%
+    %
+    \footnote{It's not a typo that $\hat{L}_{fe}$ and $H_{fe}$ are the same---at least not in the code.
+        I should try to show why this is the case.
+        }
+    %
+$$
+H(x,y,\tau_1; \hat{x}, \hat{y}, \hat{w},\tau_2) \approx -2.07919739112
+$$
+and
+$$
+\max_{x,y,w} \ L(x,y,w,\tau_2) \ge -4.15839478224 > -4.25313560433 \ge \max_{x,y,w} \ L(x,y,w,\tau_1)
+$$
+where site patterns are generated from the Farris zone topology.
+Confirming, run \texttt{python}'s numerical optimization solver for $\{.25, .25\}$.
+Since we are computing a constrained maximization, we can find a global maximum using basin hopping, and doing so yields
+$$
+\max_{x,y,w} \ L(x,y,w,\tau_1) \approx -4.5101771294218045
+$$
+and
+$$
+\max_{x,y,w} \ L(x,y,w,\tau_2) \approx -4.1544859851260281.
+$$
+Woohoo!
+
 \bibliographystyle{plain}

### 2017-03-21 joint_inf.tex
Infrastructurizing and renaming.

--- 2017-03-21 main.tex
@@
 
-Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies under these two topologies, calculated using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{semplesteel}.
+Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies under these two topologies, calculated using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}.
 Table~\ref{tab:likelihoods} contains calculations of likelihood values for fixed site patterns and topologies, with the maximum values over ancestral state patterns noted.
@@
 \bibliographystyle{plain}
-\bibliography{sample}
+\bibliography{joint_inf}
 

### 2017-03-21 joint_inf.tex
Some intro and discussion.

@@
 
+\section*{Introduction}
+
+Classical maximum-likelihood estimation in phylogenetics operates by integrating out possible ancestral states at the internal nodes of the tree.
+Recently Neher et al have suggested using an approximation to ML inference in which likelihood is maximized jointly across model parameters and ancestral sequences.
+This is attractive from a computational perspective, because joint ML inference can proceed according to an EM-like algorithm in which ML ancestral sequences are first inferred and then model parameters are optimized conditioned on the ancestral sequences.
+This latter step is much simpler and more computationally efficient when conditioning on ancestral sequences compared to the case in which ancestral sequences are marginalized out.
+
+However, existing consistency proofs for phylogenetics [cite] do not apply to the setting of joint inference of trees and ancestral sequences, raising the question of if this procedure enjoys good statistical properties.
+There are examples in statistics, such as the Neyman-Scott paradox \cite{Neyman1948-tt}, in which additional parameters destroy consistency.
+
+In this paper we show that the joint inference of trees and ancestral sequences is not consistent.
+To do so, we follow a similar strategy of using simple models and four taxon trees.
+Here we use a binary symmetric model and simulate on the ``Farris zone'' \cite{Siddall1998-hq} tree.
+We use bounds on likelihood to demarcate a sizeable area of long branch lengths in which joint inference gives the wrong tree.
+
 \section{Ancestral state reconstruction using maximum likelihood}
 
-\subsection{The usual story}
+\subsection{Classical maximum-likelihood inference in phylogenetics}
 
@@
 
+\section{Discussion}
+
+Neyman-Scott paradox.
+
+Interesting that here we are simulating on the Farris tree and end up with the Felsenstein tree.
+For maximum parsimony it's the opposite \cite{Felsenstein1978-rr}.
+Discussion of number of parameters of each.
+
+However, note that \cite{Siddall1998-hq} get things going the same way, although \cite{Swofford2001-hr} show that the problem is that they didn't simulate long enough sequences.
+
 \bibliographystyle{plain}

### 2017-03-22 deprecated.tex
better notation/careful statements

@@
 
+$$
+a = p(0100) + p(1000) + p(0011) + p(1001) = (4-2x^2-2y^2)/8,
+$$
+$$
+b = p(1000) + p(0010) + p(0011) + p(1001) = (4-2x^2-2y^2)/8,
+$$
+$$
+c = p(0001) + p(0010) + p(0100) + p(1000) = (4-4x^2y^2)/8
+$$
+and
+$$
+d = p(0101) + p(1001) = (2-4xy^2+2x^2y^2)/8.
+$$
 Some scratch work:

### 2017-03-22 joint_inf.tex
better notation/careful statements

--- 2017-03-21 joint_inf.tex
@@
 
+\section*{Introduction}
+
+Classical maximum-likelihood (ML) estimation in phylogenetics operates by integrating out possible ancestral states at the internal nodes of the tree.
+Recently \cite{Neher2017} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences.
+This is attractive from a computational perspective: joint ML inference can proceed according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
+% I would say classical maximum likelihood is closer to EM, no? At some point in EM we'd have to marginalize/take the expectation, which joint ML doesn't do. Curious to talk more about it though...
+This latter step is much simpler and more computationally efficient when conditioning on ancestral sequences compared to the case in which ancestral sequences are marginalized out.
+
+However, existing consistency proofs for phylogenetics \cite{ArindamRoyChoudhury?} do not apply to the setting of joint inference of trees and ancestral sequences, raising the question of whether this procedure enjoys good statistical properties.
+There are examples in statistics, such as the Neyman-Scott paradox \cite{Neyman1948-tt}, in which additional parameters destroy consistency.
+
+In this paper we show that the joint inference of trees and ancestral sequences is not consistent in general.
+To do so, we follow a strategy of using simple models and four taxa trees.
+Here we use a binary symmetric model and simulate on the ``Farris zone'' \cite{Siddall1998-hq} tree.
+We use bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference converges on the wrong tree.
+
 \section{Ancestral state reconstruction using maximum likelihood}
 
-\subsection{The usual story}
+\subsection{Classical maximum-likelihood inference in phylogenetics}
 
-We have observed character data $\mathbf{y}=[y_1,\ldots,y_n]\in\mathcal{Y}^n$ and corresponding unknown ancestral states $\boldsymbol\xi=[\xi_1,\ldots,\xi_n]\in\Xi^n$.
+We observe character data $\mathbf{y}=[y_1,\ldots,y_n]\in\mathcal{Y}^n$ and corresponding unknown ancestral states $\boldsymbol\xi=[\xi_1,\ldots,\xi_n]\in\Xi^n$.
 Both $\mathcal{Y}$ and $\Xi$ are discrete sets.
@@
 One tantalizing approach is to do away with the marginalization and the two-step maximization, directly estimating the maximum likelihood parameters from the full likelihood in \eqref{eq:full_likelihood}.
-We can perform this iteratively, now computing a profile likelihood
+We can perform this by computing the profile likelihood
 \begin{equation}
 \label{eq:profile_likelihood}
-L_n'(\mathbf{y};\tau, t) = \max_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t)
+L_n'(\mathbf{y};\tau, t) = \max_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
 \end{equation}
-and estimating the topology and branch lengths via
-$$
+then estimating the topology and branch lengths via
+\begin{equation}
+\label{eq:profile_likelihood_topology_bl}
 (\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ L_n'(\mathbf{y};\tau, t)
-$$
+\end{equation}
 while using $\hat{\boldsymbol\xi}$ from \eqref{eq:profile_likelihood} as an estimate for $\boldsymbol\xi$.
-
-\textbf{Claim to prove}: Is it obvious that this particular iterative approach can be done?
-In other words, do $(\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t})$ from \eqref{eq:profile_likelihood} correspond to those of \eqref{eq:full_likelihood}?
-I believe since $\Xi$ is a discrete set that this is true, but it may be worth a more rigorous argument.
+Since $\Xi$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
+In the general case, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
+For this reason, in practice joint ML estimates $\hat{\boldsymbol\xi}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\boldsymbol\xi}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.
 
@@
 
-Now, since $\mathcal{Y}$ and $\Xi$ are discrete sets, we define unique site patterns $\mathbf{s}=[s_1,\ldots,s_q]\in\mathcal{S}$ and ancestral state categories $\mathbf{z}=[z_1,\ldots,z_q]\in\mathcal{Z}$ such that
+Since $\mathcal{Y}$ and $\Xi$ are discrete sets, we define unique site patterns $\mathbf{s}=[s_1,\ldots,s_q]\in\mathcal{S}$ and ancestral state patterns $\mathbf{z}=[z_1,\ldots,z_q]\in\mathcal{Z}$ such that
 \begin{align}
@@
 $$
-Now assume $n$ observations were generated from a model with parameters $(\tau^*, t^*)$.
+Assume $n$ observations were generated from a model with parameters $(\tau^*, t^*)$.
 We have
@@
 \begin{align}
-\frac{1}{n} \log L_n'(\mathbf{y};\tau, t) &= \max_{\mathbf{z}} \ \frac{1}{n} \log L_n''(\mathbf{s};\mathbf{z},\tau,t) \\
-    &\rightarrow \max_{z_1, \ldots, z_q} \ \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)] \\
-    &= \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \max_{z_j} \ \log P(z_j | s_j, \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
+\frac{1}{n} \log L_n'(\mathbf{y};\tau, t) &= \max_{\mathbf{z}} \ \frac{1}{n} \log L_n''(\mathbf{s};\mathbf{z},\tau,t) \nonumber \\
+    &\rightarrow \max_{z_1, \ldots, z_q} \ \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)] \nonumber \\
+    &= \max_{\mathbf{z}} \ \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)] \nonumber \\
+    &= \max_{\mathbf{z}} \ \left\{\sum_{j} P(s_j | \tau^*, t^*) \log P(s_j | \tau, t)\right\} \nonumber \\
+    &\qquad + \left\{\sum_{j} P(s_j | \tau^*, t^*) \log P(z_j | s_j, \tau, t)\right\}. \label{eq:site_pattern_profile_likelihood_mean}
 \end{align}
@@
 Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
-Our approach will be to decompose the likelihood in \eqref{eq:site_pattern_profile_likelihood_mean} as
+Our approach will be to decompose the likelihood inside the maximization in \eqref{eq:site_pattern_profile_likelihood_mean} as
 \begin{equation}
 \label{eq:partial-likelihood}
-L^{(i)}(x', y', w') = H(x,y,\tau_1;x',y',w', \tau') + L^{p,(i)}(x',y',w')
+L(x', y', w'; \mathbf{z},\tau') = H(x,y,\tau_1;x',y',w', \tau') + L_{p}(x',y',w';\mathbf{z},\tau')
 \end{equation}
-where the first term is
-\begin{align}
-\label{eq:information}
-H(x,y,\tau_1;x',y',w',\tau') &= \sum_{j} P(s_j | \tau_1, \{x,y,y\}) \cdot \log P(s_j | \tau', \{x',y',w'\}) \nonumber \\
-&:= \sum_{j} p(s_j) \cdot \log P(s_j | \tau', \{x',y',w'\}).
-\end{align}
-and the second term will be a weighted sum of the $\log$s of values from Table~\ref{tab:likelihoods}, the superscript indicating which choices for ancestral state likelihood values we choose.
-
-To bound the first term, we can make use of Gibbs inequality
+where different values for $\mathbf{z}$ will alter the form of $L_{p}(x',y',w';\mathbf{z},\tau')$.
+We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
 $$
-H(x, y, \tau_1;x', y', w', \tau') \le H(x, y, \tau_1;x, y, y, \tau_1),
+\max_{\{x',y',w',\mathbf{z}\}} \ L(x',y',w';\mathbf{z},\tau') \le H(x,y,\tau_1;x,y,y,\tau_1) + L_{p}(\hat{x}, \hat{y}, \hat{w};\hat{\mathbf{z}},\tau')
 $$
-though this bound can be made tighter.
-For the second term, we seek an upper bound for the $\tau_1$ case and a lower bound for the $\tau_2$ case.
-Define
+using Gibbs inequality
 $$
-\{\hat{x}, \hat{y}, \hat{w}, \hat{i}\} = \arg\max_{\{x,y,w,i\}} \ L^{p,(i)}(x,y,w,\tau').
+H(x, y, \tau_1;x', y', w', \tau') \le H(x, y, \tau_1;x, y, y, \tau_1)
 $$
-We have upper and lower bounds for \eqref{eq:partial-likelihood} in the form
+and
 $$
-H(x,y,\tau_1; \hat{x}, \hat{y}, \hat{w}, \tau') + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w},\tau_1) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(x,y,\tau_1;x,y,y,\tau_1) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w},\tau_1).
+\{\hat{x}, \hat{y}, \hat{w}, \hat{\mathbf{z}}\} = \arg\max_{\{x,y,w,\mathbf{z}\}} \ L_{p}(x,y,w;\mathbf{z},\tau').
+$$
+Similarly, we have a lower bound as
+$$
+\max_{\{x',y',w',\mathbf{z}\}} \ L(x',y',w';\mathbf{z},\tau') \ge H(x,y,\tau_1; \hat{x}, \hat{y}, \hat{w}, \tau') + L_{p}(\hat{x}, \hat{y}, \hat{w},\hat{\mathbf{z}},\tau').
 $$
 
+% vvvv Remove these? I think they're obvious now, but maybe they can be made more explicit above?
 \textbf{Argument for lower bound}: This makes intuitive sense, though may be difficult to show in general.
@@
 $$
-a = p(0100) + p(1000) + p(0011) + p(1001)
+a = P(0100|\tau_1,\{x,y,y\}) + P(1000|\tau_1,\{x,y,y\}) + P(0011|\tau_1,\{x,y,y\}) + P(1001|\tau_1,\{x,y,y\})
 $$
@@
 $$
-b = p(1000) + p(0010) + p(0011) + p(1001).
+b = P(1000|\tau_1,\{x,y,y\}) + P(0010|\tau_1,\{x,y,y\}) + P(0011|\tau_1,\{x,y,y\}) + P(1001|\tau_1,\{x,y,y\}).
 $$
-Then $L^{p,(i)}(x',y',w',\tau_1)$ takes the form
+Then the ancestral state maximizing the likelihood is $00$ and $L_{p}(x',y',w';\mathbf{z},\tau_1)$ is bounded above by
 $$
-L^{p}(x',y',w') = (2-a)\log(1+x')+a\log(1-x')+(2-b)\log(1+y')+b\log(1-y')+\log(1+w')
+L_{p}(x',y',w';00,\tau_1) \le (2-a)\log(1+x')+a\log(1-x')+(2-b)\log(1+y')+b\log(1-y')+\log(1+w')
 $$
@@
 $$
-\hat{L}^{p}(a,b,\tau_1) = (2-a)\log(2-a)+a\log(a)+(2-b)\log(2-b)+b\log(b)+\log(2)
+\hat{L}_{p}(1-a,1-b,1;00,\tau_1) \le (2-a)\log(2-a)+a\log(a)+(2-b)\log(2-b)+b\log(b)+\log(2)
 $$
@@
 $$
-c = p(0001) + p(0010) + p(0100) + p(1000)
+c = P(0001|\tau_1,\{x,y,y\}) + P(0010|\tau_1,\{x,y,y\}) + P(0100|\tau_1,\{x,y,y\}) + P(1000|\tau_1,\{x,y,y\})
 $$
@@
 $$
-d = p(0101) + p(1001).
+d = P(0101|\tau_1,\{x,y,y\}) + P(1001|\tau_1,\{x,y,y\}).
 $$
-The likelihood in this case takes one of six forms:
+The lower bounds on the likelihoods in this case take one of six forms:
+\textbf{TODO}: ancestral states where these likelihoods occur
 $$
-L^{(1)}(x',y',w') = 2\log(1+x')+(2-c-2d)\log(1+y')+(c+2d)\log(1-y')+\log(1-w'),
+L_{p}(x',y',w';z,\tau_2) \ge 2\log(1+x')+(2-c-2d)\log(1+y')+(c+2d)\log(1-y')+\log(1-w'),
 $$
 $$
-L^{(2)}(x',y',w') = (2-c-2d)\log(1+x')+(c+2d)\log(1-x')+2\log(1+y')+\log(1-w'),
+L_{p}(x',y',w';z,\tau_2) \ge (2-c-2d)\log(1+x')+(c+2d)\log(1-x')+2\log(1+y')+\log(1-w'),
 $$
 $$
-L^{(3)}(x',y',w') = (2-2d)\log(1+x')+2d\log(1-x')+(2-c)\log(1+y')+c\log(1-y')+\log(1-w'),
+L_{p}(x',y',w';z,\tau_2) \ge (2-2d)\log(1+x')+2d\log(1-x')+(2-c)\log(1+y')+c\log(1-y')+\log(1-w'),
 $$
 $$
-L^{(4)}(x',y',w') = (2-c)\log(1+x')+c\log(1-x')+(2-2d)\log(1+y')+2d\log(1-y')+\log(1-w'),
+L_{p}(x',y',w';z,\tau_2) \ge (2-c)\log(1+x')+c\log(1-x')+(2-2d)\log(1+y')+2d\log(1-y')+\log(1-w'),
 $$
 $$
-L^{(5)}(x',y',w') = (2-d)\log(1+x')+d\log(1-x')+(2-c-d)\log(1+y')+(c+d)\log(1-y')+\log(1-w')
+L_{p}(x',y',w';z,\tau_2) \ge (2-d)\log(1+x')+d\log(1-x')+(2-c-d)\log(1+y')+(c+d)\log(1-y')+\log(1-w')
 $$
@@
 $$
-L^{(6)}(x',y',w') = (2-c-d)\log(1+x')+(c+d)\log(1-x')+(2-d)\log(1+y')+d\log(1-y')+\log(1-w').
+L_{p}(x',y',w';z,\tau_2) \ge (2-c-d)\log(1+x')+(c+d)\log(1-x')+(2-d)\log(1+y')+d\log(1-y')+\log(1-w').
 $$
-These have maxima
+Since the maxima for pairs of lower bounds above each have a symmetry in $x$ and $y$, they are bounded below by one of
 $$
-\hat{L}^{(1)}(c,d) = \hat{L}^{(2)}(c,d) = 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d),
+\hat{L}_{p}(1,1-c-2d,0;z,\tau_2) \ge 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d),
 $$
 $$
-\hat{L}^{(3)}(c,d) = \hat{L}^{(4)}(c,d) = (2-2d)\log(2-2d)+2d\log(2d)+(2-c)\log(2-c)+c\log(c)
+\hat{L}_{p}(1-2d,1-c,0;z,\tau_2) \ge (2-2d)\log(2-2d)+2d\log(2d)+(2-c)\log(2-c)+c\log(c)
 $$
@@
 $$
-\hat{L}^{(5)}(c,d) = \hat{L}^{(6)}(c,d) = (2-d)\log(2-d)+d\log(d)+(2-c-d)\log(2-c-d)+(c+d)\log(c+d)
+\hat{L}_{p}(1-d,1-c-d,0;z,\tau_2) \ge (2-d)\log(2-d)+d\log(d)+(2-c-d)\log(2-c-d)+(c+d)\log(c+d)
 $$
@@
 $$
-\hat{L}^{p}(c,d,\tau_2) = 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d).
+\hat{L}_{p}(1,1-c-2d,0;z,\tau_2) = 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d).
 $$
@@
 $$
-H_{fe}(x,y; 1, 1-c-2d, 0) + \hat{L}_{fe}(c,d) \ge H_{fa}(x,y; x, y, y) + \hat{L}_{fa}(a,b).
+H(x,y,\tau_1; 1, 1-c-2d, 0, \tau_2) + \hat{L}_{p}(1,1-c-2d,0;z,\tau_2) \ge H(x,y,\tau_1; x, y, y,\tau_1) + \hat{L}_{p}(1-a,1-b,1;00,\tau_1).
 $$
@@
 \begin{align*}
-    0 &\ge H_{fa}(x,y; x, y, y) + \hat{L}_{fa}(a,b) - H_{fe}(x,y; 1, 1-c-2d, 0) - \hat{L}_{fe}(c,d) \\
+    0 &\ge H(x,y,\tau_1; x, y, y,\tau_1) + \hat{L}_{p}(1-a,1-b,1;00,\tau_1) \\
+      &\qquad - H(x,y,\tau_1; 1, 1-c-2d, 0, \tau_2) - \hat{L}_{p}(1,1-c-2d,0;\tau_2) \\
       & = 2(12+2x^2+2y^2)\log((12+2x^2+2y^2)/8) \\
@@
 \end{align*}
+
+\textbf{TODO}: can we use sage to simplify the above so it looks nicer?
+Write it as functions of $a,b,c,d$ so it's less messy?
+What would be best?
+
 The \texttt{sage} function in listing~\ref{list:sage} will generate Fig.~\ref{fig:inconsistency-farris}.

### 2017-03-22 joint_inf.tex
update figs
Mucking about with notation.

@@
 
-However, existing consistency proofs for phylogenetics \cite{ArindamRoyChoudhury?} do not apply to the setting of joint inference of trees and ancestral sequences, raising the question of whether this procedure enjoys good statistical properties.
-There are examples in statistics, such as the Neyman-Scott paradox \cite{Neyman1948-tt}, in which additional parameters destroy consistency.
+However, existing consistency proofs for phylogenetics \cite{RoyChoudhury2015-ta} do not apply to the setting of joint inference of trees and ancestral sequences.
+There are examples in statistics, such as the Neyman-Scott paradox \cite{Neyman1948-tt}, in which estimation of additional parameters destroys consistency.
+This raises the question of whether joint inference of trees and ancestral sequences enjoys good statistical properties.
 
@@
 
-We observe character data $\mathbf{y}=[y_1,\ldots,y_n]\in\mathcal{Y}^n$ and corresponding unknown ancestral states $\boldsymbol\xi=[\xi_1,\ldots,\xi_n]\in\Xi^n$.
-Both $\mathcal{Y}$ and $\Xi$ are discrete sets.
+Let $\mathcal{Y}$ be the set of assignments of characters to tips of our tree, and $\Xi$ the set of assignments of characters to the internal nodes of the tree.
+Both $\mathcal{Y}$ and $\Xi$ are finite sets.
+Assume we observe $n$ samples of character data (i.e.\ an alignment with $n$ columns) $\mathbf{y}=[y_1,\ldots,y_n]\in\mathcal{Y}^n$ and corresponding unknown ancestral states $\boldsymbol\xi=[\xi_1,\ldots,\xi_n]\in\Xi^n$.
+%EM Hm, this doesn't make sense compared to the above.
 Without loss of generality, assume $\mathcal{Y}=\Xi=\{0,1\}$.
-For a topology $\tau$ and branch lengths $t$ we form the likelihood
+For a topology $\tau$ and branch lengths $t$ we form the fully-observed likelihood (making the usual independence assumption between sites):
 \begin{equation}
@@
 \end{equation}
-In particular, we are interested in
-$$
-(\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
-$$
-which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
-
-Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$, the typical approach to estimate these parameters involves computing the marginal likelihood
+% In particular, we are interested in
+% $$
+% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
+% $$
+% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
+% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
+The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves computing the marginal likelihood
 \begin{equation}
@@
 $$
-The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
-More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.
+% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
+% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.
 
@@
 
-One tantalizing approach is to do away with the marginalization and the two-step maximization, directly estimating the maximum likelihood parameters from the full likelihood in \eqref{eq:full_likelihood}.
+One tantalizing approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the full likelihood in \eqref{eq:full_likelihood}.
 We can perform this by computing the profile likelihood
@@
 \end{equation}
-while using $\hat{\boldsymbol\xi}$ from \eqref{eq:profile_likelihood} as an estimate for $\boldsymbol\xi$.
+while using $\hat{\boldsymbol\xi}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\boldsymbol\xi$.
 Since $\Xi$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
 In the general case, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
-For this reason, in practice joint ML estimates $\hat{\boldsymbol\xi}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\boldsymbol\xi}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.
+For this reason, in practice the joint ML strategy estimates $\hat{\boldsymbol\xi}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\boldsymbol\xi}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.
 
@@
 
-Since $\mathcal{Y}$ and $\Xi$ are discrete sets, we define unique site patterns $\mathbf{s}=[s_1,\ldots,s_q]\in\mathcal{S}$ and ancestral state patterns $\mathbf{z}=[z_1,\ldots,z_q]\in\mathcal{Z}$ such that
+Since we have a finite character alphabet, there are a finite number of possibilities of assignments of characters to tips $y_i$ or internal nodes $\xi_i$.
+These are commonly called ``site patterns'' in the literature.
+%EM in general aren't a different number of internal site patterns than tip site patterns?
+We enumerate the unique site patterns $\mathbf{s}=[s_1,\ldots,s_q]\in\mathcal{S}$ and ancestral state patterns $\mathbf{z}=[z_1,\ldots,z_q]\in\mathcal{Z}$; these are defined by the number of leaves and the character alphabet.
+This enables us to group identical terms, reformulating the likelihood equation
 \begin{align}
@@
     &= \max_{\boldsymbol\xi} \ \prod_{i=1}^{n} \ P(y_i, \xi_i | \tau, t) \\
-    &= \max_{\mathbf{z}} \ \prod_{j=1}^{q} \ P(s_j, z_j | \tau, t)^{n_j} \label{eq:site_pattern_likelihood}
+%EM Can you think a little about z_j? The indexing seems a little problematic now that we are thinking of everything as a function of y.
+    &= \max_{\mathbf{z}} \ \prod_{j=1}^{q} \ P(s_j, z_j | \tau, t)^{n_j(\mathbf{y})} \label{eq:site_pattern_likelihood}
 \end{align}
-where $n_j$ is the number of observations in the sample where site pattern $s_j$ was seen.
-Since $\mathcal{S}$ and $\mathcal{Z}$ are also discrete sets, the value of each factor $P(y_i, \xi_i | \tau, t)$ will be maximized at one or more values of $\xi_i$, though independent of $i$ given $j$.
+where $n_j(\mathbf{y})$ is the number of observations in the sample where site pattern $s_j$ was seen in $\mathbf{y}$.
+%EM It's a small point, but note that the argument below justified the equation above, which is a little upside down.
+Note that since $\mathcal{S}$ and $\mathcal{Z}$ are also finite sets, the value of each factor $P(y_i, \xi_i | \tau, t)$ will be maximized at one or more values of $\xi_i$, though independent of $i$ given $j$.
 This independence allows us to group factors by those where $\xi_i$ equals $z_j$ so that $P(s_j, z_j | \tau, t)$ is maximized.
@@
 
-We are now interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood}.
+We are now interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
 Let
 $$
-L_n''(\mathbf{s};\mathbf{z},\tau,t) = \prod_{j=1}^q \ P(s_j, z_j | \tau, t)^{n_j}
+L_n''(\mathbf{y};\mathbf{z},\tau,t) = \prod_{j=1}^q \ P(s_j, z_j | \tau, t)^{n_j(\mathbf{y})}
 $$
@@
 $$
-\max_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t) = \max_{\mathbf{z}} \ L_n''(\mathbf{s};\mathbf{z},\tau,t).
+\max_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t) = \max_{\mathbf{z}} \ L_n''(\mathbf{y};\mathbf{z},\tau,t).
 $$
-Assume $n$ observations were generated from a model with parameters $(\tau^*, t^*)$.
+Assume our $n$ observations were generated from a model with parameters $(\tau^*, t^*)$.
 We have
 \begin{align}
-\frac{1}{n} \log L_n''(\mathbf{s};\mathbf{z},\tau,t) &= \sum_{j} \frac{n_j}{n}\cdot \log P(s_j, z_j | \tau, t) \\
-  &= \sum_{j} \frac{n_j}{n}\cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)].
+\frac{1}{n} \log L_n''(\mathbf{y};\mathbf{z},\tau,t) &= \sum_{j} \frac{n_j(\mathbf{y})}{n}\cdot \log P(s_j, z_j | \tau, t) \\
+  &= \sum_{j} \frac{n_j(\mathbf{y})}{n}\cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)].
 \end{align}
-In the limit of infinite data we see
+In the limit of infinite data we have
 $$
-\frac{1}{n} \log L_n''(\mathbf{s};\mathbf{z},\tau,t) \rightarrow \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)]
+\frac{1}{n} \log L_n''(\mathbf{y};\mathbf{z},\tau,t) \rightarrow \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)]
 $$
@@
 \includegraphics[width=.9\textwidth]{analytic-inconsistency}
-\caption{Regions of inconsistency}
+\caption{
+    Region of inconsistency.
+    Due to the looseness of the upper and lower bounds, the parameters in the top right do not necessarily indicate consistency, though all parameters in the labeled region will result in an inconsistency.
+}
 \label{fig:inconsistency-farris}

### 2017-03-23 joint_inf.tex
notation notation notation

@@
 \usepackage{amsfonts}
+\usepackage{amssymb}
 \usepackage{graphicx}
@@
 \graphicspath{ {figures/} }
+
+\input{commands}
 
@@
 
-Let $\mathcal{Y}$ be the set of assignments of characters to tips of our tree, and $\Xi$ the set of assignments of characters to the internal nodes of the tree.
-Both $\mathcal{Y}$ and $\Xi$ are finite sets.
-Assume we observe $n$ samples of character data (i.e.\ an alignment with $n$ columns) $\mathbf{y}=[y_1,\ldots,y_n]\in\mathcal{Y}^n$ and corresponding unknown ancestral states $\boldsymbol\xi=[\xi_1,\ldots,\xi_n]\in\Xi^n$.
-%EM Hm, this doesn't make sense compared to the above.
-Without loss of generality, assume $\mathcal{Y}=\Xi=\{0,1\}$.
+Let $\fullAlignment$ be the set of assignments of characters to tips of our tree, and $\fullAncestralStates$ the set of assignments of characters to the internal nodes of the tree.
+Assume we observe $n$ samples of character data (i.e.\ an alignment with $n$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_n]\in\alignmentSpace^{\nSiteRows\times n}$ and corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_n]\in\ancestralStateSpace^{\nAncestralStateRows\times n}$.
+Without loss of generality, assume $\alignmentSpace=\ancestralStateSpace=\{0,1\}$.
 For a topology $\tau$ and branch lengths $t$ we form the fully-observed likelihood (making the usual independence assumption between sites):
@@
 \label{eq:full_likelihood}
-L_n(\mathbf{y};\boldsymbol\xi, \tau, t) = \prod_{i=1}^{n} \ P(y_i, \xi_i | \tau, t).
+L_n(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{n} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t).
 \end{equation}
@@
 \label{eq:marginal_likelihood}
-\tilde{L}_n(\mathbf{y}; \tau, t) = \sum_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t).
+\tilde{L}_n(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t)
 \end{equation}
@@
 $$
-(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_n(\mathbf{y}; \tau, t).
+(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_n(\fullAlignment; \tau, t).
 $$
@@
 \label{eq:profile_likelihood}
-L_n'(\mathbf{y};\tau, t) = \max_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
+L_n'(\fullAlignment;\tau, t) = \max_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t),
 \end{equation}
@@
 \label{eq:profile_likelihood_topology_bl}
-(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ L_n'(\mathbf{y};\tau, t)
+(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ L_n'(\fullAlignment;\tau, t)
 \end{equation}
-while using $\hat{\boldsymbol\xi}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\boldsymbol\xi$.
-Since $\Xi$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
+while using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
+Since $\ancestralStateSpace$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
 In the general case, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
-For this reason, in practice the joint ML strategy estimates $\hat{\boldsymbol\xi}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\boldsymbol\xi}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.
+For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.
 
@@
 
-Since we have a finite character alphabet, there are a finite number of possibilities of assignments of characters to tips $y_i$ or internal nodes $\xi_i$.
+Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
 These are commonly called ``site patterns'' in the literature.
 %EM in general aren't a different number of internal site patterns than tip site patterns?
-We enumerate the unique site patterns $\mathbf{s}=[s_1,\ldots,s_q]\in\mathcal{S}$ and ancestral state patterns $\mathbf{z}=[z_1,\ldots,z_q]\in\mathcal{Z}$; these are defined by the number of leaves and the character alphabet.
-This enables us to group identical terms, reformulating the likelihood equation
+%das: Yes, I moved the paragraph at the end to here so it's hopefully less confusing.
+Maybe writing out the example would help?
+I also changed from "pattern" to "category" which might make it a little more clear.
+We enumerate the unique site patterns $\mathbf{s}=[s_1,\ldots,s_q]\in\mathcal{S}$; these are defined by the number of leaves and the character alphabet.
+Corresponding to each of these site patterns will be an ancestral state category that is most likely given a particular likelihood.
+Since $\alignmentSpace$ and $\ancestralStateSpace$ are finite sets, the value of each factor $P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t)$ in \eqref{eq:full_likelihood} is maximized at one or more values of $\ancestralStateColumn_i$, though independent of $i$ given $j$.
+This independence allows us to group factors by those where $\ancestralStateColumn_i$ belongs to an ancestral state category $\ancestralStateCategory_j$ so that, if $\alignmentColumn_i=\sitePattern_j$ and $\ancestralStateColumn_i\in \ancestralStateCategory_j$, $P(\sitePattern_j, \ancestralStateCategory_j | \tau, t)$ is maximized.
+
+For ancestral state categories $\fullAncestralStateCategories=[\ancestralStateCategory_1,\ldots,\ancestralStateCategory_q]\in\ancestralStateCategorySpace$ we reformulate the likelihood equation
 \begin{align}
-L_n'(\mathbf{y};\tau, t) &= \max_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t) \\
-    &= \max_{\boldsymbol\xi} \ \prod_{i=1}^{n} \ P(y_i, \xi_i | \tau, t) \\
+L_n'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t) \\
+                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{n} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t) \\
 %EM Can you think a little about z_j? The indexing seems a little problematic now that we are thinking of everything as a function of y.
-    &= \max_{\mathbf{z}} \ \prod_{j=1}^{q} \ P(s_j, z_j | \tau, t)^{n_j(\mathbf{y})} \label{eq:site_pattern_likelihood}
+                             &= \max_{\fullAncestralStateCategories} \ \prod_{j=1}^{q} \ P(\sitePattern_j, \ancestralStateCategory_j | \tau, t)^{n_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
 \end{align}
-where $n_j(\mathbf{y})$ is the number of observations in the sample where site pattern $s_j$ was seen in $\mathbf{y}$.
-%EM It's a small point, but note that the argument below justified the equation above, which is a little upside down.
-Note that since $\mathcal{S}$ and $\mathcal{Z}$ are also finite sets, the value of each factor $P(y_i, \xi_i | \tau, t)$ will be maximized at one or more values of $\xi_i$, though independent of $i$ given $j$.
-This independence allows us to group factors by those where $\xi_i$ equals $z_j$ so that $P(s_j, z_j | \tau, t)$ is maximized.
+where $n_j(\fullAlignment)$ is the number of observations in the sample where site pattern $\sitePattern_j$ was seen in $\fullAlignment$.
 
@@
 
-\subsection{The case of infinite data}
+\subsection{Showing inconsistency}
 
@@
 $$
-L_n''(\mathbf{y};\mathbf{z},\tau,t) = \prod_{j=1}^q \ P(s_j, z_j | \tau, t)^{n_j(\mathbf{y})}
+L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t) = \prod_{j=1}^q \ P(\sitePattern_j, \ancestralStateCategory_j | \tau, t)^{n_j(\fullAlignment)}
 $$
@@
 $$
-\max_{\boldsymbol\xi} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t) = \max_{\mathbf{z}} \ L_n''(\mathbf{y};\mathbf{z},\tau,t).
+\max_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t) =
+    \max_{\fullAncestralStateCategories} \ L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t).
 $$
@@
 \begin{align}
-\frac{1}{n} \log L_n''(\mathbf{y};\mathbf{z},\tau,t) &= \sum_{j} \frac{n_j(\mathbf{y})}{n}\cdot \log P(s_j, z_j | \tau, t) \\
-  &= \sum_{j} \frac{n_j(\mathbf{y})}{n}\cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)].
+    \frac{1}{n} \log L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t)
+        &= \sum_{j} \frac{n_j(\fullAlignment)}{n}\cdot  \log P(\sitePattern_j, \ancestralStateCategory_j | \tau, t) \\
+        &= \sum_{j} \frac{n_j(\fullAlignment)}{n}\cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)].
 \end{align}
@@
 $$
-\frac{1}{n} \log L_n''(\mathbf{y};\mathbf{z},\tau,t) \rightarrow \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)]
+\frac{1}{n} \log L_n''(\fullAlignment;,\tau,t) \rightarrow \sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)]
 $$
-and that, since $z_j$ appears only in the last term, the log of \eqref{eq:site_pattern_likelihood} in the same limit is proportional to
+and that, since $\ancestralStateCategory_j$ appears only in the last term, the log of \eqref{eq:site_pattern_likelihood} in the same limit is proportional to
 \begin{align}
-\frac{1}{n} \log L_n'(\mathbf{y};\tau, t) &= \max_{\mathbf{z}} \ \frac{1}{n} \log L_n''(\mathbf{s};\mathbf{z},\tau,t) \nonumber \\
-    &\rightarrow \max_{z_1, \ldots, z_q} \ \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)] \nonumber \\
-    &= \max_{\mathbf{z}} \ \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \log P(z_j | s_j, \tau, t)] \nonumber \\
-    &= \max_{\mathbf{z}} \ \left\{\sum_{j} P(s_j | \tau^*, t^*) \log P(s_j | \tau, t)\right\} \nonumber \\
-    &\qquad + \left\{\sum_{j} P(s_j | \tau^*, t^*) \log P(z_j | s_j, \tau, t)\right\}. \label{eq:site_pattern_profile_likelihood_mean}
+\frac{1}{n} \log L_n'(\fullAlignment;\fullAncestralStateCategories,\tau, t)
+    &= \max_{\fullAncestralStateCategories} \ \frac{1}{n} \log L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t) \nonumber \\
+    &\rightarrow \max_{\fullAncestralStateCategories} \ \sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)] \nonumber \\
+    &= \max_{\fullAncestralStateCategories} \ \sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)] \nonumber \\
+    &= \max_{\fullAncestralStateCategories} \ \left\{\sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot \left[\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)\right]\right\}.
+\label{eq:site_pattern_profile_likelihood_mean}
 \end{align}
-
-\subsection{Showing inconsistency}
-
-For clarity, define \eqref{eq:site_pattern_profile_likelihood_mean} as
+Define the the information quantity
 $$
-J(\tau, t; \tau^*, t^*) = \sum_{j} P(s_j | \tau^*, t^*) \cdot [\log P(s_j | \tau, t) + \max_{z_j} \ \log P(z_j | s_j, \tau, t)]
+H_{\tau^*,t^*}(\tau,t) = \sum_{j} P(\sitePattern_j | \tau^*, t^*) \log P(\sitePattern_j | \tau, t),
 $$
-where $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
+and the partial log likelihood
+$$
+\ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = \sum_{j} P(\sitePattern_j | \tau^*, t^*) \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)
+$$
+so that \eqref{eq:site_pattern_profile_likelihood_mean} is written as
+\begin{equation}
+    \label{eq:log_likelihood_simplified}
+\ell_{\tau^*,t^*}(\tau,t) = \max_{\fullAncestralStateCategories} \ \left\{H_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t)\right\}.
+\end{equation}
+Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
 For data generated from $(\tau^*, t^*)$, if
@@
 \label{eq:inconsistency_inequality}
-J(\tau', \hat{t}_1; \tau^*, t^*) > J(\tau^*, \hat{t}_2; \tau^*, t^*)
+\ell_{\tau^*,t^*}(\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\tau^*,\hat{t}_2)
 \end{equation}
@@
 \section{Inconsistency of joint maximum likelihood}
+
+\subsection{Bounding the likelihoods}
+
+Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified} as
+\begin{equation}
+\label{eq:partial-likelihood}
+\ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = H_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
+\end{equation}
+We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
+$$
+\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \le
+    H_{\tau^*,t^*}(\tau^*,t^*)
+    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t})
+$$
+using Gibbs's inequality
+$$
+H_{\tau^*,t^*}(\tau,t) \le H_{\tau^*,t^*}(\tau^*,t^*)
+$$
+and
+$$
+\{\hat{\fullAncestralStateCategories},\hat{t}\} = \arg\max_{\{\fullAncestralStateCategories,t\}} \ \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
+$$
+Similarly, we have a lower bound as
+$$
+\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \ge
+    H_{\tau^*,t^*}(\tau,\hat{t})
+    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t}).
+$$
+
+% vvvv Remove these? I think they're obvious now, but maybe they can be made more explicit above?
+\textbf{Argument for lower bound}: This makes intuitive sense, though may be difficult to show in general.
+Can we not just say $\max_u f(u) + g(u)$ is greater than $f(u') + g(u')$ for any $u'$ that doesn't maximize $f(u) + g(u)$ just by definition?
+There could be multiple maxima, but that's why we have greater than or equal to.
+I think the only thing we'd need to show is that a maximum exists, but since $\{x,y,w\}$ are supported on a closed set this is certainly true.
+I must be missing something.
+
+\textbf{Argument for upper bound}: I think this is true mostly because $\max_u f(u) + g(u)$ should be less than $\max_u f(u) + \max_u g(u)$ by some triangle inequality argument.
+If we were to actually maximize the full likelihood, we should get something smaller than maximizing some form of it that we consider by looking at the parts separately.
+For a fixed $\{x^*, y^*\}$ we can calculate this upper bound directly---can we obtain it as a useful-looking function of these parameters in the general case?
 
@@
 The branch length parameters are such that the probability of a change in character along the top two branches is $p_x=1-2x$, with corresponding equalities for the other two branches.
-Here, $\{p_x,p_y,p_z\}\in[0,1/2]^3$ and $\{x,y,z\}\in[0,1]^3$.
+Here, $\{p_x,p_y,p_w\}\in[0,1/2]^3$ and $\{x,y,w\}\in[0,1]^3$.
 
@@
     \hline
-$s_j$   &$P(s_j|\tau_1,t^*)$&$P(s_j|\tau_2,t^*)$\\
+$\sitePattern_j$   &$P(\sitePattern_j|\tau_1,t^*)$&$P(\sitePattern_j|\tau_2,t^*)$\\
     \hline
@@
     \hline
-$s_j$   &$P(s_j|\tau_1,t)$&$P(s_j|\tau_2,t)$\\
+$\sitePattern_j$   &$P(\sitePattern_j|\tau_1,t)$&$P(\sitePattern_j|\tau_2,t)$\\
     \hline
@@
 \begin{tabular}{|l|ll|}
-\multicolumn{3}{c}{$P(z_j|s_j,\tau_1,t)$}\\
+\multicolumn{3}{c}{$P(\ancestralStateCategory_j|\sitePattern_j,\tau_1,t)$}\\
 \hline
-& \multicolumn{2}{|c|}{$z_j$}\\
+& \multicolumn{2}{|c|}{$\ancestralStateCategory_j$}\\
     \hline
-$s_j$    &00                              &01\\
+$\sitePattern_j$    &00                              &01\\
     \hline
@@
     \hline
-    \multicolumn{3}{c}{$P(z_j|s_j,\tau_2,t)$}\\
+    \multicolumn{3}{c}{$P(\ancestralStateCategory_j|\sitePattern_j,\tau_2,t)$}\\
 \hline
-& \multicolumn{2}{|c|}{$z_j$}\\
+& \multicolumn{2}{|c|}{$\ancestralStateCategory_j$}\\
     \hline
-$s_j$    &00                              &01\\
+$\sitePattern_j$    &00                              &01\\
     \hline
@@
 
-\subsection{Bounding the likelihoods}
+\subsubsection{Upper bound for Farris partial likelihood}
 
 Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
-Our approach will be to decompose the likelihood inside the maximization in \eqref{eq:site_pattern_profile_likelihood_mean} as
-\begin{equation}
-\label{eq:partial-likelihood}
-L(x', y', w'; \mathbf{z},\tau') = H(x,y,\tau_1;x',y',w', \tau') + L_{p}(x',y',w';\mathbf{z},\tau')
-\end{equation}
-where different values for $\mathbf{z}$ will alter the form of $L_{p}(x',y',w';\mathbf{z},\tau')$.
-We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
+To obtain an upper bound for the Farris partial likelihood---the terms in the top half of Table~\ref{tab:likelihoods}---we use the term $(1+x)^2(1+w)(1+y)^2$ for the $0101$ site pattern.
+Since all cases have a $(1+x)^2$ term except $\{0100, 1000, 0011, 1001\}$, and all cases except $\{0001, 0010, 0011, 1001\}$ have a $(1+y)^2$, we can add and subtract the $\log$ of these terms weighted by each of their generating probabilities.
+In other words, define
+\begin{align*}
+    a_{xy} &= P(0100|\tau=\tau_1,t=\{x,y,y\}) + P(1000|\tau=\tau_1,t=\{x,y,y\}) \\
+           &\qquad + P(0011|\tau=\tau_1,t=\{x,y,y\}) + P(1001|\tau=\tau_1,t=\{x,y,y\})
+\end{align*}
+and
+\begin{align*}
+    b_{xy} &= P(1000|\tau=\tau_1,t=\{x,y,y\}) + P(0010|\tau=\tau_1,t=\{x,y,y\}) \\
+           &\qquad + P(0011|\tau=\tau_1,t=\{x,y,y\}) + P(1001|\tau=\tau_1,t=\{x,y,y\}).
+\end{align*}
+Then the ancestral state maximizing the likelihood is $00$ for each site pattern and the partial likelihood is bounded above
+\begin{align*}
+    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,\{x',y',w'\}) \\
+    &\qquad \le (2-a_{xy})\log(1+x')+a_{xy}\log(1-x')+(2-b_{xy})\log(1+y')+b_{xy}\log(1-y')+\log(1+w') \\
+    &\qquad \le (2-a_{xy})\log(2-a_{xy})+a_{xy}\log(a_{xy})+(2-b_{xy})\log(2-b_{xy})+b_{xy}\log(b_{xy})+\log(2) \\
+    &\qquad \triangleq L^p_{\tau_1,\tau_1}(a_{xy}, b_{xy})
+\end{align*}
+where $a_{xy}$ and $b_{xy}$ are functions of the true generating branch parameters $\{x,y\}$.
+
+\subsubsection{Lower bound for Felsenstein partial likelihood}
+
+We construct a similar lower bound on the Felsenstein partial likelihood, though here we replace all $(1+w)$ terms with $(1-w)$.
+Let
+\begin{align*}
+    c_{xy} &= P(0001|\tau=\tau_1,t=\{x,y,y\}) + P(0010|\tau=\tau_1,t=\{x,y,y\}) \\
+           &\qquad + P(0100|\tau=\tau_1,t=\{x,y,y\}) + P(1000|\tau=\tau_1,t=\{x,y,y\})
+\end{align*}
+and
 $$
-\max_{\{x',y',w',\mathbf{z}\}} \ L(x',y',w';\mathbf{z},\tau') \le H(x,y,\tau_1;x,y,y,\tau_1) + L_{p}(\hat{x}, \hat{y}, \hat{w};\hat{\mathbf{z}},\tau')
+    d_{xy} = P(0101|\tau=\tau_1,t=\{x,y,y\}) + P(1001|\tau=\tau_1,t=\{x,y,y\}).
 $$
-using Gibbs inequality
+The lower bounds on the likelihoods in this case take one of six forms:
+Since the maxima for pairs of lower bounds above each have a symmetry in $x$ and $y$, they are bounded below by one of
+\textbf{TODO}: ancestral states where these likelihoods occur
+
+\begin{align*}
+    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,\{x',y',w'\}) \\
+    &\qquad \ge 2\log(1+x')+(2-c_{xy}-2d_{xy})\log(1+y')+(c_{xy}+2d_{xy})\log(1-y')+\log(1-w') \\
+    &\qquad \ge 2\log(2)+(2-c_{xy}-2d_{xy})\log(2-c_{xy}-2d_{xy})+(c_{xy}+2d_{xy})\log(c_{xy}+2d_{xy}) \\
+    &\qquad \triangleq L^{p,1}_{\tau_1,\tau_2}(c_{xy}, d_{xy}),
+\end{align*}
+\begin{align*}
+    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,\{x',y',w'\}) \\
+    &\qquad \ge (2-2d_{xy})\log(1+x')+2d_{xy}\log(1-x') + (2-c_{xy})\log(1+y')+c_{xy}\log(1-y')+\log(1-w') \\
+    &\qquad \ge (2-2d_{xy})\log(2-2d_{xy})+2d_{xy}\log(2d_{xy}) + (2-c_{xy})\log(2-c_{xy})+c_{xy}\log(c_{xy}) \\
+    &\qquad \triangleq L^{p,2}_{\tau_1,\tau_2}(c_{xy}, d_{xy})
+\end{align*}
+or
+\begin{align*}
+    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,\{x',y',w'\}) \\
+    &\qquad \ge (2-d_{xy})\log(1+x')+d_{xy}\log(1-x') + (2-c_{xy}-d_{xy})\log(1+y')+(c_{xy}+d_{xy})\log(1-y')+\log(1-w') \\
+    &\qquad \ge (2-d_{xy})\log(2-d_{xy})+d_{xy}\log(d_{xy}) + (2-c_{xy}-d_{xy})\log(2-c_{xy}-d_{xy})+(c_{xy}+d_{xy})\log(c_{xy}+d_{xy}) \\
+    &\qquad \triangleq L^{p,3}_{\tau_1,\tau_2}(c_{xy}, d_{xy})
+\end{align*}
+where $c_{xy}$ and $d_{xy}$ are also functions of the true generating branch parameters $\{x,y\}$.
+
+\textbf{Claim to prove}:
 $$
-H(x, y, \tau_1;x', y', w', \tau') \le H(x, y, \tau_1;x, y, y, \tau_1)
+L^{p,1}_{\tau_1,\tau_2}(c_{xy}, d_{xy}) \ge L^{p,2}_{\tau_1,\tau_2}(c_{xy}, d_{xy})
 $$
@@
 $$
-\{\hat{x}, \hat{y}, \hat{w}, \hat{\mathbf{z}}\} = \arg\max_{\{x,y,w,\mathbf{z}\}} \ L_{p}(x,y,w;\mathbf{z},\tau').
+L^{p,1}_{\tau_1,\tau_2}(c_{xy}, d_{xy}) \ge L^{p,3}_{\tau_1,\tau_2}(c_{xy}, d_{xy}).
 $$
-Similarly, we have a lower bound as
-$$
-\max_{\{x',y',w',\mathbf{z}\}} \ L(x',y',w';\mathbf{z},\tau') \ge H(x,y,\tau_1; \hat{x}, \hat{y}, \hat{w}, \tau') + L_{p}(\hat{x}, \hat{y}, \hat{w},\hat{\mathbf{z}},\tau').
-$$
+I think we can use information theory results to show this.
+It is definitely true---will work on it.
 
-% vvvv Remove these? I think they're obvious now, but maybe they can be made more explicit above?
-\textbf{Argument for lower bound}: This makes intuitive sense, though may be difficult to show in general.
-Can we not just say $\max_u f(u) + g(u)$ is greater than $f(u') + g(u')$ for any $u'$ that doesn't maximize $f(u) + g(u)$ just by definition?
-There could be multiple maxima, but that's why we have greater than or equal to.
-I think the only thing we'd need to show is that a maximum exists, but since $\{x,y,w\}$ are supported on a closed set this is certainly true.
-I must be missing something.
-
-\textbf{Argument for upper bound}: I think this is true mostly because $\max_u f(u) + g(u)$ should be less than $\max_u f(u) + \max_u g(u)$ by some triangle inequality argument.
-If we were to actually maximize the full likelihood, we should get something smaller than maximizing some form of it that we consider by looking at the parts separately.
-For a fixed $\{x^*, y^*\}$ we can calculate this upper bound directly---can we obtain it as a useful-looking function of these parameters in the general case?
-
-\subsubsection{Upper bound for Farris likelihood}
-
-To obtain an upper bound for the Farris likelihood---the terms in the top half of Table~\ref{tab:likelihoods}---we use the term $(1+x)^2(1+w)(1+y)^2$ for the $0101$ site pattern.
-Since all cases have a $(1+x)^2$ term except $\{0100, 1000, 0011, 1001\}$, and all cases except $\{0001, 0010, 0011, 1001\}$ have a $(1+y)^2$, we can add and subtract the $\log$ of these terms weighted by each of their generating probabilities.
-In other words, define
-$$
-a = P(0100|\tau_1,\{x,y,y\}) + P(1000|\tau_1,\{x,y,y\}) + P(0011|\tau_1,\{x,y,y\}) + P(1001|\tau_1,\{x,y,y\})
-$$
-and
-$$
-b = P(1000|\tau_1,\{x,y,y\}) + P(0010|\tau_1,\{x,y,y\}) + P(0011|\tau_1,\{x,y,y\}) + P(1001|\tau_1,\{x,y,y\}).
-$$
-Then the ancestral state maximizing the likelihood is $00$ and $L_{p}(x',y',w';\mathbf{z},\tau_1)$ is bounded above by
-$$
-L_{p}(x',y',w';00,\tau_1) \le (2-a)\log(1+x')+a\log(1-x')+(2-b)\log(1+y')+b\log(1-y')+\log(1+w')
-$$
-with a maximum
-$$
-\hat{L}_{p}(1-a,1-b,1;00,\tau_1) \le (2-a)\log(2-a)+a\log(a)+(2-b)\log(2-b)+b\log(b)+\log(2)
-$$
-where $a$ and $b$ are functions of the true generating branch parameters $\{x,y\}$.
-
-\subsubsection{Lower bound for Felsenstein likelihood}
-
-We construct a similar lower bound on the Felsenstein likelihood, though here we replace all $(1+w)$ terms with $(1-w)$.
-Let
-$$
-c = P(0001|\tau_1,\{x,y,y\}) + P(0010|\tau_1,\{x,y,y\}) + P(0100|\tau_1,\{x,y,y\}) + P(1000|\tau_1,\{x,y,y\})
-$$
-and
-$$
-d = P(0101|\tau_1,\{x,y,y\}) + P(1001|\tau_1,\{x,y,y\}).
-$$
-The lower bounds on the likelihoods in this case take one of six forms:
-\textbf{TODO}: ancestral states where these likelihoods occur
-$$
-L_{p}(x',y',w';z,\tau_2) \ge 2\log(1+x')+(2-c-2d)\log(1+y')+(c+2d)\log(1-y')+\log(1-w'),
-$$
-$$
-L_{p}(x',y',w';z,\tau_2) \ge (2-c-2d)\log(1+x')+(c+2d)\log(1-x')+2\log(1+y')+\log(1-w'),
-$$
-$$
-L_{p}(x',y',w';z,\tau_2) \ge (2-2d)\log(1+x')+2d\log(1-x')+(2-c)\log(1+y')+c\log(1-y')+\log(1-w'),
-$$
-$$
-L_{p}(x',y',w';z,\tau_2) \ge (2-c)\log(1+x')+c\log(1-x')+(2-2d)\log(1+y')+2d\log(1-y')+\log(1-w'),
-$$
-$$
-L_{p}(x',y',w';z,\tau_2) \ge (2-d)\log(1+x')+d\log(1-x')+(2-c-d)\log(1+y')+(c+d)\log(1-y')+\log(1-w')
-$$
-or
-$$
-L_{p}(x',y',w';z,\tau_2) \ge (2-c-d)\log(1+x')+(c+d)\log(1-x')+(2-d)\log(1+y')+d\log(1-y')+\log(1-w').
-$$
-Since the maxima for pairs of lower bounds above each have a symmetry in $x$ and $y$, they are bounded below by one of
-$$
-\hat{L}_{p}(1,1-c-2d,0;z,\tau_2) \ge 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d),
-$$
-$$
-\hat{L}_{p}(1-2d,1-c,0;z,\tau_2) \ge (2-2d)\log(2-2d)+2d\log(2d)+(2-c)\log(2-c)+c\log(c)
-$$
-and
-$$
-\hat{L}_{p}(1-d,1-c-d,0;z,\tau_2) \ge (2-d)\log(2-d)+d\log(d)+(2-c-d)\log(2-c-d)+(c+d)\log(c+d)
-$$
-where $c$ and $d$ are also functions of the true generating branch parameters $\{x,y\}$.
-
-\textbf{TODO}: cases for $\{x, y\}$ where each of the above is the corresponding maximum.
-I think the first is always greater than or equal to the other two, but I haven't been able to show it yet.
-It may take a little information theory--fu, or just standard application of the bounds in the introduction to this section.
-
-Thus, let
-$$
-\hat{L}_{p}(1,1-c-2d,0;z,\tau_2) = 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d).
-$$
+The values of $(x', y', w')$ maximizing $L^{p,1}_{\tau_1,\tau_2}(c_{xy}, d_{xy})$ are $(1, 1-c_{xy}-2d_{xy},0)$.
 
@@
 $$
-H(x,y,\tau_1; 1, 1-c-2d, 0, \tau_2) + \hat{L}_{p}(1,1-c-2d,0;z,\tau_2) \ge H(x,y,\tau_1; x, y, y,\tau_1) + \hat{L}_{p}(1-a,1-b,1;00,\tau_1).
+H_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-c_{xy}-2d_{xy}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(c_{xy},d_{xy}) \ge H_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p}_{\tau_1,\tau_1}(a_{xy},b_{xy}).
 $$
@@
 \begin{align*}
-    0 &\ge H(x,y,\tau_1; x, y, y,\tau_1) + \hat{L}_{p}(1-a,1-b,1;00,\tau_1) \\
-      &\qquad - H(x,y,\tau_1; 1, 1-c-2d, 0, \tau_2) - \hat{L}_{p}(1,1-c-2d,0;\tau_2) \\
-      & = 2(12+2x^2+2y^2)\log((12+2x^2+2y^2)/8) \\
-      &\qquad + 2(4-2x^2-2y^2)\log((4-2x^2-2y^2)/8)  + 8\log2 \\
-      &\qquad + 2(1+x^2-y^2-x^2y^2)\log(1+x^2-y^2-x^2y^2) \\
+    0 &\ge H_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p}_{\tau_1,\tau_1}(a_{xy},b_{xy}) \\
+      &\qquad - H_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-c_{xy}-2d_{xy}, 0\}) - L^{p,1}_{\tau_1,\tau_2}(c_{xy},d_{xy}) \\
+%
+      &\qquad = 2(1+x^2-y^2-x^2y^2)\log(1+x^2-y^2-x^2y^2) \\
       &\qquad + 2(1-x^2+y^2-x^2y^2)\log(1-x^2+y^2-x^2y^2) \\
@@
       &\qquad + (1+x^2+y^2-4xy^2+x^2y^2)\log(1+x^2+y^2-4xy^2+x^2y^2) \\
-      &\qquad - 16\log2 - 8(1+xy^2)\log(1+xy^2) - 8(1-xy^2)\log(1-xy^2) \\
-      &\qquad - (2-4xy^2+2x^2y^2)\log(1-2xy^2+x^2y^4) \\
-      &\qquad - (2+4xy^2+2x^2y^2)\log(1+2xy^2+x^2y^4) \\
-      &\qquad - (4-4x^2y^2)\log(1-x^2y^4).
+%
+      &\qquad + (2-a_{xy})\log(2-a_{xy})+a_{xy}\log(a_{xy})+(2-b_{xy})\log(2-b_{xy})+b_{xy}\log(b_{xy})+\log(2) \\
+%
+      &\qquad - 2(1+x^2-y^2-x^2y^2)\log(1+x'^2w'-y'^2w'-x'^2y'^2) \\
+      &\qquad - 2(1-x^2+y^2-x^2y^2)\log(1-x'^2w'+y'^2w'-x'^2y'^2) \\
+      &\qquad - (1-x^2-y^2+x^2y^2)\log(1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2) \\
+      &\qquad - (1-x^2-y^2+x^2y^2)\log(1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2) \\
+      &\qquad - (1+x^2+y^2+4xy^2+x^2y^2)\log(1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2) \\
+      &\qquad - (1+x^2+y^2-4xy^2+x^2y^2)\log(1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2) \\
+%
+      &\qquad - 2\log(2)-(2-c_{xy}-2d_{xy})\log(2-c_{xy}-2d_{xy})-(c_{xy}+2d_{xy})\log(c_{xy}+2d_{xy})
 \end{align*}
-
-\textbf{TODO}: can we use sage to simplify the above so it looks nicer?
-Write it as functions of $a,b,c,d$ so it's less messy?
-What would be best?
+where
+$$
+a_{xy} = b_{xy} = \frac{4-2x^2-2y^2}{8},
+$$
+$$
+c_{xy} = \frac{4-4x^2y^2}{8},
+$$
+$$
+d_{xy} = \frac{2-4xy^2+2x^2y^2}{8}
+$$
+and
+$$
+x' = 1, \ y' = 1-c_{xy}-2d_{xy}, \ w' = 0.
+$$
 
@@
 var('x,y')
-inconsistency = 2*(12+2*x^2+2*y^2)*log((12+2*x^2+2*y^2)/8) \
-    + 2*(4-2*x^2-2*y^2)*log((4-2*x^2-2*y^2)/8) + 8*log(2.) \
-    + 2*(1+x^2-y^2-x^2*y^2)*log(1+x^2-y^2-x^2*y^2) \
+a = (4-2*x^2-2*y^2)/8
+b = (4-2*x^2-2*y^2)/8
+c = (4-4*x^2*y^2)/8
+d = (2-4*x*y^2+2*x^2*y^2)/8
+xp = 1.
+yp = 1-c-2*d
+wp = 0.
+
+inconsistency = 2*(1+x^2-y^2-x^2*y^2)*log(1+x^2-y^2-x^2*y^2) \
     + 2*(1-x^2+y^2-x^2*y^2)*log(1-x^2+y^2-x^2*y^2) \
@@
     + (1+x^2+y^2-4*x*y^2+x^2*y^2)*log(1+x^2+y^2-4*x*y^2+x^2*y^2) \
-    - 16*log(2.) - 8*(1+x*y^2)*log(1+x*y^2) - 8*(1-x*y^2)*log(1-x*y^2) \
-    - (2-4*x*y^2+2*x^2*y^2)*log(1-2*x*y^2+x^2*y^4) \
-    - (2+4*x*y^2+2*x^2*y^2)*log(1+2*x*y^2+x^2*y^4) - (4-4*x^2*y^2)*log(1-x^2*y^4)
+    + (2-a)*log(2-a)+a*log(a)+(2-b)*log(2-b)+b*log(b)+log(2.) \
+    - 2*(1+x^2-y^2-x^2*y^2)*log(1+xp^2*wp-yp^2*wp-xp^2*yp^2) \
+    - 2*(1-x^2+y^2-x^2*y^2)*log(1-xp^2*wp+yp^2*wp-xp^2*yp^2) \
+    - (1-x^2-y^2+x^2*y^2)*log(1+2*xp*yp-2*xp*yp*wp-xp^2*wp-yp^2*wp+xp^2*yp^2) \
+    - (1-x^2-y^2+x^2*y^2)*log(1-2*xp*yp+2*xp*yp*wp-xp^2*wp-yp^2*wp+xp^2*yp^2) \
+    - (1+x^2+y^2+4*x*y^2+x^2*y^2)*log(1+2*xp*yp+2*xp*yp*wp+xp^2*wp+yp^2*wp+xp^2*yp^2) \
+    - (1+x^2+y^2-4*x*y^2+x^2*y^2)*log(1-2*xp*yp-2*xp*yp*wp+xp^2*wp+yp^2*wp+xp^2*yp^2) \
+    - 2*log(2.)-(2-c-2*d)*log(2-c-2*d)-(c+2*d)*log(c+2*d)
 
-inconsistency_plot = implicit_plot(
+incons_plot = implicit_plot(
     inconsistency(x,y)==0,
@@
 $$
-a = b = 0.46875
+a_{xy} = b_{xy} = 0.46875
 $$
@@
 $$
-\hat{L}^{p}(\hat{x}, \hat{y}, \hat{w},\tau_1) \approx -2.17803560694
+L^{p}_{\tau_1,\tau_1}(a_{xy},b_{xy}) \approx -2.17803560694157
 $$
@@
 $$
-H(x,y,\tau_1;x,y,y,\tau_1) \approx -2.07509999739.
+H_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\})  \approx -2.04470918734314.
 $$
@@
 $$
-\max_{x,y,w} \ L(x,y,w,\tau_1) \le -4.25313560433
+\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,t) \le -4.22274479428471
 $$
@@
 $$
-\max_{x,y,w} \ L(x,y,w,\tau_2) \ge -4.25313560433
+\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,t) \ge -4.22274479428471.
 $$
@@
 $$
-c = 0.498046875, \ d = 0.2431640625,
+c_{xy} = 0.498046875, \ d_{xy} = 0.2431640625,
 $$
@@
 $$
-\hat{L}(\hat{x}, \hat{y}, \hat{w},\tau_2) \approx -2.07919739112
-$$
-Calculating shows%
-    %
-    \footnote{It's not a typo that $\hat{L}_{fe}$ and $H_{fe}$ are the same---at least not in the code.
-        I should try to show why this is the case.
-        }
-    %
-$$
-H(x,y,\tau_1; \hat{x}, \hat{y}, \hat{w},\tau_2) \approx -2.07919739112
+L^{p,1}_{\tau_1,\tau_2}(c_{xy},d_{xy}) \approx -2.07968569223991
 $$
@@
 $$
-\max_{x,y,w} \ L(x,y,w,\tau_2) \ge -4.15839478224 > -4.25313560433 \ge \max_{x,y,w} \ L(x,y,w,\tau_1)
+H_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-c_{xy}-2d_{xy}, 0\}) \approx -2.07748833719921.
+$$
+The lower bound in this case becomes
+$$
+\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,t) \ge -4.22274479428471 \ge -4.22274479428471 \ge \max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,t)
 $$
@@
 $$
-\max_{x,y,w} \ L(x,y,w,\tau_1) \approx -4.5101771294218045
+\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,t) \approx -4.5101771294218045
 $$
@@
 $$
-\max_{x,y,w} \ L(x,y,w,\tau_2) \approx -4.1544859851260281.
+\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,t) \approx -4.1544859851260281.
 $$

### 2017-03-24 joint_inf.tex
Edits.

@@
 
+We will be considering phylogenetic likelihood on unrooted phylogenetic trees.
+We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
+Let $m$ be the number of tips of the tree, and $p = m-2$ the number of internal nodes.
+%EM Commented out below because it's not right-- fullAlignment is not the set of assignments of characters to tips of the tree, it's a single such.
+%Let $\fullAlignment$ be the set of assignments of characters to tips of our tree, and $\fullAncestralStates$ the set of assignments of characters to the internal nodes of the tree.
+Assume we observe $n$ samples of character data (i.e.\ an alignment with $n$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_n]\in\alphabet^{\nSiteRows\times n}$.
+There are also corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_n]\in\alphabet^{\nAncestralStateRows\times n}$.
+
 \subsection{Classical maximum-likelihood inference in phylogenetics}
 
-Let $\fullAlignment$ be the set of assignments of characters to tips of our tree, and $\fullAncestralStates$ the set of assignments of characters to the internal nodes of the tree.
-Assume we observe $n$ samples of character data (i.e.\ an alignment with $n$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_n]\in\alignmentSpace^{\nSiteRows\times n}$ and corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_n]\in\ancestralStateSpace^{\nAncestralStateRows\times n}$.
-Without loss of generality, assume $\alignmentSpace=\ancestralStateSpace=\{0,1\}$.
 For a topology $\tau$ and branch lengths $t$ we form the fully-observed likelihood (making the usual independence assumption between sites):
@@
 while using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
-Since $\ancestralStateSpace$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
+Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
 In the general case, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
-For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.
+For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence cite{Neher2017}.
 
@@
 I also changed from "pattern" to "category" which might make it a little more clear.
+%EM Seems good this is an important concept so it's worth defining it more properly.
 We enumerate the unique site patterns $\mathbf{s}=[s_1,\ldots,s_q]\in\mathcal{S}$; these are defined by the number of leaves and the character alphabet.
 Corresponding to each of these site patterns will be an ancestral state category that is most likely given a particular likelihood.
-Since $\alignmentSpace$ and $\ancestralStateSpace$ are finite sets, the value of each factor $P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t)$ in \eqref{eq:full_likelihood} is maximized at one or more values of $\ancestralStateColumn_i$, though independent of $i$ given $j$.
+Since $\alphabet$ is a finite set, the value of each factor $P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t)$ in \eqref{eq:full_likelihood} is maximized at one or more values of $\ancestralStateColumn_i$, though independent of $i$ given $j$.
 This independence allows us to group factors by those where $\ancestralStateColumn_i$ belongs to an ancestral state category $\ancestralStateCategory_j$ so that, if $\alignmentColumn_i=\sitePattern_j$ and $\ancestralStateColumn_i\in \ancestralStateCategory_j$, $P(\sitePattern_j, \ancestralStateCategory_j | \tau, t)$ is maximized.
@@
                              &= \max_{\fullAncestralStates} \ \prod_{i=1}^{n} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t) \\
-%EM Can you think a little about z_j? The indexing seems a little problematic now that we are thinking of everything as a function of y.
+%EM Can you think a little about z_j?
+%EM Even though it's a category now, it's still a function of tau and t, right?
                              &= \max_{\fullAncestralStateCategories} \ \prod_{j=1}^{q} \ P(\sitePattern_j, \ancestralStateCategory_j | \tau, t)^{n_j(\fullAlignment)} \label{eq:site_pattern_likelihood}

### 2017-03-28 joint_inf.tex
Bringing commands into main text; easier to find.
updates

@@
 
-\input{commands}
+% Commands
+\newcommand{\alphabet}{\mathcal{A}}
+\newcommand{\fullAlignment}{\mathbf{Y}}
+\newcommand{\alignmentColumn}{\mathbf{y}}
+\newcommand{\sitePattern}{s}
+\newcommand{\siteSplit}{S'}
+\newcommand{\sitePatternSet}{S}
+\newcommand{\fullSitePatterns}{\mathbf{s}}
+\newcommand{\sitePatternSpace}{\mathcal{S}}
+\newcommand{\fullAncestralStates}{\mathbf{H}}
+\newcommand{\ancestralStateSpace}{\mathcal{H}}
+\newcommand{\ancestralStateColumn}{\boldsymbol\eta}
+\newcommand{\ancestralStateCategory}{Z}
+\newcommand{\ancestralSplit}{H'}
+\newcommand{\ancestralStateSet}{H}
+\newcommand{\fullAncestralStateCategories}{\mathbb{Z}}
+\newcommand{\ancestralStateCategorySpace}{\mathcal{Z}}
+
+\newcommand{\nCols}{n}
+\newcommand{\nSiteRows}{m}
+\newcommand{\nAncestralStateRows}{p}
+\newcommand{\nSiteSplits}{q}
+\newcommand{\nAncestralSplits}{r}
+\newcommand{\nMostLikelyAncestralSplits}{\tilde{r}}
 
@@
 We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
-Let $m$ be the number of tips of the tree, and $p = m-2$ the number of internal nodes.
+Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
 %EM Commented out below because it's not right-- fullAlignment is not the set of assignments of characters to tips of the tree, it's a single such.
 %Let $\fullAlignment$ be the set of assignments of characters to tips of our tree, and $\fullAncestralStates$ the set of assignments of characters to the internal nodes of the tree.
-Assume we observe $n$ samples of character data (i.e.\ an alignment with $n$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_n]\in\alphabet^{\nSiteRows\times n}$.
-There are also corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_n]\in\alphabet^{\nAncestralStateRows\times n}$.
+Assume we observe $\nCols$ samples of character data (i.e.\ an alignment with $\nCols$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$.
+There are also corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$.
 
@@
 \label{eq:full_likelihood}
-L_n(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{n} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t).
+L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t).
 \end{equation}
@@
 \label{eq:marginal_likelihood}
-\tilde{L}_n(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t)
+\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
 \end{equation}
@@
 $$
-(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_n(\fullAlignment; \tau, t).
+(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
 $$
@@
 \label{eq:profile_likelihood}
-L_n'(\fullAlignment;\tau, t) = \max_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t),
+L_\nCols'(\fullAlignment;\tau, t) = \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t),
 \end{equation}
@@
 \label{eq:profile_likelihood_topology_bl}
-(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ L_n'(\fullAlignment;\tau, t)
+(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ L_\nCols'(\fullAlignment;\tau, t)
 \end{equation}
@@
 In the general case, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
-For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence cite{Neher2017}.
+For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.
 
-\subsection{Site pattern formulation}
+\subsection{Site split formulation}
 
 Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
-These are commonly called ``site patterns'' in the literature.
+Following Section 8.6 of \cite{Semple2003}, $\tau$ induces an ordered set of tip labels $\{1,\ldots,\nSiteRows\}$ across sites.
+For a binary character model, a realization of $\alignmentColumn_i$ can be described as a subset of complementary characters $\siteSplit\subseteq\sitePatternSet\triangleq\{1,\ldots,\nSiteRows-1\}$.
+The topology $\tau$ induces an ordered set $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ with a similar subset of complementary characters $\ancestralSplit\subseteq\ancestralStateSet\triangleq\{1,\ldots,\nAncestralStateRows\}$ describing a realization of $\ancestralStateColumn_i$.
+Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\sitePatternSet)|$ and unique ancestral state splits as $\nAncestralSplits=|\mathcal{P}(\ancestralStateSet)|$ where $\mathcal{P}$ denotes the powerset.
+
 %EM in general aren't a different number of internal site patterns than tip site patterns?
@@
 %EM Seems good this is an important concept so it's worth defining it more properly.
-We enumerate the unique site patterns $\mathbf{s}=[s_1,\ldots,s_q]\in\mathcal{S}$; these are defined by the number of leaves and the character alphabet.
-Corresponding to each of these site patterns will be an ancestral state category that is most likely given a particular likelihood.
-Since $\alphabet$ is a finite set, the value of each factor $P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t)$ in \eqref{eq:full_likelihood} is maximized at one or more values of $\ancestralStateColumn_i$, though independent of $i$ given $j$.
-This independence allows us to group factors by those where $\ancestralStateColumn_i$ belongs to an ancestral state category $\ancestralStateCategory_j$ so that, if $\alignmentColumn_i=\sitePattern_j$ and $\ancestralStateColumn_i\in \ancestralStateCategory_j$, $P(\sitePattern_j, \ancestralStateCategory_j | \tau, t)$ is maximized.
 
-For ancestral state categories $\fullAncestralStateCategories=[\ancestralStateCategory_1,\ldots,\ancestralStateCategory_q]\in\ancestralStateCategorySpace$ we reformulate the likelihood equation
+Consider the probability $P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t)$ from \eqref{eq:full_likelihood} of observing a particular column $i$.
+Given $(\tau, t)$ and maximizing over $\ancestralStateColumn_i$, if $\alignmentColumn_i=\siteSplit_j$ then there exists a partition $\ancestralStateCategory_j$ of $\mathcal{P}(\ancestralStateSet)$ of most likely ancestral states where
+\begin{align*}
+\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralStateSet)} \ P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i=\ancestralSplit_k | \tau, t) &= P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i\in\ancestralStateCategory_j | \tau, t) \\
+     &= P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t).
+\end{align*}
+This equivalence is helpful in writing the full likelihood of the joint inference problem in terms of site splits; for most likely ancestral state partitions $\fullAncestralStateCategories=\{\ancestralStateCategory_1,\ldots,\ancestralStateCategory_\nSiteSplits\}$ we reformulate the likelihood equation
 \begin{align}
-L_n'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t) \\
-                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{n} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t) \\
+L_\nCols'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) \\
+                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{\nCols} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t) \\
+                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralStateSet)} \ P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i=\ancestralSplit_k | \tau, t) \\
 %EM Can you think a little about z_j?
 %EM Even though it's a category now, it's still a function of tau and t, right?
-                             &= \max_{\fullAncestralStateCategories} \ \prod_{j=1}^{q} \ P(\sitePattern_j, \ancestralStateCategory_j | \tau, t)^{n_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
+                             &= \prod_{i=1}^{\nCols} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t) \\
+                             &= \prod_{j=1}^{\nSiteSplits} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t)^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
 \end{align}
-where $n_j(\fullAlignment)$ is the number of observations in the sample where site pattern $\sitePattern_j$ was seen in $\fullAlignment$.
-
-\textbf{Example:} This might be a good time for a toy example showing this for a very simple topology.
+where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ appears.
 
@@
 
-We are now interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
-Let
+We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
+Define \eqref{eq:site_pattern_likelihood} as
 $$
-L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t) = \prod_{j=1}^q \ P(\sitePattern_j, \ancestralStateCategory_j | \tau, t)^{n_j(\fullAlignment)}
+L_\nCols''(\fullAlignment;\fullAncestralStateCategories,\tau,t) = \prod_{j=1}^{\nSiteSplits} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t)^{\nCols_j(\fullAlignment)} .
 $$
-so that
-$$
-\max_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t) =
-    \max_{\fullAncestralStateCategories} \ L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t).
-$$
-Assume our $n$ observations were generated from a model with parameters $(\tau^*, t^*)$.
+Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
 We have
 \begin{align}
-    \frac{1}{n} \log L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t)
-        &= \sum_{j} \frac{n_j(\fullAlignment)}{n}\cdot  \log P(\sitePattern_j, \ancestralStateCategory_j | \tau, t) \\
-        &= \sum_{j} \frac{n_j(\fullAlignment)}{n}\cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)].
+    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralStateCategories,\tau,t)
+        &= \sum_{j}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t) \\
+        &= \sum_{j}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\alignmentColumn=\siteSplit_j | \tau, t) + 
+            \log P(\ancestralStateColumn\in\ancestralStateCategory_j | \alignmentColumn=\siteSplit_j , \tau, t)]
 \end{align}
-In the limit of infinite data we have
-$$
-\frac{1}{n} \log L_n''(\fullAlignment;,\tau,t) \rightarrow \sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)]
-$$
-and that, since $\ancestralStateCategory_j$ appears only in the last term, the log of \eqref{eq:site_pattern_likelihood} in the same limit is proportional to
-\begin{align}
-\frac{1}{n} \log L_n'(\fullAlignment;\fullAncestralStateCategories,\tau, t)
-    &= \max_{\fullAncestralStateCategories} \ \frac{1}{n} \log L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t) \nonumber \\
-    &\rightarrow \max_{\fullAncestralStateCategories} \ \sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)] \nonumber \\
-    &= \max_{\fullAncestralStateCategories} \ \sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)] \nonumber \\
-    &= \max_{\fullAncestralStateCategories} \ \left\{\sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot \left[\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)\right]\right\}.
+so that, in the limit of infinite data,
+\begin{equation}
+\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j}^\nSiteSplits P(\alignmentColumn=\siteSplit_j | \tau^*, t^*) \cdot [\log P(\alignmentColumn=\siteSplit_j | \tau, t) + \log P(\ancestralStateColumn\in\ancestralStateCategory_j | \alignmentColumn=\siteSplit_j , \tau, t)].
 \label{eq:site_pattern_profile_likelihood_mean}
-\end{align}
+\end{equation}
 Define the the information quantity
 $$
-H_{\tau^*,t^*}(\tau,t) = \sum_{j} P(\sitePattern_j | \tau^*, t^*) \log P(\sitePattern_j | \tau, t),
+h_{\tau^*,t^*}(\tau,t) = \sum_{j}^\nSiteSplits P(\alignmentColumn=\siteSplit_j | \tau^*, t^*) \log P(\siteSplit_j | \tau, t),
 $$
@@
 $$
-\ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = \sum_{j} P(\sitePattern_j | \tau^*, t^*) \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)
+\ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = \sum_{j}^\nSiteSplits P(\alignmentColumn=\siteSplit_j | \tau^*, t^*) \log P(\ancestralStateColumn\in\ancestralStateCategory_j | \siteSplit_j, \tau, t)
 $$
@@
     \label{eq:log_likelihood_simplified}
-\ell_{\tau^*,t^*}(\tau,t) = \max_{\fullAncestralStateCategories} \ \left\{H_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t)\right\}.
+\ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = h_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
 \end{equation}
@@
 \label{eq:inconsistency_inequality}
-\ell_{\tau^*,t^*}(\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\tau^*,\hat{t}_2)
+\ell_{\tau^*,t^*}(\hat{\fullAncestralStateCategories}_1,\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralStateCategories}_2,\tau^*,\hat{t}_2)
 \end{equation}
-for $\tau'\neq\tau^*$ with $\hat{t}_1$ and $\hat{t}_2$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
+for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralStateCategories}_1\}$ and $\{\hat{t}_2,\hat{\fullAncestralStateCategories}_2\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
 
@@
 \label{eq:partial-likelihood}
-\ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = H_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
+\ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = h_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
 \end{equation}
@@
 \max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \le
-    H_{\tau^*,t^*}(\tau^*,t^*)
+    h_{\tau^*,t^*}(\tau^*,t^*)
     + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t})
@@
 $$
-H_{\tau^*,t^*}(\tau,t) \le H_{\tau^*,t^*}(\tau^*,t^*)
+h_{\tau^*,t^*}(\tau,t) \le h_{\tau^*,t^*}(\tau^*,t^*)
 $$
@@
 \max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \ge
-    H_{\tau^*,t^*}(\tau,\hat{t})
+    h_{\tau^*,t^*}(\tau,\hat{t})
     + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t}).
@@
     \hline
-$\sitePattern_j$   &$P(\sitePattern_j|\tau_1,t^*)$&$P(\sitePattern_j|\tau_2,t^*)$\\
+$\siteSplit_j$   &$P(\alignmentColumn=\siteSplit_j|\tau_1,t^*)$&$P(\alignmentColumn=\siteSplit_j|\tau_2,t^*)$\\
     \hline
-0000&$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
-0001&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
-0010&$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
-0100&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
-1000&$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
-0011&$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
-0101&$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
-1001&$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
+    $\emptyset$&$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
+    $\{1\}$    &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
+    $\{2\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
+    $\{3\}$    &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
+    $\{1,2\}$  &$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
+    $\{1,3\}$  &$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
+    $\{2,3\}$  &$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
+    $\{1,2,3\}$&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
     \hline
@@
     \hline
-$\sitePattern_j$   &$P(\sitePattern_j|\tau_1,t)$&$P(\sitePattern_j|\tau_2,t)$\\
+$\siteSplit_j$   &$P(\alignmentColumn=\siteSplit_j|\tau_1,t)$&$P(\alignmentColumn=\siteSplit_j|\tau_2,t)$\\
     \hline
-0000&$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
-0001&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
-0010&$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
-0100&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
-1000&$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
-0011&$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
-0101&$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
-1001&$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
+    $\emptyset$&$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
+    $\{1\}$    &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
+    $\{2\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
+    $\{3\}$    &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
+    $\{1,2\}$  &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
+    $\{1,3\}$  &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
+    $\{2,3\}$  &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
+    $\{1,2,3\}$&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
     \hline
@@
 \begin{tabular}{|l|ll|}
-\multicolumn{3}{c}{$P(\ancestralStateCategory_j|\sitePattern_j,\tau_1,t)$}\\
+    \multicolumn{3}{c}{$\tau=\tau_1$}\\
     \hline
-& \multicolumn{2}{|c|}{$\ancestralStateCategory_j$}\\
+    $\siteSplit_j$    & $\ancestralStateCategory_j$ & $P(\ancestralStateColumn\in\ancestralStateCategory_j|\alignmentColumn=\siteSplit_j,\tau,t)$\\
     \hline
-$\sitePattern_j$    &00                              &01\\
+    $\emptyset$&$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)^2$\\
+     $\{1\}$    &$\{\emptyset\}$&$(1+x)(1-x)(1+w)(1+y)^2$\\
+     $\{2\}$    &$\{\emptyset\}$&$(1+x)(1-x)(1+w)(1+y)^2$\\
+     $\{3\}$    &$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)(1-y)$\\
+    $\{1,2\}$  &$\{\emptyset,\{1,2,3\}\}$&$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+    $\{1,3\}$  &$\{\emptyset,\{2\}, \{1,2\}\}$&
+    $\left\{\begin{array}{l}
+                    (1+x)^2   (1+w)(1-y)^2\\
+                    (1+x)^2   (1-w)(1+y)^2\\
+                    (1-x)^2   (1+w)(1+y)^2
+                \end{array}\right.$\\
+     $\{2,3\}$  &$\{\emptyset,\{1,2,3\}\}$&$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+    $\{1,2,3\}$&$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)(1-y)$\\
     \hline
-0000&$(1+x)^2   (1+w)(1+y)^{2*}$          &$(1+x)^2   (1-w)(1-y)^2$\\
-0001&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
-0010&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
-0100&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
-1000&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
-0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$   &$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
-1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+    \multicolumn{3}{c}{$\tau=\tau_2$}\\
     \hline
+    $\siteSplit_j$    & $\ancestralStateCategory_j$ & $P(\ancestralStateColumn\in\ancestralStateCategory_j|\alignmentColumn=\siteSplit_j,\tau,t)$\\
     \hline
-&10                           &11\\
-    \hline
-0000&$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
-0001&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-0010&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-0100&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
-1000&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
-0011&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
-0101&$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
-1001&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
-    \hline
-    \multicolumn{3}{c}{$P(\ancestralStateCategory_j|\sitePattern_j,\tau_2,t)$}\\
-\hline
-& \multicolumn{2}{|c|}{$\ancestralStateCategory_j$}\\
-    \hline
-$\sitePattern_j$    &00                              &01\\
-    \hline
-0000&$(1+x)^2   (1+w)(1+y)^{2*}$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-0001&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$\\
-0010&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$\\
-0100&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1-y)^2$\\
-1000&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1-x)^2   (1-w)(1+y)(1-y)$\\
-0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
-0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$    &$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$\\
-1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1-x)^2   (1-w)(1+y)^{2\ddagger}$\\
-    \hline
-    \hline
-&10                           &11\\
-    \hline
-0000&$(1+x)(1-x)(1-w)(1+y)(1-y)$             &$(1-x)^2   (1+w)(1-y)^2$\\
-0001&$(1+x)(1-x)(1-w)(1-y)^2$                &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-0010&$(1-x)^2   (1-w)(1+y)(1-y)$             &$(1+x)(1-x)(1+w)(1-y)^2$\\
-0100&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$      &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-1000&$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$  &$(1+x)(1-x)(1+w)(1-y)^2$\\
-0011&$(1-x)^2   (1-w)(1-y)^2$                &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
-0101&$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$  &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
-1001&$(1+x)^2   (1-w)(1-y)^{2\ddagger}$      &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
+ $\emptyset$       &$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)^2$\\
+    $\{1\}$          &$\{\emptyset,\{1\}\}$&
+    $\left\{\begin{array}{l}
+                        (1+x)(1-x)(1+w)(1+y)^2\\
+                        (1+x)^2   (1-w)(1+y)(1-y)
+                    \end{array}\right.$\\
+      $\{2\}$          &$\{\emptyset,\{1\}\}$&
+    $\left\{\begin{array}{l}
+                    (1+x)^2   (1+w)(1+y)(1-y)\\
+                    (1+x)(1-x)(1-w)(1+y)^2
+                    \end{array}\right.$\\
+      $\{3\}$          &$\{\emptyset,\{2\}\}$&
+    $\left\{\begin{array}{l}
+                    (1+x)(1-x)(1+w)(1+y)^2\\
+                    (1+x)^2   (1-w)(1+y)(1-y)
+                    \end{array}\right.$\\
+     $\{1,2\}$         &$\{\emptyset,\{2\},\{1,2\}\}$&
+    $\left\{\begin{array}{l}
+                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
+                    (1+x)^2   (1-w)(1+y)^2\\
+                    (1+x)(1-x)(1+w)(1+y)(1-y)
+                    \end{array}\right.$\\
+     $\{1,3\}$         &$\{\emptyset,\{1\},\{2\},\{1,2\}\}$&
+    $\left\{\begin{array}{l}
+                    (1+x)^2   (1+w)(1-y)^2\\
+                    (1+x)(1-x)(1-w)(1+y)(1-y)\\
+                    (1+x)(1-x)(1-w)(1+y)(1-y)\\
+                    (1-x)^2   (1+w)(1+y)^2
+                    \end{array}\right.$\\
+      $\{2,3\}$        &$\{\emptyset,\{1\},\{2\},\{1,2\}\}$&
+    $\left\{\begin{array}{l}
+                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
+                    (1-x)^2   (1-w)(1+y)^2\\
+                    (1+x)^2   (1-w)(1-y)^2\\
+                    (1+x)(1-x)(1+w)(1+y)(1-y)
+                    \end{array}\right.$\\
+     $\{1,2,3\}$       &$\{\emptyset,\{2\}\}$&
+    $\left\{\begin{array}{l}
+                    (1+x)^2   (1+w)(1+y)(1-y)\\
+                    (1+x)(1-x)(1-w)(1+y)^2
+                    \end{array}\right.$\\
     \hline
 \end{tabular}
-\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
-Maxima determined row-wise (i.e., by site pattern).
+\caption{Likelihood calculations for all site patterns and ancestral state categories.
 All values multiplied by $1/32$.
-Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
+Likelihoods with multiple entries will have maxima determined by unknown branch length parameters.}
 \label{tab:likelihoods}
@@
 Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
-To obtain an upper bound for the Farris partial likelihood---the terms in the top half of Table~\ref{tab:likelihoods}---we use the term $(1+x)^2(1+w)(1+y)^2$ for the $0101$ site pattern.
-Since all cases have a $(1+x)^2$ term except $\{0100, 1000, 0011, 1001\}$, and all cases except $\{0001, 0010, 0011, 1001\}$ have a $(1+y)^2$, we can add and subtract the $\log$ of these terms weighted by each of their generating probabilities.
+To obtain an upper bound for the Farris partial likelihood---the terms in the top half of Table~\ref{tab:likelihoods}---we use the term $(1+x)^2(1+w)(1+y)^2$ for the $\{1,3\}$ split.
+Since all cases have a $(1+x)^2$ term except $\{\{1\}, \{2\}, \{1,2\}, \{2,3\}\}$, and all cases except $\{\{3\}, \{1,2\}, \{2,3\}, \{1,2,3\}\}$ have a $(1+y)^2$, we can add and subtract the $\log$ of these terms weighted by each of their generating probabilities.
 In other words, define
 \begin{align*}
-    a_{xy} &= P(0100|\tau=\tau_1,t=\{x,y,y\}) + P(1000|\tau=\tau_1,t=\{x,y,y\}) \\
-           &\qquad + P(0011|\tau=\tau_1,t=\{x,y,y\}) + P(1001|\tau=\tau_1,t=\{x,y,y\})
+    a_{xy} &= P(\alignmentColumn=\{1\}|\tau=\tau_1,t=\{x,y,y\})
+    +        P(\alignmentColumn=\{2\}|\tau=\tau_1,t=\{x,y,y\}) \\
+    &\qquad + P(\alignmentColumn=\{1,2\}|\tau=\tau_1,t=\{x,y,y\})
+    + P(\alignmentColumn=\{2,3\}|\tau=\tau_1,t=\{x,y,y\})
 \end{align*}
@@
 \begin{align*}
-    b_{xy} &= P(1000|\tau=\tau_1,t=\{x,y,y\}) + P(0010|\tau=\tau_1,t=\{x,y,y\}) \\
-           &\qquad + P(0011|\tau=\tau_1,t=\{x,y,y\}) + P(1001|\tau=\tau_1,t=\{x,y,y\}).
+    b_{xy} &= P(\alignmentColumn=\{3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{1,2\}|\tau=\tau_1,t=\{x,y,y\}) \\
+           &\qquad + P(\alignmentColumn=\{2,3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{1,2,3\}|\tau=\tau_1,t=\{x,y,y\}).
 \end{align*}
@@
 \begin{align*}
-    c_{xy} &= P(0001|\tau=\tau_1,t=\{x,y,y\}) + P(0010|\tau=\tau_1,t=\{x,y,y\}) \\
-           &\qquad + P(0100|\tau=\tau_1,t=\{x,y,y\}) + P(1000|\tau=\tau_1,t=\{x,y,y\})
+    c_{xy} &= P(\alignmentColumn=\{1\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{2\}|\tau=\tau_1,t=\{x,y,y\}) \\
+           &\qquad + P(\alignmentColumn=\{3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{1,2,3\}|\tau=\tau_1,t=\{x,y,y\})
 \end{align*}
@@
 $$
-    d_{xy} = P(0101|\tau=\tau_1,t=\{x,y,y\}) + P(1001|\tau=\tau_1,t=\{x,y,y\}).
+d_{xy} = P(\alignmentColumn=\{1,3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{2,3\}|\tau=\tau_1,t=\{x,y,y\}).
 $$
@@
 $$
-H_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-c_{xy}-2d_{xy}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(c_{xy},d_{xy}) \ge H_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p}_{\tau_1,\tau_1}(a_{xy},b_{xy}).
+h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-c_{xy}-2d_{xy}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(c_{xy},d_{xy}) \ge h_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p}_{\tau_1,\tau_1}(a_{xy},b_{xy}).
 $$
@@
 \begin{align*}
-    0 &\ge H_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p}_{\tau_1,\tau_1}(a_{xy},b_{xy}) \\
-      &\qquad - H_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-c_{xy}-2d_{xy}, 0\}) - L^{p,1}_{\tau_1,\tau_2}(c_{xy},d_{xy}) \\
+    0 &\ge h_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p}_{\tau_1,\tau_1}(a_{xy},b_{xy}) \\
+      &\qquad - h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-c_{xy}-2d_{xy}, 0\}) - L^{p,1}_{\tau_1,\tau_2}(c_{xy},d_{xy}) \\
 %
@@
 $$
-H_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\})  \approx -2.04470918734314.
+h_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\})  \approx -2.04470918734314.
 $$
@@
 $$
-H_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-c_{xy}-2d_{xy}, 0\}) \approx -2.07748833719921.
+h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-c_{xy}-2d_{xy}, 0\}) \approx -2.07748833719921.
 $$

### 2017-03-29 joint_inf.tex
Tweaks and notes.

@@
 
-For a topology $\tau$ and branch lengths $t$ we form the fully-observed likelihood (making the usual independence assumption between sites):
+For a topology $\tau$ and branch lengths $t$ we start with the fully-observed likelihood (making the usual independence assumption between sites):
 \begin{equation}
@@
 % Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
-The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves computing the marginal likelihood
+The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across ancestral states
 \begin{equation}
@@
 Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
-Following Section 8.6 of \cite{Semple2003}, $\tau$ induces an ordered set of tip labels $\{1,\ldots,\nSiteRows\}$ across sites.
+Following Section 8.6 of \cite{Semple2003-em}, $\tau$ induces an ordered set of tip labels $\{1,\ldots,\nSiteRows\}$ across sites.
+%EM What are complementary characters?
 For a binary character model, a realization of $\alignmentColumn_i$ can be described as a subset of complementary characters $\siteSplit\subseteq\sitePatternSet\triangleq\{1,\ldots,\nSiteRows-1\}$.
+%EM Why nSiteRows-1 in the def of siteSplit above, but nAncestralStateRows below in the def of ancestralSplit?
+%EM There are two ways to do this, if you are collapsing such that 001 and 110 are the same or not, and we should be clear which one you are doing.
 The topology $\tau$ induces an ordered set $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ with a similar subset of complementary characters $\ancestralSplit\subseteq\ancestralStateSet\triangleq\{1,\ldots,\nAncestralStateRows\}$ describing a realization of $\ancestralStateColumn_i$.
@@
 
-%EM in general aren't a different number of internal site patterns than tip site patterns?
-%das: Yes, I moved the paragraph at the end to here so it's hopefully less confusing.
-Maybe writing out the example would help?
-I also changed from "pattern" to "category" which might make it a little more clear.
-%EM Seems good this is an important concept so it's worth defining it more properly.
+%das: Maybe writing out the example would help?
+%EM: Yes, I think so. e.g. the need for partitions \ancestralStateCategory may not be obvious right away.
 
 Consider the probability $P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t)$ from \eqref{eq:full_likelihood} of observing a particular column $i$.
+%EM Indexing \siteSplit_j$ not defined.
+%EM But, more importantly, as we discussed Z_j is defined only with respect to a single tree and branch lengths.
+This need to be made abundantly clear.
+Personally, I think that using the same index j for them is confusing.
 Given $(\tau, t)$ and maximizing over $\ancestralStateColumn_i$, if $\alignmentColumn_i=\siteSplit_j$ then there exists a partition $\ancestralStateCategory_j$ of $\mathcal{P}(\ancestralStateSet)$ of most likely ancestral states where

### 2017-03-29 deprecated.tex
updates
reworking some bounds/notation

@@
 \section{Temporarily unused text I may want later to show more general results on the likelihood estimates}
+
+\begin{align*}
+    0 &\ge h_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p}_{\tau_1,\tau_1}(a_{xy},b_{xy}) \\
+      &\qquad - h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-d_{xy}-2e_{xy}, 0\}) - L^{p,1}_{\tau_1,\tau_2}(d_{xy},e_{xy}) \\
+%
+      &\qquad = 2(1+x^2-y^2-x^2y^2)\log(1+x^2-y^2-x^2y^2) \\
+      &\qquad + 2(1-x^2+y^2-x^2y^2)\log(1-x^2+y^2-x^2y^2) \\
+      &\qquad + 2(1-x^2-y^2+x^2y^2)\log(1-x^2-y^2+x^2y^2) \\
+      &\qquad + (1+x^2+y^2+4xy^2+x^2y^2)\log(1+x^2+y^2+4xy^2+x^2y^2) \\
+      &\qquad + (1+x^2+y^2-4xy^2+x^2y^2)\log(1+x^2+y^2-4xy^2+x^2y^2) \\
+%
+      &\qquad + (2-a_{xy})\log(2-a_{xy})+a_{xy}\log(a_{xy})+(2-b_{xy})\log(2-b_{xy})+b_{xy}\log(b_{xy})+\log(2) \\
+%
+      &\qquad - 2(1+x^2-y^2-x^2y^2)\log(1+x'^2w'-y'^2w'-x'^2y'^2) \\
+      &\qquad - 2(1-x^2+y^2-x^2y^2)\log(1-x'^2w'+y'^2w'-x'^2y'^2) \\
+      &\qquad - (1-x^2-y^2+x^2y^2)\log(1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2) \\
+      &\qquad - (1-x^2-y^2+x^2y^2)\log(1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2) \\
+      &\qquad - (1+x^2+y^2+4xy^2+x^2y^2)\log(1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2) \\
+      &\qquad - (1+x^2+y^2-4xy^2+x^2y^2)\log(1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2) \\
+%
+      &\qquad - 2\log(2)-(2-d_{xy}-2e_{xy})\log(2-d_{xy}-2e_{xy})-(d_{xy}+2e_{xy})\log(d_{xy}+2e_{xy})
+\end{align*}
+where
+$$
+a_{xy} = b_{xy} = \frac{4-2x^2-2y^2}{8},
+$$
+$$
+d_{xy} = \frac{4-4x^2y^2}{8},
+$$
+$$
+e_{xy} = \frac{2-4xy^2+2x^2y^2}{8}
+$$
+and
+$$
+x' = 1, \ y' = 1-d_{xy}-2e_{xy}, \ w' = 0.
+$$
+
+The \texttt{sage} function in listing~\ref{list:sage} will generate Fig.~\ref{fig:inconsistency-farris}.
+
+\begin{table}
+    \begin{center}
+\begin{verbatim}
+var('x,y')
+a = (4-2*x^2-2*y^2)/8
+b = (4-2*x^2-2*y^2)/8
+c = (4-4*x^2*y^2)/8
+d = (2-4*x*y^2+2*x^2*y^2)/8
+xp = 1.
+yp = 1-c-2*d
+wp = 0.
+
+inconsistency = 2*(1+x^2-y^2-x^2*y^2)*log(1+x^2-y^2-x^2*y^2) \
+    + 2*(1-x^2+y^2-x^2*y^2)*log(1-x^2+y^2-x^2*y^2) \
+    + 2*(1-x^2-y^2+x^2*y^2)*log(1-x^2-y^2+x^2*y^2) \
+    + (1+x^2+y^2+4*x*y^2+x^2*y^2)*log(1+x^2+y^2+4*x*y^2+x^2*y^2) \
+    + (1+x^2+y^2-4*x*y^2+x^2*y^2)*log(1+x^2+y^2-4*x*y^2+x^2*y^2) \
+    + (2-a)*log(2-a)+a*log(a)+(2-b)*log(2-b)+b*log(b)+log(2.) \
+    - 2*(1+x^2-y^2-x^2*y^2)*log(1+xp^2*wp-yp^2*wp-xp^2*yp^2) \
+    - 2*(1-x^2+y^2-x^2*y^2)*log(1-xp^2*wp+yp^2*wp-xp^2*yp^2) \
+    - (1-x^2-y^2+x^2*y^2)*log(1+2*xp*yp-2*xp*yp*wp-xp^2*wp-yp^2*wp+xp^2*yp^2) \
+    - (1-x^2-y^2+x^2*y^2)*log(1-2*xp*yp+2*xp*yp*wp-xp^2*wp-yp^2*wp+xp^2*yp^2) \
+    - (1+x^2+y^2+4*x*y^2+x^2*y^2)*log(1+2*xp*yp+2*xp*yp*wp+xp^2*wp+yp^2*wp+xp^2*yp^2) \
+    - (1+x^2+y^2-4*x*y^2+x^2*y^2)*log(1-2*xp*yp-2*xp*yp*wp+xp^2*wp+yp^2*wp+xp^2*yp^2) \
+    - 2*log(2.)-(2-c-2*d)*log(2-c-2*d)-(c+2*d)*log(c+2*d)
+
+incons_plot = implicit_plot(
+    inconsistency(x,y)==0,
+    (x,0,1),
+    (y,0,1),
+    axes_labels=['$x$', '$y$'],
+    title='Region of inconsistency for Farris-generating topology',
+    color='black'
+)
+\end{verbatim}
+    \end{center}
+\caption{Sage code to generate Fig.~\ref{fig:inconsistency-farris}}
+\label{list:sage}
+\end{table}
+
+To obtain an upper bound for the Farris partial likelihood---the terms in the top half of Table~\ref{tab:likelihoods}---we use the term $(1+x)^2(1+w)(1+y)^2$ for the $\{1,3\}$ split.
+Since all cases have a $(1+x)^2$ term except $\{\{1\}, \{2\}, \{1,2\}, \{2,3\}\}$, and all cases except $\{\{3\}, \{1,2\}, \{2,3\}, \{1,2,3\}\}$ have a $(1+y)^2$, we can add and subtract the $\log$ of these terms weighted by each of their generating probabilities.
+In other words, define
+\begin{align*}
+    a_{xy} &= P(\alignmentColumn=\{1\}|\tau=\tau_1,t=\{x,y,y\})
+    +        P(\alignmentColumn=\{2\}|\tau=\tau_1,t=\{x,y,y\}) \\
+    &\qquad + P(\alignmentColumn=\{1,2\}|\tau=\tau_1,t=\{x,y,y\})
+    + P(\alignmentColumn=\{2,3\}|\tau=\tau_1,t=\{x,y,y\}),
+\end{align*}
+\begin{align*}
+    b_{xy} &= P(\alignmentColumn=\{3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{1,2\}|\tau=\tau_1,t=\{x,y,y\}) \\
+           &\qquad + P(\alignmentColumn=\{2,3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{1,2,3\}|\tau=\tau_1,t=\{x,y,y\})
+\end{align*}
+and
+\begin{equation*}
+    c_{xy} = P(\alignmentColumn=\{1,3\}|\tau=\tau_1,t=\{x,y,y\}).
+\end{equation*}
+Then the partial likelihood is bounded above by
+\begin{align*}
+    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,\{x',y',w'\}) \\
+    &\qquad \le (2-a_{xy})\log(1+x')+a_{xy}\log(1-x')\\
+    &\qquad \qquad +(2-b_{xy})\log(1+y')+b_{xy}\log(1-y')\\
+    &\qquad \qquad +(1-c_{xy})\log(1+w')+c_{xy}\log(1-w') \\
+    &\qquad \le (2-a_{xy})\log(2-a_{xy})+a_{xy}\log(a_{xy})\\
+    &\qquad \qquad +(2-b_{xy})\log(2-b_{xy})+b_{xy}\log(b_{xy})\\
+    &\qquad \qquad +(1-c_{xy})\log(1-c_{xy})+c_{xy}\log(c_{xy})+\log(2) \\
+    &\qquad \triangleq L^{p,1}_{\tau_1,\tau_1}(a_{xy}, b_{xy})
+\end{align*}
+or
+\begin{align*}
+    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,\{x',y',w'\}) \\
+    &\qquad \le (2-a_{xy}-2c_{xy})\log(1+x')+(a_{xy}+2c_{xy})\log(1-x')\\
+    &\qquad \qquad +(2-b_{xy})\log(1+y')+b_{xy}\log(1-y')+\log(1+w')\\
+    &\qquad \le (2-a_{xy}-2c_{xy})\log(2-a_{xy}-2c_{xy})+(a_{xy}+2c_{xy})\log(a_{xy}+2c_{xy})\\
+    &\qquad \qquad +(2-b_{xy})\log(2-b_{xy})+b_{xy}\log(b_{xy})+\log(2) \\
+    &\qquad \triangleq L^{p,2}_{\tau_1,\tau_1}(a_{xy}, b_{xy})
+\end{align*}
+where $a_{xy},b_{xy}$ and $c_{xy}$ are functions of the true generating branch parameters $\{x,y\}$.
+
+\begin{table}
+\centering
+\begin{tabular}{|l|ll|}
+\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k|\alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
+\hline
+& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
+    \hline
+    $\siteSplit_j$    &$\emptyset$  00                              &$\{2\}$  01\\
+    \hline
+     $\emptyset$   0000&$(1+x)^2   (1+w)(1+y)^{2*}$          &$(1+x)^2   (1-w)(1-y)^2$\\
+     $\{1\}$       1000&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
+     $\{2\}$       0100&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
+     $\{3\}$       0010&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
+     $\{1,2\}$     0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+     $\{1,3\}$     0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$   &$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
+     $\{2,3\}$     1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+     $\{1,2,3\}$   0001&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
+    \hline
+    \hline
+    &$\{1\}$  10                           &$\{1,2\}$  11\\
+    \hline
+     $\emptyset$   0000&$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
+     $\{1\}$       1000&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{2\}$       0100&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{3\}$       0010&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+     $\{1,2\}$     0011&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
+     $\{1,3\}$     0101&$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
+     $\{2,3\}$     1001&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
+     $\{1,2,3\}$   0001&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+    \hline
+\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k|\alignmentColumn=\siteSplit_j,\tau_2,t)$}\\
+\hline
+& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
+    \hline
+    $\siteSplit_j$    &$\emptyset$  00                              &$\{2\}$  01\\
+    \hline
+     $\emptyset$   0000&$(1+x)^2   (1+w)(1+y)^{2*}$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+     $\{1\}$       1000&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1-x)^2   (1-w)(1+y)(1-y)$\\
+     $\{2\}$       0100&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1-y)^2$\\
+     $\{3\}$       0010&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$\\
+     $\{1,2\}$     0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
+     $\{1,3\}$     0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$    &$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$\\
+     $\{2,3\}$     1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1-x)^2   (1-w)(1+y)^{2\ddagger}$\\
+     $\{1,2,3\}$   0001&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$\\
+    \hline
+    \hline
+    &$\{1\}$  10                           &$\{1,2\}$  11\\
+    \hline
+     $\emptyset$   0000&$(1+x)(1-x)(1-w)(1+y)(1-y)$             &$(1-x)^2   (1+w)(1-y)^2$\\
+     $\{1\}$       1000&$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$  &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{2\}$       0100&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$      &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+     $\{3\}$       0010&$(1-x)^2   (1-w)(1+y)(1-y)$             &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{1,2\}$     0011&$(1-x)^2   (1-w)(1-y)^2$                &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
+     $\{1,3\}$     0101&$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$  &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
+     $\{2,3\}$     1001&$(1+x)^2   (1-w)(1-y)^{2\ddagger}$      &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
+     $\{1,2,3\}$   0001&$(1+x)(1-x)(1-w)(1-y)^2$                &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+\hline
+\end{tabular}
+\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
+Maxima determined row-wise (i.e., by site pattern).
+All values multiplied by $1/32$.
+Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
+\label{tab:likelihoods}
+\end{table}
 

### 2017-03-29 joint_inf.tex
reworking some bounds/notation

--- 2017-03-28 joint_inf.tex
@@
 $$
-\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \le
+\max_{t,\fullAncestralStateCategories} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \le
     h_{\tau^*,t^*}(\tau^*,t^*)
@@
 $$
-\{\hat{\fullAncestralStateCategories},\hat{t}\} = \arg\max_{\{\fullAncestralStateCategories,t\}} \ \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
+\{\hat{t},\hat{\fullAncestralStateCategories}\} = \arg\max_{t,\fullAncestralStateCategories} \ \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
 $$
@@
 $$
-\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \ge
+\max_{t,\fullAncestralStateCategories} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \ge
     h_{\tau^*,t^*}(\tau,\hat{t})
@@
     \hline
-    $\emptyset$&$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
+    $\{\emptyset\}$&$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
     $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
@@
     $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
-    $\{1,2,3\}$&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
+    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
     \hline
@@
     \hline
-    $\emptyset$&$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
+    $\{\emptyset\}$&$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
     $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
@@
     $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
-    $\{1,2,3\}$&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
+    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
     \hline
@@
     \hline
-    $\emptyset$&$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)^2$\\
-     $\{1\}$    &$\{\emptyset\}$&$(1+x)(1-x)(1+w)(1+y)^2$\\
-     $\{2\}$    &$\{\emptyset\}$&$(1+x)(1-x)(1+w)(1+y)^2$\\
-     $\{3\}$    &$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)(1-y)$\\
-    $\{1,2\}$  &$\{\emptyset,\{1,2,3\}\}$&$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-    $\{1,3\}$  &$\{\emptyset,\{2\}, \{1,2\}\}$&
+    $\{\emptyset\}$&
+    $\{\emptyset\}$&
+    $(1+x)^2   (1+w)(1+y)^2$\\
+     $\{1\}$    &
+    $\{\emptyset\}$&
+    $(1+x)(1-x)(1+w)(1+y)^2$\\
+     $\{2\}$    &
+    $\{\emptyset\}$&
+    $(1+x)(1-x)(1+w)(1+y)^2$\\
+     $\{3\}$    &
+    $\{\emptyset\}$&
+    $(1+x)^2   (1+w)(1+y)(1-y)$\\
+    $\{1,2\}$  &
+    $\{\{\emptyset\},\{1,2,3\}\}$&
+    $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+    $\{1,3\}$  &
     $\left\{\begin{array}{l}
+                    \{\emptyset\}\\
+                    \{2\}\\
+                    \{1,2\}
+                \end{array}\right.$&
+    $\begin{array}{l}
                     (1+x)^2   (1+w)(1-y)^2\\
@@
                     (1-x)^2   (1+w)(1+y)^2
-                \end{array}\right.$\\
-     $\{2,3\}$  &$\{\emptyset,\{1,2,3\}\}$&$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-    $\{1,2,3\}$&$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)(1-y)$\\
+                \end{array}$\\
+    $\{2,3\}$  &
+                $\{\{\emptyset\},\{1,2,3\}\}$&
+                $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+    $\{1,2,3\}$&
+                $\{\emptyset\}$&
+                $(1+x)^2   (1+w)(1+y)(1-y)$\\
     \hline
@@
     \hline
- $\emptyset$       &$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)^2$\\
-    $\{1\}$          &$\{\emptyset,\{1\}\}$&
+    $\{\emptyset\}$       &$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)^2$\\
+    $\{1\}$          &
     $\left\{\begin{array}{l}
+                    \{\emptyset\}\\
+                    \{1\}
+                \end{array}\right.$&
+    $\begin{array}{l}
                         (1+x)(1-x)(1+w)(1+y)^2\\
                         (1+x)^2   (1-w)(1+y)(1-y)
-                    \end{array}\right.$\\
-      $\{2\}$          &$\{\emptyset,\{1\}\}$&
+                    \end{array}$\\
+      $\{2\}$          &
     $\left\{\begin{array}{l}
+                    \{\emptyset\}\\
+                    \{1\}
+                \end{array}\right.$&
+    $\begin{array}{l}
                     (1+x)^2   (1+w)(1+y)(1-y)\\
                     (1+x)(1-x)(1-w)(1+y)^2
-                    \end{array}\right.$\\
-      $\{3\}$          &$\{\emptyset,\{2\}\}$&
+                    \end{array}$\\
+      $\{3\}$          &
     $\left\{\begin{array}{l}
+                    \{\emptyset\}\\
+                    \{2\}
+                \end{array}\right.$&
+    $\begin{array}{l}
                     (1+x)(1-x)(1+w)(1+y)^2\\
                     (1+x)^2   (1-w)(1+y)(1-y)
-                    \end{array}\right.$\\
-     $\{1,2\}$         &$\{\emptyset,\{2\},\{1,2\}\}$&
+                    \end{array}$\\
+     $\{1,2\}$         &
     $\left\{\begin{array}{l}
+                    \{\{\emptyset\},\{1,2\}\}\\
+                    \{2\}
+                \end{array}\right.$&
+    $\begin{array}{l}
                     (1+x)(1-x)(1+w)(1+y)(1-y)\\
-                    (1+x)^2   (1-w)(1+y)^2\\
-                    (1+x)(1-x)(1+w)(1+y)(1-y)
-                    \end{array}\right.$\\
-     $\{1,3\}$         &$\{\emptyset,\{1\},\{2\},\{1,2\}\}$&
+                    (1+x)^2   (1-w)(1+y)^2
+                    \end{array}$\\
+     $\{1,3\}$         &
     $\left\{\begin{array}{l}
+                    \{\emptyset\}\\
+                    \{\{1\},\{2\}\}\\
+                    \{1,2\}
+                \end{array}\right.$&
+    $\begin{array}{l}
                     (1+x)^2   (1+w)(1-y)^2\\
                     (1+x)(1-x)(1-w)(1+y)(1-y)\\
-                    (1+x)(1-x)(1-w)(1+y)(1-y)\\
                     (1-x)^2   (1+w)(1+y)^2
-                    \end{array}\right.$\\
-      $\{2,3\}$        &$\{\emptyset,\{1\},\{2\},\{1,2\}\}$&
+                    \end{array}$\\
+      $\{2,3\}$        &
     $\left\{\begin{array}{l}
+                    \{\{\emptyset\},\{1,2\}\}\\
+                    \{1\}\\
+                    \{2\}
+                \end{array}\right.$&
+    $\begin{array}{l}
                     (1+x)(1-x)(1+w)(1+y)(1-y)\\
                     (1-x)^2   (1-w)(1+y)^2\\
-                    (1+x)^2   (1-w)(1-y)^2\\
-                    (1+x)(1-x)(1+w)(1+y)(1-y)
-                    \end{array}\right.$\\
-     $\{1,2,3\}$       &$\{\emptyset,\{2\}\}$&
+                    (1+x)^2   (1-w)(1-y)^2
+                    \end{array}$\\
+     $\{1,2,3\}$       &
     $\left\{\begin{array}{l}
+                    \{\emptyset\}\\
+                    \{2\}
+                \end{array}\right.$&
+    $\begin{array}{l}
                     (1+x)^2   (1+w)(1+y)(1-y)\\
                     (1+x)(1-x)(1-w)(1+y)^2
-                    \end{array}\right.$\\
+                    \end{array}$\\
     \hline
@@
 Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
-To obtain an upper bound for the Farris partial likelihood---the terms in the top half of Table~\ref{tab:likelihoods}---we use the term $(1+x)^2(1+w)(1+y)^2$ for the $\{1,3\}$ split.
-Since all cases have a $(1+x)^2$ term except $\{\{1\}, \{2\}, \{1,2\}, \{2,3\}\}$, and all cases except $\{\{3\}, \{1,2\}, \{2,3\}, \{1,2,3\}\}$ have a $(1+y)^2$, we can add and subtract the $\log$ of these terms weighted by each of their generating probabilities.
-In other words, define
-\begin{align*}
-    a_{xy} &= P(\alignmentColumn=\{1\}|\tau=\tau_1,t=\{x,y,y\})
-    +        P(\alignmentColumn=\{2\}|\tau=\tau_1,t=\{x,y,y\}) \\
-    &\qquad + P(\alignmentColumn=\{1,2\}|\tau=\tau_1,t=\{x,y,y\})
-    + P(\alignmentColumn=\{2,3\}|\tau=\tau_1,t=\{x,y,y\})
-\end{align*}
-and
-\begin{align*}
-    b_{xy} &= P(\alignmentColumn=\{3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{1,2\}|\tau=\tau_1,t=\{x,y,y\}) \\
-           &\qquad + P(\alignmentColumn=\{2,3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{1,2,3\}|\tau=\tau_1,t=\{x,y,y\}).
-\end{align*}
-Then the ancestral state maximizing the likelihood is $00$ for each site pattern and the partial likelihood is bounded above
+Define, e.g., $p_0 = P(\alignmentColumn=\{\emptyset\}|\tau=\tau_1,t=\{x,y,y\})$.
+Let
+$$
+    a_{0} = p_{13},
+$$
+$$
+    a_{1} = p_1+p_2+p_{12}+p_{23},
+$$
+$$
+    a_{2} = p_3+p_{12}+p_{23}+p_{123},
+$$
+$$
+    a_{3} = p_1+p_3+p_{12}+p_{23},
+$$
+$$
+    a_{4} = p_2+p_{12}+p_{23}+p_{123}.
+$$
+Then the full likelihood is bounded above by one of three functions:
 \begin{align*}
     & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,\{x',y',w'\}) \\
-    &\qquad \le (2-a_{xy})\log(1+x')+a_{xy}\log(1-x')+(2-b_{xy})\log(1+y')+b_{xy}\log(1-y')+\log(1+w') \\
-    &\qquad \le (2-a_{xy})\log(2-a_{xy})+a_{xy}\log(a_{xy})+(2-b_{xy})\log(2-b_{xy})+b_{xy}\log(b_{xy})+\log(2) \\
-    &\qquad \triangleq L^p_{\tau_1,\tau_1}(a_{xy}, b_{xy})
+    &\qquad \le (3-a_{1})\log(1+x')+(a_{1}+a_{3})\log(1-x')\\
+    &\qquad \qquad +(3-a_{2})\log(1+y')+(a_{2}+a_{4})\log(1-y')\\
+    &\qquad \qquad +(1-a_{0})\log(1+w')+a_{0}\log(1-w')+p_{0}\log(4)\\
+    &\qquad \le (3-a_{1})\log\frac{2(3-a_{1})}{3+a_{3}}+(a_{1}+a_{3})\log\frac{2(a_{1}+a_{3})}{3+a_{3}}\\
+    &\qquad \qquad +(3-a_{2})\log\frac{2(3-a_{2})}{3+a_{4}}+(a_{2}+a_{4})\log\frac{2(a_{2}+a_{4})}{3+a_{4}}\\
+    &\qquad \qquad +(1-a_{0})\log(1+w')+a_{0}\log(1-w')+p_{0}\log(4)\\
+    &\qquad \triangleq L^{p,1}_{\tau_1,\tau_1}(\mathbf{a}_{xy});
 \end{align*}
-where $a_{xy}$ and $b_{xy}$ are functions of the true generating branch parameters $\{x,y\}$.
+\begin{align*}
+    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,\{x',y',w'\}) \\
+    &\qquad \le (3-a_{1}-2a_{0})\log(1+x')+(a_{1}+a_{3}+2a_{0})\log(1-x')\\
+    &\qquad \qquad +(3-a_{2})\log(1+y')+(a_{2}+a_{4})\log(1-y')+\log(1+w')+p_{0}\log(4)\\
+    &\qquad \le (3-a_{1}-2a_{0})\log\frac{2(3-a_{1}-2a_{0})}{3+a_{3}}+(a_{1}+a_{3}+2a_{0})\log\frac{2(a_{1}+a_{3}+2a_{0})}{3+a_{3}}\\
+    &\qquad \qquad +(3-a_{2})\log\frac{2(3-a_{2})}{3+a_{4}}+(a_{2}+a_{4})\log\frac{2(a_{2}+a_{4})}{3+a_{4}}\\
+    &\qquad \qquad +\log(2)+p_{0}\log(4)\\
+    &\qquad \triangleq L^{p,2}_{\tau_1,\tau_1}(\mathbf{a}_{xy});
+\end{align*}
+\begin{align*}
+    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,\{x',y',w'\}) \\
+    &\qquad \le (3-a_{1})\log(1+x')+(a_{1}+a_{3})\log(1-x')\\
+    &\qquad \qquad +(3-a_{2}-2a_{0})\log(1+y')+(a_{2}+a_{4})+2a_{0}\log(1-y')+\log(1+w')+p_{0}\log(4)\\
+    &\qquad \le (3-a_{1})\log\frac{2(3-a_{1})}{3+a_{3}}+(a_{1}+a_{3})\log\frac{2(a_{1}+a_{3})}{3+a_{3}}\\
+    &\qquad \qquad +(3-a_{2}-2a_{0})\log\frac{2(3-a_{2}-2a_{0})}{3+a_{4}}+(a_{2}+a_{4}+2a_{0})\log\frac{2(a_{2}+a_{4}+2a_{0})}{3+a_{4}}\\
+    &\qquad \qquad +\log(2)+p_{0}\log(4)\\
+    &\qquad \triangleq L^{p,3}_{\tau_1,\tau_1}(\mathbf{a}_{xy}).
+\end{align*}
 
@@
 Let
-\begin{align*}
-    c_{xy} &= P(\alignmentColumn=\{1\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{2\}|\tau=\tau_1,t=\{x,y,y\}) \\
-           &\qquad + P(\alignmentColumn=\{3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{1,2,3\}|\tau=\tau_1,t=\{x,y,y\})
-\end{align*}
+$$
+b_1 = p_1+p_2+p_3+p_{123}
+$$
 and
 $$
-d_{xy} = P(\alignmentColumn=\{1,3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{2,3\}|\tau=\tau_1,t=\{x,y,y\}).
+b_2 = p_{13}+p_{23}.
 $$
-The lower bounds on the likelihoods in this case take one of six forms:
+
+\textbf{TODO}: ancestral states where these likelihoods occur
 Since the maxima for pairs of lower bounds above each have a symmetry in $x$ and $y$, they are bounded below by one of
-\textbf{TODO}: ancestral states where these likelihoods occur
-
 \begin{align*}
     & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,\{x',y',w'\}) \\
-    &\qquad \ge 2\log(1+x')+(2-c_{xy}-2d_{xy})\log(1+y')+(c_{xy}+2d_{xy})\log(1-y')+\log(1-w') \\
-    &\qquad \ge 2\log(2)+(2-c_{xy}-2d_{xy})\log(2-c_{xy}-2d_{xy})+(c_{xy}+2d_{xy})\log(c_{xy}+2d_{xy}) \\
-    &\qquad \triangleq L^{p,1}_{\tau_1,\tau_2}(c_{xy}, d_{xy}),
+    &\qquad \ge 2\log(1+x')+(2-b_1-2b_2)\log(1+y')+(b_1+2b_2)\log(1-y')+\log(1-w'),
 \end{align*}
@@
     & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,\{x',y',w'\}) \\
-    &\qquad \ge (2-2d_{xy})\log(1+x')+2d_{xy}\log(1-x') + (2-c_{xy})\log(1+y')+c_{xy}\log(1-y')+\log(1-w') \\
-    &\qquad \ge (2-2d_{xy})\log(2-2d_{xy})+2d_{xy}\log(2d_{xy}) + (2-c_{xy})\log(2-c_{xy})+c_{xy}\log(c_{xy}) \\
-    &\qquad \triangleq L^{p,2}_{\tau_1,\tau_2}(c_{xy}, d_{xy})
+    &\qquad \ge (2-2b_2)\log(1+x')+2b_2\log(1-x') + (2-b_1)\log(1+y')+b_1\log(1-y')+\log(1-w')
 \end{align*}
@@
     & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,\{x',y',w'\}) \\
-    &\qquad \ge (2-d_{xy})\log(1+x')+d_{xy}\log(1-x') + (2-c_{xy}-d_{xy})\log(1+y')+(c_{xy}+d_{xy})\log(1-y')+\log(1-w') \\
-    &\qquad \ge (2-d_{xy})\log(2-d_{xy})+d_{xy}\log(d_{xy}) + (2-c_{xy}-d_{xy})\log(2-c_{xy}-d_{xy})+(c_{xy}+d_{xy})\log(c_{xy}+d_{xy}) \\
-    &\qquad \triangleq L^{p,3}_{\tau_1,\tau_2}(c_{xy}, d_{xy})
+    &\qquad \ge (2-b_2)\log(1+x')+b_2\log(1-x') + (2-b_1-b_2)\log(1+y')+(b_1+b_2)\log(1-y')+\log(1-w')
 \end{align*}
-where $c_{xy}$ and $d_{xy}$ are also functions of the true generating branch parameters $\{x,y\}$.
+where $b_1$ and $b_2$ are also functions of the true generating branch parameters $\{x,y\}$.
+Taking the maximum of each yields
+$$
+    L^{p,1}_{\tau_1,\tau_2}(b_1, b_2) \triangleq 2\log(2)+(2-b_1-2b_2)\log(2-b_1-2b_2)+(b_1+2b_2)\log(b_1+2b_2),
+$$
+$$
+    L^{p,2}_{\tau_1,\tau_2}(b_1, b_2) \triangleq (2-2b_2)\log(2-2b_2)+2b_2\log(2b_2) + (2-b_1)\log(2-b_1)+b_1\log(b_1)
+$$
+and
+$$
+    L^{p,3}_{\tau_1,\tau_2}(b_1, b_2) \triangleq (2-b_2)\log(2-b_2)+b_2\log(b_2) + (2-b_1-b_2)\log(2-b_1-b_2)+(b_1+b_2)\log(b_1+b_2).
+$$
 
@@
 $$
-L^{p,1}_{\tau_1,\tau_2}(c_{xy}, d_{xy}) \ge L^{p,2}_{\tau_1,\tau_2}(c_{xy}, d_{xy})
+L^{p,1}_{\tau_1,\tau_2}(b_1, b_2) \ge L^{p,2}_{\tau_1,\tau_2}(b_1, b_2)
 $$
@@
 $$
-L^{p,1}_{\tau_1,\tau_2}(c_{xy}, d_{xy}) \ge L^{p,3}_{\tau_1,\tau_2}(c_{xy}, d_{xy}).
+L^{p,1}_{\tau_1,\tau_2}(b_1, b_2) \ge L^{p,3}_{\tau_1,\tau_2}(b_1, b_2).
 $$
@@
 
-The values of $(x', y', w')$ maximizing $L^{p,1}_{\tau_1,\tau_2}(c_{xy}, d_{xy})$ are $(1, 1-c_{xy}-2d_{xy},0)$.
+The values of $(x', y', w')$ maximizing $L^{p,1}_{\tau_1,\tau_2}(b_1, b_2)$ are $(1, 1-b_1-2b_2,0)$.
 
@@
 
-We want to find values for $\{x,y\}$ where
+We want to find values for $\{x,y\}$ where all of
 $$
-h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-c_{xy}-2d_{xy}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(c_{xy},d_{xy}) \ge h_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p}_{\tau_1,\tau_1}(a_{xy},b_{xy}).
-$$
-Since this is a single inequality in two variables, we can plot it analytically; we need the region where
-\begin{align*}
-    0 &\ge h_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p}_{\tau_1,\tau_1}(a_{xy},b_{xy}) \\
-      &\qquad - h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-c_{xy}-2d_{xy}, 0\}) - L^{p,1}_{\tau_1,\tau_2}(c_{xy},d_{xy}) \\
-%
-      &\qquad = 2(1+x^2-y^2-x^2y^2)\log(1+x^2-y^2-x^2y^2) \\
-      &\qquad + 2(1-x^2+y^2-x^2y^2)\log(1-x^2+y^2-x^2y^2) \\
-      &\qquad + 2(1-x^2-y^2+x^2y^2)\log(1-x^2-y^2+x^2y^2) \\
-      &\qquad + (1+x^2+y^2+4xy^2+x^2y^2)\log(1+x^2+y^2+4xy^2+x^2y^2) \\
-      &\qquad + (1+x^2+y^2-4xy^2+x^2y^2)\log(1+x^2+y^2-4xy^2+x^2y^2) \\
-%
-      &\qquad + (2-a_{xy})\log(2-a_{xy})+a_{xy}\log(a_{xy})+(2-b_{xy})\log(2-b_{xy})+b_{xy}\log(b_{xy})+\log(2) \\
-%
-      &\qquad - 2(1+x^2-y^2-x^2y^2)\log(1+x'^2w'-y'^2w'-x'^2y'^2) \\
-      &\qquad - 2(1-x^2+y^2-x^2y^2)\log(1-x'^2w'+y'^2w'-x'^2y'^2) \\
-      &\qquad - (1-x^2-y^2+x^2y^2)\log(1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2) \\
-      &\qquad - (1-x^2-y^2+x^2y^2)\log(1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2) \\
-      &\qquad - (1+x^2+y^2+4xy^2+x^2y^2)\log(1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2) \\
-      &\qquad - (1+x^2+y^2-4xy^2+x^2y^2)\log(1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2) \\
-%
-      &\qquad - 2\log(2)-(2-c_{xy}-2d_{xy})\log(2-c_{xy}-2d_{xy})-(c_{xy}+2d_{xy})\log(c_{xy}+2d_{xy})
-\end{align*}
-where
-$$
-a_{xy} = b_{xy} = \frac{4-2x^2-2y^2}{8},
+h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-b_{1}-2b_{2}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(b_{1},b_{2}) \ge h_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p,1}_{\tau_1,\tau_1}(\mathbf{a}_{xy}),
 $$
 $$
-c_{xy} = \frac{4-4x^2y^2}{8},
-$$
-$$
-d_{xy} = \frac{2-4xy^2+2x^2y^2}{8}
+h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-b_{1}-2b_{2}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(b_{1},b_{2}) \ge h_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p,2}_{\tau_1,\tau_1}(\mathbf{a}_{xy})
 $$
@@
 $$
-x' = 1, \ y' = 1-c_{xy}-2d_{xy}, \ w' = 0.
+h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-b_{1}-2b_{2}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(b_{1},b_{2}) \ge h_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p,3}_{\tau_1,\tau_1}(\mathbf{a}_{xy})
 $$
-
-The \texttt{sage} function in listing~\ref{list:sage} will generate Fig.~\ref{fig:inconsistency-farris}.
-
-\begin{table}
-    \begin{center}
-\begin{verbatim}
-var('x,y')
-a = (4-2*x^2-2*y^2)/8
-b = (4-2*x^2-2*y^2)/8
-c = (4-4*x^2*y^2)/8
-d = (2-4*x*y^2+2*x^2*y^2)/8
-xp = 1.
-yp = 1-c-2*d
-wp = 0.
-
-inconsistency = 2*(1+x^2-y^2-x^2*y^2)*log(1+x^2-y^2-x^2*y^2) \
-    + 2*(1-x^2+y^2-x^2*y^2)*log(1-x^2+y^2-x^2*y^2) \
-    + 2*(1-x^2-y^2+x^2*y^2)*log(1-x^2-y^2+x^2*y^2) \
-    + (1+x^2+y^2+4*x*y^2+x^2*y^2)*log(1+x^2+y^2+4*x*y^2+x^2*y^2) \
-    + (1+x^2+y^2-4*x*y^2+x^2*y^2)*log(1+x^2+y^2-4*x*y^2+x^2*y^2) \
-    + (2-a)*log(2-a)+a*log(a)+(2-b)*log(2-b)+b*log(b)+log(2.) \
-    - 2*(1+x^2-y^2-x^2*y^2)*log(1+xp^2*wp-yp^2*wp-xp^2*yp^2) \
-    - 2*(1-x^2+y^2-x^2*y^2)*log(1-xp^2*wp+yp^2*wp-xp^2*yp^2) \
-    - (1-x^2-y^2+x^2*y^2)*log(1+2*xp*yp-2*xp*yp*wp-xp^2*wp-yp^2*wp+xp^2*yp^2) \
-    - (1-x^2-y^2+x^2*y^2)*log(1-2*xp*yp+2*xp*yp*wp-xp^2*wp-yp^2*wp+xp^2*yp^2) \
-    - (1+x^2+y^2+4*x*y^2+x^2*y^2)*log(1+2*xp*yp+2*xp*yp*wp+xp^2*wp+yp^2*wp+xp^2*yp^2) \
-    - (1+x^2+y^2-4*x*y^2+x^2*y^2)*log(1-2*xp*yp-2*xp*yp*wp+xp^2*wp+yp^2*wp+xp^2*yp^2) \
-    - 2*log(2.)-(2-c-2*d)*log(2-c-2*d)-(c+2*d)*log(c+2*d)
-
-incons_plot = implicit_plot(
-    inconsistency(x,y)==0,
-    (x,0,1),
-    (y,0,1),
-    axes_labels=['$x$', '$y$'],
-    title='Region of inconsistency for Farris-generating topology',
-    color='black'
-)
-\end{verbatim}
-    \end{center}
-\caption{Sage code to generate Fig.~\ref{fig:inconsistency-farris}}
-\label{list:sage}
-\end{table}
+hold.
+Since these are three inequalities in two variables, we can plot them analytically---see Fig.~\ref{fig:inconsistency-farris}.
 
@@
 \caption{
-    Region of inconsistency.
+    Regions of inconsistency.
     Due to the looseness of the upper and lower bounds, the parameters in the top right do not necessarily indicate consistency, though all parameters in the labeled region will result in an inconsistency.
@@
 
-\paragraph{Example}
+\paragraph{Examples}
+
+See Table~\ref{tab:examples} for parameters, bounds and true likelihood values obtained via basin hopping.
 
@@
 $$
-a_{xy} = b_{xy} = 0.46875
+L^{p,1}_{\tau_1,\tau_1}(\mathbf{a}_{xy}) \approx -2.17803560694157,
 $$
-resulting in
 $$
-L^{p}_{\tau_1,\tau_1}(a_{xy},b_{xy}) \approx -2.17803560694157
+L^{p,2}_{\tau_1,\tau_1}(\mathbf{a}_{xy}) \approx -2.17803560694157
 $$
@@
 $$
-h_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\})  \approx -2.04470918734314.
+L^{p,3}_{\tau_1,\tau_1}(\mathbf{a}_{xy}) \approx -2.17803560694157.
 $$
@@
 $$
-\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,t) \le -4.22274479428471
+\max_{t,\fullAncestralStateCategories} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,t) \le -4.22274479428471
 $$
@@
 $$
-\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,t) \ge -4.22274479428471.
+\max_{t,\fullAncestralStateCategories} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,t) \ge -4.22274479428471.
 $$
@@
 $$
-c_{xy} = 0.498046875, \ d_{xy} = 0.2431640625,
+d_{xy} = 0.498046875, \ e_{xy} = 0.2431640625,
 $$
@@
 $$
-L^{p,1}_{\tau_1,\tau_2}(c_{xy},d_{xy}) \approx -2.07968569223991
+L^{p,1}_{\tau_1,\tau_2}(d_{xy},e_{xy}) \approx -2.07968569223991
 $$
@@
 $$
-h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-c_{xy}-2d_{xy}, 0\}) \approx -2.07748833719921.
+h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-d_{xy}-2e_{xy}, 0\}) \approx -2.07748833719921.
 $$
@@
 $$
-\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,t) \ge -4.22274479428471 \ge -4.22274479428471 \ge \max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,t)
+\max_{t,\fullAncestralStateCategories} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,t) \ge -4.15717402943912
+\ge -4.22274479428471 \ge \max_{t,\fullAncestralStateCategories} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,t)
 $$
@@
 $$
-\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,t) \approx -4.5101771294218045
+\max_{t,\fullAncestralStateCategories} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,t) \approx -4.5101771294218045
 $$
@@
 $$
-\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,t) \approx -4.1544859851260281.
+\max_{t,\fullAncestralStateCategories} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,t) \approx -4.1544859851260281.
 $$
-Woohoo!
+It looks like the lower bound is tighter than the upper bound, so we may try focusing there.
 

### 2017-03-30 joint_inf.tex
a lot of notation updates

@@
 \newcommand{\alignmentColumn}{\mathbf{y}}
-\newcommand{\sitePattern}{s}
-\newcommand{\siteSplit}{S'}
-\newcommand{\sitePatternSet}{S}
-\newcommand{\fullSitePatterns}{\mathbf{s}}
-\newcommand{\sitePatternSpace}{\mathcal{S}}
+\newcommand{\alignmentColumnRV}{Y}
+\newcommand{\siteSplit}{\tilde{s}}
+\newcommand{\siteSplitSet}{\mathcal{S}}
 \newcommand{\fullAncestralStates}{\mathbf{H}}
-\newcommand{\ancestralStateSpace}{\mathcal{H}}
-\newcommand{\ancestralStateColumn}{\boldsymbol\eta}
-\newcommand{\ancestralStateCategory}{Z}
-\newcommand{\ancestralSplit}{H'}
-\newcommand{\ancestralStateSet}{H}
-\newcommand{\fullAncestralStateCategories}{\mathbb{Z}}
-\newcommand{\ancestralStateCategorySpace}{\mathcal{Z}}
+\newcommand{\ancestralStateColumn}{\mathbf{h}}
+\newcommand{\ancestralStateColumnRV}{H}
+\newcommand{\ancestralSplit}{\tilde{h}}
+\newcommand{\ancestralSplitSet}{\mathcal{H}}
+\newcommand{\ancestralSplitPartition}{\eta}
+\newcommand{\fullAncestralSplitPartitions}{\boldsymbol\eta}
+
+\newcommand{\patternToSplit}{\psi}
+\newcommand{\ancestralToSplit}{\xi}
+
+\newcommand{\siteSplitRV}{\Psi}
+\newcommand{\ancestralSplitRV}{\Xi}
 
@@
 \newcommand{\nAncestralSplits}{r}
-\newcommand{\nMostLikelyAncestralSplits}{\tilde{r}}
+
+\newcommand{\shannonDivergence}{D}
+
+\allowdisplaybreaks
 
@@
 %Let $\fullAlignment$ be the set of assignments of characters to tips of our tree, and $\fullAncestralStates$ the set of assignments of characters to the internal nodes of the tree.
-Assume we observe $\nCols$ samples of character data (i.e.\ an alignment with $\nCols$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$.
-There are also corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$.
+Assume we observe $\nCols$ samples of character data (i.e.\ an alignment with $\nCols$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
+There are also corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ distributed as $\ancestralStateColumnRV$.
 
@@
 \label{eq:full_likelihood}
-L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t).
+L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i | \tau, t).
 \end{equation}
@@
 % Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
-The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across ancestral states
+The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across all possible ancestral states
 \begin{equation}
 \label{eq:marginal_likelihood}
-\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
+\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
 \end{equation}
@@
 Following Section 8.6 of \cite{Semple2003-em}, $\tau$ induces an ordered set of tip labels $\{1,\ldots,\nSiteRows\}$ across sites.
-%EM What are complementary characters?
-For a binary character model, a realization of $\alignmentColumn_i$ can be described as a subset of complementary characters $\siteSplit\subseteq\sitePatternSet\triangleq\{1,\ldots,\nSiteRows-1\}$.
+For a binary character model, a realization of $\alignmentColumn_i$ can be described as a subset of indices $\siteSplit\subseteq\siteSplitSet\triangleq\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
+In the binary character setting, the probability of observing a particular site split is equivalent to the probability of its complement, resulting in one fewer index in $\siteSplitSet$ than the number of tips.
+Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ where $\mathcal{P}$ denotes the powerset.
+The topology $\tau$ also induces an ordered set of internal nodes $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ with a similar subset of characters $\ancestralSplit\subseteq\ancestralSplitSet\triangleq\{1,\ldots,\nAncestralStateRows\}$ describing a realization of $\ancestralStateColumn_i$.
+The probability of observing an ancestral state split, conditional on a site split, will not be invariant to taking its complement, and thus the need for enumerating the entire set of internal nodes.
+The number of unique ancestral state splits is $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$.
+
 %EM Why nSiteRows-1 in the def of siteSplit above, but nAncestralStateRows below in the def of ancestralSplit?
 %EM There are two ways to do this, if you are collapsing such that 001 and 110 are the same or not, and we should be clear which one you are doing.
-The topology $\tau$ induces an ordered set $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ with a similar subset of complementary characters $\ancestralSplit\subseteq\ancestralStateSet\triangleq\{1,\ldots,\nAncestralStateRows\}$ describing a realization of $\ancestralStateColumn_i$.
-Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\sitePatternSet)|$ and unique ancestral state splits as $\nAncestralSplits=|\mathcal{P}(\ancestralStateSet)|$ where $\mathcal{P}$ denotes the powerset.
 
-%das: Maybe writing out the example would help?
-%EM: Yes, I think so. e.g. the need for partitions \ancestralStateCategory may not be obvious right away.
-
-Consider the probability $P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t)$ from \eqref{eq:full_likelihood} of observing a particular column $i$.
+With this split formulation, define the mappings from patterns to splits as
+$$
+\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet), \ \ancestralToSplit:\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet)
+$$
+so that, for
+$$
+\patternToSplit(\alignmentColumn_i)=\siteSplit_j, \ \ancestralToSplit(\ancestralStateColumn_i)=\ancestralSplit_k
+$$
+and
+$$
+\patternToSplit(\alignmentColumnRV)=\siteSplitRV, \ \ancestralToSplit(\ancestralStateColumnRV)=\ancestralSplitRV,
+$$
+we write the $i$th factor of \eqref{eq:full_likelihood} as
+$$
+P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i | \tau, t)=P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k | \tau, t).
+$$
+Taking the maximum of the probability of observing a particular column $i$ yields
+\begin{align*}
+\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k | \tau, t) &=
+\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j| \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k |\siteSplitRV=\siteSplit_j, \tau, t) \\
+&= P(\siteSplitRV=\siteSplit_j| \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k |\siteSplitRV=\siteSplit_j, \tau, t).
+\end{align*}
 %EM Indexing \siteSplit_j$ not defined.
@@
 Personally, I think that using the same index j for them is confusing.
-Given $(\tau, t)$ and maximizing over $\ancestralStateColumn_i$, if $\alignmentColumn_i=\siteSplit_j$ then there exists a partition $\ancestralStateCategory_j$ of $\mathcal{P}(\ancestralStateSet)$ of most likely ancestral states where
+Given $(\tau, t)$ and $\siteSplitRV=\siteSplit_j$, there exists a partition $\ancestralSplitPartition(\siteSplit_j, \tau, t)=\ancestralSplitPartition_j(\tau, t)$ of $\mathcal{P}(\ancestralSplitSet)$ of most likely ancestral states where
 \begin{align*}
-\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralStateSet)} \ P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i=\ancestralSplit_k | \tau, t) &= P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i\in\ancestralStateCategory_j | \tau, t) \\
-     &= P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t).
+\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k | \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t) \\
+     &= P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t).
 \end{align*}
-This equivalence is helpful in writing the full likelihood of the joint inference problem in terms of site splits; for most likely ancestral state partitions $\fullAncestralStateCategories=\{\ancestralStateCategory_1,\ldots,\ancestralStateCategory_\nSiteSplits\}$ we reformulate the likelihood equation
+
+For most likely ancestral state partitions $\fullAncestralSplitPartitions(\tau, t)=\{\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)\}$ we reformulate the likelihood equation
 \begin{align}
 L_\nCols'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) \\
-                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{\nCols} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t) \\
-                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralStateSet)} \ P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i=\ancestralSplit_k | \tau, t) \\
+                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i | \tau, t) \\
+                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k | \tau, t) \\
 %EM Can you think a little about z_j?
 %EM Even though it's a category now, it's still a function of tau and t, right?
-                             &= \prod_{i=1}^{\nCols} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t) \\
-                             &= \prod_{j=1}^{\nSiteSplits} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t)^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
+                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j| \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k |\siteSplitRV=\siteSplit_j, \tau, t) \\
+                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j | \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t) \\
+                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j | \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
 \end{align}
 where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ appears.
+
+\paragraph{Example}
+
+Consider a fixed, binary four-tip tree $\tau$.
+The space of all possible character assignments, in terms of site splits, is
+\begin{align*}
+\mathcal{P}(\{1,2,3,4\}) &= \{\{\emptyset\}, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
+                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
+\end{align*}
+where each set indicates the tips at which a character is assigned the character $1$.
+For example, $\{\emptyset\}$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
+Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
+The unique site splits, collapsing complements, are
+$$
+\mathcal{P}(\siteSplitSet) = \{\{\emptyset\}, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\}.
+$$
+Had we not identified character complements, we would need to consider the additional splits
+$$
+\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
+$$
+but the symmetry of the binary character model allows us to focus just on the elements of $\mathcal{P}(\siteSplitSet)$.
+This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
+$$
+\mathcal{P}(\ancestralSplitSet) = \{\{\emptyset\}, \{1\}, \{2\}, \{1,2\}\}.
+$$
+We cannot similarly identify complements in this case since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
+$$
+P(\ancestralSplitRV=\{\emptyset\}|\siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\}|\siteSplitRV=\siteSplit, \tau, t)
+$$
+in general.
+
+For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
+A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; define $\ancestralSplitPartition(\siteSplit, \tau, t)$ as the partition of $\mathcal{P}(\ancestralSplitSet)$ into those most likely ancestral splits for that particular site split, topology and set of branch lengths.
+Assume all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
+The probabilities of observing ancestral splits for site split $\{\emptyset\}$ are 
+$$
+P(\ancestralSplitRV=\{\emptyset\}|\siteSplitRV=\{\emptyset\}, \tau, t) = 
+(1-p)^5,
+$$
+$$
+P(\ancestralSplitRV=\{1\}|\siteSplitRV=\{\emptyset\}, \tau, t) =
+P(\ancestralSplitRV=\{2\}|\siteSplitRV=\{\emptyset\}, \tau, t) = 
+p^2(1-p)^3,
+$$
+$$
+P(\ancestralSplitRV=\{1,2\}|\siteSplitRV=\{\emptyset\}, \tau, t) = 
+p^4(1-p).
+$$
+Here, the partition into most likely ancestral states will proceed as
+$$
+P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p)|\siteSplitRV=\{\emptyset\}, \tau, t) = 
+(1-p)^5
+$$
+where the partition is the single element $\ancestralSplitPartition_j(\tau, p)=\{\emptyset\}$.
+For the site split $\{1,3\}$ we have
+$$
+P(\ancestralSplitRV=\{\emptyset\}|\siteSplitRV=\{1,3\}, \tau, t) = 
+P(\ancestralSplitRV=\{1,2\}|\siteSplitRV=\{1,3\}, \tau, t) =
+p^2(1-p)^3,
+$$
+$$
+P(\ancestralSplitRV=\{1\}|\siteSplitRV=\{1,3\}, \tau, t) = 
+P(\ancestralSplitRV=\{2\}|\siteSplitRV=\{1,3\}, \tau, t) =
+p^3(1-p)^2,
+$$
+where
+$$
+P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p)|\siteSplitRV=\{1,3\}, \tau, t) = 
+p^2(1-p)^3
+$$
+with the most likely partition $\ancestralSplitPartition_j(\tau, p)=\{\{\emptyset\},\{1,2\}\}.$
 
@@
 $$
-L_\nCols''(\fullAlignment;\fullAncestralStateCategories,\tau,t) = \prod_{j=1}^{\nSiteSplits} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t)^{\nCols_j(\fullAlignment)} .
+L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j | \tau, t) \cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
 $$
@@
 \begin{align}
-    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralStateCategories,\tau,t)
-        &= \sum_{j}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t) \\
-        &= \sum_{j}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\alignmentColumn=\siteSplit_j | \tau, t) +
-            \log P(\ancestralStateColumn\in\ancestralStateCategory_j | \alignmentColumn=\siteSplit_j , \tau, t)]
+    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau,t),\tau,t)
+        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \tau, t) \\
+        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j | \tau, t) +
+            \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j , \tau, t)]
 \end{align}
@@
 \begin{equation}
-\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j}^\nSiteSplits P(\alignmentColumn=\siteSplit_j | \tau^*, t^*) \cdot [\log P(\alignmentColumn=\siteSplit_j | \tau, t) + \log P(\ancestralStateColumn\in\ancestralStateCategory_j | \alignmentColumn=\siteSplit_j , \tau, t)].
+\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j | \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j | \tau, t) + \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j , \tau, t)].
 \label{eq:site_pattern_profile_likelihood_mean}
 \end{equation}
-Define the the information quantity
+Define the the divergence quantity
 $$
-h_{\tau^*,t^*}(\tau,t) = \sum_{j}^\nSiteSplits P(\alignmentColumn=\siteSplit_j | \tau^*, t^*) \log P(\siteSplit_j | \tau, t),
+\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j | \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j | \tau, t),
 $$
@@
 $$
-\ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = \sum_{j}^\nSiteSplits P(\alignmentColumn=\siteSplit_j | \tau^*, t^*) \log P(\ancestralStateColumn\in\ancestralStateCategory_j | \siteSplit_j, \tau, t)
+\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j | \tau^*, t^*)\cdot\log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplit_j, \tau, t)
 $$
@@
     \label{eq:log_likelihood_simplified}
-\ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = h_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
+    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
 \end{equation}
@@
 \label{eq:inconsistency_inequality}
-\ell_{\tau^*,t^*}(\hat{\fullAncestralStateCategories}_1,\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralStateCategories}_2,\tau^*,\hat{t}_2)
+\ell_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau',t_1)},\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau^*,t_2)},\tau^*,\hat{t}_2)
 \end{equation}
-for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralStateCategories}_1\}$ and $\{\hat{t}_2,\hat{\fullAncestralStateCategories}_2\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
+for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\widehat{\fullAncestralSplitPartitions(\tau',t_1)}\}$ and $\{\hat{t}_2,\widehat{\fullAncestralSplitPartitions(\tau^*,t_2)}\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
 
@@
 \label{eq:partial-likelihood}
-\ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = h_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
+\ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
 \end{equation}
@@
 $$
-\max_{t,\fullAncestralStateCategories} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \le
-    h_{\tau^*,t^*}(\tau^*,t^*)
-    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t})
+\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \le
+    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
+    + \ell^p_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau,t)},\tau,\hat{t})
 $$
@@
 $$
-h_{\tau^*,t^*}(\tau,t) \le h_{\tau^*,t^*}(\tau^*,t^*)
+\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
 $$
@@
 $$
-\{\hat{t},\hat{\fullAncestralStateCategories}\} = \arg\max_{t,\fullAncestralStateCategories} \ \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
+\{\hat{t},\widehat{\fullAncestralSplitPartitions(\tau,t)}\} = \arg\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
 $$
@@
 $$
-\max_{t,\fullAncestralStateCategories} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \ge
-    h_{\tau^*,t^*}(\tau,\hat{t})
-    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t}).
+\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \ge
+    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
+    + \ell^p_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau,t)},\tau,\hat{t}).
 $$
@@
     \hline
-$\siteSplit_j$   &$P(\alignmentColumn=\siteSplit_j|\tau_1,t^*)$&$P(\alignmentColumn=\siteSplit_j|\tau_2,t^*)$\\
+$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j|\tau_1,t^*)$&$P(\siteSplitRV=\siteSplit_j|\tau_2,t^*)$\\
     \hline
@@
     \hline
-$\siteSplit_j$   &$P(\alignmentColumn=\siteSplit_j|\tau_1,t)$&$P(\alignmentColumn=\siteSplit_j|\tau_2,t)$\\
+$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j|\tau_1,t)$&$P(\siteSplitRV=\siteSplit_j|\tau_2,t)$\\
     \hline
@@
     \hline
-    $\siteSplit_j$    & $\ancestralStateCategory_j$ & $P(\ancestralStateColumn\in\ancestralStateCategory_j|\alignmentColumn=\siteSplit_j,\tau,t)$\\
+    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t)|\siteSplitRV=\siteSplit_j,\tau,t)$\\
     \hline
@@
 \end{tabular}
-\caption{Likelihood calculations for all site patterns and ancestral state categories.
+\caption{Likelihood calculations for all site patterns and ancestral state partitions.
 All values multiplied by $1/32$.
@@
 Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
-Define, e.g., $p_0 = P(\alignmentColumn=\{\emptyset\}|\tau=\tau_1,t=\{x,y,y\})$.
+Define, e.g., $p_0 = P(\siteSplitRV=\{\emptyset\}|\tau=\tau_1,t=\{x,y,y\})$.
 Let
@@
 \begin{align*}
-    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,\{x',y',w'\}) \\
+    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\}) \\
     &\qquad \le (3-a_{1})\log(1+x')+(a_{1}+a_{3})\log(1-x')\\
@@
 \begin{align*}
-    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,\{x',y',w'\}) \\
+    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\}) \\
     &\qquad \le (3-a_{1}-2a_{0})\log(1+x')+(a_{1}+a_{3}+2a_{0})\log(1-x')\\
@@
 \begin{align*}
-    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,\{x',y',w'\}) \\
+    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\}) \\
     &\qquad \ge 2\log(1+x')+(2-b_1-2b_2)\log(1+y')+(b_1+2b_2)\log(1-y')+\log(1-w'),
@@
 \begin{align*}
-    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,\{x',y',w'\}) \\
+    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\}) \\
     &\qquad \ge (2-2b_2)\log(1+x')+2b_2\log(1-x') + (2-b_1)\log(1+y')+b_1\log(1-y')+\log(1-w')
@@
 \begin{align*}
-    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,\{x',y',w'\}) \\
+    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\}) \\
     &\qquad \ge (2-b_2)\log(1+x')+b_2\log(1-x') + (2-b_1-b_2)\log(1+y')+(b_1+b_2)\log(1-y')+\log(1-w')
@@
 $$
-h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-b_{1}-2b_{2}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(b_{1},b_{2}) \ge h_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p,1}_{\tau_1,\tau_1}(\mathbf{a}_{xy}),
+\shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-b_{1}-2b_{2}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(b_{1},b_{2}) \ge \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p,1}_{\tau_1,\tau_1}(\mathbf{a}_{xy}),
 $$
 $$
-h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-b_{1}-2b_{2}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(b_{1},b_{2}) \ge h_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p,2}_{\tau_1,\tau_1}(\mathbf{a}_{xy})
+\shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-b_{1}-2b_{2}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(b_{1},b_{2}) \ge \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p,2}_{\tau_1,\tau_1}(\mathbf{a}_{xy})
 $$
@@
 $$
-h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-b_{1}-2b_{2}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(b_{1},b_{2}) \ge h_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p,3}_{\tau_1,\tau_1}(\mathbf{a}_{xy})
+\shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-b_{1}-2b_{2}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(b_{1},b_{2}) \ge \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p,3}_{\tau_1,\tau_1}(\mathbf{a}_{xy})
 $$
@@
 $$
-\max_{t,\fullAncestralStateCategories} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,t) \le -4.22274479428471
+\max_{t,\fullAncestralSplitPartitions(\tau_1,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,t),\tau_1,t) \le -4.22274479428471
 $$
@@
 $$
-\max_{t,\fullAncestralStateCategories} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,t) \ge -4.22274479428471.
+\max_{t,\fullAncestralSplitPartitions(\tau_2,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,t),\tau_2,t) \ge -4.22274479428471.
 $$
@@
 $$
-h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-d_{xy}-2e_{xy}, 0\}) \approx -2.07748833719921.
+\shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-d_{xy}-2e_{xy}, 0\}) \approx -2.07748833719921.
 $$
@@
 $$
-\max_{t,\fullAncestralStateCategories} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,t) \ge -4.15717402943912
-\ge -4.22274479428471 \ge \max_{t,\fullAncestralStateCategories} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,t)
+\max_{t,\fullAncestralSplitPartitions(\tau_2,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,t),\tau_2,t) \ge -4.15717402943912
+\ge -4.22274479428471 \ge \max_{t,\fullAncestralSplitPartitions(\tau_1,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,t),\tau_1,t)
 $$
@@
 $$
-\max_{t,\fullAncestralStateCategories} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,t) \approx -4.5101771294218045
+\max_{t,\fullAncestralSplitPartitions(\tau_1,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,t),\tau_1,t) \approx -4.5101771294218045
 $$
@@
 $$
-\max_{t,\fullAncestralStateCategories} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,t) \approx -4.1544859851260281.
+\max_{t,\fullAncestralSplitPartitions(\tau_2,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,t),\tau_2,t) \approx -4.1544859851260281.
 $$

### 2017-03-30 joint_inf.tex
Pass through the beginning.

@@
 Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
+%EM Not sure how tau induces an ordered set of tip labels-- are you saying that the tree defines the labels or something about its structure?
 Following Section 8.6 of \cite{Semple2003-em}, $\tau$ induces an ordered set of tip labels $\{1,\ldots,\nSiteRows\}$ across sites.
-For a binary character model, a realization of $\alignmentColumn_i$ can be described as a subset of indices $\siteSplit\subseteq\siteSplitSet\triangleq\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
-In the binary character setting, the probability of observing a particular site split is equivalent to the probability of its complement, resulting in one fewer index in $\siteSplitSet$ than the number of tips.
+For the binary symmetric model, a realization of $\alignmentColumn_i$ can be described as a subset of indices $\siteSplit\subseteq\siteSplitSet\triangleq\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
+Indeed, we can start by including those tips in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
+Then, because the probability of observing a particular site split is equivalent to the probability of its complement under the binary symmetric model, we can take the complement if needed to exclude tip $\nSiteRows$ from $\siteSplit$.
 Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ where $\mathcal{P}$ denotes the powerset.
+%EM inducing again.
 The topology $\tau$ also induces an ordered set of internal nodes $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ with a similar subset of characters $\ancestralSplit\subseteq\ancestralSplitSet\triangleq\{1,\ldots,\nAncestralStateRows\}$ describing a realization of $\ancestralStateColumn_i$.
@@
 
-%EM Why nSiteRows-1 in the def of siteSplit above, but nAncestralStateRows below in the def of ancestralSplit?
-%EM There are two ways to do this, if you are collapsing such that 001 and 110 are the same or not, and we should be clear which one you are doing.
-
-With this split formulation, define the mappings from patterns to splits as
+With this split formulation, define the mappings from patterns to splits as described above
 $$
@@
 $$
-so that, for
+% the following seemed not necessary and a little confusing (what are j and k?)
+% so that, for
+% $$
+% \patternToSplit(\alignmentColumn_i)=\siteSplit_j, \ \ancestralToSplit(\ancestralStateColumn_i)=\ancestralSplit_k.
+% $$
+Given a site-pattern-valued random variable $H$ and an ancestral-state-valued random variable $H$, define random variables
 $$
-\patternToSplit(\alignmentColumn_i)=\siteSplit_j, \ \ancestralToSplit(\ancestralStateColumn_i)=\ancestralSplit_k
+\siteSplitRV := \patternToSplit(\alignmentColumnRV), \ \ancestralSplitRV := \ancestralToSplit(\ancestralStateColumnRV).
 $$
-and
+We can now write the $i$th factor of \eqref{eq:full_likelihood} as
 $$
-\patternToSplit(\alignmentColumnRV)=\siteSplitRV, \ \ancestralToSplit(\ancestralStateColumnRV)=\ancestralSplitRV,
+P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)=P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)
 $$
-we write the $i$th factor of \eqref{eq:full_likelihood} as
-$$
-P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i | \tau, t)=P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k | \tau, t).
-$$
+for some $j$ and $k$.
 Taking the maximum of the probability of observing a particular column $i$ yields
@@
 \end{align*}
-%EM Indexing \siteSplit_j$ not defined.
-%EM But, more importantly, as we discussed Z_j is defined only with respect to a single tree and branch lengths.
-This need to be made abundantly clear.
-Personally, I think that using the same index j for them is confusing.
-Given $(\tau, t)$ and $\siteSplitRV=\siteSplit_j$, there exists a partition $\ancestralSplitPartition(\siteSplit_j, \tau, t)=\ancestralSplitPartition_j(\tau, t)$ of $\mathcal{P}(\ancestralSplitSet)$ of most likely ancestral states where
+%EM Just noticed that eta doesn't have to be a real partition, because they don't have to be disjoint, right?
+Given $(\tau, t)$ there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
 \begin{align*}
-\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k | \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t) \\
-     &= P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t).
+\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi \mid \siteSplitRV=\siteSplit_j, \tau, t).
 \end{align*}
 
-For most likely ancestral state partitions $\fullAncestralSplitPartitions(\tau, t)=\{\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)\}$ we reformulate the likelihood equation
 \begin{align}
 L_\nCols'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) \\
-                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i | \tau, t) \\
-                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k | \tau, t) \\
-%EM Can you think a little about z_j?
-%EM Even though it's a category now, it's still a function of tau and t, right?
-                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j| \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k |\siteSplitRV=\siteSplit_j, \tau, t) \\
-                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j | \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t) \\
-                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j | \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
+%EM I guess the ith component of H is h_i?
+                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
+%EM Indexing broken here: we are taking a product over i, but i doesn't appear in the expression.
+                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) \\
+                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
+%EM As I fixed above and described in person, \ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) is not what you want here, because it will inflate the value by |\ancestralSplitPartition_j(\tau, t)| .
+This also appears below.
+                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
+                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
 \end{align}
@@
 \paragraph{Example}
-
+%EM My long term suggestion is for you to put this in example in the main text and the above in the appendix.
 Consider a fixed, binary four-tip tree $\tau$.
-The space of all possible character assignments, in terms of site splits, is
+The set of all possible character assignments can be described as
 \begin{align*}
@@
 For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
+%EM I dropped the \ancestralSplitPartition(\siteSplit, \tau, t) notation above. You can revert it if you like, but I prefer the bold eta notation as an ordered list of sets.
 A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; define $\ancestralSplitPartition(\siteSplit, \tau, t)$ as the partition of $\mathcal{P}(\ancestralSplitSet)$ into those most likely ancestral splits for that particular site split, topology and set of branch lengths.
-Assume all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
-The probabilities of observing ancestral splits for site split $\{\emptyset\}$ are 
+As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
+The probabilities of observing ancestral splits for site split $\emptyset$ are
 $$
@@
 $$
-P(\ancestralSplitRV=\{1\}|\siteSplitRV=\{\emptyset\}, \tau, t) =
-P(\ancestralSplitRV=\{2\}|\siteSplitRV=\{\emptyset\}, \tau, t) = 
+%EM Seems to me that this just requires one character change so would be p(1-p)^4.
+P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
+P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
 p^2(1-p)^3,
@@
 $$
-P(\ancestralSplitRV=\{1,2\}|\siteSplitRV=\{\emptyset\}, \tau, t) = 
+%EM Have you told us in this example what the labeled topology is?
+%Erick stopped here.
+P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
 p^4(1-p).

### 2017-03-30 joint_inf.tex
using \mid and fixing emptyset thing

@@
 \label{eq:full_likelihood}
-L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i | \tau, t).
+L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
 \end{equation}
@@
 \begin{align*}
-\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k | \tau, t) &=
-\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j| \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k |\siteSplitRV=\siteSplit_j, \tau, t) \\
-&= P(\siteSplitRV=\siteSplit_j| \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k |\siteSplitRV=\siteSplit_j, \tau, t).
+\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &=
+\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
+&= P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t).
 \end{align*}
@@
 \begin{align*}
-\mathcal{P}(\{1,2,3,4\}) &= \{\{\emptyset\}, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
+\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                          &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
@@
 where each set indicates the tips at which a character is assigned the character $1$.
-For example, $\{\emptyset\}$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
+For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
 Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
@@
 $$
-\mathcal{P}(\siteSplitSet) = \{\{\emptyset\}, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\}.
+\mathcal{P}(\siteSplitSet) = \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\}.
 $$
@@
 $$
-\mathcal{P}(\ancestralSplitSet) = \{\{\emptyset\}, \{1\}, \{2\}, \{1,2\}\}.
+\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
 $$
@@
 $$
-P(\ancestralSplitRV=\{\emptyset\}|\siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\}|\siteSplitRV=\siteSplit, \tau, t)
+P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
 $$
@@
 $$
-P(\ancestralSplitRV=\{\emptyset\}|\siteSplitRV=\{\emptyset\}, \tau, t) = 
+P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
 (1-p)^5,
@@
 $$
-P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p)|\siteSplitRV=\{\emptyset\}, \tau, t) = 
+P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p) \mid \siteSplitRV=\emptyset, \tau, t) =
 (1-p)^5
 $$
-where the partition is the single element $\ancestralSplitPartition_j(\tau, p)=\{\emptyset\}$.
+where the partition is the single element $\ancestralSplitPartition_j(\tau, p)=\emptyset$.
 For the site split $\{1,3\}$ we have
 $$
-P(\ancestralSplitRV=\{\emptyset\}|\siteSplitRV=\{1,3\}, \tau, t) = 
-P(\ancestralSplitRV=\{1,2\}|\siteSplitRV=\{1,3\}, \tau, t) =
+P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,3\}, \tau, t) =
+P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
 p^2(1-p)^3,
@@
 $$
-P(\ancestralSplitRV=\{1\}|\siteSplitRV=\{1,3\}, \tau, t) = 
-P(\ancestralSplitRV=\{2\}|\siteSplitRV=\{1,3\}, \tau, t) =
+P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
+P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
 p^3(1-p)^2,
@@
 $$
-P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p)|\siteSplitRV=\{1,3\}, \tau, t) = 
+P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p) \mid \siteSplitRV=\{1,3\}, \tau, t) =
 p^2(1-p)^3
 $$
-with the most likely partition $\ancestralSplitPartition_j(\tau, p)=\{\{\emptyset\},\{1,2\}\}.$
+with the most likely partition $\ancestralSplitPartition_j(\tau, p)=\{\emptyset,\{1,2\}\}.$
 
@@
 $$
-L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j | \tau, t) \cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
+L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
 $$
@@
     \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau,t),\tau,t)
-        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \tau, t) \\
-        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j | \tau, t) +
-            \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j , \tau, t)]
+        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \tau, t) \\
+        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
+            \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j , \tau, t)]
 \end{align}
@@
 \begin{equation}
-\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j | \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j | \tau, t) + \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j , \tau, t)].
+\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j , \tau, t)].
 \label{eq:site_pattern_profile_likelihood_mean}
@@
 $$
-\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j | \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j | \tau, t),
+\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
 $$
@@
 $$
-\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j | \tau^*, t^*)\cdot\log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplit_j, \tau, t)
+\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplit_j, \tau, t)
 $$
@@
     \hline
-$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j|\tau_1,t^*)$&$P(\siteSplitRV=\siteSplit_j|\tau_2,t^*)$\\
+$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t^*)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t^*)$\\
     \hline
-    $\{\emptyset\}$&$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
+    $\emptyset$    &$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
     $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
@@
     \hline
-$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j|\tau_1,t)$&$P(\siteSplitRV=\siteSplit_j|\tau_2,t)$\\
+$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
     \hline
-    $\{\emptyset\}$&$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
+    $\emptyset$     &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
     $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
@@
     \hline
-    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t)|\siteSplitRV=\siteSplit_j,\tau,t)$\\
+    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
     \hline
-    $\{\emptyset\}$&
-    $\{\emptyset\}$&
+    $\emptyset$&
+    $\emptyset$&
     $(1+x)^2   (1+w)(1+y)^2$\\
      $\{1\}$    &
-    $\{\emptyset\}$&
+    $\emptyset$&
     $(1+x)(1-x)(1+w)(1+y)^2$\\
      $\{2\}$    &
-    $\{\emptyset\}$&
+    $\emptyset$&
     $(1+x)(1-x)(1+w)(1+y)^2$\\
      $\{3\}$    &
-    $\{\emptyset\}$&
+    $\emptyset$&
     $(1+x)^2   (1+w)(1+y)(1-y)$\\
     $\{1,2\}$  &
-    $\{\{\emptyset\},\{1,2,3\}\}$&
+    $\{\emptyset,\{1,2,3\}\}$&
     $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
@@
     $\left\{\begin{array}{l}
-                    \{\emptyset\}\\
+                    \emptyset\\
                     \{2\}\\
@@
     $\{2,3\}$  &
-                $\{\{\emptyset\},\{1,2,3\}\}$&
+                $\{\emptyset,\{1,2,3\}\}$&
                 $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$&
-                $\{\emptyset\}$&
+                $\emptyset$&
                 $(1+x)^2   (1+w)(1+y)(1-y)$\\
@@
     \hline
-    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t)|\siteSplitRV=\siteSplit_j,\tau,t)$\\
+    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
     \hline
-    $\{\emptyset\}$       &$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)^2$\\
+    $\emptyset$       &$\emptyset$&$(1+x)^2   (1+w)(1+y)^2$\\
     $\{1\}$          &
     $\left\{\begin{array}{l}
-                    \{\emptyset\}\\
+                    \emptyset\\
                     \{1\}
@@
     $\left\{\begin{array}{l}
-                    \{\emptyset\}\\
+                    \emptyset\\
                     \{1\}
@@
     $\left\{\begin{array}{l}
-                    \{\emptyset\}\\
+                    \emptyset\\
                     \{2\}
@@
     $\left\{\begin{array}{l}
-                    \{\{\emptyset\},\{1,2\}\}\\
+                    \{\emptyset,\{1,2\}\}\\
                     \{2\}
@@
     $\left\{\begin{array}{l}
-                    \{\emptyset\}\\
+                    \emptyset\\
                     \{\{1\},\{2\}\}\\
@@
     $\left\{\begin{array}{l}
-                    \{\{\emptyset\},\{1,2\}\}\\
+                    \{\emptyset,\{1,2\}\}\\
                     \{1\}\\
@@
 Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
-Define, e.g., $p_0 = P(\siteSplitRV=\{\emptyset\}|\tau=\tau_1,t=\{x,y,y\})$.
+Define, e.g., $p_0 = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})$.
 Let

### 2017-03-31 joint_inf.tex
writing out explicit steps

@@
 %EM Not sure how tau induces an ordered set of tip labels-- are you saying that the tree defines the labels or something about its structure?
-Following Section 8.6 of \cite{Semple2003-em}, $\tau$ induces an ordered set of tip labels $\{1,\ldots,\nSiteRows\}$ across sites.
-For the binary symmetric model, a realization of $\alignmentColumn_i$ can be described as a subset of indices $\siteSplit\subseteq\siteSplitSet\triangleq\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
+Following Section 8.6 of \cite{Semple2003-em}, we define an ordered set of tip labels $\{1,\ldots,\nSiteRows\}$ on $\tau$ across sites.
+For the binary symmetric model, a realization of $\alignmentColumn_i$ can be described as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
 Indeed, we can start by including those tips in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
@@
 %EM inducing again.
-The topology $\tau$ also induces an ordered set of internal nodes $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ with a similar subset of characters $\ancestralSplit\subseteq\ancestralSplitSet\triangleq\{1,\ldots,\nAncestralStateRows\}$ describing a realization of $\ancestralStateColumn_i$.
+For topology $\tau$, we also define an ordered set of internal nodes $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ with a similar subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ describing a realization of $\ancestralStateColumn_i$.
 The probability of observing an ancestral state split, conditional on a site split, will not be invariant to taking its complement, and thus the need for enumerating the entire set of internal nodes.
@@
 % $$
-Given a site-pattern-valued random variable $H$ and an ancestral-state-valued random variable $H$, define random variables
+Given a site-pattern-valued random variable $\alignmentColumnRV$ and an ancestral-state-valued random variable $\ancestralStateColumnRV$, define random variables
 $$
@@
 %EM Just noticed that eta doesn't have to be a real partition, because they don't have to be disjoint, right?
+%das: I think they do have to be disjoint. The partition is into "most likely" and "not most likely," which is disjoint and exhausts the set. Am I missing something?
 Given $(\tau, t)$ there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
@@
 %EM I guess the ith component of H is h_i?
+    %das: Not sure what the question/comment is here.
                              &= \max_{\fullAncestralStates} \ \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
 %EM Indexing broken here: we are taking a product over i, but i doesn't appear in the expression.
+    %das: This is intentional, though: it allows us to proceed to the following line, grouping factors by site pattern since their probabilities don't depend on $i$.
+I can try to think of how to make this more clear, but I thought it was a pretty key step to be able to pass to site patterns, and the reason for the machinery involved in defining ancestral state partitions.
                              &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) \\
@@
 %EM Seems to me that this just requires one character change so would be p(1-p)^4.
+%das: I think it's wrong, but in the other direction: two character changes to get to the differing internal state, then one between the two internal states.
 P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
 P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
-p^2(1-p)^3,
+p^3(1-p)^2,
 $$
@@
 %EM Have you told us in this example what the labeled topology is?
+%das: Nope! I wanted to be simple, but it seems maybe too simple. Maybe I should move the Felsenstein/Farris bit to this example?
 %Erick stopped here.
@@
 Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
-Define, e.g., $p_0 = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})$.
-Let
+For compactness, define, e.g., $p_0 = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})$, keeping in mind it is a function of parameters $x,y$.
+The full Farris log likelihood can take one of three values---assume for now the we are only interested in the ancestral state split $\{2\}$ for the site split $\{1,3\}$.
+We then write the log likelihood as
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
+    &=        p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{1}  \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
+    &\qquad + p_{2}  \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
+    &\qquad + p_{3}  \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
+    &\qquad + p_{12} \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
+    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{23} \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
+    &\qquad + p_{123}\cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
+    &\qquad + p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
+    &\qquad + p_{1}  \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
+    &\qquad + p_{2}  \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
+    &\qquad + p_{3}  \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
+    &\qquad + p_{12} \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
+    &\qquad + p_{13} \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
+    &\qquad + p_{23} \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
+    &\qquad + p_{123}\cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
+\end{align*}
+Seeking an upper bound, we make use of the fact that, for $x',y'\in[0,1]$,
 $$
-    a_{0} = p_{13},
+(1-x'^2+y'^2-x'^2y'^2) = (1-x'^2)(1+y'^2) = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y),
+$$
+with similar substitutions for $(1+x'^2-y'^2-x'^2y'^2)$ and $(1-x'^2-y'^2+x'^2y'^2)$.
+Doing so results in
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
+    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{1}  \cdot\log((1+x')(1-x')(1+y')) \\
+    &\qquad + p_{2}  \cdot\log((1+x')(1+y')(1-y')) \\
+    &\qquad + p_{3}  \cdot\log((1+x')(1-x')(1+y')) \\
+    &\qquad + p_{12} \cdot\log((1+x')(1-x')(1+y')(1-y')) \\
+    &\qquad + p_{23} \cdot\log((1+x')(1-x')(1+y')(1-y')) \\
+    &\qquad + p_{123}\cdot\log((1+x')(1+y')(1-y')) \\
+    &\qquad + p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
+    &\qquad + p_{1}  \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
+    &\qquad + p_{2}  \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
+    &\qquad + p_{3}  \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
+    &\qquad + p_{12} \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
+    &\qquad + p_{13} \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
+    &\qquad + p_{23} \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
+    &\qquad + p_{123}\cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
+\end{align*}
+Grouping together like terms,
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
+    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + (1-p_{0}-p_{13})\cdot\log(1+x') \\
+    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
+    &\qquad + (1-p_{0}-p_{13})\cdot\log(1+y') \\
+    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
+    &\qquad + (2-p_{1}-p_{2}-p_{12}-p_{23})\cdot\log(1+x') \\
+    &\qquad + (p_{1}+p_{2}+p_{12}+p_{23})\cdot\log(1-x') \\
+    &\qquad + (2-p_{3}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\
+    &\qquad + (p_{3}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
+    &\qquad + (1-p_{13})\cdot\log(1+w') \\
+    &\qquad + p_{13}\cdot\log(1-w'),
+\end{align*}
+which equals
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
+    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + (3-p_{0}-p_{13}-p_{1}-p_{2}-p_{12}-p_{23})\cdot\log(1+x') \\
+    &\qquad + (p_{2}+p_{3}+2(p_{1}+p_{3}+p_{12}+p_{23}))\cdot\log(1-x') \\
+    &\qquad + (3-p_{0}-p_{13}-p_{3}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\
+    &\qquad + (p_{2}+p_{3}+2(p_{12}+p_{23}+p_{123}))\cdot\log(1-y') \\
+    &\qquad + (1-p_{13})\cdot\log(1+w') \\
+    &\qquad + p_{13}\cdot\log(1-w').
+\end{align*}
+To make matters further compact, let
+$$
+a_{1} = 3-p_{0}-p_{13}-p_{1}-p_{2}-p_{12}-p_{23},
 $$
 $$
-    a_{1} = p_1+p_2+p_{12}+p_{23},
+a_{2} = p_{2}+p_{3}+2(p_{1}+p_{3}+p_{12}+p_{23}),
 $$
 $$
-    a_{2} = p_3+p_{12}+p_{23}+p_{123},
+a_{3} = 3-p_{0}-p_{13}-p_{3}-p_{12}-p_{23}-p_{123},
 $$
 $$
-    a_{3} = p_1+p_3+p_{12}+p_{23},
+a_{4} = p_{2}+p_{3}+2(p_{12}+p_{23}+p_{123}),
+$$
+where
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
+    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + a_{1}\cdot\log(1+x') \\
+    &\qquad + a_{2}\cdot\log(1-x') \\
+    &\qquad + a_{3}\cdot\log(1+y') \\
+    &\qquad + a_{4}\cdot\log(1-y') \\
+    &\qquad + (1-p_{13})\cdot\log(1+w') \\
+    &\qquad + p_{13}\cdot\log(1-w').
+\end{align*}
+Applying Gibbs' inequality to the first two summands, and maximizing over the remaining terms yields
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
+    &\le      p_{0}  \cdot\log(p_{0}) \\ 
+    &\qquad + p_{13} \cdot\log(p_{13}) \\
+    &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
+    &\qquad + a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
+    &\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
+    &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
+    &\qquad + (1-p_{13})\cdot\log(2(1-p_{13})) \\
+    &\qquad + p_{13}\cdot\log(2p_{13}).
+    &\qquad := L^{p,1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
+\end{align*}
+
+The other two possibilties for the likelihood of the ancestral state split $\{2\}$ for the site split $\{1,3\}$ admit similar simplifications.
+Call
+$$
+a_{1}' = a_{1}-2p_{13},
 $$
 $$
-    a_{4} = p_2+p_{12}+p_{23}+p_{123}.
+a_{2}' = a_{2}+2p_{13},
 $$
-Then the full likelihood is bounded above by one of three functions:
+$$
+a_{3}' = a_{3}-2p_{13},
+$$
+$$
+a_{4}' = a_{4}+2p_{13}.
+$$
+Then, for these cases,
 \begin{align*}
-    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\}) \\
-    &\qquad \le (3-a_{1})\log(1+x')+(a_{1}+a_{3})\log(1-x')\\
-    &\qquad \qquad +(3-a_{2})\log(1+y')+(a_{2}+a_{4})\log(1-y')\\
-    &\qquad \qquad +(1-a_{0})\log(1+w')+a_{0}\log(1-w')+p_{0}\log(4)\\
-    &\qquad \le (3-a_{1})\log\frac{2(3-a_{1})}{3+a_{3}}+(a_{1}+a_{3})\log\frac{2(a_{1}+a_{3})}{3+a_{3}}\\
-    &\qquad \qquad +(3-a_{2})\log\frac{2(3-a_{2})}{3+a_{4}}+(a_{2}+a_{4})\log\frac{2(a_{2}+a_{4})}{3+a_{4}}\\
-    &\qquad \qquad +(1-a_{0})\log(1+w')+a_{0}\log(1-w')+p_{0}\log(4)\\
-    &\qquad \triangleq L^{p,1}_{\tau_1,\tau_1}(\mathbf{a}_{xy});
+    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
+    &\le      p_{0}  \cdot\log(p_{0}) \\ 
+    &\qquad + p_{13} \cdot\log(p_{13}) \\
+    &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\
+    &\qquad + a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
+    &\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
+    &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
+    &\qquad + \log(2),
+    &\qquad := L^{p,2}_{\tau_1,\tau_1}(\mathbf{p}_{xy});
 \end{align*}
 \begin{align*}
-    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\}) \\
-    &\qquad \le (3-a_{1}-2a_{0})\log(1+x')+(a_{1}+a_{3}+2a_{0})\log(1-x')\\
-    &\qquad \qquad +(3-a_{2})\log(1+y')+(a_{2}+a_{4})\log(1-y')+\log(1+w')+p_{0}\log(4)\\
-    &\qquad \le (3-a_{1}-2a_{0})\log\frac{2(3-a_{1}-2a_{0})}{3+a_{3}}+(a_{1}+a_{3}+2a_{0})\log\frac{2(a_{1}+a_{3}+2a_{0})}{3+a_{3}}\\
-    &\qquad \qquad +(3-a_{2})\log\frac{2(3-a_{2})}{3+a_{4}}+(a_{2}+a_{4})\log\frac{2(a_{2}+a_{4})}{3+a_{4}}\\
-    &\qquad \qquad +\log(2)+p_{0}\log(4)\\
-    &\qquad \triangleq L^{p,2}_{\tau_1,\tau_1}(\mathbf{a}_{xy});
+    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
+    &\le      p_{0}  \cdot\log(p_{0}) \\ 
+    &\qquad + p_{13} \cdot\log(p_{13}) \\
+    &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
+    &\qquad + a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
+    &\qquad + a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'} \\
+    &\qquad + a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'} \\
+    &\qquad + \log(2).
+    &\qquad := L^{p,3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
 \end{align*}
-\begin{align*}
-    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\}) \\
-    &\qquad \le (3-a_{1})\log(1+x')+(a_{1}+a_{3})\log(1-x')\\
-    &\qquad \qquad +(3-a_{2}-2a_{0})\log(1+y')+(a_{2}+a_{4})+2a_{0}\log(1-y')+\log(1+w')+p_{0}\log(4)\\
-    &\qquad \le (3-a_{1})\log\frac{2(3-a_{1})}{3+a_{3}}+(a_{1}+a_{3})\log\frac{2(a_{1}+a_{3})}{3+a_{3}}\\
-    &\qquad \qquad +(3-a_{2}-2a_{0})\log\frac{2(3-a_{2}-2a_{0})}{3+a_{4}}+(a_{2}+a_{4}+2a_{0})\log\frac{2(a_{2}+a_{4}+2a_{0})}{3+a_{4}}\\
-    &\qquad \qquad +\log(2)+p_{0}\log(4)\\
-    &\qquad \triangleq L^{p,3}_{\tau_1,\tau_1}(\mathbf{a}_{xy}).
-\end{align*}
+All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.
 
@@
 
-We construct a similar lower bound on the Felsenstein partial likelihood, though here we replace all $(1+w)$ terms with $(1-w)$.
-Let
+We construct a similar lower bound on the Felsenstein partial likelihood.
+Replace all $(1+w)$ terms with $(1-w)$ and focus just on the partial likelihood.
+Then if $(1+x)(1-y) > (1-x)(1+y)$,
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    &=      + p_{0}  \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
+    &\qquad + p_{1}  \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
+    &\qquad + p_{2}  \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
+    &\qquad + p_{3}  \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
+    &\qquad + p_{12} \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
+    &\qquad + p_{123}\cdot\log((1+x')^2   (1-w')(1+y')(1-y'))\\
+    &\qquad + p_{13} \cdot\log((1+x')^2   (1-w')(1-y')^2) \\
+    &\qquad + p_{23} \cdot\log((1+x')^2   (1-w')(1-y')^2),
+\end{align*}
+otherwise
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    &=      + p_{0}  \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
+    &\qquad + p_{1}  \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
+    &\qquad + p_{2}  \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
+    &\qquad + p_{3}  \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
+    &\qquad + p_{12} \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
+    &\qquad + p_{123}\cdot\log((1+x')(1-x')(1-w')(1+y')^2)\\
+    &\qquad + p_{13} \cdot\log((1-x')^2    (1-w')(1+y')^2) \\
+    &\qquad + p_{23} \cdot\log((1-x')^2    (1-w')(1+y')^2).
+\end{align*}
+Letting
 $$
-b_1 = p_1+p_2+p_3+p_{123}
+b_{1} = p_{1}+p_{2}+p_{3}+p_{123}
 $$
-and
-$$
-b_2 = p_{13}+p_{23}.
-$$
-
-\textbf{TODO}: ancestral states where these likelihoods occur
-Since the maxima for pairs of lower bounds above each have a symmetry in $x$ and $y$, they are bounded below by one of
+yields
 \begin{align*}
-    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\}) \\
-    &\qquad \ge 2\log(1+x')+(2-b_1-2b_2)\log(1+y')+(b_1+2b_2)\log(1-y')+\log(1-w'),
+    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    &=      + 2\cdot\log(1+x') \\
+    &\qquad + (2-b_{1})  \cdot\log(1+y') \\
+    &\qquad + b_{1}      \cdot\log(1-y') \\
+    &\qquad + \log(1-w')
 \end{align*}
+for the first lower bound, but you'll notice that, upon maximizing, we obtain the same value for both cases, namely
 \begin{align*}
-    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\}) \\
-    &\qquad \ge (2-2b_2)\log(1+x')+2b_2\log(1-x') + (2-b_1)\log(1+y')+b_1\log(1-y')+\log(1-w')
+    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    &=      + 2\cdot\log(2) \\
+    &\qquad + (2-b_{1})  \cdot\log(2-b_{1}) \\
+    &\qquad + b_{1}      \cdot\log(b_{1})
 \end{align*}
-or
+maximized at $\{1,1-b_{1},0\}$, meaning
 \begin{align*}
-    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\}) \\
-    &\qquad \ge (2-b_2)\log(1+x')+b_2\log(1-x') + (2-b_1-b_2)\log(1+y')+(b_1+b_2)\log(1-y')+\log(1-w')
+    L^{p,1}_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b_{1},0\}) + 2\cdot\log(2) + (2-b_{1})\cdot\log(2-b_{1})+b_{1}\cdot\log(b_{1}) \\
+                                             &= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1-b_{1},1,0\}) + 2\cdot\log(2) + (2-b_{1})\cdot\log(2-b_{1})+b_{1}\cdot\log(b_{1}).
 \end{align*}
-where $b_1$ and $b_2$ are also functions of the true generating branch parameters $\{x,y\}$.
-Taking the maximum of each yields
-$$
-    L^{p,1}_{\tau_1,\tau_2}(b_1, b_2) \triangleq 2\log(2)+(2-b_1-2b_2)\log(2-b_1-2b_2)+(b_1+2b_2)\log(b_1+2b_2),
-$$
-$$
-    L^{p,2}_{\tau_1,\tau_2}(b_1, b_2) \triangleq (2-2b_2)\log(2-2b_2)+2b_2\log(2b_2) + (2-b_1)\log(2-b_1)+b_1\log(b_1)
-$$
-and
-$$
-    L^{p,3}_{\tau_1,\tau_2}(b_1, b_2) \triangleq (2-b_2)\log(2-b_2)+b_2\log(b_2) + (2-b_1-b_2)\log(2-b_1-b_2)+(b_1+b_2)\log(b_1+b_2).
-$$
-
-\textbf{Claim to prove}:
-$$
-L^{p,1}_{\tau_1,\tau_2}(b_1, b_2) \ge L^{p,2}_{\tau_1,\tau_2}(b_1, b_2)
-$$
-and
-$$
-L^{p,1}_{\tau_1,\tau_2}(b_1, b_2) \ge L^{p,3}_{\tau_1,\tau_2}(b_1, b_2).
-$$
-I think we can use information theory results to show this.
-It is definitely true---will work on it.
-
-The values of $(x', y', w')$ maximizing $L^{p,1}_{\tau_1,\tau_2}(b_1, b_2)$ are $(1, 1-b_1-2b_2,0)$.
 

### 2017-04-01 joint_inf.tex
Cleaning out notes, minor tweaks.

@@
 Recently \cite{Neher2017} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences.
-This is attractive from a computational perspective: joint ML inference can proceed according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
-% I would say classical maximum likelihood is closer to EM, no? At some point in EM we'd have to marginalize/take the expectation, which joint ML doesn't do. Curious to talk more about it though...
-This latter step is much simpler and more computationally efficient when conditioning on ancestral sequences compared to the case in which ancestral sequences are marginalized out.
+This is attractive from a computational perspective: such joint ML inference can proceed according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
+This latter step is simpler and more computationally efficient when conditioning on ancestral sequences compared to the case in which ancestral sequences are marginalized out.
 
@@
 Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
-%EM Commented out below because it's not right-- fullAlignment is not the set of assignments of characters to tips of the tree, it's a single such.
-%Let $\fullAlignment$ be the set of assignments of characters to tips of our tree, and $\fullAncestralStates$ the set of assignments of characters to the internal nodes of the tree.
 Assume we observe $\nCols$ samples of character data (i.e.\ an alignment with $\nCols$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
@@
 
-One tantalizing approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the full likelihood in \eqref{eq:full_likelihood}.
+One tantalizing approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
 We can perform this by computing the profile likelihood
@@
 Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
-%EM Not sure how tau induces an ordered set of tip labels-- are you saying that the tree defines the labels or something about its structure?
-Following Section 8.6 of \cite{Semple2003-em}, we define an ordered set of tip labels $\{1,\ldots,\nSiteRows\}$ on $\tau$ across sites.
+Following Section 8.6 of \cite{Semple2003-em}, we take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
 For the binary symmetric model, a realization of $\alignmentColumn_i$ can be described as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
-Indeed, we can start by including those tips in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
-Then, because the probability of observing a particular site split is equivalent to the probability of its complement under the binary symmetric model, we can take the complement if needed to exclude tip $\nSiteRows$ from $\siteSplit$.
-Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ where $\mathcal{P}$ denotes the powerset.
-%EM inducing again.
-For topology $\tau$, we also define an ordered set of internal nodes $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ with a similar subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ describing a realization of $\ancestralStateColumn_i$.
+Indeed, we can start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
+Then, because the probability of observing a particular site split is equivalent to the probability of its complement under the binary symmetric model, we can take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.
+Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ where $\mathcal{P}$ denotes the power set.
+For topology $\tau$, we also define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization of $\ancestralStateColumn_i$.
 The probability of observing an ancestral state split, conditional on a site split, will not be invariant to taking its complement, and thus the need for enumerating the entire set of internal nodes.
@@
 
-With this split formulation, define the mappings from patterns to splits as described above
+With this split formulation, define mappings from patterns to splits as described above
 $$
-\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet), \ \ancestralToSplit:\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet)
+\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet), \ \ancestralToSplit:\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet).
 $$
-% the following seemed not necessary and a little confusing (what are j and k?)
-% so that, for
-% $$
-% \patternToSplit(\alignmentColumn_i)=\siteSplit_j, \ \ancestralToSplit(\ancestralStateColumn_i)=\ancestralSplit_k.
-% $$
 Given a site-pattern-valued random variable $\alignmentColumnRV$ and an ancestral-state-valued random variable $\ancestralStateColumnRV$, define random variables
@@
 for some $j$ and $k$.
-Taking the maximum of the probability of observing a particular column $i$ yields
+Taking the maximum across internal states yields
 \begin{align*}
@@
 %das: I think they do have to be disjoint. The partition is into "most likely" and "not most likely," which is disjoint and exhausts the set. Am I missing something?
+%EM Perhaps we're talking about partition in different senses. I was saying that eta_i can intersect with another eta_j. Let's chat IRL on Monday.
 Given $(\tau, t)$ there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
@@
 L_\nCols'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) \\
-%EM I guess the ith component of H is h_i?
-    %das: Not sure what the question/comment is here.
-                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
+                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
 %EM Indexing broken here: we are taking a product over i, but i doesn't appear in the expression.
@@
 I can try to think of how to make this more clear, but I thought it was a pretty key step to be able to pass to site patterns, and the reason for the machinery involved in defining ancestral state partitions.
+%EM My suggestion is to use \psi(y_i) until you switch to a product indexed by the j's.
                              &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) \\

### 2017-04-25 joint_inf.tex
fiddling
update notation

@@
 We can now write the $i$th factor of \eqref{eq:full_likelihood} as
-$$
-P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)=P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)
-$$
+\begin{align*}
+    P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
+    &= P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) \\
+\end{align*}
 for some $j$ and $k$.
@@
 %EM Perhaps we're talking about partition in different senses. I was saying that eta_i can intersect with another eta_j. Let's chat IRL on Monday.
-Given $(\tau, t)$ there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
+Given $(\tau, t)$ there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
 \begin{align*}
-\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi \mid \siteSplitRV=\siteSplit_j, \tau, t).
+\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
 \end{align*}
-
+This allows us to write the likelihood in \eqref{eq:profile_likelihood} as a product over site patterns as opposed to sites, i.e.,
 \begin{align}
@@
 %EM My suggestion is to use \psi(y_i) until you switch to a product indexed by the j's.
-                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) \\
-                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
-%EM As I fixed above and described in person, \ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) is not what you want here, because it will inflate the value by |\ancestralSplitPartition_j(\tau, t)| .
-This also appears below.
-                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
-                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
+                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralToSplit(\ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
+                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralToSplit(\ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} P(\ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
+                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
 \end{align}
@@
 %EM My long term suggestion is for you to put this in example in the main text and the above in the appendix.
-Consider a fixed, binary four-tip tree $\tau$.
+\begin{figure}
+    \centering
+    \includegraphics[width=.45\textwidth]{unrooted_four_taxa}
+    \caption{Four taxa tree}
+\end{figure}
+
+Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}.
 The set of all possible character assignments can be described as
@@
 %EM I dropped the \ancestralSplitPartition(\siteSplit, \tau, t) notation above. You can revert it if you like, but I prefer the bold eta notation as an ordered list of sets.
-A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; define $\ancestralSplitPartition(\siteSplit, \tau, t)$ as the partition of $\mathcal{P}(\ancestralSplitSet)$ into those most likely ancestral splits for that particular site split, topology and set of branch lengths.
+A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; for the $j$th site split, define $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ as the set of most likely ancestral splits for that particular site split, topology and set of branch lengths.
 As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
-The probabilities of observing ancestral splits for site split $\emptyset$ are
+The probabilities of observing ancestral splits for $\siteSplit_j=\emptyset$ are
 $$
@@
 %das: I think it's wrong, but in the other direction: two character changes to get to the differing internal state, then one between the two internal states.
+Another set of eyes might make a liar out of me, though.
 P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
@@
 %EM Have you told us in this example what the labeled topology is?
-%das: Nope! I wanted to be simple, but it seems maybe too simple. Maybe I should move the Felsenstein/Farris bit to this example?
 %Erick stopped here.
@@
 $$
-Here, the partition into most likely ancestral states will proceed as
+The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_j(\tau, p)=\{\emptyset\}$.
+Then, taking $\xi_j\in\ancestralSplitPartition_j(\tau, p)$ we have
 $$
-P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p) \mid \siteSplitRV=\emptyset, \tau, t) =
-(1-p)^5
+P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\emptyset, \tau, t) =
+(1-p)^5.
 $$
-where the partition is the single element $\ancestralSplitPartition_j(\tau, p)=\emptyset$.
-For the site split $\{1,3\}$ we have
+For $\siteSplit_j=\{1,3\}$ we have
 $$
@@
 P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
-p^3(1-p)^2,
+p^3(1-p)^2.
 $$
-where
+In this case, the set of most likely ancestral states is $\ancestralSplitPartition_j(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_j\in\ancestralSplitPartition_j(\tau, p)$,
 $$
-P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p) \mid \siteSplitRV=\{1,3\}, \tau, t) =
-p^2(1-p)^3
+P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\{1,3\}, \tau, t) =
+p^2(1-p)^3.
 $$
-with the most likely partition $\ancestralSplitPartition_j(\tau, p)=\{\emptyset,\{1,2\}\}.$
 
@@
 $$
-L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
+L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}.
 $$
@@
     \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau,t),\tau,t)
-        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \tau, t) \\
+        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
         &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
-            \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j , \tau, t)]
+            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
 \end{align}
@@
 \begin{equation}
-\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j , \tau, t)].
+\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)].
 \label{eq:site_pattern_profile_likelihood_mean}
@@
 $$
-\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplit_j, \tau, t)
+\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
 $$
-so that \eqref{eq:site_pattern_profile_likelihood_mean} is written as
+so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
 \begin{equation}
@@
 \label{eq:inconsistency_inequality}
-\ell_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau',t_1)},\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau^*,t_2)},\tau^*,\hat{t}_2)
+\ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau',t_1),\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau^*,t_2),\tau^*,\hat{t}_2)
 \end{equation}
-for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\widehat{\fullAncestralSplitPartitions(\tau',t_1)}\}$ and $\{\hat{t}_2,\widehat{\fullAncestralSplitPartitions(\tau^*,t_2)}\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
+for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}(\tau',t_1)\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}(\tau^*,t_2)\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
 
@@
     \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
-    + \ell^p_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau,t)},\tau,\hat{t})
++ \ell^p_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau,t),\tau,\hat{t})
 $$
@@
 $$
-\{\hat{t},\widehat{\fullAncestralSplitPartitions(\tau,t)}\} = \arg\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
+\{\hat{t},\hat{\fullAncestralSplitPartitions}(\tau,t)\} = \arg\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
 $$
@@
     \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
-    + \ell^p_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau,t)},\tau,\hat{t}).
++ \ell^p_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau,t),\tau,\hat{t}).
 $$
@@
     \hline
-    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
+    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
     \hline
@@
     $\emptyset$&
-    $(1+x)(1-x)(1+w)(1+y)^2$\\
+    $(1+x)^2   (1+w)(1+y)(1-y)$\\
      $\{3\}$    &
     $\emptyset$&
-    $(1+x)^2   (1+w)(1+y)(1-y)$\\
+    $(1+x)(1-x)(1+w)(1+y)^2$\\
     $\{1,2\}$  &
@@
     &\qquad + p_{1}  \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
-    &\qquad + p_{2}  \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
-    &\qquad + p_{3}  \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
+    &\qquad + p_{2}  \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
+    &\qquad + p_{3}  \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
     &\qquad + p_{12} \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
@@
     &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
-    &\qquad + (2-p_{1}-p_{2}-p_{12}-p_{23})\cdot\log(1+x') \\
-    &\qquad + (p_{1}+p_{2}+p_{12}+p_{23})\cdot\log(1-x') \\
-    &\qquad + (2-p_{3}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\
-    &\qquad + (p_{3}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
+    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x') \\
+    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
+    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\
+    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
     &\qquad + (1-p_{13})\cdot\log(1+w') \\
@@
     &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
-    &\qquad + (3-p_{0}-p_{13}-p_{1}-p_{2}-p_{12}-p_{23})\cdot\log(1+x') \\
-    &\qquad + (p_{2}+p_{3}+2(p_{1}+p_{3}+p_{12}+p_{23}))\cdot\log(1-x') \\
-    &\qquad + (3-p_{0}-p_{13}-p_{3}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\
-    &\qquad + (p_{2}+p_{3}+2(p_{12}+p_{23}+p_{123}))\cdot\log(1-y') \\
+    &\qquad + (3-p_{0}-p_{13}-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x') \\
+    &\qquad + 2(p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
+    &\qquad + (3-p_{0}-p_{13}-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\
+    &\qquad + 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
     &\qquad + (1-p_{13})\cdot\log(1+w') \\
@@
 $$
-a_{1} = 3-p_{0}-p_{13}-p_{1}-p_{2}-p_{12}-p_{23},
+a_{1} = 3-p_{0}-p_{13}-p_{1}-p_{3}-p_{12}-p_{23},
 $$
 $$
-a_{2} = p_{2}+p_{3}+2(p_{1}+p_{3}+p_{12}+p_{23}),
+a_{2} = 2(p_{1}+p_{3}+p_{12}+p_{23}),
 $$
 $$
-a_{3} = 3-p_{0}-p_{13}-p_{3}-p_{12}-p_{23}-p_{123},
+a_{3} = 3-p_{0}-p_{13}-p_{2}-p_{12}-p_{23}-p_{123},
 $$
 $$
-a_{4} = p_{2}+p_{3}+2(p_{12}+p_{23}+p_{123}),
+a_{4} = 2(p_{2}+p_{12}+p_{23}+p_{123}),
 $$
@@
 \end{align*}
-Applying Gibbs' inequality to the first two summands, and maximizing over the remaining terms yields
+Applying Gibbs' inequality to the first two summands and maximizing over the remaining terms yields
 \begin{align*}
@@
     &\qquad + (1-p_{13})\cdot\log(2(1-p_{13})) \\
-    &\qquad + p_{13}\cdot\log(2p_{13}).
+    &\qquad + p_{13}\cdot\log(2p_{13}) \\
     &\qquad := L^{p,1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
@@
     &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
-    &\qquad + \log(2),
+    &\qquad + \log(2) \\
     &\qquad := L^{p,2}_{\tau_1,\tau_1}(\mathbf{p}_{xy});
@@
     &\qquad + a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'} \\
-    &\qquad + \log(2).
+    &\qquad + \log(2) \\
     &\qquad := L^{p,3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
@@
 \begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
-    &=      + p_{0}  \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
+    \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    &=      + p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
     &\qquad + p_{1}  \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
@@
 \begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
-    &=      + p_{0}  \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
+    \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    &=      + p_{0}  \cdot\log((1+x')^2    (1+w')(1+y')^2) \\
     &\qquad + p_{1}  \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
@@
 \begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
     &=      + 2\cdot\log(1+x') \\
@@
     &\qquad + b_{1}      \cdot\log(1-y') \\
-    &\qquad + \log(1-w')
+    &\qquad + p_{0}\cdot\log(1+w') \\
+    &\qquad + (1-p_{0})\cdot\log(1-w')
 \end{align*}
-for the first lower bound, but you'll notice that, upon maximizing, we obtain the same value for both cases, namely
+for the first lower bound, but, upon maximizing, we obtain the same value for both cases, namely
 \begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
     &=      + 2\cdot\log(2) \\
     &\qquad + (2-b_{1})  \cdot\log(2-b_{1}) \\
-    &\qquad + b_{1}      \cdot\log(b_{1})
+    &\qquad + b_{1}      \cdot\log(b_{1}) \\
+    &\qquad + p_{0}\cdot\log(2p_{0}) \\
+    &\qquad + (1-p_{0})\cdot\log(2(1-p_{0}))
 \end{align*}
-maximized at $\{1,1-b_{1},0\}$, meaning
-\begin{align*}
-    L^{p,1}_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b_{1},0\}) + 2\cdot\log(2) + (2-b_{1})\cdot\log(2-b_{1})+b_{1}\cdot\log(b_{1}) \\
-                                             &= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1-b_{1},1,0\}) + 2\cdot\log(2) + (2-b_{1})\cdot\log(2-b_{1})+b_{1}\cdot\log(b_{1}).
-\end{align*}
+maximized at $\{1,1-b_{1},1-2p_{0}\}$.
+
+...
+
+$$
+    L^{p,1}_{\tau_1,\tau_2}(\mathbf{p}_{xy}) := \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b_{1},1-2p_{0}\}) + 
+$$
 

### 2017-04-25 joint_inf.tex
Pass through to start of second section.

@@
 
-We will be considering phylogenetic likelihood on unrooted phylogenetic trees.
+We will be considering the standard phylogenetic likelihood on unrooted phylogenetic trees (cite Felsenstein book).
 We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
@@
 \label{eq:full_likelihood}
+%EM given that the H is on the lhs of the conditioning, does it make sense to have it on the rhs of the semicolon? I don't know the convention.
 L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
@@
 
-Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
-Following Section 8.6 of \cite{Semple2003-em}, we take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
-For the binary symmetric model, a realization of $\alignmentColumn_i$ can be described as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
+Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation (we follow Section 8.6 of \cite{Semple2003-em}).
+Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
+For likelihood calculation under the binary symmetric model, a given $\alignmentColumn_i$ can be described as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
 Indeed, we can start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
-Then, because the probability of observing a particular site split is equivalent to the probability of its complement under the binary symmetric model, we can take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.
+Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we can take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.
+
 Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ where $\mathcal{P}$ denotes the power set.
@@
 
-With this split formulation, define mappings from patterns to splits as described above
+With this split formulation, define mappings from patterns to splits as described above:
 $$
@@
 $$
+% Need to introduce \siteSplit_j and \ancestralSplit_k , no?
 We can now write the $i$th factor of \eqref{eq:full_likelihood} as
@@
     P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
-    &= P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) \\
+    &= P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)
 \end{align*}
@@
 \end{align*}
-%EM Just noticed that eta doesn't have to be a real partition, because they don't have to be disjoint, right?
-%das: I think they do have to be disjoint. The partition is into "most likely" and "not most likely," which is disjoint and exhausts the set. Am I missing something?
-%EM Perhaps we're talking about partition in different senses. I was saying that eta_i can intersect with another eta_j. Let's chat IRL on Monday.
 Given $(\tau, t)$ there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
@@
 \end{align*}
+Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
 This allows us to write the likelihood in \eqref{eq:profile_likelihood} as a product over site patterns as opposed to sites, i.e.,
@@
 \end{align}
+%EM think we should say "where the site split or its complement appears."?
 where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ appears.
@@
 \paragraph{Example}
-%EM My long term suggestion is for you to put this in example in the main text and the above in the appendix.
+%EM My long term suggestion is for you to put this in example in the main text and the above in the appendix. (But we should probably check with Vladimir.)
 \begin{figure}
@@
     \includegraphics[width=.45\textwidth]{unrooted_four_taxa}
-    \caption{Four taxa tree}
+    \caption{Four taxon tree}
+\label{fig:four-taxa-tree}
 \end{figure}
@@
 For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
-%EM I dropped the \ancestralSplitPartition(\siteSplit, \tau, t) notation above. You can revert it if you like, but I prefer the bold eta notation as an ordered list of sets.
 A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; for the $j$th site split, define $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ as the set of most likely ancestral splits for that particular site split, topology and set of branch lengths.
 As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
+%EM Would it be worth enumerating the site splits ahead of time? It seems a little funky to be calling them j each time.
 The probabilities of observing ancestral splits for $\siteSplit_j=\emptyset$ are
@@
 $$
-%EM Seems to me that this just requires one character change so would be p(1-p)^4.
-%das: I think it's wrong, but in the other direction: two character changes to get to the differing internal state, then one between the two internal states.
-Another set of eyes might make a liar out of me, though.
 P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
@@
 $$
-%EM Have you told us in this example what the labeled topology is?
-%Erick stopped here.
 P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
@@
 $$
-The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_j(\tau, p)=\{\emptyset\}$.
+Thus the set of most likely ancestral states contains a single element in this case, here $\ancestralSplitPartition_j(\tau, p)=\{\emptyset\}$.
 Then, taking $\xi_j\in\ancestralSplitPartition_j(\tau, p)$ we have
@@
 We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
+%EM As described IRL, a little strange to be redefining (9) with an extra variable.
+After reading the below, I think it makes most sense to just refer to (9) indirectly but make this a stand-alone definition.
 Define \eqref{eq:site_pattern_likelihood} as
@@
 \begin{equation}
+%EM did you mean to drop the eta from the first term of this equation?
 \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)].
@@
 $$
+%EM IMHO \ell^p is so standard that it's confusing to see it in another setting. Perhaps \tilde{\ell}?
 \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)

### 2017-04-25 joint_inf.tex
Quibbles.

@@
 \usepackage[colorlinks=true, allcolors=blue]{hyperref}
+\usepackage{showlabels}
 
@@
 \newcommand{\shannonDivergence}{D}
+
+\DeclareMathOperator*{\argmax}{argmax}
 
@@
 % $$
-% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
+% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \argmax_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
 % $$
@@
 $$
-(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
+(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
 $$
@@
 \label{eq:profile_likelihood_topology_bl}
-(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ L_\nCols'(\fullAlignment;\tau, t)
+(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\fullAlignment;\tau, t)
 \end{equation}
@@
 \label{eq:inconsistency_inequality}
+%EM Suggest \hat{\fullAncestralSplitPartitions}_1, etc.
 \ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau',t_1),\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau^*,t_2),\tau^*,\hat{t}_2)
@@
 for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}(\tau',t_1)\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}(\tau^*,t_2)\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
+By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any potential example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
+In the next section, we show that such examples exist.
 
@@
 \begin{equation}
+%EM Seems a little strange to give a new equation number to an equation that already exists with an assigned number.
 \label{eq:partial-likelihood}
@@
 $$
-\{\hat{t},\hat{\fullAncestralSplitPartitions}(\tau,t)\} = \arg\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
+\{\hat{t},\hat{\fullAncestralSplitPartitions}(\tau,t)\} = \argmax_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
 $$

### 2017-04-26 joint_inf.tex
Getting to just before the calculations.

@@
 % vvvv Remove these? I think they're obvious now, but maybe they can be made more explicit above?
+%EM I agree-- quite clear now!
 \textbf{Argument for lower bound}: This makes intuitive sense, though may be difficult to show in general.
@@
 Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
+%EM clarify that we aren't assuming that the x and y in t are the same as those appearing in t^*?
 Call $t=\{x,y,w\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ is the case where the bottom three branches all share the same parameter, the classical construction of this topology.
-The branch length parameters are such that the probability of a change in character along the top two branches is $p_x=1-2x$, with corresponding equalities for the other two branches.
+
+We will parameterize these trees \textbf{not with branch lengths}, but rather such that the probability of a substitution on a branch with parameter $x$ is $p_x=1-2x$.
+%EM (Scratching head) isn't p_x=1-2x backwards? Then p_x will take values in the range -1 to 1.
 Here, $\{p_x,p_y,p_w\}\in[0,1/2]^3$ and $\{x,y,w\}\in[0,1]^3$.
+This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
+Fidelities have nice algebraic properties, as can be seen via the Hadamard transform.
 
 \textbf{Justification for estimand:} It might be worth arguing why this form of the estimand $t$ is necessary.
+%EM definitely. This is a big TODO.
 I think the idea is that for either of these topologies, any maximum obtained will occur where the top two and bottom two branches are equal, though not necessarily the middle branch.

### 2017-04-28 appendix.tex
add appendix; address most issues

--- 2017-03-21 deprecated.tex
@@
-\section{Temporarily unused text I may want later to show more general results on the likelihood estimates}
-
-Some scratch work:
-$$
-a = b = (4-2x^2-2y^2)/8,
-$$
-$$
-c+2d = 1-xy^2, \ 2-c-2d = 1+xy^2
-$$
-$$
-H_{fe}(x,y; 1, 0, 1-c-2d) \propto (2-4xy^2+2x^2y^2)\log(1-2xy^2+x^2y^4) + (2+4xy^2+2x^2y^2)\log(1+2xy^2+x^2y^4)+(4-4x^2y^2)\log(1-x^2y^4)
-$$
-$$
-H_{fa}(x,y; x, y, y) \propto (2-4xy^2+2x^2y^2)\log(1-2xy^2+x^2y^4) + (2+4xy^2+2x^2y^2)\log(1+2xy^2+x^2y^4)+(4-4x^2y^2)\log(1-x^2y^4)
-$$
-$$
-\hat{L}_{fe} \propto 2\log(2) + (1+xy^2)\log(1+xy^2) + (1-xy^2)\log(1-xy^2)
-$$
-$$
-\hat{L}_{fa} \propto \log(2) + (4-2a)\log(2-a)+2a\log(a).
-$$
-
-\subsubsection{Refinements}
-
-It will be an undertaking, but if we want to approach this using the refinements we might get something pretty nifty.
-Everything will be quite messy, but really the entire approach boils down to a series of nested inequalities in two variables.
-We may need a program that can reduce symbolic equations, but this will at least get us somewhere that isn't a lot of loose approximations---initial results using numerical optimization show the region in~\ref{fig:inconsistency-farris} is conservative, and very much so.
-See Fig.~\ref{fig:numerical-inconsistency-farris}.
-
-Perhaps something along the lines of a more general results would be interesting?
-Like, could we calculate the measure/area of inconsistency using cooler techniques?
-Could we do this for a general topology?
-Would that even be interesting since we've already found a counterexample?
-
-It also seems that, in the $\{x, y\}=\{.25, .25\}$ example, the lower bound is tight while the upper bound is looser.
-This surprises me since we replaced a lot in the likelihood, but this might indicate that the $H$ term is actually the one that varies more than the $L$ term.
-I guess this is intuitive: the difference between the lower and upper bound on $L$ is $\log2\approx0.69$.
-Maybe getting a tighter upper bound will be useful here?
-
-\begin{figure}
-\centering
-\includegraphics[width=.9\textwidth]{ineqs-num-far-gen}
-\caption{Regions of inconsistency---numerically obtained}
-\label{fig:numerical-inconsistency-farris}
-\end{figure}
-
-\subsubsection{Tighter upper bound}
-
-Call
-$$
-e = p(0001) + p(0100) = \frac{1+x^2-y^2-x^2y^2}{4},
-$$
-$$
-f = p(0010) + p(1000) = \frac{1-x^2+y^2-x^2y^2}{4}.
-$$
-Then some algebra and standard $\log$ bounds can be used to get an upper bound for the full Farris likelihood $L^{(i)}$ as
-$$
-(2-a+e)\log(2-\frac{2a}{2+e}) + a\log\frac{2a}{2+e} + (2-b+f)\log(2-\frac{2b}{2+f}) + a\log\frac{2b}{2+f}.
-$$
-In our example of $\{x,y\}=\{.25,.25\}$ this bound is $a$ compared to $-4.25313560433$ before.
-
-\subsubsection{A counterexample in branch lengths}
-
-Let $(\tau^*, t^*)=(\tau_1, \{x,y,x\})$.
-We claim there exist $\{x,y\}$ and $\{x, \tilde{y}\}$ such that
-$$
-L'(\tau^*, t^*; \tau^*, \{x, \tilde{y}\}) > L'(\tau^*, t^*; \tau^*, t^*).
-$$
-The model is given by values in the Table.
-Set $x=0$ to obtain
-$$
-L'(\tau^*, t^*; \tau^*, t^*) \propto 8\log(1+y)
-$$
-and
-$$
-L'(\tau^*, t^*; \tau^*, \{0, \tilde{y}\}) \propto 8\log(1+\tilde{y}).
-$$
-Therefore, if $\tilde{y} > y$ we have an inconsistency.
-
-\begin{table}
-\centering
-\begin{tabular}{|l|l|l|}
-    \hline
-$s_j$   &$P(s_j|\tau_1,t)$&$\max_{z_j} \ P(s_j,z_j|\tau_1,t)$\\
-    \hline
-0000&$1+2x^2+4x^2y+x^4$&$(1+x)^4(1+y)$\\
-0001&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
-0010&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
-0100&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
-1000&$1-x^4$&$(1+x)^3(1+y)(1-x)$\\
-0011&$1+2x^2-4x^2y+x^4$&max$\{(1+x)^2(1+y)(1-x)^2,(1+x)^4(1-y)\}$\\
-0101&$1-2x^2+x^4$&$(1+x)^2(1+y)(1-x)^2$\\
-1001&$1-2x^2+x^4$&$(1+x)^2(1+y)(1-x)^2$\\
-    \hline
-\end{tabular}    
-\caption{Counterexample 1.}
-\label{tab:sitepatprob_case1}
-\end{table}
-
-\subsubsection{A counterexample in topologies}
-
-Let $(\tau^*, t^*)=(\tau_1, \{x,y,y\})$.
-We claim there exist $\{x,y\}$ such that
-$$
-L'(\tau^*, t^*; \tau_2, t^*) > L'(\tau^*, t^*; \tau^*, t^*).
-$$
-Set $y=0$ to obtain parameters given in Table.
-We have
-$$
-L'(\tau^*, t^*; \tau^*, t^*) \propto 4(1+x^2)\log(1+x^2)+4(1-x^2)\log(1-x^2)+4(1+x^2)\log((1+x)^2)+4(1-x^2)\log((1+x)(1-x))
-$$
-and
-$$
-L'(\tau^*, t^*; \tau_2, t^*) \propto 3(1+x^2)\log((1+x)^2)+3(1-x^2)\log((1+x)^2)+(1+x^2)\log((1+x)(1-x))+(1-x^2)\log((1+x)(1-x)).
-$$
-The difference being greater than zero, i.e.,
-$$
-L'(\tau^*, t^*; \tau_2, t^*) - L'(\tau^*, t^*; \tau^*, t^*) > 0,
-$$
-simplifies to
-$$
-4x^2(\log(1-x^2)-\log(1+x^2)+\log(1-x)-\log(1+x)) - 2(2\log(1+x^2)+2\log(1-x^2)-\log(1+x)+\log(1-x)) > 0.
-$$
-Let $0 < x < 1/\sqrt{2}$ so that the above inequality becomes
-$$
--\log(1-x^2)-3\log(1+x^2) > 0
-$$
-or
-$$
-(1+x^2)^3(1-x^2) < 1.
-$$
-Applying $x<1/\sqrt{2}$ shows if
-$$
-x < \sqrt{-1+2^{1/3}} \approx 0.5098
-$$
-we have an inconsistency.
-
-\begin{table}
-\centering
-\begin{tabular}{|l|l|l||l|l|}
-    \hline
-$s_j$   &$P(s_j|\tau_1,t)$&$\max_{z_j} \ P(s_j,z_j|\tau_1,t)$&$P(s_j|\tau_2,t)$&$\max_{z_j} \ P(s_j,z_j|\tau_2,t)$\\
-    \hline
-0000&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)^2$\\
-0001&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)^2$\\
-0010&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)^2$\\
-0100&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)^2$\\
-1000&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)^2$\\
-0011&    $1+x^2$ & $(1+x)^2$ & 1 & $(1+x)(1-x)$\\
-0101&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)^2$\\
-1001&    $1-x^2$ & $(1+x)(1-x)$ & 1 & $(1+x)(1-x)$\\
-    \hline
-\end{tabular}    
-\caption{Counterexample 2.}
-\label{tab:sitepatprob_case2}
-\end{table}
-
-\subsubsection{More fleshed out counterexamples}
-
-The above counterexamples are highly pathological, but they show there exist whole spaces of parameters yielding inconsistencies.
-It seems as long as $x$ and $y$ are ``small enough'' we see inconsistency.
-
-\subsection{start}
-
-We can calculate the probability of a particular site pattern under this topology from the probability that all character states are equal as
-$$
-P(0000|\tau^*,t) = \frac{1}{8}(1+x^2+z^2+4xyz+x^2z^2)
-$$
-where the probability of any site pattern will be the same as above with the corresponding branch length parameter replaced with its negative.
-For example, if we have the pattern $0001$, since the last branch has length $z$ we replace it with $-z$ and obtain
-$$
-P(0001|\tau^*,t) = \frac{1}{8}(1+x^2-z^2-x^2z^2).
-$$
-All possible site pattern probabilities are given in Table~\ref{tab:sitepatprob}.
-All likelihoods and their maximal values are given in Table~\ref{tab:likelihoods}.
-
-It can be shown (I'll do it later) that for our $t^*$ the right-hand term above is $-16\log 8$.
-Moreover, for $t=\{x,y,z\}$
-\begin{equation}
-\sum_{s} \log P(s | \tau^*, t) + \log P(\xi | s, \tau^*, t) = -16\log 8 + f(x,y,z) + g(x,y,z)
-\end{equation}
-where
-\begin{align}
-f(x,y,z) &= 4\log(1+x)+4\log(1-x)+4\log(1+z)+4\log(1-z)+\\
-& 3\log(1+x^2)+3\log(1+z^2)+\\
-& \log((1+x^2)(1+z^2)+4xyz) + \log((1+x^2)(1+z^2)-4xyz).
-\end{align}
-and
-\begin{align}
-g(x,y,z) &= 10\log(1+x)+10\log(1+z)+4\log(1-x)+4\log(1-z)+7\log(1+y)+\\
-& \left\{ 
-  \begin{array}{l}
-  2\log(1+x)+2\log(1-z)+\log(1+y)\\
-  2\log(1+x)+2\log(1+z)+\log(1-y)\\
-  2\log(1-x)+2\log(1-z)+\log(1+y)\\
-  \end{array} \right.
-\end{align}
-whose component will be chosen based on whichever yields the maximum value for $\log L''$.
-
-\subsection{Two fixed topologies}
-
-\subsection{Special case: $x=z=0$}
-
-Consider now setting all data-generating branch length parameters $t^*$ to zero; this yields equal probabilities of $1/8$ for each site pattern.
-By the inequality of Gibbs we have
-$$
-\sum_{s} P(s|\tau^*,t^*)\log P(s|\tau^*,t^*) \geq \sum_{s} P(s|\tau^*,t^*)\log P(s|\tau^*,t)
-$$
-summing over all site patterns, which is equivalent to
-$$
-\sum_{s} \log P(s|\tau^*,t^*) \geq \sum_{s} \log P(s|\tau^*,t).
-$$
-We must show, then, that there exist $t\neq t^*$ such that
-$$
-\max_{\xi_s} \ \sum_{s} \log P(s | \tau^*, t) + \log P(\xi_s | s, \tau^*, t) > \max_{\xi_s} \sum_{s} \log P(s | \tau^*, t^*) + \log P(\xi_s | s, \tau^*, t^*).
-$$
-Assume for $t$ that $x=z=0$ but $y\neq 0$.
-Clearly site pattern probabilities are still all equal, and the first summand of each side of the inequality can be ignored in this case.
-Call
-$$
-f(x,y,z) = \sum_{s} \log P(\xi_s | s, \tau^*, t=\{x,y,z\}).
-$$
-For branch lengths $\{x,y,z\}$, by considering all eight cases (see Table~\ref{tab:likelihoods}) of site patterns we can show that in terms of branch length \emph{probabilities}%
-	%
-    \footnote{The function $f$ has inverse $f^{-1}(u)=(1-u)/2$ so that the probability of a change along a branch of length $x$ maps to $(1-x)/2$ and the probability of no change maps to $(1+x)/2$.}
-    %
-\begin{align}
-f(x,y,z) \propto & \ 10\log(1+x)+10\log(1+z)+4\log(1-x)+4\log(1-z)+7\log(1+y)+\\
-& \left\{ 
-  \begin{array}{l}
-  2\log(1+x)+2\log(1-z)+\log(1+y)\\
-  2\log(1+x)+2\log(1+z)+\log(1-y)\\
-  2\log(1-x)+2\log(1-z)+\log(1+y)\\
-  \end{array} \right.
-\end{align}
-whose last summand will be chosen based on whichever yields the maximum value and where the proportionality is due to requiring branch length parameters $\{x,y,z\}$ to be multiplied by $1/2$ to map them to probabilities.
-We see that
-$$
-f(x,y,z)|_{x=z=0} = 8\log(1+y)
-$$
-and thus require the constraint
-$$
-8\log(1+y) > 0
-$$
-or that $y$ is any value bounded away from zero, and will obtain a maximum at $y=1$ or $p_y=0$.
-
-\subsection{Special case: $x=z\neq 0$}
-
-Fixing $x=z=0$ is a little too pathological---anything can happen when we're flipping coins to determine terminal characters.
-Let's investigate a relaxed yet still special case where branches to the tips are all equal but allowed to be nonzero.
-So we set $\tau^*=\tau$, $t^*=\{x,x,y\}$ and $t=\{x,x,\tilde{y}\}$ where the same terminal branches are used for both parameter sets with only the internal branch different.
-Removing terms involving only $x$ and redundant $y$ shows that we need only concern ourselves with the objective functions
-\begin{align*}
-L_1(u) =& 8\log(1+u) + \\
-  &(1+2x^2+4x^2y+x^4)\log(1+2x^2+4x^2u+x^4) + \\
-  &(1+2x^2-4x^2y+x^4)\log(1+2x^2-4x^2u+x^4)
-\end{align*}
-if $(1+u)(1-x)^2 > (1-u)(1+x)^2$---or, more simply, if $u > 2x / (1+x^2)$---and
-\begin{align*}
-L_2(u) =& (7-2x^2+4x^2y-x^4)\log(1+u) + \\
-  &(1+2x^2-4x^2y+x^4)\log(1-u) + \\
-  &(1+2x^2+4x^2y+x^4)\log(1+2x^2+4x^2u+x^4) + \\
-  &(1+2x^2-4x^2y+x^4)\log(1+2x^2-4x^2u+x^4)
-\end{align*}
-otherwise.
-We can rule out the case where $\tilde{y} < y$ in $L_1$ by the inequality of Gibbs and using a similar argument along with the inequality
-$$
-1+2x^2-4x^2y+x^4 \le 7-2x^2+4x^2y-x^4
-$$
-implies we can rule out $\tilde{y} < y$ in $L_2$ as well (actually I'm not sure if I believe this anymore; the case of $L_1$ definitely holds, though).
-This leaves us with four sets of inequalities where an inconsistency will arise,
-$$
-\left\{\tilde{y} > y > \frac{2x}{1+x^2}, \ L_1(\tilde{y}) > L_1(y)\right\}, 
-$$
-$$
-\left\{\tilde{y} > \frac{2x}{1+x^2} > y, \ L_1(\tilde{y}) > L_2(y)\right\}, 
-$$
-$$
-\left\{\frac{2x}{1+x^2} > \tilde{y} > y, \ L_2(\tilde{y}) > L_2(y)\right\},
-$$
-and
-$$
-\left\{\frac{2x}{1+x^2} > y > \tilde{y}, \ L_2(\tilde{y}) > L_2(y)\right\}.
-$$
-Simplifying shows
-$$
-L_2(u) = L_1(u) + (1+2x^2-4x^2y+x^4)\log\frac{1-u}{1+u} \le L_1(u).
-$$
-Can we use this fact?
-I believe in my heart of hearts, perhaps foolishly, that the first set of inequalities holds categorically.
-The fact above will then imply the second set of inequalities must also hold.
-The tricky bits are the last two sets of inequalities.
-
-If you'll allow me to indulge in a bit of cheating, I provide a few plots showing---for fixed $x$---the regions where inconsistencies arise.
-It looks like there is some regularity that we must be able to exploit somehow, if only to show that for a fixed $x$ there are inequalities in $y$ and $\tilde{y}$ governing consistency.
-
-%\begin{figure}
-%\centering
-%\includegraphics[width=.3\textwidth]{ineqs-x-0_25}\hfill
-%\includegraphics[width=.3\textwidth]{ineqs-x-0_75}\hfill
-%\includegraphics[width=.3\textwidth]{ineqs-x-0_9}
-%\caption{Various regions of inconsistency for fixed $x$}
-%\end{figure}
-
-Say $\tau_1$ is the ``Felsenstein zone'' tree and $\tau_2$ is the ``Farris zone'' tree, but both trees have a different value for its middle, connecting branch.
-%EM widdershins!! This still depends on how we draw them, no?
-The tips are labeled widdershins proceeding from the top-left tip.
-\begin{verbatim}
-Felsenstein:
-1\      /3
-   >--< 
-2/      \4
-
-Farris:
-1\ /3
-  | 
-2/ \4
-\end{verbatim}
-Our branch length parameters $t=\{x,y,z\}$ will consist of transformed probabilities where $y$ corresponds to this middle branch.
-For $p_x\in[0,1/2]$ defined as
-$$p_x = Pr(\mbox{change occurred along branch})$$
-then $x = 1-2p_x$.
-The parameter $x$ takes values in $[0,1]$ and will appear in likelihood calculations as either $(1+x)$ or $(1-x)$.
-Properties for $y$ and $z$ are identical.
-Assume for the moment that $\tau^*=\tau_1$ and $t^*=\{x,y,z\}$.
-
-\subsubsection{Example site pattern: \texttt{0101}}
-
-It can be shown that the only $A$ for which there is more than one partition for $\tau_2$ is for the site pattern \texttt{0101}.
-In this case
-$$A_1(\tau_1) = \{(1+x)^2(1-z)^2 = (1-x)^2(1+z)^2\},$$
-$$A_2(\tau_1) = \{(1+x)(1+y)(1-z) = (1-x)(1-y)(1+z)\},$$
-$$A_3(\tau_1) = \{(1-x)(1+y)(1+z) = (1+x)(1-y)(1-z)\}$$
-and
-$$A_1(\tau_2) = \{(1-z)^2(1+y) = (1+z)^2(1-y)\},$$
-$$A_2(\tau_2) = \{(1-z)^2(1+x)^2 = (1+z)^2(1-x)^2\},$$
-$$A_3(\tau_2) = \{(1+x)^2(1-y) = (1-x)^2(1+y)\}.$$
-
-Now, the probability of generating this site pattern under $\tau^*$ is given as
-$$
-Pr(\texttt{0101} | \tau^*, t^*) = \frac{1}{8}(1-2xz+y(x^2+z^2-2xz)+x^2z^2).
-$$
-
-\subsubsection{Example site pattern: \texttt{0000}}
-
-In the case of site pattern \texttt{0000} the refinements $A$ for both topologies are single-element partitions.
-In other words, no configuration of branch lengths will result in an ancestral state other than \texttt{00}.
-Here we have the probability
-$$
-Pr(\texttt{0000} | \tau^*, t^*) = \frac{1}{8}(1+2xz+y(x^2+z^2+2xz)+x^2z^2)
-$$
-and the maximum value for $Pr(\xi | \texttt{0000}, \tau, t^*)$ for both $\tau_1$ and $\tau_2$ is
-$$
-Pr(\xi | \texttt{0000}, \tau, t^*) = (1+x)^2(1+z)^2(1+y).
-$$
-
-\subsubsection{All refinements with multi-element partitions}
-
-(show calculations for all refinements that have multiple partitions)
-
-\subsection{Numerical sanity check}
-
-Let's suppose site pattern \texttt{0101} is the only one that drives some disconcert between the generating tree and the maximum likelihood tree.
-For a number of combinations of $\{x,y,z\}$ we need to calculate first
-$$
-Pr(\texttt{0101} | \tau^*, t^*) = \frac{1}{8}(1-2xz+y(x^2+z^2-2xz)+x^2z^2)
-$$
-and the maximum value over each $A_i(\tau)$ of $Pr(\xi | \texttt{0101}, \tau, t^*)$ for $\tau\in\{\tau_1, \tau_2\}$.
-In other words, say we have the triplet $\{x,y,z\}=\{0.8,0.5,0.8\}$.
-Then we have
-$$
-\max_{\xi} Pr(\xi | \texttt{0101}, \tau_1, t^*) = 0.0243, \ \max_{\xi} Pr(\xi | \texttt{0101}, \tau_2, t^*) = 0.6561, \ Pr(\texttt{0101} | \tau^*, t^*) = 0.0162.
-$$
-This doesn't yet tell us anything, as the probabilities for other site patterns may overwhelm this particular one, but gives a general flavor of how these computations will be performed.
-Moreover, a different triplet of $\{x,y,z\}$ could obtain a different maximum value for $Pr(\xi | s, \tau, t^*)$.
-This can be seen with the pattern \texttt{0011} where
-$$
-\max_{\xi} Pr(\xi | \texttt{0011}, \tau_1, t^*) = 0.6561, \ \max_{\xi} Pr(\xi | \texttt{0011}, \tau_2, t^*) = 0.0243, \ Pr(\texttt{0011} | \tau^*, t^*) = 0.1762
-$$
-meaning that if all else were equal---i.e., $\max_{\xi} Pr(\xi | s, \tau, t^*)$ were the same for $\tau_1$ and $\tau_2$ for all $s$---and that the maximum value actually occurs at this particular triplet $\{x,y,z\}$ then
-$$
-L(\tau_1) = 0.928 > 0.119 = L(\tau_2)
-$$
-and we have no contradiction.
-That's a lot of ``if'' statements, so let's investigate in a little more generality.
-
-\subsection{Differing branch lengths}
-
-First, we do not wish to assume $\{x,y,z\}$ are the same for both trees.
-
-\subsubsection{General case}
-
-The generating probability $Pr(s | \tau^*, t^*)$ has a specific form given a Hadamard transformation of the branch length parameters, while the term proportional to $Pr(\xi | s, \tau, t^*)$ relies on taking the maximum over a function with multiple components.
-All of this is easy to write into a computer program, and we can numerically find the maximum value over $\{\theta_1, \ldots, \theta_5\}\in[0,1]$ of each $L(\tau_1)$ and $L(\tau_2)$.
-If $L(\tau_1) < L(\tau_2)$ we see that jointly maximizing branch lengths and ancestral states results in a contradiction.
-
-In sufficient enough generality, we would look for any combination of $\{\tau_1,\tau_2\}$ that would result in this contradiction, but let's first focus on the two from the previous section.
-
-\subsubsection{Classic formulations of topologies}
+\section{Appendix}
 
@@
 \begin{tabular}{|l|ll|}
+\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k|\alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
 \hline
-    &00                              &01\\
+& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
     \hline
-0000&$(1+x)^2(1+y)^{3*}$          &$(1+x)^2(1-y)^3$\\
-0001&$(1+x)^2(1+y)^2(1-y)^*$       &$(1+x)^2(1+y)(1-y)^2$\\
-0010&$(1+x)^2(1+y)^2(1-y)^*$       &$(1+x)^2(1+y)(1-y)^2$\\
-0100&$(1+x)(1-x)(1+y)^{3*}$       &$(1+x)(1-x)(1-y)^3$\\
-1000&$(1+x)(1-x)(1+y)^{3*}$       &$(1+x)(1-x)(1-y)^3$\\
-0011&$(1+x)^2(1+y)(1-y)^{2}$          &$(1+x)^2(1-y)(1+y)^{2\ddagger}$\\
-0101&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
-1001&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
+    $\siteSplit_j$    &$\emptyset$  00                              &$\{2\}$  01\\
+    \hline
+     $\emptyset$   0000&$(1+x)^2   (1+w)(1+y)^{2*}$          &$(1+x)^2   (1-w)(1-y)^2$\\
+     $\{1\}$       1000&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
+     $\{2\}$       0010&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
+     $\{3\}$       0100&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
+     $\{1,2\}$     0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+     $\{1,3\}$     0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$   &$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
+     $\{2,3\}$     1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+     $\{1,2,3\}$   0001&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     \hline
     \hline
-&10                           &11\\
+    &$\{1\}$  10                           &$\{1,2\}$  11\\
     \hline
-0000&$(1-x)^2(1+y)^2(1-y)$        &$(1-x)^2(1+y)(1-y)^2$\\
-0001&$(1-x)^2(1+y)(1-y)^2$     &$(1-x)^2(1+y)^2(1-y)$\\
-0010&$(1-x)^2(1+y)(1-y)^2$     &$(1-x)^2(1+y)^2(1-y)$\\
-0100&$(1+x)(1-x)(1+y)^2(1-y)$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
-1000&$(1+x)(1-x)(1+y)^2(1-y)$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
-0011&$(1-x)^2(1-y)^3$       &$(1-x)^2(1+y)^{3\ddagger}$\\
-0101&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
-1001&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
+     $\emptyset$   0000&$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
+     $\{1\}$       1000&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{2\}$       0010&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+     $\{3\}$       0100&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{1,2\}$     0011&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
+     $\{1,3\}$     0101&$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
+     $\{2,3\}$     1001&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
+     $\{1,2,3\}$   0001&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+    \hline
+\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k|\alignmentColumn=\siteSplit_j,\tau_2,t)$}\\
+\hline
+& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
+    \hline
+    $\siteSplit_j$    &$\emptyset$  00                              &$\{2\}$  01\\
+    \hline
+     $\emptyset$   0000&$(1+x)^2   (1+w)(1+y)^{2*}$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+     $\{1\}$       1000&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1-x)^2   (1-w)(1+y)(1-y)$\\
+     $\{2\}$       0100&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1-y)^2$\\
+     $\{3\}$       0010&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$\\
+     $\{1,2\}$     0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
+     $\{1,3\}$     0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$    &$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$\\
+     $\{2,3\}$     1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1-x)^2   (1-w)(1+y)^{2\ddagger}$\\
+     $\{1,2,3\}$   0001&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$\\
+    \hline
+    \hline
+    &$\{1\}$  10                           &$\{1,2\}$  11\\
+    \hline
+     $\emptyset$   0000&$(1+x)(1-x)(1-w)(1+y)(1-y)$             &$(1-x)^2   (1+w)(1-y)^2$\\
+     $\{1\}$       1000&$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$  &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{2\}$       0100&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$      &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+     $\{3\}$       0010&$(1-x)^2   (1-w)(1+y)(1-y)$             &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{1,2\}$     0011&$(1-x)^2   (1-w)(1-y)^2$                &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
+     $\{1,3\}$     0101&$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$  &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
+     $\{2,3\}$     1001&$(1+x)^2   (1-w)(1-y)^{2\ddagger}$      &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
+     $\{1,2,3\}$   0001&$(1+x)(1-x)(1-w)(1-y)^2$                &$(1-x)^2   (1+w)(1+y)(1-y)$\\
 \hline
@@
 Maxima determined row-wise (i.e., by site pattern).
+All values multiplied by $1/32$.
 Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
+\label{tab:likelihoods}
 \end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{|l|ll|}
-    \hline
-    &00                              &01\\
-    \hline
-0000&$(1+x)^2(1+y)^{3*}$          &$(1+x)(1-x)(1+y)(1-y)^2$\\
-0001&$(1+x)^2(1+y)^2(1-y)^{*}$       &$(1+x)(1-x)(1+y)^2(1-y)$\\
-0010&$(1+x)(1-x)(1+y)^{3\ddagger}$       &$(1+x)^2(1+y)(1-y)^{2\ddagger}$\\
-0100&$(1+x)^2(1+y)^2(1-y)^*$       &$(1+x)(1-x)(1-y)^3$\\
-1000&$(1+x)(1-x)(1+y)^{3\ddagger}$       &$(1-x)^2(1+y)(1-y)^2$\\
-0011&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$          &$(1+x)^2(1+y)^2(1-y)$\\
-0101&$(1+x)^2(1+y)(1-y)^{2\ddagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
-1001&$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$    &$(1+x)(1-x)(1+y)(1-y)^2$\\
-    \hline
-    \hline
-&10                           &11\\
-    \hline
-0000&$(1+x)(1-x)(1+y)(1-y)^2$        &$(1-x)^2(1+y)(1-y)^2$\\
-0001&$(1+x)(1-x)(1-y)^3$     &$(1-x)^2(1+y)^2(1-y)$\\
-0010&$(1-x)^2(1+y)(1-y)^2$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
-0100&$(1+x)(1-x)(1+y)^2(1-y)^{\ddagger}$     &$(1-x)^2(1+y)^2(1-y)$\\
-1000&$(1+x)^2(1+y)(1-y)^{2\ddagger}$     &$(1+x)(1-x)(1+y)(1-y)^2$\\
-0011&$(1-x)^2(1-y)^3$       &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
-0101&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1-x)^2(1+y)^{3\ddagger}$\\
-1001&$(1+x)(1-x)(1+y)(1-y)^2$  &$(1+x)(1-x)(1+y)^2(1-y)^{\dagger}$\\
-    \hline
-\end{tabular}    
-\caption{Likelihood calculations for all site patterns and internal states of Felsenstein topology.
-Maxima determined row-wise (i.e., by site pattern).
-Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
-\label{tab:likelihoods-fels}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{|l|l|}
-    \hline
-s   &$P(s)$\\
-    \hline
-0000&$1+x^2+z^2+2xyz+x^2y+z^2y+x^2z^2$\\
-0001&$1+x^2-z^2+x^2y-z^2y-x^2z^2$\\
-0010&$1-x^2+z^2-x^2y+z^2y-x^2z^2$\\
-0100&$1+x^2-z^2+x^2y-z^2y-x^2z^2$\\
-1000&$1-x^2+z^2-x^2y+z^2y-x^2z^2$\\
-0011&$1-x^2-z^2-2xyz-x^2y-z^2y+x^2z^2$\\
-0101&$1+x^2+z^2-2xyz+x^2y+z^2y+x^2z^2$\\
-1001&$1-x^2-z^2-2xyz-x^2y-z^2y+x^2z^2$\\
-    \hline
-\end{tabular}    
-\caption{Site pattern probabilities for Felsenstein zone topology.
-All values are multiplied by $1/8$.}
-\label{tab:sitepatprob-fels}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{|l|ll|}
-    \hline
-    &00                              &01\\
-    \hline
-0000&$(1+x)^2(1+y)(1+z)^{2*}$          &$(1+x)(1-x)(1-y)(1+z)(1-z)$\\
-0001&$(1+x)^2(1+y)(1+z)(1-z)^{\ddagger}$       &$(1+x)(1-x)(1-y)(1+z)^{2\ddagger}$\\
-0010&$(1+x)(1-x)(1+y)(1+z)^{2\ddagger}$       &$(1+x)^2(1-y)(1+z)(1-z)^{\ddagger}$\\
-0100&$(1+x)^2(1+y)(1+z)(1-z)^{\ddagger}$       &$(1+x)(1-x)(1-y)(1-z)^2$\\
-1000&$(1+x)(1-x)(1+y)(1+z)^{2\ddagger}$       &$(1-x)^2(1-y)(1+z)(1-z)$\\
-0011&$(1+x)(1-x)(1+y)(1+z)(1-z)^{\ddagger}$          &$(1+x)^2(1-y)(1+z)^{2\ddagger}$\\
-0101&$(1+x)^2(1+y)(1-z)^{2\ddagger}$    &$(1+x)(1-x)(1-y)(1+z)(1-z)^{\ddagger}$\\
-1001&$(1+x)(1-x)(1+y)(1+z)(1-z)^{\dagger}$    &$(1+x)(1-x)(1-y)(1+z)(1-z)$\\
-    \hline
-    \hline
-&10                           &11\\
-    \hline
-0000&$(1+x)(1-x)(1-y)(1+z)(1-z)$        &$(1-x)^2(1+y)(1-z)^2$\\
-0001&$(1+x)(1-x)(1-y)(1-z)^2$     &$(1-x)^2(1+y)(1+z)(1-z)$\\
-0010&$(1-x)^2(1-y)(1+z)(1-z)$     &$(1+x)(1-x)(1+y)(1-z)^2$\\
-0100&$(1+x)(1-x)(1-y)(1+z)^{2\ddagger}$     &$(1-x)^2(1+y)(1+z)(1-z)$\\
-1000&$(1+x)^2(1-y)(1+z)(1-z)^{\ddagger}$     &$(1+x)(1-x)(1+y)(1-z)^2$\\
-0011&$(1-x)^2(1-y)(1-z)^2$       &$(1+x)(1-x)(1+y)(1+z)(1-z)^{\ddagger}$\\
-0101&$(1+x)(1-x)(1-y)(1+z)(1-z)^{\ddagger}$  &$(1-x)^2(1+y)(1+z)^{2\ddagger}$\\
-1001&$(1+x)(1-x)(1-y)(1+z)(1-z)$  &$(1+x)(1-x)(1+y)(1+z)(1-z)^{\dagger}$\\
-    \hline
-\end{tabular}    
-\caption{Likelihood calculations for all site patterns and internal states of Felsenstein topology.
-Maxima determined row-wise (i.e., by site pattern).
-Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
-\label{tab:likelihoods-fels}
-\end{table}
-
-In the first case we are in a similar situation to the case of $x=z=0$ where $\tilde{y}>y$ implies $L(\tilde{y})>L(y)$.
-The second case is more interesting.
-We still require $\tilde{y}>y$, but we now have a further restriction.
-We are interested in the simultaneous inequalities
-$$
-\frac{2x}{1+x^2} > \tilde{y} > y,
-$$
-$$
-[7-2x^2+4x^2y-x^4]\log\frac{1+\tilde{y}}{1+y}+[1+2x^2-4x^2y+x^4]\log\frac{1-\tilde{y}}{1-y} > 0
-$$
-which can be written
-$$
-8\log\frac{1+\tilde{y}}{1+y}+[1+2x^2-4x^2y+x^4]\log\frac{(1-\tilde{y})(1+y)}{(1+\tilde{y})(1-y)} > 0.
-$$
-What can we say about this?
-
-Now terms involving only $P(s|\tau,t)$ differ between $t^*$ and $t$ for the site patterns 0000 and 0101 but no others, allowing us to ignore them.
-Additionally, all terms involving $P(\xi_s|s,\tau,t)$ will differ between $t^*$ and $t$ yielding a search for parameters such that
-$$
-\sum_{s\in\{0000,0101\}} P(s|\tau^*, t^*)\log P(s|\tau^*, t) + \sum_{s} P(s | \tau^*, t^*) \log P(\xi_s | s, \tau^*, t)
-$$
-is maximized at $t\neq t^*$.
-Since $\log P(\xi_s|s,\tau,t)$ breaks up into a sum of terms involving $x$ and $y$ we have
-$$
-(6-4x^2-2x^4)\log(1+y)+\sum_{s\in\{0000,0101\}} P(s|\tau^*, t^*)[\log P(s|\tau^*, t)+\log P(\xi_s|s, \tau^*, t)].
-$$
-For a bit of clarity, let
-$$
-a(x,y) = (1+2x^2+4x^2y+x^4),
-$$
-$$
-b(x,y) = (1+2x^2-4x^2y+x^4)
-$$
-so that we have
-$$
-(6-4x^2-2x^4)\log(1+y)+a(x,y)\log a(x,y)+b(x,y)\log b(x,y) + \sum_{s\in\{0000,0101\}} P(s|\tau^*, t^*)\log P(\xi_s|s, \tau^*, t).
-$$
-Canceling more terms shows the right-hand sum is
-$$
-(1+2x^2+4x^2y+x^4)\log(1+y) + 16x^2y\log(1+x) + P(0101|\tau^*, t^*)\log P(\xi_s|0101, \tau^*, t).
-$$
-That final right-hand term above is actually a component function since we are maximizing a likelihood, and the 0101 row in Table~\ref{tab:likelihoods} does not have a unique maximum value.
-This value is either
-$$
-c_1(x,y) = (1+x)^4(1-y),
-$$
-$$
-c_2(x,y) = (1+x)^2(1-x)^2(1+y).
-$$
-We are interested in two inequalities; if
-$$
-c_2(x,y) > c_1(x,y)
-$$
-then our whole objective function simplifies since many terms with $\log(1+y)$ cancel.
-We can then focus on a second inequality, namely whether
-$$
-8\log(1+y) + x^2y(4\log(1+y)+16\log(1+x)) + a(x,y)\log a(x,y) + b(x,y)\log b(x,y) + c_2(x,y)...blah
-$$
-is greater than
-$$
-8\log(1+\tilde{y}) + x^2y(4\log(1+\tilde{y})+16\log(1+x)) + a(x,y) + b(x,y) + c_2(x,y)...
-$$
-
-\subsection{Dashed intuition}
-
-I am having a tough time seeing the intuition behind this result.
-If $x=z=0$ then we are flipping a coin for what happens at the internal node proceeding from the leaf.
-Why would effectively constraining the interior nodes to be equal result in a larger value for the likelihood?
-It doesn't have anything to do with efficiency, does it (i.e., by constraining the parameters to be equal we are using the same amount of data to estimate fewer parameters)?
-Can we show that constraining any $\xi_i,\xi_j$ to be equal in $L_n$ will also result in problems?
-
-I think what may help my sanity is proceeding with the above for the usual maximum likelihood story, that is, for
-$$
-\log L''' = \sum_{s} P(s | \tau^*, t^*) \cdot [\log P(s | \tau^*, t) + \log\sum_{\xi} P(\xi_s | s, \tau^*, t)]
-$$
-show
-$$
-L'''(\tau^*, t^*) > L'''(\tau^*, t)
-$$
-for $t\neq t^*$.
-At least for $x=z=0$ we can see that these two quantities will be equal for all $y$ due to the symmetry of choosing the internal nodes.
-Is this satisfying?
-
-UPDATE: thinking about this more, it actually makes sense.
-If we are maximizing over internal states, we will always prefer a branch length probability that will indicate no change occurs between them, effectively constraining all internal states to be the same given a particular site pattern.
-As we can see in the table, the maximum value almost always occurs for internal states that are equivalent.
-Seeking harmony between internal states will always increase the likelihood except in cases where branches to terminating leaves are sufficiently small.
-So the question might be, now, whether or not we can prove a general result: since internal states seem to always maximize the likelihood by being constrained to be equal, what conditions are needed on the terminal branches (here $x$ and $z$) that will yield (in)consistent estimates?
-
-\subsection{Plan of attack}
-
-We recall \eqref{eq:inconsistency_inequality} and set $\tau^*=\tau_1$ and $\tau'=\tau_2$.
-In other words, data are generated from the ``Farris zone'' topology, and we seek all values for $t^*$ where we can obtain a higher likelihood by proceeding with joint maximization assuming the ``Felsenstein zone'' topology.
-
-\subsection{The cheater's approach}
-
-Using \texttt{python}, we can numerically maximize \eqref{eq:site_pattern_profile_likelihood_mean} over $t$ separately for fixed topologies $\tau^*$ and $\tau'$.
-For $t^*$, we can calculate these values over a grid of values for $(x^*, y^*)\in[0,1]^2$.
-At worst, this will show us if we're on a fool's errand, but if not then we could benefit by gaining insight into what form a refinement of branch lengths yielding inconsistency would take.

### 2017-04-28 deprecated.tex
add appendix; address most issues

@@
 \section{Temporarily unused text I may want later to show more general results on the likelihood estimates}
+
+\paragraph{Examples}
+
+See Table~\ref{tab:examples} for parameters, bounds and true likelihood values obtained via basin hopping.
+
+Set $\{x, y\} = \{0.25, 0.25\}$.
+For a Farris zone generating topology, we have
+$$
+L^{p,1}_{\tau_1,\tau_1}(\mathbf{a}_{xy}) \approx -2.17803560694157,
+$$
+$$
+L^{p,2}_{\tau_1,\tau_1}(\mathbf{a}_{xy}) \approx -2.17803560694157
+$$
+and
+$$
+L^{p,3}_{\tau_1,\tau_1}(\mathbf{a}_{xy}) \approx -2.17803560694157.
+$$
+Computing the upper bound yields
+$$
+\max_{t,\fullAncestralSplitPartitions(\tau_1,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,t),\tau_1,t) \le -4.22274479428471
+$$
+and it remains to show that
+$$
+\max_{t,\fullAncestralSplitPartitions(\tau_2,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,t),\tau_2,t) \ge -4.22274479428471.
+$$
+
+For the same generating parameters,
+$$
+d_{xy} = 0.498046875, \ e_{xy} = 0.2431640625,
+$$
+resulting in
+$$
+L^{p,1}_{\tau_1,\tau_2}(d_{xy},e_{xy}) \approx -2.07968569223991
+$$
+and
+$$
+\shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-d_{xy}-2e_{xy}, 0\}) \approx -2.07748833719921.
+$$
+The lower bound in this case becomes
+$$
+\max_{t,\fullAncestralSplitPartitions(\tau_2,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,t),\tau_2,t) \ge -4.15717402943912
+\ge -4.22274479428471 \ge \max_{t,\fullAncestralSplitPartitions(\tau_1,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,t),\tau_1,t)
+$$
+where site patterns are generated from the Farris zone topology.
+Confirming, run \texttt{python}'s numerical optimization solver for $\{.25, .25\}$.
+Since we are computing a constrained maximization, we can find a global maximum using basin hopping, and doing so yields
+$$
+\max_{t,\fullAncestralSplitPartitions(\tau_1,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,t),\tau_1,t) \approx -4.5101771294218045
+$$
+and
+$$
+\max_{t,\fullAncestralSplitPartitions(\tau_2,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,t),\tau_2,t) \approx -4.1544859851260281.
+$$
+It looks like the lower bound is tighter than the upper bound, so we may try focusing there.
 
@@
 where $a_{xy},b_{xy}$ and $c_{xy}$ are functions of the true generating branch parameters $\{x,y\}$.
-
-\begin{table}
-\centering
-\begin{tabular}{|l|ll|}
-\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k|\alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
-\hline
-& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
-    \hline
-    $\siteSplit_j$    &$\emptyset$  00                              &$\{2\}$  01\\
-    \hline
-     $\emptyset$   0000&$(1+x)^2   (1+w)(1+y)^{2*}$          &$(1+x)^2   (1-w)(1-y)^2$\\
-     $\{1\}$       1000&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
-     $\{2\}$       0100&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
-     $\{3\}$       0010&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
-     $\{1,2\}$     0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{1,3\}$     0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$   &$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
-     $\{2,3\}$     1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{1,2,3\}$   0001&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
-    \hline
-    \hline
-    &$\{1\}$  10                           &$\{1,2\}$  11\\
-    \hline
-     $\emptyset$   0000&$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
-     $\{1\}$       1000&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{2\}$       0100&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{3\}$       0010&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-     $\{1,2\}$     0011&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
-     $\{1,3\}$     0101&$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
-     $\{2,3\}$     1001&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
-     $\{1,2,3\}$   0001&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-    \hline
-\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k|\alignmentColumn=\siteSplit_j,\tau_2,t)$}\\
-\hline
-& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
-    \hline
-    $\siteSplit_j$    &$\emptyset$  00                              &$\{2\}$  01\\
-    \hline
-     $\emptyset$   0000&$(1+x)^2   (1+w)(1+y)^{2*}$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{1\}$       1000&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1-x)^2   (1-w)(1+y)(1-y)$\\
-     $\{2\}$       0100&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1-y)^2$\\
-     $\{3\}$       0010&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$\\
-     $\{1,2\}$     0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
-     $\{1,3\}$     0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$    &$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$\\
-     $\{2,3\}$     1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1-x)^2   (1-w)(1+y)^{2\ddagger}$\\
-     $\{1,2,3\}$   0001&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$\\
-    \hline
-    \hline
-    &$\{1\}$  10                           &$\{1,2\}$  11\\
-    \hline
-     $\emptyset$   0000&$(1+x)(1-x)(1-w)(1+y)(1-y)$             &$(1-x)^2   (1+w)(1-y)^2$\\
-     $\{1\}$       1000&$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$  &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{2\}$       0100&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$      &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-     $\{3\}$       0010&$(1-x)^2   (1-w)(1+y)(1-y)$             &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{1,2\}$     0011&$(1-x)^2   (1-w)(1-y)^2$                &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
-     $\{1,3\}$     0101&$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$  &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
-     $\{2,3\}$     1001&$(1+x)^2   (1-w)(1-y)^{2\ddagger}$      &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
-     $\{1,2,3\}$   0001&$(1+x)(1-x)(1-w)(1-y)^2$                &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-\hline
-\end{tabular}
-\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
-Maxima determined row-wise (i.e., by site pattern).
-All values multiplied by $1/32$.
-Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
-\label{tab:likelihoods}
-\end{table}
 

### 2017-04-28 joint_inf.tex
add appendix; address most issues

@@
 %EM given that the H is on the lhs of the conditioning, does it make sense to have it on the rhs of the semicolon? I don't know the convention.
-L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
+%das: I forgot that standard likelihoods are written with the parameters on the left.
+%Since H is like a hidden variable, I went for an expectation-maximization notation, grouping them with the observations.
+%It seems a little wonky, though, when we start having maximum likelihood ancestral states depending on \tau,t.
+%Oh well...
+L_\nCols(\tau, t; \fullAlignment,\fullAncestralStates) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
 \end{equation}
@@
 \label{eq:marginal_likelihood}
-\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
+\tilde{L}_\nCols(\tau, t; \fullAlignment) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates)
 \end{equation}
@@
 $$
-(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
+(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \  \tilde{L}_\nCols(\tau, t; \fullAlignment).
 $$
@@
 \label{eq:profile_likelihood}
-L_\nCols'(\fullAlignment;\tau, t) = \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t),
+L_\nCols'(\tau, t; \fullAlignment) = \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates),
 \end{equation}
@@
 \label{eq:profile_likelihood_topology_bl}
-(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\fullAlignment;\tau, t)
+(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment)
 \end{equation}
@@
 \begin{align}
-L_\nCols'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) \\
+L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \\
                              &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
@@
 %EM think we should say "where the site split or its complement appears."?
-where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ appears.
+where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.
 
@@
 The unique site splits, collapsing complements, are
-$$
-\mathcal{P}(\siteSplitSet) = \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\}.
-$$
+\begin{align*}
+    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
+& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
+\end{align*}
 Had we not identified character complements, we would need to consider the additional splits
@@
 %EM Would it be worth enumerating the site splits ahead of time? It seems a little funky to be calling them j each time.
-The probabilities of observing ancestral splits for $\siteSplit_j=\emptyset$ are
+The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
 $$
@@
 $$
-Thus the set of most likely ancestral states contains a single element in this case, here $\ancestralSplitPartition_j(\tau, p)=\{\emptyset\}$.
-Then, taking $\xi_j\in\ancestralSplitPartition_j(\tau, p)$ we have
+Thus the set of most likely ancestral states contains a single element in this case, here $\ancestralSplitPartition_1(\tau, p)=\{\emptyset\}$.
+Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, p)$ we have
 $$
-P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\emptyset, \tau, t) =
+P(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
 (1-p)^5.
 $$
-For $\siteSplit_j=\{1,3\}$ we have
+For $\siteSplit_6=\{1,3\}$ we have
 $$
@@
 $$
-In this case, the set of most likely ancestral states is $\ancestralSplitPartition_j(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_j\in\ancestralSplitPartition_j(\tau, p)$,
+In this case, the set of most likely ancestral states is $\ancestralSplitPartition_6(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_6\in\ancestralSplitPartition_6(\tau, p)$,
 $$
-P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\{1,3\}, \tau, t) =
+P(\ancestralSplitRV=\xi_6 \mid \siteSplitRV=\{1,3\}, \tau, t) =
 p^2(1-p)^3.
@@
 After reading the below, I think it makes most sense to just refer to (9) indirectly but make this a stand-alone definition.
-Define \eqref{eq:site_pattern_likelihood} as
+Let
 $$
-L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}.
+L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
 $$
+be the final product in \eqref{eq:site_pattern_likelihood}.
 Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
@@
 \begin{align}
-    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau,t),\tau,t)
+    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t))
         &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
@@
 %EM did you mean to drop the eta from the first term of this equation?
-\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)].
+    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)].
 \label{eq:site_pattern_profile_likelihood_mean}
@@
 %EM IMHO \ell^p is so standard that it's confusing to see it in another setting. Perhaps \tilde{\ell}?
-\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
+\tilde{\ell}_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
 $$
@@
     \label{eq:log_likelihood_simplified}
-    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
+    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
 \end{equation}
@@
 %EM Suggest \hat{\fullAncestralSplitPartitions}_1, etc.
-\ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau',t_1),\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau^*,t_2),\tau^*,\hat{t}_2)
+\ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}_1,\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}_2,\tau^*,\hat{t}_2)
 \end{equation}
-for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}(\tau',t_1)\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}(\tau^*,t_2)\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
+for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}_1\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}_2\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
 By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any potential example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
@@
 
-Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified} as
-\begin{equation}
-%EM Seems a little strange to give a new equation number to an equation that already exists with an assigned number.
-\label{eq:partial-likelihood}
-\ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
-\end{equation}
-We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
+Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified}.
+We obtain an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} as
 $$
@@
     \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
-+ \ell^p_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau,t),\tau,\hat{t})
++ \tilde{\ell}_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions},\tau,\hat{t})
 $$
@@
 $$
-\{\hat{t},\hat{\fullAncestralSplitPartitions}(\tau,t)\} = \argmax_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
+\{\hat{t},\hat{\fullAncestralSplitPartitions}\} = \argmax_{t,\fullAncestralSplitPartitions(\tau,t)} \ \tilde{\ell}_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
 $$
@@
     \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
-+ \ell^p_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau,t),\tau,\hat{t}).
++ \tilde{\ell}_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions},\tau,\hat{t}).
 $$
-
-% vvvv Remove these? I think they're obvious now, but maybe they can be made more explicit above?
-%EM I agree-- quite clear now!
-\textbf{Argument for lower bound}: This makes intuitive sense, though may be difficult to show in general.
-Can we not just say $\max_u f(u) + g(u)$ is greater than $f(u') + g(u')$ for any $u'$ that doesn't maximize $f(u) + g(u)$ just by definition?
-There could be multiple maxima, but that's why we have greater than or equal to.
-I think the only thing we'd need to show is that a maximum exists, but since $\{x,y,w\}$ are supported on a closed set this is certainly true.
-I must be missing something.
-
-\textbf{Argument for upper bound}: I think this is true mostly because $\max_u f(u) + g(u)$ should be less than $\max_u f(u) + \max_u g(u)$ by some triangle inequality argument.
-If we were to actually maximize the full likelihood, we should get something smaller than maximizing some form of it that we consider by looking at the parts separately.
-For a fixed $\{x^*, y^*\}$ we can calculate this upper bound directly---can we obtain it as a useful-looking function of these parameters in the general case?
 
@@
 %EM clarify that we aren't assuming that the x and y in t are the same as those appearing in t^*?
-Call $t=\{x,y,w\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ is the case where the bottom three branches all share the same parameter, the classical construction of this topology.
+Call $t=\{x',y',w'\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ is the case where the bottom three branches all share the same parameter, the classical construction of this topology.
 
-We will parameterize these trees \textbf{not with branch lengths}, but rather such that the probability of a substitution on a branch with parameter $x$ is $p_x=1-2x$.
+We will parameterize these trees \textbf{not with branch lengths}, but rather such that the probability of a substitution on a branch with parameter $x'$ is $p_x'=(1-x')/2$.
 %EM (Scratching head) isn't p_x=1-2x backwards? Then p_x will take values in the range -1 to 1.
-Here, $\{p_x,p_y,p_w\}\in[0,1/2]^3$ and $\{x,y,w\}\in[0,1]^3$.
+Here, $\{p_x',p_y',p_w'\}\in[0,1/2]^3$ and $\{x',y',w'\}\in[0,1]^3$.
 This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
@@
 
-\textbf{Justification for estimand:} It might be worth arguing why this form of the estimand $t$ is necessary.
-%EM definitely. This is a big TODO.
-I think the idea is that for either of these topologies, any maximum obtained will occur where the top two and bottom two branches are equal, though not necessarily the middle branch.
-This would be good to show explicitly.
+%das: did this on paper, but can write it out.
+\textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\{x_1,y_1,x_2,y_2,w\}$ counting from the top left.
+The likelihood $\ell(x_1,y_1,x_2,y_2,w)$ can be written as a function of $\{x_1x_2, x_1+x_2, y_1y_2, y_1+y_2, w\}$ only.
+In this case, $x_1$ and $x_2$ can be exchanged for one another without changing the value of $\ell$, and any maximum will require $x_1=x_2$.
+A similar argument can be made for $y_1$ and $y_2$.
+For the Felsenstein topolgy this is not the case, but, since we are interested in a lower bound for this topology, we will simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
 
@@
     \hline
-    $\emptyset$     &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
-    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
-    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
-    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
-    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
-    $\{1,3\}$      &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
-    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
-    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
+    $\emptyset$     &$1+x'^2+y'^2+4x'y'w'+x'^2y'^2$&$1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
+    $\{1\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
+    $\{2\}$        &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
+    $\{3\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
+    $\{1,2\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
+    $\{1,3\}$      &$1+x'^2+y'^2-4x'y'w'+x'^2y'^2$&$1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
+    $\{2,3\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
+    $\{1,2,3\}$    &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
     \hline
@@
 
-\subsubsection{Upper bound for Farris partial likelihood}
+\subsubsection{Upper bound for Farris likelihood}
 
@@
     &\qquad + p_{13}\cdot\log(2p_{13}) \\
-    &\qquad := L^{p,1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
+    &\qquad := L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
 \end{align*}
 
-The other two possibilties for the likelihood of the ancestral state split $\{2\}$ for the site split $\{1,3\}$ admit similar simplifications.
+The other two possibilties of ancestral state split for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
 Call
@@
     &\qquad + \log(2) \\
-    &\qquad := L^{p,2}_{\tau_1,\tau_1}(\mathbf{p}_{xy});
+    &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy});
 \end{align*}
@@
     &\qquad + \log(2) \\
-    &\qquad := L^{p,3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
+    &\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
 \end{align*}
@@
 \begin{align*}
-    \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
-    &=      + p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
+    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    &=      p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
     &\qquad + p_{1}  \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
@@
 \begin{align*}
-    \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
-    &=      + p_{0}  \cdot\log((1+x')^2    (1+w')(1+y')^2) \\
+    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    &=      p_{0}  \cdot\log((1+x')^2    (1+w')(1+y')^2) \\
     &\qquad + p_{1}  \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
@@
 $$
-b_{1} = p_{1}+p_{2}+p_{3}+p_{123}
+b = p_{1}+p_{2}+p_{3}+p_{123}
 $$
@@
 \begin{align*}
-    \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
-    &=      + 2\cdot\log(1+x') \\
-    &\qquad + (2-b_{1})  \cdot\log(1+y') \\
-    &\qquad + b_{1}      \cdot\log(1-y') \\
+    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    &=      2\cdot\log(1+x') \\
+    &\qquad + (2-b)  \cdot\log(1+y') \\
+    &\qquad + b      \cdot\log(1-y') \\
     &\qquad + p_{0}\cdot\log(1+w') \\
@@
 \begin{align*}
-    \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
-    &=      + 2\cdot\log(2) \\
-    &\qquad + (2-b_{1})  \cdot\log(2-b_{1}) \\
-    &\qquad + b_{1}      \cdot\log(b_{1}) \\
+    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    &=      2\cdot\log(2) \\
+    &\qquad + (2-b)  \cdot\log(2-b) \\
+    &\qquad + b      \cdot\log(b) \\
     &\qquad + p_{0}\cdot\log(2p_{0}) \\
@@
 \end{align*}
-maximized at $\{1,1-b_{1},1-2p_{0}\}$.
-
-...
-
-$$
-    L^{p,1}_{\tau_1,\tau_2}(\mathbf{p}_{xy}) := \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b_{1},1-2p_{0}\}) +
-$$
+maximized at $\{1,1-b,1-2p_{0}\}$.
+Our lower bound is
+\begin{align*}
+    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,1-2p_{0}\}) \\
+                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b) + p_{0}\cdot\log(2p_{0}) + (1-p_{0})\cdot\log(2(1-p_{0})).
+\end{align*}
 
@@
 
-We want to find values for $\{x,y\}$ where all of
+Ideally, we want the region where
 $$
-\shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-b_{1}-2b_{2}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(b_{1},b_{2}) \ge \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p,1}_{\tau_1,\tau_1}(\mathbf{a}_{xy}),
+L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) \ge \max(L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}), L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}),L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy})).
+$$
+Currently being slightly lazy, I'm considering values for $\{x,y\}$ where all of
+$$
+L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) \ge L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}),
 $$
 $$
-\shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-b_{1}-2b_{2}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(b_{1},b_{2}) \ge \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p,2}_{\tau_1,\tau_1}(\mathbf{a}_{xy})
+L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) \ge L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy})
 $$
@@
 $$
-\shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-b_{1}-2b_{2}, 0\}) + L^{p,1}_{\tau_1,\tau_2}(b_{1},b_{2}) \ge \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p,3}_{\tau_1,\tau_1}(\mathbf{a}_{xy})
+L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) \ge L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy})
 $$
@@
 \end{figure}
-
-\paragraph{Examples}
-
-See Table~\ref{tab:examples} for parameters, bounds and true likelihood values obtained via basin hopping.
-
-Set $\{x, y\} = \{0.25, 0.25\}$.
-For a Farris zone generating topology, we have
-$$
-L^{p,1}_{\tau_1,\tau_1}(\mathbf{a}_{xy}) \approx -2.17803560694157,
-$$
-$$
-L^{p,2}_{\tau_1,\tau_1}(\mathbf{a}_{xy}) \approx -2.17803560694157
-$$
-and
-$$
-L^{p,3}_{\tau_1,\tau_1}(\mathbf{a}_{xy}) \approx -2.17803560694157.
-$$
-Computing the upper bound yields
-$$
-\max_{t,\fullAncestralSplitPartitions(\tau_1,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,t),\tau_1,t) \le -4.22274479428471
-$$
-and it remains to show that
-$$
-\max_{t,\fullAncestralSplitPartitions(\tau_2,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,t),\tau_2,t) \ge -4.22274479428471.
-$$
-
-For the same generating parameters,
-$$
-d_{xy} = 0.498046875, \ e_{xy} = 0.2431640625,
-$$
-resulting in
-$$
-L^{p,1}_{\tau_1,\tau_2}(d_{xy},e_{xy}) \approx -2.07968569223991
-$$
-and
-$$
-\shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-d_{xy}-2e_{xy}, 0\}) \approx -2.07748833719921.
-$$
-The lower bound in this case becomes
-$$
-\max_{t,\fullAncestralSplitPartitions(\tau_2,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,t),\tau_2,t) \ge -4.15717402943912
-\ge -4.22274479428471 \ge \max_{t,\fullAncestralSplitPartitions(\tau_1,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,t),\tau_1,t)
-$$
-where site patterns are generated from the Farris zone topology.
-Confirming, run \texttt{python}'s numerical optimization solver for $\{.25, .25\}$.
-Since we are computing a constrained maximization, we can find a global maximum using basin hopping, and doing so yields
-$$
-\max_{t,\fullAncestralSplitPartitions(\tau_1,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,t),\tau_1,t) \approx -4.5101771294218045
-$$
-and
-$$
-\max_{t,\fullAncestralSplitPartitions(\tau_2,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,t),\tau_2,t) \approx -4.1544859851260281.
-$$
-It looks like the lower bound is tighter than the upper bound, so we may try focusing there.
 

### 2017-04-28 joint_inf.tex
fixing (partial) log likelihoods

@@
 %EM IMHO \ell^p is so standard that it's confusing to see it in another setting. Perhaps \tilde{\ell}?
-\tilde{\ell}_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
+\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
 $$
@@
     \label{eq:log_likelihood_simplified}
-    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
+    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
 \end{equation}
@@
 %EM Suggest \hat{\fullAncestralSplitPartitions}_1, etc.
-\ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}_1,\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}_2,\tau^*,\hat{t}_2)
+\ell_{\tau^*,t^*}(\tau', \hat{t}_1; \hat{\fullAncestralSplitPartitions}_1) > \ell_{\tau^*,t^*}(\tau^*, \hat{t}_2; \hat{\fullAncestralSplitPartitions}_2)
 \end{equation}
@@
 $$
-\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \le
+\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \le
     \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
-+ \tilde{\ell}_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions},\tau,\hat{t})
+    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \hat{\fullAncestralSplitPartitions})
 $$
@@
 $$
-\{\hat{t},\hat{\fullAncestralSplitPartitions}\} = \argmax_{t,\fullAncestralSplitPartitions(\tau,t)} \ \tilde{\ell}_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
+\{\hat{t},\hat{\fullAncestralSplitPartitions}\} = \argmax_{t,\fullAncestralSplitPartitions(\tau,t)} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
 $$
@@
 $$
-\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \ge
+\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \ge
     \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
-+ \tilde{\ell}_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions},\tau,\hat{t}).
+    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \hat{\fullAncestralSplitPartitions}).
 $$
@@
 %das: did this on paper, but can write it out.
-\textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\{x_1,y_1,x_2,y_2,w\}$ counting from the top left.
-The likelihood $\ell(x_1,y_1,x_2,y_2,w)$ can be written as a function of $\{x_1x_2, x_1+x_2, y_1y_2, y_1+y_2, w\}$ only.
+\textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ counting from the top left.
+The log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$ can be written as a function of $\{x_1x_2, x_1+x_2, y_1y_2, y_1+y_2, w\}$ only.
 In this case, $x_1$ and $x_2$ can be exchanged for one another without changing the value of $\ell$, and any maximum will require $x_1=x_2$.
@@
 \begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
     &=        p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
@@
 \begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
     &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
@@
 \begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
     &\le      p_{0}  \cdot\log(p_{0}) \\
@@
 \begin{align*}
-    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
     &=      p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
@@
 \begin{align*}
-    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
     &=      p_{0}  \cdot\log((1+x')^2    (1+w')(1+y')^2) \\
@@
 \begin{align*}
-    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
     &=      2\cdot\log(1+x') \\
@@
 \begin{align*}
-    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
+    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
     &=      2\cdot\log(2) \\

### 2017-05-02 appendix.tex
text in appendix and figures

@@
 \section{Appendix}
+
+\subsection{Likelihood compuations}
+
+To compute the likelihood of observing a set of data, we need $P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
+The probability of a character change along a branch with fidelity parameter $x$ is $(1-x)/2$, while the probability of a character remaining the same is $(1+x)/2$.
+See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern.
+Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods}.
+Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}
+
+\begin{figure}
+\centering
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[width=.95\textwidth]{farris_like00}
+\caption[short]{$P(\ancestralStateColumn=\emptyset\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
+\end{subfigure}
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[width=.95\textwidth]{farris_like01}
+\caption[short]{$P(\ancestralStateColumn=\{2\}\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
+\end{subfigure}
+\vskip\baselineskip
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[width=.95\textwidth]{farris_like10}
+\caption[short]{$P(\ancestralStateColumn=\{1\}\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
+\end{subfigure}
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[width=.95\textwidth]{farris_like11}
+\caption[short]{$P(\ancestralStateColumn=\{1,2\}\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
+\end{subfigure}
+\caption{Example likelihood computation}
+\label{fig:example_likelihoods}
+\end{figure}
 
@@
 \centering
-\begin{tabular}{|l|l|}
+\begin{tabular}{|l|ll|}
+\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
 \hline
-s   &$P(s)$\\
+& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
     \hline
-0000&$1+x^2+y^2+4xy^2+x^2y^2$\\
-0001&$1+x^2-y^2-x^2y^2$\\
-0010&$1+x^2-y^2-x^2y^2$\\
-0100&$1-x^2+y^2-x^2y^2$\\
-1000&$1-x^2+y^2-x^2y^2$\\
-0011&$1+x^2+y^2-4xy^2+x^2y^2$\\
-0101&$1-x^2-y^2+x^2y^2$\\
-1001&$1-x^2-y^2+x^2y^2$\\
+    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
+    \hline
+     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$          &$(1+x)^2   (1-w)(1-y)^2$\\
+     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
+     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
+     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
+     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+     $\{1,3\}$     &$(1+x)^2   (1+w)(1-y)^2$          &$(1+x)^2   (1-w)(1+y)^2$\\
+     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+     $\{1,2,3\}$   &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
+    \hline
+    \hline
+    &$\{1\}$                             &$\{1,2\}$  \\
+    \hline
+     $\emptyset$   &$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
+     $\{1\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{2\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+     $\{3\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+     $\{1,3\}$     &$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^2$\\
+     $\{2,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+     $\{1,2,3\}$   &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     \hline
 \end{tabular}
-\caption{Site pattern probabilities for Farris zone topology.
-All values are multiplied by $1/8$.}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{|l|l|}
-    \hline
-s   &$P(s)$\\
-    \hline
-0000&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
-0001&$1+x^2y-y^3-x^2y^2$\\
-0010&$1-x^2y+y^3-x^2y^2$\\
-0100&$1+x^2y-y^3-x^2y^2$\\
-1000&$1-x^2y+y^3-x^2y^2$\\
-0011&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
-0101&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
-1001&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
-    \hline
-\end{tabular}    
-\caption{Site pattern probabilities for Felsenstein zone topology.
-All values are multiplied by $1/8$.}
-\label{tab:sitepatprob-fels}
+\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
+All values multiplied by $1/32$.}
+\label{tab:farris_likelihoods}
 \end{table}
@@
 \begin{tabular}{|l|ll|}
-\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k|\alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
+\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_2,t)$}\\
 \hline
@@
     \hline
-    $\siteSplit_j$    &$\emptyset$  00                              &$\{2\}$  01\\
+    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
     \hline
-     $\emptyset$   0000&$(1+x)^2   (1+w)(1+y)^{2*}$          &$(1+x)^2   (1-w)(1-y)^2$\\
-     $\{1\}$       1000&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
-     $\{2\}$       0010&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
-     $\{3\}$       0100&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
-     $\{1,2\}$     0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{1,3\}$     0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$   &$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
-     $\{2,3\}$     1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{1,2,3\}$   0001&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
+     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1-x)^2   (1-w)(1+y)(1-y)$\\
+     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1-y)^2$\\
+     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1+x)^2   (1-w)(1+y)(1-y)$\\
+     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1+x)^2   (1-w)(1+y)^2$\\
+     $\{1,3\}$     &$(1+x)^2   (1+w)(1-y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1-x)^2   (1-w)(1+y)^2$\\
+     $\{1,2,3\}$   &$(1+x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1+y)^2$\\
     \hline
     \hline
-    &$\{1\}$  10                           &$\{1,2\}$  11\\
+    &$\{1\}$                             &$\{1,2\}$  \\
     \hline
-     $\emptyset$   0000&$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
-     $\{1\}$       1000&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{2\}$       0010&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-     $\{3\}$       0100&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{1,2\}$     0011&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
-     $\{1,3\}$     0101&$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
-     $\{2,3\}$     1001&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
-     $\{1,2,3\}$   0001&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-    \hline
-\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k|\alignmentColumn=\siteSplit_j,\tau_2,t)$}\\
-\hline
-& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
-    \hline
-    $\siteSplit_j$    &$\emptyset$  00                              &$\{2\}$  01\\
-    \hline
-     $\emptyset$   0000&$(1+x)^2   (1+w)(1+y)^{2*}$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{1\}$       1000&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1-x)^2   (1-w)(1+y)(1-y)$\\
-     $\{2\}$       0100&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1-y)^2$\\
-     $\{3\}$       0010&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$\\
-     $\{1,2\}$     0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
-     $\{1,3\}$     0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$    &$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$\\
-     $\{2,3\}$     1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1-x)^2   (1-w)(1+y)^{2\ddagger}$\\
-     $\{1,2,3\}$   0001&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$\\
-    \hline
-    \hline
-    &$\{1\}$  10                           &$\{1,2\}$  11\\
-    \hline
-     $\emptyset$   0000&$(1+x)(1-x)(1-w)(1+y)(1-y)$             &$(1-x)^2   (1+w)(1-y)^2$\\
-     $\{1\}$       1000&$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$  &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{2\}$       0100&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$      &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-     $\{3\}$       0010&$(1-x)^2   (1-w)(1+y)(1-y)$             &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{1,2\}$     0011&$(1-x)^2   (1-w)(1-y)^2$                &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
-     $\{1,3\}$     0101&$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$  &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
-     $\{2,3\}$     1001&$(1+x)^2   (1-w)(1-y)^{2\ddagger}$      &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
-     $\{1,2,3\}$   0001&$(1+x)(1-x)(1-w)(1-y)^2$                &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+     $\emptyset$   &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1-x)^2   (1+w)(1-y)^2$\\
+     $\{1\}$       &$(1+x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{2\}$       &$(1+x)(1-x)(1-w)(1+y)^2$           &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+     $\{3\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{1,2\}$     &$(1-x)^2   (1-w)(1-y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+     $\{1,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1-x)^2   (1+w)(1+y)^2$\\
+     $\{2,3\}$     &$(1+x)^2   (1-w)(1-y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+     $\{1,2,3\}$   &$(1+x)(1-x)(1-w)(1-y)^2$           &$(1-x)^2   (1+w)(1+y)(1-y)$\\
 \hline
 \end{tabular}
-\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
-Maxima determined row-wise (i.e., by site pattern).
-All values multiplied by $1/32$.
-Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
-\label{tab:likelihoods}
+\caption{Likelihood calculations for all site patterns and internal states of Felsenstein topology.
+All values multiplied by $1/32$.}
+\label{tab:fels_likelihoods}
 \end{table}

### 2017-05-02 joint_inf.tex
text in appendix and figures

@@
 \begin{tabular}{|l|l|l|}
-\multicolumn{3}{c}{Classical}\\
-    \hline
-$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t^*)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t^*)$\\
-    \hline
-    $\emptyset$    &$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
-    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
-    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
-    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
-    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
-    $\{1,3\}$      &$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
-    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
-    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
-    \hline
-\multicolumn{3}{c}{Modified}\\
     \hline
@@
 All values multiplied by $1/32$.
-Likelihoods with multiple entries will have maxima determined by unknown branch length parameters.}
+Likelihoods with multiple entries will have maxima determined by unknown branch length parameters.
+See Appendix for full calculations.}
 \label{tab:likelihoods}
@@
 Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
-For compactness, define, e.g., $p_0 = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})$, keeping in mind it is a function of parameters $x,y$.
+For compactness, define
+$$
+p_0 = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})
+$$
+with similar definitions for the remaining generating probabilities.
 The full Farris log likelihood can take one of three values---assume for now the we are only interested in the ancestral state split $\{2\}$ for the site split $\{1,3\}$.
@@
     \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &=      p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
+    &\ge      p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
     &\qquad + p_{1}  \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
@@
     \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &=      p_{0}  \cdot\log((1+x')^2    (1+w')(1+y')^2) \\
+    &\ge      p_{0}  \cdot\log((1+x')^2    (1+w')(1+y')^2) \\
     &\qquad + p_{1}  \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
@@
     \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &=      2\cdot\log(1+x') \\
+    &\ge      2\cdot\log(1+x') \\
     &\qquad + (2-b)  \cdot\log(1+y') \\
@@
     \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &=      2\cdot\log(2) \\
+    &\ge      2\cdot\log(2) \\
     &\qquad + (2-b)  \cdot\log(2-b) \\
@@
 
-Ideally, we want the region where
+We want the region where
 $$
@@
 $$
-Currently being slightly lazy, I'm considering values for $\{x,y\}$ where all of
-$$
-L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) \ge L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}),
-$$
-$$
-L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) \ge L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy})
-$$
-and
-$$
-L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) \ge L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy})
-$$
-hold.
-Since these are three inequalities in two variables, we can plot them analytically---see Fig.~\ref{fig:inconsistency-farris}.
+Since this is an inequality in two variables, we can plot them analytically---see Fig.~\ref{fig:inconsistency-farris}.
 
@@
 
+\input{appendix}
+
 \bibliographystyle{plain}

### 2017-05-24 joint_inf.tex
Moving appendix to very end.
pass through

@@
 Classical maximum-likelihood (ML) estimation in phylogenetics operates by integrating out possible ancestral states at the internal nodes of the tree.
-Recently \cite{Neher2017} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences.
-This is attractive from a computational perspective: such joint ML inference can proceed according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
-This latter step is simpler and more computationally efficient when conditioning on ancestral sequences compared to the case in which ancestral sequences are marginalized out.
+Recently, \cite{Neher2017} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences.
+This is attractive from a computational perspective: such joint ML inference proceeds according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
+This latter conditional optimization is simpler and more computationally efficient than marginalizing out ancestral sequences and then performing optimization.
 
-However, existing consistency proofs for phylogenetics \cite{RoyChoudhury2015-ta} do not apply to the setting of joint inference of trees and ancestral sequences.
-There are examples in statistics, such as the Neyman-Scott paradox \cite{Neyman1948-tt}, in which estimation of additional parameters destroys consistency.
-This raises the question of whether joint inference of trees and ancestral sequences enjoys good statistical properties.
+Existing consistency proofs for phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
+These proofs do not readily extend to include estimating ancestral states.
+Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-ta} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
 
-In this paper we show that the joint inference of trees and ancestral sequences is not consistent in general.
+In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
 To do so, we follow a strategy of using simple models and four taxa trees.
-Here we use a binary symmetric model and simulate on the ``Farris zone'' \cite{Siddall1998-hq} tree.
-We use bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference converges on the wrong tree.
+We use a binary symmetric model and simulate on the ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference converges to the wrong tree.
 
@@
 
-We will be considering the standard phylogenetic likelihood on unrooted phylogenetic trees (cite Felsenstein book).
+We consider the standard phylogenetic likelihood on unrooted phylogenetic trees \cite{Felsenstein2004}.
 We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
 Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
-Assume we observe $\nCols$ samples of character data (i.e.\ an alignment with $\nCols$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
-There are also corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ distributed as $\ancestralStateColumnRV$.
+Assume we observe $\nCols$ samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
+The corresponding unobserved ancestral states are $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ and distributed as $\ancestralStateColumnRV$.
 
@@
 
-For a topology $\tau$ and branch lengths $t$ we start with the fully-observed likelihood (making the usual independence assumption between sites):
+For a topology $\tau$ and branch lengths $t$, we make the usual assumption of independence between sites, obtaining the likelihood
 \begin{equation}
@@
 One tantalizing approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
-We can perform this by computing the profile likelihood
+We compute the profile likelihood
 \begin{equation}
@@
 \end{equation}
-then estimating the topology and branch lengths via
+then estimate the topology and branch lengths via
 \begin{equation}
 \label{eq:profile_likelihood_topology_bl}
-(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment)
+(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment),
 \end{equation}
-while using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
-Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
-In the general case, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
+using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
+Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and \eqref{eq:profile_likelihood_topology_bl} recovers the joint ML values of the unknown parameters.
+In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
 For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.
@@
 Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
-For likelihood calculation under the binary symmetric model, a given $\alignmentColumn_i$ can be described as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
-Indeed, we can start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
-Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we can take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.
+For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
+Indeed, we start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
+Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.
 
 Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ where $\mathcal{P}$ denotes the power set.
-For topology $\tau$, we also define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization of $\ancestralStateColumn_i$.
-The probability of observing an ancestral state split, conditional on a site split, will not be invariant to taking its complement, and thus the need for enumerating the entire set of internal nodes.
+For topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization of $\ancestralStateColumn_i$.
+The probability of observing an ancestral state split, conditional on a site split, is not invariant to taking its complement, and thus the need for enumerating the entire set of internal nodes.
 The number of unique ancestral state splits is $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$.
 
-With this split formulation, define mappings from patterns to splits as described above:
+With this split formulation, define mappings from patterns to splits as
 $$
@@
 $$
-Given a site-pattern-valued random variable $\alignmentColumnRV$ and an ancestral-state-valued random variable $\ancestralStateColumnRV$, define random variables
+Given a site pattern--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, define random variables
 $$
-\siteSplitRV := \patternToSplit(\alignmentColumnRV), \ \ancestralSplitRV := \ancestralToSplit(\ancestralStateColumnRV).
+\siteSplitRV := \patternToSplit(\alignmentColumnRV), \ \ancestralSplitRV := \ancestralToSplit(\ancestralStateColumnRV)
 $$
-% Need to introduce \siteSplit_j and \ancestralSplit_k , no?
-We can now write the $i$th factor of \eqref{eq:full_likelihood} as
+that take corresponding realizations $\siteSplit_j$ and $\ancestralSplit_k$ for some $j$ and $k$.
+We write the $i$th factor of \eqref{eq:full_likelihood} as
 \begin{align*}
     P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
-    &= P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)
+    &= P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t).
 \end{align*}
-for some $j$ and $k$.
 Taking the maximum across internal states yields
@@
 \end{align*}
-Given $(\tau, t)$ there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
+Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
 \begin{align*}
@@
 Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
-This allows us to write the likelihood in \eqref{eq:profile_likelihood} as a product over site patterns as opposed to sites, i.e.,
+Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
 \begin{align}
@@
                              &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
-%EM Indexing broken here: we are taking a product over i, but i doesn't appear in the expression.
-    %das: This is intentional, though: it allows us to proceed to the following line, grouping factors by site pattern since their probabilities don't depend on $i$.
-I can try to think of how to make this more clear, but I thought it was a pretty key step to be able to pass to site patterns, and the reason for the machinery involved in defining ancestral state partitions.
-%EM My suggestion is to use \psi(y_i) until you switch to a product indexed by the j's.
                              &= \prod_{i=1}^{\nCols} \ \max_{\ancestralToSplit(\ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
@@
 \end{align}
-%EM think we should say "where the site split or its complement appears."?
 where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.
@@
 Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}.
-The set of all possible character assignments can be described as
+The set of all possible character assignments is
 \begin{align*}
@@
 \end{align*}
-Had we not identified character complements, we would need to consider the additional splits
+Since we identify character complements, we do not consider the additional splits
 $$
@@
 $$
-but the symmetry of the binary character model allows us to focus just on the elements of $\mathcal{P}(\siteSplitSet)$.
+the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
 This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
@@
 $$
-We cannot similarly identify complements in this case since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
+We cannot similarly identify complements for ancestral states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
 $$
@@
 For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
-A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; for the $j$th site split, define $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ as the set of most likely ancestral splits for that particular site split, topology and set of branch lengths.
+A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; for the $j$th site split, define $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ as the set of most likely ancestral splits for that particular site split, topology and set of branch lengths.
 As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
-%EM Would it be worth enumerating the site splits ahead of time? It seems a little funky to be calling them j each time.
 The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
@@
 $$
-Thus the set of most likely ancestral states contains a single element in this case, here $\ancestralSplitPartition_1(\tau, p)=\{\emptyset\}$.
+The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, p)=\{\emptyset\}$.
 Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, p)$ we have
@@
 $$
-In this case, the set of most likely ancestral states is $\ancestralSplitPartition_6(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_6\in\ancestralSplitPartition_6(\tau, p)$,
+Here, the set of most likely ancestral states is $\ancestralSplitPartition_6(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_6\in\ancestralSplitPartition_6(\tau, p)$,
 $$
@@
 We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
-%EM As described IRL, a little strange to be redefining (9) with an extra variable.
-After reading the below, I think it makes most sense to just refer to (9) indirectly but make this a stand-alone definition.
 Let
@@
 \begin{equation}
-%EM did you mean to drop the eta from the first term of this equation?
     \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)].
@@
 $$
-%EM IMHO \ell^p is so standard that it's confusing to see it in another setting. Perhaps \tilde{\ell}?
 \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
@@
 \end{equation}
-Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
 For data generated from $(\tau^*, t^*)$, if
@@
 \label{eq:inconsistency_inequality}
-%EM Suggest \hat{\fullAncestralSplitPartitions}_1, etc.
 \ell_{\tau^*,t^*}(\tau', \hat{t}_1; \hat{\fullAncestralSplitPartitions}_1) > \ell_{\tau^*,t^*}(\tau^*, \hat{t}_2; \hat{\fullAncestralSplitPartitions}_2)
@@
 for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}_1\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}_2\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
-By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any potential example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
+By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
 In the next section, we show that such examples exist.
@@
 
-Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified}.
-We obtain an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} as
+Our approach seeks bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified}.
+We obtain
 $$
@@
 $$
-using Gibbs's inequality
+as an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
 $$
@@
 $$
-Similarly, we have a lower bound as
+Similarly, we have the lower bound
 $$
@@
 Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
-%EM clarify that we aren't assuming that the x and y in t are the same as those appearing in t^*?
-Call $t=\{x',y',w'\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ is the case where the bottom three branches all share the same parameter, the classical construction of this topology.
+Call $t=\{x',y',w'\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ constrains the bottom three branches to share the same parameter, the classical construction of this topology.
 
-We will parameterize these trees \textbf{not with branch lengths}, but rather such that the probability of a substitution on a branch with parameter $x'$ is $p_x'=(1-x')/2$.
-%EM (Scratching head) isn't p_x=1-2x backwards? Then p_x will take values in the range -1 to 1.
+We parameterize these trees \textbf{not with branch lengths}, but, rather, such that the probability of a substitution on a branch with parameter $x'$ is $p_x'=(1-x')/2$.
 Here, $\{p_x',p_y',p_w'\}\in[0,1/2]^3$ and $\{x',y',w'\}\in[0,1]^3$.
 This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
-Fidelities have nice algebraic properties, as can be seen via the Hadamard transform.
+Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.
 
@@
 \textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ counting from the top left.
-The log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$ can be written as a function of $\{x_1x_2, x_1+x_2, y_1y_2, y_1+y_2, w\}$ only.
-In this case, $x_1$ and $x_2$ can be exchanged for one another without changing the value of $\ell$, and any maximum will require $x_1=x_2$.
-A similar argument can be made for $y_1$ and $y_2$.
-For the Felsenstein topolgy this is not the case, but, since we are interested in a lower bound for this topology, we will simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
+The log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$ is a function of $\{x_1x_2, x_1+x_2, y_1y_2, y_1+y_2, w\}$ only.
+By symmetry, exchanging $x_1$ and $x_2$ does not change the value of $\ell$, requiring $x_1=x_2$ at the maximum.
+A similar property holds for $y_1$ and $y_2$.
+The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
 
 Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies under these two topologies, calculated using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}.
-Table~\ref{tab:likelihoods} contains calculations of likelihood values for fixed site patterns and topologies, with the maximum values over ancestral state patterns noted.
+Table~\ref{tab:likelihoods} contains calculations of likelihood values for fixed site patterns and topologies that have been maximized over ancestral state patterns.
+Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} contain the values of the likelihood for all possible ancestral states.
 
@@
 All values multiplied by $1/32$.
-Likelihoods with multiple entries will have maxima determined by unknown branch length parameters.
+Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
 See Appendix for full calculations.}
@@
 with similar definitions for the remaining generating probabilities.
-The full Farris log likelihood can take one of three values---assume for now the we are only interested in the ancestral state split $\{2\}$ for the site split $\{1,3\}$.
-We then write the log likelihood as
+The full Farris log likelihood takes one of three values---assume for now the we are only interested in the ancestral state split $\{2\}$ for the site split $\{1,3\}$.
+The log likelihood is
 \begin{align*}
@@
 \end{align*}
-Seeking an upper bound, we make use of the fact that, for $x',y'\in[0,1]$,
+Seeking an upper bound, we use the fact that, for $x',y'\in[0,1]$,
 $$
@@
 
-The other two possibilties of ancestral state split for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
+The other two possibile ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
 Call
@@
 $$
-Then, for these cases,
+Then,
 \begin{align*}
@@
 Replace all $(1+w)$ terms with $(1-w)$ and focus just on the partial likelihood.
-Then if $(1+x)(1-y) > (1-x)(1+y)$,
+Then, if $(1+x)(1-y) > (1-x)(1+y)$,
 \begin{align*}
@@
 \end{align*}
-for the first lower bound, but, upon maximizing, we obtain the same value for both cases, namely
+for the first lower bound, but, upon maximizing, we obtain
 \begin{align*}
@@
 \end{align*}
-maximized at $\{1,1-b,1-2p_{0}\}$.
+maximized at $\{1,1-b,1-2p_{0}\}$, which is the same if $(1+x)(1-y) > (1-x)(1+y)$ or $(1+x)(1-y) < (1-x)(1+y)$.
 Our lower bound is
@@
 $$
-Since this is an inequality in two variables, we can plot them analytically---see Fig.~\ref{fig:inconsistency-farris}.
+Since this is an inequality in two variables, we plot them analytically---see Fig.~\ref{fig:inconsistency-farris}.
 
@@
     Regions of inconsistency.
-    Due to the looseness of the upper and lower bounds, the parameters in the top right do not necessarily indicate consistency, though all parameters in the labeled region will result in an inconsistency.
+    Due to the looseness of the upper and lower bounds, the parameters in the top right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
 }

### 2017-06-06 joint_inf.tex
Question

@@
 Recently, \cite{Neher2017} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences.
-This is attractive from a computational perspective: such joint ML inference proceeds according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
+This is attractive from a computational perspective: such joint ML inference can proceed according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
 This latter conditional optimization is simpler and more computationally efficient than marginalizing out ancestral sequences and then performing optimization.
@@
 These proofs do not readily extend to include estimating ancestral states.
-Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-ta} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
+Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
 
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we follow a strategy of using simple models and four taxa trees.
-We use a binary symmetric model and simulate on the ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference converges to the wrong tree.
+To do so, we use a binary symmetric model and imagine data being generated on the four taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the large-data limit.
 
@@
 
-One tantalizing approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
-We compute the profile likelihood
+An alternative approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
+One can compute the profile likelihood
 \begin{equation}
@@
 
-Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ where $\mathcal{P}$ denotes the power set.
 For topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization of $\ancestralStateColumn_i$.
-The probability of observing an ancestral state split, conditional on a site split, is not invariant to taking its complement, and thus the need for enumerating the entire set of internal nodes.
-The number of unique ancestral state splits is $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$.
+In this case the entire set of internal nodes must be enumerated: the probability of observing an ancestral state split, conditional on a site split, is not invariant to taking its complement.
 
-With this split formulation, define mappings from patterns to splits as
+We enumerate the site splits $\siteSplit_j$ of which there are $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ in total where $\mathcal{P}$ denotes the power set.
+Similarly we enumerate ancestral splits $\ancestralSplit_k$ of which there are $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$ in total.
+
+This split formulation defines mappings from patterns to splits as
 $$
@@
 \begin{align*}
+%EM Wait, is this first statement actually true? It seems impossible because multiple y_i's collaps to a single psi(y_i). At least it needs a factor of 2, but it seems more complex than that.
     P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
-    &= P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t).
+    &= P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)
 \end{align*}
+for some $j$ and $k$.
 Taking the maximum across internal states yields
@@
                              &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
+%EM My suggestion is to use \psi(y_i) until you switch to a product indexed by the j's.
                              &= \prod_{i=1}^{\nCols} \ \max_{\ancestralToSplit(\ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
@@
 
-The other two possibile ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
+The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
 Call
@@
 
-\input{appendix}
-
 \bibliographystyle{plain}
@@
 
+\newpage
+
+\input{appendix}
+
 \end{document}

### 2017-06-07 joint_inf.tex
I think I fixed it?

@@
  $$
-\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet), \ \ancestralToSplit:\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet).
+ \patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet).
  $$
-Given a site pattern--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, define random variables
+Given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
 $$
-\siteSplitRV := \patternToSplit(\alignmentColumnRV), \ \ancestralSplitRV := \ancestralToSplit(\ancestralStateColumnRV)
+\siteSplitRV := \patternToSplit(\alignmentColumnRV)
 $$
-that take corresponding realizations $\siteSplit_j$ and $\ancestralSplit_k$ for some $j$ and $k$.
-We write the $i$th factor of \eqref{eq:full_likelihood} as
+that takes corresponding realizations $\siteSplit_j$ for some $j$.
+
+%EM Wait, is this first statement actually true? It seems impossible because multiple y_i's collaps to a single psi(y_i). At least it needs a factor of 2, but it seems more complex than that.
+The $i$th factor of \eqref{eq:full_likelihood} has the property
+$$
+P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
+$$
+As a consequence of assuming a binary symmetric model, taking complements yields
 \begin{align*}
-%EM Wait, is this first statement actually true? It seems impossible because multiple y_i's collaps to a single psi(y_i). At least it needs a factor of 2, but it seems more complex than that.
-    P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
-    &= P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)
+    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
+                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
+\end{align*}
+for some $j$.
+For the other term, we need to define
+$$
+\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
+$$
+that takes both the ancestral state column and site column as input.
+Given $\alignmentColumnRV$ as well as an ancestral state--valued random variable $\ancestralStateColumnRV$ define
+$$
+\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV).
+$$
+Then
+\begin{align*}
+    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
+                                                                                                           &= P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)
 \end{align*}
 for some $j$ and $k$.
+
 Taking the maximum across internal states yields
@@
 %EM My suggestion is to use \psi(y_i) until you switch to a product indexed by the j's.
-                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralToSplit(\ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
-                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralToSplit(\ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} P(\ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
+                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
+                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                              &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}

### 2017-06-07 joint_inf.tex
slight rewording

@@
 for some $j$.
-For the other term, we need to define
+For the other term, we can define
 $$
@@
 $$
-that takes both the ancestral state column and site column as input.
-Given $\alignmentColumnRV$ as well as an ancestral state--valued random variable $\ancestralStateColumnRV$ define
+and
 $$
@@
 $$
-Then
+for an ancestral state--valued random variable $\ancestralStateColumnRV$ such that
 \begin{align*}

### 2017-06-07 joint_inf.tex
add example

@@
 \end{align*}
-where each set indicates the tips at which a character is assigned the character $1$.
+where each set indicates the tips assigned the character $1$.
 For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
@@
 $$
-We cannot similarly identify complements for ancestral states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
+The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
+For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
+Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
+We cannot identify complements for ancestral states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
 $$

### 2017-06-08 joint_inf.tex
a few additions

@@
 %EM Wait, is this first statement actually true? It seems impossible because multiple y_i's collaps to a single psi(y_i). At least it needs a factor of 2, but it seems more complex than that.
-The $i$th factor of \eqref{eq:full_likelihood} has the property
+For the $i$th factor of \eqref{eq:full_likelihood},
 $$
@@
 for some $j$ and $k$.
+Then, the $i$th factor of \eqref{eq:full_likelihood} has the property
+$$
+P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \propto P(\siteSplitRV=\siteSplit_j, \tau, t) \cdot P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t).
+$$
 
@@
 \begin{align*}
-\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &=
-\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
-&= P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t).
+    \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) &=
+P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t).
 \end{align*}

### 2017-06-20 joint_inf.tex
More notation suggestions.

@@
 
-%EM Wait, is this first statement actually true? It seems impossible because multiple y_i's collaps to a single psi(y_i). At least it needs a factor of 2, but it seems more complex than that.
 For the $i$th factor of \eqref{eq:full_likelihood},
@@
 For the other term, we can define
+%EM Need a definition for \ancestralToSplit here.
 $$
@@
 \end{equation}
-Define the the divergence quantity
+Define the divergence quantity
 $$
@@
 \end{equation}
-For data generated from $(\tau^*, t^*)$, if
+For data generated from $(\tau^*, t^*)$, we have an inconsistency if
 \begin{equation}
@@
 \end{equation}
-for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}_1\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}_2\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
+for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}_1\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}_2\}$ estimated by maximizing $\tilde{\ell}_{\tau^*,t^*}$ with respect to the branch lengths and the internal states for $\tau = \tau'$ and $\tau^*$, respectively.
 By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
@@
 We obtain
+%EM Does it really make sense to talk about taking the max over \fullAncestralSplitPartitions(\tau,t)?
+This is already defined.
+I suggest just taking the max over \fullAncestralSplitPartitions as a variable.
 $$

### 2017-06-20 joint_inf.tex
definition and modifying etas

@@
 for some $j$.
-For the other term, we can define
+For the other term, we define
 %EM Need a definition for \ancestralToSplit here.
@@
 $$
-and
+as a mapping of ancestral states to ancestral state splits.
+In this case, the ancestral states we consider depend on whether their corresponding tip labels had their complements taken or not: if we exclude a set of tip labels $\alignmentColumn$ from the set of site patterns by taking its complement, we let $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ be the complement of $\ancestralStateColumn$; otherwise $\ancestralStateColumn$ is unchanged.
+We define
 $$
-\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV).
+\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
 $$
@@
 \label{eq:inconsistency_inequality}
-\ell_{\tau^*,t^*}(\tau', \hat{t}_1; \hat{\fullAncestralSplitPartitions}_1) > \ell_{\tau^*,t^*}(\tau^*, \hat{t}_2; \hat{\fullAncestralSplitPartitions}_2)
+\ell_{\tau^*,t^*}(\tau', \hat{t}_1; \fullAncestralSplitPartitions(\tau', \hat{t}_1)) > \ell_{\tau^*,t^*}(\tau^*, \hat{t}_2; \fullAncestralSplitPartitions(\tau^*, \hat{t}_2))
 \end{equation}
-for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}_1\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}_2\}$ estimated by maximizing $\tilde{\ell}_{\tau^*,t^*}$ with respect to the branch lengths and the internal states for $\tau = \tau'$ and $\tau^*$, respectively.
+for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\fullAncestralSplitPartitions(\tau', \hat{t}_1)\}$ and $\{\hat{t}_2,\fullAncestralSplitPartitions(\tau^*, \hat{t}_2)\}$ estimated by maximizing $\tilde{\ell}_{\tau^*,t^*}$ with respect to the branch lengths and the internal states for $\tau = \tau'$ and $\tau^*$, respectively.
 By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
@@
 $$
-\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \le
+\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \le
     \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
-    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \hat{\fullAncestralSplitPartitions})
+    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t}))
 $$
@@
 $$
-\{\hat{t},\hat{\fullAncestralSplitPartitions}\} = \argmax_{t,\fullAncestralSplitPartitions(\tau,t)} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
+\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
 $$
@@
 $$
-\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \ge
+\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \ge
     \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
-    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \hat{\fullAncestralSplitPartitions}).
+    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t})).
 $$

### 2017-06-23 joint_inf.tex
A pass through site split formulation.

@@
-\documentclass[a4paper]{article}
+\documentclass{article}
 
@@
 \usepackage[T1]{fontenc}
-
-%% Sets page size and margins
-\usepackage[a4paper,top=3cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
 
@@
 For the other term, we define
-%EM Need a definition for \ancestralToSplit here.
 $$
@@
 $$
-as a mapping of ancestral states to ancestral state splits.
-In this case, the ancestral states we consider depend on whether their corresponding tip labels had their complements taken or not: if we exclude a set of tip labels $\alignmentColumn$ from the set of site patterns by taking its complement, we let $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ be the complement of $\ancestralStateColumn$; otherwise $\ancestralStateColumn$ is unchanged.
+as a mapping of ancestral states and tip states to ancestral state splits.
+It is defined by if the tip states have their complements taken or not: if we exclude a set of tip labels $\alignmentColumn$ from the set of site patterns by taking its complement, we let $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ be the complement of $\ancestralStateColumn$; otherwise it is simply $\ancestralStateColumn$.
+Note $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective for each $\alignmentColumn_i$.
 We define
@@
 $$
-for an ancestral state--valued random variable $\ancestralStateColumnRV$ such that
+for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
+By considering the case requiring complementing and that not requiring complementing separately,
 \begin{align*}
-    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
-                                                                                                           &= P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)
-\end{align*}
-for some $j$ and $k$.
-Then, the $i$th factor of \eqref{eq:full_likelihood} has the property
-$$
-P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \propto P(\siteSplitRV=\siteSplit_j, \tau, t) \cdot P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t).
-$$
-
-Taking the maximum across internal states yields
-\begin{align*}
-    \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) &=
-P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t).
+    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
 \end{align*}
@@
 \end{align*}
+By the surjectivity of $\ancestralToSplit(\alignmentColumn_i, \cdot)$, we also have
+\begin{align*}
+\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
+\end{align*}
+
 Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
@@
                              &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
-%EM My suggestion is to use \psi(y_i) until you switch to a product indexed by the j's.
-                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
-                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
+                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
+                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                              &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
@@
 \paragraph{Example}
-%EM My long term suggestion is for you to put this in example in the main text and the above in the appendix. (But we should probably check with Vladimir.)
 \begin{figure}
@@
 $$
-\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
+\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
 $$
@@
 Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
-Call $t=\{x',y',w'\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ constrains the bottom three branches to share the same parameter, the classical construction of this topology.
-
-We parameterize these trees \textbf{not with branch lengths}, but, rather, such that the probability of a substitution on a branch with parameter $x'$ is $p_x'=(1-x')/2$.
-Here, $\{p_x',p_y',p_w'\}\in[0,1/2]^3$ and $\{x',y',w'\}\in[0,1]^3$.
+We parameterize the branches of these trees \textbf{not with branch lengths}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$.
 This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
@@
 
+Let $\{x',y',w'\}\in[0,1]^3$ be three fidelities and $\{p_x',p_y',p_w'\}\in[0,1/2]^3$ the corresponding substitution probabilities.
+Call $t=\{x',y',w'\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ constrains the bottom three branches to share the same parameter, the classical construction of this topology.
+
 %das: did this on paper, but can write it out.
-\textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ counting from the top left.
+\textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ with edge ordering in the order of the taxa, then the internal branch last.
+%EM Yeah, this next statement feels like it could use some justification, and or maybe you could skip to the following statement.
 The log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$ is a function of $\{x_1x_2, x_1+x_2, y_1y_2, y_1+y_2, w\}$ only.
@@
 $$
+%EM I'd suggest p_\emptyset for this particular one, but perhaps that will throw off your spacing below.
 p_0 = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})

### 2017-06-24 joint_inf.tex
Minor changes and suggestions.

@@
 as a mapping of ancestral states and tip states to ancestral state splits.
-It is defined by if the tip states have their complements taken or not: if we exclude a set of tip labels $\alignmentColumn$ from the set of site patterns by taking its complement, we let $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ be the complement of $\ancestralStateColumn$; otherwise it is simply $\ancestralStateColumn$.
+It is defined by if the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is not contained in $\siteSplitSet$ and thus we must take a complement to become a site split, we let $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ be the complement of $\ancestralStateColumn$; otherwise it is simply $\ancestralStateColumn$.
 Note $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective for each $\alignmentColumn_i$.
@@
 \begin{align}
-L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \\
-                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
-                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
-                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
+L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
+                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
+                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
+                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                              &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
@@
 the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
+%EM Is it indicated anywhere how the internal nodes are numbered? I would assume that 1 would be connected to leaf 1. If that's true please indicate here.
 This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
@@
 Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
-We parameterize the branches of these trees \textbf{not with branch lengths}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$.
+We parameterize the branches of these trees \textbf{not with the standard notion of branch length in terms of number of substitutions per site}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$.
 This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
@@
 \begin{table}
+%EM I wonder why you use primes on all variables here but not in tab:likelihoods.
 \centering
@@
                     \emptyset\\
+                    %EM Shouldn't this be \{1\}\\ ? Or perhaps the ordering of internal nodes is not what I thought?
                     \{2\}\\
@@
 
-\subsubsection{Upper bound for Farris likelihood}
+\subsubsection{Upper bound for Farris tree likelihood}
 
@@
 with similar definitions for the remaining generating probabilities.
-The full Farris log likelihood takes one of three values---assume for now the we are only interested in the ancestral state split $\{2\}$ for the site split $\{1,3\}$.
-The log likelihood is
+The full Farris tree log likelihood takes one of three values depending on branch lengths---assume for now the we are only interested in the ancestral state split $\{2\}$ for the site split $\{1,3\}$.
+Directly substituting in \eqref{eq:site_pattern_likelihood}, the log likelihood is
 \begin{align*}
@@
 \end{align*}
-Seeking an upper bound, we use the fact that, for $x',y'\in[0,1]$,
-$$
-(1-x'^2+y'^2-x'^2y'^2) = (1-x'^2)(1+y'^2) = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y),
-$$
-with similar substitutions for $(1+x'^2-y'^2-x'^2y'^2)$ and $(1-x'^2-y'^2+x'^2y'^2)$.
-Doing so results in
+Seeking an upper bound, we use the facts that, for $x',y'\in[0,1]$,
+\begin{align*}
+(1-x'^2+y'^2-x'^2y'^2) & = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y') \\
+(1+x'^2-y'^2-x'^2y'^2) & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y').
+\end{align*}
+Factoring and making these substitutions results in
 \begin{align*}
@@
     &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
-    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
     &\qquad + p_{1}  \cdot\log((1+x')(1-x')(1+y')) \\
@@
     &\qquad + p_{12} \cdot\log((1+x')(1-x')(1+y')(1-y')) \\
+    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
     &\qquad + p_{23} \cdot\log((1+x')(1-x')(1+y')(1-y')) \\
@@
 
-\subsubsection{Lower bound for Felsenstein partial likelihood}
+\subsubsection{Lower bound for Felsenstein tree partial likelihood}
 
-We construct a similar lower bound on the Felsenstein partial likelihood.
+We construct a similar lower bound on the Felsenstein tree partial likelihood.
 Replace all $(1+w)$ terms with $(1-w)$ and focus just on the partial likelihood.

### 2017-06-24 joint_inf.tex
Adding TreeTime reference

@@
 Classical maximum-likelihood (ML) estimation in phylogenetics operates by integrating out possible ancestral states at the internal nodes of the tree.
-Recently, \cite{Neher2017} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences.
+Recently, \cite{Sagulenko2017-jo} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences.
 This is attractive from a computational perspective: such joint ML inference can proceed according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
@@
 Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
+Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$ and limit themselves to optimizing branch lengths using this method, this motivates understanding when exactly this approximation breaks down, and what happens when we optimize branch lengths and tree topology using this approximation.
 
@@
 In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
-For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.
+For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.
+
+%EM It seems straightforward to come up with an example of where branch length estimation is inconsistent.
+It seems like any time a state at an internal node is uncertain we will get different branch lengths by fixing it to the ML state.
 

### 2017-07-07 joint_inf.tex
Halp
oops, gibbs actually doesn't apply
notation and form of likelihood

@@
 $$
+Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{2\}$ and internal node $\{2\}$ connected to leaves $\{3\}$ and $\{4\}$.
 The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
@@
 %EM Yeah, this next statement feels like it could use some justification, and or maybe you could skip to the following statement.
-The log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$ is a function of $\{x_1x_2, x_1+x_2, y_1y_2, y_1+y_2, w\}$ only.
-By symmetry, exchanging $x_1$ and $x_2$ does not change the value of $\ell$, requiring $x_1=x_2$ at the maximum.
+Using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}, we calculate the generating probabilities the Farris tree topology.
+For site split $\emptyset$,
+\begin{align*}
+    P(\siteSplit=\emptyset|\tau_1, \tilde{t}) & = \frac{1}{8} \cdot (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
+                                              & = \frac{1}{8} \cdot (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
+                                              & = \frac{1}{8} \cdot (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
+\end{align*}
+and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
+All other generating probabilities will differ only in the signs of each term.
+For example, for site split $\{1\}$ we have
+\begin{align*}
+    P(\siteSplit=\{1\}|\tau_1, \tilde{t}) & = \frac{1}{8} \cdot (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
+\end{align*}
+and for site split $\{3\}$ we have
+\begin{align*}
+    P(\siteSplit=\{3\}|\tau_1, \tilde{t}) & = \frac{1}{8} \cdot (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
+\end{align*}
+meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
+The corresponding possibilities for the likelihood values are
+$$
+\left\{\begin{array}{l}
+        (1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2) \\
+        (1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2) \\
+        (1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2) \\
+        (1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2)
+\end{array}\right.
+$$
+for site split $\{1\}$ and
+$$
+\left\{\begin{array}{l}
+        (1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2) \\
+        (1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2) \\
+        (1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2) \\
+        (1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2)
+\end{array}\right.
+$$
+for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.
+
+The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
+The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
+By symmetry exchanging $x_1$ and $x_2$ does not change the value of $\ell$, requiring $x_1=x_2$ at the maximum.
 A similar property holds for $y_1$ and $y_2$.
@@
 
-Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies under these two topologies, calculated using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}.
+Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies under these two topologies.
 Table~\ref{tab:likelihoods} contains calculations of likelihood values for fixed site patterns and topologies that have been maximized over ancestral state patterns.
@@
     $\emptyset$&
-    $(1+x)^2   (1+w)(1+y)^2$\\
+    $(1+x')^2   (1+w')(1+y')^2$\\
      $\{1\}$    &
     $\emptyset$&
-    $(1+x)(1-x)(1+w)(1+y)^2$\\
+    $(1+x')(1-x')(1+w')(1+y')^2$\\
      $\{2\}$    &
     $\emptyset$&
-    $(1+x)^2   (1+w)(1+y)(1-y)$\\
+    $(1+x')^2   (1+w')(1+y')(1-y')$\\
      $\{3\}$    &
     $\emptyset$&
-    $(1+x)(1-x)(1+w)(1+y)^2$\\
+    $(1+x')(1-x')(1+w')(1+y')^2$\\
     $\{1,2\}$  &
-    $\{\emptyset,\{1,2,3\}\}$&
-    $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+    $\{\emptyset,\{1,2\}\}$&
+    $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
     $\{1,3\}$  &
@@
                     %EM Shouldn't this be \{1\}\\ ? Or perhaps the ordering of internal nodes is not what I thought?
-                    \{2\}\\
+                    \{1\}\\
                     \{1,2\}
@@
     $\begin{array}{l}
-                    (1+x)^2   (1+w)(1-y)^2\\
-                    (1+x)^2   (1-w)(1+y)^2\\
-                    (1-x)^2   (1+w)(1+y)^2
+                    (1-x')^2   (1+w')(1+y')^2\\
+                    (1+x')^2   (1-w')(1+y')^2\\
+                    (1+x')^2   (1+w')(1-y')^2
                 \end{array}$\\
     $\{2,3\}$  &
-                $\{\emptyset,\{1,2,3\}\}$&
-                $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+                $\{\emptyset,\{1,2\}\}$&
+                $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
     $\{1,2,3\}$&
-                $\emptyset$&
-                $(1+x)^2   (1+w)(1+y)(1-y)$\\
+                $\{1,2\}$&
+                $(1+x')^2   (1+w')(1+y')(1-y')$\\
     \hline
@@
     \hline
-    $\emptyset$       &$\emptyset$&$(1+x)^2   (1+w)(1+y)^2$\\
+    $\emptyset$       &$\emptyset$&$(1+x')^2   (1+w')(1+y')^2$\\
     $\{1\}$          &
@@
     $\begin{array}{l}
-                        (1+x)(1-x)(1+w)(1+y)^2\\
-                        (1+x)^2   (1-w)(1+y)(1-y)
+                        (1+x')(1-x')(1+w')(1+y')^2\\
+                        (1+x')^2   (1-w')(1+y')(1-y')
                     \end{array}$\\
@@
     $\begin{array}{l}
-                    (1+x)^2   (1+w)(1+y)(1-y)\\
-                    (1+x)(1-x)(1-w)(1+y)^2
+                    (1+x')^2   (1+w')(1+y')(1-y')\\
+                    (1+x')(1-x')(1-w')(1+y')^2
                     \end{array}$\\
@@
     $\begin{array}{l}
-                    (1+x)(1-x)(1+w)(1+y)^2\\
-                    (1+x)^2   (1-w)(1+y)(1-y)
+                    (1+x')(1-x')(1+w')(1+y')^2\\
+                    (1+x')^2   (1-w')(1+y')(1-y')
                     \end{array}$\\
@@
                     \{\emptyset,\{1,2\}\}\\
-                    \{2\}
+                    \{1\}
                 \end{array}\right.$&
     $\begin{array}{l}
-                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
-                    (1+x)^2   (1-w)(1+y)^2
+                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
+                    (1+x')^2   (1-w')(1+y')^2
                     \end{array}$\\
@@
     $\begin{array}{l}
-                    (1+x)^2   (1+w)(1-y)^2\\
-                    (1+x)(1-x)(1-w)(1+y)(1-y)\\
-                    (1-x)^2   (1+w)(1+y)^2
+                    (1-x')^2   (1+w')(1+y')^2\\
+                    (1+x')(1-x')(1-w')(1+y')(1-y')\\
+                    (1+x')^2   (1+w')(1-y')^2
                     \end{array}$\\
@@
     $\begin{array}{l}
-                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
-                    (1-x)^2   (1-w)(1+y)^2\\
-                    (1+x)^2   (1-w)(1-y)^2
+                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
+                    (1-x')^2   (1-w')(1+y')^2\\
+                    (1+x')^2   (1-w')(1-y')^2
                     \end{array}$\\
@@
     $\left\{\begin{array}{l}
-                    \emptyset\\
-                    \{2\}
+                    \{2\} \\
+                    \{1,2\}
                 \end{array}\right.$&
     $\begin{array}{l}
-                    (1+x)^2   (1+w)(1+y)(1-y)\\
-                    (1+x)(1-x)(1-w)(1+y)^2
+                    (1+x')(1-x')(1-w')(1+y')^2 \\
+                    (1+x')^2   (1+w')(1+y')(1-y')
                     \end{array}$\\
@@
 %EM I'd suggest p_\emptyset for this particular one, but perhaps that will throw off your spacing below.
-p_0 = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})
+p_{\emptyset} = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})
 $$
 with similar definitions for the remaining generating probabilities.
-The full Farris tree log likelihood takes one of three values depending on branch lengths---assume for now the we are only interested in the ancestral state split $\{2\}$ for the site split $\{1,3\}$.
+The full Farris tree log likelihood takes one of three values depending on branch lengths---assume for now the we are only interested in the ancestral state split $\{1\}$ for the site split $\{1,3\}$.
 Directly substituting in \eqref{eq:site_pattern_likelihood}, the log likelihood is
@@
     \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &=        p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
     &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
@@
     &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
-    &\qquad + p_{123}\cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
-    &\qquad + p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
+    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
+    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
     &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
@@
     &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
-    &\qquad + p_{123}\cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
+    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
 \end{align*}
@@
 (1-x'^2+y'^2-x'^2y'^2) & = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y') \\
-(1+x'^2-y'^2-x'^2y'^2) & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y').
+(1+x'^2-y'^2-x'^2y'^2) & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y') \\
+(1+x'^2+y'^2+x'^2y'^2) & = (1+x'^2)(1+y'^2) \le (1+x')(1+y') \\
+4x'y' & = 2x' \cdot 2y' \le (1+x'^2)(1+y'^2) \le (1+x')(1+y')
 \end{align*}
@@
     \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\le      p_{\emptyset}  \cdot\log(2(1+x')(1+y')) \\
     &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+y')) \\
@@
     &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+y')(1-y')) \\
-    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{13}         \cdot\log((1+x')(1+y')) \\
     &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+y')(1-y')) \\
-    &\qquad + p_{123}\cdot\log((1+x')(1+y')(1-y')) \\
-    &\qquad + p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
+    &\qquad + p_{123}        \cdot\log((1+x')(1+y')(1-y')) \\
+    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
     &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
@@
     \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
-    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
-    &\qquad + (1-p_{0}-p_{13})\cdot\log(1+x') \\
+    &\le      p_{\emptyset}  \cdot\log(2) \\
+    &\qquad + \log(1+x') \\
     &\qquad + (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
-    &\qquad + (1-p_{0}-p_{13})\cdot\log(1+y') \\
+    &\qquad + \log(1+y') \\
     &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
@@
     \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
-    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
-    &\qquad + (3-p_{0}-p_{13}-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x') \\
+    &\le      p_{\emptyset}  \cdot\log(2) \\
+    &\qquad + (3-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x') \\
     &\qquad + 2(p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
-    &\qquad + (3-p_{0}-p_{13}-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\
+    &\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\
     &\qquad + 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
@@
 $$
-a_{1} = 3-p_{0}-p_{13}-p_{1}-p_{3}-p_{12}-p_{23},
+a_{1} = 3-p_{1}-p_{3}-p_{12}-p_{23},
 $$
@@
 $$
-a_{3} = 3-p_{0}-p_{13}-p_{2}-p_{12}-p_{23}-p_{123},
+a_{3} = 3-p_{2}-p_{12}-p_{23}-p_{123},
 $$
@@
     \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
-    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\le      p_{\emptyset}  \cdot\log(2) \\
     &\qquad + a_{1}\cdot\log(1+x') \\
@@
 \end{align*}
-Applying Gibbs' inequality to the first two summands and maximizing over the remaining terms yields
+%EM I need help understanding this application of Gibbs. p_{0} + p_{13} is not 1 if I understand correctly, and so I don't see how Gibbs applies.
+%You are totally right!
+That was an oversight on my part.
+I have redrawn the bounds on this term, and it will most likely affect the region of inconsistency, which is good, because I suspected something was making the inconsistency region larger than it should be.
+Though the bounds are not as tight as they could be now.
+Thanks!
+Maximizing over the unknown terms yields
 \begin{align*}
     \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &\le      p_{0}  \cdot\log(p_{0}) \\
-    &\qquad + p_{13} \cdot\log(p_{13}) \\
+    &\le      p_{\emptyset}  \cdot\log(2) \\
     &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
@@
     \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &\le      p_{0}  \cdot\log(p_{0}) \\
-    &\qquad + p_{13} \cdot\log(p_{13}) \\
+    &\le      p_{\emptyset}  \cdot\log(2) \\
+%EM Sorry to be slow, but could you provide a little more detail on why this follows? Is there some convexity of entropy argument going on here?
+%Not sure which part is confusing, but:
+    %1) a_1 log(1+x) + a_2 log(1-x) is maximized at x = (a_1 - a_2) / (a_1 + a_2).
+    %The below equation follows by rewriting 1+x and 1-x for the above x.
+    %2) a_1' and a_2' are just defined by rewriting the likelihood.
+There are now two fewer log(1+x) terms and two more log(1-x) terms since we replaced a (1+x)^2 in the likelihood with a (1-x)^2.
+That's why we have a -2p_{13} and a +2p_{13} respectively.
+    %Happy to clarify somewhere in the text but not sure what part would most need exposition.
     &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\
@@
     \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &\le      p_{0}  \cdot\log(p_{0}) \\
-    &\qquad + p_{13} \cdot\log(p_{13}) \\
+    &\le      p_{\emptyset}  \cdot\log(2) \\
     &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
@@
     \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &\ge      p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
+    &\ge      p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
     &\qquad + p_{1}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
@@
     &\qquad + p_{12}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
-    &\qquad + p_{123}\cdot\log((1+x')^2   (1-w')(1+y')(1-y'))\\
+    &\qquad + p_{123}        \cdot\log((1+x')^2   (1-w')(1+y')(1-y'))\\
     &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1-y')^2) \\
@@
     \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &\ge      p_{0}  \cdot\log((1+x')^2    (1+w')(1+y')^2) \\
+    &\ge      p_{\emptyset}  \cdot\log((1+x')^2    (1+w')(1+y')^2) \\
     &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
@@
     &\qquad + p_{12}         \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
-    &\qquad + p_{123}\cdot\log((1+x')(1-x')(1-w')(1+y')^2)\\
+    &\qquad + p_{123}        \cdot\log((1+x')(1-x')(1-w')(1+y')^2)\\
     &\qquad + p_{13}         \cdot\log((1-x')^2    (1-w')(1+y')^2) \\
@@
     &\qquad + b      \cdot\log(1-y') \\
-    &\qquad + p_{0}\cdot\log(1+w') \\
-    &\qquad + (1-p_{0})\cdot\log(1-w')
+    &\qquad + p_{\emptyset}\cdot\log(1+w') \\
+    &\qquad + (1-p_{\emptyset})\cdot\log(1-w')
 \end{align*}
@@
     &\qquad + b      \cdot\log(b) \\
-    &\qquad + p_{0}\cdot\log(2p_{0}) \\
-    &\qquad + (1-p_{0})\cdot\log(2(1-p_{0}))
+    &\qquad + p_{\emptyset}\cdot\log(2p_{\emptyset}) \\
+    &\qquad + (1-p_{\emptyset})\cdot\log(2(1-p_{\emptyset}))
 \end{align*}
-maximized at $\{1,1-b,1-2p_{0}\}$, which is the same if $(1+x)(1-y) > (1-x)(1+y)$ or $(1+x)(1-y) < (1-x)(1+y)$.
+maximized at $\{1,1-b,1-2p_{\emptyset}\}$, which is the same if $(1+x)(1-y) > (1-x)(1+y)$ or $(1+x)(1-y) < (1-x)(1+y)$.
 Our lower bound is
 \begin{align*}
-    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,1-2p_{0}\}) \\
-                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b) + p_{0}\cdot\log(2p_{0}) + (1-p_{0})\cdot\log(2(1-p_{0})).
+    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,1-2p_{\emptyset}\}) \\
+                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b) + p_{\emptyset}\cdot\log(2p_{\emptyset}) + (1-p_{\emptyset})\cdot\log(2(1-p_{\emptyset})).
 \end{align*}

### 2017-07-07 joint_inf.tex
Still working on understanding bounds

@@
 the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
-%EM Is it indicated anywhere how the internal nodes are numbered? I would assume that 1 would be connected to leaf 1. If that's true please indicate here.
 This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
@@
 Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
-We parameterize the branches of these trees \textbf{not with the standard notion of branch length in terms of number of substitutions per site}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$.
+We parameterize the branches of these trees \textbf{not with the standard notion of branch length in terms of number of substitutions per site}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
 This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
@@
 
-%das: did this on paper, but can write it out.
 \textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ with edge ordering in the order of the taxa, then the internal branch last.
-%EM Yeah, this next statement feels like it could use some justification, and or maybe you could skip to the following statement.
-Using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}, we calculate the generating probabilities the Farris tree topology.
+%EM complete to introduce: We now note that exchanging $x_1$ and $x_2$, or $y_1$ or $y_2$, does not change ...
+%EM please be a little more specific than just $ell$, because there are lots of variants-- e.g. likelihood with what data?
+Using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}, we calculate the generating probabilities on the Farris tree topology.
 For site split $\emptyset$,
@@
 The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
-By symmetry exchanging $x_1$ and $x_2$ does not change the value of $\ell$, requiring $x_1=x_2$ at the maximum.
-A similar property holds for $y_1$ and $y_2$.
+In summary, exchanging $x_1$ and $x_2$ does not change the value of $\ell$, and thus $x_1=x_2$ at the maximum.
+The analogous statement holds for $y_1$ and $y_2$.
 The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
@@
 \begin{table}
-%EM I wonder why you use primes on all variables here but not in tab:likelihoods.
 \centering
@@
      $\{1,2,3\}$       &
-    $\left\{\begin{array}{l}
+
                     \{2\} \\
@@
 $$
-%EM I'd suggest p_\emptyset for this particular one, but perhaps that will throw off your spacing below.
 p_{\emptyset} = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})
@@
     \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+%EM Afraid I don't see how all of the p_emptyset terms are bounded above here.
     &\le      p_{\emptyset}  \cdot\log(2) \\
@@
     %Happy to clarify somewhere in the text but not sure what part would most need exposition.
+%EM Ha! Still working on understanding. It seems to me that the statement is equivalent to saying
+%EM a_{1} \log a_{1} + a_{2} \log a_{2} + (1-p_{13}) \log(1-p_{13}) + p_{13}\cdot\log(2p_{13}) = a_{1}' \log a_{1}' + a_{2}' \log a_{2}'
+%EM am I thinking about this wrong? Is this what you are saying in your 2)?
     &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\

### 2017-07-07 joint_inf.tex
add some clarification but still unsure of where confusion arises

--- 2017-07-07 joint_inf.tex
@@
 
-%das: did this on paper, but can write it out.
-\textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ with edge ordering in the order of the taxa, then the internal branch last.
-%EM Yeah, this next statement feels like it could use some justification, and or maybe you could skip to the following statement.
-Using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}, we calculate the generating probabilities the Farris tree topology.
+We now show that, in the case of the Farris tree topology, fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal does not change the value of the likelihood.
+Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ with edge ordering in the order of the taxa, then the internal branch last.
+%EM complete to introduce: We now note that exchanging $x_1$ and $x_2$, or $y_1$ or $y_2$, does not change ...
+%EM please be a little more specific than just $ell$, because there are lots of variants-- e.g. likelihood with what data?
+Using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}, we calculate the generating probabilities on the Farris tree topology.
 For site split $\emptyset$,
 \begin{align*}
-    P(\siteSplit=\emptyset|\tau_1, \tilde{t}) & = \frac{1}{8} \cdot (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
-                                              & = \frac{1}{8} \cdot (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
-                                              & = \frac{1}{8} \cdot (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
+    P(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
+                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
+                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
 \end{align*}
@@
 \begin{align*}
-    P(\siteSplit=\{1\}|\tau_1, \tilde{t}) & = \frac{1}{8} \cdot (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
+    P(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
 \end{align*}
@@
 \begin{align*}
-    P(\siteSplit=\{3\}|\tau_1, \tilde{t}) & = \frac{1}{8} \cdot (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
+    P(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
 \end{align*}
@@
 The corresponding possibilities for the likelihood values are
-$$
-\left\{\begin{array}{l}
-        (1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2) \\
-        (1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2) \\
-        (1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2) \\
-        (1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2)
-\end{array}\right.
-$$
+\begin{align*}
+    P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
+    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
+    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
+    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
+\end{align*}
 for site split $\{1\}$ and
-$$
-\left\{\begin{array}{l}
-        (1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2) \\
-        (1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2) \\
-        (1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2) \\
-        (1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2)
-\end{array}\right.
-$$
+\begin{align*}
+        P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
+    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
+    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
+    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
+\end{align*}
 for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.
@@
 The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
-By symmetry exchanging $x_1$ and $x_2$ does not change the value of $\ell$, requiring $x_1=x_2$ at the maximum.
-A similar property holds for $y_1$ and $y_2$.
+
+In summary, exchanging $x_1$ and $x_2$ does not change the value of the log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, and thus $x_1=x_2$ at the maximum.
+The analogous statement holds for $y_1$ and $y_2$.
 The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
@@
 (1+x'^2+y'^2+x'^2y'^2) & = (1+x'^2)(1+y'^2) \le (1+x')(1+y') \\
-4x'y' & = 2x' \cdot 2y' \le (1+x'^2)(1+y'^2) \le (1+x')(1+y')
+4x'y' & = 2x' \cdot 2y' \le (1+x'^2)(1+y'^2) \le (1+x')(1+y').
+\end{align*}
+The term $p_{\emptyset}$ is bounded as
+\begin{align*}
+    p_{\emptyset} & = \log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+                  & \le \log(1+x'^2+y'^2+4x'y'+x'^2y'^2) \\
+                  & \le \log(2(1+x')(1+y'))
+\end{align*}
+and $p_{13}$ is bounded as
+\begin{align*}
+    p_{13} & = \log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+                  & \le \log(1+x'^2+y'^2+x'^2y'^2) \\
+                  & \le \log((1+x')(1+y')).
 \end{align*}
@@
     %Happy to clarify somewhere in the text but not sure what part would most need exposition.
+%EM Ha! Still working on understanding. It seems to me that the statement is equivalent to saying
+%EM a_{1} \log a_{1} + a_{2} \log a_{2} + (1-p_{13}) \log(1-p_{13}) + p_{13}\cdot\log(2p_{13}) = a_{1}' \log a_{1}' + a_{2}' \log a_{2}'
+%EM am I thinking about this wrong? Is this what you are saying in your 2)?
+    %Hmm, I don't know if I see what you're saying now.
+    %The way I think about it is like this: a_1 is the coefficient for the log(1+x) term and a_2 for log(1-x).
+    %It is the sum of all the generating probabilities that had a log(1+x) term in it.
+    %Looking at Table 2, since our first bound (we calculated it for \eta_j = {1}) was constructed with the p_{13} term of
+    %   (1+x)^2(1-w)(1+y)^2,
+    %our second bound (that for \eta_j = \emptyset) will be constructed with the p_{13} term of
+    %   (1-x)^2(1+w)(1+y)^2.
+    %So we lose 2 log(1+x)s (hence a_1' = a_1 - 2p_{13}) and gain two log(1-x)s (hence a2' = a2 + 2p_{13}).
+    %Now every generating probability has log(1+w) so we replace the last two terms (p_{13}log(1-w) + (1-p_{13})log(1+w)) with just log(1+w)---maximized at log(2).
     &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\

### 2017-07-08 joint_inf.tex
Making progress understanding :pooh:

@@
                     \emptyset\\
-                    %EM Shouldn't this be \{1\}\\ ? Or perhaps the ordering of internal nodes is not what I thought?
                     \{1\}\\
@@
 with similar definitions for the remaining generating probabilities.
-The full Farris tree log likelihood takes one of three values depending on branch lengths---assume for now the we are only interested in the ancestral state split $\{1\}$ for the site split $\{1,3\}$.
+The full Farris tree log likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see Table~\ref{tab:likelihoods} for $\tau = \tau_1$).
+We will begin by writing out the middle case: for the ancestral state split $\{1\}$.
 Directly substituting in \eqref{eq:site_pattern_likelihood}, the log likelihood is
@@
 \begin{align*}
-(1-x'^2+y'^2-x'^2y'^2) & = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y') \\
-(1+x'^2-y'^2-x'^2y'^2) & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y') \\
-(1+x'^2+y'^2+x'^2y'^2) & = (1+x'^2)(1+y'^2) \le (1+x')(1+y') \\
+1-x'^2+y'^2-x'^2y'^2 & = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y') \\
+1+x'^2-y'^2-x'^2y'^2 & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y') \\
+1+x'^2+y'^2+x'^2y'^2 & = (1+x'^2)(1+y'^2) \le (1+x')(1+y') \\
 4x'y' & = 2x' \cdot 2y' \le (1+x'^2)(1+y'^2) \le (1+x')(1+y').
@@
     \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-%EM Afraid I don't see how all of the p_emptyset terms are bounded above here.
     &\le      p_{\emptyset}  \cdot\log(2) \\
@@
 \end{align*}
-%EM I need help understanding this application of Gibbs. p_{0} + p_{13} is not 1 if I understand correctly, and so I don't see how Gibbs applies.
-%You are totally right!
-That was an oversight on my part.
-I have redrawn the bounds on this term, and it will most likely affect the region of inconsistency, which is good, because I suspected something was making the inconsistency region larger than it should be.
-Though the bounds are not as tight as they could be now.
-Thanks!
 Maximizing over the unknown terms yields
@@
 $$
-Then,
+Then, to bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
 \begin{align*}
@@
     &\le      p_{\emptyset}  \cdot\log(2) \\
-%EM Sorry to be slow, but could you provide a little more detail on why this follows? Is there some convexity of entropy argument going on here?
-%Not sure which part is confusing, but:
-    %1) a_1 log(1+x) + a_2 log(1-x) is maximized at x = (a_1 - a_2) / (a_1 + a_2).
-    %The below equation follows by rewriting 1+x and 1-x for the above x.
-    %2) a_1' and a_2' are just defined by rewriting the likelihood.
-There are now two fewer log(1+x) terms and two more log(1-x) terms since we replaced a (1+x)^2 in the likelihood with a (1-x)^2.
-That's why we have a -2p_{13} and a +2p_{13} respectively.
-    %Happy to clarify somewhere in the text but not sure what part would most need exposition.
-%EM Ha! Still working on understanding. It seems to me that the statement is equivalent to saying
-%EM a_{1} \log a_{1} + a_{2} \log a_{2} + (1-p_{13}) \log(1-p_{13}) + p_{13}\cdot\log(2p_{13}) = a_{1}' \log a_{1}' + a_{2}' \log a_{2}'
-%EM am I thinking about this wrong? Is this what you are saying in your 2)?
-    %Hmm, I don't know if I see what you're saying now.
-    %The way I think about it is like this: a_1 is the coefficient for the log(1+x) term and a_2 for log(1-x).
-    %It is the sum of all the generating probabilities that had a log(1+x) term in it.
-    %Looking at Table 2, since our first bound (we calculated it for \eta_j = {1}) was constructed with the p_{13} term of
-    %   (1+x)^2(1-w)(1+y)^2,
-    %our second bound (that for \eta_j = \emptyset) will be constructed with the p_{13} term of
-    %   (1-x)^2(1+w)(1+y)^2.
-    %So we lose 2 log(1+x)s (hence a_1' = a_1 - 2p_{13}) and gain two log(1-x)s (hence a2' = a2 + 2p_{13}).
-    %Now every generating probability has log(1+w) so we replace the last two terms (p_{13}log(1-w) + (1-p_{13})log(1+w)) with just log(1+w)---maximized at log(2).
     &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\
@@
     &\qquad + \log(2) \\
-    &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy});
+    &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
 \end{align*}
+To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
 \begin{align*}

### 2017-07-08 joint_inf.tex
Checking a few things in Felsenstein lower bounds.

@@
 I suggest just taking the max over \fullAncestralSplitPartitions as a variable.
+%EM Dave, did you already address this comment? It doesn't seem to make sense anymore.
 $$
@@
 We construct a similar lower bound on the Felsenstein tree partial likelihood.
-Replace all $(1+w)$ terms with $(1-w)$ and focus just on the partial likelihood.
-Then, if $(1+x)(1-y) > (1-x)(1+y)$,
+In the Felsenstein case, there are various optimal internal states depending on branch lengths (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).
+We can bound these below replacing $1+w$ with $1-w$, and by splitting into two cases.
+%EM Is this inequality flipped?
+If $(1+x)(1-y) > (1-x)(1+y)$,
 \begin{align*}
@@
     &\qquad + p_{3}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
+%EM it seems to me that the following line should be (1+x')(1-x')(1-w')(1+y')(1-y')
     &\qquad + p_{12}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
@@
     &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
+%EM it seems to me that the following line should be (1+x')(1-x')(1-w')(1+y')(1-y')
     &\qquad + p_{12}         \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
@@
 \end{align*}
+%EM check below after the above is adjusted.
 Letting

### 2017-07-10 joint_inf.tex
add fels bound clarifications

@@
 %EM Dave, did you already address this comment? It doesn't seem to make sense anymore.
+%Ah, yes, I believe it was commit 209f62b4ec37751a60c72edaa160b80d827bad04 (if that helps, for reference)
 $$
@@
 We can bound these below replacing $1+w$ with $1-w$, and by splitting into two cases.
+The value of the likelihood for the site split $\{1,2\}$ is $(1+x')^2   (1-w')(1+y')^2$ since
+$$
+(1+x')^2   (1-w')(1+y')^2 > (1+x')(1-x')(1-w')(1+y')(1-y')
+$$
+always.
 %EM Is this inequality flipped?
-If $(1+x)(1-y) > (1-x)(1+y)$,
+If $(1+x')(1-y') > (1-x')(1+y')$, then
+$$
+(1+x')^2(1-w')(1+y')(1-y') > (1+x')(1-x')(1-w')(1+y')^2
+$$
+meaning the cases $\{1\}, \{2\}, \{3\}$ and $\{1,2,3\}$ will have likelihoods equal to
+$$
+(1+x')^2(1-w')(1+y')(1-y').
+$$
+For the cases $\{1,3\}$ and $\{2,3\}$ and the same condition, we have
+$$
+(1+x')^2(1-w')(1-y')^2 > (1-x')^2(1-w')(1+y')^2
+$$
+and
+$$
+(1+x')^2(1-w')(1-y')^2 > (1+x')(1-x')(1-w')(1+y')(1-y'),
+$$
+yielding
+$$
+(1+x')^2(1-w')(1-y')^2
+$$
+as the likelihood in these cases.
+The full likelihood in this case is then bounded below as
 \begin{align*}
@@
 %EM it seems to me that the following line should be (1+x')(1-x')(1-w')(1+y')(1-y')
+    %I've added some clarification to the text. Let me know if there is still an issue.
     &\qquad + p_{12}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
@@
     &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1-y')^2) \\
-    &\qquad + p_{23}         \cdot\log((1+x')^2   (1-w')(1-y')^2),
+    &\qquad + p_{23}         \cdot\log((1+x')^2   (1-w')(1-y')^2).
 \end{align*}
-otherwise
+When $(1+x')(1-y') < (1-x')(1+y')$, similar arguments yield the lower bound
 \begin{align*}
@@
 %EM it seems to me that the following line should be (1+x')(1-x')(1-w')(1+y')(1-y')
+    %See above.
     &\qquad + p_{12}         \cdot\log((1+x')^2    (1-w')(1+y')^2) \\

### 2017-07-10 joint_inf.tex
Complete first pass :woot:

@@
 
-%comments from Erick:
-% >  This is great!
-% > - would be great to have all the calculations showing the "refinement" structure written out for the various tree structures (just done by hand and scanned is just fine)
-% > - I think of the "refinement" as a partition, so perhaps rather than "empty refinement" we have a "single-element partition"
-% > - in order to interpret the site patterns I need to figure out in what order the leaves are labeled
-% > - how is ell calculated?
-% > - on the numerical sanity check, it doesn't seem clear that the branch lengths can be different than the inferred ones-- they can!
-% > - re "general case", agreed though I don't think that we need numerical optimization of branch lengths in a set of a branch length partition-- we can just get the optimal branch length directly.
-Namely, for each branch, we know all hidden states (given that we are in a set of a partition) and so can count the number of mutations across that branch.
-From there we have a simple likelihood function, which either has an optimal in the interior or on the boundary of the partition.
-
 \renewcommand{\arraystretch}{1.2} % because otherwise exponents get eaten by \hline
@@
 Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
-Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$ and limit themselves to optimizing branch lengths using this method, this motivates understanding when exactly this approximation breaks down, and what happens when we optimize branch lengths and tree topology using this approximation.
+Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$ and limit themselves to optimizing branch lengths using this method, their work motivates understanding when exactly this approximation breaks down, and what happens when we optimize branch lengths and tree topology using this approximation.
 
@@
 We obtain
-%EM Does it really make sense to talk about taking the max over \fullAncestralSplitPartitions(\tau,t)?
-This is already defined.
-I suggest just taking the max over \fullAncestralSplitPartitions as a variable.
-%EM Dave, did you already address this comment? It doesn't seem to make sense anymore.
-%Ah, yes, I believe it was commit 209f62b4ec37751a60c72edaa160b80d827bad04 (if that helps, for reference)
 $$
@@
 In the Felsenstein case, there are various optimal internal states depending on branch lengths (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).
-We can bound these below replacing $1+w$ with $1-w$, and by splitting into two cases.
-The value of the likelihood for the site split $\{1,2\}$ is $(1+x')^2   (1-w')(1+y')^2$ since
+We can bound these below replacing $1+w$ with $1-w$, for example as follows.
+For the site split $\{1,2\}$, we would like to find a lower bound for
+$$
+\max\left((1+x')^2 (1-w')(1+y')^2, (1+x')(1-x')(1+w')(1+y')(1-y')\right).
+$$
+We observe that
 $$
@@
 $$
-always.
-%EM Is this inequality flipped?
+always, and thus the maximum can be bounded below by $(1+x')^2 (1-w')(1+y')^2$.
+For the other site splits we consider two cases.
 If $(1+x')(1-y') > (1-x')(1+y')$, then
@@
 $$
-meaning the cases $\{1\}, \{2\}, \{3\}$ and $\{1,2,3\}$ will have likelihoods equal to
+meaning the cases $\{1\}, \{2\}, \{3\}$ and $\{1,2,3\}$ will have likelihoods bounded below by
 $$
@@
 $$
-yielding
+yielding lower bounds of
 $$
@@
 $$
-as the likelihood in these cases.
+in these cases.
 The full likelihood in this case is then bounded below as
@@
     &\qquad + p_{3}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
-%EM it seems to me that the following line should be (1+x')(1-x')(1-w')(1+y')(1-y')
-    %I've added some clarification to the text. Let me know if there is still an issue.
     &\qquad + p_{12}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
@@
     &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
-%EM it seems to me that the following line should be (1+x')(1-x')(1-w')(1+y')(1-y')
     %See above.
@@
 \end{align*}
-%EM check below after the above is adjusted.
 Letting
@@
 \end{align*}
-for the first lower bound, but, upon maximizing, we obtain
+for the first lower bound, and we get the equivalent expression for the second lower bound by exchanging $x'$ and $y'$.
+Maximizing either lower bound over $\{x',y',w'\}$, we obtain
 \begin{align*}
@@
     &\qquad + p_{\emptyset}\cdot\log(2p_{\emptyset}) \\
-    &\qquad + (1-p_{\emptyset})\cdot\log(2(1-p_{\emptyset}))
+    &\qquad + (1-p_{\emptyset})\cdot\log(2(1-p_{\emptyset})),
 \end{align*}
-maximized at $\{1,1-b,1-2p_{\emptyset}\}$, which is the same if $(1+x)(1-y) > (1-x)(1+y)$ or $(1+x)(1-y) < (1-x)(1+y)$.
+%EM Shouldn't this be 2p_{\emptyset}-1 ? It doesn't matter for the bound, but...
+i.e.\ the maximum is achieved at $\{1,1-b,1-2p_{\emptyset}\}$.
 Our lower bound is

### 2017-07-11 joint_inf_plot.py
plotting code---first draft

@@ -1,0 +1,160 @@
+import matplotlib.pyplot
+from numpy import log, meshgrid, arange, maximum
+
+ALL_PATTERNS = [
+        '0000',
+        '1000',
+        '0100',
+        '0010',
+        '1110',
+        '1100',
+        '1010',
+        '0110'
+]
+
+PATTERN_TO_SPLIT_DICT = {
+        '0000': '0',
+        '1000': '1',
+        '0100': '2',
+        '0010': '3',
+        '1110': '123',
+        '1100': '12',
+        '1010': '13',
+        '0110': '23'
+}
+
+SPLIT_TO_PATTERN_DICT = {
+        '0':  '0000',
+        '1':  '1000',
+        '2':  '0100',
+        '3':  '0010',
+        '123':'1110',
+        '12': '1100',
+        '13': '1010',
+        '23': '0110'
+}
+
+def P_FARRIS(theta0, site_pattern):
+    """
+    Site pattern frequencies for Farris tree
+    """
+    # fifth branch will always be positive
+    theta = [-t if s=='1' else t \
+            for t,s in zip(theta0, site_pattern+'0')]
+    return 0.125 * (1 + \
+         theta[0]*theta[2] + \
+         theta[1]*theta[3] + \
+         theta[0]*theta[1]*theta[4] + \
+         theta[0]*theta[3]*theta[4] + \
+         theta[1]*theta[2]*theta[4] + \
+         theta[2]*theta[3]*theta[4] + \
+         theta[0]*theta[1]*theta[2]*theta[3])
+
+def P_FELSENSTEIN(theta0, site_pattern):
+    """
+    Site pattern frequencies for Felsenstein tree
+    """
+    # fifth branch will always be positive
+    theta = [-t if s=='1' else t \
+            for t,s in zip(theta0, site_pattern+'0')]
+    return 0.125 * (1 + \
+         theta[0]*theta[1] + \
+         theta[2]*theta[3] + \
+         theta[0]*theta[2]*theta[4] + \
+         theta[1]*theta[2]*theta[4] + \
+         theta[0]*theta[3]*theta[4] + \
+         theta[1]*theta[3]*theta[4] + \
[…]

### 2017-07-13 joint_inf_plot.py
cleaning up a bit

@@ -1,4 +1,4 @@
-import matplotlib.pyplot
+import matplotlib.pyplot as plt
 from numpy import log, meshgrid, arange, maximum
 
 ALL_PATTERNS = [
@@ -66,9 +66,6 @@ def P_FELSENSTEIN(theta0, site_pattern):
          theta[1]*theta[3]*theta[4] + \
          theta[0]*theta[1]*theta[2]*theta[3])
 
-def split_to_gen(pattern, x, y):
-    return P_FARRIS([x, y, x, y, y], pattern)
-
 ## Generating probabilities
 # Left column of tab:sitepatprob
 def gen_prob(split, x, y):
@@ -76,19 +73,22 @@ def gen_prob(split, x, y):
 
 ## Upper bound for Farris tree topology
 # Constants from eq:a_const
+def sum_of_four(split_list, x, y):
+    return sum([gen_prob(split, x, y) for split in split_list])
+
 def a_const_1(x, y):
-    return 3. - gen_prob('1', x, y) - gen_prob('3', x, y) - gen_prob('12', x, y) - gen_prob('23', x, y)
+    return 3. - sum_of_four(['1', '3', '12', '23'], x, y)
 
 def a_const_2(x, y):
-    return 2. * (gen_prob('1', x, y) + gen_prob('3', x, y) + gen_prob('12', x, y) + gen_prob('23', x, y))
+    return 2. * sum_of_four(['1', '3', '12', '23'], x, y)
 
 def a_const_3(x, y):
-    return 3. - gen_prob('2', x, y) - gen_prob('12', x, y) - gen_prob('23', x, y) - gen_prob('123', x, y)
+    return 3. - sum_of_four(['2', '12', '23', '123'], x, y)
 
 def a_const_4(x, y):
-    return 2. * (gen_prob('2', x, y) + gen_prob('12', x, y) + gen_prob('23', x, y) + gen_prob('123', x, y))
+    return 2. * sum_of_four(['2', '12', '23', '123'], x, y)
 
-# Constants from eq:a_const
+# Constants from eq:a_const_prime
 def a_const_1_prime(x, y):
     return a_const_1(x, y) - 2.*gen_prob('13', x, y)
 
@@ -101,41 +101,41 @@ def a_const_3_prime(x, y):
 def a_const_4_prime(x, y):
[…]

### 2017-07-13 joint_inf.tex
add eqn number to block
add normalizing factor and verify x/y are maxima

@@
     &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
+    &\qquad - \log(8) \\
     &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
@@
     &\qquad + p_{123}        \cdot\log((1+x')(1+y')(1-y')) \\
+    &\qquad - \log(8) \\
     &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
@@
     &\qquad + (1-p_{13})\cdot\log(1+w') \\
-    &\qquad + p_{13}\cdot\log(1-w'),
+    &\qquad + p_{13}\cdot\log(1-w') \\
+    &\qquad - \log(8),
 \end{align*}
@@
     &\qquad + (1-p_{13})\cdot\log(1+w') \\
-    &\qquad + p_{13}\cdot\log(1-w').
+    &\qquad + p_{13}\cdot\log(1-w') \\
+    &\qquad - \log(8).
 \end{align*}
 To make matters further compact, let
-$$
-a_{1} = 3-p_{1}-p_{3}-p_{12}-p_{23},
-$$
-$$
-a_{2} = 2(p_{1}+p_{3}+p_{12}+p_{23}),
-$$
-$$
-a_{3} = 3-p_{2}-p_{12}-p_{23}-p_{123},
-$$
-$$
-a_{4} = 2(p_{2}+p_{12}+p_{23}+p_{123}),
-$$
+\begin{equation}
+    \begin{aligned}
+        a_{1} &= 3-p_{1}-p_{3}-p_{12}-p_{23}, \\
+        a_{2} &= 2(p_{1}+p_{3}+p_{12}+p_{23}), \\
+        a_{3} &= 3-p_{2}-p_{12}-p_{23}-p_{123}, \\
+        a_{4} &= 2(p_{2}+p_{12}+p_{23}+p_{123}), \\
+    \end{aligned}
+    \label{eq:a_const}
+\end{equation}
 where
@@
     &\qquad + (1-p_{13})\cdot\log(1+w') \\
-    &\qquad + p_{13}\cdot\log(1-w').
+    &\qquad + p_{13}\cdot\log(1-w') \\
+    &\qquad - \log(8).
 \end{align*}
 Maximizing over the unknown terms yields
+$$
+\hat{x}' = \frac{a_{1}-a_{2}}{a_{1}+a_{2}}, \ \hat{y}' = \frac{a_{3}-a_{4}}{a_{3}+a_{4}}.
+$$
+We check that $0 \le \hat{x}' \le 1$ and $0 \le \hat{y}' \le 1$ so these maximizers are valid.
+Letting
+$$
+\tilde{p} = p_{1}+p_{3}+p_{12}+p_{23}
+$$
+we have
+$$
+\hat{x}' = \frac{3-\tilde{p} - 2\tilde{p}}{3-\tilde{p}+2\tilde{p}} = \frac{3-3\tilde{p}}{3+\tilde{p}},
+$$
+and, since $0 \le \tilde{p} \le 1/2$, $0 \le \hat{x}' \le 1$.
+The same approach works for $\hat{y}'$.
+The upper bound for the likelihood is then maximized at
 \begin{align*}
@@
     &\qquad + p_{13}\cdot\log(2p_{13}) \\
+    &\qquad - \log(8) \\
     &\qquad := L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
@@
 Call
+\begin{equation}
+    \begin{aligned}
+        a_{1}' &= a_{1}-2p_{13}, \\
+        a_{2}' &= a_{2}+2p_{13}, \\
+        a_{3}' &= a_{3}-2p_{13}, \\
+        a_{4}' &= a_{4}+2p_{13}. \\
+    \end{aligned}
+    \label{eq:a_const_prime}
+\end{equation}
+We also now need to check that
 $$
-a_{1}' = a_{1}-2p_{13},
+\hat{x}' = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y}' = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
 $$
+are both between zero and one.
+It will suffice to show
 $$
-a_{2}' = a_{2}+2p_{13},
+0 \le \tilde{p} - 2p_{13} \le 1
 $$
+and
 $$
-a_{3}' = a_{3}-2p_{13},
+0 \le \tilde{p} + 2p_{13} \le 1
 $$
+for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
+Using Table~\ref{tab:sitepatprob},
 $$
-a_{4}' = a_{4}+2p_{13}.
+p_{13} = \frac{1}{8}\left(1 + x^2 + y^2 - 4xy^2 + x^2y^2\right).
 $$
-Then, to bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
+We see that
+$$
+1 + x^2 + y^2 - 4xy^2 + x^2y^2 \le 2 \iff y^2\left(1 - 4x + x^2\right) \le 1 - x^2
+$$
+and
+$$
+y^2\left(1 - 4x + x^2\right) \le 1 - 4x + x^2 \le 1 - x^2,
+$$
+the last inequality holding if $x \le 2$.
+Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}'$ and $\hat{y}'$ are valid.
+
+To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
 \begin{align*}
@@
     &\qquad + \log(2) \\
-    &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
+    &\qquad - \log(8) \\
+    &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}),
 \end{align*}
-To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
+and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
 \begin{align*}
@@
     &\qquad + \log(2) \\
+    &\qquad - \log(8) \\
     &\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
@@
 in these cases.
-The full likelihood in this case is then bounded below as
+The full likelihood in this case is then bounded below by
 \begin{align*}
     \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &\ge      p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
+    &\ge      p_{\emptyset}  \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
     &\qquad + p_{1}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
@@
     \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &\ge      p_{\emptyset}  \cdot\log((1+x')^2    (1+w')(1+y')^2) \\
+    &\ge      p_{\emptyset}  \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
     &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
@@
     &\qquad + b      \cdot\log(1-y') \\
-    &\qquad + p_{\emptyset}\cdot\log(1+w') \\
-    &\qquad + (1-p_{\emptyset})\cdot\log(1-w')
+    &\qquad + \log(1-w')
 \end{align*}
@@
     &\qquad + (2-b)  \cdot\log(2-b) \\
-    &\qquad + b      \cdot\log(b) \\
-    &\qquad + p_{\emptyset}\cdot\log(2p_{\emptyset}) \\
-    &\qquad + (1-p_{\emptyset})\cdot\log(2(1-p_{\emptyset})),
+    &\qquad + b      \cdot\log(b)
 \end{align*}
 %EM Shouldn't this be 2p_{\emptyset}-1 ? It doesn't matter for the bound, but...
-i.e.\ the maximum is achieved at $\{1,1-b,1-2p_{\emptyset}\}$.
+%das If we call p_{\emptyset} just p, we're maximizing
+%   p log (1 + w) + (1 - p) log (1 - w).
+%As before, a_1 log(1+x) + a_2 log(1-x) is maximized at x = (a_1 - a_2) / (a_1 + a_2).
+%So then w = (p - (1 - p)) / (p + (1 - p)) = (2p - 1) / 1 = 2p - 1.
+%So you're right. Good catch.
+%
+%Unfortunately, I realized the above does not work for us since 2p - 1 is not guaranteed to be between 0 and 1, as x should be (see the stuff I added justifying this above).
+%Rather than do a constrained optimization, I just bounded the p_{\emptyset}\log(1+w) term below by p_{\emptyset}\log(1-w).
+i.e.\ the maximum is achieved at $\{1,1-b,0\}$.
 Our lower bound is
 \begin{align*}
-    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,1-2p_{\emptyset}\}) \\
-                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b) + p_{\emptyset}\cdot\log(2p_{\emptyset}) + (1-p_{\emptyset})\cdot\log(2(1-p_{\emptyset})).
+    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,0\}) \\
+                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b).
 \end{align*}

### 2017-07-25 joint_inf_plot_bl.py
branch length estimation plotting

@@ -1,0 +1,45 @@
+import matplotlib
+matplotlib.use('pdf')
+from numpy import log, arange, vectorize
+from scipy.optimize import basinhopping
+from pylab import meshgrid,cm,imshow,title,savefig,xlabel,ylabel,colorbar
+
+def p13(x, y, w):
+    return 0.125 * (1 + x*x + y*y + x*x*y*y - 4*x*y*w)
+
+def p0(x, y, w):
+    return 0.125 * (1 + x*x + y*y + x*x*y*y + 4*x*y*w)
+
+def likelihood(x, y, w):
+    if (1 - x) * (1 + w) * (1 + y) > (1 + x) * (1 - w) * (1 - y) or \
+        (1 + x) * (1 + w) * (1 - y) > (1 - x) * (1 - w) * (1 + y):
+        return p0(x, y, y) * log(p0(x, y, w)) + p13(x, y, y) * log(p13(x, y, w)) + \
+            log(1 + w)
+    else:
+        return p0(x, y, y) * log(p0(x, y, w)) + p13(x, y, y) * log(p13(x, y, w)) + \
+            p13(x, y, y) * log(1 - w) + (1 - p13(x, y, y)) * log(1 + w)
+
+def max_likelihood(x, y):
+    """
+    basin hopping to find maximum likelihood estimate of interior branch length
+    """
+    bnds = [(0,1)]
+    minimizer_kwargs = dict(method="L-BFGS-B", bounds=bnds)
+    what = basinhopping(lambda w: -likelihood(x, y, w),
+        x0=[0.25],
+        minimizer_kwargs=minimizer_kwargs)
+    return(what.x)
+
+delta = 0.01
+x_range = arange(0.0, 1.0, delta)
+y_range = arange(0.0, 1.0, delta)
+X, Y = meshgrid(x_range,y_range)
+
+Z = vectorize(lambda x, y: max_likelihood(x, y))(X, Y)
+im = imshow(Z, cmap=cm.gray, origin='lower')
+colorbar()
+title('Value of $\hat{w}$')
+xlabel('$x$')
+ylabel('$y$')
+savefig('figures/w_hat_heatmap.svg')
+

### 2017-07-25 joint_inf.tex
A few notes.
address comments

@@
 
-We now show that, in the case of the Farris tree topology, fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal does not change the value of the likelihood.
+%EM This sentence should be made more precise-- it certainly changes the value of the log likelihood-- you mean this restriction does not change the ML.
 Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ with edge ordering in the order of the taxa, then the internal branch last.
+We now show that, in the case of the Farris tree topology, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
 %EM complete to introduce: We now note that exchanging $x_1$ and $x_2$, or $y_1$ or $y_2$, does not change ...
@@
 For maximum parsimony it's the opposite \cite{Felsenstein1978-rr}.
+%EM We should be clear here that thus this is not "long branch attraction".
+I haven't read the Farris papers, but I wonder what his argument is and if it relates to this, in that there are too many parameters.
+Probably something should go in the intro as well pointing out that this is not LBA.
+%das: Noted. I will work on this.
 Discussion of number of parameters of each.

### 2017-07-25 joint_inf_plot_bl.py
fix tickmarks

@@ -1,8 +1,8 @@
 import matplotlib
 matplotlib.use('pdf')
+import matplotlib.pylab as plt
 from numpy import log, arange, vectorize
 from scipy.optimize import basinhopping
-from pylab import meshgrid,cm,imshow,title,savefig,xlabel,ylabel,colorbar
 
 def p13(x, y, w):
     return 0.125 * (1 + x*x + y*y + x*x*y*y - 4*x*y*w)
@@ -33,13 +33,18 @@ def max_likelihood(x, y):
 delta = 0.01
 x_range = arange(0.0, 1.0, delta)
 y_range = arange(0.0, 1.0, delta)
-X, Y = meshgrid(x_range,y_range)
+X, Y = plt.meshgrid(x_range,y_range)
 
 Z = vectorize(lambda x, y: max_likelihood(x, y))(X, Y)
-im = imshow(Z, cmap=cm.gray, origin='lower')
-colorbar()
-title('Value of $\hat{w}$')
-xlabel('$x$')
-ylabel('$y$')
-savefig('figures/w_hat_heatmap.svg')
+im = plt.imshow(Z, cmap=plt.cm.gray, origin='lower')
+plt.colorbar()
+plt.title('Value of $\hat{w}$')
+plt.xlabel('$x$')
+plt.ylabel('$y$')
+ax = plt.gca()
+ax.set_xticks(arange(0, 1/delta, .2/delta))
+ax.set_yticks(arange(0, 1/delta, .2/delta))
+ax.set_xticklabels(arange(0, 1, .2))
+ax.set_yticklabels(arange(0, 1, .2))
+plt.savefig('figures/w_hat_heatmap.svg')
 

### 2017-07-31 joint_inf.tex
branch length code and minor explanation

@@
 
+\section{Branch lengths}
+
+Consider only the Farris-tree topology.
+Assume we know the true values of $x'$ and $y'$ but not $w'$.
+In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
+In this case, the likelihood as a function of $w'$ is proportional to either
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
+    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
+    &\qquad + \log((1+w'))
+\end{align*}
+or
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
+    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
+    &\qquad + (1 - p_{13})\cdot\log((1+w')) \\
+    &\qquad + p_{13}\cdot\log((1-w')).
+\end{align*}
+When computing the global minimum for $0 \le w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}$ in Figure~\ref{fig:w_hat}.
+The estimate for $\hat{w}$ will typically be an overestimate, and in fact is almost always identically equal to 1, indicating no change along that particular branch.
+
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{w_hat_heatmap}
+\caption{
+    Estimating $\hat{w}$ for $x'$ and $y'$ fixed at their true values.
+}
+\label{fig:w_hat}
+\end{figure}
+
+Since we do not know the true values of $x$ and $y$, standard methods will not obtain the situation in Figure~\ref{fig:w_hat}.
+The form of the likelihood will alter the fitted values of $x$ and $y$ so that even with perfect information we will not necessarily obtain the true parameters back.
+If we estimate $(x', y', w')$ jointly using the same approach above, we obtain the fit $\hat{w}$ in Figure~\ref{fig:w_hat_joint}.
+
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{w_hat_heatmap_joint}
+\caption{
+    Estimating $(\hat{x}, \hat{y}, \hat{w})$.
+}
+\label{fig:w_hat}
+\end{figure}
+
+We no longer consistently overestimate $\hat{w}$, but the estimates are surprisingly unstable, with estimates of 1 for $\hat{w}$ scattered throughout the $(x,y)$ lattice.
+
 \section{Discussion}

### 2017-07-31 joint_inf_plot_bl.py

@@ -11,8 +11,8 @@ def p0(x, y, w):
     return 0.125 * (1 + x*x + y*y + x*x*y*y + 4*x*y*w)
 
 def likelihood(x, y, w):
-    if (1 - x) * (1 + w) * (1 + y) > (1 + x) * (1 - w) * (1 - y) or \
-        (1 + x) * (1 + w) * (1 - y) > (1 - x) * (1 - w) * (1 + y):
+    if (1 - x) * (1 - x) * (1 + w) > (1 + x) * (1 + x) * (1 - w) or \
+        (1 - y) * (1 - y) * (1 + w) > (1 + y) * (1 + y) * (1 - w):
         return p0(x, y, y) * log(p0(x, y, w)) + p13(x, y, y) * log(p13(x, y, w)) + \
             log(1 + w)
     else:

### 2017-07-31 joint_inf.tex
update plot

@@
 
-Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
+Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Fig.~\ref{fig:farris-fels-top}.
 We parameterize the branches of these trees \textbf{not with the standard notion of branch length in terms of number of substitutions per site}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
@@
 \end{align*}
-When computing the global minimum for $0 \le w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}$ in Figure~\ref{fig:w_hat}.
+When computing the global minimum for $0 \le w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}$ in Fig.~\ref{fig:w_hat}.
 The estimate for $\hat{w}$ will typically be an overestimate, and in fact is almost always identically equal to 1, indicating no change along that particular branch.
@@
 \centering
-\includegraphics[width=.9\textwidth]{w_hat_heatmap}
+    \begin{tabular}{cc}
+\includegraphics[width=.45\textwidth]{w_hat_heatmap} &
+\includegraphics[width=.45\textwidth]{w_hat_heatmap_no_joint}\\
+Joint inference & Marginal inference \\
+    \end{tabular}
 \caption{
@@
 
-Since we do not know the true values of $x$ and $y$, standard methods will not obtain the situation in Figure~\ref{fig:w_hat}.
-The form of the likelihood will alter the fitted values of $x$ and $y$ so that even with perfect information we will not necessarily obtain the true parameters back.
+Since we do not know the true values of $x$ and $y$, standard methods will not be in the situation in Figure~\ref{fig:w_hat}.
+The form of the likelihood will alter the fitted values of $x$ and $y$ so that even with perfect information we will not necessarily estimate $\hat{x}=x$ and $\hat{y}=y$.
 If we estimate $(x', y', w')$ jointly using the same approach above, we obtain the fit $\hat{w}$ in Figure~\ref{fig:w_hat_joint}.
@@
 }
-\label{fig:w_hat}
+\label{fig:w_hat_joint}
 \end{figure}
@@
 We no longer consistently overestimate $\hat{w}$, but the estimates are surprisingly unstable, with estimates of 1 for $\hat{w}$ scattered throughout the $(x,y)$ lattice.
+Even given the correct Farris topology we obtain poor estimates for the value of the interior branch length.
 

### 2017-08-01 joint_inf.tex
Mostly just cleaning up; new title :D

@@
 
-\title{Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States}
+\title{On joint estimation of ancestral states and phylogenetic model parameters}
 \author{Shaw, Matsen, Minin}
@@
 
-%EM This sentence should be made more precise-- it certainly changes the value of the log likelihood-- you mean this restriction does not change the ML.
 Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ with edge ordering in the order of the taxa, then the internal branch last.
-We now show that, in the case of the Farris tree topology, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
-%EM complete to introduce: We now note that exchanging $x_1$ and $x_2$, or $y_1$ or $y_2$, does not change ...
-%EM please be a little more specific than just $ell$, because there are lots of variants-- e.g. likelihood with what data?
+We now show that, in the case of the Farris tree topology, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
 Using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}, we calculate the generating probabilities on the Farris tree topology.
@@
 Consider only the Farris-tree topology.
-Assume we know the true values of $x'$ and $y'$ but not $w'$.
+We will consider two settings, corresponding to whether some branch lengths are known ahead of time or not.
+
+First, assume we know the true values of $x'$ and $y'$ but not $w'$.
 In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
@@
 When computing the global minimum for $0 \le w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}$ in Fig.~\ref{fig:w_hat}.
-The estimate for $\hat{w}$ will typically be an overestimate, and in fact is almost always identically equal to 1, indicating no change along that particular branch.
+We find that the estimate for $\hat{w}$ will typically be an overestimate, and in fact is almost always identically equal to 1, indicating no change along that particular branch.
 
@@
 
-Since we do not know the true values of $x$ and $y$, standard methods will not be in the situation in Figure~\ref{fig:w_hat}.
+Second, we consider the more standard case where the true values of $x$ and $y$ are not known.
 The form of the likelihood will alter the fitted values of $x$ and $y$ so that even with perfect information we will not necessarily estimate $\hat{x}=x$ and $\hat{y}=y$.
@@
 We no longer consistently overestimate $\hat{w}$, but the estimates are surprisingly unstable, with estimates of 1 for $\hat{w}$ scattered throughout the $(x,y)$ lattice.
-Even given the correct Farris topology we obtain poor estimates for the value of the interior branch length.
+In summary, even given the correct Farris topology we obtain poor estimates for the value of the interior branch length.
 

### 2017-10-02 deprecated.tex
text in appendix and figures
fixed case done? @_@ very close...

@@
+\section{Disjointed notes}
+
+\subsection{Branch lengths with $x$ and $y$ known}
+
+In this case, the likelihood as a function of $w'$ is proportional to either
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
+    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
+    &\qquad + \log(1+w')
+\end{align*}
+or
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
+    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
+    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
+    &\qquad + p_{13}\cdot\log(1-w').
+\end{align*}
+Some computer work shows that the first likelihood will always be chosen over the second when maximizing ancestral states, so we focus on that.
+The likelihood can be written
+$$
+\ell = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + log(1+w')
+$$
+where
+$$
+\alpha_1 = 1+x^2+y^2+4xy^2+x^2y^2,
+\alpha_2 = 1+x^2+y^2-4xy^2+x^2y^2,
+\beta = 1+x^2+y^2+x^2y^2,
+\gamma = 4xy.
+$$
+The derivative of this is quadratic in w', with
+$$
+\ell' = (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2) + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) + w^2(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2),
+$$
+so we have a closed form solution for $w'$ via the quadratic formula, where
+$$
+a = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
+b = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
+c = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
+$$
+and
+$$
+\hat{w}_1' = (-b + \sqrt{b^2 - 4ac}) / 2a,
+\hat{w}_2' = (-b - \sqrt{b^2 - 4ac}) / 2a.
+$$
+These are all functions of $x$ and $y$.
+Let's plan to get these in as simplified a form as possible.
+Then, we need cases where either $\hat{w}_1'\in[0,1]$ or $\hat{w}_2'\in[0,1]$, and, if not, we set them equal to their extrema.
+Whichever yields a higher value of the likelihood will then be the maximizer.
+The interesting case, to me, is that where $\hat{w}_2' \ge 1$, which is the case where we estimate $w'$ to be one.
+We can plot this in two variables, as we did before.
+Let's see.
+
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{bl-inconsistency}
+\caption{Region of inconsistency with $x$ and $y$ fixed}
+\label{fig:bl-inconsistency}
+\end{figure}
+
+\subsection{Below is wrong}
+
+My initial hunch is that $w'$ is always equal to 1, though the empirical fits don't show this.
+I argue that $w'=1$ always like so: consider just the last term of the likelihood $\log(1+w')$.
+This is obviously maximized when $w'=1$.
+For the remaining two terms, $p_{\emptyset} \ge p_{13}$, with equality only when $x=y=0$.
+So, excepting the case of infinite branch lengths, the term $\log(1+x^2+y^2+4xyw'+x^2y^2)$ will always be weighted more than $\log(1+x^2+y^2-4xyw'+x^2y^2)$.
+Additionally, these two terms can be written as
+$$
+c_1 \cdot f(x, y, w') + c_2 \cdot f(x, y, -w')
+$$
+where $c_1 > c_2$.
+Clearly $\hat{w}_1'=1$ where
+$$
+\hat{w}_1' = \arg\max \ f(x, y, w')
+$$
+and
+$\hat{w}_2'=0$ where
+$$
+\hat{w}_2' = \arg\max \ f(x, y, -w').
+$$
+I would think that $\hat{w}'=1$ where
+$$
+\hat{w}_2' = \arg\max \ c_1 \cdot f(x, y, w') + c_2 \cdot f(x, y, -w').
+$$
+This is clearly true if $f$ is linear in $w'$.
+Here, though, $f$ is $log(a+b*w')$.
+We have ranges $a\in[1,4]$ and $b\in[0,4]$ where $a$ increases quartically with $x$ and $y$ while $b$ increases quadratically.
+
 \section{Temporarily unused text I may want later to show more general results on the likelihood estimates}
+
+\begin{table}
+\centering
+\begin{tabular}{|l|l|l|}
+\multicolumn{3}{c}{Classical}\\
+    \hline
+$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t^*)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t^*)$\\
+    \hline
+    $\emptyset$    &$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
+    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
+    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
+    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
+    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
+    $\{1,3\}$      &$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
+    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
+    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
+    \hline
+\multicolumn{3}{c}{Modified}\\
+    \hline
+$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
+    \hline
+    $\emptyset$     &$1+x'^2+y'^2+4x'y'w'+x'^2y'^2$&$1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
+    $\{1\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
+    $\{2\}$        &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
+    $\{3\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
+    $\{1,2\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
+    $\{1,3\}$      &$1+x'^2+y'^2-4x'y'w'+x'^2y'^2$&$1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
+    $\{2,3\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
+    $\{1,2,3\}$    &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
+    \hline
+\end{tabular}
+\caption{Site pattern probabilities.
+All values are multiplied by $1/8$.}
+\label{tab:sitepatprob}
+\end{table}
 

### 2017-10-02 joint_inf.tex
fixed case done? @_@ very close...

@@
 \centering
-\includegraphics[width=.9\textwidth]{analytic-inconsistency}
+\includegraphics[width=.9\textwidth]{analytic-inconsistency-legend}
 \caption{
@@
     &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
-    &\qquad + \log((1+w'))
+    &\qquad + \log(1+w')
 \end{align*}
@@
     &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
-    &\qquad + (1 - p_{13})\cdot\log((1+w')) \\
-    &\qquad + p_{13}\cdot\log((1-w')).
-\end{align*}
-When computing the global minimum for $0 \le w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}$ in Fig.~\ref{fig:w_hat}.
+    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
+    &\qquad + p_{13}\cdot\log(1-w').
+\end{align*}
+
+\subsection{Analytic}
+
+In this case, the likelihood as a function of $w'$ is proportional to either
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
+    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
+    &\qquad + \log(1+w')
+\end{align*}
+or
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
+    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
+    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
+    &\qquad + p_{13}\cdot\log(1-w').
+\end{align*}
+At the risk of sounding circular, if $\hat{w}'$ is estimated to be one, then the first likelihood will be chosen, so the region in which that likelihood estimates $\hat{w}'$ to be one will always do so; let's focus on the first likelihood then.
+The likelihood can be written
+$$
+\ell = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
+$$
+where
+$$
+\alpha_1 = 1+x^2+y^2+4xy^2+x^2y^2,
+$$
+$$
+\alpha_2 = 1+x^2+y^2-4xy^2+x^2y^2,
+$$
+$$
+\beta = 1+x^2+y^2+x^2y^2,
+$$
+$$
+\gamma = 4xy.
+$$
+The derivative of this is quadratic in w', with
+$$
+\ell' = (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2) + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) + w^2(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2),
+$$
+so we have a closed form solution for $w'$ via the quadratic formula, where
+$$
+a = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
+$$
+$$
+b = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
+$$
+$$
+c = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
+$$
+and
+$$
+\hat{w}_1' = (-b + \sqrt{b^2 - 4ac}) / 2a,
+$$
+$$
+\hat{w}_2' = (-b - \sqrt{b^2 - 4ac}) / 2a.
+$$
+These are all functions of $x$ and $y$.
+Let's plan to get these in as simplified a form as possible.
+Then, we need cases where either $\hat{w}_1'\in[0,1]$ or $\hat{w}_2'\in[0,1]$, and, if not, we set them equal to their extrema.
+Whichever yields a higher value of the likelihood will then be the maximizer.
+The interesting case, to me, is that where $\hat{w}_2' \ge 1$, which is the case where we guarantee $w'$ will be one.
+We can plot this in two variables, as we did before.
+Let's see.
+
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{bl-inconsistency}
+\caption{Region of inconsistency with $x$ and $y$ fixed}
+\label{fig:bl-inconsistency}
+\end{figure}
+
+\subsection{Numerical}
+
+When computing the global maximum for $0 \le w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}$ in Fig.~\ref{fig:w_hat}.
 We find that the estimate for $\hat{w}$ will typically be an overestimate, and in fact is almost always identically equal to 1, indicating no change along that particular branch.

### 2017-10-04 deprecated.tex
fix up arguments for restricted case; disjoint ideas for general case

@@
 \end{figure}
+
+\subsection{Branch lengths with $x$ and $y$ unknown}
+
+Our plan of attack is to focus on the likelihood
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
+    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
+    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
+    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
+    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
+    &\qquad - \log(8) \\
+    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
+    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
+    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
+    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
+    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
+    &\qquad + p_{13}         \cdot\log((1-x')^2   (1+w')(1+y')^2) \\
+    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
+    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
+\end{align*}
+This simplifies to
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + (p_{1} + p_{3})          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
+    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
+    &\qquad + (p_{12} + p_{23})         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
+    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
+    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
+    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
+    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
+    &\qquad + \log(1+w').
+\end{align*}
+This likelihood occurs when $\hat{w}' = 1$ and $\hat{y}' > \hat{x}'$, the latter coming from
+$$
+(1-x'^2)(1+y'^2) > (1+x'^2)(1-y'^2)
+$$
+only if $y' > x'$.
+We simply need to obtain
+$$
+(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x, y, w} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+$$
+as functions of $x$ and $y$ and verify the region where $\hat{w}'(x, y) = 1$ and $\hat{y}'(x, y) > \hat{x}'(x, y)$.
+This region will be consistent with one always estimating $\hat{w}'$ to be one.
+This can similarly be done for the other likelihood term with $\hat{x}'(x, y) > \hat{y}'(x, y)$.
+
+To do is to find these functions---or subsets of their varieties---in as closed a form as possible.
+
+When we are in the region where $\hat{w}'$ is one, necessarily (I think) our region is going to be a subset of the case where $x'$ and $y'$ are fixed.
 

### 2017-10-04 joint_inf.tex
fix up arguments for restricted case; disjoint ideas for general case

--- 2017-08-01 joint_inf.tex
@@
 \centering
-\includegraphics[width=.9\textwidth]{analytic-inconsistency}
+\includegraphics[width=.9\textwidth]{topology-inconsistency-legend}
 \caption{
@@
     &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
-    &\qquad + (1 - p_{13})\cdot\log((1+w')) \\
-    &\qquad + p_{13}\cdot\log((1-w')).
-\end{align*}
-When computing the global minimum for $0 \le w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}$ in Fig.~\ref{fig:w_hat}.
-We find that the estimate for $\hat{w}$ will typically be an overestimate, and in fact is almost always identically equal to 1, indicating no change along that particular branch.
+    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
+    &\qquad + p_{13}\cdot\log(1-w').
+\end{align*}
+Since
+$$
+\log(1+w') >= (1 - p_{13})\cdot\log(1+w') + p_{13}\cdot\log(1-w')
+$$
+with equality only when $w'=0$, we focus on the first likelihood.
+Simplifying and suppressing some variables, this likelihood can be written
+$$
+\ell(w') = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
+$$
+where
+$$
+\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
+$$
+$$
+\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
+$$
+$$
+\beta := \beta(x, y) = 1+x^2+y^2+x^2y^2,
+$$
+$$
+\gamma := \gamma(x, y) = 4xy.
+$$
+Since $x$ and $y$ fall between zero and one, we make use of the inequalities (excepting the cases of $x=y=0$ and $x=y=1$)
+$$
+\alpha_1 > \alpha_2
+$$
+and
+$$
+\beta > \gamma.
+$$
+% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
+The derivative is
+$$
+\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} + \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
+$$
+The inequality $\beta > \gamma$ implies that any inequality relating $\ell'$ and 0 is equivalent to the same inequality relating 0 and the quadratic function
+\begin{align*}
+f(w') &= w^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
+      &\qquad + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
+      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2), \\
+\end{align*}
+implying $\ell'$ has two zeros according to the quadratic formula.
+By $\alpha_1 > \alpha_2$, $\ell'(0) > 0$ implies $\ell$ is increasing at $w'=0$.
+We use this fact to show that, if the smaller of the zeros of $f(w')$ is greater than one, $\hat{w}' \equiv 1$.
+Using the quadratic formula with
+$$
+a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
+$$
+$$
+b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
+$$
+$$
+c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
+$$
+the smaller zero is
+$$
+\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)},
+$$
+which is a function of the generating parameters $x$ and $y$.
+We plot the region where $\hat{w}'(x, y) \ge 1$ in Figure~\ref{fig:bl-inconsistency}.
 
@@
 \centering
-    \begin{tabular}{cc}
-\includegraphics[width=.45\textwidth]{w_hat_heatmap} &
-\includegraphics[width=.45\textwidth]{w_hat_heatmap_no_joint}\\
-Joint inference & Marginal inference \\
-    \end{tabular}
-\caption{
-    Estimating $\hat{w}$ for $x'$ and $y'$ fixed at their true values.
-}
-\label{fig:w_hat}
+\includegraphics[width=.9\textwidth]{branch-length-inconsistency-legend}
+\caption{Region of inconsistency with $x$ and $y$ fixed}
+\label{fig:bl-inconsistency}
 \end{figure}
 
-Second, we consider the more standard case where the true values of $x$ and $y$ are not known.
-The form of the likelihood will alter the fitted values of $x$ and $y$ so that even with perfect information we will not necessarily estimate $\hat{x}=x$ and $\hat{y}=y$.
-If we estimate $(x', y', w')$ jointly using the same approach above, we obtain the fit $\hat{w}$ in Figure~\ref{fig:w_hat_joint}.
+We consider the more standard case where the true values of $x$ and $y$ are not known.
+The form of the likelihood will alter the fitted values of $x$ and $y$ so that even with perfect information we will not necessarily estimate $\hat{x}'=x$ and $\hat{y}'=y$.
+When computing the global maximum for $0 \le x',y',w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}'$ in Fig.~\ref{fig:w_hat_joint}.
+The region in which $\hat{w}'$ is estimated to be one---indicating no change along that particular branch---is now much smaller, but still nontrivial.
+Additionally, the estimates are surprisingly unstable, with estimates of 1 for $\hat{w}'$ scattered throughout the $(x,y)$ lattice.
+Even given the correct Farris topology we have a region of the space of true parameters where we will always estimate the interior branch parameter to be one.
 
@@
 \caption{
-    Estimating $(\hat{x}, \hat{y}, \hat{w})$.
+    Estimating $(\hat{x}', \hat{y}', \hat{w}')$.
 }
@@
 \end{figure}
-
-We no longer consistently overestimate $\hat{w}$, but the estimates are surprisingly unstable, with estimates of 1 for $\hat{w}$ scattered throughout the $(x,y)$ lattice.
-In summary, even given the correct Farris topology we obtain poor estimates for the value of the interior branch length.
 

### 2017-10-09 deprecated.tex
a few ideas trying to work through general case

@@
 only if $y' > x'$.
+
 We simply need to obtain
@@
 
+\subsubsection{Maximization of functionals---calculus of variations?}
+
+Call the likelihood $\ell$.
+The likelihood is maximized at functions of $x$ and $y$ with the property that $\hat{x}'(x,y) = \hat{y}'(y,x)$.
+Let's call $\hat{y}'(x, y) := g(x, y)$.
+Look at
+\begin{align*}
+    \ell
+    &=        p_{\emptyset}  \cdot\log(1+g(y, x)^2+g(x, y)^2+4g(y, x)g(x, y)w'+g(y, x)^2g(x, y)^2) \\
+    &\qquad   p_{13}  \cdot\log(1+g(y, x)^2+g(x, y)^2-4g(y, x)g(x, y)w'+g(y, x)^2g(x, y)^2) \\
+    &\qquad + (p_{1} + p_{3})          \cdot\log(1-g(y, x)^2+g(x, y)^2-g(y, x)^2g(x, y)^2) \\
+    &\qquad + (p_{2} + p_{123})          \cdot\log(1+g(y, x)^2-g(x, y)^2-g(y, x)^2g(x, y)^2) \\
+    &\qquad + (p_{12} + p_{23})         \cdot\log(1-g(y, x)^2-g(x, y)^2+g(y, x)^2g(x, y)^2) \\
+    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+g(y, x)) \\
+    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-g(y, x)) \\
+    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+g(x, y)) \\
+    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-g(x, y)) \\
+    &\qquad + \log(1+w')
+\end{align*}
+with the restriction that $g(x, y) > g(y, x)$.
+We now have two problems: 1) what to do with $w'$?; and how to solve the two-variable functional maximization?
+We may be able to take $\hat{w}'(x, y)$ to be like our quadratic formula version but now with 
+$$
+\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
+$$
+$$
+\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
+$$
+$$
+\beta := \beta(x, y) = 1+g(y, x)^2+g(x, y)^2+g(y, x)^2g(x, y)^2,
+$$
+$$
+\gamma := \gamma(x, y) = 4g(y, x)g(x, y).
+$$
+So in a way we want again the region where $\hat{w}'(x, y) \ge 1$ with
+$$
+\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)}.
+$$
+We have a functional maximization of one two-variable function such that
+\begin{align*}
+    \ell
+    &=        p_{\emptyset}  \cdot\log(1+g(y, x)^2+g(x, y)^2+4g(y, x)g(x, y)\hat{w}'(g)+g(y, x)^2g(x, y)^2) \\
+    &\qquad   p_{13}  \cdot\log(1+g(y, x)^2+g(x, y)^2-4g(y, x)g(x, y)\hat{w}'(g)+g(y, x)^2g(x, y)^2) \\
+    &\qquad + (p_{1} + p_{3})          \cdot\log(1-g(y, x)^2+g(x, y)^2-g(y, x)^2g(x, y)^2) \\
+    &\qquad + (p_{2} + p_{123})          \cdot\log(1+g(y, x)^2-g(x, y)^2-g(y, x)^2g(x, y)^2) \\
+    &\qquad + (p_{12} + p_{23})         \cdot\log(1-g(y, x)^2-g(x, y)^2+g(y, x)^2g(x, y)^2) \\
+    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+g(y, x)) \\
+    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-g(y, x)) \\
+    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+g(x, y)) \\
+    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-g(x, y)) \\
+    &\qquad + \log(1+\hat{w}'(g)).
+\end{align*}
+
+\subsection{Below is wrong}
+
 When we are in the region where $\hat{w}'$ is one, necessarily (I think) our region is going to be a subset of the case where $x'$ and $y'$ are fixed.
-
-\subsection{Below is wrong}
 

### 2017-10-10 joint_inf.tex
a few ideas trying to work through general case
clarify and update plot

@@
 This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
+% In Semple/Steel, there are some assumptions that are necessary for some of the results, called M1, M2 and M3.
+% M1 and M2 seem definitely necessary, which pretty much just imply x and y can't be zero.
+% M3 implies x and y can't be one?
+% (das TODO: double check these; currently working from notes and memory)
 Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.
@@
 Directly substituting in \eqref{eq:site_pattern_likelihood}, the log likelihood is
-\begin{align*}
+\begin{align}
+    \label{eq:farris_likelihood}
     \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
-    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
-    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
-    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
-    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
-    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
-    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
-    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
-    &\qquad - \log(8) \\
-    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
-    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
-    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
-    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
-    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
-    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
-    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
+    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \nonumber \\
+    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
+    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\
+    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
+    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
+    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \nonumber \\
+    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
+    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\
+    &\qquad - \log(8) \nonumber \\
+    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \nonumber \\
+    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
+    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \nonumber \\
+    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
+    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\
+    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \nonumber \\
+    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\
     &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
-\end{align*}
+\end{align}
 Seeking an upper bound, we use the facts that, for $x',y'\in[0,1]$,
@@
 In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
-In this case, the likelihood as a function of $w'$ is proportional to either
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
-    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
-    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
-    &\qquad + \log(1+w')
-\end{align*}
-or
+Recalling Table~\ref{tab:likelihoods}, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood for $(x',y',w')$.
+If $x'$ and $y'$ are fixed, the likelihood in \eqref{eq:farris_likelihood} as a function of $w'$ is proportional to
 \begin{align*}
@@
     &\qquad + (1 - p_{13})\cdot\log(1+w') \\
-    &\qquad + p_{13}\cdot\log(1-w').
+    &\qquad + p_{13}\cdot\log(1-w')
+\end{align*}
+by collapsing like terms and ignoring terms with only $x'$ and $y'$.
+Similarly, the other two possibilities for the likelihood are both proportional to
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
+    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
+    &\qquad + \log(1+w').
 \end{align*}
@@
 $$
-with equality only when $w'=0$, we focus on the first likelihood.
+with equality only when $w'=0$, we focus on the second likelihood.
 Simplifying and suppressing some variables, this likelihood can be written
@@
 $$
-\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} + \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
+\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} - \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
 $$

### 2017-10-10 joint_inf_bl_analytic.py
code for plotting branch length inconsistency region

@@ -1,0 +1,35 @@
+# Branch length inconsistency
+
+import matplotlib.pyplot as plt
+from numpy import sqrt, log, meshgrid, arange, maximum
+
+def get_what(x, y):
+    alpha_1 = 1+x**2+y**2+4*x*y**2+x**2*y**2
+    alpha_2 = 1+x**2+y**2-4*x*y**2+x**2*y**2
+    beta =    1+x**2+y**2+x**2*y**2
+    gamma = 4*x*y
+
+    a = -gamma**2*alpha_1 - gamma**2*alpha_2 - gamma**2
+    b = gamma*alpha_1*beta - gamma**2*alpha_1 - gamma*alpha_2*beta - gamma**2*alpha_2
+    c = gamma*alpha_1*beta - gamma*alpha_2*beta + beta**2
+
+    what_1 = (-b + sqrt(b**2 - 4*a*c)) / (2*a)
+    what_2 = (-b - sqrt(b**2 - 4*a*c)) / (2*a)
+
+    return (what_2 >= 1)
+
+# Want cases where what_2 is greater than one for x and y between 0 and 1
+
+delta = 0.001
+x_range = arange(0.0, 1.0, delta)
+y_range = arange(0.0, 1.0, delta)
+X, Y = meshgrid(x_range,y_range)
+
+region = get_what(X, Y)
+
+plt.contour(X, Y, region, [0])
+plt.xlabel(r'x', fontsize=16)
+plt.ylabel(r'y', fontsize=16)
+plt.title(r'Branch length inconsistency')
+plt.savefig('figures/branch-length-inconsistency.svg')
+

### 2017-10-13 deprecated.tex
more thoughts

@@
 only if $y' > x'$.
-
 We simply need to obtain
@@
 This can similarly be done for the other likelihood term with $\hat{x}'(x, y) > \hat{y}'(x, y)$.
-
 To do is to find these functions---or subsets of their varieties---in as closed a form as possible.
+
+Simplifying further,
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2) \\
+    &\qquad + (p_{1} + p_{3} + p_{12} + p_{23})          \cdot\log(1-x'^2) \\
+    &\qquad + (p_{1} + p_{3})          \cdot\log(1+y'^2) \\
+    &\qquad + (p_{2} + p_{123} + p_{12} + p_{23})          \cdot\log(1-y'^2) \\
+    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
+    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
+    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
+    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
+    &\qquad + \log(1+w').
+\end{align*}
+% das: there might be something with 1+y^2 < 1+y here? I'm not sure how specifically it would affect the maximization, and it might be too loose of a bound, but could work to try
+Say we had estimated $\hat{x}'$ and $\hat{y}'$.
+We can use a similar argument as in the restricted case, though now
+$$
+\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
+$$
+$$
+\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
+$$
+$$
+\beta := \beta(x', y') = 1+x'^2+y'^2+x'^2y'^2,
+$$
+$$
+\gamma := \gamma(x', y') = 4x'y'.
+$$
+So instead of the lower bound for $\hat{w}'$ being a function of $x$ and $y$ through $\alpha_1,\alpha_2,\beta$ and $\gamma$, it is a function of $x$ and $y$ through $\beta$ and $\gamma$ via $\hat{x}'$ and $\hat{y}'$.
+Operating on the assumption $\hat{y}' > \hat{x}'$, which we needed to specify the form of the likelihood, we now see
+$$
+1 < 1 + 2x'^2 + x'^4 < \beta < 1+2y'^2+y'^4 < 4
+$$
+and
+$$
+0 < 4x'^2 < \gamma < 4y'^2 < 4.
+$$
+We may be able to use this to restrict our attention to fewer variables, but we need to determine how $a,b$ and $c$ are varying with $\beta$ and $\gamma$ to see which inequalities are useful.
+Recall
+$$
+\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)}
+$$
+where
+$$
+a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
+$$
+$$
+b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
+$$
+$$
+c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2.
+$$
+We know for sure $a$ is decreasing with $\gamma$ and $c$ is increasing with $\gamma$, while $c$ is increasing with $\beta$.
 

### 2017-10-13 deprecated.tex
move most recent results out of deprecated

--- 2017-10-09 deprecated.tex
@@
 
-\subsection{Branch lengths with $x$ and $y$ known}
-
-In this case, the likelihood as a function of $w'$ is proportional to either
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
-    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
-    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
-    &\qquad + \log(1+w')
-\end{align*}
-or
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
-    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
-    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
-    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
-    &\qquad + p_{13}\cdot\log(1-w').
-\end{align*}
-Some computer work shows that the first likelihood will always be chosen over the second when maximizing ancestral states, so we focus on that.
-The likelihood can be written
-$$
-\ell = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + log(1+w')
-$$
-where
-$$
-\alpha_1 = 1+x^2+y^2+4xy^2+x^2y^2,
-\alpha_2 = 1+x^2+y^2-4xy^2+x^2y^2,
-\beta = 1+x^2+y^2+x^2y^2,
-\gamma = 4xy.
-$$
-The derivative of this is quadratic in w', with
-$$
-\ell' = (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2) + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) + w^2(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2),
-$$
-so we have a closed form solution for $w'$ via the quadratic formula, where
-$$
-a = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
-b = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
-c = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
-$$
-and
-$$
-\hat{w}_1' = (-b + \sqrt{b^2 - 4ac}) / 2a,
-\hat{w}_2' = (-b - \sqrt{b^2 - 4ac}) / 2a.
-$$
-These are all functions of $x$ and $y$.
-Let's plan to get these in as simplified a form as possible.
-Then, we need cases where either $\hat{w}_1'\in[0,1]$ or $\hat{w}_2'\in[0,1]$, and, if not, we set them equal to their extrema.
-Whichever yields a higher value of the likelihood will then be the maximizer.
-The interesting case, to me, is that where $\hat{w}_2' \ge 1$, which is the case where we estimate $w'$ to be one.
-We can plot this in two variables, as we did before.
-Let's see.
-
-\begin{figure}
-\centering
-\includegraphics[width=.9\textwidth]{bl-inconsistency}
-\caption{Region of inconsistency with $x$ and $y$ fixed}
-\label{fig:bl-inconsistency}
-\end{figure}
-
 \subsection{Branch lengths with $x$ and $y$ unknown}
-
-Our plan of attack is to focus on the likelihood
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
-    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
-    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
-    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
-    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
-    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
-    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
-    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
-    &\qquad - \log(8) \\
-    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
-    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
-    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
-    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
-    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
-    &\qquad + p_{13}         \cdot\log((1-x')^2   (1+w')(1+y')^2) \\
-    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
-    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
-\end{align*}
-This simplifies to
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
-    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
-    &\qquad + (p_{1} + p_{3})          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
-    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
-    &\qquad + (p_{12} + p_{23})         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
-    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
-    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
-    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
-    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
-    &\qquad + \log(1+w').
-\end{align*}
-This likelihood occurs when $\hat{w}' = 1$ and $\hat{y}' > \hat{x}'$, the latter coming from
-$$
-(1-x'^2)(1+y'^2) > (1+x'^2)(1-y'^2)
-$$
-only if $y' > x'$.
-
-We simply need to obtain
-$$
-(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x, y, w} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-$$
-as functions of $x$ and $y$ and verify the region where $\hat{w}'(x, y) = 1$ and $\hat{y}'(x, y) > \hat{x}'(x, y)$.
-This region will be consistent with one always estimating $\hat{w}'$ to be one.
-This can similarly be done for the other likelihood term with $\hat{x}'(x, y) > \hat{y}'(x, y)$.
-
-To do is to find these functions---or subsets of their varieties---in as closed a form as possible.
 

### 2017-10-13 joint_inf.tex
add most recent general case thoughts to main document

@@
 
+\subsection{Fixed $x'$ and $y'$}
+
 First, assume we know the true values of $x'$ and $y'$ but not $w'$.
@@
 
+\subsection{General case}
+
 We consider the more standard case where the true values of $x$ and $y$ are not known.
 The form of the likelihood will alter the fitted values of $x$ and $y$ so that even with perfect information we will not necessarily estimate $\hat{x}'=x$ and $\hat{y}'=y$.
+
+Our plan of attack is to determine inequalities at which a particular form of likelihood is necessary and plot the regions of $x$ and $y$ where these occur in addition to the region where $\hat{w}'$ is estimated to be one.
+Since we are interested in the region where $\hat{w}'=1$, we choose either the term $(1-x')^2(1+w')(1+y')^2$ or $(1+x')^2(1+w')(1-y')^2$ in Table~\ref{tab:likelihoods} for $\siteSplit_j=\{1,3\}$.
+Now,
+$$
+(1-x'^2)(1+y'^2) > (1+x'^2)(1-y'^2),
+$$
+so we are also interested in the inequality $\hat{y}' > \hat{x}'$ if we use the likelihood
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
+    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
+    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
+    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
+    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
+    &\qquad - \log(8) \\
+    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
+    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
+    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
+    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
+    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
+    &\qquad + p_{13}         \cdot\log((1-x')^2   (1+w')(1+y')^2) \\
+    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
+    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
+\end{align*}
+This simplifies to
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + (p_{1} + p_{3})          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
+    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
+    &\qquad + (p_{12} + p_{23})         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
+    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
+    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
+    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
+    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
+    &\qquad + \log(1+w').
+\end{align*}
+We need to obtain
+$$
+(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x, y, w} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+$$
+as functions of $x$ and $y$ and verify the region where $\hat{w}'(x, y) = 1$ and $\hat{y}'(x, y) > \hat{x}'(x, y)$.
+This region will be consistent with one always estimating $\hat{w}'$ to be one.
+This can similarly be done for the other likelihood term with $\hat{x}'(x, y) > \hat{y}'(x, y)$.
+To do is to find these functions---or subsets of their varieties---in as closed a form as possible.
+
+As a first attempt, simplify further to
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2) \\
+    &\qquad + (p_{1} + p_{3} + p_{12} + p_{23})          \cdot\log(1-x'^2) \\
+    &\qquad + (p_{1} + p_{3})          \cdot\log(1+y'^2) \\
+    &\qquad + (p_{2} + p_{123} + p_{12} + p_{23})          \cdot\log(1-y'^2) \\
+    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
+    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
+    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
+    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
+    &\qquad + \log(1+w').
+\end{align*}
+% das: there might be something with 1+y^2 < 1+y here?
+% I'm not sure how specifically it would affect the maximization, and it might be too loose of a bound, but could work to try.
+% The first two terms kind of gunk up the works a little.
+Say we had estimated $\hat{x}'$ and $\hat{y}'$.
+We can use a similar argument as in the restricted case, though now
+$$
+\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
+$$
+$$
+\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
+$$
+$$
+\beta := \beta(x', y') = 1+x'^2+y'^2+x'^2y'^2,
+$$
+$$
+\gamma := \gamma(x', y') = 4x'y'.
+$$
+So instead of the lower bound for $\hat{w}'$ being a function of $x$ and $y$ through $\alpha_1,\alpha_2,\beta$ and $\gamma$, it is a function of $x$ and $y$ through $\beta$ and $\gamma$ via $\hat{x}'$ and $\hat{y}'$.
+Operating on the assumption $\hat{y}' > \hat{x}'$, which we needed to specify the form of the likelihood, we now see
+$$
+1 < 1 + 2x'^2 + x'^4 < \beta < 1+2y'^2+y'^4 < 4
+$$
+and
+$$
+0 < 4x'^2 < \gamma < 4y'^2 < 4.
+$$
+We may be able to use this to restrict our attention to fewer variables, but we need to determine how $a,b$ and $c$ are varying with $\beta$ and $\gamma$ to see which inequalities are useful.
+Recall
+$$
+\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)}
+$$
+where
+$$
+a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
+$$
+$$
+b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
+$$
+$$
+c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2.
+$$
+We know for sure $a$ is decreasing with $\gamma$ and $c$ is increasing with $\gamma$, while $c$ is increasing with $\beta$.
+
+\subsection{Empirical results for the general case}
+
 When computing the global maximum for $0 \le x',y',w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}'$ in Fig.~\ref{fig:w_hat_joint}.

### 2017-10-14 joint_inf.tex
pass through the fixed x' and y' part.

@@
 
-Consider only the Farris-tree topology.
-We will consider two settings, corresponding to whether some branch lengths are known ahead of time or not.
+We will now consider the problem of branch length estimation on the correct tree using joint ancestral state and branch length estimation.
+We will consider two settings on the Farris topology, corresponding to whether some branch lengths are known ahead of time or not.
+As above, we will assume that data is generated with two sister branches of fidelity $x$, with all other branches of length $y$.
+%EM We will show XXX... and introduce \hat{w'} here.
 
@@
 Recalling Table~\ref{tab:likelihoods}, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood for $(x',y',w')$.
-If $x'$ and $y'$ are fixed, the likelihood in \eqref{eq:farris_likelihood} as a function of $w'$ is proportional to
+%EM You are talking about a log likelihood here, right? So a likelihood being proportional to means that there is some additive constant. Makes things a little confusing.
+If $x'$ and $y'$ are fixed, by collapsing like terms and ignoring terms with only $x'$ and $y'$, the likelihood in \eqref{eq:farris_likelihood} as a function of $w'$ is proportional to
 \begin{align*}
@@
     &\qquad + (1 - p_{13})\cdot\log(1+w') \\
-    &\qquad + p_{13}\cdot\log(1-w')
-\end{align*}
-by collapsing like terms and ignoring terms with only $x'$ and $y'$.
+    &\qquad + p_{13}\cdot\log(1-w').
+\end{align*}
 Similarly, the other two possibilities for the likelihood are both proportional to
@@
 $$
-\log(1+w') >= (1 - p_{13})\cdot\log(1+w') + p_{13}\cdot\log(1-w')
-$$
-with equality only when $w'=0$, we focus on the second likelihood.
+\log(1+w') \geq (1 - p_{13})\cdot\log(1+w') + p_{13}\cdot\log(1-w')
+$$
+with equality only when $w'=0$, the second likelihood is the likelihood with the maximum-likelihood ancestral state, and so we need only analyze that case.
+
+Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
+%EM Not sure what's being suppressed here. Sounds naughty.
 Simplifying and suppressing some variables, this likelihood can be written
@@
 % Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
-The derivative is
+The derivative of the log likelihood is
 $$
@@
 $$
-The inequality $\beta > \gamma$ implies that any inequality relating $\ell'$ and 0 is equivalent to the same inequality relating 0 and the quadratic function
-\begin{align*}
-f(w') &= w^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
-      &\qquad + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
-      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2), \\
-\end{align*}
+The inequality $\beta > \gamma$ implies that this function stays finite.
+%EM I don't think you mean "any inequality"... what about sin(ell') > 0?
+any inequality relating $\ell'$ and 0 is equivalent to the same inequality relating 0 and the quadratic function
+\begin{align*}
+f(w') &= (w')^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
+      &\qquad + w'\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
+      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2),
+\end{align*}
+%EM Is this obvious at this point? Couldn't b^2 - 4ac be zero in principle?
 implying $\ell'$ has two zeros according to the quadratic formula.
-By $\alpha_1 > \alpha_2$, $\ell'(0) > 0$ implies $\ell$ is increasing at $w'=0$.
-We use this fact to show that, if the smaller of the zeros of $f(w')$ is greater than one, $\hat{w}' \equiv 1$.
+Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w'=0$.
+This implies that if the smaller of the zeros of $f(w')$ is greater than one, then $\hat{w}' \equiv 1$.
 Using the quadratic formula with

### 2017-10-16 joint_inf.tex
Minor points.
address some comments and a little attempt

@@
 %EM Not sure what's being suppressed here. Sounds naughty.
-Simplifying and suppressing some variables, this likelihood can be written
+%das: I meant suppressing terms not involving $w'$, but I guess I already did that...
+Simplifying, the likelihood can be written
 $$
@@
 % Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
+%EM Rather confusing to be using ' for two things here.
+What about using a dot over the l?
+In retrospect something other than ' would have been good because of the proliferation of w^{'2} etc.
+Prob not worth changing now though.
 The derivative of the log likelihood is
@@
 $$
-The inequality $\beta > \gamma$ implies that this function stays finite.
 %EM I don't think you mean "any inequality"... what about sin(ell') > 0?
-any inequality relating $\ell'$ and 0 is equivalent to the same inequality relating 0 and the quadratic function
+The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w') \lessgtr 0$, we equivalently consider $f(w') \lessgtr 0$ where $f$ is the quadratic function
 \begin{align*}
@@
       &\qquad + w'\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
-      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2),
+      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
 \end{align*}
 %EM Is this obvious at this point? Couldn't b^2 - 4ac be zero in principle?
-implying $\ell'$ has two zeros according to the quadratic formula.
+%das: only if x or y is zero (though one could be nonzero).
+%b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero.
+%c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
+%Added clarification.
+Excepting all the cases where $x=0$ or where $y=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
 Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w'=0$.
@@
 Since we are interested in the region where $\hat{w}'=1$, we choose either the term $(1-x')^2(1+w')(1+y')^2$ or $(1+x')^2(1+w')(1-y')^2$ in Table~\ref{tab:likelihoods} for $\siteSplit_j=\{1,3\}$.
-Now,
-$$
-(1-x'^2)(1+y'^2) > (1+x'^2)(1-y'^2),
-$$
-so we are also interested in the inequality $\hat{y}' > \hat{x}'$ if we use the likelihood
+%EM I don't think I follow the thread of argument here-- what's being assumed when? I think you are assuming $\hat{y}' > \hat{x}'$ ?
+Now, assume that, having estimated $\hat{x}'$ and $\hat{y}'$, $\hat{y}' > \hat{x}'$.
+In this case,
+$$
+(1-\hat{x}'^2)(1+\hat{y}'^2) > (1+\hat{x}'^2)(1-\hat{y}'^2),
+$$
+so we use the likelihood
 \begin{align*}
@@
     &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
+    &\qquad + \log(1+w')
+\end{align*}
+and
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2) \\
+    &\qquad + (p_{1} + p_{3})          \cdot\log(1+y'^2) \\
+    &\qquad + 2(1-p_{13})\log(1+x') \\
+    &\qquad + 2(p_{1}+p_{3}+p_{12}+p_{23}+p_{13})\log(1-x') \\
+    &\qquad + 2\log(1+y') \\
+    &\qquad + 2(p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
     &\qquad + \log(1+w').
 \end{align*}
+Say we had estimated $\hat{w}'=1$.
+Then we need is the region where $\hat{y}'>\hat{x}'$ and $0<\hat{x}',\hat{y}'<1$---the case for $\hat{x}'>\hat{y}'$ will be a reflection of this case.
+Let's consider first $x'=y'$; if we can show that for $x,y<\epsilon$ where it looks like $\epsilon\approx 0.4$ then $\hat{w}'=1$ we are at least getting close to something.
+So fix $x'=y'=u$ and get
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, u; \fullAncestralSplitPartitions(\tau_1,u))
+    &\propto   p_{\emptyset}  \cdot\log(1+6u^2+u^4) \\
+    &\qquad + p_{13}         \cdot\log(1-2u^2+u^4) \\
+    &\qquad + (p_{1} + p_{2} + p_{3} + p_{123})          \cdot\log(1+u^2) \\
+    &\qquad + (4-2p_{13})\log(1+u) \\
+    &\qquad + 2(p_{1}+p_{2}+p_{3}+p_{123}+p_{13}+2p_{12}+2p_{23})\log(1-u) \\
+\end{align*}
+which is
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, u; \fullAncestralSplitPartitions(\tau_1,u))
+    &\propto   p_{\emptyset}  \cdot\log(1+6u^2+u^4) \\
+    &\qquad + (p_{13} + p_{1} + p_{2} + p_{3} + p_{123})          \cdot\log(1+u^2) \\
+    &\qquad + (4-p_{13})\log(1+u) \\
+    &\qquad + (2p_{1}+2p_{2}+2p_{3}+2p_{123}+3p_{13}+4p_{12}+4p_{23})\log(1-u). \\
+\end{align*}
+
 % das: there might be something with 1+y^2 < 1+y here?

### 2017-10-16 joint_inf.tex
Cleaning out addressed notes.

@@
 Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
-%EM Not sure what's being suppressed here. Sounds naughty.
-%das: I meant suppressing terms not involving $w'$, but I guess I already did that...
 Simplifying, the likelihood can be written
@@
 $$
-%EM I don't think you mean "any inequality"... what about sin(ell') > 0?
 The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w') \lessgtr 0$, we equivalently consider $f(w') \lessgtr 0$ where $f$ is the quadratic function
@@
 \end{align*}
-%EM Is this obvious at this point? Couldn't b^2 - 4ac be zero in principle?
-%das: only if x or y is zero (though one could be nonzero).
-%b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero.
-%c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
-%Added clarification.
+% Note from das: b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero.
+c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
 Excepting all the cases where $x=0$ or where $y=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
@@
 Since we are interested in the region where $\hat{w}'=1$, we choose either the term $(1-x')^2(1+w')(1+y')^2$ or $(1+x')^2(1+w')(1-y')^2$ in Table~\ref{tab:likelihoods} for $\siteSplit_j=\{1,3\}$.
-%EM I don't think I follow the thread of argument here-- what's being assumed when? I think you are assuming $\hat{y}' > \hat{x}'$ ?
 Now, assume that, having estimated $\hat{x}'$ and $\hat{y}'$, $\hat{y}' > \hat{x}'$.

### 2017-10-23 joint_inf_bl_analytic.py
forgot a scaling factor!

@@ -4,8 +4,8 @@ import matplotlib.pyplot as plt
 from numpy import sqrt, log, meshgrid, arange, maximum
 
 def get_what(x, y):
-    alpha_1 = 1+x**2+y**2+4*x*y**2+x**2*y**2
-    alpha_2 = 1+x**2+y**2-4*x*y**2+x**2*y**2
+    alpha_1 = 1./8 * (1+x**2+y**2+4*x*y**2+x**2*y**2)
+    alpha_2 = 1./8 * (1+x**2+y**2-4*x*y**2+x**2*y**2)
     beta =    1+x**2+y**2+x**2*y**2
     gamma = 4*x*y
 

### 2017-10-30 joint_inf_bl_analytic.py
clean up wd and consolidate plotting code

@@ -1,35 +1,0 @@
-# Branch length inconsistency
-
-import matplotlib.pyplot as plt
-from numpy import sqrt, log, meshgrid, arange, maximum
-
-def get_what(x, y):
-    alpha_1 = 1./8 * (1+x**2+y**2+4*x*y**2+x**2*y**2)
-    alpha_2 = 1./8 * (1+x**2+y**2-4*x*y**2+x**2*y**2)
-    beta =    1+x**2+y**2+x**2*y**2
-    gamma = 4*x*y
-
-    a = -gamma**2*alpha_1 - gamma**2*alpha_2 - gamma**2
-    b = gamma*alpha_1*beta - gamma**2*alpha_1 - gamma*alpha_2*beta - gamma**2*alpha_2
-    c = gamma*alpha_1*beta - gamma*alpha_2*beta + beta**2
-
-    what_1 = (-b + sqrt(b**2 - 4*a*c)) / (2*a)
-    what_2 = (-b - sqrt(b**2 - 4*a*c)) / (2*a)
-
-    return (what_2 >= 1)
-
-# Want cases where what_2 is greater than one for x and y between 0 and 1
-
-delta = 0.001
-x_range = arange(0.0, 1.0, delta)
-y_range = arange(0.0, 1.0, delta)
-X, Y = meshgrid(x_range,y_range)
-
-region = get_what(X, Y)
-
-plt.contour(X, Y, region, [0])
-plt.xlabel(r'x', fontsize=16)
-plt.ylabel(r'y', fontsize=16)
-plt.title(r'Branch length inconsistency')
-plt.savefig('figures/branch-length-inconsistency.svg')
-

### 2017-10-30 joint_inf_plot.py

@@ -1,5 +1,14 @@
 import matplotlib.pyplot as plt
-from numpy import log, meshgrid, arange, maximum
+import numpy as np
+import argparse
+import sys
+
+from operator import mul
+from multiprocessing import Pool
+from scipy.optimize import differential_evolution
+
+# ~~~~~~~~~
+# global variables
 
 ALL_PATTERNS = [
         '0000',
@@ -12,17 +21,6 @@ ALL_PATTERNS = [
         '0110'
 ]
 
-PATTERN_TO_SPLIT_DICT = {
-        '0000': '0',
-        '1000': '1',
-        '0100': '2',
-        '0010': '3',
-        '1110': '123',
-        '1100': '12',
-        '1010': '13',
-        '0110': '23'
-}
-
 SPLIT_TO_PATTERN_DICT = {
         '0':  '0000',
         '1':  '1000',
@@ -33,6 +31,52 @@ SPLIT_TO_PATTERN_DICT = {
         '13': '1010',
         '23': '0110'
 }
+
+# ~~~~~~~~~
+
+def parse_args():
+    parser = argparse.ArgumentParser(description=__doc__)
+
+    parser.add_argument(
+        '--plot-name',
+        type=str,
+        help='file name for where to save plot',
+        default='temporary-plot.svg',
+    )
+    parser.add_argument(
+        '--delta',
+        type=float,
+        help='coarseness parameter for plotting',
+        default=0.01,
+    )
+    parser.add_argument(
+        '--n-jobs',
+        type=int,
+        help='number of processes to use for empirically computing maxima',
+        default=1,
+    )
+    parser.add_argument(
+        '--topology',
+        action="store_true",
+        help='topology plot',
+    )
+    parser.add_argument(
+        '--restricted-branch-lengths',
+        action="store_true",
+        help='restricted branch lengths plot',
+    )
+    parser.add_argument(
+        '--general-branch-lengths',
+        action="store_true",
+        help='general branch lengths plot',
+    )
+
+    args = parser.parse_args()
+
+    return args
+
+# ~~~~~~~~~
+# global functions for exact likelihood computation
 
[…]

### 2017-10-30 joint_inf_plot_bl.py

@@ -1,50 +1,0 @@
-import matplotlib
-matplotlib.use('pdf')
-import matplotlib.pylab as plt
-from numpy import log, arange, vectorize
-from scipy.optimize import basinhopping
-
-def p13(x, y, w):
-    return 0.125 * (1 + x*x + y*y + x*x*y*y - 4*x*y*w)
-
-def p0(x, y, w):
-    return 0.125 * (1 + x*x + y*y + x*x*y*y + 4*x*y*w)
-
-def likelihood(x, y, w):
-    if (1 - x) * (1 - x) * (1 + w) > (1 + x) * (1 + x) * (1 - w) or \
-        (1 - y) * (1 - y) * (1 + w) > (1 + y) * (1 + y) * (1 - w):
-        return p0(x, y, y) * log(p0(x, y, w)) + p13(x, y, y) * log(p13(x, y, w)) + \
-            log(1 + w)
-    else:
-        return p0(x, y, y) * log(p0(x, y, w)) + p13(x, y, y) * log(p13(x, y, w)) + \
-            p13(x, y, y) * log(1 - w) + (1 - p13(x, y, y)) * log(1 + w)
-
-def max_likelihood(x, y):
-    """
-    basin hopping to find maximum likelihood estimate of interior branch length
-    """
-    bnds = [(0,1)]
-    minimizer_kwargs = dict(method="L-BFGS-B", bounds=bnds)
-    what = basinhopping(lambda w: -likelihood(x, y, w),
-        x0=[0.25],
-        minimizer_kwargs=minimizer_kwargs)
-    return(what.x)
-
-delta = 0.01
-x_range = arange(0.0, 1.0, delta)
-y_range = arange(0.0, 1.0, delta)
-X, Y = plt.meshgrid(x_range,y_range)
-
-Z = vectorize(lambda x, y: max_likelihood(x, y))(X, Y)
-im = plt.imshow(Z, cmap=plt.cm.gray, origin='lower')
-plt.colorbar()
-plt.title('Value of $\hat{w}$')
-plt.xlabel('$x$')
-plt.ylabel('$y$')
-ax = plt.gca()
-ax.set_xticks(arange(0, 1/delta, .2/delta))
-ax.set_yticks(arange(0, 1/delta, .2/delta))
-ax.set_xticklabels(arange(0, 1, .2))
-ax.set_yticklabels(arange(0, 1, .2))
-plt.savefig('figures/w_hat_heatmap.svg')
-

### 2017-10-30 joint_inf.tex
forgot a scaling factor!
add new optimization plot

--- 2017-10-16 joint_inf.tex
@@
 \centering
+% obtained by running
+% python joint_inf_plot.py --topology --delta .001 --plot-name figures/topology-inconsistency.svg
 \includegraphics[width=.9\textwidth]{topology-inconsistency-legend}
@@
 $$
-\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
-$$
-$$
-\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
+\alpha_1 := \alpha_1(x, y) = \frac{1}{8} (1+x^2+y^2+4xy^2+x^2y^2),
+$$
+$$
+\alpha_2 := \alpha_2(x, y) = \frac{1}{8}(1+x^2+y^2-4xy^2+x^2y^2),
 $$
@@
 \centering
+% obtained by running
+% python joint_inf_plot.py --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
 \includegraphics[width=.9\textwidth]{branch-length-inconsistency-legend}
@@
 \subsection{General case}
+
+\subsubsection{Restricted general limiting case}
+
+Let $x=y=u$ and $x'=y'=w'=u'$, a ``restricted'' Farris topology.
+We will show that, as $u\rightarrow 0$, $\hat{u'}\rightarrow u_0 > 0$, which is less than ideal.
+Our first mode of approximation is to let $u'^p\approx 0$ for all $p > 1$, preserving only linear terms.
+After some algebra, we show the likelihood is equal to
+\begin{align*}
+    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
+    &=        p_{\emptyset}  \cdot\log(1+2u'^2+4u'^3+u'^4) \\
+    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log(1-u'^4) \\
+    &\qquad + (p_{12}+p_{23})         \cdot\log(1-2u'^2+u'^4) \\
+    &\qquad + p_{13}         \cdot\log(1+2u'^2-4u'^3+u'^4) \\
+    &\qquad - \log(8) \\
+    &\qquad + p_{\emptyset}  \cdot\log((1+u')^5)\\
+    &\qquad + p_{1}          \cdot\log((1+u')^4(1-u'))\\
+    &\qquad + p_{2}          \cdot\log((1+u')^4(1-u'))\\
+    &\qquad + p_{3}          \cdot\log((1+u')^4(1-u'))\\
+    &\qquad + p_{12}         \cdot\log((1+u')^3(1-u')^2)\\
+    &\qquad + p_{13}         \cdot\log((1+u')^4(1-u'))\\
+    &\qquad + p_{23}         \cdot\log((1+u')^3(1-u')^2)\\
+    &\qquad + p_{123}        \cdot\log((1+u')^4(1-u')).
+\end{align*}
+Expanding and simplifying,
+\begin{align*}
+    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
+    &=        p_{\emptyset}  \cdot\log((1+u')(1-u'+3u'^2+u'^3))\\
+    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log(1+u'^2) \\
+    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log((1+u')(1-u')) \\
+    &\qquad + 2(p_{12}+p_{23})         \cdot\log((1+u')(1-u')) \\
+    &\qquad   p_{13}  \cdot\log((1-u')(1+u'+3u'^2-u'^3))\\
+    &\qquad - \log(8) \\
+    &\qquad + 5\cdot\log(1+u')\\
+    &\qquad - (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1+u') \\
+    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1-u').
+\end{align*}
+Now, applying $u'^p\rightarrow 0$ for $p > 1$, this approaches
+\begin{align*}
+    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
+    &\rightarrow        p_{\emptyset}  \cdot\log((1+u')(1-u'))\\
+    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log((1+u')(1-u')) \\
+    &\qquad + 2(p_{12}+p_{23})         \cdot\log((1+u')(1-u')) \\
+    &\qquad   p_{13}  \cdot\log((1-u')(1+u'))\\
+    &\qquad - \log(8) \\
+    &\qquad + 5\cdot\log(1+u')\\
+    &\qquad - (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1+u') \\
+    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1-u'),
+\end{align*}
+and
+\begin{align*}
+    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
+    &\rightarrow (1+p_{12}+p_{23})\cdot\log((1+u')(1-u'))\\
+    &\qquad - \log(8) \\
+    &\qquad + 5\cdot\log(1+u')\\
+    &\qquad - (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1+u') \\
+    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1-u').
+\end{align*}
+This is of the form
+$$
+\alpha\log(1+u') + \beta\log(1-u')
+$$
+and maximized at $\hat{u}' = (\alpha - \beta) / (\alpha + \beta)$ provided $\hat{u}' \in [0,1]$.
+As $u\rightarrow 0$, $\alpha\rightarrow 21/4$ and $\beta\rightarrow 9/4$ so $\hat{u}'\rightarrow 2/5.$
+I don't know if this is interesting yet.
+
+\subsubsection{General general case}
 
@@
 
-When computing the global maximum for $0 \le x',y',w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}'$ in Fig.~\ref{fig:w_hat_joint}.
-The region in which $\hat{w}'$ is estimated to be one---indicating no change along that particular branch---is now much smaller, but still nontrivial.
-Additionally, the estimates are surprisingly unstable, with estimates of 1 for $\hat{w}'$ scattered throughout the $(x,y)$ lattice.
-Even given the correct Farris topology we have a region of the space of true parameters where we will always estimate the interior branch parameter to be one.
+When seeking a global maximum for $0 \le x',y',w' \le 1$, we use the method of differential evolution \cite{Storn1997}---we obtain the fit $\hat{w}'$ in Fig.~\ref{fig:w_hat_joint}.
+The estimates for $\hat{w}'$ when $x$ and $y$ are large seem to be unbiased, with $\hat{w}'$ increasing as $y$ increases, though the region where $\hat{w}'$ is estimated to be one---indicating no change along that particular branch---is similar to that in the restricted case.
+Although the restricted and general case are similar, comparing Figures~\ref{fig:bl-inconsistency} and~\ref{fig:w_hat_joint} shows that the general case has a smaller region where $\hat{w}'$ is one, as expected.
+Nevertheless, this region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have a region of the space of true parameters where we will always estimate the interior branch parameter to be one.
 
@@
 \centering
-\includegraphics[width=.9\textwidth]{w_hat_heatmap_joint}
+% obtained by running
+% python joint_inf_plot.py --general-branch-lengths --delta .001 --plot-name figures/w-hat-heatmap-joint-diff-evol.svg --n-jobs 10
+\includegraphics[width=.9\textwidth]{w-hat-heatmap-joint-diff-evol}
 \caption{

### 2017-10-30 joint_inf.tex
addressing some comments

@@
 
-\section{Inconsistency of joint maximum likelihood}
+\section{Inconsistency of joint maximum likelihood in estimating topology}
 
@@
 
-\section{Branch lengths}
-
-We will now consider the problem of branch length estimation on the correct tree using joint ancestral state and branch length estimation.
-We will consider two settings on the Farris topology, corresponding to whether some branch lengths are known ahead of time or not.
+\section{Inconsistency of joint maximum likelihood in estimating branch fidelities}
+
+We will now consider the problem of branch fidelity estimation on the correct tree using joint ancestral state and branch fidelity estimation.
+We will consider two settings on the Farris topology, corresponding to whether some branch fidelities are known ahead of time or not.
 As above, we will assume that data is generated with two sister branches of fidelity $x$, with all other branches of length $y$.
+We will show that, for a nontrivial subset of possible values for $x$ and $y$, the interior branch fidelity parameter $w'$ will be consistently overestimated as exactly equal to one instead of its true value of $y$.
+That is, for the restricted case, with
+$$
+\hat{w}' = \arg\max_{w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x,y,w'\}; \fullAncestralSplitPartitions(\tau_1,\{x,y,w'\})),
+$$
+we show there are values for $x$ and $y$ where $\hat{w}'\equiv 1$.
+In the general case, we do the same but with
+$$
+(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x',y',w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})).
+$$
 %EM We will show XXX... and introduce \hat{w'} here.
-
-\subsection{Fixed $x'$ and $y'$}
+%das: updated
+
+\subsection{Restricted case: fixed $x'$ and $y'$}
 
@@
 %EM You are talking about a log likelihood here, right? So a likelihood being proportional to means that there is some additive constant. Makes things a little confusing.
-If $x'$ and $y'$ are fixed, by collapsing like terms and ignoring terms with only $x'$ and $y'$, the likelihood in \eqref{eq:farris_likelihood} as a function of $w'$ is proportional to
+%das: updated
+If $x'$ and $y'$ are fixed, by collapsing like terms and ignoring terms with only $x'$ and $y'$, the log-likelihood in \eqref{eq:farris_likelihood} as a function of $w'$ is
 \begin{align*}
     \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
-    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
+    &\sim     p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
     &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
     &\qquad + (1 - p_{13})\cdot\log(1+w') \\
-    &\qquad + p_{13}\cdot\log(1-w').
-\end{align*}
-Similarly, the other two possibilities for the likelihood are both proportional to
+    &\qquad + p_{13}\cdot\log(1-w')
+\end{align*}
+where $\sim$ denotes equality up to an additive constant.
+Similarly, the other two possibilities for the likelihood both satisfy
 \begin{align*}
     \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
-    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
+    &\sim     p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
     &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
@@
     \ell_{\tau_1,\{x,y,y\}}(\tau_1, u; \fullAncestralSplitPartitions(\tau_1,u))
-    &\propto   p_{\emptyset}  \cdot\log(1+6u^2+u^4) \\
+    &\sim     p_{\emptyset}  \cdot\log(1+6u^2+u^4) \\
     &\qquad + p_{13}         \cdot\log(1-2u^2+u^4) \\
@@
     \ell_{\tau_1,\{x,y,y\}}(\tau_1, u; \fullAncestralSplitPartitions(\tau_1,u))
-    &\propto   p_{\emptyset}  \cdot\log(1+6u^2+u^4) \\
+    &\sim     p_{\emptyset}  \cdot\log(1+6u^2+u^4) \\
     &\qquad + (p_{13} + p_{1} + p_{2} + p_{3} + p_{123})          \cdot\log(1+u^2) \\

### 2017-11-06 joint_inf.tex
clean out note: p.198 2nd para seems to imply we don't need these
sort of loose bound general case---possibly interesting?

@@
 This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
-% In Semple/Steel, there are some assumptions that are necessary for some of the results, called M1, M2 and M3.
-% M1 and M2 seem definitely necessary, which pretty much just imply x and y can't be zero.
-% M3 implies x and y can't be one?
-% (das TODO: double check these; currently working from notes and memory)
 Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.
@@
 
+\subsubsection{General case: loose bounds}
+
+In the general case, $\hat{w}'$ is a function of $x$, $y$, $\hat{x}'$, and $\hat{y}'$.
+Looking to the previous section, $\hat{w}'$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}'$ and $\hat{y}'$ instead of $x'$ and $y'$.
+Now, suppressing function arguments and given that $a < 0$, we are interested in the cases where
+$$
+\frac{-b - \sqrt{b^2 - 4ac}}{2a} \ge 1,
+$$
+$$
+-b - \sqrt{b^2 - 4ac} \le 2a,
+$$
+$$
+-b - 2a \le \sqrt{b^2 - 4ac},
+$$
+$$
+b^2 + 4ab + 4a^2 \le b^2 - 4ac,
+$$
+$$
+a^2 + ab + ac \le 0,
+$$
+$$
+a + b + c \ge 0.
+$$
+Simplifying, this is
+$$
+-\gamma^2(x', y')(1 + \frac{1}{2}\beta(x, y)) + 2\gamma(x', y')\beta(x', y')xy^2 + \beta^2(x', y') \ge 0.
+$$
+If we can find
+$$
+\gamma_{L}(x, y) \le \gamma(x', y') \le \gamma_{U}(x, y),
+$$
+$$
+\beta_{L}(x, y) \le \beta(x', y'),
+$$
+then we are interested in
+$$
+-\gamma_{U}^2(x, y)(1 + \frac{1}{2}\beta(x, y)) + 2\gamma_{L}(x, y)\beta_{L}(x, y)xy^2 + \beta_{L}^2(x, y) \ge 0,
+$$
+an inequality of two variables.
+The loosest possible bounds are $\gamma_{L} = 0$, $\beta_{L} = 1$ and $\gamma_{U} = 4$ which gives
+$$
+-16(1 + \frac{1}{2}\beta(x, y)) + 1 \ge 0,
+$$
+which is never true.
+The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma(x, y)$ and $\beta_{L} = \beta(x, y)$, which gives Figure~\ref{fig:bl-inconsistency}.
+
+Let's say you are willing to tolerate a specified level of error in your estimates for $\hat{x}'$ and $\hat{y}'$.
+If we're generous, let's call this error $.1$ in either direction; this would be a substantial deviation from the truth in many applications.
+In this case, we have the left side of Figure~\ref{fig:bl-loose-inconsistency} where we either estimate $\hat{w}'$ to be one or we poorly estimate $\hat{x}'$ or $\hat{y}'$.
+
+\begin{figure}
+\centering
+% obtained by running
+% python joint_inf_plot.py --general-branch-lengths --analytic --delta .001 --plot-name figures/bl-loose-inconsistency.svg
+\includegraphics[width=.9\textwidth]{bl-loose-inconsistency}
+\caption{Region of inconsistency with $x-.1 \le \hat{x}' \le x+.1$ and $y-.1 \le \hat{y}' \le y+.1$.}
+\label{fig:bl-loose-inconsistency}
+\end{figure}
+
 \subsubsection{Restricted general limiting case}

### 2017-11-06 joint_inf_plot.py

@@ -2,6 +2,7 @@ import matplotlib.pyplot as plt
 import numpy as np
 import argparse
 import sys
+import pickle
 
 from operator import mul
 from multiprocessing import Pool
@@ -56,6 +57,13 @@ def parse_args():
         default=1,
     )
     parser.add_argument(
+        '--param-index',
+        type=int,
+        default=2,
+        choices=(0, 1, 2),
+        help='which parameter to optimize; 0: x, 1: y, 2: w',
+    )
+    parser.add_argument(
         '--topology',
         action="store_true",
         help='topology plot',
@@ -69,6 +77,21 @@ def parse_args():
         '--general-branch-lengths',
         action="store_true",
         help='general branch lengths plot',
+    )
+    parser.add_argument(
+        '--general-likelihood',
+        action="store_true",
+        help='general likelihood pickle',
+    )
+    parser.add_argument(
+        '--analytic',
+        action="store_true",
+        help='plot analytic',
+    )
+    parser.add_argument(
+        '--empirical',
+        action="store_true",
+        help='compute empirical',
     )
 
     args = parser.parse_args()
@@ -261,9 +284,28 @@ def what_lower_minus_one(x, y):
 
     what_lower = (-b - np.sqrt(b**2 - 4*a*c)) / (2*a)
 
-    return (what_lower - 1)
+    # a + b + c = -gamma**2 * (1+.5*beta) + 2*gamma*beta*x*y**2 + beta**2
+    # for general case, gamma are fns of (x', y') and 2nd and 3rd betas are too
+    #return -gamma**2 * (1+.5*beta) + 2*gamma*beta*x*y**2 + beta**2
 
-def max_likelihood(xy):
+    #return (what_lower - 1)
+    #return -gamma**2 * (1+.5*beta) + 2*gamma*beta*x*y**2 + beta**2
+    return (a + b + c)
+
+def what_lower_general(x, y):
+    beta = 1+x**2+y**2+x**2*y**2
+    gamma = 4*x*y
+    x_l = x-.1
+    x_u = x+.1
+    y_l = y-.1
+    y_u = y+.1
+    gamma_l = 4*x_l*y_l
+    gamma_u = 4*x_u*y_u
+    beta_l = 1+x_l**2+y_l**2+x_l**2*y_l**2
+
+    return -gamma_u**2 * (1+.5*beta) + 2*gamma_l*beta_l*x*y**2 + beta_l**2
+
+def max_likelihood(xy, param_index):
     """
[…]

### 2017-11-07 deprecated.tex
update argument and remove dijoint ideas

@@
 \subsection{Branch lengths with $x$ and $y$ unknown}
+
+\subsubsection{Restricted general limiting case}
+
+Let $x=y=u$ and $x'=y'=w'=u'$, a ``restricted'' Farris topology.
+We will show that, as $u\rightarrow 0$, $\hat{u'}\rightarrow u_0 > 0$, which is less than ideal.
+Our first mode of approximation is to let $u'^p\approx 0$ for all $p > 1$, preserving only linear terms.
+After some algebra, we show the likelihood is equal to
+\begin{align*}
+    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
+    &=        p_{\emptyset}  \cdot\log(1+2u'^2+4u'^3+u'^4) \\
+    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log(1-u'^4) \\
+    &\qquad + (p_{12}+p_{23})         \cdot\log(1-2u'^2+u'^4) \\
+    &\qquad + p_{13}         \cdot\log(1+2u'^2-4u'^3+u'^4) \\
+    &\qquad - \log(8) \\
+    &\qquad + p_{\emptyset}  \cdot\log((1+u')^5)\\
+    &\qquad + p_{1}          \cdot\log((1+u')^4(1-u'))\\
+    &\qquad + p_{2}          \cdot\log((1+u')^4(1-u'))\\
+    &\qquad + p_{3}          \cdot\log((1+u')^4(1-u'))\\
+    &\qquad + p_{12}         \cdot\log((1+u')^3(1-u')^2)\\
+    &\qquad + p_{13}         \cdot\log((1+u')^4(1-u'))\\
+    &\qquad + p_{23}         \cdot\log((1+u')^3(1-u')^2)\\
+    &\qquad + p_{123}        \cdot\log((1+u')^4(1-u')).
+\end{align*}
+Expanding and simplifying,
+\begin{align*}
+    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
+    &=        p_{\emptyset}  \cdot\log((1+u')(1-u'+3u'^2+u'^3))\\
+    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log(1+u'^2) \\
+    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log((1+u')(1-u')) \\
+    &\qquad + 2(p_{12}+p_{23})         \cdot\log((1+u')(1-u')) \\
+    &\qquad   p_{13}  \cdot\log((1-u')(1+u'+3u'^2-u'^3))\\
+    &\qquad - \log(8) \\
+    &\qquad + 5\cdot\log(1+u')\\
+    &\qquad - (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1+u') \\
+    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1-u').
+\end{align*}
+Now, applying $u'^p\rightarrow 0$ for $p > 1$, this approaches
+\begin{align*}
+    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
+    &\rightarrow        p_{\emptyset}  \cdot\log((1+u')(1-u'))\\
+    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log((1+u')(1-u')) \\
+    &\qquad + 2(p_{12}+p_{23})         \cdot\log((1+u')(1-u')) \\
+    &\qquad   p_{13}  \cdot\log((1-u')(1+u'))\\
+    &\qquad - \log(8) \\
+    &\qquad + 5\cdot\log(1+u')\\
+    &\qquad - (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1+u') \\
+    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1-u'),
+\end{align*}
+and
+\begin{align*}
+    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
+    &\rightarrow (1+p_{12}+p_{23})\cdot\log((1+u')(1-u'))\\
+    &\qquad - \log(8) \\
+    &\qquad + 5\cdot\log(1+u')\\
+    &\qquad - (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1+u') \\
+    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1-u').
+\end{align*}
+This is of the form
+$$
+\alpha\log(1+u') + \beta\log(1-u')
+$$
+and maximized at $\hat{u}' = (\alpha - \beta) / (\alpha + \beta)$ provided $\hat{u}' \in [0,1]$.
+As $u\rightarrow 0$, $\alpha\rightarrow 21/4$ and $\beta\rightarrow 9/4$ so $\hat{u}'\rightarrow 2/5.$
+I don't know if this is interesting yet.
+
+\subsubsection{General general case}
+
+We consider the more standard case where the true values of $x$ and $y$ are not known.
+The form of the likelihood will alter the fitted values of $x$ and $y$ so that even with perfect information we will not necessarily estimate $\hat{x}'=x$ and $\hat{y}'=y$.
+
+Our plan of attack is to determine inequalities at which a particular form of likelihood is necessary and plot the regions of $x$ and $y$ where these occur in addition to the region where $\hat{w}'$ is estimated to be one.
+Since we are interested in the region where $\hat{w}'=1$, we choose either the term $(1-x')^2(1+w')(1+y')^2$ or $(1+x')^2(1+w')(1-y')^2$ in Table~\ref{tab:likelihoods} for $\siteSplit_j=\{1,3\}$.
+Now, assume that, having estimated $\hat{x}'$ and $\hat{y}'$, $\hat{y}' > \hat{x}'$.
+In this case,
+$$
+(1-\hat{x}'^2)(1+\hat{y}'^2) > (1+\hat{x}'^2)(1-\hat{y}'^2),
+$$
+so we use the likelihood
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
+    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
+    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
+    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
+    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
+    &\qquad - \log(8) \\
+    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
+    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
+    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
+    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
+    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
+    &\qquad + p_{13}         \cdot\log((1-x')^2   (1+w')(1+y')^2) \\
+    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
+    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
+\end{align*}
+This simplifies to
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + (p_{1} + p_{3})          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
+    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
+    &\qquad + (p_{12} + p_{23})         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
+    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
+    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
+    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
+    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
+    &\qquad + \log(1+w').
+\end{align*}
+We need to obtain
+$$
+(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x, y, w} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+$$
+as functions of $x$ and $y$ and verify the region where $\hat{w}'(x, y) = 1$ and $\hat{y}'(x, y) > \hat{x}'(x, y)$.
+This region will be consistent with one always estimating $\hat{w}'$ to be one.
+This can similarly be done for the other likelihood term with $\hat{x}'(x, y) > \hat{y}'(x, y)$.
+To do is to find these functions---or subsets of their varieties---in as closed a form as possible.
+
+As a first attempt, simplify further to
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2) \\
+    &\qquad + (p_{1} + p_{3} + p_{12} + p_{23})          \cdot\log(1-x'^2) \\
+    &\qquad + (p_{1} + p_{3})          \cdot\log(1+y'^2) \\
+    &\qquad + (p_{2} + p_{123} + p_{12} + p_{23})          \cdot\log(1-y'^2) \\
+    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
+    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
+    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
+    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
+    &\qquad + \log(1+w')
+\end{align*}
+and
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2) \\
+    &\qquad + (p_{1} + p_{3})          \cdot\log(1+y'^2) \\
+    &\qquad + 2(1-p_{13})\log(1+x') \\
+    &\qquad + 2(p_{1}+p_{3}+p_{12}+p_{23}+p_{13})\log(1-x') \\
+    &\qquad + 2\log(1+y') \\
+    &\qquad + 2(p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
+    &\qquad + \log(1+w').
+\end{align*}
+Say we had estimated $\hat{w}'=1$.
+Then we need is the region where $\hat{y}'>\hat{x}'$ and $0<\hat{x}',\hat{y}'<1$---the case for $\hat{x}'>\hat{y}'$ will be a reflection of this case.
+Let's consider first $x'=y'$; if we can show that for $x,y<\epsilon$ where it looks like $\epsilon\approx 0.4$ then $\hat{w}'=1$ we are at least getting close to something.
+So fix $x'=y'=u$ and get
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, u; \fullAncestralSplitPartitions(\tau_1,u))
+    &\sim     p_{\emptyset}  \cdot\log(1+6u^2+u^4) \\
+    &\qquad + p_{13}         \cdot\log(1-2u^2+u^4) \\
+    &\qquad + (p_{1} + p_{2} + p_{3} + p_{123})          \cdot\log(1+u^2) \\
+    &\qquad + (4-2p_{13})\log(1+u) \\
+    &\qquad + 2(p_{1}+p_{2}+p_{3}+p_{123}+p_{13}+2p_{12}+2p_{23})\log(1-u) \\
+\end{align*}
+which is
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, u; \fullAncestralSplitPartitions(\tau_1,u))
+    &\sim     p_{\emptyset}  \cdot\log(1+6u^2+u^4) \\
+    &\qquad + (p_{13} + p_{1} + p_{2} + p_{3} + p_{123})          \cdot\log(1+u^2) \\
+    &\qquad + (4-p_{13})\log(1+u) \\
+    &\qquad + (2p_{1}+2p_{2}+2p_{3}+2p_{123}+3p_{13}+4p_{12}+4p_{23})\log(1-u). \\
+\end{align*}
+
+% das: there might be something with 1+y^2 < 1+y here?
+% I'm not sure how specifically it would affect the maximization, and it might be too loose of a bound, but could work to try.
+% The first two terms kind of gunk up the works a little.
+Say we had estimated $\hat{x}'$ and $\hat{y}'$.
+We can use a similar argument as in the restricted case, though now
+$$
+\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
+$$
+$$
+\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
+$$
+$$
+\beta := \beta(x', y') = 1+x'^2+y'^2+x'^2y'^2,
+$$
+$$
+\gamma := \gamma(x', y') = 4x'y'.
+$$
+So instead of the lower bound for $\hat{w}'$ being a function of $x$ and $y$ through $\alpha_1,\alpha_2,\beta$ and $\gamma$, it is a function of $x$ and $y$ through $\beta$ and $\gamma$ via $\hat{x}'$ and $\hat{y}'$.
+Operating on the assumption $\hat{y}' > \hat{x}'$, which we needed to specify the form of the likelihood, we now see
+$$
+1 < 1 + 2x'^2 + x'^4 < \beta < 1+2y'^2+y'^4 < 4
+$$
+and
+$$
+0 < 4x'^2 < \gamma < 4y'^2 < 4.
+$$
+We may be able to use this to restrict our attention to fewer variables, but we need to determine how $a,b$ and $c$ are varying with $\beta$ and $\gamma$ to see which inequalities are useful.
+Recall
+$$
+\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)}
+$$
+where
+$$
+a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
+$$
+$$
+b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
+$$
+$$
+c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2.
+$$
+We know for sure $a$ is decreasing with $\gamma$ and $c$ is increasing with $\gamma$, while $c$ is increasing with $\beta$.
 

### 2017-11-07 joint_inf.tex
update argument and remove dijoint ideas

--- 2017-10-30 joint_inf.tex
@@
 
-\subsubsection{Restricted general limiting case}
-
-Let $x=y=u$ and $x'=y'=w'=u'$, a ``restricted'' Farris topology.
-We will show that, as $u\rightarrow 0$, $\hat{u'}\rightarrow u_0 > 0$, which is less than ideal.
-Our first mode of approximation is to let $u'^p\approx 0$ for all $p > 1$, preserving only linear terms.
-After some algebra, we show the likelihood is equal to
-\begin{align*}
-    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
-    &=        p_{\emptyset}  \cdot\log(1+2u'^2+4u'^3+u'^4) \\
-    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log(1-u'^4) \\
-    &\qquad + (p_{12}+p_{23})         \cdot\log(1-2u'^2+u'^4) \\
-    &\qquad + p_{13}         \cdot\log(1+2u'^2-4u'^3+u'^4) \\
-    &\qquad - \log(8) \\
-    &\qquad + p_{\emptyset}  \cdot\log((1+u')^5)\\
-    &\qquad + p_{1}          \cdot\log((1+u')^4(1-u'))\\
-    &\qquad + p_{2}          \cdot\log((1+u')^4(1-u'))\\
-    &\qquad + p_{3}          \cdot\log((1+u')^4(1-u'))\\
-    &\qquad + p_{12}         \cdot\log((1+u')^3(1-u')^2)\\
-    &\qquad + p_{13}         \cdot\log((1+u')^4(1-u'))\\
-    &\qquad + p_{23}         \cdot\log((1+u')^3(1-u')^2)\\
-    &\qquad + p_{123}        \cdot\log((1+u')^4(1-u')).
-\end{align*}
-Expanding and simplifying,
-\begin{align*}
-    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
-    &=        p_{\emptyset}  \cdot\log((1+u')(1-u'+3u'^2+u'^3))\\
-    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log(1+u'^2) \\
-    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log((1+u')(1-u')) \\
-    &\qquad + 2(p_{12}+p_{23})         \cdot\log((1+u')(1-u')) \\
-    &\qquad   p_{13}  \cdot\log((1-u')(1+u'+3u'^2-u'^3))\\
-    &\qquad - \log(8) \\
-    &\qquad + 5\cdot\log(1+u')\\
-    &\qquad - (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1+u') \\
-    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1-u').
-\end{align*}
-Now, applying $u'^p\rightarrow 0$ for $p > 1$, this approaches
-\begin{align*}
-    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
-    &\rightarrow        p_{\emptyset}  \cdot\log((1+u')(1-u'))\\
-    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log((1+u')(1-u')) \\
-    &\qquad + 2(p_{12}+p_{23})         \cdot\log((1+u')(1-u')) \\
-    &\qquad   p_{13}  \cdot\log((1-u')(1+u'))\\
-    &\qquad - \log(8) \\
-    &\qquad + 5\cdot\log(1+u')\\
-    &\qquad - (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1+u') \\
-    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1-u'),
-\end{align*}
+\subsubsection{General case: loose bounds}
+
+In the general case, $\hat{w}'$ is a function of $x$, $y$, $\hat{x}'$, and $\hat{y}'$.
+Looking to the previous section, $\hat{w}'$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}'$ and $\hat{y}'$ instead of $x'$ and $y'$.
+Assume we know $\hat{x}'$ and $\hat{y}'$ as functions of $x$ and $y$ only.
+Suppressing function arguments and given that $a < 0$, we are interested in the cases where
+$$
+\frac{-b - \sqrt{b^2 - 4ac}}{2a} \ge 1,
+$$
+which, after some algebra, is equivalent to
+%-b - \sqrt{b^2 - 4ac} \le 2a,
+%-b - 2a \le \sqrt{b^2 - 4ac},
+%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
+%a^2 + ab + ac \le 0,
+$$
+a + b + c \ge 0.
+$$
+Simplifying, this is
+$$
+-\gamma^2(\hat{x}', \hat{y}')(1 + \frac{1}{2}\beta(x, y)) + 2\gamma(\hat{x}', \hat{y}')\beta(\hat{x}', \hat{y}')xy^2 + \beta^2(\hat{x}', \hat{y}') \ge 0.
+$$
+If we can find
+$$
+\gamma_{L}(x, y) \le \gamma(\hat{x}', \hat{y}') \le \gamma_{U}(x, y)
+$$
 and
-\begin{align*}
-    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
-    &\rightarrow (1+p_{12}+p_{23})\cdot\log((1+u')(1-u'))\\
-    &\qquad - \log(8) \\
-    &\qquad + 5\cdot\log(1+u')\\
-    &\qquad - (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1+u') \\
-    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1-u').
-\end{align*}
-This is of the form
-$$
-\alpha\log(1+u') + \beta\log(1-u')
-$$
-and maximized at $\hat{u}' = (\alpha - \beta) / (\alpha + \beta)$ provided $\hat{u}' \in [0,1]$.
-As $u\rightarrow 0$, $\alpha\rightarrow 21/4$ and $\beta\rightarrow 9/4$ so $\hat{u}'\rightarrow 2/5.$
-I don't know if this is interesting yet.
-
-\subsubsection{General general case}
-
-We consider the more standard case where the true values of $x$ and $y$ are not known.
-The form of the likelihood will alter the fitted values of $x$ and $y$ so that even with perfect information we will not necessarily estimate $\hat{x}'=x$ and $\hat{y}'=y$.
-
-Our plan of attack is to determine inequalities at which a particular form of likelihood is necessary and plot the regions of $x$ and $y$ where these occur in addition to the region where $\hat{w}'$ is estimated to be one.
-Since we are interested in the region where $\hat{w}'=1$, we choose either the term $(1-x')^2(1+w')(1+y')^2$ or $(1+x')^2(1+w')(1-y')^2$ in Table~\ref{tab:likelihoods} for $\siteSplit_j=\{1,3\}$.
-Now, assume that, having estimated $\hat{x}'$ and $\hat{y}'$, $\hat{y}' > \hat{x}'$.
-In this case,
-$$
-(1-\hat{x}'^2)(1+\hat{y}'^2) > (1+\hat{x}'^2)(1-\hat{y}'^2),
-$$
-so we use the likelihood
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
-    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
-    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
-    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
-    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
-    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
-    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
-    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
-    &\qquad - \log(8) \\
-    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
-    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
-    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
-    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
-    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
-    &\qquad + p_{13}         \cdot\log((1-x')^2   (1+w')(1+y')^2) \\
-    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
-    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
-\end{align*}
-This simplifies to
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
-    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
-    &\qquad + (p_{1} + p_{3})          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
-    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
-    &\qquad + (p_{12} + p_{23})         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
-    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
-    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
-    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
-    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
-    &\qquad + \log(1+w').
-\end{align*}
-We need to obtain
-$$
-(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x, y, w} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-$$
-as functions of $x$ and $y$ and verify the region where $\hat{w}'(x, y) = 1$ and $\hat{y}'(x, y) > \hat{x}'(x, y)$.
-This region will be consistent with one always estimating $\hat{w}'$ to be one.
-This can similarly be done for the other likelihood term with $\hat{x}'(x, y) > \hat{y}'(x, y)$.
-To do is to find these functions---or subsets of their varieties---in as closed a form as possible.
-
-As a first attempt, simplify further to
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
-    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
-    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2) \\
-    &\qquad + (p_{1} + p_{3} + p_{12} + p_{23})          \cdot\log(1-x'^2) \\
-    &\qquad + (p_{1} + p_{3})          \cdot\log(1+y'^2) \\
-    &\qquad + (p_{2} + p_{123} + p_{12} + p_{23})          \cdot\log(1-y'^2) \\
-    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
-    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
-    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
-    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
-    &\qquad + \log(1+w')
-\end{align*}
+$$
+\beta_{L}(x, y) \le \beta(\hat{x}', \hat{y}')
+$$
+then we are interested in the region where
+$$
+-\gamma_{U}^2(x, y)(1 + \frac{1}{2}\beta(x, y)) + 2\gamma_{L}(x, y)\beta_{L}(x, y)xy^2 + \beta_{L}^2(x, y) \ge 0,
+$$
+an inequality involving only $x$ and $y$.
+The loosest possible bounds are $\gamma_{L} = 0$, $\beta_{L} = 1$ and $\gamma_{U} = 4$, giving
+$$
+-16(1 + \frac{1}{2}\beta(x, y)) + 1 \ge 0,
+$$
+which is never true.
+The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma(x, y)$ and $\beta_{L} = \beta(x, y)$, which is the case in Figure~\ref{fig:bl-inconsistency}.
+
+Say we are willing to tolerate a specified level of error in our estimates for $\hat{x}'$ and $\hat{y}'$ so that
+$$
+x-x_{L} \le \hat{x}' \le x+x_{U}
+$$
 and
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
-    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
-    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2) \\
-    &\qquad + (p_{1} + p_{3})          \cdot\log(1+y'^2) \\
-    &\qquad + 2(1-p_{13})\log(1+x') \\
-    &\qquad + 2(p_{1}+p_{3}+p_{12}+p_{23}+p_{13})\log(1-x') \\
-    &\qquad + 2\log(1+y') \\
-    &\qquad + 2(p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
-    &\qquad + \log(1+w').
-\end{align*}
-Say we had estimated $\hat{w}'=1$.
-Then we need is the region where $\hat{y}'>\hat{x}'$ and $0<\hat{x}',\hat{y}'<1$---the case for $\hat{x}'>\hat{y}'$ will be a reflection of this case.
-Let's consider first $x'=y'$; if we can show that for $x,y<\epsilon$ where it looks like $\epsilon\approx 0.4$ then $\hat{w}'=1$ we are at least getting close to something.
-So fix $x'=y'=u$ and get
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, u; \fullAncestralSplitPartitions(\tau_1,u))
-    &\sim     p_{\emptyset}  \cdot\log(1+6u^2+u^4) \\
-    &\qquad + p_{13}         \cdot\log(1-2u^2+u^4) \\
-    &\qquad + (p_{1} + p_{2} + p_{3} + p_{123})          \cdot\log(1+u^2) \\
-    &\qquad + (4-2p_{13})\log(1+u) \\
-    &\qquad + 2(p_{1}+p_{2}+p_{3}+p_{123}+p_{13}+2p_{12}+2p_{23})\log(1-u) \\
-\end{align*}
-which is
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, u; \fullAncestralSplitPartitions(\tau_1,u))
-    &\sim     p_{\emptyset}  \cdot\log(1+6u^2+u^4) \\
-    &\qquad + (p_{13} + p_{1} + p_{2} + p_{3} + p_{123})          \cdot\log(1+u^2) \\
-    &\qquad + (4-p_{13})\log(1+u) \\
-    &\qquad + (2p_{1}+2p_{2}+2p_{3}+2p_{123}+3p_{13}+4p_{12}+4p_{23})\log(1-u). \\
-\end{align*}
-
-% das: there might be something with 1+y^2 < 1+y here?
-% I'm not sure how specifically it would affect the maximization, and it might be too loose of a bound, but could work to try.
-% The first two terms kind of gunk up the works a little.
-Say we had estimated $\hat{x}'$ and $\hat{y}'$.
-We can use a similar argument as in the restricted case, though now
-$$
-\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
-$$
-$$
-\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
-$$
-$$
-\beta := \beta(x', y') = 1+x'^2+y'^2+x'^2y'^2,
-$$
-$$
-\gamma := \gamma(x', y') = 4x'y'.
-$$
-So instead of the lower bound for $\hat{w}'$ being a function of $x$ and $y$ through $\alpha_1,\alpha_2,\beta$ and $\gamma$, it is a function of $x$ and $y$ through $\beta$ and $\gamma$ via $\hat{x}'$ and $\hat{y}'$.
-Operating on the assumption $\hat{y}' > \hat{x}'$, which we needed to specify the form of the likelihood, we now see
-$$
-1 < 1 + 2x'^2 + x'^4 < \beta < 1+2y'^2+y'^4 < 4
-$$
-and
-$$
-0 < 4x'^2 < \gamma < 4y'^2 < 4.
-$$
-We may be able to use this to restrict our attention to fewer variables, but we need to determine how $a,b$ and $c$ are varying with $\beta$ and $\gamma$ to see which inequalities are useful.
-Recall
-$$
-\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)}
-$$
-where
-$$
-a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
-$$
-$$
-b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
-$$
-$$
-c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2.
-$$
-We know for sure $a$ is decreasing with $\gamma$ and $c$ is increasing with $\gamma$, while $c$ is increasing with $\beta$.
+$$
+y-y_{L} \le \hat{y}' \le y+y_{U}.
+$$
+As an example, let this error be $.1$ in either direction, i.e., $x_{L}=x_{U}=y_{L}=y_{U}=.1$.
+In this case, the region to the left side of the curve in Figure~\ref{fig:bl-loose-inconsistency} will either estimate $\hat{w}'$ to be one or estimate either $\hat{x}'$ or $\hat{y}'$ to be more than $.1$ away from its true value.
+
+\begin{figure}
+\centering
+% obtained by running
+% python joint_inf_plot.py --general-branch-lengths --analytic --delta .001 --plot-name figures/bl-loose-inconsistency.svg
+\includegraphics[width=.9\textwidth]{bl-loose-inconsistency}
+\caption{Region of inconsistency with $x-.1 \le \hat{x}' \le x+.1$ and $y-.1 \le \hat{y}' \le y+.1$.}
+\label{fig:bl-loose-inconsistency}
+\end{figure}
 

### 2017-11-08 joint_inf_plot.py
nicer plots with legends added

@@ -1,4 +1,11 @@
+from __future__ import unicode_literals
+import matplotlib
+matplotlib.rcParams['text.usetex'] = True
+matplotlib.rcParams['text.latex.unicode'] = True
 import matplotlib.pyplot as plt
+import matplotlib.patches as mpatches
+import seaborn as sns
+
 import numpy as np
 import argparse
 import sys
@@ -6,7 +13,10 @@ import pickle
 
 from operator import mul
 from multiprocessing import Pool
-from scipy.optimize import differential_evolution
+from scipy.optimize import differential_evolution, minimize, basinhopping
+
+sns.set_style('white')
+sns.set_style('ticks')
 
 # ~~~~~~~~~
 # global variables
@@ -79,7 +89,7 @@ def parse_args():
         help='general branch lengths plot',
     )
     parser.add_argument(
-        '--general-likelihood',
+        '--output-likelihood',
         action="store_true",
         help='general likelihood pickle',
     )
@@ -156,9 +166,15 @@ def FARRIS_LIKELIHOOD(x, y, xp, yp, wp, joint=True):
     theta0 = [x, y, x, y, y]
     theta1 = [xp, yp, xp, yp, wp]
     for site_pattern in ALL_PATTERNS:
-        likelihood += P_FARRIS(theta0, site_pattern) * \
+        if P_FARRIS(theta0, site_pattern) != 0:
+            if np.isclose(P_FARRIS(theta1, site_pattern), 0) or np.isclose(L_FARRIS(theta1, site_pattern, joint=joint), 0):
+                likelihood = -np.inf
+                break
+
+            likelihood += P_FARRIS(theta0, site_pattern) * \
                 (np.log(L_FARRIS(theta1, site_pattern, joint=joint)) + \
                 np.log(P_FARRIS(theta1, site_pattern)))
+
     return likelihood
 
 def P_FELSENSTEIN(theta0, site_pattern):
@@ -305,15 +321,43 @@ def what_lower_general(x, y):
 
     return -gamma_u**2 * (1+.5*beta) + 2*gamma_l*beta_l*x*y**2 + beta_l**2
 
+#def max_likelihood(xy, param_index):
+#    """
+#    differential evolution to find maximum likelihood estimate of interior branch length
+#    """
+#    bnds = [(0,1), (0,1), (0,1)]
+#    what = differential_evolution(
+#        lambda z: -FARRIS_LIKELIHOOD(xy[0], xy[1], z[0], z[1], z[2]),
[…]

### 2017-11-08 joint_inf_plot.py
aspect ratio

@@ -395,8 +395,8 @@ class LegendHandler(object):
     def legend_artist(self, legend, orig_handle, fontsize, handlebox):
         x0, y0 = handlebox.xdescent, handlebox.ydescent
         width, height = handlebox.width, handlebox.height
-        patch = mpatches.Rectangle([x0, y0], width, height, facecolor='k',
-                                   edgecolor='black', lw=1, alpha=.2,
+        patch = mpatches.Rectangle([x0, y0], width, height, facecolor='w',
+                                   edgecolor='black', lw=1,
                                    transform=handlebox.get_transform())
         handlebox.add_artist(patch)
         return patch
@@ -423,20 +423,19 @@ def main(args=sys.argv[1:]):
             legendtext = r'$\hat{w}=1$ or $\hat{x}$ or $\hat{y}$ poorly estimated'
 
         ct = plt.contour(X, Y, region, [0], colors='k', linewidths=1)
+        plt.axis('scaled')
         vec = ct.collections[0].get_paths()[0].vertices
-        if args.topology:
-            x = vec[:,0]
-            y = vec[:,1]
-            plt.fill_between(x, args.delta, y, alpha=.2, color="k")
+        plt.xlabel(r'$x$')
+        plt.ylabel(r'$y$')
+        ttl = plt.title(plottitle, fontsize=14)
+        ttl.set_position([.5, 1.02])
+        if args.restricted_branch_lengths:
+            plt.legend([Legend()], [legendtext],
+                handler_map={Legend: LegendHandler()},
+                bbox_to_anchor=(1.05, 1), loc=2, borderaxespad=0.)
         else:
-            x = np.append(vec[:,0], args.delta)
-            y = np.append(vec[:,1], 1-args.delta)
-            plt.fill_between(x, y, args.delta, alpha=.2, color="k")
-        plt.xlabel(r'$x$', fontsize=16)
-        plt.ylabel(r'$y$', fontsize=16)
-        plt.title(plottitle, fontsize=16)
-        plt.legend([Legend()], [legendtext],
-            handler_map={Legend: LegendHandler()})
+            plt.legend([Legend()], [legendtext],
+                handler_map={Legend: LegendHandler()})
         sns.despine()
         plt.savefig(args.plot_name)
     elif args.empirical:

### 2017-11-08 joint_inf_plot.py
hatch waves

@@ -429,6 +429,11 @@ def main(args=sys.argv[1:]):
         plt.ylabel(r'$y$')
         ttl = plt.title(plottitle, fontsize=14)
         ttl.set_position([.5, 1.02])
+        if not args.topology:
+            x = np.concatenate(([1-args.delta], vec[:,0], [args.delta]))
+            y = np.concatenate(([args.delta], vec[:,1], [1-args.delta]))
+            plt.plot(x, y, '-', alpha=.15, lw=.5)
+
         if args.restricted_branch_lengths:
             plt.legend([Legend()], [legendtext],
                 handler_map={Legend: LegendHandler()},
@@ -436,6 +441,7 @@ def main(args=sys.argv[1:]):
         else:
             plt.legend([Legend()], [legendtext],
                 handler_map={Legend: LegendHandler()})
+
         sns.despine()
         plt.savefig(args.plot_name)
     elif args.empirical:

### 2017-11-09 joint_inf_plot.py
numerical precision handling and testing restricted vs general likelihoods

@@ -154,9 +154,9 @@ def L_FARRIS(theta, site_pattern, joint=True):
         1
     )
     if joint:
-        return 0.125 * max(s1, s2, s3, s4)
+        return 0.03125 * max(s1, s2, s3, s4)
     else:
-        return 0.125 * sum([s1, s2, s3, s4])
+        return 0.03125 * sum([s1, s2, s3, s4])
 
 def FARRIS_LIKELIHOOD(x, y, xp, yp, wp, joint=True):
     """
@@ -166,7 +166,7 @@ def FARRIS_LIKELIHOOD(x, y, xp, yp, wp, joint=True):
     theta0 = [x, y, x, y, y]
     theta1 = [xp, yp, xp, yp, wp]
     for site_pattern in ALL_PATTERNS:
-        if P_FARRIS(theta0, site_pattern) != 0:
+        if not np.isclose(P_FARRIS(theta0, site_pattern), 0):
             if np.isclose(P_FARRIS(theta1, site_pattern), 0) or np.isclose(L_FARRIS(theta1, site_pattern, joint=joint), 0):
                 likelihood = -np.inf
                 break
@@ -347,18 +347,23 @@ def what_lower_general(x, y):
 #
 #    return what.x[param_index]
 
-def max_likelihood(xy, param_index):
+def max_likelihood(xy, param_index, delta):
     """
     max likelihood as function of marginal likelihood
     """
-    bnds = [(0,1)]*3
-    what = minimize(lambda z: -FARRIS_LIKELIHOOD(xy[0], xy[1], z[0], z[1], z[2]),
-        x0=[xy[0], xy[1], xy[1]],
-        method="SLSQP",
-        bounds=bnds,
-    )
 
-    return what.x[param_index]
+    what_gen = differential_evolution(lambda z: -FARRIS_LIKELIHOOD(xy[0], xy[1], z[0], z[1], z[2]),
+        bounds=[(delta,1-delta)]*3,
+    )
+    what_res = differential_evolution(lambda z: -FARRIS_LIKELIHOOD(xy[0], xy[1], z[0], z[1], 1.-delta),
+        bounds=[(delta,1-delta)]*2,
+    )
+    if -what_res.fun > -what_gen.fun:
+        what = 1.-delta
+    else:
+        what = what_gen.x[param_index]
+
+    return what
 
 def max_likelihood_all(xy):
     """
@@ -383,10 +388,11 @@ def max_likelihood_all_w(xy):
     return what
[…]

### 2017-11-09 joint_inf_plot.py
remove comments; add pickle because I messed up a caption :(

@@ -55,6 +55,12 @@ def parse_args():
         default='temporary-plot.svg',
     )
     parser.add_argument(
+        '--pkl-name',
+        type=str,
+        help='file name for where to save pkl',
+        default='temporary-plot.pkl',
+    )
+    parser.add_argument(
         '--delta',
         type=float,
         help='coarseness parameter for plotting',
@@ -304,8 +310,6 @@ def what_lower_minus_one(x, y):
     # for general case, gamma are fns of (x', y') and 2nd and 3rd betas are too
     #return -gamma**2 * (1+.5*beta) + 2*gamma*beta*x*y**2 + beta**2
 
-    #return (what_lower - 1)
-    #return -gamma**2 * (1+.5*beta) + 2*gamma*beta*x*y**2 + beta**2
     return (a + b + c)
 
 def what_lower_general(x, y):
@@ -321,33 +325,7 @@ def what_lower_general(x, y):
 
     return -gamma_u**2 * (1+.5*beta) + 2*gamma_l*beta_l*x*y**2 + beta_l**2
 
-#def max_likelihood(xy, param_index):
-#    """
-#    differential evolution to find maximum likelihood estimate of interior branch length
-#    """
-#    bnds = [(0,1), (0,1), (0,1)]
-#    what = differential_evolution(
-#        lambda z: -FARRIS_LIKELIHOOD(xy[0], xy[1], z[0], z[1], z[2]),
-#        bounds=bnds
-#    )
-#
-#def max_likelihood(xy, param_index):
-#    """
-#    max likelihood as function of marginal likelihood
-#    """
-#    bnds = [(0,1)]*3
-#    minimizer_kwargs = dict(
-#        method="SLSQP",
-#        bounds=bnds,
-#    )
-#    what = basinhopping(lambda z: -FARRIS_LIKELIHOOD(xy[0], xy[1], z[0], z[1], z[2]),
-#        x0=[xy[0], xy[1], xy[1]],
-#        minimizer_kwargs=minimizer_kwargs
-#    )
-#
-#    return what.x[param_index]
-
-def max_likelihood(xy, param_index, delta):
+def max_likelihood(xy, delta):
     """
     max likelihood as function of marginal likelihood
     """
@@ -361,38 +339,15 @@ def max_likelihood(xy, param_index, delta):
     if -what_res.fun > -what_gen.fun:
         what = 1.-delta
     else:
-        what = what_gen.x[param_index]
+        what = what_gen.x[2]
 
-    return what
-
-def max_likelihood_all(xy):
[…]

### 2017-11-09 joint_inf_plot.py
add input pickle, too

@@ -55,10 +55,16 @@ def parse_args():
         default='temporary-plot.svg',
     )
     parser.add_argument(
-        '--pkl-name',
+        '--in-pkl-name',
         type=str,
         help='file name for where to save pkl',
-        default='temporary-plot.pkl',
+        default=None,
+    )
+    parser.add_argument(
+        '--out-pkl-name',
+        type=str,
+        help='file name for where to save pkl',
+        default='output.pkl',
     )
     parser.add_argument(
         '--delta',
@@ -385,12 +391,12 @@ def main(args=sys.argv[1:]):
 
         ct = plt.contour(X, Y, region, [0], colors='k', linewidths=1)
         plt.axis('scaled')
-        vec = ct.collections[0].get_paths()[0].vertices
         plt.xlabel(r'$x$')
         plt.ylabel(r'$y$')
         ttl = plt.title(plottitle, fontsize=14)
         ttl.set_position([.5, 1.02])
         if not args.topology:
+            vec = ct.collections[0].get_paths()[0].vertices
             x = np.concatenate(([1-args.delta], vec[:,0], [args.delta]))
             y = np.concatenate(([args.delta], vec[:,1], [1-args.delta]))
             plt.plot(x, y, '-', alpha=.15, lw=.5)
@@ -407,26 +413,33 @@ def main(args=sys.argv[1:]):
         plt.savefig(args.plot_name)
     elif args.empirical:
         if args.general_branch_lengths:
-            if args.n_jobs > 1:
-                p = Pool(processes=args.n_jobs)
-                Z_pool = p.map( MaxLikelihood(args.delta) , [(x, y) for x, y in zip(X.ravel(), Y.ravel())])
-                Z = np.reshape(Z_pool, (int(1. / args.delta) - 1, int(1. / args.delta) - 1))
+            if args.in_pkl_name is None:
+                if args.n_jobs > 1:
+                    p = Pool(processes=args.n_jobs)
+                    Z_pool = p.map( MaxLikelihood(args.delta) , [(x, y) for x, y in zip(X.ravel(), Y.ravel())])
+                    Z = np.reshape(Z_pool, (int(1. / args.delta) - 1, int(1. / args.delta) - 1))
+                else:
+                    Z = np.vectorize(lambda x, y: max_likelihood((x, y), delta=args.delta))(X, Y)
+
+                with open(args.out_pkl_name, 'w') as f:
+                    pickle.dump([X, Y, Z], f)
[…]

### 2017-11-13 joint_inf_plot.py
more consistent plotting

@@ -422,13 +422,12 @@ def main(args=sys.argv[1:]):
                     Z = np.vectorize(lambda x, y: max_likelihood((x, y), delta=args.delta))(X, Y)
 
                 with open(args.out_pkl_name, 'w') as f:
-                    pickle.dump([X, Y, Z], f)
+                    pickle.dump((X, Y, Z), f)
             else:
-                with open(args.in_pkl_name, 'w') as f:
+                with open(args.in_pkl_name, 'r') as f:
                     X, Y, Z = pickle.load(f)
 
             im = plt.imshow(Z, cmap=plt.cm.gray, origin='lower')
-            plt.colorbar()
             plt.xlabel(r'$x$')
             plt.ylabel(r'$y$')
             ttl = plt.title(r'Value of $\hat{w}$', fontsize=14)
@@ -436,9 +435,10 @@ def main(args=sys.argv[1:]):
             ax = plt.gca()
             ax.set_xticks(np.arange(0, 1/args.delta, .2/args.delta))
             ax.set_yticks(np.arange(0, 1/args.delta, .2/args.delta))
-            ax.set_xticklabels(np.arange(0, 1, .2))
-            ax.set_yticklabels(np.arange(0, 1, .2))
+            ax.set_xticklabels([r'$0.0$', r'$0.2$', r'$0.4$', r'$0.6$', r'$0.8$'])
+            ax.set_yticklabels([r'$0.0$', r'$0.2$', r'$0.4$', r'$0.6$', r'$0.8$'])
             sns.despine()
+            plt.colorbar()
             plt.savefig(args.plot_name)
     else:
         print "No plotting argument given!"

### 2017-11-13 joint_inf.tex
Re-outline.
update analytic plots (still no hatching)
hatch waves
update plot---will redo for y-axis
pass through until lower bound---need to reorganize

@@
 
-\title{On joint estimation of ancestral states and phylogenetic model parameters}
+\title{Joint estimation of ancestral states and phylogenetic model parameters is inconsistent}
 \author{Shaw, Matsen, Minin}
@@
 
+%EM clarify that we aren't doing sequence-level simulation somewhere.
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
 To do so, we use a binary symmetric model and imagine data being generated on the four taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the large-data limit.
+We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known.
+
+%EM Potential re-outline--
+%
+%EM 1. Describe setup with Fels and Farris trees, and how we are talking about the case of infinite data.
+%EM 2. Describe fidelities
+%EM 3. State theorems
+%EM 4. Plots
+%EM 5. Discussion, already below. This motivates a more thorough investigation under simulation.
 
@@
 
-We consider the standard phylogenetic likelihood on unrooted phylogenetic trees \cite{Felsenstein2004}.
+We consider the standard phylogenetic likelihood on unrooted trees \cite{Felsenstein2004}.
 We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
 Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
-Assume we observe $\nCols$ samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
+We observe $\nCols$ samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
 The corresponding unobserved ancestral states are $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ and distributed as $\ancestralStateColumnRV$.
@@
 
-An alternative approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
+An alternative approach does away with the marginalization and directly estimates the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
 One can compute the profile likelihood
@@
 
-Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation (we follow Section 8.6 of \cite{Semple2003-em}).
+Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation (we follow Section 8.6 of \cite{Semple2003-em}).
 Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
@@
 as a mapping of ancestral states and tip states to ancestral state splits.
-It is defined by if the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is not contained in $\siteSplitSet$ and thus we must take a complement to become a site split, we let $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ be the complement of $\ancestralStateColumn$; otherwise it is simply $\ancestralStateColumn$.
-Note $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective for each $\alignmentColumn_i$.
+It is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.
 We define
@@
 $$
-for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
-By considering the case requiring complementing and that not requiring complementing separately,
+for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, so that
+%By considering the case requiring complementing and that not requiring complementing separately,
+%das: did I write the above? I don't remember the significance of it...
 \begin{align*}
@@
 \end{align*}
-By the surjectivity of $\ancestralToSplit(\alignmentColumn_i, \cdot)$, we also have
+In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
+For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this have
 \begin{align*}
@@
     \centering
-    \includegraphics[width=.45\textwidth]{unrooted_four_taxa}
-    \caption{Four taxon tree}
+    \includegraphics[width=.45\textwidth]{unrooted_farris_tree}
+    \caption{Four taxon ``Farris zone'' tree.}
 \label{fig:four-taxa-tree}
@@
 
-Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}.
+Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}---this is commonly known as the ``Farris zone'' topology.
 The set of all possible character assignments is
@@
 Since we identify character complements, we do not consider the additional splits
-$$
-\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
-$$
+\begin{equation*}
+\begin{split}
+& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
+&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
+\end{split}
+\end{equation*}
 the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
@@
 $$
-Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{2\}$ and internal node $\{2\}$ connected to leaves $\{3\}$ and $\{4\}$.
+Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{3\}$ and internal node $\{2\}$ connected to leaves $\{2\}$ and $\{4\}$.
 The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
@@
 Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
-We cannot identify complements for ancestral states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
+We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
 $$
@@
 For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
-A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; for the $j$th site split, define $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ as the set of most likely ancestral splits for that particular site split, topology and set of branch lengths.
+A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
 As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
@@
 P(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
+P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
 (1-p)^5.
 $$
-For $\siteSplit_6=\{1,3\}$ we have
-$$
-P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,3\}, \tau, t) =
-P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
+For $\siteSplit_5=\{1,2\}$ we have
+$$
+P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
+P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
 p^2(1-p)^3,
@@
 $$
-P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
-P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
+P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
+P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
 p^3(1-p)^2.
 $$
-Here, the set of most likely ancestral states is $\ancestralSplitPartition_6(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_6\in\ancestralSplitPartition_6(\tau, p)$,
-$$
-P(\ancestralSplitRV=\xi_6 \mid \siteSplitRV=\{1,3\}, \tau, t) =
+Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, p)$,
+$$
+P(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
 p^2(1-p)^3.
@@
 be the final product in \eqref{eq:site_pattern_likelihood}.
-Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
+Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
 We have
-\begin{align}
-    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t))
-        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
-        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
+\begin{equation*}
+\begin{split}
+&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t)) \\
+&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
+&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
             \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
-\end{align}
+\end{split}
+\end{equation*}
 so that, in the limit of infinite data,
 \begin{equation}
-    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)].
+\begin{split}
+&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \\
+&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)].
 \label{eq:site_pattern_profile_likelihood_mean}
+\end{split}
 \end{equation}
@@
 \end{equation}
-For data generated from $(\tau^*, t^*)$, we have an inconsistency if
+For data generated from $(\tau^*, t^*)$, we have an inconsistency in estimating topology if
 \begin{equation}
@@
 Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Fig.~\ref{fig:farris-fels-top}.
-We parameterize the branches of these trees \textbf{not with the standard notion of branch length in terms of number of substitutions per site}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
+We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
 This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
-Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.
+Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform (see either 8.6 of Semple and Steel \cite{Semple2003-em} or Appendix (TODO)).
 
@@
 We now show that, in the case of the Farris tree topology, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
-Using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}, we calculate the generating probabilities on the Farris tree topology.
+Using the Hadamard transform approach, we calculate the generating probabilities on the Farris tree topology.
 For site split $\emptyset$,
@@
 
-In summary, exchanging $x_1$ and $x_2$ does not change the value of the log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, and thus $x_1=x_2$ at the maximum.
+Since exchanging $x_1$ and $x_2$ does not change the value of the log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
 The analogous statement holds for $y_1$ and $y_2$.
@@
 The full Farris tree log likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see Table~\ref{tab:likelihoods} for $\tau = \tau_1$).
-We will begin by writing out the middle case: for the ancestral state split $\{1\}$.
+We begin by writing out the middle case: for the ancestral state split $\{1\}$.
 Directly substituting in \eqref{eq:site_pattern_likelihood}, the log likelihood is
@@
 \begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &\le      p_{\emptyset}  \cdot\log(2(1+x')(1+y')) \\
-    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+y')) \\
-    &\qquad + p_{2}          \cdot\log((1+x')(1+y')(1-y')) \\
-    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+y')) \\
-    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+y')(1-y')) \\
-    &\qquad + p_{13}         \cdot\log((1+x')(1+y')) \\
-    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+y')(1-y')) \\
-    &\qquad + p_{123}        \cdot\log((1+x')(1+y')(1-y')) \\
-    &\qquad - \log(8) \\
-    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
-    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
-    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
-    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
-    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
-    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
-    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
-    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
-\end{align*}
-Grouping together like terms,
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &\le      p_{\emptyset}  \cdot\log(2) \\
-    &\qquad + \log(1+x') \\
-    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
-    &\qquad + \log(1+y') \\
-    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
-    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x') \\
-    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
-    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\
-    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
-    &\qquad + (1-p_{13})\cdot\log(1+w') \\
-    &\qquad + p_{13}\cdot\log(1-w') \\
-    &\qquad - \log(8),
-\end{align*}
-which equals
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &\le      p_{\emptyset}  \cdot\log(2) \\
-    &\qquad + (3-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x') \\
-    &\qquad + 2(p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
-    &\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\
-    &\qquad + 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
-    &\qquad + (1-p_{13})\cdot\log(1+w') \\
-    &\qquad + p_{13}\cdot\log(1-w') \\
-    &\qquad - \log(8).
+&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
+&\qquad \le      p_{\emptyset}  \cdot\log(2(1+x')(1+y'))
++ p_{1}          \cdot\log((1+x')(1-x')(1+y')) \\
+    &\qquad\qquad + p_{2}          \cdot\log((1+x')(1+y')(1-y'))
++ p_{3}          \cdot\log((1+x')(1-x')(1+y')) \\
+    &\qquad\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+y')(1-y'))
++ p_{13}         \cdot\log((1+x')(1+y')) \\
+    &\qquad\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+y')(1-y'))
++ p_{123}        \cdot\log((1+x')(1+y')(1-y'))
+- \log(8) \\
+    &\qquad\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2)
++ p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
+    &\qquad\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y'))
++ p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
+    &\qquad\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y'))
++ p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
+    &\qquad\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y'))
++ p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
+&\qquad =      p_{\emptyset}  \cdot\log(2)
++ \log(1+x')
++ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
+&\qquad\qquad + \log(1+y')
++ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
+&\qquad\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x')
++ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
+&\qquad\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y')
++ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
+&\qquad\qquad + (1-p_{13})\cdot\log(1+w')
++ p_{13}\cdot\log(1-w')
+- \log(8) \\
+&\qquad =      p_{\emptyset}  \cdot\log(2)
++ (3-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x')
++ 2(p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
+&\qquad\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y')
++ 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
+&\qquad\qquad + (1-p_{13})\cdot\log(1+w')
++ p_{13}\cdot\log(1-w')
+- \log(8).
 \end{align*}
@@
 where
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &\le      p_{\emptyset}  \cdot\log(2) \\
-    &\qquad + a_{1}\cdot\log(1+x') \\
-    &\qquad + a_{2}\cdot\log(1-x') \\
-    &\qquad + a_{3}\cdot\log(1+y') \\
-    &\qquad + a_{4}\cdot\log(1-y') \\
-    &\qquad + (1-p_{13})\cdot\log(1+w') \\
-    &\qquad + p_{13}\cdot\log(1-w') \\
-    &\qquad - \log(8).
-\end{align*}
+\begin{equation*}
+\begin{split}
+&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
+&\qquad \le      p_{\emptyset}  \cdot\log(2)
++ a_{1}\cdot\log(1+x')
++ a_{2}\cdot\log(1-x')
++ a_{3}\cdot\log(1+y') \\
+&\qquad\qquad + a_{4}\cdot\log(1-y')
++ (1-p_{13})\cdot\log(1+w')
++ p_{13}\cdot\log(1-w')
+- \log(8).
+\end{split}
+\end{equation*}
 Maximizing over the unknown terms yields
@@
 \begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &\le      p_{\emptyset}  \cdot\log(2) \\
-    &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
-    &\qquad + a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
-    &\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
-    &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
-    &\qquad + (1-p_{13})\cdot\log(2(1-p_{13})) \\
-    &\qquad + p_{13}\cdot\log(2p_{13}) \\
-    &\qquad - \log(8) \\
+&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
+&\qquad\le      p_{\emptyset}  \cdot\log(2)
++ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
++ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}}
++ a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
+&\qquad\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
++ (1-p_{13})\cdot\log(2(1-p_{13}))
++ p_{13}\cdot\log(2p_{13})
+- \log(8) \\
 &\qquad := L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
@@
 \begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &\le      p_{\emptyset}  \cdot\log(2) \\
-    &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\
-    &\qquad + a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
-    &\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
-    &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
-    &\qquad + \log(2) \\
-    &\qquad - \log(8) \\
+&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
+&\qquad\le      p_{\emptyset}  \cdot\log(2)
++ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
++ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
+&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
++ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
++ \log(2)
+- \log(8) \\
 &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}),
@@
 \begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &\le      p_{\emptyset}  \cdot\log(2) \\
-    &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
-    &\qquad + a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
-    &\qquad + a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'} \\
-    &\qquad + a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'} \\
-    &\qquad + \log(2) \\
-    &\qquad - \log(8) \\
+&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
+&\qquad\le      p_{\emptyset}  \cdot\log(2)
++ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
++ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
+&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
++ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
++ \log(2)
+- \log(8) \\
 &\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
@@
 In the Felsenstein case, there are various optimal internal states depending on branch lengths (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).
-We can bound these below replacing $1+w$ with $1-w$, for example as follows.
+We can bound these below replacing $1+w$ with $1-w$ as follows.
 For the site split $\{1,2\}$, we would like to find a lower bound for
@@
 % obtained by running
-% python joint_inf_plot.py --topology --delta .001 --plot-name figures/topology-inconsistency.svg
-\includegraphics[width=.9\textwidth]{topology-inconsistency-legend}
+% python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
+\includegraphics[width=.9\textwidth]{topology-inconsistency-inkscape}
 \caption{
@@
 \subsection{Restricted case: fixed $x'$ and $y'$}
+\label{sec:restricted-case}
 
@@
 % obtained by running
-% python joint_inf_plot.py --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
-\includegraphics[width=.9\textwidth]{branch-length-inconsistency-legend}
+% python joint_inf_plot.py --analytic --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
+\includegraphics[width=.9\textwidth]{branch-length-inconsistency-inkscape}
 \caption{Region of inconsistency with $x$ and $y$ fixed}
@@
 
-\subsubsection{General case: loose bounds}
+\subsubsection{Loose bounds}
 
@@
 % obtained by running
-% python joint_inf_plot.py --general-branch-lengths --analytic --delta .001 --plot-name figures/bl-loose-inconsistency.svg
-\includegraphics[width=.9\textwidth]{bl-loose-inconsistency}
+% python joint_inf_plot.py --analytic --general-branch-lengths --delta .001 --plot-name figures/bl-loose-inconsistency.svg
+\includegraphics[width=.9\textwidth]{bl-loose-inconsistency-inkscape}
 \caption{Region of inconsistency with $x-.1 \le \hat{x}' \le x+.1$ and $y-.1 \le \hat{y}' \le y+.1$.}
@@
 
-\subsection{Empirical results for the general case}
-
-When seeking a global maximum for $0 \le x',y',w' \le 1$, we use the method of differential evolution \cite{Storn1997}---we obtain the fit $\hat{w}'$ in Fig.~\ref{fig:w_hat_joint}.
-The estimates for $\hat{w}'$ when $x$ and $y$ are large seem to be unbiased, with $\hat{w}'$ increasing as $y$ increases, though the region where $\hat{w}'$ is estimated to be one---indicating no change along that particular branch---is similar to that in the restricted case.
-Although the restricted and general case are similar, comparing Figures~\ref{fig:bl-inconsistency} and~\ref{fig:w_hat_joint} shows that the general case has a smaller region where $\hat{w}'$ is one, as expected.
-Nevertheless, this region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have a region of the space of true parameters where we will always estimate the interior branch parameter to be one.
+\subsubsection{Empirical results}
+
+The region of inconsistency in Figure~\ref{fig:bl-loose-inconsistency} is potentially a conservative estimate of where $\hat{w}=1$.
+To determine how conservative, we use the method of differential evolution \cite{Storn1997} to perform joint maximum likelihood estimation.
+Since empirical estimates can be somewhat unstable, we perform two maximizations---one over $0 \le x',y',w' \le 1$ and one over $0 \le x',y' \le 1$ with $w'=1$---and take the value of $\hat{w}'$ with the higher likelihood.
+We compute these maxima over a lattice in steps of $1e-3$ for $x$ and $y$ from $1e-3$ to $1-1e-3$---we do not include zero or one in our lattice to further stabilize the fits, recalling from Section~\ref{sec:restricted-case} that cases where $x$ or $y$ are equal to zero or one can result in pathologies.
+From this process we obtain Fig.~\ref{fig:bl-general-inconsistency}.
+
+The estimates for $\hat{w}'$ when $x$ and $y$ are both large seem to be unbiased, with $\hat{w}'$ increasing as $y$ increases, though the region where $\hat{w}'$ is estimated to be one---indicating no change along that particular branch---is similar to that in the restricted case.
+Although the restricted and general case are similar, comparing Figures~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case has a slightly different character, possibly due to potential looseness of the bound in the restricted case.
+%das: TODO double check this; do the same empirical plot for the restricted case
+This region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have a region of the space of true parameters where we will always estimate the interior branch parameter to be one.
 
@@
 % obtained by running
-% python joint_inf_plot.py --general-branch-lengths --delta .001 --plot-name figures/w-hat-heatmap-joint-diff-evol.svg --n-jobs 10
-\includegraphics[width=.9\textwidth]{w-hat-heatmap-joint-diff-evol}
+% python joint_inf_plot.py --empirical --general-branch-lengths --delta .001 --plot-name figures/w-hat-empirical-001.svg --pkl-name figures/w-hat-empirical-001.pkl --n-jobs 16
+% or
+% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-001.svg --in-pkl-name figures/w-hat-empirical-001.pkl
+% if fit already
+\includegraphics[width=.9\textwidth]{w-hat-empirical-001}
 \caption{
+    %das: note to update caption
     Estimating $(\hat{x}', \hat{y}', \hat{w}')$.
 }
-\label{fig:w_hat_joint}
+\label{fig:bl-general-inconsistnecy}
 \end{figure}

### 2017-11-13 joint_inf_plot.py

@@ -422,10 +422,10 @@ def main(args=sys.argv[1:]):
                     Z = np.vectorize(lambda x, y: max_likelihood((x, y), delta=args.delta))(X, Y)
 
                 with open(args.out_pkl_name, 'w') as f:
-                    pickle.dump((X, Y, Z), f)
+                    pickle.dump((X, Y, Z, args.delta), f)
             else:
                 with open(args.in_pkl_name, 'r') as f:
-                    X, Y, Z = pickle.load(f)
+                    X, Y, Z, args.delta = pickle.load(f)
 
             im = plt.imshow(Z, cmap=plt.cm.gray, origin='lower')
             plt.xlabel(r'$x$')

### 2017-11-14 joint_inf.tex
substantial restructure

@@
 % Commands
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}[theorem]{Lemma}
+
 \newcommand{\alphabet}{\mathcal{A}}
@@
 
-\section{Ancestral state reconstruction using maximum likelihood}
+\section{Phylogenetic maximum likelihood}
+
+Set up mild notation needed to state likelihood
 
@@
 
-\subsection{Classical maximum-likelihood inference in phylogenetics}
+\subsection{Phylogenetic parameters}
+
+\subsubsection{Topology}
+
+\begin{figure}
+\centering
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[width=.95\textwidth]{farris_blank}
+\caption[short]{Farris topology $\tau_1$}
+\end{subfigure}
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[width=.95\textwidth]{felsenstein_blank}
+\caption[short]{Felsenstein topology $\tau_2$}
+\end{subfigure}
+\caption{Two simple topologies}
+\label{fig:farris-fels-top}
+\end{figure}
+
+Consider the fixed, binary four-tip tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}---this is commonly known as the ``Farris zone'' topology.
+
+\subsubsection{Fidelities}
+
+We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
+This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
+Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform (see either 8.6 of Semple and Steel \cite{Semple2003-em} or Appendix (TODO)).
+
+\subsection{Classical maximum likelihood}
 
@@
 
-\subsection{Joint maximization}
+\subsection{Joint maximum likelihood}
 
@@
 
-\subsection{Site split formulation}
-
-Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation (we follow Section 8.6 of \cite{Semple2003-em}).
-Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
-For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
-Indeed, we start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
-Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.
-
-For topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization of $\ancestralStateColumn_i$.
-In this case the entire set of internal nodes must be enumerated: the probability of observing an ancestral state split, conditional on a site split, is not invariant to taking its complement.
-
-We enumerate the site splits $\siteSplit_j$ of which there are $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ in total where $\mathcal{P}$ denotes the power set.
-Similarly we enumerate ancestral splits $\ancestralSplit_k$ of which there are $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$ in total.
-
-This split formulation defines mappings from patterns to splits as
- $$
- \patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet).
- $$
-Given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
-$$
-\siteSplitRV := \patternToSplit(\alignmentColumnRV)
-$$
-that takes corresponding realizations $\siteSplit_j$ for some $j$.
-
-For the $i$th factor of \eqref{eq:full_likelihood},
-$$
-P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
-$$
-As a consequence of assuming a binary symmetric model, taking complements yields
-\begin{align*}
-    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
-                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
-\end{align*}
-for some $j$.
-For the other term, we define
-$$
-\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
-$$
-as a mapping of ancestral states and tip states to ancestral state splits.
-It is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.
-We define
-$$
-\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
-$$
-for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, so that
-%By considering the case requiring complementing and that not requiring complementing separately,
-%das: did I write the above? I don't remember the significance of it...
-\begin{align*}
-    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
-\end{align*}
-Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
-\begin{align*}
-\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
-\end{align*}
-In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
-For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this have
-\begin{align*}
-\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
-\end{align*}
-
-Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
-Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
-\begin{align}
-L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
-                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
-                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
-                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
-                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
-\end{align}
-where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.
-
-\paragraph{Example}
-\begin{figure}
-    \centering
-    \includegraphics[width=.45\textwidth]{unrooted_farris_tree}
-    \caption{Four taxon ``Farris zone'' tree.}
-\label{fig:four-taxa-tree}
-\end{figure}
-
-Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}---this is commonly known as the ``Farris zone'' topology.
-The set of all possible character assignments is
-\begin{align*}
-\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
-                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
-\end{align*}
-where each set indicates the tips assigned the character $1$.
-For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
-Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
-The unique site splits, collapsing complements, are
-\begin{align*}
-    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
-& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
-\end{align*}
-Since we identify character complements, we do not consider the additional splits
-\begin{equation*}
-\begin{split}
-& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
-&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
-\end{split}
-\end{equation*}
-the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
-This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
-$$
-\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
-$$
-Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{3\}$ and internal node $\{2\}$ connected to leaves $\{2\}$ and $\{4\}$.
-The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
-For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
-Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
-We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
-$$
-P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
-$$
-in general.
-
-For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
-A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
-As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
-The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
-$$
-P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
-(1-p)^5,
-$$
-$$
-P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
-P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
-p^3(1-p)^2,
-$$
-$$
-P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
-p^4(1-p).
-$$
-The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, p)=\{\emptyset\}$.
-Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, p)$ we have
-$$
-P(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
-P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
-(1-p)^5.
-$$
-For $\siteSplit_5=\{1,2\}$ we have
-$$
-P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
-P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
-p^2(1-p)^3,
-$$
-$$
-P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
-P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
-p^3(1-p)^2.
-$$
-Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, p)$,
-$$
-P(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
-p^2(1-p)^3.
-$$
-
-\subsection{Showing inconsistency}
-
-We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
-Let
-$$
-L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
-$$
-be the final product in \eqref{eq:site_pattern_likelihood}.
-Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
-We have
-\begin{equation*}
-\begin{split}
-&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t)) \\
-&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
-&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
-            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
-\end{split}
-\end{equation*}
-so that, in the limit of infinite data,
-\begin{equation}
-\begin{split}
-&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \\
-&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)].
-\label{eq:site_pattern_profile_likelihood_mean}
-\end{split}
-\end{equation}
-Define the divergence quantity
-$$
-\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
-$$
-and the partial log likelihood
-$$
-\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
-$$
-so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
-\begin{equation}
-    \label{eq:log_likelihood_simplified}
-    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
-\end{equation}
-For data generated from $(\tau^*, t^*)$, we have an inconsistency in estimating topology if
-\begin{equation}
-\label{eq:inconsistency_inequality}
-\ell_{\tau^*,t^*}(\tau', \hat{t}_1; \fullAncestralSplitPartitions(\tau', \hat{t}_1)) > \ell_{\tau^*,t^*}(\tau^*, \hat{t}_2; \fullAncestralSplitPartitions(\tau^*, \hat{t}_2))
-\end{equation}
-for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\fullAncestralSplitPartitions(\tau', \hat{t}_1)\}$ and $\{\hat{t}_2,\fullAncestralSplitPartitions(\tau^*, \hat{t}_2)\}$ estimated by maximizing $\tilde{\ell}_{\tau^*,t^*}$ with respect to the branch lengths and the internal states for $\tau = \tau'$ and $\tau^*$, respectively.
-By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
-In the next section, we show that such examples exist.
-
-\section{Inconsistency of joint maximum likelihood in estimating topology}
-
-\subsection{Bounding the likelihoods}
-
-Our approach seeks bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified}.
-We obtain
-$$
-\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \le
-    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
-    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t}))
-$$
-as an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
-$$
-\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
-$$
-and
-$$
-\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
-$$
-Similarly, we have the lower bound
-$$
-\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \ge
-    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
-    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t})).
-$$
+\section{Inconsistency of joint inference}
 
@@
 
-\begin{figure}
-\centering
-\begin{subfigure}{.45\linewidth}
-\centering
-\includegraphics[width=.95\textwidth]{farris_blank}
-\caption[short]{Farris topology $\tau_1$}
-\end{subfigure}
-\begin{subfigure}{.45\linewidth}
-\centering
-\includegraphics[width=.95\textwidth]{felsenstein_blank}
-\caption[short]{Felsenstein topology $\tau_2$}
-\end{subfigure}
-\caption{Two simple topologies}
-\label{fig:farris-fels-top}
-\end{figure}
+Consider the Farris topology with arbitrary branch lengths, i.e.,
+$$
+\tilde{t}=\{x_1,y_1,x_2,y_2,w\}
+$$
+with edge ordering in the order of the taxa, then the internal branch last.
+We now show that, in the case of the Farris tree topology, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
 
-Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Fig.~\ref{fig:farris-fels-top}.
-We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
-This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
-Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform (see either 8.6 of Semple and Steel \cite{Semple2003-em} or Appendix (TODO)).
-
-Let $\{x',y',w'\}\in[0,1]^3$ be three fidelities and $\{p_x',p_y',p_w'\}\in[0,1/2]^3$ the corresponding substitution probabilities.
-Call $t=\{x',y',w'\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ constrains the bottom three branches to share the same parameter, the classical construction of this topology.
-
-Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ with edge ordering in the order of the taxa, then the internal branch last.
-We now show that, in the case of the Farris tree topology, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
-Using the Hadamard transform approach, we calculate the generating probabilities on the Farris tree topology.
-For site split $\emptyset$,
-\begin{align*}
-    P(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
-                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
-                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
-\end{align*}
-and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
-All other generating probabilities will differ only in the signs of each term.
-For example, for site split $\{1\}$ we have
-\begin{align*}
-    P(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
-\end{align*}
-and for site split $\{3\}$ we have
-\begin{align*}
-    P(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
-\end{align*}
-meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
-The corresponding possibilities for the likelihood values are
-\begin{align*}
-    P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
-    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
-    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
-    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
-\end{align*}
-for site split $\{1\}$ and
-\begin{align*}
-        P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
-    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
-    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
-    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
-\end{align*}
-for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.
-
-The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
-The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
+\begin{lemma}
+For $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ and $t=\{x,y,x,y,w\}$,
+$$\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t})) = \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)).$$
+\end{lemma}
 
@@
 
-Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies under these two topologies.
-Table~\ref{tab:likelihoods} contains calculations of likelihood values for fixed site patterns and topologies that have been maximized over ancestral state patterns.
-Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} contain the values of the likelihood for all possible ancestral states.
+\subsection{Inconsistency in topology estimation}
 
-\begin{table}
-\centering
-\begin{tabular}{|l|l|l|}
-    \hline
-$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
-    \hline
-    $\emptyset$    &$1+x'^2+y'^2+4x'y'w'+x'^2y'^2$&$1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
-    $\{1\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
-    $\{2\}$        &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
-    $\{3\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
-    $\{1,2\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
-    $\{1,3\}$      &$1+x'^2+y'^2-4x'y'w'+x'^2y'^2$&$1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
-    $\{2,3\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
-    $\{1,2,3\}$    &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
-    \hline
-\end{tabular}
-\caption{Site pattern probabilities.
-All values are multiplied by $1/8$.}
-\label{tab:sitepatprob}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{|l|ll|}
-    \multicolumn{3}{c}{$\tau=\tau_1$}\\
-    \hline
-    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
-    \hline
-    $\emptyset$&
-    $\emptyset$&
-    $(1+x')^2   (1+w')(1+y')^2$\\
-     $\{1\}$    &
-    $\emptyset$&
-    $(1+x')(1-x')(1+w')(1+y')^2$\\
-     $\{2\}$    &
-    $\emptyset$&
-    $(1+x')^2   (1+w')(1+y')(1-y')$\\
-     $\{3\}$    &
-    $\emptyset$&
-    $(1+x')(1-x')(1+w')(1+y')^2$\\
-    $\{1,2\}$  &
-    $\{\emptyset,\{1,2\}\}$&
-    $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
-    $\{1,3\}$  &
-    $\left\{\begin{array}{l}
-                    \emptyset\\
-                    \{1\}\\
-                    \{1,2\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                    (1-x')^2   (1+w')(1+y')^2\\
-                    (1+x')^2   (1-w')(1+y')^2\\
-                    (1+x')^2   (1+w')(1-y')^2
-                \end{array}$\\
-    $\{2,3\}$  &
-                $\{\emptyset,\{1,2\}\}$&
-                $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
-    $\{1,2,3\}$&
-                $\{1,2\}$&
-                $(1+x')^2   (1+w')(1+y')(1-y')$\\
-    \hline
-    \multicolumn{3}{c}{$\tau=\tau_2$}\\
-    \hline
-    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
-    \hline
-    $\emptyset$       &$\emptyset$&$(1+x')^2   (1+w')(1+y')^2$\\
-    $\{1\}$          &
-    $\left\{\begin{array}{l}
-                    \emptyset\\
-                    \{1\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                        (1+x')(1-x')(1+w')(1+y')^2\\
-                        (1+x')^2   (1-w')(1+y')(1-y')
-                    \end{array}$\\
-      $\{2\}$          &
-    $\left\{\begin{array}{l}
-                    \emptyset\\
-                    \{1\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                    (1+x')^2   (1+w')(1+y')(1-y')\\
-                    (1+x')(1-x')(1-w')(1+y')^2
-                    \end{array}$\\
-      $\{3\}$          &
-    $\left\{\begin{array}{l}
-                    \emptyset\\
-                    \{2\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                    (1+x')(1-x')(1+w')(1+y')^2\\
-                    (1+x')^2   (1-w')(1+y')(1-y')
-                    \end{array}$\\
-     $\{1,2\}$         &
-    $\left\{\begin{array}{l}
-                    \{\emptyset,\{1,2\}\}\\
-                    \{1\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
-                    (1+x')^2   (1-w')(1+y')^2
-                    \end{array}$\\
-     $\{1,3\}$         &
-    $\left\{\begin{array}{l}
-                    \emptyset\\
-                    \{\{1\},\{2\}\}\\
-                    \{1,2\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                    (1-x')^2   (1+w')(1+y')^2\\
-                    (1+x')(1-x')(1-w')(1+y')(1-y')\\
-                    (1+x')^2   (1+w')(1-y')^2
-                    \end{array}$\\
-      $\{2,3\}$        &
-    $\left\{\begin{array}{l}
-                    \{\emptyset,\{1,2\}\}\\
-                    \{1\}\\
-                    \{2\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
-                    (1-x')^2   (1-w')(1+y')^2\\
-                    (1+x')^2   (1-w')(1-y')^2
-                    \end{array}$\\
-     $\{1,2,3\}$       &
-    $\left\{\begin{array}{l}
-                    \{2\} \\
-                    \{1,2\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                    (1+x')(1-x')(1-w')(1+y')^2 \\
-                    (1+x')^2   (1+w')(1+y')(1-y')
-                    \end{array}$\\
-    \hline
-\end{tabular}
-\caption{Likelihood calculations for all site patterns and ancestral state partitions.
-All values multiplied by $1/32$.
-Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
-See Appendix for full calculations.}
-\label{tab:likelihoods}
-\end{table}
-
-\subsubsection{Upper bound for Farris tree likelihood}
-
-Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
-For compactness, define
+Define $\ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t))$ as
 $$
-p_{\emptyset} = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})
+\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
 $$
-with similar definitions for the remaining generating probabilities.
-The full Farris tree log likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see Table~\ref{tab:likelihoods} for $\tau = \tau_1$).
-We begin by writing out the middle case: for the ancestral state split $\{1\}$.
-Directly substituting in \eqref{eq:site_pattern_likelihood}, the log likelihood is
-\begin{align}
-    \label{eq:farris_likelihood}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \nonumber \\
-    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
-    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\
-    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
-    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
-    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \nonumber \\
-    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
-    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\
-    &\qquad - \log(8) \nonumber \\
-    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \nonumber \\
-    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
-    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \nonumber \\
-    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
-    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\
-    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \nonumber \\
-    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\
-    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
-\end{align}
-Seeking an upper bound, we use the facts that, for $x',y'\in[0,1]$,
-\begin{align*}
-1-x'^2+y'^2-x'^2y'^2 & = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y') \\
-1+x'^2-y'^2-x'^2y'^2 & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y') \\
-1+x'^2+y'^2+x'^2y'^2 & = (1+x'^2)(1+y'^2) \le (1+x')(1+y') \\
-4x'y' & = 2x' \cdot 2y' \le (1+x'^2)(1+y'^2) \le (1+x')(1+y').
-\end{align*}
-The term $p_{\emptyset}$ is bounded as
-\begin{align*}
-    p_{\emptyset} & = \log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
-                  & \le \log(1+x'^2+y'^2+4x'y'+x'^2y'^2) \\
-                  & \le \log(2(1+x')(1+y'))
-\end{align*}
-and $p_{13}$ is bounded as
-\begin{align*}
-    p_{13} & = \log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
-                  & \le \log(1+x'^2+y'^2+x'^2y'^2) \\
-                  & \le \log((1+x')(1+y')).
-\end{align*}
-Factoring and making these substitutions results in
-\begin{align*}
-&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
-&\qquad \le      p_{\emptyset}  \cdot\log(2(1+x')(1+y'))
-+ p_{1}          \cdot\log((1+x')(1-x')(1+y')) \\
-    &\qquad\qquad + p_{2}          \cdot\log((1+x')(1+y')(1-y'))
-+ p_{3}          \cdot\log((1+x')(1-x')(1+y')) \\
-    &\qquad\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+y')(1-y'))
-+ p_{13}         \cdot\log((1+x')(1+y')) \\
-    &\qquad\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+y')(1-y'))
-+ p_{123}        \cdot\log((1+x')(1+y')(1-y'))
-- \log(8) \\
-    &\qquad\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2)
-+ p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
-    &\qquad\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y'))
-+ p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
-    &\qquad\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y'))
-+ p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
-    &\qquad\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y'))
-+ p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
-&\qquad =      p_{\emptyset}  \cdot\log(2)
-+ \log(1+x')
-+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
-&\qquad\qquad + \log(1+y')
-+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
-&\qquad\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x')
-+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
-&\qquad\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y')
-+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
-&\qquad\qquad + (1-p_{13})\cdot\log(1+w')
-+ p_{13}\cdot\log(1-w')
-- \log(8) \\
-&\qquad =      p_{\emptyset}  \cdot\log(2)
-+ (3-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x')
-+ 2(p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
-&\qquad\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y')
-+ 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
-&\qquad\qquad + (1-p_{13})\cdot\log(1+w')
-+ p_{13}\cdot\log(1-w')
-- \log(8).
-\end{align*}
-To make matters further compact, let
-\begin{equation}
-    \begin{aligned}
-        a_{1} &= 3-p_{1}-p_{3}-p_{12}-p_{23}, \\
-        a_{2} &= 2(p_{1}+p_{3}+p_{12}+p_{23}), \\
-        a_{3} &= 3-p_{2}-p_{12}-p_{23}-p_{123}, \\
-        a_{4} &= 2(p_{2}+p_{12}+p_{23}+p_{123}), \\
-    \end{aligned}
-    \label{eq:a_const}
-\end{equation}
-where
-\begin{equation*}
-\begin{split}
-&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
-&\qquad \le      p_{\emptyset}  \cdot\log(2)
-+ a_{1}\cdot\log(1+x')
-+ a_{2}\cdot\log(1-x')
-+ a_{3}\cdot\log(1+y') \\
-&\qquad\qquad + a_{4}\cdot\log(1-y')
-+ (1-p_{13})\cdot\log(1+w')
-+ p_{13}\cdot\log(1-w')
-- \log(8).
-\end{split}
-\end{equation*}
-Maximizing over the unknown terms yields
+We show the following.
+\begin{theorem}
+There exist $L_0$ and $L_1$ such that
 $$
-\hat{x}' = \frac{a_{1}-a_{2}}{a_{1}+a_{2}}, \ \hat{y}' = \frac{a_{3}-a_{4}}{a_{3}+a_{4}}.
+\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \le L_0,
 $$
-We check that $0 \le \hat{x}' \le 1$ and $0 \le \hat{y}' \le 1$ so these maximizers are valid.
-Letting
 $$
-\tilde{p} = p_{1}+p_{3}+p_{12}+p_{23}
+\max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t; \fullAncestralSplitPartitions(\tau_2,t)) \ge L_1
 $$
-we have
-$$
-\hat{x}' = \frac{3-\tilde{p} - 2\tilde{p}}{3-\tilde{p}+2\tilde{p}} = \frac{3-3\tilde{p}}{3+\tilde{p}},
-$$
-and, since $0 \le \tilde{p} \le 1/2$, $0 \le \hat{x}' \le 1$.
-The same approach works for $\hat{y}'$.
-The upper bound for the likelihood is then maximized at
-\begin{align*}
-&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
-&\qquad\le      p_{\emptyset}  \cdot\log(2)
-+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
-+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}}
-+ a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
-&\qquad\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
-+ (1-p_{13})\cdot\log(2(1-p_{13}))
-+ p_{13}\cdot\log(2p_{13})
-- \log(8) \\
-&\qquad := L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
-\end{align*}
-
-The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
-Call
-\begin{equation}
-    \begin{aligned}
-        a_{1}' &= a_{1}-2p_{13}, \\
-        a_{2}' &= a_{2}+2p_{13}, \\
-        a_{3}' &= a_{3}-2p_{13}, \\
-        a_{4}' &= a_{4}+2p_{13}. \\
-    \end{aligned}
-    \label{eq:a_const_prime}
-\end{equation}
-We also now need to check that
-$$
-\hat{x}' = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y}' = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
-$$
-are both between zero and one.
-It will suffice to show
-$$
-0 \le \tilde{p} - 2p_{13} \le 1
-$$
-and
-$$
-0 \le \tilde{p} + 2p_{13} \le 1
-$$
-for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
-Using Table~\ref{tab:sitepatprob},
-$$
-p_{13} = \frac{1}{8}\left(1 + x^2 + y^2 - 4xy^2 + x^2y^2\right).
-$$
-We see that
-$$
-1 + x^2 + y^2 - 4xy^2 + x^2y^2 \le 2 \iff y^2\left(1 - 4x + x^2\right) \le 1 - x^2
-$$
-and
-$$
-y^2\left(1 - 4x + x^2\right) \le 1 - 4x + x^2 \le 1 - x^2,
-$$
-the last inequality holding if $x \le 2$.
-Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}'$ and $\hat{y}'$ are valid.
-
-To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
-\begin{align*}
-&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
-&\qquad\le      p_{\emptyset}  \cdot\log(2)
-+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
-+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
-&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
-+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
-+ \log(2)
-- \log(8) \\
-&\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}),
-\end{align*}
-and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
-\begin{align*}
-&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
-&\qquad\le      p_{\emptyset}  \cdot\log(2)
-+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
-+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
-&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
-+ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
-+ \log(2)
-- \log(8) \\
-&\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
-\end{align*}
-All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.
-
-\subsubsection{Lower bound for Felsenstein tree partial likelihood}
-
-We construct a similar lower bound on the Felsenstein tree partial likelihood.
-In the Felsenstein case, there are various optimal internal states depending on branch lengths (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).
-We can bound these below replacing $1+w$ with $1-w$ as follows.
-For the site split $\{1,2\}$, we would like to find a lower bound for
-$$
-\max\left((1+x')^2 (1-w')(1+y')^2, (1+x')(1-x')(1+w')(1+y')(1-y')\right).
-$$
-We observe that
-$$
-(1+x')^2 (1-w')(1+y')^2 > (1+x')(1-x')(1-w')(1+y')(1-y')
-$$
-always, and thus the maximum can be bounded below by $(1+x')^2 (1-w')(1+y')^2$.
-For the other site splits we consider two cases.
-If $(1+x')(1-y') > (1-x')(1+y')$, then
-$$
-(1+x')^2(1-w')(1+y')(1-y') > (1+x')(1-x')(1-w')(1+y')^2
-$$
-meaning the cases $\{1\}, \{2\}, \{3\}$ and $\{1,2,3\}$ will have likelihoods bounded below by
-$$
-(1+x')^2(1-w')(1+y')(1-y').
-$$
-For the cases $\{1,3\}$ and $\{2,3\}$ and the same condition, we have
-$$
-(1+x')^2(1-w')(1-y')^2 > (1-x')^2(1-w')(1+y')^2
-$$
-and
-$$
-(1+x')^2(1-w')(1-y')^2 > (1+x')(1-x')(1-w')(1+y')(1-y'),
-$$
-yielding lower bounds of
-$$
-(1+x')^2(1-w')(1-y')^2
-$$
-in these cases.
-The full likelihood in this case is then bounded below by
-\begin{align*}
-    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &\ge      p_{\emptyset}  \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
-    &\qquad + p_{1}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
-    &\qquad + p_{2}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
-    &\qquad + p_{3}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
-    &\qquad + p_{12}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
-    &\qquad + p_{123}        \cdot\log((1+x')^2   (1-w')(1+y')(1-y'))\\
-    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1-y')^2) \\
-    &\qquad + p_{23}         \cdot\log((1+x')^2   (1-w')(1-y')^2).
-\end{align*}
-When $(1+x')(1-y') < (1-x')(1+y')$, similar arguments yield the lower bound
-\begin{align*}
-    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &\ge      p_{\emptyset}  \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
-    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
-    &\qquad + p_{2}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
-    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
-    %See above.
-    &\qquad + p_{12}         \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
-    &\qquad + p_{123}        \cdot\log((1+x')(1-x')(1-w')(1+y')^2)\\
-    &\qquad + p_{13}         \cdot\log((1-x')^2    (1-w')(1+y')^2) \\
-    &\qquad + p_{23}         \cdot\log((1-x')^2    (1-w')(1+y')^2).
-\end{align*}
-Letting
-$$
-b = p_{1}+p_{2}+p_{3}+p_{123}
-$$
-yields
-\begin{align*}
-    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &\ge      2\cdot\log(1+x') \\
-    &\qquad + (2-b)  \cdot\log(1+y') \\
-    &\qquad + b      \cdot\log(1-y') \\
-    &\qquad + \log(1-w')
-\end{align*}
-for the first lower bound, and we get the equivalent expression for the second lower bound by exchanging $x'$ and $y'$.
-Maximizing either lower bound over $\{x',y',w'\}$, we obtain
-\begin{align*}
-    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &\ge      2\cdot\log(2) \\
-    &\qquad + (2-b)  \cdot\log(2-b) \\
-    &\qquad + b      \cdot\log(b)
-\end{align*}
-%EM Shouldn't this be 2p_{\emptyset}-1 ? It doesn't matter for the bound, but...
-%das If we call p_{\emptyset} just p, we're maximizing
-%   p log (1 + w) + (1 - p) log (1 - w).
-%As before, a_1 log(1+x) + a_2 log(1-x) is maximized at x = (a_1 - a_2) / (a_1 + a_2).
-%So then w = (p - (1 - p)) / (p + (1 - p)) = (2p - 1) / 1 = 2p - 1.
-%So you're right. Good catch.
-%
-%Unfortunately, I realized the above does not work for us since 2p - 1 is not guaranteed to be between 0 and 1, as x should be (see the stuff I added justifying this above).
-%Rather than do a constrained optimization, I just bounded the p_{\emptyset}\log(1+w) term below by p_{\emptyset}\log(1-w).
-i.e.\ the maximum is achieved at $\{1,1-b,0\}$.
-Our lower bound is
-\begin{align*}
-    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,0\}) \\
-                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b).
-\end{align*}
-
-\subsubsection{Region of inconsistency}
-
-We want the region where
-$$
-L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) \ge \max(L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}), L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}),L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy})).
-$$
-Since this is an inequality in two variables, we plot them analytically---see Fig.~\ref{fig:inconsistency-farris}.
+and $L_1 > L_0$.
+\end{theorem}
 
@@
 
-\section{Inconsistency of joint maximum likelihood in estimating branch fidelities}
+\subsection{Inconsistency in branch length estimation}
 
@@
 
-\subsection{Restricted case: fixed $x'$ and $y'$}
-\label{sec:restricted-case}
+\subsubsection{Restricted case}
 
@@
 In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
-Recalling Table~\ref{tab:likelihoods}, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood for $(x',y',w')$.
-%EM You are talking about a log likelihood here, right? So a likelihood being proportional to means that there is some additive constant. Makes things a little confusing.
-%das: updated
-If $x'$ and $y'$ are fixed, by collapsing like terms and ignoring terms with only $x'$ and $y'$, the log-likelihood in \eqref{eq:farris_likelihood} as a function of $w'$ is
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
-    &\sim     p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
-    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
-    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
-    &\qquad + p_{13}\cdot\log(1-w')
-\end{align*}
-where $\sim$ denotes equality up to an additive constant.
-Similarly, the other two possibilities for the likelihood both satisfy
-\begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
-    &\sim     p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
-    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
-    &\qquad + \log(1+w').
-\end{align*}
-Since
-$$
-\log(1+w') \geq (1 - p_{13})\cdot\log(1+w') + p_{13}\cdot\log(1-w')
-$$
-with equality only when $w'=0$, the second likelihood is the likelihood with the maximum-likelihood ancestral state, and so we need only analyze that case.
-
-Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
-Simplifying, the likelihood can be written
-$$
-\ell(w') = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
-$$
-where
+Let
 $$
@@
 $$
-Since $x$ and $y$ fall between zero and one, we make use of the inequalities (excepting the cases of $x=y=0$ and $x=y=1$)
+\begin{theorem}
+\label{thm:restricted-bl}
+The maximum likelihood value $\hat{w}' = 1$ if
 $$
-\alpha_1 > \alpha_2
+\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
 $$
-and
-$$
-\beta > \gamma.
-$$
-% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
-%EM Rather confusing to be using ' for two things here.
-What about using a dot over the l?
-In retrospect something other than ' would have been good because of the proliferation of w^{'2} etc.
-Prob not worth changing now though.
-The derivative of the log likelihood is
-$$
-\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} - \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
-$$
-The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w') \lessgtr 0$, we equivalently consider $f(w') \lessgtr 0$ where $f$ is the quadratic function
-\begin{align*}
-f(w') &= (w')^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
-      &\qquad + w'\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
-      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
-\end{align*}
-% Note from das: b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero.
-c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
-Excepting all the cases where $x=0$ or where $y=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
-Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w'=0$.
-This implies that if the smaller of the zeros of $f(w')$ is greater than one, then $\hat{w}' \equiv 1$.
-Using the quadratic formula with
-$$
-a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
-$$
-$$
-b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
-$$
-$$
-c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
-$$
-the smaller zero is
-$$
-\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)},
-$$
-which is a function of the generating parameters $x$ and $y$.
-We plot the region where $\hat{w}'(x, y) \ge 1$ in Figure~\ref{fig:bl-inconsistency}.
+\end{theorem}
+The inequality in Theorem~\ref{thm:restricted-bl} is a function of the generating parameters $x$ and $y$.
+This region is shown in Fig.~\ref{fig:bl-inconsistency}.
 
@@
 \includegraphics[width=.9\textwidth]{branch-length-inconsistency-inkscape}
-\caption{Region of inconsistency with $x$ and $y$ fixed}
+\caption{Region of inconsistency with $x$ and $y$ fixed.}
 \label{fig:bl-inconsistency}
@@
 
-\subsection{General case}
-
-\subsubsection{Loose bounds}
+\subsubsection{General case}
 
@@
 Assume we know $\hat{x}'$ and $\hat{y}'$ as functions of $x$ and $y$ only.
-Suppressing function arguments and given that $a < 0$, we are interested in the cases where
-$$
-\frac{-b - \sqrt{b^2 - 4ac}}{2a} \ge 1,
-$$
-which, after some algebra, is equivalent to
-%-b - \sqrt{b^2 - 4ac} \le 2a,
-%-b - 2a \le \sqrt{b^2 - 4ac},
-%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
-%a^2 + ab + ac \le 0,
-$$
-a + b + c \ge 0.
-$$
-Simplifying, this is
-$$
--\gamma^2(\hat{x}', \hat{y}')(1 + \frac{1}{2}\beta(x, y)) + 2\gamma(\hat{x}', \hat{y}')\beta(\hat{x}', \hat{y}')xy^2 + \beta^2(\hat{x}', \hat{y}') \ge 0.
-$$
-If we can find
+\begin{theorem}
+\label{thm:general-bl}
+For
 $$
@@
 $$
-then we are interested in the region where
+then the maximum likelihood value $\hat{w}=1$ when
 $$
--\gamma_{U}^2(x, y)(1 + \frac{1}{2}\beta(x, y)) + 2\gamma_{L}(x, y)\beta_{L}(x, y)xy^2 + \beta_{L}^2(x, y) \ge 0,
+-\gamma_{U}^2(x, y)(1 + \frac{1}{2}\beta(x, y)) + 2\gamma_{L}(x, y)\beta_{L}(x, y)xy^2 + \beta_{L}^2(x, y) \ge 0.
 $$
-an inequality involving only $x$ and $y$.
+\end{theorem}
+Similar to Theorem~\ref{thm:restricted-bl}, the inequality in Theorem~\ref{thm:general-bl} is also a function of only $x$ and $y$.
 The loosest possible bounds are $\gamma_{L} = 0$, $\beta_{L} = 1$ and $\gamma_{U} = 4$, giving
@@
 The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma(x, y)$ and $\beta_{L} = \beta(x, y)$, which is the case in Figure~\ref{fig:bl-inconsistency}.
-
-Say we are willing to tolerate a specified level of error in our estimates for $\hat{x}'$ and $\hat{y}'$ so that
-$$
-x-x_{L} \le \hat{x}' \le x+x_{U}
-$$
-and
-$$
-y-y_{L} \le \hat{y}' \le y+y_{U}.
-$$
-As an example, let this error be $.1$ in either direction, i.e., $x_{L}=x_{U}=y_{L}=y_{U}=.1$.
+As an example middle case, let the error in estimating both $\hat{x}'$ and $\hat{y}'$ be $.1$ in either direction, i.e., $x-.1\le\hat{x}'\le x+.1$ and $y-.1\le\hat{y}'\le y+.1$.
 In this case, the region to the left side of the curve in Figure~\ref{fig:bl-loose-inconsistency} will either estimate $\hat{w}'$ to be one or estimate either $\hat{x}'$ or $\hat{y}'$ to be more than $.1$ away from its true value.
@@
 
-\subsubsection{Empirical results}
+\subsubsection{Empirical validation}
 
@@
 Since empirical estimates can be somewhat unstable, we perform two maximizations---one over $0 \le x',y',w' \le 1$ and one over $0 \le x',y' \le 1$ with $w'=1$---and take the value of $\hat{w}'$ with the higher likelihood.
-We compute these maxima over a lattice in steps of $1e-3$ for $x$ and $y$ from $1e-3$ to $1-1e-3$---we do not include zero or one in our lattice to further stabilize the fits, recalling from Section~\ref{sec:restricted-case} that cases where $x$ or $y$ are equal to zero or one can result in pathologies.
+We compute these maxima over a lattice in steps of $10^{-3}$ for $x$ and $y$ from $10^{-3}$ to $1-10^{-3}$---we do not include zero or one in our lattice to further stabilize the fits, recalling from Section~\ref{sec:restricted-case} that cases where $x$ or $y$ are equal to zero or one can result in pathologies.
 From this process we obtain Fig.~\ref{fig:bl-general-inconsistency}.
@@
 }
-\label{fig:bl-general-inconsistnecy}
+\label{fig:bl-general-inconsistency}
 \end{figure}

### 2017-11-14 old_thing.tex
substantial restructure

\section{Ancestral state reconstruction using maximum likelihood}

We consider the standard phylogenetic likelihood on unrooted trees \cite{Felsenstein2004}.
We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
We observe $\nCols$ samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
The corresponding unobserved ancestral states are $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ and distributed as $\ancestralStateColumnRV$.

\subsection{Classical maximum-likelihood inference in phylogenetics}

For a topology $\tau$ and branch lengths $t$, we make the usual assumption of independence between sites, obtaining the likelihood
\begin{equation}
\label{eq:full_likelihood}
%EM given that the H is on the lhs of the conditioning, does it make sense to have it on the rhs of the semicolon? I don't know the convention.
%das: I forgot that standard likelihoods are written with the parameters on the left.
%Since H is like a hidden variable, I went for an expectation-maximization notation, grouping them with the observations.
%It seems a little wonky, though, when we start having maximum likelihood ancestral states depending on \tau,t.
%Oh well...
L_\nCols(\tau, t; \fullAlignment,\fullAncestralStates) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \argmax_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across all possible ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\tau, t; \fullAlignment) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \  \tilde{L}_\nCols(\tau, t; \fullAlignment).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

\subsection{Joint maximization}

An alternative approach does away with the marginalization and directly estimates the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
One can compute the profile likelihood
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates),
\end{equation}
then estimate the topology and branch lengths via
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment),
\end{equation}
using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and \eqref{eq:profile_likelihood_topology_bl} recovers the joint ML values of the unknown parameters.
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.

%EM It seems straightforward to come up with an example of where branch length estimation is inconsistent. It seems like any time a state at an internal node is uncertain we will get different branch lengths by fixing it to the ML state.

\subsection{Site split formulation}

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation (we follow Section 8.6 of \cite{Semple2003-em}).
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
Indeed, we start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.

For topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization of $\ancestralStateColumn_i$.
In this case the entire set of internal nodes must be enumerated: the probability of observing an ancestral state split, conditional on a site split, is not invariant to taking its complement.

We enumerate the site splits $\siteSplit_j$ of which there are $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ in total where $\mathcal{P}$ denotes the power set.
Similarly we enumerate ancestral splits $\ancestralSplit_k$ of which there are $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$ in total.

This split formulation defines mappings from patterns to splits as
 $$
 \patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet).
 $$
Given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
$$
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
$$
that takes corresponding realizations $\siteSplit_j$ for some $j$.

For the $i$th factor of \eqref{eq:full_likelihood},
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
\begin{align*}
    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$.
For the other term, we define
$$
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
as a mapping of ancestral states and tip states to ancestral state splits.
It is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.
We define
$$
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
$$
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, so that
%By considering the case requiring complementing and that not requiring complementing separately,
%das: did I write the above? I don't remember the significance of it...
\begin{align*}
    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this have
\begin{align*}
\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.

\paragraph{Example}
\begin{figure}
    \centering
    \includegraphics[width=.45\textwidth]{unrooted_farris_tree}
    \caption{Four taxon ``Farris zone'' tree.}
\label{fig:four-taxa-tree}
\end{figure}

Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}---this is commonly known as the ``Farris zone'' topology.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{3\}$ and internal node $\{2\}$ connected to leaves $\{2\}$ and $\{4\}$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, p)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, p)$ we have
$$
P(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_5=\{1,2\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^3(1-p)^2.
$$
Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, p)$,
$$
P(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3.
$$

\subsection{Showing inconsistency}

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Let
$$
L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t)) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the limit of infinite data,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
\end{equation}
For data generated from $(\tau^*, t^*)$, we have an inconsistency in estimating topology if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\tau', \hat{t}_1; \fullAncestralSplitPartitions(\tau', \hat{t}_1)) > \ell_{\tau^*,t^*}(\tau^*, \hat{t}_2; \fullAncestralSplitPartitions(\tau^*, \hat{t}_2))
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\fullAncestralSplitPartitions(\tau', \hat{t}_1)\}$ and $\{\hat{t}_2,\fullAncestralSplitPartitions(\tau^*, \hat{t}_2)\}$ estimated by maximizing $\tilde{\ell}_{\tau^*,t^*}$ with respect to the branch lengths and the internal states for $\tau = \tau'$ and $\tau^*$, respectively.
By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
In the next section, we show that such examples exist.

\section{Inconsistency of joint maximum likelihood in estimating topology}

\subsection{Bounding the likelihoods}

Our approach seeks bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified}.
We obtain
$$
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t}))
$$
as an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
$$
Similarly, we have the lower bound
$$
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t})).
$$

\subsection{Parameter setting}

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_blank}
\caption[short]{Farris topology $\tau_1$}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{felsenstein_blank}
\caption[short]{Felsenstein topology $\tau_2$}
\end{subfigure}
\caption{Two simple topologies}
\label{fig:farris-fels-top}
\end{figure}

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Fig.~\ref{fig:farris-fels-top}.
We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform (see either 8.6 of Semple and Steel \cite{Semple2003-em} or Appendix (TODO)).

Let $\{x',y',w'\}\in[0,1]^3$ be three fidelities and $\{p_x',p_y',p_w'\}\in[0,1/2]^3$ the corresponding substitution probabilities.
Call $t=\{x',y',w'\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ constrains the bottom three branches to share the same parameter, the classical construction of this topology.

Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ with edge ordering in the order of the taxa, then the internal branch last.
We now show that, in the case of the Farris tree topology, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
Using the Hadamard transform approach, we calculate the generating probabilities on the Farris tree topology.
For site split $\emptyset$,
\begin{align*}
    P(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities will differ only in the signs of each term.
For example, for site split $\{1\}$ we have
\begin{align*}
    P(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    P(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
\begin{align*}
    P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.

Since exchanging $x_1$ and $x_2$ does not change the value of the log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
The analogous statement holds for $y_1$ and $y_2$.
The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies under these two topologies.
Table~\ref{tab:likelihoods} contains calculations of likelihood values for fixed site patterns and topologies that have been maximized over ancestral state patterns.
Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} contain the values of the likelihood for all possible ancestral states.

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$    &$1+x'^2+y'^2+4x'y'w'+x'^2y'^2$&$1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{1\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{2\}$        &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    $\{3\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{1,2\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,3\}$      &$1+x'^2+y'^2-4x'y'w'+x'^2y'^2$&$1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{2,3\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,2,3\}$    &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \multicolumn{3}{c}{$\tau=\tau_1$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$&
    $\emptyset$&
    $(1+x')^2   (1+w')(1+y')^2$\\
     $\{1\}$    &
    $\emptyset$&
    $(1+x')(1-x')(1+w')(1+y')^2$\\
     $\{2\}$    &
    $\emptyset$&
    $(1+x')^2   (1+w')(1+y')(1-y')$\\
     $\{3\}$    &
    $\emptyset$&
    $(1+x')(1-x')(1+w')(1+y')^2$\\
    $\{1,2\}$  &
    $\{\emptyset,\{1,2\}\}$&
    $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
    $\{1,3\}$  &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x')^2   (1+w')(1+y')^2\\
                    (1+x')^2   (1-w')(1+y')^2\\
                    (1+x')^2   (1+w')(1-y')^2
                \end{array}$\\
    $\{2,3\}$  &
                $\{\emptyset,\{1,2\}\}$&
                $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
    $\{1,2,3\}$&
                $\{1,2\}$&
                $(1+x')^2   (1+w')(1+y')(1-y')$\\
    \hline
    \multicolumn{3}{c}{$\tau=\tau_2$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$       &$\emptyset$&$(1+x')^2   (1+w')(1+y')^2$\\
    $\{1\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                        (1+x')(1-x')(1+w')(1+y')^2\\
                        (1+x')^2   (1-w')(1+y')(1-y')
                    \end{array}$\\
      $\{2\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')^2   (1+w')(1+y')(1-y')\\
                    (1+x')(1-x')(1-w')(1+y')^2
                    \end{array}$\\
      $\{3\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1+w')(1+y')^2\\
                    (1+x')^2   (1-w')(1+y')(1-y')
                    \end{array}$\\
     $\{1,2\}$         &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
                    (1+x')^2   (1-w')(1+y')^2
                    \end{array}$\\
     $\{1,3\}$         &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{\{1\},\{2\}\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x')^2   (1+w')(1+y')^2\\
                    (1+x')(1-x')(1-w')(1+y')(1-y')\\
                    (1+x')^2   (1+w')(1-y')^2
                    \end{array}$\\
      $\{2,3\}$        &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
                    (1-x')^2   (1-w')(1+y')^2\\
                    (1+x')^2   (1-w')(1-y')^2
                    \end{array}$\\
     $\{1,2,3\}$       &
    $\left\{\begin{array}{l}
                    \{2\} \\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1-w')(1+y')^2 \\
                    (1+x')^2   (1+w')(1+y')(1-y')
                    \end{array}$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and ancestral state partitions.
All values multiplied by $1/32$.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
See Appendix for full calculations.}
\label{tab:likelihoods}
\end{table}

\subsubsection{Upper bound for Farris tree likelihood}

Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
For compactness, define
$$
p_{\emptyset} = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})
$$
with similar definitions for the remaining generating probabilities.
The full Farris tree log likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see Table~\ref{tab:likelihoods} for $\tau = \tau_1$).
We begin by writing out the middle case: for the ancestral state split $\{1\}$.
Directly substituting in \eqref{eq:site_pattern_likelihood}, the log likelihood is
\begin{align}
    \label{eq:farris_likelihood}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \nonumber \\
    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\
    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \nonumber \\
    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\
    &\qquad - \log(8) \nonumber \\
    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \nonumber \\
    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \nonumber \\
    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\
    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \nonumber \\
    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\
    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
\end{align}
Seeking an upper bound, we use the facts that, for $x',y'\in[0,1]$,
\begin{align*}
1-x'^2+y'^2-x'^2y'^2 & = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y') \\
1+x'^2-y'^2-x'^2y'^2 & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y') \\
1+x'^2+y'^2+x'^2y'^2 & = (1+x'^2)(1+y'^2) \le (1+x')(1+y') \\
4x'y' & = 2x' \cdot 2y' \le (1+x'^2)(1+y'^2) \le (1+x')(1+y').
\end{align*}
The term $p_{\emptyset}$ is bounded as
\begin{align*}
    p_{\emptyset} & = \log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
                  & \le \log(1+x'^2+y'^2+4x'y'+x'^2y'^2) \\
                  & \le \log(2(1+x')(1+y'))
\end{align*}
and $p_{13}$ is bounded as
\begin{align*}
    p_{13} & = \log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
                  & \le \log(1+x'^2+y'^2+x'^2y'^2) \\
                  & \le \log((1+x')(1+y')).
\end{align*}
Factoring and making these substitutions results in
\begin{align*}
&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
&\qquad \le      p_{\emptyset}  \cdot\log(2(1+x')(1+y'))
+ p_{1}          \cdot\log((1+x')(1-x')(1+y')) \\
    &\qquad\qquad + p_{2}          \cdot\log((1+x')(1+y')(1-y'))
+ p_{3}          \cdot\log((1+x')(1-x')(1+y')) \\
    &\qquad\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+y')(1-y'))
+ p_{13}         \cdot\log((1+x')(1+y')) \\
    &\qquad\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+y')(1-y'))
+ p_{123}        \cdot\log((1+x')(1+y')(1-y'))
- \log(8) \\
    &\qquad\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2)
+ p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y'))
+ p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y'))
+ p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
    &\qquad\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y'))
+ p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
&\qquad =      p_{\emptyset}  \cdot\log(2)
+ \log(1+x')
+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
&\qquad\qquad + \log(1+y')
+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
&\qquad\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x')
+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
&\qquad\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y')
+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
&\qquad\qquad + (1-p_{13})\cdot\log(1+w')
+ p_{13}\cdot\log(1-w')
- \log(8) \\
&\qquad =      p_{\emptyset}  \cdot\log(2)
+ (3-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x')
+ 2(p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
&\qquad\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y')
+ 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
&\qquad\qquad + (1-p_{13})\cdot\log(1+w')
+ p_{13}\cdot\log(1-w')
- \log(8).
\end{align*}
To make matters further compact, let
\begin{equation}
    \begin{aligned}
        a_{1} &= 3-p_{1}-p_{3}-p_{12}-p_{23}, \\
        a_{2} &= 2(p_{1}+p_{3}+p_{12}+p_{23}), \\
        a_{3} &= 3-p_{2}-p_{12}-p_{23}-p_{123}, \\
        a_{4} &= 2(p_{2}+p_{12}+p_{23}+p_{123}), \\
    \end{aligned}
    \label{eq:a_const}
\end{equation}
where
\begin{equation*}
\begin{split}
&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
&\qquad \le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log(1+x')
+ a_{2}\cdot\log(1-x')
+ a_{3}\cdot\log(1+y') \\
&\qquad\qquad + a_{4}\cdot\log(1-y')
+ (1-p_{13})\cdot\log(1+w')
+ p_{13}\cdot\log(1-w')
- \log(8).
\end{split}
\end{equation*}
Maximizing over the unknown terms yields
$$
\hat{x}' = \frac{a_{1}-a_{2}}{a_{1}+a_{2}}, \ \hat{y}' = \frac{a_{3}-a_{4}}{a_{3}+a_{4}}.
$$
We check that $0 \le \hat{x}' \le 1$ and $0 \le \hat{y}' \le 1$ so these maximizers are valid.
Letting
$$
\tilde{p} = p_{1}+p_{3}+p_{12}+p_{23}
$$
we have
$$
\hat{x}' = \frac{3-\tilde{p} - 2\tilde{p}}{3-\tilde{p}+2\tilde{p}} = \frac{3-3\tilde{p}}{3+\tilde{p}},
$$
and, since $0 \le \tilde{p} \le 1/2$, $0 \le \hat{x}' \le 1$.
The same approach works for $\hat{y}'$.
The upper bound for the likelihood is then maximized at
\begin{align*}
&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}}
+ a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
&\qquad\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ (1-p_{13})\cdot\log(2(1-p_{13}))
+ p_{13}\cdot\log(2p_{13})
- \log(8) \\
&\qquad := L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
\end{align*}

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}. \\
    \end{aligned}
    \label{eq:a_const_prime}
\end{equation}
We also now need to check that
$$
\hat{x}' = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y}' = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
$$
are both between zero and one.
It will suffice to show
$$
0 \le \tilde{p} - 2p_{13} \le 1
$$
and
$$
0 \le \tilde{p} + 2p_{13} \le 1
$$
for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
Using Table~\ref{tab:sitepatprob},
$$
p_{13} = \frac{1}{8}\left(1 + x^2 + y^2 - 4xy^2 + x^2y^2\right).
$$
We see that
$$
1 + x^2 + y^2 - 4xy^2 + x^2y^2 \le 2 \iff y^2\left(1 - 4x + x^2\right) \le 1 - x^2
$$
and
$$
y^2\left(1 - 4x + x^2\right) \le 1 - 4x + x^2 \le 1 - x^2,
$$
the last inequality holding if $x \le 2$.
Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}'$ and $\hat{y}'$ are valid.

To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
\begin{align*}
&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ \log(2)
- \log(8) \\
&\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}),
\end{align*}
and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
\begin{align*}
&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
+ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
+ \log(2)
- \log(8) \\
&\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
\end{align*}
All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.

\subsubsection{Lower bound for Felsenstein tree partial likelihood}

We construct a similar lower bound on the Felsenstein tree partial likelihood.
In the Felsenstein case, there are various optimal internal states depending on branch lengths (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).
We can bound these below replacing $1+w$ with $1-w$ as follows.
For the site split $\{1,2\}$, we would like to find a lower bound for
$$
\max\left((1+x')^2 (1-w')(1+y')^2, (1+x')(1-x')(1+w')(1+y')(1-y')\right).
$$
We observe that
$$
(1+x')^2 (1-w')(1+y')^2 > (1+x')(1-x')(1-w')(1+y')(1-y')
$$
always, and thus the maximum can be bounded below by $(1+x')^2 (1-w')(1+y')^2$.
For the other site splits we consider two cases.
If $(1+x')(1-y') > (1-x')(1+y')$, then
$$
(1+x')^2(1-w')(1+y')(1-y') > (1+x')(1-x')(1-w')(1+y')^2
$$
meaning the cases $\{1\}, \{2\}, \{3\}$ and $\{1,2,3\}$ will have likelihoods bounded below by
$$
(1+x')^2(1-w')(1+y')(1-y').
$$
For the cases $\{1,3\}$ and $\{2,3\}$ and the same condition, we have
$$
(1+x')^2(1-w')(1-y')^2 > (1-x')^2(1-w')(1+y')^2
$$
and
$$
(1+x')^2(1-w')(1-y')^2 > (1+x')(1-x')(1-w')(1+y')(1-y'),
$$
yielding lower bounds of
$$
(1+x')^2(1-w')(1-y')^2
$$
in these cases.
The full likelihood in this case is then bounded below by
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      p_{\emptyset}  \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
    &\qquad + p_{1}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{2}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{3}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{12}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
    &\qquad + p_{123}        \cdot\log((1+x')^2   (1-w')(1+y')(1-y'))\\
    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1-y')^2) \\
    &\qquad + p_{23}         \cdot\log((1+x')^2   (1-w')(1-y')^2).
\end{align*}
When $(1+x')(1-y') < (1-x')(1+y')$, similar arguments yield the lower bound
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      p_{\emptyset}  \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{2}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    %See above.
    &\qquad + p_{12}         \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
    &\qquad + p_{123}        \cdot\log((1+x')(1-x')(1-w')(1+y')^2)\\
    &\qquad + p_{13}         \cdot\log((1-x')^2    (1-w')(1+y')^2) \\
    &\qquad + p_{23}         \cdot\log((1-x')^2    (1-w')(1+y')^2).
\end{align*}
Letting
$$
b = p_{1}+p_{2}+p_{3}+p_{123}
$$
yields
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      2\cdot\log(1+x') \\
    &\qquad + (2-b)  \cdot\log(1+y') \\
    &\qquad + b      \cdot\log(1-y') \\
    &\qquad + \log(1-w')
\end{align*}
for the first lower bound, and we get the equivalent expression for the second lower bound by exchanging $x'$ and $y'$.
Maximizing either lower bound over $\{x',y',w'\}$, we obtain
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      2\cdot\log(2) \\
    &\qquad + (2-b)  \cdot\log(2-b) \\
    &\qquad + b      \cdot\log(b)
\end{align*}
%EM Shouldn't this be 2p_{\emptyset}-1 ? It doesn't matter for the bound, but...
%das If we call p_{\emptyset} just p, we're maximizing
%   p log (1 + w) + (1 - p) log (1 - w).
%As before, a_1 log(1+x) + a_2 log(1-x) is maximized at x = (a_1 - a_2) / (a_1 + a_2).
%So then w = (p - (1 - p)) / (p + (1 - p)) = (2p - 1) / 1 = 2p - 1.
%So you're right. Good catch.
%
%Unfortunately, I realized the above does not work for us since 2p - 1 is not guaranteed to be between 0 and 1, as x should be (see the stuff I added justifying this above).
%Rather than do a constrained optimization, I just bounded the p_{\emptyset}\log(1+w) term below by p_{\emptyset}\log(1-w).
i.e.\ the maximum is achieved at $\{1,1-b,0\}$.
Our lower bound is
\begin{align*}
    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,0\}) \\
                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b).
\end{align*}

\subsubsection{Region of inconsistency}

We want the region where
$$
L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) \ge \max(L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}), L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}),L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy})).
$$
Since this is an inequality in two variables, we plot them analytically---see Fig.~\ref{fig:inconsistency-farris}.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
\includegraphics[width=.9\textwidth]{topology-inconsistency-inkscape}
\caption{
    Regions of inconsistency.
    Due to the looseness of the upper and lower bounds, the parameters in the top right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

\section{Inconsistency of joint maximum likelihood in estimating branch fidelities}

We will now consider the problem of branch fidelity estimation on the correct tree using joint ancestral state and branch fidelity estimation.
We will consider two settings on the Farris topology, corresponding to whether some branch fidelities are known ahead of time or not.
As above, we will assume that data is generated with two sister branches of fidelity $x$, with all other branches of length $y$.
We will show that, for a nontrivial subset of possible values for $x$ and $y$, the interior branch fidelity parameter $w'$ will be consistently overestimated as exactly equal to one instead of its true value of $y$.
That is, for the restricted case, with
$$
\hat{w}' = \arg\max_{w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x,y,w'\}; \fullAncestralSplitPartitions(\tau_1,\{x,y,w'\})),
$$
we show there are values for $x$ and $y$ where $\hat{w}'\equiv 1$.
In the general case, we do the same but with
$$
(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x',y',w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})).
$$
%EM We will show XXX... and introduce \hat{w'} here.
%das: updated

\subsection{Restricted case: fixed $x'$ and $y'$}
\label{sec:restricted-case}

First, assume we know the true values of $x'$ and $y'$ but not $w'$.
In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
Recalling Table~\ref{tab:likelihoods}, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood for $(x',y',w')$.
%EM You are talking about a log likelihood here, right? So a likelihood being proportional to means that there is some additive constant. Makes things a little confusing.
%das: updated
If $x'$ and $y'$ are fixed, by collapsing like terms and ignoring terms with only $x'$ and $y'$, the log-likelihood in \eqref{eq:farris_likelihood} as a function of $w'$ is
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\sim     p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
    &\qquad + p_{13}\cdot\log(1-w')
\end{align*}
where $\sim$ denotes equality up to an additive constant.
Similarly, the other two possibilities for the likelihood both satisfy
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\sim     p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log(1+w').
\end{align*}
Since
$$
\log(1+w') \geq (1 - p_{13})\cdot\log(1+w') + p_{13}\cdot\log(1-w')
$$
with equality only when $w'=0$, the second likelihood is the likelihood with the maximum-likelihood ancestral state, and so we need only analyze that case.

Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
Simplifying, the likelihood can be written
$$
\ell(w') = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
$$
where
$$
\alpha_1 := \alpha_1(x, y) = \frac{1}{8} (1+x^2+y^2+4xy^2+x^2y^2),
$$
$$
\alpha_2 := \alpha_2(x, y) = \frac{1}{8}(1+x^2+y^2-4xy^2+x^2y^2),
$$
$$
\beta := \beta(x, y) = 1+x^2+y^2+x^2y^2,
$$
$$
\gamma := \gamma(x, y) = 4xy.
$$
Since $x$ and $y$ fall between zero and one, we make use of the inequalities (excepting the cases of $x=y=0$ and $x=y=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
%EM Rather confusing to be using ' for two things here. What about using a dot over the l? In retrospect something other than ' would have been good because of the proliferation of w^{'2} etc. Prob not worth changing now though.
The derivative of the log likelihood is
$$
\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} - \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
$$
The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w') \lessgtr 0$, we equivalently consider $f(w') \lessgtr 0$ where $f$ is the quadratic function
\begin{align*}
f(w') &= (w')^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w'\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
\end{align*}
% Note from das: b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero. c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
Excepting all the cases where $x=0$ or where $y=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w'=0$.
This implies that if the smaller of the zeros of $f(w')$ is greater than one, then $\hat{w}' \equiv 1$.
Using the quadratic formula with
$$
a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)},
$$
which is a function of the generating parameters $x$ and $y$.
We plot the region where $\hat{w}'(x, y) \ge 1$ in Figure~\ref{fig:bl-inconsistency}.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
\includegraphics[width=.9\textwidth]{branch-length-inconsistency-inkscape}
\caption{Region of inconsistency with $x$ and $y$ fixed}
\label{fig:bl-inconsistency}
\end{figure}

\subsection{General case}

\subsubsection{Loose bounds}

In the general case, $\hat{w}'$ is a function of $x$, $y$, $\hat{x}'$, and $\hat{y}'$.
Looking to the previous section, $\hat{w}'$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}'$ and $\hat{y}'$ instead of $x'$ and $y'$.
Assume we know $\hat{x}'$ and $\hat{y}'$ as functions of $x$ and $y$ only.
Suppressing function arguments and given that $a < 0$, we are interested in the cases where
$$
\frac{-b - \sqrt{b^2 - 4ac}}{2a} \ge 1,
$$
which, after some algebra, is equivalent to
%-b - \sqrt{b^2 - 4ac} \le 2a,
%-b - 2a \le \sqrt{b^2 - 4ac},
%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
%a^2 + ab + ac \le 0,
$$
a + b + c \ge 0.
$$
Simplifying, this is
$$
-\gamma^2(\hat{x}', \hat{y}')(1 + \frac{1}{2}\beta(x, y)) + 2\gamma(\hat{x}', \hat{y}')\beta(\hat{x}', \hat{y}')xy^2 + \beta^2(\hat{x}', \hat{y}') \ge 0.
$$
If we can find
$$
\gamma_{L}(x, y) \le \gamma(\hat{x}', \hat{y}') \le \gamma_{U}(x, y)
$$
and
$$
\beta_{L}(x, y) \le \beta(\hat{x}', \hat{y}')
$$
then we are interested in the region where
$$
-\gamma_{U}^2(x, y)(1 + \frac{1}{2}\beta(x, y)) + 2\gamma_{L}(x, y)\beta_{L}(x, y)xy^2 + \beta_{L}^2(x, y) \ge 0,
$$
an inequality involving only $x$ and $y$.
The loosest possible bounds are $\gamma_{L} = 0$, $\beta_{L} = 1$ and $\gamma_{U} = 4$, giving
$$
-16(1 + \frac{1}{2}\beta(x, y)) + 1 \ge 0,
$$
which is never true.
The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma(x, y)$ and $\beta_{L} = \beta(x, y)$, which is the case in Figure~\ref{fig:bl-inconsistency}.

Say we are willing to tolerate a specified level of error in our estimates for $\hat{x}'$ and $\hat{y}'$ so that
$$
x-x_{L} \le \hat{x}' \le x+x_{U}
$$
and
$$
y-y_{L} \le \hat{y}' \le y+y_{U}.
$$
As an example, let this error be $.1$ in either direction, i.e., $x_{L}=x_{U}=y_{L}=y_{U}=.1$.
In this case, the region to the left side of the curve in Figure~\ref{fig:bl-loose-inconsistency} will either estimate $\hat{w}'$ to be one or estimate either $\hat{x}'$ or $\hat{y}'$ to be more than $.1$ away from its true value.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --general-branch-lengths --delta .001 --plot-name figures/bl-loose-inconsistency.svg
\includegraphics[width=.9\textwidth]{bl-loose-inconsistency-inkscape}
\caption{Region of inconsistency with $x-.1 \le \hat{x}' \le x+.1$ and $y-.1 \le \hat{y}' \le y+.1$.}
\label{fig:bl-loose-inconsistency}
\end{figure}

\subsubsection{Empirical results}

The region of inconsistency in Figure~\ref{fig:bl-loose-inconsistency} is potentially a conservative estimate of where $\hat{w}=1$.
To determine how conservative, we use the method of differential evolution \cite{Storn1997} to perform joint maximum likelihood estimation.
Since empirical estimates can be somewhat unstable, we perform two maximizations---one over $0 \le x',y',w' \le 1$ and one over $0 \le x',y' \le 1$ with $w'=1$---and take the value of $\hat{w}'$ with the higher likelihood.
We compute these maxima over a lattice in steps of $1e-3$ for $x$ and $y$ from $1e-3$ to $1-1e-3$---we do not include zero or one in our lattice to further stabilize the fits, recalling from Section~\ref{sec:restricted-case} that cases where $x$ or $y$ are equal to zero or one can result in pathologies.
From this process we obtain Fig.~\ref{fig:bl-general-inconsistency}.

The estimates for $\hat{w}'$ when $x$ and $y$ are both large seem to be unbiased, with $\hat{w}'$ increasing as $y$ increases, though the region where $\hat{w}'$ is estimated to be one---indicating no change along that particular branch---is similar to that in the restricted case.
Although the restricted and general case are similar, comparing Figures~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case has a slightly different character, possibly due to potential looseness of the bound in the restricted case.
%das: TODO double check this; do the same empirical plot for the restricted case
This region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have a region of the space of true parameters where we will always estimate the interior branch parameter to be one.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .001 --plot-name figures/w-hat-empirical-001.svg --pkl-name figures/w-hat-empirical-001.pkl --n-jobs 16
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-001.svg --in-pkl-name figures/w-hat-empirical-001.pkl
% if fit already
\includegraphics[width=.9\textwidth]{w-hat-empirical-001}
\caption{
    %das: note to update caption
    Estimating $(\hat{x}', \hat{y}', \hat{w}')$.
}
\label{fig:bl-general-inconsistency}
\end{figure}

\section{Discussion}

Neyman-Scott paradox.

Interesting that here we are simulating on the Farris tree and end up with the Felsenstein tree.
For maximum parsimony it's the opposite \cite{Felsenstein1978-rr}.
%EM We should be clear here that thus this is not "long branch attraction". I haven't read the Farris papers, but I wonder what his argument is and if it relates to this, in that there are too many parameters. Probably something should go in the intro as well pointing out that this is not LBA.
%das: Noted. I will work on this.
Discussion of number of parameters of each.

However, note that \cite{Siddall1998-hq} get things going the same way, although \cite{Swofford2001-hr} show that the problem is that they didn't simulate long enough sequences.

### 2017-11-14 joint_inf.tex
update caption and add description

@@
 \begin{theorem}
-There exist $L_0$ and $L_1$ such that
+There exist $L_0(x, y)$ and $L_1(x, y)$ such that
 $$
-\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \le L_0,
+\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \le L_0(x, y),
 $$
 $$
-\max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t; \fullAncestralSplitPartitions(\tau_2,t)) \ge L_1
+\max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t; \fullAncestralSplitPartitions(\tau_2,t)) \ge L_1(x, y)
 $$
-and $L_1 > L_0$.
+and $L_1(x, y) > L_0(x, y)$.
 \end{theorem}
+Fig.~\ref{fig:inconsistency-farris} shows the region where $L_1(x, y) > L_0(x, y)$.
 
@@
     %das: note to update caption
-    Estimating $(\hat{x}', \hat{y}', \hat{w}')$.
+    Region where $\hat{w}'=1$ computing empirical estimates of $(\hat{x}', \hat{y}', \hat{w}')$.
 }

### 2017-11-14 joint_inf.tex
Simplification and prepping for a bio future.

@@
 
-%% Language and font encodings
-\usepackage[english]{babel}
-\usepackage[utf8x]{inputenc}
-\usepackage[T1]{fontenc}
-
-%% Useful packages
 \usepackage{amsmath}
@@
 \usepackage{showlabels}
+\usepackage{palatino}
+\usepackage{hyphenat}
+\usepackage{thm-restate}
 
 \graphicspath{ {figures/} }
+
+% http://bytesizebio.net/2013/03/11/adding-supplementary-tables-and-figures-in-latex/
+\newcommand{\beginsupplement}{%
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{S\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{S\arabic{figure}}%
+     }
 
@@
 
-Existing consistency proofs for phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
+Existing consistency proofs for maximum-likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
 These proofs do not readily extend to include estimating ancestral states.
@@
 
-%EM clarify that we aren't doing sequence-level simulation somewhere.
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model and imagine data being generated on the four taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the large-data limit.
+To do so, we use a binary symmetric model and imagine data being generated on the four taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the limit of infinite data.
 We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known.
 
-%EM Potential re-outline--
-%
-%EM 1. Describe setup with Fels and Farris trees, and how we are talking about the case of infinite data.
-%EM 2. Describe fidelities
-%EM 3. State theorems
-%EM 4. Plots
-%EM 5. Discussion, already below. This motivates a more thorough investigation under simulation.
+\section*{Phylogenetic maximum likelihood}
 
-\section{Phylogenetic maximum likelihood}
-
-Set up mild notation needed to state likelihood
-
-We consider the standard phylogenetic likelihood on unrooted trees \cite{Felsenstein2004}.
-We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
+We begin by fixing some notation needed to be precise about the notions of likelihood examined in this paper.
+We assume the binary symmetric model, namely with a character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
 Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
@@
 
-\subsection{Phylogenetic parameters}
-
-\subsubsection{Topology}
+We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``Farris'' and ``Felsenstein'' trees (Fig.~\ref{fig:farris-fels-top}).
 
@@
 \centering
+%EM Wouldn't it be better to just parameterize these with x,y,w without primes?
+Also, it looks super pro if we can use tex fonts in this figure too.
+I love the mathjax fonts, which I think you can get https://packages.ubuntu.com/trusty/fonts/fonts-mathjax .
+LMK if you run into trouble.
+The font I use is "MathJax_Math".
 \includegraphics[width=.95\textwidth]{farris_blank}
@@
 
-Consider the fixed, binary four-tip tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}---this is commonly known as the ``Farris zone'' topology.
-
-\subsubsection{Fidelities}
-
 We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
 This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
-Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform (see either 8.6 of Semple and Steel \cite{Semple2003-em} or Appendix (TODO)).
+Fidelities have useful algebraic properties, which mean that the generating probabilities using the Hadamard transform have an especially simple form (see either 8.6 of Semple and Steel \cite{Semple2003-em} or Appendix (TODO)).
 
-\subsection{Classical maximum likelihood}
+\subsection*{Two paths to maximum likelihood}
 
-For a topology $\tau$ and branch lengths $t$, we make the usual assumption of independence between sites, obtaining the likelihood
+The standard phylogenetic likelihood approach on unrooted trees under the usual assumption of independence between sites is as follows.
+For a topology $\tau$ and branch lengths $t$ the likelihood given observed ancestral states $\fullAncestralStates$ is
 \begin{equation}
 \label{eq:full_likelihood}
-%EM given that the H is on the lhs of the conditioning, does it make sense to have it on the rhs of the semicolon? I don't know the convention.
-%das: I forgot that standard likelihoods are written with the parameters on the left.
-%Since H is like a hidden variable, I went for an expectation-maximization notation, grouping them with the observations.
-%It seems a little wonky, though, when we start having maximum likelihood ancestral states depending on \tau,t.
-%Oh well...
-L_\nCols(\tau, t; \fullAlignment,\fullAncestralStates) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
+L_\nCols(\tau, t; \fullAlignment,\fullAncestralStates) = \prod_{i=1}^{\nCols} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
 \end{equation}
-% In particular, we are interested in
-% $$
-% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \argmax_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
-% $$
-% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
-% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
-The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across all possible ancestral states
+The probability $\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)$ is a product of transition probabilities determined by $\fullAlignment$, $\fullAncestralStates$, $\tau$, and $t$ \cite{Felsenstein2004}.
+
+The typical approach to is to maximize the likelihood marginalized across ancestral states
 \begin{equation}
@@
 \end{equation}
-and maximizing over the topology and branch lengths to obtain
-$$
-(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \  \tilde{L}_\nCols(\tau, t; \fullAlignment).
-$$
-% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
-% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.
+to estimate the tree $\tau$ and branch lengths $t$.
 
-\subsection{Joint maximum likelihood}
-
-An alternative approach does away with the marginalization and directly estimates the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
-One can compute the profile likelihood
+An alternative approach used by \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully\hyp{}observed likelihood in \eqref{eq:full_likelihood}.
+This is known as a profile likelihood \cite{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
 \begin{equation}
 \label{eq:profile_likelihood}
-L_\nCols'(\tau, t; \fullAlignment) = \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates),
+L_\nCols'(\tau, t; \fullAlignment) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
 \end{equation}
-then estimate the topology and branch lengths via
+%EM Does this have to be unique? Can we just say "denote an estimate"?
+We will use $\hat{\fullAncestralStates}$ to denote the estimate for $\fullAncestralStates$ obtained by maximization \eqref{eq:profile_likelihood}.
+One can then estimate a topology and branch lengths using this profile likelihood:
 \begin{equation}
 \label{eq:profile_likelihood_topology_bl}
-(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment),
+(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
 \end{equation}
-using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
-Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and \eqref{eq:profile_likelihood_topology_bl} recovers the joint ML values of the unknown parameters.
 In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
@@
 
-%EM It seems straightforward to come up with an example of where branch length estimation is inconsistent.
-It seems like any time a state at an internal node is uncertain we will get different branch lengths by fixing it to the ML state.
+\section*{Inconsistency of joint inference}
 
-\section{Inconsistency of joint inference}
+We can now state our results on inconsistency of joint inference.
+All proofs are deferred to the Appendix.
 
-\subsection{Parameter setting}
-
-Consider the Farris topology with arbitrary branch lengths, i.e.,
+%EM I think that this is a technicality.
+I propose that we just state that we can restrict to the parameterization shown in the figure (with certain branch lengths equal) and that a formal statement and proof is in the appendix.
+Consider the Farris tree but with arbitrary branch lengths, i.e.,
 $$
@@
 with edge ordering in the order of the taxa, then the internal branch last.
-We now show that, in the case of the Farris tree topology, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
+We now show that, in the case of the Farris tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
 
@@
 For $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ and $t=\{x,y,x,y,w\}$,
-$$\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t})) = \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)).$$
+$$
+\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t})) = \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)).
+$$
 \end{lemma}
@@
 
-\subsection{Inconsistency in topology estimation}
+\subsection*{Inconsistency in topology estimation}
 
@@
 We show the following.
-\begin{theorem}
+%EM Here's an example of restatable in action. See the end of the appendix.
+\begin{restatable}{theorem}{topoInconsist}
 There exist $L_0(x, y)$ and $L_1(x, y)$ such that
@@
 and $L_1(x, y) > L_0(x, y)$.
-\end{theorem}
+\end{restatable}
+
 Fig.~\ref{fig:inconsistency-farris} shows the region where $L_1(x, y) > L_0(x, y)$.
@@
 % python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
+%EM Sorry I didn't notice this before, but the font size everywhere in the plots should be at least doubled before submission.
+Fine if you want to wait one more cycle.
+The goal is that it should be the same size as the font in the text.
 \includegraphics[width=.9\textwidth]{topology-inconsistency-inkscape}
@@
 
-\subsection{Inconsistency in branch length estimation}
+\subsection*{Inconsistency in branch length estimation}
 
@@
 $$
-%EM We will show XXX... and introduce \hat{w'} here.
-%das: updated
 
-\subsubsection{Restricted case}
+\subsubsection*{Restricted case}
 
@@
 
-\subsubsection{General case}
+\subsubsection*{General case}
 
@@
 
-\subsubsection{Empirical validation}
+\subsubsection*{Empirical validation}
 
@@
 Since empirical estimates can be somewhat unstable, we perform two maximizations---one over $0 \le x',y',w' \le 1$ and one over $0 \le x',y' \le 1$ with $w'=1$---and take the value of $\hat{w}'$ with the higher likelihood.
-We compute these maxima over a lattice in steps of $10^{-3}$ for $x$ and $y$ from $10^{-3}$ to $1-10^{-3}$---we do not include zero or one in our lattice to further stabilize the fits, recalling from Section~\ref{sec:restricted-case} that cases where $x$ or $y$ are equal to zero or one can result in pathologies.
+%EM XXXX below because no bio journals have numbered sections.
+We compute these maxima over a lattice in steps of $10^{-3}$ for $x$ and $y$ from $10^{-3}$ to $1-10^{-3}$---we do not include zero or one in our lattice to further stabilize the fits, recalling from Section~XXXX that cases where $x$ or $y$ are equal to zero or one can result in pathologies.
 From this process we obtain Fig.~\ref{fig:bl-general-inconsistency}.
@@
 
-\section{Discussion}
+\section*{Discussion}
+
+Note that this two step thing is known in statistics as profile likelihood \cite{Murphy2000-ry}.
+People have worked on its properties.
 
@@
 
+\section*{Acknowledgements}
+We would like to thank Richard Neher for helpful discussions.
+
+David Shaw, Vladimir Minin, and Frederick Matsen were supported by NIH grants U19-AI117891 and R01-GM113246; David Shaw and Frederick Matsen were also supported by R01-AI12096.
+The research of Frederick Matsen was supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation.
+
 \bibliographystyle{plain}
@@
 \newpage
+\beginsupplement
 

### 2017-11-15 appendix.tex
fix up arguments for restricted case; disjoint ideas for general case
a few ideas trying to work through general case
Simplification and prepping for a bio future.
get all content into main doc

@@
-\section{Appendix}
+\section*{Appendix}
 
-\subsection{Likelihood compuations}
+\subsection*{Hadamard transform}
+
+Define: $\mathcal{T}$, $X$, $A$, $n$, $P$ and $\theta$.
+Let
+$$
+\theta(e) = 1-2p(e).
+$$
+Then
+$$
+p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subset X : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e \in P(\mathcal{T}, Y)} \ \theta(e) \right]
+$$
+
+\subsection*{Site split formulation}
+
+Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation (we follow Section 8.6 of \cite{Semple2003-em}).
+Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
+For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
+Indeed, we start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
+Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.
+
+For topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization of $\ancestralStateColumn_i$.
+In this case the entire set of internal nodes must be enumerated: the probability of observing an ancestral state split, conditional on a site split, is not invariant to taking its complement.
+
+We enumerate the site splits $\siteSplit_j$ of which there are $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ in total where $\mathcal{P}$ denotes the power set.
+Similarly we enumerate ancestral splits $\ancestralSplit_k$ of which there are $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$ in total.
+
+This split formulation defines mappings from patterns to splits as
+ $$
+ \patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet).
+ $$
+Given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
+$$
+\siteSplitRV := \patternToSplit(\alignmentColumnRV)
+$$
+that takes corresponding realizations $\siteSplit_j$ for some $j$.
+
+For the $i$th factor of \eqref{eq:full_likelihood},
+$$
+P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
+$$
+As a consequence of assuming a binary symmetric model, taking complements yields
+\begin{align*}
+    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
+                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
+\end{align*}
+for some $j$.
+For the other term, we define
+$$
+\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
+$$
+as a mapping of ancestral states and tip states to ancestral state splits.
+It is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.
+We define
+$$
+\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
+$$
+for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, so that
+%By considering the case requiring complementing and that not requiring complementing separately,
+%das: did I write the above? I don't remember the significance of it...
+\begin{align*}
+    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
+\end{align*}
+Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
+\begin{align*}
+\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
+\end{align*}
+In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
+For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this have
+\begin{align*}
+\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
+\end{align*}
+
+Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
+Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
+\begin{align}
+L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
+                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
+                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
+                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
+                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
+\end{align}
+where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.
+
+\paragraph{Example}
+\begin{figure}
+    \centering
+    \includegraphics[width=.45\textwidth]{unrooted_farris_tree}
+    \caption{Four taxon ``Farris zone'' tree.}
+\label{fig:four-taxa-tree}
+\end{figure}
+
+Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}---this is commonly known as the ``Farris zone'' topology.
+The set of all possible character assignments is
+\begin{align*}
+\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
+                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
+\end{align*}
+where each set indicates the tips assigned the character $1$.
+For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
+Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
+The unique site splits, collapsing complements, are
+\begin{align*}
+    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
+& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
+\end{align*}
+Since we identify character complements, we do not consider the additional splits
+\begin{equation*}
+\begin{split}
+& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
+&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
+\end{split}
+\end{equation*}
+the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
+This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
+$$
+\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
+$$
+Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{3\}$ and internal node $\{2\}$ connected to leaves $\{2\}$ and $\{4\}$.
+The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
+For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
+Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
+We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
+$$
+P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
+$$
+in general.
+
+For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
+A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
+As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
+The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
+$$
+P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
+(1-p)^5,
+$$
+$$
+P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
+P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
+p^3(1-p)^2,
+$$
+$$
+P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
+p^4(1-p).
+$$
+The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, p)=\{\emptyset\}$.
+Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, p)$ we have
+$$
+P(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
+P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
+(1-p)^5.
+$$
+For $\siteSplit_5=\{1,2\}$ we have
+$$
+P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
+P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
+p^2(1-p)^3,
+$$
+$$
+P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
+P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
+p^3(1-p)^2.
+$$
+Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, p)$,
+$$
+P(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
+p^2(1-p)^3.
+$$
+
+\subsection*{Likelihood computations}
+
+Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies under these two topologies.
+Table~\ref{tab:likelihoods} contains calculations of likelihood values for fixed site patterns and topologies that have been maximized over ancestral state patterns.
+Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} contain the values of the likelihood for all possible ancestral states.
+
+\begin{table}
+\centering
+\begin{tabular}{|l|l|l|}
+    \hline
+$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
+    \hline
+    $\emptyset$    &$1+x'^2+y'^2+4x'y'w'+x'^2y'^2$&$1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
+    $\{1\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
+    $\{2\}$        &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
+    $\{3\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
+    $\{1,2\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
+    $\{1,3\}$      &$1+x'^2+y'^2-4x'y'w'+x'^2y'^2$&$1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
+    $\{2,3\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
+    $\{1,2,3\}$    &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
+    \hline
+\end{tabular}
+\caption{Site pattern probabilities.
+All values are multiplied by $1/8$.}
+\label{tab:sitepatprob}
+\end{table}
+
+\begin{table}
+\centering
+\begin{tabular}{|l|ll|}
+    \multicolumn{3}{c}{$\tau=\tau_1$}\\
+    \hline
+    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
+    \hline
+    $\emptyset$&
+    $\emptyset$&
+    $(1+x')^2   (1+w')(1+y')^2$\\
+     $\{1\}$    &
+    $\emptyset$&
+    $(1+x')(1-x')(1+w')(1+y')^2$\\
+     $\{2\}$    &
+    $\emptyset$&
+    $(1+x')^2   (1+w')(1+y')(1-y')$\\
+     $\{3\}$    &
+    $\emptyset$&
+    $(1+x')(1-x')(1+w')(1+y')^2$\\
+    $\{1,2\}$  &
+    $\{\emptyset,\{1,2\}\}$&
+    $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
+    $\{1,3\}$  &
+    $\left\{\begin{array}{l}
+                    \emptyset\\
+                    \{1\}\\
+                    \{1,2\}
+                \end{array}\right.$&
+    $\begin{array}{l}
+                    (1-x')^2   (1+w')(1+y')^2\\
+                    (1+x')^2   (1-w')(1+y')^2\\
+                    (1+x')^2   (1+w')(1-y')^2
+                \end{array}$\\
+    $\{2,3\}$  &
+                $\{\emptyset,\{1,2\}\}$&
+                $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
+    $\{1,2,3\}$&
+                $\{1,2\}$&
+                $(1+x')^2   (1+w')(1+y')(1-y')$\\
+    \hline
+    \multicolumn{3}{c}{$\tau=\tau_2$}\\
+    \hline
+    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
+    \hline
+    $\emptyset$       &$\emptyset$&$(1+x')^2   (1+w')(1+y')^2$\\
+    $\{1\}$          &
+    $\left\{\begin{array}{l}
+                    \emptyset\\
+                    \{1\}
+                \end{array}\right.$&
+    $\begin{array}{l}
+                        (1+x')(1-x')(1+w')(1+y')^2\\
+                        (1+x')^2   (1-w')(1+y')(1-y')
+                    \end{array}$\\
+      $\{2\}$          &
+    $\left\{\begin{array}{l}
+                    \emptyset\\
+                    \{1\}
+                \end{array}\right.$&
+    $\begin{array}{l}
+                    (1+x')^2   (1+w')(1+y')(1-y')\\
+                    (1+x')(1-x')(1-w')(1+y')^2
+                    \end{array}$\\
+      $\{3\}$          &
+    $\left\{\begin{array}{l}
+                    \emptyset\\
+                    \{2\}
+                \end{array}\right.$&
+    $\begin{array}{l}
+                    (1+x')(1-x')(1+w')(1+y')^2\\
+                    (1+x')^2   (1-w')(1+y')(1-y')
+                    \end{array}$\\
+     $\{1,2\}$         &
+    $\left\{\begin{array}{l}
+                    \{\emptyset,\{1,2\}\}\\
+                    \{1\}
+                \end{array}\right.$&
+    $\begin{array}{l}
+                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
+                    (1+x')^2   (1-w')(1+y')^2
+                    \end{array}$\\
+     $\{1,3\}$         &
+    $\left\{\begin{array}{l}
+                    \emptyset\\
+                    \{\{1\},\{2\}\}\\
+                    \{1,2\}
+                \end{array}\right.$&
+    $\begin{array}{l}
+                    (1-x')^2   (1+w')(1+y')^2\\
+                    (1+x')(1-x')(1-w')(1+y')(1-y')\\
+                    (1+x')^2   (1+w')(1-y')^2
+                    \end{array}$\\
+      $\{2,3\}$        &
+    $\left\{\begin{array}{l}
+                    \{\emptyset,\{1,2\}\}\\
+                    \{1\}\\
+                    \{2\}
+                \end{array}\right.$&
+    $\begin{array}{l}
+                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
+                    (1-x')^2   (1-w')(1+y')^2\\
+                    (1+x')^2   (1-w')(1-y')^2
+                    \end{array}$\\
+     $\{1,2,3\}$       &
+    $\left\{\begin{array}{l}
+                    \{2\} \\
+                    \{1,2\}
+                \end{array}\right.$&
+    $\begin{array}{l}
+                    (1+x')(1-x')(1-w')(1+y')^2 \\
+                    (1+x')^2   (1+w')(1+y')(1-y')
+                    \end{array}$\\
+    \hline
+\end{tabular}
+\caption{Likelihood calculations for all site patterns and ancestral state partitions.
+All values multiplied by $1/32$.
+Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
+See Appendix for full calculations.}
+\label{tab:likelihoods}
+\end{table}
 
@@
 \end{table}
+
+\subsection*{Form of the likelihood}
+
+Using the Hadamard transform approach, we calculate the generating probabilities on the Farris tree topology.
+For site split $\emptyset$,
+\begin{align*}
+    P(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
+                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
+                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
+\end{align*}
+and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
+All other generating probabilities will differ only in the signs of each term.
+For example, for site split $\{1\}$ we have
+\begin{align*}
+    P(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
+\end{align*}
+and for site split $\{3\}$ we have
+\begin{align*}
+    P(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
+\end{align*}
+meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
+The corresponding possibilities for the likelihood values are
+\begin{align*}
+    P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
+    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
+    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
+    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
+\end{align*}
+for site split $\{1\}$ and
+\begin{align*}
+        P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
+    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
+    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
+    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
+\end{align*}
+for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.
+
+The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
+The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
+
+Since exchanging $x_1$ and $x_2$ does not change the value of the log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
+The analogous statement holds for $y_1$ and $y_2$.
+The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
+
+%EM Here we are restating.
+\topoInconsist*
+
+Our approach seeks bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified}.
+We obtain
+$$
+\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \le
+    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
+    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t}))
+$$
+as an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
+$$
+\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
+$$
+and
+$$
+\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
+$$
+Similarly, we have the lower bound
+$$
+\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \ge
+    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
+    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t})).
+$$
+
+Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
+For compactness, define
+$$
+p_{\emptyset} = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})
+$$
+with similar definitions for the remaining generating probabilities.
+The full Farris tree log likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see Table~\ref{tab:likelihoods} for $\tau = \tau_1$).
+We begin by writing out the middle case: for the ancestral state split $\{1\}$.
+Directly substituting in \eqref{eq:site_pattern_likelihood}, the log likelihood is
+\begin{align}
+    \label{eq:farris_likelihood}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
+    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \nonumber \\
+    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
+    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\
+    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
+    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
+    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \nonumber \\
+    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
+    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\
+    &\qquad - \log(8) \nonumber \\
+    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \nonumber \\
+    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
+    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \nonumber \\
+    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
+    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\
+    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \nonumber \\
+    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\
+    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
+\end{align}
+Seeking an upper bound, we use the facts that, for $x',y'\in[0,1]$,
+\begin{align*}
+1-x'^2+y'^2-x'^2y'^2 & = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y') \\
+1+x'^2-y'^2-x'^2y'^2 & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y') \\
+1+x'^2+y'^2+x'^2y'^2 & = (1+x'^2)(1+y'^2) \le (1+x')(1+y') \\
+4x'y' & = 2x' \cdot 2y' \le (1+x'^2)(1+y'^2) \le (1+x')(1+y').
+\end{align*}
+The term $p_{\emptyset}$ is bounded as
+\begin{align*}
+    p_{\emptyset} & = \log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
+                  & \le \log(1+x'^2+y'^2+4x'y'+x'^2y'^2) \\
+                  & \le \log(2(1+x')(1+y'))
+\end{align*}
+and $p_{13}$ is bounded as
+\begin{align*}
+    p_{13} & = \log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
+                  & \le \log(1+x'^2+y'^2+x'^2y'^2) \\
+                  & \le \log((1+x')(1+y')).
+\end{align*}
+Factoring and making these substitutions results in
+\begin{align*}
+&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
+&\qquad \le      p_{\emptyset}  \cdot\log(2(1+x')(1+y'))
++ p_{1}          \cdot\log((1+x')(1-x')(1+y')) \\
+    &\qquad\qquad + p_{2}          \cdot\log((1+x')(1+y')(1-y'))
++ p_{3}          \cdot\log((1+x')(1-x')(1+y')) \\
+    &\qquad\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+y')(1-y'))
++ p_{13}         \cdot\log((1+x')(1+y')) \\
+    &\qquad\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+y')(1-y'))
++ p_{123}        \cdot\log((1+x')(1+y')(1-y'))
+- \log(8) \\
+    &\qquad\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2)
++ p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
+    &\qquad\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y'))
++ p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
+    &\qquad\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y'))
++ p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
+    &\qquad\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y'))
++ p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
+&\qquad =      p_{\emptyset}  \cdot\log(2)
++ \log(1+x')
++ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
+&\qquad\qquad + \log(1+y')
++ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
+&\qquad\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x')
++ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
+&\qquad\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y')
++ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
+&\qquad\qquad + (1-p_{13})\cdot\log(1+w')
++ p_{13}\cdot\log(1-w')
+- \log(8) \\
+&\qquad =      p_{\emptyset}  \cdot\log(2)
++ (3-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x')
++ 2(p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
+&\qquad\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y')
++ 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
+&\qquad\qquad + (1-p_{13})\cdot\log(1+w')
++ p_{13}\cdot\log(1-w')
+- \log(8).
+\end{align*}
+To make matters further compact, let
+\begin{equation}
+    \begin{aligned}
+        a_{1} &= 3-p_{1}-p_{3}-p_{12}-p_{23}, \\
+        a_{2} &= 2(p_{1}+p_{3}+p_{12}+p_{23}), \\
+        a_{3} &= 3-p_{2}-p_{12}-p_{23}-p_{123}, \\
+        a_{4} &= 2(p_{2}+p_{12}+p_{23}+p_{123}), \\
+    \end{aligned}
+    \label{eq:a_const}
+\end{equation}
+where
+\begin{equation*}
+\begin{split}
+&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
+&\qquad \le      p_{\emptyset}  \cdot\log(2)
++ a_{1}\cdot\log(1+x')
++ a_{2}\cdot\log(1-x')
++ a_{3}\cdot\log(1+y') \\
+&\qquad\qquad + a_{4}\cdot\log(1-y')
++ (1-p_{13})\cdot\log(1+w')
++ p_{13}\cdot\log(1-w')
+- \log(8).
+\end{split}
+\end{equation*}
+Maximizing over the unknown terms yields
+$$
+\hat{x}' = \frac{a_{1}-a_{2}}{a_{1}+a_{2}}, \ \hat{y}' = \frac{a_{3}-a_{4}}{a_{3}+a_{4}}.
+$$
+We check that $0 \le \hat{x}' \le 1$ and $0 \le \hat{y}' \le 1$ so these maximizers are valid.
+Letting
+$$
+\tilde{p} = p_{1}+p_{3}+p_{12}+p_{23}
+$$
+we have
+$$
+\hat{x}' = \frac{3-\tilde{p} - 2\tilde{p}}{3-\tilde{p}+2\tilde{p}} = \frac{3-3\tilde{p}}{3+\tilde{p}},
+$$
+and, since $0 \le \tilde{p} \le 1/2$, $0 \le \hat{x}' \le 1$.
+The same approach works for $\hat{y}'$.
+The upper bound for the likelihood is then maximized at
+\begin{align*}
+&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
+&\qquad\le      p_{\emptyset}  \cdot\log(2)
++ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
++ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}}
++ a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
+&\qquad\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
++ (1-p_{13})\cdot\log(2(1-p_{13}))
++ p_{13}\cdot\log(2p_{13})
+- \log(8) \\
+&\qquad := L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
+\end{align*}
+
+The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
+Call
+\begin{equation}
+    \begin{aligned}
+        a_{1}' &= a_{1}-2p_{13}, \\
+        a_{2}' &= a_{2}+2p_{13}, \\
+        a_{3}' &= a_{3}-2p_{13}, \\
+        a_{4}' &= a_{4}+2p_{13}. \\
+    \end{aligned}
+    \label{eq:a_const_prime}
+\end{equation}
+We also now need to check that
+$$
+\hat{x}' = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y}' = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
+$$
+are both between zero and one.
+It will suffice to show
+$$
+0 \le \tilde{p} - 2p_{13} \le 1
+$$
+and
+$$
+0 \le \tilde{p} + 2p_{13} \le 1
+$$
+for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
+Using Table~\ref{tab:sitepatprob},
+$$
+p_{13} = \frac{1}{8}\left(1 + x^2 + y^2 - 4xy^2 + x^2y^2\right).
+$$
+We see that
+$$
+1 + x^2 + y^2 - 4xy^2 + x^2y^2 \le 2 \iff y^2\left(1 - 4x + x^2\right) \le 1 - x^2
+$$
+and
+$$
+y^2\left(1 - 4x + x^2\right) \le 1 - 4x + x^2 \le 1 - x^2,
+$$
+the last inequality holding if $x \le 2$.
+Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}'$ and $\hat{y}'$ are valid.
+
+To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
+\begin{align*}
+&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
+&\qquad\le      p_{\emptyset}  \cdot\log(2)
++ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
++ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
+&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
++ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
++ \log(2)
+- \log(8) \\
+&\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}),
+\end{align*}
+and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
+\begin{align*}
+&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
+&\qquad\le      p_{\emptyset}  \cdot\log(2)
++ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
++ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
+&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
++ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
++ \log(2)
+- \log(8) \\
+&\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
+\end{align*}
+All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.
+
+We construct a similar lower bound on the Felsenstein tree partial likelihood.
+In the Felsenstein case, there are various optimal internal states depending on branch lengths (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).
+We can bound these below replacing $1+w$ with $1-w$ as follows.
+For the site split $\{1,2\}$, we would like to find a lower bound for
+$$
+\max\left((1+x')^2 (1-w')(1+y')^2, (1+x')(1-x')(1+w')(1+y')(1-y')\right).
+$$
+We observe that
+$$
+(1+x')^2 (1-w')(1+y')^2 > (1+x')(1-x')(1-w')(1+y')(1-y')
+$$
+always, and thus the maximum can be bounded below by $(1+x')^2 (1-w')(1+y')^2$.
+For the other site splits we consider two cases.
+If $(1+x')(1-y') > (1-x')(1+y')$, then
+$$
+(1+x')^2(1-w')(1+y')(1-y') > (1+x')(1-x')(1-w')(1+y')^2
+$$
+meaning the cases $\{1\}, \{2\}, \{3\}$ and $\{1,2,3\}$ will have likelihoods bounded below by
+$$
+(1+x')^2(1-w')(1+y')(1-y').
+$$
+For the cases $\{1,3\}$ and $\{2,3\}$ and the same condition, we have
+$$
+(1+x')^2(1-w')(1-y')^2 > (1-x')^2(1-w')(1+y')^2
+$$
+and
+$$
+(1+x')^2(1-w')(1-y')^2 > (1+x')(1-x')(1-w')(1+y')(1-y'),
+$$
+yielding lower bounds of
+$$
+(1+x')^2(1-w')(1-y')^2
+$$
+in these cases.
+The full likelihood in this case is then bounded below by
+\begin{align*}
+    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
+    &\ge      p_{\emptyset}  \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
+    &\qquad + p_{1}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
+    &\qquad + p_{2}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
+    &\qquad + p_{3}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
+    &\qquad + p_{12}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
+    &\qquad + p_{123}        \cdot\log((1+x')^2   (1-w')(1+y')(1-y'))\\
+    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1-y')^2) \\
+    &\qquad + p_{23}         \cdot\log((1+x')^2   (1-w')(1-y')^2).
+\end{align*}
+When $(1+x')(1-y') < (1-x')(1+y')$, similar arguments yield the lower bound
+\begin{align*}
+    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
+    &\ge      p_{\emptyset}  \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
+    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
+    &\qquad + p_{2}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
+    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
+    %See above.
+    &\qquad + p_{12}         \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
+    &\qquad + p_{123}        \cdot\log((1+x')(1-x')(1-w')(1+y')^2)\\
+    &\qquad + p_{13}         \cdot\log((1-x')^2    (1-w')(1+y')^2) \\
+    &\qquad + p_{23}         \cdot\log((1-x')^2    (1-w')(1+y')^2).
+\end{align*}
+Letting
+$$
+b = p_{1}+p_{2}+p_{3}+p_{123}
+$$
+yields
+\begin{align*}
+    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
+    &\ge      2\cdot\log(1+x') \\
+    &\qquad + (2-b)  \cdot\log(1+y') \\
+    &\qquad + b      \cdot\log(1-y') \\
+    &\qquad + \log(1-w')
+\end{align*}
+for the first lower bound, and we get the equivalent expression for the second lower bound by exchanging $x'$ and $y'$.
+Maximizing either lower bound over $\{x',y',w'\}$, we obtain
+\begin{align*}
+    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
+    &\ge      2\cdot\log(2) \\
+    &\qquad + (2-b)  \cdot\log(2-b) \\
+    &\qquad + b      \cdot\log(b)
+\end{align*}
+%EM Shouldn't this be 2p_{\emptyset}-1 ? It doesn't matter for the bound, but...
+%das If we call p_{\emptyset} just p, we're maximizing
+%   p log (1 + w) + (1 - p) log (1 - w).
+%As before, a_1 log(1+x) + a_2 log(1-x) is maximized at x = (a_1 - a_2) / (a_1 + a_2).
+%So then w = (p - (1 - p)) / (p + (1 - p)) = (2p - 1) / 1 = 2p - 1.
+%So you're right. Good catch.
+%
+%Unfortunately, I realized the above does not work for us since 2p - 1 is not guaranteed to be between 0 and 1, as x should be (see the stuff I added justifying this above).
+%Rather than do a constrained optimization, I just bounded the p_{\emptyset}\log(1+w) term below by p_{\emptyset}\log(1-w).
+i.e.\ the maximum is achieved at $\{1,1-b,0\}$.
+Our lower bound is
+\begin{align*}
+    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,0\}) \\
+                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b).
+\end{align*}
+
+\restrictedBranchInconsist*
+
+Recalling Table~\ref{tab:likelihoods}, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood for $(x',y',w')$.
+%EM You are talking about a log likelihood here, right? So a likelihood being proportional to means that there is some additive constant. Makes things a little confusing.
+%das: updated
+If $x'$ and $y'$ are fixed, by collapsing like terms and ignoring terms with only $x'$ and $y'$, the log-likelihood in \eqref{eq:farris_likelihood} as a function of $w'$ is
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
+    &\sim     p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
+    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
+    &\qquad + p_{13}\cdot\log(1-w')
+\end{align*}
+where $\sim$ denotes equality up to an additive constant.
+Similarly, the other two possibilities for the likelihood both satisfy
+\begin{align*}
+    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
+    &\sim     p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
+    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
+    &\qquad + \log(1+w').
+\end{align*}
+Since
+$$
+\log(1+w') \geq (1 - p_{13})\cdot\log(1+w') + p_{13}\cdot\log(1-w')
+$$
+with equality only when $w'=0$, the second likelihood is the likelihood with the maximum-likelihood ancestral state, and so we need only analyze that case.
+
+Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
+Simplifying, the likelihood can be written
+$$
+\ell(w') = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
+$$
+where
+$$
+\alpha_1 := \alpha_1(x, y) = \frac{1}{8} (1+x^2+y^2+4xy^2+x^2y^2),
+$$
+$$
+\alpha_2 := \alpha_2(x, y) = \frac{1}{8}(1+x^2+y^2-4xy^2+x^2y^2),
+$$
+$$
+\beta := \beta(x, y) = 1+x^2+y^2+x^2y^2,
+$$
+$$
+\gamma := \gamma(x, y) = 4xy.
+$$
+Since $x$ and $y$ fall between zero and one, we make use of the inequalities (excepting the cases of $x=y=0$ and $x=y=1$)
+$$
+\alpha_1 > \alpha_2
+$$
+and
+$$
+\beta > \gamma.
+$$
+% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
+%EM Rather confusing to be using ' for two things here.
+What about using a dot over the l?
+In retrospect something other than ' would have been good because of the proliferation of w^{'2} etc.
+Prob not worth changing now though.
+The derivative of the log likelihood is
+$$
+\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} - \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
+$$
+The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w') \lessgtr 0$, we equivalently consider $f(w') \lessgtr 0$ where $f$ is the quadratic function
+\begin{align*}
+f(w') &= (w')^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
+      &\qquad + w'\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
+      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
+\end{align*}
+% Note from das: b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero.
+c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
+Excepting all the cases where $x=0$ or where $y=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
+Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w'=0$.
+This implies that if the smaller of the zeros of $f(w')$ is greater than one, then $\hat{w}' \equiv 1$.
+Using the quadratic formula with
+$$
+a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
+$$
+$$
+b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
+$$
+$$
+c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
+$$
+the smaller zero is
+$$
+\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)},
+$$
+which is a function of the generating parameters $x$ and $y$.
+
+\generalBranchInconsist*
+
+In the general case, $\hat{w}'$ is a function of $x$, $y$, $\hat{x}'$, and $\hat{y}'$.
+Looking to the previous section, $\hat{w}'$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}'$ and $\hat{y}'$ instead of $x'$ and $y'$.
+Assume we know $\hat{x}'$ and $\hat{y}'$ as functions of $x$ and $y$ only.
+Suppressing function arguments and given that $a < 0$, we are interested in the cases where
+$$
+\frac{-b - \sqrt{b^2 - 4ac}}{2a} \ge 1,
+$$
+which, after some algebra, is equivalent to
+%-b - \sqrt{b^2 - 4ac} \le 2a,
+%-b - 2a \le \sqrt{b^2 - 4ac},
+%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
+%a^2 + ab + ac \le 0,
+$$
+a + b + c \ge 0.
+$$
+Simplifying, this is
+$$
+-\gamma^2(\hat{x}', \hat{y}')(1 + \frac{1}{2}\beta(x, y)) + 2\gamma(\hat{x}', \hat{y}')\beta(\hat{x}', \hat{y}')xy^2 + \beta^2(\hat{x}', \hat{y}') \ge 0.
+$$
+If we can find
+$$
+\gamma_{L}(x, y) \le \gamma(\hat{x}', \hat{y}') \le \gamma_{U}(x, y)
+$$
+and
+$$
+\beta_{L}(x, y) \le \beta(\hat{x}', \hat{y}')
+$$
+then we are interested in the region where
+$$
+-\gamma_{U}^2(x, y)(1 + \frac{1}{2}\beta(x, y)) + 2\gamma_{L}(x, y)\beta_{L}(x, y)xy^2 + \beta_{L}^2(x, y) \ge 0,
+$$
+an inequality involving only $x$ and $y$.
+The loosest possible bounds are $\gamma_{L} = 0$, $\beta_{L} = 1$ and $\gamma_{U} = 4$, giving
+$$
+-16(1 + \frac{1}{2}\beta(x, y)) + 1 \ge 0,
+$$
+which is never true.
+The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma(x, y)$ and $\beta_{L} = \beta(x, y)$, which is the case in Figure~\ref{fig:bl-inconsistency}.
+
+Say we are willing to tolerate a specified level of error in our estimates for $\hat{x}'$ and $\hat{y}'$ so that
+$$
+x-x_{L} \le \hat{x}' \le x+x_{U}
+$$
+and
+$$
+y-y_{L} \le \hat{y}' \le y+y_{U}.
+$$

### 2017-11-15 joint_inf.tex
get all content into main doc

@@
 
+Assume $\fullAlignment$ is generated from topology $\tau^*$ and branch lengths $t^*$.
+Define $\ell_{\tau^*,t^*}(\tau, t)$ such that
+$$
+\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t).
+$$
+We give $\ell$ explicitly in the Appendix.
 %EM I think that this is a technicality.
@@
 with edge ordering in the order of the taxa, then the internal branch last.
-We now show that, in the case of the Farris tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
-
-\begin{lemma}
-For $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ and $t=\{x,y,x,y,w\}$,
-$$
-\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t})) = \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)).
-$$
-\end{lemma}
-
-Since exchanging $x_1$ and $x_2$ does not change the value of the log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
-The analogous statement holds for $y_1$ and $y_2$.
-The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
+We show in the Appendix that, in the case of the Farris tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
 
@@
 
-Define $\ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t))$ as
-$$
-\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
-$$
 We show the following.
@@
 \begin{restatable}{theorem}{topoInconsist}
-There exist $L_0(x, y)$ and $L_1(x, y)$ such that
+There exist $L_0(x, y), L_1(x, y)$ and $0 < x', y' < 1$ such that
 $$
-\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \le L_0(x, y),
+\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le L_0(x, y),
 $$
 $$
-\max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t; \fullAncestralSplitPartitions(\tau_2,t)) \ge L_1(x, y)
+\max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t) \ge L_1(x, y)
 $$
-and $L_1(x, y) > L_0(x, y)$.
+and $L_1(x', y') > L_0(x', y')$.
 \end{restatable}
-
-Fig.~\ref{fig:inconsistency-farris} shows the region where $L_1(x, y) > L_0(x, y)$.
+Fig.~\ref{fig:inconsistency-farris} shows the values of $x$ and $y$ where $L_1(x, y) > L_0(x, y)$.
 
@@
 $$
-\hat{w}' = \arg\max_{w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x,y,w'\}; \fullAncestralSplitPartitions(\tau_1,\{x,y,w'\})),
+\hat{w}' = \arg\max_{w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x,y,w'\}),
 $$
@@
 $$
-(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x',y',w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})).
+(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x',y',w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}).
 $$
@@
 $$
-\begin{theorem}
+\begin{restatable}{theorem}{restrictedBranchInconsist}
 \label{thm:restricted-bl}
@@
 $$
-\end{theorem}
+\end{restatable}
 The inequality in Theorem~\ref{thm:restricted-bl} is a function of the generating parameters $x$ and $y$.
@@
 Assume we know $\hat{x}'$ and $\hat{y}'$ as functions of $x$ and $y$ only.
-\begin{theorem}
+\begin{restatable}{theorem}{generalBranchInconsist}
 \label{thm:general-bl}
@@
 $$
-\end{theorem}
+\end{restatable}
 Similar to Theorem~\ref{thm:restricted-bl}, the inequality in Theorem~\ref{thm:general-bl} is also a function of only $x$ and $y$.

### 2017-11-15 appendix.tex
reorganize some appendix and address comments

@@
 \section*{Appendix}
-
-\subsection*{Hadamard transform}
-
-Define: $\mathcal{T}$, $X$, $A$, $n$, $P$ and $\theta$.
-Let
-$$
-\theta(e) = 1-2p(e).
-$$
-Then
-$$
-p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subset X : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e \in P(\mathcal{T}, Y)} \ \theta(e) \right]
-$$
 
@@
 
+\subsection*{Site split likelihood}
+
 Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
@@
 where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.
+
+Let
+$$
+L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
+$$
+be the final product in \eqref{eq:site_pattern_likelihood}.
+Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
+We have
+\begin{equation*}
+\begin{split}
+&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t)) \\
+&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
+&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
+            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
+\end{split}
+\end{equation*}
+so that, in the limit of infinite data,
+\begin{equation}
+\begin{split}
+&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \\
+&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)].
+\label{eq:site_pattern_profile_likelihood_mean}
+\end{split}
+\end{equation}
+Define the divergence quantity
+$$
+\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
+$$
+and the partial log likelihood
+$$
+\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
+$$
+so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
+\begin{equation}
+    \label{eq:log_likelihood_simplified}
+    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
+\end{equation}
+
+\subsection*{Hadamard transform}
+
+Define: $\mathcal{T}$, $X$, $A$, $n$, $P$ and $\theta$.
+Let
+$$
+\theta(e) = 1-2p(e).
+$$
+Then
+\begin{equation}
+\label{eq:hadamard_probability}
+p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subset X : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e \in P(\mathcal{T}, Y)} \ \theta(e) \right]
+\end{equation}
 
@@
 Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} contain the values of the likelihood for all possible ancestral states.
+
+To compute the likelihood of observing a set of data, we need $P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
+The probability of a character change along a branch with fidelity parameter $x$ is $(1-x)/2$, while the probability of a character remaining the same is $(1+x)/2$.
+See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern.
+Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods}.
+Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}
+
+\begin{figure}
+\centering
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[width=.95\textwidth]{farris_like00}
+\caption[short]{$P(\ancestralStateColumn=\emptyset\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
+\end{subfigure}
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[width=.95\textwidth]{farris_like01}
+\caption[short]{$P(\ancestralStateColumn=\{2\}\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
+\end{subfigure}
+\vskip\baselineskip
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[width=.95\textwidth]{farris_like10}
+\caption[short]{$P(\ancestralStateColumn=\{1\}\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
+\end{subfigure}
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[width=.95\textwidth]{farris_like11}
+\caption[short]{$P(\ancestralStateColumn=\{1,2\}\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
+\end{subfigure}
+\caption{Example likelihood computation}
+\label{fig:example_likelihoods}
+\end{figure}
+
+\begin{table}
+\centering
+\begin{tabular}{|l|ll|}
+\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
+\hline
+& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
+    \hline
+    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
+    \hline
+     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$          &$(1+x)^2   (1-w)(1-y)^2$\\
+     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
+     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
+     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
+     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+     $\{1,3\}$     &$(1+x)^2   (1+w)(1-y)^2$          &$(1+x)^2   (1-w)(1+y)^2$\\
+     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+     $\{1,2,3\}$   &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
+    \hline
+    \hline
+    &$\{1\}$                             &$\{1,2\}$  \\
+    \hline
+     $\emptyset$   &$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
+     $\{1\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{2\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+     $\{3\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+     $\{1,3\}$     &$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^2$\\
+     $\{2,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+     $\{1,2,3\}$   &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+    \hline
+\end{tabular}
+\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
+All values multiplied by $1/32$.}
+\label{tab:farris_likelihoods}
+\end{table}
+
+\begin{table}
+\centering
+\begin{tabular}{|l|ll|}
+\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_2,t)$}\\
+\hline
+& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
+    \hline
+    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
+    \hline
+     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1-x)^2   (1-w)(1+y)(1-y)$\\
+     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1-y)^2$\\
+     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1+x)^2   (1-w)(1+y)(1-y)$\\
+     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1+x)^2   (1-w)(1+y)^2$\\
+     $\{1,3\}$     &$(1+x)^2   (1+w)(1-y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1-x)^2   (1-w)(1+y)^2$\\
+     $\{1,2,3\}$   &$(1+x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1+y)^2$\\
+    \hline
+    \hline
+    &$\{1\}$                             &$\{1,2\}$  \\
+    \hline
+     $\emptyset$   &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1-x)^2   (1+w)(1-y)^2$\\
+     $\{1\}$       &$(1+x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{2\}$       &$(1+x)(1-x)(1-w)(1+y)^2$           &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+     $\{3\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
+     $\{1,2\}$     &$(1-x)^2   (1-w)(1-y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+     $\{1,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1-x)^2   (1+w)(1+y)^2$\\
+     $\{2,3\}$     &$(1+x)^2   (1-w)(1-y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+     $\{1,2,3\}$   &$(1+x)(1-x)(1-w)(1-y)^2$           &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+\hline
+\end{tabular}
+\caption{Likelihood calculations for all site patterns and internal states of Felsenstein topology.
+All values multiplied by $1/32$.}
+\label{tab:fels_likelihoods}
+\end{table}
 
@@
 
-To compute the likelihood of observing a set of data, we need $P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
-The probability of a character change along a branch with fidelity parameter $x$ is $(1-x)/2$, while the probability of a character remaining the same is $(1+x)/2$.
-See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern.
-Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods}.
-Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}
-
-\begin{figure}
-\centering
-\begin{subfigure}{.45\linewidth}
-\centering
-\includegraphics[width=.95\textwidth]{farris_like00}
-\caption[short]{$P(\ancestralStateColumn=\emptyset\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
-\end{subfigure}
-\begin{subfigure}{.45\linewidth}
-\centering
-\includegraphics[width=.95\textwidth]{farris_like01}
-\caption[short]{$P(\ancestralStateColumn=\{2\}\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
-\end{subfigure}
-\vskip\baselineskip
-\begin{subfigure}{.45\linewidth}
-\centering
-\includegraphics[width=.95\textwidth]{farris_like10}
-\caption[short]{$P(\ancestralStateColumn=\{1\}\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
-\end{subfigure}
-\begin{subfigure}{.45\linewidth}
-\centering
-\includegraphics[width=.95\textwidth]{farris_like11}
-\caption[short]{$P(\ancestralStateColumn=\{1,2\}\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
-\end{subfigure}
-\caption{Example likelihood computation}
-\label{fig:example_likelihoods}
-\end{figure}
-
-\begin{table}
-\centering
-\begin{tabular}{|l|ll|}
-\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
-\hline
-& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
-    \hline
-    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
-    \hline
-     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$          &$(1+x)^2   (1-w)(1-y)^2$\\
-     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
-     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
-     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
-     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{1,3\}$     &$(1+x)^2   (1+w)(1-y)^2$          &$(1+x)^2   (1-w)(1+y)^2$\\
-     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{1,2,3\}$   &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
-    \hline
-    \hline
-    &$\{1\}$                             &$\{1,2\}$  \\
-    \hline
-     $\emptyset$   &$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
-     $\{1\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{2\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-     $\{3\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-     $\{1,3\}$     &$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^2$\\
-     $\{2,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-     $\{1,2,3\}$   &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-    \hline
-\end{tabular}
-\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
-All values multiplied by $1/32$.}
-\label{tab:farris_likelihoods}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{|l|ll|}
-\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_2,t)$}\\
-\hline
-& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
-    \hline
-    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
-    \hline
-     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1-x)^2   (1-w)(1+y)(1-y)$\\
-     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1-y)^2$\\
-     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1+x)^2   (1-w)(1+y)(1-y)$\\
-     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1+x)^2   (1-w)(1+y)^2$\\
-     $\{1,3\}$     &$(1+x)^2   (1+w)(1-y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1-x)^2   (1-w)(1+y)^2$\\
-     $\{1,2,3\}$   &$(1+x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1+y)^2$\\
-    \hline
-    \hline
-    &$\{1\}$                             &$\{1,2\}$  \\
-    \hline
-     $\emptyset$   &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1-x)^2   (1+w)(1-y)^2$\\
-     $\{1\}$       &$(1+x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{2\}$       &$(1+x)(1-x)(1-w)(1+y)^2$           &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-     $\{3\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{1,2\}$     &$(1-x)^2   (1-w)(1-y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-     $\{1,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1-x)^2   (1+w)(1+y)^2$\\
-     $\{2,3\}$     &$(1+x)^2   (1-w)(1-y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-     $\{1,2,3\}$   &$(1+x)(1-x)(1-w)(1-y)^2$           &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-\hline
-\end{tabular}
-\caption{Likelihood calculations for all site patterns and internal states of Felsenstein topology.
-All values multiplied by $1/32$.}
-\label{tab:fels_likelihoods}
-\end{table}
-
 \subsection*{Form of the likelihood}
@@
 
+\begin{proof}
 Our approach seeks bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified}.
@@
 \end{align*}
+\end{proof}
 
@@
 
+\begin{proof}
 Recalling Table~\ref{tab:likelihoods}, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood for $(x',y',w')$.
@@
 which is a function of the generating parameters $x$ and $y$.
+\end{proof}
 
@@
 
+\begin{proof}
 In the general case, $\hat{w}'$ is a function of $x$, $y$, $\hat{x}'$, and $\hat{y}'$.
@@
 $$
+\end{proof}

### 2017-11-15 joint_inf.tex
reorganize some appendix and address comments

@@
 \usepackage{amssymb}
+\usepackage{amsthm}
 \usepackage{graphicx}
@@
 The font I use is "MathJax_Math".
+%das: I think I'm either going to go with something other than the primes, like z_1, z_2, z_3, or replace x and y with greek letters or whatever, but it'll be a little bit of a tedious edit.
 \includegraphics[width=.95\textwidth]{farris_blank}
@@
 This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
-Fidelities have useful algebraic properties, which mean that the generating probabilities using the Hadamard transform have an especially simple form (see either 8.6 of Semple and Steel \cite{Semple2003-em} or Appendix (TODO)).
+Fidelities have useful algebraic properties, which mean that the generating probabilities using the Hadamard transform have an especially simple form (see either 8.6 of Semple and Steel \cite{Semple2003-em} or \eqref{eq:hadamard_probability} in the Appendix).
 
@@
 %EM Does this have to be unique? Can we just say "denote an estimate"?
+%das: what does this refer to? the deleted sentence about the alphabet being discrete? I had that in just to show that there will be a maximum---i.e., no degeneracy---but maybe it's not necessary.
 We will use $\hat{\fullAncestralStates}$ to denote the estimate for $\fullAncestralStates$ obtained by maximization \eqref{eq:profile_likelihood}.
@@
 $$
-We give $\ell$ explicitly in the Appendix.
+We give $\ell$ explicitly in \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
 %EM I think that this is a technicality.
@@
 %EM XXXX below because no bio journals have numbered sections.
+%das: what should we do you think? I ran into this with the hadamard transform and went with the equation number instead of section, but not sure what to do here.
 We compute these maxima over a lattice in steps of $10^{-3}$ for $x$ and $y$ from $10^{-3}$ to $1-10^{-3}$---we do not include zero or one in our lattice to further stabilize the fits, recalling from Section~XXXX that cases where $x$ or $y$ are equal to zero or one can result in pathologies.

### 2017-11-15 joint_inf.tex
Pass through comments.
quick pass

@@
 We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``Farris'' and ``Felsenstein'' trees (Fig.~\ref{fig:farris-fels-top}).
+In the classical configuration of each of these topologies, the interior branch length parameters are equal to the bottom two parameters.
+We show in the Appendix that, in the case of the Farris tree, fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
 
@@
 %das: I think I'm either going to go with something other than the primes, like z_1, z_2, z_3, or replace x and y with greek letters or whatever, but it'll be a little bit of a tedious edit.
+%EM You mean globally in the whole document? Let's kick around ideas if so.
+%das: I think we just need the true generating parameters---x^* and y^*---and the estimands---x', y', w'.
+I've been bad about using x and y instead of x^* and y^* for function arguments because it's cleaner, and I think there is inconsistency in using the primes and not.
+How about Greek for unknown (truth) and Latin for estimands?
+\alpha and \beta for generating parameters and x, y, z for estimands?
+Would have to change some other notation but shouldn't be too hard I hope.
 \includegraphics[width=.95\textwidth]{farris_blank}
@@
 %das: what does this refer to? the deleted sentence about the alphabet being discrete? I had that in just to show that there will be a maximum---i.e., no degeneracy---but maybe it's not necessary.
-We will use $\hat{\fullAncestralStates}$ to denote the estimate for $\fullAncestralStates$ obtained by maximization \eqref{eq:profile_likelihood}.
+%EM Damn, sorry to be unclear. The following statement says "denote THE estimate" which makes the ancestral state sound unique. Are you saying that it is? It's not obvious to me.
+%das: ah, got it. It's not unique, and why we needed the $\eta$s; hopefully "an" is better than "the?"
+We will use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximization \eqref{eq:profile_likelihood}.
 One can then estimate a topology and branch lengths using this profile likelihood:
@@
 I propose that we just state that we can restrict to the parameterization shown in the figure (with certain branch lengths equal) and that a formal statement and proof is in the appendix.
-Consider the Farris tree but with arbitrary branch lengths, i.e.,
-$$
-\tilde{t}=\{x_1,y_1,x_2,y_2,w\}
-$$
-with edge ordering in the order of the taxa, then the internal branch last.
-We show in the Appendix that, in the case of the Farris tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
+%EM :) I think that we can make this even simpler by skipping the following statement and cutting straight to "fixing both the top two branch lengths..."
 
@@
 We show the following.
-%EM Here's an example of restatable in action. See the end of the appendix.
 \begin{restatable}{theorem}{topoInconsist}
@@
 %das: what should we do you think? I ran into this with the hadamard transform and went with the equation number instead of section, but not sure what to do here.
-We compute these maxima over a lattice in steps of $10^{-3}$ for $x$ and $y$ from $10^{-3}$ to $1-10^{-3}$---we do not include zero or one in our lattice to further stabilize the fits, recalling from Section~XXXX that cases where $x$ or $y$ are equal to zero or one can result in pathologies.
+%EM In fact I wasn't even sure exactly to what you were referring when the section number was in there. It doesn't seem to me to need a whole lot of justification though.
+We compute these maxima over a lattice in steps of $10^{-3}$ for $x$ and $y$ from $10^{-3}$ to $1-10^{-3}$---we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
 From this process we obtain Fig.~\ref{fig:bl-general-inconsistency}.

### 2017-11-16 joint_inf_plot.py
larger fonts

@@ -21,6 +21,8 @@ sns.set_style('ticks')
 # ~~~~~~~~~
 # global variables
 
+FONT_SIZE = 16
+
 ALL_PATTERNS = [
         '0000',
         '1000',
@@ -387,14 +389,15 @@ def main(args=sys.argv[1:]):
         elif args.general_branch_lengths:
             region = what_lower_general(X, Y)
             plottitle = r'Region of inconsistent branch parameter estimation (general case)'
-            legendtext = r'$\hat{w}=1$ or $\hat{x}$ or $\hat{y}$ poorly estimated'
+            legendtext = r'$\hat{w}=1$ or $\hat{x}$ or $\hat{y}$''\n'r'poorly estimated'
 
         ct = plt.contour(X, Y, region, [0], colors='k', linewidths=1)
         plt.axis('scaled')
-        plt.xlabel(r'$x$')
-        plt.ylabel(r'$y$')
-        ttl = plt.title(plottitle, fontsize=14)
+        plt.xlabel(r'$x$', fontsize=FONT_SIZE)
+        plt.ylabel(r'$y$', fontsize=FONT_SIZE)
+        ttl = plt.title(plottitle, fontsize=FONT_SIZE+2)
         ttl.set_position([.5, 1.02])
+        ct.ax.tick_params(labelsize=FONT_SIZE-2)
         if not args.topology:
             vec = ct.collections[0].get_paths()[0].vertices
             x = np.concatenate(([1-args.delta], vec[:,0], [args.delta]))
@@ -404,10 +407,10 @@ def main(args=sys.argv[1:]):
         if args.restricted_branch_lengths:
             plt.legend([Legend()], [legendtext],
                 handler_map={Legend: LegendHandler()},
-                bbox_to_anchor=(1.05, 1), loc=2, borderaxespad=0.)
+                bbox_to_anchor=(1.05, 1), loc=2, borderaxespad=0., fontsize=FONT_SIZE)
         else:
             plt.legend([Legend()], [legendtext],
-                handler_map={Legend: LegendHandler()})
+                handler_map={Legend: LegendHandler()}, fontsize=FONT_SIZE)
 
         sns.despine()
         plt.savefig(args.plot_name)
@@ -428,17 +431,18 @@ def main(args=sys.argv[1:]):
                     X, Y, Z, args.delta = pickle.load(f)
 
[…]

### 2017-11-16 joint_inf_plot.py
title position and update plots (not inkscape yet)

@@ -396,7 +396,7 @@ def main(args=sys.argv[1:]):
         plt.xlabel(r'$x$', fontsize=FONT_SIZE)
         plt.ylabel(r'$y$', fontsize=FONT_SIZE)
         ttl = plt.title(plottitle, fontsize=FONT_SIZE+2)
-        ttl.set_position([.5, 1.02])
+        ttl.set_position([.5, 1.05])
         ct.ax.tick_params(labelsize=FONT_SIZE-2)
         if not args.topology:
             vec = ct.collections[0].get_paths()[0].vertices
@@ -434,7 +434,7 @@ def main(args=sys.argv[1:]):
             plt.xlabel(r'$x$', fontsize=FONT_SIZE)
             plt.ylabel(r'$y$', fontsize=FONT_SIZE)
             ttl = plt.title(r'Value of $\hat{w}$', fontsize=FONT_SIZE+2)
-            ttl.set_position([.5, 1.02])
+            ttl.set_position([.5, 1.05])
             ax = plt.gca()
             ax.set_xticks(np.arange(0, 1/args.delta, .2/args.delta))
             ax.set_yticks(np.arange(0, 1/args.delta, .2/args.delta))

### 2017-11-17 joint_inf.tex
title position and update plots (not inkscape yet)
quick pass; first crack at discussion and intuition

@@
 
-We begin by fixing some notation needed to be precise about the notions of likelihood examined in this paper.
-We assume the binary symmetric model, namely with a character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
+Assume the binary symmetric model, namely with a character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
+%das: @EM is there a good reference to binary symmetric model, or should we trot out the definition and all implications/corollaries of it? I guess it depends where we'd submit?
 Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
@@
 In the classical configuration of each of these topologies, the interior branch length parameters are equal to the bottom two parameters.
-We show in the Appendix that, in the case of the Farris tree, fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
+We show in the Appendix that, in the case of the Farris tree, fixing both the top two branch parameters to be equal and the bottom two branch parameters to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch parameters.
 
@@
 
-We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
-This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
-Fidelities have useful algebraic properties, which mean that the generating probabilities using the Hadamard transform have an especially simple form (see either 8.6 of Semple and Steel \cite{Semple2003-em} or \eqref{eq:hadamard_probability} in the Appendix).
+We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$ where $0 \le \theta \le 1$.
+We follow \cite{Matsen2007-jq} in calling this edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge.
+Fidelities have useful algebraic properties, and generating probabilities using the Hadamard transform have an especially simple form (see either 8.6 of Semple and Steel \cite{Semple2003-em} or \eqref{eq:hadamard_probability} in the Appendix).
 
@@
 The standard phylogenetic likelihood approach on unrooted trees under the usual assumption of independence between sites is as follows.
-For a topology $\tau$ and branch lengths $t$ the likelihood given observed ancestral states $\fullAncestralStates$ is
+For a topology $\tau$ and branch fidelities $t$ the likelihood given observed ancestral states $\fullAncestralStates$ is
 \begin{equation}
@@
 \end{equation}
-to estimate the tree $\tau$ and branch lengths $t$.
+to estimate the tree $\tau$ and branch fidelities $t$.
 
-An alternative approach used by \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully\hyp{}observed likelihood in \eqref{eq:full_likelihood}.
+An alternative approach \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully\hyp{}observed likelihood in \eqref{eq:full_likelihood}.
 This is known as a profile likelihood \cite{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
@@
 We will use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximization \eqref{eq:profile_likelihood}.
-One can then estimate a topology and branch lengths using this profile likelihood:
+One can then estimate a topology and branch fidelities using this profile likelihood:
 \begin{equation}
@@
 In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
-For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.
+For this reason, in practice, the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.
 
@@
 
-Assume $\fullAlignment$ is generated from topology $\tau^*$ and branch lengths $t^*$.
+Assume $\fullAlignment$ is generated from topology $\tau^*$ and branch fidelities $t^*$.
 Define $\ell_{\tau^*,t^*}(\tau, t)$ such that
@@
 
-We show the following.
 \begin{restatable}{theorem}{topoInconsist}
@@
 Fig.~\ref{fig:inconsistency-farris} shows the values of $x$ and $y$ where $L_1(x, y) > L_0(x, y)$.
+Intuitively, $\tau_2$ is favored in performing joint maximum likelihood since the likelihood for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its likelihood while $\tau_2$ has many more.
+This results in more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.
 
@@
 
-\subsection*{Inconsistency in branch length estimation}
+\subsection*{Inconsistency in branch fidelity estimation}
 
-We will now consider the problem of branch fidelity estimation on the correct tree using joint ancestral state and branch fidelity estimation.
-We will consider two settings on the Farris topology, corresponding to whether some branch fidelities are known ahead of time or not.
-As above, we will assume that data is generated with two sister branches of fidelity $x$, with all other branches of length $y$.
-We will show that, for a nontrivial subset of possible values for $x$ and $y$, the interior branch fidelity parameter $w'$ will be consistently overestimated as exactly equal to one instead of its true value of $y$.
+We now consider the problem of branch fidelity estimation on the correct tree using joint ancestral state and branch fidelity estimation.
+We analyze two settings on the Farris topology, corresponding to whether some branch fidelities are fixed at their true values or not.
+As above, assume that data is generated with two sister branches of fidelity $x$, with all other branches of fidelity $y$.
+We show that, for a nontrivial subset of possible values for $x$ and $y$, the interior branch fidelity parameter $w'$ will be consistently overestimated as exactly equal to one instead of its true value of $y$.
 That is, for the restricted case, with
@@
 $$
+These situations are in contrast to the approach using marginal likelihood where necessarily $\hat{w}'=y$.
 
@@
 
-First, assume we know the true values of $x'$ and $y'$ but not $w'$.
-In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
+Fix $x'=x$ and $y'=y$ to their true, generating values and proceed to estimate the internal branch parameter $w'$.
 Let
 $$
-\alpha_1 := \alpha_1(x, y) = \frac{1}{8} (1+x^2+y^2+4xy^2+x^2y^2),
+\alpha_1 := \alpha_1(x, y) = \frac{1}{8} \left(1+x^2+y^2+4xy^2+x^2y^2\right),
 $$
 $$
-\alpha_2 := \alpha_2(x, y) = \frac{1}{8}(1+x^2+y^2-4xy^2+x^2y^2),
+\alpha_2 := \alpha_2(x, y) = \frac{1}{8}\left(1+x^2+y^2-4xy^2+x^2y^2\right),
 $$
@@
 
+Intuitively, since we perform maximization over internal states, branch fidelities will have a tendency to be overestimated.
+Seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.
+
 \subsubsection*{General case}
@@
 In the general case, $\hat{w}'$ is a function of $x$, $y$, $\hat{x}'$, and $\hat{y}'$.
-Looking to the previous section, $\hat{w}'$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}'$ and $\hat{y}'$ instead of $x'$ and $y'$.
+Looking to the previous section, $\hat{w}'$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}'$ and $\hat{y}'$ instead of $x$ and $y$.
 Assume we know $\hat{x}'$ and $\hat{y}'$ as functions of $x$ and $y$ only.
@@
 $$
-\beta_{L}(x, y) \le \beta(\hat{x}', \hat{y}')
+\beta_{L}(x, y) \le \beta(\hat{x}', \hat{y}'),
 $$
-then the maximum likelihood value $\hat{w}=1$ when
+the maximum likelihood value $\hat{w}=1$ when
 $$
--\gamma_{U}^2(x, y)(1 + \frac{1}{2}\beta(x, y)) + 2\gamma_{L}(x, y)\beta_{L}(x, y)xy^2 + \beta_{L}^2(x, y) \ge 0.
+-\gamma_{U}^2(x, y)\left(1 + \frac{1}{2}\beta(x, y)\right) + 2\gamma_{L}(x, y)\beta_{L}(x, y)xy^2 + \beta_{L}^2(x, y) \ge 0.
 $$
@@
 $$
--16(1 + \frac{1}{2}\beta(x, y)) + 1 \ge 0,
+-16\left(1 + \frac{1}{2}\beta(x, y)\right) + 1 \ge 0,
 $$
@@
 The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma(x, y)$ and $\beta_{L} = \beta(x, y)$, which is the case in Figure~\ref{fig:bl-inconsistency}.
-As an example middle case, let the error in estimating both $\hat{x}'$ and $\hat{y}'$ be $.1$ in either direction, i.e., $x-.1\le\hat{x}'\le x+.1$ and $y-.1\le\hat{y}'\le y+.1$.
-In this case, the region to the left side of the curve in Figure~\ref{fig:bl-loose-inconsistency} will either estimate $\hat{w}'$ to be one or estimate either $\hat{x}'$ or $\hat{y}'$ to be more than $.1$ away from its true value.
+As an example middle case, let the error in estimating both $\hat{x}'$ and $\hat{y}'$ be $0.1$ in either direction, i.e.,
+$$
+x-0.1\le\hat{x}'\le x+0.1,
+$$
+$$
+y-0.1\le\hat{y}'\le y+0.1.
+$$
+In this case, the region to the left side of the curve in Figure~\ref{fig:bl-loose-inconsistency} will either estimate $\hat{w}'$ to be one or estimate either $\hat{x}'$ or $\hat{y}'$ to be more than $0.1$ away from its true value.
 
@@
 \includegraphics[width=.9\textwidth]{bl-loose-inconsistency-inkscape}
-\caption{Region of inconsistency with $x-.1 \le \hat{x}' \le x+.1$ and $y-.1 \le \hat{y}' \le y+.1$.}
+\caption{Region of inconsistency with $x-0.1 \le \hat{x}' \le x+0.1$ and $y-0.1 \le \hat{y}' \le y+0.1$.}
 \label{fig:bl-loose-inconsistency}
@@
 
-The region of inconsistency in Figure~\ref{fig:bl-loose-inconsistency} is potentially a conservative estimate of where $\hat{w}=1$.
-To determine how conservative, we use the method of differential evolution \cite{Storn1997} to perform joint maximum likelihood estimation.
-Since empirical estimates can be somewhat unstable, we perform two maximizations---one over $0 \le x',y',w' \le 1$ and one over $0 \le x',y' \le 1$ with $w'=1$---and take the value of $\hat{w}'$ with the higher likelihood.
+The region of inconsistency in Figure~\ref{fig:bl-loose-inconsistency} is a conservative estimate of where $\hat{w}=1$.
+To determine how conservative, we use the method of differential evolution \cite{Storn1997} to perform joint maximum likelihood estimation using numerical optimization.
+Since empirical estimates can be unstable, we perform two maximizations---one over $0 \le x',y',w' \le 1$ and one over $0 \le x',y' \le 1$ with $w'=1$---and take the value of $\hat{w}'$ with the higher likelihood.
 %EM XXXX below because no bio journals have numbered sections.
@@
 % if fit already
-\includegraphics[width=.9\textwidth]{w-hat-empirical-001}
+\includegraphics[width=\textwidth]{w-hat-empirical-001}
 \caption{
     %das: note to update caption
-    Region where $\hat{w}'=1$ computing empirical estimates of $(\hat{x}', \hat{y}', \hat{w}')$.
+    Estimates for $\hat{w}'$ when computing empirical estimates of $(\hat{x}', \hat{y}', \hat{w}')$ using differential evolution \cite{Storn1997}.
 }
@@
 
-Note that this two step thing is known in statistics as profile likelihood \cite{Murphy2000-ry}.
-People have worked on its properties.
+The method of profile likleihood is well-studied \cite{Murphy2000-ry}.
+Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known \cite{Geman1982}, and joint inference of phylogenetic parameters and ancestral states \cite{Sagulenko2017-jo} is a similar approach to profile likelihood.
+Elsewhere in statistics literature, undesirable properties of maximum likelihood estimates in the case of infinite parameters have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though the fact of parameters tending to infinity is not necessarily the reason for inconsistency \cite{LeCam1990}.
+Until recently \cite{RoyChoudhury2015-ta}, results on the properties of standard maximum likelihood estimates of phylogeny as in \eqref{eq:marginal_likelihood} have been difficult to prove, and no results have been obtained for profile likelihood.
+We have furthered progress to understand the limitations of this joint maximum likelihood approach in the case of phylogenetics.
 
-Neyman-Scott paradox.
+In phylogenetics, counterexamples showing methods of phylogenetic inference are inconsistent have been shown using the same four-taxa topologies used here \cite{Felsenstein1978-rr}.
+In this case, the converse is shown: when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely.
+Looking at the figure in \cite{Felsenstein1978-rr}, the region in Fig.~\ref{fig:inconsistency-farris} has a similar character, albeit a smaller region.
+Difficulties in phylogenetic estimation when the ``Farris zone'' generates characters is used in \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.
 
-Interesting that here we are simulating on the Farris tree and end up with the Felsenstein tree.
-For maximum parsimony it's the opposite \cite{Felsenstein1978-rr}.
 %EM We should be clear here that thus this is not "long branch attraction".
@@
 %das: Noted. I will work on this.
-Discussion of number of parameters of each.
-
-However, note that \cite{Siddall1998-hq} get things going the same way, although \cite{Swofford2001-hr} show that the problem is that they didn't simulate long enough sequences.
+%das: still not quite sure about what to say re LBA, but I'm thinking about it
+%Discussion of number of parameters of each.
 
 \section*{Acknowledgements}
-We would like to thank Richard Neher for helpful discussions.
+We thank Richard Neher for helpful discussions.
 

### 2017-11-17 appendix.tex
more explicit hadamard and some rewrites

@@
 
-Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation (we follow Section 8.6 of \cite{Semple2003-em}).
+Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation.
 Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
 For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
-Indeed, we start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
+We start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
 Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.
@@
 
-This split formulation defines mappings from patterns to splits as
+\begin{definition}
+Let the mapping from site patterns to site splits be
 $$
- \patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet).
+\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
 $$
-Given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
+and the mapping from ancestral states and tip states to ancestral state splits be
+$$
+\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet).
+$$
+Then, given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
 $$
@@
 $$
-that takes corresponding realizations $\siteSplit_j$ for some $j$.
+that takes corresponding realizations $\siteSplit_j$ for some $j$, and
+$$
+\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
+$$
+for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
+\end{definition}
+The mapping $\patternToSplit$ takes the complement of site patterns to obtain a site split in $\mathcal{P}(\siteSplitSet)$.
+The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.
 
@@
 \end{align*}
-for some $j$.
-For the other term, we define
-$$
-\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
-$$
-as a mapping of ancestral states and tip states to ancestral state splits.
-It is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.
-We define
-$$
-\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
-$$
-for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, so that
-%By considering the case requiring complementing and that not requiring complementing separately,
-%das: did I write the above? I don't remember the significance of it...
+for some $j$ and
 \begin{align*}
@@
 
-\subsection*{Hadamard transform}
+\subsection*{Hadamard representation}
 
-Define: $\mathcal{T}$, $X$, $A$, $n$, $P$ and $\theta$.
-Let
+We follow Section 8.6 of \cite{Semple2003-em} to state the Hadamard representation of site split generating probabilities.
+%das: @EM thoughts on including all of this in full generality? should we just reference using the sentence above and be done with it?
+For each edge $e$ define the edge fidelity for that edge as
 $$
@@
 $$
-Then
+For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
+For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
 \begin{equation}
 \label{eq:hadamard_probability}
-p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subset X : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e \in P(\mathcal{T}, Y)} \ \theta(e) \right]
+p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
 \end{equation}
+By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
+We follow with an expository example computing these probabilities and likelihoods.
 
@@
 To compute the likelihood of observing a set of data, we need $P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
-The probability of a character change along a branch with fidelity parameter $x$ is $(1-x)/2$, while the probability of a character remaining the same is $(1+x)/2$.
+The probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
 See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern.
@@
 Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
-For compactness, define
+For compactness, we follow the convention in \eqref{eq:hadamard_probability} and define
 $$

### 2017-11-17 joint_inf.tex
more explicit hadamard and some rewrites

@@
 \newtheorem{lemma}[theorem]{Lemma}
+\newtheorem*{definition}{Definition}
 
@@
 We follow \cite{Matsen2007-jq} in calling this edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge.
-Fidelities have useful algebraic properties, and generating probabilities using the Hadamard transform have an especially simple form (see either 8.6 of Semple and Steel \cite{Semple2003-em} or \eqref{eq:hadamard_probability} in the Appendix).
+Fidelities have useful algebraic properties, and generating probabilities using the Hadamard transform have an especially simple form (see \eqref{eq:hadamard_probability} in the Appendix).
 
@@
 Fix $x'=x$ and $y'=y$ to their true, generating values and proceed to estimate the internal branch parameter $w'$.
+\begin{restatable}{theorem}{restrictedBranchInconsist}
+\label{thm:restricted-bl}
 Let
@@
 $$
-\begin{restatable}{theorem}{restrictedBranchInconsist}
-\label{thm:restricted-bl}
 The maximum likelihood value $\hat{w}' = 1$ if
@@
 This region is shown in Fig.~\ref{fig:bl-inconsistency}.
+Intuitively, since we perform maximization over internal states, branch fidelities will have a tendency to be overestimated.
+Seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.
 
@@
 
-Intuitively, since we perform maximization over internal states, branch fidelities will have a tendency to be overestimated.
-Seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.
-
 \subsubsection*{General case}
@@
 For
+$$
+\beta(x, y) = 1+x^2+y^2+x^2y^2,
+$$
+$$
+\gamma(x, y) = 4xy,
+$$
+and bounds
 $$
@@
 Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known \cite{Geman1982}, and joint inference of phylogenetic parameters and ancestral states \cite{Sagulenko2017-jo} is a similar approach to profile likelihood.
-Elsewhere in statistics literature, undesirable properties of maximum likelihood estimates in the case of infinite parameters have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though the fact of parameters tending to infinity is not necessarily the reason for inconsistency \cite{LeCam1990}.
+Elsewhere in statistics literature, undesirable properties of maximum likelihood estimates in the case of infinite parameters have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \cite{LeCam1990}.
 Until recently \cite{RoyChoudhury2015-ta}, results on the properties of standard maximum likelihood estimates of phylogeny as in \eqref{eq:marginal_likelihood} have been difficult to prove, and no results have been obtained for profile likelihood.

### 2017-11-18 appendix.tex
tidying up proofs a bit

@@
 $$
-and the partial log likelihood
+and the partial log-likelihood
 $$
@@
 
-Since exchanging $x_1$ and $x_2$ does not change the value of the log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
+Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
 The analogous statement holds for $y_1$ and $y_2$.
@@
 \begin{proof}
-Our approach seeks bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified}.
-We obtain
+In general,
 $$
@@
 $$
-as an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
+is an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
 $$
@@
 $$
-Similarly, we have the lower bound
+Similarly,
 $$
@@
     \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
-    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t})).
+    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t}))
 $$
+is a lower bound.
 
 Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
-For compactness, we follow the convention in \eqref{eq:hadamard_probability} and define
+Following \eqref{eq:hadamard_probability},
 $$
-p_{\emptyset} = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})
+p_{\emptyset} = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\}),
 $$
 with similar definitions for the remaining generating probabilities.
-The full Farris tree log likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see Table~\ref{tab:likelihoods} for $\tau = \tau_1$).
-We begin by writing out the middle case: for the ancestral state split $\{1\}$.
-Directly substituting in \eqref{eq:site_pattern_likelihood}, the log likelihood is
+The full Farris tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see Table~\ref{tab:likelihoods} for $\tau = \tau_1$).
+We begin by writing out the case for the ancestral state split $\{1\}$.
+Directly substituting in \eqref{eq:site_pattern_likelihood}, the log-likelihood is
 \begin{align}
@@
 \end{align}
-Seeking an upper bound, we use the facts that, for $x',y'\in[0,1]$,
+We use the facts that, for $x',y'\in[0,1]$,
 \begin{align*}
@@
 \end{align*}
-To make matters further compact, let
+To simplify, let
 \begin{equation}
@@
 \end{equation}
-where
+so that
 \begin{equation*}
@@
 All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.
+Call the upper bound
+$$
+L_0(x, y) := max(L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}), L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}), L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy})).
+$$
 
@@
 Our lower bound is
-\begin{align*}
-    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,0\}) \\
-                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b).
-\end{align*}
+$$
+L_1(x, y) := \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,0\}) + 2\cdot\log(2) + (2-b)\cdot\log(2-b) + b\cdot\log(b).
+$$
+See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < x', y' < 1$ where $L_1(x', y') > L_0(x', y')$ and joint inference in inconsistent.
 \end{proof}
@@
 Prob not worth changing now though.
-The derivative of the log likelihood is
+The derivative of the log-likelihood is
 $$
@@
 which is a function of the generating parameters $x$ and $y$.
+We see that
+$$
+\hat{w}'(x, y) \ge 1
+$$
+is equivalent to
+%-b - \sqrt{b^2 - 4ac} \le 2a,
+%-b - 2a \le \sqrt{b^2 - 4ac},
+%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
+%a^2 + ab + ac \le 0,
+$$
+a(x, y) + b(x, y) + c(x, y) \ge 0,
+$$
+which, simplifying as functions of $\alpha_1, \alpha_2, \gamma$ and $\beta$ yields
+\begin{equation}
+\label{eq:restricted-bl-result}
+\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
+\end{equation}
 \end{proof}
@@
 Assume we know $\hat{x}'$ and $\hat{y}'$ as functions of $x$ and $y$ only.
-Suppressing function arguments and given that $a < 0$, we are interested in the cases where
+Since
 $$
-\frac{-b - \sqrt{b^2 - 4ac}}{2a} \ge 1,
+\alpha_1 + \alpha_2 = \frac{1}{2}\beta
 $$
-which, after some algebra, is equivalent to
-%-b - \sqrt{b^2 - 4ac} \le 2a,
-%-b - 2a \le \sqrt{b^2 - 4ac},
-%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
-%a^2 + ab + ac \le 0,
+and
 $$
-a + b + c \ge 0.
+\alpha_1 - \alpha_2 = xy^2,
 $$
-Simplifying, this is
-$$
+simplifying \eqref{eq:restricted-bl-result} further yields
+\begin{equation}
+\label{eq:general-bl-result}
 -\gamma^2(\hat{x}', \hat{y}')(1 + \frac{1}{2}\beta(x, y)) + 2\gamma(\hat{x}', \hat{y}')\beta(\hat{x}', \hat{y}')xy^2 + \beta^2(\hat{x}', \hat{y}') \ge 0.
-$$
-If we can find
+\end{equation}
+Given the bounds
 $$
@@
 $$
-\beta_{L}(x, y) \le \beta(\hat{x}', \hat{y}')
+\beta_{L}(x, y) \le \beta(\hat{x}', \hat{y}'),
 $$
-then we are interested in the region where
+if
 $$
@@
 $$
-an inequality involving only $x$ and $y$.
+then \eqref{general-bl-result} holds, and this is an inequality involving only $x$ and $y$.
 The loosest possible bounds are $\gamma_{L} = 0$, $\beta_{L} = 1$ and $\gamma_{U} = 4$, giving
@@
 which is never true.
-The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma(x, y)$ and $\beta_{L} = \beta(x, y)$, which is the case in Figure~\ref{fig:bl-inconsistency}.
+The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma(x, y)$ and $\beta_{L} = \beta(x, y)$, which is the case in Fig.~\ref{fig:bl-inconsistency}.
 
-Say we are willing to tolerate a specified level of error in our estimates for $\hat{x}'$ and $\hat{y}'$ so that
+For an intermediate bound, define a specified level of error in estimates for $\hat{x}'$ and $\hat{y}'$ so that
 $$
@@
 $$
+Then $\gamma_L(x, y), \gamma_U(x, y)$ and $\beta_L(x, y)$ are defined directly from the bounds on $\hat{x}'$ and $\hat{y}'$.
+To get a sense of the region of overestimation, Fig.~\ref{fig:bl-loose-inconsistency} plots the case where $x_L=x_U=y_L=y_U=0.1$.
 \end{proof}

### 2017-11-18 appendix.tex
full pass main text; remove redundancy appendix

@@
 then \eqref{general-bl-result} holds, and this is an inequality involving only $x$ and $y$.
-The loosest possible bounds are $\gamma_{L} = 0$, $\beta_{L} = 1$ and $\gamma_{U} = 4$, giving
-$$
--16(1 + \frac{1}{2}\beta(x, y)) + 1 \ge 0,
-$$
-which is never true.
-The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma(x, y)$ and $\beta_{L} = \beta(x, y)$, which is the case in Fig.~\ref{fig:bl-inconsistency}.
-
-For an intermediate bound, define a specified level of error in estimates for $\hat{x}'$ and $\hat{y}'$ so that
-$$
-x-x_{L} \le \hat{x}' \le x+x_{U}
-$$
-and
-$$
-y-y_{L} \le \hat{y}' \le y+y_{U}.
-$$
-Then $\gamma_L(x, y), \gamma_U(x, y)$ and $\beta_L(x, y)$ are defined directly from the bounds on $\hat{x}'$ and $\hat{y}'$.
-To get a sense of the region of overestimation, Fig.~\ref{fig:bl-loose-inconsistency} plots the case where $x_L=x_U=y_L=y_U=0.1$.
+To get a sense of the region of overestimation, Fig.~\ref{fig:bl-loose-inconsistency} plots an intermediate case.
 \end{proof}

### 2017-11-18 joint_inf.tex
full pass main text; remove redundancy appendix

@@
 
+\begin{abstract}
+TODO
+\end{abstract}
+
 \section*{Introduction}
@@
 This is attractive from a computational perspective: such joint ML inference can proceed according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
-This latter conditional optimization is simpler and more computationally efficient than marginalizing out ancestral sequences and then performing optimization.
+This latter conditional optimization is simpler and more computationally efficient than marginalizing over ancestral sequences and then performing optimization.
 
@@
 Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
-Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$ and limit themselves to optimizing branch lengths using this method, their work motivates understanding when exactly this approximation breaks down, and what happens when we optimize branch lengths and tree topology using this approximation.
+Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$ and limit themselves to optimizing branch lengths using this method, their work motivates understanding when exactly this approximation breaks down, and what happens when optimizing branch lengths and tree topology using this approximation.
 
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model and imagine data being generated on the four taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the limit of infinite data.
+To do so, we use a binary symmetric model and imagine data being generated on the four-taxa ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the limit of infinite data.
 We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known.
@@
 \end{subfigure}
-\caption{Two simple topologies}
+\caption{Two four-taxa topologies}
 \label{fig:farris-fels-top}
@@
 
-An alternative approach \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully\hyp{}observed likelihood in \eqref{eq:full_likelihood}.
+An alternative approach \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
 This is known as a profile likelihood \cite{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
@@
 We analyze two settings on the Farris topology, corresponding to whether some branch fidelities are fixed at their true values or not.
-As above, assume that data is generated with two sister branches of fidelity $x$, with all other branches of fidelity $y$.
+As above, assume that data is generated from the ``Farris zone'' with two top branches of fidelity $x$ and all other branches of fidelity $y$.
 We show that, for a nontrivial subset of possible values for $x$ and $y$, the interior branch fidelity parameter $w'$ will be consistently overestimated as exactly equal to one instead of its true value of $y$.
-That is, for the restricted case, with
+That is, if we know $x'$ and $y'$, then for
 $$
@@
 $$
-These situations are in contrast to the approach using marginal likelihood where necessarily $\hat{w}'=y$.
+These situations are in contrast to the approach using marginal likelihood where necessarily $\hat{w}'=y$ in the limit of infinite data.
 
@@
 \end{restatable}
-Similar to Theorem~\ref{thm:restricted-bl}, the inequality in Theorem~\ref{thm:general-bl} is also a function of only $x$ and $y$.
+Similar to Theorem~\ref{thm:restricted-bl}, the inequality in Theorem~\ref{thm:general-bl} is also a function of only $x$ and $y$, though general forms for $\gamma_L(x, y)$, $\gamma_U(x, y)$, and $\beta_L(x, y)$ are difficult to obtain.
 The loosest possible bounds are $\gamma_{L} = 0$, $\beta_{L} = 1$ and $\gamma_{U} = 4$, giving
@@
 which is never true.
-The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma(x, y)$ and $\beta_{L} = \beta(x, y)$, which is the case in Figure~\ref{fig:bl-inconsistency}.
-As an example middle case, let the error in estimating both $\hat{x}'$ and $\hat{y}'$ be $0.1$ in either direction, i.e.,
+The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma(x, y)$ and $\beta_{L} = \beta(x, y)$, which is the case in Fig.~\ref{fig:bl-inconsistency}.
+For an intermediate bound, define a specified level of error in estimates for $\hat{x}'$ and $\hat{y}'$ so that
 $$
-x-0.1\le\hat{x}'\le x+0.1,
+x-x_{L} \le \hat{x}' \le x+x_{U}
 $$
+and
 $$
-y-0.1\le\hat{y}'\le y+0.1.
+y-y_{L} \le \hat{y}' \le y+y_{U}.
 $$
-In this case, the region to the left side of the curve in Figure~\ref{fig:bl-loose-inconsistency} will either estimate $\hat{w}'$ to be one or estimate either $\hat{x}'$ or $\hat{y}'$ to be more than $0.1$ away from its true value.
+Then $\gamma_L(x, y), \gamma_U(x, y)$ and $\beta_L(x, y)$ are defined directly from the bounds on $\hat{x}'$ and $\hat{y}'$.
+As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
+In this case, we will either estimate $\hat{w}'$ to be one or estimate either $\hat{x}'$ or $\hat{y}'$ to be more than $0.1$ away from its true value.
 
@@
 
-The region of inconsistency in Figure~\ref{fig:bl-loose-inconsistency} is a conservative estimate of where $\hat{w}=1$.
+The region of inconsistency in Fig.~\ref{fig:bl-loose-inconsistency} is a conservative estimate of where $\hat{w}=1$.
 To determine how conservative, we use the method of differential evolution \cite{Storn1997} to perform joint maximum likelihood estimation using numerical optimization.
@@
 
-The estimates for $\hat{w}'$ when $x$ and $y$ are both large seem to be unbiased, with $\hat{w}'$ increasing as $y$ increases, though the region where $\hat{w}'$ is estimated to be one---indicating no change along that particular branch---is similar to that in the restricted case.
-Although the restricted and general case are similar, comparing Figures~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case has a slightly different character, possibly due to potential looseness of the bound in the restricted case.
+The estimates for $\hat{w}'$ when $x$ and $y$ are both large look unbiased, with $\hat{w}'$ increasing as $y$ increases, though the region where $\hat{w}'$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
+Although the restricted and general case are similar, comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case has a slightly different character, possibly due to potential looseness of the bound in the restricted case.
 %das: TODO double check this; do the same empirical plot for the restricted case
-This region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have a region of the space of true parameters where we will always estimate the interior branch parameter to be one.
+This region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have many situations where we will always estimate the interior branch parameter to be one.
 
@@
 The method of profile likleihood is well-studied \cite{Murphy2000-ry}.
-Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known \cite{Geman1982}, and joint inference of phylogenetic parameters and ancestral states \cite{Sagulenko2017-jo} is a similar approach to profile likelihood.
-Elsewhere in statistics literature, undesirable properties of maximum likelihood estimates in the case of infinite parameters have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \cite{LeCam1990}.
+Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., \cite{Geman1982}.
+Joint inference of phylogenetic parameters and ancestral states \cite{Sagulenko2017-jo} is a similar approach to profile likelihood.
+Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates in the case of infinite parameters have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \cite{LeCam1990}.
 Until recently \cite{RoyChoudhury2015-ta}, results on the properties of standard maximum likelihood estimates of phylogeny as in \eqref{eq:marginal_likelihood} have been difficult to prove, and no results have been obtained for profile likelihood.
-We have furthered progress to understand the limitations of this joint maximum likelihood approach in the case of phylogenetics.
+We have furthered progress in understanding the limitations of this joint maximum likelihood approach.
 
 In phylogenetics, counterexamples showing methods of phylogenetic inference are inconsistent have been shown using the same four-taxa topologies used here \cite{Felsenstein1978-rr}.
-In this case, the converse is shown: when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely.
+In this case, the converse is shown: when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
 Looking at the figure in \cite{Felsenstein1978-rr}, the region in Fig.~\ref{fig:inconsistency-farris} has a similar character, albeit a smaller region.
-Difficulties in phylogenetic estimation when the ``Farris zone'' generates characters is used in \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.
+Difficulties in phylogenetic estimation when the ``Farris zone'' generates characters have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.
 

### 2017-11-20 joint_inf.tex
Responding to queries.

@@
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model and imagine data being generated on the four-taxa ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the limit of infinite data.
+To do so, we use a binary symmetric model and imagine data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the limit of infinite data.
 We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known.
@@
 %das: @EM is there a good reference to binary symmetric model, or should we trot out the definition and all implications/corollaries of it? I guess it depends where we'd submit?
+%EM I think that this is enough. In the lcfit paper we referenced the Semple and Stell phylogenetics book.
 Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
@@
 We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``Farris'' and ``Felsenstein'' trees (Fig.~\ref{fig:farris-fels-top}).
+%EM Do you know that "topology" means tree without branch lengths? Therefore I think "tree" is more appropriate in the following sentence and elsewhere.
 In the classical configuration of each of these topologies, the interior branch length parameters are equal to the bottom two parameters.
@@
 Would have to change some other notation but shouldn't be too hard I hope.
+%EM Wow, this is going to be some work.
+Sounds like a good plan, but I'd prefer for a system in which it's easy to remember which greek goes with which latin.
+could we do a, b, g and alpha, beta, gamma or something?
+Let's keep discussing before making the big change.
 \includegraphics[width=.95\textwidth]{farris_blank}
@@
 \end{subfigure}
-\caption{Two four-taxa topologies}
+\caption{Two four-taxon topologies}
 \label{fig:farris-fels-top}
@@
 \end{equation}
-%EM Does this have to be unique? Can we just say "denote an estimate"?
-%das: what does this refer to? the deleted sentence about the alphabet being discrete? I had that in just to show that there will be a maximum---i.e., no degeneracy---but maybe it's not necessary.
-%EM Damn, sorry to be unclear. The following statement says "denote THE estimate" which makes the ancestral state sound unique. Are you saying that it is? It's not obvious to me.
-%das: ah, got it. It's not unique, and why we needed the $\eta$s; hopefully "an" is better than "the?"
 We will use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximization \eqref{eq:profile_likelihood}.
@@
 We give $\ell$ explicitly in \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
-%EM I think that this is a technicality.
-I propose that we just state that we can restrict to the parameterization shown in the figure (with certain branch lengths equal) and that a formal statement and proof is in the appendix.
-%EM :) I think that we can make this even simpler by skipping the following statement and cutting straight to "fixing both the top two branch lengths..."
 
@@
 
-\subsection*{Inconsistency in branch fidelity estimation}
+\subsection*{Inconsistency in branch length estimation}
 
 We now consider the problem of branch fidelity estimation on the correct tree using joint ancestral state and branch fidelity estimation.
+As described above, we will use the equivalent but different notion of branch fidelities.
 We analyze two settings on the Farris topology, corresponding to whether some branch fidelities are fixed at their true values or not.
@@
 Since empirical estimates can be unstable, we perform two maximizations---one over $0 \le x',y',w' \le 1$ and one over $0 \le x',y' \le 1$ with $w'=1$---and take the value of $\hat{w}'$ with the higher likelihood.
-%EM XXXX below because no bio journals have numbered sections.
-%das: what should we do you think? I ran into this with the hadamard transform and went with the equation number instead of section, but not sure what to do here.
-%EM In fact I wasn't even sure exactly to what you were referring when the section number was in there. It doesn't seem to me to need a whole lot of justification though.
 We compute these maxima over a lattice in steps of $10^{-3}$ for $x$ and $y$ from $10^{-3}$ to $1-10^{-3}$---we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
@@
 In this case, the converse is shown: when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
+%EM Re LBA I was only thinking of putting a little sentence here that says that thus this is _not_ long branch attraction.
+%EM I don't know what "similar character" means here. I guess you are talking about his Figure 2?
 Looking at the figure in \cite{Felsenstein1978-rr}, the region in Fig.~\ref{fig:inconsistency-farris} has a similar character, albeit a smaller region.

### 2017-11-20 appendix.tex
Responding to queries.
marginal optim (also switch back to basin hopping for stability)

@@
 %das: @EM thoughts on including all of this in full generality? should we just reference using the sentence above and be done with it?
+%EM You already have it in here, and so I don't see a reason to yank it now.
 For each edge $e$ define the edge fidelity for that edge as
@@
 The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
+
+\subsection*{Theorems and proofs}
 
@@
 $$
-then \eqref{general-bl-result} holds, and this is an inequality involving only $x$ and $y$.
+then \eqref{eq:general-bl-result} holds, and this is an inequality involving only $x$ and $y$.
 To get a sense of the region of overestimation, Fig.~\ref{fig:bl-loose-inconsistency} plots an intermediate case.
 \end{proof}
+
+%EM Wanna add a short section about numerical optimization? You've seen that it's not entirely trivial.

### 2017-11-20 joint_inf.tex
marginal optim (also switch back to basin hopping for stability)

@@
 %EM Do you know that "topology" means tree without branch lengths? Therefore I think "tree" is more appropriate in the following sentence and elsewhere.
-In the classical configuration of each of these topologies, the interior branch length parameters are equal to the bottom two parameters.
+In the classical configuration of each of these trees, the interior branch length parameters are equal to the bottom two parameters.
 We show in the Appendix that, in the case of the Farris tree, fixing both the top two branch parameters to be equal and the bottom two branch parameters to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch parameters.
@@
 \includegraphics[width=.95\textwidth]{farris_blank}
-\caption[short]{Farris topology $\tau_1$}
+\caption[short]{Farris tree $\tau_1$}
 \end{subfigure}
@@
 \includegraphics[width=.95\textwidth]{felsenstein_blank}
-\caption[short]{Felsenstein topology $\tau_2$}
+\caption[short]{Felsenstein tree $\tau_2$}
 \end{subfigure}
-\caption{Two four-taxon topologies}
+\caption{Two four-taxon trees}
 \label{fig:farris-fels-top}
@@
 Although the restricted and general case are similar, comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case has a slightly different character, possibly due to potential looseness of the bound in the restricted case.
-%das: TODO double check this; do the same empirical plot for the restricted case
 This region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have many situations where we will always estimate the interior branch parameter to be one.
@@
 \centering
+\begin{subfigure}{.45\linewidth}
+\centering
 % obtained by running
-% python joint_inf_plot.py --empirical --general-branch-lengths --delta .001 --plot-name figures/w-hat-empirical-001.svg --pkl-name figures/w-hat-empirical-001.pkl --n-jobs 16
+% python joint_inf_plot.py --empirical --general-branch-lengths --delta .001 --plot-name figures/w-hat-empirical-001-marginal.svg --out-pkl-name figures/w-hat-empirical-001-marginal.pkl --n-jobs 16 --marginal
+% or
+% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-001-marginal.svg --in-pkl-name figures/w-hat-empirical-001-marginal.pkl
+% if fit already
+% !! TODO: update with finer-grained plot; this is a placeholder as the global optimization runs...
+\includegraphics[width=\textwidth]{w-hat-empirical-001-marginal}
+\caption[short]{Marginal inference optimizing \eqref{eq:marginal_likelihood}}
+\end{subfigure}
+\begin{subfigure}{.45\linewidth}
+\centering
+% obtained by running
+% python joint_inf_plot.py --empirical --general-branch-lengths --delta .001 --plot-name figures/w-hat-empirical-001.svg --out-pkl-name figures/w-hat-empirical-001.pkl --n-jobs 16
 % or
@@
 \includegraphics[width=\textwidth]{w-hat-empirical-001}
+\caption[short]{Joint inference optimizing \eqref{eq:profile_likelihood}}
+\end{subfigure}
 \caption{
@@
 %EM I don't know what "similar character" means here. I guess you are talking about his Figure 2?
-Looking at the figure in \cite{Felsenstein1978-rr}, the region in Fig.~\ref{fig:inconsistency-farris} has a similar character, albeit a smaller region.
+Looking at the figure in \cite{Felsenstein1978-rr} we see similarities in curve shape to the region in Fig.~\ref{fig:inconsistency-farris}, though the latter has a smaller region of inconsistency.
 Difficulties in phylogenetic estimation when the ``Farris zone'' generates characters have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.

### 2017-11-20 joint_inf_plot.py

@@ -1,15 +1,16 @@
 from __future__ import unicode_literals
 import matplotlib
+matplotlib.use('SVG')
 matplotlib.rcParams['text.usetex'] = True
 matplotlib.rcParams['text.latex.unicode'] = True
 import matplotlib.pyplot as plt
 import matplotlib.patches as mpatches
 import seaborn as sns
-
 import numpy as np
 import argparse
 import sys
 import pickle
+import time
 
 from operator import mul
 from multiprocessing import Pool
@@ -116,6 +117,11 @@ def parse_args():
         '--empirical',
         action="store_true",
         help='compute empirical',
+    )
+    parser.add_argument(
+        '--marginal',
+        action="store_true",
+        help='do marginal likelihood instead of joint inference',
     )
 
     args = parser.parse_args()
@@ -333,16 +339,20 @@ def what_lower_general(x, y):
 
     return -gamma_u**2 * (1+.5*beta) + 2*gamma_l*beta_l*x*y**2 + beta_l**2
 
-def max_likelihood(xy, delta):
+def max_likelihood(xy, delta, joint=True):
     """
     max likelihood as function of marginal likelihood
     """
 
-    what_gen = differential_evolution(lambda z: -FARRIS_LIKELIHOOD(xy[0], xy[1], z[0], z[1], z[2]),
-        bounds=[(delta,1-delta)]*3,
+    minimizer_kwargs = dict(method="L-BFGS-B", bounds=[(delta,1-delta)]*3)
+    what_gen = basinhopping(lambda z: -FARRIS_LIKELIHOOD(xy[0], xy[1], z[0], z[1], z[2], joint=joint),
+        x0=[xy[0], xy[1], xy[1]],
+        minimizer_kwargs=minimizer_kwargs
     )
-    what_res = differential_evolution(lambda z: -FARRIS_LIKELIHOOD(xy[0], xy[1], z[0], z[1], 1.-delta),
-        bounds=[(delta,1-delta)]*2,
+    minimizer_kwargs = dict(method="L-BFGS-B", bounds=[(delta,1-delta)]*2)
+    what_res = basinhopping(lambda z: -FARRIS_LIKELIHOOD(xy[0], xy[1], z[0], z[1], 1.-delta, joint=joint),
+        x0=[xy[0], xy[1]],
+        minimizer_kwargs=minimizer_kwargs
     )
     if -what_res.fun > -what_gen.fun:
         what = 1.-delta
[…]

### 2017-11-20 appendix.tex
optimization section of appendix and marginal plot stuff

@@
 %EM Wanna add a short section about numerical optimization? You've seen that it's not entirely trivial.
+
+\subsection*{Empirical validation}
+
+We use basin-hopping \cite{Wales1997} implemented in \texttt{scipy} \cite{Jones2001} to obtain the plots in Fig.~\ref{fig:bl-general-inconsistency}.
+This method randomly perturbs candidate solutions and accepts or rejects them based on the nearby likelihood surface, eventually obtaining an estimate of the ``global'' optimum.
+While it is not guaranteed to always obtain a global optimum---or even always converge---its implementation here leads us to believe that we can estimate $\hat{w}'$ as well as any optimization method available.
+
+We take various safeguards to ensure convergence and stability of the procedure.
+Since this method involves randomness, we take precautions to not evaluate the likelihood near or outside of boundary conditions, i.e., when $x$ or $y$ are either $0$ or $1$.
+As is apparent from Table~\ref{tab:sitepatprob}, if $0 < x, y < 1$ and we try to evaluate the case of $x'=y'=w'=1$ we have a likelihood of exactly zero.
+Worse still, in Table~\ref{tab:likelihoods} for the Farris topology, in the cases of $\{2,\}, \{1,2\}, \{2,3\},$ and $\{1,2,3\}$, any attempt to evaluate $y'=1$ will similarly yield a zero likelihood.
+In these cases of zero likelihood, methods that evaluate many candidate parameter values can fail to converge when near these boundaries.
+We sidestep these computational issues by restricting ourselves to the region $x,y\in [10^{-3},1-10^{-3}]^2$.
+
+We initialize our optimization procedure at the true branch parameter vector $\{x, y, y\}$.
+We posit that $\hat{w}'$ is equal to one for regions of $x$ and $y$ small; methods that use local optimization may never see candidate values for large $x$ and $y$ if our initial value is small, so we perform two maximizations, one with $w'$ fixed and one where it is estimated.
+We take the value of $\hat{w}'$ with the larger likelihood as our estimate.

### 2017-11-20 joint_inf.tex
optimization section of appendix and marginal plot stuff

@@
 The region of inconsistency in Fig.~\ref{fig:bl-loose-inconsistency} is a conservative estimate of where $\hat{w}=1$.
-To determine how conservative, we use the method of differential evolution \cite{Storn1997} to perform joint maximum likelihood estimation using numerical optimization.
+To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint maximum likelihood estimation using numerical optimization, described in the Appendix.
 Since empirical estimates can be unstable, we perform two maximizations---one over $0 \le x',y',w' \le 1$ and one over $0 \le x',y' \le 1$ with $w'=1$---and take the value of $\hat{w}'$ with the higher likelihood.
 We compute these maxima over a lattice in steps of $10^{-3}$ for $x$ and $y$ from $10^{-3}$ to $1-10^{-3}$---we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
-From this process we obtain Fig.~\ref{fig:bl-general-inconsistency}.
+From this process we obtain the plots in Fig.~\ref{fig:bl-general-inconsistency}.
 
-The estimates for $\hat{w}'$ when $x$ and $y$ are both large look unbiased, with $\hat{w}'$ increasing as $y$ increases, though the region where $\hat{w}'$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
+The case of marginal inference in as expected, where $\hat{w}'$ is equal to $y$ regardless of the value of $x$.
+For joint inference, the estimates for $\hat{w}'$ when $x$ and $y$ are both large look unbiased, with $\hat{w}'$ increasing as $y$ increases, though the region where $\hat{w}'$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
 Although the restricted and general case are similar, comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case has a slightly different character, possibly due to potential looseness of the bound in the restricted case.
@@
 % if fit already
-% !! TODO: update with finer-grained plot; this is a placeholder as the global optimization runs...
 \includegraphics[width=\textwidth]{w-hat-empirical-001-marginal}
@@
     %das: note to update caption
-    Estimates for $\hat{w}'$ when computing empirical estimates of $(\hat{x}', \hat{y}', \hat{w}')$ using differential evolution \cite{Storn1997}.
+    Estimates for $\hat{w}'$ when computing empirical estimates of $(\hat{x}', \hat{y}', \hat{w}')$ using basin-hopping \cite{Wales1997}.
 }

### 2017-11-20 joint_inf.tex
note on LBA

@@
 %EM I don't know what "similar character" means here. I guess you are talking about his Figure 2?
+%das: took a stab at it
 Looking at the figure in \cite{Felsenstein1978-rr} we see similarities in curve shape to the region in Fig.~\ref{fig:inconsistency-farris}, though the latter has a smaller region of inconsistency.
+Moreover, the inconsistency in \cite{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
+This is not the case here; for our case, the inconsistency generally occurs when all branches are long, and has more to do with the choices of the form of the likelihood than from the interplay of long and short branches.
 Difficulties in phylogenetic estimation when the ``Farris zone'' generates characters have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.

### 2017-11-21 appendix.tex
@.@

--- 2017-11-20 appendix.tex
@@
     \hline
-    $\emptyset$    &$1+x'^2+y'^2+4x'y'w'+x'^2y'^2$&$1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
-    $\{1\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
-    $\{2\}$        &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
-    $\{3\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
-    $\{1,2\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
-    $\{1,3\}$      &$1+x'^2+y'^2-4x'y'w'+x'^2y'^2$&$1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
-    $\{2,3\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
-    $\{1,2,3\}$    &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
+    $\emptyset$    &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
+    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
+    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
+    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
+    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
+    $\{1,3\}$      &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
+    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
+    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
     \hline
@@
     $\emptyset$&
-    $(1+x')^2   (1+w')(1+y')^2$\\
+    $(1+x)^2   (1+w)(1+y)^2$\\
      $\{1\}$    &
     $\emptyset$&
-    $(1+x')(1-x')(1+w')(1+y')^2$\\
+    $(1+x)(1-x)(1+w)(1+y)^2$\\
      $\{2\}$    &
     $\emptyset$&
-    $(1+x')^2   (1+w')(1+y')(1-y')$\\
+    $(1+x)^2   (1+w)(1+y)(1-y)$\\
      $\{3\}$    &
     $\emptyset$&
-    $(1+x')(1-x')(1+w')(1+y')^2$\\
+    $(1+x)(1-x)(1+w)(1+y)^2$\\
     $\{1,2\}$  &
     $\{\emptyset,\{1,2\}\}$&
-    $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
+    $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$  &
@@
     $\begin{array}{l}
-                    (1-x')^2   (1+w')(1+y')^2\\
-                    (1+x')^2   (1-w')(1+y')^2\\
-                    (1+x')^2   (1+w')(1-y')^2
+                    (1-x)^2   (1+w)(1+y)^2\\
+                    (1+x)^2   (1-w)(1+y)^2\\
+                    (1+x)^2   (1+w)(1-y)^2
                 \end{array}$\\
@@
                 $\{\emptyset,\{1,2\}\}$&
-                $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
+                $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$&
                 $\{1,2\}$&
-                $(1+x')^2   (1+w')(1+y')(1-y')$\\
+                $(1+x)^2   (1+w)(1+y)(1-y)$\\
     \hline
@@
     \hline
-    $\emptyset$       &$\emptyset$&$(1+x')^2   (1+w')(1+y')^2$\\
+    $\emptyset$       &$\emptyset$&$(1+x)^2   (1+w)(1+y)^2$\\
     $\{1\}$          &
@@
     $\begin{array}{l}
-                        (1+x')(1-x')(1+w')(1+y')^2\\
-                        (1+x')^2   (1-w')(1+y')(1-y')
+                        (1+x)(1-x)(1+w)(1+y)^2\\
+                        (1+x)^2   (1-w)(1+y)(1-y)
                     \end{array}$\\
@@
     $\begin{array}{l}
-                    (1+x')^2   (1+w')(1+y')(1-y')\\
-                    (1+x')(1-x')(1-w')(1+y')^2
+                    (1+x)^2   (1+w)(1+y)(1-y)\\
+                    (1+x)(1-x)(1-w)(1+y)^2
                     \end{array}$\\
@@
     $\begin{array}{l}
-                    (1+x')(1-x')(1+w')(1+y')^2\\
-                    (1+x')^2   (1-w')(1+y')(1-y')
+                    (1+x)(1-x)(1+w)(1+y)^2\\
+                    (1+x)^2   (1-w)(1+y)(1-y)
                     \end{array}$\\
@@
     $\begin{array}{l}
-                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
-                    (1+x')^2   (1-w')(1+y')^2
+                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
+                    (1+x)^2   (1-w)(1+y)^2
                     \end{array}$\\
@@
     $\begin{array}{l}
-                    (1-x')^2   (1+w')(1+y')^2\\
-                    (1+x')(1-x')(1-w')(1+y')(1-y')\\
-                    (1+x')^2   (1+w')(1-y')^2
+                    (1-x)^2   (1+w)(1+y)^2\\
+                    (1+x)(1-x)(1-w)(1+y)(1-y)\\
+                    (1+x)^2   (1+w)(1-y)^2
                     \end{array}$\\
@@
     $\begin{array}{l}
-                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
-                    (1-x')^2   (1-w')(1+y')^2\\
-                    (1+x')^2   (1-w')(1-y')^2
+                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
+                    (1-x)^2   (1-w)(1+y)^2\\
+                    (1+x)^2   (1-w)(1-y)^2
                     \end{array}$\\
@@
     $\begin{array}{l}
-                    (1+x')(1-x')(1-w')(1+y')^2 \\
-                    (1+x')^2   (1+w')(1+y')(1-y')
+                    (1+x)(1-x)(1-w)(1+y)^2 \\
+                    (1+x)^2   (1+w)(1+y)(1-y)
                     \end{array}$\\
@@
 
-Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
-Following \eqref{eq:hadamard_probability},
+Assume $\tau^*=\tau_1$ and $t^*=\{x^*,y^*,y^*\}$.
+Following \eqref{eq:hadamard_probability}, let
 $$
-p_{\emptyset} = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\}),
+p_{\emptyset} := p_{\emptyset}(x^*, y^*) = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=t^*),
 $$
@@
     \label{eq:farris_likelihood}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
-    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \nonumber \\
-    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
-    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\
-    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
-    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
-    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \nonumber \\
-    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
-    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\
+    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t))
+    &=        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw+x^2y^2) \nonumber \\
+    &\qquad + p_{1}          \cdot\log(1-x^2+y^2-x^2y^2) \nonumber \\
+    &\qquad + p_{2}          \cdot\log(1+x^2-y^2-x^2y^2) \nonumber \\
+    &\qquad + p_{3}          \cdot\log(1-x^2+y^2-x^2y^2) \nonumber \\
+    &\qquad + p_{12}         \cdot\log(1-x^2-y^2+x^2y^2) \nonumber \\
+    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw+x^2y^2) \nonumber \\
+    &\qquad + p_{23}         \cdot\log(1-x^2-y^2+x^2y^2) \nonumber \\
+    &\qquad + p_{123}        \cdot\log(1+x^2-y^2-x^2y^2) \nonumber \\
     &\qquad - \log(8) \nonumber \\
-    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \nonumber \\
-    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
-    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \nonumber \\
-    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
-    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\
-    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \nonumber \\
-    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\
-    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
+    &\qquad + p_{\emptyset}  \cdot\log((1+x)^2   (1+w)(1+y)^2) \nonumber \\
+    &\qquad + p_{1}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \nonumber \\
+    &\qquad + p_{2}          \cdot\log((1+x)^2   (1+w)(1+y)(1-y)) \nonumber \\
+    &\qquad + p_{3}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \nonumber \\
+    &\qquad + p_{12}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y)) \nonumber \\
+    &\qquad + p_{13}         \cdot\log((1+x)^2   (1-w)(1+y)^2) \nonumber \\
+    &\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y)) \nonumber \\
+    &\qquad + p_{123}        \cdot\log((1+x)^2   (1+w)(1+y)(1-y)).
 \end{align}
-We use the facts that, for $x',y'\in[0,1]$,
+We use the facts that, for $x,y\in[0,1]$,
 \begin{align*}
-1-x'^2+y'^2-x'^2y'^2 & = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y') \\
-1+x'^2-y'^2-x'^2y'^2 & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y') \\
-1+x'^2+y'^2+x'^2y'^2 & = (1+x'^2)(1+y'^2) \le (1+x')(1+y') \\
-4x'y' & = 2x' \cdot 2y' \le (1+x'^2)(1+y'^2) \le (1+x')(1+y').
+1-x^2+y^2-x^2y^2 & = (1+x)(1-x)(1+y^2) \le (1+x)(1-x)(1+y) \\
+1+x^2-y^2-x^2y^2 & = (1+x^2)(1+y)(1-y) \le (1+x)(1+y)(1-y) \\
+1+x^2+y^2+x^2y^2 & = (1+x^2)(1+y^2) \le (1+x)(1+y) \\
+4xy & = 2x \cdot 2y \le (1+x^2)(1+y^2) \le (1+x)(1+y).
 \end{align*}
@@
 \begin{align*}
-    p_{\emptyset} & = \log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
-                  & \le \log(1+x'^2+y'^2+4x'y'+x'^2y'^2) \\
-                  & \le \log(2(1+x')(1+y'))
+    p_{\emptyset} & = \log(1+x^2+y^2+4xyw+x^2y^2) \\
+                  & \le \log(1+x^2+y^2+4xy+x^2y^2) \\
+                  & \le \log(2(1+x)(1+y))
 \end{align*}
@@
 \begin{align*}
-    p_{13} & = \log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
-                  & \le \log(1+x'^2+y'^2+x'^2y'^2) \\
-                  & \le \log((1+x')(1+y')).
+    p_{13} & = \log(1+x^2+y^2-4xyw+x^2y^2) \\
+                  & \le \log(1+x^2+y^2+x^2y^2) \\
+                  & \le \log((1+x)(1+y)).
 \end{align*}
@@
 \begin{align*}
-&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
-&\qquad \le      p_{\emptyset}  \cdot\log(2(1+x')(1+y'))
-+ p_{1}          \cdot\log((1+x')(1-x')(1+y')) \\
-    &\qquad\qquad + p_{2}          \cdot\log((1+x')(1+y')(1-y'))
-+ p_{3}          \cdot\log((1+x')(1-x')(1+y')) \\
-    &\qquad\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+y')(1-y'))
-+ p_{13}         \cdot\log((1+x')(1+y')) \\
-    &\qquad\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+y')(1-y'))
-+ p_{123}        \cdot\log((1+x')(1+y')(1-y'))
+&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
+&\qquad \le      p_{\emptyset}  \cdot\log(2(1+x)(1+y))
++ p_{1}          \cdot\log((1+x)(1-x)(1+y)) \\
+    &\qquad\qquad + p_{2}          \cdot\log((1+x)(1+y)(1-y))
++ p_{3}          \cdot\log((1+x)(1-x)(1+y)) \\
+    &\qquad\qquad + p_{12}         \cdot\log((1+x)(1-x)(1+y)(1-y))
++ p_{13}         \cdot\log((1+x)(1+y)) \\
+    &\qquad\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+y)(1-y))
++ p_{123}        \cdot\log((1+x)(1+y)(1-y))
 - \log(8) \\
-    &\qquad\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2)
-+ p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
-    &\qquad\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y'))
-+ p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
-    &\qquad\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y'))
-+ p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
-    &\qquad\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y'))
-+ p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
+    &\qquad\qquad + p_{\emptyset}  \cdot\log((1+x)^2   (1+w)(1+y)^2)
++ p_{1}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \\
+    &\qquad\qquad + p_{2}          \cdot\log((1+x)^2   (1+w)(1+y)(1-y))
++ p_{3}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \\
+    &\qquad\qquad + p_{12}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y))
++ p_{13}         \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
+    &\qquad\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y))
++ p_{123}        \cdot\log((1+x)^2   (1+w)(1+y)(1-y)) \\
 &\qquad =      p_{\emptyset}  \cdot\log(2)
-+ \log(1+x')
-+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
-&\qquad\qquad + \log(1+y')
-+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
-&\qquad\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x')
-+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
-&\qquad\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y')
-+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
-&\qquad\qquad + (1-p_{13})\cdot\log(1+w')
-+ p_{13}\cdot\log(1-w')
++ \log(1+x)
++ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x) \\
+&\qquad\qquad + \log(1+y)
++ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
+&\qquad\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x)
++ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x) \\
+&\qquad\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y)
++ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
+&\qquad\qquad + (1-p_{13})\cdot\log(1+w)
++ p_{13}\cdot\log(1-w)
 - \log(8) \\
 &\qquad =      p_{\emptyset}  \cdot\log(2)
-+ (3-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x')
-+ 2(p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
-&\qquad\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y')
-+ 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
-&\qquad\qquad + (1-p_{13})\cdot\log(1+w')
-+ p_{13}\cdot\log(1-w')
++ (3-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x)
++ 2(p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x) \\
+&\qquad\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y)
++ 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
+&\qquad\qquad + (1-p_{13})\cdot\log(1+w)
++ p_{13}\cdot\log(1-w)
 - \log(8).
@@
 \begin{split}
-&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
+&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
 &\qquad \le      p_{\emptyset}  \cdot\log(2)
-+ a_{1}\cdot\log(1+x')
-+ a_{2}\cdot\log(1-x')
-+ a_{3}\cdot\log(1+y') \\
-&\qquad\qquad + a_{4}\cdot\log(1-y')
-+ (1-p_{13})\cdot\log(1+w')
-+ p_{13}\cdot\log(1-w')
++ a_{1}\cdot\log(1+x)
++ a_{2}\cdot\log(1-x)
++ a_{3}\cdot\log(1+y) \\
+&\qquad\qquad + a_{4}\cdot\log(1-y)
++ (1-p_{13})\cdot\log(1+w)
++ p_{13}\cdot\log(1-w)
 - \log(8).
@@
 $$
-\hat{x}' = \frac{a_{1}-a_{2}}{a_{1}+a_{2}}, \ \hat{y}' = \frac{a_{3}-a_{4}}{a_{3}+a_{4}}.
+\hat{x} = \frac{a_{1}-a_{2}}{a_{1}+a_{2}}, \ \hat{y} = \frac{a_{3}-a_{4}}{a_{3}+a_{4}}, \ \hat{w} = 1-2p_{13}.
 $$
-We check that $0 \le \hat{x}' \le 1$ and $0 \le \hat{y}' \le 1$ so these maximizers are valid.
+Trivially $0 \le \hat{w} \le 1$.
+We check that $0 \le \hat{x} \le 1$ and $0 \le \hat{y} \le 1$ so these maximizers are valid.
 Letting
@@
 $$
-\hat{x}' = \frac{3-\tilde{p} - 2\tilde{p}}{3-\tilde{p}+2\tilde{p}} = \frac{3-3\tilde{p}}{3+\tilde{p}},
+\hat{x} = \frac{3-\tilde{p} - 2\tilde{p}}{3-\tilde{p}+2\tilde{p}} = \frac{3-3\tilde{p}}{3+\tilde{p}},
 $$
-and, since $0 \le \tilde{p} \le 1/2$, $0 \le \hat{x}' \le 1$.
-The same approach works for $\hat{y}'$.
+and, since $0 \le \tilde{p} \le 1/2$, $0 \le \hat{x} \le 1$.
+The same approach works for $\hat{y}$.
 The upper bound for the likelihood is then maximized at
 \begin{align*}
-&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
+&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
 &\qquad\le      p_{\emptyset}  \cdot\log(2)
@@
 - \log(8) \\
-&\qquad := L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
+&\qquad := L^{1}_{\tau_1,\tau_1}(x^*, y^*).
 \end{align*}
@@
 $$
-\hat{x}' = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y}' = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
+\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
 $$
@@
 for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
-Using Table~\ref{tab:sitepatprob},
+Using Table~\ref{tab:sitepatprob} and $t^*$,
 $$
-p_{13} = \frac{1}{8}\left(1 + x^2 + y^2 - 4xy^2 + x^2y^2\right).
+p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right).
 $$
@@
 $$
-1 + x^2 + y^2 - 4xy^2 + x^2y^2 \le 2 \iff y^2\left(1 - 4x + x^2\right) \le 1 - x^2
+1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 2 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - (x^*)^2
 $$
@@
 $$
-y^2\left(1 - 4x + x^2\right) \le 1 - 4x + x^2 \le 1 - x^2,
+(y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - 4x^* + (x^*)^2 \le 1 - (x^*)^2,
 $$
-the last inequality holding if $x \le 2$.
-Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}'$ and $\hat{y}'$ are valid.
+the last inequality holding if $x^* \le 2$.
+Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}$ and $\hat{y}$ are valid.
 
@@
 \begin{align*}
-&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
+&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
 &\qquad\le      p_{\emptyset}  \cdot\log(2)
@@
 - \log(8) \\
-&\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}),
+&\qquad := L^{2}_{\tau_1,\tau_1}(x^*, y^*),
 \end{align*}
@@
 - \log(8) \\
-&\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
+&\qquad := L^{3}_{\tau_1,\tau_1}(x^*, y^*).
 \end{align*}
-All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.
+All bounds are functions only of $x,y$ through the true generating probabilities $x^*, y^*$.
 Call the upper bound
 $$
-L_0(x, y) := max(L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}), L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy}), L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy})).
+L_0(x^*, y^*) := \max(L^{1}_{\tau_1,\tau_1}(x^*, y^*), L^{2}_{\tau_1,\tau_1}(x^*, y^*), L^{3}_{\tau_1,\tau_1}(x^*, y^*)).
 $$
@@
 $$
-\max\left((1+x')^2 (1-w')(1+y')^2, (1+x')(1-x')(1+w')(1+y')(1-y')\right).
+\max\left((1+x)^2 (1-w)(1+y)^2, (1+x)(1-x)(1+w)(1+y)(1-y)\right).
 $$
@@
 $$
-(1+x')^2 (1-w')(1+y')^2 > (1+x')(1-x')(1-w')(1+y')(1-y')
+(1+x)^2 (1-w)(1+y)^2 > (1+x)(1-x)(1-w)(1+y)(1-y)
 $$
-always, and thus the maximum can be bounded below by $(1+x')^2 (1-w')(1+y')^2$.
+always, and thus the maximum can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
 For the other site splits we consider two cases.
-If $(1+x')(1-y') > (1-x')(1+y')$, then
+If $(1+x)(1-y) > (1-x)(1+y)$, then
 $$
-(1+x')^2(1-w')(1+y')(1-y') > (1+x')(1-x')(1-w')(1+y')^2
+(1+x)^2(1-w)(1+y)(1-y) > (1+x)(1-x)(1-w)(1+y)^2
 $$
@@
 $$
-(1+x')^2(1-w')(1+y')(1-y').
+(1+x)^2(1-w)(1+y)(1-y).
 $$
@@
 $$
-(1+x')^2(1-w')(1-y')^2 > (1-x')^2(1-w')(1+y')^2
+(1+x)^2(1-w)(1-y)^2 > (1-x)^2(1-w)(1+y)^2
 $$
@@
 $$
-(1+x')^2(1-w')(1-y')^2 > (1+x')(1-x')(1-w')(1+y')(1-y'),
+(1+x)^2(1-w)(1-y)^2 > (1+x)(1-x)(1-w)(1+y)(1-y),
 $$
@@
 $$
-(1+x')^2(1-w')(1-y')^2
+(1+x)^2(1-w)(1-y)^2
 $$
@@
 \begin{align*}
-    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &\ge      p_{\emptyset}  \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
-    &\qquad + p_{1}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
-    &\qquad + p_{2}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
-    &\qquad + p_{3}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
-    &\qquad + p_{12}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
-    &\qquad + p_{123}        \cdot\log((1+x')^2   (1-w')(1+y')(1-y'))\\
-    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1-y')^2) \\
-    &\qquad + p_{23}         \cdot\log((1+x')^2   (1-w')(1-y')^2).
+    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t; \fullAncestralSplitPartitions(\tau_2,t))
+    &\ge      p_{\emptyset}  \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
+    &\qquad + p_{1}          \cdot\log((1+x)^2   (1-w)(1+y)(1-y)) \\
+    &\qquad + p_{2}          \cdot\log((1+x)^2   (1-w)(1+y)(1-y)) \\
+    &\qquad + p_{3}          \cdot\log((1+x)^2   (1-w)(1+y)(1-y)) \\
+    &\qquad + p_{12}         \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
+    &\qquad + p_{123}        \cdot\log((1+x)^2   (1-w)(1+y)(1-y))\\
+    &\qquad + p_{13}         \cdot\log((1+x)^2   (1-w)(1-y)^2) \\
+    &\qquad + p_{23}         \cdot\log((1+x)^2   (1-w)(1-y)^2).
 \end{align*}
-When $(1+x')(1-y') < (1-x')(1+y')$, similar arguments yield the lower bound
+When $(1+x)(1-y) < (1-x)(1+y)$, similar arguments yield the lower bound
 \begin{align*}
-    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &\ge      p_{\emptyset}  \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
-    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
-    &\qquad + p_{2}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
-    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
+    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t; \fullAncestralSplitPartitions(\tau_2,t))
+    &\ge      p_{\emptyset}  \cdot\log((1+x)^2    (1-w)(1+y)^2) \\
+    &\qquad + p_{1}          \cdot\log((1+x)(1-x)(1-w)(1+y)^2) \\
+    &\qquad + p_{2}          \cdot\log((1+x)(1-x)(1-w)(1+y)^2) \\
+    &\qquad + p_{3}          \cdot\log((1+x)(1-x)(1-w)(1+y)^2) \\
     %See above.
-    &\qquad + p_{12}         \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
-    &\qquad + p_{123}        \cdot\log((1+x')(1-x')(1-w')(1+y')^2)\\
-    &\qquad + p_{13}         \cdot\log((1-x')^2    (1-w')(1+y')^2) \\
-    &\qquad + p_{23}         \cdot\log((1-x')^2    (1-w')(1+y')^2).
+    &\qquad + p_{12}         \cdot\log((1+x)^2    (1-w)(1+y)^2) \\
+    &\qquad + p_{123}        \cdot\log((1+x)(1-x)(1-w)(1+y)^2)\\
+    &\qquad + p_{13}         \cdot\log((1-x)^2    (1-w)(1+y)^2) \\
+    &\qquad + p_{23}         \cdot\log((1-x)^2    (1-w)(1+y)^2).
 \end{align*}
@@
 yields
-\begin{align*}
-    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &\ge      2\cdot\log(1+x') \\
-    &\qquad + (2-b)  \cdot\log(1+y') \\
-    &\qquad + b      \cdot\log(1-y') \\
-    &\qquad + \log(1-w')
-\end{align*}
-for the first lower bound, and we get the equivalent expression for the second lower bound by exchanging $x'$ and $y'$.
-Maximizing either lower bound over $\{x',y',w'\}$, we obtain
-\begin{align*}
-    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
-    &\ge      2\cdot\log(2) \\
-    &\qquad + (2-b)  \cdot\log(2-b) \\
-    &\qquad + b      \cdot\log(b)
-\end{align*}
+$$
+    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t; \fullAncestralSplitPartitions(\tau_2,t))
+    \ge      2\cdot\log(1+x)
+    + (2-b)  \cdot\log(1+y)
+    + b      \cdot\log(1-y)
+    + \log(1-w)
+$$
+for the first lower bound, and we get the equivalent expression for the second lower bound by exchanging $x$ and $y$.
+Maximizing either lower bound over $t$, we obtain
+$$
+\tilde{\ell}_{\tau_1,t^*}(\tau_2, t; \fullAncestralSplitPartitions(\tau_2,t))
+\ge      2\cdot\log(2)
++ (2-b)  \cdot\log(2-b)
++ b      \cdot\log(b)
+$$
 %EM Shouldn't this be 2p_{\emptyset}-1 ? It doesn't matter for the bound, but...
@@
 $$
-L_1(x, y) := \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,0\}) + 2\cdot\log(2) + (2-b)\cdot\log(2-b) + b\cdot\log(b).
+L_1(x^*, y^*) := \shannonDivergence_{\tau_1,t^*}(\tau_2,\{1,1-b,0\}) + 2\cdot\log(2) + (2-b)\cdot\log(2-b) + b\cdot\log(b).
 $$
-See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < x', y' < 1$ where $L_1(x', y') > L_0(x', y')$ and joint inference in inconsistent.
+See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < x^*, y^* < 1$ where $L_1(x^*, y^*) > L_0(x^*, y^*)$ and joint inference in inconsistent.
 \end{proof}
@@
 \begin{proof}
-Recalling Table~\ref{tab:likelihoods}, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood for $(x',y',w')$.
+Recalling Table~\ref{tab:likelihoods}, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood over $(x,y,w)$.
 %EM You are talking about a log likelihood here, right? So a likelihood being proportional to means that there is some additive constant. Makes things a little confusing.
 %das: updated
-If $x'$ and $y'$ are fixed, by collapsing like terms and ignoring terms with only $x'$ and $y'$, the log-likelihood in \eqref{eq:farris_likelihood} as a function of $w'$ is
+If $x=x^*$ and $y=y^*$ are fixed, by collapsing like terms and ignoring terms with only $x^*$ and $y^*$, the log-likelihood in \eqref{eq:farris_likelihood} as a function of $w$ is
 \begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
-    &\sim     p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
-    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
-    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
-    &\qquad + p_{13}\cdot\log(1-w')
+    \ell_{\tau_1,t^*}(\tau_1, w; \fullAncestralSplitPartitions(\tau_1,w))
+    &\sim     p_{\emptyset}  \cdot\log(1+(x^*)^2+(y^*)^2+4x^*y^*w+(x^*)^2(y^*)^2) \\
+    &\qquad + p_{13}         \cdot\log(1+(x^*)^2+(y^*)^2-4x^*y^*w+(x^*)^2(y^*)^2) \\
+    &\qquad + (1 - p_{13})\cdot\log(1+w) \\
+    &\qquad + p_{13}\cdot\log(1-w)
 \end{align*}
@@
 \begin{align*}
-    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
-    &\sim     p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
-    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
-    &\qquad + \log(1+w').
+    \ell_{\tau_1,t^*}(\tau_1, w; \fullAncestralSplitPartitions(\tau_1,w))
+    &\sim     p_{\emptyset}  \cdot\log(1+(x^*)^2+(y^*)^2+4x^*y^*w+(x^*)^2(y^*)^2) \\
+    &\qquad + p_{13}         \cdot\log(1+(x^*)^2+(y^*)^2-4x^*y^*w+(x^*)^2(y^*)^2) \\
+    &\qquad + \log(1+w).
 \end{align*}
@@
 $$
-\log(1+w') \geq (1 - p_{13})\cdot\log(1+w') + p_{13}\cdot\log(1-w')
+\log(1+w) \geq (1 - p_{13})\cdot\log(1+w) + p_{13}\cdot\log(1-w)
 $$
-with equality only when $w'=0$, the second likelihood is the likelihood with the maximum-likelihood ancestral state, and so we need only analyze that case.
+with equality only when $w=0$, the second likelihood is the likelihood with the maximum-likelihood ancestral state, and so we need only analyze that case.
 
@@
 $$
-\ell(w') = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
+\ell(w) = \alpha_1\log(\beta+\gamma w) + \alpha_2\log(\beta-\gamma w) + \log(1+w)
 $$
@@
 $$
-\alpha_1 := \alpha_1(x, y) = \frac{1}{8} (1+x^2+y^2+4xy^2+x^2y^2),
+\alpha_1 := \alpha_1(x^*, y^*) = \frac{1}{8} \left(1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
 $$
 $$
-\alpha_2 := \alpha_2(x, y) = \frac{1}{8}(1+x^2+y^2-4xy^2+x^2y^2),
+\alpha_2 := \alpha_2(x^*, y^*) = \frac{1}{8}\left(1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
 $$
 $$
-\beta := \beta(x, y) = 1+x^2+y^2+x^2y^2,
+\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
 $$
 $$
-\gamma := \gamma(x, y) = 4xy.
+\gamma := \gamma(x^*, y^*) = 4x^*y^*.
 $$
-Since $x$ and $y$ fall between zero and one, we make use of the inequalities (excepting the cases of $x=y=0$ and $x=y=1$)
+Since $x^*$ and $y^*$ fall between zero and one, we make use of the inequalities (excepting the cases of $x^*=y^*=0$ and $x^*=y^*=1$)
 $$
@@
 Prob not worth changing now though.
-The derivative of the log-likelihood is
+The derivative of the log-likelihood with respect to $w$ is
 $$
-\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} - \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
+\ell'(w) := \frac{d}{dw} \ell(w) = \frac{\alpha_1 \gamma}{\beta+\gamma w} - \frac{\alpha_2 \gamma}{\beta-\gamma w} + \frac{1}{1+w}.
 $$
-The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w') \lessgtr 0$, we equivalently consider $f(w') \lessgtr 0$ where $f$ is the quadratic function
+The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w) \lessgtr 0$, we equivalently consider $f(w) \lessgtr 0$ where $f$ is the quadratic function
 \begin{align*}
-f(w') &= (w')^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
-      &\qquad + w'\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
+f(w) &= (w)^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
+      &\qquad + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
       &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
@@
 c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
-Excepting all the cases where $x=0$ or where $y=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
-Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w'=0$.
-This implies that if the smaller of the zeros of $f(w')$ is greater than one, then $\hat{w}' \equiv 1$.
+Excepting all the cases where $x^*=0$ or where $y^*=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
+Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w=0$.
+This implies that if the smaller of the zeros of $f(w)$ is greater than one, then $\hat{w} \equiv 1$.
 Using the quadratic formula with
 $$
-a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
+a = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
 $$
 $$
-b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
+b = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
 $$
 $$
-c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
+c = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
 $$
@@
 $$
-\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)},
+\hat{w} = \frac{-b - \sqrt{b^2 - 4ac}}{2a},
 $$
-which is a function of the generating parameters $x$ and $y$.
+which is a function of the generating parameters $x^*$ and $y^*$.
 We see that
 $$
-\hat{w}'(x, y) \ge 1
+\hat{w} \ge 1
 $$
@@
 $$
-a(x, y) + b(x, y) + c(x, y) \ge 0,
+a + b + c \ge 0,
 $$
@@
 \begin{proof}
-In the general case, $\hat{w}'$ is a function of $x$, $y$, $\hat{x}'$, and $\hat{y}'$.
-Looking to the previous section, $\hat{w}'$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}'$ and $\hat{y}'$ instead of $x'$ and $y'$.
-Assume we know $\hat{x}'$ and $\hat{y}'$ as functions of $x$ and $y$ only.
+In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
+Looking to the previous section, $\hat{w}$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
+Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
 Since
@@
 $$
-\alpha_1 - \alpha_2 = xy^2,
+\alpha_1 - \alpha_2 = x^*(y^*)^2,
 $$
@@
 \label{eq:general-bl-result}
--\gamma^2(\hat{x}', \hat{y}')(1 + \frac{1}{2}\beta(x, y)) + 2\gamma(\hat{x}', \hat{y}')\beta(\hat{x}', \hat{y}')xy^2 + \beta^2(\hat{x}', \hat{y}') \ge 0.
+-\gamma^2(\hat{x}, \hat{y})(1 + \frac{1}{2}\beta) + 2\gamma(\hat{x}, \hat{y})\beta(\hat{x}, \hat{y})x^*(y^*)^2 + \beta^2(\hat{x}, \hat{y}) \ge 0.
 \end{equation}
@@
 $$
-\gamma_{L}(x, y) \le \gamma(\hat{x}', \hat{y}') \le \gamma_{U}(x, y)
+\gamma_{L} \le \gamma(\hat{x}, \hat{y}) \le \gamma_{U}
 $$
@@
 $$
-\beta_{L}(x, y) \le \beta(\hat{x}', \hat{y}'),
+\beta_{L} \le \beta(\hat{x}, \hat{y}),
 $$
@@
 $$
--\gamma_{U}^2(x, y)(1 + \frac{1}{2}\beta(x, y)) + 2\gamma_{L}(x, y)\beta_{L}(x, y)xy^2 + \beta_{L}^2(x, y) \ge 0,
+-\gamma_{U}^2(1 + \frac{1}{2}\beta) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0,
 $$
-then \eqref{eq:general-bl-result} holds, and this is an inequality involving only $x$ and $y$.
+then \eqref{eq:general-bl-result} holds, and this is an inequality involving only $x^*$ and $y^*$.
 To get a sense of the region of overestimation, Fig.~\ref{fig:bl-loose-inconsistency} plots an intermediate case.
@@
 %EM Wanna add a short section about numerical optimization? You've seen that it's not entirely trivial.
+
+\subsection*{Empirical validation}
+
+We use basin-hopping \cite{Wales1997} implemented in \texttt{scipy} \cite{Jones2001} to obtain the plots in Fig.~\ref{fig:bl-general-inconsistency}.
+This method randomly perturbs candidate solutions and accepts or rejects them based on the nearby likelihood surface, eventually obtaining an estimate of the ``global'' optimum.
+While it is not guaranteed to always obtain a global optimum---or even always converge---its implementation here leads us to believe that we can estimate $\hat{w}$ as well as any optimization method available.
+
+We take various safeguards to ensure convergence and stability of the procedure.
+Since this method involves randomness, we take precautions to not evaluate the likelihood near or outside of boundary conditions, i.e., when $x^*$ or $y^*$ are either $0$ or $1$.
+As is apparent from Table~\ref{tab:sitepatprob}, if $0 < x^*, y^* < 1$ and we try to evaluate the candidate $\hat{x}=\hat{y}=\hat{w}=1$ we have a likelihood of exactly zero.
+Worse still, in Table~\ref{tab:likelihoods} for the Farris topology, in the cases of $\{2,\}, \{1,2\}, \{2,3\},$ and $\{1,2,3\}$, any attempt to evaluate $\hat{y}=1$ will similarly yield a zero likelihood.
+In these cases of zero likelihood, methods that evaluate many candidate parameter values can fail to converge when near these boundaries.
+We sidestep these computational issues by restricting ourselves to the region $x^*,y^*\in [10^{-3},1-10^{-3}]^2$.
+
+We initialize our optimization procedure at the true branch parameter vector $\{x^*,y^*,y^*\}$.
+We posit that $\hat{w}$ is equal to one for regions of $x^*$ and $y^*$ small; methods that use local optimization may never see candidate values for large $x$ and $y$ if our initial value is small, so we perform two maximizations, one with $w$ fixed and one where it is estimated.
+We take the value of $\hat{w}$ with the larger likelihood as our estimate.

### 2017-11-21 joint_inf.tex
@.@

@@
 \begin{restatable}{theorem}{topoInconsist}
-There exist $L_0(x, y), L_1(x, y)$ and $0 < x', y' < 1$ such that
+Let $t^*=\{x^*, y^*, y^*\}$ and $t=\{x, y, w\}$.
+There exist $L_0(x^*, y^*), L_1(x^*, y^*)$ and $0 < x^*, y^* < 1$ such that
 $$
-\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le L_0(x, y),
+\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le L_0(x^*, y^*),
 $$
 $$
-\max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t) \ge L_1(x, y)
+\max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t) \ge L_1(x^*, y^*)
 $$
-and $L_1(x', y') > L_0(x', y')$.
+with $L_1(x^*, y^*) > L_0(x^*, y^*)$.
 \end{restatable}
-Fig.~\ref{fig:inconsistency-farris} shows the values of $x$ and $y$ where $L_1(x, y) > L_0(x, y)$.
+Fig.~\ref{fig:inconsistency-farris} shows the values of $x^*$ and $y^*$ where $L_1(x^*, y^*) > L_0(x^*, y^*)$.
 Intuitively, $\tau_2$ is favored in performing joint maximum likelihood since the likelihood for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its likelihood while $\tau_2$ has many more.
@@
 We analyze two settings on the Farris topology, corresponding to whether some branch fidelities are fixed at their true values or not.
-As above, assume that data is generated from the ``Farris zone'' with two top branches of fidelity $x$ and all other branches of fidelity $y$.
-We show that, for a nontrivial subset of possible values for $x$ and $y$, the interior branch fidelity parameter $w'$ will be consistently overestimated as exactly equal to one instead of its true value of $y$.
-That is, if we know $x'$ and $y'$, then for
+As above, assume that data is generated from the ``Farris zone'' with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$.
+We show that, for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one instead of its true value of $y^*$.
+That is, if we know $x^*$ and $y^*$, then for
 $$
-\hat{w}' = \arg\max_{w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x,y,w'\}),
+\hat{w} = \arg\max_{w} \ \ell_{\tau_1,\{x^*,y^*,y^*\}}(\tau_1, \{x^*,y^*,w\}),
 $$
-we show there are values for $x$ and $y$ where $\hat{w}'\equiv 1$.
+we show there are values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
 In the general case, we do the same but with
 $$
-(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x',y',w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}).
+(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,\{x^*,y^*,y^*\}}(\tau_1, \{x,y,w\}).
 $$
-These situations are in contrast to the approach using marginal likelihood where necessarily $\hat{w}'=y$ in the limit of infinite data.
+These situations are in contrast to the approach using marginal likelihood where necessarily $\hat{w}=y^*$ in the limit of infinite data.
 
@@
 
-Fix $x'=x$ and $y'=y$ to their true, generating values and proceed to estimate the internal branch parameter $w'$.
+Fix $x=x^*$ and $y=y^*$ to their true, generating values and proceed to estimate the internal branch parameter $w$.
 \begin{restatable}{theorem}{restrictedBranchInconsist}
@@
 $$
-\alpha_1 := \alpha_1(x, y) = \frac{1}{8} \left(1+x^2+y^2+4xy^2+x^2y^2\right),
+\alpha_1 := \alpha_1(x^*, y^*) = \frac{1}{8} \left(1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
 $$
 $$
-\alpha_2 := \alpha_2(x, y) = \frac{1}{8}\left(1+x^2+y^2-4xy^2+x^2y^2\right),
+\alpha_2 := \alpha_2(x^*, y^*) = \frac{1}{8}\left(1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
 $$
 $$
-\beta := \beta(x, y) = 1+x^2+y^2+x^2y^2,
+\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
 $$
 $$
-\gamma := \gamma(x, y) = 4xy.
+\gamma := \gamma(x^*, y^*) = 4x^*y^*.
 $$
-The maximum likelihood value $\hat{w}' = 1$ if
+The maximum likelihood value $\hat{w} = 1$ if
 $$
@@
 \end{restatable}
-The inequality in Theorem~\ref{thm:restricted-bl} is a function of the generating parameters $x$ and $y$.
+The inequality in Theorem~\ref{thm:restricted-bl} is a function of the generating parameters $x^*$ and $y^*$.
 This region is shown in Fig.~\ref{fig:bl-inconsistency}.
@@
 \includegraphics[width=.9\textwidth]{branch-length-inconsistency-inkscape}
-\caption{Region of inconsistency with $x$ and $y$ fixed.}
+\caption{Region of inconsistency with $x=x^*$ and $y=y^*$ fixed.}
 \label{fig:bl-inconsistency}
@@
 
-In the general case, $\hat{w}'$ is a function of $x$, $y$, $\hat{x}'$, and $\hat{y}'$.
-Looking to the previous section, $\hat{w}'$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}'$ and $\hat{y}'$ instead of $x$ and $y$.
-Assume we know $\hat{x}'$ and $\hat{y}'$ as functions of $x$ and $y$ only.
+In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
+Looking to the previous section, $\hat{w}$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
+Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
 \begin{restatable}{theorem}{generalBranchInconsist}
@@
 $$
-\beta(x, y) = 1+x^2+y^2+x^2y^2,
+\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
 $$
 $$
-\gamma(x, y) = 4xy,
+\gamma := \gamma(x^*, y^*) = 4x^*y^*,
 $$
@@
 $$
-\gamma_{L}(x, y) \le \gamma(\hat{x}', \hat{y}') \le \gamma_{U}(x, y)
+\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
+$$
+$$
+\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
 $$
@@
 $$
-\beta_{L}(x, y) \le \beta(\hat{x}', \hat{y}'),
+\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}),
 $$
@@
 $$
--\gamma_{U}^2(x, y)\left(1 + \frac{1}{2}\beta(x, y)\right) + 2\gamma_{L}(x, y)\beta_{L}(x, y)xy^2 + \beta_{L}^2(x, y) \ge 0.
+-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
 $$
 \end{restatable}
-Similar to Theorem~\ref{thm:restricted-bl}, the inequality in Theorem~\ref{thm:general-bl} is also a function of only $x$ and $y$, though general forms for $\gamma_L(x, y)$, $\gamma_U(x, y)$, and $\beta_L(x, y)$ are difficult to obtain.
+Similar to Theorem~\ref{thm:restricted-bl}, the inequality in Theorem~\ref{thm:general-bl} is also a function of only $x^*$ and $y^*$, though general forms for $\gamma_L$, $\gamma_U$, and $\beta_L$ are difficult to obtain.
 The loosest possible bounds are $\gamma_{L} = 0$, $\beta_{L} = 1$ and $\gamma_{U} = 4$, giving
 $$
--16\left(1 + \frac{1}{2}\beta(x, y)\right) + 1 \ge 0,
+-16\left(1 + \frac{1}{2}\beta\right) + 1 \ge 0,
 $$
 which is never true.
-The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma(x, y)$ and $\beta_{L} = \beta(x, y)$, which is the case in Fig.~\ref{fig:bl-inconsistency}.
-For an intermediate bound, define a specified level of error in estimates for $\hat{x}'$ and $\hat{y}'$ so that
+The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the case in Fig.~\ref{fig:bl-inconsistency}.
+For an intermediate bound, define a specified level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
 $$
-x-x_{L} \le \hat{x}' \le x+x_{U}
+x^*-x_{L} \le \hat{x} \le x^*+x_{U}
 $$
@@
 $$
-y-y_{L} \le \hat{y}' \le y+y_{U}.
+y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
 $$
-Then $\gamma_L(x, y), \gamma_U(x, y)$ and $\beta_L(x, y)$ are defined directly from the bounds on $\hat{x}'$ and $\hat{y}'$.
+Then $\gamma_L, \gamma_U$ and $\beta_L$ are defined directly from the bounds on $\hat{x}$ and $\hat{y}$.
 As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
-In this case, we will either estimate $\hat{w}'$ to be one or estimate either $\hat{x}'$ or $\hat{y}'$ to be more than $0.1$ away from its true value.
+In this case, we will either estimate $\hat{w}$ to be one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true value.
 
@@
 \includegraphics[width=.9\textwidth]{bl-loose-inconsistency-inkscape}
-\caption{Region of inconsistency with $x-0.1 \le \hat{x}' \le x+0.1$ and $y-0.1 \le \hat{y}' \le y+0.1$.}
+\caption{Region of inconsistency with $x^*-0.1 \le \hat{x} \le x^*+0.1$ and $y^*-0.1 \le \hat{y} \le y^*+0.1$.}
 \label{fig:bl-loose-inconsistency}
@@
 To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint maximum likelihood estimation using numerical optimization, described in the Appendix.
-Since empirical estimates can be unstable, we perform two maximizations---one over $0 \le x',y',w' \le 1$ and one over $0 \le x',y' \le 1$ with $w'=1$---and take the value of $\hat{w}'$ with the higher likelihood.
-We compute these maxima over a lattice in steps of $10^{-3}$ for $x$ and $y$ from $10^{-3}$ to $1-10^{-3}$---we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
+Since empirical estimates can be unstable, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher likelihood.
+We compute these maxima over a lattice in steps of $10^{-3}$ for $x^*$ and $y^*$ from $10^{-3}$ to $1-10^{-3}$---we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
 From this process we obtain the plots in Fig.~\ref{fig:bl-general-inconsistency}.
 
-The case of marginal inference in as expected, where $\hat{w}'$ is equal to $y$ regardless of the value of $x$.
-For joint inference, the estimates for $\hat{w}'$ when $x$ and $y$ are both large look unbiased, with $\hat{w}'$ increasing as $y$ increases, though the region where $\hat{w}'$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
+The case of marginal inference in as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$.
+For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look unbiased, with $\hat{w}$ increasing as $y^*$ increases, though the region where $\hat{w}$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
 Although the restricted and general case are similar, comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case has a slightly different character, possibly due to potential looseness of the bound in the restricted case.
@@
     %das: note to update caption
-    Estimates for $\hat{w}'$ when computing empirical estimates of $(\hat{x}', \hat{y}', \hat{w}')$ using basin-hopping \cite{Wales1997}.
+    Estimates for $\hat{w}$ when computing empirical estimates of $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997}.
 }

### 2017-11-21 joint_inf_plot.py
use stars

@@ -405,8 +405,8 @@ def main(args=sys.argv[1:]):
 
         ct = plt.contour(X, Y, region, [0], colors='k', linewidths=1)
         plt.axis('scaled')
-        plt.xlabel(r'$x$', fontsize=FONT_SIZE)
-        plt.ylabel(r'$y$', fontsize=FONT_SIZE)
+        plt.xlabel(r'$x^*$', fontsize=FONT_SIZE)
+        plt.ylabel(r'$y^*$', fontsize=FONT_SIZE)
         ttl = plt.title(plottitle, fontsize=FONT_SIZE+2)
         ttl.set_position([.5, 1.05])
         ct.ax.tick_params(labelsize=FONT_SIZE-2)
@@ -443,8 +443,8 @@ def main(args=sys.argv[1:]):
                     X, Y, Z, args.delta = pickle.load(f)
 
             im = plt.imshow(Z, cmap=plt.cm.gray, origin='lower')
-            plt.xlabel(r'$x$', fontsize=FONT_SIZE)
-            plt.ylabel(r'$y$', fontsize=FONT_SIZE)
+            plt.xlabel(r'$x^*$', fontsize=FONT_SIZE)
+            plt.ylabel(r'$y^*$', fontsize=FONT_SIZE)
             ttl = plt.title(r'Value of $\hat{w}$', fontsize=FONT_SIZE+2)
             ttl.set_position([.5, 1.05])
             ax = plt.gca()

### 2017-11-21 joint_inf_plot.py
update plots

@@ -3,6 +3,7 @@ import matplotlib
 matplotlib.use('SVG')
 matplotlib.rcParams['text.usetex'] = True
 matplotlib.rcParams['text.latex.unicode'] = True
+matplotlib.rcParams['figure.autolayout'] = True
 import matplotlib.pyplot as plt
 import matplotlib.patches as mpatches
 import seaborn as sns
@@ -22,7 +23,7 @@ sns.set_style('ticks')
 # ~~~~~~~~~
 # global variables
 
-FONT_SIZE = 16
+FONT_SIZE = 20
 
 ALL_PATTERNS = [
         '0000',
@@ -396,11 +397,11 @@ def main(args=sys.argv[1:]):
             legendtext = r'joint inference inconsistent'
         elif args.restricted_branch_lengths:
             region = what_lower_minus_one(X, Y)
-            plottitle = r'Region of inconsistent branch parameter estimation (restricted case)'
+            plottitle = r'Region of inconsistent branch parameter estimation''\n'r'(restricted case)'
             legendtext = r'$\hat{w}=1$'
         elif args.general_branch_lengths:
             region = what_lower_general(X, Y)
-            plottitle = r'Region of inconsistent branch parameter estimation (general case)'
+            plottitle = r'Region of inconsistent branch parameter estimation''\n'r'(general case)'
             legendtext = r'$\hat{w}=1$ or $\hat{x}$ or $\hat{y}$''\n'r'poorly estimated'
 
         ct = plt.contour(X, Y, region, [0], colors='k', linewidths=1)
@@ -419,11 +420,16 @@ def main(args=sys.argv[1:]):
         if args.restricted_branch_lengths:
             plt.legend([Legend()], [legendtext],
                 handler_map={Legend: LegendHandler()},
-                bbox_to_anchor=(1.05, 1), loc=2, borderaxespad=0., fontsize=FONT_SIZE)
+                bbox_to_anchor=(1.05, 1), loc=2, borderaxespad=0., fontsize=FONT_SIZE-4)
         else:
             plt.legend([Legend()], [legendtext],
-                handler_map={Legend: LegendHandler()}, fontsize=FONT_SIZE)
+                handler_map={Legend: LegendHandler()}, fontsize=FONT_SIZE-4)
 
+        ax = plt.gca()
[…]

### 2017-11-21 appendix.tex
move around plots

@@
 \includegraphics[width=.95\textwidth]{farris_like00}
-\caption[short]{$P(\ancestralStateColumn=\emptyset\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
+\caption[short]{}
 \end{subfigure}
@@
 \centering
-\includegraphics[width=.95\textwidth]{farris_like01}
-\caption[short]{$P(\ancestralStateColumn=\{2\}\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
+\includegraphics[width=.95\textwidth]{farris_like10}
+\caption[short]{}
 \end{subfigure}
-\vskip\baselineskip
-\begin{subfigure}{.45\linewidth}
-\centering
-\includegraphics[width=.95\textwidth]{farris_like10}
-\caption[short]{$P(\ancestralStateColumn=\{1\}\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
-\end{subfigure}
-\begin{subfigure}{.45\linewidth}
-\centering
-\includegraphics[width=.95\textwidth]{farris_like11}
-\caption[short]{$P(\ancestralStateColumn=\{1,2\}\mid \alignmentColumn=\{2,3\},\tau_1,\{x,y,w\})$}
-\end{subfigure}
-\caption{Example likelihood computation}
+\caption{
+    Example likelihood computations given the above branch probability parameterizations.
+    In (a), we compute the product to obtain $P(\ancestralStateColumn=\emptyset\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1+w)/32$.
+    In (b), the same process yields $P(\ancestralStateColumn=\{1\}\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1-w)/32$.
+}
 \label{fig:example_likelihoods}
@@
 
+Consider the Farris tree with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
+We now show that, in the case of the Farris tree topology, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
 Using the Hadamard transform approach, we calculate the generating probabilities on the Farris tree topology.
@@
 
-Assume $\tau^*=\tau_1$ and $t^*=\{x^*,y^*,y^*\}$.
+Assume $\tau^*=\tau_1$.
 Following \eqref{eq:hadamard_probability}, let
@@
 $$
-L_0(x^*, y^*) := \max(L^{1}_{\tau_1,\tau_1}(x^*, y^*), L^{2}_{\tau_1,\tau_1}(x^*, y^*), L^{3}_{\tau_1,\tau_1}(x^*, y^*)).
+L_0(x^*, y^*) := \max\left(L^{1}_{\tau_1,\tau_1}(x^*, y^*), L^{2}_{\tau_1,\tau_1}(x^*, y^*), L^{3}_{\tau_1,\tau_1}(x^*, y^*)\right).
 $$
@@
 In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
-Looking to the previous section, $\hat{w}$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
+From the previous section, $\hat{w}$ is given by the quadratic formula, though now with $\gamma$ and $\beta$ as functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
 Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
@@
 
-We use basin-hopping \cite{Wales1997} implemented in \texttt{scipy} \cite{Jones2001} to obtain the plots in Fig.~\ref{fig:bl-general-inconsistency}.
+We use basin-hopping \cite{Wales1997} implemented in \texttt{scipy} \cite{Jones2001} to obtain the plots in Fig.~\ref{fig:bl-general-inconsistency} and Fig.~\ref{fig:bl-general-marginal}.
 This method randomly perturbs candidate solutions and accepts or rejects them based on the nearby likelihood surface, eventually obtaining an estimate of the ``global'' optimum.
@@
 In these cases of zero likelihood, methods that evaluate many candidate parameter values can fail to converge when near these boundaries.
-We sidestep these computational issues by restricting ourselves to the region $x^*,y^*\in [10^{-3},1-10^{-3}]^2$.
-
-We initialize our optimization procedure at the true branch parameter vector $\{x^*,y^*,y^*\}$.
+We sidestep these computational issues by restricting ourselves to the region $x^*,y^*\in [10^{-2},1-10^{-2}]^2$.
+
+We initialize our optimization procedure at the true branch parameters $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
 We posit that $\hat{w}$ is equal to one for regions of $x^*$ and $y^*$ small; methods that use local optimization may never see candidate values for large $x$ and $y$ if our initial value is small, so we perform two maximizations, one with $w$ fixed and one where it is estimated.
 We take the value of $\hat{w}$ with the larger likelihood as our estimate.
+
+\begin{figure}
+\centering
+% obtained by running
+% python joint_inf_plot.py --empirical --general-branch-lengths --delta .01 --plot-name figures/w-hat-empirical-01-marginal.svg --out-pkl-name figures/w-hat-empirical-01-marginal.pkl --n-jobs 16 --marginal
+% or
+% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-01-marginal.svg --in-pkl-name figures/w-hat-empirical-01-marginal.pkl
+% if fit already
+\includegraphics[width=\textwidth]{w-hat-empirical-01-marginal}
+\caption{
+    Estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:marginal_likelihood}.
+}
+\label{fig:bl-general-marginal}
+\end{figure}

### 2017-11-21 joint_inf.tex
Cleaning out addressed notes.
move around plots

@@
 
-Assume the binary symmetric model, namely with a character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
-%das: @EM is there a good reference to binary symmetric model, or should we trot out the definition and all implications/corollaries of it? I guess it depends where we'd submit?
-%EM I think that this is enough. In the lcfit paper we referenced the Semple and Stell phylogenetics book.
+Assume the binary symmetric model, namely with a character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution \cite{Semple2003-em}.
 Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
@@
 Fidelities have useful algebraic properties, and generating probabilities using the Hadamard transform have an especially simple form (see \eqref{eq:hadamard_probability} in the Appendix).
+For the trees in Fig.~\ref{fig:farris-fels-top}, define the general branch fidelity parameter $t=\{\theta_1,\theta_2,\theta_3,\theta_4,\theta_5\}$ where fidelities are ordered in the order of the taxa with the internal branch last.
 
@@
 \begin{restatable}{theorem}{topoInconsist}
-Let $t^*=\{x^*, y^*, y^*\}$ and $t=\{x, y, w\}$.
+Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
 There exist $L_0(x^*, y^*), L_1(x^*, y^*)$ and $0 < x^*, y^* < 1$ such that
@@
 The goal is that it should be the same size as the font in the text.
-\includegraphics[width=.9\textwidth]{topology-inconsistency-inkscape}
+\includegraphics[width=\textwidth]{topology-inconsistency-inkscape}
 \caption{
@@
 $$
-\hat{w} = \arg\max_{w} \ \ell_{\tau_1,\{x^*,y^*,y^*\}}(\tau_1, \{x^*,y^*,w\}),
+\hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
 $$
@@
 $$
-(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,\{x^*,y^*,y^*\}}(\tau_1, \{x,y,w\}).
+(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\}).
 $$
@@
 % python joint_inf_plot.py --analytic --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
-\includegraphics[width=.9\textwidth]{branch-length-inconsistency-inkscape}
+\includegraphics[width=\textwidth]{branch-length-inconsistency-inkscape}
 \caption{Region of inconsistency with $x=x^*$ and $y=y^*$ fixed.}
@@
 % python joint_inf_plot.py --analytic --general-branch-lengths --delta .001 --plot-name figures/bl-loose-inconsistency.svg
-\includegraphics[width=.9\textwidth]{bl-loose-inconsistency-inkscape}
+\includegraphics[width=\textwidth]{bl-loose-inconsistency-inkscape}
 \caption{Region of inconsistency with $x^*-0.1 \le \hat{x} \le x^*+0.1$ and $y^*-0.1 \le \hat{y} \le y^*+0.1$.}
@@
 Since empirical estimates can be unstable, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher likelihood.
-We compute these maxima over a lattice in steps of $10^{-3}$ for $x^*$ and $y^*$ from $10^{-3}$ to $1-10^{-3}$---we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
-From this process we obtain the plots in Fig.~\ref{fig:bl-general-inconsistency}.
+We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$---we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
+From this process we obtain the plots in Fig.~\ref{fig:bl-general-inconsistency} and Fig.~\ref{fig:bl-general-marginal}.
 
@@
 For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look unbiased, with $\hat{w}$ increasing as $y^*$ increases, though the region where $\hat{w}$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
-Although the restricted and general case are similar, comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case has a slightly different character, possibly due to potential looseness of the bound in the restricted case.
+Comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case is similar to that of the general case.
 This region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have many situations where we will always estimate the interior branch parameter to be one.
+As a positive control, Fig.~\ref{fig:bl-general-marginal} shows the result of optimizing \eqref{eq:marginal_likelihood} using the same procedure.
+Here we obtain the correct $\hat{w}=y^*$.
 
@@
 \centering
-\begin{subfigure}{.45\linewidth}
-\centering
 % obtained by running
-% python joint_inf_plot.py --empirical --general-branch-lengths --delta .001 --plot-name figures/w-hat-empirical-001-marginal.svg --out-pkl-name figures/w-hat-empirical-001-marginal.pkl --n-jobs 16 --marginal
+% python joint_inf_plot.py --empirical --general-branch-lengths --delta .01 --plot-name figures/w-hat-empirical-01.svg --out-pkl-name figures/w-hat-empirical-01.pkl --n-jobs 16
 % or
-% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-001-marginal.svg --in-pkl-name figures/w-hat-empirical-001-marginal.pkl
+% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-01.svg --in-pkl-name figures/w-hat-empirical-01.pkl
 % if fit already
-\includegraphics[width=\textwidth]{w-hat-empirical-001-marginal}
-\caption[short]{Marginal inference optimizing \eqref{eq:marginal_likelihood}}
-\end{subfigure}
-\begin{subfigure}{.45\linewidth}
-\centering
-% obtained by running
-% python joint_inf_plot.py --empirical --general-branch-lengths --delta .001 --plot-name figures/w-hat-empirical-001.svg --out-pkl-name figures/w-hat-empirical-001.pkl --n-jobs 16
-% or
-% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-001.svg --in-pkl-name figures/w-hat-empirical-001.pkl
-% if fit already
-\includegraphics[width=\textwidth]{w-hat-empirical-001}
-\caption[short]{Joint inference optimizing \eqref{eq:profile_likelihood}}
-\end{subfigure}
+\includegraphics[width=\textwidth]{w-hat-empirical-01}
 \caption{
-    %das: note to update caption
-    Estimates for $\hat{w}$ when computing empirical estimates of $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997}.
+    Estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:profile_likelihood}.
 }
@@
 In this case, the converse is shown: when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
-%EM Re LBA I was only thinking of putting a little sentence here that says that thus this is _not_ long branch attraction.
-%EM I don't know what "similar character" means here. I guess you are talking about his Figure 2?
-%das: took a stab at it
 Looking at the figure in \cite{Felsenstein1978-rr} we see similarities in curve shape to the region in Fig.~\ref{fig:inconsistency-farris}, though the latter has a smaller region of inconsistency.
@@
 Difficulties in phylogenetic estimation when the ``Farris zone'' generates characters have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.
-
-%EM We should be clear here that thus this is not "long branch attraction".
-I haven't read the Farris papers, but I wonder what his argument is and if it relates to this, in that there are too many parameters.
-Probably something should go in the intro as well pointing out that this is not LBA.
-%das: Noted. I will work on this.
-%das: still not quite sure about what to say re LBA, but I'm thinking about it
-%Discussion of number of parameters of each.
 

### 2017-11-27 joint_inf.tex
update fig and add to discussion

@@
 Fidelities have useful algebraic properties, and generating probabilities using the Hadamard transform have an especially simple form (see \eqref{eq:hadamard_probability} in the Appendix).
-For the trees in Fig.~\ref{fig:farris-fels-top}, define the general branch fidelity parameter $t=\{\theta_1,\theta_2,\theta_3,\theta_4,\theta_5\}$ where fidelities are ordered in the order of the taxa with the internal branch last.
+For a four-taxon tree, define the general branch fidelity parameter $t=\{\theta_1,\theta_2,\theta_3,\theta_4,\theta_5\}$ where fidelities are ordered in the order of the taxa with the internal branch last.
+The trees in Fig.~\ref{fig:farris-fels-top} have $\theta_1=\theta_3=x$, $\theta_2=\theta_4=y$, and $\theta_5=w$.
 
@@
 
-The case of marginal inference in as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$.
+The case of marginal inference is as expected, where $\hat{w}$ is equal to $y^*$ in the limit of infinite data regardless of the value of $x^*$.
 For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look unbiased, with $\hat{w}$ increasing as $y^*$ increases, though the region where $\hat{w}$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
-Comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case is similar to that of the general case.
-This region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have many situations where we will always estimate the interior branch parameter to be one.
+Comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case is similar to that of the restricted case.
+This region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have many situations where we will estimate the interior branch parameter to be one.
 As a positive control, Fig.~\ref{fig:bl-general-marginal} shows the result of optimizing \eqref{eq:marginal_likelihood} using the same procedure.
-Here we obtain the correct $\hat{w}=y^*$.
+Here we obtain the correct $\hat{w}\rightarrow y^*$.
 
@@
 \section*{Discussion}
+
+We have shown that the method of joint inference when estimating phylogenetic parameters is not consistent in general.
+Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, inferring the incorrect topology and, second, estimating severely biased branch lengths even given the correct topology.
+In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities are closer to zero than to one.
 

### 2017-11-28 joint_inf.tex
rough cut abstract and consistent wording

@@
 \begin{abstract}
-TODO
+Maximum-likelihood phylogenetics is an approach to estimate the phylogenetic parameters of topology and branch length given observed character data.
+Since a phylogeny has unknown intermediate states, computing a maximum likelihood solution requires dealing with these latent variables.
+Classically, the method of marginal likelihood is employed: the unknown intermediate states typically come from a small set, and so averaging over their finite space allows for consistent estimation of the parameters of interest.
+Recently, a computationally-efficient approach has been proposed to maximize over these unknown states as opposed to averaging over them.
+While this method of maximum likelihood has the potential to obtain estimates more efficiently, its properties are not well-understood.
+We show that, like other methods of profile likelihood where the number of unknowns grows with observations, this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
+There are a number of cases when generating under a four-taxon tree results in a joint estimation procedure converging on the wrong topology.
+If the topology is known a priori, there are similar cases where the branch lengths are consistently overestimated.
+While these results do not show joint inference is always inconsistent, they indicate further study into where care can be taken when using this approach.
 \end{abstract}
@@
 
-Classical maximum-likelihood (ML) estimation in phylogenetics operates by integrating out possible ancestral states at the internal nodes of the tree.
+Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out possible ancestral states at the internal nodes of the tree.
 Recently, \cite{Sagulenko2017-jo} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences.
@@
 
-Existing consistency proofs for maximum-likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
+Existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
 These proofs do not readily extend to include estimating ancestral states.

### 2017-11-28 joint_inf.tex
quick edit

@@
 \begin{abstract}
-Maximum-likelihood phylogenetics is an approach to estimate the phylogenetic parameters of topology and branch length given observed character data.
+Maximum likelihood phylogenetics is an approach to estimate the phylogenetic parameters of topology and branch length given observed character data.
 Since a phylogeny has unknown intermediate states, computing a maximum likelihood solution requires dealing with these latent variables.
-Classically, the method of marginal likelihood is employed: the unknown intermediate states typically come from a small set, and so averaging over their finite space allows for consistent estimation of the parameters of interest.
+Classically, the method of marginal likelihood is employed: the unknown intermediate states are averaged over, allowing for consistent estimation of the parameters of interest.
 Recently, a computationally-efficient approach has been proposed to maximize over these unknown states as opposed to averaging over them.
@@
 We show that, like other methods of profile likelihood where the number of unknowns grows with observations, this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
-There are a number of cases when generating under a four-taxon tree results in a joint estimation procedure converging on the wrong topology.
+There are a number of cases where generating under a four-taxon tree results in a joint estimation procedure converging on the wrong topology.
 If the topology is known a priori, there are similar cases where the branch lengths are consistently overestimated.
-While these results do not show joint inference is always inconsistent, they indicate further study into where care can be taken when using this approach.
+While these results do not imply joint inference is always inconsistent, they indicate care must be taken when using this approach.
 \end{abstract}

### 2017-11-30 appendix.tex
minor edits; removing redundancy

@@
 \end{equation*}
-so that, in the limit of infinite data,
+so that, as $\nCols\rightarrow\infty$,
 \begin{equation}
@@
     \includegraphics[width=.45\textwidth]{unrooted_farris_tree}
-    \caption{Four taxon ``Farris zone'' tree.}
+    \caption{Four taxon ``Farris zone'' topology.}
 \label{fig:four-taxa-tree}
@@
 \end{tabular}
-\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
+\caption{Likelihood calculations for all site patterns and internal states of Farris tree.
 All values multiplied by $1/32$.}
@@
 \end{tabular}
-\caption{Likelihood calculations for all site patterns and internal states of Felsenstein topology.
+\caption{Likelihood calculations for all site patterns and internal states of Felsenstein tree.
 All values multiplied by $1/32$.}
@@
 Consider the Farris tree with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
-We now show that, in the case of the Farris tree topology, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
-Using the Hadamard transform approach, we calculate the generating probabilities on the Farris tree topology.
+We now show that, in the case of the Farris tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
+Using the Hadamard transform approach, we calculate the generating probabilities on the Farris tree.
 For site split $\emptyset$,
@@
 The analogous statement holds for $y_1$ and $y_2$.
-The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
+The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
 
@@
 As is apparent from Table~\ref{tab:sitepatprob}, if $0 < x^*, y^* < 1$ and we try to evaluate the candidate $\hat{x}=\hat{y}=\hat{w}=1$ we have a likelihood of exactly zero.
-Worse still, in Table~\ref{tab:likelihoods} for the Farris topology, in the cases of $\{2,\}, \{1,2\}, \{2,3\},$ and $\{1,2,3\}$, any attempt to evaluate $\hat{y}=1$ will similarly yield a zero likelihood.
+Worse still, in Table~\ref{tab:likelihoods} for the Farris tree, in the cases of $\{2,\}, \{1,2\}, \{2,3\},$ and $\{1,2,3\}$, any attempt to evaluate $\hat{y}=1$ will similarly yield a zero likelihood.
 In these cases of zero likelihood, methods that evaluate many candidate parameter values can fail to converge when near these boundaries.

### 2017-11-30 joint_inf.tex
authors
minor edits; removing redundancy

@@
 \title{Joint estimation of ancestral states and phylogenetic model parameters is inconsistent}
-\author{Shaw, Matsen, Minin}
+\author{David A. Shaw \and Frederick A. Matsen IV}
 
@@
 \begin{abstract}
-Maximum likelihood phylogenetics is an approach to estimate the phylogenetic parameters of topology and branch length given observed character data.
-Since a phylogeny has unknown intermediate states, computing a maximum likelihood solution requires dealing with these latent variables.
-Classically, the method of marginal likelihood is employed: the unknown intermediate states are averaged over, allowing for consistent estimation of the parameters of interest.
-Recently, a computationally-efficient approach has been proposed to maximize over these unknown states as opposed to averaging over them.
-While this method of maximum likelihood has the potential to obtain estimates more efficiently, its properties are not well-understood.
+Maximum likelihood is a widespread and well-understood approach to estimate the phylogenetic parameters of topology and branch length given observed character data.
+Since a phylogeny has unknown ancestral states, computing a maximum likelihood solution becomes a problem involving latent variables.
+Classically, the method of marginal likelihood is employed for this problem: the unknown intermediate states are averaged over, allowing for consistent estimation of the parameters of interest.
+Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
+While this method of profile maximum likelihood can obtain estimates more efficiently, it has the potential to result in undesired properties of the estimates.
 We show that, like other methods of profile likelihood where the number of unknowns grows with observations, this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
 There are a number of cases where generating under a four-taxon tree results in a joint estimation procedure converging on the wrong topology.
-If the topology is known a priori, there are similar cases where the branch lengths are consistently overestimated.
+If the topology is known a priori, there are similar cases where the branch lengths are consistently overestimated in the limit of infinite data.
 While these results do not imply joint inference is always inconsistent, they indicate care must be taken when using this approach.
@@
 Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
-Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$ and limit themselves to optimizing branch lengths using this method, their work motivates understanding when exactly this approximation breaks down, and what happens when optimizing branch lengths and tree topology using this approximation.
+Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximation breaks down.
+%das: @EM do we need to point out that they limit themselves to BLs? this seems like text that was written before we showed the fixed-topology inconsistency
+Additionally, they limit themselves to optimizing branch lengths holding the topology fixed, a limitation we relax by considering both topology and branch length estimation.
 
@@
 %EM Do you know that "topology" means tree without branch lengths? Therefore I think "tree" is more appropriate in the following sentence and elsewhere.
+%das: I believe all of this terminology is fixed now
 In the classical configuration of each of these trees, the interior branch length parameters are equal to the bottom two parameters.
@@
 As described above, we will use the equivalent but different notion of branch fidelities.
-We analyze two settings on the Farris topology, corresponding to whether some branch fidelities are fixed at their true values or not.
+We analyze two settings on the Farris tree, corresponding to whether some branch fidelities are fixed at their true values or not.
 As above, assume that data is generated from the ``Farris zone'' with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$.
@@
 $$
-These situations are in contrast to the approach using marginal likelihood where necessarily $\hat{w}=y^*$ in the limit of infinite data.
+These situations are in contrast to the approach using marginal likelihood where necessarily $\hat{w}$ converges to $y^*$ as the number of observations grows.
 
@@
 
-The case of marginal inference is as expected, where $\hat{w}$ is equal to $y^*$ in the limit of infinite data regardless of the value of $x^*$.
+The case of marginal inference is as expected, where $\hat{w}$ is equal to $y^*$ in the limit data regardless of the value of $x^*$.
 For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look unbiased, with $\hat{w}$ increasing as $y^*$ increases, though the region where $\hat{w}$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
 Comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case is similar to that of the restricted case.
-This region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have many situations where we will estimate the interior branch parameter to be one.
+This region encompasses the majority of the parameter space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch parameter to be one.
 As a positive control, Fig.~\ref{fig:bl-general-marginal} shows the result of optimizing \eqref{eq:marginal_likelihood} using the same procedure.
@@
 \section*{Acknowledgements}
-We thank Richard Neher for helpful discussions.
+We thank Richard Neher and Vladimir Minin for helpful discussions.
 
-David Shaw, Vladimir Minin, and Frederick Matsen were supported by NIH grants U19-AI117891 and R01-GM113246; David Shaw and Frederick Matsen were also supported by R01-AI12096.
+David Shaw and Frederick Matsen were supported by NIH grants U19-AI117891 and R01-GM113246 and by R01-AI12096.
 The research of Frederick Matsen was supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation.

### 2017-12-01 joint_inf.tex
A quick pass through the main text.

@@
 \begin{abstract}
+%EM This abstract is really nice! :+1:
 Maximum likelihood is a widespread and well-understood approach to estimate the phylogenetic parameters of topology and branch length given observed character data.
 Since a phylogeny has unknown ancestral states, computing a maximum likelihood solution becomes a problem involving latent variables.
-Classically, the method of marginal likelihood is employed for this problem: the unknown intermediate states are averaged over, allowing for consistent estimation of the parameters of interest.
+Classically, these unknown intermediate states are averaged out, which leads to consistent estimation of the topology and continuous model parameters.
 Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
+%EM I think that should either use "joint" or "profile" just to have a single word to describe things for most of the text, except when we connect it to the stats literature.
 While this method of profile maximum likelihood can obtain estimates more efficiently, it has the potential to result in undesired properties of the estimates.
@@
 If the topology is known a priori, there are similar cases where the branch lengths are consistently overestimated in the limit of infinite data.
-While these results do not imply joint inference is always inconsistent, they indicate care must be taken when using this approach.
+Although joint maximum likelihood may turn out to be a useful tool in practice, care should be taken when using this approach.
 \end{abstract}
@@
 Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out possible ancestral states at the internal nodes of the tree.
-Recently, \cite{Sagulenko2017-jo} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences.
+Recently, \cite{Sagulenko2017-jo} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences on a fixed topology.
 This is attractive from a computational perspective: such joint ML inference can proceed according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
@@
 Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
-Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximation breaks down.
-%das: @EM do we need to point out that they limit themselves to BLs? this seems like text that was written before we showed the fixed-topology inconsistency
-Additionally, they limit themselves to optimizing branch lengths holding the topology fixed, a limitation we relax by considering both topology and branch length estimation.
+Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximate technique breaks down, even in the limit of infinite sequence data.
 
@@
 We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``Farris'' and ``Felsenstein'' trees (Fig.~\ref{fig:farris-fels-top}).
-%EM Do you know that "topology" means tree without branch lengths? Therefore I think "tree" is more appropriate in the following sentence and elsewhere.
-%das: I believe all of this terminology is fixed now
 In the classical configuration of each of these trees, the interior branch length parameters are equal to the bottom two parameters.
@@
 \centering
-%EM Wouldn't it be better to just parameterize these with x,y,w without primes?
-Also, it looks super pro if we can use tex fonts in this figure too.
-I love the mathjax fonts, which I think you can get https://packages.ubuntu.com/trusty/fonts/fonts-mathjax .
-LMK if you run into trouble.
-The font I use is "MathJax_Math".
-%das: I think I'm either going to go with something other than the primes, like z_1, z_2, z_3, or replace x and y with greek letters or whatever, but it'll be a little bit of a tedious edit.
-%EM You mean globally in the whole document? Let's kick around ideas if so.
-%das: I think we just need the true generating parameters---x^* and y^*---and the estimands---x', y', w'.
-I've been bad about using x and y instead of x^* and y^* for function arguments because it's cleaner, and I think there is inconsistency in using the primes and not.
-How about Greek for unknown (truth) and Latin for estimands?
-\alpha and \beta for generating parameters and x, y, z for estimands?
-Would have to change some other notation but shouldn't be too hard I hope.
-%EM Wow, this is going to be some work.
-Sounds like a good plan, but I'd prefer for a system in which it's easy to remember which greek goes with which latin.
-could we do a, b, g and alpha, beta, gamma or something?
-Let's keep discussing before making the big change.
-\includegraphics[width=.95\textwidth]{farris_blank}
+\includegraphics[width=\textwidth]{farris_blank}
 \caption[short]{Farris tree $\tau_1$}
@@
 \centering
-\includegraphics[width=.95\textwidth]{felsenstein_blank}
+\includegraphics[width=\textwidth]{felsenstein_blank}
 \caption[short]{Felsenstein tree $\tau_2$}
@@
 
-We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$ where $0 \le \theta \le 1$.
-We follow \cite{Matsen2007-jq} in calling this edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge.
+We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with an alternate formulation called ``fidelity.''
+The probability of a substitution on a branch with fidelity $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$ where $0 \le \theta \le 1$.
+This parameter thus quantifies the fidelity of transmission of the ancestral state across an edge \cite{Matsen2007-jq}.
+
 Fidelities have useful algebraic properties, and generating probabilities using the Hadamard transform have an especially simple form (see \eqref{eq:hadamard_probability} in the Appendix).
@@
 An alternative approach \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
-This is known as a profile likelihood \cite{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
+This is known in statistics as a profile likelihood \cite{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
 \begin{equation}
@@
 \end{equation}
-We will use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximization \eqref{eq:profile_likelihood}.
+We will use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximization of \eqref{eq:profile_likelihood}.
 One can then estimate a topology and branch fidelities using this profile likelihood:
@@
 Assume $\fullAlignment$ is generated from topology $\tau^*$ and branch fidelities $t^*$.
-Define $\ell_{\tau^*,t^*}(\tau, t)$ such that
+Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log likelihood, which can be thought of as the infinite-length sequence limit
 $$
@@
 \begin{restatable}{theorem}{topoInconsist}
+%EM Wow this is hard to parse!
+%EM 1. The main message is that there is a region on which the wrong topology is preferred. This should be stated in words before the theorem, with the link to the figure.
+%EM 2. To see that in the theorem, we have to wind our way through inequalities facing different ways. Can we just say that there is a region? What do these L's tell us?
+%EM 3. L_0 and L_1 are totally different that L_n, but with very similar notation.
 Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
@@
 % python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
-%EM Sorry I didn't notice this before, but the font size everywhere in the plots should be at least doubled before submission.
-Fine if you want to wait one more cycle.
-The goal is that it should be the same size as the font in the text.
 \includegraphics[width=\textwidth]{topology-inconsistency-inkscape}
@@
     Regions of inconsistency.
-    Due to the looseness of the upper and lower bounds, the parameters in the top right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
+    Due to the looseness of the upper and lower bounds, the parameters in the upper right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
 }
@@
 
-We now consider the problem of branch fidelity estimation on the correct tree using joint ancestral state and branch fidelity estimation.
+We now consider the problem of branch length estimation on the correct tree using joint ancestral state and branch length estimation.
 As described above, we will use the equivalent but different notion of branch fidelities.
@@
 As above, assume that data is generated from the ``Farris zone'' with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$.
-We show that, for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one instead of its true value of $y^*$.
+In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one instead of its true value of $y^*$.
 That is, if we know $x^*$ and $y^*$, then for
@@
 $$
-we show there are values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
+we show there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
 In the general case, we do the same but with
@@
 $$
+Again, we find a region for which the inferred values do not converge to the generating values.
 These situations are in contrast to the approach using marginal likelihood where necessarily $\hat{w}$ converges to $y^*$ as the number of observations grows.
@@
 
+The general case is more challenging to analyze and so we will get somewhat weaker bounds.
 In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
+%EM This is the first we're hearing of the quadratic formula in this version.
 Looking to the previous section, $\hat{w}$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
 Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
+%EM Here would be a good place to introduce the concept that youre going to be saying that either the pendant branch lengths are bad, or what is going to be 1.
+%EM differing notions of "bad" will lead to different theorems.
 \begin{restatable}{theorem}{generalBranchInconsist}
@@
 Similar to Theorem~\ref{thm:restricted-bl}, the inequality in Theorem~\ref{thm:general-bl} is also a function of only $x^*$ and $y^*$, though general forms for $\gamma_L$, $\gamma_U$, and $\beta_L$ are difficult to obtain.
+%EM I think you put this loosest bound in for exposition's sake, but I don't think it helps clarify.
 The loosest possible bounds are $\gamma_{L} = 0$, $\beta_{L} = 1$ and $\gamma_{U} = 4$, giving
@@
 which is never true.
+%EM On the other hand, I think that explaining that the restricted case fits in this framework is great.
 The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the case in Fig.~\ref{fig:bl-inconsistency}.
@@
 \includegraphics[width=\textwidth]{bl-loose-inconsistency-inkscape}
+%EM If you read this caption without reading the text, would you be able to understand what's going on?
+I think I'd be confused because in the caption you are stating something that looks good, and then the figure describes xhat and yhat as being poorly estimated.
 \caption{Region of inconsistency with $x^*-0.1 \le \hat{x} \le x^*+0.1$ and $y^*-0.1 \le \hat{y} \le y^*+0.1$.}
@@
 Since empirical estimates can be unstable, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher likelihood.
-We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$---we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
+We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$: we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
+%EM Noo, I need to make a name for this sort of sentence-- the figure announcement sentence? Sentences should read "Here is a statement (Fig. X)."
+%EM There is a big bomb to drop here, which is that what is 1 all over the place!
+%EM Question: it seems to me that even in the non-white region, the BL inferences aren't correct.
+Is that true?
+We should say so if.
+I don't want to make extra work for you, but a figure in which you take the difference between the correct value and the inferred value might be nice.
 From this process we obtain the plots in Fig.~\ref{fig:bl-general-inconsistency} and Fig.~\ref{fig:bl-general-marginal}.
 
-The case of marginal inference is as expected, where $\hat{w}$ is equal to $y^*$ in the limit data regardless of the value of $x^*$.
+Marginal inference performs is as expected, where $\hat{w}$ is equal to $y^*$ in the limit data regardless of the value of $x^*$.
 For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look unbiased, with $\hat{w}$ increasing as $y^*$ increases, though the region where $\hat{w}$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
@@
 
-The method of profile likleihood is well-studied \cite{Murphy2000-ry}.
+The method of profile likelihood is well-studied \cite{Murphy2000-ry}.
 Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., \cite{Geman1982}.
@@
 In phylogenetics, counterexamples showing methods of phylogenetic inference are inconsistent have been shown using the same four-taxa topologies used here \cite{Felsenstein1978-rr}.
+%EM Seems like a sentence is missing here-- what's the usual setting?
 In this case, the converse is shown: when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
-Looking at the figure in \cite{Felsenstein1978-rr} we see similarities in curve shape to the region in Fig.~\ref{fig:inconsistency-farris}, though the latter has a smaller region of inconsistency.
+Looking at Fig~2 of \cite{Felsenstein1978-rr} we see similarities in curve shape to the region in Fig.~\ref{fig:inconsistency-farris}, though the latter has a smaller region of inconsistency.
 Moreover, the inconsistency in \cite{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
@@
 Difficulties in phylogenetic estimation when the ``Farris zone'' generates characters have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.
+
+%EM We need to describe limitations (we weren't able to write out equations getting the region exactly right, and we are using a simple model of subs.
+%EM We should also describe the need for extensive simulations to see the extent of bias.
 

### 2017-12-01 appendix.tex
adding some clarifications

@@
 $$
-L_0(x^*, y^*) := \max\left(L^{1}_{\tau_1,\tau_1}(x^*, y^*), L^{2}_{\tau_1,\tau_1}(x^*, y^*), L^{3}_{\tau_1,\tau_1}(x^*, y^*)\right).
+C_0(x^*, y^*) := \max\left(L^{1}_{\tau_1,\tau_1}(x^*, y^*), L^{2}_{\tau_1,\tau_1}(x^*, y^*), L^{3}_{\tau_1,\tau_1}(x^*, y^*)\right).
 $$
@@
 $$
-L_1(x^*, y^*) := \shannonDivergence_{\tau_1,t^*}(\tau_2,\{1,1-b,0\}) + 2\cdot\log(2) + (2-b)\cdot\log(2-b) + b\cdot\log(b).
-$$
-See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < x^*, y^* < 1$ where $L_1(x^*, y^*) > L_0(x^*, y^*)$ and joint inference in inconsistent.
+C_1(x^*, y^*) := \shannonDivergence_{\tau_1,t^*}(\tau_2,\{1,1-b,0\}) + 2\cdot\log(2) + (2-b)\cdot\log(2-b) + b\cdot\log(b).
+$$
+See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < x^*, y^* < 1$ where $C_1(x^*, y^*) > C_0(x^*, y^*)$ and joint inference in inconsistent.
 \end{proof}

### 2017-12-01 joint_inf.tex
adding some clarifications

@@
 %EM I think that should either use "joint" or "profile" just to have a single word to describe things for most of the text, except when we connect it to the stats literature.
-While this method of profile maximum likelihood can obtain estimates more efficiently, it has the potential to result in undesired properties of the estimates.
-We show that, like other methods of profile likelihood where the number of unknowns grows with observations, this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
+While this method of joint maximum likelihood estimation can obtain estimates more efficiently, it has the potential to result in undesired properties of the estimates.
+We show that, like other methods of maximum likelihood where the number of unknowns grows with observations, this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
 There are a number of cases where generating under a four-taxon tree results in a joint estimation procedure converging on the wrong topology.
@@
 
+To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the Farris topology.
+We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology than for the true, generating Farris topology.
+To do so, we construct an upper bound for the likelihood given the Farris topology as a function of $x^*$ and $y^*$, and, similarly, we construct a lower bound for the likelihood given the Felsenstein topology.
+The values of $x^*$ and $y^*$ where the likelihood in the Felsenstein case is larger than the likelihood in the Farris case indicate an inconsistency (Fig.~\ref{fig:inconsistency-farris}).
 \begin{restatable}{theorem}{topoInconsist}
@@
 %EM 3. L_0 and L_1 are totally different that L_n, but with very similar notation.
+%das: is this better, or did you want to get rid of the upper/lower bound constants entirely?
+I have them since they are a closed form function of x* and y* so it's clear that all we need is a region where some f(x*, y*) > 0.
+Are you saying the paragraph I just put before the theorem is a better statement of the theorem?
 Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
-There exist $L_0(x^*, y^*), L_1(x^*, y^*)$ and $0 < x^*, y^* < 1$ such that
+There exist $C_0(x^*, y^*), C_1(x^*, y^*)$ and $0 < x^*, y^* < 1$ such that
 $$
-\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le L_0(x^*, y^*),
+\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le C_0(x^*, y^*),
 $$
 $$
-\max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t) \ge L_1(x^*, y^*)
+\max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t) \ge C_1(x^*, y^*)
 $$
-with $L_1(x^*, y^*) > L_0(x^*, y^*)$.
+with $C_1(x^*, y^*) > C_0(x^*, y^*)$.
 \end{restatable}
-Fig.~\ref{fig:inconsistency-farris} shows the values of $x^*$ and $y^*$ where $L_1(x^*, y^*) > L_0(x^*, y^*)$.
+Fig.~\ref{fig:inconsistency-farris} shows the values of $x^*$ and $y^*$ where $C_1(x^*, y^*) > C_0(x^*, y^*)$.
 Intuitively, $\tau_2$ is favored in performing joint maximum likelihood since the likelihood for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its likelihood while $\tau_2$ has many more.
@@
 %EM This is the first we're hearing of the quadratic formula in this version.
-Looking to the previous section, $\hat{w}$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
+Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
 Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
@@
 %EM differing notions of "bad" will lead to different theorems.
+We show that there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
+We fix bounds on these estimates and show that, in the general case, we either estimate $\hat{w}$ to be one or fall outside of the specified bounds for $\hat{x}$ or $\hat{y}$, indicating a poor estimate in at least one of the three unknown branch parameters.
 \begin{restatable}{theorem}{generalBranchInconsist}
@@
 %EM I think you put this loosest bound in for exposition's sake, but I don't think it helps clarify.
-The loosest possible bounds are $\gamma_{L} = 0$, $\beta_{L} = 1$ and $\gamma_{U} = 4$, giving
-$$
--16\left(1 + \frac{1}{2}\beta\right) + 1 \ge 0,
-$$
-which is never true.
 %EM On the other hand, I think that explaining that the restricted case fits in this framework is great.
-The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the case in Fig.~\ref{fig:bl-inconsistency}.
+The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section and given in Fig.~\ref{fig:bl-inconsistency}.
 For an intermediate bound, define a specified level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
@@
 I think I'd be confused because in the caption you are stating something that looks good, and then the figure describes xhat and yhat as being poorly estimated.
-\caption{Region of inconsistency with $x^*-0.1 \le \hat{x} \le x^*+0.1$ and $y^*-0.1 \le \hat{y} \le y^*+0.1$.}
+\caption{
+    Region of inconsistency.
+    In the marked region, either $\hat{w}=1$ or one of $x^*-0.1 \le \hat{x} \le x^*+0.1$ or $y^*-0.1 \le \hat{y} \le y^*+0.1$ will not be true, resulting in poor estimation of the noninterior branch parameters.
+}
 \label{fig:bl-loose-inconsistency}
@@
 Since empirical estimates can be unstable, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher likelihood.
-We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$: we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
+We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$ (Fig.~\ref{fig:bl-general-inconsistency}): we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
 %EM Noo, I need to make a name for this sort of sentence-- the figure announcement sentence? Sentences should read "Here is a statement (Fig. X)."
@@
 I don't want to make extra work for you, but a figure in which you take the difference between the correct value and the inferred value might be nice.
-From this process we obtain the plots in Fig.~\ref{fig:bl-general-inconsistency} and Fig.~\ref{fig:bl-general-marginal}.
+%das: it seemed like the estimates were decent to me, but I can do this super easily (output is pickled).
+Though are you talking about another plot or just a check that we can add a sentence like "even in cases where \hat{w} is not one w is overestimated/poorly estimated."
 
-Marginal inference performs is as expected, where $\hat{w}$ is equal to $y^*$ in the limit data regardless of the value of $x^*$.
+Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ in the limit data regardless of the value of $x^*$.
 For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look unbiased, with $\hat{w}$ increasing as $y^*$ increases, though the region where $\hat{w}$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
@@
 %EM Seems like a sentence is missing here-- what's the usual setting?
-In this case, the converse is shown: when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
-Looking at Fig~2 of \cite{Felsenstein1978-rr} we see similarities in curve shape to the region in Fig.~\ref{fig:inconsistency-farris}, though the latter has a smaller region of inconsistency.
+In \cite{Felsenstein1978-rr}, when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
+This is the converse of what we have shown for joint inference, where the Felsenstein topology is more likely than the Farris topology.
+Looking at Fig.~2 of \cite{Felsenstein1978-rr} we see similarities in curve shape to the region in Fig.~\ref{fig:inconsistency-farris}, though the latter has a smaller region of inconsistency.
 Moreover, the inconsistency in \cite{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
@@
 %EM We should also describe the need for extensive simulations to see the extent of bias.
+
+While we have shown inconsistency in both topology and branch parameter estimation, these results have their limitations.
+The techniques used to obtain upper and lower bounds for the likelihoods in the topology estimation case provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
+Similarly, for the general case in estimating branch lengths, we were only able to provide a conservative region of overestimation.
+Empirical validation shows that the general case is not unlike the restricted case.
+All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to apply these results to more complicated general cases.
 

### 2017-12-02 joint_inf.tex
errors in estimation

@@
 \end{equation}
-We will use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximization of \eqref{eq:profile_likelihood}.
+We will use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}.
 One can then estimate a topology and branch fidelities using this profile likelihood:
@@
 Here we obtain the correct $\hat{w}\rightarrow y^*$.
+To understand the quality of each fit, we report the range of $\hat{w}-y^*$ where $\hat{w}\neq 1$.
+For joint inference, the errors range from $[-7\times 10^{-3}, 8\times 10^{-2}]$ and for marginal inference, $[-8\times 10^{-8}, 5\times 10^{-7}]$ showing that, even in cases where joint inference does not estimate $\hat{w}$ to be exactly one, it still fails to achieve a low error from truth.
 

### 2017-12-02 joint_inf_plot.py

@@ -462,6 +462,8 @@ def main(args=sys.argv[1:]):
             cbar = plt.colorbar()
             cbar.ax.tick_params(labelsize=FONT_SIZE-2)
             plt.savefig(args.plot_name)
+            errs = [z - y for y_row, z_row in zip(Y, Z) for y, z in zip(y_row, z_row) if z != 1-args.delta]
+            print "Error range [%.0E, %.0E]" % (min(errs), max(errs))
     else:
         print "No plotting argument given!"
 

### 2017-12-04 appendix.tex
picking nits in the appendix

@@
 $$
-\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
+\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t)
 $$
@@
 Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies under these two topologies.
+Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} contain the values of the likelihood for all possible ancestral states.
 Table~\ref{tab:likelihoods} contains calculations of likelihood values for fixed site patterns and topologies that have been maximized over ancestral state patterns.
-Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} contain the values of the likelihood for all possible ancestral states.
 
@@
 The probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
-See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern.
+See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the Farris tree in Fig.~\ref{fig:farris-fels-top}.
 Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods}.
@@
 \centering
+\begin{tabular}{|l|l|l|}
+    \hline
+$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
+    \hline
+    $\emptyset$    &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
+    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
+    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
+    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
+    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
+    $\{1,3\}$      &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
+    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
+    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
+    \hline
+\end{tabular}
+\caption{Site pattern probabilities.
+All values are multiplied by $1/8$.}
+\label{tab:sitepatprob}
+\end{table}
+
+\begin{table}
+\centering
 \begin{tabular}{|l|ll|}
@@
 \label{tab:fels_likelihoods}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{|l|l|l|}
-    \hline
-$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
-    \hline
-    $\emptyset$    &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
-    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
-    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
-    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
-    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
-    $\{1,3\}$      &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
-    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
-    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
-    \hline
-\end{tabular}
-\caption{Site pattern probabilities.
-All values are multiplied by $1/8$.}
-\label{tab:sitepatprob}
 \end{table}
@@
 - \log(8) \\
-&\qquad := L^{1}_{\tau_1,\tau_1}(x^*, y^*).
+&\qquad := C^{1}_{\tau_1,\tau_1}(x^*, y^*).
 \end{align*}
@@
 - \log(8) \\
-&\qquad := L^{2}_{\tau_1,\tau_1}(x^*, y^*),
+&\qquad := C^{2}_{\tau_1,\tau_1}(x^*, y^*),
 \end{align*}
@@
 - \log(8) \\
-&\qquad := L^{3}_{\tau_1,\tau_1}(x^*, y^*).
+&\qquad := C^{3}_{\tau_1,\tau_1}(x^*, y^*).
 \end{align*}
@@
 $$
-C_0(x^*, y^*) := \max\left(L^{1}_{\tau_1,\tau_1}(x^*, y^*), L^{2}_{\tau_1,\tau_1}(x^*, y^*), L^{3}_{\tau_1,\tau_1}(x^*, y^*)\right).
+C_0(x^*, y^*) := \max\left(C^{1}_{\tau_1,\tau_1}(x^*, y^*), C^{2}_{\tau_1,\tau_1}(x^*, y^*), C^{3}_{\tau_1,\tau_1}(x^*, y^*)\right).
 $$
@@
 $$
-See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < x^*, y^* < 1$ where $C_1(x^*, y^*) > C_0(x^*, y^*)$ and joint inference in inconsistent.
+See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < x^*, y^* < 1$ where $C_1(x^*, y^*) > C_0(x^*, y^*)$ and joint inference is inconsistent.
 \end{proof}
@@
 \label{eq:general-bl-result}
--\gamma^2(\hat{x}, \hat{y})(1 + \frac{1}{2}\beta) + 2\gamma(\hat{x}, \hat{y})\beta(\hat{x}, \hat{y})x^*(y^*)^2 + \beta^2(\hat{x}, \hat{y}) \ge 0.
+-\gamma^2(\hat{x}, \hat{y})\left(1 + \frac{1}{2}\beta\right) + 2\gamma(\hat{x}, \hat{y})\beta(\hat{x}, \hat{y})x^*(y^*)^2 + \beta^2(\hat{x}, \hat{y}) \ge 0.
 \end{equation}
@@
 $$
--\gamma_{U}^2(1 + \frac{1}{2}\beta) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0,
+-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0,
 $$
@@
 As is apparent from Table~\ref{tab:sitepatprob}, if $0 < x^*, y^* < 1$ and we try to evaluate the candidate $\hat{x}=\hat{y}=\hat{w}=1$ we have a likelihood of exactly zero.
-Worse still, in Table~\ref{tab:likelihoods} for the Farris tree, in the cases of $\{2,\}, \{1,2\}, \{2,3\},$ and $\{1,2,3\}$, any attempt to evaluate $\hat{y}=1$ will similarly yield a zero likelihood.
+Worse still, in Table~\ref{tab:likelihoods} for the Farris tree, in the cases of $\{2\}, \{1,2\}, \{2,3\},$ and $\{1,2,3\}$, any attempt to evaluate $\hat{y}=1$ will similarly yield a zero likelihood.
 In these cases of zero likelihood, methods that evaluate many candidate parameter values can fail to converge when near these boundaries.
@@
 We initialize our optimization procedure at the true branch parameters $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
-We posit that $\hat{w}$ is equal to one for regions of $x^*$ and $y^*$ small; methods that use local optimization may never see candidate values for large $x$ and $y$ if our initial value is small, so we perform two maximizations, one with $w$ fixed and one where it is estimated.
+We posit that $\hat{w}$ is equal to one for regions of $x^*$ and $y^*$ small; methods that use local optimization may never see candidate values for large $x$ and $y$ if our initial guess is small, so we perform two maximizations, one with $w=1$ fixed and one where $w$ is estimated.
 We take the value of $\hat{w}$ with the larger likelihood as our estimate.

### 2017-12-07 joint_inf.tex
Going through main text again :+1:

@@
 \begin{abstract}
-%EM This abstract is really nice! :+1:
 Maximum likelihood is a widespread and well-understood approach to estimate the phylogenetic parameters of topology and branch length given observed character data.
@@
 Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
-%EM I think that should either use "joint" or "profile" just to have a single word to describe things for most of the text, except when we connect it to the stats literature.
 While this method of joint maximum likelihood estimation can obtain estimates more efficiently, it has the potential to result in undesired properties of the estimates.
@@
 Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
+In this case those additional parameters come in the form of the bases at ancestral sequences.
 Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximate technique breaks down, even in the limit of infinite sequence data.
@@
 To do so, we use a binary symmetric model and imagine data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the limit of infinite data.
-We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known.
+We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.
 
@@
 In the classical configuration of each of these trees, the interior branch length parameters are equal to the bottom two parameters.
-We show in the Appendix that, in the case of the Farris tree, fixing both the top two branch parameters to be equal and the bottom two branch parameters to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch parameters.
+We show in the Appendix that, in the case of the Farris tree, doing inference fixing both the top two branch parameters to be equal and the bottom two branch parameters to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch parameters.
 
@@
 \end{subfigure}
-\caption{Two four-taxon trees}
+\caption{Two four-taxon trees with fidelities as labeled: $\theta_1=\theta_3=x$, $\theta_2=\theta_4=y$, and $\theta_5=w$.
+}
 \label{fig:farris-fels-top}
@@
 Fidelities have useful algebraic properties, and generating probabilities using the Hadamard transform have an especially simple form (see \eqref{eq:hadamard_probability} in the Appendix).
-For a four-taxon tree, define the general branch fidelity parameter $t=\{\theta_1,\theta_2,\theta_3,\theta_4,\theta_5\}$ where fidelities are ordered in the order of the taxa with the internal branch last.
-The trees in Fig.~\ref{fig:farris-fels-top} have $\theta_1=\theta_3=x$, $\theta_2=\theta_4=y$, and $\theta_5=w$.
+For a four-taxon tree, define the general branch fidelity parameter $t=\{\theta_1,\theta_2,\theta_3,\theta_4,\theta_5\}$ where fidelities are ordered in the order of the taxa with the internal branch last (Fig.~\ref{fig:farris-fels-top}).
 
@@
 
-The typical approach to is to maximize the likelihood marginalized across ancestral states
+The classical approach to is to maximize the likelihood marginalized across ancestral states
 \begin{equation}
@@
 $$
-We give $\ell$ explicitly in \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
+We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
 
@@
 We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology than for the true, generating Farris topology.
-To do so, we construct an upper bound for the likelihood given the Farris topology as a function of $x^*$ and $y^*$, and, similarly, we construct a lower bound for the likelihood given the Felsenstein topology.
-The values of $x^*$ and $y^*$ where the likelihood in the Felsenstein case is larger than the likelihood in the Farris case indicate an inconsistency (Fig.~\ref{fig:inconsistency-farris}).
+To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the Farris topology as a function of $x^*$ and $y^*$, and, similarly, we construct a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
+When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the Farris case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
 \begin{restatable}{theorem}{topoInconsist}
-%EM Wow this is hard to parse!
-%EM 1. The main message is that there is a region on which the wrong topology is preferred. This should be stated in words before the theorem, with the link to the figure.
-%EM 2. To see that in the theorem, we have to wind our way through inequalities facing different ways. Can we just say that there is a region? What do these L's tell us?
-%EM 3. L_0 and L_1 are totally different that L_n, but with very similar notation.
-%das: is this better, or did you want to get rid of the upper/lower bound constants entirely?
-I have them since they are a closed form function of x* and y* so it's clear that all we need is a region where some f(x*, y*) > 0.
-Are you saying the paragraph I just put before the theorem is a better statement of the theorem?
 Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
-There exist $C_0(x^*, y^*), C_1(x^*, y^*)$ and $0 < x^*, y^* < 1$ such that
+There exist $C_0(x^*, y^*), C_1(x^*, y^*)$ and a set of $0 < x^*, y^* < 1$ such that
 $$
@@
 $$
-\max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t) \ge C_1(x^*, y^*)
+C_1(x^*, y^*) \le \max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t)
 $$
-with $C_1(x^*, y^*) > C_0(x^*, y^*)$.
+with $C_0(x^*, y^*) < C_1(x^*, y^*)$.
 \end{restatable}
-Fig.~\ref{fig:inconsistency-farris} shows the values of $x^*$ and $y^*$ where $C_1(x^*, y^*) > C_0(x^*, y^*)$.
-Intuitively, $\tau_2$ is favored in performing joint maximum likelihood since the likelihood for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its likelihood while $\tau_2$ has many more.
-This results in more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.
+The proof of this theorem is by a detailed examination of inequalities.
+The only intuition we can offer is that $\tau_2$ is favored in performing joint maximum likelihood since the likelihood for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its likelihood while $\tau_2$ has many more.
+This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.
 
@@
 \caption{
-    Regions of inconsistency.
+    Regions of topological inconsistency in terms of fidelities.
     Due to the looseness of the upper and lower bounds, the parameters in the upper right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
@@
 As above, assume that data is generated from the ``Farris zone'' with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$.
-In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one instead of its true value of $y^*$.
+In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
 That is, if we know $x^*$ and $y^*$, then for
@@
 Again, we find a region for which the inferred values do not converge to the generating values.
-These situations are in contrast to the approach using marginal likelihood where necessarily $\hat{w}$ converges to $y^*$ as the number of observations grows.
+These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows.
 
@@
 
-Fix $x=x^*$ and $y=y^*$ to their true, generating values and proceed to estimate the internal branch parameter $w$.
+Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and proceed to estimate the internal branch parameter $w$.
 \begin{restatable}{theorem}{restrictedBranchInconsist}
@@
 $$
-The maximum likelihood value $\hat{w} = 1$ if
+The maximum likelihood value $\hat{w}$ is equal to 1 if
 $$
@@
 \includegraphics[width=\textwidth]{branch-length-inconsistency-inkscape}
-\caption{Region of inconsistency with $x=x^*$ and $y=y^*$ fixed.}
+\caption{Region of branch parameter inconsistency in terms of fidelities $x=x^*$ and $y=y^*$ that are fixed to their correct values.}
 \label{fig:bl-inconsistency}
@@
 In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
-%EM This is the first we're hearing of the quadratic formula in this version.
 Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
 Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
-%EM Here would be a good place to introduce the concept that youre going to be saying that either the pendant branch lengths are bad, or what is going to be 1.
-%EM differing notions of "bad" will lead to different theorems.
 We show that there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
-We fix bounds on these estimates and show that, in the general case, we either estimate $\hat{w}$ to be one or fall outside of the specified bounds for $\hat{x}$ or $\hat{y}$, indicating a poor estimate in at least one of the three unknown branch parameters.
+We fix bounds on these estimates and show that, in the general case, we either estimate $\hat{w}$ to be one or fall outside of specified bounds for $\hat{x}$ or $\hat{y}$, indicating a poor estimate in at least one of the three unknown branch parameters.
 \begin{restatable}{theorem}{generalBranchInconsist}
 \label{thm:general-bl}
+Define $\gamma(x, y) = 4xy$.
 For
@@
 $$
-\gamma := \gamma(x^*, y^*) = 4x^*y^*,
+\gamma := \gamma(x^*, y^*),
 $$
-and bounds
+bounds
 $$
@@
 \end{restatable}
-Similar to Theorem~\ref{thm:restricted-bl}, the inequality in Theorem~\ref{thm:general-bl} is also a function of only $x^*$ and $y^*$, though general forms for $\gamma_L$, $\gamma_U$, and $\beta_L$ are difficult to obtain.
-%EM I think you put this loosest bound in for exposition's sake, but I don't think it helps clarify.
-%EM On the other hand, I think that explaining that the restricted case fits in this framework is great.
-The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section and given in Fig.~\ref{fig:bl-inconsistency}.
-For an intermediate bound, define a specified level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
+Similar to Theorem~\ref{thm:restricted-bl}, the inequality in Theorem~\ref{thm:general-bl} is also a function of only $x^*$ and $y^*$.
+Although general forms for $\gamma_L$, $\gamma_U$, and $\beta_L$ are difficult to obtain, we can make progress by bounding the error in branch length estimation for the pendant or the central branch.
+If we don't want to allow any error in pendant branches, we can use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section and given in Fig.~\ref{fig:bl-inconsistency}.
+For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
 $$
@@
 As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
-In this case, we will either estimate $\hat{w}$ to be one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true value.
+In this case, joint inference will either estimate $\hat{w}$ to be one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true value.
 
@@
 \includegraphics[width=\textwidth]{bl-loose-inconsistency-inkscape}
-%EM If you read this caption without reading the text, would you be able to understand what's going on?
-I think I'd be confused because in the caption you are stating something that looks good, and then the figure describes xhat and yhat as being poorly estimated.
 \caption{
-    Region of inconsistency.
+    Region of branch parameter inconsistency in terms of fidelities.
     In the marked region, either $\hat{w}=1$ or one of $x^*-0.1 \le \hat{x} \le x^*+0.1$ or $y^*-0.1 \le \hat{y} \le y^*+0.1$ will not be true, resulting in poor estimation of the noninterior branch parameters.
@@
 
+Direct numerical optimization confirms our theoretically-derived bounds and provides more detailed picture.
 The region of inconsistency in Fig.~\ref{fig:bl-loose-inconsistency} is a conservative estimate of where $\hat{w}=1$.
-To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint maximum likelihood estimation using numerical optimization, described in the Appendix.
+To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint maximum likelihood estimation using numerical optimization, as described in the Appendix.
+%EM This could be considered a little unfair.
+Optimization of 3 variables is harder than optimizing two, so when there is an optimization problem with the 3-variable case then w=1 has an advantage.
+I don't have a suggestion to change things.
 Since empirical estimates can be unstable, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher likelihood.
 We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$ (Fig.~\ref{fig:bl-general-inconsistency}): we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
-%EM Noo, I need to make a name for this sort of sentence-- the figure announcement sentence? Sentences should read "Here is a statement (Fig. X)."
-%EM There is a big bomb to drop here, which is that what is 1 all over the place!
 %EM Question: it seems to me that even in the non-white region, the BL inferences aren't correct.
@@
 Though are you talking about another plot or just a check that we can add a sentence like "even in cases where \hat{w} is not one w is overestimated/poorly estimated."
+%EM I'm pretty sure I can see that there's bias in the estimates even in the non-white area.
+A plot of the difference between the two would be cool.
+Can we zoom into the region with x* and y* greater than 0.6 so that the scale doesn't get thrown off by big differences where x* and y* are zero?
 
@@
 
-We have shown that the method of joint inference when estimating phylogenetic parameters is not consistent in general.
-Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, inferring the incorrect topology and, second, estimating severely biased branch lengths even given the correct topology.
+We have shown that the method of joint inference when estimating phylogenetic parameters \cite{Sagulenko2017-jo} is not consistent in general.
+Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference infers the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
 In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities are closer to zero than to one.
 
-The method of profile likelihood is well-studied \cite{Murphy2000-ry}.
+Joint inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \cite{Murphy2000-ry}.
 Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., \cite{Geman1982}.
-Joint inference of phylogenetic parameters and ancestral states \cite{Sagulenko2017-jo} is a similar approach to profile likelihood.
-Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates in the case of infinite parameters have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \cite{LeCam1990}.
+Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \cite{LeCam1990}.
 Until recently \cite{RoyChoudhury2015-ta}, results on the properties of standard maximum likelihood estimates of phylogeny as in \eqref{eq:marginal_likelihood} have been difficult to prove, and no results have been obtained for profile likelihood.
@@
 In phylogenetics, counterexamples showing methods of phylogenetic inference are inconsistent have been shown using the same four-taxa topologies used here \cite{Felsenstein1978-rr}.
-%EM Seems like a sentence is missing here-- what's the usual setting?
 In \cite{Felsenstein1978-rr}, when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
 This is the converse of what we have shown for joint inference, where the Felsenstein topology is more likely than the Farris topology.
-Looking at Fig.~2 of \cite{Felsenstein1978-rr} we see similarities in curve shape to the region in Fig.~\ref{fig:inconsistency-farris}, though the latter has a smaller region of inconsistency.
+Looking at Fig.~2 of \cite{Felsenstein1978-rr} we see some similarities in curve shape to the region in Fig.~\ref{fig:inconsistency-farris}, though the latter has a smaller region of inconsistency.
 Moreover, the inconsistency in \cite{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
 This is not the case here; for our case, the inconsistency generally occurs when all branches are long, and has more to do with the choices of the form of the likelihood than from the interplay of long and short branches.
-Difficulties in phylogenetic estimation when the ``Farris zone'' generates characters have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.
-
-%EM We need to describe limitations (we weren't able to write out equations getting the region exactly right, and we are using a simple model of subs.
-%EM We should also describe the need for extensive simulations to see the extent of bias.
+Difficulties in phylogenetic estimation when generating data on the ``Farris zone'' tree have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.
 
@@
 Empirical validation shows that the general case is not unlike the restricted case.
-All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to apply these results to more complicated general cases.
+All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
 

### 2017-12-07 joint_inf.tex
More work on the beginning.

@@
 
-\title{Joint estimation of ancestral states and phylogenetic model parameters is inconsistent}
+%EM I think this is punchier. Thoughts?
+\title{Joint estimation of phylogenies and ancestral states is not consistent}
+%\title{Joint estimation of ancestral states and phylogenetic model parameters is inconsistent}
 \author{David A. Shaw \and Frederick A. Matsen IV}
@@
 \begin{abstract}
-Maximum likelihood is a widespread and well-understood approach to estimate the phylogenetic parameters of topology and branch length given observed character data.
-Since a phylogeny has unknown ancestral states, computing a maximum likelihood solution becomes a problem involving latent variables.
+Maximum likelihood is a widely used and well understood approach to estimate the phylogenetic topology and branch length given observed character data.
+Any phylogenetic maximum likelihood solution must propose a means of handling unknown ancestral states.
 Classically, these unknown intermediate states are averaged out, which leads to consistent estimation of the topology and continuous model parameters.
@@
 While this method of joint maximum likelihood estimation can obtain estimates more efficiently, it has the potential to result in undesired properties of the estimates.
-We show that, like other methods of maximum likelihood where the number of unknowns grows with observations, this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
-There are a number of cases where generating under a four-taxon tree results in a joint estimation procedure converging on the wrong topology.
-If the topology is known a priori, there are similar cases where the branch lengths are consistently overestimated in the limit of infinite data.
+In this paper, we show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
+We derive a region where generating data under a four-taxon tree results in a joint estimation procedure converging on the wrong topology.
+If the topology is known a priori, there are similar cases where the branch lengths are consistently overestimated in the case of infinite-length sequences.
 Although joint maximum likelihood may turn out to be a useful tool in practice, care should be taken when using this approach.
@@
 Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
-In this case those additional parameters come in the form of the bases at ancestral sequences.
-Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximate technique breaks down, even in the limit of infinite sequence data.
+In this case those additional parameters come in the form of the states of ancestral sequences.
+Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is ``perfect''.
 
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model and imagine data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the limit of infinite data.
+To do so, we use a binary symmetric model and imagine data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
 We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.
@@
 Assume $\fullAlignment$ is generated from topology $\tau^*$ and branch fidelities $t^*$.
-Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log likelihood, which can be thought of as the infinite-length sequence limit
+Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log likelihood, which can be thought of as the infinite-length sequence case
 $$
@@
 \caption{
-    Regions of topological inconsistency in terms of fidelities.
-    Due to the looseness of the upper and lower bounds, the parameters in the upper right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
+    An analytically-derived region of topological inconsistency in terms of fidelities for ``perfect'' data generated on the Farris topology (Fig.~\ref{fig:farris-fels-top}) with $w^* = y^*$.
+    Due to the looseness of the upper and lower bounds, the parameters in the non-shaded region do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
 }
@@
 \includegraphics[width=\textwidth]{branch-length-inconsistency-inkscape}
-\caption{Region of branch parameter inconsistency in terms of fidelities $x=x^*$ and $y=y^*$ that are fixed to their correct values.}
+\caption{Analytically-derived region of branch parameter inconsistency in terms of fidelities $x=x^*$ and $y=y^*$ that are fixed to their correct values with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
+Here again, the shaded region is guaranteed to give inconsistent estimation, while the non-shaded region may or may not do so.}
 \label{fig:bl-inconsistency}
@@
 Similar to Theorem~\ref{thm:restricted-bl}, the inequality in Theorem~\ref{thm:general-bl} is also a function of only $x^*$ and $y^*$.
-Although general forms for $\gamma_L$, $\gamma_U$, and $\beta_L$ are difficult to obtain, we can make progress by bounding the error in branch length estimation for the pendant or the central branch.
+Although general forms for $\gamma_L$, $\gamma_U$, and $\beta_L$ are difficult to obtain, we can make progress by bounding the error in branch length estimation for the pendant and/or the central branch.
 If we don't want to allow any error in pendant branches, we can use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section and given in Fig.~\ref{fig:bl-inconsistency}.
@@
 $$
-Then $\gamma_L, \gamma_U$ and $\beta_L$ are defined directly from the bounds on $\hat{x}$ and $\hat{y}$.
+Then useful $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
 As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
-In this case, joint inference will either estimate $\hat{w}$ to be one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true value.
+In this case, joint inference will either estimate $\hat{w}$ to have fidelity 1 or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true value.
 
@@
 \caption{
-    Region of branch parameter inconsistency in terms of fidelities.
-    In the marked region, either $\hat{w}=1$ or one of $x^*-0.1 \le \hat{x} \le x^*+0.1$ or $y^*-0.1 \le \hat{y} \le y^*+0.1$ will not be true, resulting in poor estimation of the noninterior branch parameters.
+    Analytically-derived region of branch parameter inconsistency in terms of fidelities with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
+    In the marked region, either $\hat{w}=1$ or one of $x^*-0.1 \le \hat{x} \le x^*+0.1$ or $y^*-0.1 \le \hat{y} \le y^*+0.1$ will not be true, resulting in poor estimation of the pendant branch parameters.
+Here again, the shaded region is guaranteed to give such poor estimation, while the non-shaded region may or may not do so.
 }
@@
 
-Direct numerical optimization confirms our theoretically-derived bounds and provides more detailed picture.
-The region of inconsistency in Fig.~\ref{fig:bl-loose-inconsistency} is a conservative estimate of where $\hat{w}=1$.
-To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint maximum likelihood estimation using numerical optimization, as described in the Appendix.
+Direct numerical optimization confirms our theoretically-derived bounds and provides more detailed picture compared to the conservative analytically-derived region of Fig.~\ref{fig:bl-loose-inconsistency}.
+To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint maximum likelihood estimation using numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}).
+The region in which $\hat{w}$ is numerically estimated to be 1 is an unusual shape containing the analytically-derived region but is not the same as it.
+We see that the region of inconsistency in the general case is similar to that of the restricted case (compare Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency}).
+This region encompasses the majority of the parameter space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch parameter to be one.
+
+Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same procedure.
+
 %EM This could be considered a little unfair.
@@
 I don't have a suggestion to change things.
-Since empirical estimates can be unstable, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher likelihood.
-We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$ (Fig.~\ref{fig:bl-general-inconsistency}): we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
+We provide a full description of our optimization procedure in the Appendix, but briefly, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher likelihood.
+We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$.
+We do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
 %EM Question: it seems to me that even in the non-white region, the BL inferences aren't correct.
@@
 Can we zoom into the region with x* and y* greater than 0.6 so that the scale doesn't get thrown off by big differences where x* and y* are zero?
-
-Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ in the limit data regardless of the value of $x^*$.
-For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look unbiased, with $\hat{w}$ increasing as $y^*$ increases, though the region where $\hat{w}$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
-Comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case is similar to that of the restricted case.
-This region encompasses the majority of the parameter space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch parameter to be one.
-As a positive control, Fig.~\ref{fig:bl-general-marginal} shows the result of optimizing \eqref{eq:marginal_likelihood} using the same procedure.
-Here we obtain the correct $\hat{w}\rightarrow y^*$.
+% For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look unbiased, with $\hat{w}$ increasing as $y^*$ increases, though the region where $\hat{w}$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
+% Here we obtain the correct $\hat{w}\rightarrow y^*$.
 To understand the quality of each fit, we report the range of $\hat{w}-y^*$ where $\hat{w}\neq 1$.
@@
     Estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:profile_likelihood}.
+    Data generated as in Fig.~\ref{fig:inconsistency-farris}.
 }
@@
 Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference infers the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
-In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities are closer to zero than to one.
+In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.
 
@@
 Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \cite{LeCam1990}.
-Until recently \cite{RoyChoudhury2015-ta}, results on the properties of standard maximum likelihood estimates of phylogeny as in \eqref{eq:marginal_likelihood} have been difficult to prove, and no results have been obtained for profile likelihood.
+Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are in fact quite recent \cite{RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
 We have furthered progress in understanding the limitations of this joint maximum likelihood approach.
@@
 In phylogenetics, counterexamples showing methods of phylogenetic inference are inconsistent have been shown using the same four-taxa topologies used here \cite{Felsenstein1978-rr}.
-In \cite{Felsenstein1978-rr}, when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
+In this previous work, when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
 This is the converse of what we have shown for joint inference, where the Felsenstein topology is more likely than the Farris topology.
-Looking at Fig.~2 of \cite{Felsenstein1978-rr} we see some similarities in curve shape to the region in Fig.~\ref{fig:inconsistency-farris}, though the latter has a smaller region of inconsistency.
 Moreover, the inconsistency in \cite{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
 This is not the case here; for our case, the inconsistency generally occurs when all branches are long, and has more to do with the choices of the form of the likelihood than from the interplay of long and short branches.
+%EM Given the above, do you think that the next sentence forms a meaningful comparison? The shapes aren't _so_ similar, really...
+Looking at Fig.~2 of \cite{Felsenstein1978-rr} we see some similarities in curve shape to the region in Fig.~\ref{fig:inconsistency-farris}, though the latter has a smaller region of inconsistency.
 Difficulties in phylogenetic estimation when generating data on the ``Farris zone'' tree have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.
 
-While we have shown inconsistency in both topology and branch parameter estimation, these results have their limitations.
+While we have shown inconsistency in both topology and branch parameter estimation, there is substantial scope for future work to make these results more precise and more general.
 The techniques used to obtain upper and lower bounds for the likelihoods in the topology estimation case provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
-Similarly, for the general case in estimating branch lengths, we were only able to provide a conservative region of overestimation.
+Similarly, for the general case in estimating branch lengths, we were only able to provide a conservative region of overestimation, and the unusual shape we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) begs further investigation.
 Empirical validation shows that the general case is not unlike the restricted case.

### 2017-12-07 appendix.tex
bias plot; minor issues

@@
 
-We use basin-hopping \cite{Wales1997} implemented in \texttt{scipy} \cite{Jones2001} to obtain the plots in Fig.~\ref{fig:bl-general-inconsistency} and Fig.~\ref{fig:bl-general-marginal}.
+We use basin-hopping \cite{Wales1997} implemented in \texttt{scipy} \cite{Jones2001} to obtain the plots in Figs.~\ref{fig:bl-general-inconsistency}, \ref{fig:bl-general-marginal}, and \ref{fig:bl-general-bias}.
 This method randomly perturbs candidate solutions and accepts or rejects them based on the nearby likelihood surface, eventually obtaining an estimate of the ``global'' optimum.
@@
 \end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{w-hat-empirical-01-bias}
+\caption{
+    Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:marginal_likelihood}.
+}
+\label{fig:bl-general-bias}
+\end{figure}

### 2017-12-07 joint_inf.tex
bias plot; minor issues

@@
 %EM I think this is punchier. Thoughts?
+%das: <3 <3
 \title{Joint estimation of phylogenies and ancestral states is not consistent}
@@
 In this case those additional parameters come in the form of the states of ancestral sequences.
+%das: is "perfect" a standard term in phylogenetics? Does it just mean infinite length with independent sites?
 Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is ``perfect''.
@@
 
-Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same procedure.
-
 %EM This could be considered a little unfair.
@@
 I don't have a suggestion to change things.
+%das: I sort of gave a justification for this in the appendix (line 998); I agree it may not be the most fair, but other than also including a plot of just the 3-variable case (which recall doing at some point but not how it looks) then I'm not sure what else we could say either.
+Maybe I'll just run the 3-variable case now to have it in our pocket just in case?
 We provide a full description of our optimization procedure in the Appendix, but briefly, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher likelihood.
@@
 We do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
+
 %EM Question: it seems to me that even in the non-white region, the BL inferences aren't correct.
@@
 Can we zoom into the region with x* and y* greater than 0.6 so that the scale doesn't get thrown off by big differences where x* and y* are zero?
-% For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look unbiased, with $\hat{w}$ increasing as $y^*$ increases, though the region where $\hat{w}$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
+Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same procedure.
+For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look reasonable, with $\hat{w}$ increasing as $y^*$ increases, though Fig.~\ref{fig:bl-general-bias} shows there is a systematic positive bias in this procedure.
 % Here we obtain the correct $\hat{w}\rightarrow y^*$.
@@
 %EM Given the above, do you think that the next sentence forms a meaningful comparison? The shapes aren't _so_ similar, really...
-Looking at Fig.~2 of \cite{Felsenstein1978-rr} we see some similarities in curve shape to the region in Fig.~\ref{fig:inconsistency-farris}, though the latter has a smaller region of inconsistency.
+%das: Yah, I think I was just staring at clouds. Removed.
 Difficulties in phylogenetic estimation when generating data on the ``Farris zone'' tree have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.

### 2017-12-07 joint_inf_plot.py

@@ -462,8 +462,27 @@ def main(args=sys.argv[1:]):
             cbar = plt.colorbar()
             cbar.ax.tick_params(labelsize=FONT_SIZE-2)
             plt.savefig(args.plot_name)
-            errs = [z - y for y_row, z_row in zip(Y, Z) for y, z in zip(y_row, z_row) if z != 1-args.delta]
-            print "Error range [%.0E, %.0E]" % (min(errs), max(errs))
+            plt.close()
+            bias = Z - Y
+            ones_mask = (Z != 1 - args.delta)
+            print "Error range [%.0E, %.0E], Mean: %.2E" % (min(bias[ones_mask]), max(bias[ones_mask]), np.mean(bias[ones_mask]))
+
+            # plot bias
+            bias_to_plot = np.where(ones_mask, bias, .09)
+            im = plt.imshow(bias_to_plot[int(.6/args.delta):, int(.6/args.delta):], cmap=plt.cm.gray, origin='lower')
+            plt.xlabel(r'$x^*$', fontsize=FONT_SIZE)
+            plt.ylabel(r'$y^*$', fontsize=FONT_SIZE)
+            ttl = plt.title(r'Bias: $\hat{w}-y^*$', fontsize=FONT_SIZE+2)
+            ttl.set_position([.5, 1.05])
+            ax = plt.gca()
+            ax.set_xticks(np.arange(0, .4/args.delta, .1/args.delta))
+            ax.set_yticks(np.arange(0, .4/args.delta, .1/args.delta))
+            ax.set_xticklabels([r'$0.6$', r'$0.7$', r'$0.8$', r'$0.9$'], fontsize=FONT_SIZE-2)
+            ax.set_yticklabels([r'$0.6$', r'$0.7$', r'$0.8$', r'$0.9$'], fontsize=FONT_SIZE-2)
+            sns.despine()
+            cbar = plt.colorbar()
+            cbar.ax.tick_params(labelsize=FONT_SIZE-2)
+            plt.savefig(args.plot_name.replace('.svg', '_bias.svg'))
     else:
         print "No plotting argument given!"
 

### 2017-12-07 appendix.tex
Nits. So tasty.

@@
 Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.
+%EM is there a logic to what gets bolded or no? siteSplit isn't, but a site pattern is.
 
@@
 
+We first fix notation for maps that are defined just below.
 \begin{definition}
@@
 The mapping $\patternToSplit$ takes the complement of site patterns to obtain a site split in $\mathcal{P}(\siteSplitSet)$.
-The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.
+The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is defined to be the complement of $\ancestralStateColumn$.
 
@@
 As a consequence of assuming a binary symmetric model, taking complements yields
+%EM continuing to nitpick, is there anything that is just s_j? Seems funny to only have the tilde s_j form.
 \begin{align*}
@@
 In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
-For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this have
+For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this we have
 \begin{align*}
@@
 $$
+%EM It seems a little strange that eta appears here as an argument when it didn't appear in eq:site_pattern_likelihood.
+In fact it's not really an argument, right?
+Everything is determined by t and tau it seems to me.
 L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
@@
 \end{equation*}
-so that, as $\nCols\rightarrow\infty$,
+so that, in the $\nCols\rightarrow\infty$ limit,
 \begin{equation}
@@
 Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
-We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
+As described above we cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
 $$
@@
 $$
+%EM Nitpick-- eta takes a tau and a t as arguments. Here we give it a p.
 The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, p)=\{\emptyset\}$.
@@
 To compute the likelihood of observing a set of data, we need $P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
-The probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
+We remind the reader that the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
 See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the Farris tree in Fig.~\ref{fig:farris-fels-top}.
+%EM Isn't some of this redundant with the prev paragraph?
 Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods}.
-Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}
+Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}.
 
@@
 \caption{
-    Example likelihood computations given the above branch probability parameterizations.
+    Example likelihood computations given fidelities $x$, $y$, and $w$.
     In (a), we compute the product to obtain $P(\ancestralStateColumn=\emptyset\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1+w)/32$.
@@
      $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+     %EM I was doing some spot-checking here, and this line seems strange.
+     %EM I'd think that this left hand entry has everything in the upper half of the tree in the "on" state, which means that it should be (1+x)^2   (1-w)(1+y)^2, no?
      $\{1,3\}$     &$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^2$\\
@@
 \end{tabular}
-\caption{Likelihood calculations for all site patterns and internal states of Farris tree.
+\caption{Likelihood calculations for all site patterns and internal states of the Farris tree.
 All values multiplied by $1/32$.}

### 2017-12-07 joint_inf.tex
clarify

@@
 Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same procedure.
-For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look reasonable, with $\hat{w}$ increasing as $y^*$ increases, though Fig.~\ref{fig:bl-general-bias} shows there is a systematic positive bias in this procedure.
+For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look reasonable, with $\hat{w}$ increasing as $y^*$ increases, though Fig.~\ref{fig:bl-general-bias} shows there is a systematic positive bias in this procedure when $\hat{w}$ is not estimated to be 1.
 % Here we obtain the correct $\hat{w}\rightarrow y^*$.
@@
 All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
+%EM Any intuition of if the effect will go away on larger trees? What about more complex models?
 

### 2017-12-13 appendix.tex
clarify
phew lots of hand calculations but tables should be good
Topic sentences, reorg, and a few Q's.

@@
 We follow Section 8.6 of \cite{Semple2003-em} to state the Hadamard representation of site split generating probabilities.
-%das: @EM thoughts on including all of this in full generality? should we just reference using the sentence above and be done with it?
-%EM You already have it in here, and so I don't see a reason to yank it now.
 For each edge $e$ define the edge fidelity for that edge as
@@
 By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
+Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies for our two topologies (Fig.~\ref{fig:farris-fels-top}).
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{|l|l|l|}
+    \hline
+$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
+    \hline
+    $\emptyset$    &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
+    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
+    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
+    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
+    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
+    $\{1,3\}$      &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
+    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
+    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
+    \hline
+\end{tabular}
+\caption{Site pattern probabilities $\siteSplit_j$ on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
+All values multiplied by $1/8$.}
+\label{tab:sitepatprob}
+\end{table}
+
+\subsection*{Example}
 We follow with an expository example computing these probabilities and likelihoods.
-
-\paragraph{Example}
-\begin{figure}
-    \centering
-    \includegraphics[width=.45\textwidth]{unrooted_farris_tree}
-    \caption{Four taxon ``Farris zone'' topology.}
-\label{fig:four-taxa-tree}
-\end{figure}
-
-Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}---this is commonly known as the ``Farris zone'' topology.
+Consider the fixed, binary four-tip tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}(a)---this is commonly known as the ``Farris zone'' topology.
 The set of all possible character assignments is
@@
 
-Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies under these two topologies.
-Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} contain the values of the likelihood for all possible ancestral states.
-Table~\ref{tab:likelihoods} contains calculations of likelihood values for fixed site patterns and topologies that have been maximized over ancestral state patterns.
-
 To compute the likelihood of observing a set of data, we need $P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
 We remind the reader that the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
-See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the Farris tree in Fig.~\ref{fig:farris-fels-top}.
-%EM Isn't some of this redundant with the prev paragraph?
+See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the Farris tree.
 Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods}.
@@
 \caption{
-    Example likelihood computations given fidelities $x$, $y$, and $w$.
+    Example likelihood computations on the Farris tree $\tau_1$ for fidelities $x$, $y$, and $w$.
+    Edges labeled by the probability of substitution along that edge.
     In (a), we compute the product to obtain $P(\ancestralStateColumn=\emptyset\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1+w)/32$.
@@
 \centering
-\begin{tabular}{|l|l|l|}
-    \hline
-$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
-    \hline
-    $\emptyset$    &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
-    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
-    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
-    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
-    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
-    $\{1,3\}$      &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
-    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
-    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
-    \hline
-\end{tabular}
-\caption{Site pattern probabilities.
-All values are multiplied by $1/8$.}
-\label{tab:sitepatprob}
-\end{table}
-
-\begin{table}
-\centering
 \begin{tabular}{|l|ll|}
@@
      $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{1,3\}$     &$(1+x)^2   (1+w)(1-y)^2$          &$(1+x)^2   (1-w)(1+y)^2$\\
+     $\{1,3\}$     &$(1-x)^2   (1+w)(1+y)^2$          &$(1-x)^2   (1-w)(1-y)^2$\\
      $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{1,2,3\}$   &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
+     $\{1,2,3\}$   &$(1-x)^2   (1+w)(1+y)(1-y)$       &$(1-x)^2   (1-w)(1+y)(1-y)$\\
     \hline
@@
      $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-     %EM I was doing some spot-checking here, and this line seems strange.
-     %EM I'd think that this left hand entry has everything in the upper half of the tree in the "on" state, which means that it should be (1+x)^2   (1-w)(1+y)^2, no?
-     $\{1,3\}$     &$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^2$\\
+     $\{1,3\}$     &$(1+x)^2   (1-w)(1+y)^2$     &$(1+x)^2   (1+w)(1-y)^2$\\
      $\{2,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-     $\{1,2,3\}$   &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+     $\{1,2,3\}$   &$(1+x)^2   (1-w)(1+y)(1-y)$  &$(1+x)^2   (1+w)(1+y)(1-y)$\\
     \hline
 \end{tabular}
-\caption{Likelihood calculations for all site patterns and internal states of the Farris tree.
+\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the Farris tree.
 All values multiplied by $1/32$.}
@@
      $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1+x)^2   (1-w)(1+y)(1-y)$\\
-     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1+x)^2   (1-w)(1+y)^2$\\
-     $\{1,3\}$     &$(1+x)^2   (1+w)(1-y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1-x)^2   (1-w)(1+y)^2$\\
-     $\{1,2,3\}$   &$(1+x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1+y)^2$\\
+     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1-x)^2   (1-w)(1-y)^2$\\
+     $\{1,3\}$     &$(1-x)^2   (1+w)(1+y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
+     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1+x)^2   (1-w)(1-y)^2$\\
+     $\{1,2,3\}$   &$(1-x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1-y)^2$\\
     \hline
@@
      $\{3\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{1,2\}$     &$(1-x)^2   (1-w)(1-y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-     $\{1,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1-x)^2   (1+w)(1+y)^2$\\
-     $\{2,3\}$     &$(1+x)^2   (1-w)(1-y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-     $\{1,2,3\}$   &$(1+x)(1-x)(1-w)(1-y)^2$           &$(1-x)^2   (1+w)(1+y)(1-y)$\\
+     $\{1,2\}$     &$(1+x)^2   (1-w)(1+y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+     $\{1,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1+x)^2   (1+w)(1-y)^2$\\
+     $\{2,3\}$     &$(1-x)^2   (1-w)(1+y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
+     $\{1,2,3\}$   &$(1+x)(1-x)(1-w)(1+y)^2$           &$(1+x)^2   (1+w)(1+y)(1-y)$\\
 \hline
 \end{tabular}
-\caption{Likelihood calculations for all site patterns and internal states of Felsenstein tree.
+\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the Felsenstein tree.
 All values multiplied by $1/32$.}
@@
 \end{tabular}
-\caption{Likelihood calculations for all site patterns and ancestral state partitions.
+\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$.
 All values multiplied by $1/32$.
@@
 
-Consider the Farris tree with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
-We now show that, in the case of the Farris tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
+Consider the Farris tree with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
+We now show that, in the case of the Farris tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch lengths to be equal and the bottom two branch lengths to be equal during inference will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
 Using the Hadamard transform approach, we calculate the generating probabilities on the Farris tree.
@@
 
+%EM I think this following statement could use a little more justification.
+If we write out a symmetric function f in two variables and take the derivative of f(x+delta, x) = f(x, x+delta) with respect to delta and set it to zero we can see that this must be a stationary point of f.
+Do we need to say more to ensure that this is a maximum?
 Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
@@
 
-%EM Here we are restating.
 \topoInconsist*
@@
 Recalling Table~\ref{tab:likelihoods}, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood over $(x,y,w)$.
-%EM You are talking about a log likelihood here, right? So a likelihood being proportional to means that there is some additive constant. Makes things a little confusing.
-%das: updated
 If $x=x^*$ and $y=y^*$ are fixed, by collapsing like terms and ignoring terms with only $x^*$ and $y^*$, the log-likelihood in \eqref{eq:farris_likelihood} as a function of $w$ is
@@
 % Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
-%EM Rather confusing to be using ' for two things here.
-What about using a dot over the l?
-In retrospect something other than ' would have been good because of the proliferation of w^{'2} etc.
-Prob not worth changing now though.
 The derivative of the log-likelihood with respect to $w$ is
@@
     Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:marginal_likelihood}.
+    Plot focuses on $0.6 < x^*, y^* < 1$ where $\hat{w}$ is estimated to be a value different from 1.
 }

### 2017-12-13 joint_inf.tex
Topic sentences, reorg, and a few Q's.

--- 2017-12-07 joint_inf.tex
@@
 
-%EM I think this is punchier. Thoughts?
 \title{Joint estimation of phylogenies and ancestral states is not consistent}
-%\title{Joint estimation of ancestral states and phylogenetic model parameters is inconsistent}
 \author{David A. Shaw \and Frederick A. Matsen IV}
@@
 This latter conditional optimization is simpler and more computationally efficient than marginalizing over ancestral sequences and then performing optimization.
+But is it statistically consistent?
 
@@
 In this case those additional parameters come in the form of the states of ancestral sequences.
-Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is ``perfect''.
+Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is generated without sampling error according to the exact model used for inference.
 
@@
     An analytically-derived region of topological inconsistency in terms of fidelities for ``perfect'' data generated on the Farris topology (Fig.~\ref{fig:farris-fels-top}) with $w^* = y^*$.
-    Due to the looseness of the upper and lower bounds, the parameters in the non-shaded region do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
+    Due to the looseness of the upper and lower bounds, the parameters in the non-shaded region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
 }
@@
 We analyze two settings on the Farris tree, corresponding to whether some branch fidelities are fixed at their true values or not.
-As above, assume that data is generated from the ``Farris zone'' with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$.
+As above, assume that data is generated from the Farris tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Figure~\ref{fig:farris-fels-top}).
 In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
@@
 Again, we find a region for which the inferred values do not converge to the generating values.
-These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows.
+These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \cite{RoyChoudhury2015-ta}.
 
@@
 \end{restatable}
-The inequality in Theorem~\ref{thm:restricted-bl} is a function of the generating parameters $x^*$ and $y^*$.
-This region is shown in Fig.~\ref{fig:bl-inconsistency}.
-Intuitively, since we perform maximization over internal states, branch fidelities will have a tendency to be overestimated.
-Seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.
+This theorem allows us to demarcate a region of bad internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
+Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch, and then a branch length is estimated to be zero (i.e.\ have fidelity $\hat{w} = 1$).
+As an intuition for the theoretical development, seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.
 
@@
 \caption{Analytically-derived region of branch parameter inconsistency in terms of fidelities $x=x^*$ and $y=y^*$ that are fixed to their correct values with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
+The shaded region shows the area in which the internal branch length is estimated to be 0, i.e.\ the estimated fidelity $\hat{w} = 1$, even though the generating fidelity is set to be that of $y^*$.
 Here again, the shaded region is guaranteed to give inconsistent estimation, while the non-shaded region may or may not do so.}
@@
 In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
+%EM This doesn't look like the same inequality to me.
 Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
@@
 We show that there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
-We fix bounds on these estimates and show that, in the general case, we either estimate $\hat{w}$ to be one or fall outside of specified bounds for $\hat{x}$ or $\hat{y}$, indicating a poor estimate in at least one of the three unknown branch parameters.
+We fix bounds on these estimates and show that, in the general case, joint estimation either estimates $\hat{w}$ to be 1 or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
 \begin{restatable}{theorem}{generalBranchInconsist}
@@
 \end{restatable}
-Similar to Theorem~\ref{thm:restricted-bl}, the inequality in Theorem~\ref{thm:general-bl} is also a function of only $x^*$ and $y^*$.
-Although general forms for $\gamma_L$, $\gamma_U$, and $\beta_L$ are difficult to obtain, we can make progress by bounding the error in branch length estimation for the pendant and/or the central branch.
-If we don't want to allow any error in pendant branches, we can use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section and given in Fig.~\ref{fig:bl-inconsistency}.
+This theorem can be used to show incorrect branch parameter estimates as follows.
+If we don't want to allow any error in pendant branches, we can use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section (Fig.~\ref{fig:bl-inconsistency}).
 For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
@@
 As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
-In this case, joint inference will either estimate $\hat{w}$ to have fidelity 1 or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true value.
+In this case, joint inference will either estimate $\hat{w}$ to have fidelity 1 or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.
 
@@
 
-Direct numerical optimization confirms our theoretically-derived bounds and provides more detailed picture compared to the conservative analytically-derived region of Fig.~\ref{fig:bl-loose-inconsistency}.
+Direct numerical optimization confirms our theoretically-derived bounds and provides more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:bl-loose-inconsistency}).
 To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint maximum likelihood estimation using numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}).
@@
 We see that the region of inconsistency in the general case is similar to that of the restricted case (compare Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency}).
-This region encompasses the majority of the parameter space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch parameter to be one.
+This region encompasses the majority of the branch fidelity space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch fidelity to be 1.
 
-Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same procedure.
-
-%EM This could be considered a little unfair.
-Optimization of 3 variables is harder than optimizing two, so when there is an optimization problem with the 3-variable case then w=1 has an advantage.
-I don't have a suggestion to change things.
 We provide a full description of our optimization procedure in the Appendix, but briefly, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher likelihood.
@@
 We do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
-%EM Question: it seems to me that even in the non-white region, the BL inferences aren't correct.
-Is that true?
-We should say so if.
-I don't want to make extra work for you, but a figure in which you take the difference between the correct value and the inferred value might be nice.
-%das: it seemed like the estimates were decent to me, but I can do this super easily (output is pickled).
-Though are you talking about another plot or just a check that we can add a sentence like "even in cases where \hat{w} is not one w is overestimated/poorly estimated."
-%EM I'm pretty sure I can see that there's bias in the estimates even in the non-white area.
-A plot of the difference between the two would be cool.
-Can we zoom into the region with x* and y* greater than 0.6 so that the scale doesn't get thrown off by big differences where x* and y* are zero?
-% For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look unbiased, with $\hat{w}$ increasing as $y^*$ increases, though the region where $\hat{w}$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
+%EM Need GH link here.
+
+Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same procedure.
+For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look reasonable, with $\hat{w}$ increasing as $y^*$ increases, though Fig.~\ref{fig:bl-general-bias} shows there is a systematic positive bias in this procedure even when $\hat{w}$ is not estimated to be 1.
 % Here we obtain the correct $\hat{w}\rightarrow y^*$.
@@
 \caption{
-    Estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:profile_likelihood}.
+    Numerical estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:profile_likelihood}.
     Data generated as in Fig.~\ref{fig:inconsistency-farris}.
@@
 
-In phylogenetics, counterexamples showing methods of phylogenetic inference are inconsistent have been shown using the same four-taxa topologies used here \cite{Felsenstein1978-rr}.
-In this previous work, when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
+Previous work in phylogenetics has developed consistency counterexamples using the same four-taxa topologies used here \cite{Felsenstein1978-rr}.
+In this previous work, when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more optimal when performing a particular estimation procedure.
 This is the converse of what we have shown for joint inference, where the Felsenstein topology is more likely than the Farris topology.
@@
 This is not the case here; for our case, the inconsistency generally occurs when all branches are long, and has more to do with the choices of the form of the likelihood than from the interplay of long and short branches.
-%EM Given the above, do you think that the next sentence forms a meaningful comparison? The shapes aren't _so_ similar, really...
-Looking at Fig.~2 of \cite{Felsenstein1978-rr} we see some similarities in curve shape to the region in Fig.~\ref{fig:inconsistency-farris}, though the latter has a smaller region of inconsistency.
 Difficulties in phylogenetic estimation when generating data on the ``Farris zone'' tree have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.
@@
 All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
+%EM Any intuition of if the effect will go away on larger trees? What about more complex models?
+%EM Also, is there any hope that an alternate profile likelihood formulation could lead to more consistent estimation? It'd be nice to end with something like that.
 
@@
 
-David Shaw and Frederick Matsen were supported by NIH grants U19-AI117891 and R01-GM113246 and by R01-AI12096.
+This work supported by National Institutes of Health grants R01-AI12096, U19-AI117891, and U54-GM111274 as well as National Science Foundation grants CISE-1561334 and CISE-1564137.
 The research of Frederick Matsen was supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation.

### 2017-12-13 appendix.tex
More work on the appendix.

@@
 with similar definitions for the remaining generating probabilities.
-The full Farris tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see Table~\ref{tab:likelihoods} for $\tau = \tau_1$).
+The full Farris tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see Table~\ref{tab:likelihoods}).
 We begin by writing out the case for the ancestral state split $\{1\}$.
-Directly substituting in \eqref{eq:site_pattern_likelihood}, the log-likelihood is
+%EM Is eq:site_pattern_likelihood the best ref? It still has counts. It would be nice to have something that's exactly this before plugging in the values from the tables.
+Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_likelihood}, the log-likelihood is
+%EM This is the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1\}$. You address the other cases later, but at this point it's not clear that you're going to. Can you introduce with this in mind?
 \begin{align}
@@
 \begin{align*}
+1-x^2-y^2-x^2y^2 & = (1+x)(1-x)(1+y)(1-y) \\
 1-x^2+y^2-x^2y^2 & = (1+x)(1-x)(1+y^2) \le (1+x)(1-x)(1+y) \\
@@
 \end{align*}
-The term $p_{\emptyset}$ is bounded as
+The $p_{\emptyset}$ term is bounded as
 \begin{align*}
@@
 Call
+%EM It would seem helful for the below check to express these a_1' etc in terms of tilde{p}.
 \begin{equation}
@@
 \end{equation}
+%EM This check would make a lot more sense if you put the C^{2}_{\tau_1,\tau_1}(x^*, y^*)  and the other analogous definition first.
+%EM It would also be nice to do one of the calculations to get us from a_1 world to a_1' world. Just staring at table S4 doesn't make this obvious.
 We also now need to check that
@@
 We see that
-$$
-1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 2 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - (x^*)^2
-$$
-and
-$$
-(y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - 4x^* + (x^*)^2 \le 1 - (x^*)^2,
-$$
-the last inequality holding if $x^* \le 2$.
+\begin{equation}
+1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 2 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - (x^*)^2.
+\label{eq:lenny}
+\end{equation}
+Note
+$$
+(y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - 4x^* + (x^*)^2;
+$$
+showing that \eqref{eq:lenny} holds if $x^* \le 2$, which is satisfied by definition.
 Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}$ and $\hat{y}$ are valid.
@@
 We construct a similar lower bound on the Felsenstein tree partial likelihood.
-In the Felsenstein case, there are various optimal internal states depending on branch lengths (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).
+In the Felsenstein case, there are quite a few optimal internal states depending on branch lengths (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).
 We can bound these below replacing $1+w$ with $1-w$ as follows.
@@
 $$
+%EM Hm, the following doesn't look right to me.
 always, and thus the maximum can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
@@
 
-%EM Wanna add a short section about numerical optimization? You've seen that it's not entirely trivial.
-
 \subsection*{Empirical validation}
@@
 We initialize our optimization procedure at the true branch parameters $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
-We posit that $\hat{w}$ is equal to one for regions of $x^*$ and $y^*$ small; methods that use local optimization may never see candidate values for large $x$ and $y$ if our initial guess is small, so we perform two maximizations, one with $w=1$ fixed and one where $w$ is estimated.
+We posit that $\hat{w}$ is equal to 1 for when $x^*$ and $y^*$ are small; methods that use local optimization may never see candidate values for large $x$ and $y$ if our initial guess is small, so we perform two maximizations, one with $w=1$ fixed and one where $w$ is estimated.
 We take the value of $\hat{w}$ with the larger likelihood as our estimate.

### 2017-12-14 appendix.tex
address (almost all) comments

@@
 %EM is there a logic to what gets bolded or no? siteSplit isn't, but a site pattern is.
+%das: I was thinking bold for matrices/vectors (here of character data) then a vector of character data maps to a set (site split), and so I didn't bold the set.
 
@@
 %EM continuing to nitpick, is there anything that is just s_j? Seems funny to only have the tilde s_j form.
+%das: I think s_j used to be the site pattern and then \tilde{s}_j became the split?
+It's still the case for h (no tilde is ancestral state pattern, tilde is ancestral state split).
+I changed to \tilde{y} to be consistent with this, but mabye we can think of something better?
 \begin{align*}
@@
 Everything is determined by t and tau it seems to me.
+%das: We added it to clarify that the likelihood now depends on \xi, which is an element of \eta.
+It seems dirty to ignore it.
+At the risk of adding more notation, we could add \hat{H} to the original likelihood?
+Or just axe the etas...
 L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
@@
 A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
-As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
+As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch, with $t=\{p,p,p,p,p\}$.
 The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
@@
 %EM Nitpick-- eta takes a tau and a t as arguments. Here we give it a p.
-The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, p)=\{\emptyset\}$.
-Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, p)$ we have
+%das: nit unpicked
+The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
+Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, t)$ we have
 $$
@@
 $$
-Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, p)$,
+Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, t)$,
 $$
@@
 Do we need to say more to ensure that this is a maximum?
+%das: I wasn't even thinking derivatives.
+Let f(x, y) be symmetric so that f(x, y) = f(y, x).
+If f is maximized at some (x_0, y_0) then f(x_0, y_0) > f(x, y) for all x, y.
+But then by symmetry f(y_0, x_0) is also a maximum.
+Now *if* a unique maximum exists, x_0 = y_0 should necessarily be this maximum by the properties defined above, otherwise, f(x_0, y_0) would not be equal to f(y_0, x_0).
+I think there are probably counterexamples where there is not a unique maximum where this statement isn't necessarily true (trivially, because you'd just say x_0 = y_1 if y_1 is also a maximum and voila) but I don't think this discounts the fact that x_0 = y_0 will still be *a* maximum in this case.
+In that case, it's reasonable to assume so, in my opinion, especially for the topological inconsistency (we don't use the branch length values), but even for the branch length inconsistency because the case of no unique solution is just as bad if not worse than inconsistency.
+%das: Should some or all of that be in the text? Can you see anything wrong with it?
 Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
@@
 The full Farris tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see Table~\ref{tab:likelihoods}).
-We begin by writing out the case for the ancestral state split $\{1\}$.
+We begin by writing out the case for the ancestral state split $\{1\}$, and show the other two cases follow a similar argument.
 %EM Is eq:site_pattern_likelihood the best ref? It still has counts. It would be nice to have something that's exactly this before plugging in the values from the tables.
-Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_likelihood}, the log-likelihood is
+%das: gah, you're right
+Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_profile_likelihood_mean}, the log-likelihood is
 %EM This is the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1\}$. You address the other cases later, but at this point it's not clear that you're going to. Can you introduce with this in mind?
+%das: done
 \begin{align}
@@
 %EM It would also be nice to do one of the calculations to get us from a_1 world to a_1' world. Just staring at table S4 doesn't make this obvious.
+%das:todo
 We also now need to check that
@@
 $$
-We observe that
+Replacing all $1+w$ with $1-w$ in Table~\ref{tab:likelihoods} for $\tau = \tau_2$ will result in a lower bound on the maximum, so we proceed with that first.
+Then, for site split $\{1,2\}$ we have
+$$
+\max\left((1+x)^2 (1-w)(1+y)^2, (1+x)(1-x)(1-w)(1+y)(1-y)\right)
+$$
+where since
 $$
@@
 %EM Hm, the following doesn't look right to me.
-always, and thus the maximum can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
+%das: Okay, think I missed the point about 1+w for 1-w that I added back.
+always, the maximum can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
 For the other site splits we consider two cases.
@@
 %Rather than do a constrained optimization, I just bounded the p_{\emptyset}\log(1+w) term below by p_{\emptyset}\log(1-w).
+%das: This is old and I believe the text has been updated to fix it. Can you verify?
 i.e.\ the maximum is achieved at $\{1,1-b,0\}$.
@@
 $$
-which, simplifying as functions of $\alpha_1, \alpha_2, \gamma$ and $\beta$ yields
+which, using
+$$
+\alpha_1 + \alpha_2 = \frac{1}{2}\beta
+$$
+and
+$$
+\alpha_1 - \alpha_2 = x^*(y^*)^2,
+$$
+and simplifying as functions of $\gamma$ and $\beta$ yields
 \begin{equation}
 \label{eq:restricted-bl-result}
-\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
+%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
+-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0.
 \end{equation}

### 2017-12-14 joint_inf.tex
address (almost all) comments

@@
 \newcommand{\alignmentColumnRV}{Y}
-\newcommand{\siteSplit}{\tilde{s}}
-\newcommand{\siteSplitSet}{\mathcal{S}}
+\newcommand{\siteSplit}{\tilde{y}}
+\newcommand{\siteSplitSet}{\mathcal{Y}}
 \newcommand{\fullAncestralStates}{\mathbf{H}}
@@
 $$
-\alpha_1 := \alpha_1(x^*, y^*) = \frac{1}{8} \left(1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
-$$
-$$
-\alpha_2 := \alpha_2(x^*, y^*) = \frac{1}{8}\left(1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
-$$
-$$
 \beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
@@
 $$
-\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
+%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
+-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0.
 $$
@@
 %EM This doesn't look like the same inequality to me.
+%das: updated
 Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
@@
 We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$.
-We do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
+We do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies\footnote{\url{https://github.com/matsengrp/joint-inf/blob/master/joint_inf_plot.py}}.
 %EM Need GH link here.
@@
 %EM Also, is there any hope that an alternate profile likelihood formulation could lead to more consistent estimation? It'd be nice to end with something like that.
+%das: wow, that would be an undertaking.
+The only thing I know off the top of my head are things like sieves, which I recall Paul Edlefsen being jazzed about but don't know if he uses them in this way.
+I can think more but it might just be a vague throwaway line.
+%das:todo
 

### 2017-12-15 appendix.tex
more explanation of a-primes

@@
 The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
-Call
+The previous upper bound is for the ancestral state split $\{1\}$.
+For example, for the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term and gain a $(1+w)$ term.
+In this case, all terms involving $w$ simplify to $\log(1+w)$, maximized at $\hat{w}=1$.
+For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
 %EM It would seem helful for the below check to express these a_1' etc in terms of tilde{p}.
@@
         a_{2}' &= a_{2}+2p_{13}, \\
+    \end{aligned}
+    \label{eq:a_const_prime_x}
+\end{equation}
+for site split $\emptyset$ and
+\begin{equation}
+    \begin{aligned}
         a_{3}' &= a_{3}-2p_{13}, \\
-        a_{4}' &= a_{4}+2p_{13}. \\
+        a_{4}' &= a_{4}+2p_{13}, \\
     \end{aligned}
-    \label{eq:a_const_prime}
+    \label{eq:a_const_prime_y}
 \end{equation}
+for site split $\{1,2\}$.
 %EM This check would make a lot more sense if you put the C^{2}_{\tau_1,\tau_1}(x^*, y^*)  and the other analogous definition first.
 %EM It would also be nice to do one of the calculations to get us from a_1 world to a_1' world. Just staring at table S4 doesn't make this obvious.
-%das:todo
-We also now need to check that
-$$
-\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
-$$
-are both between zero and one.
-It will suffice to show
-$$
-0 \le \tilde{p} - 2p_{13} \le 1
-$$
-and
-$$
-0 \le \tilde{p} + 2p_{13} \le 1
-$$
-for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
-Using Table~\ref{tab:sitepatprob} and $t^*$,
-$$
-p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right).
-$$
-We see that
-\begin{equation}
-1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 2 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - (x^*)^2.
-\label{eq:lenny}
-\end{equation}
-Note
-$$
-(y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - 4x^* + (x^*)^2;
-$$
-showing that \eqref{eq:lenny} holds if $x^* \le 2$, which is satisfied by definition.
-Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}$ and $\hat{y}$ are valid.
-
+%das: is this a better explanation or did you want it written out explicitly?
 To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
@@
 \end{align*}
+
+We now need to check that
+$$
+\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
+$$
+are both between zero and one.
+It will suffice to show
+$$
+0 \le \tilde{p} - 2p_{13} \le 1
+$$
+and
+$$
+0 \le \tilde{p} + 2p_{13} \le 1
+$$
+for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
+Using Table~\ref{tab:sitepatprob} and $t^*$,
+$$
+p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right).
+$$
+We see that
+\begin{equation}
+1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 2 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - (x^*)^2.
+\label{eq:lenny}
+\end{equation}
+Note
+$$
+(y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - 4x^* + (x^*)^2;
+$$
+showing that \eqref{eq:lenny} holds if $x^* \le 2$, which is satisfied by definition.
+Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}$ and $\hat{y}$ are valid.
+
 All bounds are functions only of $x,y$ through the true generating probabilities $x^*, y^*$.

### 2017-12-15 appendix.tex
Comment pass.

@@
 Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.
-%EM is there a logic to what gets bolded or no? siteSplit isn't, but a site pattern is.
-%das: I was thinking bold for matrices/vectors (here of character data) then a vector of character data maps to a set (site split), and so I didn't bold the set.
 
@@
 Or just axe the etas...
+%EM I think I must be missing something:
+%EM The likelihood doesn't belong on the specific choice of \xi. Besides, \xi already appears in eq:site_pattern_likelihood.
+%EM Eta is a deterministic function of t and tau, so isn't having it redundant?
 L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
@@
 $$
-%EM Nitpick-- eta takes a tau and a t as arguments. Here we give it a p.
-%das: nit unpicked
 The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
@@
 %das: Should some or all of that be in the text? Can you see anything wrong with it?
+%EM I don't see why this works.
+In fact, imagine that x=y is a ridge of minima, and that we have maxima at (1,2) and at (2,1).
+My argument shows that we must have a stationary point at x=y, but that doesn't mean that it has to be a maximum!
 Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
@@
 We begin by writing out the case for the ancestral state split $\{1\}$, and show the other two cases follow a similar argument.
-%EM Is eq:site_pattern_likelihood the best ref? It still has counts. It would be nice to have something that's exactly this before plugging in the values from the tables.
-%das: gah, you're right
 Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_profile_likelihood_mean}, the log-likelihood is
-%EM This is the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1\}$. You address the other cases later, but at this point it's not clear that you're going to. Can you introduce with this in mind?
-%das: done
 \begin{align}
@@
 The previous upper bound is for the ancestral state split $\{1\}$.
-For example, for the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term and gain a $(1+w)$ term.
+%EM Yay! Would be even better if you specified which upper bound (eqref) you want us to be looking at... there are many.
+For the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term and gain a $(1+w)$ term.
 In this case, all terms involving $w$ simplify to $\log(1+w)$, maximized at $\hat{w}=1$.
@@
 %das: Okay, think I missed the point about 1+w for 1-w that I added back.
+%EM No, the lower bound is (1+x)(1-x)(1-w)(1+y)(1-y).
 always, the maximum can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
@@
 $$
-%EM Shouldn't this be 2p_{\emptyset}-1 ? It doesn't matter for the bound, but...
-%das If we call p_{\emptyset} just p, we're maximizing
-%   p log (1 + w) + (1 - p) log (1 - w).
-%As before, a_1 log(1+x) + a_2 log(1-x) is maximized at x = (a_1 - a_2) / (a_1 + a_2).
-%So then w = (p - (1 - p)) / (p + (1 - p)) = (2p - 1) / 1 = 2p - 1.
-%So you're right. Good catch.
-%
-%Unfortunately, I realized the above does not work for us since 2p - 1 is not guaranteed to be between 0 and 1, as x should be (see the stuff I added justifying this above).
-%Rather than do a constrained optimization, I just bounded the p_{\emptyset}\log(1+w) term below by p_{\emptyset}\log(1-w).
-%das: This is old and I believe the text has been updated to fix it. Can you verify?
 i.e.\ the maximum is achieved at $\{1,1-b,0\}$.

### 2017-12-15 joint_inf.tex
more explanation of a-primes
Comment pass.

@@
 In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
-%EM This doesn't look like the same inequality to me.
-%das: updated
 Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
@@
 We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$.
-We do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies\footnote{\url{https://github.com/matsengrp/joint-inf/blob/master/joint_inf_plot.py}}.
-%EM Need GH link here.
+We do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
+Our optimization code can be found at \url{https://github.com/matsengrp/joint-inf/}.
 
@@
 %EM Any intuition of if the effect will go away on larger trees? What about more complex models?
-%EM Also, is there any hope that an alternate profile likelihood formulation could lead to more consistent estimation? It'd be nice to end with something like that.
-%das: wow, that would be an undertaking.
-The only thing I know off the top of my head are things like sieves, which I recall Paul Edlefsen being jazzed about but don't know if he uses them in this way.
-I can think more but it might just be a vague throwaway line.
-%das:todo
 

### 2017-12-15 appendix.tex
counter-Qs

--- 2017-12-15 appendix.tex
@@
 Or just axe the etas...
+%EM I think I must be missing something:
+%EM The likelihood doesn't belong on the specific choice of \xi. Besides, \xi already appears in eq:site_pattern_likelihood.
+%EM Eta is a deterministic function of t and tau, so isn't having it redundant?
+%das: The likelihood doesn't depend on the specific choice of \xi, but it does depend on \eta.
+I don't know, I guess redundancy isn't always the worst thing?
+I just see it as a tradeoff between how explicit we want to be and how clean.
+Like, in the first line of eq:site_pattern_likelihood we could have as an argument \hat{H} instead of suppressing it, even though it's totally determined by the inputs as-is.
+Anyway, that was my justification for including \eta.
+I'm fine with removing it though if you think it's weird.
 L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
@@
 %das: Should some or all of that be in the text? Can you see anything wrong with it?
+%EM I don't see why this works.
+In fact, imagine that x=y is a ridge of minima, and that we have maxima at (1,2) and at (2,1).
+My argument shows that we must have a stationary point at x=y, but that doesn't mean that it has to be a maximum!
+%das: Wouldn't the assumption of there existing a unique maximum preclude your case?
+My argument is that if there is a unique maximum, it necessarily has to have x=y.
+If there is not a unique maximum then the ML problem is ill-posed.
+I now don't see why we have to have a unique maximum other than to say we hope the problem is well-posed, but it may not be.
+Do you have any thoughts on where to take this?
 Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
@@
 The upper bound for the likelihood is then maximized at
-\begin{align*}
+\begin{align}
+\begin{split}
 &    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
@@
 &\qquad := C^{1}_{\tau_1,\tau_1}(x^*, y^*).
-\end{align*}
+\end{split}
+\label{eq:farris-upper-bound}
+\end{align}
 
 The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
-The previous upper bound is for the ancestral state split $\{1\}$.
-For example, for the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term and gain a $(1+w)$ term.
+The upper bound in \eqref{eq:farris-upper-bound} is for the ancestral state split $\{1\}$.
+%EM Yay! Would be even better if you specified which upper bound (eqref) you want us to be looking at... there are many.
+For the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term and gain a $(1+w)$ term.
 In this case, all terms involving $w$ simplify to $\log(1+w)$, maximized at $\hat{w}=1$.
@@
 %EM It would seem helful for the below check to express these a_1' etc in terms of tilde{p}.
+%das: not quite sure what this comment means actually.
 \begin{equation}
@@
 %das: Okay, think I missed the point about 1+w for 1-w that I added back.
+%EM No, the lower bound is (1+x)(1-x)(1-w)(1+y)(1-y).
+%das: I agree that that's *a* lower bound, yeah.
+So maybe tell me what's wrong with the following: L is the general likelihood in the joint inf problem for the Felsenstein topology, obtained by doing the usual maximization over ancestral states in tab:likelihoods.
+Now let's obtain L' via the same process but we've now replaced all (1+w)s in that table with (1-w)s.
+L \ge L', so then L' is a lower bound for the likelihood L, yes?
+So looking at, let's call it the lower bound version of tab:likelihoods where everything is a (1-w), maximizing over ancestral states then results in the above inequality.
+Maybe I'm being careless?
 always, the maximum can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.

### 2017-12-18 appendix.tex
fix etas

@@
 I'm fine with removing it though if you think it's weird.
-L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
+L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
 $$
@@
 \begin{split}
-&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t)) \\
+&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
 &\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
@@
 \begin{split}
-&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \\
+&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
 &\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)].
@@
 $$
-\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
+\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
 $$
@@
     \label{eq:log_likelihood_simplified}
-    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
+    \ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
 \end{equation}
@@
 Do you have any thoughts on where to take this?
-Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
+Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, $x_1=x_2$ at the maximum.
 The analogous statement holds for $y_1$ and $y_2$.
@@
 $$
-\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \le
+\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \le
     \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
-    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t}))
+    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
 $$
@@
 $$
-\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
+\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t).
 $$
@@
 $$
-\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \ge
+\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \ge
     \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
-    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t}))
+    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
 $$
@@
     \label{eq:farris_likelihood}
-    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t))
+    \ell_{\tau_1,t^*}(\tau_1, t)
     &=        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw+x^2y^2) \nonumber \\
@@
 \begin{align*}
-&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
+&    \ell_{\tau_1,t^*}(\tau_1, t) \\
 &\qquad \le      p_{\emptyset}  \cdot\log(2(1+x)(1+y))
@@
 \begin{split}
-&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
+&    \ell_{\tau_1,t^*}(\tau_1, t) \\
 &\qquad \le      p_{\emptyset}  \cdot\log(2)
@@
 \begin{split}
-&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
+&    \ell_{\tau_1,t^*}(\tau_1, t) \\
 &\qquad\le      p_{\emptyset}  \cdot\log(2)
@@
 \begin{align*}
-&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
+&    \ell_{\tau_1,t^*}(\tau_1, t) \\
 &\qquad\le      p_{\emptyset}  \cdot\log(2)
@@
 In the Felsenstein case, there are quite a few optimal internal states depending on branch lengths (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).
-We can bound these below replacing $1+w$ with $1-w$ as follows.
-For the site split $\{1,2\}$, we would like to find a lower bound for
-$$
-\max\left((1+x)^2 (1-w)(1+y)^2, (1+x)(1-x)(1+w)(1+y)(1-y)\right).
-$$
-Replacing all $1+w$ with $1-w$ in Table~\ref{tab:likelihoods} for $\tau = \tau_2$ will result in a lower bound on the maximum, so we proceed with that first.
+We can bound these below replacing $1+w$ with $1-w$ in Table~\ref{tab:likelihoods} for $\tau = \tau_2$, as this will result in a lower bound on the maximum, so we proceed with that first.
 Then, for site split $\{1,2\}$ we have
@@
 \begin{align*}
-    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t; \fullAncestralSplitPartitions(\tau_2,t))
+    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
     &\ge      p_{\emptyset}  \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
@@
 \begin{align*}
-    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t; \fullAncestralSplitPartitions(\tau_2,t))
+    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
     &\ge      p_{\emptyset}  \cdot\log((1+x)^2    (1-w)(1+y)^2) \\
@@
 $$
-    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t; \fullAncestralSplitPartitions(\tau_2,t))
+    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
     \ge      2\cdot\log(1+x)
@@
 $$
-\tilde{\ell}_{\tau_1,t^*}(\tau_2, t; \fullAncestralSplitPartitions(\tau_2,t))
+\tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
 \ge      2\cdot\log(2)
@@
 \begin{align*}
-    \ell_{\tau_1,t^*}(\tau_1, w; \fullAncestralSplitPartitions(\tau_1,w))
+    \ell_{\tau_1,t^*}(\tau_1, w)
     &\sim     p_{\emptyset}  \cdot\log(1+(x^*)^2+(y^*)^2+4x^*y^*w+(x^*)^2(y^*)^2) \\

### 2017-12-18 joint_inf.tex
discussion and quick pass

@@
 Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
-We observe $\nCols$ samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
+%das: @EM I added iid here but maybe it sounds redundant? I feel it's important and we don't bring up independence until the next section, and only implicitly assume i.d. through \alignmentColumnRV
+We observe $\nCols$ independent and identically distributed samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
 The corresponding unobserved ancestral states are $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ and distributed as $\ancestralStateColumnRV$.
@@
 Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., \cite{Geman1982}.
+A potential solution in this case using the method of sieves could be to project the column-wise ancestral state patterns into a lower-dimensional space by requiring certain columns to be equal to one another, allowing the degrees of freedom in the ancestral state columns to grow with $\nCols$, albeit more slowly than $O(\nCols)$.
 Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \cite{LeCam1990}.
@@
 All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
+Given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \cite{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
 %EM Any intuition of if the effect will go away on larger trees? What about more complex models?
+%das: my brash intution is that it could potentially be much worse in more complex cases because we have a larger number of degrees of freedom, which is my intuition for why we have inconsistency now.
+I'm not sure I can justify this though.
 

### 2017-12-18 appendix.tex
remove etas, clarify lower bound, identifiability

@@
 I'm fine with removing it though if you think it's weird.
+%das: I changed all this.
 L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
@@
 Do you have any thoughts on where to take this?
-Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, $x_1=x_2$ at the maximum.
-The analogous statement holds for $y_1$ and $y_2$.
-The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
+%das: So there are references for identifiability, but I'm a little nervous using them because they're for the usual marginal ML problem.
+How cagey does it sound to say (as I do below) "here are references for identifiability in the usual case, so either joint inf will be not identifiable (bad) or it (likely?) will be and we can use this (which turns out to be inconsistent later)?" Can you think of a better approach?
+%das: some other references I found for identifiability are Rogers 1997, RoyChoudhury 2014...
+The maximum likelihood procedure in phylogenetics results in identifiable parameter estimation under mild conditions \cite{Chang1996}.
+The joint inference procedure may not necessarily result in the same identifiability properties, but we claim the following.
+If joint inference obtains unique parameter estimates, then, since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, $x_1=x_2$ at the maximum, with the analogous statement holding for $y_1$ and $y_2$.
+The case in which we do not obtain unique parameter estimates results in an ill-posed likelihood problem, and thus by design an inconsistency since multiple maxima exist that will not equal the true, generating parameters.
+Given this, we let $x_1=x_2=x$ and $y_1=y_2=y$.
+The Felsenstein tree does not admit this property of symmetry of function inputs, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
 
@@
 We construct a similar lower bound on the Felsenstein tree partial likelihood.
-In the Felsenstein case, there are quite a few optimal internal states depending on branch lengths (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).
-We can bound these below replacing $1+w$ with $1-w$ in Table~\ref{tab:likelihoods} for $\tau = \tau_2$, as this will result in a lower bound on the maximum, so we proceed with that first.
-Then, for site split $\{1,2\}$ we have
+In the Felsenstein case, there are many more optimal internal states depending on branch lengths than in the Farris case (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).
+We first bound the entire likelihood below and then proceed with joint inference.
+First, we obtain a lower bound for the likelihood by replacing $1+w$ with $1-w$ in Table~\ref{tab:likelihoods} for $\tau = \tau_2$; in this case, we resolve many of the ambiguous likelihood terms.
+For example, for site split $\{1,2\}$, after we replace the $1+w$ term for $\ancestralSplitPartition_j(\tau, t)=\{\emptyset, \{1,2\}\}$ with $1-w$, we have
 $$
@@
 Maybe I'm being careless?
-always, the maximum can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
+%das: Is this better?
+always, the maximum for this site split can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
 For the other site splits we consider two cases.

### 2017-12-19 joint_inf.tex
a little looky

@@
 Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
-%das: @EM I added iid here but maybe it sounds redundant? I feel it's important and we don't bring up independence until the next section, and only implicitly assume i.d. through \alignmentColumnRV
 We observe $\nCols$ independent and identically distributed samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
@@
 Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., \cite{Geman1982}.
-A potential solution in this case using the method of sieves could be to project the column-wise ancestral state patterns into a lower-dimensional space by requiring certain columns to be equal to one another, allowing the degrees of freedom in the ancestral state columns to grow with $\nCols$, albeit more slowly than $O(\nCols)$.
+A potential solution in this case using the method of sieves could be to project the column-wise ancestral state patterns into a lower-dimensional space, allowing the degrees of freedom in the ancestral state columns to grow with $\nCols$, albeit more slowly than $O(\nCols)$.
 Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \cite{LeCam1990}.
@@
 Given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \cite{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
-%EM Any intuition of if the effect will go away on larger trees? What about more complex models?
-%das: my brash intution is that it could potentially be much worse in more complex cases because we have a larger number of degrees of freedom, which is my intuition for why we have inconsistency now.
-I'm not sure I can justify this though.
 

### 2017-12-22 appendix.tex
napkin math in the comments
moved comments to words

@@
 %das: some other references I found for identifiability are Rogers 1997, RoyChoudhury 2014...
-The maximum likelihood procedure in phylogenetics results in identifiable parameter estimation under mild conditions \cite{Chang1996}.
-The joint inference procedure may not necessarily result in the same identifiability properties, but we claim the following.
-If joint inference obtains unique parameter estimates, then, since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, $x_1=x_2$ at the maximum, with the analogous statement holding for $y_1$ and $y_2$.
-The case in which we do not obtain unique parameter estimates results in an ill-posed likelihood problem, and thus by design an inconsistency since multiple maxima exist that will not equal the true, generating parameters.
-Given this, we let $x_1=x_2=x$ and $y_1=y_2=y$.
+Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, $x_1=x_2$ at the maximum provided a unique maximum exists.
+An analogous statement holds for $y_1$ and $y_2$.
+It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
+Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference, similar arguments holding for the remaining parameters.
+We decompose the likelihood into the entropy and partial likelihood terms as in \eqref{eq:log_likelihood_simplified}.
+By Gibbs's inequality, the entropy term $\shannonDivergence_{\tau^*,t^*}(\tau^*,t)$ has a unique maximum over $t$---namely $t^*$.
+For the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$, we can compute the partial likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ as in Table~\ref{tab:farris_likelihoods}.
+Clearly this partial likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $1/2 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in[0,1]$ and thus has a unique maximum.
+Since both summands of the likelihood are concave on $x_1\in[0,1]$, the likelihood is as well, and there exists a single $x_1\in[0,1]$ maximizing the likelihood.
+Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
 The Felsenstein tree does not admit this property of symmetry of function inputs, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
@@
     Plot focuses on $0.6 < x^*, y^* < 1$ where $\hat{w}$ is estimated to be a value different from 1.
+    We do not compute the bias for the white region where $\hat{w}=1$.
 }

### 2017-12-26 appendix.tex
noticed minor error; fix plus add clarification

@@
 \end{tabular}
-\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$.
+\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$ after maximizing over ancestral states.
 All values multiplied by $1/32$.
 Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
-See Appendix for full calculations.}
+See Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} for full calculations.}
 \label{tab:likelihoods}
@@
 We begin by writing out the case for the ancestral state split $\{1\}$, and show the other two cases follow a similar argument.
-Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_profile_likelihood_mean}, the log-likelihood is
+Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_profile_likelihood_mean}, the log-likelihood is, suppressing normalizing constants $-\log 8$ and $-\log 32$,
 \begin{align}
@@
     &\qquad + p_{123}        \cdot\log(1+x^2-y^2-x^2y^2) \nonumber \\
-    &\qquad - \log(8) \nonumber \\
     &\qquad + p_{\emptyset}  \cdot\log((1+x)^2   (1+w)(1+y)^2) \nonumber \\
@@
 \end{align}
-We use the facts that, for $x,y\in[0,1]$,
-\begin{align*}
-1-x^2-y^2-x^2y^2 & = (1+x)(1-x)(1+y)(1-y) \\
-1-x^2+y^2-x^2y^2 & = (1+x)(1-x)(1+y^2) \le (1+x)(1-x)(1+y) \\
-1+x^2-y^2-x^2y^2 & = (1+x^2)(1+y)(1-y) \le (1+x)(1+y)(1-y) \\
+Our plan is to simplify the likelihood to remove log-of-quadratic terms of $x,y,$ and $w$, obtaining a likelihood that has a closed-form maximum in each variable.
+To bound the generating probabilities, we use the facts that, for $x,y\in[0,1]$,
+\begin{align*}
+p_{12} = p_{23} = 1-x^2-y^2+x^2y^2 & = (1+x)(1-x)(1+y)(1-y) \\
+p_{1} = p_{3} = 1-x^2+y^2-x^2y^2 & = (1+x)(1-x)(1+y^2) \le (1+x)(1-x)(1+y) \\
+p_{2} = p_{123} = 1+x^2-y^2-x^2y^2 & = (1+x^2)(1+y)(1-y) \le (1+x)(1+y)(1-y).
+\end{align*}
+To bound the remaining $p_{\emptyset}$ and $p_{13}$, we use
+\begin{align*}
 1+x^2+y^2+x^2y^2 & = (1+x^2)(1+y^2) \le (1+x)(1+y) \\
-4xy & = 2x \cdot 2y \le (1+x^2)(1+y^2) \le (1+x)(1+y).
-\end{align*}
-The $p_{\emptyset}$ term is bounded as
+4xy & = 2x \cdot 2y \le (1+x^2)(1+y^2) \le (1+x)(1+y)
+\end{align*}
+so that the $p_{\emptyset}$ term is bounded as
 \begin{align*}
@@
     &\qquad\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+y)(1-y))
-+ p_{123}        \cdot\log((1+x)(1+y)(1-y))
-- \log(8) \\
++ p_{123}        \cdot\log((1+x)(1+y)(1-y)) \\
     &\qquad\qquad + p_{\emptyset}  \cdot\log((1+x)^2   (1+w)(1+y)^2)
@@
 &\qquad\qquad + (1-p_{13})\cdot\log(1+w)
-+ p_{13}\cdot\log(1-w)
-- \log(8) \\
++ p_{13}\cdot\log(1-w) \\
 &\qquad =      p_{\emptyset}  \cdot\log(2)
@@
 &\qquad\qquad + (1-p_{13})\cdot\log(1+w)
-+ p_{13}\cdot\log(1-w)
-- \log(8).
++ p_{13}\cdot\log(1-w).
 \end{align*}
@@
 + (1-p_{13})\cdot\log(1+w)
-+ p_{13}\cdot\log(1-w)
-- \log(8).
++ p_{13}\cdot\log(1-w).
 \end{split}
@@
 $$
-Trivially $0 \le \hat{w} \le 1$.
-We check that $0 \le \hat{x} \le 1$ and $0 \le \hat{y} \le 1$ so these maximizers are valid.
-Letting
+Trivially $0 \le \hat{w} \le 1$ and $\hat{x}, \hat{y} \le 1$.
+We check that $\hat{x}, \hat{y} \ge 0$ so these maximizers are valid.
+For $\hat{x} \ge 0$ we need $a_1 \ge a_2$; letting
 $$
@@
 $$
-we have
-$$
-\hat{x} = \frac{3-\tilde{p} - 2\tilde{p}}{3-\tilde{p}+2\tilde{p}} = \frac{3-3\tilde{p}}{3+\tilde{p}},
-$$
-and, since $0 \le \tilde{p} \le 1/2$, $0 \le \hat{x} \le 1$.
+we have from \eqref{eq:a_const} that $a_1 \ge a_2$ only if $3 \ge 3\tilde{p}$, and, since $0 \le \tilde{p} \le 1/2$, $\hat{x} \ge 0$.
 The same approach works for $\hat{y}$.
@@
 + (1-p_{13})\cdot\log(2(1-p_{13}))
-+ p_{13}\cdot\log(2p_{13})
-- \log(8) \\
++ p_{13}\cdot\log(2p_{13}) \\
 &\qquad := C^{1}_{\tau_1,\tau_1}(x^*, y^*).
@@
 %EM Yay! Would be even better if you specified which upper bound (eqref) you want us to be looking at... there are many.
-For the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term and gain a $(1+w)$ term.
+For the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.
 In this case, all terms involving $w$ simplify to $\log(1+w)$, maximized at $\hat{w}=1$.
@@
 + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
-+ \log(2)
-- \log(8) \\
++ \log(2) \\
 &\qquad := C^{2}_{\tau_1,\tau_1}(x^*, y^*),
@@
 + a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
-+ \log(2)
-- \log(8) \\
++ \log(2) \\
 &\qquad := C^{3}_{\tau_1,\tau_1}(x^*, y^*).
@@
 are both between zero and one.
-It will suffice to show
-$$
-0 \le \tilde{p} - 2p_{13} \le 1
-$$
-and
-$$
-0 \le \tilde{p} + 2p_{13} \le 1
-$$
-for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
+Using a similar argument as that where $\ancestralSplitPartition_j(\tau_1, t) = \{1,3\}$, we only need to show $a_1' \ge a_2'$, which is true if $3-2p_{13} \ge 3\tilde{p}+2p_{13}$.
+Some rearranging and the fact that $0 \le \tilde{p} \le 1/2$ shows this is equivalent to showing $p_{13} \le 3/8$.
 Using Table~\ref{tab:sitepatprob} and $t^*$,
 $$
-p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right).
-$$
+p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right),
+$$
+meaning we are interested in whether
+\begin{equation}
+1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3.
+\label{eq:generating_ineq}
+\end{equation}
 We see that
 \begin{equation}
-1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 2 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - (x^*)^2.
+1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 2 - (x^*)^2.
 \label{eq:lenny}
 \end{equation}
-Note
-$$
-(y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - 4x^* + (x^*)^2;
-$$
-showing that \eqref{eq:lenny} holds if $x^* \le 2$, which is satisfied by definition.
-Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}$ and $\hat{y}$ are valid.
+and that
+\begin{align*}
+(y^*)^2\left(1 - 4x^* + (x^*)^2\right) &\le (y^*)^2\left(1 - 2x^* + (x^*)^2\right) \\
+                                       &=   (y^*)^2\left(1 - x^*\right)^2 \\
+                                       &\le        \left(1 - x^*\right)^2 \\
+                                       &\le              2 - (x^*)^2,
+\end{align*}
+with the last inequality holding if $0 \le x^* \le 1$, showing that \eqref{eq:lenny} holds.
+Thus, $p_{13} \le 3/8$ and our $\hat{x}$ and $\hat{y}$ are valid.
 
@@
 in these cases.
-The full likelihood in this case is then bounded below by
+The partial likelihood in this case is then bounded below by
 \begin{align*}
@@
 + (2-b)  \cdot\log(2-b)
-+ b      \cdot\log(b)
-$$
-i.e.\ the maximum is achieved at $\{1,1-b,0\}$.
++ b      \cdot\log(b),
+$$
+i.e., the maximum is achieved at $\{1,1-b,0\}$.
 Our lower bound is
@@
 which is a function of the generating parameters $x^*$ and $y^*$.
-We see that
+We see that $a \le 0$ and
 $$
@@
 %-b - \sqrt{b^2 - 4ac} \le 2a,
-%-b - 2a \le \sqrt{b^2 - 4ac},
+%0 \ge -b - 2a \le \sqrt{b^2 - 4ac},
 %b^2 + 4ab + 4a^2 \le b^2 - 4ac,

### 2017-12-31 appendix.tex
Finished pass through appendix.

@@
 \subsection*{Site split formulation}
+We begin by introducing ``site splits,'' that simply formalize the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
+This is a standard step in the description of the Hadamard transform (Section 8.6 of \cite{Semple2003-em}), although we have to be a little careful here due to the extra bookkeeping of ancestral states.
 
@@
 For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
-We start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
-Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.
+We define the site split $\siteSplit$ for a $\alignmentColumn_i$ as simply $\alignmentColumn_i$ if the label $\nSiteRows$ is not in $\alignmentColumn_i$, and as its complement otherwise.
+Taking such a complement simplifies but does not change the result of likelihood computation because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model.
 
@@
 As a consequence of assuming a binary symmetric model, taking complements yields
-%EM continuing to nitpick, is there anything that is just s_j? Seems funny to only have the tilde s_j form.
-%das: I think s_j used to be the site pattern and then \tilde{s}_j became the split?
-It's still the case for h (no tilde is ancestral state pattern, tilde is ancestral state split).
-I changed to \tilde{y} to be consistent with this, but mabye we can think of something better?
 \begin{align*}
@@
 \end{align}
-where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.
+where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.
 
@@
 $$
-%EM It seems a little strange that eta appears here as an argument when it didn't appear in eq:site_pattern_likelihood.
-In fact it's not really an argument, right?
-Everything is determined by t and tau it seems to me.
-%das: We added it to clarify that the likelihood now depends on \xi, which is an element of \eta.
-It seems dirty to ignore it.
-At the risk of adding more notation, we could add \hat{H} to the original likelihood?
-Or just axe the etas...
-%EM I think I must be missing something:
-%EM The likelihood doesn't belong on the specific choice of \xi. Besides, \xi already appears in eq:site_pattern_likelihood.
-%EM Eta is a deterministic function of t and tau, so isn't having it redundant?
-%das: The likelihood doesn't depend on the specific choice of \xi, but it does depend on \eta.
-I don't know, I guess redundancy isn't always the worst thing?
-I just see it as a tradeoff between how explicit we want to be and how clean.
-Like, in the first line of eq:site_pattern_likelihood we could have as an argument \hat{H} instead of suppressing it, even though it's totally determined by the inputs as-is.
-Anyway, that was my justification for including \eta.
-I'm fine with removing it though if you think it's weird.
-%das: I changed all this.
 L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
@@
 
-We follow Section 8.6 of \cite{Semple2003-em} to state the Hadamard representation of site split generating probabilities.
-For each edge $e$ define the edge fidelity for that edge as
+We now state the Hadamard representation of site split generating probabilities, following Section 8.6 of \cite{Semple2003-em}.
+For each edge $e$ define the edge \emph{fidelity} for that edge as
 $$
@@
 \end{tabular}
+%EM When I came back to this after the break I found it confusing that we refer to these as p_A below (eg in proof of Theorem 1) but that notation doesn't appear in this table. Can we use it here?
 \caption{Site pattern probabilities $\siteSplit_j$ on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
@@
 \end{tabular}
-\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the Farris tree.
+\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the Farris tree $\tau_1$.
 All values multiplied by $1/32$.}
@@
 \end{tabular}
-\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the Felsenstein tree.
+\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the Felsenstein tree $\tau_2$.
 All values multiplied by $1/32$.}
@@
 \begin{tabular}{|l|ll|}
-    \multicolumn{3}{c}{$\tau=\tau_1$}\\
+    \multicolumn{3}{c}{Farris tree ($\tau=\tau_1$)}\\
     \hline
@@
     \hline
-    \multicolumn{3}{c}{$\tau=\tau_2$}\\
+    \multicolumn{3}{c}{Felsenstein tree ($\tau=\tau_2$)}\\
     \hline
@@
 \end{tabular}
-\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$ after maximizing over ancestral states.
+\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$ after maximizing over ancestral states on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$.
 All values multiplied by $1/32$.
@@
 
-%EM I think this following statement could use a little more justification.
-If we write out a symmetric function f in two variables and take the derivative of f(x+delta, x) = f(x, x+delta) with respect to delta and set it to zero we can see that this must be a stationary point of f.
-Do we need to say more to ensure that this is a maximum?
-%das: I wasn't even thinking derivatives.
-Let f(x, y) be symmetric so that f(x, y) = f(y, x).
-If f is maximized at some (x_0, y_0) then f(x_0, y_0) > f(x, y) for all x, y.
-But then by symmetry f(y_0, x_0) is also a maximum.
-Now *if* a unique maximum exists, x_0 = y_0 should necessarily be this maximum by the properties defined above, otherwise, f(x_0, y_0) would not be equal to f(y_0, x_0).
-I think there are probably counterexamples where there is not a unique maximum where this statement isn't necessarily true (trivially, because you'd just say x_0 = y_1 if y_1 is also a maximum and voila) but I don't think this discounts the fact that x_0 = y_0 will still be *a* maximum in this case.
-In that case, it's reasonable to assume so, in my opinion, especially for the topological inconsistency (we don't use the branch length values), but even for the branch length inconsistency because the case of no unique solution is just as bad if not worse than inconsistency.
-%das: Should some or all of that be in the text? Can you see anything wrong with it?
-%EM I don't see why this works.
-In fact, imagine that x=y is a ridge of minima, and that we have maxima at (1,2) and at (2,1).
-My argument shows that we must have a stationary point at x=y, but that doesn't mean that it has to be a maximum!
-%das: Wouldn't the assumption of there existing a unique maximum preclude your case?
-My argument is that if there is a unique maximum, it necessarily has to have x=y.
-If there is not a unique maximum then the ML problem is ill-posed.
-I now don't see why we have to have a unique maximum other than to say we hope the problem is well-posed, but it may not be.
-Do you have any thoughts on where to take this?
-%das: So there are references for identifiability, but I'm a little nervous using them because they're for the usual marginal ML problem.
-How cagey does it sound to say (as I do below) "here are references for identifiability in the usual case, so either joint inf will be not identifiable (bad) or it (likely?) will be and we can use this (which turns out to be inconsistent later)?" Can you think of a better approach?
-%das: some other references I found for identifiability are Rogers 1997, RoyChoudhury 2014...
-Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, $x_1=x_2$ at the maximum provided a unique maximum exists.
+Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood we will have $x_1=x_2$ at the maximum.
 An analogous statement holds for $y_1$ and $y_2$.
 It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
-Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference, similar arguments holding for the remaining parameters.
+Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments holding for the remaining parameters.
 We decompose the likelihood into the entropy and partial likelihood terms as in \eqref{eq:log_likelihood_simplified}.
@@
 For the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$, we can compute the partial likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ as in Table~\ref{tab:farris_likelihoods}.
+%EM why is 1/2 \le p obvious?
 Clearly this partial likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $1/2 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in[0,1]$ and thus has a unique maximum.
@@
 The Felsenstein tree does not admit this property of symmetry of function inputs, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
+%EM WOOT
 
@@
 with similar definitions for the remaining generating probabilities.
-The full Farris tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see Table~\ref{tab:likelihoods}).
+The full Farris tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see the upper table of Table~\ref{tab:likelihoods}).
+
 We begin by writing out the case for the ancestral state split $\{1\}$, and show the other two cases follow a similar argument.
-Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_profile_likelihood_mean}, the log-likelihood is, suppressing normalizing constants $-\log 8$ and $-\log 32$,
+Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_profile_likelihood_mean}, the log-likelihood is, suppressing normalizing constants $-\log 8$ (from the first additive term of \eqref{eq:site_pattern_profile_likelihood_mean}) and $-\log 32$ (from the second),
 \begin{align}
@@
 \end{align*}
-Factoring and making these substitutions results in
+Factoring and making these substitutions (again, under the assumption that the ancestral state split is $\{1\}$) results in
 \begin{align*}
@@
 Trivially $0 \le \hat{w} \le 1$ and $\hat{x}, \hat{y} \le 1$.
-We check that $\hat{x}, \hat{y} \ge 0$ so these maximizers are valid.
+We check that $\hat{x}, \hat{y} \ge 0$ to ensure these maxima are valid.
 For $\hat{x} \ge 0$ we need $a_1 \ge a_2$; letting
@@
 $$
+%EM Do we get $0 \le \tilde{p} \le 1/2$ by just adding together terms in Table 1?
 we have from \eqref{eq:a_const} that $a_1 \ge a_2$ only if $3 \ge 3\tilde{p}$, and, since $0 \le \tilde{p} \le 1/2$, $\hat{x} \ge 0$.
@@
 The upper bound in \eqref{eq:farris-upper-bound} is for the ancestral state split $\{1\}$.
-%EM Yay! Would be even better if you specified which upper bound (eqref) you want us to be looking at... there are many.
 For the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.
@@
 \begin{equation}
-1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 2 - (x^*)^2.
+1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 2 - (x^*)^2
 \label{eq:lenny}
@@
 We construct a similar lower bound on the Felsenstein tree partial likelihood.
-In the Felsenstein case, there are many more optimal internal states depending on branch lengths than in the Farris case (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).
+In the Felsenstein case, there are many more optimal internal states depending on branch lengths than in the Farris case (see the lower table of Table~\ref{tab:likelihoods}).
 We first bound the entire likelihood below and then proceed with joint inference.
@@
 $$
-%EM Hm, the following doesn't look right to me.
-%das: Okay, think I missed the point about 1+w for 1-w that I added back.
-%EM No, the lower bound is (1+x)(1-x)(1-w)(1+y)(1-y).
-%das: I agree that that's *a* lower bound, yeah.
-So maybe tell me what's wrong with the following: L is the general likelihood in the joint inf problem for the Felsenstein topology, obtained by doing the usual maximization over ancestral states in tab:likelihoods.
-Now let's obtain L' via the same process but we've now replaced all (1+w)s in that table with (1-w)s.
-L \ge L', so then L' is a lower bound for the likelihood L, yes?
-So looking at, let's call it the lower bound version of tab:likelihoods where everything is a (1-w), maximizing over ancestral states then results in the above inequality.
-Maybe I'm being careless?
-%das: Is this better?
 always, the maximum for this site split can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
@@
 \begin{proof}
-Recalling Table~\ref{tab:likelihoods}, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood over $(x,y,w)$.
+Recalling Table~\ref{tab:likelihoods} for the Farris tree $\tau = \tau_1$, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood over $(x,y,w)$.
 If $x=x^*$ and $y=y^*$ are fixed, by collapsing like terms and ignoring terms with only $x^*$ and $y^*$, the log-likelihood in \eqref{eq:farris_likelihood} as a function of $w$ is
@@
 c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
-Excepting all the cases where $x^*=0$ or where $y^*=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
-Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w=0$.
+Excepting all the cases where $x^*=0$ or where $y^*=0$, this implies $\ell'$ has two zeros according to the quadratic formula and $\alpha_1 > \alpha_2$.
+Also because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w=0$.
 This implies that if the smaller of the zeros of $f(w)$ is greater than one, then $\hat{w} \equiv 1$.
@@
 We initialize our optimization procedure at the true branch parameters $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
-We posit that $\hat{w}$ is equal to 1 for when $x^*$ and $y^*$ are small; methods that use local optimization may never see candidate values for large $x$ and $y$ if our initial guess is small, so we perform two maximizations, one with $w=1$ fixed and one where $w$ is estimated.
+Because our analysis has shown that that $\hat{w}$ can be equal to 1 for when $x^*$ and $y^*$ are small, and local optimization may have a hard time obtaining this value if our initial guess is small, we perform two maximizations, one with $w=1$ fixed and one where $w$ is estimated.
 We take the value of $\hat{w}$ with the larger likelihood as our estimate.
@@
 \caption{
-    Estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:marginal_likelihood}.
+    Estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing the classical marginal likelihood \eqref{eq:marginal_likelihood} (rather than a joint optimization procedure).
 }
@@
 \caption{
+    %EM I don't think you mean eq:marginal_likelihood here.
     Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:marginal_likelihood}.
     Plot focuses on $0.6 < x^*, y^* < 1$ where $\hat{w}$ is estimated to be a value different from 1.
-    We do not compute the bias for the white region where $\hat{w}=1$.
+    We do not compute the bias for the white region where $\hat{w}=1$ to preserve a useful scale for the rest of the plot.
 }

### 2018-01-01 appendix.tex
A little more detail and a factor of 2?

@@
 which is a function of the generating parameters $x^*$ and $y^*$.
-We see that $a \le 0$ and
-$$
-\hat{w} \ge 1
-$$
-is equivalent to
+We see that $a \le 0$ and, by a small calculation, $2a+b \leq 0$.
+With this we have,
+$$
+\hat{w} \ge 1 \iff
+|2a + b| \leq \sqrt{b^2 - 4ac} \iff
 %-b - \sqrt{b^2 - 4ac} \le 2a,
@@
 %a^2 + ab + ac \le 0,
-$$
-a + b + c \ge 0,
-$$
-which, using
+a + b + c \ge 0.
+$$
+Using
 $$
@@
 $$
-and simplifying as functions of $\gamma$ and $\beta$ yields
+and simplifying as functions of $\gamma$ and $\beta$ shows that $a + b + c \ge 0$ is equivalent to
 \begin{equation}
@@
 %\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
+%EM Something's not quite right here. If we take the previous comment as truth we should get a
+%EM -\gamma^2(1 + \beta) term which is what I get via independent calculation.
 -\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0.
@@
 Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
-Since
-$$
-\alpha_1 + \alpha_2 = \frac{1}{2}\beta
-$$
-and
-$$
-\alpha_1 - \alpha_2 = x^*(y^*)^2,
-$$
-simplifying \eqref{eq:restricted-bl-result} further yields
+The same derivation as for \eqref{eq:restricted-bl-result} further yields
 \begin{equation}
@@
 then \eqref{eq:general-bl-result} holds, and this is an inequality involving only $x^*$ and $y^*$.
+\end{proof}
 To get a sense of the region of overestimation, Fig.~\ref{fig:bl-loose-inconsistency} plots an intermediate case.
-\end{proof}
-
 \subsection*{Empirical validation}

### 2018-01-01 joint_inf.tex
noticed minor error; fix plus add clarification
A little more detail and a factor of 2?

@@
 %\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
--\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0.
+-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0
 $$
+and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
 \end{restatable}
@@
 $$
+%EM It's a little surprising that "naked" beta shows up here.
+While it seems like a correct statement, why can we have beta here to make the plots, while in another term we have to bound it with beta_L?
+Seems like we'd need a beta_U.
 -\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
@@
 
-We have shown that the method of joint inference when estimating phylogenetic parameters \cite{Sagulenko2017-jo} is not consistent in general.
+We have shown that maximum-likelihood joint inference when estimating phylogenetic parameters \cite{Sagulenko2017-jo} is not consistent in general.
 Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference infers the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
@@
 
-Joint inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \cite{Murphy2000-ry}.
+Joint maximum-likelihood inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \cite{Murphy2000-ry}.
 Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., \cite{Geman1982}.
@@
 
+As a final reminder, we note that all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.
+
 \section*{Acknowledgements}

### 2018-01-02 appendix.tex
address comments

@@
 %EM When I came back to this after the break I found it confusing that we refer to these as p_A below (eg in proof of Theorem 1) but that notation doesn't appear in this table. Can we use it here?
+%das: Does L540 not help with this? Are you thinking we should add another column with "Notation" or something?
 \caption{Site pattern probabilities $\siteSplit_j$ on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
@@
 %EM why is 1/2 \le p obvious?
-Clearly this partial likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $1/2 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in[0,1]$ and thus has a unique maximum.
+%das: I think it's true, but might take some annoying exposition to show. However, p is clearly between zero and one, which still passes the second derivative test.
+Clearly this partial likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in[0,1]$ and thus has a unique maximum.
 Since both summands of the likelihood are concave on $x_1\in[0,1]$, the likelihood is as well, and there exists a single $x_1\in[0,1]$ maximizing the likelihood.
@@
 %EM Do we get $0 \le \tilde{p} \le 1/2$ by just adding together terms in Table 1?
-we have from \eqref{eq:a_const} that $a_1 \ge a_2$ only if $3 \ge 3\tilde{p}$, and, since $0 \le \tilde{p} \le 1/2$, $\hat{x} \ge 0$.
+%das: Yah, if you add them it's 1/8 (4-4x^2) for 0<x<1.
+we have from \eqref{eq:a_const} that $a_1 \ge a_2$ only if $3 \ge 3\tilde{p}$.
+Since $\tilde{p} = 1/8\cdot(4-4(x^*)^2)$, $0 \le \tilde{p} \le 1/2$ implying $\hat{x} \ge 0$.
 The same approach works for $\hat{y}$.
@@
 $$
-\alpha_1 + \alpha_2 = \frac{1}{2}\beta
+\alpha_1 + \alpha_2 = \frac{1}{4}\beta
 $$
@@
 %EM -\gamma^2(1 + \beta) term which is what I get via independent calculation.
+%das: I missed a factor of 1/2 in the \alpha_1+\alpha_2 term. Oops. I think below is right given that.
 -\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0.
@@
 \label{eq:general-bl-result}
--\gamma^2(\hat{x}, \hat{y})\left(1 + \frac{1}{2}\beta\right) + 2\gamma(\hat{x}, \hat{y})\beta(\hat{x}, \hat{y})x^*(y^*)^2 + \beta^2(\hat{x}, \hat{y}) \ge 0.
+-\gamma^2(\hat{x}, \hat{y})\left(1 + \frac{1}{2}\beta\right) + 2\gamma(\hat{x}, \hat{y})\beta(\hat{x}, \hat{y})x^*(y^*)^2 + \beta^2(\hat{x}, \hat{y}) \ge 0
 \end{equation}
+where, since $\alpha_1$ and $\alpha_2$ are still functions of $x^*$ and $y^*$, $\left(1+1/2\cdot\beta\right)$ and $x^*(y^*)^2$ from \eqref{eq:restricted-bl-result} remain unchanged.
 Given the bounds
@@
     %EM I don't think you mean eq:marginal_likelihood here.
-    Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:marginal_likelihood}.
+    %das: Ha, you're right.
+    Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:profile_likelihood}.
     Plot focuses on $0.6 < x^*, y^* < 1$ where $\hat{w}$ is estimated to be a value different from 1.

### 2018-01-02 appendix.tex
consistent Pr() notation and update table

@@
 $$
-P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
+\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
 $$
@@
 \begin{align*}
-    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
-                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
+    2\cdot \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
+                                                                 &= \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)
 \end{align*}
@@
 \begin{align*}
-    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
+    \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
 \end{align*}
@@
 \begin{align*}
-\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
+\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
 \end{align*}
@@
 \begin{align*}
-\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
+\max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
 \end{align*}
@@
 L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
-                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
-                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
-                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
-                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
+                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
+                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
+                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
+                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
 \end{align}
@@
 $$
-L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
+L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
 $$
@@
 &    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
-&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
-&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
-            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
+&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
+&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) +
+            \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
 \end{split}
@@
 &    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
-&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)].
+&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)].
 \label{eq:site_pattern_profile_likelihood_mean}
@@
 $$
-\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t)
+\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)
 $$
@@
 $$
-\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
+\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
 $$
@@
 By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
+For notational convenience, let
+$$
+p_{\emptyset} := \Pr(\siteSplitRV=\emptyset \mid \tau_1,t),
+$$
+with similar definitions for the remaining generating probabilities.
 Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies for our two topologies (Fig.~\ref{fig:farris-fels-top}).
@@
 \centering
-\begin{tabular}{|l|l|l|}
-    \hline
-$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
-    \hline
-    $\emptyset$    &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
-    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
-    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
-    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
-    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
-    $\{1,3\}$      &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
-    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
-    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
+\begin{tabular}{|l|l|l|l|}
+    \hline
+$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
+    \hline
+    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
+    $\{1\}$     & $p_{1}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
+    $\{2\}$     & $p_{2}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
+    $\{3\}$     & $p_{3}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
+    $\{1,2\}$   & $p_{12}$   &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
+    $\{1,3\}$   & $p_{13}$   &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
+    $\{2,3\}$   & $p_{23}$   &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
+    $\{1,2,3\}$ & $p_{123}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
     \hline
@@
 %EM When I came back to this after the break I found it confusing that we refer to these as p_A below (eg in proof of Theorem 1) but that notation doesn't appear in this table. Can we use it here?
-%das: Does L540 not help with this? Are you thinking we should add another column with "Notation" or something?
-\caption{Site pattern probabilities $\siteSplit_j$ on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
+\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
 All values multiplied by $1/8$.}
@@
 $$
-P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
+\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
 $$
@@
 $$
-P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
+\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
 (1-p)^5,
@@
 $$
-P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
-P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
+\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
+\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
 p^3(1-p)^2,
@@
 $$
-P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
+\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
 p^4(1-p).
@@
 $$
-P(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
-P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
+\Pr(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
+\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
 (1-p)^5.
@@
 $$
-P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
-P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
+\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
+\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
 p^2(1-p)^3,
@@
 $$
-P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
-P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
+\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
+\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
 p^3(1-p)^2.
@@
 $$
-P(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
+\Pr(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
 p^2(1-p)^3.
@@
 
-To compute the likelihood of observing a set of data, we need $P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
+To compute the likelihood of observing a set of data, we need $\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
 We remind the reader that the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
@@
     Edges labeled by the probability of substitution along that edge.
-    In (a), we compute the product to obtain $P(\ancestralStateColumn=\emptyset\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1+w)/32$.
-    In (b), the same process yields $P(\ancestralStateColumn=\{1\}\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1-w)/32$.
+    In (a), we compute the product to obtain $\Pr(\ancestralStateColumn=\emptyset\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1+w)/32$.
+    In (b), the same process yields $\Pr(\ancestralStateColumn=\{1\}\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1-w)/32$.
 }
@@
 \begin{tabular}{|l|ll|}
-\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
+\multicolumn{3}{c}{$\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
 \hline
@@
 \begin{tabular}{|l|ll|}
-\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_2,t)$}\\
+\multicolumn{3}{c}{$\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_2,t)$}\\
 \hline
@@
     \hline
-    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
+    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
     \hline
@@
 \begin{align*}
-    P(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
+    \Pr(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                               & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
@@
 \begin{align*}
-    P(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
+    \Pr(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
 \end{align*}
@@
 \begin{align*}
-    P(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
+    \Pr(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
 \end{align*}
@@
 \begin{align*}
-    P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
-    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
-    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
-    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
+    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
+    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
+    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
+    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
 \end{align*}
@@
 \begin{align*}
-        P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
-    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
-    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
-    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
+        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
+    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
+    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
+    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
 \end{align*}
@@
 Assume $\tau^*=\tau_1$.
-Following \eqref{eq:hadamard_probability}, let
-$$
-p_{\emptyset} := p_{\emptyset}(x^*, y^*) = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=t^*),
-$$
-with similar definitions for the remaining generating probabilities.
 The full Farris tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see the upper table of Table~\ref{tab:likelihoods}).

### 2018-01-02 appendix.tex
:woot: :woot: :woot:

@@
 $$
-p_{\emptyset} := \Pr(\siteSplitRV=\emptyset \mid \tau_1,t),
-$$
-with similar definitions for the remaining generating probabilities.
+p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
+$$
+for any site split $\siteSplit_j$.
 Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies for our two topologies (Fig.~\ref{fig:farris-fels-top}).
@@
 \end{tabular}
-%EM When I came back to this after the break I found it confusing that we refer to these as p_A below (eg in proof of Theorem 1) but that notation doesn't appear in this table. Can we use it here?
 \caption{Site pattern probabilities $p_{\siteSplit_j}$ on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
@@
 For the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$, we can compute the partial likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ as in Table~\ref{tab:farris_likelihoods}.
-%EM why is 1/2 \le p obvious?
-%das: I think it's true, but might take some annoying exposition to show. However, p is clearly between zero and one, which still passes the second derivative test.
 Clearly this partial likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in[0,1]$ and thus has a unique maximum.
@@
 The Felsenstein tree does not admit this property of symmetry of function inputs, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
-%EM WOOT
 
@@
 $$
-%EM Do we get $0 \le \tilde{p} \le 1/2$ by just adding together terms in Table 1?
-%das: Yah, if you add them it's 1/8 (4-4x^2) for 0<x<1.
 we have from \eqref{eq:a_const} that $a_1 \ge a_2$ only if $3 \ge 3\tilde{p}$.
-Since $\tilde{p} = 1/8\cdot(4-4(x^*)^2)$, $0 \le \tilde{p} \le 1/2$ implying $\hat{x} \ge 0$.
+Since $\tilde{p} = 1/8\cdot(4-4(x^*)^2)$, we have $0 \le \tilde{p} \le 1/2$ implying $\hat{x} \ge 0$.
 The same approach works for $\hat{y}$.
@@
 For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
-%EM It would seem helful for the below check to express these a_1' etc in terms of tilde{p}.
-%das: not quite sure what this comment means actually.
 \begin{equation}
@@
 for site split $\{1,2\}$.
-%EM This check would make a lot more sense if you put the C^{2}_{\tau_1,\tau_1}(x^*, y^*)  and the other analogous definition first.
-%EM It would also be nice to do one of the calculations to get us from a_1 world to a_1' world. Just staring at table S4 doesn't make this obvious.
-%das: is this a better explanation or did you want it written out explicitly?
 To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
@@
 %\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-%EM Something's not quite right here. If we take the previous comment as truth we should get a
-%EM -\gamma^2(1 + \beta) term which is what I get via independent calculation.
-%das: I missed a factor of 1/2 in the \alpha_1+\alpha_2 term. Oops. I think below is right given that.
 -\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0.
@@
 \caption{
-    %EM I don't think you mean eq:marginal_likelihood here.
-    %das: Ha, you're right.
     Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:profile_likelihood}.

### 2018-01-02 joint_inf.tex
address comments
:woot: :woot: :woot:
Main text wordsmithing.

@@
 \begin{abstract}
-Maximum likelihood is a widely used and well understood approach to estimate the phylogenetic topology and branch length given observed character data.
+Maximum likelihood is a widely used and well understood approach to estimate phylogenetic topology and branch length given observed character data.
 Any phylogenetic maximum likelihood solution must propose a means of handling unknown ancestral states.
@@
 Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
-While this method of joint maximum likelihood estimation can obtain estimates more efficiently, it has the potential to result in undesired properties of the estimates.
+While this method of joint maximum likelihood estimation can obtain estimates more efficiently, its properties as an estimator are not yet clear.
 In this paper, we show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
-We derive a region where generating data under a four-taxon tree results in a joint estimation procedure converging on the wrong topology.
-If the topology is known a priori, there are similar cases where the branch lengths are consistently overestimated in the case of infinite-length sequences.
-Although joint maximum likelihood may turn out to be a useful tool in practice, care should be taken when using this approach.
+We derive a region where generating data under a four-taxon tree results in a joint estimation procedure that gives incorrect topological inferences in the limit of infinite-length sequences.
+For branch length estimation on a correct topology, there are similar cases where the branch lengths are consistently overestimated in the case of infinite-length sequences.
+Thus although joint maximum likelihood may turn out to be a useful tool in practice, care should be taken when using this approach.
 \end{abstract}
@@
 
-Existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
+Statistical consistency means that an estimator gives the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
 These proofs do not readily extend to include estimating ancestral states.
@@
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model and imagine data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
+To do so, we use a binary symmetric model with data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
 We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.
@@
 
-To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the Farris topology.
+To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the Farris topology (Fig.~\ref{fig:farris-fels-top}a).
 We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology than for the true, generating Farris topology.
@@
 $$
-%EM It's a little surprising that "naked" beta shows up here.
-While it seems like a correct statement, why can we have beta here to make the plots, while in another term we have to bound it with beta_L?
-Seems like we'd need a beta_U.
 -\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
@@
 
-Previous work in phylogenetics has developed consistency counterexamples using the same four-taxa topologies used here \cite{Felsenstein1978-rr}.
+Previous work in phylogenetics has developed consistency counterexamples using the same four-taxon topologies used here \cite{Felsenstein1978-rr}.
 In this previous work, when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more optimal when performing a particular estimation procedure.

### 2018-01-02 joint_inf.tex
just some typos, inconsistencies, and one clarification

@@
 \begin{abstract}
-Maximum likelihood is a widely used and well understood approach to estimate phylogenetic topology and branch length given observed character data.
+Maximum likelihood is a widely used and well understood approach to estimate phylogenetic topology and branch lengths given observed character data.
 Any phylogenetic maximum likelihood solution must propose a means of handling unknown ancestral states.
@@
 In this case those additional parameters come in the form of the states of ancestral sequences.
-Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is generated without sampling error according to the exact model used for inference.
+Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.
 
@@
 We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with an alternate formulation called ``fidelity.''
-The probability of a substitution on a branch with fidelity $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$ where $0 \le \theta \le 1$.
+The probability of a substitution on a branch with fidelity $\theta$ is $(1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$ where $0 \le \theta \le 1$.
 This parameter thus quantifies the fidelity of transmission of the ancestral state across an edge \cite{Matsen2007-jq}.
@@
 
-The classical approach to is to maximize the likelihood marginalized across ancestral states
+The classical approach is to maximize the likelihood marginalized across ancestral states
 \begin{equation}
@@
 Assume $\fullAlignment$ is generated from topology $\tau^*$ and branch fidelities $t^*$.
-Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log likelihood, which can be thought of as the infinite-length sequence case
+Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
 $$
@@
 Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
-There exist $C_0(x^*, y^*), C_1(x^*, y^*)$ and a set of $0 < x^*, y^* < 1$ such that
+There exist $C_0(x^*, y^*), C_1(x^*, y^*),$ and a set of $0 < x^*, y^* < 1$ such that
 $$
@@
 We now consider the problem of branch length estimation on the correct tree using joint ancestral state and branch length estimation.
-As described above, we will use the equivalent but different notion of branch fidelities.
+As described above, we use the equivalent but different notion of branch fidelities.
 We analyze two settings on the Farris tree, corresponding to whether some branch fidelities are fixed at their true values or not.
-As above, assume that data is generated from the Farris tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Figure~\ref{fig:farris-fels-top}).
+As above, assume that data is generated from the Farris tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
 In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
@@
 This theorem allows us to demarcate a region of bad internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
-Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch, and then a branch length is estimated to be zero (i.e.\ have fidelity $\hat{w} = 1$).
+Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch, and then a branch length is estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
 As an intuition for the theoretical development, seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.
@@
 
-Direct numerical optimization confirms our theoretically-derived bounds and provides more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:bl-loose-inconsistency}).
+Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:bl-loose-inconsistency}).
 To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint maximum likelihood estimation using numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}).
@@
 Joint maximum-likelihood inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \cite{Murphy2000-ry}.
-Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., \cite{Geman1982}.
+Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties \cite{Geman1982}.
 A potential solution in this case using the method of sieves could be to project the column-wise ancestral state patterns into a lower-dimensional space, allowing the degrees of freedom in the ancestral state columns to grow with $\nCols$, albeit more slowly than $O(\nCols)$.

### 2018-01-02 joint_inf.tex
forgot :w

@@
 Joint maximum-likelihood inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \cite{Murphy2000-ry}.
-Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties \cite{Geman1982}.
+Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., the method of sieves \cite{Geman1982}.
 A potential solution in this case using the method of sieves could be to project the column-wise ancestral state patterns into a lower-dimensional space, allowing the degrees of freedom in the ancestral state columns to grow with $\nCols$, albeit more slowly than $O(\nCols)$.
@@
 Previous work in phylogenetics has developed consistency counterexamples using the same four-taxon topologies used here \cite{Felsenstein1978-rr}.
-In this previous work, when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more optimal when performing a particular estimation procedure.
+In this previous work, when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
 This is the converse of what we have shown for joint inference, where the Felsenstein topology is more likely than the Farris topology.

### 2018-01-26 joint_inf.tex
abstract

@@
 \begin{abstract}
-Maximum likelihood is a widely used and well understood approach to estimate phylogenetic topology and branch lengths given observed character data.
-Any phylogenetic maximum likelihood solution must propose a means of handling unknown ancestral states.
-Classically, these unknown intermediate states are averaged out, which leads to consistent estimation of the topology and continuous model parameters.
-Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
-While this method of joint maximum likelihood estimation can obtain estimates more efficiently, its properties as an estimator are not yet clear.
+Maximum likelihood estimation of phylogeny requires a means of handling unknown ancestral states.
+Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
+Recently, a computationally-efficient approach was proposed to jointly maximize over these unknown states and phylogenetic parameters.
 In this paper, we show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
-We derive a region where generating data under a four-taxon tree results in a joint estimation procedure that gives incorrect topological inferences in the limit of infinite-length sequences.
-For branch length estimation on a correct topology, there are similar cases where the branch lengths are consistently overestimated in the case of infinite-length sequences.
-Thus although joint maximum likelihood may turn out to be a useful tool in practice, care should be taken when using this approach.
+We derive a set of parameters that generate data under a four-taxon tree and result in this method estimating an incorrect topology in the limit of infinite-length sequences.
+For branch length estimation on the correct topology, we outline similar cases where branch lengths are consistently and heavily biased.
 \end{abstract}

### 2018-01-29 joint_inf.tex
main text

@@
 \begin{abstract}
-Maximum likelihood estimation of phylogeny requires a means of handling unknown ancestral states.
+Maximum likelihood estimation in phylogenetics requires a means of handling unknown ancestral states.
 Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
@@
 
-Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out possible ancestral states at the internal nodes of the tree.
-Recently, \cite{Sagulenko2017-jo} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences on a fixed topology.
-This is attractive from a computational perspective: such joint ML inference can proceed according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
-This latter conditional optimization is simpler and more computationally efficient than marginalizing over ancestral sequences and then performing optimization.
+Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out latent ancestral states at the internal nodes of the tree.
+In a recent proposal, \cite{Sagulenko2017-jo} suggest using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences on a fixed topology.
+This is attractive from a computational perspective: such joint inference can proceed according to an iterative procedure in which ancestral sequences are first estimated and model parameters are optimized conditional on these estimates.
+This latter conditional optimization is simpler and more computationally efficient than optimizing the marginal likelihood.
 But is it statistically consistent?
 
-Statistical consistency means that an estimator gives the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
+An estimator is statistically consistent if it converges to the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
 These proofs do not readily extend to include estimating ancestral states.
-Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
+Moreover, examples of inconsistency arising from problems where the number of parameters increases with the amount of data \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
 In this case those additional parameters come in the form of the states of ancestral sequences.
-Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.
+Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case where ``branch lengths are short and only a minority of sites change on a give branch,'' their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.
 
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model with data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
+To do so, we use a binary symmetric model with data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint objective function to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
 We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.
@@
 We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``Farris'' and ``Felsenstein'' trees (Fig.~\ref{fig:farris-fels-top}).
-In the classical configuration of each of these trees, the interior branch length parameters are equal to the bottom two parameters.
-We show in the Appendix that, in the case of the Farris tree, doing inference fixing both the top two branch parameters to be equal and the bottom two branch parameters to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch parameters.
+In the standard configuration of each of these trees, the interior branch length parameters are equal to the bottom two parameters.
+We show in the Appendix that, in the case of the Farris tree, performing inference fixing both the top two branch parameters to be equal and the bottom two branch parameters to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch parameters.
 
@@
 The probability of a substitution on a branch with fidelity $\theta$ is $(1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$ where $0 \le \theta \le 1$.
-This parameter thus quantifies the fidelity of transmission of the ancestral state across an edge \cite{Matsen2007-jq}.
+This parameter quantifies the fidelity of transmission of the ancestral state across an edge \cite{Matsen2007-jq}.
 
@@
 \label{eq:marginal_likelihood}
-\tilde{L}_\nCols(\tau, t; \fullAlignment) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates)
+\tilde{L}_\nCols(\tau, t; \fullAlignment) = \prod_{i=1}^{\nCols} \ \sum_{\ancestralStateColumn_i\in\alphabet^{\nAncestralStateRows}} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)
 \end{equation}
@@
 
-An alternative approach \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
+The alternative approach \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
 This is known in statistics as a profile likelihood \cite{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
@@
 \label{eq:profile_likelihood}
-L_\nCols'(\tau, t; \fullAlignment) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
+L_\nCols'(\tau, t; \fullAlignment) = \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i\in\alphabet^{\nAncestralStateRows}} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
 \end{equation}
-We will use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}.
-One can then estimate a topology and branch fidelities using this profile likelihood:
+We use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}, and estimate a topology and branch fidelities using this profile likelihood as
 \begin{equation}
@@
 In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
-For this reason, in practice, the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.
+For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.
 
@@
 
-We can now state our results on inconsistency of joint inference.
+We now state our results on the inconsistency of joint inference.
 All proofs are deferred to the Appendix.
@@
 We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology than for the true, generating Farris topology.
-To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the Farris topology as a function of $x^*$ and $y^*$, and, similarly, we construct a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
+To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the Farris topology as a function of $x^*$ and $y^*$ and, similarly, a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
 When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the Farris case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
@@
 The proof of this theorem is by a detailed examination of inequalities.
-The only intuition we can offer is that $\tau_2$ is favored in performing joint maximum likelihood since the likelihood for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its likelihood while $\tau_2$ has many more.
+Intuitively, $\tau_2$ is favored in performing joint inference since the objective function for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its objective function while $\tau_2$ has many more.
 This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.
@@
     An analytically-derived region of topological inconsistency in terms of fidelities for ``perfect'' data generated on the Farris topology (Fig.~\ref{fig:farris-fels-top}) with $w^* = y^*$.
-    Due to the looseness of the upper and lower bounds, the parameters in the non-shaded region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
+    Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
 }
@@
 
-We now consider the problem of branch length estimation on the correct tree using joint ancestral state and branch length estimation.
+We now consider the problem of branch length estimation on the correct tree using joint estimation.
 As described above, we use the equivalent but different notion of branch fidelities.
@@
 In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
-That is, if we know $x^*$ and $y^*$, then for
+That is, if we know $x$ and $y$ and estimate them to be the truth, then, for
 $$
@@
 $$
-we show there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
+there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
 In the general case, we do the same but with
 $$
-(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\}).
+(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\})
 $$
-Again, we find a region for which the inferred values do not converge to the generating values.
+to find a region where the inferred values do not converge to the generating values.
 These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \cite{RoyChoudhury2015-ta}.
@@
 
-Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and proceed to estimate the internal branch parameter $w$.
+Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and estimate the internal branch parameter $w$.
 \begin{restatable}{theorem}{restrictedBranchInconsist}
@@
 \end{restatable}
-This theorem allows us to demarcate a region of bad internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
-Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch, and then a branch length is estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
+This theorem allows us to demarcate a region of biased internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
+Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch with this branch length estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
 As an intuition for the theoretical development, seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.
+%das: @EM Joe pointed out that this is in effect another instance of converging to the wrong topology since once the branch length goes to zero we are no longer in a bifurcating tree case.
+I don't recall if he had more intuition than the above sentence but he seemed to think it made sense.
+Not sure if you have any thoughts.
 
@@
 \caption{Analytically-derived region of branch parameter inconsistency in terms of fidelities $x=x^*$ and $y=y^*$ that are fixed to their correct values with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
-The shaded region shows the area in which the internal branch length is estimated to be 0, i.e.\ the estimated fidelity $\hat{w} = 1$, even though the generating fidelity is set to be that of $y^*$.
-Here again, the shaded region is guaranteed to give inconsistent estimation, while the non-shaded region may or may not do so.}
+The shaded region shows the area in which the internal branch length is estimated to be 0, i.e.\ the estimated fidelity $\hat{w} = 1$, even though the generating fidelity is $y^*$.
+Here again, the shaded region is guaranteed to give inconsistent estimation, while the white region may or may not do so.}
 \label{fig:bl-inconsistency}
@@
 
-The general case is more challenging to analyze and so we will get somewhat weaker bounds.
-In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
+The general case is more challenging to analyze and so we obtain weaker bounds.
+Here, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
 Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
-Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
-We show that there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
-We fix bounds on these estimates and show that, in the general case, joint estimation either estimates $\hat{w}$ to be 1 or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
+Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$.
+We show there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
+We fix bounds on these estimates and show that, in the general case, joint estimation either estimates $\hat{w}$ to be one or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
 \begin{restatable}{theorem}{generalBranchInconsist}
@@
 \end{restatable}
-This theorem can be used to show incorrect branch parameter estimates as follows.
-If we don't want to allow any error in pendant branches, we can use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section (Fig.~\ref{fig:bl-inconsistency}).
+We use this theorem to show incorrect branch parameter estimates as follows.
+If we do not tolerate any error in pendant branches, we use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section (Fig.~\ref{fig:bl-inconsistency}).
 For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
@@
 $$
-Then useful $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
+Then $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
 As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
-In this case, joint inference will either estimate $\hat{w}$ to have fidelity 1 or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.
+In this case, joint inference will either estimate $\hat{w}$ to have fidelity one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.
 
@@
     In the marked region, either $\hat{w}=1$ or one of $x^*-0.1 \le \hat{x} \le x^*+0.1$ or $y^*-0.1 \le \hat{y} \le y^*+0.1$ will not be true, resulting in poor estimation of the pendant branch parameters.
-Here again, the shaded region is guaranteed to give such poor estimation, while the non-shaded region may or may not do so.
+    As in previous plots, the shaded and white regions are loose indications of inconsistency.
 }
@@
 Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:bl-loose-inconsistency}).
-To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint maximum likelihood estimation using numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}).
-The region in which $\hat{w}$ is numerically estimated to be 1 is an unusual shape containing the analytically-derived region but is not the same as it.
+To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
 We see that the region of inconsistency in the general case is similar to that of the restricted case (compare Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency}).
-This region encompasses the majority of the branch fidelity space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch fidelity to be 1.
+This region encompasses the majority of the branch fidelity space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch fidelity to be one.
 
-We provide a full description of our optimization procedure in the Appendix, but briefly, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher likelihood.
+We provide a full description of our optimization procedure in the Appendix, but briefly, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher objective function.
 We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$.
@@
 Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same procedure.
-For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look reasonable, with $\hat{w}$ increasing as $y^*$ increases, though Fig.~\ref{fig:bl-general-bias} shows there is a systematic positive bias in this procedure even when $\hat{w}$ is not estimated to be 1.
-% Here we obtain the correct $\hat{w}\rightarrow y^*$.
+For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look reasonable, with $\hat{w}$ increasing as $y^*$ increases, though Fig.~\ref{fig:bl-general-bias} shows there is a systematic positive bias in this procedure even when $\hat{w}$ is not estimated to be one.
 To understand the quality of each fit, we report the range of $\hat{w}-y^*$ where $\hat{w}\neq 1$.
@@
 
-We have shown that maximum-likelihood joint inference when estimating phylogenetic parameters \cite{Sagulenko2017-jo} is not consistent in general.
-Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference infers the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
+We have shown that jointly inferring ancestral states and phylogenetic parameters \cite{Sagulenko2017-jo} is not consistent in general.
+Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference converges on the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
 In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.
 
-Joint maximum-likelihood inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \cite{Murphy2000-ry}.
+Joint inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \cite{Murphy2000-ry}.
 Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., the method of sieves \cite{Geman1982}.
@@
 Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \cite{LeCam1990}.
-Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are in fact quite recent \cite{RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
-We have furthered progress in understanding the limitations of this joint maximum likelihood approach.
+Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \cite{RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
+We have furthered progress in understanding the limitations of this joint optimization procedure.
 
@@
 Given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \cite{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
-
-As a final reminder, we note that all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.
+Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.
 
@@
 
-This work supported by National Institutes of Health grants R01-AI12096, U19-AI117891, and U54-GM111274 as well as National Science Foundation grants CISE-1561334 and CISE-1564137.
+This work was supported by National Institutes of Health grants R01-AI12096, U19-AI117891, and U54-GM111274 as well as National Science Foundation grants CISE-1561334 and CISE-1564137.
 The research of Frederick Matsen was supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation.

### 2018-01-29 appendix.tex
appendix

@@
 \subsection*{Site split formulation}
-We begin by introducing ``site splits,'' that simply formalize the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
-This is a standard step in the description of the Hadamard transform (Section 8.6 of \cite{Semple2003-em}), although we have to be a little careful here due to the extra bookkeeping of ancestral states.
+We begin by introducing ``site splits,'' formalizing the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
+This is a standard step in the description of the Hadamard transform (Section 8.6 of \cite{Semple2003-em}), although our approach is complicated slightly by the inclusion of ancestral states.
 
@@
 
-For topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization of $\ancestralStateColumn_i$.
-In this case the entire set of internal nodes must be enumerated: the probability of observing an ancestral state split, conditional on a site split, is not invariant to taking its complement.
+For topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
+In this case the entire set of internal nodes must be enumerated: the probability of observing an ancestral state split conditional on a site split is not invariant to taking its complement.
 
@@
 
-We first fix notation for maps that are defined just below.
+We first fix notation.
 \begin{definition}
@@
 The mapping $\patternToSplit$ takes the complement of site patterns to obtain a site split in $\mathcal{P}(\siteSplitSet)$.
-The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is defined to be the complement of $\ancestralStateColumn$.
+The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.
 
@@
 
-Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
+Let $\xi_j$ be such a choice for each $1 \leq j \leq q$.
 Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
@@
 $$
-so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
+so that \eqref{eq:site_pattern_profile_likelihood_mean} is
 \begin{equation}
@@
 
-We now state the Hadamard representation of site split generating probabilities, following Section 8.6 of \cite{Semple2003-em}.
-For each edge $e$ define the edge \emph{fidelity} for that edge as
+We state the Hadamard representation of site split generating probabilities, following Section 8.6 of \cite{Semple2003-em}.
+For each edge $e$ define the edge ``fidelity'' for that edge as
 $$
@@
 for any site split $\siteSplit_j$.
-Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies for our two topologies (Fig.~\ref{fig:farris-fels-top}).
+Table~\ref{tab:sitepatprob} contains calculations of site pattern probabilities for our two topologies (Fig.~\ref{fig:farris-fels-top}).
 
@@
 We follow with an expository example computing these probabilities and likelihoods.
-Consider the fixed, binary four-tip tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}(a)---this is commonly known as the ``Farris zone'' topology.
+Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a---this is commonly known as the ``Farris zone'' topology.
 The set of all possible character assignments is
@@
 Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
-As described above we cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
+We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
 $$
@@
 To compute the likelihood of observing a set of data, we need $\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
-We remind the reader that the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
+Using branch fidelities, the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
 See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the Farris tree.
@@
 Consider the Farris tree with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
-We now show that, in the case of the Farris tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch lengths to be equal and the bottom two branch lengths to be equal during inference will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
-Using the Hadamard transform approach, we calculate the generating probabilities on the Farris tree.
+We now show that, in the case of the Farris tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch parameters to be equal and the bottom two branch parameters to be equal during inference obtains the same maximum likelihood estimate as in the case of arbitrary branch parameters.
+Using the Hadamard transform, we calculate the generating probabilities on the Farris tree.
 For site split $\emptyset$,
@@
 It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
-Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments holding for the remaining parameters.
+Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
 We decompose the likelihood into the entropy and partial likelihood terms as in \eqref{eq:log_likelihood_simplified}.
@@
 Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
-The Felsenstein tree does not admit this property of symmetry of function inputs, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
+The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
 
@@
 Assume $\tau^*=\tau_1$.
-The full Farris tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see the upper table of Table~\ref{tab:likelihoods}).
-
-We begin by writing out the case for the ancestral state split $\{1\}$, and show the other two cases follow a similar argument.
+The Farris tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see the upper table of Table~\ref{tab:likelihoods}).
+
+We write out the case for the ancestral state split $\{1\}$, and show the other two cases follow a similar argument.
 Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_profile_likelihood_mean}, the log-likelihood is, suppressing normalizing constants $-\log 8$ (from the first additive term of \eqref{eq:site_pattern_profile_likelihood_mean}) and $-\log 32$ (from the second),
@@
 \end{align*}
-Factoring and making these substitutions (again, under the assumption that the ancestral state split is $\{1\}$) results in
+Factoring and making these substitutions results in
 \begin{align*}
@@
 We first bound the entire likelihood below and then proceed with joint inference.
-First, we obtain a lower bound for the likelihood by replacing $1+w$ with $1-w$ in Table~\ref{tab:likelihoods} for $\tau = \tau_2$; in this case, we resolve many of the ambiguous likelihood terms.
+To obtain a lower bound for the likelihood we replace $1+w$ with $1-w$ in Table~\ref{tab:likelihoods} for $\tau = \tau_2$; in this case, we resolve many of the ambiguous likelihood terms.
 For example, for site split $\{1,2\}$, after we replace the $1+w$ term for $\ancestralSplitPartition_j(\tau, t)=\{\emptyset, \{1,2\}\}$ with $1-w$, we have
@@
 $$
-(1+x)^2 (1-w)(1+y)^2 > (1+x)(1-x)(1-w)(1+y)(1-y)
-$$
-always, the maximum for this site split can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
+(1+x)^2 (1-w)(1+y)^2 > (1+x)(1-x)(1-w)(1+y)(1-y),
+$$
+the maximum for this site split can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
 For the other site splits we consider two cases.
@@
 $$
-with equality only when $w=0$, the second likelihood is the likelihood with the maximum-likelihood ancestral state, and so we need only analyze that case.
+with equality only when $w=0$, the second likelihood is the likelihood with the maximum-likelihood ancestral state, and so we only analyze that case.
 
@@
 c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
-Excepting all the cases where $x^*=0$ or where $y^*=0$, this implies $\ell'$ has two zeros according to the quadratic formula and $\alpha_1 > \alpha_2$.
-Also because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w=0$.
+Excepting all the cases where $x^*=0$ or where $y^*=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
+Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w=0$.
 This implies that if the smaller of the zeros of $f(w)$ is greater than one, then $\hat{w} \equiv 1$.
@@
 We initialize our optimization procedure at the true branch parameters $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
-Because our analysis has shown that that $\hat{w}$ can be equal to 1 for when $x^*$ and $y^*$ are small, and local optimization may have a hard time obtaining this value if our initial guess is small, we perform two maximizations, one with $w=1$ fixed and one where $w$ is estimated.
-We take the value of $\hat{w}$ with the larger likelihood as our estimate.
+Because our analysis shows that $\hat{w}$ can be equal to one when $x^*$ and $y^*$ are small, and local optimization may have a hard time obtaining this value if our initial guess is small, we perform two maximizations, one with $w=1$ fixed and one where $w$ is estimated.
+We take the value of $\hat{w}$ with the larger objective function as our estimate.
 

### 2018-01-30 appendix.tex
Pass through edits.

@@
 \subsection*{Site split formulation}
-We begin by introducing ``site splits,'' formalizing the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
+We begin by introducing ``site splits,'' which formalize the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
 This is a standard step in the description of the Hadamard transform (Section 8.6 of \cite{Semple2003-em}), although our approach is complicated slightly by the inclusion of ancestral states.
@@
 \end{align*}
-Factoring and making these substitutions results in
+%EM You dropped the below parenthetical reminder. Did you think that it was clear?
+Factoring and making these substitutions (again, under the assumption that the ancestral state split is $\{1\}$) results in
 \begin{align*}

### 2018-01-30 joint_inf.tex
Pass through edits.

@@
 Recently, a computationally-efficient approach was proposed to jointly maximize over these unknown states and phylogenetic parameters.
+Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
 In this paper, we show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
-We derive a set of parameters that generate data under a four-taxon tree and result in this method estimating an incorrect topology in the limit of infinite-length sequences.
-For branch length estimation on the correct topology, we outline similar cases where branch lengths are consistently and heavily biased.
+We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates an incorrect topology in the limit of infinite-length sequences.
+For branch length estimation on the correct topology, we outline similar cases where branch length estimates are consistently and heavily biased.
 \end{abstract}
@@
 Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out latent ancestral states at the internal nodes of the tree.
-In a recent proposal, \cite{Sagulenko2017-jo} suggest using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences on a fixed topology.
+In a recent paper, \cite{Sagulenko2017-jo} suggest using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences on a fixed topology.
 This is attractive from a computational perspective: such joint inference can proceed according to an iterative procedure in which ancestral sequences are first estimated and model parameters are optimized conditional on these estimates.
@@
 
-An estimator is statistically consistent if it converges to the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
+An estimator is said to be statistically consistent if it converges to the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
 These proofs do not readily extend to include estimating ancestral states.
@@
 In this case those additional parameters come in the form of the states of ancestral sequences.
+%EM I don't see this sentence in the most recent version of the paper via search. Can you point me to a section?
 Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case where ``branch lengths are short and only a minority of sites change on a give branch,'' their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.
@@
 In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
-That is, if we know $x$ and $y$ and estimate them to be the truth, then, for
+That is, if we estimate $x$ and $y$ correctly, then, for
 $$
@@
 As an intuition for the theoretical development, seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.
-%das: @EM Joe pointed out that this is in effect another instance of converging to the wrong topology since once the branch length goes to zero we are no longer in a bifurcating tree case.
-I don't recall if he had more intuition than the above sentence but he seemed to think it made sense.
-Not sure if you have any thoughts.
+If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.
 
@@
 \section*{Acknowledgements}
-We thank Richard Neher and Vladimir Minin for helpful discussions.
+We thank Richard Neher, Vladimir Minin and Joe Felsenstein for helpful discussions.
 

### 2018-01-30 joint_inf.tex
comments

@@
 %EM I don't see this sentence in the most recent version of the paper via search. Can you point me to a section?
-Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case where ``branch lengths are short and only a minority of sites change on a give branch,'' their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.
+%das: changed give to given in case you were ctrl+f-ing, but it's here https://academic.oup.com/ve/article/4/1/vex042/4794731
+%page 2 section 2.1 at the top of the second column first paragraph second sentence
+Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.
 

### 2018-02-03 cover-letter/supervisor.tex
cover letter

\documentclass[letterpaper,10pt]{article}

\usepackage{cite}
\usepackage{enumerate}
\usepackage{url}

\hyphenation{Ge-nome Ge-nomes through-put re-comb-in-ation}

\input{FHletterdefs}

\renewcommand\FHtoname{}
\renewcommand\FHtoaddress{}

\renewcommand\FHopening{Dear MBE Editorial Board,}

\begin{document}

Please consider our manuscript entitled ``Joint estimation of phylogenies and ancestral states is not consistent'' for publication in the \textit{Molecular Biology and Evolution}.
In this paper, my co-author David Shaw and I investigate the statistical properties of joint maximum-likelihood estimation of trees and ancestral sequences.
We were motivated to do so by recent work from Richard Neher's group (\url{http://dx.doi.org/10.1093/ve/vex042}) that employs such joint optimization as an efficient alternative to the standard Felsenstein marginalization.

We show that such joint inference is not consistent, both for topology inference and branch length optimization.
Indeed, there are sizeable regions of parameter space in which this joint inference gives the wrong topology and branch lengths in the limit of large data.
This is not a case of long branch attraction.
Indeed, the inconsistency in Felsenstein 1978 requires a combination of long and short branches, whereas our inconsistency comes from all branches being long.

Although the proofs in our paper require detailed examination of likelihood functions, we have written the paper to be broadly accessible.
The paper has five illustrations showing various regions of inconsistency, and provides intuitive summaries of the results in the main text.
A complete set of proofs appears in the appendix.

We feel that it is important for this paper to get a critical review by a mathematically-oriented phylogenetics expert.
We suggest:
\begin{itemize}
\item Elizabeth Allman \texttt{esallman@alaska.edu}
\item Laura Kubatko \texttt{lkubatko@stat.osu.edu}
\item Mike Steel \texttt{mike.steel@canterbury.ac.nz} (although Mike is somewhat conflicted as my postdoc advisor, he would do an excellent job reviewing this paper)
\item Ed Susko \texttt{susko@mathstat.dal.ca}
\end{itemize}

Thank you for your consideration.
\end{document}

### 2018-02-05 cover-letter/supervisor.tex
minor additions

@@
 
-Please consider our manuscript entitled ``Joint estimation of phylogenies and ancestral states is not consistent'' for publication in the \textit{Molecular Biology and Evolution}.
+Please consider our manuscript entitled ``Joint maximum-likelihood of phylogenies and ancestral states is not consistent'' for publication in \textit{Molecular Biology and Evolution}.
 In this paper, my co-author David Shaw and I investigate the statistical properties of joint maximum-likelihood estimation of trees and ancestral sequences.

### 2018-02-05 joint_inf.tex
:+1: for paper reference
Lineno
Proposed title change!
minor additions

@@
 \usepackage{thm-restate}
+\usepackage{lineno}
+\linenumbers
 
@@
 
-\title{Joint estimation of phylogenies and ancestral states is not consistent}
+\title{Joint maximum-likelihood on phylogenies and ancestral states is not consistent}
 \author{David A. Shaw \and Frederick A. Matsen IV}
@@
 In this case those additional parameters come in the form of the states of ancestral sequences.
-%EM I don't see this sentence in the most recent version of the paper via search. Can you point me to a section?
-%das: changed give to given in case you were ctrl+f-ing, but it's here https://academic.oup.com/ve/article/4/1/vex042/4794731
 %page 2 section 2.1 at the top of the second column first paragraph second sentence
@@
 
-The alternative approach \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
+The alternative approach \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}, though only branch lengths are estimated as the topology is fixed in advance.
 This is known in statistics as a profile likelihood \cite{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
@@
 In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.
+This inconsistency in the case of long branches concurs with some empirical findings in \cite{Sagulenko2017-jo}, namely their Figures 2 and 3.
 

### 2018-02-05 joint_inf.tex
Moving the fixed topology up.

--- 2018-01-30 joint_inf.tex
@@
 In this case those additional parameters come in the form of the states of ancestral sequences.
-%EM I don't see this sentence in the most recent version of the paper via search. Can you point me to a section?
-Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case where ``branch lengths are short and only a minority of sites change on a give branch,'' their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.
+%page 2 section 2.1 at the top of the second column first paragraph second sentence
+Although the software described in \cite{Sagulenko2017-jo} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
+In particular, one would like to know when this approximate technique breaks down for both topology and branch length inference, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.
 
@@
 In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.
+This inconsistency in the case of long branches concurs with some empirical findings in \cite{Sagulenko2017-jo}, namely their Figures~2 and 3.
 

### 2018-02-06 appendix.tex
comments
:+1: for paper reference
references, spacing, cover page

@@
 We begin by introducing ``site splits,'' which formalize the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
-This is a standard step in the description of the Hadamard transform (Section 8.6 of \cite{Semple2003-em}), although our approach is complicated slightly by the inclusion of ancestral states.
+This is a standard step in the description of the Hadamard transform (Section 8.6 of \citet{Semple2003-em}), although our approach is complicated slightly by the inclusion of ancestral states.
 
@@
 
-We state the Hadamard representation of site split generating probabilities, following Section 8.6 of \cite{Semple2003-em}.
+We state the Hadamard representation of site split generating probabilities, following Section 8.6 of \citet{Semple2003-em}.
 For each edge $e$ define the edge ``fidelity'' for that edge as
@@
 \end{align*}
-%EM You dropped the below parenthetical reminder. Did you think that it was clear?
 Factoring and making these substitutions (again, under the assumption that the ancestral state split is $\{1\}$) results in
@@
 
-We use basin-hopping \cite{Wales1997} implemented in \texttt{scipy} \cite{Jones2001} to obtain the plots in Figs.~\ref{fig:bl-general-inconsistency}, \ref{fig:bl-general-marginal}, and \ref{fig:bl-general-bias}.
+We use basin-hopping \citep{Wales1997} implemented in \texttt{scipy} \citep{Jones2001} to obtain the plots in Figs.~\ref{fig:bl-general-inconsistency}, \ref{fig:bl-general-marginal}, and \ref{fig:bl-general-bias}.
 This method randomly perturbs candidate solutions and accepts or rejects them based on the nearby likelihood surface, eventually obtaining an estimate of the ``global'' optimum.
@@
 \caption{
-    Estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing the classical marginal likelihood \eqref{eq:marginal_likelihood} (rather than a joint optimization procedure).
+    Estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing the classical marginal likelihood \eqref{eq:marginal_likelihood} (rather than a joint optimization procedure).
 }
@@
 \caption{
-    Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:profile_likelihood}.
+    Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing \eqref{eq:profile_likelihood}.
     Plot focuses on $0.6 < x^*, y^* < 1$ where $\hat{w}$ is estimated to be a value different from 1.

### 2018-02-06 joint_inf.tex
references, spacing, cover page

@@
 \linenumbers
+\usepackage{natbib}
+\usepackage{setspace}
+\doublespacing
+\usepackage{authblk}
+\usepackage{fancyhdr}
 
@@
 
-\title{Joint maximum-likelihood on phylogenies and ancestral states is not consistent}
-\author{David A. Shaw \and Frederick A. Matsen IV}
+\fancyhead[R]{ submission intended as Article in Discoveries section of \emph{MBE} }
+
+\title{Joint maximum-likelihood of phylogenies and ancestral states is not consistent}
+
+\author[1]{David A. Shaw}
+\author[1]{Frederick A. Matsen IV\thanks{Corresponding author, email: \url{matsen@fredhutch.org}}}
+\affil[1]{Computational Biology Program, Fred Hutchinson Cancer Research Center\\ Seattle, WA, USA}
+\date{}
 
 \begin{document}
-\maketitle
 
 \renewcommand{\arraystretch}{1.2} % because otherwise exponents get eaten by \hline
+
+\maketitle
+\thispagestyle{fancy}
 
@@
 
+\newpage
+
 \section*{Introduction}
@@
 Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out latent ancestral states at the internal nodes of the tree.
-In a recent paper, \cite{Sagulenko2017-jo} suggest using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences on a fixed topology.
+In a recent paper, \citet{Sagulenko2017-jo} suggest using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences on a fixed topology.
 This is attractive from a computational perspective: such joint inference can proceed according to an iterative procedure in which ancestral sequences are first estimated and model parameters are optimized conditional on these estimates.
@@
 
-An estimator is said to be statistically consistent if it converges to the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
+An estimator is said to be statistically consistent if it converges to the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
 These proofs do not readily extend to include estimating ancestral states.
-Moreover, examples of inconsistency arising from problems where the number of parameters increases with the amount of data \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
+Moreover, examples of inconsistency arising from problems where the number of parameters increases with the amount of data \citep{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
 In this case those additional parameters come in the form of the states of ancestral sequences.
 %page 2 section 2.1 at the top of the second column first paragraph second sentence
-Although the software described in \cite{Sagulenko2017-jo} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
+Although the software described in \citet{Sagulenko2017-jo} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
 In particular, one would like to know when this approximate technique breaks down for both topology and branch length inference, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.
@@
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model with data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint objective function to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
+To do so, we use a binary symmetric model with data being generated on the four-taxon ``Farris zone'' \citep{Siddall1998-hq} tree, and we construct bounds on the joint objective function to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
 We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.
@@
 
-Assume the binary symmetric model, namely with a character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution \cite{Semple2003-em}.
+Assume the binary symmetric model, namely with a character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution \citep{Semple2003-em}.
 Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
@@
 The probability of a substitution on a branch with fidelity $\theta$ is $(1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$ where $0 \le \theta \le 1$.
-This parameter quantifies the fidelity of transmission of the ancestral state across an edge \cite{Matsen2007-jq}.
+This parameter quantifies the fidelity of transmission of the ancestral state across an edge \citep{Matsen2007-jq}.
 
@@
 \end{equation}
-The probability $\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)$ is a product of transition probabilities determined by $\fullAlignment$, $\fullAncestralStates$, $\tau$, and $t$ \cite{Felsenstein2004}.
+The probability $\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)$ is a product of transition probabilities determined by $\fullAlignment$, $\fullAncestralStates$, $\tau$, and $t$ \citep{Felsenstein2004}.
 
@@
 
-The alternative approach \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
-This is known in statistics as a profile likelihood \cite{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
+The alternative approach \citep{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
+This is known in statistics as a profile likelihood \citep{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
 \begin{equation}
@@
 In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
-For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.
+For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2017-jo}.
 
@@
 to find a region where the inferred values do not converge to the generating values.
-These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \cite{RoyChoudhury2015-ta}.
+These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \citep{RoyChoudhury2015-ta}.
 
@@
 Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:bl-loose-inconsistency}).
-To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
+To determine how conservative, we use the method of basin-hopping \citep{Wales1997} to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
 We see that the region of inconsistency in the general case is similar to that of the restricted case (compare Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency}).
@@
 \caption{
-    Numerical estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:profile_likelihood}.
+    Numerical estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing \eqref{eq:profile_likelihood}.
     Data generated as in Fig.~\ref{fig:inconsistency-farris}.
@@
 
-We have shown that jointly inferring ancestral states and phylogenetic parameters \cite{Sagulenko2017-jo} is not consistent in general.
+We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2017-jo} is not consistent in general.
 Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference converges on the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
 In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.
-This inconsistency in the case of long branches concurs with some empirical findings in \cite{Sagulenko2017-jo}, namely their Figures~2 and 3.
+This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2017-jo}, namely their Figures~2 and 3.
 
-Joint inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \cite{Murphy2000-ry}.
-Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., the method of sieves \cite{Geman1982}.
+Joint inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \citep{Murphy2000-ry}.
+Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., the method of sieves \citep{Geman1982}.
 A potential solution in this case using the method of sieves could be to project the column-wise ancestral state patterns into a lower-dimensional space, allowing the degrees of freedom in the ancestral state columns to grow with $\nCols$, albeit more slowly than $O(\nCols)$.
-Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \cite{LeCam1990}.
-Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \cite{RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
+Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \citep{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \citep{LeCam1990}.
+Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \citep{RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
 We have furthered progress in understanding the limitations of this joint optimization procedure.
 
-Previous work in phylogenetics has developed consistency counterexamples using the same four-taxon topologies used here \cite{Felsenstein1978-rr}.
+Previous work in phylogenetics has developed consistency counterexamples using the same four-taxon topologies used here \citep{Felsenstein1978-rr}.
 In this previous work, when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
 This is the converse of what we have shown for joint inference, where the Felsenstein topology is more likely than the Farris topology.
-Moreover, the inconsistency in \cite{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
+Moreover, the inconsistency in \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
 This is not the case here; for our case, the inconsistency generally occurs when all branches are long, and has more to do with the choices of the form of the likelihood than from the interplay of long and short branches.
-Difficulties in phylogenetic estimation when generating data on the ``Farris zone'' tree have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.
+Difficulties in phylogenetic estimation when generating data on the ``Farris zone'' tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.
 
@@
 All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
-Given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \cite{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
+Given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
 Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.
@@
 \section*{Acknowledgements}
-We thank Richard Neher, Vladimir Minin and Joe Felsenstein for helpful discussions.
+We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.
 
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{plainnat}
 \bibliography{joint_inf}

### 2018-02-06 joint_inf.tex
final edit

@@
 This is the converse of what we have shown for joint inference, where the Felsenstein topology is more likely than the Farris topology.
-Moreover, the inconsistency in \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
+Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
 This is not the case here; for our case, the inconsistency generally occurs when all branches are long, and has more to do with the choices of the form of the likelihood than from the interplay of long and short branches.

### 2018-03-02 joint_inf.tex
MBE submission requirements
Adding B cell grant.

@@
-\documentclass{article}
+\documentclass[11pt]{article}
 
@@
 \usepackage{setspace}
-\doublespacing
+\onehalfspacing
 \usepackage{authblk}
@@
 
-This work was supported by National Institutes of Health grants R01-AI12096, U19-AI117891, and U54-GM111274 as well as National Science Foundation grants CISE-1561334 and CISE-1564137.
+This work was supported by National Institutes of Health grants R01-GM113246, R01-AI12096, U19-AI117891, and U54-GM111274 as well as National Science Foundation grants CISE-1561334 and CISE-1564137.
 The research of Frederick Matsen was supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation.

### 2018-04-16 versions/response.v1.txt
Adding v1 and draft response.

> Associate Editor
> Editors’ comments to the author:
> All four reviewers agreed that the work is worth publishing and
> provides an important insight into the behavior of the joint
> ML/ancestral states estimation method. However, all reviewers asked
> for what amounts to major revisions to the text and presentation. You
> should especially focus on clarity and accessibility, provide the
> requested examples, and address suggestions on how the results could
> be even further improved.

> Finally, please pay VERY careful attention to improving your
> literature review (reviewer 4) and making a very clear case about how
> your work IMPROVES and EXTENDS previous results. And, please, do get
> rid of the Farris zone. The name "Farris" should not be attached to
> results that have to do with ML estimation, given the long and
> torturous efforts of cladists to undermine the entire line of inquiry.

Felsenstein's book has a subsection entitled "The Farris zone" spanning two pages (pp 120-121), a figure entitled "A tree in the Farris zone.", and a table with a caption "Properties of the Felsenstein and Farris zones..."
The term "Farris zone" has almost 7 times as many citations as the term "inverse Felsenstein zone" on Google Scholar.
Thus, although we realize that the term causes offense, it is standard.

> Reviewer: 1

> Comments to the Author
> The authors show that when ancestral states are considered as
> parameters, there are zones where topologies and edge-lengths are
> inconsistently estimated. As is evidenced by any ML package, it is
> widely accepted that ancestral states should not be parameters; the
> large number of parameters suggests the approach will be problematic.
> Still, to the best of my knowledge, it has not been shown that
> treating ancestral states as parameters actually leads to poor
> performance, or how poor. Moreover, the authors point to Sagulenko et
> al (2018) - now published in Virus Evolution - as an example
> implementation. I found the paper interesting and I believe a more
> succinct presentation will be of interest.

> A main concern with the paper is that the shaded regions of
> topological inconsistency or poor estimation in Figs 2-4, are only
> subsets of the actual regions where topological inconsistency or poor
> estimation occur; and yet determining the actual regions seems
> feasible. The contrast between Fig 5 and 4 illustrates how far off the
> subsets are in giving the actual regions where estimation is
> problematic. The white region in Fig 5 is the actual region where hat
> w=1 whereas the shaded region of Fig 3 is the subset of the actual
> region that the authors were able to determine through analytical
> arguments. For most readers, out of Fig 3 and Fig 5, only Fig 5 is of
> interest.

> Getting exact regions is what I think the authors should do. The
> approach used to obtain Fig 5 is generally applicable. Explicit,
> relatively simple, expressions are given in the appendix for long-run
> lnls, l_{tau_1,t*}(tau_2,w,x,y). For any given (x*,y*) on a fine grid
> in Figs 2-4, l_{tau_1,t*}(tau_j,w,x,y), can be maximized using
> numerical algorithms over the three parameters w,x,y. Doing so for
> tau_j=tau_1, tau_2 would determine the actual regions of problematic
> estimation corresponding to Figs 2-4. The only potential difficulty
> that I can see is that numerical algorithms might not converge. This
> can either be avoided by considering a cubic grid of starting (w,x,y),
> or through something like the following. For any given set of
> ancestral character state assignments for the distinct set of 8
> patterns 1111, 1110,... (0000 can treated as the same as 1111), the
> long-run lnl is of the form

> l(x,y,w)=a log(1-x) + b log(1+x) + c log(1-y) + d log(1+y) + e log(1-w) + f log(1+w)

> where a,..f>=0 are constants determined by (x*,y*) and what ancestral
> states have been assigned. The function l(x,y,w) has an explicit
> simple maximizer. Thus the overall maximizer is determined as the
> largest maximized l(x,y,w) over all 4^8 possible ancestral state
> assignments: a finite number of comparisons. The actual number of
> possible ancestral state assignments is quite a bit smaller than 4^8
> (for instance 11 is always the ancestral reconstruction for 1111) and
> the authors have a lot of familiarity with what the ancestral
> reconstructions are. I believe Fig 2 can be replaced with an exact (up
> to grid approximation) region of topological inconsistency and Figs
> 3-5 with heat maps of limiting (x,y,w) for tau_1.

Theory leads to generalization.

> The other main concern that I have is with the restriction that
> t=(x,y,x,y,w) in estimation rather than t=(t_1,...,t_5). On the one
> hand, usually one expects that incorporating knowledge about the true
> underlying tree (t_1=t_3 & t_2=t_4 in the present case) should be
> helpful for a good estimation procedure. Thus the fact that treating
> ancestral states as parameters gives inconsistent topological
> estimation in this setting demonstrates that the approach is
> problematic. On the other hand, allowing t=(t_1,...,t_5) is what would
> be done in any practical implementation and it seems quite possible to
> me that results would be radically different in this case. For
> instance, from Fig 2, the generating tree with all equal edge-lengths
> (y*=x*=0.2) provides a case where topology tau_2 is inconsistently
> estimated with the restriction that estimated t=(x,y,x,y,w). Because
> of the symmetry of the problem, I am doubtful y*=x*=0.2 would be a
> point of inconsistency if t=(t_1,...,t_5) were unrestricted in
> estimation. The issue deserves discussion because I don't think it is
> transparent.

> I don't think the restricted case of p8-9 will be of interest to
> readers. Only the general case.

> Minor Comments:

> -l78-88: Calculations are more natural with `fidelities' and this
> might remain in the appendix but it is much more conventional in the
> literature for x,y,w to be edge-lengths or probabilities of a
> different character states at the tips of the edge. It takes quite a
> bit of getting used to realize that (0,0) corresponds to infinite
> edge-lengths and it seems unnecessary with a little work on the part
> of the authors to do the conversions that will make the figs more
> natural for readers.

> Reviewer: 2

> Comments to the Author
> In this manuscript  the authors show that joint inference of phylogeny
> and ancestral characters is not consistent in general by providing a
> counter-example, in which as the number of sites grows, they show that
> the wrong tree topology can have a higher likelihood than the correct
> topology under the joint inference model.
> The result is important for the community to understand, in particular
> because TreeTime (the joint inference method) is used in high-profile
> analyses on Ebola, influenza and other viruses on the nextflu /
> nextstrain websites. These are beautiful and accessible sites aiming
> to make analyses on viruses open to exploration by the community, but
> this comes at the cost of suppressing the details of the analysis. If
> indeed there are deep statistical issues or limitations it is
> important that the community knows what these are and what their
> implications are.

> Accordingly my comments are focussed on where I think the manuscript
> could be more clear (and where I did not quite follow the logic). My
> major suggestion would be to go even further towards demonstrating in
> a simple example how the joint inference can be wrong – for example,
> providing a toy multiple sequence alignment or even one of some
> influenza HA sequences – you could find 2 sets of 4 closely-related
> sequences that have the Farris and Felsenstein topologies, maybe? And
> demonstrate that the inference gives the wrong result by explicitly
> computing the likelihood.

> Also, laying out the theory a little bit more step-by-step so that a
> general quantitative reader doesn't struggle too much to follow would
> help (suggestions below).

> I know that it's not feasible to explicitly compute (analytically)
> these likelihoods for bigger trees. But one of the main motivations
> and applications for tools like TreeTime is surely that the number of
> sequences grows (not the number of sites, which is fixed by the viral
> sequence length in the case of viruses). What happens when the number
> of tips grows? Are there more opportunities for the kind of issue that
> gives the wrong value for w? Many influenza trees have many 0 (or near
> 0) internal branch lengths, and it would be interesting to understand
> the implications for the ancestral state reconstructions (which in
> turn are used to infer spatial spread, transmission etc).

> Some detailed comments and questions:
> – line 85 some cut/paste or sentence issue
> - line 97 below (2) define A^p
> - l106: why is the functional form of (3) determined by inequalites;
> what are they?
> - l 118-124: clarify are tau1 and tau2 Farris and Felsenstein?
> L 132 clarify “equivalent but different notion” – different from what?
> - line 146 “seeing no change along a branch more likely increases the
> likelihood by introducing a term of (1 + θ) instead of (1 − θ)” – I
> don't understand this
> line 185++ and figure5: true value of w?

> L 334 missing “the”

> L332++ I got confused about the complements and factors of 2, though I
> see the point that the site split has symmetry. As it's fundamental to
> the rest of the work, a more clear statement of where the 2 comes from
> in the second eq below 358 would be very helpful.

> L354 why does psi take the complement of site patterns? Just above it
> is defined as a random variable psi (Y) where Y is a site pattern

> What are the elements of the sets eta(tau, t)?

> Why is xi surjective? I don't understand the equation before line 359,
> and then I don't understand what “such a choice” is in line 359
> line 364 define p(e)

> line 366  I don't understand (8) but perhaps that's just background I
> don't have – a quick explanation or reference would be handy
> What's a 'generating probability' ? Is that the fidelity? Or a
> generator?

> Reviewer: 3

> Comments to the Author
> Shaw and Matsen show that a joint maximum-likelihood inference of
> phylogenies and ancestral states is not consistent when branch lengths
> are long. We have recently proposed such a joint estimation for
> sequences that recently diverged and are hence similar (Sagulenko et
> al). The results presented here are anticipated in Sagulenko et al
> "For more divergent sequences, however, subleading states of internal
> nodes make a substantial contribution and the iterative optimization
> will underestimate the branch lengths."
> and
> "The iterative optimization procedures are not expected to be accurate
> for trees were many sites are saturated. In scenarios with extensive
> uncertainty of ancestral states and tree topology, convergence of the
> iterative steps cannot be guaranteed."

> In this light, it should be pointed out that joint estimation
> procedure was explicitly suggested as an approximation and hence never
> expected to be consistent in a strict sense. Furthermore, it was
> suggested as an approximation when branch lengths are short -- the
> opposite of the problematic regions identified here.
> But systematic analysis of the behavior of this approximation is
> useful and I believe the work is correct (but haven't checked the
> proofs). The work by Shaw and Matsen suggests the radius of
> convergence of this procedure is somewhere around branch length of
> order 10-20%, which is consistent with results reported by Sagulenko
> et al.

> At the same time, these results are not particularly surprising.
> Assigning ML ancestral states will always result in shorter branches.
> For short branches, this correction is linear in branch length, for
> longer branches higher order effects start to contribute. If the
> typical branch length around a bifurcating node is x<<1, a fraction ~x
> of the sites mutated along one of the three adjacent branches, while a
> fraction ~x^2 mutated on two branches and not the third. The latter
> case is ignored in this joint inference procedure. Not having to sum
> over these subleading states is what makes joint inference faster, but
> such truncation will first lead to inaccuracies and at longer
> distances to qualitatively different solutions (like Taylor expanding
> 1/(1-x) ).

> There are other obvious cases where joint inference will fail. For
> sequences with a heavily biased in nucleotide composition (say biased
> towards A). For sufficiently long terminal branches, all ancestral
> states will be estimated as all 'A' and subsequent branch length
> estimate will shrink all internal branches to zero. This happens for
> the very same reason that the most likely outcome of multiple flips of
> a biased coin is all heads -- a very atypical yet most likely outcome.

> Currently, the work largely shows that areas of inconsistencies exist.
> It would be useful if the authors could present Taylor expansions for
> the ML branch length estimates around x=y=w=1. This would more
> explicitly show the inaccuracies accumulate with increasing branch
> length and how the inaccuracies depend on asymmetries in the tree. To
> me, this would be much more useful than to know that joint inference
> will fail if branches are long.

> Specific comments:
>  - when y=x=w, the two topologies are indistinguishable other than for
>  the labeling of the leaves. What happens in Fig 2 on the diagonal?
>  Does it degenerate into a star?

>  - line 50: overestimates branch lengths? It overestimates fidelity ->
>  underestimates branch length.  As expected.  Again line 249. Please
>  check. The back and forth between fidelity and branch length can be
>  confusing.

>  - Fig 5: it is surprising that \hat{w} seems completely flat as a
>  function of x* until it jumps discontinuously to 0. Why not include
>  the analytical bound in Fig 5?

>  - It would be nice if the abstract pointed out that these
>  inconsistencies arise when branch lengths are long and that the
>  approximate joint maximization procedure was suggested for precisely
>  the opposite limit. Similar at the beginning of the discussion, I
>  think it would be more appropriate if this would say smth along the
>  lines "Sagulenko et al suggested an approximation for short
>  branches... we show that this doesn't work for long branches..."

> Richard Neher

> Reviewer: 4

> Comments to the Author
> Review of Shaw, D. and Matsen, F.  "Joint maximum-likelihood of
> phylogenies and ancestral states is not consistent"

> Shaw and Matsen demonstrate that the approach of jointly estimating
> ancestral states and tree topologies (or branch lengths on a fixed
> topology) is not statistically consistent.  I believe that their
> results are correct, and that their result is important to
> disseminate.  The paper is clearly written, given the complexity of
> the material, and I have no reservation about recommending publication
> virtually as is.

> However, although my original intent was to carefully work through the
> math (including the derivations and proofs in the appendix), I have
> not found the time to do so, and because of the lateness of my review
> at this point, I will provide just a few comments in the hope that
> they will be useful.  Hopefully, other reviewers were able to check
> the correctness of the mathematical results more carefully.

> The only real criticism I have of the paper is that the authors could
> have done a better job in connecting their new work to previous
> results.  For example, Goldman (1990, Syst. Biol. 39:345-361) showed
> that joint estimation of topology and ancestral states by maximum
> likelihood was equivalent to maximum parsimony analysis for the same
> 2-state model considered by Shaw and Matsen when all branches are
> forced to have the same length.  They noted the statistical
> inconsistency of this approach, which was necessarily true because
> parsimony was already known to be inconsistent (the famous result of
> Felsenstein 1978).  Shaw and Matsen extend this result to the case of
> unequal branch lengths, which while not particularly surprising, is
> nonetheless an important extension.  The authors should also cite key
> relevant earlier work on the identifiability ot the tree topology with
> "classical" maximum likelihood phylogenetic estimation (Allman et al
> 2008, Adv Appl Prob 40:229–249; Chai and Housworth 2011, Syst Biol
> 60:713–718) in addition to crediting Roychoudhury 2015 for
> establishing ML consistency.

%EM Can you chat me links to these refs when you find them? I can get
%the pdfs for you as needed.

> Finally, let me just add that I hate the term "Farris Zone".  The
> "Felsenstein Zone" was a name applied to Felsenstein's important
> demonstration of a tree and branch-length configuration where
> parsimony would be an inconsistent estimator of the phylogeny.  On the
> contrary, Farris had nothing to do with the "Farris zone"; it was
> simply a name given by Mark Siddall to a different configuration in
> which parsimony outperforms likelihood, thinking incorrectly that
> likelihood was inconsistent in this zone.  Swofford et al. (2001,
> cited by the authors) referred to this zone as the "inverse
> Felsenstein zone", which I think is a better name for it.

### 2018-04-17 joint_inf.tex
fixing reference to Julie's grant

@@
 
-This work was supported by National Institutes of Health grants R01-GM113246, R01-AI12096, U19-AI117891, and U54-GM111274 as well as National Science Foundation grants CISE-1561334 and CISE-1564137.
+This work was supported by National Institutes of Health grants R01-GM113246, R01-AI120961, U19-AI117891, and U54-GM111274 as well as National Science Foundation grants CISE-1561334 and CISE-1564137.
 The research of Frederick Matsen was supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation.

### 2018-04-30 appendix.tex
clarify, be explicit, address (most of) reviewer 2

@@
 
-Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation.
+Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simpler likelihood calculation.
 Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
@@
 \begin{definition}
-Let the mapping from site patterns to site splits be
+Define the complement of $\alignmentColumn$ as $\overline{\alignmentColumn}$.
+Let the mapping from site patterns to site splits
 $$
@@
 $$
-and the mapping from ancestral states and tip states to ancestral state splits be
-$$
-\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet).
+be
+$$
+\patternToSplit(\alignmentColumn) =
+\left\{
+    \begin{array}{ll}
+        \alignmentColumn  & \mbox{if } \alignmentColumn\in\mathcal{P}(\siteSplitSet) \\
+        \overline{\alignmentColumn}  & \mbox{if } \alignmentColumn\notin\mathcal{P}(\siteSplitSet)
+    \end{array}
+\right.
+$$
+and the mapping from ancestral states and tip states to ancestral state splits
+$$
+\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
+$$
+be
+$$
+\ancestralToSplit(\alignmentColumn, \ancestralStateColumn) =
+\left\{
+    \begin{array}{ll}
+        \ancestralStateColumn  & \mbox{if } \alignmentColumn\in\mathcal{P}(\siteSplitSet) \\
+        \overline{\ancestralStateColumn}  & \mbox{if } \alignmentColumn\notin\mathcal{P}(\siteSplitSet).
+    \end{array}
+\right.
 $$
@@
 \end{definition}
-The mapping $\patternToSplit$ takes the complement of site patterns to obtain a site split in $\mathcal{P}(\siteSplitSet)$.
+The mapping $\patternToSplit$ operates by returning the complement of a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if it is not in $\mathcal{P}(\siteSplitSet)$.
 The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.
@@
 $$
-As a consequence of assuming a binary symmetric model, taking complements yields
-\begin{align*}
-    2\cdot \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
-                                                                 &= \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)
-\end{align*}
-for some $j$ and
+As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
+\begin{align*}
+    \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t)\\
+    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
+    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
+    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t)
+\end{align*}
+where $\overline{Y}$ is the complement of the site-patterned random variable $Y$ and $\overline{Y}\sim Y$.
+For ancestral states and for some $\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)$
 \begin{align*}
@@
 \end{align*}
-Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
+Given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
 \begin{align*}
@@
 In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
-For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this we have
-\begin{align*}
-\max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
-\end{align*}
+For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it maps all possible values from $\alphabet^\nAncestralStateRows$ to an element in $\mathcal{P}(\ancestralSplitSet)$, and from this we have
+\begin{align*}
+    \max_{\ancestralStateColumn_i} \ \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t)\\
+    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)\\
+    &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
+\end{align*}
+Thus, each term in the likelihood can be collapsed into terms relating only to site splits, indexed by $j$,  as opposed to each individual observation, indexed by $i$.
 
@@
 
-Let $\xi_j$ be such a choice for each $1 \leq j \leq q$.
-Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
+The likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
 \begin{align}
@@
 \end{align}
-where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.
+for some $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.
 
@@
 
-We state the Hadamard representation of site split generating probabilities, following Section 8.6 of \citet{Semple2003-em}.
+We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
 For each edge $e$ define the edge ``fidelity'' for that edge as
 $$
-\theta(e) = 1-2p(e).
-$$
+\theta(e) = 1-2p(e)
+$$
+where $p(e)$ is the probability of a character change along edge $e$.
 For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.

### 2018-04-30 joint_inf.tex
clarify, be explicit, address (most of) reviewer 2

@@
 We observe $\nCols$ independent and identically distributed samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
-The corresponding unobserved ancestral states are $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ and distributed as $\ancestralStateColumnRV$.
+The corresponding unobserved ancestral states are $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ and distributed as $\ancestralStateColumnRV$ with each $\ancestralStateColumn_i\in\alphabet^\nAncestralStateRows$.
 
@@
 
-Fidelities have useful algebraic properties, and generating probabilities using the Hadamard transform have an especially simple form (see \eqref{eq:hadamard_probability} in the Appendix).
+Fidelities have useful algebraic properties.
+As data becomes plentiful, we use the Hadamard transform (see \eqref{eq:hadamard_probability} in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form.
 For a four-taxon tree, define the general branch fidelity parameter $t=\{\theta_1,\theta_2,\theta_3,\theta_4,\theta_5\}$ where fidelities are ordered in the order of the taxa with the internal branch last (Fig.~\ref{fig:farris-fels-top}).
@@
 \end{equation}
-In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
+In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima row-wise in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} to obtain the likelihood term for each most likely ancestral state, these terms depending on the unknown $(\tau, t)$.
 For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2017-jo}.
@@
 To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the Farris topology (Fig.~\ref{fig:farris-fels-top}a).
-We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology than for the true, generating Farris topology.
+We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology ($\tau_2$) than for the true, generating Farris topology ($\tau_1$).
 To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the Farris topology as a function of $x^*$ and $y^*$ and, similarly, a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
@@
 We now consider the problem of branch length estimation on the correct tree using joint estimation.
-As described above, we use the equivalent but different notion of branch fidelities.
+As described above, we use the notion of fidelities to analyze what happens as changes occur along each branch as opposed to the standard branch length formulation.
 We analyze two settings on the Farris tree, corresponding to whether some branch fidelities are fixed at their true values or not.
@@
 Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch with this branch length estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
-As an intuition for the theoretical development, seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.
+As an intuition for the theoretical development, if there is no change along a branch with fidelity $\theta$, the likelihood will more likely include a term of $(1+\theta)$ instead of $(1-\theta)$ for this event, resulting in positive bias of branch fidelities.
 If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.
@@
 
+As data were generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
 Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same procedure.

### 2018-05-01 versions/response.v1.txt
Adding responses.

@@
 > – line 85 some cut/paste or sentence issue
+
+TODO: not sure what this is?
+
 > - line 97 below (2) define A^p
+
+We have clarified this with an explicit definition of A^p where we introduce A, at the beginning of the section on Phylogenetic maximum likelihood.
+
 > - l106: why is the functional form of (3) determined by inequalites;
 > what are they?
+
+We have made this more precise and directly referenced the inequalities.
+
 > - l 118-124: clarify are tau1 and tau2 Farris and Felsenstein?
+
+That is correct, and we've added parenthetical notes to reminde the reader, thanks.
+
 > L 132 clarify “equivalent but different notion” – different from what?
+
+We have reformulated this sentence to avoid the comparison.
+
 > - line 146 “seeing no change along a branch more likely increases the
@@
 > don't understand this
+
+TODO: this is changed, but I don't see this being a lot clearer.
+
 > line 185++ and figure5: true value of w?
 
+We have added clarifying sentence:
+"As data were generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$."
+
 > L 334 missing “the”
+
+Fixed via simpler language.
 
@@
 
+We have now made this much more explicit with three additional display equations spelling out the intermediate steps.
+
 > L354 why does psi take the complement of site patterns? Just above it
@@
 
+This wasn't as clear as it should have been. We have revised to:
+"The mapping $\patternToSplit$ operates by returning the complement of a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if it is not in $\mathcal{P}(\siteSplitSet)$."
+
 > What are the elements of the sets eta(tau, t)?
+
+TODO: fill in
 
@@
 > and then I don't understand what “such a choice” is in line 359
+
+We have made this explicit now, with two additional intermediate steps and an explicit explanation in terms of collapsing terms.
+
 > line 364 define p(e)
+
+Thanks for reminding us to define p(e), which is now described as "where $p(e)$ is the probability of a character change along edge $e$."
 
@@
 > generator?
+
+TODO: just say tough nuts?
 

### 2018-05-01 appendix.tex
Making a partial pass through Appendix.

@@
 \subsection*{Site split formulation}
-We begin by introducing ``site splits,'' which formalize the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
+We begin by introducing ``site splits.''
+One application of site splits is to formalize the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
 This is a standard step in the description of the Hadamard transform (Section 8.6 of \citet{Semple2003-em}), although our approach is complicated slightly by the inclusion of ancestral states.
 
-Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simpler likelihood calculation.
+Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
 Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
@@
 
-For topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
-In this case the entire set of internal nodes must be enumerated: the probability of observing an ancestral state split conditional on a site split is not invariant to taking its complement.
+For a fixed topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
+In this case we cannot use the same complement trick as before: the probability of observing an ancestral state split conditional on a site split is not invariant to taking its complement.
 
@@
 \end{align*}
-where $\overline{Y}$ is the complement of the site-patterned random variable $Y$ and $\overline{Y}\sim Y$.
-For ancestral states and for some $\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)$
+where $\overline{Y}$ is the complement of the site-patterned random variable $Y$ and so has the same distribution as $Y$.
+Note
 \begin{align*}
@@
 \end{align*}
-Given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
+Thus given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
 \begin{align*}
@@
 In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
+%EM Wait, does this establish surjectivity? We need to be able to get all values of P(H).
 For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it maps all possible values from $\alphabet^\nAncestralStateRows$ to an element in $\mathcal{P}(\ancestralSplitSet)$, and from this we have
@@
     &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)\\
-    &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
-\end{align*}
+    &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)
+\end{align*}
+for some $j$ depending on $i$.
 Thus, each term in the likelihood can be collapsed into terms relating only to site splits, indexed by $j$,  as opposed to each individual observation, indexed by $i$.
@@
 
-The likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
+Thus the likelihood in \eqref{eq:profile_likelihood} can be written as a product over site patterns as opposed to sites:
 \begin{align}
@@
 where $p(e)$ is the probability of a character change along edge $e$.
-For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
+For an even-sized subset of $Y\subseteq\mathcal{S}$, let the path set $P(Y)$ be the set of edges in the path connecting both elements of $Y$.
 For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
@@
 $$
-Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{3\}$ and internal node $\{2\}$ connected to leaves $\{2\}$ and $\{4\}$.
+Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
 The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
@@
 See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the Farris tree.
-Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods}.
+Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} (for the Farris tree) and~\ref{tab:fels_likelihoods} (for the Felsenstein tree).
 Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}.
@@
 Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
-We decompose the likelihood into the entropy and partial likelihood terms as in \eqref{eq:log_likelihood_simplified}.
+We decompose the likelihood into the entropy and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
+%EM I'm thinking about this wrong for some reason: the sum of two things, each of which has a unique maximum, does not necessarily have a unique maximum.
 By Gibbs's inequality, the entropy term $\shannonDivergence_{\tau^*,t^*}(\tau^*,t)$ has a unique maximum over $t$---namely $t^*$.
-For the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$, we can compute the partial likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ as in Table~\ref{tab:farris_likelihoods}.
-Clearly this partial likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in[0,1]$ and thus has a unique maximum.
+%EM There's not good notational overlap between Table S2 and the definition of partial log-likelihood. Could you fix that?
+For the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$, we can compute the partial log-likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ as in Table~\ref{tab:farris_likelihoods}.
+Clearly this partial log-likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in[0,1]$ and thus has a unique maximum.
 Since both summands of the likelihood are concave on $x_1\in[0,1]$, the likelihood is as well, and there exists a single $x_1\in[0,1]$ maximizing the likelihood.
@@
 \end{align}
+%EM Shouldn't this read "obtaining an upper bound"?
 Our plan is to simplify the likelihood to remove log-of-quadratic terms of $x,y,$ and $w$, obtaining a likelihood that has a closed-form maximum in each variable.
@@
 + (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x) \\
+%EM Shouldn't this be
+%&\qquad\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y)
 &\qquad\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y)
+%EM And this be
+%+ 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
 + (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
@@
 The upper bound in \eqref{eq:farris-upper-bound} is for the ancestral state split $\{1\}$.
+%EM Let's be more specific, giving an equation number, rather than just saying "the upper bound" here.
 For the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.

### 2018-05-02 versions/response.v1.txt
Minor changes and suggestions; draft response.

@@
 
+Outline of revisions:
+1.
+2.
+
+Some of the reviewers have expressed their opinion that we reduce our emphasis on mathematical development in exchange for a more numerical-optimization-based approach.
+Our primary goal was to mathematically prove inconsistency of joint inference, which is what we have done.
+As pointed out by reviewer 4, this answers an open question, which has special relevance given the recent work of Sagulenko et al.
+We provide some numerical results to confirm our derivations and to provide a comparison.
+
+As we now emphasize (TODO XXX) this work is not the final word on the joint inference question from an applied perspective.
+Even if we were to greatly expand the sections doing numerical optimization, these results would only be for the binary symmetric model, which is never applied in practice.
+An investigation of these aspects deserves another paper.
+
 > Finally, please pay VERY careful attention to improving your
@@
 
-Felsenstein's book has a subsection entitled "The Farris zone" spanning two pages (pp 120-121), a figure entitled "A tree in the Farris zone.", and a table with a caption "Properties of the Felsenstein and Farris zones..."
+We are grateful to reviewer 4 for pointing us to literature of which we were not aware.
+We have greatly expanded ...
+
+We have complied with this request regarding terminology, despite it going against current practice.
+Indeed, Felsenstein's book has a subsection entitled "The Farris zone" spanning two pages (pp 120-121), a figure entitled "A tree in the Farris zone.", and a table with a caption "Properties of the Felsenstein and Farris zones...".
 The term "Farris zone" has almost 7 times as many citations as the term "inverse Felsenstein zone" on Google Scholar.
-Thus, although we realize that the term causes offense, it is standard.
 
@@
 > interest.
+
+As described above, our intent is to mathematically establish inconsistency and poor estimation.
+This must be done with analytic arguments rather than optimization.
+We could have only shown this bad behavior in a very small area to establish that it happens.
+In contrast to this reviewer, we believe that Fig 5 is optional to this line of argument, and we provide it just for interest.
+A true investigation into the practical implications of our work will require another paper.
 
@@
 
-Theory leads to generalization.
+Please see above for our response.
 
@@
 
+TODO: make the argument more clear in the main text.
+
 > I don't think the restricted case of p8-9 will be of interest to
 > readers. Only the general case.
+
+%EM I think we could move this to the appendix to throw this reviewer a bone.
 
@@
 
+We agree that understanding the implications of our derivation is a very interesting avenue for future work.
+However, to do such a project justice would require a separate paper.
+We are sure that once our paper demonstrates this inconsistency, others will quickly follow to write such a paper.
+
 > Also, laying out the theory a little bit more step-by-step so that a
@@
 > help (suggestions below).
+
+Thank you very much for reading the derivations and for your suggestions. As noted inline we have revised the text accordingly.
 
@@
 > turn are used to infer spatial spread, transmission etc).
+
+%EM David-- does it seem like we could extend these results to bigger trees containing a 4 taxon tree?
 
@@
 
+Thank you for your thoughts, and we have worked hard to not oversell our results.
+TODO we should make some changes to at least say that we've further clarified the implications of our results.
+The TreeTime paper does present joint optimization as an approximate approach, and described that it is for short branch lengths.
+However, in our assessment the word "approximate" doesn't completely describe a situation in which there is significant bias.
+The TreeTime paper says that its use is appropriate when "ancestral sequence inference has little ambiguity" but even in these cases the bias in branch lengths persists.
+
 > At the same time, these results are not particularly surprising.
@@
 > 1/(1-x) ).
-
+>
 > There are other obvious cases where joint inference will fail. For
@@
 > a biased coin is all heads -- a very atypical yet most likely outcome.
-
+>
 > Currently, the work largely shows that areas of inconsistencies exist.
@@
 > will fail if branches are long.
+
+This is an interesting suggestion but is different than our goal, which was to establish inconsistency.
+TODO describe why Taylor isn't a straightforward thing.
+TODO refer to idea of applied paper again?
 
@@
 >  branches... we show that this doesn't work for long branches..."
+
+TODO we should do some change, but make it clear that it's consistently biased even for short branches.
 
@@
 
+Thank you for your assessment.
+We have worked hard in our initial submission and in our revision to write out the argument in great (indeed, excruciating) detail, and we would appreciate it if you could take a look at the revision.
+
 > The only real criticism I have of the paper is that the authors could
@@
 
-%EM Can you chat me links to these refs when you find them? I can get
-%the pdfs for you as needed.
+We are very grateful to this reviewer for pointing out the Goldman paper.
+We were not aware of this paper and now have a paragraph describing it and comparing to our work.
+It's also a good idea to cite additional papers about identifiability, and we have done so.
+
+%EM David, heh, can you point me to the place where the equal branch length result was done?
+%EM Also, I think you said that he says that the varying branch length thing is an open question-- it would be great to quote that.
 
@@
 > Felsenstein zone", which I think is a better name for it.
+
+As described at the beginning, we have made this change, thanks.

### 2018-05-02 appendix.tex
Minor changes and suggestions; draft response.

@@
 Taking such a complement simplifies but does not change the result of likelihood computation because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model.
+%EM "ancestral state splits" are an important item in the development but don't get the same level of definition as "site splits". Let's define them here.
 
@@
 We enumerate the site splits $\siteSplit_j$ of which there are $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ in total where $\mathcal{P}$ denotes the power set.
+%EM I think that we should either use "ancestral splits" everywhere, or "ancestral state splits."
 Similarly we enumerate ancestral splits $\ancestralSplit_k$ of which there are $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$ in total.
@@
 + p_{13}         \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
-    &\qquad\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y))
-+ p_{123}        \cdot\log((1+x)^2   (1+w)(1+y)(1-y)) \\
+    &\qquad\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y)) \\
+    &\qquad\qquad + p_{123}        \cdot\log((1+x)^2   (1+w)(1+y)(1-y)) \\
 &\qquad =      p_{\emptyset}  \cdot\log(2)
@@
 
-The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
+The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications, as we now describe.
 The upper bound in \eqref{eq:farris-upper-bound} is for the ancestral state split $\{1\}$.
@@
 For the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.
-In this case, all terms involving $w$ simplify to $\log(1+w)$, maximized at $\hat{w}=1$.
+In this case, all terms involving $w$ simplify to $\log(1+w)$, which is maximized at $\hat{w}=1$.
 For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
@@
 for site split $\{1,2\}$.
-To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
+To bound the case of ancestral state split $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
 \begin{align*}
@@
 \end{align*}
-and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
+and, for the case of ancestral state split $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
 \begin{align*}
@@
 Some rearranging and the fact that $0 \le \tilde{p} \le 1/2$ shows this is equivalent to showing $p_{13} \le 3/8$.
+%EM Using what about t^*?
 Using Table~\ref{tab:sitepatprob} and $t^*$,
@@
 \begin{equation}
-1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 2 - (x^*)^2
+\begin{split}
+& 1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3 \\
+& \qquad \qquad \qquad \qquad \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 2 - (x^*)^2
+\end{split}
 \label{eq:lenny}
@@
 
-We construct a similar lower bound on the Felsenstein tree partial likelihood.
+We next construct a lower bound on the Felsenstein tree partial likelihood.
 In the Felsenstein case, there are many more optimal internal states depending on branch lengths than in the Farris case (see the lower table of Table~\ref{tab:likelihoods}).

### 2018-05-02 appendix.tex
fix deprecated $$; tex kinda sux :(

@@
 Let the mapping from site patterns to site splits
-$$
+\[
 \patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
-$$
+\]
 be
-$$
+\[
 \patternToSplit(\alignmentColumn) =
@@
 \right.
-$$
+\]
 and the mapping from ancestral states and tip states to ancestral state splits
-$$
+\[
 \ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
-$$
+\]
 be
-$$
+\[
 \ancestralToSplit(\alignmentColumn, \ancestralStateColumn) =
@@
 \right.
-$$
+\]
 Then, given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
-$$
+\[
 \siteSplitRV := \patternToSplit(\alignmentColumnRV)
-$$
+\]
 that takes corresponding realizations $\siteSplit_j$ for some $j$, and
-$$
+\[
 \ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
-$$
+\]
 for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
@@
 For the $i$th factor of \eqref{eq:full_likelihood},
-$$
+\[
 \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
-$$
+\]
 As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
@@
 Let
-$$
+\[
 L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
-$$
+\]
 be the final product in \eqref{eq:site_pattern_likelihood}.
@@
 Define the divergence quantity
-$$
+\[
 \shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)
-$$
+\]
 and the partial log-likelihood
-$$
+\[
 \tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
-$$
+\]
 so that \eqref{eq:site_pattern_profile_likelihood_mean} is
@@
 For each edge $e$ define the edge ``fidelity'' for that edge as
-$$
+\[
 \theta(e) = 1-2p(e)
-$$
+\]
 where $p(e)$ is the probability of a character change along edge $e$.
@@
 For notational convenience, let
-$$
+\[
 p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
-$$
+\]
 for any site split $\siteSplit_j$.
@@
 This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
-$$
+\[
 \mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
-$$
+\]
 Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
@@
 We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
-$$
+\[
 \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
-$$
+\]
 in general.
@@
 The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
-$$
+\[
 \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
 (1-p)^5,
-$$
-$$
+\]
+\[
 \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
@@
 p^3(1-p)^2,
-$$
-$$
+\]
+\[
 \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
 p^4(1-p).
-$$
+\]
 The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
 Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, t)$ we have
-$$
+\[
 \Pr(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
@@
 (1-p)^5.
-$$
+\]
 For $\siteSplit_5=\{1,2\}$ we have
-$$
+\[
 \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
@@
 p^2(1-p)^3,
-$$
-$$
+\]
+\[
 \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
@@
 p^3(1-p)^2.
-$$
+\]
 Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, t)$,
-$$
+\[
 \Pr(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
 p^2(1-p)^3.
-$$
+\]
 
@@
 In general,
-$$
+\[
 \max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \le
@@
     + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
-$$
+\]
 is an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
-$$
+\[
 \shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
-$$
+\]
 and
-$$
+\[
 \hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t).
-$$
+\]
 Similarly,
-$$
+\[
 \max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \ge
@@
     + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
-$$
+\]
 is a lower bound.
@@
 Maximizing over the unknown terms yields
-$$
+\[
 \hat{x} = \frac{a_{1}-a_{2}}{a_{1}+a_{2}}, \ \hat{y} = \frac{a_{3}-a_{4}}{a_{3}+a_{4}}, \ \hat{w} = 1-2p_{13}.
-$$
+\]
 Trivially $0 \le \hat{w} \le 1$ and $\hat{x}, \hat{y} \le 1$.
@@
 For $\hat{x} \ge 0$ we need $a_1 \ge a_2$; letting
-$$
+\[
 \tilde{p} = p_{1}+p_{3}+p_{12}+p_{23}
-$$
+\]
 we have from \eqref{eq:a_const} that $a_1 \ge a_2$ only if $3 \ge 3\tilde{p}$.
@@
 We now need to check that
-$$
+\[
 \hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
-$$
+\]
 are both between zero and one.
@@
 Using Table~\ref{tab:sitepatprob} and $t^*$,
-$$
+\[
 p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right),
-$$
+\]
 meaning we are interested in whether
@@
 Call the upper bound
-$$
+\[
 C_0(x^*, y^*) := \max\left(C^{1}_{\tau_1,\tau_1}(x^*, y^*), C^{2}_{\tau_1,\tau_1}(x^*, y^*), C^{3}_{\tau_1,\tau_1}(x^*, y^*)\right).
-$$
+\]
 
@@
 For example, for site split $\{1,2\}$, after we replace the $1+w$ term for $\ancestralSplitPartition_j(\tau, t)=\{\emptyset, \{1,2\}\}$ with $1-w$, we have
-$$
+\[
 \max\left((1+x)^2 (1-w)(1+y)^2, (1+x)(1-x)(1-w)(1+y)(1-y)\right)
-$$
+\]
 where since
-$$
+\[
 (1+x)^2 (1-w)(1+y)^2 > (1+x)(1-x)(1-w)(1+y)(1-y),
-$$
+\]
 the maximum for this site split can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
@@
 If $(1+x)(1-y) > (1-x)(1+y)$, then
-$$
+\[
 (1+x)^2(1-w)(1+y)(1-y) > (1+x)(1-x)(1-w)(1+y)^2
-$$
+\]
 meaning the cases $\{1\}, \{2\}, \{3\}$ and $\{1,2,3\}$ will have likelihoods bounded below by
-$$
+\[
 (1+x)^2(1-w)(1+y)(1-y).
-$$
+\]
 For the cases $\{1,3\}$ and $\{2,3\}$ and the same condition, we have
-$$
+\[
 (1+x)^2(1-w)(1-y)^2 > (1-x)^2(1-w)(1+y)^2
-$$
+\]
 and
-$$
+\[
 (1+x)^2(1-w)(1-y)^2 > (1+x)(1-x)(1-w)(1+y)(1-y),
-$$
+\]
 yielding lower bounds of
-$$
+\[
 (1+x)^2(1-w)(1-y)^2
-$$
+\]
 in these cases.
@@
 Letting
-$$
+\[
 b = p_{1}+p_{2}+p_{3}+p_{123}
-$$
+\]
 yields
-$$
+\[
     \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
@@
     + \log(1-w)
-$$
+\]
 for the first lower bound, and we get the equivalent expression for the second lower bound by exchanging $x$ and $y$.
 Maximizing either lower bound over $t$, we obtain
-$$
+\[
 \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
@@
 + b      \cdot\log(b),
-$$
+\]
 i.e., the maximum is achieved at $\{1,1-b,0\}$.
 Our lower bound is
-$$
+\[
 C_1(x^*, y^*) := \shannonDivergence_{\tau_1,t^*}(\tau_2,\{1,1-b,0\}) + 2\cdot\log(2) + (2-b)\cdot\log(2-b) + b\cdot\log(b).
-$$
+\]
 See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < x^*, y^* < 1$ where $C_1(x^*, y^*) > C_0(x^*, y^*)$ and joint inference is inconsistent.
@@
 Since
-$$
+\[
 \log(1+w) \geq (1 - p_{13})\cdot\log(1+w) + p_{13}\cdot\log(1-w)
-$$
+\]
 with equality only when $w=0$, the second likelihood is the likelihood with the maximum-likelihood ancestral state, and so we only analyze that case.
@@
 Simplifying, the likelihood can be written
-$$
+\[
 \ell(w) = \alpha_1\log(\beta+\gamma w) + \alpha_2\log(\beta-\gamma w) + \log(1+w)
-$$
+\]
 where
-$$
+\[
 \alpha_1 := \alpha_1(x^*, y^*) = \frac{1}{8} \left(1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
-$$
-$$
+\]
+\[
 \alpha_2 := \alpha_2(x^*, y^*) = \frac{1}{8}\left(1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
-$$
-$$
+\]
+\[
 \beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
-$$
-$$
+\]
+\[
 \gamma := \gamma(x^*, y^*) = 4x^*y^*.
-$$
+\]
 Since $x^*$ and $y^*$ fall between zero and one, we make use of the inequalities (excepting the cases of $x^*=y^*=0$ and $x^*=y^*=1$)
-$$
+\[
 \alpha_1 > \alpha_2
-$$
+\]
 and
-$$
+\[
 \beta > \gamma.
-$$
+\]
 % Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
 The derivative of the log-likelihood with respect to $w$ is
-$$
+\[
 \ell'(w) := \frac{d}{dw} \ell(w) = \frac{\alpha_1 \gamma}{\beta+\gamma w} - \frac{\alpha_2 \gamma}{\beta-\gamma w} + \frac{1}{1+w}.
-$$
+\]
 The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w) \lessgtr 0$, we equivalently consider $f(w) \lessgtr 0$ where $f$ is the quadratic function
@@
 Using the quadratic formula with
-$$
+\[
 a = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
-$$
-$$
+\]
+\[
 b = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
-$$
-$$
+\]
+\[
 c = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
-$$
+\]
 the smaller zero is
-$$
+\[
 \hat{w} = \frac{-b - \sqrt{b^2 - 4ac}}{2a},
-$$
+\]
 which is a function of the generating parameters $x^*$ and $y^*$.
@@
 With this we have,
-$$
+\[
 \hat{w} \ge 1 \iff
@@
 a + b + c \ge 0.
-$$
+\]
 Using
-$$
+\[
 \alpha_1 + \alpha_2 = \frac{1}{4}\beta
-$$
+\]
 and
-$$
+\[
 \alpha_1 - \alpha_2 = x^*(y^*)^2,
-$$
+\]
 and simplifying as functions of $\gamma$ and $\beta$ shows that $a + b + c \ge 0$ is equivalent to
@@
 Given the bounds
-$$
+\[
 \gamma_{L} \le \gamma(\hat{x}, \hat{y}) \le \gamma_{U}
-$$
+\]
 and
-$$
+\[
 \beta_{L} \le \beta(\hat{x}, \hat{y}),
-$$
+\]
 if
-$$
+\[
 -\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0,
-$$
+\]
 then \eqref{eq:general-bl-result} holds, and this is an inequality involving only $x^*$ and $y^*$.

### 2018-05-02 joint_inf.tex
Minor changes and suggestions; draft response.
fix deprecated $$; tex kinda sux :(

@@
 \usepackage{thm-restate}
-\usepackage{lineno}
-\linenumbers
 \usepackage{natbib}
@@
 \usepackage{authblk}
-\usepackage{fancyhdr}
+
+% Line numbers and some weirdness
+\usepackage{lineno}
+\linenumbers
+\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
+  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
+  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
+  \renewenvironment{#1}%
+     {\linenomath\csname old#1\endcsname}%
+     {\csname oldend#1\endcsname\endlinenomath}}% 
+\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
+  \patchAmsMathEnvironmentForLineno{#1}%
+  \patchAmsMathEnvironmentForLineno{#1*}}%
+\AtBeginDocument{%
+\patchBothAmsMathEnvironmentsForLineno{equation}%
+\patchBothAmsMathEnvironmentsForLineno{align}%
+\patchBothAmsMathEnvironmentsForLineno{flalign}%
+\patchBothAmsMathEnvironmentsForLineno{alignat}%
+\patchBothAmsMathEnvironmentsForLineno{gather}%
+\patchBothAmsMathEnvironmentsForLineno{multline}%
+}
+
+% commented for arXiv submission
+%\usepackage{fancyhdr}
+% \fancyhead[R]{ submission intended as Article in Discoveries section of \emph{MBE} }
 
@@
 
-\fancyhead[R]{ submission intended as Article in Discoveries section of \emph{MBE} }
-
 \title{Joint maximum-likelihood of phylogenies and ancestral states is not consistent}
@@
 \author[1]{David A. Shaw}
-\author[1]{Frederick A. Matsen IV\thanks{Corresponding author, email: \url{matsen@fredhutch.org}}}
+\author[1]{Frederick A. Matsen IV\thanks{Corresponding author. Email: \url{matsen@fredhutch.org}}}
 \affil[1]{Computational Biology Program, Fred Hutchinson Cancer Research Center\\ Seattle, WA, USA}
@@
 \maketitle
-\thispagestyle{fancy}
+% commented for arXiv submission
+%\thispagestyle{fancy}
 
@@
 Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
-$$
+\[
 \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t).
-$$
+\]
 We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
@@
 There exist $C_0(x^*, y^*), C_1(x^*, y^*),$ and a set of $0 < x^*, y^* < 1$ such that
-$$
+\[
 \max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le C_0(x^*, y^*),
-$$
-$$
+\]
+\[
 C_1(x^*, y^*) \le \max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t)
-$$
+\]
 with $C_0(x^*, y^*) < C_1(x^*, y^*)$.
@@
 
+%EM Line numbering has a huge gap here. Could you look around and see if you see anything about this on the web?
 We now consider the problem of branch length estimation on the correct tree using joint estimation.
@@
 That is, if we estimate $x$ and $y$ correctly, then, for
-$$
+\[
 \hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
-$$
+\]
 there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
 In the general case, we do the same but with
-$$
+\[
 (\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\})
-$$
+\]
 to find a region where the inferred values do not converge to the generating values.
@@
 Let
-$$
+\[
 \beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
-$$
-$$
+\]
+\[
 \gamma := \gamma(x^*, y^*) = 4x^*y^*.
-$$
+\]
 The maximum likelihood value $\hat{w}$ is equal to 1 if
-$$
+\[
 %\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
 -\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0
-$$
+\]
 and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
@@
 For
-$$
+\[
 \beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
-$$
-$$
+\]
+\[
 \gamma := \gamma(x^*, y^*),
-$$
+\]
 bounds
-$$
+\[
 \gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
-$$
-$$
+\]
+\[
 \gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
-$$
+\]
 and
-$$
+\[
 \beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}),
-$$
+\]
 the maximum likelihood value $\hat{w}=1$ when
-$$
+\[
 -\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
-$$
+\]
 \end{restatable}
@@
 For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
-$$
+\[
 x^*-x_{L} \le \hat{x} \le x^*+x_{U}
-$$
+\]
 and
-$$
+\[
 y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
-$$
+\]
 Then $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.

### 2018-05-02 versions/response.v1.txt
dave's responses

@@
 %EM I think we could move this to the appendix to throw this reviewer a bone.
+%das: does this reviewer mean both the general case and the empirical case or just the empirical case?
+%Their review makes me think they take the empirical case the be the "general" case.
+%You think we should move both analytic cases to the appendix or just the one?
 
@@
 %EM David-- does it seem like we could extend these results to bigger trees containing a 4 taxon tree?
+%das: I don't see why not, though I haven't thought it through very much.
+%It could take a lot of work or not much work at all.
+%Do you know of any other mathematical phylogenetics papers that do something like this?
+%Like induction on trees?
+%I can think a bit and get back to you...
 
@@
 TODO: not sure what this is?
+%das: this reviewer was confused about the phrase "generating probabilities," thinking we were generating the probabilities and not that they were the probabilities of generating a particular data configuration.
+%I changed:
+%Fidelities have useful algebraic properties, and generating probabilities using the Hadamard transform have an especially simple form (see \eqref{eq:hadamard_probability} in the Appendix).
+%to:
+%Fidelities have useful algebraic properties.
+%As data becomes plentiful, we use the Hadamard transform (see \eqref{eq:hadamard_probability} in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form.
+
+We have added the definition of a "generating probability" to ease confusion in this sentence.
 
@@
 TODO: fill in
+%das: This reviewer was confused about the bold eta, which was an ordered list of sets eta_1, eta_2, ...
+%I noticed we never actually used this ordered list so got rid of it and wrote:
+%Thus given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies...
+
+We removed the ordered list of sets as it did not add to any understanding of the concept and now refer to each of these sets and their relevance directly.
 
@@
 TODO: just say tough nuts?
+%das: I defined "generating probability" earlier, but, I mean, come on, a reference? We give a section of a chapter! I'd vote for tough nuts:
+%We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
+
+We have updated the text to explicitly define "generating probability." Background for equation (8) can be found in Section 8.6 in the Semple and Steel text.
+This is referenced at the beginning of the section.
 
@@
 TODO we should make some changes to at least say that we've further clarified the implications of our results.
+%das: can you clarify this TODO?
+%Do you mean we should add more text about the bias in Fig. S3, or we should quote Richard more and show how we think things might still be bad?
 The TreeTime paper does present joint optimization as an approximate approach, and described that it is for short branch lengths.
@@
 %EM Also, I think you said that he says that the varying branch length thing is an open question-- it would be great to quote that.
+%das: below are some notes and quotes. I will work on paring these down and adding some to the text, maybe with a little more diplomacy.
+%The equal branch length example occurs in Example 2, starting halfway down the second column of page 352 in Goldman (1990) and continuing until the top of the second column on page 354.
+%Here he only talks about how parsimony is inconsistent because it may result in equally parsimonious trees (this is the last paragraph of the example section).
+%It's not until the bottom of page 356 that he punts the inconsistency to Felsenstein and Hendy/Penny, and the bottom of page 357 where he wonders how small branch lengths have to be for parsimony to be consistent in general.
+
+---
+
+Goldman (1990) concerns estimating both structural (e.g., topology, branch lengths) and incidental (e.g., ancestral states) parameters simultaneously, though the only worked example is in the case of a topology with equal branch lengths.
+Goldman shows that for certain special cases, "maximum relative likelihood," i.e., performing maximum likelihood over both topology and ancestral states, is equivalent to parsimony, and thus inconsistent.
+He says
+"Consistent estimates of incidental parameters are unobtainable, as it is not possible to increase indefinitely the number of observations (data) pertinent to each."
+However, he also says "results may be unpredictable" and "estimates of the structural parameters (e.g., the tree T) may also be inconsistent."
+He does not discuss the inconsistency of maximum relative likelihood in general, and proceeds to show that parsimony analysis is a specific instance of maximum relative likelihood, implying maximum relative likelihood of topology given equal branch lengths cannot be consistent in general, but stops there.
+We show cases where the estimates of the topology will be inconsistent if we are estimating both topology and branch lengths, and, moreover, even if we know the correct topology, we show cases where branch lengths will be inconsistent.
+Finally, just prior to his conclusion, he discusses when parsimony will give the same answer as ML, concluding that the question is ill-posed since parsimony estimates different parameters than ML, i.e., it assumes equal branch lengths.
+He states that if the expected amount of change along a branch is "very small" then parsimony will recover the same tree as ML (under a Poisson process model).
+We are able to quantify how small is "very small" in the case of general joint ML---estimating topology and branch lengths together---and not just parsimony in particular.
+
+quote:
+"The second approach is to estimate the random variables as though they were parameters of the model.
+However, in this case they will be incidental parameters: as the amount of data (i.e., the number of characters) increases, the number of parameters also increases.
+For each additional character (labeled a, for instance), there are additional data xa, consisting of the states for all n species, and additional parameters Ya, consisting of the states for all the internal nodes {NI: I = 1,2, .
+.
+, m}.
+Although this method is the only one possible if we wish to estimate the internal node character states, the results may be unpredictable.
+As noted above, these estimates cannot be consistent.
+Further, the estimates of the structural parameters (e.g., the tree T) may also be inconsistent.
+Indeed, under certain models, this approach renders inferences virtually impossible (Cavalli-Sforza et al., 1964; Cavalli-Sforza and Edwards, 1966; Kendall and Stuart, 1979).
+Examples of both approaches are investigated below."
+
+---
 

### 2018-05-02 appendix.tex
everybody hates farris

@@
 \end{tabular}
-\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
+\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
 All values multiplied by $1/8$.}
@@
 We follow with an expository example computing these probabilities and likelihoods.
-Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a---this is commonly known as the ``Farris zone'' topology.
+Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
 The set of all possible character assignments is
@@
 Using branch fidelities, the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
-See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the Farris tree.
-Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} (for the Farris tree) and~\ref{tab:fels_likelihoods} (for the Felsenstein tree).
+See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the InvFels tree.
+Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} (for the InvFels tree) and~\ref{tab:fels_likelihoods} (for the Felsenstein tree).
 Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}.
@@
 \caption{
-    Example likelihood computations on the Farris tree $\tau_1$ for fidelities $x$, $y$, and $w$.
+    Example likelihood computations on the InvFels tree $\tau_1$ for fidelities $x$, $y$, and $w$.
     Edges labeled by the probability of substitution along that edge.
@@
 \end{tabular}
-\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the Farris tree $\tau_1$.
+\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
 All values multiplied by $1/32$.}
@@
 \begin{tabular}{|l|ll|}
-    \multicolumn{3}{c}{Farris tree ($\tau=\tau_1$)}\\
+    \multicolumn{3}{c}{InvFels tree ($\tau=\tau_1$)}\\
     \hline
@@
 \end{tabular}
-\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$ after maximizing over ancestral states on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$.
+\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$ after maximizing over ancestral states on the InvFels tree $\tau_1$ and the Felsenstein tree $\tau_2$.
 All values multiplied by $1/32$.
@@
 
-Consider the Farris tree with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
-We now show that, in the case of the Farris tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch parameters to be equal and the bottom two branch parameters to be equal during inference obtains the same maximum likelihood estimate as in the case of arbitrary branch parameters.
-Using the Hadamard transform, we calculate the generating probabilities on the Farris tree.
+Consider the InvFels tree with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
+We now show that, in the case of the InvFels tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch parameters to be equal and the bottom two branch parameters to be equal during inference obtains the same maximum likelihood estimate as in the case of arbitrary branch parameters.
+Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
 For site split $\emptyset$,
@@
 Assume $\tau^*=\tau_1$.
-The Farris tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see the upper table of Table~\ref{tab:likelihoods}).
+The InvFels tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see the upper table of Table~\ref{tab:likelihoods}).
 
@@
 We next construct a lower bound on the Felsenstein tree partial likelihood.
-In the Felsenstein case, there are many more optimal internal states depending on branch lengths than in the Farris case (see the lower table of Table~\ref{tab:likelihoods}).
+In the Felsenstein case, there are many more optimal internal states depending on branch lengths than in the InvFels case (see the lower table of Table~\ref{tab:likelihoods}).
 We first bound the entire likelihood below and then proceed with joint inference.
@@
 \begin{proof}
-Recalling Table~\ref{tab:likelihoods} for the Farris tree $\tau = \tau_1$, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood over $(x,y,w)$.
+Recalling Table~\ref{tab:likelihoods} for the InvFels tree $\tau = \tau_1$, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood over $(x,y,w)$.
 If $x=x^*$ and $y=y^*$ are fixed, by collapsing like terms and ignoring terms with only $x^*$ and $y^*$, the log-likelihood in \eqref{eq:farris_likelihood} as a function of $w$ is
@@
 As is apparent from Table~\ref{tab:sitepatprob}, if $0 < x^*, y^* < 1$ and we try to evaluate the candidate $\hat{x}=\hat{y}=\hat{w}=1$ we have a likelihood of exactly zero.
-Worse still, in Table~\ref{tab:likelihoods} for the Farris tree, in the cases of $\{2\}, \{1,2\}, \{2,3\},$ and $\{1,2,3\}$, any attempt to evaluate $\hat{y}=1$ will similarly yield a zero likelihood.
+Worse still, in Table~\ref{tab:likelihoods} for the InvFels tree, in the cases of $\{2\}, \{1,2\}, \{2,3\},$ and $\{1,2,3\}$, any attempt to evaluate $\hat{y}=1$ will similarly yield a zero likelihood.
 In these cases of zero likelihood, methods that evaluate many candidate parameter values can fail to converge when near these boundaries.

### 2018-05-02 joint_inf.tex
everybody hates farris

@@
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model with data being generated on the four-taxon ``Farris zone'' \citep{Siddall1998-hq} tree, and we construct bounds on the joint objective function to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
+To do so, we use a binary symmetric model with data being generated on the four-taxon ``inverse Felsenstein (InvFels)'' tree---also called the ``Farris'' tree \citep{Siddall1998-hq, Felsenstein2004}---and we construct bounds on the joint objective function to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
 We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.
@@
 
-We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``Farris'' and ``Felsenstein'' trees (Fig.~\ref{fig:farris-fels-top}).
+We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``InvFels'' and ``Felsenstein'' trees (Fig.~\ref{fig:farris-fels-top}).
 In the standard configuration of each of these trees, the interior branch length parameters are equal to the bottom two parameters.
-We show in the Appendix that, in the case of the Farris tree, performing inference fixing both the top two branch parameters to be equal and the bottom two branch parameters to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch parameters.
+We show in the Appendix that, in the case of the InvFels tree, performing inference fixing both the top two branch parameters to be equal and the bottom two branch parameters to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch parameters.
 
@@
 \includegraphics[width=\textwidth]{farris_blank}
-\caption[short]{Farris tree $\tau_1$}
+\caption[short]{InvFels tree $\tau_1$}
 \end{subfigure}
@@
 
-To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the Farris topology (Fig.~\ref{fig:farris-fels-top}a).
-We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology ($\tau_2$) than for the true, generating Farris topology ($\tau_1$).
-To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the Farris topology as a function of $x^*$ and $y^*$ and, similarly, a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
-When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the Farris case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
+To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the InvFels topology (Fig.~\ref{fig:farris-fels-top}a).
+We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology ($\tau_2$) than for the true, generating InvFels topology ($\tau_1$).
+To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the InvFels topology as a function of $x^*$ and $y^*$ and, similarly, a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
+When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the InvFels case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
 \begin{restatable}{theorem}{topoInconsist}
@@
 
+%das: TODO, rerun this figure through inkscape with updated title
 \begin{figure}
@@
 \caption{
-    An analytically-derived region of topological inconsistency in terms of fidelities for ``perfect'' data generated on the Farris topology (Fig.~\ref{fig:farris-fels-top}) with $w^* = y^*$.
+    An analytically-derived region of topological inconsistency in terms of fidelities for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $w^* = y^*$.
     Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
@@
 As described above, we use the notion of fidelities to analyze what happens as changes occur along each branch as opposed to the standard branch length formulation.
-We analyze two settings on the Farris tree, corresponding to whether some branch fidelities are fixed at their true values or not.
-As above, assume that data is generated from the Farris tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
+We analyze two settings on the InvFels tree, corresponding to whether some branch fidelities are fixed at their true values or not.
+As above, assume that data is generated from the InvFels tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
 In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
@@
 Previous work in phylogenetics has developed consistency counterexamples using the same four-taxon topologies used here \citep{Felsenstein1978-rr}.
-In this previous work, when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
-This is the converse of what we have shown for joint inference, where the Felsenstein topology is more likely than the Farris topology.
+In this previous work, when simulating data under the Felsenstein topology $\tau_2$, as the number of observations increases, the InvFels topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
+This is the converse of what we have shown for joint inference, where the Felsenstein topology is more likely than the InvFels topology.
 Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
 This is not the case here; for our case, the inconsistency generally occurs when all branches are long, and has more to do with the choices of the form of the likelihood than from the interplay of long and short branches.
-Difficulties in phylogenetic estimation when generating data on the ``Farris zone'' tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.
+Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.
 

### 2018-05-02 joint_inf_plot.py

@@ -132,9 +132,9 @@ def parse_args():
 # ~~~~~~~~~
 # global functions for exact likelihood computation
 
-def P_FARRIS(theta0, site_pattern):
+def P_INVFELS(theta0, site_pattern):
     """
-    Site pattern frequencies for Farris tree
+    Site pattern frequencies for inverse Felsenstein tree
     """
     # fifth branch will always be positive
     theta = [-t if s=='1' else t \
@@ -148,9 +148,9 @@ def P_FARRIS(theta0, site_pattern):
          theta[2]*theta[3]*theta[4] + \
          theta[0]*theta[1]*theta[2]*theta[3])
 
-def L_FARRIS(theta, site_pattern, joint=True):
+def L_INVFELS(theta, site_pattern, joint=True):
     """
-    Likelihood for Farris zone topology
+    Likelihood for inverse Felsenstein topology
     """
     s1 = reduce(
         mul,
@@ -179,7 +179,7 @@ def L_FARRIS(theta, site_pattern, joint=True):
     else:
         return 0.03125 * sum([s1, s2, s3, s4])
 
-def FARRIS_LIKELIHOOD(x, y, xp, yp, wp, joint=True):
+def INVFELS_LIKELIHOOD(x, y, xp, yp, wp, joint=True):
     """
     Total likelihood as function of marginal likelihood
     """
@@ -187,14 +187,14 @@ def FARRIS_LIKELIHOOD(x, y, xp, yp, wp, joint=True):
     theta0 = [x, y, x, y, y]
     theta1 = [xp, yp, xp, yp, wp]
     for site_pattern in ALL_PATTERNS:
-        if not np.isclose(P_FARRIS(theta0, site_pattern), 0):
-            if np.isclose(P_FARRIS(theta1, site_pattern), 0) or np.isclose(L_FARRIS(theta1, site_pattern, joint=joint), 0):
+        if not np.isclose(P_INVFELS(theta0, site_pattern), 0):
+            if np.isclose(P_INVFELS(theta1, site_pattern), 0) or np.isclose(L_INVFELS(theta1, site_pattern, joint=joint), 0):
                 likelihood = -np.inf
                 break
 
-            likelihood += P_FARRIS(theta0, site_pattern) * \
-                (np.log(L_FARRIS(theta1, site_pattern, joint=joint)) + \
[…]

### 2018-05-04 appendix.tex
answering EMs

@@
 For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
-We define the site split $\siteSplit$ for a $\alignmentColumn_i$ as simply $\alignmentColumn_i$ if the label $\nSiteRows$ is not in $\alignmentColumn_i$, and as its complement otherwise.
+For the binary symmetric model $\alphabet=\{0,1\}$.
+Define the complement of $\alignmentColumn$ as $\overline{\alignmentColumn}$, and let $\alignmentColumn_{i,i'}$ be the label of the $i'$th tip in the $i$th alignment column.
+We define the site split $\siteSplit$ for a $\alignmentColumn_i$ as the set of tips labeled with $1$ in $\alignmentColumn_i$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of tips labeled with $1$ in $\overline{\alignmentColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
 Taking such a complement simplifies but does not change the result of likelihood computation because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model.
+We define an ``ancestral state split'' in a similar way, where the ancestral state split $\ancestralSplit$ for an internal node $\ancestralStateColumn_i$ is the set of internal nodes labeled with $1$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of tips labeled with $1$ in $\overline{\ancestralStateColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$
+In other words, both site splits and ancestral state splits are defined by whether the $\nSiteRows$th element of $\alignmentColumn_i$ is labeled as $1$.
 %EM "ancestral state splits" are an important item in the development but don't get the same level of definition as "site splits". Let's define them here.
+%das: done
 
@@
 %EM I think that we should either use "ancestral splits" everywhere, or "ancestral state splits."
-Similarly we enumerate ancestral splits $\ancestralSplit_k$ of which there are $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$ in total.
+%das: yah, agreed
+Similarly we enumerate ancestral state splits $\ancestralSplit_k$ of which there are $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$ in total.
 
@@
 \begin{definition}
-Define the complement of $\alignmentColumn$ as $\overline{\alignmentColumn}$.
 Let the mapping from site patterns to site splits
@@
     \begin{array}{ll}
-        \alignmentColumn  & \mbox{if } \alignmentColumn\in\mathcal{P}(\siteSplitSet) \\
-        \overline{\alignmentColumn}  & \mbox{if } \alignmentColumn\notin\mathcal{P}(\siteSplitSet)
+        \{i'\in\{1,\ldots,\nSiteRows\}: \alignmentColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}\neq 1,\\
+        \{i'\in\{1,\ldots,\nSiteRows\}: \overline{\alignmentColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1,
     \end{array}
@@
 \[
-\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
+\ancestralToSplit:\alphabet^\nSiteRows\times\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet)
 \]
@@
     \begin{array}{ll}
-        \ancestralStateColumn  & \mbox{if } \alignmentColumn\in\mathcal{P}(\siteSplitSet) \\
-        \overline{\ancestralStateColumn}  & \mbox{if } \alignmentColumn\notin\mathcal{P}(\siteSplitSet).
+        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \ancestralStateColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}\neq 1,\\
+        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \overline{\ancestralStateColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1.
     \end{array}
@@
 \]
-that takes corresponding realizations $\siteSplit_j$ for some $j$, and
+and
 \[
@@
 \end{definition}
-The mapping $\patternToSplit$ operates by returning the complement of a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if it is not in $\mathcal{P}(\siteSplitSet)$.
-The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.
+The mapping $\patternToSplit$ operates by returning the tips labeled as $1$ in a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if the set of tips labeled $1$ is not in $\mathcal{P}(\siteSplitSet)$.
+The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if the set of tips labeled $1$ in $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the set of tips labeled $1$ in $\ancestralStateColumn$; otherwise, the set of tips labeled $1$ in $\overline{\alignmentColumn}$ necessarily is in $\siteSplitSet$.
 
@@
 \end{align*}
-In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
+In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
 %EM Wait, does this establish surjectivity? We need to be able to get all values of P(H).
-For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it maps all possible values from $\alphabet^\nAncestralStateRows$ to an element in $\mathcal{P}(\ancestralSplitSet)$, and from this we have
+%das: ah crap, I don't know why I wrote the above. I changed a bit of the notation (sorry!) but I hope this will be more clear?
+For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
+This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,i'}\neq 1$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
+The same can be done for the case of $\alignmentColumn_{i,i'}=1$, implying $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective.
+From this we have
 \begin{align*}
@@
 For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
-A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
+A maximum occurs at one of possibly several ancestral state splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
 As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch, with $t=\{p,p,p,p,p\}$.
-The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
+The probabilities of observing ancestral state splits for $\siteSplit_1=\emptyset$ are
 \[
@@
 
-To compute the likelihood of observing a set of data, we need $\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
+To compute the likelihood of observing a set of data, we need $\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
 Using branch fidelities, the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
@@
     Edges labeled by the probability of substitution along that edge.
-    In (a), we compute the product to obtain $\Pr(\ancestralStateColumn=\emptyset\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1+w)/32$.
-    In (b), the same process yields $\Pr(\ancestralStateColumn=\{1\}\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1-w)/32$.
+    In (a), we compute the product to obtain $\Pr(\ancestralSplitRV=\emptyset\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1+w)/32$.
+    In (b), the same process yields $\Pr(\ancestralSplitRV=\{1\}\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1-w)/32$.
 }
@@
 \begin{tabular}{|l|ll|}
-\multicolumn{3}{c}{$\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
+\multicolumn{3}{c}{$\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_1, t)$}\\
 \hline
@@
 \begin{tabular}{|l|ll|}
-\multicolumn{3}{c}{$\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_2,t)$}\\
+\multicolumn{3}{c}{$\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_2, t)$}\\
 \hline
@@
 \end{table}
+%TODO: the caption of this overlaps with the page number---will this be a problem or will MBE handle it?
 
@@
 %EM There's not good notational overlap between Table S2 and the definition of partial log-likelihood. Could you fix that?
+%das: I think fixed, or did you mean something in addition to changing h-->\Xi and y-->\Psi?
 For the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$, we can compute the partial log-likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ as in Table~\ref{tab:farris_likelihoods}.
@@
 %EM Shouldn't this read "obtaining an upper bound"?
-Our plan is to simplify the likelihood to remove log-of-quadratic terms of $x,y,$ and $w$, obtaining a likelihood that has a closed-form maximum in each variable.
+%das: yes
+Our plan is to simplify the likelihood to remove log-of-quadratic terms of $x,y,$ and $w$, obtaining an upper bound on the likelihood that has a closed-form maximum in each variable.
 To bound the generating probabilities, we use the facts that, for $x,y\in[0,1]$,
@@
 %+ 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
+%das: I don't think so---the first two lines starting at the equals sign are bringing together the first four lines starting above at the <= sign.
+%Should I remove this intermediate equation?
+%I thought it would help clarify but maybe not?
 + (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
@@
 + p_{13}\cdot\log(2p_{13}) \\
-&\qquad := C^{1}_{\tau_1,\tau_1}(x^*, y^*).
+&\qquad := C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*).
 \end{split}
-\label{eq:farris-upper-bound}
+\label{eq:farris-upper-bound-1}
 \end{align}
@@
 The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications, as we now describe.
-The upper bound in \eqref{eq:farris-upper-bound} is for the ancestral state split $\{1\}$.
+The upper bound $C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*)$ in \eqref{eq:farris-upper-bound-1} is for the ancestral state split $\{1\}$.
 %EM Let's be more specific, giving an equation number, rather than just saying "the upper bound" here.
-For the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.
+%das: is this better? there were a few other typos, so I changed "site split" to "ancestral state split" where needed and updated the constant definitions
+For the ancestral state split $\emptyset$ we can derive an upper bound $C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*)$ in \eqref{eq:farris-upper-bound-empty} in a similar fashion, and $C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*)$ will be the same as $C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*)$ except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.
 In this case, all terms involving $w$ simplify to $\log(1+w)$, which is maximized at $\hat{w}=1$.
@@
 \end{equation}
-for site split $\emptyset$ and
+for ancestral state split $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$, yielding the upper bound
+\begin{align}
+\begin{split}
+&    \ell_{\tau_1,t^*}(\tau_1, t) \\
+&\qquad\le      p_{\emptyset}  \cdot\log(2)
++ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
++ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
+&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
++ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
++ \log(2) \\
+&\qquad := C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*).
+\end{split}
+\label{eq:farris-upper-bound-empty}
+\end{align}
+
+For ancestral state split $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$, we have constants
 \begin{equation}
@@
 \end{equation}
-for site split $\{1,2\}$.
-To bound the case of ancestral state split $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
-\begin{align*}
-&    \ell_{\tau_1,t^*}(\tau_1, t) \\
-&\qquad\le      p_{\emptyset}  \cdot\log(2)
-+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
-+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
-&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
-+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
-+ \log(2) \\
-&\qquad := C^{2}_{\tau_1,\tau_1}(x^*, y^*),
-\end{align*}
-and, for the case of ancestral state split $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
-\begin{align*}
+and the upper bound
+\begin{align}
+\begin{split}
 &    \ell_{\tau_1,t^*}(\tau_1, t) \\
@@
 + \log(2) \\
-&\qquad := C^{3}_{\tau_1,\tau_1}(x^*, y^*).
-\end{align*}
+&\qquad := C^{\{1,2\}}_{\tau_1,\tau_1}(x^*, y^*).
+\end{split}
+\label{eq:farris-upper-bound-12}
+\end{align}
 
@@
 are both between zero and one.
-Using a similar argument as that where $\ancestralSplitPartition_j(\tau_1, t) = \{1,3\}$, we only need to show $a_1' \ge a_2'$, which is true if $3-2p_{13} \ge 3\tilde{p}+2p_{13}$.
+Using a similar argument as that where $\ancestralSplitPartition_j(\tau_1, t) = \{1\}$, we only need to show $a_1' \ge a_2'$, which is true if $3-2p_{13} \ge 3\tilde{p}+2p_{13}$.
 Some rearranging and the fact that $0 \le \tilde{p} \le 1/2$ shows this is equivalent to showing $p_{13} \le 3/8$.
 %EM Using what about t^*?
-Using Table~\ref{tab:sitepatprob} and $t^*$,
+%das: updated
+Using Table~\ref{tab:sitepatprob} while setting $t=t^*=\{x^*,y^*,x^*,y^*,y^*\}$,
 \[
@@
 \[
-C_0(x^*, y^*) := \max\left(C^{1}_{\tau_1,\tau_1}(x^*, y^*), C^{2}_{\tau_1,\tau_1}(x^*, y^*), C^{3}_{\tau_1,\tau_1}(x^*, y^*)\right).
+C_0(x^*, y^*) := \max\left(C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*), C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*), C^{\{1,2\}}_{\tau_1,\tau_1}(x^*, y^*)\right).
 \]

### 2018-05-07 appendix.tex
unique maximum! and a little rearranging

@@
 for any site split $\siteSplit_j$.
-Table~\ref{tab:sitepatprob} contains calculations of site pattern probabilities for our two topologies (Fig.~\ref{fig:farris-fels-top}).
+Table~\ref{tab:gen-sitepatprob} contains calculations of site pattern probabilities for the InvFels topology with arbitrary branch lengths and Table~\ref{tab:sitepatprob} site pattern probabilities for the InvFels topology in Fig.~\ref{fig:farris-fels-top}a.
 
@@
 \centering
-\begin{tabular}{|l|l|l|l|}
-    \hline
-$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
-    \hline
-    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
-    $\{1\}$     & $p_{1}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
-    $\{2\}$     & $p_{2}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
-    $\{3\}$     & $p_{3}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
-    $\{1,2\}$   & $p_{12}$   &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
-    $\{1,3\}$   & $p_{13}$   &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
-    $\{2,3\}$   & $p_{23}$   &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
-    $\{1,2,3\}$ & $p_{123}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
+\begin{tabular}{|l|l|l|}
+    \hline
+$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\})$\\
+    \hline
+    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
+    $\{1\}$     & $p_{\{1\}}$       &$1 - x_1x_2 +  y_1y_2 +  w[- x_1 + x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
+    $\{2\}$     & $p_{\{2\}}$       &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][- y_1 + y_2] - x_1y_1x_2y_2$\\
+    $\{3\}$     & $p_{\{3\}}$       &$1 - x_1x_2 +  y_1y_2 +  w[  x_1 - x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
+    $\{1,2\}$   & $p_{\{1,2\}}$     &$1 - x_1x_2 -  y_1y_2 +  w[- x_1 + x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
+    $\{1,3\}$   & $p_{\{1,3\}}$     &$1 + x_1x_2 +  y_1y_2 +  w[- x_1 - x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
+    $\{2,3\}$   & $p_{\{2,3\}}$     &$1 - x_1x_2 -  y_1y_2 +  w[  x_1 - x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
+    $\{1,2,3\}$ & $p_{\{1,2,3\}}$   &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][  y_1 - y_2] - x_1y_1x_2y_2$\\
     \hline
 \end{tabular}
-\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
+\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
+All values multiplied by $1/8$.}
+\label{tab:gen-sitepatprob}
+\end{table}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{|l|l|l|}
+    \hline
+$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x,y,x,y,w\})$\\
+    \hline
+    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xyw+x^2y^2$\\
+    $\{1\}$     & $p_{1}$   &$1-x^2+y^2-x^2y^2$\\
+    $\{2\}$     & $p_{2}$   &$1+x^2-y^2-x^2y^2$\\
+    $\{3\}$     & $p_{3}$   &$1-x^2+y^2-x^2y^2$\\
+    $\{1,2\}$   & $p_{12}$   &$1-x^2-y^2+x^2y^2$\\
+    $\{1,3\}$   & $p_{13}$   &$1+x^2+y^2-4xyw+x^2y^2$\\
+    $\{2,3\}$   & $p_{23}$   &$1-x^2-y^2+x^2y^2$\\
+    $\{1,2,3\}$ & $p_{123}$   &$1+x^2-y^2-x^2y^2$\\
+    \hline
+\end{tabular}
+\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x,y,x,y,w\}$ obtained using the Hadamard transform.
 All values multiplied by $1/8$.}
@@
 %EM I'm thinking about this wrong for some reason: the sum of two things, each of which has a unique maximum, does not necessarily have a unique maximum.
-By Gibbs's inequality, the entropy term $\shannonDivergence_{\tau^*,t^*}(\tau^*,t)$ has a unique maximum over $t$---namely $t^*$.
+%das: I added the table of general generating probabilities, but it's pretty straightforward I think to see below the second derivative of the likelihood will be negative on the whole interval, implying we have a unique maximum.
+%I think we're good.
+%Let me know if I should say anything about the continuity of the likelihood, but I think it's obvious this gets us to where we need to be?
+Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$, and the map $g(x_1) \mapsto -g(x_1)\log g(x_1)$.
+We restrict our attention to $x_1,y_1,x_2,y_2,w\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
+By straightforward calculus, the second derivative of this map is
+\[
+-g''(x_1)(1+\log g(x_1)) - \frac{g'(x_1)^2}{g(x_1)}.
+\]
+Using each functional form in Table~\ref{tab:gen-sitepatprob} we see in all cases $g''(x_1)=0$ and $g(x_1) > 0$ for $x_1\in(0,1).$
+This implies the above second derivative is negative for each component of $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t})$, and the second derivative of $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t})$ itself is negative.
 %EM There's not good notational overlap between Table S2 and the definition of partial log-likelihood. Could you fix that?
 %das: I think fixed, or did you mean something in addition to changing h-->\Xi and y-->\Psi?
-For the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$, we can compute the partial log-likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ as in Table~\ref{tab:farris_likelihoods}.
-Clearly this partial log-likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in[0,1]$ and thus has a unique maximum.
-Since both summands of the likelihood are concave on $x_1\in[0,1]$, the likelihood is as well, and there exists a single $x_1\in[0,1]$ maximizing the likelihood.
+
+We can compute the partial log-likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ as in Table~\ref{tab:farris_likelihoods}.
+Clearly this partial log-likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in(0,1)$ and thus has a unique maximum.
+Since both summands of the likelihood are concave on $x_1\in(0,1)$, the likelihood is as well, and there exists a single $x_1\in(0,1)$ maximizing the likelihood.
 Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
@@
 \begin{proof}
-In general,
-\[
-\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \le
-    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
-    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
-\]
-is an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
-\[
-\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
-\]
-and
-\[
-\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t).
-\]
-Similarly,
-\[
-\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \ge
-    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
-    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
-\]
-is a lower bound.
-
 Assume $\tau^*=\tau_1$.
@@
 We next construct a lower bound on the Felsenstein tree partial likelihood.
+%das: we never actually use the Gibbs inequality trick for the upper bound---just some log-of-polynomial bounds
+In general, taking
+\[
+\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t).
+\]
+we see
+\[
+\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \ge
+    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
+    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
+\]
+is a lower bound for the joint maximum of \eqref{eq:log_likelihood_simplified}.
 In the Felsenstein case, there are many more optimal internal states depending on branch lengths than in the InvFels case (see the lower table of Table~\ref{tab:likelihoods}).

### 2018-05-08 appendix.tex
typos, redundancies, and moved example earlier

@@
 Taking such a complement simplifies but does not change the result of likelihood computation because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model.
-We define an ``ancestral state split'' in a similar way, where the ancestral state split $\ancestralSplit$ for an internal node $\ancestralStateColumn_i$ is the set of internal nodes labeled with $1$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of tips labeled with $1$ in $\overline{\ancestralStateColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$
+
+For a fixed topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
+In this case we cannot use the same complement trick as before: the probability of observing an ancestral state split conditional on a site split is not invariant to taking its complement.
+We define an ``ancestral state split'' in a similar way to a site split, where the ancestral state split $\ancestralSplit$ for an internal node $\ancestralStateColumn_i$ is the set of internal nodes labeled with $1$ if the $\nSiteRows$th tip---\emph{not} internal node---is not labeled with $1$, and as the set of internal nodes labeled with $1$ in $\overline{\ancestralStateColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
 In other words, both site splits and ancestral state splits are defined by whether the $\nSiteRows$th element of $\alignmentColumn_i$ is labeled as $1$.
@@
 %das: done
-
-For a fixed topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
-In this case we cannot use the same complement trick as before: the probability of observing an ancestral state split conditional on a site split is not invariant to taking its complement.
 
@@
     \begin{array}{ll}
-        \{i'\in\{1,\ldots,\nSiteRows\}: \alignmentColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}\neq 1,\\
-        \{i'\in\{1,\ldots,\nSiteRows\}: \overline{\alignmentColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1,
+        \{i'\in\{1,\ldots,\nSiteRows-1\}: \alignmentColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}\neq 1,\\
+        \{i'\in\{1,\ldots,\nSiteRows-1\}: \overline{\alignmentColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1,
     \end{array}
@@
 \]
-Then, given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
+Then, given a site pattern--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, define the random variables
 \[
@@
 \[
-\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
-\]
-for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
+\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV).
+\]
 \end{definition}
 The mapping $\patternToSplit$ operates by returning the tips labeled as $1$ in a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if the set of tips labeled $1$ is not in $\mathcal{P}(\siteSplitSet)$.
-The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if the set of tips labeled $1$ in $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the set of tips labeled $1$ in $\ancestralStateColumn$; otherwise, the set of tips labeled $1$ in $\overline{\alignmentColumn}$ necessarily is in $\siteSplitSet$.
+The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if the set of tips labeled $1$ in $\alignmentColumn$ is in $\mathcal{P}(\siteSplitSet)$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the set of tips labeled $1$ in $\ancestralStateColumn$; otherwise, the set of tips labeled $1$ in $\overline{\alignmentColumn}$ necessarily is in $\mathcal{P}(\siteSplitSet)$.
 
@@
 \end{align*}
-where $\overline{Y}$ is the complement of the site-patterned random variable $Y$ and so has the same distribution as $Y$.
+where $\overline{Y}$ is the complement of the site pattern--valued random variable $Y$ and has the same distribution as $Y$.
 Note
@@
 \end{align*}
-In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
+In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subseteq\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, i.e., $\ancestralSplitPartition_j(\tau, t)$ is a set of sets of most likely internal node labels.
+Here, $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
 %EM Wait, does this establish surjectivity? We need to be able to get all values of P(H).
@@
 For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
-This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,i'}\neq 1$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
-The same can be done for the case of $\alignmentColumn_{i,i'}=1$, implying $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective.
+This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,\nSiteRows}\neq 1$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
+The same can be done for the case of $\alignmentColumn_{i,\nSiteRows}=1$, implying $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective.
 From this we have
@@
 \end{align*}
-for some $j$ depending on $i$.
-Thus, each term in the likelihood can be collapsed into terms relating only to site splits, indexed by $j$,  as opposed to each individual observation, indexed by $i$.
+for some $j$.
+Thus, each term in the likelihood can be collapsed into terms relating only to site splits and ancestral state splits, indexed by $j$, as opposed to individual observations, indexed by $i$.
+
+\subsection*{Example}
+We follow with an expository example computing these probabilities and likelihoods.
+Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
+The set of all possible character assignments is
+\begin{align*}
+\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
+                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}
+\end{align*}
+where each set indicates the tips assigned the character $1$.
+For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
+Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
+The unique site splits, collapsing complements, are
+\begin{align*}
+    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
+& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
+\end{align*}
+Since we identify character complements, we do not consider the additional splits
+\begin{equation*}
+\begin{split}
+& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
+&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
+\end{split}
+\end{equation*}
+the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
+This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
+\[
+\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
+\]
+Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
+The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
+For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
+Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
+We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
+\[
+\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
+\]
+in general.
+
+For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
+A maximum occurs at one of possibly several ancestral state splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
+As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch, with $t=\{p,p,p,p,p\}$.
+The probabilities of observing ancestral state splits for $\siteSplit_1=\emptyset$ are
+\[
+\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
+(1-p)^5,
+\]
+\[
+\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
+\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
+p^3(1-p)^2,
+\]
+\[
+\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
+p^4(1-p).
+\]
+The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
+Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, t)$ we have
+\[
+\Pr(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
+\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
+(1-p)^5.
+\]
+For $\siteSplit_5=\{1,2\}$ we have
+\[
+\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
+\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
+p^2(1-p)^3,
+\]
+\[
+\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
+\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
+p^3(1-p)^2.
+\]
+Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, t)$,
+\[
+\Pr(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
+p^2(1-p)^3.
+\]
 
@@
 
-Thus the likelihood in \eqref{eq:profile_likelihood} can be written as a product over site patterns as opposed to sites:
+The likelihood in \eqref{eq:profile_likelihood} can be written as
 \begin{align}
@@
 \end{align}
-for some $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.
+for $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ and some $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.
 
@@
 for any site split $\siteSplit_j$.
-Table~\ref{tab:gen-sitepatprob} contains calculations of site pattern probabilities for the InvFels topology with arbitrary branch lengths and Table~\ref{tab:sitepatprob} site pattern probabilities for the InvFels topology in Fig.~\ref{fig:farris-fels-top}a.
+Table~\ref{tab:gen-sitepatprob} contains calculations of site split probabilities for the InvFels topology with arbitrary branch lengths and Table~\ref{tab:sitepatprob} site split probabilities for the InvFels topology in Fig.~\ref{fig:farris-fels-top}a.
 
@@
 \end{tabular}
-\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
+\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
 All values multiplied by $1/8$.}
@@
 \end{tabular}
-\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x,y,x,y,w\}$ obtained using the Hadamard transform.
+\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x,y,x,y,w\}$ obtained using the Hadamard transform.
 All values multiplied by $1/8$.}
@@
 
-\subsection*{Example}
-We follow with an expository example computing these probabilities and likelihoods.
-Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
-The set of all possible character assignments is
-\begin{align*}
-\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
-                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
-\end{align*}
-where each set indicates the tips assigned the character $1$.
-For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
-Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
-The unique site splits, collapsing complements, are
-\begin{align*}
-    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
-& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
-\end{align*}
-Since we identify character complements, we do not consider the additional splits
-\begin{equation*}
-\begin{split}
-& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
-&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
-\end{split}
-\end{equation*}
-the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
-This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
-\[
-\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
-\]
-Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
-The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
-For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
-Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
-We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
-\[
-\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
-\]
-in general.
-
-For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
-A maximum occurs at one of possibly several ancestral state splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
-As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch, with $t=\{p,p,p,p,p\}$.
-The probabilities of observing ancestral state splits for $\siteSplit_1=\emptyset$ are
-\[
-\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
-(1-p)^5,
-\]
-\[
-\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
-\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
-p^3(1-p)^2,
-\]
-\[
-\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
-p^4(1-p).
-\]
-The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
-Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, t)$ we have
-\[
-\Pr(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
-\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
-(1-p)^5.
-\]
-For $\siteSplit_5=\{1,2\}$ we have
-\[
-\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
-\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
-p^2(1-p)^3,
-\]
-\[
-\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
-\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
-p^3(1-p)^2.
-\]
-Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, t)$,
-\[
-\Pr(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
-p^2(1-p)^3.
-\]
-
 \subsection*{Likelihood computations}
@@
 See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the InvFels tree.
-Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} (for the InvFels tree) and~\ref{tab:fels_likelihoods} (for the Felsenstein tree).
+Likelihood computations for all site splits and ancestral state splits are in Tables~\ref{tab:farris_likelihoods} (for the InvFels tree) and~\ref{tab:fels_likelihoods} (for the Felsenstein tree).
 Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}.
@@
 \end{tabular}
-\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
+\caption{Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
 All values multiplied by $1/32$.}
@@
 \end{tabular}
-\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the Felsenstein tree $\tau_2$.
+\caption{Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the Felsenstein tree $\tau_2$.
 All values multiplied by $1/32$.}
@@
 \end{tabular}
-\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$ after maximizing over ancestral states on the InvFels tree $\tau_1$ and the Felsenstein tree $\tau_2$.
+\caption{Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$ after maximizing over ancestral states on the InvFels tree $\tau_1$ and the Felsenstein tree $\tau_2$.
 All values multiplied by $1/32$.
@@
                                               & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
-                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
+                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
 \end{align*}
 and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
-All other generating probabilities will differ only in the signs of each term.
+All other generating probabilities will differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
 For example, for site split $\{1\}$ we have
@@
 Using each functional form in Table~\ref{tab:gen-sitepatprob} we see in all cases $g''(x_1)=0$ and $g(x_1) > 0$ for $x_1\in(0,1).$
-This implies the above second derivative is negative for each component of $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t})$, and the second derivative of $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t})$ itself is negative.
+This implies the above second derivative is negative for each component of $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t})$.
 %EM There's not good notational overlap between Table S2 and the definition of partial log-likelihood. Could you fix that?
@@
 
-We can compute the partial log-likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ as in Table~\ref{tab:farris_likelihoods}.
+We can compute the partial log-likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ using Table~\ref{tab:farris_likelihoods}.
 Clearly this partial log-likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in(0,1)$ and thus has a unique maximum.

### 2018-05-09 versions/response.v1.txt
more clarifications

@@
 Please see above for our response.
+%das: I added some stuff below to maybe alleviate some confusion and to show why what this reviewer thinks we should do won't work, but not sure how much is necessary.
+%Maybe it adds to why we didn't do this.
+It is true that the form given above for l(x,y,w) has an explicit and simple maximizer.
+However, the long-run log-likelihood is slightly more complicated than this function, as seen in equation (9).
+For both topologies, we have a number of equations of this form that we would need to maximize in a grid search.
+The function of the form in l(x,y,w) above was our approximation to the upper and lower bounds of the log-likelihood, and was our choice for these bounds precisely because it has an explicit and simple maximizer.
+Doing a grid search over all ancestral states would not be trivial, and we would need to perform numerical optimization for all possible ancestral states to compute the maximum.
 
@@
 TODO: make the argument more clear in the main text.
+%das: Added sentence at L92. Do you think it needs more?
 
@@
 > natural for readers.
+
+%das: @EM how do you feel about just plotting 1-x and 1-y and explaining that this is intuitively more like branch lengths?
 
@@
 TODO: this is changed, but I don't see this being a lot clearer.
+%das: changed, is it better?
 

### 2018-05-09 joint_inf.tex
remove farris from plot
answering EMs
more clarifications

@@
 For a four-taxon tree, define the general branch fidelity parameter $t=\{\theta_1,\theta_2,\theta_3,\theta_4,\theta_5\}$ where fidelities are ordered in the order of the taxa with the internal branch last (Fig.~\ref{fig:farris-fels-top}).
+We show in the Appendix that, for the InvFels tree, the maximum likelihood fidelities require $\theta_1=\theta_3$ and $\theta_2=\theta_4$, so we subsequently fix fidelity parameters to $t=\{x,y,x,y,w\}$.
 
@@
 
-%das: TODO, rerun this figure through inkscape with updated title
 \begin{figure}
@@
 %EM Line numbering has a huge gap here. Could you look around and see if you see anything about this on the web?
+%das: fixed
 We now consider the problem of branch length estimation on the correct tree using joint estimation.
@@
 Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch with this branch length estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
-As an intuition for the theoretical development, if there is no change along a branch with fidelity $\theta$, the likelihood will more likely include a term of $(1+\theta)$ instead of $(1-\theta)$ for this event, resulting in positive bias of branch fidelities.
+We provide the following as an intuition for the theoretical development.
+For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral states.
+For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
+A maximization procedure will more likely introduce a $(1+w)$ term than a $(1-w)$ term, though this will not be guaranteed as the maximum depends on the values of $x$ and $y$.
+This tendency to more likley include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities.
 If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.

### 2018-05-11 appendix.tex
move special case to appendix

@@
 
-\restrictedBranchInconsist*
+\generalBranchInconsist*
 
@@
 Recalling Table~\ref{tab:likelihoods} for the InvFels tree $\tau = \tau_1$, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood over $(x,y,w)$.
-If $x=x^*$ and $y=y^*$ are fixed, by collapsing like terms and ignoring terms with only $x^*$ and $y^*$, the log-likelihood in \eqref{eq:farris_likelihood} as a function of $w$ is
+In general, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
+Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
+By collapsing like terms and ignoring terms without $w$, the log-likelihood in \eqref{eq:farris_likelihood} as a function of $w$ is
 \begin{align*}
     \ell_{\tau_1,t^*}(\tau_1, w)
-    &\sim     p_{\emptyset}  \cdot\log(1+(x^*)^2+(y^*)^2+4x^*y^*w+(x^*)^2(y^*)^2) \\
-    &\qquad + p_{13}         \cdot\log(1+(x^*)^2+(y^*)^2-4x^*y^*w+(x^*)^2(y^*)^2) \\
+    &\sim     p_{\emptyset}  \cdot\log(1+(\hat{x})^2+(\hat{y})^2+4\hat{x}\hat{y}w+(\hat{x})^2(\hat{y})^2) \\
+    &\qquad + p_{13}         \cdot\log(1+(\hat{x})^2+(\hat{y})^2-4\hat{x}\hat{y}w+(\hat{x})^2(\hat{y})^2) \\
     &\qquad + (1 - p_{13})\cdot\log(1+w) \\
@@
     \ell_{\tau_1,t^*}(\tau_1, w)
-    &\sim     p_{\emptyset}  \cdot\log(1+(x^*)^2+(y^*)^2+4x^*y^*w+(x^*)^2(y^*)^2) \\
-    &\qquad + p_{13}         \cdot\log(1+(x^*)^2+(y^*)^2-4x^*y^*w+(x^*)^2(y^*)^2) \\
+    &\sim     p_{\emptyset}  \cdot\log(1+(\hat{x})^2+(\hat{y})^2+4\hat{x}\hat{y}w+(\hat{x})^2(\hat{y})^2) \\
+    &\qquad + p_{13}         \cdot\log(1+(\hat{x})^2+(\hat{y})^2-4\hat{x}\hat{y}w+(\hat{x})^2(\hat{y})^2) \\
     &\qquad + \log(1+w).
@@
 \[
-\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
-\]
-\[
-\gamma := \gamma(x^*, y^*) = 4x^*y^*.
-\]
-Since $x^*$ and $y^*$ fall between zero and one, we make use of the inequalities (excepting the cases of $x^*=y^*=0$ and $x^*=y^*=1$)
+\beta := \beta(\hat{x}, \hat{y}) = 1+(\hat{x})^2+(\hat{y})^2+(\hat{x})^2(\hat{y})^2,
+\]
+\[
+\gamma := \gamma(\hat{x}, \hat{y}) = 4\hat{x}\hat{y}.
+\]
+%das: TODO double check restrictions on \hat{x} and \hat{y} are reasonable
+Since $x^*,y^*,\hat{x},$ and $\hat{y}$ all fall between zero and one, we make use of the inequalities (excepting the cases of $x^*=y^*=0$ and $x^*=y^*=1$)
 \[
@@
 \[
-\alpha_1 + \alpha_2 = \frac{1}{4}\beta
+\alpha_1 + \alpha_2 = \frac{1}{4}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)
 \]
@@
 and simplifying as functions of $\gamma$ and $\beta$ shows that $a + b + c \ge 0$ is equivalent to
+\[
+-\gamma^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0,
+\]
+which, writing out $\gamma$ and $\beta$ explicitly yields
+\begin{equation}
+\label{eq:general-bl-result}
+-\gamma^2(\hat{x}, \hat{y})\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma(\hat{x}, \hat{y})\beta(\hat{x}, \hat{y})x^*(y^*)^2 + \beta^2(\hat{x}, \hat{y}) \ge 0.
+\end{equation}
+Given the bounds
+\[
+\gamma_{L} \le \gamma(\hat{x}, \hat{y}) \le \gamma_{U}
+\]
+and
+\[
+\beta_{L} \le \beta(\hat{x}, \hat{y}),
+\]
+if
+\[
+-\gamma_{U}^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0,
+\]
+then \eqref{eq:general-bl-result} holds, and this is an inequality involving only $x^*$ and $y^*$.
+\end{proof}
+To get a sense of the region of overestimation, Fig.~\ref{fig:bl-loose-inconsistency} plots an intermediate case.
+
+\subsubsection*{Restricted case}
+
+Theorem~\ref{thm:general-bl} allows us to see where inconsistency in branch length estimation arises when performing general joint maximization over all of the branch parameters, though its conclusion is a statement about both the inconsistency of the internal branch estimate as well as about poor estimation of the other branch parameter estimates.
+To obtain an analytic result indicating where only the internal branch parameter is inconsistently estimated, we consider a restricted case of Theorem~\ref{thm:general-bl}.
+Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and estimate the internal branch parameter $w$.
+%das: @EM I made this not a theorem since it's now a special case. Should it stay a theorem?
+Let
+\[
+\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
+\]
+\[
+\gamma := \gamma(x^*, y^*) = 4x^*y^*.
+\]
+The maximum likelihood value $\hat{w}$ is equal to 1 if
 \begin{equation}
@@
 %\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
--\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0.
+-\gamma^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0,
 \end{equation}
-\end{proof}
-
-\generalBranchInconsist*
-
-\begin{proof}
-In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
-From the previous section, $\hat{w}$ is given by the quadratic formula, though now with $\gamma$ and $\beta$ as functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
-Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
-The same derivation as for \eqref{eq:restricted-bl-result} further yields
-\begin{equation}
-\label{eq:general-bl-result}
--\gamma^2(\hat{x}, \hat{y})\left(1 + \frac{1}{2}\beta\right) + 2\gamma(\hat{x}, \hat{y})\beta(\hat{x}, \hat{y})x^*(y^*)^2 + \beta^2(\hat{x}, \hat{y}) \ge 0
-\end{equation}
-where, since $\alpha_1$ and $\alpha_2$ are still functions of $x^*$ and $y^*$, $\left(1+1/2\cdot\beta\right)$ and $x^*(y^*)^2$ from \eqref{eq:restricted-bl-result} remain unchanged.
-Given the bounds
-\[
-\gamma_{L} \le \gamma(\hat{x}, \hat{y}) \le \gamma_{U}
-\]
-and
-\[
-\beta_{L} \le \beta(\hat{x}, \hat{y}),
-\]
-if
-\[
--\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0,
-\]
-then \eqref{eq:general-bl-result} holds, and this is an inequality involving only $x^*$ and $y^*$.
-\end{proof}
-To get a sense of the region of overestimation, Fig.~\ref{fig:bl-loose-inconsistency} plots an intermediate case.
+and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
+This follows in a straightforward manner from \eqref{eq:general-bl-result}.
+If $\hat{x}=x^*$ and $\hat{y}=y^*$, then $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$ and \eqref{eq:general-bl-result} becomes \eqref{eq:restricted-bl-result}.
+This allows us to demarcate a region of biased internal branch length estimation by plotting where the inequality in \eqref{eq:restricted-bl-result} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
+
+\begin{figure}
+\centering
+% obtained by running
+% python joint_inf_plot.py --analytic --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
+\includegraphics[width=\textwidth]{branch-length-inconsistency-inkscape}
+\caption{Analytically-derived region of branch parameter inconsistency in terms of fidelities $x=x^*$ and $y=y^*$ that are fixed to their correct values with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
+The shaded region shows the area in which the internal branch length is estimated to be 0, i.e.\ the estimated fidelity $\hat{w} = 1$, even though the generating fidelity is $y^*$.
+Here again, the shaded region is guaranteed to give inconsistent estimation, while the white region may or may not do so.}
+\label{fig:bl-inconsistency}
+\end{figure}
+
 \subsection*{Empirical validation}

### 2018-05-11 joint_inf.tex
move special case to appendix

@@
 As above, assume that data is generated from the InvFels tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
-In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
+
+In the general case, we do the same but with
+\begin{equation}
+\label{eq:general-bl-opt}
+(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\})
+\end{equation}
+to find a region where the inferred values do not converge to the generating values.
+In the Appendix we consider a ``restricted'' case where we show that, for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
 That is, if we estimate $x$ and $y$ correctly, then, for
@@
 there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
-In the general case, we do the same but with
-\[
-(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\})
-\]
-to find a region where the inferred values do not converge to the generating values.
 These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \citep{RoyChoudhury2015-ta}.
 
-\subsubsection*{Restricted case}
-
-Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and estimate the internal branch parameter $w$.
-\begin{restatable}{theorem}{restrictedBranchInconsist}
-\label{thm:restricted-bl}
-Let
+%Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
+%We show there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
+The estimate $\hat{w}$ in \eqref{eq:general-bl-opt} is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
+Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$.
+We fix bounds on these estimates and show that, in optimizing \eqref{eq:general-bl-opt}, joint estimation either estimates $\hat{w}$ to be one or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
+\begin{restatable}{theorem}{generalBranchInconsist}
+\label{thm:general-bl}
+Define
 \[
-\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
+\gamma(x, y) = 4xy,
 \]
 \[
-\gamma := \gamma(x^*, y^*) = 4x^*y^*.
+\beta(x, y) = 1+x^2+y^2+x^2y^2,
 \]
-The maximum likelihood value $\hat{w}$ is equal to 1 if
-\[
-%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
--\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0
-\]
-and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
-\end{restatable}
-This theorem allows us to demarcate a region of biased internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
-Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch with this branch length estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
-We provide the following as an intuition for the theoretical development.
-For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral states.
-For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
-A maximization procedure will more likely introduce a $(1+w)$ term than a $(1-w)$ term, though this will not be guaranteed as the maximum depends on the values of $x$ and $y$.
-This tendency to more likley include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities.
-If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.
-
-\begin{figure}
-\centering
-% obtained by running
-% python joint_inf_plot.py --analytic --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
-\includegraphics[width=\textwidth]{branch-length-inconsistency-inkscape}
-\caption{Analytically-derived region of branch parameter inconsistency in terms of fidelities $x=x^*$ and $y=y^*$ that are fixed to their correct values with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
-The shaded region shows the area in which the internal branch length is estimated to be 0, i.e.\ the estimated fidelity $\hat{w} = 1$, even though the generating fidelity is $y^*$.
-Here again, the shaded region is guaranteed to give inconsistent estimation, while the white region may or may not do so.}
-\label{fig:bl-inconsistency}
-\end{figure}
-
-\subsubsection*{General case}
-
-The general case is more challenging to analyze and so we obtain weaker bounds.
-Here, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
-Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
-Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$.
-We show there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
-We fix bounds on these estimates and show that, in the general case, joint estimation either estimates $\hat{w}$ to be one or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
-\begin{restatable}{theorem}{generalBranchInconsist}
-\label{thm:general-bl}
-Define $\gamma(x, y) = 4xy$.
-For
-\[
-\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
-\]
-\[
-\gamma := \gamma(x^*, y^*),
-\]
-bounds
+and bounds
 \[
@@
 \[
-\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}),
+\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}).
 \]
-the maximum likelihood value $\hat{w}=1$ when
+The maximum likelihood value $\hat{w}=1$ when
 \[
--\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
+-\gamma_{U}^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
 \]
@@
 We use this theorem to show incorrect branch parameter estimates as follows.
-If we do not tolerate any error in pendant branches, we use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section (Fig.~\ref{fig:bl-inconsistency}).
+If we do not tolerate any error in pendant branches, we use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$; this corresponds to the restricted case, analyzed in the Appendix (Fig.~\ref{fig:bl-inconsistency}).
 For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
@@
 In this case, joint inference will either estimate $\hat{w}$ to have fidelity one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.
+
+%das: TODO update this intuition
+Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch with this branch length estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
+We provide the following as an intuition for the theoretical development.
+For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral states.
+For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
+A maximization procedure will more likely introduce a $(1+w)$ term than a $(1-w)$ term, though this will not be guaranteed as the maximum depends on the values of $x$ and $y$.
+This tendency to more likley include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities.
+If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.
 

### 2018-05-11 appendix.tex
fidelities --> branch lengths

@@
 \]
-See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < x^*, y^* < 1$ where $C_1(x^*, y^*) > C_0(x^*, y^*)$ and joint inference is inconsistent.
+See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < p_{x^*}, p_{y^*} < 1$ with $p_{x^*}=(1-x^*)/2$ and $p_{y^*}=(1-y^*)/2$ where $C_1(x^*, y^*) > C_0(x^*, y^*)$ and joint inference is inconsistent.
 \end{proof}
@@
 \includegraphics[width=\textwidth]{branch-length-inconsistency-inkscape}
-\caption{Analytically-derived region of branch parameter inconsistency in terms of fidelities $x=x^*$ and $y=y^*$ that are fixed to their correct values with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
-The shaded region shows the area in which the internal branch length is estimated to be 0, i.e.\ the estimated fidelity $\hat{w} = 1$, even though the generating fidelity is $y^*$.
+\caption{Analytically-derived region of branch parameter inconsistency where $p_x=p_{x^*}$ and $p_y=p_{y^*}$ are fixed to their correct values with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
+The shaded region shows the area in which the internal branch length $\hat{p}_w$ is estimated to be 0, i.e.\ the estimated fidelity $\hat{w} = 1$, even though the generating fidelity is $y^*$.
 Here again, the shaded region is guaranteed to give inconsistent estimation, while the white region may or may not do so.}

### 2018-05-11 joint_inf.tex
fidelities --> branch lengths

@@
 \caption{
-    An analytically-derived region of topological inconsistency in terms of fidelities for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $w^* = y^*$.
+    An analytically-derived region of topological inconsistency in terms of fidelities for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $w^* = y^*$ ($p_{w^*}=p_{y^*}$).
     Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
@@
 
-%das: TODO update this intuition
-Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch with this branch length estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
 We provide the following as an intuition for the theoretical development.
@@
     Analytically-derived region of branch parameter inconsistency in terms of fidelities with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
-    In the marked region, either $\hat{w}=1$ or one of $x^*-0.1 \le \hat{x} \le x^*+0.1$ or $y^*-0.1 \le \hat{y} \le y^*+0.1$ will not be true, resulting in poor estimation of the pendant branch parameters.
+    In the marked region, either $\hat{p}_w=0$ or, for $p_{x^*}=(1-x^*)/2$ and $p_{y^*}=(1-y^*)/2$, one of $x^*-0.1 \le \hat{x} \le x^*+0.1$ or $y^*-0.1 \le \hat{y} \le y^*+0.1$ will not be true, resulting in poor estimation of the pendant branch parameters.
     As in previous plots, the shaded and white regions are loose indications of inconsistency.
@@
 % or
-% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-01.svg --in-pkl-name figures/w-hat-empirical-01.pkl
+% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-01.svg --in-pkl-name ../intermed/w-hat-empirical-01.pkl
 % if fit already
@@
 \caption{
-    Numerical estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing \eqref{eq:profile_likelihood}.
+    Numerical estimates for $\hat{p}_w=(1-\hat{w})/2$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing \eqref{eq:profile_likelihood}.
     Data generated as in Fig.~\ref{fig:inconsistency-farris}.

### 2018-05-11 joint_inf_plot.py

@@ -386,9 +386,12 @@ def main(args=sys.argv[1:]):
     args = parse_args()
 
     st_time = time.time()
-    x_range = np.arange(args.delta, 1.0, args.delta)
-    y_range = np.arange(args.delta, 1.0, args.delta)
-    X, Y = np.meshgrid(x_range,y_range)
+    # plot probabilities of substitution, not fidelities
+    px_range = np.arange(args.delta, 0.5, args.delta)
+    py_range = np.arange(args.delta, 0.5, args.delta)
+    PX, PY = np.meshgrid(px_range,py_range)
+    X = 1-2*PX
+    Y = 1-2*PY
 
     if args.analytic:
         if args.topology:
@@ -398,23 +401,28 @@ def main(args=sys.argv[1:]):
         elif args.restricted_branch_lengths:
             region = what_lower_minus_one(X, Y)
             plottitle = r'Region of inconsistent branch parameter estimation''\n'r'(restricted case)'
-            legendtext = r'$\hat{w}=1$'
+            legendtext = r'$\hat{p}_w=0$'
         elif args.general_branch_lengths:
             region = what_lower_general(X, Y)
             plottitle = r'Region of inconsistent branch parameter estimation''\n'r'(general case)'
-            legendtext = r'$\hat{w}=1$ or $\hat{x}$ or $\hat{y}$''\n'r'poorly estimated'
+            legendtext = r'$\hat{p}_w=0$ or $\hat{p}_x$ or $\hat{p}_y$''\n'r'poorly estimated'
 
-        ct = plt.contour(X, Y, region, [0], colors='k', linewidths=1)
+        ct = plt.contour(PX, PY, region, [0], colors='k', linewidths=1)
         plt.axis('scaled')
-        plt.xlabel(r'$x^*$', fontsize=FONT_SIZE)
-        plt.ylabel(r'$y^*$', fontsize=FONT_SIZE)
+        plt.xlabel(r'$p_{x^*}$', fontsize=FONT_SIZE-4)
+        plt.ylabel(r'$p_{y^*}$', fontsize=FONT_SIZE-4)
         ttl = plt.title(plottitle, fontsize=FONT_SIZE+2)
         ttl.set_position([.5, 1.05])
         ct.ax.tick_params(labelsize=FONT_SIZE-2)
-        if not args.topology:
+        if args.topology:
[…]

### 2018-05-11 joint_inf.tex
add goldman and allman/chai

@@
 
-An estimator is said to be statistically consistent if it converges to the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
+An estimator is said to be statistically consistent if it converges to the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
 These proofs do not readily extend to include estimating ancestral states.
@@
 Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \citep{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \citep{LeCam1990}.
-Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \citep{RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
+Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \citep{Allman2004-wd,Chai2011-ff,RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
 We have furthered progress in understanding the limitations of this joint optimization procedure.
@@
 Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.
+
+The case of joint inference of a phylogenetic likelihood was discussed in \citet{Goldman1990-dk}.
+In it, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general.
+We show cases where the estimates of the topology will be inconsistent if we are estimating both topology and branch lengths, and, moreover, even if we know the correct topology, we show cases where branch lengths will also be inconsistent.
+Finally, just prior to his conclusion, he discusses when parsimony will give the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.
+The joint inference procedure outlined here estimates the same parameters as classical maximum likelihood, rendering the question well-posed.
+Finally, we are able to provide much more detail on how large branch lengths must be for general joint inference---estimating topology and branch lengths together---to fail to be consistent.
 

### 2018-05-11 appendix.tex
I was more explicit in the theorem about cases that were exceptions

@@
 \]
-%das: TODO double check restrictions on \hat{x} and \hat{y} are reasonable
-Since $x^*,y^*,\hat{x},$ and $\hat{y}$ all fall between zero and one, we make use of the inequalities (excepting the cases of $x^*=y^*=0$ and $x^*=y^*=1$)
+%das: added as assumption to theorem
+Since $x^*,y^*,\hat{x},$ and $\hat{y}$ all fall between zero and one, we make use of the inequalities
 \[
@@
 \[
-\beta > \gamma.
-\]
+\beta > \gamma
+\]
+using the assumptions that $x^*=y^*=0$, $\hat{x}=\hat{y}=0$, and $\hat{x}=\hat{y}=1$.
 % Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
@@
 c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
-Excepting all the cases where $x^*=0$ or where $y^*=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
+Using the assumption that $\hat{x}\neq 0$ and $\hat{y}\neq 0$, this implies $\ell'$ has two zeros according to the quadratic formula.
 Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w=0$.

### 2018-05-11 joint_inf.tex
clarify approximation/bias
I was more explicit in the theorem about cases that were exceptions

@@
 \]
-The maximum likelihood value $\hat{w}=1$ when
+Assume $\hat{x}\neq 0,\hat{y}\neq 0,$ $\hat{x}$ and $\hat{y}$ are not both simultaneously one, and $x^*$ and $y^*$ are not both simultaneously zero.
+Then, the maximum likelihood value $\hat{w}=1$ when
 \[
@@
 This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2017-jo}, namely their Figures~2 and 3.
+\citet{Sagulenko2017-jo} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:bl-general-bias} shows even when branches are short there is a consistent bias.
 

### 2018-05-12 appendix.tex
Still a few points of confusion :pooh:

@@
 Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
+For the binary symmetric model, the alphabet $\alphabet$ is $\{0,1\}$.
 Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
-For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
-For the binary symmetric model $\alphabet=\{0,1\}$.
-Define the complement of $\alignmentColumn$ as $\overline{\alignmentColumn}$, and let $\alignmentColumn_{i,i'}$ be the label of the $i'$th tip in the $i$th alignment column.
+For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$, commonly called a ``site split.''
+Define the complement of $\alignmentColumn$ as $\overline{\alignmentColumn}$, and let $\alignmentColumn_{i,k}$ be the label of the $k$th tip in the $i$th alignment column.
 We define the site split $\siteSplit$ for a $\alignmentColumn_i$ as the set of tips labeled with $1$ in $\alignmentColumn_i$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of tips labeled with $1$ in $\overline{\alignmentColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
@@
 In this case we cannot use the same complement trick as before: the probability of observing an ancestral state split conditional on a site split is not invariant to taking its complement.
-We define an ``ancestral state split'' in a similar way to a site split, where the ancestral state split $\ancestralSplit$ for an internal node $\ancestralStateColumn_i$ is the set of internal nodes labeled with $1$ if the $\nSiteRows$th tip---\emph{not} internal node---is not labeled with $1$, and as the set of internal nodes labeled with $1$ in $\overline{\ancestralStateColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
-In other words, both site splits and ancestral state splits are defined by whether the $\nSiteRows$th element of $\alignmentColumn_i$ is labeled as $1$.
-%EM "ancestral state splits" are an important item in the development but don't get the same level of definition as "site splits". Let's define them here.
-%das: done
+We thus define an ``ancestral state split'' $\ancestralSplit$ for an internal node $\ancestralStateColumn_i$ to be the set of internal nodes labeled with $1$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of internal nodes labeled with $1$ in $\overline{\ancestralStateColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
+We emphasize that the ancestral state split complementing procedure depends on tip states, not ancestral states: both site splits and ancestral state splits are defined by whether the $\nSiteRows$th element of $\alignmentColumn_i$ is labeled as $1$.
 
 We enumerate the site splits $\siteSplit_j$ of which there are $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ in total where $\mathcal{P}$ denotes the power set.
-%EM I think that we should either use "ancestral splits" everywhere, or "ancestral state splits."
-%das: yah, agreed
 Similarly we enumerate ancestral state splits $\ancestralSplit_k$ of which there are $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$ in total.
@@
 Here, $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
-%EM Wait, does this establish surjectivity? We need to be able to get all values of P(H).
-%das: ah crap, I don't know why I wrote the above. I changed a bit of the notation (sorry!) but I hope this will be more clear?
 For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
@@
 %TODO: the caption of this overlaps with the page number---will this be a problem or will MBE handle it?
+%EM There is a latex package for that, which I tried out and then removed for some reason. It's no bigs.
 
@@
 Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
-We decompose the likelihood into the entropy and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
-%EM I'm thinking about this wrong for some reason: the sum of two things, each of which has a unique maximum, does not necessarily have a unique maximum.
-%das: I added the table of general generating probabilities, but it's pretty straightforward I think to see below the second derivative of the likelihood will be negative on the whole interval, implying we have a unique maximum.
-%I think we're good.
-%Let me know if I should say anything about the continuity of the likelihood, but I think it's obvious this gets us to where we need to be?
-Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$, and the map $g(x_1) \mapsto -g(x_1)\log g(x_1)$.
-We restrict our attention to $x_1,y_1,x_2,y_2,w\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
-By straightforward calculus, the second derivative of this map is
+We decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
+Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2,w\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
+%EM It seems to me that this only works if we know that \tilde{t} = t^*$. Otherwise the divergence won't have this form.
+%EM Also, I tried to make the connection between the divergence and g a little more explicit.
+The second derivative of the divergence term $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t}) = -g(x_1)\log g(x_1)$ is
 \[
@@
 This implies the above second derivative is negative for each component of $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t})$.
-%EM There's not good notational overlap between Table S2 and the definition of partial log-likelihood. Could you fix that?
-%das: I think fixed, or did you mean something in addition to changing h-->\Xi and y-->\Psi?
 
 We can compute the partial log-likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ using Table~\ref{tab:farris_likelihoods}.
+%EM This falls short of being "clearly"--- I can easily see that we get some log(1+x_1) terms and some log(1-x_1) terms, and that the below follows if the sum of the coefficients for the two terms is < 1.
+Perhaps add a sentence here?
 Clearly this partial log-likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in(0,1)$ and thus has a unique maximum.
@@
 \end{align}
-%EM Shouldn't this read "obtaining an upper bound"?
-%das: yes
 Our plan is to simplify the likelihood to remove log-of-quadratic terms of $x,y,$ and $w$, obtaining an upper bound on the likelihood that has a closed-form maximum in each variable.
@@
 + (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x) \\
-%EM Shouldn't this be
-%&\qquad\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y)
 &\qquad\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y)
-%EM And this be
-%+ 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
-%das: I don't think so---the first two lines starting at the equals sign are bringing together the first four lines starting above at the <= sign.
-%Should I remove this intermediate equation?
-%I thought it would help clarify but maybe not?
 + (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
@@
 The upper bound $C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*)$ in \eqref{eq:farris-upper-bound-1} is for the ancestral state split $\{1\}$.
-%EM Let's be more specific, giving an equation number, rather than just saying "the upper bound" here.
-%das: is this better? there were a few other typos, so I changed "site split" to "ancestral state split" where needed and updated the constant definitions
 For the ancestral state split $\emptyset$ we can derive an upper bound $C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*)$ in \eqref{eq:farris-upper-bound-empty} in a similar fashion, and $C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*)$ will be the same as $C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*)$ except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.
@@
 Some rearranging and the fact that $0 \le \tilde{p} \le 1/2$ shows this is equivalent to showing $p_{13} \le 3/8$.
-%EM Using what about t^*?
-%das: updated
 Using Table~\ref{tab:sitepatprob} while setting $t=t^*=\{x^*,y^*,x^*,y^*,y^*\}$,
@@
 We next construct a lower bound on the Felsenstein tree partial likelihood.
-%das: we never actually use the Gibbs inequality trick for the upper bound---just some log-of-polynomial bounds
 In general, taking

### 2018-05-12 joint_inf.tex
Still a few points of confusion :pooh:

@@
 \usepackage[colorlinks=true, allcolors=blue]{hyperref}
-\usepackage{showlabels}
+% \usepackage{showlabels}
 \usepackage{palatino}
@@
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model with data being generated on the four-taxon ``inverse Felsenstein (InvFels)'' tree---also called the ``Farris'' tree \citep{Siddall1998-hq, Felsenstein2004}---and we construct bounds on the joint objective function to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
+To do so, we use a binary symmetric model with data being generated on the four-taxon ``inverse Felsenstein'' tree (a.k.a.\ the ``Farris'' tree) and we construct bounds on the joint objective function to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
 We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.
@@
 
-We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``InvFels'' and ``Felsenstein'' trees (Fig.~\ref{fig:farris-fels-top}).
+We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``Inverse Felsenstein (InvFels)'' tree and the ``Felsenstein'' tree (Fig.~\ref{fig:farris-fels-top}).
+The ``inverse Felsenstein'' terminology comes from \citet{Swofford2001-hr}, although it is also called the ``Farris'' tree \citep{Siddall1998-hq, Felsenstein2004}.
 In the standard configuration of each of these trees, the interior branch length parameters are equal to the bottom two parameters.
@@
 
-%EM Line numbering has a huge gap here. Could you look around and see if you see anything about this on the web?
-%das: fixed
 We now consider the problem of branch length estimation on the correct tree using joint estimation.
@@
 A maximization procedure will more likely introduce a $(1+w)$ term than a $(1-w)$ term, though this will not be guaranteed as the maximum depends on the values of $x$ and $y$.
-This tendency to more likley include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities.
+This tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities.
 If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.
@@
 
+%EM Let's set a time in the first half of this week for you to walk me through the Goldman paper and this paragraph. Thanks!
 The case of joint inference of a phylogenetic likelihood was discussed in \citet{Goldman1990-dk}.

### 2018-05-14 appendix.tex
more explicit (and hopefully correct) likelihood form

@@
 We decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
-Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2,w\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
+Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
 %EM It seems to me that this only works if we know that \tilde{t} = t^*$. Otherwise the divergence won't have this form.
+%das: Heh, yeah oops, that was super sloppy.
+%Should be fixed, and a lot more explicit now; thanks for keeping me honest!
+%Also there's no restriction on w being not equal to zero or one since we're only interested in x and y, so that was removed, too.
 %EM Also, I tried to make the connection between the divergence and g a little more explicit.
-The second derivative of the divergence term $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t}) = -g(x_1)\log g(x_1)$ is
-\[
--g''(x_1)(1+\log g(x_1)) - \frac{g'(x_1)^2}{g(x_1)}.
-\]
-Using each functional form in Table~\ref{tab:gen-sitepatprob} we see in all cases $g''(x_1)=0$ and $g(x_1) > 0$ for $x_1\in(0,1).$
-This implies the above second derivative is negative for each component of $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t})$.
-
-We can compute the partial log-likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ using Table~\ref{tab:farris_likelihoods}.
+Define
+\[
+g_j(x_1) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}),
+\]
+i.e., the generating probability of the $j$th site split as a function of $x_1$.
+Recall the divergence
+\[
+\shannonDivergence_{\tau^*,t^*}(\tau,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \tilde{t}),
+\]
+which, suppressing notation for clarity, we define as a function of $x_1$ as
+\[
+D(x_1) = \sum_{j=1}^\nSiteSplits c_j \cdot \log g_j(x_1)
+\]
+where $c_j \ge 0$ for $j=1,\ldots,\nSiteSplits$.
+It remains to show $D''(x_1) < 0$ for $x_1,y_1,x_2,y_2\in(0,1)$.
+By straightforward calculus,
+\[
+D''(x_1) = \sum_{j=1}^\nSiteSplits c_j \cdot \left(\frac{g_j(x_1)g_j''(x_1) - g_j'(x_1)^2}{g_j(x_1)^2}\right).
+\]
+Using each functional form in Table~\ref{tab:gen-sitepatprob} we see $g_j''(x_1)=0$ and $g_j(x_1)>0$ for $x_1,y_1,x_2,y_2\in(0,1)$ and $j=1,\ldots,\nSiteSplits$ so that
+\[
+D''(x_1) = -\sum_{j=1}^\nSiteSplits c_j \cdot \left(\frac{g_j'(x_1)}{g_j(x_1)}\right)^2.
+\]
+Finally, we require at least one $j$ where $c_j \neq 0$ and $g_j'(x_1) \neq 0$ so that $D''(x_1) \not\equiv 0$.
+Consider the case $j=1$ where $\siteSplit_1=\emptyset$ so that, ignoring the $1/8$ constant in each term,
+\[
+g_1(x_1) \propto 1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2,
+\]
+\[
+g_1'(x_1) \propto x_2 + y_1w + y_2w + y_1x_2y_2,
+\]
+and
+\[
+c_1 \propto 1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2.
+\]
+We have $c_1 > 0$ by $x^*,y^*\in[0,1]$ and, by assumption, $g_1'(x_1) > 0$ for $x_1,y_1,x_2,y_2\in(0,1)$.
+This implies the above second derivative is negative.
+
 %EM This falls short of being "clearly"--- I can easily see that we get some log(1+x_1) terms and some log(1-x_1) terms, and that the below follows if the sum of the coefficients for the two terms is < 1.
 Perhaps add a sentence here?
-Clearly this partial log-likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in(0,1)$ and thus has a unique maximum.
-Since both summands of the likelihood are concave on $x_1\in(0,1)$, the likelihood is as well, and there exists a single $x_1\in(0,1)$ maximizing the likelihood.
+%das: you're right, I missed a factor of two; hope it's more clear now.
+We now focus on the partial log-likelihood term.
+Recall the partial log-likelihood takes the form
+\[
+\tilde{\ell}_{\tau^*,t^*}(\tau, \tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, \tilde{t}).
+\]
+As above, define the partial log-likelihood as a function of $x_1$ as
+\[
+\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_j(x_1)
+\]
+where
+\[
+h_j(x_1) = \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\}).
+\]
+Using the procedure in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, we see each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, will be
+\[
+h_j(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
+\]
+for $e_j\in\{0,1\}$, which, simplifying, results in
+\begin{align}
+\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
+                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1).
+\end{align}
+This has a second derivative
+\[
+\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
+\]
+Again, by assumption $x_1,y_1,x_2,y_2\in(0,1)$, and we need only $0 \le \sum_j c_je_j \le 1$ so that, by the second derivative test, the sum of $D(x_1)$ and $\tilde{\ell}(x_1)$ has a unique maximum.
+Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$, thus, $D(x_1) + \tilde{\ell}(x_1)$ is concave as a function of $x_1\in(0,1)$, and there exists a single $x_1\in(0,1)$ maximizing the likelihood.
 Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.

### 2018-05-14 joint_inf.tex
fix reference

@@
 Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \citep{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \citep{LeCam1990}.
-Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \citep{Allman2004-wd,Chai2011-ff,RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
+Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
 We have furthered progress in understanding the limitations of this joint optimization procedure.

### 2018-05-15 appendix.tex
A little more nit-picking

@@
 Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
-%EM It seems to me that this only works if we know that \tilde{t} = t^*$. Otherwise the divergence won't have this form.
-%das: Heh, yeah oops, that was super sloppy.
-%Should be fixed, and a lot more explicit now; thanks for keeping me honest!
-%Also there's no restriction on w being not equal to zero or one since we're only interested in x and y, so that was removed, too.
-%EM Also, I tried to make the connection between the divergence and g a little more explicit.
+%EM This looks much better now. Can we get a subscript on tau for these calculations?
 Define
@@
 \]
-which, suppressing notation for clarity, we define as a function of $x_1$ as
+which, simplifying notation for clarity, we define as a function of $x_1$ as
 \[
@@
 
-%EM This falls short of being "clearly"--- I can easily see that we get some log(1+x_1) terms and some log(1-x_1) terms, and that the below follows if the sum of the coefficients for the two terms is < 1.
-Perhaps add a sentence here?
-%das: you're right, I missed a factor of two; hope it's more clear now.
 We now focus on the partial log-likelihood term.
@@
 \]
-Using the procedure in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, we see each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, will be
+Doing calculation as in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, we see each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, will be
 \[
@@
 Again, by assumption $x_1,y_1,x_2,y_2\in(0,1)$, and we need only $0 \le \sum_j c_je_j \le 1$ so that, by the second derivative test, the sum of $D(x_1)$ and $\tilde{\ell}(x_1)$ has a unique maximum.
+%EM Shouldn't you at least mention that the second deriv isn't exactly zero as you did before?
 Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$, thus, $D(x_1) + \tilde{\ell}(x_1)$ is concave as a function of $x_1\in(0,1)$, and there exists a single $x_1\in(0,1)$ maximizing the likelihood.

### 2018-05-15 joint_inf.tex
A little more nit-picking

@@
 Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out latent ancestral states at the internal nodes of the tree.
-In a recent paper, \citet{Sagulenko2017-jo} suggest using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences on a fixed topology.
+In a recent paper, \citet{Sagulenko2018-xl} suggest using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences on a fixed topology.
 This is attractive from a computational perspective: such joint inference can proceed according to an iterative procedure in which ancestral sequences are first estimated and model parameters are optimized conditional on these estimates.
@@
 %page 2 section 2.1 at the top of the second column first paragraph second sentence
-Although the software described in \citet{Sagulenko2017-jo} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
+Although the software described in \citet{Sagulenko2018-xl} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
 In particular, one would like to know when this approximate technique breaks down for both topology and branch length inference, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.
@@
 
-The alternative approach \citep{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
+The alternative approach \citep{Sagulenko2018-xl} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
 This is known in statistics as a profile likelihood \citep{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
@@
 In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima row-wise in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} to obtain the likelihood term for each most likely ancestral state, these terms depending on the unknown $(\tau, t)$.
-For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2017-jo}.
+For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2018-xl}.
 
@@
 \caption{
+%EM I'm sure you're getting to this, but we need to change around wording to compensate for the use of probabilities vs. fidelities now.
     An analytically-derived region of topological inconsistency in terms of fidelities for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $w^* = y^*$ ($p_{w^*}=p_{y^*}$).
@@
 
-We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2017-jo} is not consistent in general.
+We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
 Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference converges on the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
 In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.
-This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2017-jo}, namely their Figures~2 and 3.
-\citet{Sagulenko2017-jo} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:bl-general-bias} shows even when branches are short there is a consistent bias.
+This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and 3.
+\citet{Sagulenko2018-xl} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:bl-general-bias} shows even when branches are short there is a consistent bias.
 

### 2018-05-15 appendix.tex
addressing comments

@@
 
-Consider the InvFels tree with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
+Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
 We now show that, in the case of the InvFels tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch parameters to be equal and the bottom two branch parameters to be equal during inference obtains the same maximum likelihood estimate as in the case of arbitrary branch parameters.
@@
 %EM This looks much better now. Can we get a subscript on tau for these calculations?
+%das: I changed all of them that were already subscripts, but did you also want subscripts on ones that didn't have them, e.g., D'' and g_1, etc.?
 Define
 \[
-g_j(x_1) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}),
+g_j(x_1) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \{x_1,y_1,x_2,y_2,w\}),
 \]
@@
 \[
-\shannonDivergence_{\tau^*,t^*}(\tau,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \tilde{t}),
+\shannonDivergence_{\tau_1,t^*}(\tau_1,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}),
 \]
@@
 \[
-\tilde{\ell}_{\tau^*,t^*}(\tau, \tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, \tilde{t}).
+\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, \tilde{t}).
 \]
@@
 \[
-h_j(x_1) = \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\}).
+h_j(x_1) = \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, \{x_1,y_1,x_2,y_2,w\}).
 \]
@@
 %EM Shouldn't you at least mention that the second deriv isn't exactly zero as you did before?
-Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$, thus, $D(x_1) + \tilde{\ell}(x_1)$ is concave as a function of $x_1\in(0,1)$, and there exists a single $x_1\in(0,1)$ maximizing the likelihood.
+%das: added a sentence
+Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
+Thus, with $D''(x_1) < 0$, the sum $D''(x_1) + \tilde{\ell}''(x_1) < 0$, implying $D(x_1) + \tilde{\ell}(x_1)$ is concave as a function of $x_1\in(0,1)$, and there exists a single $x_1\in(0,1)$ maximizing the likelihood.
 Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.

### 2018-05-15 joint_inf.tex
addressing comments

@@
 %EM I'm sure you're getting to this, but we need to change around wording to compensate for the use of probabilities vs. fidelities now.
-    An analytically-derived region of topological inconsistency in terms of fidelities for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $w^* = y^*$ ($p_{w^*}=p_{y^*}$).
+%das: You just mean for the figure captions, right?
+I changed it in the figures and a few places in the text, but maybe you meant we should change around the whole text to be in terms of branch lengths?
+Should we change the InvFels/Fels tree figure, too?
+    An analytically-derived region of topological inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
     Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
@@
 Then $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
-As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
+As an example, Fig.~\ref{fig:bl-loose-inconsistency} shows the case where we have poor estimation of one of the three branch parameters when $x_L=x_U=y_L=y_U=0.1$.
 In this case, joint inference will either estimate $\hat{w}$ to have fidelity one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.
@@
 A maximization procedure will more likely introduce a $(1+w)$ term than a $(1-w)$ term, though this will not be guaranteed as the maximum depends on the values of $x$ and $y$.
-This tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities.
+This tendency to include $(1+w)$ terms in the likelihood results in a bias toward one of branch fidelities.
 If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.
@@
 \caption{
-    Analytically-derived region of branch parameter inconsistency in terms of fidelities with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
-    In the marked region, either $\hat{p}_w=0$ or, for $p_{x^*}=(1-x^*)/2$ and $p_{y^*}=(1-y^*)/2$, one of $x^*-0.1 \le \hat{x} \le x^*+0.1$ or $y^*-0.1 \le \hat{y} \le y^*+0.1$ will not be true, resulting in poor estimation of the pendant branch parameters.
+    Analytically-derived region of branch parameter inconsistency in terms of probabilities of a character change with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
+    In the marked region, either $\hat{p}_w=0$ or one of $p_{x^*}-.05 \le \hat{p}_x \le p_{x^*}+.05$ or $p_{y^*}-.05 \le \hat{p}_y \le p_{y^*}+.05$ will not be true, resulting in poor estimation of the pendant branch parameters.
+    In terms of fidelities, these bounds are $x^*-0.1 \le \hat{x} \le x^*+0.1$ and $y^*-0.1 \le \hat{y} \le y^*+0.1$.
     As in previous plots, the shaded and white regions are loose indications of inconsistency.
@@
 We see that the region of inconsistency in the general case is similar to that of the restricted case (compare Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency}).
-This region encompasses the majority of the branch fidelity space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch fidelity to be one.
+This region encompasses the majority of the branch parameter space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch length to be zero.
 
@@
 Finally, just prior to his conclusion, he discusses when parsimony will give the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.
-The joint inference procedure outlined here estimates the same parameters as classical maximum likelihood, rendering the question well-posed.
+The joint inference procedure outlined here estimates the same parameters as classical maximum likelihood---albeit implicitly estimating ancestral states as well---rendering the question well-posed.
 Finally, we are able to provide much more detail on how large branch lengths must be for general joint inference---estimating topology and branch lengths together---to fail to be consistent.

### 2018-05-15 joint_inf.tex
Adding a clarifying sentence, and clearing out :+1: comments.

@@
 
+Although we use fidelities exclusively for our theoretical development, we have made our figures in terms of probabilities of substitution $p_{\theta} = (1-\theta)/2$ as they are easier to interpret.
+
 \subsection*{Two paths to maximum likelihood}
@@
 \caption{
-%EM I'm sure you're getting to this, but we need to change around wording to compensate for the use of probabilities vs. fidelities now.
-%das: You just mean for the figure captions, right?
-I changed it in the figures and a few places in the text, but maybe you meant we should change around the whole text to be in terms of branch lengths?
-Should we change the InvFels/Fels tree figure, too?
     An analytically-derived region of topological inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
@@
 
-%EM Let's set a time in the first half of this week for you to walk me through the Goldman paper and this paragraph. Thanks!
 The case of joint inference of a phylogenetic likelihood was discussed in \citet{Goldman1990-dk}.

### 2018-05-22 appendix.tex
Adding a clarifying sentence, and clearing out :+1: comments.
update form argument and add proptos

@@
 It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
-Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
+We show the Hessian of the joint likelihood term is negative definite, which implies there is a unique maximizer of \eqref{eq:profile_likelihood}.
 We decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
+
+%Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
 Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
-%EM This looks much better now. Can we get a subscript on tau for these calculations?
-%das: I changed all of them that were already subscripts, but did you also want subscripts on ones that didn't have them, e.g., D'' and g_1, etc.?
-Define
-\[
-g_j(x_1) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \{x_1,y_1,x_2,y_2,w\}),
-\]
-i.e., the generating probability of the $j$th site split as a function of $x_1$.
 Recall the divergence
 \[
-\shannonDivergence_{\tau_1,t^*}(\tau_1,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}),
-\]
-which, simplifying notation for clarity, we define as a function of $x_1$ as
-\[
-D(x_1) = \sum_{j=1}^\nSiteSplits c_j \cdot \log g_j(x_1)
-\]
-where $c_j \ge 0$ for $j=1,\ldots,\nSiteSplits$.
-It remains to show $D''(x_1) < 0$ for $x_1,y_1,x_2,y_2\in(0,1)$.
-By straightforward calculus,
-\[
-D''(x_1) = \sum_{j=1}^\nSiteSplits c_j \cdot \left(\frac{g_j(x_1)g_j''(x_1) - g_j'(x_1)^2}{g_j(x_1)^2}\right).
-\]
-Using each functional form in Table~\ref{tab:gen-sitepatprob} we see $g_j''(x_1)=0$ and $g_j(x_1)>0$ for $x_1,y_1,x_2,y_2\in(0,1)$ and $j=1,\ldots,\nSiteSplits$ so that
-\[
-D''(x_1) = -\sum_{j=1}^\nSiteSplits c_j \cdot \left(\frac{g_j'(x_1)}{g_j(x_1)}\right)^2.
-\]
-Finally, we require at least one $j$ where $c_j \neq 0$ and $g_j'(x_1) \neq 0$ so that $D''(x_1) \not\equiv 0$.
-Consider the case $j=1$ where $\siteSplit_1=\emptyset$ so that, ignoring the $1/8$ constant in each term,
-\[
-g_1(x_1) \propto 1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2,
-\]
-\[
-g_1'(x_1) \propto x_2 + y_1w + y_2w + y_1x_2y_2,
-\]
-and
-\[
-c_1 \propto 1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2.
-\]
-We have $c_1 > 0$ by $x^*,y^*\in[0,1]$ and, by assumption, $g_1'(x_1) > 0$ for $x_1,y_1,x_2,y_2\in(0,1)$.
-This implies the above second derivative is negative.
+\shannonDivergence_{\tau_1,t^*}(\tau_1,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}).
+\]
+By Gibbs inequality, this function attains a unique maximum when
+\[
+\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)
+\]
+for all $j=1,\ldots,\nSiteSplits$.
+Referring to Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
+The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
+\[
+-(x^*)^2 = -x_1x_2+y_1w(x_1-x_2)
+\]
+and for $\{1\}$ and $\{1,2\}$ yields
+\[
+-(x^*)^2 = -x_1x_2+y_2w(x_2-x_1).
+\]
+The difference of these two equations simplifies to
+\[
+0 = (y_1w+y_2w)(x_1-x_2)
+\]
+and, since $y_1w+y_2w > 0$ by assumption, we have $x_1=x_2$.
+This along with the previous equalities imply $x_1=x_2=x^*$.
+We use the same process to show that, at the maximum, $y_1=y_2=y^*$, and thus that the divergence term attains a unique maximum over $x_1,y_1,x_2,y_2\in(0,1)$.
 
@@
 \]
-As above, define the partial log-likelihood as a function of $x_1$ as
-\[
-\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_j(x_1)
-\]
-where
-\[
-h_j(x_1) = \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, \{x_1,y_1,x_2,y_2,w\}).
+The partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
+\[
+\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits c_j\cdot\left(\log h_{j,1}(x_1)+\log h_{j,2}(y_1)+\log h_{j,3}(x_2)+\log h_{j,4}(y_2)+\log h_{j,5}(w)\right).
+\]
+Due to this additive form, any off-diagonal terms of the Hessian for this function will be zero, so it remains to show that the diagonal terms are nonpositive.
+Without loss of generality we focus on the variable $x_1$ and the partial log-likelihood proportional to
+\[
+\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,1}(x_1).
 \]
@@
 \[
-h_j(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
+h_{j,1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
 \]
@@
 \]
-Again, by assumption $x_1,y_1,x_2,y_2\in(0,1)$, and we need only $0 \le \sum_j c_je_j \le 1$ so that, by the second derivative test, the sum of $D(x_1)$ and $\tilde{\ell}(x_1)$ has a unique maximum.
-%EM Shouldn't you at least mention that the second deriv isn't exactly zero as you did before?
-%das: added a sentence
+Again, by assumption $x_1,y_1,x_2,y_2\in(0,1)$, and we need only $0 \le \sum_j c_je_j \le 1$ so that the diagonal term of the Hessian will be nonpositive.
 Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
-Thus, with $D''(x_1) < 0$, the sum $D''(x_1) + \tilde{\ell}''(x_1) < 0$, implying $D(x_1) + \tilde{\ell}(x_1)$ is concave as a function of $x_1\in(0,1)$, and there exists a single $x_1\in(0,1)$ maximizing the likelihood.
+Applying similar arguments to the other variables, the Hessian for the partial log-likelihood will have nonpositive diagonal terms and off-diagonal terms equal to zero.
+Thus, the Hessian of the sum of the divergence and the partial log-likelihood will be negative definite, as the Hessian for the divergence is negative definite, and there exists a unique maximum to the likelihood over $x_1,x_2,y_1,y_2\in(0,1)$.
 Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
@@
 \begin{align*}
-p_{12} = p_{23} = 1-x^2-y^2+x^2y^2 & = (1+x)(1-x)(1+y)(1-y) \\
-p_{1} = p_{3} = 1-x^2+y^2-x^2y^2 & = (1+x)(1-x)(1+y^2) \le (1+x)(1-x)(1+y) \\
-p_{2} = p_{123} = 1+x^2-y^2-x^2y^2 & = (1+x^2)(1+y)(1-y) \le (1+x)(1+y)(1-y).
+p_{12} = p_{23} \propto 1-x^2-y^2+x^2y^2 & = (1+x)(1-x)(1+y)(1-y) \\
+p_{1} = p_{3} \propto 1-x^2+y^2-x^2y^2 & = (1+x)(1-x)(1+y^2) \le (1+x)(1-x)(1+y) \\
+p_{2} = p_{123} \propto 1+x^2-y^2-x^2y^2 & = (1+x^2)(1+y)(1-y) \le (1+x)(1+y)(1-y).
 \end{align*}
@@
 \begin{align*}
-    p_{\emptyset} & = \log(1+x^2+y^2+4xyw+x^2y^2) \\
-                  & \le \log(1+x^2+y^2+4xy+x^2y^2) \\
-                  & \le \log(2(1+x)(1+y))
+    p_{\emptyset} & \propto 1+x^2+y^2+4xyw+x^2y^2 \\
+                  & \le 1+x^2+y^2+4xy+x^2y^2 \\
+                  & \le 2(1+x)(1+y)
 \end{align*}
@@
 \begin{align*}
-    p_{13} & = \log(1+x^2+y^2-4xyw+x^2y^2) \\
-                  & \le \log(1+x^2+y^2+x^2y^2) \\
-                  & \le \log((1+x)(1+y)).
+    p_{13} & \propto 1+x^2+y^2-4xyw+x^2y^2 \\
+                  & \le 1+x^2+y^2+x^2y^2 \\
+                  & \le (1+x)(1+y).
 \end{align*}
@@
 Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and estimate the internal branch parameter $w$.
-%das: @EM I made this not a theorem since it's now a special case. Should it stay a theorem?
 Let

### 2018-05-24 appendix.tex
had to add edge case per vu

@@
 Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
+Additionally, we exclude the infinite branch length case of $w=0$.
 Recall the divergence
@@
 \end{align}
+Assume for now the functional form of the partial likelihood is fixed---in other words, each site split has a maximum at a single ancestral state split.
 This has a second derivative
@@
 Applying similar arguments to the other variables, the Hessian for the partial log-likelihood will have nonpositive diagonal terms and off-diagonal terms equal to zero.
-Thus, the Hessian of the sum of the divergence and the partial log-likelihood will be negative definite, as the Hessian for the divergence is negative definite, and there exists a unique maximum to the likelihood over $x_1,x_2,y_1,y_2\in(0,1)$.
-Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
+Thus, the Hessian of the sum of the divergence and the partial log-likelihood will be negative definite, as the Hessian for the divergence is negative definite, provided there is a single maximal ancestral state split for each site split.
+
+We now consider the case where there may be more than one ancestral state split that maximizes the partial likelihood for a site split.
+For site split $\{1\}$, the four possible forms of the partial likelihood are
+\[
+(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2),
+\]
+\[
+(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2),
+\]
+\[
+(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2),
+\]
+\[
+(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2).
+\]
+We ignore the second function as it will always be less than the first except for the case $y_1=y_2=w=0$.
+Comparing the remaining functions reveals that there will be an ambiguity in the form of the likelihood only when either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$.
+We reach the same conclusions doing these comparisons for site splits $\emptyset$, $\{2\}$, $\{3\}$, and $\{1,2,3\}$.
+The remaining site splits $\{1,2\}$, $\{1,3\}$, and $\{2,3\}$ require the following argument.
+For site split $\{1,2\}$, the four possible forms of the partial likelihood are
+\[
+(1-x_1)(1+x_2)(1+w)(1-y_1)(1+y_2),
+\]
+\[
+(1-x_1)(1-x_2)(1-w)(1+y_1)(1-y_2),
+\]
+\[
+(1+x_1)(1-x_2)(1-w)(1-y_1)(1+y_2),
+\]
+\[
+(1+x_1)(1-x_2)(1+w)(1+y_1)(1-y_2).
+\]
+We ignore cases where either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$, and see that an ambiguity in this maximization will occur when
+\[
+(1-x_1)(1+x_2)(1-y_1)(1+y_2) = (1+x_1)(1-x_2)(1+y_1)(1-y_2).
+\]
+Note that this can happen when $x_1=y_2=u_1$ and $x_2=y_1=u_2$ but $u_1 \neq u_2$.
+However, when $x_1=x_2=u_1$ and $y_1=y_2=u_2$ the partial likelihood takes the same value.
+To resolve this ambiguity, we look to the divergence quanitity and recall it will be maximized when $x_1=x_2$ and $y_1=y_2$.
+Thus, the full joint likelihood will obtain a unique maximum, and this maximum will occur when $x_1=x_2$ and $y_1=y_2$.
+
+Given the above, there exists a unique maximum to the likelihood over $x_1,y_2,x_1,y_2\in(0,1)$ with $w\neq 0$, and, with the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
 The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

### 2018-05-25 appendix.tex
put eqns in table and change some grammar

--- 2018-05-22 appendix.tex
@@
 %Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
-Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
+We restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
+Additionally, we exclude the infinite branch length case of $w=0$.
 Recall the divergence
@@
 and, since $y_1w+y_2w > 0$ by assumption, we have $x_1=x_2$.
-This along with the previous equalities imply $x_1=x_2=x^*$.
-We use the same process to show that, at the maximum, $y_1=y_2=y^*$, and thus that the divergence term attains a unique maximum over $x_1,y_1,x_2,y_2\in(0,1)$.
+This along with the previous equalities imply $x_1=x_2=x^*$ is the unique maximum of the divergence term for $x_1$ and $x_2$.
+We use the same process to show that, at the maximum, $y_1=y_2=y^*$, and thus the divergence term attains a unique maximum over $x_1,y_1,x_2,y_2\in(0,1)$ with $w \neq 0$.
 
@@
 The partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
-\[
-\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits c_j\cdot\left(\log h_{j,1}(x_1)+\log h_{j,2}(y_1)+\log h_{j,3}(x_2)+\log h_{j,4}(y_2)+\log h_{j,5}(w)\right).
-\]
-Due to this additive form, any off-diagonal terms of the Hessian for this function will be zero, so it remains to show that the diagonal terms are nonpositive.
+\begin{align*}
+\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,2}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,3}(x_2)\\
+&\qquad +\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,4}(y_2)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,5}(w).
+\end{align*}
+Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
 Without loss of generality we focus on the variable $x_1$ and the partial log-likelihood proportional to
@@
 \]
-Doing calculation as in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, we see each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, will be
+Doing calculation as in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, is
 \[
@@
 \end{align}
-This has a second derivative
+Assume for now the functional form of the partial likelihood is fixed---in other words, each site split has a maximum at a single ancestral state split.
+We have the second derivative
 \[
@@
 \]
-Again, by assumption $x_1,y_1,x_2,y_2\in(0,1)$, and we need only $0 \le \sum_j c_je_j \le 1$ so that the diagonal term of the Hessian will be nonpositive.
+As $x_1,y_1,x_2,y_2\in(0,1)$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
 Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
-Applying similar arguments to the other variables, the Hessian for the partial log-likelihood will have nonpositive diagonal terms and off-diagonal terms equal to zero.
-Thus, the Hessian of the sum of the divergence and the partial log-likelihood will be negative definite, as the Hessian for the divergence is negative definite, and there exists a unique maximum to the likelihood over $x_1,x_2,y_1,y_2\in(0,1)$.
-Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
+Applying similar arguments to the other variables, the Hessian for the partial log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero.
+Thus, the Hessian of the sum of the divergence and the partial log-likelihood is negative definite, as the Hessian for the divergence is negative definite, provided there is a single maximal ancestral state split for each site split.
+
+\begin{table}
+\centering
+\begin{tabular}{|l|ll|}
+    \hline
+    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau_1, \tilde{t})$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,\tilde{t})$\\
+    \hline
+     $\{1\}$    &
+    $\begin{array}{l}
+                    \emptyset\\
+                    \{1\}\\
+                    \{2\}\\
+                    \{1,2\}
+                \end{array}$&
+    $\begin{array}{l}
+                    (1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2)\\
+                    (1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2)\\
+                    (1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2)\\
+                    (1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2)
+                \end{array}$\\
+    \hline
+     $\{1,2\}$    &
+    $\begin{array}{l}
+                    \emptyset\\
+                    \{1\}\\
+                    \{2\}\\
+                    \{1,2\}
+                \end{array}$&
+    $\begin{array}{l}
+                    (1-x_1)(1+x_2)(1+w)(1-y_1)(1+y_2)\\
+                    (1-x_1)(1-x_2)(1-w)(1+y_1)(1-y_2)\\
+                    (1+x_1)(1-x_2)(1-w)(1-y_1)(1+y_2)\\
+                    (1+x_1)(1-x_2)(1+w)(1+y_1)(1-y_2)
+                \end{array}$\\
+    \hline
+\end{tabular}
+\caption{Likelihood calculations for site splits $\{1\}$ and $\{1,2\}$ and ancestral state partitions $\ancestralSplitPartition_j$ on the InvFels tree $\tau_1$ with general branch parameters $\tilde{t}$.
+All values multiplied by $1/32$.}
+\label{tab:edge-cases}
+\end{table}
+
+We now consider the case where there may be more than one ancestral state split that maximizes the partial likelihood for a site split.
+Site split $\emptyset$ always attains a maximum at ancestral state split $\emptyset$ provided at least one branch fidelity is not exactly zero.
+Table~\ref{tab:edge-cases} gives the possible forms for the partial likelihood for two cases of site splits.
+For site split $\{1\}$ we ignore ancestral state split $\{2\}$ as its partial likelihood is always less than the ancestral state split $\emptyset$ except for the case $y_1=y_2=w=0$.
+Comparing the remaining ancestral state splits reveals that there is an ambiguity in the form of the likelihood only when either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$.
+We reach the same conclusions doing these comparisons for site splits $\{2\}$, $\{3\}$, and $\{1,2,3\}$.
+
+The remaining site splits $\{1,2\}$, $\{1,3\}$, and $\{2,3\}$ require the following argument.
+For site split $\{1,2\}$, we ignore cases where either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$, and see that an ambiguity in this maximization occurs when
+\[
+(1-x_1)(1+x_2)(1-y_1)(1+y_2) = (1+x_1)(1-x_2)(1+y_1)(1-y_2),
+\]
+or, in other words, when the ancestral state splits $\emptyset$ and $\{1,2\}$ both maximize the partial likelihood.
+This can happen when $x_1=y_2=u_1$ and $x_2=y_1=u_2$ but $u_1 \neq u_2$, though, when $x_1=x_2=u_1$ and $y_1=y_2=u_2$ the partial likelihood takes the same value.
+To resolve this ambiguity, we look to the divergence quanitity that is maximized when $x_1=x_2$ and $y_1=y_2$, and, thus, the full joint likelihood obtains a unique maximum when $x_1=x_2$ and $y_1=y_2$.
+
+Given the above, there exists a unique maximum to the likelihood over $x_1,y_2,x_1,y_2\in(0,1)$ with $w\neq 0$, and, with the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
 The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

### 2018-05-29 appendix.tex
A bit of confusion :pooh:
add a little more clarification, and fix typos

@@
 Additionally, we exclude the infinite branch length case of $w=0$.
+%EM This looks good, but remind us what x^* is here.
+%das: fixed
 Recall the divergence
@@
 \]
+where data is generated on the InvFels tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
 By Gibbs inequality, this function attains a unique maximum when
@@
 \]
-for all $j=1,\ldots,\nSiteSplits$.
+for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
 Referring to Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
@@
 This along with the previous equalities imply $x_1=x_2=x^*$ is the unique maximum of the divergence term for $x_1$ and $x_2$.
-We use the same process to show that, at the maximum, $y_1=y_2=y^*$, and thus the divergence term attains a unique maximum over $x_1,y_1,x_2,y_2\in(0,1)$ with $w \neq 0$.
+We use the same process to show that the divergence has a unique maximum for $y_1$ and $y_2$ at $y_1=y_2=y^*$, and thus the divergence term attains a unique maximum over $x_1,y_1,x_2,y_2\in(0,1)$ with $w \neq 0$.
+This implies the Hessian of $\shannonDivergence$ is negative definite.
+%EM IIUC we only use this fact to show that we have a _negative_ second derivative at the ^* parameter values.
+It would be nice to emphasize that fact, which would also keep people from worrying that we're falling into the "sum of two things with unique optima has a unique optimum."
+%das: Ah, right.
 
@@
     $\begin{array}{l}
-                    (1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2)\\
-                    (1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2)\\
-                    (1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2)\\
-                    (1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2)
+    %EM I must be confused here-- these aren't the values I get at all!
+    % e.g. for this first one everything is 0 except for the first taxon, which is 1. So shouldn't it be
+                   %(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2)\\ ?
+    %das: Ha! I must have written out a different case. Went through everything with a fine-tooth comb and noticed one or two slip-ups but everything is basically the same.
+                    (1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2)\\
+                    (1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2)\\
+                    (1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2)\\
+                    (1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2)
                 \end{array}$\\
@@
                     (1-x_1)(1+x_2)(1+w)(1-y_1)(1+y_2)\\
-                    (1-x_1)(1-x_2)(1-w)(1+y_1)(1-y_2)\\
                     (1+x_1)(1-x_2)(1-w)(1-y_1)(1+y_2)\\
+                    (1-x_1)(1+x_2)(1-w)(1+y_1)(1-y_2)\\
                     (1+x_1)(1-x_2)(1+w)(1+y_1)(1-y_2)
@@
 For site split $\{1\}$ we ignore ancestral state split $\{2\}$ as its partial likelihood is always less than the ancestral state split $\emptyset$ except for the case $y_1=y_2=w=0$.
-Comparing the remaining ancestral state splits reveals that there is an ambiguity in the form of the likelihood only when either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$.
+Comparing the remaining ancestral state splits reveals that there is an ambiguity in the form of the likelihood only when one of $x_1,y_1,x_2,y_2,$ or $w$ is equal to zero, which are cases we exclude.
+For example, in comparing ancestral state splits $\emptyset$ and $\{1\}$ an ambiguity arises when
+\[
+(1-x_1)(1+x_2)(1+w) = (1+x_1)(1-x_2)(1-w),
+\]
+which holds either if $x_2=0$ and $w=x_1$ or $w=0$ and $x_1=x_2$.
 We reach the same conclusions doing these comparisons for site splits $\{2\}$, $\{3\}$, and $\{1,2,3\}$.
@@
 The remaining site splits $\{1,2\}$, $\{1,3\}$, and $\{2,3\}$ require the following argument.
-For site split $\{1,2\}$, we ignore cases where either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$, and see that an ambiguity in this maximization occurs when
+For site split $\{1,2\}$, we ignore cases where any of $x_1,y_1,x_2,y_2,$ or $w$ are zero and see that an ambiguity in this maximization can occur when
 \[
@@
 This can happen when $x_1=y_2=u_1$ and $x_2=y_1=u_2$ but $u_1 \neq u_2$, though, when $x_1=x_2=u_1$ and $y_1=y_2=u_2$ the partial likelihood takes the same value.
+%das: @EM does the below make sense?
+I know you were worried earlier about arguing the max of the sum of two things, but I think to resolve this ambiguity it's necessary.
+Basically, we want to show we get a unique t from (max_t f(t) + (max_h g(t,h))) for t branch lengths and h ancestral states.
+To do this, if max_h g(t,h) has more than one value we can look at f(t) + max_h g(t,h).
+I.e., max_h g(t,h) happens if either {x_1=y_2,x_2=y_1} or {x_1=x_2,y_1=y_2}, but f(t)|_{x_1=x_2,y_1=y_2} > f(t)|_{x_1=y_2,x_2=y_1} (where |_{} is "restricted to") so the full likelihood will have a unique maximum, and that's what we're concerned about.
 To resolve this ambiguity, we look to the divergence quanitity that is maximized when $x_1=x_2$ and $y_1=y_2$, and, thus, the full joint likelihood obtains a unique maximum when $x_1=x_2$ and $y_1=y_2$.
+It is possible that ancestral state splits $\{1\}$ and $\{2\}$ both maximize the partial likelihood as well, but this case follows a similar argument.
 

### 2018-05-29 appendix.tex
More things to fix

@@
 Additionally, we exclude the infinite branch length case of $w=0$.
-%EM This looks good, but remind us what x^* is here.
-%das: fixed
 Recall the divergence
@@
 This implies the Hessian of $\shannonDivergence$ is negative definite.
-%EM IIUC we only use this fact to show that we have a _negative_ second derivative at the ^* parameter values.
-It would be nice to emphasize that fact, which would also keep people from worrying that we're falling into the "sum of two things with unique optima has a unique optimum."
-%das: Ah, right.
 
@@
     $\begin{array}{l}
-    %EM I must be confused here-- these aren't the values I get at all!
-    % e.g. for this first one everything is 0 except for the first taxon, which is 1. So shouldn't it be
-                   %(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2)\\ ?
-    %das: Ha! I must have written out a different case. Went through everything with a fine-tooth comb and noticed one or two slip-ups but everything is basically the same.
                     (1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2)\\
@@
     $\begin{array}{l}
+    %EM This doesn't seem right to me-- for all of these, shouldn't x_1 and x_2 have the same coefficient?
                     (1-x_1)(1+x_2)(1+w)(1-y_1)(1+y_2)\\
@@
 \]
+%EM It sounds like you are saying that this is the only case in which the above equality can happen, which is not the case, right? Isn't there a whole 2-dimensional set that has this equality?
 which holds either if $x_2=0$ and $w=x_1$ or $w=0$ and $x_1=x_2$.
@@
 I.e., max_h g(t,h) happens if either {x_1=y_2,x_2=y_1} or {x_1=x_2,y_1=y_2}, but f(t)|_{x_1=x_2,y_1=y_2} > f(t)|_{x_1=y_2,x_2=y_1} (where |_{} is "restricted to") so the full likelihood will have a unique maximum, and that's what we're concerned about.
-To resolve this ambiguity, we look to the divergence quanitity that is maximized when $x_1=x_2$ and $y_1=y_2$, and, thus, the full joint likelihood obtains a unique maximum when $x_1=x_2$ and $y_1=y_2$.
+To resolve this ambiguity, we look to the divergence quantity that is maximized when $x_1=x_2$ and $y_1=y_2$, and, thus, the full joint likelihood obtains a unique maximum when $x_1=x_2$ and $y_1=y_2$.
 It is possible that ancestral state splits $\{1\}$ and $\{2\}$ both maximize the partial likelihood as well, but this case follows a similar argument.

### 2018-06-01 appendix.tex
changed argument to identifiability

--- 2018-05-25 appendix.tex
@@
     $\emptyset$ & $p_{\emptyset}$   &$1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
-    $\{1\}$     & $p_{\{1\}}$       &$1 - x_1x_2 +  y_1y_2 +  w[- x_1 + x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
-    $\{2\}$     & $p_{\{2\}}$       &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][- y_1 + y_2] - x_1y_1x_2y_2$\\
-    $\{3\}$     & $p_{\{3\}}$       &$1 - x_1x_2 +  y_1y_2 +  w[  x_1 - x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
-    $\{1,2\}$   & $p_{\{1,2\}}$     &$1 - x_1x_2 -  y_1y_2 +  w[- x_1 + x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
-    $\{1,3\}$   & $p_{\{1,3\}}$     &$1 + x_1x_2 +  y_1y_2 +  w[- x_1 - x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
-    $\{2,3\}$   & $p_{\{2,3\}}$     &$1 - x_1x_2 -  y_1y_2 +  w[  x_1 - x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
-    $\{1,2,3\}$ & $p_{\{1,2,3\}}$   &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][  y_1 - y_2] - x_1y_1x_2y_2$\\
+    $\{1\}$     & $p_{1}$       &$1 - x_1x_2 +  y_1y_2 +  w[- x_1 + x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
+    $\{2\}$     & $p_{2}$       &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][- y_1 + y_2] - x_1y_1x_2y_2$\\
+    $\{3\}$     & $p_{3}$       &$1 - x_1x_2 +  y_1y_2 +  w[  x_1 - x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
+    $\{1,2\}$   & $p_{12}$     &$1 - x_1x_2 -  y_1y_2 +  w[- x_1 + x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
+    $\{1,3\}$   & $p_{13}$     &$1 + x_1x_2 +  y_1y_2 +  w[- x_1 - x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
+    $\{2,3\}$   & $p_{23}$     &$1 - x_1x_2 -  y_1y_2 +  w[  x_1 - x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
+    $\{1,2,3\}$ & $p_{123}$   &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][  y_1 - y_2] - x_1y_1x_2y_2$\\
     \hline
@@
 
-\subsection*{Form of the likelihood}
+\subsection*{Identifiability criteria and form of the likelihood}
 
 Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
-We now show that, in the case of the InvFels tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch parameters to be equal and the bottom two branch parameters to be equal during inference obtains the same maximum likelihood estimate as in the case of arbitrary branch parameters.
+In order for our maximum likelihood problem to be well-posed, we require identifiability conditions on the branch parameters.
+We show that, in the case of the InvFels tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch parameters to be equal and the bottom two branch parameters to be equal during inference yields an identifiable maximum likelihood inference procedure.
+
+We first show that the existence of a unique maximizer implies $x_1=x_2$ and $y_1=y_2$.
 Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
@@
 and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
-All other generating probabilities will differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
+All other generating probabilities differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
 For example, for site split $\{1\}$ we have
@@
 The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
-
-Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood we will have $x_1=x_2$ at the maximum.
+Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood then $x_1=x_2$ at the maximum.
 An analogous statement holds for $y_1$ and $y_2$.
-It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
-We show the Hessian of the joint likelihood term is negative definite, which implies there is a unique maximizer of \eqref{eq:profile_likelihood}.
-We decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
-
-%Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
-We restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
+
+It remains to show the converse: if $x_1=x_2$ and $y_1=y_2$ then we obtain a unique estimate of $\tilde{t}$ when performing joint inference maximizing \eqref{eq:profile_likelihood}.
+To do so, we show the Hessian of the joint likelihood term is negative definite.
+First, we decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
+Assume $x_1=x_2=x$ and $y_1=y_2=y$ so that $t=\{x,y,x,y,w\}$.
+We restrict our attention to $x,y\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
 Additionally, we exclude the infinite branch length case of $w=0$.
@@
 \[
-\shannonDivergence_{\tau_1,t^*}(\tau_1,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}).
-\]
+\shannonDivergence_{\tau_1,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
+\]
+where data is generated on the InvFels tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
 By Gibbs inequality, this function attains a unique maximum when
 \[
-\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)
-\]
-for all $j=1,\ldots,\nSiteSplits$.
-Referring to Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
+\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)
+\]
+for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
+Referring to Table~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
 The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
 \[
--(x^*)^2 = -x_1x_2+y_1w(x_1-x_2)
-\]
-and for $\{1\}$ and $\{1,2\}$ yields
-\[
--(x^*)^2 = -x_1x_2+y_2w(x_2-x_1).
-\]
-The difference of these two equations simplifies to
-\[
-0 = (y_1w+y_2w)(x_1-x_2)
-\]
-and, since $y_1w+y_2w > 0$ by assumption, we have $x_1=x_2$.
-This along with the previous equalities imply $x_1=x_2=x^*$ is the unique maximum of the divergence term for $x_1$ and $x_2$.
-We use the same process to show that, at the maximum, $y_1=y_2=y^*$, and thus the divergence term attains a unique maximum over $x_1,y_1,x_2,y_2\in(0,1)$ with $w \neq 0$.
+2-2(x^*)^2 = 2-2x^2
+\]
+and for $\{2\}$ and $\{1,2\}$ yields
+\[
+2-2(y^*)^2 = 2-2y^2.
+\]
+implying $x=x^*$ and $y=y^*$.
+Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
+\[
+8x^*(y^*)^2 = 8x^*y^*w
+\]
+implying $w=x^*$.
+The unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and thus the Hessian of $\shannonDivergence$ is negative definite.
 
@@
 \[
-\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, \tilde{t}).
-\]
-The partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
-\begin{align*}
-\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,2}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,3}(x_2)\\
-&\qquad +\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,4}(y_2)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,5}(w).
+\tilde{\ell}_{\tau_1,t^*}(\tau_1, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, t).
+\]
+Assume for now the functional form of the partial likelihood is fixed---in other words, each site split has a single partial likelihood function when maximizing over ancestral state splits.
+Then, the partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
+\begin{align*}
+\tilde{\ell}_{\tau_1,t^*}(\tau_1, t) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x}(x)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y}(y)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,w}(w).
 \end{align*}
 Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
-Without loss of generality we focus on the variable $x_1$ and the partial log-likelihood proportional to
-\[
-\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,1}(x_1).
-\]
-Doing calculation as in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, is
-\[
-h_{j,1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
+Without loss of generality we focus on the variable $x$ and the partial log-likelihood proportional to
+\[
+\tilde{\ell}(x) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x}(x).
+\]
+Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to $x$ and the initial $1/32$ constant, is
+\[
+h_{j,x}(x) \propto (1+x)^{e_j}(1-x)^{1-e_j}
 \]
@@
 \begin{align}
-\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
-                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1).
+\tilde{\ell}(x) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x)\\
+                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x). \label{eq:partial-x-form}
 \end{align}
-Assume for now the functional form of the partial likelihood is fixed---in other words, each site split has a maximum at a single ancestral state split.
 We have the second derivative
 \[
-\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
-\]
-As $x_1,y_1,x_2,y_2\in(0,1)$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
-Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
+\tilde{\ell}''(x) = -\left(\frac{\sum_j c_je_j}{(1+x)^2}+\frac{1-\sum_j c_je_j}{(1-x)^2}\right).
+\]
+As $x,y\in(0,1)$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
+Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x) \le 0$.
 Applying similar arguments to the other variables, the Hessian for the partial log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero.
@@
 
-\begin{table}
-\centering
-\begin{tabular}{|l|ll|}
-    \hline
-    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau_1, \tilde{t})$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,\tilde{t})$\\
-    \hline
-     $\{1\}$    &
-    $\begin{array}{l}
-                    \emptyset\\
-                    \{1\}\\
-                    \{2\}\\
-                    \{1,2\}
-                \end{array}$&
-    $\begin{array}{l}
-                    (1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2)\\
-                    (1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2)\\
-                    (1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2)\\
-                    (1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2)
-                \end{array}$\\
-    \hline
-     $\{1,2\}$    &
-    $\begin{array}{l}
-                    \emptyset\\
-                    \{1\}\\
-                    \{2\}\\
-                    \{1,2\}
-                \end{array}$&
-    $\begin{array}{l}
-                    (1-x_1)(1+x_2)(1+w)(1-y_1)(1+y_2)\\
-                    (1-x_1)(1-x_2)(1-w)(1+y_1)(1-y_2)\\
-                    (1+x_1)(1-x_2)(1-w)(1-y_1)(1+y_2)\\
-                    (1+x_1)(1-x_2)(1+w)(1+y_1)(1-y_2)
-                \end{array}$\\
-    \hline
-\end{tabular}
-\caption{Likelihood calculations for site splits $\{1\}$ and $\{1,2\}$ and ancestral state partitions $\ancestralSplitPartition_j$ on the InvFels tree $\tau_1$ with general branch parameters $\tilde{t}$.
-All values multiplied by $1/32$.}
-\label{tab:edge-cases}
-\end{table}
-
-We now consider the case where there may be more than one ancestral state split that maximizes the partial likelihood for a site split.
-Site split $\emptyset$ always attains a maximum at ancestral state split $\emptyset$ provided at least one branch fidelity is not exactly zero.
-Table~\ref{tab:edge-cases} gives the possible forms for the partial likelihood for two cases of site splits.
-For site split $\{1\}$ we ignore ancestral state split $\{2\}$ as its partial likelihood is always less than the ancestral state split $\emptyset$ except for the case $y_1=y_2=w=0$.
-Comparing the remaining ancestral state splits reveals that there is an ambiguity in the form of the likelihood only when either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$.
-We reach the same conclusions doing these comparisons for site splits $\{2\}$, $\{3\}$, and $\{1,2,3\}$.
-
-The remaining site splits $\{1,2\}$, $\{1,3\}$, and $\{2,3\}$ require the following argument.
-For site split $\{1,2\}$, we ignore cases where either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$, and see that an ambiguity in this maximization occurs when
-\[
-(1-x_1)(1+x_2)(1-y_1)(1+y_2) = (1+x_1)(1-x_2)(1+y_1)(1-y_2),
-\]
-or, in other words, when the ancestral state splits $\emptyset$ and $\{1,2\}$ both maximize the partial likelihood.
-This can happen when $x_1=y_2=u_1$ and $x_2=y_1=u_2$ but $u_1 \neq u_2$, though, when $x_1=x_2=u_1$ and $y_1=y_2=u_2$ the partial likelihood takes the same value.
-To resolve this ambiguity, we look to the divergence quanitity that is maximized when $x_1=x_2$ and $y_1=y_2$, and, thus, the full joint likelihood obtains a unique maximum when $x_1=x_2$ and $y_1=y_2$.
-
-Given the above, there exists a unique maximum to the likelihood over $x_1,y_2,x_1,y_2\in(0,1)$ with $w\neq 0$, and, with the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
+We now consider the case where there may be more than one functional form that maximizes the partial likelihood over ancestral state splits for a site split.
+Table~\ref{tab:likelihoods} shows that an ambiguity in the form of the partial likelihood occurs in one of four cases:
+\[
+1: (1-x)^2(1+w)(1+y)^2 = (1+x)^2(1-w)(1+y)^2 > (1+x)^2(1+w)(1-y)^2,
+\]
+\[
+2: (1+x)^2(1-w)(1+y)^2 = (1+x)^2(1+w)(1-y)^2 > (1-x)^2(1+w)(1+y)^2,
+\]
+\[
+3: (1-x)^2(1+w)(1+y)^2 = (1+x)^2(1+w)(1-y)^2 > (1+x)^2(1-w)(1+y)^2,
+\]
+or
+\[
+4: (1-x)^2(1+w)(1+y)^2 = (1+x)^2(1-w)(1+y)^2 = (1+x)^2(1+w)(1-y)^2.
+\]
+In case 1, the equality implies $w=2x/(1+x^2)$.
+Plugging this into Table~\ref{tab:likelihoods}, we see the partial log-likelihood, again suppressing constants with respect to $x$ and the initial $1/32$ constant, is
+\begin{align*}
+\tilde{\ell}(x) &\propto (2-p_1-p_3-p_{12}-p_{23}-2p_{13})\log(1+x) \\
+&\qquad + (p_1+p_3+p_{12}+p_{23}+2p_{13})\log(1-x) \\
+&\qquad + 2\log(1+x)-\log(1+x^2).
+\end{align*}
+Call
+\[
+c := p_1+p_3+p_{12}+p_{23}+2p_{13}
+\]
+so that straightforward calculus reveals
+\begin{align*}
+\tilde{\ell}''(x) &= \frac{4x^2}{(1+x^2)^2}-\frac{4-c}{(1+x)^2}-\frac{c}{(1-x)^2}\\
+                  &= \left(\frac{4x^2}{(1+x^2)^2}-\frac{4}{(1+x)^2}\right)+c\left(\frac{1}{(1+x)^2}-\frac{1}{(1-x)^2}\right).
+\end{align*}
+The term on the right is negative as $c \ge 0$ and $0 < x < 1$.
+The term on the left can be written as
+\[
+\frac{4}{(\frac{1}{x}+x)^2}-\frac{4}{(1+x)^2},
+\]
+which for $0<x<1$ is negative so that $\tilde{\ell}''(x) \le 0$.
+Case 2 follows the same argument but for $y$.
+In case 3, the ambiguity implies $x=y$, which yields a partial log-likelihood of the form \eqref{eq:partial-x-form} multiplied by two, and thus has negative second derivative.
+Finally, case 4 implies $x=y$ and $w=2x/(1+x^2)$, and a combination of the above arguments shows we again have a negative second derivative.
+Thus, for $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, there is a unique maximum of $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$ if and only if $x_1=x_2$ and $y_1=y_2$, yielding our identifiability criteria.
+
+Given the above, with the symmetry of the likelihood we let $x_1=x_2=x$ and $y_1=y_2=y$.
 The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

### 2018-06-15 appendix.tex
argument updates

@@
     $\{3\}$     & $p_{3}$           &$1-x^2+y^2-x^2y^2$\\
+    $\{1,2,3\}$ & $p_{123}$         &$1+x^2-y^2-x^2y^2$\\
     $\{1,2\}$   & $p_{12}$          &$1-x^2-y^2+x^2y^2$\\
+    $\{2,3\}$   & $p_{23}$          &$1-x^2-y^2+x^2y^2$\\
     $\{1,3\}$   & $p_{13}$          &$1+x^2+y^2-4xyw+x^2y^2$\\
-    $\{2,3\}$   & $p_{23}$   &$1-x^2-y^2+x^2y^2$\\
-    $\{1,2,3\}$ & $p_{123}$   &$1+x^2-y^2-x^2y^2$\\
     \hline
@@
 
-It remains to show the converse: if $x_1=x_2$ and $y_1=y_2$ then we obtain a unique estimate of $\tilde{t}$ when performing joint inference maximizing \eqref{eq:profile_likelihood}.
+We now show the converse: if $x_1=x_2$ and $y_1=y_2$ then we obtain a unique estimate of $\tilde{t}$ when performing joint inference maximizing \eqref{eq:profile_likelihood}.
 To do so, we show the Hessian of the joint likelihood term is negative definite.
@@
 \]
-implying $w=x^*$.
+implying $w=y^*$.
 The unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and thus the Hessian of $\shannonDivergence$ is negative definite.
@@
 
+We have shown above that, for each ancestral state split, we obtain a unique maximum performing joint maximization of the likelihood.
+It may be the case that separate ancestral state splits yield different estimates with the same likelihood value.
+As joint inference obtains estimates by maximizing over latent variables, if a single functional form is maximal then there is a single solution that is feasible for the joint optimization procedure to recover.
 We now consider the case where there may be more than one functional form that maximizes the partial likelihood over ancestral state splits for a site split.
-Table~\ref{tab:likelihoods} shows that an ambiguity in the form of the partial likelihood occurs in one of four cases:
+To show joint optimzation results in a unique maximum, it suffices to consider the cases where an ambiguity in the form of the partial likelihood occurs.
+Table~\ref{tab:likelihoods} shows this occurs in one of four cases:
 \[
@@
 \]
-In case 1, the equality implies $w=2x/(1+x^2)$.
-Plugging this into Table~\ref{tab:likelihoods}, we see the partial log-likelihood, again suppressing constants with respect to $x$ and the initial $1/32$ constant, is
-\begin{align*}
-\tilde{\ell}(x) &\propto (2-p_1-p_3-p_{12}-p_{23}-2p_{13})\log(1+x) \\
-&\qquad + (p_1+p_3+p_{12}+p_{23}+2p_{13})\log(1-x) \\
-&\qquad + 2\log(1+x)-\log(1+x^2).
-\end{align*}
-Call
-\[
-c := p_1+p_3+p_{12}+p_{23}+2p_{13}
-\]
-so that straightforward calculus reveals
-\begin{align*}
-\tilde{\ell}''(x) &= \frac{4x^2}{(1+x^2)^2}-\frac{4-c}{(1+x)^2}-\frac{c}{(1-x)^2}\\
-                  &= \left(\frac{4x^2}{(1+x^2)^2}-\frac{4}{(1+x)^2}\right)+c\left(\frac{1}{(1+x)^2}-\frac{1}{(1-x)^2}\right).
-\end{align*}
-The term on the right is negative as $c \ge 0$ and $0 < x < 1$.
-The term on the left can be written as
-\[
-\frac{4}{(\frac{1}{x}+x)^2}-\frac{4}{(1+x)^2},
-\]
-which for $0<x<1$ is negative so that $\tilde{\ell}''(x) \le 0$.
-Case 2 follows the same argument but for $y$.
-In case 3, the ambiguity implies $x=y$, which yields a partial log-likelihood of the form \eqref{eq:partial-x-form} multiplied by two, and thus has negative second derivative.
-Finally, case 4 implies $x=y$ and $w=2x/(1+x^2)$, and a combination of the above arguments shows we again have a negative second derivative.
-Thus, for $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, there is a unique maximum of $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$ if and only if $x_1=x_2$ and $y_1=y_2$, yielding our identifiability criteria.
-
-Given the above, with the symmetry of the likelihood we let $x_1=x_2=x$ and $y_1=y_2=y$.
+In case 1, the equality implies $w=2x/(1+x^2)$, and in case 2 $w=2y/(1+y^2)$.
+Case 3 implies $x=y$, and, finally, case 4 is a combination of these.
+These cases may or may not indicate an identifiability issue with the joint inference procedure, as showing a unique estimate arises in each case is nontrivial.
+While we may lose identifiability, the set of parameters for which these equalities hold and, additionally, do not yield a unique maximum will encompass a minority of the parameter space.
+
+Thus, for $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, there is a unique maximum of $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$ if and only if $x_1=x_2$ and $y_1=y_2$ for the majority of the parameter space, yielding our identifiability criteria.
+With the symmetry of the likelihood we let $x_1=x_2=x$ and $y_1=y_2=y$.
 The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
@@
 the maximum for this site split can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
-For the other site splits we consider two cases.
+For the other site splits we consider three cases.
 If $(1+x)(1-y) > (1-x)(1+y)$, then
@@
 \end{align*}
+Finally, if $(1+x)(1-y) = (1-x)(1+y)$, $x=y$ and the lower bound will be less than the other cases.
 Letting
@@
 for the first lower bound, and we get the equivalent expression for the second lower bound by exchanging $x$ and $y$.
+Moreover, the divergence quanitity is unchanged when exchanging $x$ and $y$ provided $w=0$.
 Maximizing either lower bound over $t$, we obtain

### 2018-06-22 appendix.tex
major rewrite to appendix for general branch lengths

@@
     \hline
-$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x,y,x,y,w\})$\\
+$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x,y,x,y,y\})$\\
     \hline
-    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xyw+x^2y^2$\\
+    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xy^2+x^2y^2$\\
     $\{1\}$     & $p_{1}$           &$1-x^2+y^2-x^2y^2$\\
@@
     $\{2,3\}$   & $p_{23}$          &$1-x^2-y^2+x^2y^2$\\
-    $\{1,3\}$   & $p_{13}$          &$1+x^2+y^2-4xyw+x^2y^2$\\
+    $\{1,3\}$   & $p_{13}$          &$1+x^2+y^2-4xy^2+x^2y^2$\\
     \hline
 \end{tabular}
-\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x,y,x,y,w\}$ obtained using the Hadamard transform.
+\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x,y,x,y,y\}$ obtained using the Hadamard transform.
 All values multiplied by $1/8$.}
@@
 See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the InvFels tree.
-Likelihood computations for all site splits and ancestral state splits are in Tables~\ref{tab:farris_likelihoods} (for the InvFels tree) and~\ref{tab:fels_likelihoods} (for the Felsenstein tree).
-Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}.
+Likelihood computations for all site splits and ancestral state splits are in Table~\ref{tab:farris_likelihoods} for the InvFels tree.
+Taking maxima over ancestral state splits results in Table~\ref{tab:likelihoods}.
 
@@
 \caption{
-    Example likelihood computations on the InvFels tree $\tau_1$ for fidelities $x$, $y$, and $w$.
+    Example likelihood computations on the InvFels tree $\tau_1$ for fidelities $\{x_1,y_1,x_2,y_2,w\}$.
     Edges labeled by the probability of substitution along that edge.
-    In (a), we compute the product to obtain $\Pr(\ancestralSplitRV=\emptyset\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1+w)/32$.
-    In (b), the same process yields $\Pr(\ancestralSplitRV=\{1\}\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1-w)/32$.
+    In (a), we compute the product to obtain $\Pr(\ancestralSplitRV=\emptyset\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x_1)(1-x_2)(1+y_1)(1-y_2)(1+w)/32$.
+    In (b), the same process yields $\Pr(\ancestralSplitRV=\{1\}\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x_1)(1-x_2)(1+y_1)(1-y_2)(1-w)/32$.
 }
@@
 \end{figure}
+%TODO: update this figure with general branch lengths
 
@@
 \centering
-\begin{tabular}{|l|ll|}
-\multicolumn{3}{c}{$\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_1, t)$}\\
+\begin{tabular}{|ll|l|}
 \hline
-& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
+$\siteSplit_j$ & $\ancestralSplit_k$ & $\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_1, t)$\\
 \hline
-    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
-    \hline
-     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$          &$(1+x)^2   (1-w)(1-y)^2$\\
-     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
-     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
-     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
-     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{1,3\}$     &$(1-x)^2   (1+w)(1+y)^2$          &$(1-x)^2   (1-w)(1-y)^2$\\
-     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{1,2,3\}$   &$(1-x)^2   (1+w)(1+y)(1-y)$       &$(1-x)^2   (1-w)(1+y)(1-y)$\\
-    \hline
-    \hline
-    &$\{1\}$                             &$\{1,2\}$  \\
-    \hline
-     $\emptyset$   &$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
-     $\{1\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{2\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-     $\{3\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-     $\{1,3\}$     &$(1+x)^2   (1-w)(1+y)^2$     &$(1+x)^2   (1+w)(1-y)^2$\\
-     $\{2,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-     $\{1,2,3\}$   &$(1+x)^2   (1-w)(1+y)(1-y)$  &$(1+x)^2   (1+w)(1+y)(1-y)$\\
+$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$\\
+&$\{2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$\\
+&$\{1,2\}$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$\\
+
+$\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$\\
+&$\{2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$\\
+&$\{1,2\}$&$(1+x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$\\
+
+$\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1-w)$\\
+&$\{2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1-w)$\\
+&$\{1,2\}$&$(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1+w)$\\
+
+$\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
+&$\{2\}$&$(1+x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$\\
+&$\{1,2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
+
+$\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
+&$\{2\}$&$(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1-w)$\\
+&$\{1,2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
+
+$\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1-w)$\\
+&$\{2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1-w)$\\
+&$\{1,2\}$&$(1+x_1)(1+y_1)(1-x_2)(1-y_2)(1+w)$\\
+
+$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
+&$\{2\}$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$\\
+&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
+
+$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
+&$\{2\}$&$(1+x_1)(1+y_1)(1-x_2)(1-y_2)(1-w)$\\
+&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
 \hline
@@
 \centering
-\begin{tabular}{|l|ll|}
-\multicolumn{3}{c}{$\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_2, t)$}\\
-\hline
-& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
-    \hline
-    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
-    \hline
-     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1-x)^2   (1-w)(1+y)(1-y)$\\
-     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1-y)^2$\\
-     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1+x)^2   (1-w)(1+y)(1-y)$\\
-     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1-x)^2   (1-w)(1-y)^2$\\
-     $\{1,3\}$     &$(1-x)^2   (1+w)(1+y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
-     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1+x)^2   (1-w)(1-y)^2$\\
-     $\{1,2,3\}$   &$(1-x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1-y)^2$\\
-    \hline
-    \hline
-    &$\{1\}$                             &$\{1,2\}$  \\
-    \hline
-     $\emptyset$   &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1-x)^2   (1+w)(1-y)^2$\\
-     $\{1\}$       &$(1+x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{2\}$       &$(1+x)(1-x)(1-w)(1+y)^2$           &$(1-x)^2   (1+w)(1+y)(1-y)$\\
-     $\{3\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
-     $\{1,2\}$     &$(1+x)^2   (1-w)(1+y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-     $\{1,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1+x)^2   (1+w)(1-y)^2$\\
-     $\{2,3\}$     &$(1-x)^2   (1-w)(1+y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-     $\{1,2,3\}$   &$(1+x)(1-x)(1-w)(1+y)^2$           &$(1+x)^2   (1+w)(1+y)(1-y)$\\
-\hline
-\end{tabular}
-\caption{Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the Felsenstein tree $\tau_2$.
-All values multiplied by $1/32$.}
-\label{tab:fels_likelihoods}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{|l|ll|}
-    \multicolumn{3}{c}{InvFels tree ($\tau=\tau_1$)}\\
+\begin{tabular}{|ll|l|}
 \hline
@@
 \hline
-    $\emptyset$&
-    $\emptyset$&
-    $(1+x)^2   (1+w)(1+y)^2$\\
-     $\{1\}$    &
-    $\emptyset$&
-    $(1+x)(1-x)(1+w)(1+y)^2$\\
-     $\{2\}$    &
-    $\emptyset$&
-    $(1+x)^2   (1+w)(1+y)(1-y)$\\
-     $\{3\}$    &
-    $\emptyset$&
-    $(1+x)(1-x)(1+w)(1+y)^2$\\
-    $\{1,2\}$  &
-    $\{\emptyset,\{1,2\}\}$&
-    $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-    $\{1,3\}$  &
-    $\left\{\begin{array}{l}
-                    \emptyset\\
-                    \{1\}\\
-                    \{1,2\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                    (1-x)^2   (1+w)(1+y)^2\\
-                    (1+x)^2   (1-w)(1+y)^2\\
-                    (1+x)^2   (1+w)(1-y)^2
-                \end{array}$\\
-    $\{2,3\}$  &
-                $\{\emptyset,\{1,2\}\}$&
-                $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
-    $\{1,2,3\}$&
-                $\{1,2\}$&
-                $(1+x)^2   (1+w)(1+y)(1-y)$\\
-    \hline
-    \multicolumn{3}{c}{Felsenstein tree ($\tau=\tau_2$)}\\
-    \hline
-    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
-    \hline
-    $\emptyset$       &$\emptyset$&$(1+x)^2   (1+w)(1+y)^2$\\
-    $\{1\}$          &
-    $\left\{\begin{array}{l}
-                    \emptyset\\
-                    \{1\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                        (1+x)(1-x)(1+w)(1+y)^2\\
-                        (1+x)^2   (1-w)(1+y)(1-y)
-                    \end{array}$\\
-      $\{2\}$          &
-    $\left\{\begin{array}{l}
-                    \emptyset\\
-                    \{1\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                    (1+x)^2   (1+w)(1+y)(1-y)\\
-                    (1+x)(1-x)(1-w)(1+y)^2
-                    \end{array}$\\
-      $\{3\}$          &
-    $\left\{\begin{array}{l}
-                    \emptyset\\
-                    \{2\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                    (1+x)(1-x)(1+w)(1+y)^2\\
-                    (1+x)^2   (1-w)(1+y)(1-y)
-                    \end{array}$\\
-     $\{1,2\}$         &
-    $\left\{\begin{array}{l}
-                    \{\emptyset,\{1,2\}\}\\
-                    \{1\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
-                    (1+x)^2   (1-w)(1+y)^2
-                    \end{array}$\\
-     $\{1,3\}$         &
-    $\left\{\begin{array}{l}
-                    \emptyset\\
-                    \{\{1\},\{2\}\}\\
-                    \{1,2\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                    (1-x)^2   (1+w)(1+y)^2\\
-                    (1+x)(1-x)(1-w)(1+y)(1-y)\\
-                    (1+x)^2   (1+w)(1-y)^2
-                    \end{array}$\\
-      $\{2,3\}$        &
-    $\left\{\begin{array}{l}
-                    \{\emptyset,\{1,2\}\}\\
-                    \{1\}\\
-                    \{2\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
-                    (1-x)^2   (1-w)(1+y)^2\\
-                    (1+x)^2   (1-w)(1-y)^2
-                    \end{array}$\\
-     $\{1,2,3\}$       &
-    $\left\{\begin{array}{l}
-                    \{2\} \\
-                    \{1,2\}
-                \end{array}\right.$&
-    $\begin{array}{l}
-                    (1+x)(1-x)(1-w)(1+y)^2 \\
-                    (1+x)^2   (1+w)(1+y)(1-y)
-                    \end{array}$\\
+$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
+
+$\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
+
+$\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
+
+$\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
+&$\{1,2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
+
+$\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
+&$\{1,2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
+
+$\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
+
+$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
+&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
+
+$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
+&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
 \hline
 \end{tabular}
-\caption{Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$ after maximizing over ancestral states on the InvFels tree $\tau_1$ and the Felsenstein tree $\tau_2$.
+\caption{
+Due to the symmetry of the likelihood, WLOG we assume $x_2 > x_1$ and $y_2 > y_1$ and maximize over ancestral states to reduce the number of possible functional forms to consider.
 All values multiplied by $1/32$.
 Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
-See Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} for full calculations.}
+See Table~\ref{tab:farris_likelihoods} for full calculations.}
 \label{tab:likelihoods}
@@
 
-\subsection*{Identifiability criteria and form of the likelihood}
+\subsection*{Properties of the joint objective function}
 
 Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
-In order for our maximum likelihood problem to be well-posed, we require identifiability conditions on the branch parameters.
-We show that, in the case of the InvFels tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch parameters to be equal and the bottom two branch parameters to be equal during inference yields an identifiable maximum likelihood inference procedure.
-
-We first show that the existence of a unique maximizer implies $x_1=x_2$ and $y_1=y_2$.
+We show that the likelihood remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
 Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
@@
 
-We now show the converse: if $x_1=x_2$ and $y_1=y_2$ then we obtain a unique estimate of $\tilde{t}$ when performing joint inference maximizing \eqref{eq:profile_likelihood}.
+We now show that, if there exists a unique maximal ancestral state split for each site split then there exists a a unique estimate of $\tilde{t}$ when performing joint inference maximizing \eqref{eq:profile_likelihood}.
 To do so, we show the Hessian of the joint likelihood term is negative definite.
 First, we decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
-Assume $x_1=x_2=x$ and $y_1=y_2=y$ so that $t=\{x,y,x,y,w\}$.
-We restrict our attention to $x,y\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
-Additionally, we exclude the infinite branch length case of $w=0$.
 Recall the divergence
@@
 for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
+Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the divergence, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
 Referring to Table~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
@@
 \begin{align*}
-\tilde{\ell}_{\tau_1,t^*}(\tau_1, t) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x}(x)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y}(y)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,w}(w).
+\tilde{\ell}_{\tau_1,t^*}(\tau_1, t) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_2}(x_2)\\
+&\qquad + \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,w}(w).
 \end{align*}
 Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
-Without loss of generality we focus on the variable $x$ and the partial log-likelihood proportional to
+Without loss of generality we focus on the variable $x_1$ and the partial log-likelihood proportional to
 \[
-\tilde{\ell}(x) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x}(x).
+\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1).
 \]
-Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to $x$ and the initial $1/32$ constant, is
+Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, is
 \[
-h_{j,x}(x) \propto (1+x)^{e_j}(1-x)^{1-e_j}
+h_{j,x_1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
 \]
@@
 \begin{align}
-\tilde{\ell}(x) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x)\\
-                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x). \label{eq:partial-x-form}
+\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
+                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1). \label{eq:partial-x-form}
 \end{align}
@@
 \[
-\tilde{\ell}''(x) = -\left(\frac{\sum_j c_je_j}{(1+x)^2}+\frac{1-\sum_j c_je_j}{(1-x)^2}\right).
+\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
 \]
-As $x,y\in(0,1)$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
-Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x) \le 0$.
+As $x_1\in(0,1)$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
+Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
 Applying similar arguments to the other variables, the Hessian for the partial log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero.
@@
 
-We have shown above that, for each ancestral state split, we obtain a unique maximum performing joint maximization of the likelihood.
-It may be the case that separate ancestral state splits yield different estimates with the same likelihood value.
-As joint inference obtains estimates by maximizing over latent variables, if a single functional form is maximal then there is a single solution that is feasible for the joint optimization procedure to recover.
-We now consider the case where there may be more than one functional form that maximizes the partial likelihood over ancestral state splits for a site split.
-To show joint optimzation results in a unique maximum, it suffices to consider the cases where an ambiguity in the form of the partial likelihood occurs.
-Table~\ref{tab:likelihoods} shows this occurs in one of four cases:
-\[
-1: (1-x)^2(1+w)(1+y)^2 = (1+x)^2(1-w)(1+y)^2 > (1+x)^2(1+w)(1-y)^2,
-\]
-\[
-2: (1+x)^2(1-w)(1+y)^2 = (1+x)^2(1+w)(1-y)^2 > (1-x)^2(1+w)(1+y)^2,
-\]
-\[
-3: (1-x)^2(1+w)(1+y)^2 = (1+x)^2(1+w)(1-y)^2 > (1+x)^2(1-w)(1+y)^2,
-\]
-or
-\[
-4: (1-x)^2(1+w)(1+y)^2 = (1+x)^2(1-w)(1+y)^2 = (1+x)^2(1+w)(1-y)^2.
-\]
-In case 1, the equality implies $w=2x/(1+x^2)$, and in case 2 $w=2y/(1+y^2)$.
-Case 3 implies $x=y$, and, finally, case 4 is a combination of these.
-These cases may or may not indicate an identifiability issue with the joint inference procedure, as showing a unique estimate arises in each case is nontrivial.
-While we may lose identifiability, the set of parameters for which these equalities hold and, additionally, do not yield a unique maximum will encompass a minority of the parameter space.
-
-Thus, for $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, there is a unique maximum of $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$ if and only if $x_1=x_2$ and $y_1=y_2$ for the majority of the parameter space, yielding our identifiability criteria.
-With the symmetry of the likelihood we let $x_1=x_2=x$ and $y_1=y_2=y$.
-The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
-
 \subsection*{Theorems and proofs}
 
-\topoInconsist*
+Tentative theorem 1: 
+\begin{restatable}{theorem}{topoInconsist2}
+Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$.
+There exists a set of $0 < x^*, y^* < 1$ such that
+\[
+\{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\} = \arg\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t)
+\]
+and $\hat{w}\equiv 1$.
+\end{restatable}
 
-\begin{proof}
-Assume $\tau^*=\tau_1$.
-The InvFels tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see the upper table of Table~\ref{tab:likelihoods}).
+In words, this implies an inconsistency as we converge on the incorrect topology, estimating $\hat{w}$ to be $1$ and not $y^*$.
 
-We write out the case for the ancestral state split $\{1\}$, and show the other two cases follow a similar argument.
-Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_profile_likelihood_mean}, the log-likelihood is, suppressing normalizing constants $-\log 8$ (from the first additive term of \eqref{eq:site_pattern_profile_likelihood_mean}) and $-\log 32$ (from the second),
-\begin{align}
-    \label{eq:farris_likelihood}
-    \ell_{\tau_1,t^*}(\tau_1, t)
-    &=        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw+x^2y^2) \nonumber \\
-    &\qquad + p_{1}          \cdot\log(1-x^2+y^2-x^2y^2) \nonumber \\
-    &\qquad + p_{2}          \cdot\log(1+x^2-y^2-x^2y^2) \nonumber \\
-    &\qquad + p_{3}          \cdot\log(1-x^2+y^2-x^2y^2) \nonumber \\
-    &\qquad + p_{12}         \cdot\log(1-x^2-y^2+x^2y^2) \nonumber \\
-    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw+x^2y^2) \nonumber \\
-    &\qquad + p_{23}         \cdot\log(1-x^2-y^2+x^2y^2) \nonumber \\
-    &\qquad + p_{123}        \cdot\log(1+x^2-y^2-x^2y^2) \nonumber \\
-    &\qquad + p_{\emptyset}  \cdot\log((1+x)^2   (1+w)(1+y)^2) \nonumber \\
-    &\qquad + p_{1}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \nonumber \\
-    &\qquad + p_{2}          \cdot\log((1+x)^2   (1+w)(1+y)(1-y)) \nonumber \\
-    &\qquad + p_{3}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \nonumber \\
-    &\qquad + p_{12}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y)) \nonumber \\
-    &\qquad + p_{13}         \cdot\log((1+x)^2   (1-w)(1+y)^2) \nonumber \\
-    &\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y)) \nonumber \\
-    &\qquad + p_{123}        \cdot\log((1+x)^2   (1+w)(1+y)(1-y)).
-\end{align}
-Our plan is to simplify the likelihood to remove log-of-quadratic terms of $x,y,$ and $w$, obtaining an upper bound on the likelihood that has a closed-form maximum in each variable.
-To bound the generating probabilities, we use the facts that, for $x,y\in[0,1]$,
+Tentative lemma:
+\begin{lemma}
+Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$.
+There exists a set of $0 < x^*, y^* < 1$ such that $\emptyset$ is the maximal ancestral state for all site splits.
+\end{lemma}
+
+This is useful as it will result in the largest estimates of $w$ and $y_2$ and a single objective function with a unique maximum.
+Empirically, it looks as if, for small $x$ and large $y$, this results in $\hat{w} = \hat{y}_2 = 1$.
+We can see easily that, if this is the case, $\hat{x}_1=\hat{x}_2=x*(y^*)^2$ and $\hat{y}_1=(y^*)^2$.
+We then need to show
+\begin{enumerate}
+\item there exist $x^*,y^*$ that give us this form for the likelihood
+\item there exist $x^*,y^*$ such that this likelihood for $w$ and $y_2$ has a maximum on the boundary
+\item the intersection of these sets of $x^*,y^*$ is nonempty
+\end{enumerate}
+
+Note that the above will hold if we swap $\hat{y}_2$ and $\hat{y}_1$, as the likelihood is invariant to this.
+
+\subsection*{Empirical validation}
+
+Consider Table~\ref{tab:likelihoods}.
+There are 81 separate possible forms for the likelihood, corresponding to the three possible most likely ancestral states for each of the four ambiguous site splits.
+Moreover, as shown previously, each of these objective functions has a unique maximum and is concave on $[0,1]^5$.
+
+Assume $w=1$ and $y_2=1$.
+Then, there is a closed form solution maximizing the joint optimization procedure.
+In this case, the likelihood has no ambiguity, and has terms given in Table~\ref{tab:likelihoods-restricted}.
+
+\begin{table}
+\centering
+\begin{tabular}{|ll|l|}
+\hline
+$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
+\hline
+$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)$\\
+
+$\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)$\\
+
+$\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)$\\
+
+$\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)$\\
+
+$\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)$\\
+
+$\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)$\\
+
+$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)$\\
+
+$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)$\\
+\hline
+\end{tabular}
+\caption{
+The maximal likelihood values given $w=1$ and $y_2=1$.
+All values multiplied by $1/32$.}
+\label{tab:likelihoods-restricted}
+\end{table}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{|l|l|l|}
+    \hline
+$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,1,1\})$\\
+    \hline
+    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1x_2 +  y_1 +  [  x_1 + x_2][  y_1 + 1] + x_1y_1x_2 = (1+x_1)(1+y_1)(1+x_2)$\\
+    $\{1\}$     & $p_{1}$       &$1 - x_1x_2 +  y_1 +  [- x_1 + x_2][  y_1 + 1] - x_1y_1x_2 = (1-x_1)(1+y_1)(1+x_2)$\\
+    $\{2\}$     & $p_{2}$       &$1 + x_1x_2 -  y_1 +  [  x_1 + x_2][- y_1 + 1] - x_1y_1x_2 = (1+x_1)(1-y_1)(1+x_2)$\\
+    $\{3\}$     & $p_{3}$       &$1 - x_1x_2 +  y_1 +  [  x_1 - x_2][  y_1 + 1] - x_1y_1x_2 = (1+x_1)(1+y_1)(1-x_2)$\\
+    $\{1,2,3\}$ & $p_{123}$   &$1 + x_1x_2 -  y_1 +  [  x_1 + x_2][  y_1 - 1] - x_1y_1x_2 = (1-x_1)(1-y_1)(1-x_2)$\\
+    $\{1,2\}$   & $p_{12}$     &$1 - x_1x_2 -  y_1 +  [- x_1 + x_2][- y_1 + 1] + x_1y_1x_2 = (1-x_1)(1-y_1)(1+x_2)$\\
+    $\{1,3\}$   & $p_{13}$     &$1 + x_1x_2 +  y_1 +  [- x_1 - x_2][  y_1 + 1] + x_1y_1x_2 = (1-x_1)(1+y_1)(1-x_2)$\\
+    $\{2,3\}$   & $p_{23}$     &$1 - x_1x_2 -  y_1 +  [  x_1 - x_2][- y_1 + 1] + x_1y_1x_2 = (1+x_1)(1-y_1)(1-x_2)$\\
+    \hline
+\end{tabular}
+\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,1,1\}$ obtained using the Hadamard transform.
+All values multiplied by $1/8$.}
+\label{tab:gen-sitepatprob-restricted}
+\end{table}
+
+Additionally, we simplify the generating probabilities as in Table~\ref{tab:gen-sitepatprob-restricted}.
+We see that, on simplifying, for each site split the generating probability is the same functional form as the likelihood.
+Thus, the joint optimization procedure has a unique, closed form solution solving
 \begin{align*}
-p_{12} = p_{23} \propto 1-x^2-y^2+x^2y^2 & = (1+x)(1-x)(1+y)(1-y) \\
-p_{1} = p_{3} \propto 1-x^2+y^2-x^2y^2 & = (1+x)(1-x)(1+y^2) \le (1+x)(1-x)(1+y) \\
-p_{2} = p_{123} \propto 1+x^2-y^2-x^2y^2 & = (1+x^2)(1+y)(1-y) \le (1+x)(1+y)(1-y).
+&\arg\max_{x_1,y_y,x_2} \ (p_{\emptyset} + p_2 + p_3 + p_{23}) \log(1+x_1) + (p_1 + p_{123} + p_{12} + p_{13}) \log(1-x_1)\\
+               &\qquad + (p_{\emptyset} + p_1 + p_3 + p_{13}) \log(1+y_1) + (p_2 + p_{123} + p_{12} + p_{23}) \log(1-y_1)\\
+               &\qquad + (p_{\emptyset} + p_1 + p_2 + p_{12}) \log(1+x_2) + (p_3 + p_{123} + p_{13} + p_{23}) \log(1-x_2)\\
+&\propto\arg\max_{x_1,y_y,x_2} \ (4+4x*(y^*)^2) \log(1+x_1) + (4-4x*(y^*)^2) \log(1-x_1)\\
+               &\qquad + (4+4(y^*)^2) \log(1+y_1) + (4-4(y^*)^2) \log(1-y_1)\\
+               &\qquad + (4+4x*(y^*)^2) \log(1+x_2) + (4-4x*(y^*)^2) \log(1-x_2).
 \end{align*}
-To bound the remaining $p_{\emptyset}$ and $p_{13}$, we use
-\begin{align*}
-1+x^2+y^2+x^2y^2 & = (1+x^2)(1+y^2) \le (1+x)(1+y) \\
-4xy & = 2x \cdot 2y \le (1+x^2)(1+y^2) \le (1+x)(1+y)
-\end{align*}
-so that the $p_{\emptyset}$ term is bounded as
-\begin{align*}
-    p_{\emptyset} & \propto 1+x^2+y^2+4xyw+x^2y^2 \\
-                  & \le 1+x^2+y^2+4xy+x^2y^2 \\
-                  & \le 2(1+x)(1+y)
-\end{align*}
-and $p_{13}$ is bounded as
-\begin{align*}
-    p_{13} & \propto 1+x^2+y^2-4xyw+x^2y^2 \\
-                  & \le 1+x^2+y^2+x^2y^2 \\
-                  & \le (1+x)(1+y).
-\end{align*}
-Factoring and making these substitutions (again, under the assumption that the ancestral state split is $\{1\}$) results in
-\begin{align*}
-&    \ell_{\tau_1,t^*}(\tau_1, t) \\
-&\qquad \le      p_{\emptyset}  \cdot\log(2(1+x)(1+y))
-+ p_{1}          \cdot\log((1+x)(1-x)(1+y)) \\
-    &\qquad\qquad + p_{2}          \cdot\log((1+x)(1+y)(1-y))
-+ p_{3}          \cdot\log((1+x)(1-x)(1+y)) \\
-    &\qquad\qquad + p_{12}         \cdot\log((1+x)(1-x)(1+y)(1-y))
-+ p_{13}         \cdot\log((1+x)(1+y)) \\
-    &\qquad\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+y)(1-y))
-+ p_{123}        \cdot\log((1+x)(1+y)(1-y)) \\
-    &\qquad\qquad + p_{\emptyset}  \cdot\log((1+x)^2   (1+w)(1+y)^2)
-+ p_{1}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \\
-    &\qquad\qquad + p_{2}          \cdot\log((1+x)^2   (1+w)(1+y)(1-y))
-+ p_{3}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \\
-    &\qquad\qquad + p_{12}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y))
-+ p_{13}         \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
-    &\qquad\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y)) \\
-    &\qquad\qquad + p_{123}        \cdot\log((1+x)^2   (1+w)(1+y)(1-y)) \\
-&\qquad =      p_{\emptyset}  \cdot\log(2)
-+ \log(1+x)
-+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x) \\
-&\qquad\qquad + \log(1+y)
-+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
-&\qquad\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x)
-+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x) \\
-&\qquad\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y)
-+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
-&\qquad\qquad + (1-p_{13})\cdot\log(1+w)
-+ p_{13}\cdot\log(1-w) \\
-&\qquad =      p_{\emptyset}  \cdot\log(2)
-+ (3-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x)
-+ 2(p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x) \\
-&\qquad\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y)
-+ 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
-&\qquad\qquad + (1-p_{13})\cdot\log(1+w)
-+ p_{13}\cdot\log(1-w).
-\end{align*}
-To simplify, let
-\begin{equation}
-    \begin{aligned}
-        a_{1} &= 3-p_{1}-p_{3}-p_{12}-p_{23}, \\
-        a_{2} &= 2(p_{1}+p_{3}+p_{12}+p_{23}), \\
-        a_{3} &= 3-p_{2}-p_{12}-p_{23}-p_{123}, \\
-        a_{4} &= 2(p_{2}+p_{12}+p_{23}+p_{123}), \\
-    \end{aligned}
-    \label{eq:a_const}
-\end{equation}
-so that
-\begin{equation*}
-\begin{split}
-&    \ell_{\tau_1,t^*}(\tau_1, t) \\
-&\qquad \le      p_{\emptyset}  \cdot\log(2)
-+ a_{1}\cdot\log(1+x)
-+ a_{2}\cdot\log(1-x)
-+ a_{3}\cdot\log(1+y) \\
-&\qquad\qquad + a_{4}\cdot\log(1-y)
-+ (1-p_{13})\cdot\log(1+w)
-+ p_{13}\cdot\log(1-w).
-\end{split}
-\end{equation*}
-Maximizing over the unknown terms yields
+Using the fact that $a_1\log(1+x)+a_2\log(1-x)$ attains a maximum at $\hat{x} = (a_1-a_2)/(a_1+a_2)$ we have the parameters
 \[
-\hat{x} = \frac{a_{1}-a_{2}}{a_{1}+a_{2}}, \ \hat{y} = \frac{a_{3}-a_{4}}{a_{3}+a_{4}}, \ \hat{w} = 1-2p_{13}.
+\{\hat{x}_1,\hat{y}_1,\hat{x}_2\} = \{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2\}
 \]
-Trivially $0 \le \hat{w} \le 1$ and $\hat{x}, \hat{y} \le 1$.
-We check that $\hat{x}, \hat{y} \ge 0$ to ensure these maxima are valid.
-For $\hat{x} \ge 0$ we need $a_1 \ge a_2$; letting
-\[
-\tilde{p} = p_{1}+p_{3}+p_{12}+p_{23}
-\]
-we have from \eqref{eq:a_const} that $a_1 \ge a_2$ only if $3 \ge 3\tilde{p}$.
-Since $\tilde{p} = 1/8\cdot(4-4(x^*)^2)$, we have $0 \le \tilde{p} \le 1/2$ implying $\hat{x} \ge 0$.
-The same approach works for $\hat{y}$.
-The upper bound for the likelihood is then maximized at
-\begin{align}
-\begin{split}
-&    \ell_{\tau_1,t^*}(\tau_1, t) \\
-&\qquad\le      p_{\emptyset}  \cdot\log(2)
-+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
-+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}}
-+ a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
-&\qquad\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
-+ (1-p_{13})\cdot\log(2(1-p_{13}))
-+ p_{13}\cdot\log(2p_{13}) \\
-&\qquad := C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*).
-\end{split}
-\label{eq:farris-upper-bound-1}
-\end{align}
+and a closed form function of $x^*,y^*$ for this likelihood.
 
-The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications, as we now describe.
-The upper bound $C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*)$ in \eqref{eq:farris-upper-bound-1} is for the ancestral state split $\{1\}$.
-For the ancestral state split $\emptyset$ we can derive an upper bound $C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*)$ in \eqref{eq:farris-upper-bound-empty} in a similar fashion, and $C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*)$ will be the same as $C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*)$ except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.
-In this case, all terms involving $w$ simplify to $\log(1+w)$, which is maximized at $\hat{w}=1$.
-For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
-\begin{equation}
-    \begin{aligned}
-        a_{1}' &= a_{1}-2p_{13}, \\
-        a_{2}' &= a_{2}+2p_{13}, \\
-    \end{aligned}
-    \label{eq:a_const_prime_x}
-\end{equation}
-for ancestral state split $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$, yielding the upper bound
-\begin{align}
-\begin{split}
-&    \ell_{\tau_1,t^*}(\tau_1, t) \\
-&\qquad\le      p_{\emptyset}  \cdot\log(2)
-+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
-+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
-&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
-+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
-+ \log(2) \\
-&\qquad := C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*).
-\end{split}
-\label{eq:farris-upper-bound-empty}
-\end{align}
-
-For ancestral state split $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$, we have constants
-\begin{equation}
-    \begin{aligned}
-        a_{3}' &= a_{3}-2p_{13}, \\
-        a_{4}' &= a_{4}+2p_{13}, \\
-    \end{aligned}
-    \label{eq:a_const_prime_y}
-\end{equation}
-and the upper bound
-\begin{align}
-\begin{split}
-&    \ell_{\tau_1,t^*}(\tau_1, t) \\
-&\qquad\le      p_{\emptyset}  \cdot\log(2)
-+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
-+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
-&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
-+ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
-+ \log(2) \\
-&\qquad := C^{\{1,2\}}_{\tau_1,\tau_1}(x^*, y^*).
-\end{split}
-\label{eq:farris-upper-bound-12}
-\end{align}
-
-We now need to check that
-\[
-\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
-\]
-are both between zero and one.
-Using a similar argument as that where $\ancestralSplitPartition_j(\tau_1, t) = \{1\}$, we only need to show $a_1' \ge a_2'$, which is true if $3-2p_{13} \ge 3\tilde{p}+2p_{13}$.
-Some rearranging and the fact that $0 \le \tilde{p} \le 1/2$ shows this is equivalent to showing $p_{13} \le 3/8$.
-Using Table~\ref{tab:sitepatprob} while setting $t=t^*=\{x^*,y^*,x^*,y^*,y^*\}$,
-\[
-p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right),
-\]
-meaning we are interested in whether
-\begin{equation}
-1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3.
-\label{eq:generating_ineq}
-\end{equation}
-We see that
-\begin{equation}
-\begin{split}
-& 1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3 \\
-& \qquad \qquad \qquad \qquad \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 2 - (x^*)^2
-\end{split}
-\label{eq:lenny}
-\end{equation}
-and that
-\begin{align*}
-(y^*)^2\left(1 - 4x^* + (x^*)^2\right) &\le (y^*)^2\left(1 - 2x^* + (x^*)^2\right) \\
-                                       &=   (y^*)^2\left(1 - x^*\right)^2 \\
-                                       &\le        \left(1 - x^*\right)^2 \\
-                                       &\le              2 - (x^*)^2,
-\end{align*}
-with the last inequality holding if $0 \le x^* \le 1$, showing that \eqref{eq:lenny} holds.
-Thus, $p_{13} \le 3/8$ and our $\hat{x}$ and $\hat{y}$ are valid.
-
-All bounds are functions only of $x,y$ through the true generating probabilities $x^*, y^*$.
-Call the upper bound
-\[
-C_0(x^*, y^*) := \max\left(C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*), C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*), C^{\{1,2\}}_{\tau_1,\tau_1}(x^*, y^*)\right).
-\]
-
-We next construct a lower bound on the Felsenstein tree partial likelihood.
-In general, taking
-\[
-\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t).
-\]
-we see
-\[
-\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \ge
-    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
-    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
-\]
-is a lower bound for the joint maximum of \eqref{eq:log_likelihood_simplified}.
-In the Felsenstein case, there are many more optimal internal states depending on branch lengths than in the InvFels case (see the lower table of Table~\ref{tab:likelihoods}).
-We first bound the entire likelihood below and then proceed with joint inference.
-To obtain a lower bound for the likelihood we replace $1+w$ with $1-w$ in Table~\ref{tab:likelihoods} for $\tau = \tau_2$; in this case, we resolve many of the ambiguous likelihood terms.
-For example, for site split $\{1,2\}$, after we replace the $1+w$ term for $\ancestralSplitPartition_j(\tau, t)=\{\emptyset, \{1,2\}\}$ with $1-w$, we have
-\[
-\max\left((1+x)^2 (1-w)(1+y)^2, (1+x)(1-x)(1-w)(1+y)(1-y)\right)
-\]
-where since
-\[
-(1+x)^2 (1-w)(1+y)^2 > (1+x)(1-x)(1-w)(1+y)(1-y),
-\]
-the maximum for this site split can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
-For the other site splits we consider three cases.
-If $(1+x)(1-y) > (1-x)(1+y)$, then
-\[
-(1+x)^2(1-w)(1+y)(1-y) > (1+x)(1-x)(1-w)(1+y)^2
-\]
-meaning the cases $\{1\}, \{2\}, \{3\}$ and $\{1,2,3\}$ will have likelihoods bounded below by
-\[
-(1+x)^2(1-w)(1+y)(1-y).
-\]
-For the cases $\{1,3\}$ and $\{2,3\}$ and the same condition, we have
-\[
-(1+x)^2(1-w)(1-y)^2 > (1-x)^2(1-w)(1+y)^2
-\]
-and
-\[
-(1+x)^2(1-w)(1-y)^2 > (1+x)(1-x)(1-w)(1+y)(1-y),
-\]
-yielding lower bounds of
-\[
-(1+x)^2(1-w)(1-y)^2
-\]
-in these cases.
-The partial likelihood in this case is then bounded below by
-\begin{align*}
-    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
-    &\ge      p_{\emptyset}  \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
-    &\qquad + p_{1}          \cdot\log((1+x)^2   (1-w)(1+y)(1-y)) \\
-    &\qquad + p_{2}          \cdot\log((1+x)^2   (1-w)(1+y)(1-y)) \\
-    &\qquad + p_{3}          \cdot\log((1+x)^2   (1-w)(1+y)(1-y)) \\
-    &\qquad + p_{12}         \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
-    &\qquad + p_{123}        \cdot\log((1+x)^2   (1-w)(1+y)(1-y))\\
-    &\qquad + p_{13}         \cdot\log((1+x)^2   (1-w)(1-y)^2) \\
-    &\qquad + p_{23}         \cdot\log((1+x)^2   (1-w)(1-y)^2).
-\end{align*}
-When $(1+x)(1-y) < (1-x)(1+y)$, similar arguments yield the lower bound
-\begin{align*}
-    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
-    &\ge      p_{\emptyset}  \cdot\log((1+x)^2    (1-w)(1+y)^2) \\
-    &\qquad + p_{1}          \cdot\log((1+x)(1-x)(1-w)(1+y)^2) \\
-    &\qquad + p_{2}          \cdot\log((1+x)(1-x)(1-w)(1+y)^2) \\
-    &\qquad + p_{3}          \cdot\log((1+x)(1-x)(1-w)(1+y)^2) \\
-    %See above.
-    &\qquad + p_{12}         \cdot\log((1+x)^2    (1-w)(1+y)^2) \\
-    &\qquad + p_{123}        \cdot\log((1+x)(1-x)(1-w)(1+y)^2)\\
-    &\qquad + p_{13}         \cdot\log((1-x)^2    (1-w)(1+y)^2) \\
-    &\qquad + p_{23}         \cdot\log((1-x)^2    (1-w)(1+y)^2).
-\end{align*}
-Finally, if $(1+x)(1-y) = (1-x)(1+y)$, $x=y$ and the lower bound will be less than the other cases.
-Letting
-\[
-b = p_{1}+p_{2}+p_{3}+p_{123}
-\]
-yields
-\[
-    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
-    \ge      2\cdot\log(1+x)
-    + (2-b)  \cdot\log(1+y)
-    + b      \cdot\log(1-y)
-    + \log(1-w)
-\]
-for the first lower bound, and we get the equivalent expression for the second lower bound by exchanging $x$ and $y$.
-Moreover, the divergence quanitity is unchanged when exchanging $x$ and $y$ provided $w=0$.
-Maximizing either lower bound over $t$, we obtain
-\[
-\tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
-\ge      2\cdot\log(2)
-+ (2-b)  \cdot\log(2-b)
-+ b      \cdot\log(b),
-\]
-i.e., the maximum is achieved at $\{1,1-b,0\}$.
-Our lower bound is
-\[
-C_1(x^*, y^*) := \shannonDivergence_{\tau_1,t^*}(\tau_2,\{1,1-b,0\}) + 2\cdot\log(2) + (2-b)\cdot\log(2-b) + b\cdot\log(b).
-\]
-See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < p_{x^*}, p_{y^*} < 1$ with $p_{x^*}=(1-x^*)/2$ and $p_{y^*}=(1-y^*)/2$ where $C_1(x^*, y^*) > C_0(x^*, y^*)$ and joint inference is inconsistent.
-\end{proof}
-
-\generalBranchInconsist*
-
-\begin{proof}
-Recalling Table~\ref{tab:likelihoods} for the InvFels tree $\tau = \tau_1$, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood over $(x,y,w)$.
-In general, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
-Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
-By collapsing like terms and ignoring terms without $w$, the log-likelihood in \eqref{eq:farris_likelihood} as a function of $w$ is
-\begin{align*}
-    \ell_{\tau_1,t^*}(\tau_1, w)
-    &\sim     p_{\emptyset}  \cdot\log(1+(\hat{x})^2+(\hat{y})^2+4\hat{x}\hat{y}w+(\hat{x})^2(\hat{y})^2) \\
-    &\qquad + p_{13}         \cdot\log(1+(\hat{x})^2+(\hat{y})^2-4\hat{x}\hat{y}w+(\hat{x})^2(\hat{y})^2) \\
-    &\qquad + (1 - p_{13})\cdot\log(1+w) \\
-    &\qquad + p_{13}\cdot\log(1-w)
-\end{align*}
-where $\sim$ denotes equality up to an additive constant.
-Similarly, the other two possibilities for the likelihood both satisfy
-\begin{align*}
-    \ell_{\tau_1,t^*}(\tau_1, w)
-    &\sim     p_{\emptyset}  \cdot\log(1+(\hat{x})^2+(\hat{y})^2+4\hat{x}\hat{y}w+(\hat{x})^2(\hat{y})^2) \\
-    &\qquad + p_{13}         \cdot\log(1+(\hat{x})^2+(\hat{y})^2-4\hat{x}\hat{y}w+(\hat{x})^2(\hat{y})^2) \\
-    &\qquad + \log(1+w).
-\end{align*}
-Since
-\[
-\log(1+w) \geq (1 - p_{13})\cdot\log(1+w) + p_{13}\cdot\log(1-w)
-\]
-with equality only when $w=0$, the second likelihood is the likelihood with the maximum-likelihood ancestral state, and so we only analyze that case.
-
-Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
-Simplifying, the likelihood can be written
-\[
-\ell(w) = \alpha_1\log(\beta+\gamma w) + \alpha_2\log(\beta-\gamma w) + \log(1+w)
-\]
-where
-\[
-\alpha_1 := \alpha_1(x^*, y^*) = \frac{1}{8} \left(1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
-\]
-\[
-\alpha_2 := \alpha_2(x^*, y^*) = \frac{1}{8}\left(1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
-\]
-\[
-\beta := \beta(\hat{x}, \hat{y}) = 1+(\hat{x})^2+(\hat{y})^2+(\hat{x})^2(\hat{y})^2,
-\]
-\[
-\gamma := \gamma(\hat{x}, \hat{y}) = 4\hat{x}\hat{y}.
-\]
-%das: added as assumption to theorem
-Since $x^*,y^*,\hat{x},$ and $\hat{y}$ all fall between zero and one, we make use of the inequalities
-\[
-\alpha_1 > \alpha_2
-\]
-and
-\[
-\beta > \gamma
-\]
-using the assumptions that $x^*=y^*=0$, $\hat{x}=\hat{y}=0$, and $\hat{x}=\hat{y}=1$.
-% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
-The derivative of the log-likelihood with respect to $w$ is
-\[
-\ell'(w) := \frac{d}{dw} \ell(w) = \frac{\alpha_1 \gamma}{\beta+\gamma w} - \frac{\alpha_2 \gamma}{\beta-\gamma w} + \frac{1}{1+w}.
-\]
-The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w) \lessgtr 0$, we equivalently consider $f(w) \lessgtr 0$ where $f$ is the quadratic function
-\begin{align*}
-f(w) &= (w)^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
-      &\qquad + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
-      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
-\end{align*}
-% Note from das: b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero.
-c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
-Using the assumption that $\hat{x}\neq 0$ and $\hat{y}\neq 0$, this implies $\ell'$ has two zeros according to the quadratic formula.
-Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w=0$.
-This implies that if the smaller of the zeros of $f(w)$ is greater than one, then $\hat{w} \equiv 1$.
-Using the quadratic formula with
-\[
-a = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
-\]
-\[
-b = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
-\]
-\[
-c = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
-\]
-the smaller zero is
-\[
-\hat{w} = \frac{-b - \sqrt{b^2 - 4ac}}{2a},
-\]
-which is a function of the generating parameters $x^*$ and $y^*$.
-We see that $a \le 0$ and, by a small calculation, $2a+b \leq 0$.
-With this we have,
-\[
-\hat{w} \ge 1 \iff
-|2a + b| \leq \sqrt{b^2 - 4ac} \iff
-%-b - \sqrt{b^2 - 4ac} \le 2a,
-%0 \ge -b - 2a \le \sqrt{b^2 - 4ac},
-%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
-%a^2 + ab + ac \le 0,
-a + b + c \ge 0.
-\]
-Using
-\[
-\alpha_1 + \alpha_2 = \frac{1}{4}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)
-\]
-and
-\[
-\alpha_1 - \alpha_2 = x^*(y^*)^2,
-\]
-and simplifying as functions of $\gamma$ and $\beta$ shows that $a + b + c \ge 0$ is equivalent to
-\[
--\gamma^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0,
-\]
-which, writing out $\gamma$ and $\beta$ explicitly yields
-\begin{equation}
-\label{eq:general-bl-result}
--\gamma^2(\hat{x}, \hat{y})\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma(\hat{x}, \hat{y})\beta(\hat{x}, \hat{y})x^*(y^*)^2 + \beta^2(\hat{x}, \hat{y}) \ge 0.
-\end{equation}
-Given the bounds
-\[
-\gamma_{L} \le \gamma(\hat{x}, \hat{y}) \le \gamma_{U}
-\]
-and
-\[
-\beta_{L} \le \beta(\hat{x}, \hat{y}),
-\]
-if
-\[
--\gamma_{U}^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0,
-\]
-then \eqref{eq:general-bl-result} holds, and this is an inequality involving only $x^*$ and $y^*$.
-\end{proof}
-To get a sense of the region of overestimation, Fig.~\ref{fig:bl-loose-inconsistency} plots an intermediate case.
-
-\subsubsection*{Restricted case}
-
-Theorem~\ref{thm:general-bl} allows us to see where inconsistency in branch length estimation arises when performing general joint maximization over all of the branch parameters, though its conclusion is a statement about both the inconsistency of the internal branch estimate as well as about poor estimation of the other branch parameter estimates.
-To obtain an analytic result indicating where only the internal branch parameter is inconsistently estimated, we consider a restricted case of Theorem~\ref{thm:general-bl}.
-Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and estimate the internal branch parameter $w$.
-Let
-\[
-\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
-\]
-\[
-\gamma := \gamma(x^*, y^*) = 4x^*y^*.
-\]
-The maximum likelihood value $\hat{w}$ is equal to 1 if
-\begin{equation}
-\label{eq:restricted-bl-result}
-%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
--\gamma^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0,
-\end{equation}
-and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
-This follows in a straightforward manner from \eqref{eq:general-bl-result}.
-If $\hat{x}=x^*$ and $\hat{y}=y^*$, then $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$ and \eqref{eq:general-bl-result} becomes \eqref{eq:restricted-bl-result}.
-This allows us to demarcate a region of biased internal branch length estimation by plotting where the inequality in \eqref{eq:restricted-bl-result} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
+In Fig.~\ref{fig:empirical-incons}, for various values of $x^*,y^*$, we compute the maximum of each of the 81 likelihoods, as well as the above restricted likelihood, and plot the generating values for which the above likelihood is greater than or equal to all 81 candidate likelihoods.
 
@@
 \centering
-% obtained by running
-% python joint_inf_plot.py --analytic --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
-\includegraphics[width=\textwidth]{branch-length-inconsistency-inkscape}
-\caption{Analytically-derived region of branch parameter inconsistency where $p_x=p_{x^*}$ and $p_y=p_{y^*}$ are fixed to their correct values with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
-The shaded region shows the area in which the internal branch length $\hat{p}_w$ is estimated to be 0, i.e.\ the estimated fidelity $\hat{w} = 1$, even though the generating fidelity is $y^*$.
-Here again, the shaded region is guaranteed to give inconsistent estimation, while the white region may or may not do so.}
-\label{fig:bl-inconsistency}
+\includegraphics[width=.95\textwidth]{empirical-incons}
+\caption{
+Inconsistency in joint inference.
+In the white region, one of the possibly many equally likely maxima has the property that $\hat{w}=1$ and $\hat{y}_2=1$, indicating we converge to a topology that is not the one that generated the data.
+The gray regions are edge cases where one of the generating parameters was zero or one.
+}
+\label{fig:empirical-incons}
 \end{figure}
-
-\subsection*{Empirical validation}
-
-We use basin-hopping \citep{Wales1997} implemented in \texttt{scipy} \citep{Jones2001} to obtain the plots in Figs.~\ref{fig:bl-general-inconsistency}, \ref{fig:bl-general-marginal}, and \ref{fig:bl-general-bias}.
-This method randomly perturbs candidate solutions and accepts or rejects them based on the nearby likelihood surface, eventually obtaining an estimate of the ``global'' optimum.
-While it is not guaranteed to always obtain a global optimum---or even always converge---its implementation here leads us to believe that we can estimate $\hat{w}$ as well as any optimization method available.
-
-We take various safeguards to ensure convergence and stability of the procedure.
-Since this method involves randomness, we take precautions to not evaluate the likelihood near or outside of boundary conditions, i.e., when $x^*$ or $y^*$ are either $0$ or $1$.
-As is apparent from Table~\ref{tab:sitepatprob}, if $0 < x^*, y^* < 1$ and we try to evaluate the candidate $\hat{x}=\hat{y}=\hat{w}=1$ we have a likelihood of exactly zero.
-Worse still, in Table~\ref{tab:likelihoods} for the InvFels tree, in the cases of $\{2\}, \{1,2\}, \{2,3\},$ and $\{1,2,3\}$, any attempt to evaluate $\hat{y}=1$ will similarly yield a zero likelihood.
-In these cases of zero likelihood, methods that evaluate many candidate parameter values can fail to converge when near these boundaries.
-We sidestep these computational issues by restricting ourselves to the region $x^*,y^*\in [10^{-2},1-10^{-2}]^2$.
-
-We initialize our optimization procedure at the true branch parameters $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
-Because our analysis shows that $\hat{w}$ can be equal to one when $x^*$ and $y^*$ are small, and local optimization may have a hard time obtaining this value if our initial guess is small, we perform two maximizations, one with $w=1$ fixed and one where $w$ is estimated.
-We take the value of $\hat{w}$ with the larger objective function as our estimate.
-
-\begin{figure}
-\centering
-% obtained by running
-% python joint_inf_plot.py --empirical --general-branch-lengths --delta .01 --plot-name figures/w-hat-empirical-01-marginal.svg --out-pkl-name figures/w-hat-empirical-01-marginal.pkl --n-jobs 16 --marginal
-% or
-% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-01-marginal.svg --in-pkl-name figures/w-hat-empirical-01-marginal.pkl
-% if fit already
-\includegraphics[width=\textwidth]{w-hat-empirical-01-marginal}
-\caption{
-    Estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing the classical marginal likelihood \eqref{eq:marginal_likelihood} (rather than a joint optimization procedure).
-}
-\label{fig:bl-general-marginal}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{w-hat-empirical-01-bias}
-\caption{
-    Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing \eqref{eq:profile_likelihood}.
-    Plot focuses on $0.6 < x^*, y^* < 1$ where $\hat{w}$ is estimated to be a value different from 1.
-    We do not compute the bias for the white region where $\hat{w}=1$ to preserve a useful scale for the rest of the plot.
-}
-\label{fig:bl-general-bias}
-\end{figure}

### 2018-06-22 old/old_thing_submission1.tex
major rewrite to appendix for general branch lengths

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_1, t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
    \hline
     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$          &$(1+x)^2   (1-w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1-x)^2   (1+w)(1+y)^2$          &$(1-x)^2   (1-w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1-x)^2   (1+w)(1+y)(1-y)$       &$(1-x)^2   (1-w)(1+y)(1-y)$\\
    \hline
    \hline
    &$\{1\}$                             &$\{1,2\}$  \\
    \hline
     $\emptyset$   &$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1+x)^2   (1-w)(1+y)^2$     &$(1+x)^2   (1+w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1+x)^2   (1-w)(1+y)(1-y)$  &$(1+x)^2   (1+w)(1+y)(1-y)$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
All values multiplied by $1/32$.}
\label{tab:farris_likelihoods}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_2, t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
    \hline
     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1-x)^2   (1-w)(1+y)(1-y)$\\
     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1-x)^2   (1-w)(1-y)^2$\\
     $\{1,3\}$     &$(1-x)^2   (1+w)(1+y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1+x)^2   (1-w)(1-y)^2$\\
     $\{1,2,3\}$   &$(1-x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1-y)^2$\\
    \hline
    \hline
    &$\{1\}$                             &$\{1,2\}$  \\
    \hline
     $\emptyset$   &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       &$(1+x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       &$(1+x)(1-x)(1-w)(1+y)^2$           &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)^2   (1-w)(1+y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1+x)^2   (1+w)(1-y)^2$\\
     $\{2,3\}$     &$(1-x)^2   (1-w)(1+y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1+x)(1-x)(1-w)(1+y)^2$           &$(1+x)^2   (1+w)(1+y)(1-y)$\\
\hline
\end{tabular}
\caption{Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the Felsenstein tree $\tau_2$.
All values multiplied by $1/32$.}
\label{tab:fels_likelihoods}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \multicolumn{3}{c}{InvFels tree ($\tau=\tau_1$)}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$&
    $\emptyset$&
    $(1+x)^2   (1+w)(1+y)^2$\\
     $\{1\}$    &
    $\emptyset$&
    $(1+x)(1-x)(1+w)(1+y)^2$\\
     $\{2\}$    &
    $\emptyset$&
    $(1+x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$    &
    $\emptyset$&
    $(1+x)(1-x)(1+w)(1+y)^2$\\
    $\{1,2\}$  &
    $\{\emptyset,\{1,2\}\}$&
    $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
    $\{1,3\}$  &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x)^2   (1+w)(1+y)^2\\
                    (1+x)^2   (1-w)(1+y)^2\\
                    (1+x)^2   (1+w)(1-y)^2
                \end{array}$\\
    $\{2,3\}$  &
                $\{\emptyset,\{1,2\}\}$&
                $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
    $\{1,2,3\}$&
                $\{1,2\}$&
                $(1+x)^2   (1+w)(1+y)(1-y)$\\
    \hline
    \multicolumn{3}{c}{Felsenstein tree ($\tau=\tau_2$)}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$       &$\emptyset$&$(1+x)^2   (1+w)(1+y)^2$\\
    $\{1\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                        (1+x)(1-x)(1+w)(1+y)^2\\
                        (1+x)^2   (1-w)(1+y)(1-y)
                    \end{array}$\\
      $\{2\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)^2   (1+w)(1+y)(1-y)\\
                    (1+x)(1-x)(1-w)(1+y)^2
                    \end{array}$\\
      $\{3\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)^2\\
                    (1+x)^2   (1-w)(1+y)(1-y)
                    \end{array}$\\
     $\{1,2\}$         &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1+x)^2   (1-w)(1+y)^2
                    \end{array}$\\
     $\{1,3\}$         &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{\{1\},\{2\}\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x)^2   (1+w)(1+y)^2\\
                    (1+x)(1-x)(1-w)(1+y)(1-y)\\
                    (1+x)^2   (1+w)(1-y)^2
                    \end{array}$\\
      $\{2,3\}$        &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1-x)^2   (1-w)(1+y)^2\\
                    (1+x)^2   (1-w)(1-y)^2
                    \end{array}$\\
     $\{1,2,3\}$       &
    $\left\{\begin{array}{l}
                    \{2\} \\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1-w)(1+y)^2 \\
                    (1+x)^2   (1+w)(1+y)(1-y)
                    \end{array}$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$ after maximizing over ancestral states on the InvFels tree $\tau_1$ and the Felsenstein tree $\tau_2$.
All values multiplied by $1/32$.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
See Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} for full calculations.}
\label{tab:likelihoods}
\end{table}

\topoInconsist*

\begin{proof}
Assume $\tau^*=\tau_1$.
The InvFels tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see the upper table of Table~\ref{tab:likelihoods}).

We write out the case for the ancestral state split $\{1\}$, and show the other two cases follow a similar argument.
Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_profile_likelihood_mean}, the log-likelihood is, suppressing normalizing constants $-\log 8$ (from the first additive term of \eqref{eq:site_pattern_profile_likelihood_mean}) and $-\log 32$ (from the second),
\begin{align}
    \label{eq:farris_likelihood}
    \ell_{\tau_1,t^*}(\tau_1, t)
    &=        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw+x^2y^2) \nonumber \\
    &\qquad + p_{1}          \cdot\log(1-x^2+y^2-x^2y^2) \nonumber \\
    &\qquad + p_{2}          \cdot\log(1+x^2-y^2-x^2y^2) \nonumber \\
    &\qquad + p_{3}          \cdot\log(1-x^2+y^2-x^2y^2) \nonumber \\
    &\qquad + p_{12}         \cdot\log(1-x^2-y^2+x^2y^2) \nonumber \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw+x^2y^2) \nonumber \\
    &\qquad + p_{23}         \cdot\log(1-x^2-y^2+x^2y^2) \nonumber \\
    &\qquad + p_{123}        \cdot\log(1+x^2-y^2-x^2y^2) \nonumber \\
    &\qquad + p_{\emptyset}  \cdot\log((1+x)^2   (1+w)(1+y)^2) \nonumber \\
    &\qquad + p_{1}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \nonumber \\
    &\qquad + p_{2}          \cdot\log((1+x)^2   (1+w)(1+y)(1-y)) \nonumber \\
    &\qquad + p_{3}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \nonumber \\
    &\qquad + p_{12}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y)) \nonumber \\
    &\qquad + p_{13}         \cdot\log((1+x)^2   (1-w)(1+y)^2) \nonumber \\
    &\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y)) \nonumber \\
    &\qquad + p_{123}        \cdot\log((1+x)^2   (1+w)(1+y)(1-y)).
\end{align}
Our plan is to simplify the likelihood to remove log-of-quadratic terms of $x,y,$ and $w$, obtaining an upper bound on the likelihood that has a closed-form maximum in each variable.
To bound the generating probabilities, we use the facts that, for $x,y\in[0,1]$,
\begin{align*}
p_{12} = p_{23} \propto 1-x^2-y^2+x^2y^2 & = (1+x)(1-x)(1+y)(1-y) \\
p_{1} = p_{3} \propto 1-x^2+y^2-x^2y^2 & = (1+x)(1-x)(1+y^2) \le (1+x)(1-x)(1+y) \\
p_{2} = p_{123} \propto 1+x^2-y^2-x^2y^2 & = (1+x^2)(1+y)(1-y) \le (1+x)(1+y)(1-y).
\end{align*}
To bound the remaining $p_{\emptyset}$ and $p_{13}$, we use
\begin{align*}
1+x^2+y^2+x^2y^2 & = (1+x^2)(1+y^2) \le (1+x)(1+y) \\
4xy & = 2x \cdot 2y \le (1+x^2)(1+y^2) \le (1+x)(1+y)
\end{align*}
so that the $p_{\emptyset}$ term is bounded as
\begin{align*}
    p_{\emptyset} & \propto 1+x^2+y^2+4xyw+x^2y^2 \\
                  & \le 1+x^2+y^2+4xy+x^2y^2 \\
                  & \le 2(1+x)(1+y)
\end{align*}
and $p_{13}$ is bounded as
\begin{align*}
    p_{13} & \propto 1+x^2+y^2-4xyw+x^2y^2 \\
                  & \le 1+x^2+y^2+x^2y^2 \\
                  & \le (1+x)(1+y).
\end{align*}
Factoring and making these substitutions (again, under the assumption that the ancestral state split is $\{1\}$) results in
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad \le      p_{\emptyset}  \cdot\log(2(1+x)(1+y))
+ p_{1}          \cdot\log((1+x)(1-x)(1+y)) \\
    &\qquad\qquad + p_{2}          \cdot\log((1+x)(1+y)(1-y))
+ p_{3}          \cdot\log((1+x)(1-x)(1+y)) \\
    &\qquad\qquad + p_{12}         \cdot\log((1+x)(1-x)(1+y)(1-y))
+ p_{13}         \cdot\log((1+x)(1+y)) \\
    &\qquad\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+y)(1-y))
+ p_{123}        \cdot\log((1+x)(1+y)(1-y)) \\
    &\qquad\qquad + p_{\emptyset}  \cdot\log((1+x)^2   (1+w)(1+y)^2)
+ p_{1}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \\
    &\qquad\qquad + p_{2}          \cdot\log((1+x)^2   (1+w)(1+y)(1-y))
+ p_{3}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \\
    &\qquad\qquad + p_{12}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y))
+ p_{13}         \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
    &\qquad\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y)) \\
    &\qquad\qquad + p_{123}        \cdot\log((1+x)^2   (1+w)(1+y)(1-y)) \\
&\qquad =      p_{\emptyset}  \cdot\log(2)
+ \log(1+x)
+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x) \\
&\qquad\qquad + \log(1+y)
+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
&\qquad\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x)
+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x) \\
&\qquad\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y)
+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
&\qquad\qquad + (1-p_{13})\cdot\log(1+w)
+ p_{13}\cdot\log(1-w) \\
&\qquad =      p_{\emptyset}  \cdot\log(2)
+ (3-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x)
+ 2(p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x) \\
&\qquad\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y)
+ 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
&\qquad\qquad + (1-p_{13})\cdot\log(1+w)
+ p_{13}\cdot\log(1-w).
\end{align*}
To simplify, let
\begin{equation}
    \begin{aligned}
        a_{1} &= 3-p_{1}-p_{3}-p_{12}-p_{23}, \\
        a_{2} &= 2(p_{1}+p_{3}+p_{12}+p_{23}), \\
        a_{3} &= 3-p_{2}-p_{12}-p_{23}-p_{123}, \\
        a_{4} &= 2(p_{2}+p_{12}+p_{23}+p_{123}), \\
    \end{aligned}
    \label{eq:a_const}
\end{equation}
so that
\begin{equation*}
\begin{split}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad \le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log(1+x)
+ a_{2}\cdot\log(1-x)
+ a_{3}\cdot\log(1+y) \\
&\qquad\qquad + a_{4}\cdot\log(1-y)
+ (1-p_{13})\cdot\log(1+w)
+ p_{13}\cdot\log(1-w).
\end{split}
\end{equation*}
Maximizing over the unknown terms yields
\[
\hat{x} = \frac{a_{1}-a_{2}}{a_{1}+a_{2}}, \ \hat{y} = \frac{a_{3}-a_{4}}{a_{3}+a_{4}}, \ \hat{w} = 1-2p_{13}.
\]
Trivially $0 \le \hat{w} \le 1$ and $\hat{x}, \hat{y} \le 1$.
We check that $\hat{x}, \hat{y} \ge 0$ to ensure these maxima are valid.
For $\hat{x} \ge 0$ we need $a_1 \ge a_2$; letting
\[
\tilde{p} = p_{1}+p_{3}+p_{12}+p_{23}
\]
we have from \eqref{eq:a_const} that $a_1 \ge a_2$ only if $3 \ge 3\tilde{p}$.
Since $\tilde{p} = 1/8\cdot(4-4(x^*)^2)$, we have $0 \le \tilde{p} \le 1/2$ implying $\hat{x} \ge 0$.
The same approach works for $\hat{y}$.
The upper bound for the likelihood is then maximized at
\begin{align}
\begin{split}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}}
+ a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
&\qquad\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ (1-p_{13})\cdot\log(2(1-p_{13}))
+ p_{13}\cdot\log(2p_{13}) \\
&\qquad := C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*).
\end{split}
\label{eq:farris-upper-bound-1}
\end{align}

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications, as we now describe.
The upper bound $C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*)$ in \eqref{eq:farris-upper-bound-1} is for the ancestral state split $\{1\}$.
For the ancestral state split $\emptyset$ we can derive an upper bound $C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*)$ in \eqref{eq:farris-upper-bound-empty} in a similar fashion, and $C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*)$ will be the same as $C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*)$ except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.
In this case, all terms involving $w$ simplify to $\log(1+w)$, which is maximized at $\hat{w}=1$.
For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_x}
\end{equation}
for ancestral state split $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$, yielding the upper bound
\begin{align}
\begin{split}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ \log(2) \\
&\qquad := C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*).
\end{split}
\label{eq:farris-upper-bound-empty}
\end{align}

For ancestral state split $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$, we have constants
\begin{equation}
    \begin{aligned}
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_y}
\end{equation}
and the upper bound
\begin{align}
\begin{split}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
+ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
+ \log(2) \\
&\qquad := C^{\{1,2\}}_{\tau_1,\tau_1}(x^*, y^*).
\end{split}
\label{eq:farris-upper-bound-12}
\end{align}

We now need to check that
\[
\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
\]
are both between zero and one.
Using a similar argument as that where $\ancestralSplitPartition_j(\tau_1, t) = \{1\}$, we only need to show $a_1' \ge a_2'$, which is true if $3-2p_{13} \ge 3\tilde{p}+2p_{13}$.
Some rearranging and the fact that $0 \le \tilde{p} \le 1/2$ shows this is equivalent to showing $p_{13} \le 3/8$.
Using Table~\ref{tab:sitepatprob} while setting $t=t^*=\{x^*,y^*,x^*,y^*,y^*\}$,
\[
p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right),
\]
meaning we are interested in whether
\begin{equation}
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3.
\label{eq:generating_ineq}
\end{equation}
We see that
\begin{equation}
\begin{split}
& 1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3 \\
& \qquad \qquad \qquad \qquad \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 2 - (x^*)^2
\end{split}
\label{eq:lenny}
\end{equation}
and that
\begin{align*}
(y^*)^2\left(1 - 4x^* + (x^*)^2\right) &\le (y^*)^2\left(1 - 2x^* + (x^*)^2\right) \\
                                       &=   (y^*)^2\left(1 - x^*\right)^2 \\
                                       &\le        \left(1 - x^*\right)^2 \\
                                       &\le              2 - (x^*)^2,
\end{align*}
with the last inequality holding if $0 \le x^* \le 1$, showing that \eqref{eq:lenny} holds.
Thus, $p_{13} \le 3/8$ and our $\hat{x}$ and $\hat{y}$ are valid.

All bounds are functions only of $x,y$ through the true generating probabilities $x^*, y^*$.
Call the upper bound
\[
C_0(x^*, y^*) := \max\left(C^{\{1\}}_{\tau_1,\tau_1}(x^*, y^*), C^{\emptyset}_{\tau_1,\tau_1}(x^*, y^*), C^{\{1,2\}}_{\tau_1,\tau_1}(x^*, y^*)\right).
\]

We next construct a lower bound on the Felsenstein tree partial likelihood.
In general, taking
\[
\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\]
we see
\[
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
\]
is a lower bound for the joint maximum of \eqref{eq:log_likelihood_simplified}.
In the Felsenstein case, there are many more optimal internal states depending on branch lengths than in the InvFels case (see the lower table of Table~\ref{tab:likelihoods}).
We first bound the entire likelihood below and then proceed with joint inference.
To obtain a lower bound for the likelihood we replace $1+w$ with $1-w$ in Table~\ref{tab:likelihoods} for $\tau = \tau_2$; in this case, we resolve many of the ambiguous likelihood terms.
For example, for site split $\{1,2\}$, after we replace the $1+w$ term for $\ancestralSplitPartition_j(\tau, t)=\{\emptyset, \{1,2\}\}$ with $1-w$, we have
\[
\max\left((1+x)^2 (1-w)(1+y)^2, (1+x)(1-x)(1-w)(1+y)(1-y)\right)
\]
where since
\[
(1+x)^2 (1-w)(1+y)^2 > (1+x)(1-x)(1-w)(1+y)(1-y),
\]
the maximum for this site split can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
For the other site splits we consider three cases.
If $(1+x)(1-y) > (1-x)(1+y)$, then
\[
(1+x)^2(1-w)(1+y)(1-y) > (1+x)(1-x)(1-w)(1+y)^2
\]
meaning the cases $\{1\}, \{2\}, \{3\}$ and $\{1,2,3\}$ will have likelihoods bounded below by
\[
(1+x)^2(1-w)(1+y)(1-y).
\]
For the cases $\{1,3\}$ and $\{2,3\}$ and the same condition, we have
\[
(1+x)^2(1-w)(1-y)^2 > (1-x)^2(1-w)(1+y)^2
\]
and
\[
(1+x)^2(1-w)(1-y)^2 > (1+x)(1-x)(1-w)(1+y)(1-y),
\]
yielding lower bounds of
\[
(1+x)^2(1-w)(1-y)^2
\]
in these cases.
The partial likelihood in this case is then bounded below by
\begin{align*}
    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
    &\ge      p_{\emptyset}  \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
    &\qquad + p_{1}          \cdot\log((1+x)^2   (1-w)(1+y)(1-y)) \\
    &\qquad + p_{2}          \cdot\log((1+x)^2   (1-w)(1+y)(1-y)) \\
    &\qquad + p_{3}          \cdot\log((1+x)^2   (1-w)(1+y)(1-y)) \\
    &\qquad + p_{12}         \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
    &\qquad + p_{123}        \cdot\log((1+x)^2   (1-w)(1+y)(1-y))\\
    &\qquad + p_{13}         \cdot\log((1+x)^2   (1-w)(1-y)^2) \\
    &\qquad + p_{23}         \cdot\log((1+x)^2   (1-w)(1-y)^2).
\end{align*}
When $(1+x)(1-y) < (1-x)(1+y)$, similar arguments yield the lower bound
\begin{align*}
    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
    &\ge      p_{\emptyset}  \cdot\log((1+x)^2    (1-w)(1+y)^2) \\
    &\qquad + p_{1}          \cdot\log((1+x)(1-x)(1-w)(1+y)^2) \\
    &\qquad + p_{2}          \cdot\log((1+x)(1-x)(1-w)(1+y)^2) \\
    &\qquad + p_{3}          \cdot\log((1+x)(1-x)(1-w)(1+y)^2) \\
    %See above.
    &\qquad + p_{12}         \cdot\log((1+x)^2    (1-w)(1+y)^2) \\
    &\qquad + p_{123}        \cdot\log((1+x)(1-x)(1-w)(1+y)^2)\\
    &\qquad + p_{13}         \cdot\log((1-x)^2    (1-w)(1+y)^2) \\
    &\qquad + p_{23}         \cdot\log((1-x)^2    (1-w)(1+y)^2).
\end{align*}
Finally, if $(1+x)(1-y) = (1-x)(1+y)$, $x=y$ and the lower bound will be less than the other cases.
Letting
\[
b = p_{1}+p_{2}+p_{3}+p_{123}
\]
yields
\[
    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
    \ge      2\cdot\log(1+x)
    + (2-b)  \cdot\log(1+y)
    + b      \cdot\log(1-y)
    + \log(1-w)
\]
for the first lower bound, and we get the equivalent expression for the second lower bound by exchanging $x$ and $y$.
Moreover, the divergence quanitity is unchanged when exchanging $x$ and $y$ provided $w=0$.
Maximizing either lower bound over $t$, we obtain
\[
\tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
\ge      2\cdot\log(2)
+ (2-b)  \cdot\log(2-b)
+ b      \cdot\log(b),
\]
i.e., the maximum is achieved at $\{1,1-b,0\}$.
Our lower bound is
\[
C_1(x^*, y^*) := \shannonDivergence_{\tau_1,t^*}(\tau_2,\{1,1-b,0\}) + 2\cdot\log(2) + (2-b)\cdot\log(2-b) + b\cdot\log(b).
\]
See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < p_{x^*}, p_{y^*} < 1$ with $p_{x^*}=(1-x^*)/2$ and $p_{y^*}=(1-y^*)/2$ where $C_1(x^*, y^*) > C_0(x^*, y^*)$ and joint inference is inconsistent.
\end{proof}

\generalBranchInconsist*

\begin{proof}
Recalling Table~\ref{tab:likelihoods} for the InvFels tree $\tau = \tau_1$, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood over $(x,y,w)$.
In general, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
By collapsing like terms and ignoring terms without $w$, the log-likelihood in \eqref{eq:farris_likelihood} as a function of $w$ is
\begin{align*}
    \ell_{\tau_1,t^*}(\tau_1, w)
    &\sim     p_{\emptyset}  \cdot\log(1+(\hat{x})^2+(\hat{y})^2+4\hat{x}\hat{y}w+(\hat{x})^2(\hat{y})^2) \\
    &\qquad + p_{13}         \cdot\log(1+(\hat{x})^2+(\hat{y})^2-4\hat{x}\hat{y}w+(\hat{x})^2(\hat{y})^2) \\
    &\qquad + (1 - p_{13})\cdot\log(1+w) \\
    &\qquad + p_{13}\cdot\log(1-w)
\end{align*}
where $\sim$ denotes equality up to an additive constant.
Similarly, the other two possibilities for the likelihood both satisfy
\begin{align*}
    \ell_{\tau_1,t^*}(\tau_1, w)
    &\sim     p_{\emptyset}  \cdot\log(1+(\hat{x})^2+(\hat{y})^2+4\hat{x}\hat{y}w+(\hat{x})^2(\hat{y})^2) \\
    &\qquad + p_{13}         \cdot\log(1+(\hat{x})^2+(\hat{y})^2-4\hat{x}\hat{y}w+(\hat{x})^2(\hat{y})^2) \\
    &\qquad + \log(1+w).
\end{align*}
Since
\[
\log(1+w) \geq (1 - p_{13})\cdot\log(1+w) + p_{13}\cdot\log(1-w)
\]
with equality only when $w=0$, the second likelihood is the likelihood with the maximum-likelihood ancestral state, and so we only analyze that case.

Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
Simplifying, the likelihood can be written
\[
\ell(w) = \alpha_1\log(\beta+\gamma w) + \alpha_2\log(\beta-\gamma w) + \log(1+w)
\]
where
\[
\alpha_1 := \alpha_1(x^*, y^*) = \frac{1}{8} \left(1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
\]
\[
\alpha_2 := \alpha_2(x^*, y^*) = \frac{1}{8}\left(1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
\]
\[
\beta := \beta(\hat{x}, \hat{y}) = 1+(\hat{x})^2+(\hat{y})^2+(\hat{x})^2(\hat{y})^2,
\]
\[
\gamma := \gamma(\hat{x}, \hat{y}) = 4\hat{x}\hat{y}.
\]
%das: added as assumption to theorem
Since $x^*,y^*,\hat{x},$ and $\hat{y}$ all fall between zero and one, we make use of the inequalities
\[
\alpha_1 > \alpha_2
\]
and
\[
\beta > \gamma
\]
using the assumptions that $x^*=y^*=0$, $\hat{x}=\hat{y}=0$, and $\hat{x}=\hat{y}=1$.
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
The derivative of the log-likelihood with respect to $w$ is
\[
\ell'(w) := \frac{d}{dw} \ell(w) = \frac{\alpha_1 \gamma}{\beta+\gamma w} - \frac{\alpha_2 \gamma}{\beta-\gamma w} + \frac{1}{1+w}.
\]
The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w) \lessgtr 0$, we equivalently consider $f(w) \lessgtr 0$ where $f$ is the quadratic function
\begin{align*}
f(w) &= (w)^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
\end{align*}
% Note from das: b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero. c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
Using the assumption that $\hat{x}\neq 0$ and $\hat{y}\neq 0$, this implies $\ell'$ has two zeros according to the quadratic formula.
Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w=0$.
This implies that if the smaller of the zeros of $f(w)$ is greater than one, then $\hat{w} \equiv 1$.
Using the quadratic formula with
\[
a = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
\]
\[
b = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
\]
\[
c = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
\]
the smaller zero is
\[
\hat{w} = \frac{-b - \sqrt{b^2 - 4ac}}{2a},
\]
which is a function of the generating parameters $x^*$ and $y^*$.
We see that $a \le 0$ and, by a small calculation, $2a+b \leq 0$.
With this we have,
\[
\hat{w} \ge 1 \iff
|2a + b| \leq \sqrt{b^2 - 4ac} \iff
%-b - \sqrt{b^2 - 4ac} \le 2a,
%0 \ge -b - 2a \le \sqrt{b^2 - 4ac},
%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
%a^2 + ab + ac \le 0,
a + b + c \ge 0.
\]
Using
\[
\alpha_1 + \alpha_2 = \frac{1}{4}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)
\]
and
\[
\alpha_1 - \alpha_2 = x^*(y^*)^2,
\]
and simplifying as functions of $\gamma$ and $\beta$ shows that $a + b + c \ge 0$ is equivalent to
\[
-\gamma^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0,
\]
which, writing out $\gamma$ and $\beta$ explicitly yields
\begin{equation}
\label{eq:general-bl-result}
-\gamma^2(\hat{x}, \hat{y})\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma(\hat{x}, \hat{y})\beta(\hat{x}, \hat{y})x^*(y^*)^2 + \beta^2(\hat{x}, \hat{y}) \ge 0.
\end{equation}
Given the bounds
\[
\gamma_{L} \le \gamma(\hat{x}, \hat{y}) \le \gamma_{U}
\]
and
\[
\beta_{L} \le \beta(\hat{x}, \hat{y}),
\]
if
\[
-\gamma_{U}^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0,
\]
then \eqref{eq:general-bl-result} holds, and this is an inequality involving only $x^*$ and $y^*$.
\end{proof}
To get a sense of the region of overestimation, Fig.~\ref{fig:bl-loose-inconsistency} plots an intermediate case.

\subsubsection*{Restricted case}

Theorem~\ref{thm:general-bl} allows us to see where inconsistency in branch length estimation arises when performing general joint maximization over all of the branch parameters, though its conclusion is a statement about both the inconsistency of the internal branch estimate as well as about poor estimation of the other branch parameter estimates.
To obtain an analytic result indicating where only the internal branch parameter is inconsistently estimated, we consider a restricted case of Theorem~\ref{thm:general-bl}.
Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and estimate the internal branch parameter $w$.
Let
\[
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
\]
\[
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
\]
The maximum likelihood value $\hat{w}$ is equal to 1 if
\begin{equation}
\label{eq:restricted-bl-result}
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0,
\end{equation}
and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
This follows in a straightforward manner from \eqref{eq:general-bl-result}.
If $\hat{x}=x^*$ and $\hat{y}=y^*$, then $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$ and \eqref{eq:general-bl-result} becomes \eqref{eq:restricted-bl-result}.
This allows us to demarcate a region of biased internal branch length estimation by plotting where the inequality in \eqref{eq:restricted-bl-result} is satisfied (Fig.~\ref{fig:bl-inconsistency}).

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
\includegraphics[width=\textwidth]{branch-length-inconsistency-inkscape}
\caption{Analytically-derived region of branch parameter inconsistency where $p_x=p_{x^*}$ and $p_y=p_{y^*}$ are fixed to their correct values with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
The shaded region shows the area in which the internal branch length $\hat{p}_w$ is estimated to be 0, i.e.\ the estimated fidelity $\hat{w} = 1$, even though the generating fidelity is $y^*$.
Here again, the shaded region is guaranteed to give inconsistent estimation, while the white region may or may not do so.}
\label{fig:bl-inconsistency}
\end{figure}

\subsection*{Empirical validation}

We use basin-hopping \citep{Wales1997} implemented in \texttt{scipy} \citep{Jones2001} to obtain the plots in Figs.~\ref{fig:bl-general-inconsistency}, \ref{fig:bl-general-marginal}, and \ref{fig:bl-general-bias}.
This method randomly perturbs candidate solutions and accepts or rejects them based on the nearby likelihood surface, eventually obtaining an estimate of the ``global'' optimum.
While it is not guaranteed to always obtain a global optimum---or even always converge---its implementation here leads us to believe that we can estimate $\hat{w}$ as well as any optimization method available.

We take various safeguards to ensure convergence and stability of the procedure.
Since this method involves randomness, we take precautions to not evaluate the likelihood near or outside of boundary conditions, i.e., when $x^*$ or $y^*$ are either $0$ or $1$.
As is apparent from Table~\ref{tab:sitepatprob}, if $0 < x^*, y^* < 1$ and we try to evaluate the candidate $\hat{x}=\hat{y}=\hat{w}=1$ we have a likelihood of exactly zero.
Worse still, in Table~\ref{tab:likelihoods} for the InvFels tree, in the cases of $\{2\}, \{1,2\}, \{2,3\},$ and $\{1,2,3\}$, any attempt to evaluate $\hat{y}=1$ will similarly yield a zero likelihood.
In these cases of zero likelihood, methods that evaluate many candidate parameter values can fail to converge when near these boundaries.
We sidestep these computational issues by restricting ourselves to the region $x^*,y^*\in [10^{-2},1-10^{-2}]^2$.

We initialize our optimization procedure at the true branch parameters $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
Because our analysis shows that $\hat{w}$ can be equal to one when $x^*$ and $y^*$ are small, and local optimization may have a hard time obtaining this value if our initial guess is small, we perform two maximizations, one with $w=1$ fixed and one where $w$ is estimated.
We take the value of $\hat{w}$ with the larger objective function as our estimate.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .01 --plot-name figures/w-hat-empirical-01-marginal.svg --out-pkl-name figures/w-hat-empirical-01-marginal.pkl --n-jobs 16 --marginal
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-01-marginal.svg --in-pkl-name figures/w-hat-empirical-01-marginal.pkl
% if fit already
\includegraphics[width=\textwidth]{w-hat-empirical-01-marginal}
\caption{
    Estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing the classical marginal likelihood \eqref{eq:marginal_likelihood} (rather than a joint optimization procedure).
}
\label{fig:bl-general-marginal}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{w-hat-empirical-01-bias}
\caption{
    Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing \eqref{eq:profile_likelihood}.
    Plot focuses on $0.6 < x^*, y^* < 1$ where $\hat{w}$ is estimated to be a value different from 1.
    We do not compute the bias for the white region where $\hat{w}=1$ to preserve a useful scale for the rest of the plot.
}
\label{fig:bl-general-bias}
\end{figure}

### 2018-06-22 joint_inf_plot_empirical.py
empirical code

@@ -1,0 +1,221 @@
+from __future__ import unicode_literals
+import matplotlib
+matplotlib.use('SVG')
+matplotlib.rcParams['text.usetex'] = True
+matplotlib.rcParams['text.latex.unicode'] = True
+matplotlib.rcParams['figure.autolayout'] = True
+import matplotlib.pyplot as plt
+import matplotlib.patches as mpatches
+import seaborn as sns
+import numpy as np
+import argparse
+import sys
+import pickle
+import time
+
+from multiprocessing import Pool
+from scipy.optimize import minimize
+from itertools import product
+
+sns.set_style('white')
+sns.set_style('ticks')
+
+LOG32 = np.log(32)
+
+FONT_SIZE = 20
+    
+ALL_PATTERNS = [
+        '0000',
+        '1000',
+        '0100',
+        '0010',
+        '1110',
+        '1100',
+        '1010',
+        '0110'
+]
+
+def parse_args():
+    parser = argparse.ArgumentParser(description=__doc__)
+
+    parser.add_argument(
+        '--plot-name',
+        type=str,
+        help='file name for where to save plot',
+        default='temporary-plot.svg',
+    )
+    parser.add_argument(
+        '--delta',
+        type=float,
+        help='coarseness parameter for plotting',
+        default=0.1,
+    )
+    parser.add_argument(
+        '--n-jobs',
+        type=int,
+        help='number of processes to use for empirically computing maxima',
+        default=10,
+    )
+    parser.add_argument(
+        '--plot-curve',
+        action='store_true',
+        help='plot curve showing intuition for why this happens?',
+    )
+
+    args = parser.parse_args()
+
+    return args
+
+def P_INVFELS(theta0, site_pattern):
+    """
+    Site pattern frequencies for inverse Felsenstein tree
+    """
+    # fifth branch will always be positive
+    theta = [-t if s=='1' else t \
+            for t,s in zip(theta0, site_pattern+'0')]
+    return 0.125 * (1 + \
+         theta[0]*theta[2] + \
[…]

### 2018-06-22 joint_inf_helpers.py
didn't feel like writing everything out by hand so i made python do it

@@ -1,0 +1,159 @@
+from __future__ import unicode_literals
+import matplotlib
+matplotlib.use('SVG')
+matplotlib.rcParams['text.usetex'] = True
+matplotlib.rcParams['text.latex.unicode'] = True
+matplotlib.rcParams['figure.autolayout'] = True
+import matplotlib.pyplot as plt
+import matplotlib.patches as mpatches
+import seaborn as sns
+import numpy as np
+import argparse
+import sys
+import collections
+
+# ~~~~~~~~~
+# global variables
+
+ALL_PATTERNS = [
+    '0000',
+    '1000',
+    '0100',
+    '0010',
+    '1110',
+    '1100',
+    '1010',
+    '0110'
+]
+
+ALL_ANC_STATES = [
+    '00',
+    '10',
+    '01',
+    '11',
+]
+
+PATT2SPLIT = {
+    '0000': '$\emptyset$',
+    '1000': '$\{1\}$    ',
+    '0100': '$\{2\}$    ',
+    '0010': '$\{3\}$    ',
+    '1110': '$\{1,2,3\}$',
+    '1100': '$\{1,2\}$  ',
+    '1010': '$\{1,3\}$  ',
+    '0110': '$\{2,3\}$  ',
+}
+
+ANC2SPLIT = {
+    '00': '$\emptyset$',
+    '10': '$\{1\}$    ',
+    '01': '$\{2\}$    ',
+    '11': '$\{1,2\}$  ',
+}
+
+# ~~~~~~~~~
+
+def parse_args():
+    parser = argparse.ArgumentParser(description=__doc__)
+
+    parser.add_argument(
+        '--restricted-branch-lengths',
+        action="store_true",
+        help='restricted branch lengths plot',
+    )
+    parser.add_argument(
+        '--general-branch-lengths',
+        action="store_true",
+        help='general branch lengths plot',
+    )
+
+    args = parser.parse_args()
+
+    return args
+
+# ~~~~~~~~~
+# global functions for exact likelihood computation
+
+def P_INVFELS(theta, site_pattern):
+    """
+    Site pattern frequencies for inverse Felsenstein tree
+    """
+    # fifth branch will always be positive
+    theta_sgn = [-1 if s=='1' else 1 for s in site_pattern+'0']
+    prob_list = [
+        [theta[0], theta[2]],
+        [theta[1], theta[3]],
[…]

### 2018-06-22 joint_inf_plot.py
plots

@@ -10,12 +10,11 @@ import seaborn as sns
 import numpy as np
 import argparse
 import sys
-import pickle
 import time
 
-from operator import mul
 from multiprocessing import Pool
-from scipy.optimize import differential_evolution, minimize, basinhopping
+from scipy.optimize import minimize
+from itertools import product
 
 sns.set_style('white')
 sns.set_style('ticks')
@@ -24,6 +23,8 @@ sns.set_style('ticks')
 # global variables
 
 FONT_SIZE = 20
+
+LOG32 = np.log(32)
 
 ALL_PATTERNS = [
         '0000',
@@ -47,6 +48,19 @@ SPLIT_TO_PATTERN_DICT = {
         '23': '0110'
 }
 
+class Legend(object):
+    pass
+
+class LegendHandler(object):
+    def legend_artist(self, legend, orig_handle, fontsize, handlebox):
+        x0, y0 = handlebox.xdescent, handlebox.ydescent
+        width, height = handlebox.width, handlebox.height
+        patch = mpatches.Rectangle([x0, y0], width, height, facecolor='w',
+                                   edgecolor='black', lw=1,
+                                   transform=handlebox.get_transform())
+        handlebox.add_artist(patch)
+        return patch
+
 # ~~~~~~~~~
 
 def parse_args():
@@ -59,18 +73,6 @@ def parse_args():
         default='temporary-plot.svg',
     )
     parser.add_argument(
-        '--in-pkl-name',
-        type=str,
-        help='file name for where to save pkl',
-        default=None,
-    )
-    parser.add_argument(
-        '--out-pkl-name',
-        type=str,
-        help='file name for where to save pkl',
-        default='output.pkl',
-    )
-    parser.add_argument(
         '--delta',
         type=float,
         help='coarseness parameter for plotting',
@@ -81,33 +83,6 @@ def parse_args():
         type=int,
         help='number of processes to use for empirically computing maxima',
         default=1,
-    )
-    parser.add_argument(
-        '--param-index',
-        type=int,
-        default=2,
-        choices=(0, 1, 2),
[…]

### 2018-06-22 joint_inf_plot_empirical.py

@@ -1,221 +1,0 @@
-from __future__ import unicode_literals
-import matplotlib
-matplotlib.use('SVG')
-matplotlib.rcParams['text.usetex'] = True
-matplotlib.rcParams['text.latex.unicode'] = True
-matplotlib.rcParams['figure.autolayout'] = True
-import matplotlib.pyplot as plt
-import matplotlib.patches as mpatches
-import seaborn as sns
-import numpy as np
-import argparse
-import sys
-import pickle
-import time
-
-from multiprocessing import Pool
-from scipy.optimize import minimize
-from itertools import product
-
-sns.set_style('white')
-sns.set_style('ticks')
-
-LOG32 = np.log(32)
-
-FONT_SIZE = 20
-    
-ALL_PATTERNS = [
-        '0000',
-        '1000',
-        '0100',
-        '0010',
-        '1110',
-        '1100',
-        '1010',
-        '0110'
-]
-
-def parse_args():
-    parser = argparse.ArgumentParser(description=__doc__)
-
-    parser.add_argument(
-        '--plot-name',
-        type=str,
-        help='file name for where to save plot',
-        default='temporary-plot.svg',
-    )
-    parser.add_argument(
-        '--delta',
-        type=float,
-        help='coarseness parameter for plotting',
-        default=0.1,
-    )
-    parser.add_argument(
-        '--n-jobs',
-        type=int,
-        help='number of processes to use for empirically computing maxima',
-        default=10,
-    )
-    parser.add_argument(
-        '--plot-curve',
-        action='store_true',
-        help='plot curve showing intuition for why this happens?',
-    )
-
-    args = parser.parse_args()
-
-    return args
-
-def P_INVFELS(theta0, site_pattern):
-    """
-    Site pattern frequencies for inverse Felsenstein tree
-    """
-    # fifth branch will always be positive
-    theta = [-t if s=='1' else t \
-            for t,s in zip(theta0, site_pattern+'0')]
-    return 0.125 * (1 + \
-         theta[0]*theta[2] + \
[…]

### 2018-06-22 old/joint_inf_plot_v1.py

@@ -1,0 +1,501 @@
+from __future__ import unicode_literals
+import matplotlib
+matplotlib.use('SVG')
+matplotlib.rcParams['text.usetex'] = True
+matplotlib.rcParams['text.latex.unicode'] = True
+matplotlib.rcParams['figure.autolayout'] = True
+import matplotlib.pyplot as plt
+import matplotlib.patches as mpatches
+import seaborn as sns
+import numpy as np
+import argparse
+import sys
+import pickle
+import time
+
+from operator import mul
+from multiprocessing import Pool
+from scipy.optimize import differential_evolution, minimize, basinhopping
+
+sns.set_style('white')
+sns.set_style('ticks')
+
+# ~~~~~~~~~
+# global variables
+
+FONT_SIZE = 20
+
+ALL_PATTERNS = [
+        '0000',
+        '1000',
+        '0100',
+        '0010',
+        '1110',
+        '1100',
+        '1010',
+        '0110'
+]
+
+SPLIT_TO_PATTERN_DICT = {
+        '0':  '0000',
+        '1':  '1000',
+        '2':  '0100',
+        '3':  '0010',
+        '123':'1110',
+        '12': '1100',
+        '13': '1010',
+        '23': '0110'
+}
+
+# ~~~~~~~~~
+
+def parse_args():
+    parser = argparse.ArgumentParser(description=__doc__)
+
+    parser.add_argument(
+        '--plot-name',
+        type=str,
+        help='file name for where to save plot',
+        default='temporary-plot.svg',
+    )
+    parser.add_argument(
+        '--in-pkl-name',
+        type=str,
+        help='file name for where to save pkl',
+        default=None,
+    )
+    parser.add_argument(
+        '--out-pkl-name',
+        type=str,
+        help='file name for where to save pkl',
+        default='output.pkl',
+    )
+    parser.add_argument(
+        '--delta',
+        type=float,
+        help='coarseness parameter for plotting',
+        default=0.01,
+    )
+    parser.add_argument(
+        '--n-jobs',
+        type=int,
+        help='number of processes to use for empirically computing maxima',
+        default=1,
+    )
+    parser.add_argument(
[…]

### 2018-06-22 appendix.tex
condition on gradients and plot of intuition

@@
 
+Tentative lemma:
+\begin{lemma}
+Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$ and $\emptyset$ be the maximal ancestral state for all site splits.
+There exists a set of $0 < x^*, y^* < 1$ such that $\{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2, 1, 1\}$ is the solution to maximizing \eqref{eq:profile_likelihood}.
+\end{lemma}
+
+Sketch of proof:
+We know the gradient is zero for the partials of the first three parameters.
+We must show then that the objective function is increasing in all directions for the second two parameters at the boundary.
+If the gradients for both $w$ and $y_2$ are positive at 1, the objective will be increasing in all increasing directions---for any positive direction, i.e., unit vector with nonnegative entries, the directional derivative is positive if each gradient is positive.
+Since the objective function is concave then this must be the maximum.
+Let
+\[
+a_{\emptyset} = \frac{(x_1+x_2)(y_1+y_2)}{(1+x_1x_2)(1+y_1y_2)}
+\]
+with similar definitions for the remaining site splits.
+The objective function proportional to $w$ is
+\begin{align*}
+\ell(w) \propto \log(1+w) + \sum_j p_{\siteSplit_j} \log(1 + a_{\siteSplit_j}w)
+\end{align*}
+and has gradient with respect to $w$ as
+\begin{align*}
+\ell_w(t) = \frac{1}{1+w} + \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{1 + a_{\siteSplit_j}w}.
+\end{align*}
+Note that, $a_{\emptyset} = -a_{13}$ and $a_{2} = -a_{123}$, and, for $x_1=x_2$,
+\[
+a_{1} = a_{3} = a_{12} = a_{23} = 0.
+\]
+
+For $y_2$, we do the same, but now with
+\begin{align*}
+\ell(y_2) \propto \log(1+y_2) + \sum_j p_{\siteSplit_j} \log(1 + b_{\siteSplit_j}y_2)
+\end{align*}
+where
+\[
+b_{\emptyset} = \frac{y_1+wx_1+wx_2+x_1y_1x_2}{1+x_1x_2+wx_1y_1+wx_2y_1}.
+\]
+For $x_1=x_2$, the denominator is always positive, allowing the above simplification.
+%It is. Since x_1=x_2 at the maximum we can use an argument like 1+x^2-2y_1wx > 1+x^2-2x = (1-x)^2 > 0. We'll need to ignore cases where x=0 (i.e., x^* or y^* are zero)
+The constants $b$ do not simplify as easily.
+In Fig.~\ref{fig:positive-gradients}, we see the region where our solution is $\hat{t}=\{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2, 1, 1\}$.
+
+\begin{figure}
+\centering
+%python joint_inf_plot.py --analytic --plot-name svg-fig/positive-gradients.svg --delta .001
+\includegraphics[width=.95\textwidth]{positive-gradients}
+\caption{
+}
+\label{fig:positive-gradients}
+\end{figure}
+
 \subsection*{Empirical validation}
@@
 \centering
+%python joint_inf_plot.py --empirical --plot-name svg-fig/empirical-incons.svg --delta .01 --n-jobs 10
 \includegraphics[width=.95\textwidth]{empirical-incons}
@@
 \end{figure}
+
+\begin{figure}
+\centering
+%python joint_inf_plot.py --empirical --plot-name svg-fig/empirical-incons-curve.svg --delta .01 --n-jobs 10 --plot-curve
+\includegraphics[width=.95\textwidth]{empirical-incons-curve}
+\caption{
+Inconsistency in joint inference with intuitive curves.
+The white curve is our condition on the gradients.
+The blue curve is a condition on $x^*$ and $y^*$ such that $\emptyset$ is the most likely ancestral state.
+}
+\label{fig:empirical-incons}
+\end{figure}

### 2018-06-26 appendix.tex
updated plots... looks like we found it! time to triple check...

@@
 \caption{
-Due to the symmetry of the likelihood, WLOG we assume $x_2 > x_1$ and $y_2 > y_1$ and maximize over ancestral states to reduce the number of possible functional forms to consider.
+Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral states to reduce the number of possible functional forms to consider.
 All values multiplied by $1/32$.
@@
 \centering
-%python joint_inf_plot.py --analytic --plot-name svg-fig/positive-gradients.svg --delta .001
+%python joint_inf_plot.py --optimum-conditions --plot-name svg-fig/positive-gradients.svg --delta .001 --n-jobs 10
 \includegraphics[width=.95\textwidth]{positive-gradients}
@@
 \label{fig:positive-gradients}
+\end{figure}
+
+\begin{figure}
+\centering
+%python joint_inf_plot.py --analytic --plot-name svg-fig/incons-analytic.svg --delta .001 --n-jobs 10
+\includegraphics[width=.95\textwidth]{incons-analytic}
+\caption{
+Region of inconsistency.
+}
+\label{fig:incons-analytic}
 \end{figure}
@@
 \centering
-%python joint_inf_plot.py --empirical --plot-name svg-fig/empirical-incons-curve.svg --delta .01 --n-jobs 10 --plot-curve
-\includegraphics[width=.95\textwidth]{empirical-incons-curve}
+%python joint_inf_plot.py --empirical --plot-name svg-fig/empirical-incons-intuition.svg --delta .01 --n-jobs 10 --plot-intuition
+\includegraphics[width=.95\textwidth]{empirical-incons-intuition}
 \caption{
@@
 }
-\label{fig:empirical-incons}
+\label{fig:empirical-incons-intuition}
 \end{figure}
+
+\begin{figure}
+\centering
+%python joint_inf_plot.py --empirical --plot-name svg-fig/empirical-incons-analytic.svg --delta .01 --n-jobs 10 --plot-curve
+\includegraphics[width=.95\textwidth]{empirical-incons-analytic}
+\caption{
+Inconsistency in joint inference with bounding curve.
+}
+\label{fig:empirical-incons-analytic}
+\end{figure}

### 2018-06-26 joint_inf_plot.py

@@ -24,6 +24,7 @@ sns.set_style('ticks')
 
 FONT_SIZE = 20
 
+LOG2 = np.log(2)
 LOG32 = np.log(32)
 
 ALL_PATTERNS = [
@@ -36,17 +37,6 @@ ALL_PATTERNS = [
         '1010',
         '0110'
 ]
-
-SPLIT_TO_PATTERN_DICT = {
-        '0':  '0000',
-        '1':  '1000',
-        '2':  '0100',
-        '3':  '0010',
-        '123':'1110',
-        '12': '1100',
-        '13': '1010',
-        '23': '0110'
-}
 
 class Legend(object):
     pass
@@ -90,6 +80,11 @@ def parse_args():
         help='plot analytic',
     )
     parser.add_argument(
+        '--optimum-conditions',
+        action="store_true",
+        help='plot conditions where we have y_2=1, w=1 at optimum',
+    )
+    parser.add_argument(
         '--empirical',
         action="store_true",
         help='compute empirical',
@@ -100,9 +95,14 @@ def parse_args():
         help='do marginal likelihood instead of joint inference',
     )
     parser.add_argument(
-        '--plot-curve',
+        '--plot-intuition',
         action='store_true',
         help='plot curve showing intuition for why this happens?',
+    )
+    parser.add_argument(
+        '--plot-curve',
+        action='store_true',
+        help='plot curve showing analytic region?',
     )
 
     args = parser.parse_args()
@@ -112,8 +112,10 @@ def parse_args():
 # ~~~~~~~~~
 # Functions for exact likelihood bounds
 
-def GRADIENTS(x, y):
+def GRADIENTS(xy):
     # w gradient
+    x = xy[0]
+    y = xy[1]
     w_output = .5
     theta = [x*y*y, y*y, x*y*y, 1., 1.]
     theta0 = [x, y, x, y, y]
@@ -161,73 +163,110 @@ def safe_log(x):
 def safe_p_logq(p, logq):
     return p * logq if p > 0. else 0.
 
-def bound(theta):
+#def likelihood_lower_bound(theta):
+#    """
+#    Table S4 but with w=1 and y_2=1 (and then x_1, y_1, x_2 their optima)
+#    """
[…]

### 2018-06-28 appendix.tex
reorganizing, formalizing, and plotting

--- 2018-06-22 appendix.tex
@@
 $\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$\\
-&$\{2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$\\
-&$\{1,2\}$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$\\
+&$\{1\}^*$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$\\
+&$\{2\}^*$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$\\
+&$\{1,2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$\\
 
@@
 &$\{1\}$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$\\
-&$\{2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$\\
+&$\{2\}^*$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$\\
 &$\{1,2\}$&$(1+x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$\\
@@
 $\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1-w)$\\
+&$\{1\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1-w)$\\
 &$\{2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1-w)$\\
@@
 &$\{1\}$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{2\}$&$(1+x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$\\
+&$\{2\}^*$&$(1+x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$\\
 &$\{1,2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
@@
 &$\{1\}$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{2\}$&$(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1-w)$\\
+&$\{2\}^*$&$(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1-w)$\\
 &$\{1,2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
@@
 &$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{2\}$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$\\
+&$\{2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$\\
 &$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
@@
 \end{tabular}
-\caption{Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
-All values multiplied by $1/32$.}
+\caption{
+Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
+All values multiplied by $1/32$.
+Ancestral states with $^*$ are never maximal provided parameters are in $(0,1)$.
+There are $3^5\cdot 4^2=3,888$ possible forms for the likelihood.
+}
 \label{tab:farris_likelihoods}
@@
 \caption{
-Due to the symmetry of the likelihood, WLOG we assume $x_2 > x_1$ and $y_2 > y_1$ and maximize over ancestral states to reduce the number of possible functional forms to consider.
+Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral states to reduce the number of possible functional forms to consider.
 All values multiplied by $1/32$.
 Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
-See Table~\ref{tab:farris_likelihoods} for full calculations.}
+There are $3^4=81$ possible forms for the likelihood.
+}
 \label{tab:likelihoods}
@@
 
-We now show that, if there exists a unique maximal ancestral state split for each site split then there exists a a unique estimate of $\tilde{t}$ when performing joint inference maximizing \eqref{eq:profile_likelihood}.
-To do so, we show the Hessian of the joint likelihood term is negative definite.
-First, we decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
-Recall the divergence
+\subsection*{Theorems and proofs}
+
+%where data is generated on the InvFels tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
+
+\topoInconsist*
+
+We first prove intermediate results to show the above.
+
+\begin{lemma}
+\label{lem:concavity}
+Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, $t=\{x_1, y_1, x_2, y_2, w\}$, and $\tau^*$ be the InvFels tree.
+For a fixed $\{\xi_1,\ldots,\xi_q\}$, the objective function
+\begin{align*}
+\ell_{\tau^*,t^*}(\tau^*, t) = \sum_{j=1}^\nSiteSplits &\left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t) \right.\\
+&\qquad+ \left.\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau^*, t)\right]
+\end{align*}
+is concave and
 \[
-\shannonDivergence_{\tau_1,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
+\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau^*, t)
 \]
-where data is generated on the InvFels tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
-By Gibbs inequality, this function attains a unique maximum when
+is a unique maximum.
+\end{lemma}
+
+\begin{proof}
+We focus on the two terms separately, showing they are both concave and, thus, their sum is concave.
+Call the second term
 \[
-\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)
+\tilde{\ell}_{\tau^*,t^*}(\tau^*, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau^*, t).
 \]
-for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
-Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the divergence, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
-Referring to Table~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
-The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
-\[
-2-2(x^*)^2 = 2-2x^2
-\]
-and for $\{2\}$ and $\{1,2\}$ yields
-\[
-2-2(y^*)^2 = 2-2y^2.
-\]
-implying $x=x^*$ and $y=y^*$.
-Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
-\[
-8x^*(y^*)^2 = 8x^*y^*w
-\]
-implying $w=y^*$.
-The unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and thus the Hessian of $\shannonDivergence$ is negative definite.
-
-We now focus on the partial log-likelihood term.
-Recall the partial log-likelihood takes the form
-\[
-\tilde{\ell}_{\tau_1,t^*}(\tau_1, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, t).
-\]
-Assume for now the functional form of the partial likelihood is fixed---in other words, each site split has a single partial likelihood function when maximizing over ancestral state splits.
-Then, the partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
+For fixed $\xi_j$, the partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
 \begin{align*}
-\tilde{\ell}_{\tau_1,t^*}(\tau_1, t) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_2}(x_2)\\
+\tilde{\ell}_{\tau^*,t^*}(\tau^*, t) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_2}(x_2)\\
 &\qquad + \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,w}(w).
@@
 Applying similar arguments to the other variables, the Hessian for the partial log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero.
-Thus, the Hessian of the sum of the divergence and the partial log-likelihood is negative definite, as the Hessian for the divergence is negative definite, provided there is a single maximal ancestral state split for each site split.
 
-\subsection*{Theorems and proofs}
+We now focus on the divergence term.
+By Gibbs inequality, this function attains a unique maximum when
+\[
+\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
+\]
+for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
+Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the divergence, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
+Referring to Table~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
+The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
+\[
+2-2(x^*)^2 = 2-2x^2
+\]
+and for $\{2\}$ and $\{1,2\}$ yields
+\[
+2-2(y^*)^2 = 2-2y^2.
+\]
+implying $x=x^*$ and $y=y^*$.
+Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
+\[
+8x^*(y^*)^2 = 8x^*y^*w
+\]
+implying $w=y^*$.
+The unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$.%, and thus the Hessian of $\shannonDivergence$ is negative definite.
+...
+\textbf{
+The Hessian as a function of $\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t)$ is negative definite, but is it as a function of $t$?
+This seems intuitively obvious, but I don't know how to show it.
+How could it not be true?
+}
+%Let D(p, q) = \sum_j p_j \log q_j
+%This is concave in q_j, trivially, but we are interested in concavity in t.
+%Because t generates q_j, how can this not be true?
+%Thought:
+%- if f(x) has a unique maximum on [a,b] and g(x) is concave on [a,b], all we want to show is that f(x)+g(x) has a unique maximum on [a,b]; moreover, f(x) is concave in an open ball around its maximum.
+%- note that we have D(p, q(\theta)), where q(\theta) is this sort of complicated function involving Hadamard matrices.
+however, in the trivial case, D(p, q(\theta)) is concave, so we may be able to use induction?
+%- note f(x) + sum_h g(x,h) has a unique solution; what about f(x) + max_h g(x,h)? we still need concavity...
+%- the sum of divergence and partial likelihood may be easy to show is concave as partial likelihood clearly is.
+%- empirical results show that the gradient condition is accurate, though that might mean we need to show there are no zero crossings of the derivative or something like that
+%Oblique strategies:
+%- the condition of y_2 \ge y_1 and x_2 \ge x_1 may play a role
+%- we could also find x,y where derivative has no solutions, so fn is always increasing, though this might require some more argument
+%- we could show (if it's true) that going from fidelities to the Hadamard representation results in a concave function in t, and from there we know the divergence is concave in t.
+%   this seems hard but may be our best bet; proceed by induction seems reasonable, but, again, hard.
+%   it's true in the "trivial" n=0 case, which encourages me.
+%   moreover, the fidelities generate the probabilities, and the objective is concave in the probabilities, which is more encouragement.
+\end{proof}
 
-Tentative theorem 1: 
-\begin{restatable}{theorem}{topoInconsist2}
-Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$.
-There exists a set of $0 < x^*, y^* < 1$ such that
-\[
-\{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\} = \arg\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t)
-\]
-and $\hat{w}\equiv 1$.
-\end{restatable}
+In words, this means that, if we have a single maximal ancestral state for each site split, we have a unique maximum.
 
-In words, this implies an inconsistency as we converge on the incorrect topology, estimating $\hat{w}$ to be $1$ and not $y^*$.
-
-Tentative lemma:
 \begin{lemma}
@@
 
-This is useful as it will result in the largest estimates of $w$ and $y_2$ and a single objective function with a unique maximum.
-Empirically, it looks as if, for small $x$ and large $y$, this results in $\hat{w} = \hat{y}_2 = 1$.
-We can see easily that, if this is the case, $\hat{x}_1=\hat{x}_2=x*(y^*)^2$ and $\hat{y}_1=(y^*)^2$.
-We then need to show
-\begin{enumerate}
-\item there exist $x^*,y^*$ that give us this form for the likelihood
-\item there exist $x^*,y^*$ such that this likelihood for $w$ and $y_2$ has a maximum on the boundary
-\item the intersection of these sets of $x^*,y^*$ is nonempty
-\end{enumerate}
+\begin{proof}
+We obtain
+\[
+\max_{t} \ \ell_{\tau^*,t^*}(\tau^*, t) \le
+    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
+    + \tilde{\ell}_{\tau^*,t^*}(\tau^*, \hat{t})
+\]
+as an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
+\[
+\shannonDivergence_{\tau^*,t^*}(\tau^*,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
+\]
+and
+\[
+\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau^*, t).
+\]
+By straightforward calculus, the function $f(x)=a_1\log(1+x)+a_2\log(1-x)$ attains a maximum at $\hat{x} = (a_1-a_2)/(a_1+a_2)$, and so we have a closed form for $\hat{t}$.
+Using this estimate, we have the lower bound
+\[
+\max_{t} \ \ell_{\tau^*,t^*}(\tau^*, t) \ge
+    \shannonDivergence_{\tau^*,t^*}(\tau^*,\hat{t})
+    + \tilde{\ell}_{\tau^*,t^*}(\tau^*, \hat{t}).
+\]
+Figure~\ref{fig:max-anc-state} shows where the lower bound above is greater than all 80 other upper bounds.
+\end{proof}
 
-Note that the above will hold if we swap $\hat{y}_2$ and $\hat{y}_1$, as the likelihood is invariant to this.
+\begin{figure}
+\centering
+\includegraphics[width=.95\textwidth]{analytic-anc-state-inkscape}
+\caption{
+Region where $\emptyset$ is maximal ancestral state for each site split.
+}
+\label{fig:max-anc-state}
+\end{figure}
 
-\subsection*{Empirical validation}
+The above will hold if we swap $\hat{y}_2$ and $\hat{y}_1$, as the likelihood is invariant to this.
 
-Consider Table~\ref{tab:likelihoods}.
-There are 81 separate possible forms for the likelihood, corresponding to the three possible most likely ancestral states for each of the four ambiguous site splits.
-Moreover, as shown previously, each of these objective functions has a unique maximum and is concave on $[0,1]^5$.
+\begin{lemma}
+Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$ and $\emptyset$ be the maximal ancestral state for all site splits.
+There exists a set of $0 < x^*, y^* < 1$ such that $\{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2, 1, 1\}$ is the solution to maximizing \eqref{eq:profile_likelihood}.
+\end{lemma}
 
-Assume $w=1$ and $y_2=1$.
-Then, there is a closed form solution maximizing the joint optimization procedure.
+\begin{proof}
+Assume $\emptyset$ is the most likely ancestral state for all site splits.
 In this case, the likelihood has no ambiguity, and has terms given in Table~\ref{tab:likelihoods-restricted}.
@@
 \hline
-$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)$\\
-
-$\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)$\\
-
-$\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)$\\
-
-$\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)$\\
-
-$\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)$\\
-
-$\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)$\\
-
-$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)$\\
-
-$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)$\\
+$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
+$\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
+$\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
+$\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
+$\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
+$\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
+$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
+$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
 \hline
@@
 
-\begin{table}[ht]
-\centering
-\begin{tabular}{|l|l|l|}
-    \hline
-$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,1,1\})$\\
-    \hline
-    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1x_2 +  y_1 +  [  x_1 + x_2][  y_1 + 1] + x_1y_1x_2 = (1+x_1)(1+y_1)(1+x_2)$\\
-    $\{1\}$     & $p_{1}$       &$1 - x_1x_2 +  y_1 +  [- x_1 + x_2][  y_1 + 1] - x_1y_1x_2 = (1-x_1)(1+y_1)(1+x_2)$\\
-    $\{2\}$     & $p_{2}$       &$1 + x_1x_2 -  y_1 +  [  x_1 + x_2][- y_1 + 1] - x_1y_1x_2 = (1+x_1)(1-y_1)(1+x_2)$\\
-    $\{3\}$     & $p_{3}$       &$1 - x_1x_2 +  y_1 +  [  x_1 - x_2][  y_1 + 1] - x_1y_1x_2 = (1+x_1)(1+y_1)(1-x_2)$\\
-    $\{1,2,3\}$ & $p_{123}$   &$1 + x_1x_2 -  y_1 +  [  x_1 + x_2][  y_1 - 1] - x_1y_1x_2 = (1-x_1)(1-y_1)(1-x_2)$\\
-    $\{1,2\}$   & $p_{12}$     &$1 - x_1x_2 -  y_1 +  [- x_1 + x_2][- y_1 + 1] + x_1y_1x_2 = (1-x_1)(1-y_1)(1+x_2)$\\
-    $\{1,3\}$   & $p_{13}$     &$1 + x_1x_2 +  y_1 +  [- x_1 - x_2][  y_1 + 1] + x_1y_1x_2 = (1-x_1)(1+y_1)(1-x_2)$\\
-    $\{2,3\}$   & $p_{23}$     &$1 - x_1x_2 -  y_1 +  [  x_1 - x_2][- y_1 + 1] + x_1y_1x_2 = (1+x_1)(1-y_1)(1-x_2)$\\
-    \hline
-\end{tabular}
-\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,1,1\}$ obtained using the Hadamard transform.
-All values multiplied by $1/8$.}
-\label{tab:gen-sitepatprob-restricted}
-\end{table}
-
-Additionally, we simplify the generating probabilities as in Table~\ref{tab:gen-sitepatprob-restricted}.
+Assume $w=1$ and $y_2=1$.
+Then, there is a closed form solution maximizing the joint optimization procedure, as follows.
+We simplify the generating probabilities as in Table~\ref{tab:gen-sitepatprob-restricted}.
 We see that, on simplifying, for each site split the generating probability is the same functional form as the likelihood.
@@
 \end{align*}
-Using the fact that $a_1\log(1+x)+a_2\log(1-x)$ attains a maximum at $\hat{x} = (a_1-a_2)/(a_1+a_2)$ we have the parameters
+We have estimates
 \[
@@
 \]
-and a closed form function of $x^*,y^*$ for this likelihood.
+and the gradient is equal to zero at this point.
 
+\begin{table}[ht]
+\centering
+\begin{tabular}{|l|l|l|}
+    \hline
+$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,1,1\})$\\
+    \hline
+    $\emptyset$ & $p_{\emptyset}$   &$(1+x_1)(1+y_1)(1+x_2)$\\
+    $\{1\}$     & $p_{1}$       &$(1-x_1)(1+y_1)(1+x_2)$\\
+    $\{2\}$     & $p_{2}$       &$(1+x_1)(1-y_1)(1+x_2)$\\
+    $\{3\}$     & $p_{3}$       &$(1+x_1)(1+y_1)(1-x_2)$\\
+    $\{1,2,3\}$ & $p_{123}$   &$(1-x_1)(1-y_1)(1-x_2)$\\
+    $\{1,2\}$   & $p_{12}$     &$(1-x_1)(1-y_1)(1+x_2)$\\
+    $\{1,3\}$   & $p_{13}$     &$(1-x_1)(1+y_1)(1-x_2)$\\
+    $\{2,3\}$   & $p_{23}$     &$(1+x_1)(1-y_1)(1-x_2)$\\
+    \hline
+\end{tabular}
+\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,1,1\}$ obtained using the Hadamard transform.
+All values multiplied by $1/8$.}
+\label{tab:gen-sitepatprob-restricted}
+\end{table}
+
+Now we show there are $0 < x^*, y^* < 1$ on which the gradient with respect to $w$ and $y_2$ is increasing on the boundary for all $x_1,y_1,x_2,y_2,w$.
+We use the fact that $x_1=x_2$ as there exists a unique solution and the likelihood is invariant to swapping these two variables.
+Let
+\[
+a_{\emptyset} = \frac{(x_1+x_2)(y_1+y_2)}{(1+x_1x_2)(1+y_1y_2)}
+\]
+with similar definitions for the remaining site splits.
+The objective function proportional to $w$ is
+\begin{align*}
+\ell(w) \propto \log(1+w) + \sum_j p_{\siteSplit_j} \log(1 + a_{\siteSplit_j}w)
+\end{align*}
+and has gradient with respect to $w$ as
+\begin{align*}
+\ell_w(t) = \frac{1}{1+w} + \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{1 + a_{\siteSplit_j}w}.
+\end{align*}
+%Note that, $a_{\emptyset} = -a_{13}$ and $a_{2} = -a_{123}$, and, for $x_1=x_2$,
+%\[
+%a_{1} = a_{3} = a_{12} = a_{23} = 0.
+%\]
+For $y_2$, we do the same, but now with
+\begin{align*}
+\ell(y_2) \propto \log(1+y_2) + \sum_j p_{\siteSplit_j} \log(1 + b_{\siteSplit_j}y_2)
+\end{align*}
+where
+\[
+b_{\emptyset} = \frac{y_1+wx_1+wx_2+x_1y_1x_2}{1+x_1x_2+wx_1y_1+wx_2y_1}.
+\]
+For $x_1=x_2$, the denominator is always positive, allowing the above simplification.
+%Since x_1=x_2 at the maximum we can use an argument like 1+x^2-2y_1wx > 1+x^2-2x = (1-x)^2 > 0. We'll need to ignore cases where x=0 (i.e., x^* or y^* are zero)
+In Fig.~\ref{fig:positive-gradients}, we see the region where the gradients of both $w$ and $y_2$ are increasing at $\hat{t}=\{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2, 1, 1\}$, and, thus, the objective function is increasing in all positive directions over $(y_2,w)$ on the boundary.
+Since, from Lemma~\ref{lem:concavity}, the objective function is concave on our domain, this must be the maximum.
+\end{proof}
+
+\begin{figure}
+\centering
+\includegraphics[width=.95\textwidth]{positive-gradients-inkscape}
+\caption{
+Region where gradients w.r.t. $w$ and $y_2$ are increasing on the boundary.
+}
+\label{fig:positive-gradients}
+\end{figure}
+
+\subsection*{Empirical validation}
+
+Consider Table~\ref{tab:likelihoods}.
+There are 81 separate possible forms for the likelihood, corresponding to the three possible most likely ancestral states for each of the four ambiguous site splits.
+Moreover, as shown previously, each of these objective functions has a unique maximum on $[0,1]^5$.
 In Fig.~\ref{fig:empirical-incons}, for various values of $x^*,y^*$, we compute the maximum of each of the 81 likelihoods, as well as the above restricted likelihood, and plot the generating values for which the above likelihood is greater than or equal to all 81 candidate likelihoods.
@@
 \end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=.95\textwidth]{empirical-bias}
+\caption{
+Bias in branch length estimation.
+In regions where joint optimization should perform well, there is systematic bias toward shorter branch lengths.
+}
+\label{fig:empirical-bias}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{empirical-marginal}
+\caption{
+    Estimates for $\hat{p}_w$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using L-BFGS-B optimizing the classical marginal likelihood \eqref{eq:marginal_likelihood} (rather than a joint optimization procedure).
+}
+\label{fig:bl-general-marginal}
+\end{figure}
+%
+%\begin{figure}
+%\centering
+%\includegraphics[width=.95\textwidth]{empirical-incons-intuition}
+%\caption{
+%Inconsistency in joint inference with intuitive curves.
+%The white curve is our condition on the gradients.
+%The blue curve is a condition on $x^*$ and $y^*$ such that $\emptyset$ is the most likely ancestral state.
+%}
+%\label{fig:empirical-incons-intuition}
+%\end{figure}

### 2018-06-28 joint_inf.tex
removing/rewriting to update results in main text

@@
 In this paper, we show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
+%TODO: change this line? it's still like an inconsistency in topology
 We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates an incorrect topology in the limit of infinite-length sequences.
@@
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model with data being generated on the four-taxon ``inverse Felsenstein'' tree (a.k.a.\ the ``Farris'' tree) and we construct bounds on the joint objective function to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
-We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.
+To do so, we use a binary symmetric model with data being generated on the four-taxon ``inverse Felsenstein'' tree (a.k.a.\ the ``Farris'' tree) and we construct bounds on the joint objective function to demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
+We find similar areas where joint inference consistently underestimates interior branch lengths when the topology is known and fixed.
 
@@
 \end{equation}
-In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima row-wise in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} to obtain the likelihood term for each most likely ancestral state, these terms depending on the unknown $(\tau, t)$.
+In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral state splits in Table~\ref{tab:farris_likelihoods} to obtain the likelihood term for each most likely ancestral state, these terms depending on the unknown $(\tau, t)$.
 For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2018-xl}.
@@
 
-\subsection*{Inconsistency in topology estimation}
+To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the InvFels topology (Fig.~\ref{fig:farris-fels-top}a).
+%We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology ($\tau_2$) than for the true, generating InvFels topology ($\tau_1$).
+%To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the InvFels topology as a function of $x^*$ and $y^*$ and, similarly, a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
+%When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the InvFels case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
 
-To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the InvFels topology (Fig.~\ref{fig:farris-fels-top}a).
-We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology ($\tau_2$) than for the true, generating InvFels topology ($\tau_1$).
-To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the InvFels topology as a function of $x^*$ and $y^*$ and, similarly, a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
-When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the InvFels case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
 \begin{restatable}{theorem}{topoInconsist}
-Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
-There exist $C_0(x^*, y^*), C_1(x^*, y^*),$ and a set of $0 < x^*, y^* < 1$ such that
+Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$.
+There exists a set of $0 < x^*, y^* < 1$ such that
 \[
-\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le C_0(x^*, y^*),
+\{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\} = \arg\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t)
 \]
-\[
-C_1(x^*, y^*) \le \max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t)
-\]
-with $C_0(x^*, y^*) < C_1(x^*, y^*)$.
+and $\hat{w}\equiv 1$.
 \end{restatable}
+
+In words, this implies an inconsistency as we converge on the incorrect topology, estimating the interior branch length to be zero.
 The proof of this theorem is by a detailed examination of inequalities.
-Intuitively, $\tau_2$ is favored in performing joint inference since the objective function for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its objective function while $\tau_2$ has many more.
-This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.
+%Intuitively, $\tau_2$ is favored in performing joint inference since the objective function for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its objective function while $\tau_2$ has many more.
+%This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.
 
+%TODO: change legend of this figure---the intersection is just the region where \emptyset is maximal, so it will be the same plot
 \begin{figure}
 \centering
-% obtained by running
-% python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
-\includegraphics[width=\textwidth]{topology-inconsistency-inkscape}
+\includegraphics[width=.95\textwidth]{incons-analytic-inkscape}
 \caption{
-    An analytically-derived region of topological inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
+    An analytically-derived region of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
     Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
 }
-\label{fig:inconsistency-farris}
+\label{fig:incons-analytic}
 \end{figure}
-
-\subsection*{Inconsistency in branch length estimation}
-
-We now consider the problem of branch length estimation on the correct tree using joint estimation.
-As described above, we use the notion of fidelities to analyze what happens as changes occur along each branch as opposed to the standard branch length formulation.
-We analyze two settings on the InvFels tree, corresponding to whether some branch fidelities are fixed at their true values or not.
-As above, assume that data is generated from the InvFels tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
-
-In the general case, we do the same but with
-\begin{equation}
-\label{eq:general-bl-opt}
-(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\})
-\end{equation}
-to find a region where the inferred values do not converge to the generating values.
-In the Appendix we consider a ``restricted'' case where we show that, for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
-That is, if we estimate $x$ and $y$ correctly, then, for
-\[
-\hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
-\]
-there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
-These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \citep{RoyChoudhury2015-ta}.
-
-%Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
-%We show there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
-The estimate $\hat{w}$ in \eqref{eq:general-bl-opt} is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
-Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$.
-We fix bounds on these estimates and show that, in optimizing \eqref{eq:general-bl-opt}, joint estimation either estimates $\hat{w}$ to be one or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
-\begin{restatable}{theorem}{generalBranchInconsist}
-\label{thm:general-bl}
-Define
-\[
-\gamma(x, y) = 4xy,
-\]
-\[
-\beta(x, y) = 1+x^2+y^2+x^2y^2,
-\]
-and bounds
-\[
-\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
-\]
-\[
-\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
-\]
-and
-\[
-\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}).
-\]
-Assume $\hat{x}\neq 0,\hat{y}\neq 0,$ $\hat{x}$ and $\hat{y}$ are not both simultaneously one, and $x^*$ and $y^*$ are not both simultaneously zero.
-Then, the maximum likelihood value $\hat{w}=1$ when
-\[
--\gamma_{U}^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
-\]
-\end{restatable}
-We use this theorem to show incorrect branch parameter estimates as follows.
-If we do not tolerate any error in pendant branches, we use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$; this corresponds to the restricted case, analyzed in the Appendix (Fig.~\ref{fig:bl-inconsistency}).
-For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
-\[
-x^*-x_{L} \le \hat{x} \le x^*+x_{U}
-\]
-and
-\[
-y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
-\]
-Then $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
-As an example, Fig.~\ref{fig:bl-loose-inconsistency} shows the case where we have poor estimation of one of the three branch parameters when $x_L=x_U=y_L=y_U=0.1$.
-In this case, joint inference will either estimate $\hat{w}$ to have fidelity one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.
 
@@
 This tendency to include $(1+w)$ terms in the likelihood results in a bias toward one of branch fidelities.
+This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split (Table~\ref{tab:likelihoods} in the Appendix with $x_1=x_2=x^*$ and $y_1=y_2=y^*$).
 If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.
-
-\begin{figure}
-\centering
-% obtained by running
-% python joint_inf_plot.py --analytic --general-branch-lengths --delta .001 --plot-name figures/bl-loose-inconsistency.svg
-\includegraphics[width=\textwidth]{bl-loose-inconsistency-inkscape}
-\caption{
-    Analytically-derived region of branch parameter inconsistency in terms of probabilities of a character change with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
-    In the marked region, either $\hat{p}_w=0$ or one of $p_{x^*}-.05 \le \hat{p}_x \le p_{x^*}+.05$ or $p_{y^*}-.05 \le \hat{p}_y \le p_{y^*}+.05$ will not be true, resulting in poor estimation of the pendant branch parameters.
-    In terms of fidelities, these bounds are $x^*-0.1 \le \hat{x} \le x^*+0.1$ and $y^*-0.1 \le \hat{y} \le y^*+0.1$.
-    As in previous plots, the shaded and white regions are loose indications of inconsistency.
-}
-\label{fig:bl-loose-inconsistency}
-\end{figure}
 
@@
 
-Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:bl-loose-inconsistency}).
-To determine how conservative, we use the method of basin-hopping \citep{Wales1997} to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
-We see that the region of inconsistency in the general case is similar to that of the restricted case (compare Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency}).
-This region encompasses the majority of the branch parameter space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch length to be zero.
+Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:incons-analytic}).
+To determine how conservative, we use the well-known L-BFGS-B method to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
+We see that the region of inconsistency in the general case is much larger than in the analytic case---the analytic case required $\hat{y}_2 = 1$ and dealt with loose bounds, so this is reasonable.
+This region encompasses almost half of the branch parameter space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch length to be zero.
 
-We provide a full description of our optimization procedure in the Appendix, but briefly, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher objective function.
+We provide a full description of our optimization procedure in the Appendix, but briefly, we perform 81 maximizations for each separate likelihood function.
+For each function, we optimize the objective starting at five separate initial values to mitigate problems with small curvature of the objective surface.
 We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$.
@@
 Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same procedure.
-For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look reasonable, with $\hat{w}$ increasing as $y^*$ increases, though Fig.~\ref{fig:bl-general-bias} shows there is a systematic positive bias in this procedure even when $\hat{w}$ is not estimated to be one.
-To understand the quality of each fit, we report the range of $\hat{w}-y^*$ where $\hat{w}\neq 1$.
-For joint inference, the errors range from $[-7\times 10^{-3}, 8\times 10^{-2}]$ and for marginal inference, $[-8\times 10^{-8}, 5\times 10^{-7}]$ showing that, even in cases where joint inference does not estimate $\hat{w}$ to be exactly one, it still fails to achieve a low error from truth.
+For joint inference, the estimates for $\hat{p}_w$ when $p_{x^*}$ and $p_{y^*}$ are both small look reasonable, with $\hat{p}_w$ increasing as $p_{y^*}$ increases, though Fig.~\ref{fig:empirical-bias} shows there is a systematic negative bias in this procedure even when the true branches are short.
+To understand the quality of each fit, we report the range of $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.
+Due to the discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}), an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization.
+For joint inference, the errors range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for marginal inference, $[-3\times 10^{-17}, 3\times 10^{-17}]$ showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
 
@@
 \centering
-% obtained by running
-% python joint_inf_plot.py --empirical --general-branch-lengths --delta .01 --plot-name figures/w-hat-empirical-01.svg --out-pkl-name figures/w-hat-empirical-01.pkl --n-jobs 16
-% or
-% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-01.svg --in-pkl-name ../intermed/w-hat-empirical-01.pkl
-% if fit already
-\includegraphics[width=\textwidth]{w-hat-empirical-01}
+\includegraphics[width=\textwidth]{empirical-estimate}
 \caption{
-    Numerical estimates for $\hat{p}_w=(1-\hat{w})/2$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing \eqref{eq:profile_likelihood}.
-    Data generated as in Fig.~\ref{fig:inconsistency-farris}.
+    Numerical estimates for $\hat{p}_w=(1-\hat{w})/2$ when computing $(\hat{x}, \hat{y}, \hat{w})$ optimizing \eqref{eq:profile_likelihood}.
+    Data generated as in Fig.~\ref{fig:incons-analytic}
 }
@@
 We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
-Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference converges on the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
-In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.
+Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of inconsistency: the joint inference procedure estimates severely biased branch lengths even given the correct topology.
+These regions of inconsistency arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
 This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and 3.
-\citet{Sagulenko2018-xl} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:bl-general-bias} shows even when branches are short there is a consistent bias.
+\citet{Sagulenko2018-xl} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:empirical-bias} shows even when branches are short there is a consistent bias.
 
@@
 In this previous work, when simulating data under the Felsenstein topology $\tau_2$, as the number of observations increases, the InvFels topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
-This is the converse of what we have shown for joint inference, where the Felsenstein topology is more likely than the InvFels topology.
+We have shown cases in which, when generating from the InvFels topology, we converge to a multifurcating topology, with one or more branch lengths estimated to be zero.
 Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
-This is not the case here; for our case, the inconsistency generally occurs when all branches are long, and has more to do with the choices of the form of the likelihood than from the interplay of long and short branches.
+This is not the case here; while analytically the inconsistency occurs in the long branch attraction region of the parameter space, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
 Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.
@@
 In it, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general.
-We show cases where the estimates of the topology will be inconsistent if we are estimating both topology and branch lengths, and, moreover, even if we know the correct topology, we show cases where branch lengths will also be inconsistent.
-Finally, just prior to his conclusion, he discusses when parsimony will give the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.
+We show cases where the estimates of the topology are inconsistent if we are estimating both topology and branch lengths, and, moreover, even if we know the correct topology, we show cases where branch lengths are severely biased.
+Finally, just prior to his conclusion, he discusses when parsimony gives the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.
 The joint inference procedure outlined here estimates the same parameters as classical maximum likelihood---albeit implicitly estimating ancestral states as well---rendering the question well-posed.
@@
 The techniques used to obtain upper and lower bounds for the likelihoods in the topology estimation case provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
-Similarly, for the general case in estimating branch lengths, we were only able to provide a conservative region of overestimation, and the unusual shape we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) begs further investigation.
-Empirical validation shows that the general case is not unlike the restricted case.
+Similarly, for the general case in estimating branch lengths, we were only able to provide a conservative region of overestimation, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) begs further investigation.
+Empirical validation shows that the region of inconsistency is much larger than the one analytically derived.
 All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
@@
 We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.
+Insight from the editors and Vu Dinh greatly improved this manuscript.
 

### 2018-06-28 joint_inf_plot.py
make plot function, get what/marginal

@@ -11,6 +11,7 @@ import numpy as np
 import argparse
 import sys
 import time
+import pickle
 
 from multiprocessing import Pool
 from scipy.optimize import minimize
@@ -24,6 +25,7 @@ sns.set_style('ticks')
 
 FONT_SIZE = 20
 
+# constants to slightly speed up computation
 LOG2 = np.log(2)
 LOG32 = np.log(32)
 
@@ -75,7 +77,7 @@ def parse_args():
         default=1,
     )
     parser.add_argument(
-        '--analytic',
+        '--ancestral-state-conditions',
         action="store_true",
         help='plot analytic',
     )
@@ -85,24 +87,31 @@ def parse_args():
         help='plot conditions where we have y_2=1, w=1 at optimum',
     )
     parser.add_argument(
-        '--empirical',
+        '--empirical-inconsistency',
+        action="store_true",
+        help='compute empirical',
+    )
+    parser.add_argument(
+        '--empirical-parameter-estimate',
         action="store_true",
         help='compute empirical',
     )
     parser.add_argument(
         '--marginal',
         action="store_true",
-        help='do marginal likelihood instead of joint inference',
+        help='compute marginal',
     )
     parser.add_argument(
-        '--plot-intuition',
-        action='store_true',
-        help='plot curve showing intuition for why this happens?',
+        '--in-pkl-name',
+        type=str,
+        help='file name for where to save pkl',
+        default=None,
     )
     parser.add_argument(
-        '--plot-curve',
-        action='store_true',
-        help='plot curve showing analytic region?',
+        '--out-pkl-name',
+        type=str,
+        help='file name for where to save pkl',
+        default='output.pkl',
     )
 
     args = parser.parse_args()
@@ -111,20 +120,6 @@ def parse_args():
 
 # ~~~~~~~~~
 # Functions for exact likelihood bounds
-
-def GRADIENTS(xy):
-    # w gradient
-    x = xy[0]
-    y = xy[1]
-    w_output = .5
-    theta = [x*y*y, y*y, x*y*y, 1., 1.]
-    theta0 = [x, y, x, y, y]
[…]

### 2018-07-06 appendix.tex
remove outdated sentence
bounds on estimate arguments, new region of inconsistency

@@
 \end{figure}
-%TODO: update this figure with general branch lengths
 
@@
 
-\topoInconsist*
-
 We first prove intermediate results to show the above.
-
-\begin{lemma}
-\label{lem:concavity}
-Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, $t=\{x_1, y_1, x_2, y_2, w\}$, and $\tau^*$ be the InvFels tree.
-For a fixed $\{\xi_1,\ldots,\xi_q\}$, the objective function
-\begin{align*}
-\ell_{\tau^*,t^*}(\tau^*, t) = \sum_{j=1}^\nSiteSplits &\left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t) \right.\\
-&\qquad+ \left.\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau^*, t)\right]
-\end{align*}
-is concave and
-\[
-\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau^*, t)
-\]
-is a unique maximum.
-\end{lemma}
-
-\begin{proof}
-We focus on the two terms separately, showing they are both concave and, thus, their sum is concave.
-Call the second term
-\[
-\tilde{\ell}_{\tau^*,t^*}(\tau^*, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau^*, t).
-\]
-For fixed $\xi_j$, the partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
-\begin{align*}
-\tilde{\ell}_{\tau^*,t^*}(\tau^*, t) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_2}(x_2)\\
-&\qquad + \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,w}(w).
-\end{align*}
-Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
-Without loss of generality we focus on the variable $x_1$ and the partial log-likelihood proportional to
-\[
-\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1).
-\]
-Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, is
-\[
-h_{j,x_1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
-\]
-for $e_j\in\{0,1\}$, which, simplifying, results in
-\begin{align}
-\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
-                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1). \label{eq:partial-x-form}
-\end{align}
-We have the second derivative
-\[
-\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
-\]
-As $x_1\in(0,1)$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
-Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
-Applying similar arguments to the other variables, the Hessian for the partial log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero.
-
-We now focus on the divergence term.
-By Gibbs inequality, this function attains a unique maximum when
-\[
-\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
-\]
-for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
-Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the divergence, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
-Referring to Table~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
-The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
-\[
-2-2(x^*)^2 = 2-2x^2
-\]
-and for $\{2\}$ and $\{1,2\}$ yields
-\[
-2-2(y^*)^2 = 2-2y^2.
-\]
-implying $x=x^*$ and $y=y^*$.
-Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
-\[
-8x^*(y^*)^2 = 8x^*y^*w
-\]
-implying $w=y^*$.
-The unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$.%, and thus the Hessian of $\shannonDivergence$ is negative definite.
-...
-\textbf{
-The Hessian as a function of $\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t)$ is negative definite, but is it as a function of $t$?
-This seems intuitively obvious, but I don't know how to show it.
-How could it not be true?
-}
-%Let D(p, q) = \sum_j p_j \log q_j
-%This is concave in q_j, trivially, but we are interested in concavity in t.
-%Because t generates q_j, how can this not be true?
-%Thought:
-%- if f(x) has a unique maximum on [a,b] and g(x) is concave on [a,b], all we want to show is that f(x)+g(x) has a unique maximum on [a,b]; moreover, f(x) is concave in an open ball around its maximum.
-%- note that we have D(p, q(\theta)), where q(\theta) is this sort of complicated function involving Hadamard matrices.
-however, in the trivial case, D(p, q(\theta)) is concave, so we may be able to use induction?
-%- note f(x) + sum_h g(x,h) has a unique solution; what about f(x) + max_h g(x,h)? we still need concavity...
-%- the sum of divergence and partial likelihood may be easy to show is concave as partial likelihood clearly is.
-%- empirical results show that the gradient condition is accurate, though that might mean we need to show there are no zero crossings of the derivative or something like that
-%Oblique strategies:
-%- the condition of y_2 \ge y_1 and x_2 \ge x_1 may play a role
-%- we could also find x,y where derivative has no solutions, so fn is always increasing, though this might require some more argument
-%- we could show (if it's true) that going from fidelities to the Hadamard representation results in a concave function in t, and from there we know the divergence is concave in t.
-%   this seems hard but may be our best bet; proceed by induction seems reasonable, but, again, hard.
-%   it's true in the "trivial" n=0 case, which encourages me.
-%   moreover, the fidelities generate the probabilities, and the objective is concave in the probabilities, which is more encouragement.
-\end{proof}
-
-In words, this means that, if we have a single maximal ancestral state for each site split, we have a unique maximum.
 
@@
 Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$ and $\emptyset$ be the maximal ancestral state for all site splits.
-There exists a set of $0 < x^*, y^* < 1$ such that $\{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2, 1, 1\}$ is the solution to maximizing \eqref{eq:profile_likelihood}.
+The solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
+\[
+\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau^*, t)
+\]
+is subject to the constraints $\hat{x}_1,\hat{x}_2 \in [0, x^*]$, $\hat{y}_1 \in [0, y^*]$, and $\hat{y}_2,\hat{w} \in [y^*, 1]$.
+\label{lemma:bounds}
 \end{lemma}
@@
 \begin{proof}
+We proceed by analyzing the gradients of each component of the likelihood function.
+We first focus on the divergence term
+\[
+\shannonDivergence_{\tau^*,t^*}(\tau^*,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t).
+\]
+By Gibbs inequality, this function attains a unique maximum when
+\[
+\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
+\]
+for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
+Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the divergence, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
+Referring to Table~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
+The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
+\[
+2-2(x^*)^2 = 2-2x^2
+\]
+and for $\{2\}$ and $\{1,2\}$ yields
+\[
+2-2(y^*)^2 = 2-2y^2.
+\]
+implying $x=x^*$ and $y=y^*$.
+Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
+\[
+8x^*(y^*)^2 = 8x^*y^*w
+\]
+implying $w=y^*$.
+The unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and
+\[
+\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau^*,t) \mid_{t=t^*} = 0.
+\]
+
 Assume $\emptyset$ is the most likely ancestral state for all site splits.
-In this case, the likelihood has no ambiguity, and has terms given in Table~\ref{tab:likelihoods-restricted}.
+In this case, the partial likelihood has no ambiguity, and has terms given in Table~\ref{tab:likelihoods-restricted}.
 
@@
 
-Assume $w=1$ and $y_2=1$.
-Then, there is a closed form solution maximizing the joint optimization procedure, as follows.
-We simplify the generating probabilities as in Table~\ref{tab:gen-sitepatprob-restricted}.
-We see that, on simplifying, for each site split the generating probability is the same functional form as the likelihood.
-Thus, the joint optimization procedure has a unique, closed form solution solving
-\begin{align*}
-&\arg\max_{x_1,y_y,x_2} \ (p_{\emptyset} + p_2 + p_3 + p_{23}) \log(1+x_1) + (p_1 + p_{123} + p_{12} + p_{13}) \log(1-x_1)\\
-               &\qquad + (p_{\emptyset} + p_1 + p_3 + p_{13}) \log(1+y_1) + (p_2 + p_{123} + p_{12} + p_{23}) \log(1-y_1)\\
-               &\qquad + (p_{\emptyset} + p_1 + p_2 + p_{12}) \log(1+x_2) + (p_3 + p_{123} + p_{13} + p_{23}) \log(1-x_2)\\
-&\propto\arg\max_{x_1,y_y,x_2} \ (4+4x*(y^*)^2) \log(1+x_1) + (4-4x*(y^*)^2) \log(1-x_1)\\
-               &\qquad + (4+4(y^*)^2) \log(1+y_1) + (4-4(y^*)^2) \log(1-y_1)\\
-               &\qquad + (4+4x*(y^*)^2) \log(1+x_2) + (4-4x*(y^*)^2) \log(1-x_2).
-\end{align*}
-We have estimates
+We focus on $x_1$, similar arguments applying to the other variables.
+The partial likelihood
 \[
-\{\hat{x}_1,\hat{y}_1,\hat{x}_2\} = \{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2\}
+\tilde{\ell}_{\tau^*,t^*}(\tau^*, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau^*, t)
 \]
-and the gradient is equal to zero at this point.
+as a function of $x_1$, using Table~\ref{tab:sitepatprob}, is
+\[
+\tilde{\ell}(x_1) \propto \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1).
+\]
+By some calculus, the gradient with respect to $x_1$ is
+\[
+\tilde{\ell}_{x_1}(x_1) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2}
+\]
+which is clearly negative for all $x_1 \ge x^*$.
+%TODO: make more explicit/clear why this means \hat{x}_1 \in [0,x^*]
+This same process applied to the remaining variables shows the gradient of the full likelihood is negative for $x_1,x_2 \ge x^*$ and $y_1 \ge y^*$ and positive for $y_2,w \ge y^*$, yielding the constraints
+\[
+\hat{x}_1,\hat{x}_2 \in [0, x^*], \ \hat{y}_1 \in [0, y^*], \ \hat{y}_2,\hat{w} \in [y^*, 1].
+\]
+\end{proof}
 
-\begin{table}[ht]
-\centering
-\begin{tabular}{|l|l|l|}
-    \hline
-$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,1,1\})$\\
-    \hline
-    $\emptyset$ & $p_{\emptyset}$   &$(1+x_1)(1+y_1)(1+x_2)$\\
-    $\{1\}$     & $p_{1}$       &$(1-x_1)(1+y_1)(1+x_2)$\\
-    $\{2\}$     & $p_{2}$       &$(1+x_1)(1-y_1)(1+x_2)$\\
-    $\{3\}$     & $p_{3}$       &$(1+x_1)(1+y_1)(1-x_2)$\\
-    $\{1,2,3\}$ & $p_{123}$   &$(1-x_1)(1-y_1)(1-x_2)$\\
-    $\{1,2\}$   & $p_{12}$     &$(1-x_1)(1-y_1)(1+x_2)$\\
-    $\{1,3\}$   & $p_{13}$     &$(1-x_1)(1+y_1)(1-x_2)$\\
-    $\{2,3\}$   & $p_{23}$     &$(1+x_1)(1-y_1)(1-x_2)$\\
-    \hline
-\end{tabular}
-\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,1,1\}$ obtained using the Hadamard transform.
-All values multiplied by $1/8$.}
-\label{tab:gen-sitepatprob-restricted}
-\end{table}
+We now prove the main result.
 
-Now we show there are $0 < x^*, y^* < 1$ on which the gradient with respect to $w$ and $y_2$ is increasing on the boundary for all $x_1,y_1,x_2,y_2,w$.
-We use the fact that $x_1=x_2$ as there exists a unique solution and the likelihood is invariant to swapping these two variables.
+\topoInconsist*
+
+\begin{proof}
+%TODO: make this more explicit/clear
+We use the fact that $x_1=x_2=x$ as there exists a unique solution and the likelihood is invariant to swapping these two variables.
 Let
 \[
-a_{\emptyset} = \frac{(x_1+x_2)(y_1+y_2)}{(1+x_1x_2)(1+y_1y_2)}
+a_{\emptyset} = \frac{(2x)(y_1+y_2)}{(1+x^2)(1+y_1y_2)}
 \]
@@
 \end{align*}
-and has gradient with respect to $w$ as
+and has derivative with respect to $w$ as
 \begin{align*}
-\ell_w(t) = \frac{1}{1+w} + \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{1 + a_{\siteSplit_j}w}.
+\ell'(w) := \frac{d}{dw} \ell(w) = \frac{1}{1+w} + \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{1 + a_{\siteSplit_j}w}.
 \end{align*}
-%Note that, $a_{\emptyset} = -a_{13}$ and $a_{2} = -a_{123}$, and, for $x_1=x_2$,
-%\[
-%a_{1} = a_{3} = a_{12} = a_{23} = 0.
-%\]
-For $y_2$, we do the same, but now with
+Note that, $a_{\emptyset} = -a_{13}$, $a_{2} = -a_{123}$, $p_{2} = p_{123}$, and for $x_1=x_2$
+\[
+a_{1} = a_{3} = a_{12} = a_{23} = 0
+\]
+so
 \begin{align*}
-\ell(y_2) \propto \log(1+y_2) + \sum_j p_{\siteSplit_j} \log(1 + b_{\siteSplit_j}y_2)
+\ell'(w) = \frac{1}{1+w} + \frac{p_{\emptyset}a_{\emptyset}}{1 + a_{\emptyset}w} - \frac{p_{13}a_{\emptyset}}{1 - a_{\emptyset}w} + \frac{p_{2}a_{2}}{1 + a_{2}w} - \frac{p_{2}a_{2}}{1 - a_{2}w}.
 \end{align*}
-where
+Using the bounds on $\hat{x}_1,\hat{y}_1,\hat{y}_1$, and $\hat{y}_2$ from Lemma~\ref{lemma:bounds}, we have
 \[
-b_{\emptyset} = \frac{y_1+wx_1+wx_2+x_1y_1x_2}{1+x_1x_2+wx_1y_1+wx_2y_1}.
+0 \le a_{\emptyset} \le \frac{2x^*}{1+(x^*)^2},
 \]
-For $x_1=x_2$, the denominator is always positive, allowing the above simplification.
-%Since x_1=x_2 at the maximum we can use an argument like 1+x^2-2y_1wx > 1+x^2-2x = (1-x)^2 > 0. We'll need to ignore cases where x=0 (i.e., x^* or y^* are zero)
-In Fig.~\ref{fig:positive-gradients}, we see the region where the gradients of both $w$ and $y_2$ are increasing at $\hat{t}=\{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2, 1, 1\}$, and, thus, the objective function is increasing in all positive directions over $(y_2,w)$ on the boundary.
-Since, from Lemma~\ref{lem:concavity}, the objective function is concave on our domain, this must be the maximum.
+\[
+0 \le a_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*},
+\]
+so that $\ell'(w)$ has lower bound
+\[
+\ell'_{L}(w) = \frac{1}{1+w} + - \frac{p_{13}}{c_x^* - w} - \frac{p_{2}}{c_x^*c_y^* - w}
+\]
+with
+\[
+c_x^* = \frac{1+(x^*)^2}{2x^*},
+\]
+\[
+c_y^* = 1-y^*.
+\]
+The solution $\hat{w}$ solving $\ell'(w) = 0$ is greater than $\hat{w}'$ solving $\ell'_{L}(w) = 0$.
+As $c_x^* > 1$, provided $c_x^*c_y^* > 1$, $\ell'_{L}(w)$ stays finite and, when considering $\ell'_{L}(w) \lessgtr 0$, we equivalently consider $f(w) \lessgtr 0$ where $f$ is the quadratic function
+\begin{align*}
+f(w) &= (w)^2\cdot(1+p_{13}+p_{2}) \\
+      &\qquad + w\cdot(-c_x^*-c_x^*c_y^*-p_{13}c_x^*c_y^*+p_{13}-p_{2}c_x^*+p_{2}) \\
+      &\qquad + ((c_x^*)^2c_y^*-p_{13}c_x^*c_y^*-p_{2}c_x^*).
+\end{align*}
+We see
+\[
+\ell'_L(0) = 1 - \frac{p_{13}}{c_x^*} - \frac{p_{2}}{c_x^*c_y^*} > 0
+\]
+provided again that $c_x^*c_y^* > 1$.
+Now this quadratic has either zero, one, or two solutions.
+If there are zero solutions, the derivative with respect to $w$ is increasing on all of $[0,1]$ so that $\hat{w} = 1$ and we are done.
+If there are one or two solutions,
+if the smaller of the zeros of $f(w)$ is greater than one, then $\hat{w} \equiv 1$.
+Using the quadratic formula with
+\[
+a = 1+p_{13}+p_{2},
+\]
+\[
+b = -c_x^*-c_x^*c_y^*-p_{13}c_x^*c_y^*+p_{13}-p_{2}c_x^*+p_{2},
+\]
+\[
+c = (c_x^*)^2c_y^*-p_{13}c_x^*c_y^*-p_{2}c_x^*,
+\]
+the smaller zero is
+\[
+\hat{w} = \frac{-b - \sqrt{b^2 - 4ac}}{2a},
+\]
+which is a function of the generating parameters $x^*$ and $y^*$.
+We see that $a \ge 0$ and, provided that $2a+b \leq 0$, we have,
+\[
+\hat{w} \ge 1 \iff
+|2a + b| \leq \sqrt{b^2 - 4ac} \iff
+%-b - \sqrt{b^2 - 4ac} \le 2a,
+%0 \ge -b - 2a \le \sqrt{b^2 - 4ac},
+%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
+%a^2 + ab + ac \le 0,
+a + b + c \ge 0.
+\]
+Therefore, we have three conditions to satisfy for $\hat{w} \equiv 1$:
+\[
+\mathrm{C}1: c_x^*c_y^* > 1,
+\]
+\[
+\mathrm{C}2: 2-c_x^*c_y^*-c_x^*+p_{13}(3-c_x^*c_y^*)+p_{2}(3-c_x^*) \leq 0,
+\]
+\[
+\mathrm{C}3: 1+(c_x^*)^2c_y^*-c_x^*c_y^*-c_x^*+p_{13}(2-2c_x^*c_y^*)+p_{2}(2-2c_x^*) \ge 0.
+\]
+The set of $x^*,y^*$ satisfying these conditions as well as those for $\emptyset$ being the most likely ancestral state is given in Figure~\ref{fig:incons-analytic}.
 \end{proof}
 
-\begin{figure}
-\centering
-\includegraphics[width=.95\textwidth]{positive-gradients-inkscape}
-\caption{
-Region where gradients w.r.t. $w$ and $y_2$ are increasing on the boundary.
-}
-\label{fig:positive-gradients}
-\end{figure}
+As the region in Figure~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on the incorrect topology in these cases, resulting in an inconsistency.
 
@@
 There are 81 separate possible forms for the likelihood, corresponding to the three possible most likely ancestral states for each of the four ambiguous site splits.
-Moreover, as shown previously, each of these objective functions has a unique maximum on $[0,1]^5$.
-In Fig.~\ref{fig:empirical-incons}, for various values of $x^*,y^*$, we compute the maximum of each of the 81 likelihoods, as well as the above restricted likelihood, and plot the generating values for which the above likelihood is greater than or equal to all 81 candidate likelihoods.
-
-\begin{figure}
-\centering
-\includegraphics[width=.95\textwidth]{empirical-incons}
-\caption{
-Inconsistency in joint inference.
-In the white region, one of the possibly many equally likely maxima has the property that $\hat{w}=1$ and $\hat{y}_2=1$, indicating we converge to a topology that is not the one that generated the data.
-The gray regions are edge cases where one of the generating parameters was zero or one.
-}
-\label{fig:empirical-incons}
-\end{figure}
+In Fig.~\ref{fig:bl-general-inconsistency}, for various values of $x^*,y^*$, we compute the maximum of each of the 81 likelihoods and plot the value of $\hat{p}_w = (1-\hat{w})/2$.
 

### 2018-07-06 old/old_thing.tex
bounds on estimate arguments, new region of inconsistency

--- 2017-11-14 old_thing.tex
@@
 We now show that, in the case of the Farris tree topology, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
+
+\begin{lemma}
+For $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ and $t=\{x,y,x,y,w\}$,
+$$
+\ell_{\tau_1,t^*}(\tau_1, \tilde{t}) = \ell_{\tau_1,t^*}(\tau_1, t).
+$$
+\end{lemma}
+
+Since exchanging $x_1$ and $x_2$ does not change the value of the log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, $x_1=x_2$ at the maximum.
+The analogous statement holds for $y_1$ and $y_2$.
+The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
+
 Using the Hadamard transform approach, we calculate the generating probabilities on the Farris tree topology.
@@
 However, note that \cite{Siddall1998-hq} get things going the same way, although \cite{Swofford2001-hr} show that the problem is that they didn't simulate long enough sequences.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{lemma}
+\label{lem:concavity}
+Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, $t=\{x_1, y_1, x_2, y_2, w\}$, and $\tau^*$ be the InvFels tree.
+For a fixed $\{\xi_1,\ldots,\xi_q\}$, the objective function
+\begin{align*}
+\ell_{\tau^*,t^*}(\tau^*, t) = \sum_{j=1}^\nSiteSplits &\left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t) \right.\\
+&\qquad+ \left.\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau^*, t)\right]
+\end{align*}
+is concave and
+\[
+\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau^*, t)
+\]
+is a unique maximum.
+\end{lemma}
+
+\begin{proof}
+We focus on the two terms separately, showing they are both concave and, thus, their sum is concave.
+Call the second term
+\[
+\tilde{\ell}_{\tau^*,t^*}(\tau^*, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau^*, t).
+\]
+For fixed $\xi_j$, the partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
+\begin{align*}
+\tilde{\ell}_{\tau^*,t^*}(\tau^*, t) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_2}(x_2)\\
+&\qquad + \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,w}(w).
+\end{align*}
+Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
+Without loss of generality we focus on the variable $x_1$ and the partial log-likelihood proportional to
+\[
+\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1).
+\]
+Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, is
+\[
+h_{j,x_1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
+\]
+for $e_j\in\{0,1\}$, which, simplifying, results in
+\begin{align}
+\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
+                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1). \label{eq:partial-x-form}
+\end{align}
+We have the second derivative
+\[
+\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
+\]
+As $x_1\in(0,1)$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
+Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
+Applying similar arguments to the other variables, the Hessian for the partial log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero.
+
+We now focus on the divergence term.
+By Gibbs inequality, this function attains a unique maximum when
+\[
+\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
+\]
+for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
+Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the divergence, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
+Referring to Table~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
+The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
+\[
+2-2(x^*)^2 = 2-2x^2
+\]
+and for $\{2\}$ and $\{1,2\}$ yields
+\[
+2-2(y^*)^2 = 2-2y^2.
+\]
+implying $x=x^*$ and $y=y^*$.
+Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
+\[
+8x^*(y^*)^2 = 8x^*y^*w
+\]
+implying $w=y^*$.
+The unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$.%, and thus the Hessian of $\shannonDivergence$ is negative definite.
+...
+\textbf{
+The Hessian as a function of $\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t)$ is negative definite, but is it as a function of $t$?
+This seems intuitively obvious, but I don't know how to show it.
+How could it not be true?
+}
+%Let D(p, q) = \sum_j p_j \log q_j
+%This is concave in q_j, trivially, but we are interested in concavity in t.
+%Because t generates q_j, how can this not be true?
+%Thought:
+%- if f(x) has a unique maximum on [a,b] and g(x) is concave on [a,b], all we want to show is that f(x)+g(x) has a unique maximum on [a,b]; moreover, f(x) is concave in an open ball around its maximum.
+%- note that we have D(p, q(\theta)), where q(\theta) is this sort of complicated function involving Hadamard matrices.
+however, in the trivial case, D(p, q(\theta)) is concave, so we may be able to use induction?
+%- note f(x) + sum_h g(x,h) has a unique solution; what about f(x) + max_h g(x,h)? we still need concavity...
+%- the sum of divergence and partial likelihood may be easy to show is concave as partial likelihood clearly is.
+%- empirical results show that the gradient condition is accurate, though that might mean we need to show there are no zero crossings of the derivative or something like that
+%Oblique strategies:
+%- the condition of y_2 \ge y_1 and x_2 \ge x_1 may play a role
+%- we could also find x,y where derivative has no solutions, so fn is always increasing, though this might require some more argument
+%- we could show (if it's true) that going from fidelities to the Hadamard representation results in a concave function in t, and from there we know the divergence is concave in t.
+%   this seems hard but may be our best bet; proceed by induction seems reasonable, but, again, hard.
+%   it's true in the "trivial" n=0 case, which encourages me.
+%   moreover, the fidelities generate the probabilities, and the objective is concave in the probabilities, which is more encouragement.
+\end{proof}
+
+In words, this means that, if we have a single maximal ancestral state for each site split, we have a unique maximum.
+
+\begin{lemma}
+Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$ and $\emptyset$ be the maximal ancestral state for all site splits.
+There exists a set of $0 < x^*, y^* < 1$ such that $\{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2, 1, 1\}$ is the solution to maximizing \eqref{eq:profile_likelihood}.
+\end{lemma}
+
+\begin{proof}
+Assume $w=1$ and $y_2=1$.
+Then, there is a closed form solution maximizing the joint optimization procedure, as follows.
+We simplify the generating probabilities as in Table~\ref{tab:gen-sitepatprob-restricted}.
+We see that, on simplifying, for each site split the generating probability is the same functional form as the likelihood.
+Thus, the joint optimization procedure has a unique, closed form solution solving
+\begin{align*}
+&\arg\max_{x_1,y_y,x_2} \ (p_{\emptyset} + p_2 + p_3 + p_{23}) \log(1+x_1) + (p_1 + p_{123} + p_{12} + p_{13}) \log(1-x_1)\\
+               &\qquad + (p_{\emptyset} + p_1 + p_3 + p_{13}) \log(1+y_1) + (p_2 + p_{123} + p_{12} + p_{23}) \log(1-y_1)\\
+               &\qquad + (p_{\emptyset} + p_1 + p_2 + p_{12}) \log(1+x_2) + (p_3 + p_{123} + p_{13} + p_{23}) \log(1-x_2)\\
+&\propto\arg\max_{x_1,y_y,x_2} \ (4+4x*(y^*)^2) \log(1+x_1) + (4-4x*(y^*)^2) \log(1-x_1)\\
+               &\qquad + (4+4(y^*)^2) \log(1+y_1) + (4-4(y^*)^2) \log(1-y_1)\\
+               &\qquad + (4+4x*(y^*)^2) \log(1+x_2) + (4-4x*(y^*)^2) \log(1-x_2).
+\end{align*}
+We have estimates
+\[
+\{\hat{x}_1,\hat{y}_1,\hat{x}_2\} = \{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2\}
+\]
+and the gradient is equal to zero at this point.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{|l|l|l|}
+    \hline
+$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,1,1\})$\\
+    \hline
+    $\emptyset$ & $p_{\emptyset}$   &$(1+x_1)(1+y_1)(1+x_2)$\\
+    $\{1\}$     & $p_{1}$       &$(1-x_1)(1+y_1)(1+x_2)$\\
+    $\{2\}$     & $p_{2}$       &$(1+x_1)(1-y_1)(1+x_2)$\\
+    $\{3\}$     & $p_{3}$       &$(1+x_1)(1+y_1)(1-x_2)$\\
+    $\{1,2,3\}$ & $p_{123}$   &$(1-x_1)(1-y_1)(1-x_2)$\\
+    $\{1,2\}$   & $p_{12}$     &$(1-x_1)(1-y_1)(1+x_2)$\\
+    $\{1,3\}$   & $p_{13}$     &$(1-x_1)(1+y_1)(1-x_2)$\\
+    $\{2,3\}$   & $p_{23}$     &$(1+x_1)(1-y_1)(1-x_2)$\\
+    \hline
+\end{tabular}
+\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,1,1\}$ obtained using the Hadamard transform.
+All values multiplied by $1/8$.}
+\label{tab:gen-sitepatprob-restricted}
+\end{table}
+
+Now we show there are $0 < x^*, y^* < 1$ on which the gradient with respect to $w$ and $y_2$ is increasing on the boundary for all $x_1,y_1,x_2,y_2,w$.
+%TODO: is there a unique solution? it's not concave, but i think there may be a unique solution; need to show this... gradient argument?
+We use the fact that $x_1=x_2=x$ as there exists a unique solution and the likelihood is invariant to swapping these two variables.
+Let
+\[
+a_{\emptyset} = \frac{(2x)(y_1+y_2)}{(1+x^2)(1+y_1y_2)}
+\]
+with similar definitions for the remaining site splits.
+The objective function proportional to $w$ is
+\begin{align*}
+\ell(w) \propto \log(1+w) + \sum_j p_{\siteSplit_j} \log(1 + a_{\siteSplit_j}w)
+\end{align*}
+and has gradient with respect to $w$ as
+\begin{align*}
+\ell_w(t) = \frac{1}{1+w} + \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{1 + a_{\siteSplit_j}w}.
+\end{align*}
+Note that, $a_{\emptyset} = -a_{13}$, $a_{2} = -a_{123}$, $p_{2} = p_{123}$, and
+\[
+a_{1} = a_{3} = a_{12} = a_{23} = 0
+\]
+so
+\begin{align*}
+\ell_w(t) = \frac{1}{1+w} + \frac{p_{\emptyset}a_{\emptyset}}{1 + a_{\emptyset}w} - \frac{p_{13}a_{\emptyset}}{1 - a_{\emptyset}w} + \frac{p_{2}a_{2}}{1 + a_{2}w} - \frac{p_{2}a_{2}}{1 - a_{2}w}.
+\end{align*}
+
+For $y_2$, we do the same, but now with
+\begin{align*}
+\ell(y_2) \propto \log(1+y_2) + \sum_j p_{\siteSplit_j} \log(1 + b_{\siteSplit_j}y_2)
+\end{align*}
+where
+\[
+b_{\emptyset} = \frac{y_1+x^2y_1+2wx}{1+x^2+2wxy_1}.
+\]
+The denominator is always positive, allowing the above simplification.
+%Since x_1=x_2 at the maximum we can use an argument like 1+x^2-2y_1wx > 1+x^2-2x = (1-x)^2 > 0. We'll need to ignore cases where x=0 (i.e., x^* or y^* are zero)
+In Fig.~\ref{fig:positive-gradients}, we see the region where the gradients of both $w$ and $y_2$ are increasing at $\hat{t}=\{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2, 1, 1\}$, and, thus, the objective function is increasing in all positive directions over $(y_2,w)$ on the boundary.
+Since, from Lemma~\ref{lem:concavity}, the objective function is concave on our domain, this must be the maximum.
+
+Ok, the feasible set for the solution is actuall not $[0,1]^5$ because of assumptions we've made.
+It's actually
+\begin{itemize}
+\item $x_1 \in [0,x_2]$
+\item $y_1 \in [0,y_2]$
+\item $x_2 \in [0,1]$
+\item $y_2 \in [y^*,1]$
+\item $w \in [y^*,1]$
+\end{itemize}
+Now, it's possible that our objective function, while not concave, is still decreasing in all directions on this set from the maximum.
+How can we show this?
+Also, do we want to prove the objective is not concave?
+We might be able to have better bounds, like
+\begin{itemize}
+\item $x_1=x_2 \in [0,x^*]$
+\item $y_1 \in [0,y^*]$
+\item $y_2 \in [y^*,1]$
+\item $w \in [y^*,1]$
+\end{itemize}
+Bound:
+\begin{align*}
+\ell_w(t) = \frac{1}{1+w} + \frac{p_{\emptyset}a_{\emptyset}}{1 + a_{\emptyset}w} - \frac{p_{13}a_{\emptyset}}{1 - a_{\emptyset}w} + \frac{p_{2}a_{2}}{1 + a_{2}w} - \frac{p_{2}a_{2}}{1 - a_{2}w}.
+\end{align*}
+
+\end{proof}
+
+\begin{figure}
+\centering
+\includegraphics[width=.95\textwidth]{positive-gradients-inkscape}
+\caption{
+Region where gradients w.r.t. $w$ and $y_2$ are increasing on the boundary.
+}
+\label{fig:positive-gradients}
+\end{figure}

### 2018-07-06 joint_inf_plot.py

@@ -82,16 +82,6 @@ def parse_args():
         help='plot analytic',
     )
     parser.add_argument(
-        '--optimum-conditions',
-        action="store_true",
-        help='plot conditions where we have y_2=1, w=1 at optimum',
-    )
-    parser.add_argument(
-        '--empirical-inconsistency',
-        action="store_true",
-        help='compute empirical',
-    )
-    parser.add_argument(
         '--empirical-parameter-estimate',
         action="store_true",
         help='compute empirical',
@@ -113,6 +103,10 @@ def parse_args():
         help='file name for where to save pkl',
         default='output.pkl',
     )
+    parser.add_argument(
+        '--analytic-inconsistency',
+        action="store_true",
+    )
 
     args = parser.parse_args()
 
@@ -121,17 +115,6 @@ def parse_args():
 # ~~~~~~~~~
 # Functions for exact likelihood bounds
 
-
-def A_INVFELS(theta0, site_pattern):
-    theta = [-t if s=='1' else t \
-            for t,s in zip(theta0, site_pattern+'0')]
-    return ((theta[0]+theta[2])*(theta[1]+theta[3])) / ((1+theta[0]*theta[2])*(1+theta[1]*theta[3]))
-
-def B_INVFELS(theta0, site_pattern):
-    theta = [-t if s=='1' else t \
-            for t,s in zip(theta0, site_pattern+'0')]
-    return (theta[1] + theta[4]*theta[0] + theta[4]*theta[2] + theta[0]*theta[1]*theta[2]) / \
-           (1 + theta[0]*theta[2] + theta[0]*theta[1]*theta[4] + theta[1]*theta[2]*theta[4])
 
 def P_INVFELS(theta0, site_pattern):
     """
@@ -272,48 +255,32 @@ def get_ancestral_state_conds(xy):
         output.append(likelihood_upper_bound(theta, anc_state))
     return bnd - max(output)
 
-def get_empirical_inconsistency(xy):
+def get_analytic_inconsistency(xy):
     """
-    Plot region where likelihood of w=1 and y_2=1 is greater than or equal to all other maximized likelihoods
     """
-    # if we're in an edge case, return .5 to gray out plot
-    if xy[0] == 1. or xy[0] == 0. or xy[1] == 1. or xy[1] == 0.:
+    x = xy[0]
+    y = xy[1]
[…]

### 2018-07-09 appendix.tex
update proof of theorem

@@
 \hline
-$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
+$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
 \hline
@@
 \caption{
-The maximal likelihood values given $w=1$ and $y_2=1$.
+The maximal likelihood values given $\emptyset$ is the most likely ancestral state.
 All values multiplied by $1/32$.}
@@
 \]
-By some calculus, the gradient with respect to $x_1$ is
+By some calculus, the partial derivative with respect to $x_1$ is
 \[
@@
 \[
-\ell'_{L}(w) = \frac{1}{1+w} + - \frac{p_{13}}{c_x^* - w} - \frac{p_{2}}{c_x^*c_y^* - w}
+\ell'_{L}(w) = \frac{1}{1+w} - \frac{p_{13}}{c_x^* - w} - \frac{p_{2}}{c_x^*c_y^* - w}
 \]
@@
 \]
-The solution $\hat{w}$ solving $\ell'(w) = 0$ is greater than $\hat{w}'$ solving $\ell'_{L}(w) = 0$.
-As $c_x^* > 1$, provided $c_x^*c_y^* > 1$, $\ell'_{L}(w)$ stays finite and, when considering $\ell'_{L}(w) \lessgtr 0$, we equivalently consider $f(w) \lessgtr 0$ where $f$ is the quadratic function
-\begin{align*}
-f(w) &= (w)^2\cdot(1+p_{13}+p_{2}) \\
-      &\qquad + w\cdot(-c_x^*-c_x^*c_y^*-p_{13}c_x^*c_y^*+p_{13}-p_{2}c_x^*+p_{2}) \\
-      &\qquad + ((c_x^*)^2c_y^*-p_{13}c_x^*c_y^*-p_{2}c_x^*).
-\end{align*}
+Clearly $c_x^* > 1$.
 We see
@@
 \]
-provided again that $c_x^*c_y^* > 1$.
-Now this quadratic has either zero, one, or two solutions.
-If there are zero solutions, the derivative with respect to $w$ is increasing on all of $[0,1]$ so that $\hat{w} = 1$ and we are done.
-If there are one or two solutions,
-if the smaller of the zeros of $f(w)$ is greater than one, then $\hat{w} \equiv 1$.
-Using the quadratic formula with
+provided $c_x^*c_y^* > 1$.
+Note that $\ell'_{L}(w)$ is strictly decreasing in $w$ on the interval $(0,1)$.
+Thus, if $\ell'_{L}(1) > 0$ then $\ell(w)$ is increasing on the entire interval $(0,1)$, including at the boundary $w=1$, implying $\hat{w} \equiv 1$.
+So we want $x^*$ and $y^*$ where $c_x^*c_y^* > 1$ and $\ell_L'(1) > 0$.
+These two conditions are straightforward, the second one being
 \[
-a = 1+p_{13}+p_{2},
-\]
-\[
-b = -c_x^*-c_x^*c_y^*-p_{13}c_x^*c_y^*+p_{13}-p_{2}c_x^*+p_{2},
-\]
-\[
-c = (c_x^*)^2c_y^*-p_{13}c_x^*c_y^*-p_{2}c_x^*,
-\]
-the smaller zero is
-\[
-\hat{w} = \frac{-b - \sqrt{b^2 - 4ac}}{2a},
-\]
-which is a function of the generating parameters $x^*$ and $y^*$.
-We see that $a \ge 0$ and, provided that $2a+b \leq 0$, we have,
-\[
-\hat{w} \ge 1 \iff
-|2a + b| \leq \sqrt{b^2 - 4ac} \iff
-%-b - \sqrt{b^2 - 4ac} \le 2a,
-%0 \ge -b - 2a \le \sqrt{b^2 - 4ac},
-%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
-%a^2 + ab + ac \le 0,
-a + b + c \ge 0.
-\]
-Therefore, we have three conditions to satisfy for $\hat{w} \equiv 1$:
-\[
-\mathrm{C}1: c_x^*c_y^* > 1,
-\]
-\[
-\mathrm{C}2: 2-c_x^*c_y^*-c_x^*+p_{13}(3-c_x^*c_y^*)+p_{2}(3-c_x^*) \leq 0,
-\]
-\[
-\mathrm{C}3: 1+(c_x^*)^2c_y^*-c_x^*c_y^*-c_x^*+p_{13}(2-2c_x^*c_y^*)+p_{2}(2-2c_x^*) \ge 0.
+\ell'_L(1) = \frac{1}{2} - \frac{p_{13}}{c_x^*-1} - \frac{p_{2}}{c_x^*c_y^*-1} > 0.
 \]

### 2018-07-09 joint_inf_plot.py

@@ -268,16 +268,15 @@ def get_analytic_inconsistency(xy):
     cxcy = cx*cy
     p13 = P_INVFELS(theta, '1010')
     p2 = P_INVFELS(theta, '0100')
-    cond1 = cxcy >= 1
-    cond2 = 2 - cxcy - cx + p13*(3-cxcy) + p2*(3-cx) <= 0
-    cond3 = 1 + cx*cxcy - cxcy - cx + p13*(2-2*cxcy) + p2*(2-2*cx) >= 0
+    cond1 = cxcy > 1
+    cond2 = .5 - p13 / (cx - 1.) - p2 / (cxcy - 1.) >= 0
     bnd = likelihood_lower_bound(theta)
     output = []
     for anc_state in product('012', repeat=4):
         if ''.join(anc_state) == '0000':
             continue
         output.append(likelihood_upper_bound(theta, anc_state))
-    if bnd >= max(output) and cond1 and cond2 and cond3:
+    if bnd >= max(output) and cond1 and cond2:
         return -1.
     else:
         return 1.

### 2018-07-10 appendix.tex
remove x_1=x_2 condition

@@
 %TODO: make this more explicit/clear
-We use the fact that $x_1=x_2=x$ as there exists a unique solution and the likelihood is invariant to swapping these two variables.
 Let
 \[
-a_{\emptyset} = \frac{(2x)(y_1+y_2)}{(1+x^2)(1+y_1y_2)}
+a_{\emptyset} = \frac{(x_1+x_2)(y_1+y_2)}{(1+x_1x_2)(1+y_1y_2)}
 \]
@@
 \end{align*}
-Note that, $a_{\emptyset} = -a_{13}$, $a_{2} = -a_{123}$, $p_{2} = p_{123}$, and for $x_1=x_2$
-\[
-a_{1} = a_{3} = a_{12} = a_{23} = 0
-\]
+Note the following:
+\begin{itemize}
+\item $a_{\emptyset} = -a_{13}$, $p_{\emptyset} \ge p_{13}$;
+\item $a_{1} = -a_{3}$, $p_{1} = p_{3}$;
+\item $a_{2} = -a_{123}$, $p_{2} = p_{123}$;
+\item $a_{12} = -a_{23}$, $p_{12} = p_{23}$;
+\end{itemize}
 so
 \begin{align*}
-\ell'(w) = \frac{1}{1+w} + \frac{p_{\emptyset}a_{\emptyset}}{1 + a_{\emptyset}w} - \frac{p_{13}a_{\emptyset}}{1 - a_{\emptyset}w} + \frac{p_{2}a_{2}}{1 + a_{2}w} - \frac{p_{2}a_{2}}{1 - a_{2}w}.
+\ell'(w) &= \frac{1}{1+w} + \frac{p_{\emptyset}a_{\emptyset}}{1 + a_{\emptyset}w} - \frac{p_{13}a_{\emptyset}}{1 - a_{\emptyset}w} + \frac{p_{2}a_{2}}{1 + a_{2}w} - \frac{p_{2}a_{2}}{1 - a_{2}w} \\
+         &\qquad \frac{p_{1}a_{1}}{1 + a_{1}w} - \frac{p_{1}a_{1}}{1 - a_{1}w} + \frac{p_{12}a_{12}}{1 + a_{12}w} - \frac{p_{12}a_{12}}{1 - a_{12}w} \\
 \end{align*}
-Using the bounds on $\hat{x}_1,\hat{y}_1,\hat{y}_1$, and $\hat{y}_2$ from Lemma~\ref{lemma:bounds}, we have
+Using the bounds on $\hat{x}_1,\hat{y}_1,\hat{y}_1$, and $\hat{y}_2$ from Lemma~\ref{lemma:bounds} and our assumption of $y_2 \ge y_1$ and $x_2 \ge x_1$, we have
 \[
@@
 \]
+\[
+0 \le a_{1} \le \frac{x^*}{1-(x^*)^2}
+\]
+\[
+0 \le a_{12} \le \frac{x^*}{1-(x^*)^2}\cdot\frac{1}{1-y^*},
+\]
 so that $\ell'(w)$ has lower bound
 \[
-\ell'_{L}(w) = \frac{1}{1+w} - \frac{p_{13}}{c_x^* - w} - \frac{p_{2}}{c_x^*c_y^* - w}
+\ell'_{L}(w) = \frac{1}{1+w} - \frac{p_{13}}{c_1 - w} - \frac{p_{2}}{c_1c_2 - w} - \frac{p_{1}}{c_3 - w} - \frac{p_{12}}{c_2c_3 - w}
 \]
@@
 \[
-c_x^* = \frac{1+(x^*)^2}{2x^*},
+c_1 = \frac{1+(x^*)^2}{2x^*},
 \]
 \[
-c_y^* = 1-y^*.
+c_2 = 1-y^*,
 \]
-Clearly $c_x^* > 1$.
+\[
+c_3 = \frac{1-(x^*)^2}{x^*}.
+\]
+Clearly $c_1 > 1$.
 We see
 \[
-\ell'_L(0) = 1 - \frac{p_{13}}{c_x^*} - \frac{p_{2}}{c_x^*c_y^*} > 0
+\ell'_{L}(0) = 1 - \frac{p_{13}}{c_1} - \frac{p_{2}}{c_1c_2} - \frac{p_{1}}{c_3} - \frac{p_{12}}{c_2c_3} > 0
 \]
-provided $c_x^*c_y^* > 1$.
+provided $c_1c_2, c_3, c_2c_3 > 1$
 Note that $\ell'_{L}(w)$ is strictly decreasing in $w$ on the interval $(0,1)$.
 Thus, if $\ell'_{L}(1) > 0$ then $\ell(w)$ is increasing on the entire interval $(0,1)$, including at the boundary $w=1$, implying $\hat{w} \equiv 1$.
-So we want $x^*$ and $y^*$ where $c_x^*c_y^* > 1$ and $\ell_L'(1) > 0$.
-These two conditions are straightforward, the second one being
+So we want $x^*$ and $y^*$ where $c_1c_2, c_3, c_2c_3 > 1$ and $\ell_L'(1) > 0$.
+These conditions are straightforward, the one on the derivative being
 \[
-\ell'_L(1) = \frac{1}{2} - \frac{p_{13}}{c_x^*-1} - \frac{p_{2}}{c_x^*c_y^*-1} > 0.
+\ell'_{L}(1) = \frac{1}{2} - \frac{p_{13}}{c_1-1} - \frac{p_{2}}{c_1c_2-1} - \frac{p_{1}}{c_3-1} - \frac{p_{12}}{c_2c_3-1} > 0.
 \]

### 2018-07-10 joint_inf_plot.py

@@ -263,20 +263,26 @@ def get_analytic_inconsistency(xy):
     if x == 1. or x == 0. or y == 1. or y == 0.:
         return np.nan
     theta = [x, y, x, y, y]
-    cx = (1+x*x) / (2.*x)
-    cy = (1-y)
-    cxcy = cx*cy
+    c1 = (1+x*x) / (2.*x)
+    c2 = (1-y)
+    c3 = (1-x*x) / x
+    c1c2 = c1*c2
+    c2c3 = c2*c3
     p13 = P_INVFELS(theta, '1010')
     p2 = P_INVFELS(theta, '0100')
-    cond1 = cxcy > 1
-    cond2 = .5 - p13 / (cx - 1.) - p2 / (cxcy - 1.) >= 0
+    p1 = P_INVFELS(theta, '0010')
+    p12 = P_INVFELS(theta, '0110')
+    cond1 = c1c2 > 1
+    cond2 = c2c3 > 1
+    cond3 = c3 > 1
+    cond4 = .5 - p13 / (c1 - 1.) - p2 / (c1c2 - 1.) - p1 / (c3 - 1.) - p12 / (c2c3 - 1.) >= 0
     bnd = likelihood_lower_bound(theta)
     output = []
     for anc_state in product('012', repeat=4):
         if ''.join(anc_state) == '0000':
             continue
         output.append(likelihood_upper_bound(theta, anc_state))
-    if bnd >= max(output) and cond1 and cond2:
+    if bnd >= max(output) and cond1 and cond2 and cond3 and cond4:
         return -1.
     else:
         return 1.

### 2018-07-11 joint_inf.tex
bounds on estimate arguments, new region of inconsistency
pass thru main text

@@
 In this paper, we show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
-%TODO: change this line? it's still like an inconsistency in topology
-We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates an incorrect topology in the limit of infinite-length sequences.
-For branch length estimation on the correct topology, we outline similar cases where branch length estimates are consistently and heavily biased.
+We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates an incorrect topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
+For branch length estimation on the correct topology, we outline similar cases where branch length estimates are consistently and heavily biased and provide extensive empirical results for this setting.
 \end{abstract}
@@
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model with data being generated on the four-taxon ``inverse Felsenstein'' tree (a.k.a.\ the ``Farris'' tree) and we construct bounds on the joint objective function to demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
-We find similar areas where joint inference consistently underestimates interior branch lengths when the topology is known and fixed.
+To do so, we use a binary symmetric model with data being generated on the four-taxon ``inverse Felsenstein'' tree (a.k.a.\ the ``Farris'' tree) and we construct bounds on the joint objective function to demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
+We find similar areas through empirical means where joint inference consistently underestimates interior branch lengths when the topology is known and fixed.
 
@@
 
-We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``Inverse Felsenstein (InvFels)'' tree and the ``Felsenstein'' tree (Fig.~\ref{fig:farris-fels-top}).
+We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``inverse Felsenstein (InvFels)'' tree (Fig.~\ref{fig:farris-fels-top}).
 The ``inverse Felsenstein'' terminology comes from \citet{Swofford2001-hr}, although it is also called the ``Farris'' tree \citep{Siddall1998-hq, Felsenstein2004}.
-In the standard configuration of each of these trees, the interior branch length parameters are equal to the bottom two parameters.
-We show in the Appendix that, in the case of the InvFels tree, performing inference fixing both the top two branch parameters to be equal and the bottom two branch parameters to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch parameters.
+In the standard configuration of this tree, the interior branch length parameters are equal to the bottom two parameters.
+We use this standard configuration as our data generating process, though we do not constrain our branch lengths to be equal in considering our optimization objective function.
 
+%TODO: update this figure
 \begin{figure}
@@
 \end{subfigure}
-\caption{Two four-taxon trees with fidelities as labeled: $\theta_1=\theta_3=x$, $\theta_2=\theta_4=y$, and $\theta_5=w$.
+\caption{Two four-taxon trees with fidelities as labeled.
 }
@@
 For a four-taxon tree, define the general branch fidelity parameter $t=\{\theta_1,\theta_2,\theta_3,\theta_4,\theta_5\}$ where fidelities are ordered in the order of the taxa with the internal branch last (Fig.~\ref{fig:farris-fels-top}).
-We show in the Appendix that, for the InvFels tree, the maximum likelihood fidelities require $\theta_1=\theta_3$ and $\theta_2=\theta_4$, so we subsequently fix fidelity parameters to $t=\{x,y,x,y,w\}$.
-
 Although we use fidelities exclusively for our theoretical development, we have made our figures in terms of probabilities of substitution $p_{\theta} = (1-\theta)/2$ as they are easier to interpret.
@@
 In words, this implies an inconsistency as we converge on the incorrect topology, estimating the interior branch length to be zero.
-The proof of this theorem is by a detailed examination of inequalities.
+For example, the Felsenstein tree $\tau_2$ will achieve the same objective function value in this case as the InvFels tree $\tau_1$, as if the interior branch is zero-length these two topologies are indistinguishable.
+%TODO: more discussion on proof, intution
 %Intuitively, $\tau_2$ is favored in performing joint inference since the objective function for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its objective function while $\tau_2$ has many more.
@@
 
-%TODO: change legend of this figure---the intersection is just the region where \emptyset is maximal, so it will be the same plot
 \begin{figure}
@@
 This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split (Table~\ref{tab:likelihoods} in the Appendix with $x_1=x_2=x^*$ and $y_1=y_2=y^*$).
-If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.
+If we allow multifurcating trees in our inference, then we can think of this as an instance of converging to the wrong topology.
 
@@
 To determine how conservative, we use the well-known L-BFGS-B method to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
-We see that the region of inconsistency in the general case is much larger than in the analytic case---the analytic case required $\hat{y}_2 = 1$ and dealt with loose bounds, so this is reasonable.
+We see that the region of inconsistency in the general case is much larger than in the analytic case---the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral states, so this is reasonable.
 This region encompasses almost half of the branch parameter space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch length to be zero.
 
-We provide a full description of our optimization procedure in the Appendix, but briefly, we perform 81 maximizations for each separate likelihood function.
-For each function, we optimize the objective starting at five separate initial values to mitigate problems with small curvature of the objective surface.
-We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$.
-We do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
+We provide a full description of our optimization procedure in the Appendix, but briefly, we perform 81 maximizations for each separate likelihood function in Table~\ref{table:}.
+For each function, we optimize the objective starting at five separate initial values to mitigate problems with small curvature of the objective surface and nonconcavity.
+We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $0$ to $1$.
 Our optimization code can be found at \url{https://github.com/matsengrp/joint-inf/}.
@@
 As data were generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
-Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same procedure.
+Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same optimization code.
 For joint inference, the estimates for $\hat{p}_w$ when $p_{x^*}$ and $p_{y^*}$ are both small look reasonable, with $\hat{p}_w$ increasing as $p_{y^*}$ increases, though Fig.~\ref{fig:empirical-bias} shows there is a systematic negative bias in this procedure even when the true branches are short.
@@
 Due to the discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}), an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization.
-For joint inference, the errors range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for marginal inference, $[-3\times 10^{-17}, 3\times 10^{-17}]$ showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
+For joint inference, the errors given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for marginal inference, $[-3\times 10^{-17}, 3\times 10^{-17}]$ showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
 
@@
 \caption{
-    Numerical estimates for $\hat{p}_w=(1-\hat{w})/2$ when computing $(\hat{x}, \hat{y}, \hat{w})$ optimizing \eqref{eq:profile_likelihood}.
+    Numerical estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}.
     Data generated as in Fig.~\ref{fig:incons-analytic}
@@
 These regions of inconsistency arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
-This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and 3.
+This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and~3.
 \citet{Sagulenko2018-xl} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:empirical-bias} shows even when branches are short there is a consistent bias.
@@
 
-Previous work in phylogenetics has developed consistency counterexamples using the same four-taxon topologies used here \citep{Felsenstein1978-rr}.
+Previous work in phylogenetics has developed consistency counterexamples using similar four-taxon topologies to the one used here \citep{Felsenstein1978-rr}.
 In this previous work, when simulating data under the Felsenstein topology $\tau_2$, as the number of observations increases, the InvFels topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
@@
 This is not the case here; while analytically the inconsistency occurs in the long branch attraction region of the parameter space, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
+%TODO: add plot or at least discussion of intuition here
 Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.
@@
 
-While we have shown inconsistency in both topology and branch parameter estimation, there is substantial scope for future work to make these results more precise and more general.
+While we have shown inconsistency in topology estimation and investigated the performance of branch parameter estimation, there is substantial scope for future work to make these results more precise and more general.
 The techniques used to obtain upper and lower bounds for the likelihoods in the topology estimation case provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
-Similarly, for the general case in estimating branch lengths, we were only able to provide a conservative region of overestimation, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) begs further investigation.
+Similarly, for the general case in estimating branch lengths, we were only able to provide empirical estimates for regions of overestimation, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) beg further investigation.
 Empirical validation shows that the region of inconsistency is much larger than the one analytically derived.

### 2018-07-11 appendix.tex
change notation in plotting code; gradient argument for lemma 3

@@
 \]
+We now show the gradient of the divergence term is a decreasing function on $(0,1)$ in each variable.
+We focus on $x_1$.
+The divergence proportional to $x_1$ can be written as
+\begin{align*}
+\shannonDivergence(x_1) \propto \sum_j p_{\siteSplit_j} \log(c_{\siteSplit_j} + a_{\siteSplit_j}x_1)
+\end{align*}
+where $c_{\siteSplit_j}$ and $a_{\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
+For example,
+\[
+c_{\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
+\]
+and
+\[
+a_{\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
+\]
+with similar definitions for the remaining site splits.
+Consider the derivative with respect to $x_1$
+\begin{align*}
+\shannonDivergence'(x_1) := \frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}
+\end{align*}
+and focus on a single term
+\[
+f_j(x_1) = p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}.
+\]
+Regardless of $a_{\siteSplit_j} \lessgtr 0$, this is a decreasing function of $x_1$ provided $c_{\siteSplit_j} + a_{\siteSplit_j}x_1 > 0$, but we see
+\[
+p_{\siteSplit_j} := \frac{1}{8} (c_{\siteSplit_j} + a_{\siteSplit_j}x_1)
+\]
+and $p_{\siteSplit_j} > 0$ so long as $x^*\in(0,1)$ and $y^*\in(0,1)$.
 
+We now focus on the partial likelihood term.
 Assume $\emptyset$ is the most likely ancestral state for all site splits.
@@
 
-We focus on $x_1$, similar arguments applying to the other variables.
 The partial likelihood
@@
 %TODO: make this more explicit/clear
-Let
+Similar to the previous proof, let
 \[
-a_{\emptyset} = \frac{(x_1+x_2)(y_1+y_2)}{(1+x_1x_2)(1+y_1y_2)}
+b_{\emptyset} = \frac{(x_1+x_2)(y_1+y_2)}{(1+x_1x_2)(1+y_1y_2)}
 \]
@@
 \begin{align*}
-\ell(w) \propto \log(1+w) + \sum_j p_{\siteSplit_j} \log(1 + a_{\siteSplit_j}w)
+\ell(w) \propto \log(1+w) + \sum_j p_{\siteSplit_j} \log(1 + b_{\siteSplit_j}w)
 \end{align*}
@@
 \begin{align*}
-\ell'(w) := \frac{d}{dw} \ell(w) = \frac{1}{1+w} + \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{1 + a_{\siteSplit_j}w}.
+\ell'(w) := \frac{d}{dw} \ell(w) = \frac{1}{1+w} + \sum_j p_{\siteSplit_j}b_{\siteSplit_j}\frac{1}{1 + b_{\siteSplit_j}w}.
 \end{align*}
@@
 \begin{itemize}
-\item $a_{\emptyset} = -a_{13}$, $p_{\emptyset} \ge p_{13}$;
-\item $a_{1} = -a_{3}$, $p_{1} = p_{3}$;
-\item $a_{2} = -a_{123}$, $p_{2} = p_{123}$;
-\item $a_{12} = -a_{23}$, $p_{12} = p_{23}$;
+\item $b_{\emptyset} = -b_{13}$, $p_{\emptyset} \ge p_{13}$;
+\item $b_{1} = -b_{3}$, $p_{1} = p_{3}$;
+\item $b_{2} = -b_{123}$, $p_{2} = p_{123}$;
+\item $b_{12} = -b_{23}$, $p_{12} = p_{23}$;
 \end{itemize}
@@
 \begin{align*}
-\ell'(w) &= \frac{1}{1+w} + \frac{p_{\emptyset}a_{\emptyset}}{1 + a_{\emptyset}w} - \frac{p_{13}a_{\emptyset}}{1 - a_{\emptyset}w} + \frac{p_{2}a_{2}}{1 + a_{2}w} - \frac{p_{2}a_{2}}{1 - a_{2}w} \\
-         &\qquad \frac{p_{1}a_{1}}{1 + a_{1}w} - \frac{p_{1}a_{1}}{1 - a_{1}w} + \frac{p_{12}a_{12}}{1 + a_{12}w} - \frac{p_{12}a_{12}}{1 - a_{12}w} \\
+\ell'(w) &= \frac{1}{1+w} + \frac{p_{\emptyset}b_{\emptyset}}{1 + b_{\emptyset}w} - \frac{p_{13}b_{\emptyset}}{1 - b_{\emptyset}w} + \frac{p_{2}b_{2}}{1 + b_{2}w} - \frac{p_{2}b_{2}}{1 - b_{2}w} \\
+         &\qquad \frac{p_{1}b_{1}}{1 + b_{1}w} - \frac{p_{1}b_{1}}{1 - b_{1}w} + \frac{p_{12}b_{12}}{1 + b_{12}w} - \frac{p_{12}b_{12}}{1 - b_{12}w} \\
 \end{align*}
@@
 \[
-0 \le a_{\emptyset} \le \frac{2x^*}{1+(x^*)^2},
+0 \le b_{\emptyset} \le \frac{2x^*}{1+(x^*)^2} := b_{\emptyset}^U,
 \]
 \[
-0 \le a_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*},
+0 \le b_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*} := b_{2}^U,
 \]
 \[
-0 \le a_{1} \le \frac{x^*}{1-(x^*)^2}
+0 \le b_{1} \le \frac{x^*}{1-(x^*)^2} := b_{1}^U
 \]
 \[
-0 \le a_{12} \le \frac{x^*}{1-(x^*)^2}\cdot\frac{1}{1-y^*},
+0 \le b_{12} \le \frac{x^*}{1-(x^*)^2}\cdot\frac{1}{1-y^*} := b_{12}^U,
 \]
@@
 \[
-\ell'_{L}(w) = \frac{1}{1+w} - \frac{p_{13}}{c_1 - w} - \frac{p_{2}}{c_1c_2 - w} - \frac{p_{1}}{c_3 - w} - \frac{p_{12}}{c_2c_3 - w}
+\ell'_{L}(w) = \frac{1}{1+w} - \frac{p_{13}}{1/b_{\emptyset}^U - w} - \frac{p_{2}}{1/b_{2}^U - w} - \frac{p_{1}}{1/b_{1}^U - w} - \frac{p_{12}}{1/b_{12}^U - w}
 \]
-with
-\[
-c_1 = \frac{1+(x^*)^2}{2x^*},
-\]
-\[
-c_2 = 1-y^*,
-\]
-\[
-c_3 = \frac{1-(x^*)^2}{x^*}.
-\]
-Clearly $c_1 > 1$.
+Clearly $b_{\emptyset}^U < 1$.
 We see
 \[
-\ell'_{L}(0) = 1 - \frac{p_{13}}{c_1} - \frac{p_{2}}{c_1c_2} - \frac{p_{1}}{c_3} - \frac{p_{12}}{c_2c_3} > 0
+\ell'_{L}(0) = 1 - p_{13}b_{\emptyset}^U - p_{2}b_{2}^U - p_{1}b_{1}^U p_{12}b_{12}^U > 0
 \]
-provided $c_1c_2, c_3, c_2c_3 > 1$
-Note that $\ell'_{L}(w)$ is strictly decreasing in $w$ on the interval $(0,1)$.
+provided $b_{2}^U, b_{1}^U, b_{12}^U < 1$.
+Note that $\ell'_{L}(w)$ is strictly decreasing in $w$ on the interval $(0,1)$, as shown in the previous proof.
 Thus, if $\ell'_{L}(1) > 0$ then $\ell(w)$ is increasing on the entire interval $(0,1)$, including at the boundary $w=1$, implying $\hat{w} \equiv 1$.
-So we want $x^*$ and $y^*$ where $c_1c_2, c_3, c_2c_3 > 1$ and $\ell_L'(1) > 0$.
+So we want $x^*$ and $y^*$ where $b_{2}^U, b_{1}^U, b_{12}^U < 1$ and $\ell_L'(1) > 0$.
 These conditions are straightforward, the one on the derivative being
 \[
-\ell'_{L}(1) = \frac{1}{2} - \frac{p_{13}}{c_1-1} - \frac{p_{2}}{c_1c_2-1} - \frac{p_{1}}{c_3-1} - \frac{p_{12}}{c_2c_3-1} > 0.
+\ell'_{L}(1) = \frac{1}{2} - \frac{p_{13}b_{\emptyset}^U}{1-b_{\emptyset}^U} - \frac{p_{2}b_{2}^U}{1-b_{2}^U} - \frac{p_{1}b_{1}^U}{1-b_{1}^U} - \frac{p_{12}b_{12}^U}{1-b_{12}^U} > 0.
 \]

### 2018-07-11 joint_inf_plot.py

@@ -263,19 +263,18 @@ def get_analytic_inconsistency(xy):
     if x == 1. or x == 0. or y == 1. or y == 0.:
         return np.nan
     theta = [x, y, x, y, y]
-    c1 = (1+x*x) / (2.*x)
-    c2 = (1-y)
-    c3 = (1-x*x) / x
-    c1c2 = c1*c2
-    c2c3 = c2*c3
+    b0u = (2.*x) / (1+x*x)
+    b2u = (2.*x) / ((1+x*x) * (1-y))
+    b1u = x / (1-x*x)
+    b12u = x / ((1-x*x) * (1-y))
     p13 = P_INVFELS(theta, '1010')
     p2 = P_INVFELS(theta, '0100')
     p1 = P_INVFELS(theta, '0010')
     p12 = P_INVFELS(theta, '0110')
-    cond1 = c1c2 > 1
-    cond2 = c2c3 > 1
-    cond3 = c3 > 1
-    cond4 = .5 - p13 / (c1 - 1.) - p2 / (c1c2 - 1.) - p1 / (c3 - 1.) - p12 / (c2c3 - 1.) >= 0
+    cond1 = b2u < 1
+    cond2 = b1u < 1
+    cond3 = b12u < 1
+    cond4 = .5 - p13*b0u / (1. - b0u) - p2*b2u / (1. - b2u) - p1*b1u / (1. - b1u) - p12*b12u / (1. - b12u) > 0
     bnd = likelihood_lower_bound(theta)
     output = []
     for anc_state in product('012', repeat=4):

### 2018-07-11 appendix.tex
oops; forgot commit
a little rearranging and adding assumption

@@
 Likelihood computations for all site splits and ancestral state splits are in Table~\ref{tab:farris_likelihoods} for the InvFels tree.
-Taking maxima over ancestral state splits results in Table~\ref{tab:likelihoods}.
 
@@
 
-\begin{table}
-\centering
-\begin{tabular}{|ll|l|}
-\hline
-$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
-\hline
-$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
-
-$\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
-
-$\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
-
-$\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{1,2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
-
-$\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{1,2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
-
-$\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
-
-$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
-
-$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
-\hline
-\end{tabular}
-\caption{
-Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral states to reduce the number of possible functional forms to consider.
-All values multiplied by $1/32$.
-Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
-There are $3^4=81$ possible forms for the likelihood.
-}
-\label{tab:likelihoods}
-\end{table}
 %TODO: the caption of this overlaps with the page number---will this be a problem or will MBE handle it?
@@
 The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
-Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood then $x_1=x_2$ at the maximum.
-An analogous statement holds for $y_1$ and $y_2$.
+Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$ we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$.
+An analogous statement holds for $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral states.
+
+\begin{table}
+\centering
+\begin{tabular}{|ll|l|}
+\hline
+$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
+\hline
+$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
+
+$\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
+
+$\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
+
+$\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
+&$\{1,2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
+
+$\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
+&$\{1,2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
+
+$\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
+
+$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
+&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
+
+$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
+&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
+\hline
+\end{tabular}
+\caption{
+Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral states to reduce the number of possible functional forms to consider.
+All values multiplied by $1/32$.
+Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
+There are $3^4=81$ possible forms for the likelihood.
+}
+\label{tab:likelihoods}
+\end{table}
 
@@
 Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$ and $\emptyset$ be the maximal ancestral state for all site splits.
+Assume $0 < x^*, y^* < 1$.
 The solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
@@
 and $p_{\siteSplit_j} > 0$ so long as $x^*\in(0,1)$ and $y^*\in(0,1)$.
+The remaining variables follow similar arguments.
 
@@
 \ell'(w) &= \frac{1}{1+w} + \frac{p_{\emptyset}b_{\emptyset}}{1 + b_{\emptyset}w} - \frac{p_{13}b_{\emptyset}}{1 - b_{\emptyset}w} + \frac{p_{2}b_{2}}{1 + b_{2}w} - \frac{p_{2}b_{2}}{1 - b_{2}w} \\
-         &\qquad \frac{p_{1}b_{1}}{1 + b_{1}w} - \frac{p_{1}b_{1}}{1 - b_{1}w} + \frac{p_{12}b_{12}}{1 + b_{12}w} - \frac{p_{12}b_{12}}{1 - b_{12}w} \\
+         &\qquad \frac{p_{1}b_{1}}{1 + b_{1}w} - \frac{p_{1}b_{1}}{1 - b_{1}w} + \frac{p_{12}b_{12}}{1 + b_{12}w} - \frac{p_{12}b_{12}}{1 - b_{12}w}.\\
 \end{align*}
@@
 Clearly $b_{\emptyset}^U < 1$.
-We see
-\[
-\ell'_{L}(0) = 1 - p_{13}b_{\emptyset}^U - p_{2}b_{2}^U - p_{1}b_{1}^U p_{12}b_{12}^U > 0
-\]
-provided $b_{2}^U, b_{1}^U, b_{12}^U < 1$.
-Note that $\ell'_{L}(w)$ is strictly decreasing in $w$ on the interval $(0,1)$, as shown in the previous proof.
+Note that $\ell'_{L}(w)$ is strictly decreasing in $w$ on the interval $(0,1)$, as shown in the previous proof, provided $b_{2}^U, b_{1}^U, b_{12}^U < 1$.
 Thus, if $\ell'_{L}(1) > 0$ then $\ell(w)$ is increasing on the entire interval $(0,1)$, including at the boundary $w=1$, implying $\hat{w} \equiv 1$.

### 2018-07-11 joint_inf.tex
undefined ref

@@
 
-We provide a full description of our optimization procedure in the Appendix, but briefly, we perform 81 maximizations for each separate likelihood function in Table~\ref{table:}.
+We provide a full description of our optimization procedure in the Appendix, but briefly, we perform 81 maximizations for each separate likelihood function in Table~\ref{table:likelihoods}.
 For each function, we optimize the objective starting at five separate initial values to mitigate problems with small curvature of the objective surface and nonconcavity.

### 2018-07-11 joint_inf.tex
fix table ref

--- 2018-07-11 joint_inf.tex
@@
 
-We provide a full description of our optimization procedure in the Appendix, but briefly, we perform 81 maximizations for each separate likelihood function in Table~\ref{table:}.
+We provide a full description of our optimization procedure in the Appendix, but briefly, we perform 81 maximizations for each separate likelihood function in Table~\ref{tab:likelihoods}.
 For each function, we optimize the objective starting at five separate initial values to mitigate problems with small curvature of the objective surface and nonconcavity.

### 2018-07-16 appendix.tex
rearranging; cleaning TODOs; felsenstein argument

@@
 for any site split $\siteSplit_j$.
-Table~\ref{tab:gen-sitepatprob} contains calculations of site split probabilities for the InvFels topology with arbitrary branch lengths and Table~\ref{tab:sitepatprob} site split probabilities for the InvFels topology in Fig.~\ref{fig:farris-fels-top}a.
+Table~\ref{tab:gen-sitepatprob} contains calculations of site split probabilities for the trees in Fig.~\ref{fig:farris-fels-top}.
 
@@
 \begin{tabular}{|l|l|l|}
+    \multicolumn{3}{c}{InvFels tree $\tau=\tau^*$, $t^*=\{x^*,y^*,x^*,y^*,y^*\}$}\\
     \hline
-$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\})$\\
+$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau,t)$\\
+    \hline
+    $\emptyset$ & $p_{\emptyset}$   &$1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2$\\
+    $\{1\}$     & $p_{1}$           &$1-(x^*)^2+(y^*)^2-(x^*)^2(y^*)^2$\\
+    $\{2\}$     & $p_{2}$           &$1+(x^*)^2-(y^*)^2-(x^*)^2(y^*)^2$\\
+    $\{3\}$     & $p_{3}$           &$1-(x^*)^2+(y^*)^2-(x^*)^2(y^*)^2$\\
+    $\{1,2,3\}$ & $p_{123}$         &$1+(x^*)^2-(y^*)^2-(x^*)^2(y^*)^2$\\
+    $\{1,2\}$   & $p_{12}$          &$1-(x^*)^2-(y^*)^2+(x^*)^2(y^*)^2$\\
+    $\{2,3\}$   & $p_{23}$          &$1-(x^*)^2-(y^*)^2+(x^*)^2(y^*)^2$\\
+    $\{1,3\}$   & $p_{13}$          &$1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2$\\
+    \hline
+    \multicolumn{3}{c}{InvFels tree $\tau=\tau_1$, $t=\{x_1,y_1,x_2,y_2,w\}$}\\
+    \hline
+$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau,t)$\\
     \hline
@@
     $\{3\}$     & $p_{3}$           &$1 - x_1x_2 +  y_1y_2 +  w[  x_1 - x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
+    $\{1,2,3\}$ & $p_{123}$         &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][  y_1 - y_2] - x_1y_1x_2y_2$\\
     $\{1,2\}$   & $p_{12}$          &$1 - x_1x_2 -  y_1y_2 +  w[- x_1 + x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
+    $\{2,3\}$   & $p_{23}$          &$1 - x_1x_2 -  y_1y_2 +  w[  x_1 - x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
     $\{1,3\}$   & $p_{13}$          &$1 + x_1x_2 +  y_1y_2 +  w[- x_1 - x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
-    $\{2,3\}$   & $p_{23}$     &$1 - x_1x_2 -  y_1y_2 +  w[  x_1 - x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
-    $\{1,2,3\}$ & $p_{123}$   &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][  y_1 - y_2] - x_1y_1x_2y_2$\\
+    \hline
+    \multicolumn{3}{c}{Felsenstein tree $\tau=\tau_2$, $t=\{x_1,y_1,x_2,y_2,w\}$}\\
+    \hline
+$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau,t)$\\
+    \hline
+    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1y_1 +  x_2y_2 +  w[  x_1 + y_1][  x_2 + y_2] + x_1y_1x_2y_2$\\
+    $\{1\}$     & $p_{1}$           &$1 - x_1y_1 +  x_2y_2 +  w[ -x_1 + y_1][  x_2 + y_2] - x_1y_1x_2y_2$\\
+    $\{2\}$     & $p_{2}$           &$1 - x_1y_1 +  x_2y_2 +  w[  x_1 - y_1][  x_2 + y_2] - x_1y_1x_2y_2$\\
+    $\{3\}$     & $p_{3}$           &$1 + x_1y_1 -  x_2y_2 +  w[  x_1 + y_1][ -x_2 + y_2] - x_1y_1x_2y_2$\\
+    $\{1,2,3\}$ & $p_{123}$         &$1 + x_1y_1 -  x_2y_2 +  w[ -x_1 - y_1][ -x_2 + y_2] - x_1y_1x_2y_2$\\
+    $\{1,2\}$   & $p_{12}$          &$1 + x_1y_1 +  x_2y_2 +  w[ -x_1 - y_1][  x_2 + y_2] + x_1y_1x_2y_2$\\
+    $\{2,3\}$   & $p_{23}$          &$1 - x_1y_1 -  x_2y_2 +  w[  x_1 - y_1][ -x_2 + y_2] + x_1y_1x_2y_2$\\
+    $\{1,3\}$   & $p_{13}$          &$1 - x_1y_1 -  x_2y_2 +  w[ -x_1 + y_1][ -x_2 + y_2] + x_1y_1x_2y_2$\\
     \hline
 \end{tabular}
-\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
+\caption{Site split probabilities $p_{\siteSplit_j}$ on the true InvFels tree $\tau^*$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, and on the InvFels tree $\tau_1$ and Felsenstein tree $\tau_2$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
 All values multiplied by $1/8$.}
 \label{tab:gen-sitepatprob}
-\end{table}
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{|l|l|l|}
-    \hline
-$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x,y,x,y,y\})$\\
-    \hline
-    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xy^2+x^2y^2$\\
-    $\{1\}$     & $p_{1}$           &$1-x^2+y^2-x^2y^2$\\
-    $\{2\}$     & $p_{2}$           &$1+x^2-y^2-x^2y^2$\\
-    $\{3\}$     & $p_{3}$           &$1-x^2+y^2-x^2y^2$\\
-    $\{1,2,3\}$ & $p_{123}$         &$1+x^2-y^2-x^2y^2$\\
-    $\{1,2\}$   & $p_{12}$          &$1-x^2-y^2+x^2y^2$\\
-    $\{2,3\}$   & $p_{23}$          &$1-x^2-y^2+x^2y^2$\\
-    $\{1,3\}$   & $p_{13}$          &$1+x^2+y^2-4xy^2+x^2y^2$\\
-    \hline
-\end{tabular}
-\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x,y,x,y,y\}$ obtained using the Hadamard transform.
-All values multiplied by $1/8$.}
-\label{tab:sitepatprob}
 \end{table}
@@
 \centering
-\begin{tabular}{|ll|l|}
+\begin{tabular}{|l|ll|}
 \hline
@@
 $\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
-&$\{1\}^*$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$\\
-&$\{2\}^*$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$\\
-&$\{1,2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$\\
+&$\{1\}^*$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$             \\
+&$\{2\}^*$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$             \\
+&$\{1,2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$           \\
 
 $\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$\\
-&$\{2\}^*$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$\\
-&$\{1,2\}$&$(1+x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$\\
+&$\{1\}$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$               \\
+&$\{2\}^*$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$             \\
+&$\{1,2\}$&$(1+x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$             \\
 
 $\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
-&$\{1\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1-w)$\\
-&$\{2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1-w)$\\
-&$\{1,2\}$&$(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1+w)$\\
+&$\{1\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1-w)$             \\
+&$\{2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1-w)$               \\
+&$\{1,2\}$&$(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1+w)$             \\
 
 $\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{2\}^*$&$(1+x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$\\
-&$\{1,2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
+&$\{1\}$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$               \\
+&$\{2\}^*$&$(1+x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$             \\
+&$\{1,2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$             \\
 
 $\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{2\}^*$&$(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1-w)$\\
-&$\{1,2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
+&$\{1\}$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$               \\
+&$\{2\}^*$&$(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1-w)$             \\
+&$\{1,2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$             \\
 
 $\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1-w)$\\
-&$\{2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1-w)$\\
-&$\{1,2\}$&$(1+x_1)(1+y_1)(1-x_2)(1-y_2)(1+w)$\\
+&$\{1\}$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1-w)$               \\
+&$\{2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1-w)$               \\
+&$\{1,2\}$&$(1+x_1)(1+y_1)(1-x_2)(1-y_2)(1+w)$             \\
+
+$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$               \\
+&$\{2\}$&$(1+x_1)(1+y_1)(1-x_2)(1-y_2)(1-w)$               \\
+&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$             \\
 
 $\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$\\
-&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
-
-$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{2\}$&$(1+x_1)(1+y_1)(1-x_2)(1-y_2)(1-w)$\\
-&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
+&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$               \\
+&$\{2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$             \\
+&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$             \\
 \hline
@@
 
+$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
+&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
+&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
+
 $\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
@@
 &$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
-
-$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
 \hline
@@
 \]
-Figure~\ref{fig:max-anc-state} shows where the lower bound above is greater than all 80 other upper bounds.
+Fig.~\ref{fig:max-anc-state} shows where the lower bound above is greater than all 80 other upper bounds.
 \end{proof}
@@
 Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the divergence, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
-Referring to Table~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
+Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
 The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
@@
 \[
-p_{\siteSplit_j} := \frac{1}{8} (c_{\siteSplit_j} + a_{\siteSplit_j}x_1)
+\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) := \frac{1}{8} (c_{\siteSplit_j} + a_{\siteSplit_j}x_1)
 \]
-and $p_{\siteSplit_j} > 0$ so long as $x^*\in(0,1)$ and $y^*\in(0,1)$.
+which is strictly greater than zero provided $x_1,y_1,x_2,y_2\in(0,1)$.
 The remaining variables follow similar arguments.
@@
 $\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
+$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
 $\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
-$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
 \hline
@@
 \]
-as a function of $x_1$, using Table~\ref{tab:sitepatprob}, is
+as a function of $x_1$, using Table~\ref{tab:gen-sitepatprob}, is
 \[
@@
 which is clearly negative for all $x_1 \ge x^*$.
-%TODO: make more explicit/clear why this means \hat{x}_1 \in [0,x^*]
 This same process applied to the remaining variables shows the gradient of the full likelihood is negative for $x_1,x_2 \ge x^*$ and $y_1 \ge y^*$ and positive for $y_2,w \ge y^*$, yielding the constraints
@@
 \begin{proof}
-%TODO: make this more explicit/clear
 Similar to the previous proof, let
@@
 \end{align*}
-Note the following:
-\begin{itemize}
-\item $b_{\emptyset} = -b_{13}$, $p_{\emptyset} \ge p_{13}$;
-\item $b_{1} = -b_{3}$, $p_{1} = p_{3}$;
-\item $b_{2} = -b_{123}$, $p_{2} = p_{123}$;
-\item $b_{12} = -b_{23}$, $p_{12} = p_{23}$;
-\end{itemize}
-so
+Using the fact that $b_{\emptyset} = -b_{13}$, $b_{1} = -b_{3}$, $b_{2} = -b_{123}$, and $b_{12} = -b_{23}$ in addition to $p_{1} = p_{3}$, $p_{2} = p_{123}$, and $p_{12} = p_{23}$, we simplify the likelihood as
 \begin{align*}
@@
 \]
-The set of $x^*,y^*$ satisfying these conditions as well as those for $\emptyset$ being the most likely ancestral state is given in Figure~\ref{fig:incons-analytic}.
+The set of $x^*,y^*$ satisfying these conditions as well as those for $\emptyset$ being the most likely ancestral state is given in Fig.~\ref{fig:incons-analytic}.
 \end{proof}
 
-As the region in Figure~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on the incorrect topology in these cases, resulting in an inconsistency.
+As the region in Fig.~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on the incorrect topology in these cases, resulting in an inconsistency.
+We show this is a type of inconsistency in topology estimation by constructing a lower bound on the likelihood for the Felsenstein topology and showing it is equal to the likelihood given the InvFels topology in the demarcated region of Fig.~\ref{fig:incons-analytic}.
+As we are interested in a lower bound, assume $\ancestralSplitPartition_j(\tau_2,\{x_1,y_1,x_2,y_2,w\})=\emptyset$ for all $j$.
+In this case, the partial likelihood term for the Felsenstein tree is composed of the same functional forms given in Table~\ref{tab:likelihoods-restricted}.
+This is apparent as assuming $\emptyset$ for the ancestral state split will result in the same partial likelihood for both topologies.
+
+Assume $w=1$.
+The maximum likelihood value for the Felsenstein topology is greater than or equal to the value of the likelihood given these two constraints.
+In this case, Table~\ref{tab:gen-sitepatprob} shows that, for $w=1$, the divergence term is identical to that of the InvFels topology.
+Again, this is apparent as when the internal branch becomes zero-length, the two topologies are indistinguishable.
+Let $\mathcal{T}$ be the shaded region in Fig.~\ref{fig:incons-analytic}.
+We have
+\begin{align*}
+\max_{t} & \ \ell_{\tau^*,t^*}(\tau_2, t) \ge \\
+&\max_{t} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_2, \{x_1,y_1,x_2,y_2,1\}) \\
+&\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit_j, \tau_2, \{x_1,y_1,x_2,y_2,1\}) \\ 
+&= \max_{t} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \{x_1,y_1,x_2,y_2,1\})\\
+&\qquad \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit_j, \tau_1, \{x_1,y_1,x_2,y_2,1\})\\
+&= \max_{t : t^*\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau_1, t).
+\end{align*}
 

### 2018-07-16 joint_inf.tex
rearranging; cleaning TODOs; felsenstein argument

@@
 
-%TODO: update this figure
 \begin{figure}
 \centering
-\begin{subfigure}{.45\linewidth}
+\begin{subfigure}{.3\linewidth}
 \centering
-\includegraphics[width=\textwidth]{farris_blank}
+\includegraphics[width=\textwidth]{farris_blank_generating}
+\caption[short]{InvFels tree $\tau^*$}
+\end{subfigure}
+\begin{subfigure}{.3\linewidth}
+\centering
+\includegraphics[width=\textwidth]{farris_blank_general}
 \caption[short]{InvFels tree $\tau_1$}
 \end{subfigure}
-\begin{subfigure}{.45\linewidth}
+\begin{subfigure}{.3\linewidth}
 \centering
-\includegraphics[width=\textwidth]{felsenstein_blank}
+\includegraphics[width=\textwidth]{felsenstein_blank_general}
 \caption[short]{Felsenstein tree $\tau_2$}
 \end{subfigure}
-\caption{Two four-taxon trees with fidelities as labeled.
-}
+\caption{Three four-taxon trees with fidelities as labeled.}
 \label{fig:farris-fels-top}
@@
 
+The proof is through analyzing upper and lower bounds of various likelihood forms and the gradients of the objective function.
 In words, this implies an inconsistency as we converge on the incorrect topology, estimating the interior branch length to be zero.
 For example, the Felsenstein tree $\tau_2$ will achieve the same objective function value in this case as the InvFels tree $\tau_1$, as if the interior branch is zero-length these two topologies are indistinguishable.
-%TODO: more discussion on proof, intution
-%Intuitively, $\tau_2$ is favored in performing joint inference since the objective function for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its objective function while $\tau_2$ has many more.
-%This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.
 
@@
 This is not the case here; while analytically the inconsistency occurs in the long branch attraction region of the parameter space, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
-%TODO: add plot or at least discussion of intuition here
+This is especially apparent in the ``long $x^*$, short $y^*$'' case discussed previously.
 Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.

### 2018-07-16 joint_inf_helpers.py
add felsenstein to helpers

@@ -66,6 +66,14 @@ def parse_args():
         action="store_true",
         help='general branch lengths plot',
     )
+    parser.add_argument(
+        '--fels',
+        action="store_true",
+    )
+    parser.add_argument(
+        '--invfels',
+        action="store_true",
+    )
 
     args = parser.parse_args()
 
@@ -73,6 +81,50 @@ def parse_args():
 
 # ~~~~~~~~~
 # global functions for exact likelihood computation
+
+def P_FELS(theta, site_pattern):
+    """
+    Site pattern frequencies for Felsenstein tree
+    """
+    # fifth branch will always be positive
+    theta_sgn = [-1 if s=='1' else 1 for s in site_pattern+'0']
+    prob_list = [
+        [theta[0], theta[1]],
+        [theta[2], theta[3]],
+        [theta[0], theta[2], theta[4]],
+        [theta[1], theta[2], theta[4]],
+        [theta[0], theta[3], theta[4]],
+        [theta[1], theta[3], theta[4]],
+        [theta[0], theta[1], theta[2], theta[3]],
+    ]
+    sgn_list = [
+        theta_sgn[0]*theta_sgn[1],
+        theta_sgn[2]*theta_sgn[3],
+        theta_sgn[0]*theta_sgn[2]*theta_sgn[4],
+        theta_sgn[1]*theta_sgn[2]*theta_sgn[4],
+        theta_sgn[0]*theta_sgn[3]*theta_sgn[4],
+        theta_sgn[1]*theta_sgn[3]*theta_sgn[4],
+        theta_sgn[0]*theta_sgn[1]*theta_sgn[2]*theta_sgn[3],
+    ]
+    new_prob_list = [resolve_exp(p, pad=False) for p in prob_list]
+    out_str = ''.join(['+'+p if s > 0 else '-'+p for p,s in zip(new_prob_list, sgn_list)])
+
+    return PATT2SPLIT[site_pattern] + '&' + '(1'+out_str+')' + '\\\\'
+
+def L_FELS(theta, site_pattern):
+    """
+    Likelihood for Felsenstein topology
+    """
+    all_likes = []
+    all_likes += [''.join(['(1+'+t+')' if s=='0' else '(1-'+t+')' for t,s in zip(theta, site_pattern+'0')])]
[…]

### 2018-07-17 joint_inf_plot.py
code documentation and figure update

@@ -143,17 +143,20 @@ def safe_p_logq(p, logq):
 
 def marginal_likelihood(theta, theta_hat):
     """
-    Table S4 in manuscript: we have 81 possible likelihood functions we can compute
+    Marginal likelihood given generating theta and estimated theta_hat
     """
     probs = {pattern: P_INVFELS(theta, pattern) for pattern in ALL_PATTERNS}
     fit_probs = {pattern: P_INVFELS(theta_hat, pattern) for pattern in ALL_PATTERNS}
 
     likelihood = np.sum([safe_p_logq(p_gen, safe_log(fit_probs[pattern])) for pattern, p_gen in probs.iteritems()])
     likelihood -= LOG32
-    likelihood += get_partial_likelihood(probs, theta_hat)
+    likelihood += get_partial_marginal_likelihood(probs, theta_hat)
     return likelihood
 
-def get_partial_likelihood(probs, theta):
+def get_partial_marginal_likelihood(probs, theta):
+    """
+    Second term of the likelihood in marginal inference
+    """
     all_terms = []
     likelihood = 0.
     for t in theta:
@@ -174,6 +177,11 @@ def get_partial_likelihood(probs, theta):
     return likelihood
 
 def get_log_coefficients(probs, anc_states):
+    """
+    Function to return coefficients for (1+x_1), (1-x_1), etc.
+
+    returns a 5x2 array, with coefficient for (1+x_1) the 0,0 term, (1-x_1) the 0,1 term, etc.
+    """
     all_terms = []
     for _ in range(5):
         all_terms.append([0., 0.])
@@ -198,7 +206,7 @@ def get_log_coefficients(probs, anc_states):
 
 def likelihood_lower_bound(theta):
     """
-    Table S4 but with w=1 and y_2=1 (and then x_1, y_1, x_2 their optima)
+    Compute lower bound of the likelihood in Lemma 2
     """
     probs = {pattern: P_INVFELS(theta, pattern) for pattern in ALL_PATTERNS}
     all_terms = get_log_coefficients(probs, '0000')
@@ -214,6 +222,8 @@ def likelihood_lower_bound(theta):
 
 def likelihood_upper_bound(theta, anc_states='0000'):
     """
+    Compute upper bound of the likelihood in Lemma 2
+
     maximized at t=(a1-a2)/(a1+a2), but note a1+a2=1, so a1log(1+a1-a2) + a2log(1+a2-a1) is max
     """
[…]

### 2018-07-17 appendix.tex
clarify notation/arguments/assumptions

@@
 To compute the likelihood of observing a set of data, we need $\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
-Using branch fidelities, the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
+Using branch fidelities, the probability of a character change along a branch with fidelity parameter $x$ is $(1-x)/2$, while the probability of a character remaining the same is $(1+x)/2$.
 See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the InvFels tree.
@@
 All values multiplied by $1/32$.
-Ancestral states with $^*$ are never maximal provided parameters are in $(0,1)$.
+Ancestral states with $^*$ are never maximal provided parameters are in $(0,1]$.
 There are $3^5\cdot 4^2=3,888$ possible forms for the likelihood.
@@
 Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$ we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$.
-An analogous statement holds for $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral states.
+An analogous statement holds for $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.
 
@@
 \hline
-$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
+$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,t)$\\
 \hline
@@
 \caption{
-Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral states to reduce the number of possible functional forms to consider.
+Likelihood calculations on the InvFels tree $\tau_1$.
+Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
 All values multiplied by $1/32$.
@@
 
-We first prove intermediate results to show the above.
+We first prove intermediate results to show the main result.
 
 \begin{lemma}
-Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$.
-There exists a set of $0 < x^*, y^* < 1$ such that $\emptyset$ is the maximal ancestral state for all site splits.
+Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
+There exists a set of $0 < x^*, y^* < 1$ such that $\emptyset$ is the maximal ancestral state split for all site splits given $\tau=\tau_1$.
 \end{lemma}
@@
 \[
-\max_{t} \ \ell_{\tau^*,t^*}(\tau^*, t) \le
+\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \le
     \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
-    + \tilde{\ell}_{\tau^*,t^*}(\tau^*, \hat{t})
+    + \tilde{\ell}_{\tau^*,t^*}(\tau, \tilde{t})
 \]
@@
 \[
-\shannonDivergence_{\tau^*,t^*}(\tau^*,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
+\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
 \]
@@
 \[
-\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau^*, t).
+\tilde{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t).
 \]
-By straightforward calculus, the function $f(x)=a_1\log(1+x)+a_2\log(1-x)$ attains a maximum at $\hat{x} = (a_1-a_2)/(a_1+a_2)$, and so we have a closed form for $\hat{t}$.
+By straightforward calculus, the function $f(x)=a_1\log(1+x)+a_2\log(1-x)$ attains a maximum at $\hat{x} = (a_1-a_2)/(a_1+a_2)$, and so we have a closed form for each component of $\tilde{t}$.
 Using this estimate, we have the lower bound
 \[
-\max_{t} \ \ell_{\tau^*,t^*}(\tau^*, t) \ge
-    \shannonDivergence_{\tau^*,t^*}(\tau^*,\hat{t})
-    + \tilde{\ell}_{\tau^*,t^*}(\tau^*, \hat{t}).
+\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \ge
+    \shannonDivergence_{\tau^*,t^*}(\tau,\tilde{t})
+    + \tilde{\ell}_{\tau^*,t^*}(\tau, \tilde{t}).
 \]
-Fig.~\ref{fig:max-anc-state} shows where the lower bound above is greater than all 80 other upper bounds.
+We compute the lower bound $C_0(x^*,y^*)$ above given $\emptyset$ as the maximal ancestral state split for all site splits and $\tau=\tau_1$.
+We need only consider the remaining $u=1,\ldots,80$ possibilities for choices of ancestral state splits in Table~\ref{tab:likelihoods} upon excluding cases of infinite branch lengths (i.e., any of $x_1,y_1,x_2,y_2,w$ equal to zero) and the redundant cases of $x_1 > x_2$ and $y_1 > y_2$.
+We compute upper bounds $C_{1,u}(x^*,y^*)$ and show in Fig.~\ref{fig:max-anc-state} the region where $C_0(x^*,y^*) > \max_u C_{1,u}(x^*,y^*)$.
 \end{proof}
@@
 \caption{
-Region where $\emptyset$ is maximal ancestral state for each site split.
+Region where $\emptyset$ is maximal ancestral state split for each site split on the InvFels tree $\tau_1$.
 }
@@
 
-The above will hold if we swap $\hat{y}_2$ and $\hat{y}_1$, as the likelihood is invariant to this.
-
 \begin{lemma}
-Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$ and $\emptyset$ be the maximal ancestral state for all site splits.
-Assume $0 < x^*, y^* < 1$.
-The solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
+Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
+For $0 < x^*, y^* < 1$ and $\emptyset$ as the maximal ancestral state split for all site splits, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
 \[
-\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau^*, t)
+\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t)
 \]
-is subject to the constraints $\hat{x}_1,\hat{x}_2 \in [0, x^*]$, $\hat{y}_1 \in [0, y^*]$, and $\hat{y}_2,\hat{w} \in [y^*, 1]$.
+is subject to the constraints $\hat{x}_1,\hat{x}_2 \in (0, x^*]$, $\hat{y}_1 \in (0, y^*]$, and $\hat{y}_2,\hat{w} \in [y^*, 1]$.
 \label{lemma:bounds}
@@
 \[
-\shannonDivergence_{\tau^*,t^*}(\tau^*,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t).
+\shannonDivergence_{\tau^*,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
 \]
@@
 \[
-\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
+\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
 \]
@@
 \[
-\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau^*,t) \mid_{t=t^*} = 0.
+\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau_1,t) \mid_{t=t^*} = 0.
 \]
-We now show the gradient of the divergence term is a decreasing function on $(0,1)$ in each variable.
+We now show the gradient of the divergence term is a decreasing function on $(0,1]$ in each variable.
 We focus on $x_1$.
@@
 \]
-Regardless of $a_{\siteSplit_j} \lessgtr 0$, this is a decreasing function of $x_1$ provided $c_{\siteSplit_j} + a_{\siteSplit_j}x_1 > 0$, but we see
+We see
 \[
-\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) := \frac{1}{8} (c_{\siteSplit_j} + a_{\siteSplit_j}x_1)
+\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) := \frac{1}{8} (c_{\siteSplit_j} + a_{\siteSplit_j}x_1),
 \]
-which is strictly greater than zero provided $x_1,y_1,x_2,y_2\in(0,1)$.
+which greater than or equal to zero as it is a probability.
+Thus, regardless of $a_{\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{\siteSplit_j} + a_{\siteSplit_j}x_1 \ge 0$.
 The remaining variables follow similar arguments.
-
-We now focus on the partial likelihood term.
-Assume $\emptyset$ is the most likely ancestral state for all site splits.
-In this case, the partial likelihood has no ambiguity, and has terms given in Table~\ref{tab:likelihoods-restricted}.
 
@@
 \hline
-$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau, t)$ & $\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
+$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit_j,\tau_1,t)$\\
 \hline
@@
 \caption{
-The maximal likelihood values given $\emptyset$ is the most likely ancestral state.
+The maximal likelihood values on the InvFels tree $\tau_1$ given $\emptyset$ is the most likely ancestral state split for each site split.
 All values multiplied by $1/32$.}
@@
 
+We now focus on the partial likelihood term.
+Assume $\emptyset$ is the most likely ancestral state split for all site splits.
+In this case, the partial likelihood has no ambiguity, and has terms given in Table~\ref{tab:likelihoods-restricted}.
 The partial likelihood
 \[
-\tilde{\ell}_{\tau^*,t^*}(\tau^*, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau^*, t)
+\tilde{\ell}_{\tau^*,t^*}(\tau_1, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, t)
 \]
@@
 \]
-By some calculus, the partial derivative with respect to $x_1$ is
+The derivative with respect to $x_1$ is
 \[
-\tilde{\ell}_{x_1}(x_1) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2}
+\tilde{\ell}'(x_1) := \frac{d}{dx_1} \tilde{\ell}(x_1) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2}
 \]
-which is clearly negative for all $x_1 \ge x^*$.
+which is negative for all $x_1 \ge x^*$.
 This same process applied to the remaining variables shows the gradient of the full likelihood is negative for $x_1,x_2 \ge x^*$ and $y_1 \ge y^*$ and positive for $y_2,w \ge y^*$, yielding the constraints
@@
 \begin{proof}
-Similar to the previous proof, let
+The objective function proportional to $w$ is
+\begin{align*}
+\ell(w) \propto \log(1+w) + \sum_j p_{\siteSplit_j} \log(1 + b_{\siteSplit_j}w)
+\end{align*}
+where, as in the previous proof,
 \[
@@
 with similar definitions for the remaining site splits.
-The objective function proportional to $w$ is
-\begin{align*}
-\ell(w) \propto \log(1+w) + \sum_j p_{\siteSplit_j} \log(1 + b_{\siteSplit_j}w)
-\end{align*}
-and has derivative with respect to $w$ as
+This function has derivative with respect to $w$ as
 \begin{align*}
@@
 \end{align*}
-Using the fact that $b_{\emptyset} = -b_{13}$, $b_{1} = -b_{3}$, $b_{2} = -b_{123}$, and $b_{12} = -b_{23}$ in addition to $p_{1} = p_{3}$, $p_{2} = p_{123}$, and $p_{12} = p_{23}$, we simplify the likelihood as
+Using the fact that $b_{\emptyset} = -b_{13}$, $b_{1} = -b_{3}$, $b_{2} = -b_{123}$, and $b_{12} = -b_{23}$ in addition to $p_{1} = p_{3}$, $p_{2} = p_{123}$, and $p_{12} = p_{23}$, we simplify the derivative as
 \begin{align*}
@@
 \end{align*}
-Using the bounds on $\hat{x}_1,\hat{y}_1,\hat{y}_1$, and $\hat{y}_2$ from Lemma~\ref{lemma:bounds} and our assumption of $y_2 \ge y_1$ and $x_2 \ge x_1$, we have
+Using the bounds on $\hat{x}_1,\hat{y}_1,\hat{x}_2$, and $\hat{y}_2$ from Lemma~\ref{lemma:bounds} and our assumption of $y_2 \ge y_1$ and $x_2 \ge x_1$, we have
 \[
-0 \le b_{\emptyset} \le \frac{2x^*}{1+(x^*)^2} := b_{\emptyset}^U,
+0 < b_{\emptyset} \le \frac{2x^*}{1+(x^*)^2} := b_{\emptyset}^U,
 \]
 \[
-0 \le b_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*} := b_{2}^U,
+0 < b_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*} := b_{2}^U,
 \]
 \[
-0 \le b_{1} \le \frac{x^*}{1-(x^*)^2} := b_{1}^U
+0 < b_{1} \le \frac{x^*}{1-(x^*)^2} := b_{1}^U,
 \]
 \[
-0 \le b_{12} \le \frac{x^*}{1-(x^*)^2}\cdot\frac{1}{1-y^*} := b_{12}^U,
+0 < b_{12} \le \frac{x^*}{1-(x^*)^2}\cdot\frac{1}{1-y^*} := b_{12}^U,
 \]
@@
 \]
-Clearly $b_{\emptyset}^U < 1$.
-Note that $\ell'_{L}(w)$ is strictly decreasing in $w$ on the interval $(0,1)$, as shown in the previous proof, provided $b_{2}^U, b_{1}^U, b_{12}^U < 1$.
-Thus, if $\ell'_{L}(1) > 0$ then $\ell(w)$ is increasing on the entire interval $(0,1)$, including at the boundary $w=1$, implying $\hat{w} \equiv 1$.
-So we want $x^*$ and $y^*$ where $b_{2}^U, b_{1}^U, b_{12}^U < 1$ and $\ell_L'(1) > 0$.
-These conditions are straightforward, the one on the derivative being
+Note that $\ell'_{L}(w)$ is strictly decreasing in $w$ on the interval $(0,1]$, as shown in the previous proof, provided $b_{\emptyset}^U, b_{2}^U, b_{1}^U, b_{12}^U < 1$.
+Thus, if $\ell'_{L}(1) > 0$ then $\ell(w)$ is increasing on the entire interval $(0,1]$, including at the boundary $w=1$, and $\hat{w} \equiv 1$.
+The conditions on $x^*$ and $y^*$ where $b_{2}^U, b_{1}^U, b_{12}^U < 1$ and $\ell_L'(1) > 0$ are straightforward, the one on the derivative being
 \[
@@
 \]
-The set of $x^*,y^*$ satisfying these conditions as well as those for $\emptyset$ being the most likely ancestral state is given in Fig.~\ref{fig:incons-analytic}.
+The set of $x^*,y^*$ satisfying these conditions as well as those for $\emptyset$ being the most likely ancestral state split is given in Fig.~\ref{fig:incons-analytic}, giving the region of generating parameters where $\hat{w}\equiv 1$.
 \end{proof}
 
-As the region in Fig.~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on the incorrect topology in these cases, resulting in an inconsistency.
-We show this is a type of inconsistency in topology estimation by constructing a lower bound on the likelihood for the Felsenstein topology and showing it is equal to the likelihood given the InvFels topology in the demarcated region of Fig.~\ref{fig:incons-analytic}.
-As we are interested in a lower bound, assume $\ancestralSplitPartition_j(\tau_2,\{x_1,y_1,x_2,y_2,w\})=\emptyset$ for all $j$.
-In this case, the partial likelihood term for the Felsenstein tree is composed of the same functional forms given in Table~\ref{tab:likelihoods-restricted}.
-This is apparent as assuming $\emptyset$ for the ancestral state split will result in the same partial likelihood for both topologies.
+As the region in Fig.~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
+We show this is a type of inconsistency in topology estimation by constructing a lower bound on the likelihood for the Felsenstein topology $\tau_2$ and demonstrating it is equal to the likelihood given the InvFels topology $\tau_1$ in the demarcated region of Fig.~\ref{fig:incons-analytic}.
+As we are interested in a lower bound, assume $\ancestralSplitPartition_j(\tau_2,\{x_1,y_1,x_2,y_2,w\})=\emptyset$ for all $j=1,\ldots,\nSiteSplits$.
+In this case, the partial likelihood term for $\tau_2$ is composed of the same functional forms given in Table~\ref{tab:likelihoods-restricted}, as assuming $\emptyset$ for the ancestral state split results in the same partial likelihood for both topologies.
 
 Assume $w=1$.
-The maximum likelihood value for the Felsenstein topology is greater than or equal to the value of the likelihood given these two constraints.
-In this case, Table~\ref{tab:gen-sitepatprob} shows that, for $w=1$, the divergence term is identical to that of the InvFels topology.
-Again, this is apparent as when the internal branch becomes zero-length, the two topologies are indistinguishable.
-Let $\mathcal{T}$ be the shaded region in Fig.~\ref{fig:incons-analytic}.
+In this case, Table~\ref{tab:gen-sitepatprob} shows that, for $w=1$, the divergence term is identical for $\tau_1$ and $\tau_2$ as when the internal branch becomes zero-length the two topologies are indistinguishable.
+Let $\mathcal{T}$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}$ corresponds to $x^*$ and $y^*$ falling in the shaded region in Fig.~\ref{fig:incons-analytic}.
+The unconstrained maximum likelihood value for $\tau_2$ is greater than or equal to the value of the likelihood given these two constraints.
 We have
@@
 \max_{t} & \ \ell_{\tau^*,t^*}(\tau_2, t) \ge \\
-&\max_{t} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_2, \{x_1,y_1,x_2,y_2,1\}) \\
-&\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit_j, \tau_2, \{x_1,y_1,x_2,y_2,1\}) \\ 
-&= \max_{t} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \{x_1,y_1,x_2,y_2,1\})\\
-&\qquad \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit_j, \tau_1, \{x_1,y_1,x_2,y_2,1\})\\
-&= \max_{t : t^*\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau_1, t).
+&\max_{t: \xi_j=\emptyset, w=1, \tau=\tau_2} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}) \\
+&\qquad\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\}) \\
+&= \max_{t: \xi_j=\emptyset, w=1, \tau=\tau_1} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
+&\qquad\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\})\\
+&= \max_{t : t^*\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau_1, t),
 \end{align*}
+and thus the incorrect topology achieves a likelihood value greater than or equal to the correct one.
 
-\subsection*{Empirical validation}
-
-Consider Table~\ref{tab:likelihoods}.
-There are 81 separate possible forms for the likelihood, corresponding to the three possible most likely ancestral states for each of the four ambiguous site splits.
-In Fig.~\ref{fig:bl-general-inconsistency}, for various values of $x^*,y^*$, we compute the maximum of each of the 81 likelihoods and plot the value of $\hat{p}_w = (1-\hat{w})/2$.
+%~~~
+% Figures mentioned in main section of manuscript
 

### 2018-07-17 joint_inf.tex
clarify notation/arguments/assumptions

@@
 Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
-In this paper, we show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
+We show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
 We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates an incorrect topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
-For branch length estimation on the correct topology, we outline similar cases where branch length estimates are consistently and heavily biased and provide extensive empirical results for this setting.
+For branch length estimation on the correct topology, we outline cases where branch length estimates are consistently and heavily biased and provide extensive empirical results for this setting.
 \end{abstract}
@@
 
-An estimator is said to be statistically consistent if it converges to the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
+An estimator is said to be statistically consistent if it converges to the generating model with probability one in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
 These proofs do not readily extend to include estimating ancestral states.
@@
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model with data being generated on the four-taxon ``inverse Felsenstein'' tree (a.k.a.\ the ``Farris'' tree) and we construct bounds on the joint objective function to demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
+To do so, we use a binary symmetric model with data being generated on a four-taxon tree and we construct bounds on the joint objective function to demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
 We find similar areas through empirical means where joint inference consistently underestimates interior branch lengths when the topology is known and fixed.
@@
 
-We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``inverse Felsenstein (InvFels)'' tree (Fig.~\ref{fig:farris-fels-top}).
+We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``inverse Felsenstein (InvFels)'' tree (Figs.~\ref{fig:farris-fels-top}a and~\ref{fig:farris-fels-top}b) and the ``Felsenstein'' tree (Fig.~\ref{fig:farris-fels-top}c).
 The ``inverse Felsenstein'' terminology comes from \citet{Swofford2001-hr}, although it is also called the ``Farris'' tree \citep{Siddall1998-hq, Felsenstein2004}.
-In the standard configuration of this tree, the interior branch length parameters are equal to the bottom two parameters.
-We use this standard configuration as our data generating process, though we do not constrain our branch lengths to be equal in considering our optimization objective function.
+In the standard configuration of this tree, the interior branch parameters are equal to the bottom two parameters as in Fig.~\ref{fig:farris-fels-top}a.
+We use this standard configuration as our data generating process, though we do not constrain our branch parameters to be equal in considering our optimization objective function.
 
@@
 We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with an alternate formulation called ``fidelity.''
-The probability of a substitution on a branch with fidelity $\theta$ is $(1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$ where $0 \le \theta \le 1$.
+The probability of a substitution on a branch with fidelity $x$ is $(1-x)/2$ while the probability of no substitution is $(1+x)/2$ where $0 \le x \le 1$.
 This parameter quantifies the fidelity of transmission of the ancestral state across an edge \citep{Matsen2007-jq}.
@@
 As data becomes plentiful, we use the Hadamard transform (see \eqref{eq:hadamard_probability} in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form.
-For a four-taxon tree, define the general branch fidelity parameter $t=\{\theta_1,\theta_2,\theta_3,\theta_4,\theta_5\}$ where fidelities are ordered in the order of the taxa with the internal branch last (Fig.~\ref{fig:farris-fels-top}).
-Although we use fidelities exclusively for our theoretical development, we have made our figures in terms of probabilities of substitution $p_{\theta} = (1-\theta)/2$ as they are easier to interpret.
+For a four-taxon tree, define the general branch fidelity parameter $t=\{x_1,y_1,x_2,y_2,w\}$ where fidelities are ordered in the order of the taxa with the internal branch last (Figs.~\ref{fig:farris-fels-top}b and~\ref{fig:farris-fels-top}c).
+Although we use fidelities exclusively for our theoretical development, we have made our figures in terms of probabilities of substitution $p_{x} = (1-x)/2$ as they are easier to interpret.
 
@@
 The alternative approach \citep{Sagulenko2018-xl} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
-This is known in statistics as a profile likelihood \citep{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
+This is known in statistics as a profile likelihood \citep{Murphy2000-ry} or a relative likelihood \citep{Goldman1990-dk}, which exists here because $\alphabet$ is a finite set:
 \begin{equation}
@@
 \end{equation}
-In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral state splits in Table~\ref{tab:farris_likelihoods} to obtain the likelihood term for each most likely ancestral state, these terms depending on the unknown $(\tau, t)$.
+In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral state splits in Table~\ref{tab:farris_likelihoods} to obtain the likelihood term for each most likely ancestral state split, these terms depending on the unknown $(\tau, t)$.
 For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2018-xl}.
@@
 
-Assume $\fullAlignment$ is generated from topology $\tau^*$ and branch fidelities $t^*$.
+Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ (Fig.~\ref{fig:farris-fels-top}a).
 Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
@@
 
-To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the InvFels topology (Fig.~\ref{fig:farris-fels-top}a).
 %We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology ($\tau_2$) than for the true, generating InvFels topology ($\tau_1$).
@@
 \begin{restatable}{theorem}{topoInconsist}
-Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$.
-There exists a set of $0 < x^*, y^* < 1$ such that
+Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
+There exists a set of $0 < x^*, y^* < 1$ such that the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
 \[
-\{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\} = \arg\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t)
+\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t)
 \]
-and $\hat{w}\equiv 1$.
+has the property $\hat{w}\equiv 1$.
 \end{restatable}
 
-The proof is through analyzing upper and lower bounds of various likelihood forms and the gradients of the objective function.
-In words, this implies an inconsistency as we converge on the incorrect topology, estimating the interior branch length to be zero.
-For example, the Felsenstein tree $\tau_2$ will achieve the same objective function value in this case as the InvFels tree $\tau_1$, as if the interior branch is zero-length these two topologies are indistinguishable.
+The proof is through analyzing upper and lower bounds of the potential likelihood forms in Table~\ref{tab:farris_likelihoods} and the gradients of the objective function $\ell_{\tau^*,t^*}(\tau_1, t)$ with respect to $t$.
+This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one), and the incorrect topology attains an equal or greater likelihood at its maximum than the true topology in the limit.
+For example, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t)$ and $\ell_{\tau^*,t^*}(\tau_2, t)$ are equivalent.
+We elaborate on this point in the Appendix.
 
@@
 We provide the following as an intuition for the theoretical development.
-For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral states.
+For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral state splits.
 For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
-A maximization procedure will more likely introduce a $(1+w)$ term than a $(1-w)$ term, though this will not be guaranteed as the maximum depends on the values of $x$ and $y$.
+A maximization procedure will more likely introduce a $(1+w)$ term than a $(1-w)$ term, though this will not be guaranteed as the maximum depends on the values of the unknown branch parameters $t$.
 This tendency to include $(1+w)$ terms in the likelihood results in a bias toward one of branch fidelities.
-This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split (Table~\ref{tab:likelihoods} in the Appendix with $x_1=x_2=x^*$ and $y_1=y_2=y^*$).
-If we allow multifurcating trees in our inference, then we can think of this as an instance of converging to the wrong topology.
+This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split if we let $x_1=x_2=x^*$ and $y_1=y_2=y^*$ (Table~\ref{tab:likelihoods}).
+If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology, as the true $y^*\neq 1$.
 
@@
 Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:incons-analytic}).
-To determine how conservative, we use the well-known L-BFGS-B method to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
-We see that the region of inconsistency in the general case is much larger than in the analytic case---the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral states, so this is reasonable.
-This region encompasses almost half of the branch parameter space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch length to be zero.
+To determine how conservative, we use the L-BFGS-B method to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
+We see that the region of inconsistency in the general case is much larger than in the analytic case---the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral state splits, so this is reasonable.
+This region encompasses almost half of the branch parameter space; given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch length to be zero.
 
@@
 For each function, we optimize the objective starting at five separate initial values to mitigate problems with small curvature of the objective surface and nonconcavity.
-We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $0$ to $1$.
+We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*,y^*\in(0,1)$.
 Our optimization code can be found at \url{https://github.com/matsengrp/joint-inf/}.
@@
 To understand the quality of each fit, we report the range of $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.
-Due to the discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}), an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization.
+Due to the discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}), an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
 For joint inference, the errors given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for marginal inference, $[-3\times 10^{-17}, 3\times 10^{-17}]$ showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
@@
 We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
-Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of inconsistency: the joint inference procedure estimates severely biased branch lengths even given the correct topology.
+Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of inconsistency: the joint inference procedure estimates zero-length branches given the correct topology.
+This results in an incorrect topology having a likelihood as large or greater than that of the generating topology.
 These regions of inconsistency arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
+%TODO: double check the below sentence
 This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and~3.
@@
 Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., the method of sieves \citep{Geman1982}.
-A potential solution in this case using the method of sieves could be to project the column-wise ancestral state patterns into a lower-dimensional space, allowing the degrees of freedom in the ancestral state columns to grow with $\nCols$, albeit more slowly than $O(\nCols)$.
+A potential solution in this case using the method of sieves could be to project the column-wise ancestral states into a lower-dimensional space, allowing the degrees of freedom in the ancestral state columns to grow with $\nCols$, albeit more slowly than $O(\nCols)$.
 Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \citep{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \citep{LeCam1990}.
@@
 We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.
+%TODO: change up this line?
 Insight from the editors and Vu Dinh greatly improved this manuscript.

### 2018-07-17 appendix.tex
consistent notation and a few loose ends

@@
 \caption{
-    Example likelihood computations on the InvFels tree $\tau_1$ for fidelities $\{x_1,y_1,x_2,y_2,w\}$.
+    Example likelihood computations on the InvFels tree $\tau_1$ for fidelities $t=\{x_1,y_1,x_2,y_2,w\}$.
     Edges labeled by the probability of substitution along that edge.
@@
 
-Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
+Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
 We show that the likelihood remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
@@
 \begin{align*}
-    \Pr(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
+    \Pr(\siteSplitRV=\emptyset\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                               & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
@@
 \begin{align*}
-    \Pr(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
+    \Pr(\siteSplitRV=\{1\}\mid \tau_1, t) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
 \end{align*}
@@
 \begin{align*}
-    \Pr(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
+    \Pr(\siteSplitRV=\{3\}\mid \tau_1, t) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
 \end{align*}
@@
 \begin{align*}
-    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
-    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
-    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
-    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
+    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
+    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
+    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
+    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
 \end{align*}
@@
 \begin{align*}
-        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
-    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
-    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
-    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
+        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
+    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
+    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
+    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
 \end{align*}
@@
 The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
-Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$ we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$.
+Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t)$ we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$.
 An analogous statement holds for $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.
@@
 \centering
-\begin{tabular}{|ll|l|}
+\begin{tabular}{|lll|l|}
 \hline
-$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,t)$\\
+$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_j$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,t)$\\
 \hline
-$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
+$\emptyset$&$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
 
-$\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
+$\{1\}$    &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
 
-$\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
+$\{2\}$    &$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
 
-$\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{1,2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
+$\{3\}$    &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
+&&$\{1\}$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
+&&$\{1,2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
 
-$\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{1,2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
+$\{1,2,3\}$&$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
+&&$\{1\}$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
+&&$\{1,2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
 
-$\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
+$\{1,2\}$  &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
 
-$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
+$\{2,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
+&&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
+&&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
 
-$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
-&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
-&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
+$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
+&&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
+&&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
 \hline
@@
 \centering
-\begin{tabular}{|ll|l|}
+\begin{tabular}{|lll|l|}
 \hline
-$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit_j,\tau_1,t)$\\
+$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_j$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,t)$\\
 \hline
-$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
-$\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
-$\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
-$\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
-$\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
-$\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
-$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
-$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
+$\emptyset$&$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
+$\{1\}$    &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
+$\{2\}$    &$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
+$\{3\}$    &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
+$\{1,2,3\}$&$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
+$\{1,2\}$  &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
+$\{2,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
+$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
 \hline
@@
 \[
-\hat{x}_1,\hat{x}_2 \in [0, x^*], \ \hat{y}_1 \in [0, y^*], \ \hat{y}_2,\hat{w} \in [y^*, 1].
+\hat{x}_1,\hat{x}_2 \in (0, x^*], \ \hat{y}_1 \in (0, y^*], \ \hat{y}_2,\hat{w} \in [y^*, 1].
 \]
@@
 We show this is a type of inconsistency in topology estimation by constructing a lower bound on the likelihood for the Felsenstein topology $\tau_2$ and demonstrating it is equal to the likelihood given the InvFels topology $\tau_1$ in the demarcated region of Fig.~\ref{fig:incons-analytic}.
-As we are interested in a lower bound, assume $\ancestralSplitPartition_j(\tau_2,\{x_1,y_1,x_2,y_2,w\})=\emptyset$ for all $j=1,\ldots,\nSiteSplits$.
+As we are interested in a lower bound, assume $\xi_j=\emptyset$ for all $j=1,\ldots,\nSiteSplits$.
 In this case, the partial likelihood term for $\tau_2$ is composed of the same functional forms given in Table~\ref{tab:likelihoods-restricted}, as assuming $\emptyset$ for the ancestral state split results in the same partial likelihood for both topologies.
@@
 \caption{
-    Estimates for $\hat{p}_w$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using L-BFGS-B optimizing the classical marginal likelihood \eqref{eq:marginal_likelihood} (rather than a joint optimization procedure).
+    Estimates for $\hat{p}_w$ when computing $(\hat{x}_1, \hat{y}_1, \hat{x}_2, \hat{y}_2, \hat{w})$ using L-BFGS-B optimizing the classical marginal likelihood \eqref{eq:marginal_likelihood} rather than a joint optimization procedure.
 }

### 2018-07-17 versions/response.v1.txt
update revision comments
pass thru reviewer response

@@
 %das: I added some stuff below to maybe alleviate some confusion and to show why what this reviewer thinks we should do won't work, but not sure how much is necessary.
-%Maybe it adds to why we didn't do this.
 It is true that the form given above for l(x,y,w) has an explicit and simple maximizer.
-However, the long-run log-likelihood is slightly more complicated than this function, as seen in equation (9).
+However, the long-run log-likelihood is more complicated than this function, as seen in equation (9).
 For both topologies, we have a number of equations of this form that we would need to maximize in a grid search.
@@
 Doing a grid search over all ancestral states would not be trivial, and we would need to perform numerical optimization for all possible ancestral states to compute the maximum.
+Moreover, not assuming x_1=x_2 and y_1=y_2 (see below) results in an explosion of functions to perform a grid search over, with none of these functions being guaranteed concave.
+The analytic bounds we have provided were devised specifically to sidestep this nonconcavity.
 
@@
 TODO: make the argument more clear in the main text.
-%das: Added sentence at L92. Do you think it needs more?
+%das: BIIIIG update
+We thank this reviewer for highlighting an omission in our argument.
+We had initially simplified the results by making the erroneous assumption that t_1=t_3 and t_2=t_4.
+It can be shown that this assumption ignores potential feasible parameter sets, and we have removed it.
+This requires a change in the arguments in the theorems as well, though the conclusions are similar.
 
@@
 %EM I think we could move this to the appendix to throw this reviewer a bone.
-%das: does this reviewer mean both the general case and the empirical case or just the empirical case?
-%Their review makes me think they take the empirical case the be the "general" case.
-%You think we should move both analytic cases to the appendix or just the one?
+%das: another big update
+We have removed the restricted case as it does not reflect how inference is performed and was only included as an expository example.
+Given the change to the theorem arguments, we do not believe this case is necessary for highlighting inconsistency.
 
@@
 
-%das: @EM how do you feel about just plotting 1-x and 1-y and explaining that this is intuitively more like branch lengths?
+Thank you for this suggestion.
+We have implemented it in all the plots and have been careful to clarify that "fidelities" do not correspond to "lengths" in the text, deferring to "lengths" where appropriate.
 
@@
 %EM David-- does it seem like we could extend these results to bigger trees containing a 4 taxon tree?
-%das: I don't see why not, though I haven't thought it through very much.
-%It could take a lot of work or not much work at all.
-%Do you know of any other mathematical phylogenetics papers that do something like this?
-%Like induction on trees?
-%I can think a bit and get back to you...
+%das: I wasn't able to come up with anything interesting for this.
 
@@
 %das: this reviewer was confused about the phrase "generating probabilities," thinking we were generating the probabilities and not that they were the probabilities of generating a particular data configuration.
-%I changed:
-%Fidelities have useful algebraic properties, and generating probabilities using the Hadamard transform have an especially simple form (see \eqref{eq:hadamard_probability} in the Appendix).
-%to:
-%Fidelities have useful algebraic properties.
-%As data becomes plentiful, we use the Hadamard transform (see \eqref{eq:hadamard_probability} in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form.
-
 We have added the definition of a "generating probability" to ease confusion in this sentence.
@@
 TODO: this is changed, but I don't see this being a lot clearer.
-%das: changed, is it better?
+%das: I believe this is clear now; lmk
+We have clarified the intuition in this sentence.
 
@@
 
-TODO: fill in
-%das: This reviewer was confused about the bold eta, which was an ordered list of sets eta_1, eta_2, ...
-%I noticed we never actually used this ordered list so got rid of it and wrote:
-%Thus given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies...
-
 We removed the ordered list of sets as it did not add to any understanding of the concept and now refer to each of these sets and their relevance directly.
@@
 
-TODO: just say tough nuts?
-%das: I defined "generating probability" earlier, but, I mean, come on, a reference? We give a section of a chapter! I'd vote for tough nuts:
-%We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
-
 We have updated the text to explicitly define "generating probability." Background for equation (8) can be found in Section 8.6 in the Semple and Steel text.
 This is referenced at the beginning of the section.
+We do not believe an exhaustive explanation would provide any benefit to the reader that they could not get from referring to the text.
 
@@
 Thank you for your thoughts, and we have worked hard to not oversell our results.
-TODO we should make some changes to at least say that we've further clarified the implications of our results.
-%das: can you clarify this TODO?
-%Do you mean we should add more text about the bias in Fig. S3, or we should quote Richard more and show how we think things might still be bad?
 The TreeTime paper does present joint optimization as an approximate approach, and described that it is for short branch lengths.
@@
 The TreeTime paper says that its use is appropriate when "ancestral sequence inference has little ambiguity" but even in these cases the bias in branch lengths persists.
+
+TODO we should make some changes to at least say that we've further clarified the implications of our results.
+%das: how much more should we say than this sentence (which we already say)? We already quote them as saying they should only use it when branches are short.
+%Although the software described in \citet{Sagulenko2017-jo} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
+%and
+%In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.
+%This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2017-jo}, namely their Figures~2 and 3.
 
@@
 This is an interesting suggestion but is different than our goal, which was to establish inconsistency.
+Moreover, in the revised manuscript we see cases where assigning ML ancestral states results in longer branches, not shorter ones (Lemma 3 in the updated paper).
+We believe the inaccuracies have more to do with the specific choice of which ancestral state is maximal than asymmetries in the tree.
+Asymmetries can drive these particular choices, but Fig. 3 in the revised manuscript shows that almost half of the parameter space results in branch lengths being estimated as zero.
+We do not think providing an argument about accumulating inaccuracies would elucidate why such starkly-defined regions of inconsistency exist.
+For reasonable parameter settings, one could investigate the error accumulation given fixed ancestral states, but this was not our goal and will require another paper.
+
 TODO describe why Taylor isn't a straightforward thing.
@@
 
+It is true when y=x=w we are unable to distinguish the topologies apart from the labeling, though the labeling is what distinguishes the topologies.
+As shown in Goldman (1990), with all branch lengths fixed and equal the joint inference procedure degenerates into parsimony, which has well known inconsistency issues.
+We have added discussion of this point.
+
 >  - line 50: overestimates branch lengths? It overestimates fidelity ->
@@
 
+Thank you for bringing this to our attention.
+We have tried to use branch lengths in as many instances as we could in the discussions, using fidelities only in the computations as they allow for great simplifications.
+
 >  - Fig 5: it is surprising that \hat{w} seems completely flat as a
@@
 >  the analytical bound in Fig 5?
+
+The discontinuity is a reflection of the change in ancestral states when the generating parameters change.
+As the true generating parameters change, we become more likely to choose a different ancestral state as most likely, resulting in an entirely different objective function to be optimized.
+The analytical bound is a coarse estimate of where inconsistency arises.
+In the revised manuscript, we fix the maximal ancestral state to simplify the argument, and comparing the analytical with the empirical would not provide much added benefit to the analysis.
 
@@
 TODO we should do some change, but make it clear that it's consistently biased even for short branches.
+%das: I added this line:
+We have added the line:
+\citet{Sagulenko2017-jo} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:bl-general-bias} shows even when branches are short there is a consistent bias.
+
+We believe it's important not to mischaracterize the suggested use of joint inference, though we would like to emphasize to the reader that problems are apparent in large regions of the parameter space.
+We think the current discussion treats this optimization procedure as fairly as possible.
 
@@
 %EM Also, I think you said that he says that the varying branch length thing is an open question-- it would be great to quote that.
+%das: update I added some of the below.
+Let me know what you think.
+And re: Allman/Chai, I just added them in with RoyChoudhury; is that ok?
+They are all just examples of conditions where phylogenetics is consistent it seems.
+
 %das: below are some notes and quotes. I will work on paring these down and adding some to the text, maybe with a little more diplomacy.

### 2018-07-17 joint_inf.tex
consistent notation and a few loose ends
pass thru reviewer response

@@
 For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
-A maximization procedure will more likely introduce a $(1+w)$ term than a $(1-w)$ term, though this will not be guaranteed as the maximum depends on the values of the unknown branch parameters $t$.
-This tendency to include $(1+w)$ terms in the likelihood results in a bias toward one of branch fidelities.
+As $(1+w) > (1-w)$, a maximization procedure more likely introduces the $(1+w)$ term, though this will not be guaranteed as the maximum depends on the values of the unknown branch parameters $t$.
+This tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
 This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split if we let $x_1=x_2=x^*$ and $y_1=y_2=y^*$ (Table~\ref{tab:likelihoods}).
@@
 For joint inference, the estimates for $\hat{p}_w$ when $p_{x^*}$ and $p_{y^*}$ are both small look reasonable, with $\hat{p}_w$ increasing as $p_{y^*}$ increases, though Fig.~\ref{fig:empirical-bias} shows there is a systematic negative bias in this procedure even when the true branches are short.
+%TODO: does this bias make sense? should it be log-diff?
 To understand the quality of each fit, we report the range of $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.

### 2018-07-18 joint_inf_helpers.py
more documentation

@@ -57,22 +57,19 @@ def parse_args():
     parser = argparse.ArgumentParser(description=__doc__)
 
     parser.add_argument(
-        '--restricted-branch-lengths',
-        action="store_true",
-        help='restricted branch lengths plot',
-    )
-    parser.add_argument(
         '--general-branch-lengths',
         action="store_true",
-        help='general branch lengths plot',
+        help='general branch lengths probabilities',
     )
     parser.add_argument(
         '--fels',
         action="store_true",
+        help='felsenstein topology probabilities',
     )
     parser.add_argument(
         '--invfels',
         action="store_true",
+        help='inverse felsenstein topology probabilities',
     )
 
     args = parser.parse_args()
@@ -191,12 +188,10 @@ def main(args=sys.argv[1:]):
     if args.general_branch_lengths:
         pref = 'General'
         theta = ['x_1', 'y_1', 'x_2', 'y_2', 'w']
-    elif args.restricted_branch_lengths:
-        pref = 'Restricted'
-        theta = ['x', 'y', 'x', 'y', 'w']
     else:
         pref = 'True'
         theta = ['x*', 'y*', 'x*', 'y*', 'y*']
+
     if args.fels:
         pref += ' Fels'
         p = lambda x,y: P_FELS(x, y)

### 2018-07-18 joint_inf_plot.py

@@ -17,6 +17,8 @@ from multiprocessing import Pool
 from scipy.optimize import minimize
 from itertools import product
 
+from joint_inf_helpers import ALL_PATTERNS
+
 sns.set_style('white')
 sns.set_style('ticks')
 
@@ -29,16 +31,6 @@ FONT_SIZE = 20
 LOG2 = np.log(2)
 LOG32 = np.log(32)
 
-ALL_PATTERNS = [
-        '0000',
-        '1000',
-        '0100',
-        '0010',
-        '1110',
-        '1100',
-        '1010',
-        '0110'
-]
 
 class Legend(object):
     pass
@@ -79,33 +71,34 @@ def parse_args():
     parser.add_argument(
         '--ancestral-state-conditions',
         action="store_true",
-        help='plot analytic',
+        help='plot conditions where \emptyset is most likely ancestral state (Fig. S2)',
     )
     parser.add_argument(
         '--empirical-parameter-estimate',
         action="store_true",
-        help='compute empirical',
+        help='plot empirical estimate of \hat{w} for joint inference (Fig. 3)',
     )
     parser.add_argument(
         '--marginal',
         action="store_true",
-        help='compute marginal',
+        help='plot empirical estimate of \hat{w} for marginal likelihood (Fig. S4)',
     )
     parser.add_argument(
         '--in-pkl-name',
         type=str,
-        help='file name for where to save pkl',
+        help='file name for where saved output is, if computed, for time saving',
         default=None,
     )
     parser.add_argument(
         '--out-pkl-name',
         type=str,
-        help='file name for where to save pkl',
+        help='file name for where to save output for time saving',
         default='output.pkl',
     )
     parser.add_argument(
         '--analytic-inconsistency',
         action="store_true",
+        help='plot conditions where \emptyset is most likely ancestral state and \hat{w} is one analytically (Fig. 2)',
     )
 
     args = parser.parse_args()
@@ -135,6 +128,7 @@ def P_INVFELS(theta0, site_pattern):
 # ~~~~~~~~~
 # Functions for empirical plot
 
+# Safe log and safe p*log(q) so we do not have as many edge cases in optimization
 def safe_log(x):
     return np.log(x) if x > 0. else -np.inf
 

### 2018-07-18 joint_inf_plot.py
add cutoff and reduce redundancy

@@ -63,25 +63,38 @@ def parse_args():
         default=0.01,
     )
     parser.add_argument(
+        '--cutoff',
+        type=float,
+        help='upper bound for p_x* and p_y* (default .5)',
+        default=.5,
+    )
+    parser.add_argument(
         '--n-jobs',
         type=int,
         help='number of processes to use for empirically computing maxima',
         default=1,
     )
     parser.add_argument(
-        '--ancestral-state-conditions',
-        action="store_true",
-        help='plot conditions where \emptyset is most likely ancestral state (Fig. S2)',
+        '--plot-type',
+        type=str,
+        help="""
+             choices:
+             ancestral_state_conditions-- conditions where \emptyset is most likely ancestral state (Fig. S2),
+             analytic_inconsistency-- plot conditions where \emptyset is most likely ancestral state and \hat{w} is one analytically (Fig. 2),
+             joint_empirical-- plot empirical estimate of \hat{w} for joint inference (Fig. 3),
+             marginal_empirical-- plot empirical estimate of \hat{w} for marginal inference (Fig. S4),
+             """,
+        choices=(
+             'ancestral_state_conditions',
+             'analytic_inconsistency',
+             'joint_empirical',
+             'marginal_empirical',
+            ),
     )
     parser.add_argument(
-        '--empirical-parameter-estimate',
+        '--bias',
         action="store_true",
-        help='plot empirical estimate of \hat{w} for joint inference (Fig. 3)',
-    )
-    parser.add_argument(
-        '--marginal',
-        action="store_true",
-        help='plot empirical estimate of \hat{w} for marginal likelihood (Fig. S4)',
+        help='plot and print bias instead of parameter estimate (Fig. S3)',
     )
     parser.add_argument(
         '--in-pkl-name',
@@ -94,11 +107,6 @@ def parse_args():
         type=str,
         help='file name for where to save output for time saving',
         default='output.pkl',
-    )
-    parser.add_argument(
-        '--analytic-inconsistency',
[…]

### 2018-07-18 versions/response.v1.txt
go through TODOs and responses

@@
 Outline of revisions:
-1.
-2.
+1. generalized the results to arbitrary branch lengths and not restricting the top and bottom two branches to be equal
+2. expanded and clarified the exposition of the mathematics and the motivation for the Hadamard representation
+3. changed discussion and plots to involve branch lengths and not fidelities
 
@@
 
-As we now emphasize (TODO XXX) this work is not the final word on the joint inference question from an applied perspective.
+As we now emphasize, this work is not the final word on the joint inference question from an applied perspective.
+%das: added
+%Examples with larger trees and more realistic mutation models fall within the scope of an extension to applied models.
+%Further study is needed to determine the impact of these results on problems that are of interest to practitioners.
 Even if we were to greatly expand the sections doing numerical optimization, these results would only be for the binary symmetric model, which is never applied in practice.
@@
 We are grateful to reviewer 4 for pointing us to literature of which we were not aware.
-We have greatly expanded ...
+We have greatly expanded on the discussion in Goldman (1990) about the limitations of joint optimization of phylogenetic parameters and ancestral states.
+The previous work concerns an example where branch lengths are fixed and equal, while here we show cases where inconsistency arises when branch lengths are treated as unknown parameters.
+Additionally, we have placed joint optimization and maximum likelihood on level footing by viewing them both as optimization procedures over topologies and branch lengths.
+Goldman (1990) does not compare maximum likelihood and maximum relative likelihood as the latter assumes equal branch lengths while the former does not.
+We have added this to our discussion.
 
@@
 
-Thank you very much for reading the derivations and for your suggestions. As noted inline we have revised the text accordingly.
+Thank you very much for reading the derivations and for your suggestions.
+As noted inline we have revised the text accordingly.
 
@@
 
-This wasn't as clear as it should have been. We have revised to:
+This wasn't as clear as it should have been.
+We have revised to:
 "The mapping $\patternToSplit$ operates by returning the complement of a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if it is not in $\mathcal{P}(\siteSplitSet)$."
@@
 We believe the inaccuracies have more to do with the specific choice of which ancestral state is maximal than asymmetries in the tree.
-Asymmetries can drive these particular choices, but Fig. 3 in the revised manuscript shows that almost half of the parameter space results in branch lengths being estimated as zero.
+Asymmetries can drive these particular choices, but Fig. 3 in the revised manuscript shows that almost half of the parameter space results in an interior branch length being estimated as zero.
 We do not think providing an argument about accumulating inaccuracies would elucidate why such starkly-defined regions of inconsistency exist.

### 2018-07-18 joint_inf.tex
go through TODOs and responses

@@
 These regions of inconsistency arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
-%TODO: double check the below sentence
-This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and~3.
+This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and~3, though we show not all branches need to be long for inconsistency to arise.
 \citet{Sagulenko2018-xl} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:empirical-bias} shows even when branches are short there is a consistent bias.
@@
 All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
+Examples with larger trees and more realistic mutation models fall within the scope of an extension to applied models.
+Further study is needed to determine the impact of these results on problems that are of interest to practitioners.
 Given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.

### 2018-07-18 joint_inf_plot.py
messed up plot titles

@@ -466,14 +466,14 @@ def main(args=sys.argv[1:]):
         move_legend = True
     elif args.plot_type == 'joint_empirical':
         if args.bias:
-            plottitle = r'Relative bias: $(\hat{p}_w-p_{y^*})/p_{y^*}$'
+            plottitle = r'Bias: $\hat{p}_w-p_{y^*}$'
         else:
-            plottitle = r'Estimated $\hat{p}_w$ (marginal inference)'
+            plottitle = r'Estimated $\hat{p}_w$'
         scale = args.delta
         plotbar = True
     elif args.plot_type == 'marginal_empirical':
         if args.bias:
-            plottitle = r'Relative bias: $(\hat{p}_w-p_{y^*})/p_{y^*}$ (marginal inference)'
+            plottitle = r'Bias: $\hat{p}_w-p_{y^*}$ (marginal inference)'
         else:
             plottitle = r'Estimated $\hat{p}_w$ (marginal inference)'
         scale = args.delta

### 2018-07-18 joint_inf.tex
update bias plot/discussion

@@
 For joint inference, the estimates for $\hat{p}_w$ when $p_{x^*}$ and $p_{y^*}$ are both small look reasonable, with $\hat{p}_w$ increasing as $p_{y^*}$ increases, though Fig.~\ref{fig:empirical-bias} shows there is a systematic negative bias in this procedure even when the true branches are short.
-%TODO: does this bias make sense? should it be log-diff?
-To understand the quality of each fit, we report the range of $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.
+To understand the quality of each fit, we report the range of the bias $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.
 Due to the discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}), an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
-For joint inference, the errors given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for marginal inference, $[-3\times 10^{-17}, 3\times 10^{-17}]$ showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
+For joint inference, the errors given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for marginal inference are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
 

### 2018-07-23 joint_inf.tex
A quick pass through main text.

@@
 \centering
-\begin{subfigure}{.3\linewidth}
+\begin{subfigure}{.32\linewidth}
 \centering
@@
 \end{subfigure}
-\begin{subfigure}{.3\linewidth}
+\begin{subfigure}{.32\linewidth}
 \centering
@@
 \]
+%EM Could we give an extra sentence justifying that the limit of the inferential process will be the same as the inferences applied to the limiting data?
+Compare https://paperpile.com/shared/HohiVH in which they are rather different.
 We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
@@
 
+%EM Suggest "Our main theorem states that ... internal branch length zero ..."
 \begin{restatable}{theorem}{topoInconsist}
@@
 The proof is through analyzing upper and lower bounds of the potential likelihood forms in Table~\ref{tab:farris_likelihoods} and the gradients of the objective function $\ell_{\tau^*,t^*}(\tau_1, t)$ with respect to $t$.
+%EM You're doing some interpretation here, but biologists will skip any paragraph that starts with "proof".
+%EM If you are using "topology" to refer to the multifurcating tree structure, you need to clarify that.
 This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one), and the incorrect topology attains an equal or greater likelihood at its maximum than the true topology in the limit.
@@
 For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
-As $(1+w) > (1-w)$, a maximization procedure more likely introduces the $(1+w)$ term, though this will not be guaranteed as the maximum depends on the values of the unknown branch parameters $t$.
-This tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
+As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this will not be guaranteed because the maximum depends on the values of the unknown branch parameters $t$.
+Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
 This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split if we let $x_1=x_2=x^*$ and $y_1=y_2=y^*$ (Table~\ref{tab:likelihoods}).
@@
 
-Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:incons-analytic}).
+Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically\hyp{}derived region (Fig.~\ref{fig:incons-analytic}).
 To determine how conservative, we use the L-BFGS-B method to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
-We see that the region of inconsistency in the general case is much larger than in the analytic case---the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral state splits, so this is reasonable.
-This region encompasses almost half of the branch parameter space; given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch length to be zero.
+We see that the region of inconsistency in the general case is much larger than in the analytic case, which is unsurprising because the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral state splits.
+This region encompasses almost half of the branch fidelity space; given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch length to be zero.
 
+%EM Describe in a sentence how you can reduce things down to 81 maximizations.
 We provide a full description of our optimization procedure in the Appendix, but briefly, we perform 81 maximizations for each separate likelihood function in Table~\ref{tab:likelihoods}.
@@
 To understand the quality of each fit, we report the range of the bias $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.
+%EM I'd suggest bringing this figure into the main text. It'd fit.
 Due to the discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}), an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
@@
     Numerical estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}.
-    Data generated as in Fig.~\ref{fig:incons-analytic}
+    Data generated as in Fig.~\ref{fig:incons-analytic}.
+%EM This is fucking nuts.
+Nice work.
+Perhaps drive the point home by adding a sentence repeating that p_w is p_y and just saying the basic point that the entire white region represents inconsistent estimates?
 }
@@
 We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
-Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of inconsistency: the joint inference procedure estimates zero-length branches given the correct topology.
+Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of topological inconsistency: the joint inference procedure estimates zero-length branches given the correct topology.
+%EM Suggest "Considering this zero length branch as a multifurcating topology..."
 This results in an incorrect topology having a likelihood as large or greater than that of the generating topology.
 These regions of inconsistency arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
-This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and~3, though we show not all branches need to be long for inconsistency to arise.
+This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and~3,
+%EM I suggest flipping this around and making it stronger, so that we lead with and emphasize the differences.
+though we show not all branches need to be long for inconsistency to arise.
 \citet{Sagulenko2018-xl} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:empirical-bias} shows even when branches are short there is a consistent bias.
+%EM ...
+and that the bias is on the same order as the magnitude of the parameters.
 
@@
 The case of joint inference of a phylogenetic likelihood was discussed in \citet{Goldman1990-dk}.
-In it, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general.
+There, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general.
+%EM ...
+and indeed explicitly describes general consistency as an open problem.
+%EM Does the following sentence need revision after the proof updates?
 We show cases where the estimates of the topology are inconsistent if we are estimating both topology and branch lengths, and, moreover, even if we know the correct topology, we show cases where branch lengths are severely biased.
 Finally, just prior to his conclusion, he discusses when parsimony gives the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.
+%EM Two em-dash phrases in a row is too many.
 The joint inference procedure outlined here estimates the same parameters as classical maximum likelihood---albeit implicitly estimating ancestral states as well---rendering the question well-posed.

### 2018-07-25 joint_inf.tex
going thru comments

@@
 \end{equation}
-We use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}, and estimate a topology and branch fidelities using this profile likelihood as
+We use $\hat{\fullAncestralStates}_\nCols$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}, and estimate a topology and branch fidelities using this profile likelihood as
 \begin{equation}
 \label{eq:profile_likelihood_topology_bl}
-(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
+(\hat{\tau}_\nCols, \hat{t}_\nCols) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
 \end{equation}
 In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral state splits in Table~\ref{tab:farris_likelihoods} to obtain the likelihood term for each most likely ancestral state split, these terms depending on the unknown $(\tau, t)$.
-For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2018-xl}.
+For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}_\nCols$ for a fixed $(\tau,t)$, then $(\hat{\tau}_\nCols,\hat{t}_\nCols)$ given $\hat{\fullAncestralStates}_\nCols$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2018-xl}.
 
@@
 Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
-\[
+\begin{equation}
+\label{eq:exp-log-lik}
 \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t).
-\]
+\end{equation}
 %EM Could we give an extra sentence justifying that the limit of the inferential process will be the same as the inferences applied to the limiting data?
 Compare https://paperpile.com/shared/HohiVH in which they are rather different.
+%das: wow, good catch! This is not at all obvious to me it turns out :/ I guess I'll have to think harder on it
 We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
@@
 %EM Suggest "Our main theorem states that ... internal branch length zero ..."
+Our main theorem states that given data generated on $\tau_1$ there exist true nonzero branch lengths such that the estimator $\hat{t}$ maximizing \eqref{eq:exp-log-lik} has an internal branch of length zero.
 \begin{restatable}{theorem}{topoInconsist}
@@
 
-The proof is through analyzing upper and lower bounds of the potential likelihood forms in Table~\ref{tab:farris_likelihoods} and the gradients of the objective function $\ell_{\tau^*,t^*}(\tau_1, t)$ with respect to $t$.
 %EM You're doing some interpretation here, but biologists will skip any paragraph that starts with "proof".
 %EM If you are using "topology" to refer to the multifurcating tree structure, you need to clarify that.
-This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one), and the incorrect topology attains an equal or greater likelihood at its maximum than the true topology in the limit.
-For example, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t)$ and $\ell_{\tau^*,t^*}(\tau_2, t)$ are equivalent.
+This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one).
+As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology attains a likelihood value at its maximum equal to that of the true topology in the limit.
+In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t)$ and $\ell_{\tau^*,t^*}(\tau_2, t)$ are equivalent.
 We elaborate on this point in the Appendix.
+The proof is through analyzing upper and lower bounds of the potential likelihood forms in Table~\ref{tab:farris_likelihoods} and the gradients of the objective function $\ell_{\tau^*,t^*}(\tau_1, t)$ with respect to $t$.
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{empirical-estimate}
+\caption{
+    Numerical estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
+    Data generated as in Fig.~\ref{fig:incons-analytic}.
+    The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero, resulting in an inconsistency.
+%EM This is fucking nuts.
+Nice work.
+Perhaps drive the point home by adding a sentence repeating that p_w is p_y and just saying the basic point that the entire white region represents inconsistent estimates?
+}
+\label{fig:bl-general-inconsistency}
+\end{figure}
+
 %EM Describe in a sentence how you can reduce things down to 81 maximizations.
 We provide a full description of our optimization procedure in the Appendix, but briefly, we perform 81 maximizations for each separate likelihood function in Table~\ref{tab:likelihoods}.
+By the symmetry in the likelihood, we assume without loss of generality that $y_2 \ge y_1$ and $x_2 \ge x_1$, allowing us to consider far fewer potential ancestral state splits than the 3,888 possible given in Table~\ref{tab:farris_likelihoods}.
 For each function, we optimize the objective starting at five separate initial values to mitigate problems with small curvature of the objective surface and nonconcavity.
@@
 \centering
-\includegraphics[width=\textwidth]{empirical-estimate}
+\includegraphics[width=.95\textwidth]{empirical-bias}
 \caption{
-    Numerical estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}.
-    Data generated as in Fig.~\ref{fig:incons-analytic}.
-%EM This is fucking nuts.
-Nice work.
-Perhaps drive the point home by adding a sentence repeating that p_w is p_y and just saying the basic point that the entire white region represents inconsistent estimates?
+Bias in branch length estimation.
+In regions where joint optimization should perform well with $p_{x^*}, p_{y^*} \le .1$, there is systematic bias toward shorter branch lengths.
 }
-\label{fig:bl-general-inconsistency}
+\label{fig:empirical-bias}
 \end{figure}
@@
 %EM Suggest "Considering this zero length branch as a multifurcating topology..."
-This results in an incorrect topology having a likelihood as large or greater than that of the generating topology.
+Considering a tree with a zero-length branch as a multifurcating topology, this results in the incorrect topology having a likelihood as large or greater than that of the generating topology.
 These regions of inconsistency arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
-This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and~3,
+We see that this inconsistency occurs even if some branches are short.
+This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3).
 %EM I suggest flipping this around and making it stronger, so that we lead with and emphasize the differences.
-though we show not all branches need to be long for inconsistency to arise.
-\citet{Sagulenko2018-xl} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:empirical-bias} shows even when branches are short there is a consistent bias.
-%EM ...
-and that the bias is on the same order as the magnitude of the parameters.
+\citet{Sagulenko2018-xl} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:empirical-bias} shows even when all branches are short there is a consistent bias, and that the bias is on the same order as the magnitude of the parameters.
 
@@
 The case of joint inference of a phylogenetic likelihood was discussed in \citet{Goldman1990-dk}.
+%das: not so sure if your below edit accurately reflects Goldman; he says that, because parsimony (ie, joint inf) and ML estimate different parameters the question is not ineteresting:
+%"There has been considerable speculation
+%and analysis of when the above methods
+%give the same answer (e.g., Felsenstein,
+%1973b, 1981b, 1983a; Sober, 1983, 1985; Fel-
+%senstein and Sober, 1986; Friday, 1989).
+%Now that clear formulations are available
+%for each, this question in fact becomes less
+%interesting. The methods can now be
+%judged on the merits of their models as
+%well as on the results they produce; be-
+%cause they estimate different parameters,
+%the estimates cannot match precisely."
+%das: I don't remember reading anywhere him stating that general consistency is an open problem.
+Regarding general consistency, he says things like "results may be unpredictable" and "may also be inconsistent" and only provides a single example, but this doesn't seem strong enough to say he says it's an open problem.
+Thoughts?
 There, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general.
@@
 %EM Does the following sentence need revision after the proof updates?
-We show cases where the estimates of the topology are inconsistent if we are estimating both topology and branch lengths, and, moreover, even if we know the correct topology, we show cases where branch lengths are severely biased.
+We show cases where the incorrect topology attains an equal likelihood value at the maxium as the correct topology, and, moreover, if we know the correct topology, we show cases where branch lengths are severely biased.
 Finally, just prior to his conclusion, he discusses when parsimony gives the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.
 %EM Two em-dash phrases in a row is too many.
-The joint inference procedure outlined here estimates the same parameters as classical maximum likelihood---albeit implicitly estimating ancestral states as well---rendering the question well-posed.
-Finally, we are able to provide much more detail on how large branch lengths must be for general joint inference---estimating topology and branch lengths together---to fail to be consistent.
+We render the question well-posed: the joint inference procedure outlined here estimates the same parameters as classical maximum likelihood---topology and branch lengths---albeit implicitly estimating ancestral states as well.
+We are able to provide much more detail on how large branch lengths must be for general joint inference to fail to be consistent.
 

### 2018-07-25 joint_inf.tex
sketch of proof of inference limit ==> limit inference

@@
 \end{equation}
+We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
+
 %EM Could we give an extra sentence justifying that the limit of the inferential process will be the same as the inferences applied to the limiting data?
@@
 %das: wow, good catch! This is not at all obvious to me it turns out :/ I guess I'll have to think harder on it
-We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
+%das: update: I think I've got it, but I need to write it up a little more explicitly, and probably in the appendix.
+I think this approach maybe doesn't apply to your reference?
+But I'd like to work through it and see if that's the case, as a sanity check.
+Let
+\[
+\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t)
+\]
+and
+\[
+m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
+\]
+so that for a fixed $\tau$
+\[
+\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
+\]
+and
+\[
+\ell_{\tau^*,t^*}(\tau, t) = E[m_t].
+\]
+To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires the following:
+\begin{itemize}
+\item $\mathcal{T}$ is compact;
+\item $t \mapsto m_t(\alignmentColumn)$ is upper-semicontinuous for almost all $\alignmentColumn$;
+\item $\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)$ is measurable;
+\item $E[\sup_t m_t] < \infty$.
+\end{itemize}
+Point one is obviously true.
+Point two is true as well: for a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
+Discontinuities only occur when, if we move about in $\mathcal{T}$, a different ancestral state split becomes more likely.
+Maximizing over ancestral state splits ensures the upper-semicontinuous property of this mapping.
+For point three, we can think of $\sup_t m_t$ as fixing a maximal ancestral state split.
+Since $\alignmentColumn\in\alphabet^\nSiteRows$, the domain $\alphabet^\nSiteRows$ of the mapping $\alignmentColumn \mapsto m_t(\alignmentColumn)$ is a finite set, and so all subsets of the domain are also finite.
+This implies measurability of this mapping.
+Finally, $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, satisfying point four.
 

### 2018-07-25 appendix.tex
update bias plot/discussion
going thru comments
update notation

@@
 In this case, Table~\ref{tab:gen-sitepatprob} shows that, for $w=1$, the divergence term is identical for $\tau_1$ and $\tau_2$ as when the internal branch becomes zero-length the two topologies are indistinguishable.
-Let $\mathcal{T}$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}$ corresponds to $x^*$ and $y^*$ falling in the shaded region in Fig.~\ref{fig:incons-analytic}.
+Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the shaded region in Fig.~\ref{fig:incons-analytic}.
 The unconstrained maximum likelihood value for $\tau_2$ is greater than or equal to the value of the likelihood given these two constraints.
@@
 &\qquad\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\})\\
-&= \max_{t : t^*\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau_1, t),
+&= \max_{t : t^*\in\mathcal{T}_0} \ \ell_{\tau^*,t^*}(\tau_1, t),
 \end{align*}
@@
 % Figures mentioned in main section of manuscript
-
-\begin{figure}
-\centering
-\includegraphics[width=.95\textwidth]{empirical-bias}
-\caption{
-Bias in branch length estimation.
-In regions where joint optimization should perform well, there is systematic bias toward shorter branch lengths.
-}
-\label{fig:empirical-bias}
-\end{figure}
 

### 2018-07-25 appendix.tex
consistency graf

@@
 
+\subsection*{Convergence of branch parameters}
+
+%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
+%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
+%Also I don't think the compactness condition holds---tree space isn't compact is it?
+%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
+%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
+For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
+\[
+\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
+\]
+and
+\[
+\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t).
+\]
+Using the notation in \citet{van1998asymptotic}, we let
+\[
+m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
+\]
+so that
+\[
+\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
+\]
+and
+\[
+\ell_{\tau^*,t^*}(\tau, t) = E[m_t].
+\]
+To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
+The first is that $\mathcal{T}$ is compact, which is obviously true.
+The second is that
+\[
+E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
+\]
+and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
+The remaining conditions are on the maps
+\[
+\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
+\]
+and
+\[
+t \mapsto m_t(\alignmentColumn).
+\]
+We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
+Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
+For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
+Discontinuities occur when, if we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we are at a point of discontinuity we maximize over ancestral state splits, ensuring the upper-semicontinuous property of this mapping.
+This shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t)$.
+
 \subsection*{Properties of the joint objective function}

### 2018-07-25 joint_inf.tex
consistency graf

--- 2018-07-25 joint_inf.tex
@@
 \end{equation}
+We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
+For a fixed $\tau$, let $\hat{t}_\nCols$ maximize the left-hand side of \eqref{eq:exp-log-lik} and $\hat{t}$ maximize the right-hand side.
+We show in the Appendix that $\hat{t}_\nCols \rightarrow \hat{t}$, allowing us to focus on only the right-hand side above.
+
 %EM Could we give an extra sentence justifying that the limit of the inferential process will be the same as the inferences applied to the limiting data?
 Compare https://paperpile.com/shared/HohiVH in which they are rather different.
-%das: wow, good catch! This is not at all obvious to me it turns out :/ I guess I'll have to think harder on it
-We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
+%das: moved to appendix
 
@@
 %EM Suggest "Our main theorem states that ... internal branch length zero ..."
-Our main theorem states that given data generated on $\tau_1$ there exist true nonzero branch lengths such that the estimator $\hat{t}$ maximizing \eqref{eq:exp-log-lik} has an internal branch of length zero.
+Our main theorem states that given data generated on $\tau_1$ there exist true nonzero branch lengths such that the estimator $\hat{t}$ maximizing the right-hand side of \eqref{eq:exp-log-lik} has an internal branch of length zero.
 \begin{restatable}{theorem}{topoInconsist}

### 2018-07-26 joint_inf.tex
Main text looky.

@@
 
-%EM Could we give an extra sentence justifying that the limit of the inferential process will be the same as the inferences applied to the limiting data?
-Compare https://paperpile.com/shared/HohiVH in which they are rather different.
-%das: moved to appendix
-
 %We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology ($\tau_2$) than for the true, generating InvFels topology ($\tau_1$).
@@
 
-%EM Suggest "Our main theorem states that ... internal branch length zero ..."
 Our main theorem states that given data generated on $\tau_1$ there exist true nonzero branch lengths such that the estimator $\hat{t}$ maximizing the right-hand side of \eqref{eq:exp-log-lik} has an internal branch of length zero.
@@
 
-%EM You're doing some interpretation here, but biologists will skip any paragraph that starts with "proof".
-%EM If you are using "topology" to refer to the multifurcating tree structure, you need to clarify that.
 This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one).
@@
     The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero, resulting in an inconsistency.
-%EM This is fucking nuts.
-Nice work.
-Perhaps drive the point home by adding a sentence repeating that p_w is p_y and just saying the basic point that the entire white region represents inconsistent estimates?
 }
@@
 
-%EM Describe in a sentence how you can reduce things down to 81 maximizations.
-We provide a full description of our optimization procedure in the Appendix, but briefly, we perform 81 maximizations for each separate likelihood function in Table~\ref{tab:likelihoods}.
-By the symmetry in the likelihood, we assume without loss of generality that $y_2 \ge y_1$ and $x_2 \ge x_1$, allowing us to consider far fewer potential ancestral state splits than the 3,888 possible given in Table~\ref{tab:farris_likelihoods}.
+We provide a full description of our optimization procedure in the Appendix, but briefly, we analytically reduce the general case to 81 (see Table~\ref{tab:likelihoods}) and perform maximizations for each.
 For each function, we optimize the objective starting at five separate initial values to mitigate problems with small curvature of the objective surface and nonconcavity.
@@
 %EM Suggest "Considering this zero length branch as a multifurcating topology..."
+%EM Oops, somehow I misunderstood this the first time around.
+You're saying that the wrong topology, with internal branch shrunk to zero, will have the same likelihood.
+And that we can't exclude that it could have a greater likelihood with internal branch nonzero.
+If you want to make those points (or something about multifurcating) they need to be more explicit.
+Right now it reads like we found cases where the wrong topology has larger likelihood.
+%EM Seems like clarification might be justified in other places in the MS as well.
 Considering a tree with a zero-length branch as a multifurcating topology, this results in the incorrect topology having a likelihood as large or greater than that of the generating topology.
@@
 %EM I suggest flipping this around and making it stronger, so that we lead with and emphasize the differences.
-\citet{Sagulenko2018-xl} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:empirical-bias} shows even when all branches are short there is a consistent bias, and that the bias is on the same order as the magnitude of the parameters.
+%EM I've made this a little spicier: happy with the first part?
+However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and that the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias})
 
@@
 The case of joint inference of a phylogenetic likelihood was discussed in \citet{Goldman1990-dk}.
-%das: not so sure if your below edit accurately reflects Goldman; he says that, because parsimony (ie, joint inf) and ML estimate different parameters the question is not ineteresting:
-%"There has been considerable speculation
-%and analysis of when the above methods
-%give the same answer (e.g., Felsenstein,
-%1973b, 1981b, 1983a; Sober, 1983, 1985; Fel-
-%senstein and Sober, 1986; Friday, 1989).
-%Now that clear formulations are available
-%for each, this question in fact becomes less
-%interesting. The methods can now be
-%judged on the merits of their models as
-%well as on the results they produce; be-
-%cause they estimate different parameters,
-%the estimates cannot match precisely."
-%das: I don't remember reading anywhere him stating that general consistency is an open problem.
-Regarding general consistency, he says things like "results may be unpredictable" and "may also be inconsistent" and only provides a single example, but this doesn't seem strong enough to say he says it's an open problem.
-Thoughts?
 There, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general.
-%EM ...
-and indeed explicitly describes general consistency as an open problem.
-%EM Does the following sentence need revision after the proof updates?
-We show cases where the incorrect topology attains an equal likelihood value at the maxium as the correct topology, and, moreover, if we know the correct topology, we show cases where branch lengths are severely biased.
+We show cases where the incorrect topology attains an equal likelihood value at the maximum as the correct topology, and, moreover, if we know the correct topology, we show cases where branch lengths are severely biased.
 Finally, just prior to his conclusion, he discusses when parsimony gives the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.
-%EM Two em-dash phrases in a row is too many.
 We render the question well-posed: the joint inference procedure outlined here estimates the same parameters as classical maximum likelihood---topology and branch lengths---albeit implicitly estimating ancestral states as well.

### 2018-07-26 appendix.tex
Beginning to read appendix.

@@
 \]
+%EM Curious: is this an established use of the term "partial"?
 and the partial log-likelihood
@@
 Ancestral states with $^*$ are never maximal provided parameters are in $(0,1]$.
-There are $3^5\cdot 4^2=3,888$ possible forms for the likelihood.
+By combinations of $\ancestralSplit_k$, there are $3^5\cdot 4^2=3,888$ possible forms for the likelihood.
 }
@@
 Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
-We show that the likelihood remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
+We show that the likelihood $\ell_{\tau_1,t^*}(\tau_1, t)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
 Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
@@
 meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
+
+%EM It seems to me that the point here is that these swaps in the generating probabilities are matched by swaps in the likelihoods, meaning that things stay the same.
+A topic sentence to this effect would help if that's right.
 The corresponding possibilities for the likelihood values are
@@
 for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.
-
 The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
 The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
-Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t)$ we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$.
+Exchanging $x_1$ and $x_2$ also does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t)$.
+
+Thus, we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$.
 An analogous statement holds for $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.
@@
 Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
-There are $3^4=81$ possible forms for the likelihood.
+Because in 4 cases there are 3 possibilities for $\xi_j$, there are $3^4=81$ possible forms for the likelihood.
 }
@@
 \begin{proof}
+Recall
+\[
+\ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
+\]
 We obtain
@@
 \]
+%EM C_0 needs a better definition here. Too much going on.
 We compute the lower bound $C_0(x^*,y^*)$ above given $\emptyset$ as the maximal ancestral state split for all site splits and $\tau=\tau_1$.
 We need only consider the remaining $u=1,\ldots,80$ possibilities for choices of ancestral state splits in Table~\ref{tab:likelihoods} upon excluding cases of infinite branch lengths (i.e., any of $x_1,y_1,x_2,y_2,w$ equal to zero) and the redundant cases of $x_1 > x_2$ and $y_1 > y_2$.
+%EM Also suggest a real def for C_{1,u}.
 We compute upper bounds $C_{1,u}(x^*,y^*)$ and show in Fig.~\ref{fig:max-anc-state} the region where $C_0(x^*,y^*) > \max_u C_{1,u}(x^*,y^*)$.

### 2018-07-27 appendix.tex
a lot of notation and terminology; some clarification; addressing points

@@
 
-For the $i$th factor of \eqref{eq:full_likelihood},
+For the $i$th factor of \eqref{eq:full_likelihood}, we decompose the likelihood into its marginal and conditional components as
 \[
@@
 \end{equation}
-Define the divergence quantity
+Define the marginal log-likelihood
 \[
@@
 %EM Curious: is this an established use of the term "partial"?
-and the partial log-likelihood
+%das: not really; a partial likelihood is if the likelihood is separable in the parameters, so more accurate terms for what we're doing might be "marginal likelihood" and "conditional likelihood." I changed a lot throughout, using "integrated likelihood" so we don't confuse "marginal inference" with "marginal likelihood".
+and the conditional log-likelihood
 \[
-\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
+\ell^C_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
 \]
-so that \eqref{eq:site_pattern_profile_likelihood_mean} is
+with $\boldsymbol\xi = [\xi_j]_{j=1}^\nSiteSplits$ so that \eqref{eq:site_pattern_profile_likelihood_mean} is
 \begin{equation}
     \label{eq:log_likelihood_simplified}
-    \ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
+    \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^C_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
 \end{equation}
@@
 \[
-\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t).
+\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
 \]
@@
 \[
-\ell_{\tau^*,t^*}(\tau, t) = E[m_t].
+\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
 \]
@@
 Discontinuities occur when, if we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we are at a point of discontinuity we maximize over ancestral state splits, ensuring the upper-semicontinuous property of this mapping.
-This shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t)$.
+This shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.
 
@@
 Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
-We show that the likelihood $\ell_{\tau_1,t^*}(\tau_1, t)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
+We show that the likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
 Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
@@
 and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
+
 All other generating probabilities differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
@@
 meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
-
+We show that for site splits $\{1\}$ and $\{3\}$, exchanging $x_1$ and $x_2$ also swaps the values of the conditional likelihood terms.
 %EM It seems to me that the point here is that these swaps in the generating probabilities are matched by swaps in the likelihoods, meaning that things stay the same.
 A topic sentence to this effect would help if that's right.
-The corresponding possibilities for the likelihood values are
+%das: updated
+The corresponding possibilities for the conditional likelihood values are
 \begin{align*}
@@
 \end{align*}
-for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.
+for site split $\{3\}$, which shows the likelihood remains unchanged if $x_1$ and $x_2$ are swapped.
 The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
 The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
-Exchanging $x_1$ and $x_2$ also does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t)$.
+Exchanging $x_1$ and $x_2$ also does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$.
 
@@
 \[
-\ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
+\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^C_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
 \]
@@
 \[
-\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \le
+\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) \le
     \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
-    + \tilde{\ell}_{\tau^*,t^*}(\tau, \tilde{t})
+    + \ell^C_{\tau^*,t^*}(\tau, \tilde{t}; \boldsymbol\xi)
 \]
@@
 \[
-\tilde{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t).
+\tilde{t} = \argmax_{t} \ \ell^C_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
 \]
@@
 \[
-\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \ge
+\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) \ge
     \shannonDivergence_{\tau^*,t^*}(\tau,\tilde{t})
-    + \tilde{\ell}_{\tau^*,t^*}(\tau, \tilde{t}).
+    + \ell^C_{\tau^*,t^*}(\tau, \tilde{t}; \boldsymbol\xi).
 \]
 %EM C_0 needs a better definition here. Too much going on.
-We compute the lower bound $C_0(x^*,y^*)$ above given $\emptyset$ as the maximal ancestral state split for all site splits and $\tau=\tau_1$.
-We need only consider the remaining $u=1,\ldots,80$ possibilities for choices of ancestral state splits in Table~\ref{tab:likelihoods} upon excluding cases of infinite branch lengths (i.e., any of $x_1,y_1,x_2,y_2,w$ equal to zero) and the redundant cases of $x_1 > x_2$ and $y_1 > y_2$.
+%das: :check:
+Let $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$ and
+\[
+\tilde{t}_{\boldsymbol\xi} = \argmax_{t} \ \ell^C_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi).
+\]
+Define
+\[
+C_0(x^*, y^*) :=
+    \shannonDivergence_{\tau^*,t^*}(\tau_1,\tilde{t}_{\boldsymbol\xi_0})
+    + \ell^C_{\tau^*,t^*}(\tau_1, \tilde{t}_{\boldsymbol\xi_0}; \boldsymbol\xi_0),
+\]
+i.e., the lower bound given $\emptyset$ as the maximal ancestral state split for all site splits and $\tau=\tau_1$.
+
+We need only consider the remaining possibilities for choices of ancestral state splits in Table~\ref{tab:likelihoods} upon excluding cases of infinite branch lengths (i.e., any of $x_1,y_1,x_2,y_2,w$ equal to zero) and the redundant cases of $x_1 > x_2$ and $y_1 > y_2$, i.e.,
+\[
+C_1(x^*, y^*) :=
+    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
+    +  \max_{\boldsymbol\xi : \boldsymbol\xi \neq \boldsymbol\xi_0} \ \ell^C_{\tau^*,t^*}(\tau_1, \tilde{t}_{\boldsymbol\xi}; \boldsymbol\xi).
+\]
 %EM Also suggest a real def for C_{1,u}.
-We compute upper bounds $C_{1,u}(x^*,y^*)$ and show in Fig.~\ref{fig:max-anc-state} the region where $C_0(x^*,y^*) > \max_u C_{1,u}(x^*,y^*)$.
+%das: :check:
+We show in Fig.~\ref{fig:max-anc-state} the region where $C_0(x^*,y^*) > C_1(x^*,y^*)$.
 \end{proof}
@@
 Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
-For $0 < x^*, y^* < 1$ and $\emptyset$ as the maximal ancestral state split for all site splits, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
+For $0 < x^*, y^* < 1$ and $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
 \[
-\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t)
+\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi_0)
 \]
@@
 We proceed by analyzing the gradients of each component of the likelihood function.
-We first focus on the divergence term
+We first focus on the marginal log-likelihood
 \[
@@
 for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
-Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the divergence, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
+Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
 Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
@@
 \]
-We now show the gradient of the divergence term is a decreasing function on $(0,1]$ in each variable.
+We now show the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
 We focus on $x_1$.
-The divergence proportional to $x_1$ can be written as
+The marginal log-likelihood proportional to $x_1$ can be written as
 \begin{align*}
@@
 \begin{align*}
-\shannonDivergence'(x_1) := \frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}
+\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}
 \end{align*}
@@
 
-We now focus on the partial likelihood term.
+We now focus on the conditional log-likelihood.
 Assume $\emptyset$ is the most likely ancestral state split for all site splits.
-In this case, the partial likelihood has no ambiguity, and has terms given in Table~\ref{tab:likelihoods-restricted}.
-The partial likelihood
+In this case, the conditional log-likelihood has no ambiguity, and has terms given in Table~\ref{tab:likelihoods-restricted}.
+The conditional log-likelihood
 \[
-\tilde{\ell}_{\tau^*,t^*}(\tau_1, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, t)
+\ell^C_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi_0) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV = \siteSplit_j, \tau_1, t)
 \]
-as a function of $x_1$, using Table~\ref{tab:gen-sitepatprob}, is
+as a function of $x_1$, using Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:likelihoods-restricted}, is
 \[
-\tilde{\ell}(x_1) \propto \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1).
+\ell^C(x_1; \boldsymbol\xi_0) \propto \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1).
 \]
@@
 \[
-\tilde{\ell}'(x_1) := \frac{d}{dx_1} \tilde{\ell}(x_1) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2}
+\frac{d}{dx_1} \ell^C(x_1; \boldsymbol\xi_0) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2}
 \]
@@
 We show this is a type of inconsistency in topology estimation by constructing a lower bound on the likelihood for the Felsenstein topology $\tau_2$ and demonstrating it is equal to the likelihood given the InvFels topology $\tau_1$ in the demarcated region of Fig.~\ref{fig:incons-analytic}.
-As we are interested in a lower bound, assume $\xi_j=\emptyset$ for all $j=1,\ldots,\nSiteSplits$.
-In this case, the partial likelihood term for $\tau_2$ is composed of the same functional forms given in Table~\ref{tab:likelihoods-restricted}, as assuming $\emptyset$ for the ancestral state split results in the same partial likelihood for both topologies.
+As we are interested in a lower bound, assume $\boldsymbol\xi=\boldsymbol\xi_0$, i.e., $\emptyset$ is the maximal ancestral state split for all site splits.
+In this case, the conditional log-likelihood for $\tau_2$ is composed of the same functional forms given in Table~\ref{tab:likelihoods-restricted}, as assuming $\emptyset$ for the ancestral state split results in the same conditional log-likelihood for both topologies.
 
 Assume $w=1$.
-In this case, Table~\ref{tab:gen-sitepatprob} shows that, for $w=1$, the divergence term is identical for $\tau_1$ and $\tau_2$ as when the internal branch becomes zero-length the two topologies are indistinguishable.
+In this case, Table~\ref{tab:gen-sitepatprob} shows that, for $w=1$, the marginal log-likelihood is identical for $\tau_1$ and $\tau_2$ as when the internal branch becomes zero-length the two topologies are indistinguishable.
 Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the shaded region in Fig.~\ref{fig:incons-analytic}.
@@
 \begin{align*}
-\max_{t} & \ \ell_{\tau^*,t^*}(\tau_2, t) \ge \\
-&\max_{t: \xi_j=\emptyset, w=1, \tau=\tau_2} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}) \\
+\max_{t} & \ \ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi) \ge \\
+&\max_{t: \boldsymbol\xi=\boldsymbol\xi_0, w=1, \tau=\tau_2} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}) \\
 &\qquad\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\}) \\
-&= \max_{t: \xi_j=\emptyset, w=1, \tau=\tau_1} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
+&= \max_{t: \boldsymbol\xi=\boldsymbol\xi_0, w=1, \tau=\tau_1} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
 &\qquad\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\})\\
-&= \max_{t : t^*\in\mathcal{T}_0} \ \ell_{\tau^*,t^*}(\tau_1, t),
+&= \max_{t : t^*\in\mathcal{T}_0} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi),
 \end{align*}
@@
 \caption{
-    Estimates for $\hat{p}_w$ when computing $(\hat{x}_1, \hat{y}_1, \hat{x}_2, \hat{y}_2, \hat{w})$ using L-BFGS-B optimizing the classical marginal likelihood \eqref{eq:marginal_likelihood} rather than a joint optimization procedure.
+    Estimates for $\hat{p}_w$ when computing $(\hat{x}_1, \hat{y}_1, \hat{x}_2, \hat{y}_2, \hat{w})$ using L-BFGS-B optimizing the classical integrated likelihood \eqref{eq:marginal_likelihood} rather than a joint optimization procedure.
 }

### 2018-07-27 joint_inf.tex
a lot of notation and terminology; some clarification; addressing points

@@
 
-\newcommand{\shannonDivergence}{D}
+\newcommand{\shannonDivergence}{\ell^M}
 
@@
 
-Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out latent ancestral states at the internal nodes of the tree.
+Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out latent ancestral states at the internal nodes of the tree, obtaining an integrated likelihood \citep{Goldman1990-dk}.
 In a recent paper, \citet{Sagulenko2018-xl} suggest using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences on a fixed topology.
 This is attractive from a computational perspective: such joint inference can proceed according to an iterative procedure in which ancestral sequences are first estimated and model parameters are optimized conditional on these estimates.
-This latter conditional optimization is simpler and more computationally efficient than optimizing the marginal likelihood.
+This latter conditional optimization is simpler and more computationally efficient than optimizing the integrated likelihood.
 But is it statistically consistent?
@@
 Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ (Fig.~\ref{fig:farris-fels-top}a).
-Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
+Let $\boldsymbol\xi = [\xi_j]_{j=1}^\nSiteSplits$ be the vector of most likely ancestral state splits.
+Use $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
 \begin{equation}
 \label{eq:exp-log-lik}
-\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t).
+\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
 \end{equation}
@@
 Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
-There exists a set of $0 < x^*, y^* < 1$ such that the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
+For $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$ there exists a set of $0 < x^*, y^* < 1$ such that the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
 \[
-\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t)
+\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi_0)
 \]
@@
 As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology attains a likelihood value at its maximum equal to that of the true topology in the limit.
-In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t)$ and $\ell_{\tau^*,t^*}(\tau_2, t)$ are equivalent.
+In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
 We elaborate on this point in the Appendix.
-The proof is through analyzing upper and lower bounds of the potential likelihood forms in Table~\ref{tab:farris_likelihoods} and the gradients of the objective function $\ell_{\tau^*,t^*}(\tau_1, t)$ with respect to $t$.
+The proof is through analyzing upper and lower bounds of the potential likelihood forms in Table~\ref{tab:farris_likelihoods} and the gradients of the objective function $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ with respect to $t$.
 
@@
 For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral state splits.
-For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
+For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest conditional log-likelihood.
 As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this will not be guaranteed because the maximum depends on the values of the unknown branch parameters $t$.
@@
 %EM I'd suggest bringing this figure into the main text. It'd fit.
+%das: :check:
 Due to the discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}), an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
-For joint inference, the errors given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for marginal inference are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
+For joint inference, the errors given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for inference on the integrated likelihood are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
 
@@
 %EM Seems like clarification might be justified in other places in the MS as well.
-Considering a tree with a zero-length branch as a multifurcating topology, this results in the incorrect topology having a likelihood as large or greater than that of the generating topology.
-These regions of inconsistency arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
+%das: maybe this is better?
+Considering a tree with a zero-length branch as a multifurcating topology, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
+Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length will have a greater likelihood, though we have not found regions where this is the case.
+The regions of inconsistency we found arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
 We see that this inconsistency occurs even if some branches are short.
@@
 %EM I've made this a little spicier: happy with the first part?
+%das: I like it
 However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and that the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias})

### 2018-07-27 joint_inf_plot.py

@@ -473,9 +473,9 @@ def main(args=sys.argv[1:]):
         plotbar = True
     elif args.plot_type == 'marginal_empirical':
         if args.bias:
-            plottitle = r'Bias: $\hat{p}_w-p_{y^*}$ (marginal inference)'
+            plottitle = r'Bias: $\hat{p}_w-p_{y^*}$ (integrated likelihood)'
         else:
-            plottitle = r'Estimated $\hat{p}_w$ (marginal inference)'
+            plottitle = r'Estimated $\hat{p}_w$ (integrated likelihood)'
         scale = args.delta
         plotbar = True
     elif args.plot_type == 'analytic_inconsistency':

### 2018-07-27 joint_inf.tex
hedging and some tense changes

@@
 Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
-Recently, a computationally-efficient approach was proposed to jointly maximize over these unknown states and phylogenetic parameters.
+Recently, a computationally-efficient approach proposing to jointly maximize over these unknown states and phylogenetic parameters.
 Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
 We show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
-We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates an incorrect topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
+We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
 For branch length estimation on the correct topology, we outline cases where branch length estimates are consistently and heavily biased and provide extensive empirical results for this setting.
@@
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model with data being generated on a four-taxon tree and we construct bounds on the joint objective function to demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
+To do so, we use a binary symmetric model with data being generated on a four-taxon tree and we construct bounds on the joint objective function to demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
 We find similar areas through empirical means where joint inference consistently underestimates interior branch lengths when the topology is known and fixed.
@@
 \begin{restatable}{theorem}{topoInconsist}
+\label{thm:main_theorem}
 Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
@@
 This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one).
-As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology attains a likelihood value at its maximum equal to that of the true topology in the limit.
+As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
 In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
@@
 For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest conditional log-likelihood.
-As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this will not be guaranteed because the maximum depends on the values of the unknown branch parameters $t$.
+As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this is not guaranteed because the maximum depends on the values of the unknown branch parameters $t$.
 Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
 This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split if we let $x_1=x_2=x^*$ and $y_1=y_2=y^*$ (Table~\ref{tab:likelihoods}).
-If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology, as the true $y^*\neq 1$.
+If we allow multifurcating trees in our inference, then we can think of this as an instance of converging to the wrong topology, as the true $y^*\neq 1$.
 
@@
 We see that the region of inconsistency in the general case is much larger than in the analytic case, which is unsurprising because the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral state splits.
-This region encompasses almost half of the branch fidelity space; given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch length to be zero.
+This region encompasses almost half of the branch fidelity space; given the correct topology and performing our best possible optimization, we have many situations where we estimate the interior branch length to be zero.
 
@@
 
-As data were generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
+As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
 Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same optimization code.
@@
 %das: :check:
-Due to the discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}), an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
+There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal.
+Because of these discontinuities, an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
 For joint inference, the errors given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for inference on the integrated likelihood are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
@@
 Right now it reads like we found cases where the wrong topology has larger likelihood.
+%das: maybe this is better?
 %EM Seems like clarification might be justified in other places in the MS as well.
-%das: maybe this is better?
+%das: also hedged a bit in the remainder of the MS.
 Considering a tree with a zero-length branch as a multifurcating topology, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
-Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length will have a greater likelihood, though we have not found regions where this is the case.
+Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length is more likely to be observed, though we have not found regions where this is the case.
 The regions of inconsistency we found arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
@@
 
-The case of joint inference of a phylogenetic likelihood was discussed in \citet{Goldman1990-dk}.
+The case of joint inference of a phylogenetic likelihood is discussed in \citet{Goldman1990-dk}.
 There, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general.
@@
 
-While we have shown inconsistency in topology estimation and investigated the performance of branch parameter estimation, there is substantial scope for future work to make these results more precise and more general.
-The techniques used to obtain upper and lower bounds for the likelihoods in the topology estimation case provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
-Similarly, for the general case in estimating branch lengths, we were only able to provide empirical estimates for regions of overestimation, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) beg further investigation.
+We have shown an inconsistency when performing joint inference on branch lengths given an InvFels topology and investigated the performance of branch parameter estimation.
+Our aim is to analytically demonstrate joint inference is inconsistent, and, while this indicates an issue with this inference procedure, there is substantial scope for future work to make these results more precise and more general.
+The techniques used to obtain upper and lower bounds for the likelihoods in Lemma~\ref{lemma:bounds} provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
+Similarly, for the general case in estimating branch lengths, we are only able to provide empirical estimates for regions of overestimation, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) beg further investigation.
 Empirical validation shows that the region of inconsistency is much larger than the one analytically derived.

### 2018-07-27 versions/response.v1.txt
go thru all responses; add quotes from edits inline in response

@@
 
-As we now emphasize, this work is not the final word on the joint inference question from an applied perspective.
 %das: added
-%Examples with larger trees and more realistic mutation models fall within the scope of an extension to applied models.
-%Further study is needed to determine the impact of these results on problems that are of interest to practitioners.
+This work is not the final word on the joint inference question from an applied perspective, as we now emphasize with an expanded Discussion section as well as the addition of these lines:
+"Examples with larger trees and more realistic mutation models fall within the scope of an extension to applied models.
+Further study is needed to determine the impact of these results on problems that are of interest to practitioners."
+
 Even if we were to greatly expand the sections doing numerical optimization, these results would only be for the binary symmetric model, which is never applied in practice.
 An investigation of these aspects deserves another paper.
+We have added clarification of these points to the discussion:
+"All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
+Examples with larger trees and more realistic mutation models fall within the scope of an extension to applied models.
+Further study is needed to determine the impact of these results on problems that are of interest to practitioners."
 
@@
 We are grateful to reviewer 4 for pointing us to literature of which we were not aware.
-We have greatly expanded on the discussion in Goldman (1990) about the limitations of joint optimization of phylogenetic parameters and ancestral states.
+We have greatly expanded on the discussion in Goldman (1990) about the limitations of joint optimization of phylogenetic parameters and ancestral states (the newly added fourth paragraph of the Discussion section), and have chosen to use his terminology of "integrated likelihood" to avoid confusion.
 The previous work concerns an example where branch lengths are fixed and equal, while here we show cases where inconsistency arises when branch lengths are treated as unknown parameters.
 Additionally, we have placed joint optimization and maximum likelihood on level footing by viewing them both as optimization procedures over topologies and branch lengths.
-Goldman (1990) does not compare maximum likelihood and maximum relative likelihood as the latter assumes equal branch lengths while the former does not.
-We have added this to our discussion.
+Goldman (1990) does not compare maximum integrated likelihood and maximum relative likelihood as the latter assumes equal branch lengths while the former does not.
+We have added this to our discussion as well.
 
-We have complied with this request regarding terminology, despite it going against current practice.
+We have complied with the request regarding the use of "Farris tree," despite it going against current practice.
 Indeed, Felsenstein's book has a subsection entitled "The Farris zone" spanning two pages (pp 120-121), a figure entitled "A tree in the Farris zone.", and a table with a caption "Properties of the Felsenstein and Farris zones...".
@@
 A true investigation into the practical implications of our work will require another paper.
+We have clarified this in the Discussion section with:
+"Our aim is to analytically demonstrate joint inference is inconsistent, and, while this indicates an issue with this inference procedure, there is substantial scope for future work to make these results more precise and more general.
+The techniques used to obtain upper and lower bounds for the likelihoods in Lemma 3 provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation."
 
@@
 
-TODO: not sure what this is?
-%das: this reviewer was confused about the phrase "generating probabilities," thinking we were generating the probabilities and not that they were the probabilities of generating a particular data configuration.
-We have added the definition of a "generating probability" to ease confusion in this sentence.
+We have added the definition of a "generating probability" to ease confusion in this sentence:
+"As data becomes plentiful, we use the Hadamard transform (see (9) in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form."
 
@@
 
-We have made this more precise and directly referenced the inequalities.
+We have made this more precise and directly referenced the inequalities:
+"In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral state splits in Table~\ref{tab:farris_likelihoods} to obtain the likelihood term for each most likely ancestral state split, these terms depending on the unknown $(\tau, t)$.
+For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}_\nCols$ for a fixed $(\tau,t)$, then $(\hat{\tau}_\nCols,\hat{t}_\nCols)$ given $\hat{\fullAncestralStates}_\nCols$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2018-xl}."
 
@@
 
-That is correct, and we've added parenthetical notes to reminde the reader, thanks.
+Thank you for highlighting this omission.
+You are correct, and we have added parenthetical notes to remind the reader where appropriate.
 
@@
 
-We have reformulated this sentence to avoid the comparison.
+We have reformulated this sentence to avoid the comparison:
+"We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with an alternate formulation called "fidelity.""
 
@@
 %das: I believe this is clear now; lmk
-We have clarified the intuition in this sentence.
+We have clarified the intuition in this sentence:
+"For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest conditional log-likelihood.
+As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this is not guaranteed because the maximum depends on the values of the unknown branch parameters $t$.
+Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth."
 
@@
 We have added clarifying sentence:
-"As data were generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$."
+"As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$."
 
@@
 Fixed via simpler language.
+We removed the sentence following the semicolon entirely, as we feel it did not add anything:
+"Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$."
 
@@
 
-This wasn't as clear as it should have been.
+We apologize for the confusion, as this was not as clear as it should have been.
 We have revised to:
-"The mapping $\patternToSplit$ operates by returning the complement of a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if it is not in $\mathcal{P}(\siteSplitSet)$."
+"The mapping $\patternToSplit$ operates by returning the tips labeled as $1$ in a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if the set of tips labeled $1$ is not in $\mathcal{P}(\siteSplitSet)$.
+The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if the set of tips labeled $1$ in $\alignmentColumn$ is in $\mathcal{P}(\siteSplitSet)$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the set of tips labeled $1$ in $\ancestralStateColumn$; otherwise, the set of tips labeled $1$ in $\overline{\alignmentColumn}$ necessarily is in $\mathcal{P}(\siteSplitSet)$."
+
+Additionally, we have explicitly defined psi(Y) and xi(Y,H) prior to this clarifying sentence.
 
@@
 
-We have made this explicit now, with two additional intermediate steps and an explicit explanation in terms of collapsing terms.
+We have made this explicit now, with two additional intermediate steps and an explicit explanation in terms of collapsing terms:
+"For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
+This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,\nSiteRows}\neq 1$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
+The same can be done for the case of $\alignmentColumn_{i,\nSiteRows}=1$, implying $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective."
 
@@
 
-We have updated the text to explicitly define "generating probability." Background for equation (8) can be found in Section 8.6 in the Semple and Steel text.
-This is referenced at the beginning of the section.
+We have updated the text to explicitly define "generating probability."
+"As data becomes plentiful, we use the Hadamard transform (see (9) in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form."
+
+Background for equation (8) can be found in Section 8.6 in the Semple and Steel text.
+This is referenced at the beginning of the section:
+"We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}."
+
 We do not believe an exhaustive explanation would provide any benefit to the reader that they could not get from referring to the text.
@@
 TODO we should make some changes to at least say that we've further clarified the implications of our results.
-%das: how much more should we say than this sentence (which we already say)? We already quote them as saying they should only use it when branches are short.
-%Although the software described in \citet{Sagulenko2017-jo} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
-%and
-%In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.
-%This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2017-jo}, namely their Figures~2 and 3.
+%das: I believe this is done
+We clarify our aim and warnings from Sagulenko et al. (2017) in the Introduction:
+"Although the software described in \citet{Sagulenko2018-xl} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
+In particular, one would like to know when this approximate technique breaks down for both topology and branch length inference, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference."
+
+Additionally, we refer to how our work extends that from Sagulenko et al. (2017) in the Discussion:
+"The regions of inconsistency we found arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
+We see that this inconsistency occurs even if some branches are short.
+This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3)."
 
@@
 This is an interesting suggestion but is different than our goal, which was to establish inconsistency.
-Moreover, in the revised manuscript we see cases where assigning ML ancestral states results in longer branches, not shorter ones (Lemma 3 in the updated paper).
-We believe the inaccuracies have more to do with the specific choice of which ancestral state is maximal than asymmetries in the tree.
+Moreover, in the revised manuscript we see cases where assigning ML ancestral states results in longer branches, not shorter ones (Lemma 3).
+We believe the inaccuracies have more to do with the specific choice of which ancestral state is maximal than asymmetries in the tree:
+"This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split if we let $x_1=x_2=x^*$ and $y_1=y_2=y^*$ (Table~\ref{tab:likelihoods})."
+
 Asymmetries can drive these particular choices, but Fig. 3 in the revised manuscript shows that almost half of the parameter space results in an interior branch length being estimated as zero.
@@
 TODO refer to idea of applied paper again?
+%das: I believe these are done as well
 
@@
 As shown in Goldman (1990), with all branch lengths fixed and equal the joint inference procedure degenerates into parsimony, which has well known inconsistency issues.
-We have added discussion of this point.
+We have added discussion of this point:
+"The case of joint inference of a phylogenetic likelihood is discussed in \citet{Goldman1990-dk}.
+There, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general."
 
@@
 As the true generating parameters change, we become more likely to choose a different ancestral state as most likely, resulting in an entirely different objective function to be optimized.
+We have added the following:
+"There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal."
+
 The analytical bound is a coarse estimate of where inconsistency arises.
@@
 TODO we should do some change, but make it clear that it's consistently biased even for short branches.
-%das: I added this line:
+%das: I believe this is done
 We have added the line:
-\citet{Sagulenko2017-jo} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:bl-general-bias} shows even when branches are short there is a consistent bias.
+"Although the software described in \citet{Sagulenko2018-xl} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference."
 
@@
 We were not aware of this paper and now have a paragraph describing it and comparing to our work.
-It's also a good idea to cite additional papers about identifiability, and we have done so.
+See the fourth paragraph of the Discussion section.
+Additionally, we have used his terminology of "integrated likelihood" to avoid confusion.
+
+It's also a good idea to cite additional papers about identifiability, and we have done so in the Introduction:
+"An estimator is said to be statistically consistent if it converges to the generating model with probability one in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood."
+and Discussion sections:
+"Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood."
 

### 2018-07-30 appendix.tex
hedging and some tense changes
A handful of appendix changes.

@@
 Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
-The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
+The mapping from characters to splits in this case depends on the characters at the tips and the ancestral states.
 For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
@@
 \begin{lemma}
+\label{lemma:bounds}
 Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
@@
 for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
+We will show that this implies $t=t^*$.
 Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
@@
 implying $w=y^*$.
-The unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and
+Thus the unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and
 \[
@@
 We focus on $x_1$.
+%EM I think you mean the divergence term as a function of x_1 is proportional to...
 The marginal log-likelihood proportional to $x_1$ can be written as
 \begin{align*}
+%EM This is a little funky-- I think this is a definition of D(x_1)?
 \shannonDivergence(x_1) \propto \sum_j p_{\siteSplit_j} \log(c_{\siteSplit_j} + a_{\siteSplit_j}x_1)
@@
 \]
-which greater than or equal to zero as it is a probability.
+which, being a probability, is greater than or equal to zero.
 Thus, regardless of $a_{\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{\siteSplit_j} + a_{\siteSplit_j}x_1 \ge 0$.
@@
 
+%das: @EM did you say you wanted to get rid of these paragraphs? we never show there is a case where Fels > InvFels, just where it's equal...
 As the region in Fig.~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.

### 2018-07-31 appendix.tex
Cleaning out addressed comments.
Getting to a point of confusion.

@@
     \begin{array}{ll}
-        \{i'\in\{1,\ldots,\nSiteRows-1\}: \alignmentColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}\neq 1,\\
+        \{i'\in\{1,\ldots,\nSiteRows-1\}: \alignmentColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=0,\\
         \{i'\in\{1,\ldots,\nSiteRows-1\}: \overline{\alignmentColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1,
@@
     \begin{array}{ll}
-        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \ancestralStateColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}\neq 1,\\
+        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \ancestralStateColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=0,\\
         \{i'\in\{1,\ldots,\nAncestralStateRows\}: \overline{\ancestralStateColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1.
@@
 For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
-This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,\nSiteRows}\neq 1$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
+This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,\nSiteRows}=0$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
 The same can be done for the case of $\alignmentColumn_{i,\nSiteRows}=1$, implying $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective.
@@
 \]
-%EM Curious: is this an established use of the term "partial"?
-%das: not really; a partial likelihood is if the likelihood is separable in the parameters, so more accurate terms for what we're doing might be "marginal likelihood" and "conditional likelihood." I changed a lot throughout, using "integrated likelihood" so we don't confuse "marginal inference" with "marginal likelihood".
 and the conditional log-likelihood
@@
 For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
+%EM Do discontinuities occur? In my mind we are taking the maximum of continuous functions, which is continuous.
 Discontinuities occur when, if we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we are at a point of discontinuity we maximize over ancestral state splits, ensuring the upper-semicontinuous property of this mapping.
@@
 We show that for site splits $\{1\}$ and $\{3\}$, exchanging $x_1$ and $x_2$ also swaps the values of the conditional likelihood terms.
-%EM It seems to me that the point here is that these swaps in the generating probabilities are matched by swaps in the likelihoods, meaning that things stay the same.
-A topic sentence to this effect would help if that's right.
-%das: updated
 The corresponding possibilities for the conditional likelihood values are
@@
 \]
-%EM C_0 needs a better definition here. Too much going on.
-%das: :check:
 Let $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$ and
@@
 \]
-i.e., the lower bound given $\emptyset$ as the maximal ancestral state split for all site splits and $\tau=\tau_1$.
+i.e., the lower bound assuming $\emptyset$ is the maximal ancestral state split for all site splits and $\tau=\tau_1$.
 
@@
 \]
-%EM Also suggest a real def for C_{1,u}.
-%das: :check:
 We show in Fig.~\ref{fig:max-anc-state} the region where $C_0(x^*,y^*) > C_1(x^*,y^*)$.
@@
 \caption{
-Region where $\emptyset$ is maximal ancestral state split for each site split on the InvFels tree $\tau_1$.
+Region where $\emptyset$ is the maximal ancestral state split for each site split on the InvFels tree $\tau_1$.
 }
@@
 \]
-By Gibbs inequality, this function attains a unique maximum when
+By Gibbs' inequality, this function attains a unique maximum when
 \[
@@
 implying $w=y^*$.
-Thus the unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and
+Thus the unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and consequently
 \[
@@
 where $c_{\siteSplit_j}$ and $a_{\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
-For example,
+For example, reading from Table~\ref{tab:gen-sitepatprob},
 \[
@@
 \[
-f_j(x_1) = p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}.
+f_j(x_1) := p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}.
 \]
-We see
+Recall that
 \[
-\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) := \frac{1}{8} (c_{\siteSplit_j} + a_{\siteSplit_j}x_1),
+\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{\siteSplit_j} + a_{\siteSplit_j}x_1),
 \]
@@
 \caption{
-The maximal likelihood values on the InvFels tree $\tau_1$ given $\emptyset$ is the most likely ancestral state split for each site split.
+The maximal likelihood values on the InvFels tree $\tau_1$ assuming $\emptyset$ is the most likely ancestral state split for each site split.
 All values multiplied by $1/32$.}
@@
 \[
+%EM Can you help me understand this?
+Where did the other terms from \log \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV = \siteSplit_j, \tau_1, t) go?
+They don't disappear with the proportionality-- they will be additive terms, not multiplicative.
+Or perhaps there's a bunch of cancellation?
+If the latter, let me know and I'll pull out my paper and pad.
 \ell^C(x_1; \boldsymbol\xi_0) \propto \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1).

### 2018-07-31 joint_inf.tex
Cleaning out addressed comments.
Getting to a point of confusion.

@@
 Let $\boldsymbol\xi = [\xi_j]_{j=1}^\nSiteSplits$ be the vector of most likely ancestral state splits.
-Use $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
+Use $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case because
 \begin{equation}
@@
 To understand the quality of each fit, we report the range of the bias $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.
-%EM I'd suggest bringing this figure into the main text. It'd fit.
-%das: :check:
 There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal.
@@
 Bias in branch length estimation.
-In regions where joint optimization should perform well with $p_{x^*}, p_{y^*} \le .1$, there is systematic bias toward shorter branch lengths.
+Even in regions with short branch length ($p_{x^*}, p_{y^*} \le .1$) where joint optimization should perform well, there is systematic bias toward shorter branch lengths.
 }
@@
 Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of topological inconsistency: the joint inference procedure estimates zero-length branches given the correct topology.
-%EM Suggest "Considering this zero length branch as a multifurcating topology..."
-%EM Oops, somehow I misunderstood this the first time around.
-You're saying that the wrong topology, with internal branch shrunk to zero, will have the same likelihood.
-And that we can't exclude that it could have a greater likelihood with internal branch nonzero.
-If you want to make those points (or something about multifurcating) they need to be more explicit.
-Right now it reads like we found cases where the wrong topology has larger likelihood.
-%das: maybe this is better?
-%EM Seems like clarification might be justified in other places in the MS as well.
-%das: also hedged a bit in the remainder of the MS.
 Considering a tree with a zero-length branch as a multifurcating topology, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
@@
 This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3).
-%EM I suggest flipping this around and making it stronger, so that we lead with and emphasize the differences.
-%EM I've made this a little spicier: happy with the first part?
-%das: I like it
-However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and that the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias})
+However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and that the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).
 

### 2018-08-01 appendix.tex
Sign question.
pass through comments

@@
     \hline
-$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau,t)$\\
+$\siteSplit_j$  & $p_{\siteSplit_j}$ &$8\cdot\Pr(\siteSplitRV=\siteSplit_j \mid \tau,t)$\\
     \hline
@@
 \caption{Site split probabilities $p_{\siteSplit_j}$ on the true InvFels tree $\tau^*$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, and on the InvFels tree $\tau_1$ and Felsenstein tree $\tau_2$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
-All values multiplied by $1/8$.}
+%EM "Multiplied by 1/8"... I think this should be "need to be multiplied by 1/8" or "All values have been divided by 1/8."
+%das: added it to the table headings, too
+All values need to be multiplied by $1/8$ to be probabilities.}
 \label{tab:gen-sitepatprob}
@@
 \hline
-$\siteSplit_j$ & $\ancestralSplit_k$ & $\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_1, t)$\\
+$\siteSplit_j$ & $\ancestralSplit_k$ & $32\cdot\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_1, t)$\\
 \hline
@@
 Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
-All values multiplied by $1/32$.
+All values need to be multiplied by $1/32$ to be probabilities.
 Ancestral states with $^*$ are never maximal provided parameters are in $(0,1]$.
@@
 \hline
-$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_j$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,t)$\\
+$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_j$ & $32\cdot\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,t)$\\
 \hline
@@
 Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
-All values multiplied by $1/32$.
+All values need to be multiplied by $1/32$ to be probabilities.
 Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
@@
 \begin{lemma}
-\label{lemma:bounds}
 Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
 There exists a set of $0 < x^*, y^* < 1$ such that $\emptyset$ is the maximal ancestral state split for all site splits given $\tau=\tau_1$.
+\label{lemma:ancestral-state}
 \end{lemma}
@@
 for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
-We will show that this implies $t=t^*$.
+We show that this implies $t=t^*$.
 Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
@@
 We now show the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
-We focus on $x_1$.
 %EM I think you mean the divergence term as a function of x_1 is proportional to...
-The marginal log-likelihood proportional to $x_1$ can be written as
+%das: removed "proportional" terminology abuse
+The marginal log-likelihood is
 \begin{align*}
 %EM This is a little funky-- I think this is a definition of D(x_1)?
-\shannonDivergence(x_1) \propto \sum_j p_{\siteSplit_j} \log(c_{\siteSplit_j} + a_{\siteSplit_j}x_1)
+%das: made more explicit
+\shannonDivergence(x_1) = \sum_j p_{\siteSplit_j} \log(c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1) - \log 8
 \end{align*}
-where $c_{\siteSplit_j}$ and $a_{\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
+where $c_{x_1,\siteSplit_j}$ and $a_{x_1,\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
 For example, reading from Table~\ref{tab:gen-sitepatprob},
 \[
-c_{\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
+c_{x_1,\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
 \]
@@
 \[
-a_{\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
+a_{x_1,\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
 \]
@@
 \begin{align*}
-\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}
+\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}
 \end{align*}
-and focus on a single term
+and focus on the $j$th term in the summation
 \[
-f_j(x_1) := p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}.
+f_j(x_1) := \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}.
 \]
@@
 \[
-\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{\siteSplit_j} + a_{\siteSplit_j}x_1),
+\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
 \]
 which, being a probability, is greater than or equal to zero.
-Thus, regardless of $a_{\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{\siteSplit_j} + a_{\siteSplit_j}x_1 \ge 0$.
+Thus, regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 \ge 0$.
 The remaining variables follow similar arguments.
@@
 The maximal likelihood values on the InvFels tree $\tau_1$ assuming $\emptyset$ is the most likely ancestral state split for each site split.
-All values multiplied by $1/32$.}
+All values need to be multiplied by $1/32$ to be probabilities.}
 \label{tab:likelihoods-restricted}
@@
 In this case, the conditional log-likelihood has no ambiguity, and has terms given in Table~\ref{tab:likelihoods-restricted}.
+We focus on the derivative with respect to $x_1$, aggregating all terms involving $x_2,y_1,y_2,$ and $w$ into a constant term.
 The conditional log-likelihood
 \[
-\ell^C_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi_0) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV = \siteSplit_j, \tau_1, t)
+\ell^C_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi_0) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV = \siteSplit_j, \tau_1, t),
 \]
-as a function of $x_1$, using Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:likelihoods-restricted}, is
-\[
+using Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:likelihoods-restricted}, is
+\begin{align*}
 %EM Can you help me understand this?
@@
 If the latter, let me know and I'll pull out my paper and pad.
-\ell^C(x_1; \boldsymbol\xi_0) \propto \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1).
-\]
+%das: hope this is better?
+\ell^C(x_1; \boldsymbol\xi_0) &= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
+&\qquad + \left(\frac{1}{2}+\frac{1}{2}(y^*)^2\right)\log(1+y_1) + \left(\frac{1}{2}-\frac{1}{2}(y^*)^2\right)\log(1-y_1)\\
+&\qquad + \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_2) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_2)\\
+&\qquad + \log(1+y_2) + \log(1+w) - \log 32\\
+&:= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
+&\qquad + C_{x_1}(x_2,y_1,y_2,w).
+\end{align*}
+where $C_{x_1}$ is a function of $y_1,x_2,y_2,$ and $w$.
 The derivative with respect to $x_1$ is
@@
 \[
-\hat{x}_1,\hat{x}_2 \in (0, x^*], \ \hat{y}_1 \in (0, y^*], \ \hat{y}_2,\hat{w} \in [y^*, 1].
+\hat{x}_1,\hat{x}_2 \in (0, x^*], \ \hat{y}_1 \in (0, y^*], \text{and}\ \hat{y}_2,\hat{w} \in [y^*, 1].
 \]
@@
 \begin{proof}
-The objective function proportional to $w$ is
+%EM "Proportional" thing again.
+%das: got rid of all the proportional talk and just explicitly wrote out the likelihood
+We proceed as in the previous proof by simplifying the constants in the log-likelihood.
+We analyze the derivative with respect to $w$, so we absorb all terms of only $x_1,y_1,x_2,$ and $y_2$ into a constant that we ignore.
+The objective function is
 \begin{align*}
-\ell(w) \propto \log(1+w) + \sum_j p_{\siteSplit_j} \log(1 + b_{\siteSplit_j}w)
+\ell(w; \boldsymbol\xi_0) = \left(\sum_j p_{\siteSplit_j} \log(c_{w,\siteSplit_j} + a_{w,\siteSplit_j}w)\right) + \log(1+w) + C_w(x_1,y_1,x_2,y_2)
 \end{align*}
-where, as in the previous proof,
+where $c_{w,\siteSplit_j},a_{w,\siteSplit_j}$ and $C_w$ are functions of $x_1,y_1,x_2,$ and $y_2$.
+Again, reading from Table~\ref{tab:gen-sitepatprob},
 \[
-b_{\emptyset} = \frac{(x_1+x_2)(y_1+y_2)}{(1+x_1x_2)(1+y_1y_2)}
+c_{w,\emptyset} = 1+x_1x_2+y_1y_2+x_1y_1x_2y_2 = (1+x_1x_2)(1+y_1y_2)
+\]
+and
+\[
+a_{w,\emptyset} = x_1y_1+x_1y_2+x_2y_1+x_2y_2 = (x_1+x_2)(y_1+y_2)
+\]
+with similar definitions for the remaining site splits.
+We make a further simplification on our way to computing bounds on the likelihood.
+Let
+\begin{align*}
+%EM The conditional term is \log(1+w), right? If so, it'd be easier to have that second rather than first so that it corresponds to the order in eq:log_likelihood_simplified
+%das: done
+\ell(w; \boldsymbol\xi_0) &= \left(\sum_j p_{\siteSplit_j} \log(c_{w,\siteSplit_j} + a_{w,\siteSplit_j}w)\right) + \log(1+w) + C_w(x_1,y_1,x_2,y_2) \\
+        &= \left(\sum_j p_{\siteSplit_j} \log(c_{w,\siteSplit_j} + a_{w,\siteSplit_j}w) - p_{\siteSplit_j} \log(c_{w,\siteSplit_j})\right) + \log(1+w) \\
+        &\qquad + \left(C_w(x_1,y_1,x_2,y_2) + \sum_j p_{\siteSplit_j} \log(c_{w,\siteSplit_j})\right) \\
+        &:= \left(\sum_j p_{\siteSplit_j} \log(1 + b_{w,\siteSplit_j}w)\right) + \log(1+w) + C^*_w(x_1,y_1,x_2,y_2)
+\end{align*}
+% Hm, we didn't do a trick quite like this with the division bit.
+%das: wrote out more explicitly
+where
+\[
+b_{\emptyset} := \frac{a_{w,\emptyset}}{c_{w,\emptyset}} = \frac{(x_1+x_2)(y_1+y_2)}{(1+x_1x_2)(1+y_1y_2)}
 \]
@@
 \begin{align*}
-\ell'(w) := \frac{d}{dw} \ell(w) = \frac{1}{1+w} + \sum_j p_{\siteSplit_j}b_{\siteSplit_j}\frac{1}{1 + b_{\siteSplit_j}w}.
+\ell'(w) := \frac{d}{dw} \ell(w; \boldsymbol\xi_0) =  \left(\sum_j \frac{p_{\siteSplit_j}b_{\siteSplit_j}}{1 + b_{\siteSplit_j}w}\right) + \frac{1}{1+w}.
 \end{align*}
@@
 \begin{align*}
-\ell'(w) &= \frac{1}{1+w} + \frac{p_{\emptyset}b_{\emptyset}}{1 + b_{\emptyset}w} - \frac{p_{13}b_{\emptyset}}{1 - b_{\emptyset}w} + \frac{p_{2}b_{2}}{1 + b_{2}w} - \frac{p_{2}b_{2}}{1 - b_{2}w} \\
-         &\qquad \frac{p_{1}b_{1}}{1 + b_{1}w} - \frac{p_{1}b_{1}}{1 - b_{1}w} + \frac{p_{12}b_{12}}{1 + b_{12}w} - \frac{p_{12}b_{12}}{1 - b_{12}w}.\\
+%%EM the negative sign                                                             V here doesn't seem right. We get the negative sign from going from p13 to the empty set.
+%\ell'(w) &= \frac{1}{1+w} + \frac{p_{\emptyset}b_{\emptyset}}{1 + b_{\emptyset}w} - \frac{p_{13}b_{\emptyset}}{1 - b_{\emptyset}w} + \frac{p_{2}b_{2}}{1 + b_{2}w} - \frac{p_{2}b_{2}}{1 - b_{2}w} \\
+%         &\qquad \frac{p_{1}b_{1}}{1 + b_{1}w} - \frac{p_{1}b_{1}}{1 - b_{1}w} + \frac{p_{12}b_{12}}{1 + b_{12}w} - \frac{p_{12}b_{12}}{1 - b_{12}w}.\\
+%das: the negative sign comes from b13 -> b0. I have written all the steps out explicitly so it's more clear.
+\ell'(w) &=  \frac{p_{\emptyset}b_{\emptyset}}{1 + b_{\emptyset}w} - \frac{p_{13}b_{\emptyset}}{1 - b_{\emptyset}w} + \frac{p_{2}b_{2}}{1 + b_{2}w} - \frac{p_{2}b_{2}}{1 - b_{2}w} \\
+         &\qquad \frac{p_{1}b_{1}}{1 + b_{1}w} - \frac{p_{1}b_{1}}{1 - b_{1}w} + \frac{p_{12}b_{12}}{1 + b_{12}w} - \frac{p_{12}b_{12}}{1 - b_{12}w} + \frac{1}{1+w}.\\
 \end{align*}
@@
 \[
-\ell'_{L}(w) = \frac{1}{1+w} - \frac{p_{13}}{1/b_{\emptyset}^U - w} - \frac{p_{2}}{1/b_{2}^U - w} - \frac{p_{1}}{1/b_{1}^U - w} - \frac{p_{12}}{1/b_{12}^U - w}
+\ell'_{L}(w) =  -\frac{p_{13}}{1/b_{\emptyset}^U - w} - \frac{p_{2}}{1/b_{2}^U - w} - \frac{p_{1}}{1/b_{1}^U - w} - \frac{p_{12}}{1/b_{12}^U - w} + \frac{1}{1+w}
 \]
 Note that $\ell'_{L}(w)$ is strictly decreasing in $w$ on the interval $(0,1]$, as shown in the previous proof, provided $b_{\emptyset}^U, b_{2}^U, b_{1}^U, b_{12}^U < 1$.
-Thus, if $\ell'_{L}(1) > 0$ then $\ell(w)$ is increasing on the entire interval $(0,1]$, including at the boundary $w=1$, and $\hat{w} \equiv 1$.
+Thus, if $\ell'_{L}(1) > 0$ then $\ell(w; \boldsymbol\xi_0)$ is increasing on the entire interval $(0,1]$, including at the boundary $w=1$, and $\hat{w} \equiv 1$.
 The conditions on $x^*$ and $y^*$ where $b_{2}^U, b_{1}^U, b_{12}^U < 1$ and $\ell_L'(1) > 0$ are straightforward, the one on the derivative being
 \[
-\ell'_{L}(1) = \frac{1}{2} - \frac{p_{13}b_{\emptyset}^U}{1-b_{\emptyset}^U} - \frac{p_{2}b_{2}^U}{1-b_{2}^U} - \frac{p_{1}b_{1}^U}{1-b_{1}^U} - \frac{p_{12}b_{12}^U}{1-b_{12}^U} > 0.
+\ell'_{L}(1) =  -\frac{p_{13}b_{\emptyset}^U}{1-b_{\emptyset}^U} - \frac{p_{2}b_{2}^U}{1-b_{2}^U} - \frac{p_{1}b_{1}^U}{1-b_{1}^U} - \frac{p_{12}b_{12}^U}{1-b_{12}^U} + \frac{1}{2} > 0.
 \]
@@
 %das: @EM did you say you wanted to get rid of these paragraphs? we never show there is a case where Fels > InvFels, just where it's equal...
+%EM It doesn't seem necessary to me. If we shrink the internal branch to zero, the two topologies match, right? It just seems like we need a sentence or two expressing that.
 As the region in Fig.~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.

### 2018-08-02 appendix.tex
Getting closer.

@@
 \end{tabular}
-\caption{Site split probabilities $p_{\siteSplit_j}$ on the true InvFels tree $\tau^*$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, and on the InvFels tree $\tau_1$ and Felsenstein tree $\tau_2$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
-%EM "Multiplied by 1/8"... I think this should be "need to be multiplied by 1/8" or "All values have been divided by 1/8."
-%das: added it to the table headings, too
-All values need to be multiplied by $1/8$ to be probabilities.}
+\caption{8 times the site split probabilities $p_{\siteSplit_j}$ on the true InvFels tree $\tau^*$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, and on the InvFels tree $\tau_1$ and Felsenstein tree $\tau_2$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
+}
 \label{tab:gen-sitepatprob}
@@
 \caption{
-Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
-All values need to be multiplied by $1/32$ to be probabilities.
+%EM Here and below, it seems a bit funny to call these likelihoods, as the Xi aren't observed.
+32 times likelihoods for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
 Ancestral states with $^*$ are never maximal provided parameters are in $(0,1]$.
@@
 \caption{
-Likelihood calculations on the InvFels tree $\tau_1$.
+32 times likelihoods on the InvFels tree $\tau_1$.
 Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
-All values need to be multiplied by $1/32$ to be probabilities.
 Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
@@
 \]
-We now show the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
-%EM I think you mean the divergence term as a function of x_1 is proportional to...
-%das: removed "proportional" terminology abuse
-The marginal log-likelihood is
+We now show that the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
+The marginal log-likelihood as a function of $x_1$ is
 \begin{align*}
-%EM This is a little funky-- I think this is a definition of D(x_1)?
-%das: made more explicit
 \shannonDivergence(x_1) = \sum_j p_{\siteSplit_j} \log(c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1) - \log 8
@@
 \caption{
-The maximal likelihood values on the InvFels tree $\tau_1$ assuming $\emptyset$ is the most likely ancestral state split for each site split.
-All values need to be multiplied by $1/32$ to be probabilities.}
+32 times the maximal likelihood values on the InvFels tree $\tau_1$ assuming $\emptyset$ is the most likely ancestral state split for each site split.
+}
 \label{tab:likelihoods-restricted}
@@
 \begin{align*}
-%EM Can you help me understand this?
-Where did the other terms from \log \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV = \siteSplit_j, \tau_1, t) go?
-They don't disappear with the proportionality-- they will be additive terms, not multiplicative.
-Or perhaps there's a bunch of cancellation?
-If the latter, let me know and I'll pull out my paper and pad.
-%das: hope this is better?
 \ell^C(x_1; \boldsymbol\xi_0) &= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
@@
 \begin{proof}
-%EM "Proportional" thing again.
-%das: got rid of all the proportional talk and just explicitly wrote out the likelihood
 We proceed as in the previous proof by simplifying the constants in the log-likelihood.
-We analyze the derivative with respect to $w$, so we absorb all terms of only $x_1,y_1,x_2,$ and $y_2$ into a constant that we ignore.
+We analyze the derivative with respect to $w$, so we absorb all terms of only $x_1,y_1,x_2,$ and $y_2$ into a constant that disappears upon taking the derivative.
 The objective function is
@@
 where $c_{w,\siteSplit_j},a_{w,\siteSplit_j}$ and $C_w$ are functions of $x_1,y_1,x_2,$ and $y_2$.
-Again, reading from Table~\ref{tab:gen-sitepatprob},
+Using the middle panel of Table~\ref{tab:gen-sitepatprob}, we obtain
 \[
@@
 \begin{align*}
-%EM The conditional term is \log(1+w), right? If so, it'd be easier to have that second rather than first so that it corresponds to the order in eq:log_likelihood_simplified
-%das: done
 \ell(w; \boldsymbol\xi_0) &= \left(\sum_j p_{\siteSplit_j} \log(c_{w,\siteSplit_j} + a_{w,\siteSplit_j}w)\right) + \log(1+w) + C_w(x_1,y_1,x_2,y_2) \\
@@
         &\qquad + \left(C_w(x_1,y_1,x_2,y_2) + \sum_j p_{\siteSplit_j} \log(c_{w,\siteSplit_j})\right) \\
-        &:= \left(\sum_j p_{\siteSplit_j} \log(1 + b_{w,\siteSplit_j}w)\right) + \log(1+w) + C^*_w(x_1,y_1,x_2,y_2)
+        &=: \left(\sum_j p_{\siteSplit_j} \log(1 + b_{w,\siteSplit_j}w)\right) + \log(1+w) + C^*_w(x_1,y_1,x_2,y_2)
 \end{align*}
-% Hm, we didn't do a trick quite like this with the division bit.
-%das: wrote out more explicitly
 where
@@
 with similar definitions for the remaining site splits.
-This function has derivative with respect to $w$ as
+The derivative with respect to $w$ of this function is
 \begin{align*}
@@
 \begin{align*}
-%%EM the negative sign                                                             V here doesn't seem right. We get the negative sign from going from p13 to the empty set.
-%\ell'(w) &= \frac{1}{1+w} + \frac{p_{\emptyset}b_{\emptyset}}{1 + b_{\emptyset}w} - \frac{p_{13}b_{\emptyset}}{1 - b_{\emptyset}w} + \frac{p_{2}b_{2}}{1 + b_{2}w} - \frac{p_{2}b_{2}}{1 - b_{2}w} \\
-%         &\qquad \frac{p_{1}b_{1}}{1 + b_{1}w} - \frac{p_{1}b_{1}}{1 - b_{1}w} + \frac{p_{12}b_{12}}{1 + b_{12}w} - \frac{p_{12}b_{12}}{1 - b_{12}w}.\\
-%das: the negative sign comes from b13 -> b0. I have written all the steps out explicitly so it's more clear.
 \ell'(w) &=  \frac{p_{\emptyset}b_{\emptyset}}{1 + b_{\emptyset}w} - \frac{p_{13}b_{\emptyset}}{1 - b_{\emptyset}w} + \frac{p_{2}b_{2}}{1 + b_{2}w} - \frac{p_{2}b_{2}}{1 - b_{2}w} \\
-         &\qquad \frac{p_{1}b_{1}}{1 + b_{1}w} - \frac{p_{1}b_{1}}{1 - b_{1}w} + \frac{p_{12}b_{12}}{1 + b_{12}w} - \frac{p_{12}b_{12}}{1 - b_{12}w} + \frac{1}{1+w}.\\
+         &\qquad + \frac{p_{1}b_{1}}{1 + b_{1}w} - \frac{p_{1}b_{1}}{1 - b_{1}w} + \frac{p_{12}b_{12}}{1 + b_{12}w} - \frac{p_{12}b_{12}}{1 - b_{12}w} + \frac{1}{1+w}.\\
 \end{align*}
@@
 \[
-0 < b_{\emptyset} \le \frac{2x^*}{1+(x^*)^2} := b_{\emptyset}^U,
+0 < b_{\emptyset} \le \frac{2x^*}{1+(x^*)^2} =: b_{\emptyset}^U,
 \]
 \[
-0 < b_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*} := b_{2}^U,
+0 < b_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*} =: b_{2}^U,
 \]
 \[
-0 < b_{1} \le \frac{x^*}{1-(x^*)^2} := b_{1}^U,
+0 < b_{1} \le \frac{x^*}{1-(x^*)^2} =: b_{1}^U,
 \]
 \[
-0 < b_{12} \le \frac{x^*}{1-(x^*)^2}\cdot\frac{1}{1-y^*} := b_{12}^U,
+0 < b_{12} \le \frac{x^*}{1-(x^*)^2}\cdot\frac{1}{1-y^*} =: b_{12}^U,
 \]

### 2018-08-02 appendix.tex
Hopefully my last stupid questions.
terminology; going through comments and writing out steps explicitly

@@
 See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the InvFels tree.
-Likelihood computations for all site splits and ancestral state splits are in Table~\ref{tab:farris_likelihoods} for the InvFels tree.
+Conditional likelihood computations for all site splits and ancestral state splits are in Table~\ref{tab:farris_likelihoods} for the InvFels tree.
 
@@
 %EM Here and below, it seems a bit funny to call these likelihoods, as the Xi aren't observed.
-32 times likelihoods for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
+%das: went with "conditional likelihood"
+32 times conditional likelihood values for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
 Ancestral states with $^*$ are never maximal provided parameters are in $(0,1]$.
-By combinations of $\ancestralSplit_k$, there are $3^5\cdot 4^2=3,888$ possible forms for the likelihood.
+By combinations of $\ancestralSplit_k$, there are $3^5\cdot 4^2=3,888$ possible forms for the conditional likelihood.
 }
@@
 Thus, we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$.
-An analogous statement holds for $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.
+An analogous statement holds for $y_2 \ge y_1$, with these conditional likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.
 
@@
 \caption{
-32 times likelihoods on the InvFels tree $\tau_1$.
+32 times conditional likelihood values on the InvFels tree $\tau_1$.
 Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
 Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
-Because in 4 cases there are 3 possibilities for $\xi_j$, there are $3^4=81$ possible forms for the likelihood.
+Because in 4 cases there are 3 possibilities for $\xi_j$, there are $3^4=81$ possible forms for the conditional likelihood.
 }
@@
 which, being a probability, is greater than or equal to zero.
-Thus, regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 \ge 0$.
+As we exclude the infinite branch length cases by letting $x_1,y_1,x_2,y_2,w > 0$, we have the strict inequality $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 > 0$.
+Thus, regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$.
 The remaining variables follow similar arguments.
@@
 \caption{
-32 times the maximal likelihood values on the InvFels tree $\tau_1$ assuming $\emptyset$ is the most likely ancestral state split for each site split.
+32 times the maximal conditional likelihood values on the InvFels tree $\tau_1$ assuming $\emptyset$ is the most likely ancestral state split for each site split.
 }
@@
 Assume $\emptyset$ is the most likely ancestral state split for all site splits.
-In this case, the conditional log-likelihood has no ambiguity, and has terms given in Table~\ref{tab:likelihoods-restricted}.
+In this case, the conditional log-likelihood has no ambiguity, and has terms given as the log of those in Table~\ref{tab:likelihoods-restricted}.
 We focus on the derivative with respect to $x_1$, aggregating all terms involving $x_2,y_1,y_2,$ and $w$ into a constant term.
@@
 \end{align*}
-Using the bounds on $\hat{x}_1,\hat{y}_1,\hat{x}_2$, and $\hat{y}_2$ from Lemma~\ref{lemma:bounds} and our assumption of $y_2 \ge y_1$ and $x_2 \ge x_1$, we have
+Using the bounds on $\hat{x}_1,\hat{y}_1,\hat{x}_2$, and $\hat{y}_2$ from Lemma~\ref{lemma:bounds} and our assumption of $y_2 \ge y_1 > 0$ and $x_2 \ge x_1 > 0$, we have
 \[
@@
 \]
+%das: The lower bound for b_{\emptyset} is strict because we assume x_1,...,w > 0, but for the other three they aren't strict since x_1 could equal x_2 and/or y_1 could equal y_2.
+This doesn't affect any later results, though, b/c b -> 0 just makes that term drop out.
 \[
-0 < b_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*} =: b_{2}^U,
+0 \le b_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*} =: b_{2}^U,
 \]
 \[
-0 < b_{1} \le \frac{x^*}{1-(x^*)^2} =: b_{1}^U,
+0 \le b_{1} \le \frac{x^*}{1-(x^*)^2} =: b_{1}^U,
 \]
 \[
-0 < b_{12} \le \frac{x^*}{1-(x^*)^2}\cdot\frac{1}{1-y^*} =: b_{12}^U,
+0 \le b_{12} \le \frac{x^*}{1-(x^*)^2}\cdot\frac{1}{1-y^*} =: b_{12}^U,
 \]
@@
 \[
-\ell'_{L}(w) =  -\frac{p_{13}}{1/b_{\emptyset}^U - w} - \frac{p_{2}}{1/b_{2}^U - w} - \frac{p_{1}}{1/b_{1}^U - w} - \frac{p_{12}}{1/b_{12}^U - w} + \frac{1}{1+w}
+\ell'_{L}(w) :=  -\frac{p_{13}}{1/b_{\emptyset}^U - w} - \frac{p_{2}}{1/b_{2}^U - w} - \frac{p_{1}}{1/b_{1}^U - w} - \frac{p_{12}}{1/b_{12}^U - w} + \frac{1}{1+w}.
 \]
-Note that $\ell'_{L}(w)$ is strictly decreasing in $w$ on the interval $(0,1]$, as shown in the previous proof, provided $b_{\emptyset}^U, b_{2}^U, b_{1}^U, b_{12}^U < 1$.
+%EM Where is this shown in the previous proof?
+%das: wrote out more explicitly. It's similar to the argument starting at L612.
+Each term in $\ell'_{L}(w)$ is of the form
+\[
+f(w) = \frac{p}{1/b + w} = \frac{p\cdot b}{1 + b\cdot w}
+\]
+with $0 \le p \le 1$.
+Assume $b_{\emptyset}^U, b_{2}^U, b_{1}^U, b_{12}^U < 1$, so that $-1 < b < 1$.
+Since $b > -1$ and $w\in(0,1]$, $1+b\cdot w > 0$ and, thus, each term in $\ell'_{L}(w)$ is a decreasing function in $w$.
+%EM I agree that this condition shows that the lower bound is increasing on the entire interval. But how do we get to a conclusion about $\ell(.)$ itself?
+%das: The derivative of \ell, i.e., \ell', is a decreasing function on [0,1].
+The value of \ell' at 1 is positive, thus, \ell' > 0 on all of [0,1].
+Since \ell' > 0 on all of [0,1], \ell is increasing on all of [0,1] (I think via MVT?).
 Thus, if $\ell'_{L}(1) > 0$ then $\ell(w; \boldsymbol\xi_0)$ is increasing on the entire interval $(0,1]$, including at the boundary $w=1$, and $\hat{w} \equiv 1$.
@@
 \[
+%EM Hm... it seems like we need to exclude having the b's very close to 1. How do we do that?
+%das: I thought that was taken care of by the condition above: $b_{2}^U, b_{1}^U, b_{12}^U < 1$.
+It should also be taken care of by the below inequality, too, right?
+Otherwise as b -> 1 this will go to -\infty, so the below will fail.
+Do you think they need to be well-separated from 1 instead of just strictly less than 1?
 \ell'_{L}(1) =  -\frac{p_{13}b_{\emptyset}^U}{1-b_{\emptyset}^U} - \frac{p_{2}b_{2}^U}{1-b_{2}^U} - \frac{p_{1}b_{1}^U}{1-b_{1}^U} - \frac{p_{12}b_{12}^U}{1-b_{12}^U} + \frac{1}{2} > 0.
@@
 %EM It doesn't seem necessary to me. If we shrink the internal branch to zero, the two topologies match, right? It just seems like we need a sentence or two expressing that.
+%EM Thoughts here?
+%das: I kept the one paragraph showing they're equal, but maybe even that's too much?
 As the region in Fig.~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
-We show this is a type of inconsistency in topology estimation by constructing a lower bound on the likelihood for the Felsenstein topology $\tau_2$ and demonstrating it is equal to the likelihood given the InvFels topology $\tau_1$ in the demarcated region of Fig.~\ref{fig:incons-analytic}.
-As we are interested in a lower bound, assume $\boldsymbol\xi=\boldsymbol\xi_0$, i.e., $\emptyset$ is the maximal ancestral state split for all site splits.
-In this case, the conditional log-likelihood for $\tau_2$ is composed of the same functional forms given in Table~\ref{tab:likelihoods-restricted}, as assuming $\emptyset$ for the ancestral state split results in the same conditional log-likelihood for both topologies.
-
-Assume $w=1$.
-In this case, Table~\ref{tab:gen-sitepatprob} shows that, for $w=1$, the marginal log-likelihood is identical for $\tau_1$ and $\tau_2$ as when the internal branch becomes zero-length the two topologies are indistinguishable.
+It is easy to see that assuming $\emptyset$ for each maximal ancestral state split results in the same conditional log-likelihood for $\tau_1$ and $\tau_2$.
+Moreover, if $w=1$, the internal branch becomes zero-length and the two topologies are indistinguishable.
 Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the shaded region in Fig.~\ref{fig:incons-analytic}.
-The unconstrained maximum likelihood value for $\tau_2$ is greater than or equal to the value of the likelihood given these two constraints.
-We have
+We can see this results in the likelihood of both topologies being equal, i.e.,
 \begin{align*}
-\max_{t} & \ \ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi) \ge \\
-&\max_{t: \boldsymbol\xi=\boldsymbol\xi_0, w=1, \tau=\tau_2} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}) \\
-&\qquad\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\}) \\
+\max_{t : t^*\in\mathcal{T}_0} & \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)\\
 &= \max_{t: \boldsymbol\xi=\boldsymbol\xi_0, w=1, \tau=\tau_1} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
-&\qquad\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\})\\
-&= \max_{t : t^*\in\mathcal{T}_0} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi),
+&\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\})\\
+&= \max_{t: \boldsymbol\xi=\boldsymbol\xi_0, w=1, \tau=\tau_2} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}) \\
+&\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\}).
 \end{align*}
-and thus the incorrect topology achieves a likelihood value greater than or equal to the correct one.
 

### 2018-08-02 joint_inf.tex
terminology; going through comments and writing out steps explicitly

@@
 \end{equation}
-In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral state splits in Table~\ref{tab:farris_likelihoods} to obtain the likelihood term for each most likely ancestral state split, these terms depending on the unknown $(\tau, t)$.
+In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral state splits in Table~\ref{tab:farris_likelihoods} to obtain the conditional likelihood term for each most likely ancestral state split, these terms depending on the unknown $(\tau, t)$.
 For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}_\nCols$ for a fixed $(\tau,t)$, then $(\hat{\tau}_\nCols,\hat{t}_\nCols)$ given $\hat{\fullAncestralStates}_\nCols$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2018-xl}.
@@
 We elaborate on this point in the Appendix.
-The proof is through analyzing upper and lower bounds of the potential likelihood forms in Table~\ref{tab:farris_likelihoods} and the gradients of the objective function $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ with respect to $t$.
+The proof is through analyzing upper and lower bounds of the potential conditional likelihood forms in Table~\ref{tab:farris_likelihoods} and the gradients of the objective function $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ with respect to $t$.
 

### 2018-08-02 appendix.tex
fix error and add assumption

@@
 Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
+Assume $x_1,y_1,x_2 < 1$.
 For $0 < x^*, y^* < 1$ and $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
@@
 with similar definitions for the remaining site splits.
+By the assumption of $0 < x^*, y^* < 1$, we have $p_{\siteSplit_j} > 0$ for $j=1,\ldots,\nSiteSplits$, therefore we assume $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 \neq 0$ for all $j$ otherwise we obtain a zero value for our likelihood, i.e., $\shannonDivergence\rightarrow -\infty$.
+Assuming $x_1,y_1,x_2 < 1$ precludes all cases where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ for some $j$.
+Moreover, we have
+\[
+\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
+\]
+which, being a probability, is greater than or equal to zero.
 Consider the derivative with respect to $x_1$
@@
 \]
-Recall that
-\[
-\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
-\]
-which, being a probability, is greater than or equal to zero.
-As we exclude the infinite branch length cases by letting $x_1,y_1,x_2,y_2,w > 0$, we have the strict inequality $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 > 0$.
-Thus, regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$.
+%das: gah, had to roll back this line and add some stuff above since there are cases where some probabilities are 0.
+Maybe I'm being too careful?
+I don't think it's an issue, just an edge case that could induce wonkiness, so I included an assumption.
+Regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 > 0$.
 The remaining variables follow similar arguments.
@@
 
+Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ is reasonable.
 We now prove the main result.

### 2018-08-02 appendix.tex
A few notes.

@@
 \caption{
-%EM Here and below, it seems a bit funny to call these likelihoods, as the Xi aren't observed.
-%das: went with "conditional likelihood"
 32 times conditional likelihood values for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
@@
 %EM Do discontinuities occur? In my mind we are taking the maximum of continuous functions, which is continuous.
+%EM We don't think they occur.
 Discontinuities occur when, if we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we are at a point of discontinuity we maximize over ancestral state splits, ensuring the upper-semicontinuous property of this mapping.
@@
 Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
+%EM Perhaps we can say something about how this is not a strong assumption.
+%EM see below "By the assumption of"
 Assume $x_1,y_1,x_2 < 1$.
@@
 
+%EM Also here.
 Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ is reasonable.
@@
 \]
-%EM Where is this shown in the previous proof?
-%das: wrote out more explicitly. It's similar to the argument starting at L612.
 Each term in $\ell'_{L}(w)$ is of the form
@@
 Thus, if $\ell'_{L}(1) > 0$ then $\ell(w; \boldsymbol\xi_0)$ is increasing on the entire interval $(0,1]$, including at the boundary $w=1$, and $\hat{w} \equiv 1$.
+
+%EM Topic sentence here please.
 The conditions on $x^*$ and $y^*$ where $b_{2}^U, b_{1}^U, b_{12}^U < 1$ and $\ell_L'(1) > 0$ are straightforward, the one on the derivative being
 \[
-%EM Hm... it seems like we need to exclude having the b's very close to 1. How do we do that?
-%das: I thought that was taken care of by the condition above: $b_{2}^U, b_{1}^U, b_{12}^U < 1$.
-It should also be taken care of by the below inequality, too, right?
-Otherwise as b -> 1 this will go to -\infty, so the below will fail.
-Do you think they need to be well-separated from 1 instead of just strictly less than 1?
 \ell'_{L}(1) =  -\frac{p_{13}b_{\emptyset}^U}{1-b_{\emptyset}^U} - \frac{p_{2}b_{2}^U}{1-b_{2}^U} - \frac{p_{1}b_{1}^U}{1-b_{1}^U} - \frac{p_{12}b_{12}^U}{1-b_{12}^U} + \frac{1}{2} > 0.

### 2018-08-03 appendix.tex
be more explicit

@@
 %EM We don't think they occur.
-Discontinuities occur when, if we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we are at a point of discontinuity we maximize over ancestral state splits, ensuring the upper-semicontinuous property of this mapping.
-This shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.
+%das: done
+If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
+This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.
 
@@
 %EM see below "By the assumption of"
-Assume $x_1,y_1,x_2 < 1$.
+%das: moved below
 For $0 < x^*, y^* < 1$ and $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
@@
 By the assumption of $0 < x^*, y^* < 1$, we have $p_{\siteSplit_j} > 0$ for $j=1,\ldots,\nSiteSplits$, therefore we assume $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 \neq 0$ for all $j$ otherwise we obtain a zero value for our likelihood, i.e., $\shannonDivergence\rightarrow -\infty$.
-Assuming $x_1,y_1,x_2 < 1$ precludes all cases where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ for some $j$.
+In other words, we make the assumption that $x_1,y_1,x_2 < 1$, precluding all cases where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ for some $j$.
+This assumption is reasonable as the likelihood is never maximized at these values where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$.
 Moreover, we have
@@
 %EM Also here.
-Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ is reasonable.
+%das: reworded
+Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ does not contradict with the statement of Lemma~\ref{lemma:bounds}.
 We now prove the main result.
@@
 Since \ell' > 0 on all of [0,1], \ell is increasing on all of [0,1] (I think via MVT?).
-Thus, if $\ell'_{L}(1) > 0$ then $\ell(w; \boldsymbol\xi_0)$ is increasing on the entire interval $(0,1]$, including at the boundary $w=1$, and $\hat{w} \equiv 1$.
+Thus, if $\ell'_{L}(1) > 0$, since $\ell'_{L}(w)$ is decreasing on $(0,1]$ then $\ell'_{L}(w) > 0$ for all $w\in (0,1]$.
+Since $\ell'(w) > \ell'_{L}(w)$ for $w\in(0,1]$, this implies $\ell'(w) > 0$, which further implies $\ell(w; \boldsymbol\xi_0)$ is increasing on the entire interval $(0,1]$, including at the boundary $w=1$, and $\hat{w} \equiv 1$.
 
 %EM Topic sentence here please.
-The conditions on $x^*$ and $y^*$ where $b_{2}^U, b_{1}^U, b_{12}^U < 1$ and $\ell_L'(1) > 0$ are straightforward, the one on the derivative being
+%das: made more explicit and introduced more gently
+We now construct the conditions on $x^*$ and $y^*$ that are required for $\hat{w}\equiv 1$:
 \[
-\ell'_{L}(1) =  -\frac{p_{13}b_{\emptyset}^U}{1-b_{\emptyset}^U} - \frac{p_{2}b_{2}^U}{1-b_{2}^U} - \frac{p_{1}b_{1}^U}{1-b_{1}^U} - \frac{p_{12}b_{12}^U}{1-b_{12}^U} + \frac{1}{2} > 0.
+C1-C3: b_{2}^U, b_{1}^U, b_{12}^U < 1,
 \]
-The set of $x^*,y^*$ satisfying these conditions as well as those for $\emptyset$ being the most likely ancestral state split is given in Fig.~\ref{fig:incons-analytic}, giving the region of generating parameters where $\hat{w}\equiv 1$.
+\[
+C4: -\frac{p_{13}b_{\emptyset}^U}{1-b_{\emptyset}^U} - \frac{p_{2}b_{2}^U}{1-b_{2}^U} - \frac{p_{1}b_{1}^U}{1-b_{1}^U} - \frac{p_{12}b_{12}^U}{1-b_{12}^U} + \frac{1}{2} > 0.
+\]
+Conditions $C1-C3$ are needed so that $\ell'_{L}(w)$ is a decreasing function on $(0,1]$, while condition $C4$ is the one on $\ell'_{L}(1) > 0$, i.e., so that $\ell'(w)$ is positive on $(0,1]$.
+The set of $x^*,y^*$ satisfying conditions $C1-C4$ as well as those for $\emptyset$ being the most likely ancestral state split is given in Fig.~\ref{fig:incons-analytic}, giving the region of generating parameters where $\hat{w}\equiv 1$.
 \end{proof}

### 2018-08-03 appendix.tex
Cleaning out addressed comments.
Main text pass and half of appendix.

@@
 We begin by introducing ``site splits.''
-One application of site splits is to formalize the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
+We use site splits to formalize the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
 This is a standard step in the description of the Hadamard transform (Section 8.6 of \citet{Semple2003-em}), although our approach is complicated slightly by the inclusion of ancestral states.
@@
 The mapping $\patternToSplit$ operates by returning the tips labeled as $1$ in a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if the set of tips labeled $1$ is not in $\mathcal{P}(\siteSplitSet)$.
-The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if the set of tips labeled $1$ in $\alignmentColumn$ is in $\mathcal{P}(\siteSplitSet)$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the set of tips labeled $1$ in $\ancestralStateColumn$; otherwise, the set of tips labeled $1$ in $\overline{\alignmentColumn}$ necessarily is in $\mathcal{P}(\siteSplitSet)$.
+The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if the set of tips labeled $1$ in $\alignmentColumn$ is in $\mathcal{P}(\siteSplitSet)$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the set of tips labeled $1$ in $\ancestralStateColumn$; otherwise, the set of tips labeled $1$ in $\overline{\alignmentColumn}$ necessarily is in $\mathcal{P}(\siteSplitSet)$ and so $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\overline{\ancestralStateColumn}$.
 
@@
 \subsection*{Example}
-We follow with an expository example computing these probabilities and likelihoods.
+We follow with an example computing these probabilities and likelihoods.
 Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
@@
 \]
+%EM perhaps a section reference for this big book?
 Using the notation in \citet{van1998asymptotic}, we let
@@
 For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
-%EM Do discontinuities occur? In my mind we are taking the maximum of continuous functions, which is continuous.
-%EM We don't think they occur.
-%das: done
 If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
@@
 Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
-We show that the likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
+Next we show that the likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
 Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
@@
 meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
-We show that for site splits $\{1\}$ and $\{3\}$, exchanging $x_1$ and $x_2$ also swaps the values of the conditional likelihood terms.
-The corresponding possibilities for the conditional likelihood values are
+We show that for site splits $\{1\}$ and $\{3\}$, exchanging $x_1$ and $x_2$ also swaps the values of the conditional likelihood terms (Table~\ref{tab:farris_likelihoods}).
+Indeed, the corresponding possibilities for the conditional likelihood values are
 \begin{align*}
@@
 for site split $\{3\}$, which shows the likelihood remains unchanged if $x_1$ and $x_2$ are swapped.
-The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
+The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$
+%EM does this one need a different explanation now?
+as well as $\{1,2\}$ and $\{2,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
 The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
@@
 Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
-%EM Perhaps we can say something about how this is not a strong assumption.
-%EM see below "By the assumption of"
-%das: moved below
 For $0 < x^*, y^* < 1$ and $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
@@
 
-%EM Also here.
-%das: reworded
 Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ does not contradict with the statement of Lemma~\ref{lemma:bounds}.
@@
 Since $b > -1$ and $w\in(0,1]$, $1+b\cdot w > 0$ and, thus, each term in $\ell'_{L}(w)$ is a decreasing function in $w$.
-%EM I agree that this condition shows that the lower bound is increasing on the entire interval. But how do we get to a conclusion about $\ell(.)$ itself?
-%das: The derivative of \ell, i.e., \ell', is a decreasing function on [0,1].
-The value of \ell' at 1 is positive, thus, \ell' > 0 on all of [0,1].
-Since \ell' > 0 on all of [0,1], \ell is increasing on all of [0,1] (I think via MVT?).
 Thus, if $\ell'_{L}(1) > 0$, since $\ell'_{L}(w)$ is decreasing on $(0,1]$ then $\ell'_{L}(w) > 0$ for all $w\in (0,1]$.
@@
 
-%das: @EM did you say you wanted to get rid of these paragraphs? we never show there is a case where Fels > InvFels, just where it's equal...
-%EM It doesn't seem necessary to me. If we shrink the internal branch to zero, the two topologies match, right? It just seems like we need a sentence or two expressing that.
-%EM Thoughts here?
-%das: I kept the one paragraph showing they're equal, but maybe even that's too much?
 As the region in Fig.~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.

### 2018-08-03 joint_inf.tex
Main text pass and half of appendix.

@@
 Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
-Recently, a computationally-efficient approach proposing to jointly maximize over these unknown states and phylogenetic parameters.
+Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
 Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
@@
 Moreover, examples of inconsistency arising from problems where the number of parameters increases with the amount of data \citep{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
-In this case those additional parameters come in the form of the states of ancestral sequences.
+In this case those additional parameters are the states of ancestral sequences.
 %page 2 section 2.1 at the top of the second column first paragraph second sentence
-Although the software described in \citet{Sagulenko2018-xl} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
+Although \citet{Sagulenko2018-xl} explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
 In particular, one would like to know when this approximate technique breaks down for both topology and branch length inference, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.
@@
 In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model with data being generated on a four-taxon tree and we construct bounds on the joint objective function to demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
+To do so, we use a binary symmetric model with data generated on a four-taxon tree: we construct bounds on the joint objective function to demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
 We find similar areas through empirical means where joint inference consistently underestimates interior branch lengths when the topology is known and fixed.
@@
 Assume the binary symmetric model, namely with a character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution \citep{Semple2003-em}.
-Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
+Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ be the number of internal nodes.
 We observe $\nCols$ independent and identically distributed samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
@@
 We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with an alternate formulation called ``fidelity.''
-The probability of a substitution on a branch with fidelity $x$ is $(1-x)/2$ while the probability of no substitution is $(1+x)/2$ where $0 \le x \le 1$.
+The probability of a substitution on a branch with fidelity $x$ is $(1-x)/2$, while the probability of no substitution is $(1+x)/2$ where $0 \le x \le 1$.
 This parameter quantifies the fidelity of transmission of the ancestral state across an edge \citep{Matsen2007-jq}.
@@
 \end{equation}
+%EM ancestral state splits not introduced-- can we just say "ancestral states"? Note that splits appear below as well.
 In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral state splits in Table~\ref{tab:farris_likelihoods} to obtain the conditional likelihood term for each most likely ancestral state split, these terms depending on the unknown $(\tau, t)$.
@@
 Let $\boldsymbol\xi = [\xi_j]_{j=1}^\nSiteSplits$ be the vector of most likely ancestral state splits.
-Use $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case because
+Use $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case because, as shown in the Appendix,
 \begin{equation}
@@
 
+%EM I'm going to suggest that we move this to the appendix. The theorem says that there's a set, and we want people to nod then look at the current Figure 3.
 \begin{figure}
@@
 Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically\hyp{}derived region (Fig.~\ref{fig:incons-analytic}).
-To determine how conservative, we use the L-BFGS-B method to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
+To determine how conservative it is, we use the L-BFGS-B method to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
 We see that the region of inconsistency in the general case is much larger than in the analytic case, which is unsurprising because the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral state splits.
@@
 
+%EM This is an important paragraph but the meat is hidden in the middle.
+%EM We need to start with a topic sentence that says that For branch lengh, we find a systematic bias even in the short-branch case (Figure ref).
+%EM Then we can talk about details such as discontinuities and how marginal works great.
 As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
@@
 We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
-Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of topological inconsistency: the joint inference procedure estimates zero-length branches given the correct topology.
-Considering a tree with a zero-length branch as a multifurcating topology, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
+Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of topological inconsistency: the joint inference procedure estimates zero-length branches, which can be considered as a multifurcating topology.
+Also, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
 Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length is more likely to be observed, though we have not found regions where this is the case.
@@
 Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
+%EM I don't follow here. This doesn't seem like the long branch attraction region of parameter space-- isn't that on the Felsenstein tree, not the InvFels tree?
 This is not the case here; while analytically the inconsistency occurs in the long branch attraction region of the parameter space, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
@@
 We have shown an inconsistency when performing joint inference on branch lengths given an InvFels topology and investigated the performance of branch parameter estimation.
-Our aim is to analytically demonstrate joint inference is inconsistent, and, while this indicates an issue with this inference procedure, there is substantial scope for future work to make these results more precise and more general.
+There is substantial scope for future work to make these results more precise and more general.
+%EM I suggest compacting the next few sentences.
 The techniques used to obtain upper and lower bounds for the likelihoods in Lemma~\ref{lemma:bounds} provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
+%EM What is the "general case" here?
 Similarly, for the general case in estimating branch lengths, we are only able to provide empirical estimates for regions of overestimation, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) beg further investigation.
@@
 Further study is needed to determine the impact of these results on problems that are of interest to practitioners.
-Given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
+Also, given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
 Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.
@@
 We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.
-%TODO: change up this line?
-Insight from the editors and Vu Dinh greatly improved this manuscript.
+Insight from the reviewers, the editors, and Vu Dinh greatly improved this manuscript.
 

### 2018-08-03 appendix.tex
update swapping argument

@@
 and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
+Similarly, for site split $\{1,3\}$,
+\[
+    \Pr(\siteSplitRV=\{1,3\}\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 -  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
+\]
+which also is invariant to exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
 
@@
 \end{align*}
-meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
-We show that for site splits $\{1\}$ and $\{3\}$, exchanging $x_1$ and $x_2$ also swaps the values of the conditional likelihood terms (Table~\ref{tab:farris_likelihoods}).
+meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values, regardless of what we do with $y_1$ and $y_2$.
+We show that for site splits $\{1\}$ and $\{3\}$, exchanging $x_1$ and $x_2$ also swaps the values of the conditional likelihood terms, again independent of what happens to $y_1$ and $y_2$ (Table~\ref{tab:farris_likelihoods}).
 Indeed, the corresponding possibilities for the conditional likelihood values are
@@
 for site split $\{3\}$, which shows the likelihood remains unchanged if $x_1$ and $x_2$ are swapped.
-The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$
 %EM does this one need a different explanation now?
-as well as $\{1,2\}$ and $\{2,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
-The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
-Exchanging $x_1$ and $x_2$ also does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$.
+%das: wrote out in more detail; I realize now I maybe took some of these intermediate steps for granted.
+Effectively we want to look at exchanging xs and ys together as well as separately, because they result in different arguments of symmetry, but still don't change the likelihood value.
+The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ independent of what we do with $x_1$ and $x_2$.
 
-Thus, we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$.
-An analogous statement holds for $y_2 \ge y_1$, with these conditional likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.
+For site splits $\{1,2\}$ and $\{2,3\}$ we see the following.
+By exchanging only $x_1$ with $x_2$, the generating probabilities and conditional likelihood values swap between these two site splits.
+The same is true of the generating probabilities if we exchange only $y_1$ and $y_2$, except, for the case of the conditional likelihood values, we look at the complement of the most likely ancestral state split.
+In other words, the function value for the conditional likelihood also swaps between site splits $\{1,2\}$ and $\{2,3\}$, though the most likely ancestral state split is different.
+Now, if we exchange both $x_1$ with $x_2$ and $y_1$ with $y_2$, we see these generating probabilities remain unchanged.
+Again, for the conditional likelihood values, we look at the complement of the most likely ancestral state split and see these values also remain unchanged.
+
+Exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$.
+Thus, we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$ and $y_2 \ge y_1$, with these conditional likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.
 

### 2018-08-03 appendix.tex
address appendix comments/lemma 3 question

@@
 %EM perhaps a section reference for this big book?
-Using the notation in \citet{van1998asymptotic}, we let
+Using the notation in Section 5.2.1 in \citet{van1998asymptotic}, we let
 \[
@@
 \[
-    \Pr(\siteSplitRV=\{1,3\}\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 -  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
+    \Pr(\siteSplitRV=\{1,3\}\mid \tau_1, t) = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 -  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
 \]
@@
 for site split $\{3\}$, which shows the likelihood remains unchanged if $x_1$ and $x_2$ are swapped.
+
 %EM does this one need a different explanation now?
@@
 Effectively we want to look at exchanging xs and ys together as well as separately, because they result in different arguments of symmetry, but still don't change the likelihood value.
-The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ independent of what we do with $x_1$ and $x_2$.
+%das: okay, even gorier detail---I think it's all correct now.
+For site splits $\{2\}$ and $\{1,2,3\}$, exchanging $y_1$ and $y_2$ swaps the values of the generating probabilities, independent of what happens to $x_1$ and $x_2$.
+In the case of the conditional likelihood values, we see that the values for these site splits swap as well, though, we look at the complement of the most likely ancestral state split.
+In other words, the function value for the conditional likelihood also swaps between site splits $\{2\}$ and $\{1,2,3\}$, though the most likely ancestral state split is different.
+Indeed,
+\begin{align*}
+    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{2\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w);\\
+    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{2\}, \tau_1, t)     &= \frac{1}{32}(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1-w);\\
+    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{2\}, \tau_1, t)     &= \frac{1}{32}(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1-w);\\
+    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{2\}, \tau_1, t)   &= \frac{1}{32}(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1+w);
+\end{align*}
+for site split $\{2\}$ and
+\begin{align*}
+    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2,3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w);\\
+    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2,3\}, \tau_1, t)     &= \frac{1}{32}(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w);\\
+    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2,3\}, \tau_1, t)     &= \frac{1}{32}(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1-w);\\
+    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2,3\}, \tau_1, t)   &= \frac{1}{32}(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w);
+\end{align*}
+for site split $\{1,2,3\}$, which shows the likelihood remains unchanged if $y_1$ and $y_2$ are swapped.
 
@@
 By exchanging only $x_1$ with $x_2$, the generating probabilities and conditional likelihood values swap between these two site splits.
-The same is true of the generating probabilities if we exchange only $y_1$ and $y_2$, except, for the case of the conditional likelihood values, we look at the complement of the most likely ancestral state split.
-In other words, the function value for the conditional likelihood also swaps between site splits $\{1,2\}$ and $\{2,3\}$, though the most likely ancestral state split is different.
-Now, if we exchange both $x_1$ with $x_2$ and $y_1$ with $y_2$, we see these generating probabilities remain unchanged.
-Again, for the conditional likelihood values, we look at the complement of the most likely ancestral state split and see these values also remain unchanged.
+The same is true of the generating probabilities if we exchange only $y_1$ and $y_2$, except, for the case of the conditional likelihood values, we again look at the complement of the most likely ancestral state split as in the case of splits $\{2\}$ and $\{1,2,3\}$.
+Now, if we exchange both $x_1$ with $x_2$ and $y_1$ with $y_2$, we see these generating probabilities remain unchanged, and, for the conditional likelihood values, we look at the complement of the most likely ancestral state split and see these values also remain unchanged.
 
@@
 \[
-\frac{d}{dx_1} \ell^C(x_1; \boldsymbol\xi_0) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2}
+\frac{d}{dx_1} \ell^C(x_1; \boldsymbol\xi_0) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2},
 \]
 which is negative for all $x_1 \ge x^*$.
-This same process applied to the remaining variables shows the gradient of the full likelihood is negative for $x_1,x_2 \ge x^*$ and $y_1 \ge y^*$ and positive for $y_2,w \ge y^*$, yielding the constraints
+This same process applied to $x_2$ and $y_1$ shows the gradient of the full likelihood is negative for $x_2 \ge x^*$ and $y_1 \ge y^*$.
+
+The variables $y_2$ and $w$ require the following.
+Consider the conditional log-likelihood with respect to $w$
+\[
+\ell^C(w; \boldsymbol\xi_0) := \log(1+w) + C_{w}(x_1,y_1,x_2,y_2)
+\]
+with derivative with respect to $w$
+\[
+\frac{d}{dw} \ell^C(w; \boldsymbol\xi_0) = \frac{1}{1+w},
+\]
+which is positive for all $w \ge 0$.
+As the gradient of the marginal log-likelihood is a decreasing function in $w$ and has a zero crossing at $y^*$, the addition of the derivative of the conditional log-likelihood makes the gradient of the full log-likelihood positive for all $w \le y^*$.
+The same argument holds for $y_2$, yielding the constraints
 \[

### 2018-08-03 versions/response.v1.txt
address all comments and update responses

@@
 %das: added
-This work is not the final word on the joint inference question from an applied perspective, as we now emphasize with an expanded Discussion section as well as the addition of these lines:
-"Examples with larger trees and more realistic mutation models fall within the scope of an extension to applied models.
-Further study is needed to determine the impact of these results on problems that are of interest to practitioners."
-
+This work is not the final word on the joint inference question from an applied perspective, as we now emphasize with an expanded Discussion section.
 Even if we were to greatly expand the sections doing numerical optimization, these results would only be for the binary symmetric model, which is never applied in practice.
@@
 We have added clarification of these points to the discussion:
-"All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
-Examples with larger trees and more realistic mutation models fall within the scope of an extension to applied models.
-Further study is needed to determine the impact of these results on problems that are of interest to practitioners."
+"All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners."
 
@@
 We have made this more precise and directly referenced the inequalities:
-"In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral state splits in Table~\ref{tab:farris_likelihoods} to obtain the likelihood term for each most likely ancestral state split, these terms depending on the unknown $(\tau, t)$.
+"In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral states (Table~\ref{tab:farris_likelihoods}) to obtain each conditional likelihood term, these terms depending on the unknown $(\tau, t)$.
 For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}_\nCols$ for a fixed $(\tau,t)$, then $(\hat{\tau}_\nCols,\hat{t}_\nCols)$ given $\hat{\fullAncestralStates}_\nCols$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2018-xl}."
@@
 We believe the inaccuracies have more to do with the specific choice of which ancestral state is maximal than asymmetries in the tree:
-"This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split if we let $x_1=x_2=x^*$ and $y_1=y_2=y^*$ (Table~\ref{tab:likelihoods})."
+"This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which the most likely ancestral states are the same for each internal node letting $x_1=x_2=x^*$ and $y_1=y_2=y^*$ ($\boldsymbol\xi=\emptyset$ in Table~\ref{tab:likelihoods})."
 

### 2018-08-03 appendix.tex
Cleaning out addressed comments.
address all comments and update responses

@@
 \]
-%EM perhaps a section reference for this big book?
 Using the notation in Section 5.2.1 in \citet{van1998asymptotic}, we let
@@
 
-%EM does this one need a different explanation now?
-%das: wrote out in more detail; I realize now I maybe took some of these intermediate steps for granted.
-Effectively we want to look at exchanging xs and ys together as well as separately, because they result in different arguments of symmetry, but still don't change the likelihood value.
-%das: okay, even gorier detail---I think it's all correct now.
 For site splits $\{2\}$ and $\{1,2,3\}$, exchanging $y_1$ and $y_2$ swaps the values of the generating probabilities, independent of what happens to $x_1$ and $x_2$.
@@
 
-%EM Topic sentence here please.
-%das: made more explicit and introduced more gently
 We now construct the conditions on $x^*$ and $y^*$ that are required for $\hat{w}\equiv 1$:
@@
 \end{proof}
+
+\begin{figure}
+\centering
+\includegraphics[width=.95\textwidth]{incons-analytic-inkscape}
+\caption{
+    An analytically-derived region of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
+    Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
+}
+\label{fig:incons-analytic}
+\end{figure}
 

### 2018-08-03 joint_inf.tex
address all comments and update responses

@@
 %EM ancestral state splits not introduced-- can we just say "ancestral states"? Note that splits appear below as well.
-In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral state splits in Table~\ref{tab:farris_likelihoods} to obtain the conditional likelihood term for each most likely ancestral state split, these terms depending on the unknown $(\tau, t)$.
+%das: I removed "splits" where I could, but kept it right before the theorem. Thoughts?
+In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral states (Table~\ref{tab:farris_likelihoods}) to obtain each conditional likelihood term, these terms depending on the unknown $(\tau, t)$.
 For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}_\nCols$ for a fixed $(\tau,t)$, then $(\hat{\tau}_\nCols,\hat{t}_\nCols)$ given $\hat{\fullAncestralStates}_\nCols$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2018-xl}.
@@
 Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ (Fig.~\ref{fig:farris-fels-top}a).
-Let $\boldsymbol\xi = [\xi_j]_{j=1}^\nSiteSplits$ be the vector of most likely ancestral state splits.
+Let $\boldsymbol\xi = [\xi_j]_{j=1}^\nSiteSplits$ be the vector of most likely ancestral state splits---the explicit definition for $\boldsymbol\xi$ is given in the Appendix.
 Use $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case because, as shown in the Appendix,
@@
 %EM I'm going to suggest that we move this to the appendix. The theorem says that there's a set, and we want people to nod then look at the current Figure 3.
-\begin{figure}
-\centering
-\includegraphics[width=.95\textwidth]{incons-analytic-inkscape}
-\caption{
-    An analytically-derived region of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
-    Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
-}
-\label{fig:incons-analytic}
-\end{figure}
+%das: good idea; I think this will also appease reviewer 1.
 
 We provide the following as an intuition for the theoretical development.
-For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral state splits.
-For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest conditional log-likelihood.
+For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral states.
+For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state corresponds to the highest conditional log-likelihood.
 As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this is not guaranteed because the maximum depends on the values of the unknown branch parameters $t$.
 Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
-This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split if we let $x_1=x_2=x^*$ and $y_1=y_2=y^*$ (Table~\ref{tab:likelihoods}).
+This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which the most likely ancestral states are the same for each internal node letting $x_1=x_2=x^*$ and $y_1=y_2=y^*$ ($\boldsymbol\xi=\emptyset$ in Table~\ref{tab:likelihoods}).
 If we allow multifurcating trees in our inference, then we can think of this as an instance of converging to the wrong topology, as the true $y^*\neq 1$.
@@
 To determine how conservative it is, we use the L-BFGS-B method to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
-We see that the region of inconsistency in the general case is much larger than in the analytic case, which is unsurprising because the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral state splits.
+We see that the region of inconsistency in the general case is much larger than in the analytic case, which is unsurprising because the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral states.
 This region encompasses almost half of the branch fidelity space; given the correct topology and performing our best possible optimization, we have many situations where we estimate the interior branch length to be zero.
@@
 %EM Then we can talk about details such as discontinuities and how marginal works great.
+%das: okay, reworked.
+In estimating the interior branch length $w$, we find a systematic bias in the joint inference procedure even when the true branches are short (Fig.~\ref{fig:empirical-bias}).
 As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
-Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same optimization code.
-For joint inference, the estimates for $\hat{p}_w$ when $p_{x^*}$ and $p_{y^*}$ are both small look reasonable, with $\hat{p}_w$ increasing as $p_{y^*}$ increases, though Fig.~\ref{fig:empirical-bias} shows there is a systematic negative bias in this procedure even when the true branches are short.
-To understand the quality of each fit, we report the range of the bias $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.
-There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal.
-Because of these discontinuities, an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
-For joint inference, the errors given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for inference on the integrated likelihood are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
+There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal, so we investigate the bias in the region where $p_{x^*}$ and $p_{y^*}$ are both small, i.e., $p_{x^*}, p_{y^*} \le .1$, as these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
+The estimates for $\hat{p}_w$ in this case look reasonable at first, compared with those estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), though joint inference predictably underestimates the interior branch length.
+Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$.
+Inference on the integrated likelihood performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same optimization code.
+Moreover, the errors in this case are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
 
@@
 %EM I don't follow here. This doesn't seem like the long branch attraction region of parameter space-- isn't that on the Felsenstein tree, not the InvFels tree?
-This is not the case here; while analytically the inconsistency occurs in the long branch attraction region of the parameter space, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
+%das: yeah, it was kind of hand-wavy.
+This is not the case here; while analytically the inconsistency occurs when the top two branches are long and the bottom three are short, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
 This is especially apparent in the ``long $x^*$, short $y^*$'' case discussed previously.
+Additionally, we generate data on the InvFels tree $\tau_1$ while \citet{Felsenstein1978-rr} generates data on the Felsenstein tree $\tau_2$.
 Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.
@@
 %EM What is the "general case" here?
-Similarly, for the general case in estimating branch lengths, we are only able to provide empirical estimates for regions of overestimation, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) beg further investigation.
-Empirical validation shows that the region of inconsistency is much larger than the one analytically derived.
-All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
-Examples with larger trees and more realistic mutation models fall within the scope of an extension to applied models.
-Further study is needed to determine the impact of these results on problems that are of interest to practitioners.
+%das: did some scalpeling and rewording
+For the case of estimating branch lengths, we are able to provide empirical estimates for regions of overestimation that are tighter than the analytical bounds, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) beg further investigation.
+All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners.
 Also, given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.

### 2018-08-06 appendix.tex
:male-factory-worker:

@@
 Next we show that the likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
+Although this property should not be surprising due to symmetry, we write it out for completeness.
 Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
@@
 
-Exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$.
-Thus, we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$ and $y_2 \ge y_1$, with these conditional likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.
+Thus exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$.
+Therefore we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$ and $y_2 \ge y_1$, with these conditional likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.
 
@@
 \]
-for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
+for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$ and $\tau^*$ being the InvFels tree $\tau_1$.
 We show that this implies $t=t^*$.
@@
 where $c_{x_1,\siteSplit_j}$ and $a_{x_1,\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
-For example, reading from Table~\ref{tab:gen-sitepatprob},
+For example, reading from the middle panel of Table~\ref{tab:gen-sitepatprob},
 \[
@@
 with similar definitions for the remaining site splits.
+
+%EM Could you restructure this paragraph with a topic-sentence statement of the assumption we want to make, then evidence that it's ok? It roves around a good bit.
 By the assumption of $0 < x^*, y^* < 1$, we have $p_{\siteSplit_j} > 0$ for $j=1,\ldots,\nSiteSplits$, therefore we assume $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 \neq 0$ for all $j$ otherwise we obtain a zero value for our likelihood, i.e., $\shannonDivergence\rightarrow -\infty$.
@@
 This assumption is reasonable as the likelihood is never maximized at these values where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$.
+
 Moreover, we have
@@
 We now focus on the conditional log-likelihood.
+%EM These "Assume"s feel frightening to the reader. Can we replace with "Under our stated hypothesis"?
 Assume $\emptyset$ is the most likely ancestral state split for all site splits.
@@
 which is positive for all $w \ge 0$.
+%EM Does the zero crossing need additional argument?
+I'm guessing that you get this from the gradient-is-zero thing from the unique-maximum argument above.
+But that's just that it's zero when we are doing maximization of all of the variables at once.
+Here aren't we wanting to make an argument for lots of x_i and y_i?
 As the gradient of the marginal log-likelihood is a decreasing function in $w$ and has a zero crossing at $y^*$, the addition of the derivative of the conditional log-likelihood makes the gradient of the full log-likelihood positive for all $w \le y^*$.
@@
 
+%EM Are we saying that Lemma 3 applies in this case? I think so but it could be more specific.
 Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ does not contradict with the statement of Lemma~\ref{lemma:bounds}.

### 2018-08-06 joint_inf.tex
:male-factory-worker:

@@
 \end{equation}
-%EM ancestral state splits not introduced-- can we just say "ancestral states"? Note that splits appear below as well.
-%das: I removed "splits" where I could, but kept it right before the theorem. Thoughts?
 In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral states (Table~\ref{tab:farris_likelihoods}) to obtain each conditional likelihood term, these terms depending on the unknown $(\tau, t)$.
@@
 Our main theorem states that given data generated on $\tau_1$ there exist true nonzero branch lengths such that the estimator $\hat{t}$ maximizing the right-hand side of \eqref{eq:exp-log-lik} has an internal branch of length zero.
+%EM A funny thing: this theorem as stated is really under the hypothesis that the ancestral state is emptyset.
+We can combine this with the lemmas to remove this assumption, but it seems like we should do this for the reader.
+Time for a corollary?
 \begin{restatable}{theorem}{topoInconsist}
@@
 Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
+%EM Can we say "open set" or "set of positive measure"?
 For $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$ there exists a set of $0 < x^*, y^* < 1$ such that the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
@@
 %das: good idea; I think this will also appease reviewer 1.
+%EM On the other hand we should still refer to it in the text around here.
 
 We provide the following as an intuition for the theoretical development.
-For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral states.
+For a particular site pattern, to obtain the joint maximum likelihood function we maximize over ancestral states.
 For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state corresponds to the highest conditional log-likelihood.
@@
 We see that the region of inconsistency in the general case is much larger than in the analytic case, which is unsurprising because the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral states.
-This region encompasses almost half of the branch fidelity space; given the correct topology and performing our best possible optimization, we have many situations where we estimate the interior branch length to be zero.
+This region encompasses almost half of the branch fidelity space; given the correct topology and performing our best possible optimization, there are many situations where we estimate the interior branch length to be zero.
 
@@
 
-%EM This is an important paragraph but the meat is hidden in the middle.
-%EM We need to start with a topic sentence that says that For branch lengh, we find a systematic bias even in the short-branch case (Figure ref).
-%EM Then we can talk about details such as discontinuities and how marginal works great.
-%das: okay, reworked.
 In estimating the interior branch length $w$, we find a systematic bias in the joint inference procedure even when the true branches are short (Fig.~\ref{fig:empirical-bias}).
@@
 There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal, so we investigate the bias in the region where $p_{x^*}$ and $p_{y^*}$ are both small, i.e., $p_{x^*}, p_{y^*} \le .1$, as these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
-The estimates for $\hat{p}_w$ in this case look reasonable at first, compared with those estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), though joint inference predictably underestimates the interior branch length.
+Although the estimates for $\hat{p}_w$ are better than the estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), joint inference still predictably underestimates the interior branch length.
 Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$.
@@
 This is not the case here; while analytically the inconsistency occurs when the top two branches are long and the bottom three are short, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
+%EM Is this next sentence necessary given the previous one?
 This is especially apparent in the ``long $x^*$, short $y^*$'' case discussed previously.
@@
 There is substantial scope for future work to make these results more precise and more general.
-%EM I suggest compacting the next few sentences.
 The techniques used to obtain upper and lower bounds for the likelihoods in Lemma~\ref{lemma:bounds} provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
-%EM What is the "general case" here?
-%das: did some scalpeling and rewording
 For the case of estimating branch lengths, we are able to provide empirical estimates for regions of overestimation that are tighter than the analytical bounds, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) beg further investigation.

### 2018-08-06 appendix.tex
some restructuring mostly

@@
 32 times conditional likelihood values on the InvFels tree $\tau_1$.
-Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
+Due to the symmetry of the likelihood, WLOG we let $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
 Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
@@
 Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
-There exists a set of $0 < x^*, y^* < 1$ such that $\emptyset$ is the maximal ancestral state split for all site splits given $\tau=\tau_1$.
+There exists an open set of $0 < x^*, y^* < 1$ such that $\emptyset$ is the maximal ancestral state split for all site splits given $\tau=\tau_1$.
 \label{lemma:ancestral-state}
@@
 \]
-i.e., the lower bound assuming $\emptyset$ is the maximal ancestral state split for all site splits and $\tau=\tau_1$.
+i.e., the lower bound given $\emptyset$ is the maximal ancestral state split for all site splits and $\tau=\tau_1$.
 
@@
 
-%EM Could you restructure this paragraph with a topic-sentence statement of the assumption we want to make, then evidence that it's ok? It roves around a good bit.
-By the assumption of $0 < x^*, y^* < 1$, we have $p_{\siteSplit_j} > 0$ for $j=1,\ldots,\nSiteSplits$, therefore we assume $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 \neq 0$ for all $j$ otherwise we obtain a zero value for our likelihood, i.e., $\shannonDivergence\rightarrow -\infty$.
-In other words, we make the assumption that $x_1,y_1,x_2 < 1$, precluding all cases where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ for some $j$.
-This assumption is reasonable as the likelihood is never maximized at these values where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$.
-
-Moreover, we have
-\[
-\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
-\]
-which, being a probability, is greater than or equal to zero.
 Consider the derivative with respect to $x_1$
@@
 \]
-%das: gah, had to roll back this line and add some stuff above since there are cases where some probabilities are 0.
-Maybe I'm being too careful?
-I don't think it's an issue, just an edge case that could induce wonkiness, so I included an assumption.
-Regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 > 0$.
+Regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ so long as $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 > 0$.
+We have
+\[
+\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
+\]
+which, being a probability, is greater than or equal to zero, so we focus on those fidelities where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$.
+
+To ensure each $f_j$ is decreasing, we assume that $x_1,y_1,x_2 < 1$; these are the only such cases where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ for some $j$.
+In fact, since $0 < x^*, y^* < 1$---implying $p_{\siteSplit_j} > 0$ for all $j$---if $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ for any $j$, we obtain a zero value for our likelihood, i.e., $\shannonDivergence\rightarrow -\infty$, and so the likelihood is never maximized when $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$
+Thus, each $f_j$ is decreasing no matter the value of $x_1,y_1,$ or $x_2$, though for some values of $x_1,y_1,$ or $x_2$ equal to one the log-likelihood diverges, and the gradient of the marginal log-likelihood is a decreasing function in $x_1$.
 The remaining variables follow similar arguments.
+
+%EM Could you restructure this paragraph with a topic-sentence statement of the assumption we want to make, then evidence that it's ok? It roves around a good bit.
+%das: it was a little confusing; I merged and rearranged a few things.
 
@@
 \caption{
-32 times the maximal conditional likelihood values on the InvFels tree $\tau_1$ assuming $\emptyset$ is the most likely ancestral state split for each site split.
+32 times the maximal conditional likelihood values on the InvFels tree $\tau_1$ where $\emptyset$ is the most likely ancestral state split for each site split.
 }
@@
 %EM These "Assume"s feel frightening to the reader. Can we replace with "Under our stated hypothesis"?
-Assume $\emptyset$ is the most likely ancestral state split for all site splits.
+%das: got rid of some others in the text, too, that seemed like overkill
+By Lemma~\ref{lemma:ancestral-state}, we let $\emptyset$ be the most likely ancestral state split for all site splits.
 In this case, the conditional log-likelihood has no ambiguity, and has terms given as the log of those in Table~\ref{tab:likelihoods-restricted}.
@@
 Here aren't we wanting to make an argument for lots of x_i and y_i?
-As the gradient of the marginal log-likelihood is a decreasing function in $w$ and has a zero crossing at $y^*$, the addition of the derivative of the conditional log-likelihood makes the gradient of the full log-likelihood positive for all $w \le y^*$.
+%das: Yeah, I think how it was written before didn't state it correctly.
+The zero crossing is in all the variables, and we want to make a statement about all the variables at once (as stated in the lemma).
+We focus on a single variable at a time for the conditional likelihood (\ell^C) because it is separable in each variable, but the constraints are for the maximization over all variables.
+The gradient of the marginal log-likelihood has a zero crossing at $\{x^*,y^*,x^*,y^*,y^*\}$, and the gradient of the conditional log-likelihood is a function of only $w$.
+Moreover, the gradient of the marginal log-likelihood is decreasing in $w$, thus, the addition of the derivative of the conditional log-likelihood makes the gradient of the $w$ component of the full log-likelihood positive for all $w \le y^*$.
 The same argument holds for $y_2$, yielding the constraints
@@
 %EM Are we saying that Lemma 3 applies in this case? I think so but it could be more specific.
-Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ does not contradict with the statement of Lemma~\ref{lemma:bounds}.
+%das: I'm not quite sure I follow the question; I reworded it though because I think it was confusingly stated.
+Since $x^*, y^* < 1$, we see the bounds in Lemma~\ref{lemma:bounds} have the property that $x_1,y_1,x_2 < 1$.
 We now prove the main result.
@@
 We analyze the derivative with respect to $w$, so we absorb all terms of only $x_1,y_1,x_2,$ and $y_2$ into a constant that disappears upon taking the derivative.
-The objective function is
+Using Lemma~\ref{lemma:ancestral-state}, the objective function is
 \begin{align*}
@@
 where $c_{w,\siteSplit_j},a_{w,\siteSplit_j}$ and $C_w$ are functions of $x_1,y_1,x_2,$ and $y_2$.
-Using the middle panel of Table~\ref{tab:gen-sitepatprob}, we obtain
+From the middle panel of Table~\ref{tab:gen-sitepatprob}, we obtain
 \[
@@
 Conditions $C1-C3$ are needed so that $\ell'_{L}(w)$ is a decreasing function on $(0,1]$, while condition $C4$ is the one on $\ell'_{L}(1) > 0$, i.e., so that $\ell'(w)$ is positive on $(0,1]$.
-The set of $x^*,y^*$ satisfying conditions $C1-C4$ as well as those for $\emptyset$ being the most likely ancestral state split is given in Fig.~\ref{fig:incons-analytic}, giving the region of generating parameters where $\hat{w}\equiv 1$.
+The open set of $x^*,y^*$ satisfying conditions $C1-C4$ as well as in Lemma~\ref{lemma:ancestral-state} is given in Fig.~\ref{fig:incons-analytic}, giving the region of generating parameters where $\hat{w}\equiv 1$.
 \end{proof}
@@
 As the region in Fig.~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
-It is easy to see that assuming $\emptyset$ for each maximal ancestral state split results in the same conditional log-likelihood for $\tau_1$ and $\tau_2$.
+It is easy to see that letting $\emptyset$ be the maximal ancestral state split results in the same conditional log-likelihood for $\tau_1$ and $\tau_2$.
 Moreover, if $w=1$, the internal branch becomes zero-length and the two topologies are indistinguishable.

### 2018-08-06 joint_inf.tex
some restructuring mostly

@@
 
-\title{Joint maximum-likelihood of phylogenies and ancestral states is not consistent}
+\title{Joint maximum-likelihood of phylogeny and ancestral states is not consistent}
 
@@
 Time for a corollary?
+%das: ok, I reworded it a bit; let me know how you think it sounds.
 \begin{restatable}{theorem}{topoInconsist}
@@
 %EM Can we say "open set" or "set of positive measure"?
-For $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$ there exists a set of $0 < x^*, y^* < 1$ such that the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
+There exists an open set of $0 < x^*, y^* < 1$ such that for some $\boldsymbol\xi$ the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
 \[
-\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi_0)
+\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)
 \]
@@
 
-This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one).
+This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:incons-analytic}).
 As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
@@
 %EM Is this next sentence necessary given the previous one?
-This is especially apparent in the ``long $x^*$, short $y^*$'' case discussed previously.
 Additionally, we generate data on the InvFels tree $\tau_1$ while \citet{Felsenstein1978-rr} generates data on the Felsenstein tree $\tau_2$.

### 2018-08-07 appendix.tex
:tennis:

@@
 \]
-which, being a probability, is greater than or equal to zero, so we focus on those fidelities where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$.
+which, being a probability, is greater than or equal to zero.
+We next exclude the situation of $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$.
 
-To ensure each $f_j$ is decreasing, we assume that $x_1,y_1,x_2 < 1$; these are the only such cases where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ for some $j$.
-In fact, since $0 < x^*, y^* < 1$---implying $p_{\siteSplit_j} > 0$ for all $j$---if $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ for any $j$, we obtain a zero value for our likelihood, i.e., $\shannonDivergence\rightarrow -\infty$, and so the likelihood is never maximized when $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$
-Thus, each $f_j$ is decreasing no matter the value of $x_1,y_1,$ or $x_2$, though for some values of $x_1,y_1,$ or $x_2$ equal to one the log-likelihood diverges, and the gradient of the marginal log-likelihood is a decreasing function in $x_1$.
+To exclude this situation we assume that $x_1,y_1,x_2 < 1$; if this criterion does not hold we get zero likelihoods (which obviously are not maximal).
+In fact, since $0 < x^*, y^* < 1$---implying $p_{\siteSplit_j} > 0$ for all $j$---if $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ for any $j$, we obtain a zero value for our likelihood, i.e., $\shannonDivergence\rightarrow -\infty$.
+Hence, the likelihood is never maximized when $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ and each $f_j$ is decreasing no matter the value of $x_1,y_1,$ or $x_2$, though for some values of $x_1,y_1,$ or $x_2$ equal to one the log-likelihood diverges, and the gradient of the marginal log-likelihood is a decreasing function in $x_1$.
 The remaining variables follow similar arguments.
@@
 %das: it was a little confusing; I merged and rearranged a few things.
+%EM I made a few further changes. Seem OK?
 
@@
 %das: got rid of some others in the text, too, that seemed like overkill
+%EM "we let $\emptyset$ be the most likely ancestral state split" still seems funny to me.
+We can't "let" it be something that is defined by being a maximum.
+How about saying that it is _a_ most likely...
+% By Lemma~\ref{lemma:ancestral-state}, $\emptyset$ is a most likely ancestral state split for all site splits.
 By Lemma~\ref{lemma:ancestral-state}, we let $\emptyset$ be the most likely ancestral state split for all site splits.
@@
 We focus on a single variable at a time for the conditional likelihood (\ell^C) because it is separable in each variable, but the constraints are for the maximization over all variables.
+%EM This is definitely an improvement, though I don't know if I get it yet.
+There is a single point at which we know it's zero.
+We also know that it's decreasing in w.
+It seems like if we combine these two facts we could still have a function that isn't positive for all $w \le y^*$ (if the level sets of the function are "tilted").
+%EM Also, a "zero crossing" at a single point of a multivariate function doesn't seem like the right terminology. Perhaps "vanishes"?
 The gradient of the marginal log-likelihood has a zero crossing at $\{x^*,y^*,x^*,y^*,y^*\}$, and the gradient of the conditional log-likelihood is a function of only $w$.
@@
 
-%EM Are we saying that Lemma 3 applies in this case? I think so but it could be more specific.
-%das: I'm not quite sure I follow the question; I reworded it though because I think it was confusingly stated.
 Since $x^*, y^* < 1$, we see the bounds in Lemma~\ref{lemma:bounds} have the property that $x_1,y_1,x_2 < 1$.
@@
 As the region in Fig.~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
+%EM "letting $\emptyset$ be the maximal ancestral state split" sounds a little funny-- either it is or it isn't maximal.
 It is easy to see that letting $\emptyset$ be the maximal ancestral state split results in the same conditional log-likelihood for $\tau_1$ and $\tau_2$.

### 2018-09-19 appendix.tex
version 3...

--- 2018-08-06 appendix.tex
@@
 
-For the $i$th factor of \eqref{eq:full_likelihood}, we decompose the likelihood into its marginal and conditional components as
-\[
-\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
-\]
+We now consider the $i$th factor of \eqref{eq:full_likelihood}.
 As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
 \begin{align*}
-    \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t)\\
-    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
-    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
-    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t)
+    \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t)\\
+    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)\\
+    &= \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)\\
+    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)
 \end{align*}
 where $\overline{Y}$ is the complement of the site pattern--valued random variable $Y$ and has the same distribution as $Y$.
-Note
+Since
 \begin{align*}
-    \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
+    \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t),
 \end{align*}
-Thus given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
+given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
 \begin{align*}
-\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
+\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_j \mid \tau, t).
 \end{align*}
@@
 \begin{align*}
-    \max_{\ancestralStateColumn_i} \ \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t)\\
-    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)\\
-    &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)
+    \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t)\\
+    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)\\
+    &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_j \mid \tau, t)
 \end{align*}
@@
 \[
-\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
+\Pr(\siteSplitRV=\siteSplit, \ancestralSplitRV=\emptyset \mid \tau, t)\neq \Pr(\siteSplitRV=\siteSplit, \ancestralSplitRV=\{1,2\} \mid \tau, t)
 \]
@@
 \[
-\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
+\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\emptyset \mid \tau, t) =
 (1-p)^5,
@@
 \[
-\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
-\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
+\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\{1\} \mid \tau, t) =
+\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\{2\} \mid \tau, t) =
 p^3(1-p)^2,
@@
 \[
-\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
+\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\{1,2\} \mid \tau, t) =
 p^4(1-p).
@@
 \[
-\Pr(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
-\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
+\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\xi_1 \mid \tau, t) =
+\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\emptyset \mid \tau, t) =
 (1-p)^5.
@@
 \[
-\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
-\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
+\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\emptyset \mid \tau, t) =
+\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\{1,2\} \mid \tau, t) =
 p^2(1-p)^3,
@@
 \[
-\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
-\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
+\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\{1\} \mid \tau, t) =
+\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\{2\} \mid \tau, t) =
 p^3(1-p)^2.
@@
 \[
-\Pr(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
+\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\xi_5 \mid \tau, t) =
 p^2(1-p)^3.
@@
                              &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
-                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
+                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \nonumber \\
+                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
 \end{align}
@@
 \[
-L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
+L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)}
 \]
@@
 \begin{equation*}
-\begin{split}
-&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
-&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
-&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) +
-            \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
-\end{split}
+\frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)
 \end{equation*}
@@
 &    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
-&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)].
+&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)].
 \label{eq:site_pattern_profile_likelihood_mean}
 \end{split}
-\end{equation}
-Define the marginal log-likelihood
-\[
-\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)
-\]
-and the conditional log-likelihood
-\[
-\ell^C_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
-\]
-with $\boldsymbol\xi = [\xi_j]_{j=1}^\nSiteSplits$ so that \eqref{eq:site_pattern_profile_likelihood_mean} is
-\begin{equation}
-    \label{eq:log_likelihood_simplified}
-    \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^C_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
 \end{equation}
@@
 
-To compute the likelihood of observing a set of data, we need $\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
+To compute the likelihood of observing a set of data, we need $\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
 Using branch fidelities, the probability of a character change along a branch with fidelity parameter $x$ is $(1-x)/2$, while the probability of a character remaining the same is $(1+x)/2$.
 See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the InvFels tree.
-Conditional likelihood computations for all site splits and ancestral state splits are in Table~\ref{tab:farris_likelihoods} for the InvFels tree.
+Likelihood computations for all site splits and ancestral state splits are in Table~\ref{tab:farris_likelihoods} for the InvFels tree.
 
@@
     Edges labeled by the probability of substitution along that edge.
-    In (a), we compute the product to obtain $\Pr(\ancestralSplitRV=\emptyset\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x_1)(1-x_2)(1+y_1)(1-y_2)(1+w)/32$.
-    In (b), the same process yields $\Pr(\ancestralSplitRV=\{1\}\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x_1)(1-x_2)(1+y_1)(1-y_2)(1-w)/32$.
+    In (a), we compute the product to obtain $\Pr(\siteSplitRV=\{2,3\},\ancestralSplitRV=\emptyset\mid \tau_1,t) = (1+x_1)(1-x_2)(1+y_1)(1-y_2)(1+w)/32$.
+    In (b), the same process yields $\Pr(\siteSplitRV=\{2,3\},\ancestralSplitRV=\{1\}\mid \tau_1,t) = (1+x_1)(1-x_2)(1+y_1)(1-y_2)(1-w)/32$.
 }
@@
 \hline
-$\siteSplit_j$ & $\ancestralSplit_k$ & $32\cdot\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_1, t)$\\
+$\siteSplit_j$ & $\ancestralSplit_k$ & $32\cdot\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau_1, t)$\\
 \hline
@@
 \caption{
-32 times conditional likelihood values for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
+32 times likelihood values for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
 Ancestral states with $^*$ are never maximal provided parameters are in $(0,1]$.
-By combinations of $\ancestralSplit_k$, there are $3^5\cdot 4^2=3,888$ possible forms for the conditional likelihood.
+By combinations of $\ancestralSplit_k$, there are $3^5\cdot 4^2=3,888$ possible forms for the likelihood.
 }
@@
 \[
-m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
+m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)]
 \]
@@
 meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values, regardless of what we do with $y_1$ and $y_2$.
-We show that for site splits $\{1\}$ and $\{3\}$, exchanging $x_1$ and $x_2$ also swaps the values of the conditional likelihood terms, again independent of what happens to $y_1$ and $y_2$ (Table~\ref{tab:farris_likelihoods}).
-Indeed, the corresponding possibilities for the conditional likelihood values are
+We show that for site splits $\{1\}$ and $\{3\}$, exchanging $x_1$ and $x_2$ also swaps the values of the likelihood terms, again independent of what happens to $y_1$ and $y_2$ (Table~\ref{tab:farris_likelihoods}).
+Indeed, the corresponding possibilities for the likelihood values are
 \begin{align*}
-    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
-    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
-    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
-    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
+    \Pr(\siteSplitRV=\{1\}, \ancestralSplitRV=\emptyset \mid \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
+    \Pr(\siteSplitRV=\{1\}, \ancestralSplitRV=\{1\} \mid \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
+    \Pr(\siteSplitRV=\{1\}, \ancestralSplitRV=\{2\} \mid \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
+    \Pr(\siteSplitRV=\{1\}, \ancestralSplitRV=\{1,2\} \mid \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
 \end{align*}
@@
 \begin{align*}
-        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
-    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
-    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
-    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
+    \Pr(\siteSplitRV=\{3\}, \ancestralSplitRV=\emptyset \mid \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
+    \Pr(\siteSplitRV=\{3\}, \ancestralSplitRV=\{1\} \mid \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
+    \Pr(\siteSplitRV=\{3\}, \ancestralSplitRV=\{2\} \mid \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
+    \Pr(\siteSplitRV=\{3\}, \ancestralSplitRV=\{1,2\} \mid \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
 \end{align*}
@@
 For site splits $\{2\}$ and $\{1,2,3\}$, exchanging $y_1$ and $y_2$ swaps the values of the generating probabilities, independent of what happens to $x_1$ and $x_2$.
-In the case of the conditional likelihood values, we see that the values for these site splits swap as well, though, we look at the complement of the most likely ancestral state split.
-In other words, the function value for the conditional likelihood also swaps between site splits $\{2\}$ and $\{1,2,3\}$, though the most likely ancestral state split is different.
+In the case of the likelihood values, we see that the values for these site splits swap as well, though, we look at the complement of the most likely ancestral state split.
+In other words, the function value for the likelihood also swaps between site splits $\{2\}$ and $\{1,2,3\}$, though the most likely ancestral state split is different.
 Indeed,
 \begin{align*}
-    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{2\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w);\\
-    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{2\}, \tau_1, t)     &= \frac{1}{32}(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1-w);\\
-    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{2\}, \tau_1, t)     &= \frac{1}{32}(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1-w);\\
-    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{2\}, \tau_1, t)   &= \frac{1}{32}(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1+w);
+    \Pr(\siteSplitRV=\{2\}, \ancestralSplitRV=\emptyset \mid \tau_1, t) &= \frac{1}{32}(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w);\\
+    \Pr(\siteSplitRV=\{2\}, \ancestralSplitRV=\{1\} \mid \tau_1, t)     &= \frac{1}{32}(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1-w);\\
+    \Pr(\siteSplitRV=\{2\}, \ancestralSplitRV=\{2\} \mid \tau_1, t)     &= \frac{1}{32}(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1-w);\\
+    \Pr(\siteSplitRV=\{2\}, \ancestralSplitRV=\{1,2\} \mid \tau_1, t)   &= \frac{1}{32}(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1+w);
 \end{align*}
@@
 \begin{align*}
-    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2,3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w);\\
-    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2,3\}, \tau_1, t)     &= \frac{1}{32}(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w);\\
-    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2,3\}, \tau_1, t)     &= \frac{1}{32}(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1-w);\\
-    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2,3\}, \tau_1, t)   &= \frac{1}{32}(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w);
+    \Pr(\siteSplitRV=\{1,2,3\}, \ancestralSplitRV=\emptyset \mid \tau_1, t) &= \frac{1}{32}(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w);\\
+    \Pr(\siteSplitRV=\{1,2,3\}, \ancestralSplitRV=\{1\} \mid \tau_1, t)     &= \frac{1}{32}(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w);\\
+    \Pr(\siteSplitRV=\{1,2,3\}, \ancestralSplitRV=\{2\} \mid \tau_1, t)     &= \frac{1}{32}(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1-w);\\
+    \Pr(\siteSplitRV=\{1,2,3\}, \ancestralSplitRV=\{1,2\} \mid \tau_1, t)   &= \frac{1}{32}(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w);
 \end{align*}
@@
 For site splits $\{1,2\}$ and $\{2,3\}$ we see the following.
-By exchanging only $x_1$ with $x_2$, the generating probabilities and conditional likelihood values swap between these two site splits.
-The same is true of the generating probabilities if we exchange only $y_1$ and $y_2$, except, for the case of the conditional likelihood values, we again look at the complement of the most likely ancestral state split as in the case of splits $\{2\}$ and $\{1,2,3\}$.
-Now, if we exchange both $x_1$ with $x_2$ and $y_1$ with $y_2$, we see these generating probabilities remain unchanged, and, for the conditional likelihood values, we look at the complement of the most likely ancestral state split and see these values also remain unchanged.
+By exchanging only $x_1$ with $x_2$, the generating probabilities and likelihood values swap between these two site splits.
+The same is true of the generating probabilities if we exchange only $y_1$ and $y_2$, except, for the case of the likelihood values, we again look at the complement of the most likely ancestral state split as in the case of splits $\{2\}$ and $\{1,2,3\}$.
+Now, if we exchange both $x_1$ with $x_2$ and $y_1$ with $y_2$, we see these generating probabilities remain unchanged, and, for the likelihood values, we look at the complement of the most likely ancestral state split and see these values also remain unchanged.
 
 Thus exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$.
-Therefore we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$ and $y_2 \ge y_1$, with these conditional likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.
+Therefore we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$ and $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.
 
@@
 \hline
-$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_j$ & $32\cdot\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,t)$\\
+$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_j$ & $32\cdot\Pr(\siteSplitRV=\siteSplit_j,\ancestralSplitRV=\xi_j \mid \tau_1,t)$\\
 \hline
@@
 \caption{
-32 times conditional likelihood values on the InvFels tree $\tau_1$.
-Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
+32 times likelihood values on the InvFels tree $\tau_1$.
+Due to the symmetry of the likelihood, WLOG we let $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
 Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
-Because in 4 cases there are 3 possibilities for $\xi_j$, there are $3^4=81$ possible forms for the conditional likelihood.
+Because in 4 cases there are 3 possibilities for $\xi_j$, there are $3^4=81$ possible forms for the likelihood.
 }
@@
 
-We first prove intermediate results to show the main result.
+We begin with the branch length inconsistency.
 
-\begin{lemma}
-Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
-There exists a set of $0 < x^*, y^* < 1$ such that $\emptyset$ is the maximal ancestral state split for all site splits given $\tau=\tau_1$.
-\label{lemma:ancestral-state}
-\end{lemma}
+\blInconsist*
 
 \begin{proof}
-Recall
+For a fixed, known $\boldsymbol\xi$, there exists a closed form solution to $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ solving
 \[
-\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^C_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
+\hat{t}_{\boldsymbol\xi} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi).
 \]
-We obtain
+We show in this case that the log-likelihood $\ell$ attains a unique maximum at $\hat{t}_{\boldsymbol\xi}$.
+For fixed $\boldsymbol\xi$, the log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
+\begin{align*}
+\ell_{\tau^*,t^*}(\tau^*, t) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_2}(x_2)\\
+&\qquad + \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,w}(w).
+\end{align*}
+Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
+Without loss of generality we focus on the variable $x_1$ and the log-likelihood proportional to
 \[
-\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) \le
-    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
-    + \ell^C_{\tau^*,t^*}(\tau, \tilde{t}; \boldsymbol\xi)
+\ell(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1).
 \]
-as an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
+Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, is
 \[
-\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
+h_{j,x_1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
 \]
-and
+for $e_j\in\{0,1\}$, which, simplifying, results in
+\begin{align}
+\ell(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
+                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1). \label{eq:closed-form-x}
+\end{align}
+We have the second derivative
 \[
-\tilde{t} = \argmax_{t} \ \ell^C_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
+\ell''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
 \]
-By straightforward calculus, the function $f(x)=a_1\log(1+x)+a_2\log(1-x)$ attains a maximum at $\hat{x} = (a_1-a_2)/(a_1+a_2)$, and so we have a closed form for each component of $\tilde{t}$.
-Using this estimate, we have the lower bound
+As $x_1\in(0,1]$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
+Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\ell''(x_1) \le 0$.
+Applying similar arguments to the other variables, the Hessian for the log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero.
+
+Now, by straightforward calculus, we solve for the unique maximum $\hat{x}_1$ by setting the first derivative of \eqref{eq:closed-form-x} to zero to obtain
 \[
-\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) \ge
-    \shannonDivergence_{\tau^*,t^*}(\tau,\tilde{t})
-    + \ell^C_{\tau^*,t^*}(\tau, \tilde{t}; \boldsymbol\xi).
+\hat{x}_1 = 2\cdot\left(\sum_{j=1}^\nSiteSplits c_je_j\right) - 1
 \]
-Let $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$ and
+where
 \[
-\tilde{t}_{\boldsymbol\xi} = \argmax_{t} \ \ell^C_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi).
+\sum_{j=1}^\nSiteSplits c_je_j = \sum_{j=1}^\nSiteSplits \mathbf{1}\{\mbox{site split $j$ has term $(1+x_1)$}\} \cdot p_{\siteSplit_j}.
 \]
-Define
+As an example, Table~\ref{tab:likelihoods-restricted} shows the maximal ancestral state splits and corresponding likelihood values for $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$ and so, in this case,
 \[
-C_0(x^*, y^*) :=
-    \shannonDivergence_{\tau^*,t^*}(\tau_1,\tilde{t}_{\boldsymbol\xi_0})
-    + \ell^C_{\tau^*,t^*}(\tau_1, \tilde{t}_{\boldsymbol\xi_0}; \boldsymbol\xi_0),
+\sum_{j=1}^\nSiteSplits c_je_j = p_{\emptyset} + p_2 + p_3 + p_{23} = \frac{1}{2} + \frac{1}{2}x^*(y^*)^2
 \]
-i.e., the lower bound assuming $\emptyset$ is the maximal ancestral state split for all site splits and $\tau=\tau_1$.
-
-We need only consider the remaining possibilities for choices of ancestral state splits in Table~\ref{tab:likelihoods} upon excluding cases of infinite branch lengths (i.e., any of $x_1,y_1,x_2,y_2,w$ equal to zero) and the redundant cases of $x_1 > x_2$ and $y_1 > y_2$, i.e.,
-\[
-C_1(x^*, y^*) :=
-    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
-    +  \max_{\boldsymbol\xi : \boldsymbol\xi \neq \boldsymbol\xi_0} \ \ell^C_{\tau^*,t^*}(\tau_1, \tilde{t}_{\boldsymbol\xi}; \boldsymbol\xi).
-\]
-We show in Fig.~\ref{fig:max-anc-state} the region where $C_0(x^*,y^*) > C_1(x^*,y^*)$.
-\end{proof}
-
-\begin{figure}
-\centering
-\includegraphics[width=.95\textwidth]{analytic-anc-state-inkscape}
-\caption{
-Region where $\emptyset$ is the maximal ancestral state split for each site split on the InvFels tree $\tau_1$.
-}
-\label{fig:max-anc-state}
-\end{figure}
-
-\begin{lemma}
-Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
-For $0 < x^*, y^* < 1$ and $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
-\[
-\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi_0)
-\]
-is subject to the constraints $\hat{x}_1,\hat{x}_2 \in (0, x^*]$, $\hat{y}_1 \in (0, y^*]$, and $\hat{y}_2,\hat{w} \in [y^*, 1]$.
-\label{lemma:bounds}
-\end{lemma}
-
-\begin{proof}
-We proceed by analyzing the gradients of each component of the likelihood function.
-We first focus on the marginal log-likelihood
-\[
-\shannonDivergence_{\tau^*,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
-\]
-By Gibbs' inequality, this function attains a unique maximum when
-\[
-\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
-\]
-for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$ and $\tau^*$ being the InvFels tree $\tau_1$.
-We show that this implies $t=t^*$.
-Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
-Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
-The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
-\[
-2-2(x^*)^2 = 2-2x^2
-\]
-and for $\{2\}$ and $\{1,2\}$ yields
-\[
-2-2(y^*)^2 = 2-2y^2.
-\]
-implying $x=x^*$ and $y=y^*$.
-Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
-\[
-8x^*(y^*)^2 = 8x^*y^*w
-\]
-implying $w=y^*$.
-Thus the unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and consequently
-\[
-\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau_1,t) \mid_{t=t^*} = 0.
-\]
-We now show that the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
-The marginal log-likelihood as a function of $x_1$ is
-\begin{align*}
-\shannonDivergence(x_1) = \sum_j p_{\siteSplit_j} \log(c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1) - \log 8
-\end{align*}
-where $c_{x_1,\siteSplit_j}$ and $a_{x_1,\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
-For example, reading from the middle panel of Table~\ref{tab:gen-sitepatprob},
-\[
-c_{x_1,\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
-\]
-and
-\[
-a_{x_1,\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
-\]
-with similar definitions for the remaining site splits.
-
-%EM Could you restructure this paragraph with a topic-sentence statement of the assumption we want to make, then evidence that it's ok? It roves around a good bit.
-By the assumption of $0 < x^*, y^* < 1$, we have $p_{\siteSplit_j} > 0$ for $j=1,\ldots,\nSiteSplits$, therefore we assume $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 \neq 0$ for all $j$ otherwise we obtain a zero value for our likelihood, i.e., $\shannonDivergence\rightarrow -\infty$.
-In other words, we make the assumption that $x_1,y_1,x_2 < 1$, precluding all cases where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ for some $j$.
-This assumption is reasonable as the likelihood is never maximized at these values where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$.
-
-Moreover, we have
-\[
-\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
-\]
-which, being a probability, is greater than or equal to zero.
-Consider the derivative with respect to $x_1$
-\begin{align*}
-\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}
-\end{align*}
-and focus on the $j$th term in the summation
-\[
-f_j(x_1) := \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}.
-\]
-%das: gah, had to roll back this line and add some stuff above since there are cases where some probabilities are 0.
-Maybe I'm being too careful?
-I don't think it's an issue, just an edge case that could induce wonkiness, so I included an assumption.
-Regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 > 0$.
-The remaining variables follow similar arguments.
+and $\hat{x}_1 = x^*(y^*)^2$.
 
@@
 \caption{
-32 times the maximal conditional likelihood values on the InvFels tree $\tau_1$ assuming $\emptyset$ is the most likely ancestral state split for each site split.
+32 times the maximal likelihood values on the InvFels tree $\tau_1$ where $\emptyset$ is the most likely ancestral state split for each site split.
 }
@@
 
-We now focus on the conditional log-likelihood.
-%EM These "Assume"s feel frightening to the reader. Can we replace with "Under our stated hypothesis"?
-Assume $\emptyset$ is the most likely ancestral state split for all site splits.
-In this case, the conditional log-likelihood has no ambiguity, and has terms given as the log of those in Table~\ref{tab:likelihoods-restricted}.
-We focus on the derivative with respect to $x_1$, aggregating all terms involving $x_2,y_1,y_2,$ and $w$ into a constant term.
-The conditional log-likelihood
+We show that solutions of this form never obtain $\hat{t} = t^*$ except in cases of zero or infinite branch length.
+Given Table~\ref{tab:gen-sitepatprob}, all solutions to $\hat{x}_1$ have the form
 \[
-\ell^C_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi_0) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV = \siteSplit_j, \tau_1, t),
+\hat{x}_1 = a_{x_1,0} + a_{x_1,1}(x^*)^2 + a_{x_1,2}(y^*)^2 + a_{x_1,3} x^*(y^*)^2 + a_{x_1,4}(x^*)^2(y^*)^2.
 \]
-using Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:likelihoods-restricted}, is
-\begin{align*}
-\ell^C(x_1; \boldsymbol\xi_0) &= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
-&\qquad + \left(\frac{1}{2}+\frac{1}{2}(y^*)^2\right)\log(1+y_1) + \left(\frac{1}{2}-\frac{1}{2}(y^*)^2\right)\log(1-y_1)\\
-&\qquad + \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_2) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_2)\\
-&\qquad + \log(1+y_2) + \log(1+w) - \log 32\\
-&:= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
-&\qquad + C_{x_1}(x_2,y_1,y_2,w).
-\end{align*}
-where $C_{x_1}$ is a function of $y_1,x_2,y_2,$ and $w$.
-The derivative with respect to $x_1$ is
-\[
-\frac{d}{dx_1} \ell^C(x_1; \boldsymbol\xi_0) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2},
-\]
-which is negative for all $x_1 \ge x^*$.
-This same process applied to $x_2$ and $y_1$ shows the gradient of the full likelihood is negative for $x_2 \ge x^*$ and $y_1 \ge y^*$.
-
-The variables $y_2$ and $w$ require the following.
-Consider the conditional log-likelihood with respect to $w$
-\[
-\ell^C(w; \boldsymbol\xi_0) := \log(1+w) + C_{w}(x_1,y_1,x_2,y_2)
-\]
-with derivative with respect to $w$
-\[
-\frac{d}{dw} \ell^C(w; \boldsymbol\xi_0) = \frac{1}{1+w},
-\]
-which is positive for all $w \ge 0$.
-%EM Does the zero crossing need additional argument?
-I'm guessing that you get this from the gradient-is-zero thing from the unique-maximum argument above.
-But that's just that it's zero when we are doing maximization of all of the variables at once.
-Here aren't we wanting to make an argument for lots of x_i and y_i?
-As the gradient of the marginal log-likelihood is a decreasing function in $w$ and has a zero crossing at $y^*$, the addition of the derivative of the conditional log-likelihood makes the gradient of the full log-likelihood positive for all $w \le y^*$.
-The same argument holds for $y_2$, yielding the constraints
-\[
-\hat{x}_1,\hat{x}_2 \in (0, x^*], \ \hat{y}_1 \in (0, y^*], \text{and}\ \hat{y}_2,\hat{w} \in [y^*, 1].
-\]
+where $a_{x_1,k}$ are constants independent of $x^*$ and $y^*$---in fact, $a_{x_1,k}$ takes values in the set $\{i/8 : i=-4,-3,\ldots,7,8\}$.
+The true branch fidelity for $x_1$ is $x^*$, and the only cases when we might have $\hat{x}_1=x^*$ are when $y^*=1$ or when $(x^*)^2=x^*$, i.e., one of the generating branch parameters is zero or infinite length; the same is true for $x_2$.
+A similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent when $(y^*)^2=y^*$, i.e., $y^*=0$ or $y^*=1$.
 \end{proof}
 
-%EM Are we saying that Lemma 3 applies in this case? I think so but it could be more specific.
-Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ does not contradict with the statement of Lemma~\ref{lemma:bounds}.
-We now prove the main result.
+We now proceed to show there exist $x^*$ and $y^*$ such that the interior branch parameter $w$ is estimated as exactly one.
 
@@
 \begin{proof}
-We proceed as in the previous proof by simplifying the constants in the log-likelihood.
-We analyze the derivative with respect to $w$, so we absorb all terms of only $x_1,y_1,x_2,$ and $y_2$ into a constant that disappears upon taking the derivative.
-The objective function is
-\begin{align*}
-\ell(w; \boldsymbol\xi_0) = \left(\sum_j p_{\siteSplit_j} \log(c_{w,\siteSplit_j} + a_{w,\siteSplit_j}w)\right) + \log(1+w) + C_w(x_1,y_1,x_2,y_2)
-\end{align*}
-where $c_{w,\siteSplit_j},a_{w,\siteSplit_j}$ and $C_w$ are functions of $x_1,y_1,x_2,$ and $y_2$.
-Using the middle panel of Table~\ref{tab:gen-sitepatprob}, we obtain
+As we have a closed form solution to our likelihood problem, we compute the optimal solution given Table~\ref{tab:likelihoods}.
+Let
 \[
-c_{w,\emptyset} = 1+x_1x_2+y_1y_2+x_1y_1x_2y_2 = (1+x_1x_2)(1+y_1y_2)
+\hat{t}_{\boldsymbol\xi} = \argmax_{t} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
 \]
-and
+be the closed form solution for $t$ for a fixed maximal ancestral state split $\boldsymbol\xi$.
+We need only consider the possibilities for choices of ancestral state splits in Table~\ref{tab:likelihoods} upon excluding cases of infinite branch lengths (i.e., any of $x_1,y_1,x_2,y_2,w$ equal to zero) and the redundant cases of $x_1 > x_2$ and $y_1 > y_2$, i.e.,
 \[
-a_{w,\emptyset} = x_1y_1+x_1y_2+x_2y_1+x_2y_2 = (x_1+x_2)(y_1+y_2)
+\hat{\boldsymbol\xi} =
+    \argmax_{\boldsymbol\xi} \ \ell_{\tau^*,t^*}(\tau_1, \hat{t}_{\boldsymbol\xi}; \boldsymbol\xi).
 \]
-with similar definitions for the remaining site splits.
-We make a further simplification on our way to computing bounds on the likelihood.
-Let
-\begin{align*}
-\ell(w; \boldsymbol\xi_0) &= \left(\sum_j p_{\siteSplit_j} \log(c_{w,\siteSplit_j} + a_{w,\siteSplit_j}w)\right) + \log(1+w) + C_w(x_1,y_1,x_2,y_2) \\
-        &= \left(\sum_j p_{\siteSplit_j} \log(c_{w,\siteSplit_j} + a_{w,\siteSplit_j}w) - p_{\siteSplit_j} \log(c_{w,\siteSplit_j})\right) + \log(1+w) \\
-        &\qquad + \left(C_w(x_1,y_1,x_2,y_2) + \sum_j p_{\siteSplit_j} \log(c_{w,\siteSplit_j})\right) \\
-        &=: \left(\sum_j p_{\siteSplit_j} \log(1 + b_{w,\siteSplit_j}w)\right) + \log(1+w) + C^*_w(x_1,y_1,x_2,y_2)
-\end{align*}
-where
-\[
-b_{\emptyset} := \frac{a_{w,\emptyset}}{c_{w,\emptyset}} = \frac{(x_1+x_2)(y_1+y_2)}{(1+x_1x_2)(1+y_1y_2)}
-\]
-with similar definitions for the remaining site splits.
-The derivative with respect to $w$ of this function is
-\begin{align*}
-\ell'(w) := \frac{d}{dw} \ell(w; \boldsymbol\xi_0) =  \left(\sum_j \frac{p_{\siteSplit_j}b_{\siteSplit_j}}{1 + b_{\siteSplit_j}w}\right) + \frac{1}{1+w}.
-\end{align*}
-Using the fact that $b_{\emptyset} = -b_{13}$, $b_{1} = -b_{3}$, $b_{2} = -b_{123}$, and $b_{12} = -b_{23}$ in addition to $p_{1} = p_{3}$, $p_{2} = p_{123}$, and $p_{12} = p_{23}$, we simplify the derivative as
-\begin{align*}
-\ell'(w) &=  \frac{p_{\emptyset}b_{\emptyset}}{1 + b_{\emptyset}w} - \frac{p_{13}b_{\emptyset}}{1 - b_{\emptyset}w} + \frac{p_{2}b_{2}}{1 + b_{2}w} - \frac{p_{2}b_{2}}{1 - b_{2}w} \\
-         &\qquad + \frac{p_{1}b_{1}}{1 + b_{1}w} - \frac{p_{1}b_{1}}{1 - b_{1}w} + \frac{p_{12}b_{12}}{1 + b_{12}w} - \frac{p_{12}b_{12}}{1 - b_{12}w} + \frac{1}{1+w}.\\
-\end{align*}
-Using the bounds on $\hat{x}_1,\hat{y}_1,\hat{x}_2$, and $\hat{y}_2$ from Lemma~\ref{lemma:bounds} and our assumption of $y_2 \ge y_1 > 0$ and $x_2 \ge x_1 > 0$, we have
-\[
-0 < b_{\emptyset} \le \frac{2x^*}{1+(x^*)^2} =: b_{\emptyset}^U,
-\]
-%das: The lower bound for b_{\emptyset} is strict because we assume x_1,...,w > 0, but for the other three they aren't strict since x_1 could equal x_2 and/or y_1 could equal y_2.
-This doesn't affect any later results, though, b/c b -> 0 just makes that term drop out.
-\[
-0 \le b_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*} =: b_{2}^U,
-\]
-\[
-0 \le b_{1} \le \frac{x^*}{1-(x^*)^2} =: b_{1}^U,
-\]
-\[
-0 \le b_{12} \le \frac{x^*}{1-(x^*)^2}\cdot\frac{1}{1-y^*} =: b_{12}^U,
-\]
-so that $\ell'(w)$ has lower bound
-\[
-\ell'_{L}(w) :=  -\frac{p_{13}}{1/b_{\emptyset}^U - w} - \frac{p_{2}}{1/b_{2}^U - w} - \frac{p_{1}}{1/b_{1}^U - w} - \frac{p_{12}}{1/b_{12}^U - w} + \frac{1}{1+w}.
-\]
-Each term in $\ell'_{L}(w)$ is of the form
-\[
-f(w) = \frac{p}{1/b + w} = \frac{p\cdot b}{1 + b\cdot w}
-\]
-with $0 \le p \le 1$.
-Assume $b_{\emptyset}^U, b_{2}^U, b_{1}^U, b_{12}^U < 1$, so that $-1 < b < 1$.
-Since $b > -1$ and $w\in(0,1]$, $1+b\cdot w > 0$ and, thus, each term in $\ell'_{L}(w)$ is a decreasing function in $w$.
-Thus, if $\ell'_{L}(1) > 0$, since $\ell'_{L}(w)$ is decreasing on $(0,1]$ then $\ell'_{L}(w) > 0$ for all $w\in (0,1]$.
-Since $\ell'(w) > \ell'_{L}(w)$ for $w\in(0,1]$, this implies $\ell'(w) > 0$, which further implies $\ell(w; \boldsymbol\xi_0)$ is increasing on the entire interval $(0,1]$, including at the boundary $w=1$, and $\hat{w} \equiv 1$.
+We show the maximal ancestral states in Fig.~\ref{fig:max-anc-state}.
 
-We now construct the conditions on $x^*$ and $y^*$ that are required for $\hat{w}\equiv 1$:
-\[
-C1-C3: b_{2}^U, b_{1}^U, b_{12}^U < 1,
-\]
-\[
-C4: -\frac{p_{13}b_{\emptyset}^U}{1-b_{\emptyset}^U} - \frac{p_{2}b_{2}^U}{1-b_{2}^U} - \frac{p_{1}b_{1}^U}{1-b_{1}^U} - \frac{p_{12}b_{12}^U}{1-b_{12}^U} + \frac{1}{2} > 0.
-\]
-Conditions $C1-C3$ are needed so that $\ell'_{L}(w)$ is a decreasing function on $(0,1]$, while condition $C4$ is the one on $\ell'_{L}(1) > 0$, i.e., so that $\ell'(w)$ is positive on $(0,1]$.
-The set of $x^*,y^*$ satisfying conditions $C1-C4$ as well as those for $\emptyset$ being the most likely ancestral state split is given in Fig.~\ref{fig:incons-analytic}, giving the region of generating parameters where $\hat{w}\equiv 1$.
+\begin{figure}
+    \begin{minipage}{0.95\linewidth}
+        \centering
+        \includegraphics[width=.95\textwidth]{analytic-anc-state}
+        \label{ }
+    \end{minipage}
+    \begin{minipage}{0.95\linewidth}
+        \centering
+        \begin{tabular}{ll}
+        \hline
+        $\boldsymbol\xi_1$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset\}$\\
+        $\boldsymbol\xi_2$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \emptyset, \emptyset\}$\\
+        $\boldsymbol\xi_3$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \emptyset, \{1\}\}$\\
+        $\boldsymbol\xi_4$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \{1,2\}, \{1\}\}$\\
+        $\boldsymbol\xi_5$ & $\{\emptyset, \emptyset, \emptyset, \{1\}, \{1\}, \emptyset, \{1\}, \{1\}\}$
+        \end{tabular}
+    \end{minipage}
+\caption{
+Region with maximal ancestral state splits on the InvFels tree $\tau_1$.
+}
+\label{fig:max-anc-state}
+\end{figure}
+
+Mapping each maximal ancestral state split to each likelihood value, we see that $\hat{w}\equiv 1$ if $\boldsymbol\xi=\boldsymbol\xi_1$ or $\boldsymbol\xi=\boldsymbol\xi_2$, which encompasses the bottom-right region of Figure~\ref{fig:max-anc-state}.
+This is an analytically-derived region of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
 \end{proof}
 
-\begin{figure}
-\centering
-\includegraphics[width=.95\textwidth]{incons-analytic-inkscape}
-\caption{
-    An analytically-derived region of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
-    Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
-}
-\label{fig:incons-analytic}
-\end{figure}
-
-As the region in Fig.~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
-It is easy to see that assuming $\emptyset$ for each maximal ancestral state split results in the same conditional log-likelihood for $\tau_1$ and $\tau_2$.
+As the region of degeneracy in Fig.~\ref{fig:max-anc-state} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
+%EM "letting $\emptyset$ be the maximal ancestral state split" sounds a little funny-- either it is or it isn't maximal.
+It is easy to see that when $\emptyset$ is the maximal ancestral state split, we have the same log-likelihood for $\tau_1$ and $\tau_2$.
 Moreover, if $w=1$, the internal branch becomes zero-length and the two topologies are indistinguishable.
-Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the shaded region in Fig.~\ref{fig:incons-analytic}.
+Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the shaded region in Fig.~\ref{fig:max-anc-state}.
 We can see this results in the likelihood of both topologies being equal, i.e.,
@@
 \max_{t : t^*\in\mathcal{T}_0} & \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)\\
-&= \max_{t: \boldsymbol\xi=\boldsymbol\xi_0, w=1, \tau=\tau_1} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
-&\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\})\\
-&= \max_{t: \boldsymbol\xi=\boldsymbol\xi_0, w=1, \tau=\tau_2} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}) \\
-&\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\}).
+&= \max_{t: \boldsymbol\xi=\boldsymbol\xi_0, w=1, \tau=\tau_1} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
+&= \max_{t: \boldsymbol\xi=\boldsymbol\xi_0, w=1, \tau=\tau_2} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}).
 \end{align*}

### 2018-09-19 joint_inf.tex
:tennis:
version 3...

@@
 \author[1]{David A. Shaw}
+\author[2]{Vu C. Dinh}
 \author[1]{Frederick A. Matsen IV\thanks{Corresponding author. Email: \url{matsen@fredhutch.org}}}
 \affil[1]{Computational Biology Program, Fred Hutchinson Cancer Research Center\\ Seattle, WA, USA}
+\affil[2]{Department of Mathematical Sciences, University of Delaware\\ Newark, DE, USA}
 \date{}
@@
 We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
-For branch length estimation on the correct topology, we outline cases where branch length estimates are consistently and heavily biased and provide extensive empirical results for this setting.
+For branch length estimation on the correct topology, we show that this joint method cannot estimate consistent branch lengths except in degenerate cases, and we provide extensive empirical results for outlining the consistent bias in this setting.
 \end{abstract}
@@
 To do so, we use a binary symmetric model with data generated on a four-taxon tree: we construct bounds on the joint objective function to demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
-We find similar areas through empirical means where joint inference consistently underestimates interior branch lengths when the topology is known and fixed.
+We show that, when the topology is known and fixed, joint inference cannot be consistent except in cases of zero or infinite branch length, and we find similar areas through empirical means where joint inference consistently underestimates interior branch lengths.
 
@@
 
+\subsection*{Inconsistent branch length estimation}
+
+When the topology is known and fixed and we estimate only branch lengths, we show the following, i.e., that for almost all $x^*$ and $y^*$ any branch length estimate is consistently biased.
+
+\begin{restatable}{theorem}{blInconsist}
+\label{thm:bl_theorem}
+Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
+For all $0 < x^*, y^* < 1$, given some $\boldsymbol\xi$ the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
+\[
+\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)
+\]
+has the property $\hat{t} \neq t^*$.
+\end{restatable}
+
+%TODO: more discussion here
+In words, the joint estimation procedure never recovers the true generating $t^*$ except in cases of zero or infinite branch length.
+
+\subsection*{Convergence to degenerate topology}
+
 Our main theorem states that given data generated on $\tau_1$ there exist true nonzero branch lengths such that the estimator $\hat{t}$ maximizing the right-hand side of \eqref{eq:exp-log-lik} has an internal branch of length zero.
@@
 %das: ok, I reworded it a bit; let me know how you think it sounds.
+%EM This is better, but the notation still seems a little funny.
+Namely, xi is really something that's determined by Y, but the notation seems to hint that it can be specified a priori.
+The point is that the ML xi has problems, not that there exists a xi that has problems.
+%das: ok, edited
 \begin{restatable}{theorem}{topoInconsist}
-\label{thm:main_theorem}
+\label{thm:topo_theorem}
 Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
-%EM Can we say "open set" or "set of positive measure"?
-There exists an open set of $0 < x^*, y^* < 1$ such that for some $\boldsymbol\xi$ the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
+There exists an open set of $0 < x^*, y^* < 1$ such that the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
 \[
-\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)
+\hat{t} = \arg\max_{t} \ \max_{\boldsymbol\xi} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)
 \]
@@
 
-This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:incons-analytic}).
+This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:max-anc-state}).
 As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
@@
 %EM On the other hand we should still refer to it in the text around here.
+%das: it's referred to in the first sentence of above paragraph---good?
 
@@
 
-Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically\hyp{}derived region (Fig.~\ref{fig:incons-analytic}).
-To determine how conservative it is, we use the L-BFGS-B method to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
-We see that the region of inconsistency in the general case is much larger than in the analytic case, which is unsurprising because the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral states.
-This region encompasses almost half of the branch fidelity space; given the correct topology and performing our best possible optimization, there are many situations where we estimate the interior branch length to be zero.
+Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the analytically\hyp{}derived region (Fig.~\ref{fig:max-anc-state}).
+To verify the regions of inconsistency and obtain a clearer picture of the closed form parameter estimates, we plot the optimal $\hat{w}$ via joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
+As before, the region of inconsistency encompasses almost half of the branch fidelity space; given the correct topology and performing our best possible optimization, there are many situations where we estimate the interior branch length to be zero.
 
@@
 \caption{
-    Numerical estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
-    Data generated as in Fig.~\ref{fig:incons-analytic}.
+    Estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
+    Data generated as in Fig.~\ref{fig:max-anc-state}.
     The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero, resulting in an inconsistency.
@@
 
-We provide a full description of our optimization procedure in the Appendix, but briefly, we analytically reduce the general case to 81 (see Table~\ref{tab:likelihoods}) and perform maximizations for each.
+In our optimization procedure, we analytically reduce the general case to 81 separate cases (see Table~\ref{tab:likelihoods}) and obtain a closed form maximal value for each.
 For each function, we optimize the objective starting at five separate initial values to mitigate problems with small curvature of the objective surface and nonconcavity.
@@
 Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$.
-Inference on the integrated likelihood performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same optimization code.
-Moreover, the errors in this case are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
+
+Inference on the integrated likelihood performs as expected where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}).
+We use L-BFGS-B when optimizing \eqref{eq:marginal_likelihood}.
+The errors in this case are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
 
@@
 However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and that the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).
+In addition, we have shown there are no nontrivial generating parameters that yield consistent branch length estimates.
 
@@
 %EM I don't follow here. This doesn't seem like the long branch attraction region of parameter space-- isn't that on the Felsenstein tree, not the InvFels tree?
-%das: yeah, it was kind of hand-wavy.
+%das: yeah, it was kind of hand-wavy. edited
 This is not the case here; while analytically the inconsistency occurs when the top two branches are long and the bottom three are short, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
-%EM Is this next sentence necessary given the previous one?
 Additionally, we generate data on the InvFels tree $\tau_1$ while \citet{Felsenstein1978-rr} generates data on the Felsenstein tree $\tau_2$.
@@
 There is substantial scope for future work to make these results more precise and more general.
-The techniques used to obtain upper and lower bounds for the likelihoods in Lemma~\ref{lemma:bounds} provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
-For the case of estimating branch lengths, we are able to provide empirical estimates for regions of overestimation that are tighter than the analytical bounds, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) beg further investigation.
 All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners.
@@
 We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.
-Insight from the reviewers, the editors, and Vu Dinh greatly improved this manuscript.
+Insight from the reviewers and the editors greatly improved this manuscript.
 

### 2018-09-19 joint_inf_plot.py
v3 plotting code

@@ -30,6 +30,7 @@ FONT_SIZE = 20
 # constants to slightly speed up computation
 LOG2 = np.log(2)
 LOG32 = np.log(32)
+ANC_STATE_DICT = {'0': (0,0), '1': (1,0), '2': (1,1), '3': (0,1)}
 
 
 class Legend(object):
@@ -79,14 +80,12 @@ def parse_args():
         type=str,
         help="""
              choices:
-             ancestral_state_conditions-- conditions where \emptyset is most likely ancestral state (Fig. S2),
-             analytic_inconsistency-- plot conditions where \emptyset is most likely ancestral state and \hat{w} is one analytically (Fig. 2),
+             ancestral_state_conditions-- ,
              joint_empirical-- plot empirical estimate of \hat{w} for joint inference (Fig. 3),
              marginal_empirical-- plot empirical estimate of \hat{w} for marginal inference (Fig. S4),
              """,
         choices=(
              'ancestral_state_conditions',
-             'analytic_inconsistency',
              'joint_empirical',
              'marginal_empirical',
             ),
@@ -95,6 +94,13 @@ def parse_args():
         '--bias',
         action="store_true",
         help='plot and print bias instead of parameter estimate (Fig. S3)',
+    )
+    parser.add_argument(
+        '--generating-topology',
+        type=str,
+        choices=('invfels', 'fels'),
+        default='invfels',
+        help='which topology generated the data',
     )
     parser.add_argument(
         '--in-pkl-name',
@@ -117,21 +123,32 @@ def parse_args():
 # Functions for exact likelihood bounds
 
 
-def P_INVFELS(theta0, site_pattern):
+def get_generating_probability(theta0, site_pattern, generating_topology='invfels'):
     """
     Site pattern frequencies for inverse Felsenstein tree
     """
     # fifth branch will always be positive
     theta = [-t if s=='1' else t \
             for t,s in zip(theta0, site_pattern+'0')]
-    return 0.125 * (1 + \
-         theta[0]*theta[2] + \
-         theta[1]*theta[3] + \
-         theta[0]*theta[1]*theta[4] + \
-         theta[0]*theta[3]*theta[4] + \
-         theta[1]*theta[2]*theta[4] + \
[…]

### 2018-09-21 appendix.tex
minor changes; updated discussion

@@
     \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t)\\
-    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)\\
-    &= \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)\\
+    &= \Pr((\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i) \cup (\overline{\alignmentColumnRV}=\alignmentColumn_i, \overline{\ancestralStateColumnRV}=\ancestralStateColumn_i) \mid \tau, t)\\
+    &= \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i, \overline{\ancestralStateColumnRV}=\ancestralStateColumn_i \mid \tau, t)\\
     &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)
 \end{align*}
-where $\overline{Y}$ is the complement of the site pattern--valued random variable $Y$ and has the same distribution as $Y$.
+where $\overline{\alignmentColumnRV}$ is the complement of the site pattern--valued random variable $\alignmentColumnRV$ and has the same distribution as $\alignmentColumnRV$ (similarly for $\ancestralStateColumnRV$).
 Since
 \begin{align*}
-    \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t),
+    2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t),
 \end{align*}
@@
 \begin{align*}
-    \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t)\\
+    \max_{\ancestralStateColumn_i} \ 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t)\\
     &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)\\
@@
     \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
-& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
+& =: \{\siteSplit_1, \ldots, \siteSplit_8\}.
 \end{align*}
@@
                              &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
-                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
-                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
-                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \nonumber \\
+                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t) \nonumber \\
+                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \nonumber \\
                              &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
@@
 &    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
-&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)].
+&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t).
 \label{eq:site_pattern_profile_likelihood_mean}
@@
 \[
-m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)]
+m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)
 \]
@@
 Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
-Next we show that the likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
+Next we show that the likelihood $\ell_{\tau_1,t}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
 Although this property should not be surprising due to symmetry, we write it out for completeness.
+This holds for a general $t$, and thus holds setting $t=t^*$.
 Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
@@
 
-Thus exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$.
+Thus exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t}(\tau_1, t; \boldsymbol\xi)$.
 Therefore we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$ and $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.
@@
 
-We begin with the branch length inconsistency.
+We begin by showing an inconsistency in branch length estimation on the InvFels tree.
 
@@
 \begin{align*}
-\ell_{\tau^*,t^*}(\tau^*, t) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_2}(x_2)\\
+\ell_{\tau^*,t^*}(\tau^*, t, \boldsymbol\xi) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_2}(x_2)\\
 &\qquad + \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,w}(w).
@@
 \ell(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
-                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1). \label{eq:closed-form-x}
+                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1), \label{eq:closed-form-x}
 \end{align}
-We have the second derivative
+which has second derivative
 \[
@@
 Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\ell''(x_1) \le 0$.
-Applying similar arguments to the other variables, the Hessian for the log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero.
+Applying similar arguments to the other variables, the Hessian for the log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero, and $\hat{t}$ uniquely maximizes $\ell$.
 
@@
 \]
-As an example, Table~\ref{tab:likelihoods-restricted} shows the maximal ancestral state splits and corresponding likelihood values for $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$ and so, in this case,
+As an example, Table~\ref{tab:likelihoods-restricted} shows the maximal ancestral state splits and corresponding likelihood values for $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$.
+In this case,
 \[
@@
 where $a_{x_1,k}$ are constants independent of $x^*$ and $y^*$---in fact, $a_{x_1,k}$ takes values in the set $\{i/8 : i=-4,-3,\ldots,7,8\}$.
-The true branch fidelity for $x_1$ is $x^*$, and the only cases when we might have $\hat{x}_1=x^*$ are when $y^*=1$ or when $(x^*)^2=x^*$, i.e., one of the generating branch parameters is zero or infinite length; the same is true for $x_2$.
+The true branch fidelity for $x_1$ is $x^*$, and the only cases to possibly obtain $\hat{x}_1=x^*$ are when $y^*=1$ or when $(x^*)^2=x^*$, i.e., one of the generating branch parameters is zero or infinite length; the same is true for $x_2$.
 A similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent when $(y^*)^2=y^*$, i.e., $y^*=0$ or $y^*=1$.
@@
 
-We now proceed to show there exist $x^*$ and $y^*$ such that the interior branch parameter $w$ is estimated as exactly one.
+We now proceed to show there exist $x^*$ and $y^*$ such that the interior branch parameter $w$ is estimated as exactly one, indicating convergence to a multifurcating topology.
 
@@
 \begin{proof}
-As we have a closed form solution to our likelihood problem, we compute the optimal solution given Table~\ref{tab:likelihoods}.
+As we have a closed form solution to our likelihood problem, we compute the optimal solution given Table~\ref{tab:farris_likelihoods}.
 Let
@@
 be the closed form solution for $t$ for a fixed maximal ancestral state split $\boldsymbol\xi$.
-We need only consider the possibilities for choices of ancestral state splits in Table~\ref{tab:likelihoods} upon excluding cases of infinite branch lengths (i.e., any of $x_1,y_1,x_2,y_2,w$ equal to zero) and the redundant cases of $x_1 > x_2$ and $y_1 > y_2$, i.e.,
+We need only consider the possibilities for choices of ancestral state splits in Table~\ref{tab:likelihoods} as opposed to Table~\ref{tab:farris_likelihoods}.
+Upon excluding cases of infinite branch lengths (i.e., any of $x_1,y_1,x_2,y_2,w$ equal to zero) and the redundant cases of $x_1 > x_2$ and $y_1 > y_2$, we obtain
 \[
@@
         \hline
-        $\boldsymbol\xi_1$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset\}$\\
-        $\boldsymbol\xi_2$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \emptyset, \emptyset\}$\\
-        $\boldsymbol\xi_3$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \emptyset, \{1\}\}$\\
-        $\boldsymbol\xi_4$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \{1,2\}, \{1\}\}$\\
-        $\boldsymbol\xi_5$ & $\{\emptyset, \emptyset, \emptyset, \{1\}, \{1\}, \emptyset, \{1\}, \{1\}\}$
+        $\hat{\boldsymbol\xi_1}$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset\}$\\
+        $\hat{\boldsymbol\xi_2}$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \emptyset, \emptyset\}$\\
+        $\hat{\boldsymbol\xi_3}$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \emptyset, \{1\}\}$\\
+        $\hat{\boldsymbol\xi_4}$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \{1,2\}, \{1\}\}$\\
+        $\hat{\boldsymbol\xi_5}$ & $\{\emptyset, \emptyset, \emptyset, \{1\}, \{1\}, \emptyset, \{1\}, \{1\}\}$
         \end{tabular}
@@
 \caption{
-Region with maximal ancestral state splits on the InvFels tree $\tau_1$.
+Regions of maximal ancestral state splits on the InvFels tree $\tau_1$.
 }
@@
 
-Mapping each maximal ancestral state split to each likelihood value, we see that $\hat{w}\equiv 1$ if $\boldsymbol\xi=\boldsymbol\xi_1$ or $\boldsymbol\xi=\boldsymbol\xi_2$, which encompasses the bottom-right region of Figure~\ref{fig:max-anc-state}.
-This is an analytically-derived region of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
+Mapping each maximal ancestral state split to each likelihood value, we see that $\hat{w}\equiv 1$ if $\hat{\boldsymbol\xi}=\hat{\boldsymbol\xi}_1$ or $\hat{\boldsymbol\xi}=\hat{\boldsymbol\xi}_2$, which encompasses the bottom-right region of Figure~\ref{fig:max-anc-state}.
 \end{proof}
 
+The regions in Fig.~\ref{fig:max-anc-state} are analytically-derived regions of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
 As the region of degeneracy in Fig.~\ref{fig:max-anc-state} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
@@
 Moreover, if $w=1$, the internal branch becomes zero-length and the two topologies are indistinguishable.
-Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the shaded region in Fig.~\ref{fig:max-anc-state}.
+Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the region in Fig.~\ref{fig:max-anc-state} where $\hat{\boldsymbol\xi}=\hat{\boldsymbol\xi}_1$.
 We can see this results in the likelihood of both topologies being equal, i.e.,
@@
 \max_{t : t^*\in\mathcal{T}_0} & \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)\\
-&= \max_{t: \boldsymbol\xi=\boldsymbol\xi_0, w=1, \tau=\tau_1} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
-&= \max_{t: \boldsymbol\xi=\boldsymbol\xi_0, w=1, \tau=\tau_2} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}).
+&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_1} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
+&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_2} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}).
 \end{align*}

### 2018-09-21 joint_inf.tex
minor changes; updated discussion

@@
 
-In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model with data generated on a four-taxon tree: we construct bounds on the joint objective function to demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
+In this paper, we show that jointly inferring trees and ancestral sequences is not consistent in general.
+To do so, we use a binary symmetric model with data generated on a four-taxon tree: we compute closed form solutions to the joint objective function and demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
 We show that, when the topology is known and fixed, joint inference cannot be consistent except in cases of zero or infinite branch length, and we find similar areas through empirical means where joint inference consistently underestimates interior branch lengths.
@@
 In the standard configuration of this tree, the interior branch parameters are equal to the bottom two parameters as in Fig.~\ref{fig:farris-fels-top}a.
-We use this standard configuration as our data generating process, though we do not constrain our branch parameters to be equal in considering our optimization objective function.
+We use this standard configuration as our data generating process, though we do not constrain our branch parameters to be equal when optimizing our objective function.
 
@@
 
-Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ (Fig.~\ref{fig:farris-fels-top}a).
+Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ (Fig.~\ref{fig:farris-fels-top}a) and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$.
 Let $\boldsymbol\xi = [\xi_j]_{j=1}^\nSiteSplits$ be the vector of most likely ancestral state splits---the explicit definition for $\boldsymbol\xi$ is given in the Appendix.
@@
 
-When the topology is known and fixed and we estimate only branch lengths, we show the following, i.e., that for almost all $x^*$ and $y^*$ any branch length estimate is consistently biased.
+When the topology is known and fixed and we estimate only branch lengths, we show the following, i.e., that for all $x^*$ and $y^*$ in $(0,1)$ any branch length estimate is consistently biased.
 
@@
 Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
-For all $0 < x^*, y^* < 1$, given some $\boldsymbol\xi$ the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
+For all $0 < x^*, y^* < 1$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
 \[
-\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)
+\hat{t} = \arg\max_{t} \ \max_{\boldsymbol\xi} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)
 \]
@@
 
-%TODO: more discussion here
 In words, the joint estimation procedure never recovers the true generating $t^*$ except in cases of zero or infinite branch length.
+This is apparent given Table~\ref{tab:gen-sitepatprob}, as the solution $\hat{t}$ is a linear combination of $p_{\siteSplit_j}$ values, and no generating probability contains an $x^*$ or $y^*$ term.
 
@@
 
-Our main theorem states that given data generated on $\tau_1$ there exist true nonzero branch lengths such that the estimator $\hat{t}$ maximizing the right-hand side of \eqref{eq:exp-log-lik} has an internal branch of length zero.
+Given data generated on $\tau_1$ there exist true nonzero branch lengths such that the estimator $\hat{t}$ maximizing the right-hand side of \eqref{eq:exp-log-lik} has an internal branch of length zero.
 %EM A funny thing: this theorem as stated is really under the hypothesis that the ancestral state is emptyset.
@@
 We elaborate on this point in the Appendix.
-The proof is through analyzing upper and lower bounds of the potential conditional likelihood forms in Table~\ref{tab:farris_likelihoods} and the gradients of the objective function $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ with respect to $t$.
+The proof is through analytically reducing the general case to 81 separate cases (Table~\ref{tab:likelihoods}) to obtain a closed form maximal value for each.
 
@@
 Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
-This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which the most likely ancestral states are the same for each internal node letting $x_1=x_2=x^*$ and $y_1=y_2=y^*$ ($\boldsymbol\xi=\emptyset$ in Table~\ref{tab:likelihoods}).
+This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which the most likely ancestral states are the same for each internal node letting $x_1=x_2=x^*$ and $y_1=y_2=y^*$ ($\xi_j=\emptyset$ for all $j$ in Table~\ref{tab:likelihoods}).
 If we allow multifurcating trees in our inference, then we can think of this as an instance of converging to the wrong topology, as the true $y^*\neq 1$.
@@
 To verify the regions of inconsistency and obtain a clearer picture of the closed form parameter estimates, we plot the optimal $\hat{w}$ via joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
-As before, the region of inconsistency encompasses almost half of the branch fidelity space; given the correct topology and performing our best possible optimization, there are many situations where we estimate the interior branch length to be zero.
+As before, the region of inconsistency encompasses almost half of the branch fidelity space; given the correct topology, there are many situations where we estimate the interior branch length to be zero.
 
@@
 
-In our optimization procedure, we analytically reduce the general case to 81 separate cases (see Table~\ref{tab:likelihoods}) and obtain a closed form maximal value for each.
-For each function, we optimize the objective starting at five separate initial values to mitigate problems with small curvature of the objective surface and nonconcavity.
+In our optimization procedure, we again consider the 81 separate cases (Table~\ref{tab:likelihoods}) and, for each function, we compute the closed form solution for $\hat{t}$.
 We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*,y^*\in(0,1)$.
@@
 Although the estimates for $\hat{p}_w$ are better than the estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), joint inference still predictably underestimates the interior branch length.
-Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$.
+Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ range from $[-4\times 10^{-2}, 3\times 10^{-3}]$.
 
@@
 This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3).
-However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and that the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).
+However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).
 In addition, we have shown there are no nontrivial generating parameters that yield consistent branch length estimates.
@@
 There, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general.
-We show cases where the incorrect topology attains an equal likelihood value at the maximum as the correct topology, and, moreover, if we know the correct topology, we show cases where branch lengths are severely biased.
+We show cases where the incorrect topology attains an equal likelihood value at the maximum as the correct topology, and, moreover, if we know the correct topology, we show cases where branch lengths are severely biased and cannot be consistent.
 Finally, just prior to his conclusion, he discusses when parsimony gives the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.

### 2018-09-21 plotting/joint_inf_plot.py
update plot and code

@@ -436,7 +436,7 @@ def main(args=sys.argv[1:]):
         cnt = 1
         for idx, anc_state in enumerate(product('012', repeat=4)):
             if idx in uniques:
-                legendtext.append(r'$\xi_%d$' % cnt)
+                legendtext.append(r'$\hat{\xi}_%d$' % cnt)
                 cnt += 1
         move_legend = True
     elif args.plot_type == 'joint_empirical':

### 2018-09-26 plotting/joint_inf_helpers.py
code documentation and pub. quality figs

@@ -1,13 +1,9 @@
+"""
+Code to print generating probabilities and likelihoods for Felsenstein and
+inverse Felsenstein trees.
+"""
+
 from __future__ import unicode_literals
-import matplotlib
-matplotlib.use('SVG')
-matplotlib.rcParams['text.usetex'] = True
-matplotlib.rcParams['text.latex.unicode'] = True
-matplotlib.rcParams['figure.autolayout'] = True
-import matplotlib.pyplot as plt
-import matplotlib.patches as mpatches
-import seaborn as sns
-import numpy as np
 import argparse
 import sys
 import collections

### 2018-09-26 plotting/joint_inf_plot.py

@@ -1,9 +1,14 @@
-from __future__ import unicode_literals
+"""
+Code to generate plots in manuscript (Figs. 2, 3, S2 and S3).
+"""
+
 import matplotlib
 matplotlib.use('SVG')
 matplotlib.rcParams['text.usetex'] = True
 matplotlib.rcParams['text.latex.unicode'] = True
 matplotlib.rcParams['figure.autolayout'] = True
+matplotlib.rcParams['text.latex.preamble']=["\usepackage{amsmath}"]
+
 import matplotlib.pyplot as plt
 import matplotlib.patches as mpatches
 import seaborn as sns
@@ -21,6 +26,8 @@ from joint_inf_helpers import ALL_PATTERNS
 
 sns.set_style('white')
 sns.set_style('ticks')
+
+plt.rcParams['hatch.color'] = '#cdcdcd'
 
 # ~~~~~~~~~
 # global variables
@@ -339,16 +346,15 @@ def make_plot(plottitle, plotname, legendtext=None, ct=None, scale=1, plotbar=Fa
     plt.ylabel(r'$p_{y^*}$', fontsize=FONT_SIZE-4)
     ttl = plt.title(plottitle, fontsize=FONT_SIZE+2)
     ttl.set_position([.5, 1.05])
-    if ct is not None:
-        ct.ax.tick_params(labelsize=FONT_SIZE-2)
-        x = np.concatenate(([0.], [cutoff], [cutoff]))
-        y = np.concatenate(([cutoff], [cutoff], [0.]))
-        plt.plot(x, y, '-', alpha=.15, lw=.5)
     if legendtext is not None:
         if move_legend:
             artists, labels = ct.legend_elements()
+            for artist in artists:
+                artist.set_alpha(1.0)
+                artist.set_edgecolor('black')
+                artist.set_linewidth(1)
             plt.legend(artists, legendtext,
-                bbox_to_anchor=(1.05, 1), loc=2, borderaxespad=0., fontsize=FONT_SIZE-4, mode='expand')
+                bbox_to_anchor=(1.05, 1), loc=2, fontsize=FONT_SIZE-4)
         else:
             plt.legend([Legend()], [legendtext], loc=2,
                 handler_map={Legend: LegendHandler()}, fontsize=FONT_SIZE-4)
@@ -363,17 +369,17 @@ def make_plot(plottitle, plotname, legendtext=None, ct=None, scale=1, plotbar=Fa
[…]

### 2018-10-01 appendix.tex
oops hat problem
code documentation and pub. quality figs
make plot the same size
rename xi subscripts

@@
 \end{align*}
-given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
+given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_{\siteSplit_j}\in\ancestralSplitPartition_j(\tau, t)$ satisfies
 \begin{align*}
-\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_j \mid \tau, t).
+\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_{\siteSplit_j} \mid \tau, t).
 \end{align*}
 In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subseteq\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, i.e., $\ancestralSplitPartition_j(\tau, t)$ is a set of sets of most likely internal node labels.
-Here, $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
+Here, $\xi_{\siteSplit_j}$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
 For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
@@
     &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)\\
-    &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_j \mid \tau, t)
+    &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_{\siteSplit_j} \mid \tau, t)
 \end{align*}
@@
 The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
-Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, t)$ we have
+Then, taking $\xi_{\emptyset}\in\ancestralSplitPartition_1(\tau, t)$ we have
 \[
-\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\xi_1 \mid \tau, t) =
+\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\xi_{\emptyset} \mid \tau, t) =
 \Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\emptyset \mid \tau, t) =
@@
 \]
-Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, t)$,
+Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_{12}\in\ancestralSplitPartition_5(\tau, t)$,
 \[
-\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\xi_5 \mid \tau, t) =
+\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\xi_{12} \mid \tau, t) =
 p^2(1-p)^3.
@@
                              &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t) \nonumber \\
-                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \nonumber \\
-                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
+                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \nonumber \\
+                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
 \end{align}
-for $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ and some $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.
+for $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ and some $\xi_{\siteSplit_j}\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.
 
@@
 \[
-L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)}
+L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)}
 \]
@@
 \begin{equation*}
-\frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)
+\frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)
 \end{equation*}
@@
 &    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
-&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t).
+&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t).
 \label{eq:site_pattern_profile_likelihood_mean}
@@
 \[
-m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)
+m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)
 \]
@@
 \hline
-$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_j$ & $32\cdot\Pr(\siteSplitRV=\siteSplit_j,\ancestralSplitRV=\xi_j \mid \tau_1,t)$\\
+$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_{\siteSplit_j}$ & $32\cdot\Pr(\siteSplitRV=\siteSplit_j,\ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau_1,t)$\\
 \hline
@@
 Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
-Because in 4 cases there are 3 possibilities for $\xi_j$, there are $3^4=81$ possible forms for the likelihood.
+Because in 4 cases there are 3 possibilities for $\xi_{\siteSplit_j}$, there are $3^4=81$ possible forms for the likelihood.
 }
@@
 \begin{figure}
-    \begin{minipage}{0.95\linewidth}
+    \begin{minipage}{\linewidth}
         \centering
@@
     \end{minipage}
-    \begin{minipage}{0.95\linewidth}
+    \begin{minipage}{\linewidth}
         \centering
-        \begin{tabular}{ll}
+        Maximal ancestral state split definitions
+        \begin{tabular}{l}
         \hline
-        $\hat{\boldsymbol\xi_1}$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset\}$\\
-        $\hat{\boldsymbol\xi_2}$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \emptyset, \emptyset\}$\\
-        $\hat{\boldsymbol\xi_3}$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \emptyset, \{1\}\}$\\
-        $\hat{\boldsymbol\xi_4}$ & $\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \{1,2\}, \{1\}\}$\\
-        $\hat{\boldsymbol\xi_5}$ & $\{\emptyset, \emptyset, \emptyset, \{1\}, \{1\}, \emptyset, \{1\}, \{1\}\}$
+        $\hat{\boldsymbol\xi}=\{\xi_{\emptyset}, \xi_{1}, \xi_{2}, \xi_{3}, \xi_{123}, \xi_{12}, \xi_{23}, \xi_{13}\}$\\
+        \hline
+        $\hat{\boldsymbol\xi}_1=\{\emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset\}$\\
+        $\hat{\boldsymbol\xi}_2=\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \emptyset, \emptyset\}$\\
+        $\hat{\boldsymbol\xi}_3=\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \{1\}, \emptyset\}$\\
+        $\hat{\boldsymbol\xi}_4=\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \{1,\}, \{1,2\}\}$\\
+        $\hat{\boldsymbol\xi}_5=\{\emptyset, \emptyset, \emptyset, \{1\}, \{1\}, \emptyset, \{1\}, \{1\}\}$\\
+        \hline
         \end{tabular}
@@
 \max_{t : t^*\in\mathcal{T}_0} & \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)\\
-&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_1} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
-&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_2} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}).
+&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_1} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
+&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_2} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, \{x_1,y_1,x_2,y_2,w\}).
 \end{align*}

### 2018-10-01 plotting/joint_inf_helpers.py
probability tests (thanks Vu!) and some more checks/log output

@@ -8,6 +8,10 @@ import argparse
 import sys
 import collections
 
+from sympy import sympify, expand
+from operator import mul
+from numpy import isclose
+
 # ~~~~~~~~~
 # global variables
 
@@ -58,14 +62,22 @@ def parse_args():
         help='general branch lengths probabilities',
     )
     parser.add_argument(
-        '--fels',
+        '--true-branch-lengths',
         action="store_true",
-        help='felsenstein topology probabilities',
+        help='true branch lengths probabilities',
     )
     parser.add_argument(
-        '--invfels',
-        action="store_true",
-        help='inverse felsenstein topology probabilities',
+        '--generating-topology',
+        type=str,
+        choices=('invfels', 'fels'),
+        default='invfels',
+        help='which topology generated the data',
+    )
+    parser.add_argument(
+        '--output-file',
+        type=str,
+        default='_output/out.txt',
+        help='text file to output probabilities and likelihoods',
     )
 
     args = parser.parse_args()
@@ -75,136 +87,115 @@ def parse_args():
 # ~~~~~~~~~
 # global functions for exact likelihood computation
 
-def P_FELS(theta, site_pattern):
+def get_generating_probability(theta, site_pattern, generating_topology='invfels', symbolic=False):
     """
-    Site pattern frequencies for Felsenstein tree
-    """
-    # fifth branch will always be positive
-    theta_sgn = [-1 if s=='1' else 1 for s in site_pattern+'0']
-    prob_list = [
-        [theta[0], theta[1]],
-        [theta[2], theta[3]],
-        [theta[0], theta[2], theta[4]],
-        [theta[1], theta[2], theta[4]],
-        [theta[0], theta[3], theta[4]],
-        [theta[1], theta[3], theta[4]],
-        [theta[0], theta[1], theta[2], theta[3]],
-    ]
-    sgn_list = [
-        theta_sgn[0]*theta_sgn[1],
-        theta_sgn[2]*theta_sgn[3],
-        theta_sgn[0]*theta_sgn[2]*theta_sgn[4],
-        theta_sgn[1]*theta_sgn[2]*theta_sgn[4],
[…]

### 2018-10-01 plotting/joint_inf_plot.py

@@ -22,7 +22,7 @@ from multiprocessing import Pool
 from scipy.optimize import minimize
 from itertools import product
 
-from joint_inf_helpers import ALL_PATTERNS
+from joint_inf_helpers import *
 
 sns.set_style('white')
 sns.set_style('ticks')
@@ -37,7 +37,7 @@ FONT_SIZE = 20
 # constants to slightly speed up computation
 LOG2 = np.log(2)
 LOG32 = np.log(32)
-ANC_STATE_DICT = {'0': (0,0), '1': (1,0), '2': (1,1), '3': (0,1)}
+ANC_STATE_DICT = {'0': '00', '1': '10', '2': '11', '3': '01'}
 
 
 class Legend(object):
@@ -121,41 +121,16 @@ def parse_args():
         help='file name for where to save output for time saving',
         default='output.pkl',
     )
+    parser.add_argument(
+        '--output-file',
+        type=str,
+        help='file name for where to save log file',
+        default='output.txt',
+    )
 
     args = parser.parse_args()
 
     return args
-
-# ~~~~~~~~~
-# Functions for exact likelihood bounds
-
-
-def get_generating_probability(theta0, site_pattern, generating_topology='invfels'):
-    """
-    Site pattern frequencies for inverse Felsenstein tree
-    """
-    # fifth branch will always be positive
-    theta = [-t if s=='1' else t \
-            for t,s in zip(theta0, site_pattern+'0')]
-    if generating_topology == 'invfels':
-        prob = 0.125 * (1 + \
-            theta[0]*theta[2] + \
-            theta[1]*theta[3] + \
-            theta[0]*theta[1]*theta[4] + \
-            theta[0]*theta[3]*theta[4] + \
-            theta[1]*theta[2]*theta[4] + \
-            theta[2]*theta[3]*theta[4] + \
-            theta[0]*theta[1]*theta[2]*theta[3])
-    else:
-        prob = 0.125 * (1 + \
-            theta[0]*theta[1] + \
-            theta[2]*theta[3] + \
-            theta[0]*theta[2]*theta[4] + \
-            theta[1]*theta[2]*theta[4] + \
-            theta[0]*theta[3]*theta[4] + \
-            theta[1]*theta[3]*theta[4] + \
-            theta[0]*theta[1]*theta[2]*theta[3])
-    return prob
 
 # ~~~~~~~~~
 # Functions for empirical plot
@@ -172,26 +147,12 @@ def marginal_likelihood(theta, theta_hat, generating_topology='invfels'):
[…]

### 2018-10-01 plotting/joint_inf_plot.py
keep unambiguous states in order and add file paths to doc

@@ -39,6 +39,9 @@ LOG2 = np.log(2)
 LOG32 = np.log(32)
 ANC_STATE_DICT = {'0': '00', '1': '10', '2': '11', '3': '01'}
 
+# unambiguous patterns for invfels (Table S3)
+INVFELS_UNAMB_PATTERNS = ['0000', '1000', '0100', '1100']
+
 
 class Legend(object):
     pass
@@ -87,9 +90,9 @@ def parse_args():
         type=str,
         help="""
              choices:
-             ancestral_state_conditions-- ,
-             joint_empirical-- plot empirical estimate of \hat{w} for joint inference (Fig. 3),
-             marginal_empirical-- plot empirical estimate of \hat{w} for marginal inference (Fig. S4),
+             ancestral_state_conditions-- plot analytic regions of maximal ancestral states (Fig. S2),
+             joint_empirical-- plot empirical estimate of \hat{w} for joint inference (Fig. 2),
+             marginal_empirical-- plot empirical estimate of \hat{w} for marginal inference (Fig. S3),
              """,
         choices=(
              'ancestral_state_conditions',
@@ -100,7 +103,7 @@ def parse_args():
     parser.add_argument(
         '--bias',
         action="store_true",
-        help='plot and print bias instead of parameter estimate (Fig. S3)',
+        help='plot and print bias instead of parameter estimate (Fig. 3)',
     )
     parser.add_argument(
         '--generating-topology',
@@ -166,8 +169,9 @@ def get_log_coefficients(theta, anc_states, estimating_topology='invfels', gener
 
     # unambiguous ancestral states
     if estimating_topology == 'invfels' and generating_topology == 'invfels':
-        unamb_patterns = ['0000', '1000', '0100', '1100']
-        amb_patterns = ['0010', '1110', '1010', '0110']
+        unamb_patterns = INVFELS_UNAMB_PATTERNS
+        # need to preserve order so can't use set()
+        amb_patterns = [pattern for pattern in ALL_PATTERNS if pattern not in unamb_patterns]
     else:
         unamb_patterns = ['0000']
         amb_patterns = ['1000', '0100', '0010', '1110', '1100', '0110', '1010']
@@ -406,7 +410,6 @@ def main(args=sys.argv[1:]):
         plottitle = 'Maximal ancestral state splits'
         legendtext = []
         cnt = 1
-        unamb_patterns = ['0000', '1000', '0100', '1100']
         with open(args.output_file, 'a') as f:
[…]

### 2018-10-02 plotting/joint_inf_helpers.py
preserve tip state order and add file paths

@@ -22,8 +22,8 @@ ALL_PATTERNS = [
     '0010',
     '1110',
     '1100',
+    '0110',
     '1010',
-    '0110'
 ]
 
 ALL_ANC_STATES = [
@@ -34,21 +34,32 @@ ALL_ANC_STATES = [
 ]
 
 PATT2SPLIT = {
-    '0000': '$\emptyset$',
-    '1000': '$\{1\}$    ',
-    '0100': '$\{2\}$    ',
-    '0010': '$\{3\}$    ',
-    '1110': '$\{1,2,3\}$',
-    '1100': '$\{1,2\}$  ',
-    '1010': '$\{1,3\}$  ',
-    '0110': '$\{2,3\}$  ',
+    '0000': '\emptyset',
+    '1000': '\{1\}',
+    '0100': '\{2\}',
+    '0010': '\{3\}',
+    '1110': '\{1,2,3\}',
+    '1100': '\{1,2\}',
+    '0110': '\{2,3\}',
+    '1010': '\{1,3\}',
+}
+
+PATT2SUBSCRIPT = {
+    '0000': '\emptyset',
+    '1000': '1',
+    '0100': '2',
+    '0010': '3',
+    '1110': '123',
+    '1100': '12',
+    '0110': '23',
+    '1010': '13',
 }
 
 ANC2SPLIT = {
-    '00': '$\emptyset$',
-    '10': '$\{1\}$    ',
-    '01': '$\{2\}$    ',
-    '11': '$\{1,2\}$  ',
+    '00': '\emptyset',
+    '10': '\{1\}',
+    '01': '\{2\}',
+    '11': '\{1,2\}',
 }
 
 # ~~~~~~~~~
@@ -174,7 +185,7 @@ def main(args=sys.argv[1:]):
         for site_pattern in ALL_PATTERNS:
             p = get_generating_probability(theta, site_pattern, args.generating_topology, symbolic=symbolic)
             all_ps.append(p)
-            f.write('\t'.join([PATT2SPLIT[site_pattern], str(sympify(p))]))
+            f.write('\t'.join([format('$%s$' % PATT2SPLIT[site_pattern], '<11'), '$%s$' % str(sympify(p))]))
             f.write('\n')
 
 
@@ -182,10 +193,10 @@ def main(args=sys.argv[1:]):
         f.write('Likelihoods\n')
         all_ells = []
         for site_pattern in ALL_PATTERNS:
-            f.write(PATT2SPLIT[site_pattern] + '\n')
[…]

### 2018-10-02 plotting/joint_inf_plot.py

@@ -39,6 +39,9 @@ LOG2 = np.log(2)
 LOG32 = np.log(32)
 ANC_STATE_DICT = {'0': '00', '1': '10', '2': '11', '3': '01'}
 
+# unambiguous patterns for invfels (Table S3)
+INVFELS_UNAMB_PATTERNS = ['0000', '1000', '0100', '1100']
+
 
 class Legend(object):
     pass
@@ -87,9 +90,9 @@ def parse_args():
         type=str,
         help="""
              choices:
-             ancestral_state_conditions-- ,
-             joint_empirical-- plot empirical estimate of \hat{w} for joint inference (Fig. 3),
-             marginal_empirical-- plot empirical estimate of \hat{w} for marginal inference (Fig. S4),
+             ancestral_state_conditions-- plot analytic regions of maximal ancestral states (Fig. S2),
+             joint_empirical-- plot empirical estimate of \hat{w} for joint inference (Fig. 2),
+             marginal_empirical-- plot empirical estimate of \hat{w} for marginal inference (Fig. S3),
              """,
         choices=(
              'ancestral_state_conditions',
@@ -100,7 +103,7 @@ def parse_args():
     parser.add_argument(
         '--bias',
         action="store_true",
-        help='plot and print bias instead of parameter estimate (Fig. S3)',
+        help='plot and print bias instead of parameter estimate (Fig. 3)',
     )
     parser.add_argument(
         '--generating-topology',
@@ -166,11 +169,12 @@ def get_log_coefficients(theta, anc_states, estimating_topology='invfels', gener
 
     # unambiguous ancestral states
     if estimating_topology == 'invfels' and generating_topology == 'invfels':
-        unamb_patterns = ['0000', '1000', '0100', '1100']
-        amb_patterns = ['0010', '1110', '1010', '0110']
+        unamb_patterns = INVFELS_UNAMB_PATTERNS
+        # need to preserve order so can't use set()
+        amb_patterns = [pattern for pattern in ALL_PATTERNS if pattern not in unamb_patterns]
     else:
         unamb_patterns = ['0000']
-        amb_patterns = ['1000', '0100', '0010', '1110', '1100', '0110', '1010']
+        amb_patterns = [pattern for pattern in ALL_PATTERNS if pattern not in unamb_patterns]
 
     for pattern in unamb_patterns:
         all_terms[4][0] += probs[pattern]
@@ -406,22 +410,23 @@ def main(args=sys.argv[1:]):
[…]

### 2018-10-02 appendix.tex
keep unambiguous states in order and add file paths to doc
preserve tip state order and add file paths
measure zero and clarifications

@@
 
-We show that solutions of this form never obtain $\hat{t} = t^*$ except in cases of zero or infinite branch length.
-Given Table~\ref{tab:gen-sitepatprob}, all solutions to $\hat{x}_1$ have the form
+We show that solutions of this form never obtain $\hat{t} = t^*$ except on a set of measure zero.
+Given Table~\ref{tab:gen-sitepatprob}, all solutions $\hat{x}_1$ have the form
 \[
@@
 where $a_{x_1,k}$ are constants independent of $x^*$ and $y^*$---in fact, $a_{x_1,k}$ takes values in the set $\{i/8 : i=-4,-3,\ldots,7,8\}$.
-The true branch fidelity for $x_1$ is $x^*$, and the only cases to possibly obtain $\hat{x}_1=x^*$ are when $y^*=1$ or when $(x^*)^2=x^*$, i.e., one of the generating branch parameters is zero or infinite length; the same is true for $x_2$.
-A similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent when $(y^*)^2=y^*$, i.e., $y^*=0$ or $y^*=1$.
+The true branch fidelity for $x_1$ is $x^*$.
+It is possible to obtain $\hat{x}_1=x^*$, such as when $y^*=1$ or when $(x^*)^2=x^*$, i.e., one of the generating branch parameters is zero or infinite length, but we cannot have $\hat{x}_1=x^*$ for all $x^*$.
+The same is true for $x_2$.
+A similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent on a set of measure zero.
 \end{proof}
+
+Examples of generating parameters resulting in consistency of joint inference occur in the following.
+Let $x^*$ and $y^*$ be such that the estimate for $\hat{x}_1$ is
+\[
+\hat{x}_1 = 2 \cdot \left(p_{\emptyset} + p_2 + p_{123} + p_{13}\right) - 1 = (x^*)^2.
+\]
+There may be no $x^*$ and $y^*$ that yield this maximum (Fig.~\ref{fig:max-anc-state}), though it is a feasible solution to the general joint inference problem.
+In this case, $\hat{x}_1 = x^*$ implies consistency only when $x^*=0$ or $x^*=1$.
+Another case,
+\[
+\hat{x}_1 = 2 \cdot \left(p_{\emptyset} + p_2 + p_{3} + p_{123} + p_{23} + p_{13}\right) - 1 = \frac{1}{2} + \frac{1}{2}(x^*)^2
+\]
+shows $\hat{x}_1 = x^*$ implies consistency only when $x^*=1$.
 
@@
         \centering
+        %file: plotting/_output/invfels/ancestral_state_conditions/non-bias/plot.png
         \includegraphics[width=.95\textwidth]{analytic-anc-state}
@@
         \hline
-        $\hat{\boldsymbol\xi}=\{\xi_{\emptyset}, \xi_{1}, \xi_{2}, \xi_{3}, \xi_{123}, \xi_{12}, \xi_{23}, \xi_{13}\}$\\
+        $\hat{\boldsymbol\xi} = \{\xi_{\emptyset},\xi_{1},\xi_{2},\xi_{3},\xi_{123},\xi_{12},\xi_{23},\xi_{13}\}$\\
         \hline
-        $\hat{\boldsymbol\xi}_1=\{\emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset\}$\\
-        $\hat{\boldsymbol\xi}_2=\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \emptyset, \emptyset\}$\\
-        $\hat{\boldsymbol\xi}_3=\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \{1\}, \emptyset\}$\\
-        $\hat{\boldsymbol\xi}_4=\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \{1,\}, \{1,2\}\}$\\
-        $\hat{\boldsymbol\xi}_5=\{\emptyset, \emptyset, \emptyset, \{1\}, \{1\}, \emptyset, \{1\}, \{1\}\}$\\
+        $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$\\
+        $\hat{\boldsymbol\xi}_2 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\emptyset,\emptyset\}$\\
+        $\hat{\boldsymbol\xi}_3 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\emptyset,\{1\}\}$\\
+        $\hat{\boldsymbol\xi}_4 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\{1,2\},\{1\}\}$\\
+        $\hat{\boldsymbol\xi}_5 = \{\emptyset,\emptyset,\emptyset,\{1\},\{1\},\emptyset,\{1\},\{1\}\}$\\
         \hline
@@
 
+The boundaries determining maximal ancestral state splits (Fig.~\ref{fig:max-anc-state}) are obtained through maximizing over 81 separate functional values (Table~\ref{tab:likelihoods}).
+Using the notation in the proof of Theorem~\ref{thm:bl_theorem}, we see that, in general, our likelihood given in~\eqref{eq:site_pattern_profile_likelihood_mean} has an exact, closed form maximum of
+\begin{align}
+\ell_{\tau^*,t^*}(\tau_1, \hat{t}; \boldsymbol\xi) &= \frac{1+\hat{x}_1}{2}\log(1+\hat{x}_1) + \frac{1-\hat{x}_1}{2}\log(1-\hat{x}_1)\nonumber\\
+&\qquad + \frac{1+\hat{y}_1}{2}\log(1+\hat{y}_1) + \frac{1-\hat{y}_1}{2}\log(1-\hat{y}_1)\nonumber\\
+&\qquad + \frac{1+\hat{x}_2}{2}\log(1+\hat{x}_2) + \frac{1-\hat{x}_2}{2}\log(1-\hat{x}_2)\nonumber\\
+&\qquad + \frac{1+\hat{y}_2}{2}\log(1+\hat{y}_2) + \frac{1-\hat{y}_2}{2}\log(1-\hat{y}_2)\nonumber\\
+&\qquad + \frac{1+\hat{w}}{2}\log(1+\hat{w}) + \frac{1-\hat{w}}{2}\log(1-\hat{w}).
+\end{align}
+For example, in the lower right region of Fig.~\ref{fig:max-anc-state}, the maximal ancestral state split is $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$ so that
+\[
+\{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\} = \{x^*(y^*)^2,(y^*)^2,x^*(y^*)^2,1,1\}
+\]
+and the exact, closed form value of the likelihood is
+\begin{align}
+\ell_{U,1}(x^*,y^*) &= (1+x^*(y^*)^2)\log(1+x^*(y^*)^2) + (1-x^*(y^*)^2)\log(1-x^*(y^*)^2)\nonumber\\
+&\qquad + \frac{1+(y^*)^2}{2}\log(1+(y^*)^2) + \frac{1-(y^*)^2}{2}\log(1-(y^*)^2)\nonumber\\
+&\qquad + 2\log(2).
+\end{align}
+We obtain similar maximal values as functions of $x^*$ and $y^*$ for each ancestral state split.
+The curves delineating maximal ancestral state splits (Fig.~\ref{fig:max-anc-state}) are determined by these 81 functions.
+
 %~~~
@@
 \centering
+%file: plotting/_output/invfels/marginal_empirical/non-bias/plot.svg
 \includegraphics[width=\textwidth]{empirical-marginal}
@@
 \end{figure}
-%
-%\begin{figure}
-%\centering
-%\includegraphics[width=.95\textwidth]{empirical-incons-intuition}
-%\caption{
-%Inconsistency in joint inference with intuitive curves.
-%The white curve is our condition on the gradients.
-%The blue curve is a condition on $x^*$ and $y^*$ such that $\emptyset$ is the most likely ancestral state.
-%}
-%\label{fig:empirical-incons-intuition}
-%\end{figure}

### 2018-10-02 joint_inf.tex
preserve tip state order and add file paths
measure zero and clarifications

@@
 We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
-For branch length estimation on the correct topology, we show that this joint method cannot estimate consistent branch lengths except in degenerate cases, and we provide extensive empirical results for outlining the consistent bias in this setting.
+For branch length estimation on the correct topology, we show that this joint method cannot estimate consistent branch lengths except on a set of measure zero, and we provide extensive empirical results for outlining the consistent bias in this setting.
 \end{abstract}
@@
 To do so, we use a binary symmetric model with data generated on a four-taxon tree: we compute closed form solutions to the joint objective function and demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
-We show that, when the topology is known and fixed, joint inference cannot be consistent except in cases of zero or infinite branch length, and we find similar areas through empirical means where joint inference consistently underestimates interior branch lengths.
+We show that, when the topology is known and fixed, joint inference cannot be consistent except on a set of measure zero, such as in cases of zero or infinite branch length, and we find similar areas through empirical means where joint inference consistently underestimates interior branch lengths.
 
@@
 
-When the topology is known and fixed and we estimate only branch lengths, we show the following, i.e., that for all $x^*$ and $y^*$ in $(0,1)$ any branch length estimate is consistently biased.
+When the topology is known and fixed and we estimate only branch lengths, we show the following, i.e., that for almost all $x^*$ and $y^*$ in $(0,1)$ any branch length estimate is consistently biased.
 
@@
 Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
-For all $0 < x^*, y^* < 1$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
+For $0 < x^*, y^* < 1$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
 \[
@@
 \]
-has the property $\hat{t} \neq t^*$.
+has the property $\hat{t} \neq t^*$ everywhere except a set of measure zero.
 \end{restatable}
 
-In words, the joint estimation procedure never recovers the true generating $t^*$ except in cases of zero or infinite branch length.
+In words, the joint estimation procedure never recovers the true generating $t^*$ almost everywhere in the space of generating parameters.
 This is apparent given Table~\ref{tab:gen-sitepatprob}, as the solution $\hat{t}$ is a linear combination of $p_{\siteSplit_j}$ values, and no generating probability contains an $x^*$ or $y^*$ term.
+We may have special cases where certain $x^*$ and $y^*$ values yield consistent estimates for $\hat{t}$, but consistency does not hold in general.
+We give an example in the Appendix following the proof of Theorem~\ref{thm:bl_theorem}.
+Thus, inconsistency is guaranteed for almost all $x^*$ and $y^*$, though not all $x^*$ and $y^*$.
 
@@
 \centering
+%file: plotting/_output/invfels/joint_empirical/non-bias/plot.svg
 \includegraphics[width=\textwidth]{empirical-estimate}
@@
 \centering
+%file: plotting/_output/invfels/joint_empirical/bias/plot.svg
 \includegraphics[width=.95\textwidth]{empirical-bias}
@@
 We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
-Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of topological inconsistency: the joint inference procedure estimates zero-length branches, which can be considered as a multifurcating topology.
+Specifically, we have shown there are no sets of generating parameters other than ones of measure zero that yield consistent branch length estimates given the fixed, correct topology.
+In addition, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of topological inconsistency: the joint inference procedure estimates zero-length branches, which can be considered as a multifurcating topology.
 Also, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
@@
 However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).
-In addition, we have shown there are no nontrivial generating parameters that yield consistent branch length estimates.
 

### 2018-10-03 joint_inf.tex
pass thru

@@
 We may have special cases where certain $x^*$ and $y^*$ values yield consistent estimates for $\hat{t}$, but consistency does not hold in general.
-We give an example in the Appendix following the proof of Theorem~\ref{thm:bl_theorem}.
-Thus, inconsistency is guaranteed for almost all $x^*$ and $y^*$, though not all $x^*$ and $y^*$.
+We give an example of consistency in a degenerate case in the Appendix following the proof of Theorem~\ref{thm:bl_theorem}.
 

### 2018-10-04 joint_inf.tex
Quick look

@@
 We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
-For branch length estimation on the correct topology, we show that this joint method cannot estimate consistent branch lengths except on a set of measure zero, and we provide extensive empirical results for outlining the consistent bias in this setting.
+For branch length estimation on the correct topology, we show that this joint method cannot correctly estimate branch lengths except on a set of measure zero, and we provide extensive empirical results for branch length bias in this setting.
 \end{abstract}
@@
 To do so, we use a binary symmetric model with data generated on a four-taxon tree: we compute closed form solutions to the joint objective function and demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
-We show that, when the topology is known and fixed, joint inference cannot be consistent except on a set of measure zero, such as in cases of zero or infinite branch length, and we find similar areas through empirical means where joint inference consistently underestimates interior branch lengths.
+We show that, when the topology is known and fixed, joint inference for branch length estimation cannot be consistent except on a set of measure zero, such as in cases of zero or infinite branch length, and we find similar areas through empirical means where joint inference consistently underestimates interior branch lengths.
 
@@
 
+%EM Starting here, we only talk about branch lengths and don't mention fidelities, whereas just above we say "estimate a topology and branch fidelities using this"...
+%EM At least remind the reader that we're actually talking about fidelities as the variables.
 \subsection*{Inconsistent branch length estimation}
-
 When the topology is known and fixed and we estimate only branch lengths, we show the following, i.e., that for almost all $x^*$ and $y^*$ in $(0,1)$ any branch length estimate is consistently biased.
@@
 
-In words, the joint estimation procedure never recovers the true generating $t^*$ almost everywhere in the space of generating parameters.
+In words, the joint estimation procedure does not recover the true generating $t^*$ almost everywhere in the space of generating parameters.
+%EM Totally confused here!
+%EM 1. The generating probabilities do contain x* and y*.
+%EM 2. "This is apparent" makes it sound like it should follow directly and that it doesn't require a proof.
 This is apparent given Table~\ref{tab:gen-sitepatprob}, as the solution $\hat{t}$ is a linear combination of $p_{\siteSplit_j}$ values, and no generating probability contains an $x^*$ or $y^*$ term.
@@
 
-\subsection*{Convergence to degenerate topology}
-
+\subsection*{Convergence to the degenerate topology}
 Given data generated on $\tau_1$ there exist true nonzero branch lengths such that the estimator $\hat{t}$ maximizing the right-hand side of \eqref{eq:exp-log-lik} has an internal branch of length zero.
-%EM A funny thing: this theorem as stated is really under the hypothesis that the ancestral state is emptyset.
-We can combine this with the lemmas to remove this assumption, but it seems like we should do this for the reader.
-Time for a corollary?
-%das: ok, I reworded it a bit; let me know how you think it sounds.
-%EM This is better, but the notation still seems a little funny.
-Namely, xi is really something that's determined by Y, but the notation seems to hint that it can be specified a priori.
-The point is that the ML xi has problems, not that there exists a xi that has problems.
-%das: ok, edited
 \begin{restatable}{theorem}{topoInconsist}
@@
 
+%EM Is this the right figure reference? It seems strange-- it's about the max ancestral states.
 This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:max-anc-state}).
@@
 The proof is through analytically reducing the general case to 81 separate cases (Table~\ref{tab:likelihoods}) to obtain a closed form maximal value for each.
-
-%EM I'm going to suggest that we move this to the appendix. The theorem says that there's a set, and we want people to nod then look at the current Figure 3.
-%das: good idea; I think this will also appease reviewer 1.
-%EM On the other hand we should still refer to it in the text around here.
-%das: it's referred to in the first sentence of above paragraph---good?
 
@@
 Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
-%EM I don't follow here. This doesn't seem like the long branch attraction region of parameter space-- isn't that on the Felsenstein tree, not the InvFels tree?
-%das: yeah, it was kind of hand-wavy. edited
 This is not the case here; while analytically the inconsistency occurs when the top two branches are long and the bottom three are short, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.

### 2018-10-04 versions/response.v1.txt
update response

@@
 
-As described above, our intent is to mathematically establish inconsistency and poor estimation.
-This must be done with analytic arguments rather than optimization.
-We could have only shown this bad behavior in a very small area to establish that it happens.
-In contrast to this reviewer, we believe that Fig 5 is optional to this line of argument, and we provide it just for interest.
-A true investigation into the practical implications of our work will require another paper.
-We have clarified this in the Discussion section with:
-"Our aim is to analytically demonstrate joint inference is inconsistent, and, while this indicates an issue with this inference procedure, there is substantial scope for future work to make these results more precise and more general.
-The techniques used to obtain upper and lower bounds for the likelihoods in Lemma 3 provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation."
+Thank you for the opportunity to improve on our analytically-derived regions of inconsistency.
+Indeed, obtaining these regions in closed form is feasible, though it involves comparing a large number of two-variables functions to one another.
+Fig. S2 shows the regions where a particular likelihood equation will be maximal, and allows us to compute---in closed form---the value of the branch parameter in Fig. 2.
 
@@
 
-Please see above for our response.
-%das: I added some stuff below to maybe alleviate some confusion and to show why what this reviewer thinks we should do won't work, but not sure how much is necessary.
-It is true that the form given above for l(x,y,w) has an explicit and simple maximizer.
-However, the long-run log-likelihood is more complicated than this function, as seen in equation (9).
-For both topologies, we have a number of equations of this form that we would need to maximize in a grid search.
-The function of the form in l(x,y,w) above was our approximation to the upper and lower bounds of the log-likelihood, and was our choice for these bounds precisely because it has an explicit and simple maximizer.
-Doing a grid search over all ancestral states would not be trivial, and we would need to perform numerical optimization for all possible ancestral states to compute the maximum.
-Moreover, not assuming x_1=x_2 and y_1=y_2 (see below) results in an explosion of functions to perform a grid search over, with none of these functions being guaranteed concave.
-The analytic bounds we have provided were devised specifically to sidestep this nonconcavity.
+Another thanks for this suggestion.
+As noted, we can in fact get exact regions, and have done so.
+Our line of argument starting on page 25, line 541 in the section "Properties of the joint objective function" allows us to pare down the 4^8 possible ancestral states to just 81, and we obtain exact, analytic results in a reasonable amount of time.
 
@@
 
-TODO: make the argument more clear in the main text.
 %das: BIIIIG update
@@
 We have added the definition of a "generating probability" to ease confusion in this sentence:
-"As data becomes plentiful, we use the Hadamard transform (see (9) in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form."
+"As data becomes plentiful, we use the Hadamard transform (see (8) in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form."
 
@@
 %das: I believe this is clear now; lmk
+%das: @EM bump
 We have clarified the intuition in this sentence:
@@
 
-Fixed via simpler language.
 We removed the sentence following the semicolon entirely, as we feel it did not add anything:
@@
 We have updated the text to explicitly define "generating probability."
-"As data becomes plentiful, we use the Hadamard transform (see (9) in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form."
+"As data becomes plentiful, we use the Hadamard transform (see (8) in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form."
 
@@
 We clarify our aim and warnings from Sagulenko et al. (2017) in the Introduction:
-"Although the software described in \citet{Sagulenko2018-xl} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
+"Although \citet{Sagulenko2018-xl} explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
 In particular, one would like to know when this approximate technique breaks down for both topology and branch length inference, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference."
@@
 This is an interesting suggestion but is different than our goal, which was to establish inconsistency.
-Moreover, in the revised manuscript we see cases where assigning ML ancestral states results in longer branches, not shorter ones (Lemma 3).
+Moreover, in the revised manuscript we see cases where assigning ML ancestral states results in longer branches, not shorter ones (see the discussion after the proof of Theorem 1, page 31, line 627).
 We believe the inaccuracies have more to do with the specific choice of which ancestral state is maximal than asymmetries in the tree:
@@
 
-Asymmetries can drive these particular choices, but Fig. 3 in the revised manuscript shows that almost half of the parameter space results in an interior branch length being estimated as zero.
+Asymmetries can drive these particular choices, but Fig. 2 in the revised manuscript shows that almost half of the parameter space results in an interior branch length being estimated as zero.
 We do not think providing an argument about accumulating inaccuracies would elucidate why such starkly-defined regions of inconsistency exist.
@@
 
-The analytical bound is a coarse estimate of where inconsistency arises.
-In the revised manuscript, we fix the maximal ancestral state to simplify the argument, and comparing the analytical with the empirical would not provide much added benefit to the analysis.
+In the revised manuscript, all region bounds are analytical as we have determined a closed form solution for the joint inference problem in this case.
 
@@
 We have added the line:
-"Although the software described in \citet{Sagulenko2018-xl} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference."
+"Although \citet{Sagulenko2018-xl} explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference."
 
@@
 
+%das: @EM let me know if, given these notes, you think the discussion in the text is reasonable
+
 ---

### 2018-10-04 appendix.tex
pass thru
Quick look
pass through comments

@@
 %We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
+%EM @das I think that ^ is all resolved, right?
+%das: I think to be fair I should add that we don't actually show a real inconsistency in topology because we haven't shown this functional convergence for it, only for branch lengths with a fixed topology.
+I added a paragraph hedging a bit.
+Let me know what you think.
 For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
@@
 This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.
+
+This Wald-type consistency does not extend readily to convergence of topology, i.e., we cannot use the above arguments to say that, for
+\[
+(\hat{\tau}_\nCols, \hat{t}_\nCols) = \arg\max_{\tau, t} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
+\]
+and
+\[
+(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)
+\]
+then
+\[
+(\hat{\tau}_\nCols, \hat{t}_\nCols) \rightarrow (\hat{\tau}, \hat{t})
+\]
+as we did for $\hat{t}_\nCols$ with $\tau$ fixed due to the properties of tree space \citep{Yang1994}.
+For this reason we do not claim joint inference is inconsistent in estimating topology, only that there exist parameters that yield degenerate maxima when performing joint optimization.
 
@@
 There may be no $x^*$ and $y^*$ that yield this maximum (Fig.~\ref{fig:max-anc-state}), though it is a feasible solution to the general joint inference problem.
-In this case, $\hat{x}_1 = x^*$ implies consistency only when $x^*=0$ or $x^*=1$.
+In this case, $\hat{x}_1 = x^*$ implies consistency only occurs when $x^*=0$ or $x^*=1$.
 Another case,
@@
 Regions of maximal ancestral state splits on the InvFels tree $\tau_1$.
+Data generated as in Fig.~\ref{fig:bl-general-inconsistency}.
 }
@@
 As the region of degeneracy in Fig.~\ref{fig:max-anc-state} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
-%EM "letting $\emptyset$ be the maximal ancestral state split" sounds a little funny-- either it is or it isn't maximal.
 It is easy to see that when $\emptyset$ is the maximal ancestral state split, we have the same log-likelihood for $\tau_1$ and $\tau_2$.

### 2018-10-04 joint_inf.tex
pass through comments

@@
 %EM At least remind the reader that we're actually talking about fidelities as the variables.
-\subsection*{Inconsistent branch length estimation}
-When the topology is known and fixed and we estimate only branch lengths, we show the following, i.e., that for almost all $x^*$ and $y^*$ in $(0,1)$ any branch length estimate is consistently biased.
+%das: clarification
+\subsection*{Inconsistent branch parameter estimation}
+When the topology is known and fixed and we estimate only the branch parameters, we show that for almost all generating parameter values, any branch parameter estimate is consistently biased.
+For the branch parameter, we use branch fidelities (discussed earlier) in all statements and proofs, though these trivially extend to branch lengths via transformation.
 
@@
 %EM 2. "This is apparent" makes it sound like it should follow directly and that it doesn't require a proof.
-This is apparent given Table~\ref{tab:gen-sitepatprob}, as the solution $\hat{t}$ is a linear combination of $p_{\siteSplit_j}$ values, and no generating probability contains an $x^*$ or $y^*$ term.
+%das: ok, took a stab at clarifying
+For example, in estimating $x_1$, $\hat{x}_1$ is a linear combination of $p_{\siteSplit_j}$ values.
+For the generating probabilities (top panel of Table~\ref{tab:gen-sitepatprob}), there is no linear combination that results in an isolated $x^*$ term, as all terms are either quadratic---$(x^*)^2$ and $(x^*)^2(y^*)^2$---or have both $x^*$ and $y^*$---$x^*(y^*)^2$.
+Thus, we cannot obtain a linear combination that in the general case yields $x^*$.
 We may have special cases where certain $x^*$ and $y^*$ values yield consistent estimates for $\hat{t}$, but consistency does not hold in general.
@@
 %EM Is this the right figure reference? It seems strange-- it's about the max ancestral states.
-This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:max-anc-state}).
+%das: ha oops! you're right
+This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:bl-general-inconsistency}).
 As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
@@
     Estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
-    Data generated as in Fig.~\ref{fig:max-anc-state}.
+    Regions derived in terms of probabilities of a character change along a branch for ``perfect''data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}a).
     The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero, resulting in an inconsistency.

### 2018-10-04 joint_inf.tex
em-dash looked too much like a minus

@@
 For example, in estimating $x_1$, $\hat{x}_1$ is a linear combination of $p_{\siteSplit_j}$ values.
-For the generating probabilities (top panel of Table~\ref{tab:gen-sitepatprob}), there is no linear combination that results in an isolated $x^*$ term, as all terms are either quadratic---$(x^*)^2$ and $(x^*)^2(y^*)^2$---or have both $x^*$ and $y^*$---$x^*(y^*)^2$.
+For the generating probabilities (top panel of Table~\ref{tab:gen-sitepatprob}), there is no linear combination that results in an isolated $x^*$ term, as all terms are either quadratic, i.e., $(x^*)^2$ or $(x^*)^2(y^*)^2$, or have both $x^*$ and $y^*$, i.e., $x^*(y^*)^2$.
 Thus, we cannot obtain a linear combination that in the general case yields $x^*$.

### 2018-10-04 versions/response.v1.txt
A bit of cleaning up of response.

@@
 
-Outline of revisions:
-1. generalized the results to arbitrary branch lengths and not restricting the top and bottom two branches to be equal
-2. expanded and clarified the exposition of the mathematics and the motivation for the Hadamard representation
-3. changed discussion and plots to involve branch lengths and not fidelities
+Thank you for your helpful comments. We think that this version of the paper is significantly improved, and we hope you agree.
+
+Outline of major revisions:
+1.
+We have generalized the results to arbitrary branch lengths and not restricting the top and bottom two branches to be equal.
+The previous version presented an argument that the ML branch lengths would have this property, but this argument did not hold up to further scrutiny.
+2. We have expanded and clarified the exposition of the mathematics and the motivation for the Hadamard representation.
+3. We have changed discussion and plots to involve branch lengths and not fidelities.
 
@@
 
-%das: added
 This work is not the final word on the joint inference question from an applied perspective, as we now emphasize with an expanded Discussion section.
@@
 Thank you for the opportunity to improve on our analytically-derived regions of inconsistency.
-Indeed, obtaining these regions in closed form is feasible, though it involves comparing a large number of two-variables functions to one another.
+Indeed, obtaining these regions in closed form is feasible, though it involves comparing a large number of two-variable functions to one another.
 Fig. S2 shows the regions where a particular likelihood equation will be maximal, and allows us to compute---in closed form---the value of the branch parameter in Fig. 2.
@@
 
-%das: BIIIIG update
 We thank this reviewer for highlighting an omission in our argument.
@@
 
-%EM I think we could move this to the appendix to throw this reviewer a bone.
-%das: another big update
 We have removed the restricted case as it does not reflect how inference is performed and was only included as an expository example.
@@
 However, to do such a project justice would require a separate paper.
-We are sure that once our paper demonstrates this inconsistency, others will quickly follow to write such a paper.
+We suspect that once our paper demonstrates this inconsistency, others will quickly follow to write such a paper.
 
@@
 
-%EM David-- does it seem like we could extend these results to bigger trees containing a 4 taxon tree?
-%das: I wasn't able to come up with anything interesting for this.
+We have considered how to extend these results to larger trees but this does not appear to be straightforward.
 
@@
 
-TODO: this is changed, but I don't see this being a lot clearer.
-%das: I believe this is clear now; lmk
-%das: @EM bump
-We have clarified the intuition in this sentence:
+We have clarified the intuition in these new sentences:
 "For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest conditional log-likelihood.
@@
 
-We do not believe an exhaustive explanation would provide any benefit to the reader that they could not get from referring to the text.
-
 > Reviewer: 3
@@
 However, in our assessment the word "approximate" doesn't completely describe a situation in which there is significant bias.
-The TreeTime paper says that its use is appropriate when "ancestral sequence inference has little ambiguity" but even in these cases the bias in branch lengths persists.
+The TreeTime paper says that its use is appropriate when "ancestral sequence inference has little ambiguity" but as we show the bias in branch lengths persists even in these cases.
 
-TODO we should make some changes to at least say that we've further clarified the implications of our results.
-%das: I believe this is done
 We clarify our aim and warnings from Sagulenko et al. (2017) in the Introduction:
@@
 
-TODO describe why Taylor isn't a straightforward thing.
-TODO refer to idea of applied paper again?
-%das: I believe these are done as well
-
 > Specific comments:
@@
 
-TODO we should do some change, but make it clear that it's consistently biased even for short branches.
-%das: I believe this is done
 We have added the line:
@@
 Thank you for your assessment.
-We have worked hard in our initial submission and in our revision to write out the argument in great (indeed, excruciating) detail, and we would appreciate it if you could take a look at the revision.
+We have worked hard in our initial submission and in our revision to write out the argument in great (indeed, excruciating) detail, and hope that you will be able to read the revision more fully.
 
@@
 "Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood."
-
-%EM David, heh, can you point me to the place where the equal branch length result was done?
-%EM Also, I think you said that he says that the varying branch length thing is an open question-- it would be great to quote that.
-%das: update I added some of the below.
-Let me know what you think.
-And re: Allman/Chai, I just added them in with RoyChoudhury; is that ok?
-They are all just examples of conditions where phylogenetics is consistent it seems.
-
-%das: below are some notes and quotes. I will work on paring these down and adding some to the text, maybe with a little more diplomacy.
-%The equal branch length example occurs in Example 2, starting halfway down the second column of page 352 in Goldman (1990) and continuing until the top of the second column on page 354.
-%Here he only talks about how parsimony is inconsistent because it may result in equally parsimonious trees (this is the last paragraph of the example section).
-%It's not until the bottom of page 356 that he punts the inconsistency to Felsenstein and Hendy/Penny, and the bottom of page 357 where he wonders how small branch lengths have to be for parsimony to be consistent in general.
-
-%das: @EM let me know if, given these notes, you think the discussion in the text is reasonable
 
@@
 
-As described at the beginning, we have made this change, thanks.
+As described at the beginning, we have made this change.

### 2018-10-05 appendix.tex
Pass through.

@@
 
+%EM This is very minor compared to the below comments, so address it second in a separate commit, please, if you wish to address it.
+Curly braces are really for unordered sets, right?
+I suggest round parensI suggest round parens.
 For a fixed topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
@@
 \end{align*}
-In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subseteq\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, i.e., $\ancestralSplitPartition_j(\tau, t)$ is a set of sets of most likely internal node labels.
+In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subseteq\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, i.e., $\ancestralSplitPartition_j(\tau, t)$ is a set of sets of most likely internal node states.
 Here, $\xi_{\siteSplit_j}$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
@@
 For fixed $\boldsymbol\xi$, the log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
+%EM These c_j's and their properties need to be explained a lot more clearly. What are they, how can we know that the likelihood has this form, and why they have the properties below.
 \begin{align*}
@@
 \begin{align}
+%EM Here's this propto thing again. It's up to a constant, not propto, right?
 \ell(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
@@
 As $x_1\in(0,1]$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
+%EM We need to hear why both of these properties hold.
 Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\ell''(x_1) \le 0$.
@@
 
+%EM Isn't this table redundant given Table S3?
 \begin{table}
@@
 The true branch fidelity for $x_1$ is $x^*$.
-It is possible to obtain $\hat{x}_1=x^*$, such as when $y^*=1$ or when $(x^*)^2=x^*$, i.e., one of the generating branch parameters is zero or infinite length, but we cannot have $\hat{x}_1=x^*$ for all $x^*$.
+%EM The following sentence falls short of getting us rigorously to where we need to go, which is the measure zero statement.
+Doesn't that just follow from the number of zeros of a polynomial being finite?
+It is possible to obtain $\hat{x}_1=x^*$, such as when $y^*=1$ or when $(x^*)^2=x^*$, i.e., one of the generating branch lengths is zero or infinite, but we cannot have $\hat{x}_1=x^*$ for all $x^*$.
 The same is true for $x_2$.
@@
 
+%EM I don't see this paragraph being helpful.
+%EM It's a confusing read. First you say that we're going to look at examples, but then perhaps you cast doubt on if the examples exist "There may be no..."
 Examples of generating parameters resulting in consistency of joint inference occur in the following.
@@
 Moreover, if $w=1$, the internal branch becomes zero-length and the two topologies are indistinguishable.
+%EM The rest of this paragraph is obvious, no?
 Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the region in Fig.~\ref{fig:max-anc-state} where $\hat{\boldsymbol\xi}=\hat{\boldsymbol\xi}_1$.
@@
 The boundaries determining maximal ancestral state splits (Fig.~\ref{fig:max-anc-state}) are obtained through maximizing over 81 separate functional values (Table~\ref{tab:likelihoods}).
+%EM Are you saying that we should be able to obtain this next display equation by maximizing over the entries in Table S3? I think this needs more explanation.
 Using the notation in the proof of Theorem~\ref{thm:bl_theorem}, we see that, in general, our likelihood given in~\eqref{eq:site_pattern_profile_likelihood_mean} has an exact, closed form maximum of
@@
 \begin{align}
+%EM I wouldn't introduce the \ell_{U,1} notation unless it's really needed for some reason I don't see yet.
 \ell_{U,1}(x^*,y^*) &= (1+x^*(y^*)^2)\log(1+x^*(y^*)^2) + (1-x^*(y^*)^2)\log(1-x^*(y^*)^2)\nonumber\\

### 2018-10-05 joint_inf.tex
Is this a correct summary?
i had a bit of a hard time following, but maybe this is still a little unclear? :tennis:
A bit of cleaning up of response.
Pass through.

@@
 We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
-For branch length estimation on the correct topology, we show that this joint method cannot correctly estimate branch lengths except on a set of measure zero, and we provide extensive empirical results for branch length bias in this setting.
+For branch length estimation on the correct topology, we show that this joint method cannot correctly estimate branch lengths except on a set of measure zero, and we show empirically that branch lengths are biased in this setting even for short branches, with bias of the same order as the branch lengths themselves.
 \end{abstract}
@@
 In this paper, we show that jointly inferring trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model with data generated on a four-taxon tree: we compute closed form solutions to the joint objective function and demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
-We show that, when the topology is known and fixed, joint inference for branch length estimation cannot be consistent except on a set of measure zero, such as in cases of zero or infinite branch length, and we find similar areas through empirical means where joint inference consistently underestimates interior branch lengths.
+To do so, we use a binary symmetric model with data generated on a four-taxon tree: we compute closed form solutions to the joint objective function and demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be exactly zero.
+We show that, when the topology is known and fixed, joint inference for branch length estimation cannot be consistent except on a set of measure zero (i.e.\ a set that occupies zero volume in parameter space).
+Empirically, we find areas where joint inference consistently underestimates interior branch lengths, including regions of short branch length where bias is on the same order as the branch length.
 
@@
 
-%EM Starting here, we only talk about branch lengths and don't mention fidelities, whereas just above we say "estimate a topology and branch fidelities using this"...
-%EM At least remind the reader that we're actually talking about fidelities as the variables.
-%das: clarification
 \subsection*{Inconsistent branch parameter estimation}
@@
 \]
-has the property $\hat{t} \neq t^*$ everywhere except a set of measure zero.
+has the property that $\hat{t} \neq t^*$ everywhere except a set of measure zero.
 \end{restatable}
@@
 In words, the joint estimation procedure does not recover the true generating $t^*$ almost everywhere in the space of generating parameters.
-%EM Totally confused here!
-%EM 1. The generating probabilities do contain x* and y*.
-%EM 2. "This is apparent" makes it sound like it should follow directly and that it doesn't require a proof.
-%das: ok, took a stab at clarifying
+The proof can be seen intuitively through the fact that the estimator for a given branch fidelity cannot isolate the individual generating parameter as a linear term since the estimator itself is a combination of nonlinear functions (in $(x^*,y^*)$) of generating fidelities.
 For example, in estimating $x_1$, $\hat{x}_1$ is a linear combination of $p_{\siteSplit_j}$ values.
@@
 
-%EM Is this the right figure reference? It seems strange-- it's about the max ancestral states.
-%das: ha oops! you're right
 This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:bl-general-inconsistency}).
@@
 \caption{
-    Estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
+    Estimates for $\hat{p}_w=(1-\hat{w})/2$, the optimal probability of substitution along the inner edge for the profile likelihood \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
     Regions derived in terms of probabilities of a character change along a branch for ``perfect''data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}a).
@@
 Although the estimates for $\hat{p}_w$ are better than the estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), joint inference still predictably underestimates the interior branch length.
+%EM I don't feel like this "range" is a very strong statement. What if we just repeat that the bias is of the same order as the branch lengths and give an example?
 Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ range from $[-4\times 10^{-2}, 3\times 10^{-3}]$.
+showing that even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
 
-Inference on the integrated likelihood performs as expected where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}).
-We use L-BFGS-B when optimizing \eqref{eq:marginal_likelihood}.
-The errors in this case are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
+In contrast, inference on the integrated likelihood performs as expected, such that $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}).
+The errors in this case \eqref{eq:marginal_likelihood} via optimization with L-BFGS-B are lower than machine tolerance.
 
@@
 We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
-Specifically, we have shown there are no sets of generating parameters other than ones of measure zero that yield consistent branch length estimates given the fixed, correct topology.
-In addition, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of topological inconsistency: the joint inference procedure estimates zero-length branches, which can be considered as a multifurcating topology.
+Specifically, we have shown that the only generating parameters that yield consistent branch length estimates given the correct topology lie in a set of measure zero.
+In addition, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in topological ambiguity: the joint inference procedure estimates zero-length branches, which can be considered as a multifurcating topology.
 Also, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
 Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length is more likely to be observed, though we have not found regions where this is the case.
-The regions of inconsistency we found arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
+The regions of inconsistency we found arise when one set of sister branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
 We see that this inconsistency occurs even if some branches are short.
@@
 Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
-This is not the case here; while analytically the inconsistency occurs when the top two branches are long and the bottom three are short, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
+This is not the case here; while analytically the inconsistency occurs on a four-taxon tree when one pair of sister branches are long and the other three are short, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
 Additionally, we generate data on the InvFels tree $\tau_1$ while \citet{Felsenstein1978-rr} generates data on the Felsenstein tree $\tau_2$.
-Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.
+Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that the difficulties come from insufficient sequence length, which is not the case here.
 
@@
 Finally, just prior to his conclusion, he discusses when parsimony gives the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.
-We render the question well-posed: the joint inference procedure outlined here estimates the same parameters as classical maximum likelihood---topology and branch lengths---albeit implicitly estimating ancestral states as well.
+Our question, in contrast, is well-posed: the joint inference procedure outlined here estimates the same parameters as classical maximum likelihood---topology and branch lengths---albeit implicitly estimating ancestral states as well.
 We are able to provide much more detail on how large branch lengths must be for general joint inference to fail to be consistent.
+
+%The equal branch length example occurs in Example 2, starting halfway down the second column of page 352 in Goldman (1990) and continuing until the top of the second column on page 354.
+%Here he only talks about how parsimony is inconsistent because it may result in equally parsimonious trees (this is the last paragraph of the example section).
+%It's not until the bottom of page 356 that he punts the inconsistency to Felsenstein and Hendy/Penny, and the bottom of page 357 where he wonders how small branch lengths have to be for parsimony to be consistent in general.
 

### 2018-10-05 appendix.tex
getting there

@@
 I suggest round parensI suggest round parens.
+%das: TODO
 For a fixed topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
@@
 %EM These c_j's and their properties need to be explained a lot more clearly. What are they, how can we know that the likelihood has this form, and why they have the properties below.
+%das: heh they must have been carried over---they're actually p_yj's
 \begin{align*}
-\ell_{\tau^*,t^*}(\tau^*, t, \boldsymbol\xi) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_2}(x_2)\\
-&\qquad + \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,w}(w).
+\ell_{\tau^*,t^*}(\tau^*, t, \boldsymbol\xi) &= \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_2}(x_2)\\
+&\qquad + \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,w}(w).
 \end{align*}
 Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
-Without loss of generality we focus on the variable $x_1$ and the log-likelihood proportional to
+Without loss of generality we focus on the variable $x_1$ and the function
 \[
-\ell(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1).
+\ell(x_1) \sim \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_1}(x_1).
 \]
+where $\sim$ denotes equality up to an additive constant.
 Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, is
 \[
-h_{j,x_1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
+h_{j,x_1}(x_1) = (1+x_1)^{e_j}(1-x_1)^{1-e_j}
 \]
@@
 %EM Here's this propto thing again. It's up to a constant, not propto, right?
-\ell(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
-                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1), \label{eq:closed-form-x}
+%das: right---reverted to \sim
+\ell(x_1) &\sim \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}(1-e_j)\right) \log (1-x_1)\\
+                  &= \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log (1-x_1), \label{eq:closed-form-x}
 \end{align}
@@
 \[
-\ell''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
+\ell''(x_1) = -\left(\frac{\sum_j p_{\siteSplit_j}e_j}{(1+x_1)^2}+\frac{1-\sum_j p_{\siteSplit_j}e_j}{(1-x_1)^2}\right).
 \]
-As $x_1\in(0,1]$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
+As $x_1\in(0,1]$, we need only $0 \le \sum_j p_{\siteSplit_j}e_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
 %EM We need to hear why both of these properties hold.
-Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\ell''(x_1) \le 0$.
+%das: now that c_j is changed to the probability maybe it's more clear? e_j in {0,1} is by definition above
+Since $p_{\siteSplit_j}$ are probabilities, then $\sum_j p_{\siteSplit_j} = 1$.
+As $e_j\in\{0,1\}$ by definition, this implies $0 \le \sum_j p_{\siteSplit_j}e_j \le 1$ and $\ell''(x_1) \le 0$.
 Applying similar arguments to the other variables, the Hessian for the log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero, and $\hat{t}$ uniquely maximizes $\ell$.
@@
 \[
-\hat{x}_1 = 2\cdot\left(\sum_{j=1}^\nSiteSplits c_je_j\right) - 1
+\hat{x}_1 = 2\cdot\left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) - 1
 \]
@@
 \[
-\sum_{j=1}^\nSiteSplits c_je_j = \sum_{j=1}^\nSiteSplits \mathbf{1}\{\mbox{site split $j$ has term $(1+x_1)$}\} \cdot p_{\siteSplit_j}.
+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j = \sum_{j=1}^\nSiteSplits \mathbf{1}\{\mbox{site split $j$ has term $(1+x_1)$}\} \cdot p_{\siteSplit_j}.
 \]
-As an example, Table~\ref{tab:likelihoods-restricted} shows the maximal ancestral state splits and corresponding likelihood values for $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$.
-In this case,
-\[
-\sum_{j=1}^\nSiteSplits c_je_j = p_{\emptyset} + p_2 + p_3 + p_{23} = \frac{1}{2} + \frac{1}{2}x^*(y^*)^2
-\]
-and $\hat{x}_1 = x^*(y^*)^2$.
 
 %EM Isn't this table redundant given Table S3?
-\begin{table}
-\centering
-\begin{tabular}{|lll|l|}
-\hline
-$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_{\siteSplit_j}$ & $32\cdot\Pr(\siteSplitRV=\siteSplit_j,\ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau_1,t)$\\
-\hline
-$\emptyset$&$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
-$\{1\}$    &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
-$\{2\}$    &$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
-$\{3\}$    &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
-$\{1,2,3\}$&$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
-$\{1,2\}$  &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
-$\{2,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
-$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
-\hline
-\end{tabular}
-\caption{
-32 times the maximal likelihood values on the InvFels tree $\tau_1$ where $\emptyset$ is the most likely ancestral state split for each site split.
-}
-\label{tab:likelihoods-restricted}
-\end{table}
+%das: I wrote out the likelihood in the example after the proof
 
@@
 \[
-\hat{x}_1 = a_{x_1,0} + a_{x_1,1}(x^*)^2 + a_{x_1,2}(y^*)^2 + a_{x_1,3} x^*(y^*)^2 + a_{x_1,4}(x^*)^2(y^*)^2.
+\hat{x}_1(x^*,y^*) = a_{x_1,0} + a_{x_1,1}(x^*)^2 + a_{x_1,2}(y^*)^2 + a_{x_1,3} x^*(y^*)^2 + a_{x_1,4}(x^*)^2(y^*)^2,
 \]
 where $a_{x_1,k}$ are constants independent of $x^*$ and $y^*$---in fact, $a_{x_1,k}$ takes values in the set $\{i/8 : i=-4,-3,\ldots,7,8\}$.
-The true branch fidelity for $x_1$ is $x^*$.
+The true branch fidelity for $x_1$ is $x^*$, and so we have consistency when
+\[
+f_{y^*}(x^*) = \hat{x}_1(x^*,y^*) - x^*
+\]
+is zero.
+As the number of zeros of $f_{y^*}(x^*)$ is finite for $0 < x^*, y^* < 1$, consistency holds only on a set of measure zero.
 %EM The following sentence falls short of getting us rigorously to where we need to go, which is the measure zero statement.
 Doesn't that just follow from the number of zeros of a polynomial being finite?
-It is possible to obtain $\hat{x}_1=x^*$, such as when $y^*=1$ or when $(x^*)^2=x^*$, i.e., one of the generating branch lengths is zero or infinite, but we cannot have $\hat{x}_1=x^*$ for all $x^*$.
-The same is true for $x_2$.
-A similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent on a set of measure zero.
+%das: fixed
+The same is true for $x_2$, and a similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent on a set of measure zero.
 \end{proof}
@@
 %EM It's a confusing read. First you say that we're going to look at examples, but then perhaps you cast doubt on if the examples exist "There may be no..."
-Examples of generating parameters resulting in consistency of joint inference occur in the following.
-Let $x^*$ and $y^*$ be such that the estimate for $\hat{x}_1$ is
+%das: moved some things around so the example is outside of the proof
+As an example solution for maximizing $x_1$, the maximal ancestral state splits for $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$ (Table~\ref{tab:likelihoods}) yields the log-likelihood
+\begin{align*}
+\ell_{\tau^*,t^*}(x_1; \hat{\boldsymbol\xi}_1) &= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
+&\qquad + \left(\frac{1}{2}+\frac{1}{2}(y^*)^2\right)\log(1+y_1) + \left(\frac{1}{2}-\frac{1}{2}(y^*)^2\right)\log(1-y_1)\\
+&\qquad + \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_2) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_2)\\
+&\qquad + \log(1+y_2) + \log(1+w) - \log 32\\
+&:= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
+&\qquad + C_{x_1}(x_2,y_1,y_2,w).
+\end{align*}
+where $C_{x_1}$ is a function of $y_1,x_2,y_2,$ and $w$ and
 \[
-\hat{x}_1 = 2 \cdot \left(p_{\emptyset} + p_2 + p_{123} + p_{13}\right) - 1 = (x^*)^2.
+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j = p_{\emptyset} + p_2 + p_3 + p_{23} = \frac{1}{2} + \frac{1}{2}x^*(y^*)^2.
 \]
-There may be no $x^*$ and $y^*$ that yield this maximum (Fig.~\ref{fig:max-anc-state}), though it is a feasible solution to the general joint inference problem.
-In this case, $\hat{x}_1 = x^*$ implies consistency only occurs when $x^*=0$ or $x^*=1$.
-Another case,
-\[
-\hat{x}_1 = 2 \cdot \left(p_{\emptyset} + p_2 + p_{3} + p_{123} + p_{23} + p_{13}\right) - 1 = \frac{1}{2} + \frac{1}{2}(x^*)^2
-\]
-shows $\hat{x}_1 = x^*$ implies consistency only when $x^*=1$.
+Thus, $\hat{x}_1(x^*, y^*) = x^*(y^*)^2$, and it is possible to obtain $\hat{x}_1(x^*,y^*)=x^*$, such as when $y^*=1$, but we cannot have $\hat{x}(x^*,y^*)_1=x^*$ for all $x^*$ where $0 < x^*, y^* < 1$.
 
@@
         \centering
-        Maximal ancestral state split definitions
-        \begin{tabular}{l}
+        Maximal ancestral state split definitions and $\hat{w}$
+        \begin{tabular}{ll}
         \hline
-        $\hat{\boldsymbol\xi} = \{\xi_{\emptyset},\xi_{1},\xi_{2},\xi_{3},\xi_{123},\xi_{12},\xi_{23},\xi_{13}\}$\\
+        $\hat{\boldsymbol\xi} = \{\xi_{\emptyset},\xi_{1},\xi_{2},\xi_{3},\xi_{123},\xi_{12},\xi_{23},\xi_{13}\}$ & $\hat{w}$\\
         \hline
-        $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$\\
-        $\hat{\boldsymbol\xi}_2 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\emptyset,\emptyset\}$\\
-        $\hat{\boldsymbol\xi}_3 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\emptyset,\{1\}\}$\\
-        $\hat{\boldsymbol\xi}_4 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\{1,2\},\{1\}\}$\\
-        $\hat{\boldsymbol\xi}_5 = \{\emptyset,\emptyset,\emptyset,\{1\},\{1\},\emptyset,\{1\},\{1\}\}$\\
+        $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$ & $1$\\
+        $\hat{\boldsymbol\xi}_2 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\emptyset,\emptyset\}$ & $1$\\
+        $\hat{\boldsymbol\xi}_3 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\emptyset,\{1\}\}$ & $\frac{3}{4}+x^*(y^*)^2-\frac{1}{4}((x^*)^2+(y^*)^2+(x^*)^2(y^*)^2)$\\
+        $\hat{\boldsymbol\xi}_4 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\{1,2\},\{1\}\}$ & $\frac{3}{4}+x^*(y^*)^2-\frac{1}{4}((x^*)^2+(y^*)^2+(x^*)^2(y^*)^2)$\\
+        $\hat{\boldsymbol\xi}_5 = \{\emptyset,\emptyset,\emptyset,\{1\},\{1\},\emptyset,\{1\},\{1\}\}$ & $x^*(y^*)^2$\\
         \hline
@@
 %EM The rest of this paragraph is obvious, no?
-Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the region in Fig.~\ref{fig:max-anc-state} where $\hat{\boldsymbol\xi}=\hat{\boldsymbol\xi}_1$.
-We can see this results in the likelihood of both topologies being equal, i.e.,
-\begin{align*}
-\max_{t : t^*\in\mathcal{T}_0} & \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)\\
-&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_1} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
-&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_2} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, \{x_1,y_1,x_2,y_2,w\}).
-\end{align*}
+%das: yeah seems like it.
 
@@
 %EM Are you saying that we should be able to obtain this next display equation by maximizing over the entries in Table S3? I think this needs more explanation.
-Using the notation in the proof of Theorem~\ref{thm:bl_theorem}, we see that, in general, our likelihood given in~\eqref{eq:site_pattern_profile_likelihood_mean} has an exact, closed form maximum of
+%das: included more detail
+Referring to the proof of Theorem~\ref{thm:bl_theorem}, we see that
+\[
+\ell(x_1) \sim \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log (1-x_1)
+\]
+and
+\[
+\hat{x}_1 = 2\cdot\left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) - 1
+\]
+so that the maximal value for $\ell(x_1)$ is
+\[
+\frac{1+\hat{x}_1}{2}\log(1+\hat{x}_1) + \frac{1-\hat{x}_1}{2}\log(1-\hat{x}_1).
+\]
+With similar forms for the remaining variables our likelihood given in~\eqref{eq:site_pattern_profile_likelihood_mean} has an exact, closed form maximum of
 \begin{align}
@@
 \]
-and the exact, closed form value of the likelihood is
+and the exact, closed form value of the likelihood is, as a function of $x^*$ and $y^*$,
 \begin{align}
 %EM I wouldn't introduce the \ell_{U,1} notation unless it's really needed for some reason I don't see yet.
-\ell_{U,1}(x^*,y^*) &= (1+x^*(y^*)^2)\log(1+x^*(y^*)^2) + (1-x^*(y^*)^2)\log(1-x^*(y^*)^2)\nonumber\\
+%das: written as just a function of x* and y*
+g_{\boldsymbol\xi_1}(x^*,y^*) &= (1+x^*(y^*)^2)\log(1+x^*(y^*)^2) + (1-x^*(y^*)^2)\log(1-x^*(y^*)^2)\nonumber\\
 &\qquad + \frac{1+(y^*)^2}{2}\log(1+(y^*)^2) + \frac{1-(y^*)^2}{2}\log(1-(y^*)^2)\nonumber\\
@@
 \end{align}
-We obtain similar maximal values as functions of $x^*$ and $y^*$ for each ancestral state split.
+We obtain similar maximal values as functions of $x^*$ and $y^*$ for each ancestral state split to get $g_{\boldsymbol\xi}$ for all relevant $\boldsymbol\xi$.
 The curves delineating maximal ancestral state splits (Fig.~\ref{fig:max-anc-state}) are determined by these 81 functions.

### 2018-10-05 joint_inf.tex
getting there

@@
 In words, the joint estimation procedure does not recover the true generating $t^*$ almost everywhere in the space of generating parameters.
-The proof can be seen intuitively through the fact that the estimator for a given branch fidelity cannot isolate the individual generating parameter as a linear term since the estimator itself is a combination of nonlinear functions (in $(x^*,y^*)$) of generating fidelities.
+The proof can be seen intuitively through the fact that the estimator for a given branch fidelity cannot isolate the individual generating parameter as a linear term since the estimator itself is a combination of nonlinear functions in $(x^*,y^*)$ of generating fidelities.
 For example, in estimating $x_1$, $\hat{x}_1$ is a linear combination of $p_{\siteSplit_j}$ values.
@@
 This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:bl-general-inconsistency}).
+Although the regions seem coarse, $\hat{w}$ has a closed form in the entire space of $x^*$ and $y^*$ (table in bottom panel of Fig.~\ref{fig:max-anc-state}).
 As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
@@
 %EM I don't feel like this "range" is a very strong statement. What if we just repeat that the bias is of the same order as the branch lengths and give an example?
-Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ range from $[-4\times 10^{-2}, 3\times 10^{-3}]$.
-showing that even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
+%das: that works, though I'm not too clear on what kind of example you're talking about. Maybe Fig. 3 is enough?
+Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ are on the same order as the branch lengths (Fig.~\ref{fig:empirical-bias}), showing that even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
 

### 2018-10-14 appendix.tex
Going through revisions.

@@
 For fixed $\boldsymbol\xi$, the log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
-%EM These c_j's and their properties need to be explained a lot more clearly. What are they, how can we know that the likelihood has this form, and why they have the properties below.
-%das: heh they must have been carried over---they're actually p_yj's
+%EM The previous version switched between only looking at x_1 and the general case. Here I tried to make that more flexible and explicit.
 \begin{align*}
@@
 Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
-Without loss of generality we focus on the variable $x_1$ and the function
+For example, if we focus on the variable $x_1$, holding other variables constant, then
 \[
@@
 where $\sim$ denotes equality up to an additive constant.
-Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, is
+Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to the focal variable (here $x_1$) and the initial $1/32$ constant, takes the form
 \[
@@
 \begin{align}
-%EM Here's this propto thing again. It's up to a constant, not propto, right?
-%das: right---reverted to \sim
 \ell(x_1) &\sim \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}(1-e_j)\right) \log (1-x_1)\\
@@
 As $x_1\in(0,1]$, we need only $0 \le \sum_j p_{\siteSplit_j}e_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
-%EM We need to hear why both of these properties hold.
-%das: now that c_j is changed to the probability maybe it's more clear? e_j in {0,1} is by definition above
 Since $p_{\siteSplit_j}$ are probabilities, then $\sum_j p_{\siteSplit_j} = 1$.
@@
 
-%EM Isn't this table redundant given Table S3?
-%das: I wrote out the likelihood in the example after the proof
-
-We show that solutions of this form never obtain $\hat{t} = t^*$ except on a set of measure zero.
+Next we show that solutions of this form never obtain $\hat{t} = t^*$ except on a set of measure zero.
 Given Table~\ref{tab:gen-sitepatprob}, all solutions $\hat{x}_1$ have the form
@@
 As the number of zeros of $f_{y^*}(x^*)$ is finite for $0 < x^*, y^* < 1$, consistency holds only on a set of measure zero.
-%EM The following sentence falls short of getting us rigorously to where we need to go, which is the measure zero statement.
-Doesn't that just follow from the number of zeros of a polynomial being finite?
-%das: fixed
 The same is true for $x_2$, and a similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent on a set of measure zero.
+
+%EM This is much improved, thanks! It seems like we need a little more to complete the proof in the general case. We start the proof hypothesizing a fixed \xi. When do we relax that assumption?
 \end{proof}
@@
 \]
-Thus, $\hat{x}_1(x^*, y^*) = x^*(y^*)^2$, and it is possible to obtain $\hat{x}_1(x^*,y^*)=x^*$, such as when $y^*=1$, but we cannot have $\hat{x}(x^*,y^*)_1=x^*$ for all $x^*$ where $0 < x^*, y^* < 1$.
+Thus, $\hat{x}_1(x^*, y^*) = x^*(y^*)^2$, and it is possible to obtain $\hat{x}_1(x^*,y^*)=x^*$, such as when $y^*=1$, but we cannot have $\hat{x}_1(x^*,y^*)=x^*$ for all $x^*$ where $0 < x^*, y^* < 1$.
 
@@
 Moreover, if $w=1$, the internal branch becomes zero-length and the two topologies are indistinguishable.
-%EM The rest of this paragraph is obvious, no?
-%das: yeah seems like it.
 
 The boundaries determining maximal ancestral state splits (Fig.~\ref{fig:max-anc-state}) are obtained through maximizing over 81 separate functional values (Table~\ref{tab:likelihoods}).
-%EM Are you saying that we should be able to obtain this next display equation by maximizing over the entries in Table S3? I think this needs more explanation.
-%das: included more detail
 Referring to the proof of Theorem~\ref{thm:bl_theorem}, we see that
@@
 \]
+%EM The following sentence needs revision.... read aloud?
 With similar forms for the remaining variables our likelihood given in~\eqref{eq:site_pattern_profile_likelihood_mean} has an exact, closed form maximum of
@@
 \begin{align}
-%EM I wouldn't introduce the \ell_{U,1} notation unless it's really needed for some reason I don't see yet.
-%das: written as just a function of x* and y*
-g_{\boldsymbol\xi_1}(x^*,y^*) &= (1+x^*(y^*)^2)\log(1+x^*(y^*)^2) + (1-x^*(y^*)^2)\log(1-x^*(y^*)^2)\nonumber\\
+g_{\boldsymbol\xi_1}(x^*,y^*) &:= (1+x^*(y^*)^2)\log(1+x^*(y^*)^2) + (1-x^*(y^*)^2)\log(1-x^*(y^*)^2)\nonumber\\
 &\qquad + \frac{1+(y^*)^2}{2}\log(1+(y^*)^2) + \frac{1-(y^*)^2}{2}\log(1-(y^*)^2)\nonumber\\

### 2018-11-06 appendix.tex
going through comments

@@
 I suggest round parensI suggest round parens.
-%das: TODO
+%das: on second thought, this is using notation like in Semple/Steel.
+If we change this here, I think we'd have to change almost all curly braces in the document, including the ones that use the Semple/Steel notation.
+I'm not necessarily opposed, but I think it might be confusing, especially because we use their notation a lot throughout.
+Thoughts?
 For a fixed topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
@@
 
-The likelihood in \eqref{eq:profile_likelihood} can be written as
+The likelihood~\eqref{eq:profile_likelihood} can be written as
 \begin{align}
@@
 Let me know what you think.
+%das: bump
 For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
@@
 %EM The previous version switched between only looking at x_1 and the general case. Here I tried to make that more flexible and explicit.
+%das: looks good :+1:
 \begin{align*}
@@
 is zero.
-As the number of zeros of $f_{y^*}(x^*)$ is finite for $0 < x^*, y^* < 1$, consistency holds only on a set of measure zero.
+As the number of zeros of $f_{y^*}(x^*)$ is finite for $0 < x^*, y^* < 1$, consistency holds only on a set of measure zero for a fixed $\boldsymbol\xi$.
+Since we have inconsistency in $\hat{t}_{\boldsymbol\xi}$ for each $\boldsymbol\xi$, this implies an inconsistency when maximizing over $\boldsymbol\xi$ as it takes values on a finite set.
+
 The same is true for $x_2$, and a similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent on a set of measure zero.
@@
 %EM This is much improved, thanks! It seems like we need a little more to complete the proof in the general case. We start the proof hypothesizing a fixed \xi. When do we relax that assumption?
+%das: yah, maximizing over xi is straightforward since it takes values on a finite set. I changed a bit of notation so hope it's more clear now.
 \end{proof}
@@
 %das: moved some things around so the example is outside of the proof
+%das: bump
 As an example solution for maximizing $x_1$, the maximal ancestral state splits for $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$ (Table~\ref{tab:likelihoods}) yields the log-likelihood
@@
 \[
-\frac{1+\hat{x}_1}{2}\log(1+\hat{x}_1) + \frac{1-\hat{x}_1}{2}\log(1-\hat{x}_1).
+\frac{1+\hat{x}_1}{2}\log(1+\hat{x}_1) + \frac{1-\hat{x}_1}{2}\log(1-\hat{x}_1)
 \]
 %EM The following sentence needs revision.... read aloud?
-With similar forms for the remaining variables our likelihood given in~\eqref{eq:site_pattern_profile_likelihood_mean} has an exact, closed form maximum of
+%das: changed
+with similar forms for the remaining variables.
+Our likelihood~\eqref{eq:site_pattern_profile_likelihood_mean} has an exact, closed form maximum of
 \begin{align}

### 2019-01-11 cover-letter/supervisor.tex
added richard
spacing for one page
update cover letter

--- 2018-02-03 cover-letter/supervisor.tex
@@
 
-Please consider our manuscript entitled ``Joint estimation of phylogenies and ancestral states is not consistent'' for publication in the \textit{Molecular Biology and Evolution}.
-In this paper, my co-author David Shaw and I investigate the statistical properties of joint maximum-likelihood estimation of trees and ancestral sequences.
+Please consider our revised manuscript entitled ``Joint maximum-likelihood of phylogeny and ancestral states is not consistent'' for publication in \textit{Molecular Biology and Evolution}.
+In this paper, my co-authors, David Shaw and Vu Dinh, along with myself, investigate the statistical properties of joint maximum-likelihood estimation of trees and ancestral sequences.
 We were motivated to do so by recent work from Richard Neher's group (\url{http://dx.doi.org/10.1093/ve/vex042}) that employs such joint optimization as an efficient alternative to the standard Felsenstein marginalization.
@@
 We show that such joint inference is not consistent, both for topology inference and branch length optimization.
-Indeed, there are sizeable regions of parameter space in which this joint inference gives the wrong topology and branch lengths in the limit of large data.
+For a fixed topology, almost the entirety of the parameter space yields inconsistent branch length estimates.
+Additionally, there are sizeable regions of parameter space in which this joint inference gives the wrong topology in the limit of large data.
 This is not a case of long branch attraction.
-Indeed, the inconsistency in Felsenstein 1978 requires a combination of long and short branches, whereas our inconsistency comes from all branches being long.
+Indeed, the inconsistency in Felsenstein 1978 requires a combination of long and short branches, whereas our inconsistency arises in roughly half of the parameter space.
 
 Although the proofs in our paper require detailed examination of likelihood functions, we have written the paper to be broadly accessible.
-The paper has five illustrations showing various regions of inconsistency, and provides intuitive summaries of the results in the main text.
+The paper has two illustrations showing regions of inconsistency for topology and branch length inconsistency, respectively, and provides intuitive summaries of the results in the main text.
 A complete set of proofs appears in the appendix.
@@
 \begin{itemize}
+\setlength\itemsep{1.4pt}
 \item Elizabeth Allman \texttt{esallman@alaska.edu}
@@
 \item Ed Susko \texttt{susko@mathstat.dal.ca}
+\item Richard Neher \texttt{richard.neher@unibas.ch}
 \end{itemize}

### 2019-01-15 versions/response.v1.txt
Edit response.

@@
 
-Thank you for your helpful comments. We think that this version of the paper is significantly improved, and we hope you agree.
+We are grateful to the reviewers for their helpful comments. Here we submit a very substantial revision implementing almost all of them.
+
+We realize that this paper is being resubmitted well beyond the typical deadline for responses.
+We apologize for that, however we hope the reviewers understand that in the interim the first author has gotten an industry job, we have added an author, and we have rewritten a sizeable fraction of the proofs.
 
@@
 Even if we were to greatly expand the sections doing numerical optimization, these results would only be for the binary symmetric model, which is never applied in practice.
-An investigation of these aspects deserves another paper.
-We have added clarification of these points to the discussion:
-"All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners."
+An investigation of bigger trees and more complex models deserves another paper.
+We have added clarification of these points to the discussion: "All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners."
 
@@
 Indeed, obtaining these regions in closed form is feasible, though it involves comparing a large number of two-variable functions to one another.
-Fig. S2 shows the regions where a particular likelihood equation will be maximal, and allows us to compute---in closed form---the value of the branch parameter in Fig. 2.
+Fig.
+S2 shows the regions where a particular likelihood equation will be maximal, and allows us to compute---in closed form---the value of the branch parameter in Fig.
+2.
 
@@
 
-Thank you very much for reading the derivations and for your suggestions.
-As noted inline we have revised the text accordingly.
+Thank you very much for reading the derivations and for your suggestions. As noted inline we have revised the text accordingly.
 
@@
 
-We have added the definition of a "generating probability" to ease confusion in this sentence:
-"As data becomes plentiful, we use the Hadamard transform (see (8) in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form."
+We have added the definition of a "generating probability" to ease confusion in this sentence: "As data becomes plentiful, we use the Hadamard transform (see (8) in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form."
 
@@
 
-Thank you for highlighting this omission.
-You are correct, and we have added parenthetical notes to remind the reader where appropriate.
+Thank you for highlighting this omission. You are correct, and we have added parenthetical notes to remind the reader where appropriate.
 
@@
 
-We apologize for the confusion, as this was not as clear as it should have been.
-We have revised to:
+We apologize for the confusion, as this was not as clear as it should have been. We have revised to:
 "The mapping $\patternToSplit$ operates by returning the tips labeled as $1$ in a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if the set of tips labeled $1$ is not in $\mathcal{P}(\siteSplitSet)$.
@@
 
-Asymmetries can drive these particular choices, but Fig. 2 in the revised manuscript shows that almost half of the parameter space results in an interior branch length being estimated as zero.
+Asymmetries can drive these particular choices, but Fig.
+2 in the revised manuscript shows that almost half of the parameter space results in an interior branch length being estimated as zero.
 We do not think providing an argument about accumulating inaccuracies would elucidate why such starkly-defined regions of inconsistency exist.
@@
 We believe it's important not to mischaracterize the suggested use of joint inference, though we would like to emphasize to the reader that problems are apparent in large regions of the parameter space.
-We think the current discussion treats this optimization procedure as fairly as possible.
+We think the current discussion treats this optimization procedure as fairly as possible, although we would gladly accept any further specific suggestions.
 
@@
 Thank you for your assessment.
-We have worked hard in our initial submission and in our revision to write out the argument in great (indeed, excruciating) detail, and hope that you will be able to read the revision more fully.
+We have worked hard in our initial submission and in our revision to write out the argument in great (indeed, excruciating) detail, and hope that reviewers will be able to read the revision more fully.
 
@@
 
----
-
 Goldman (1990) concerns estimating both structural (e.g., topology, branch lengths) and incidental (e.g., ancestral states) parameters simultaneously, though the only worked example is in the case of a topology with equal branch lengths.
@@
 
----
-
 > Finally, let me just add that I hate the term "Farris Zone".  The

### 2019-03-19 versions/response.v2.txt
Adding response template and old version of ms

> The reviewers have universally viewed the revision as much improved.
> That said, 3/4 maintain that further changes are necessary prior to
> publication. It is our view that the remaining criticisms are "minor",
> and should you address them thoroughly and convincingly, we expect
> that final acceptance of the revision can be done editorially.

XXX

> Reviewers' comments:
> Reviewer: 1

> Comments to the Author
> As with the previous version, I believe the topic matter will be of
> interest. For me, however, the work is incomplete in important ways
> that I believe can be relatively easily addressed. The authors went
> beyond my expectations in calculating explicit solutions for the
> limiting edge-lengths (Fig S2) when tau1 is the putative topology of
> estimation. This allows them to draw conclusions about edge-length
> inconsistency but there is no definitive conclusion about `strict'
> topological inconsistency (in the sense of converging upon the wrong
> tree rather than a multifurcating one). From the statements near l238
> it seems the authors have completed some preliminary investigation on
> strict topological inconsistency and find no evidence for it at
> present. Assuming that remains true, the findings of the current ms
> are of a different nature than the previous version where strict
> topological inconsistency was found. In settings with a limiting
> multifurcating topology, because it is multifurcating, for finite but
> large sequence-lengths, measures of uncertainty (eg. bootstrap
> support) will suggest that there is substantial uncertainty in
> whatever topology gets estimated (this probably deserves discussion).
> The joint optimization approach thus lacks the power to estimate the
> correct topology where ML would but is not being positively mislead.
> Assuming the lack of strict topological inconsistency holds, the
> results for the method are not as negative as they were in the
> previous version.

TODO: add discussion of bootstrap

> Getting complete results about topological consistency is feasible.
> The arguments that led to the closed form solution for the limiting
> edge-length estimates for tau_1 can be repeated similarly for tau2.
> This would allow, for any given x* & y*, comparison of the limiting
> lnl per site for tau1 vs tau2. Whichever topology has the larger lnl
> is the long-run estimated topology. In cases where hat w=1 for both
> topologies, conclusions are different but definite. Thus it seems a
> relatively straightforward exercise given the work the authors have
> already done (I hope I am not missing a subtle difficulty) to have a
> computer determine for a grid of (x*,y*) whether (a) hat w=1 exactly
> for both topologies, (b) the difference in limiting lnl over
> topologies is either positive or negative or (c) hat w neq 1 for at
> least one of the topologies and the difference in limiting lnl is 0 up
> to machine precision. I would emphasize that the cases (a) and (b)
> constitute rigorous proof. The cases (c) likely are rare and
> correspond to boundary cases, so the boundary may have some slight
> uncertainty. There may be some aesthetic appeal to getting analytical
> results rather than using a computer but being able to draw clear
> conclusions about strict topological inconsistency is important.

DISCUSS: we have no evidence of topological inconsistency

> The language used to refer to the approach and references to the
> literature are awkward. The references to profile likelihood work
> suggest that there is a theoretical basis for the approach (that paper
> suggests consistency) but the settings considered in such work are
> quite different, I believe. Integrated likelihood usually refers to
> something different too. Finally, as Reviewer 3 makes clear the only
> current implementation of the approach was intended as an
> approximation to ML rather than as ML. The central issue is whether
> treating unobserved data (ancestral states here) as parameters
> sometimes works; the EM algorithm were developed with the idea that
> this is not what should be done. In any case, references directly tied
> to that issue would be more relevant. Additional suggestions include:
> 1) Make a clear statement in the intro about why the usual likelihood
> is considered the right likelihood: it is the probability of the
> *observed* data. 2) The point that large numbers of parameters need
> not defeat likelihood methods is valid but it should be made clear
> when referencing such work that the setting is different.

> -l21: Some rewording is needed. The joint method will, as n gets
> large, continue to estimate a well-resolved topology with positive
> probability; albeit with small middle edge-length.

> -l131: A reader might be able to avoid the appendix (which is
> difficult to wade into) if an example zeta in a four taxon case was
> given as illustration.

> -l150: I initially found this paragraph confusing. Table S1 is a
> result that you are using as intuition, so you can't be giving
> intuition for the entire result (otherwise it gets circular). I think
> you are trying to give intuition for the `everywhere except a set of
> measure' portion of the result rather than entire result. Perhaps `the
> proof that hat t neq t* except on a set of measure 0 can be seen
> intuitively...'.

> -l154 and (should check elsewhere): `the limiting value of the
> estimator' rather than `the estimator' is what is being referred to
> here.

> -l186: which `likelihood terms' is not that clear to me: `each site
> results in multiple factors, only one of which depends on w with
> dependence being through (1+w) or (1-w) depending ...'

> -l189: `depends on the values of other branch parameters' or some
> other rewording. I don't think it is that they are unknown. It is that
> they need optimization, so you can't optimize w independently of them.

> -l201, Figs 2 and 3: I don't understand why direct numerical
> optimization was needed. You have the exact expressions for hat w in
> Fig S2. So you could have plotted them as Fig 2. A similar comment
> applies to Fig 3. Are you trying to convince the reader that your
> proofs have no errors? That seems more an issue for supp material.

> -l252-259: I don't think this is an example of profile likelihood as
> considered in Murphy and Van der Vaart.

> -l264: Do you mean that `consistency can hold even when the numbers of
> parameters tend to infinity'? The statement `not a necessary
> condition' just means that inconsistency can arise in different ways
> (eg. lack of identifiability).

> Reviewer: 2

> Comments to the Author
> Shaw et al provided a revised version of the manuscript that I
> reviewed about a year ago showing that joint estimation of ancestral
> states and branch length is in general not consistent. The authors now
> map out the regions in which topologies are incorrectly estimated and
> include a plot that shows the bias of the internal branch length for
> different tree configurations. I think this is a useful piece of work.
> But some parts are completely unsurprising, while some empirical
> results seem inaccurate or misleading.

> The most important results/claims by Shaw et al are:

> 1) "jointly estimating phylogenetic parameters along with ancestral
> states is not consistent in general"... "only correct on a set of
> measure 0".
> It is clearly true that branch length will be biased. But the fact
> that joint estimation will not recover the true branch lengths is not
> surprising but perfectly expected. Joint estimation ignores subleading
> configurations of internal states. Since the subleading terms
> typically require more state changes, branch length estimates ignoring
> them are biased. In a similar way a Taylor expansion is almost nowhere
> "correct", but still a useful and insightful approximation (a better
> analogy is ignoring looped diagram in a diagrammatic expansion.).

> 2) In some parameter regions the internal branch of the four taxon
> tree collapses to 0.
> Again, this is true and identifying regions in parameter space where
> the internal branch collapses to zero is the most interesting part of
> the manuscript in my opinion. Unfortunately, the manuscript doesn't
> give much insight as to why the length of the internal branch shrinks
> to zero in some regions of parameter space. There is, however, a
> rather intuitive explanation for this. In the inv Fels topology, the
> internal branch collapses to 0 if
>
> p_x^2 > c p_y
>
> (the prefactor c depends on specifics like alphabet size etc). And
> this has an intuitive interpretation: if p_x^2 > p_y, it is more
> likely to observe a mutation twice on two branches of length p_x than
> once on a branch of length p_y. As a result, the joint inference will
> push p_w to zero and assign all state changes on the terminal branches
> of length p_x. But for short branches, this corresponding sector of
> parameter space is a small (tending to zero for a fixed p_x/p_y with
> p_x \to 0).
>
> My empirical tests of the approximate joint inference scheme suggest
> this region is much smaller than the one identified by Shaw et al in
> Fig 3.

> 3) "[The bias is] of the same order as the branch lengths themselves."
> This is incorrect or misleading, depending on how you read it. In
> typically cases, the error made by this procedure should be quadratic
> in branch length and hence of higher order than the branch length
> itself. All empirical test we have done via numerically optimizing
> branch lengths suggests this is true for typical trees and that the
> errors made are small provided ALL adjacent branch lengths are short.
> That said, relative error of one branch can be large if adjacant
> branches are long such as the case above. In this case the error is
> quadratic in the length of the longer adjacent branches rather than
> the branch itself.

> Reviewer: 3

> Comments to the Author
> This a resubmission of an earlier version of the paper, for which I
> was reviewer #4.  The other reviewers of the previous submission did
> an excellent job in pointing out minor problems and making suggestions
> for improvement.  It seems that Shaw et al. have addressed all of
> these suggestions in exemplary fashion, and  think the result is a
> much stronger paper.  I especially appreciate the authors' efforts to
> make the mathematical derivations and proofs more accessible to
> general readers (like myself).

> I often write long reviews but given the careful attention to detail
> in responding to the previous reviews of an already good paper, I have
> nothing further to add.

> Reviewer: 4

> Comments to the Author
> The authors have made some substantial changes addressing the concerns
> of the last round and I think that has improved the clarity and in
> some instances also strengthened the results.
> Remaining comments are mostly about the notation which I still find a
> bit hard to follow.
> Why are site and ancestral node splits carrying this distinction of
> whether the m'th taxon (or node) has a 1?

> Perhaps it's a latexdiff leftover but eq 6 in line 151 seems to have a
> line with just a ; \xi )

> Line 593 p 25 I don't understand "the mapping \psi operates by
> returning the tips labeled as 1 in a site pattern to obtain a site
> split in P(Y) if the set of tips labeled 1 is not in P(Y)" and this
> mapping seems fundamental. Perhaps you mean to refer to the m'th taxon
> (or node)..?

> The symmetric model is assumed strongly in the first part of the
> appendix, with the factor of 2; my guess is that the results hold in
> asymmetric models but the notation is harder, is that correct?

> What's q in equation 7 in the appendix (line 655) ?

> More substantitvely, the limit in (8) seems not established because as
> n -> \infty so would q and n_j, so these can't be treated as fixed .

> page 9 - at this point in the paper, unless the reader has already
> read the appendix they don't know what the estimator means or what the
> p_ytilde j are - so this text doesn't make sense at this point.

> page 30 line 667-8 what are S and Y and what makes a path? I gather
> from the math that you mean paths in the tree, since the sum is over
> ancestral states and there's a product of probabilities of changes on
> branches but (perhaps my lack of background) the form of (9) isn't
> obvious I think because I am not exactly clear on what the notation
> means.

### 2019-03-19 versions/joint_inf.v2.tex
Adding response template and old version of ms

--- 2018-10-05 joint_inf.tex
@@
 Maximum likelihood estimation in phylogenetics requires a means of handling unknown ancestral states.
-Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
+Classical maximum likelihood averages over these unknown intermediate states, leading to provably consistent estimation of the topology and continuous model parameters.
 Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
@@
 We show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
-We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
-For branch length estimation on the correct topology, we show that this joint method cannot correctly estimate branch lengths except on a set of measure zero, and we show empirically that branch lengths are biased in this setting even for short branches, with bias of the same order as the branch lengths themselves.
+We find a sizeable region of parameter space that generates data on a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
+More generally, we show that this joint method only estimates branch lengths correctly on a set of measure zero.
+We show empirically that branch length estimates are biased even for short branches, with bias of the same order as the branch lengths themselves.
 \end{abstract}
@@
 
-This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:bl-general-inconsistency}).
-Although the regions seem coarse, $\hat{w}$ has a closed form in the entire space of $x^*$ and $y^*$ (table in bottom panel of Fig.~\ref{fig:max-anc-state}).
+This result implies an inconsistency because the joint estimate of the interior branch length is zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:bl-general-inconsistency}).
+$\hat{w}$ is available in closed form in the entire space of $x^*$ and $y^*$ (table in bottom panel of Fig.~\ref{fig:max-anc-state}).
 As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
@@
 Although the estimates for $\hat{p}_w$ are better than the estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), joint inference still predictably underestimates the interior branch length.
-%EM I don't feel like this "range" is a very strong statement. What if we just repeat that the bias is of the same order as the branch lengths and give an example?
-%das: that works, though I'm not too clear on what kind of example you're talking about. Maybe Fig. 3 is enough?
 Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ are on the same order as the branch lengths (Fig.~\ref{fig:empirical-bias}), showing that even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

### 2019-03-19 versions/response.v2.txt
first pass through reviews

@@
 
-TODO: add discussion of bootstrap
+E TODO: add discussion of bootstrap
 
@@
 
+D TODO
+
 > -l21: Some rewording is needed. The joint method will, as n gets
@@
 
+It will estimate a branch length of zero which is equivalent to a multifurcating topology.
+
 > -l131: A reader might be able to avoid the appendix (which is
@@
 > given as illustration.
+
+E TODO
 
@@
 
+E TODO
+
 > -l154 and (should check elsewhere): `the limiting value of the
@@
 > here.
+
+DISCUSS
 
@@
 
+E TODO
+
 > -l189: `depends on the values of other branch parameters' or some
@@
 > they need optimization, so you can't optimize w independently of them.
+
+E TODO
 
@@
 
+E TODO
+
 > -l252-259: I don't think this is an example of profile likelihood as
 > considered in Murphy and Van der Vaart.
+
+D TODO
 
@@
 > (eg. lack of identifiability).
+
+D TODO
 
@@
 
+E TODO
+
 > 2) In some parameter regions the internal branch of the four taxon
@@
 > Fig 3.
+
+We have worked this out with Dr. Neher and found that the difference lay in that he was not optimizing the objective function from both sides.
+This also demonstrated that the above-listed intuitive explanation was incorrect.
 
@@
 
+Indeed.
+
 > Line 593 p 25 I don't understand "the mapping \psi operates by
@@
 
+E TODO add example
+
 > The symmetric model is assumed strongly in the first part of the
@@
 
+You have to have symmetry in order to have stationarity
+
 > What's q in equation 7 in the appendix (line 655) ?
+
+The number of site splits.
 
@@
 
+E TODO
+
 > page 9 - at this point in the paper, unless the reader has already
@@
 > p_ytilde j are - so this text doesn't make sense at this point.
+
+E TODO
 
@@
 > means.
+
+E TODO

### 2019-03-20 versions/response.v2.txt
Start on responses and edits.

@@
 
-XXX
+Thank you for handling the manuscript, and please thank the reviewers for their helpful comments. As stated by Reviewer 3, we believe that the review process has significantly improved the manuscript.
 
@@
 
-E TODO: add discussion of bootstrap
+%EM I honestly can't figure out what the reviewer wants to see re the bootstrap.
+Branch length inferences don't come into the bootstrap unless you contract things with zero branch lengths, which isn't generally done when doing bootstrap estimation.
+
+The reviewer is correct that this story is different than the one in the previous version of the paper.
 
@@
 
-E TODO
+We have added the suggested wording, thanks.
 
@@
 
-E TODO
+We have clarified this by specifying that we're talking about the joint likelihood expression here.
 
@@
 
-E TODO
+We have adopted the suggested wording, thanks.
 
@@
 
-E TODO
+%EM David, am I right with this?
+Although we have expressions for the likelihoods, it is not trivial to infer the branch lengths maximizing those likelihoods.
 

### 2019-03-20 joint_inf.tex
Start on responses and edits.

@@
 In words, the joint estimation procedure does not recover the true generating $t^*$ almost everywhere in the space of generating parameters.
-The proof can be seen intuitively through the fact that the estimator for a given branch fidelity cannot isolate the individual generating parameter as a linear term since the estimator itself is a combination of nonlinear functions in $(x^*,y^*)$ of generating fidelities.
+The proof that $\hat{t} \neq t^*$ everywhere except a set of measure zero can be seen intuitively through the fact that the estimator for a given branch fidelity cannot isolate the individual generating parameter as a linear term since the estimator itself is a combination of nonlinear functions in $(x^*,y^*)$ of generating fidelities.
 For example, in estimating $x_1$, $\hat{x}_1$ is a linear combination of $p_{\siteSplit_j}$ values.
@@
 For a particular site pattern, to obtain the joint maximum likelihood function we maximize over ancestral states.
-For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state corresponds to the highest conditional log-likelihood.
-As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this is not guaranteed because the maximum depends on the values of the unknown branch parameters $t$.
+For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of the terms of the joint likelihood expression depending on which ancestral state corresponds to the highest conditional log-likelihood.
+As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this is not guaranteed because the maximum depends on the values of the other branch parameters $t$.
 Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
@@
 All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners.
+Initial results by of Dr\.
+Neher using the TreeTime \citep{Sagulenko2018-xl} package has shown that the region of parameter space on which joint inference infers a zero internal branch length for 4-state DNA models is smaller than that for 2-state binary models.
 Also, given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
@@
 
+On a practical note, we would like to emphasize that TreeTime has implemented marginal mode inference since the publication of the \citet{Sagulenko2018-xl} paper.
+In January 2019 a warning was added, in all caps, that is triggered when joint inference is used for trees such that any branch length is greater than 0.15.
+
 \section*{Acknowledgements}
 We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.
+We are especially thankful to Dr.\ Neher for providing TreeTime simulation code that allowed us to test our conclusions on simulated data.
 Insight from the reviewers and the editors greatly improved this manuscript.

### 2019-03-21 versions/response.v2.txt
more progress through responses.

@@
 
+%EM Shall we strike the speculation about higher likelihoods on the wrong topology?
+
 > Getting complete results about topological consistency is feasible.
@@
 
-D TODO
+das TODO
 
@@
 
-DISCUSS
+We focus in this paper in the limiting case only.
+We have added a sentence clarifying language: (INSERT)
 
@@
 
-%EM David, am I right with this?
-Although we have expressions for the likelihoods, it is not trivial to infer the branch lengths maximizing those likelihoods.
+We have rewritten this section to clarify what is being used where.
+%EM das-- is this right?
+Specifically, although we can find the regions for which the various ancestral state splits are optimal, we cannot obtain optimal branch lengths analytically.
+
+We moved these figures around in response to a previous reviewer comment, who thought that Figure S2 was quite uninteresting while Figure 2 was relevant.
+We would prefer to leave them be.
+Figure 3 is an essential part of our argument that the internal branch length has substantial deviation from truth.
 
@@
 
-E TODO
+I'm afraid that this fact wasn't obvious to us at the outset of this project. The interest that the other reviewers have shown in the manuscript confirms that it is not broadly obvious.
 
@@
 > the branch itself.
+
+We have removed the "order" language, instead just focusing on the downward bias.
 
@@
 
-E TODO
+No, the number of taxa is fixed so this is fixed.
 

### 2019-03-21 appendix.tex
Cleaning out addressed comments.
more progress through responses.

@@
 
-%EM This is very minor compared to the below comments, so address it second in a separate commit, please, if you wish to address it.
-Curly braces are really for unordered sets, right?
-I suggest round parensI suggest round parens.
-%das: on second thought, this is using notation like in Semple/Steel.
-If we change this here, I think we'd have to change almost all curly braces in the document, including the ones that use the Semple/Steel notation.
-I'm not necessarily opposed, but I think it might be confusing, especially because we use their notation a lot throughout.
-Thoughts?
 For a fixed topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
@@
 Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
-We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
+We cannot identify complements for ancestral states We have added a sentence clarifying language:
 \[
@@
 \end{equation}
+%EM das I believe this is the definition of \ell-- this should be made more explicit.
+%EM Also, in the main text we read that \ell is the limit of L' as n goes to infinity, but here it's L''.
 
@@
 
-%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
-%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
-%Also I don't think the compactness condition holds---tree space isn't compact is it?
-%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
-%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
-%EM @das I think that ^ is all resolved, right?
-%das: I think to be fair I should add that we don't actually show a real inconsistency in topology because we haven't shown this functional convergence for it, only for branch lengths with a fixed topology.
-I added a paragraph hedging a bit.
-Let me know what you think.
-%das: bump
 For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
@@
 For fixed $\boldsymbol\xi$, the log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
-%EM The previous version switched between only looking at x_1 and the general case. Here I tried to make that more flexible and explicit.
-%das: looks good :+1:
 \begin{align*}
@@
 
-%EM This is much improved, thanks! It seems like we need a little more to complete the proof in the general case. We start the proof hypothesizing a fixed \xi. When do we relax that assumption?
-%das: yah, maximizing over xi is straightforward since it takes values on a finite set. I changed a bit of notation so hope it's more clear now.
 \end{proof}
 
-%EM I don't see this paragraph being helpful.
-%EM It's a confusing read. First you say that we're going to look at examples, but then perhaps you cast doubt on if the examples exist "There may be no..."
-%das: moved some things around so the example is outside of the proof
-%das: bump
 As an example solution for maximizing $x_1$, the maximal ancestral state splits for $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$ (Table~\ref{tab:likelihoods}) yields the log-likelihood
@@
 \]
-%EM The following sentence needs revision.... read aloud?
-%das: changed
 with similar forms for the remaining variables.

### 2019-03-21 joint_inf.tex
more progress through responses.

@@
 More generally, we show that this joint method only estimates branch lengths correctly on a set of measure zero.
-We show empirically that branch length estimates are biased even for short branches, with bias of the same order as the branch lengths themselves.
+We show empirically that branch length estimates are systematically biased downward, even for short branches.
 \end{abstract}
@@
 Although \citet{Sagulenko2018-xl} explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
-In particular, one would like to know when this approximate technique breaks down for both topology and branch length inference, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.
+One would like to know when this approximate technique breaks down, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.
 
@@
 We show that, when the topology is known and fixed, joint inference for branch length estimation cannot be consistent except on a set of measure zero (i.e.\ a set that occupies zero volume in parameter space).
-Empirically, we find areas where joint inference consistently underestimates interior branch lengths, including regions of short branch length where bias is on the same order as the branch length.
+Empirically, we find that joint inference consistently underestimates interior branch lengths, even when branch lengths are small.
 
@@
 
+Because we are interested in the properties of joint inference in the limit of $n$ going to infinity, we will focus on estimates in this case and simply refer to the right hand side of \eqref{eq:exp-log-lik} as ``the estimator.''
+
 %We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology ($\tau_2$) than for the true, generating InvFels topology ($\tau_1$).
@@
 This result implies an inconsistency because the joint estimate of the interior branch length is zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:bl-general-inconsistency}).
-$\hat{w}$ is available in closed form in the entire space of $x^*$ and $y^*$ (table in bottom panel of Fig.~\ref{fig:max-anc-state}).
+$\hat{w}$ is available in closed form in the entire space of $x^*$ and $y^*$ (Table in bottom panel of Fig.~\ref{fig:max-anc-state}).
 As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
-In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
-We elaborate on this point in the Appendix.
+In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent; we elaborate on this point in the Appendix.
 The proof is through analytically reducing the general case to 81 separate cases (Table~\ref{tab:likelihoods}) to obtain a closed form maximal value for each.
@@
 This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which the most likely ancestral states are the same for each internal node letting $x_1=x_2=x^*$ and $y_1=y_2=y^*$ ($\xi_j=\emptyset$ for all $j$ in Table~\ref{tab:likelihoods}).
-If we allow multifurcating trees in our inference, then we can think of this as an instance of converging to the wrong topology, as the true $y^*\neq 1$.
+% If we allow multifurcating trees in our inference, then we can think of this as an instance of converging to the wrong topology, as the true $y^*\neq 1$.
 
-\subsubsection*{Empirical validation}
+\subsubsection*{The outcome of joint inference is divided into distinct regions}
 
-Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the analytically\hyp{}derived region (Fig.~\ref{fig:max-anc-state}).
-To verify the regions of inconsistency and obtain a clearer picture of the closed form parameter estimates, we plot the optimal $\hat{w}$ via joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
-As before, the region of inconsistency encompasses almost half of the branch fidelity space; given the correct topology, there are many situations where we estimate the interior branch length to be zero.
+One can also use the possible cases for the form of the joint likelihood (Table~\ref{tab:likelihoods}) to understand the outcome of joint inference for various regions in parameter space.
+For example, one can plot the regions in which each of these ancestral state splits attain the maximum (Fig.~\ref{fig:max-anc-state}).
+In order to understand the optimal internal branch length in each of these cases, we optimize branch lengths numerically to obtain $\hat{w}$ under joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
+Using either technique, we see that the region of inconsistency encompasses almost half of the branch fidelity space: given the correct topology, there are many situations where the interior branch length is estimated to be zero.
 
@@
 
-In our optimization procedure, we again consider the 81 separate cases (Table~\ref{tab:likelihoods}) and, for each function, we compute the closed form solution for $\hat{t}$.
-We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*,y^*\in(0,1)$.
-Our optimization code can be found at \url{https://github.com/matsengrp/joint-inf/}.
-
 In estimating the interior branch length $w$, we find a systematic bias in the joint inference procedure even when the true branches are short (Fig.~\ref{fig:empirical-bias}).
@@
 Although the estimates for $\hat{p}_w$ are better than the estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), joint inference still predictably underestimates the interior branch length.
-Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ are on the same order as the branch lengths (Fig.~\ref{fig:empirical-bias}), showing that even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
+Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ are comparable to the branch lengths (Fig.~\ref{fig:empirical-bias}), showing that even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
 
@@
 The errors in this case \eqref{eq:marginal_likelihood} via optimization with L-BFGS-B are lower than machine tolerance.
+Our optimization code can be found at \url{https://github.com/matsengrp/joint-inf/}.
 
@@
 This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3).
-However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).
+However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and the bias is comparable to the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).
 
@@
 All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners.
-Initial results by of Dr\.
+Initial simulation-based results by Dr\.
 Neher using the TreeTime \citep{Sagulenko2018-xl} package has shown that the region of parameter space on which joint inference infers a zero internal branch length for 4-state DNA models is smaller than that for 2-state binary models.
-Also, given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
+On the other hand, when the joint inference optimization is started with a small branch length, the region ending up with a zero branch length is much larger, even for 4-state DNA models.
+In a more mathematical direction, given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
 Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.

### 2019-03-21 versions/response.v2.txt
Done with everything I can do.

@@
 
-%EM I honestly can't figure out what the reviewer wants to see re the bootstrap.
+DISCUSS I honestly can't figure out what the reviewer wants to see re the bootstrap.
 Branch length inferences don't come into the bootstrap unless you contract things with zero branch lengths, which isn't generally done when doing bootstrap estimation.
@@
 
-%EM Shall we strike the speculation about higher likelihoods on the wrong topology?
+DISCUSS Shall we strike the speculation about higher likelihoods on the wrong topology?
 
@@
 
-E TODO
+DISCUSS eliminating xi
 
@@
 
-We have worked this out with Dr. Neher and found that the difference lay in that he was not optimizing the objective function from both sides.
-This also demonstrated that the above-listed intuitive explanation was incorrect.
+We have worked this out with Dr.
+Neher and found that the difference lay in that he was not optimizing the objective function from both sides.
+This simulation work also demonstrated that the intuitive explanation by this reviewer is incorrect.
 
@@
 
+Thank you for your kind words.
+
 > Reviewer: 4
@@
 
-Indeed.
+Indeed, this was not present in the non-latexdiff pdf.
 
@@
 
-E TODO add example
+We have rewritten this confusing sentence in terms of complements, and have added a little bit of text pointing the reader to the formal definition just above.
 
@@
 
-You have to have symmetry in order to have stationarity
+Unfortunately a lot would have to change to generalize from the symmetric model.
+We would not be able to "collapse" site splits, and the Hadamard transform only works for submodels of the Kimura 3 parameter model.
+See, for example:
+
+Bryant, D. (2009). Hadamard Phylogenetic Methods and the n-taxon Process. Bulletin of Mathematical Biology, 71(2), 339–351. https://doi.org/10.1007/s11538-008-9364-8
 
@@
 
-The number of site splits.
+We now remind the reader that this is the number of site splits.
 
@@
 
-No, the number of taxa is fixed so this is fixed.
+We now emphasize that the number of taxa is fixed so q is fixed, and that although n_j goes to infinity the ratio n_j/n converges to the probability of seeing that column.
 
@@
 
-E TODO
+This is a good point! We have rephrased things, eliminating p_ytilde_j except for a parenthetical reference to the Appendix. We believe that the result is much more readable.
 
@@
 
-E TODO
+We have clarified the definitions of S and Y, and added that by path we mean graph-theoretic path.
+We are afraid that adding more detail about the Hadamard transform is beyond the scope of this paper, though we have added a link to a scanned page of the Semple and Steel book where I (Erick) have added some details on the proofs.

### 2019-03-21 appendix.tex
Done with everything I can do.

@@
 \end{definition}
-The mapping $\patternToSplit$ operates by returning the tips labeled as $1$ in a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if the set of tips labeled $1$ is not in $\mathcal{P}(\siteSplitSet)$.
+To explain the above definition in words, the mapping $\patternToSplit$ operates by taking the complement of a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if the set of tips labeled $1$ is not in $\mathcal{P}(\siteSplitSet)$.
 The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if the set of tips labeled $1$ in $\alignmentColumn$ is in $\mathcal{P}(\siteSplitSet)$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the set of tips labeled $1$ in $\ancestralStateColumn$; otherwise, the set of tips labeled $1$ in $\overline{\alignmentColumn}$ necessarily is in $\mathcal{P}(\siteSplitSet)$ and so $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\overline{\ancestralStateColumn}$.
@@
 &    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
-&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t).
-\label{eq:site_pattern_profile_likelihood_mean}
+&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \label{eq:site_pattern_profile_likelihood_mean}
 \end{split}
 \end{equation}
+because
+$\frac{\nCols_j(\fullAlignment)}{\nCols}$ converges to $\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)$.
 %EM das I believe this is the definition of \ell-- this should be made more explicit.
 %EM Also, in the main text we read that \ell is the limit of L' as n goes to infinity, but here it's L''.
+Here $j$ indexes the site splits, of which there are a fixed number $q$.
 
@@
 We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
+For additional detail on the proofs in that section, see \url{http://i.imgur.com/T5XLPLB.jpg}.
 For each edge $e$ define the edge ``fidelity'' for that edge as
@@
 where $p(e)$ is the probability of a character change along edge $e$.
-For an even-sized subset of $Y\subseteq\mathcal{S}$, let the path set $P(Y)$ be the set of edges in the path connecting both elements of $Y$.
+For an even-sized subset $Y$ of a taxon set $\mathcal{S}$, let the path set $P(Y)$ be the set of edges in the graph-theoretic path connecting both elements of $Y$.
 For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is

### 2019-03-21 joint_inf.tex
Done with everything I can do.

@@
 In words, the joint estimation procedure does not recover the true generating $t^*$ almost everywhere in the space of generating parameters.
-The proof that $\hat{t} \neq t^*$ everywhere except a set of measure zero can be seen intuitively through the fact that the estimator for a given branch fidelity cannot isolate the individual generating parameter as a linear term since the estimator itself is a combination of nonlinear functions in $(x^*,y^*)$ of generating fidelities.
-For example, in estimating $x_1$, $\hat{x}_1$ is a linear combination of $p_{\siteSplit_j}$ values.
-For the generating probabilities (top panel of Table~\ref{tab:gen-sitepatprob}), there is no linear combination that results in an isolated $x^*$ term, as all terms are either quadratic, i.e., $(x^*)^2$ or $(x^*)^2(y^*)^2$, or have both $x^*$ and $y^*$, i.e., $x^*(y^*)^2$.
+The fact that $\hat{t} \neq t^*$ everywhere except a set of measure zero can be seen intuitively through the fact that the estimator for a given branch fidelity cannot isolate the individual generating parameter as a linear term since the estimator itself is a combination of nonlinear functions in $(x^*,y^*)$ of generating fidelities.
+For example, in estimating $x_1$, $\hat{x}_1$ is a linear combination of data-generating probabilities; these probabilities are polynomials in $x^*$ and $y^*$.
+For the generating probabilities (these are called $p_{\siteSplit_j}$ values in the Appendix; see Table~\ref{tab:gen-sitepatprob} for a complete tabulation), there is no linear combination that results in an isolated $x^*$ term, as all terms are either quadratic, i.e., $(x^*)^2$ or $(x^*)^2(y^*)^2$, or have both $x^*$ and $y^*$, i.e., $x^*(y^*)^2$.
 Thus, we cannot obtain a linear combination that in the general case yields $x^*$.
-We may have special cases where certain $x^*$ and $y^*$ values yield consistent estimates for $\hat{t}$, but consistency does not hold in general.
+Although there may be special cases in which certain $x^*$ and $y^*$ values yield consistent estimates for $\hat{t}$, but consistency does not hold in general.
 We give an example of consistency in a degenerate case in the Appendix following the proof of Theorem~\ref{thm:bl_theorem}.

### 2019-03-21 joint_inf.tex
Two little fixes.

@@
 Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
-We show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
+In this paper, we show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
 We find a sizeable region of parameter space that generates data on a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
@@
 All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners.
-Initial simulation-based results by Dr\.
-Neher using the TreeTime \citep{Sagulenko2018-xl} package has shown that the region of parameter space on which joint inference infers a zero internal branch length for 4-state DNA models is smaller than that for 2-state binary models.
+Initial simulation-based results by Dr.\ Neher using the TreeTime \citep{Sagulenko2018-xl} package has shown that the region of parameter space on which joint inference infers a zero internal branch length for 4-state DNA models is smaller than that for 2-state binary models.
 On the other hand, when the joint inference optimization is started with a small branch length, the region ending up with a zero branch length is much larger, even for 4-state DNA models.

### 2019-03-24 versions/response.v2.txt
addressing TODOs

@@
 Branch length inferences don't come into the bootstrap unless you contract things with zero branch lengths, which isn't generally done when doing bootstrap estimation.
+das: I think the reviewer is saying that, since in the case of infinite data we converge on a multifurcating topology, that, for finite sequence lengths, we *might* estimate the incorrect topology to be of higher likelihood.
+The bootstrap support would just be one way to estimate how frequently this happens for different sequence lengths.
+We're only interested in the case of infinite data, though, so I don't think there's any benefit to this---it would just be more speculation, which we below say we don't want to do.
 
@@
 DISCUSS Shall we strike the speculation about higher likelihoods on the wrong topology?
+das: I am fine with striking the speculation (and have done so, so let me know what you think).
+I guess I agree the results are "not as negative," though I don't think they are positive.
+Strong topological inconsistency would have been dope, but I don't see any way we can get it out of our method, so we'll have to settle with regular old branch length inconsistency.
 
@@
 DISCUSS: we have no evidence of topological inconsistency
+das: I still don't quite understand what this reviewer is saying: w=1 yields the highest likelihood, so the difference in lnl will always be zero.
+There is no "machine precision" involved in any of it---w is one exactly.
+That's its optimal value.
+Should we change some of the discussion to hammer this home?
 
@@
 
-das TODO
+We use "integrated likelihood" in the same manner as Goldman.
+We have removed references that are unrelated to the approaches considered in the manuscript, and defer these to the discussion for the interested reader to follow up with if they so choose.
+We have made various terminology changes throughout.
+Thank you for your suggestion in point 1; we have added it to the introduction.
+das: @EM I think your discussion with Richard over an approximation being consistently biased might be helpful to answer some of the above.
 
@@
 DISCUSS eliminating xi
+das: I think we should, as you're correct that we marginalize over it.
+I redefined a version of it at the beginning of the proof.
 
@@
 Specifically, although we can find the regions for which the various ancestral state splits are optimal, we cannot obtain optimal branch lengths analytically.
+das: Yeah, the only "numerical optimization" is for the marginal likelihood.
+What's numerical in our setup is determining the regions of maximal ancestral states, and the only reason it's numerical is that analytically it would requires 72 (maybe fewer?) inequalities, so getting them "numerically" was easier.
 
@@
 
-das TODO
+We agree, and have made the requisite changes.
 
@@
 
-das TODO
+We have removed this reference in the discussion (though we keep it in the introduction).
+We use it to show a relatively well-known example where maximum likelihood fails for a similar reason as in the current case of joint inference of phylogeny.
+The discussion preceding this sentence shows that parameters tending to infinity are not a sufficient condition for inconsistency, which we believe is more clear having removed the reference to Neyman and Scott.
+We have removed the reference to the paper by LeCam.
+It explains various setting where maximum likelihood fails, including Neyman and Scott's example, but is not directly relevant to our argument.
 
@@
 We are afraid that adding more detail about the Hadamard transform is beyond the scope of this paper, though we have added a link to a scanned page of the Semple and Steel book where I (Erick) have added some details on the proofs.
+
+das: @EM are we keeping this imgur link or is this a placeholder for a specific reference we want to make? Do you think they'd accept this?

### 2019-03-24 appendix.tex
addressing TODOs

@@
 Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
-We cannot identify complements for ancestral states We have added a sentence clarifying language:
+We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
 \[
@@
                              &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
-                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t) \nonumber \\
-                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \nonumber \\
-                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
+                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \frac{1}{2}\Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t) \nonumber \\
+                             &= \prod_{i=1}^{\nCols} \ \frac{1}{2}\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \nonumber \\
+                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\frac{1}{2}\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \nonumber\\
+                             &:= \left(\frac{1}{2}\right)^n L_\nCols''(\tau, t; \fullAlignment) \label{eq:site_pattern_likelihood}
 \end{align}
@@
 
-Let
-\[
-L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)}
-\]
-be the final product in \eqref{eq:site_pattern_likelihood}.
 Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
@@
 &    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
-&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \label{eq:site_pattern_profile_likelihood_mean}
+&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \\
+&\qquad := \ell_{\tau^*,t^*}(\tau, t) \label{eq:site_pattern_profile_likelihood_mean}
 \end{split}
 \end{equation}
-because
-$\frac{\nCols_j(\fullAlignment)}{\nCols}$ converges to $\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)$.
+because $\nCols_j(\fullAlignment)/\nCols$ converges to $\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)$.
 %EM das I believe this is the definition of \ell-- this should be made more explicit.
 %EM Also, in the main text we read that \ell is the limit of L' as n goes to infinity, but here it's L''.
+%das: L'' is the value that L' is proportional to. It does read kinda awkward, so I just explicitly added the 1/2.
 Here $j$ indexes the site splits, of which there are a fixed number $q$.
@@
 \[
-\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
+\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t).
 \]
@@
 \[
-\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
+\ell_{\tau^*,t^*}(\tau, t) = E[m_t].
 \]
@@
 If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
-This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.
+This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t)$.
 
@@
 \[
-(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)
+(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ \ell_{\tau^*,t^*}(\tau, t)
 \]
@@
 Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
-Next we show that the likelihood $\ell_{\tau_1,t}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
+Next we show that the likelihood $\ell_{\tau_1,t}(\tau_1, t)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
 Although this property should not be surprising due to symmetry, we write it out for completeness.
@@
 
-Thus exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t}(\tau_1, t; \boldsymbol\xi)$.
+Thus exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t}(\tau_1, t)$.
 Therefore we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$ and $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.
@@
 \begin{proof}
-For a fixed, known $\boldsymbol\xi$, there exists a closed form solution to $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ solving
+Consider a version of $\ell$ given by
+\[
+\ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi)
+\]
+where we fix a known value of $\boldsymbol\xi$.
+Then, there exists a closed form solution to $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ solving
 \[
@@
 \begin{align*}
-\ell_{\tau^*,t^*}(\tau^*, t, \boldsymbol\xi) &= \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_2}(x_2)\\
+\ell_{\tau^*,t^*}(\tau^*, t; \boldsymbol\xi) &= \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_2}(x_2)\\
 &\qquad + \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,w}(w).

### 2019-03-24 joint_inf.tex
addressing TODOs

@@
 
-An estimator is said to be statistically consistent if it converges to the generating model with probability one in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
+An estimator is said to be statistically consistent if it converges to the generating model with probability one in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood---this is the correct approach to maximum likelihood, as we are optimizing parameters given only observed data.
 These proofs do not readily extend to include estimating ancestral states.
@@
 The alternative approach \citep{Sagulenko2018-xl} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
-This is known in statistics as a profile likelihood \citep{Murphy2000-ry} or a relative likelihood \citep{Goldman1990-dk}, which exists here because $\alphabet$ is a finite set:
+This has been described as a relative likelihood \citep{Goldman1990-dk}, which exists here because $\alphabet$ is a finite set:
 \begin{equation}
@@
 \end{equation}
-We use $\hat{\fullAncestralStates}_\nCols$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}, and estimate a topology and branch fidelities using this profile likelihood as
+We use $\hat{\fullAncestralStates}_\nCols$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}, and estimate a topology and branch fidelities using this objective function as
 \begin{equation}
@@
 Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ (Fig.~\ref{fig:farris-fels-top}a) and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$.
-Let $\boldsymbol\xi = [\xi_j]_{j=1}^\nSiteSplits$ be the vector of most likely ancestral state splits---the explicit definition for $\boldsymbol\xi$ is given in the Appendix.
-Use $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case because, as shown in the Appendix,
+Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case because, as shown in the Appendix,
 \begin{equation}
 \label{eq:exp-log-lik}
-\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
+\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t) - \log 2.
 \end{equation}
@@
 \[
-\hat{t} = \arg\max_{t} \ \max_{\boldsymbol\xi} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)
+\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t)
 \]
@@
 As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
-In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent; we elaborate on this point in the Appendix.
+In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t)$ and $\ell_{\tau^*,t^*}(\tau_2, t)$ are equivalent; we elaborate on this point in the Appendix.
 The proof is through analytically reducing the general case to 81 separate cases (Table~\ref{tab:likelihoods}) to obtain a closed form maximal value for each.
@@
 Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
-This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which the most likely ancestral states are the same for each internal node letting $x_1=x_2=x^*$ and $y_1=y_2=y^*$ ($\xi_j=\emptyset$ for all $j$ in Table~\ref{tab:likelihoods}).
+This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which the most likely ancestral states are the same for each internal node letting $x_1=x_2=x^*$ and $y_1=y_2=y^*$.
 % If we allow multifurcating trees in our inference, then we can think of this as an instance of converging to the wrong topology, as the true $y^*\neq 1$.
@@
 \caption{
-    Estimates for $\hat{p}_w=(1-\hat{w})/2$, the optimal probability of substitution along the inner edge for the profile likelihood \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
+    Estimates for $\hat{p}_w=(1-\hat{w})/2$, the optimal probability of substitution along the inner edge for the joint objective function \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
     Regions derived in terms of probabilities of a character change along a branch for ``perfect''data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}a).
@@
 Also, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
-Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length is more likely to be observed, though we have not found regions where this is the case.
 The regions of inconsistency we found arise when one set of sister branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
@@
 
-Joint inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \citep{Murphy2000-ry}.
+Joint inference of tree parameters and ancestral sequences is related to maximizing profile likelihoods, a well-studied subject in statistics \citep{Murphy2000-ry}.
 Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., the method of sieves \citep{Geman1982}.
 A potential solution in this case using the method of sieves could be to project the column-wise ancestral states into a lower-dimensional space, allowing the degrees of freedom in the ancestral state columns to grow with $\nCols$, albeit more slowly than $O(\nCols)$.
-Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \citep{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \citep{LeCam1990}.
-Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
+Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta}, and no results have been obtained for this joint maximization technique.
 We have furthered progress in understanding the limitations of this joint optimization procedure.
@@
 All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners.
+%das: @EM do we have to add "private communication" or something here?
 Initial simulation-based results by Dr.\ Neher using the TreeTime \citep{Sagulenko2018-xl} package has shown that the region of parameter space on which joint inference infers a zero internal branch length for 4-state DNA models is smaller than that for 2-state binary models.

### 2019-03-25 versions/response.v2.txt
I'm happy.

--- 2019-03-21 versions/response.v2.txt
@@
 
-DISCUSS I honestly can't figure out what the reviewer wants to see re the bootstrap.
-Branch length inferences don't come into the bootstrap unless you contract things with zero branch lengths, which isn't generally done when doing bootstrap estimation.
-
-The reviewer is correct that this story is different than the one in the previous version of the paper.
-
-DISCUSS Shall we strike the speculation about higher likelihoods on the wrong topology?
+The reviewer is correct that this story is slightly different than the one in the previous version of the paper.
+Regarding the bootstrap, we remind the reviewer that we are interested in the case of "infinite data," and thus the bootstrap doesn't apply.
+We have also taken out the speculation about higher likelihoods on the wrong topology.
 
@@
 
-DISCUSS: we have no evidence of topological inconsistency
+In our revised work we have found no evidence of topological inconsistency, and as mentioned above we have removed such speculation from the paper.
+"Machine precision" is not involved: w exactly equal to 1.
+That's its optimal value.
 
@@
 
-das TODO
+We use "integrated likelihood" in the same manner as Goldman.
+We have replaced "profile likelihood" with "joint objective function" throughout, unless we were specifically making a comparison with profile methods.
+We have removed references that are unrelated to the approaches considered in the manuscript, and defer these to the discussion for the interested reader to follow up with if they so choose.
+Thanks.
+
+Thank you for your suggestion in point 1; we have added it to the introduction.
+We have also clarified re point 2 that the joint objective function is related to profile likelihoods, but is not the same thing.
 
@@
 
-It will estimate a branch length of zero which is equivalent to a multifurcating topology.
+We have removed all instances of "multifurcating" with "zero internal branch length" except for two occurrences in the Discussion.
 
@@
 
-DISCUSS eliminating xi
+We realized that we could eliminate the xi (xi looks like zeta) from the notation in the main text. This is definitely an improvement in readability-- thanks!
 
@@
 We focus in this paper in the limiting case only.
-We have added a sentence clarifying language: (INSERT)
+We have added a sentence clarifying language: "Because we are interested in the properties of joint inference in the limit of $n$ going to infinity, we will focus on estimates in this case and simply refer to the right hand side of \eqref{eq:exp-log-lik} as ``the estimator.''"
 
@@
 We have rewritten this section to clarify what is being used where.
-%EM das-- is this right?
 Specifically, although we can find the regions for which the various ancestral state splits are optimal, we cannot obtain optimal branch lengths analytically.
 
-We moved these figures around in response to a previous reviewer comment, who thought that Figure S2 was quite uninteresting while Figure 2 was relevant.
+We moved these figures into their current locations in response to a previous reviewer comment, who thought that Figure S2 was quite uninteresting while Figure 2 was relevant.
 We would prefer to leave them be.
@@
 
-das TODO
+We agree, and have made the requisite changes: both replacing references to "profile likelihood" in the body of the text with "joint objective function", and by clarifying that they aren't quite the same things in the Discussion.
 
@@
 
-das TODO
+We have removed this reference in the Discussion.
+We use it to show a relatively well-known example where maximum likelihood fails for a similar reason as in the current case of joint inference of phylogeny.
+The discussion preceding this sentence shows that parameters tending to infinity are not a sufficient condition for inconsistency, which we believe is more clear having removed the reference to Neyman and Scott.
+We have removed the reference to the paper by LeCam.
+It explains various setting where maximum likelihood fails, including Neyman and Scott's example, but is not directly relevant to our argument.
 
@@
 > p_x \to 0).
->
+
+We have added this intuition, thank you!
+
 > My empirical tests of the approximate joint inference scheme suggest
@@
 
-We have worked this out with Dr.
-Neher and found that the difference lay in that he was not optimizing the objective function from both sides.
-This simulation work also demonstrated that the intuitive explanation by this reviewer is incorrect.
+We have worked this out with Dr. Neher and found that the difference lay in that he was not optimizing the objective function thoroughly and was getting stuck in a local optimum.
 
@@
 
-We now remind the reader that this is the number of site splits.
+We have inserted text reminding the reader that this is the number of site splits.
 
@@
 
-We now emphasize that the number of taxa is fixed so q is fixed, and that although n_j goes to infinity the ratio n_j/n converges to the probability of seeing that column.
+We have added text emphasizing that the number of taxa is fixed so q is fixed, and that although n_j goes to infinity the ratio n_j/n converges to the probability of seeing that column.
 
@@
 We have clarified the definitions of S and Y, and added that by path we mean graph-theoretic path.
-We are afraid that adding more detail about the Hadamard transform is beyond the scope of this paper, though we have added a link to a scanned page of the Semple and Steel book where I (Erick) have added some details on the proofs.
+We are afraid that adding more detail about the Hadamard transform is beyond the scope of this paper, and it is well described in Semple and Steel's book.

### 2019-03-25 appendix.tex
I'm happy.

@@
                              &= \prod_{j=1}^{\nSiteSplits} \ \left[\frac{1}{2}\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \nonumber\\
-                             &:= \left(\frac{1}{2}\right)^n L_\nCols''(\tau, t; \fullAlignment) \label{eq:site_pattern_likelihood}
+                             &=: \left(\frac{1}{2}\right)^n L_\nCols''(\tau, t; \fullAlignment) \label{eq:site_pattern_likelihood}
 \end{align}
@@
 &    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
-&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \\
-&\qquad := \ell_{\tau^*,t^*}(\tau, t) \label{eq:site_pattern_profile_likelihood_mean}
+&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \label{eq:site_pattern_profile_likelihood_mean}
 \end{split}
@@
 because $\nCols_j(\fullAlignment)/\nCols$ converges to $\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)$.
-%EM das I believe this is the definition of \ell-- this should be made more explicit.
-%EM Also, in the main text we read that \ell is the limit of L' as n goes to infinity, but here it's L''.
-%das: L'' is the value that L' is proportional to. It does read kinda awkward, so I just explicitly added the 1/2.
+We define this limit to be $\ell_{\tau^*,t^*}(\tau, t)$.
 Here $j$ indexes the site splits, of which there are a fixed number $q$.
@@
 We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
-For additional detail on the proofs in that section, see \url{http://i.imgur.com/T5XLPLB.jpg}.
 For each edge $e$ define the edge ``fidelity'' for that edge as
@@
 \]
-where we fix a known value of $\boldsymbol\xi$.
+where we fix a given value of $\boldsymbol\xi$.
 Then, there exists a closed form solution to $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ solving

### 2019-03-25 joint_inf.tex
I'm happy.

@@
 \usepackage[colorinlistoftodos]{todonotes}
-\usepackage[colorlinks=true, allcolors=blue]{hyperref}
+\usepackage[hidelinks]{hyperref}
+%\usepackage[colorlinks=true, allcolors=blue]{hyperref}
 % \usepackage{showlabels}
@@
 
-\title{Joint maximum-likelihood of phylogeny and ancestral states is not consistent}
+\title{Joint maximum likelihood of phylogeny and ancestral states is not consistent}
 
@@
 In this paper, we show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
-We find a sizeable region of parameter space that generates data on a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
+We find a sizeable region of parameter space that generates data on a four-taxon tree for which this joint method estimates the internal branch length to be exactly zero, even in the limit of infinite-length sequences.
 More generally, we show that this joint method only estimates branch lengths correctly on a set of measure zero.
@@
 
-Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out latent ancestral states at the internal nodes of the tree, obtaining an integrated likelihood \citep{Goldman1990-dk}.
+Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out latent ancestral states at the internal nodes of the tree, obtaining an integrated likelihood \citep{Felsenstein1973-ba,Goldman1990-dk}.
 In a recent paper, \citet{Sagulenko2018-xl} suggest using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences on a fixed topology.
@@
 In this paper, we show that jointly inferring trees and ancestral sequences is not consistent in general.
-To do so, we use a binary symmetric model with data generated on a four-taxon tree: we compute closed form solutions to the joint objective function and demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be exactly zero.
+To do so, we use a binary symmetric model with data generated on a four-taxon tree: we compute closed form solutions to the joint objective function and demarcate a sizeable area of branch lengths in which joint inference is guaranteed to estimate the internal branch length to be exactly zero, even in the case of perfect sequence data with an infinite number of sites.
 We show that, when the topology is known and fixed, joint inference for branch length estimation cannot be consistent except on a set of measure zero (i.e.\ a set that occupies zero volume in parameter space).
@@
 Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ are comparable to the branch lengths (Fig.~\ref{fig:empirical-bias}), showing that even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
+One can see the region in which the internal branch length is estimated to be zero under the curve in the lower right of this plot.
+Dr.\ Neher has pointed out in his review that the shape of this region can be intuitively understood as that where two mutations on the longer pendant branches are more likely ($p_{x^*}^2$) than one mutation on the internal branch ($p_{y^*}$).
 
-In contrast, inference on the integrated likelihood performs as expected, such that $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}).
+In contrast, inference on the classically-used integrated likelihood performs as expected, such that $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}).
 The errors in this case \eqref{eq:marginal_likelihood} via optimization with L-BFGS-B are lower than machine tolerance.
@@
 This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3).
-However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and the bias is comparable to the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).
+However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and the bias can be comparable to the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).
 
@@
 %das: @EM do we have to add "private communication" or something here?
+%Erick response: I don't know, but the copyeditors will tell us.
 Initial simulation-based results by Dr.\ Neher using the TreeTime \citep{Sagulenko2018-xl} package has shown that the region of parameter space on which joint inference infers a zero internal branch length for 4-state DNA models is smaller than that for 2-state binary models.
@@
 
-On a practical note, we would like to emphasize that TreeTime has implemented marginal mode inference since the publication of the \citet{Sagulenko2018-xl} paper.
+On a practical note, we would like to emphasize that TreeTime implements marginal mode inference.
 In January 2019 a warning was added, in all caps, that is triggered when joint inference is used for trees such that any branch length is greater than 0.15.
@@
 We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.
-We are especially thankful to Dr.\ Neher for providing TreeTime simulation code that allowed us to test our conclusions on simulated data.
+We are especially thankful to Dr.\ Neher for his helpful suggestions and for providing TreeTime simulation code that verified our conclusions on simulated data.
 Insight from the reviewers and the editors greatly improved this manuscript.

### 2019-03-26 appendix.tex
Doneso.

@@
 We begin by showing an inconsistency in branch length estimation on the InvFels tree.
+In this proof and other material below, we will use a version of $\ell$ given by
+\[
+\ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi)
+\]
+where we fix a given value of the ancestral state split $\boldsymbol\xi$.
 
@@
 \begin{proof}
-Consider a version of $\ell$ given by
-\[
-\ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi)
-\]
-where we fix a given value of $\boldsymbol\xi$.
-Then, there exists a closed form solution to $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ solving
+When $\boldsymbol\xi$ is fixed, there exists a closed form solution to $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ solving
 \[
@@
 \]
-We show in this case that the log-likelihood $\ell$ attains a unique maximum at $\hat{t}_{\boldsymbol\xi}$.
+We first show that this maximum is unique.
 For fixed $\boldsymbol\xi$, the log-likelihood can be decomposed into a sum of functions of each variable, i.e.,

### 2019-03-26 joint_inf.tex
Doneso.

@@
 
-An estimator is said to be statistically consistent if it converges to the generating model with probability one in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood---this is the correct approach to maximum likelihood, as we are optimizing parameters given only observed data.
+An estimator is said to be statistically consistent if it converges to the generating model with probability one in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood---this is the correct approach to maximum likelihood, as it optimizes parameters given only observed data.
 These proofs do not readily extend to include estimating ancestral states.
@@
 In order to understand the optimal internal branch length in each of these cases, we optimize branch lengths numerically to obtain $\hat{w}$ under joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
-Using either technique, we see that the region of inconsistency encompasses almost half of the branch fidelity space: given the correct topology, there are many situations where the interior branch length is estimated to be zero.
+We find that for almost half of branch fidelity space, the interior branch length is estimated to be zero given the correct topology.
 
@@
     Regions derived in terms of probabilities of a character change along a branch for ``perfect''data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}a).
-    The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero, resulting in an inconsistency.
+    The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero.
 }
@@
 One can see the region in which the internal branch length is estimated to be zero under the curve in the lower right of this plot.
-Dr.\ Neher has pointed out in his review that the shape of this region can be intuitively understood as that where two mutations on the longer pendant branches are more likely ($p_{x^*}^2$) than one mutation on the internal branch ($p_{y^*}$).
+Dr.\ Neher has pointed out during peer review that the shape of this region can be intuitively understood as that where two mutations on the longer pendant branches are more likely (order $p_{x^*}^2$) than one mutation on the internal branch (order $p_{y^*}$).
 
@@
 
-On a practical note, we would like to emphasize that TreeTime implements marginal mode inference.
-In January 2019 a warning was added, in all caps, that is triggered when joint inference is used for trees such that any branch length is greater than 0.15.
+On a practical note, we would like to emphasize that TreeTime also implements marginal mode inference, which calculates the classical integrated likelihood.
+In January 2019 a warning was added, which cannot be turned off and is in all capital letters, that is triggered when joint inference is used for trees such that any branch length is greater than 0.15.
 
@@
 We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.
-We are especially thankful to Dr.\ Neher for his helpful suggestions and for providing TreeTime simulation code that verified our conclusions on simulated data.
-Insight from the reviewers and the editors greatly improved this manuscript.
+We are especially thankful to Dr.\ Neher for his suggestions and for providing TreeTime simulation code that verified our conclusions on simulated data.
+Insight from all of the reviewers and the editors greatly improved this manuscript.
 

### 2019-03-26 treetime-sims/four_taxon.py
Adding scripts used in discussion with Richard.

@@ -1,0 +1,193 @@
+import os
+import numpy as np
+from collections import defaultdict
+from Bio import AlignIO,Phylo
+from treetime.treeanc import TreeAnc
+from treetime.gtr import GTR
+from treetime.seqgen import SeqGen
+from matplotlib import cm
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+def invFels(y,x, L=1000, alphabet='nuc_nogap'):
+    from io import StringIO
+    T = Phylo.read(StringIO("(A:%f,B:%f,(C:%f,D:%f):%f);"%(y,y,x,x,y)), "newick")
+
+    gtr = GTR(alphabet=alphabet)
+    gtr.seq_len=L
+    mySeq = SeqGen(gtr=gtr, tree=T)
+    mySeq.evolve()
+
+    return T, mySeq.get_aln()
+
+def tiny_internal(T):
+    T.root.clades[2].branch_length = 0.0001
+    T.root.clades[2].mutation_length = 0.0001
+
+def noisy_shortened(T):
+    for n in T.find_clades():
+        n.branch_length *= 0.6 + np.random.random()*0.4
+        n.mutation_length=n.branch_length
+
+def topology_only(T):
+    for n in T.find_clades():
+        n.branch_length = 0.1
+        n.mutation_length=n.branch_length
+
+def infer_branch_length(tree, aln, distance_scale = 1.0,
+                        marginal=False, IC=None, alphabet='nuc_nogap'):
+    before = {}
+    after = {}
+    tt = TreeAnc(tree=tree, aln=aln, alphabet=alphabet, verbose=0)
+    for n in tt.tree.find_clades():
+        n.branch_length   *= distance_scale
+        n.mutation_length *= distance_scale
+        before[n.name] = n.branch_length
+    total_before = tt.tree.total_branch_length()
+
+
+    # mess up branch length prior to optimizition
+    if IC:
+        IC(tt.tree)
+    tt.prepare_tree()
+
+    if marginal:
+        tt.optimize_sequences_and_branch_length(branch_length_mode='marginal',
+                                                prune_short=False, max_iter=20)
+    else:
+        tt.optimize_sequences_and_branch_length(branch_length_mode='joint',
[…]

### 2019-03-26 treetime-sims/four_taxon_twoside.py

@@ -1,0 +1,289 @@
+import os
+import numpy as np
+from collections import defaultdict
+from Bio import AlignIO,Phylo
+from treetime.treeanc import TreeAnc
+from treetime.gtr import GTR
+from treetime.seqgen import SeqGen
+from matplotlib import cm
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+def invFels(y,x, L=1000, alphabet='nuc_nogap'):
+    from io import StringIO
+    T = Phylo.read(StringIO("(A:%f,B:%f,(C:%f,D:%f):%f);"%(y,y,x,x,y)), "newick")
+
+    gtr = GTR(alphabet=alphabet)
+    gtr.seq_len=L
+    mySeq = SeqGen(gtr=gtr, tree=T)
+    mySeq.evolve()
+
+    return T, mySeq.get_aln()
+
+def tiny_internal(T):
+    T.root.clades[2].branch_length = 0.0001
+    T.root.clades[2].mutation_length = 0.0001
+
+def noisy_shortened(T):
+    for n in T.find_clades():
+        n.branch_length *= 0.6 + np.random.random()*0.4
+        n.mutation_length=n.branch_length
+
+def topology_only(T):
+    for n in T.find_clades():
+        n.branch_length = 0.1
+        n.mutation_length=n.branch_length
+
+def infer_branch_length(tree, aln, distance_scale = 1.0,
+                        marginal=False, IC=None, alphabet='nuc_nogap'):
+    before = {}
+    after = {}
+    tt = TreeAnc(tree=tree, aln=aln, alphabet=alphabet, verbose=0)
+    for n in tt.tree.find_clades():
+        n.branch_length   *= distance_scale
+        n.mutation_length *= distance_scale
+        before[n.name] = n.branch_length
+    total_before = tt.tree.total_branch_length()
+
+
+    # mess up branch length prior to optimizition
+    if IC:
+        IC(tt.tree)
+    tt.prepare_tree()
+
+    if marginal:
+        tt.optimize_sequences_and_branch_length(branch_length_mode='marginal',
+                                                prune_short=False, max_iter=20)
+    else:
+        tt.optimize_sequences_and_branch_length(branch_length_mode='joint',
+                                                prune_short=False, max_iter=20)
+
[…]

### 2019-03-29 versions/response.v2.txt
once over

@@
 
-In our revised work we have found no evidence of topological inconsistency, and as mentioned above we have removed such speculation from the paper.
-"Machine precision" is not involved: w exactly equal to 1.
-That's its optimal value.
+In our revised work we found no evidence of topological inconsistency, and as mentioned above we have removed such speculation from the paper.
+"Machine precision" is not involved: the optimal value of w is exactly equal to 1 in the limit.
 
@@
 Thank you for your suggestion in point 1; we have added it to the introduction.
-We have also clarified re point 2 that the joint objective function is related to profile likelihoods, but is not the same thing.
+We have also clarified in regards to point 2 that the joint objective function is related to profile likelihoods, but is not the same thing.
 
@@
 
-We focus in this paper in the limiting case only.
+We focus in this paper on the limiting case only.
 We have added a sentence clarifying language: "Because we are interested in the properties of joint inference in the limit of $n$ going to infinity, we will focus on estimates in this case and simply refer to the right hand side of \eqref{eq:exp-log-lik} as ``the estimator.''"
